### 2021-01-25 main/MSstatsPTM_Main.tex
Converted main to latex

\documentclass[11pt]{article}

\title{Statistical methods for relative quantification of post-translational modifications in global proteomics experiments}

\author{
  LastName1, FirstName1\\
  \texttt{first1.last1@xxxxx.com}
  \and
  LastName2, FirstName2\\
  \texttt{first2.last2@xxxxx.com}
}

\begin{document}

\maketitle

\section{Abstract}

\section{Introduction}

The signaling mechanisms that allow cells to mount a dynamic and fast response to a multitude of events are primarily facilitated by the modification of proteins at specific residues, acting as molecular on/off switches. Mass spectrometry-based label-free proteomics is broadly established as the tool-of-choice for unbiased and large-scale identification and quantification of proteins and their post-translational modifications (PTMs) using liquid chromatography coupled with mass spectrometry (LC-MS). Studies targeting the post-translationally modified proteome focus either on the accurate localization of modification sites on proteins, relative or absolute quantification of a modification site’s occupancy repertoire, or relative changes in occupancy across experimental conditions. Regardless of the question at hand, interrogating the modified proteome is challenging due to a number of reasons. First, the relatively lower abundance of modified proteo-forms dictates that a global interrogation can only be achieved through large-scale enrichment protocols with modification-specific antibodies or beads. Variability in the enrichment efficiency inevitably affects the reproducibility of the number of spectral features (e.g., peptide precursor ions or their fragments) and their intensities, which imposes challenges in both quantification and statistical modeling. Second, contrary to the often large number of identified peptides that can be used as features to model protein abundance changes, there are relatively few representative peptides that span a modification sites, which often results in sparse, and sometimes, inherently convoluted models (i.e., single versus multiple modified sites on a single peptide). Third, unless early signaling events are interrogated, the interpretation of the relative changes in modification occupancy are inherently convoluted with changes in the overall protein expression, making the interpretation of the results not straightforward. Therefore, a robust approach to estimate systematic relative changes in post-translational modifications, at scale, should not only combine the quantitative information pertaining to a PTM site over peptides and replicates in multiple conditions, but take into account various sources of variations and confounding factors present in the experiments.

Despite the important implications of PTMs in biological functions, there is a lack of general framework to summarize the available quantitative information from LC-MS data, to perform statistical inference, and to draw conclusions to characterize the quantitative properties of PTM in a statistically rigorous manner. Many investigations performed differential expression analysis of PTMs using two-sample t-test or its extensions. The approach takes as input intensities of individual features from modified peptides, or intensity ratios of modified and unmodified peptide features, and compares the mean abundance of a PTM site from one condition to another. Modifications of the t-test such as moderated t-test with limma were also proposed. While simple, the approach does not fully account for the sources of variations, and it is not directly applicable to experiments with complex designs, e.g., comparisons of multiple conditions, acquisition in multiple batches, etc. Isobar-PTM was developed for experiments with MS/MS quantitative strategies that employ isobaric labels such as tandem mass tags (TMT) and isobaric tag for relative and absolute quantification (iTRAQ).[Source] Isobar-PTM expresses MS measurements with a linear model and performs adjustment with respect to protein abundance using the difference between log-ratio of modified peptides in two channels and log-ratio of protein level. The modeling framework, however, is not applicable for either label-free workflows or experiments with complex designs. [Source]

We propose a general statistical approach, which explicitly characterizes the variations and confounding factors present in bottom-up PTM experiments. With analysis procedure developed for summarization of LC-MS data, quantitative characterization of site-specific PTMs, and adjustment with respect to protein abundance, the proposed approach is aimed at the detection of quantitative changes in PTMs between conditions. Quantitative analyses of PTMs often involve comparisons between multiple inter-related conditions of the same biological system. The general statistical framework underlying the proposed approach allows for analyzing experiments with complex designs, including those with multiple conditions and/or multiple batches.

The proposed approach was evaluated using datasets from computer simulations, benchmark controlled mixtures and biological investigations, in comparison with the commonly applied t-test. The results demonstrated that by appropriately leveraging the information from the entire dataset the proposed approach improves the reproducibility and accuracy of the estimates of PTM fold changes, resultsin a better calibrated type I error rate, and improves the statistical power of detecting changes in PTMs. For analyses of datasets with multiple batches, the proposed approach also improves the overall performance under various forms of batch effects. The proposed approach is implemented as an open source R package MSstatsPTM, which employs similar input format as in MSstats.

\section{Proposed Approach}

Figure 3 schematically illustrates a simplified version of the data structure resulting from a typical bottom-up experiment for quantitative analysis of PTMs, in which there are multiple layers of variation present. A PTM site is quantified with multiple spectral features, which vary in sequence (e.g., fully or partially cleaved peptides), ionization efficiency, charge states, etc. The number quantified features vary across replicate LC-MS/MS runs of the same sample, and across conditions. To perform adjustment with respect to protein abundance, features of unmodified peptides are used for the inference of underlying protein abundance. Typically, because of the enrichment step for PTMs, very few of those features are present in original LC-MS runs. For more accurate estimation of protein abundance, separate global proteomics data of unenriched samples are often acquired. As different levels of variability are present in the data, the log-intensities of the features for modified and unmodified peptides are modeled separately using two linear mixed models.

\subsection*{Statistical modeling and parameter estimation}

The proposed approach takes as input a list of log-transformed intensities of spectral features, identified and quantified across LC-MS runs. The features, which are precursor ions of modified or unmodified peptides, are used to characterize the identified PTM sites and proteins. For each PTM site, the feature log-intensities of the modified peptides spanning the site are expressed using a linear mixed model in consideration of the effects of condition, run, feature and interaction between run and feature.The model parameters are estimated using the split-plot approach as in MSstats[Source], where the feature log-intensities are first summarized into a single value per site per run in the subplot model, and the site-level summaries are then used for the inference of the PTM site abundance. In the site-level summarization, Tukey's median polish (TMP), a simple and robust procedure is applied to iteratively fit a two-way additive model with the effects of run and feature, which in turn summarizes the log-intensities for each site. [Source]  After the summarization, the inference of the PTM site abundance in each condition is carried out in account for the experimental design (Fig. 3a). Statistical modeling and quantification for global proteomics data are performed by the same procedure as for PTM data.

\subsection*{Detection of changes in PTMs}

Detection of differentially modified PTM sites is performed through testing the null hypothesis of ‘no change’ against the alternative. The null hypothesis states that there is no difference in log-abundance of the PTM site between conditions, adjusted with respect to protein abundance (Fig. 3b). Specifically, the adjusted difference is given by the difference in log-abundance of the PTM site, subtracted by the difference in log-abundance of the underlying protein, which is equivalent to the log of the ratio of PTM abundance difference to protein abundance difference. The estimate of the adjusted difference and the standard error (SE) of the estimate are obtained by combining the difference estimates and the associated SEs from both counterparts (for details, see Supplementary Sec. S3.1). The test statistic for the hypothesis testing is the ratio of the estimate of the adjusted difference to its SE. To determine the statistical significance of the difference in terms of p-value, the test statistic is compared against the t distribution with degrees of freedom approximated by the Satterthwaite method.Adjustment for multiple comparisons is performed using the Benjamini-Hochberg procedure to control the false discovery rate at a desired level, e.g., 0.05. The details are provided in Supplementary Sec. S3.2.

\subsection*{Analysis with multiple batches}

The proposed statistical framework allows to analyze data from experiments of complex designs such as factorial design. We discuss below a specific design commonly considered in PTM experiments, in which data are acquired in multiple batches, often as a result of multiple repeats for the antibody enrichment step. Typically, the number of spectral features and their intensities are different across batches, and the run-to-run variation sometimes varies across batches as well. The proposed approach summarizes log-intensities in each batch separately, and considers two ways to perform the statistical inference and testing. When variability in site-level summaries and in PTM changes is expected across batches, the proposed approach performs inference of the PTM site abundance and the adjusted difference for each batch separately. The evidence about differential PTM sites between conditions is averaged over batches. Alternatively, assuming identical variability in the site-level summaries and identical differences between conditions across batches (i.e., no interaction effect between condition and batch), the statistical inference and testing can be performed based on one single model with an additional fixed effect of batch. Details about these modeling considerations are discussed in Supplementary Sec. S3.4.

\section{Results}

\subsection*{Computer simulations}

The proposed statistical approach methods were evaluated using computer simulation. Specifically, their properties under adjustment with respect to protein abundance and batch effects were evaluated.

\paragraph*{Computer simulation: protein-level adjustment.} Differential intensity levels of modified peptides may be due to changes in modification, change in protein abundance, or both. The proposed approach adjusts the abundance with respect to unmodified peptides by combining the inference of modified and unmodified peptide abundances. Alternatively, two-sample t-test that takes as input the ratio between modified and unmodified peptide intensities (difference on log scale) is commonly applied for the same purpose. In real experiments, multiple inter-related conditions are often compared together. Whereas t-test uses measurements from the two conditions being compared, the proposed approach leverages measurements in all conditions for the inference of underlying abundance. We evaluated the impact with such adjustment by computer simulation, considering the following factors: with/without protein-level changes, effect size, number of replicates and number of conditions. Details of the simulation can be found in Supplementary Sec. S4.1. 
In this simulation, when there was no change in protein abundance, all the considered approaches well calibrated the Type I error rate (Fig. 4a). However, when the modification changes were entirely due to the changes in protein abundance across conditions, analysis without accounting for the protein-level changes resulted in off-target, high false positive rates (Fig. 4b). As shown in Fig. 4c-e, in detection of systematic changes in PTM abundance, the proposed approach improved statistical power with small sample sizes in almost all the considered scenarios. The advantage by using the proposed approach over t-test became more profound in the presence of missing data (Fig. 4d-e). In cases with small sample sizes (e.g., 2 replicates) as shown in Fig. 4d, performance by t-test decreased dramatically with one missing value, and increasing the log-fold change to 2 did not effectively reduce the negative impact. Two-sample t-test only used data within the groups of interest while ignoring the rest of the data. Consequently, it gave similar performance across cases with different number of conditions. In contrast, the proposed approach leveraged all available information, which resulted in improved power with increased number of conditions (see for example, the improved performance in Fig. 4e [with 4 conditions] over Fig. 4d [with 2 conditions]). Further results are provided in Supplementary Sec. S4.1.

\paragraph*{Computer simulation: batch effects.} PTM experiments often create batches during data acquisition.Whereas the proposed approach can be naturally extended to handle such scenarios by modeling various forms of batch effects; t-test is not directly applicableto a problem with batches of data. Two commonly used ad-hoc approaches are a) t-test (no batch): ignoring batch effects when applying t-test, and b) t-test (most significant batch): applying t-test in each batch and drawing conclusions based on themost significant batch. Although simple, these ad-hoc methods lack statistical justification. We characterized their statistical properties under various forms of batch effects in Supplementary Sec. S4.2. Twobatches of data were generated, where various forms of batch effects were simulated, includingdifference in signal intensities across batches, difference in variability across batches, and interactionbetween batch and condition (i.e., change between conditions affected by batch). 
The major results are summarized in Fig. 5 with additional details provided in Supplementary Sec. S4.2.In terms of estimation of fold change, the most statistically significant batch with t-test was highly variable and frequently biased (Fig. 5a). In the detection of systematic changes in PTM abundance, it also frequently resulted in higher false positive rates than the expected value (Fig. 5b). As shown Fig. 5c, the proposed approach improved power with small sample sizes in almost all thescenarios, under various forms of batch effects including difference in intensity level and difference in variability. Using t-test while ignoring batch effect gave similar performancein special cases with no difference in intensity level between batches (Supplementary Fig. SXX), but its performance dramatically decreased in general cases (Supplementary Figs. S8-S9). The negative impact was only partially reduced by increasing the sample size to 5. The proposed approach gave consistently improved performancecompared with other methods by properly characterizing batch effects and leveraging all availableinformation.

\subsubsection*{Spike-in experiment [Need to write results of this experiment]}

We evaluated our approach using a custom designed benchmark experiment, where

\subsubsection*{Dilution experiment [What was this experiment?]}

We evaluated our approach using a custom designed benchmark experiment, where...
Two sets of liver samples from five Atg16L1 deficient mice (one with KGG enrichment followed by trypsin digestion, one with trypsin digestion alone) were prepared. Following a randomized order, each sample was injected at three different concentrations 4ul, 2ul, and 1ul. The experiment resulted in 15 KGG and 15 global profiling runs. To correct for run-to-run variation, the AQUA peptide mixture was spiked into each sample. There are 4696 ubiquitinated proteins identified in this dataset. Among these, 3173 have corresponding measurements from global data. 
To simulate the effects of changes in protein abundance, we created three groups of data by matching one concentration for the KGG data with one for the global data: K1P1, K2P2, and K4P4. Systematic changes in the abundance of ubiquitinated sites between the groups are present in the KGG data. These changes are considered as artifacts driven by the changes in the abundance of their corresponding proteins. Using the comparison of K2P2 vs. K1P1 as example, there are 19355 possible comparisons with KGG data, and 3860 changes are detected. Out of the 19355 sites, 15036 have corresponding measurements from global data and are eligible for protein-level correction. The correction significantly reduced the number of false positives to 79. Similar observations were made in the other two comparisons K4P4 vs. K2P2 and K4P4 vs. K1P1, as shown in Table XX.

\subsubsection*{Ubiquitin changes in bone marrow derived macrophages from Atg16L1 deficient mice Re-analysis of published dataset: USP30}

Compare results with/without adjustment\\
Very few measurements of unmodified peptides are available\\
Compare results with original findings] [Need original findings\\
Include USP30OE vs. Ctrl in addition to Combo vs. CCCP?\\
Comment on features, degrees of freedom\\
Different changes across batches – P54652, Q5VYK3

\subsubsection*{Re-analysis of published dataset: RIP1 [Need this dataset and original findings]}

\section{Conclusion and discussion}

We proposed a general statistical modeling framework for PTM characterization. The framework is designed for bottom-up MS workflows, which are characterized with variations from multiple convoluted sources, frequent missing data, and associated uncertainty in the conclusions. The framework is general and is applicable to a variety of experimental designs. It outperforms the ad-hoc methods underlying the t-test, and yields accurate results in the broad type of experimental circumstances, including the presence of missing values, changes in protein abundance, and batch effects. The framework allows us to plan for subsequent experiments, and choose the appropriate number of replicates in consideration of adjustment with respect to protein abundance.

Our results show that when measurements from multiple related conditions are available, the proposed approach for joint modeling and summarization of all the LC-MS/MS runs leads to more sensitive PTM significance analysis and more accurate and precise quantification than when separately analyzing conditions of runs. The gain is due to a more efficient use of the data, and to a more accurate understanding of the systematic and random variations. The proposed framework can be extended beyond the experimental designs with multiple batches and conditions discussed above. For example, it can represent experimental designs with even more complex structures, such as time series or factorial investigations.

A potential limitation of the proposed framework is the assumption that all the peptides are correctly mapped to the underlying proteins and PTM sites, and the features are informative of the abundances of underlying protein and PTM. Also, characterizing PTMs with current data-dependent acquisition workflows is prone to being undersampled, leading to a sparse dataset with a large number of missing values for the analysis. Statistical methods accounting for effects due to experimental units and missing values introduced in this manuscript help interpret the data in a more objective manner. The latest development of targeted acquisition and data-independent acquisition methods are expected to further alleviate these issues.

Overall, the proposed approach balances accuracy and practicality, and enables the analysis of complex experiments in high throughput. Future work is to carry out the inference and testing for not only the relative change of PTM abundance, but also the fraction of the protein that is modified at the particular site (site occupancy, or stoichiometry). We are also interested in characterizing the interplay of PTMs at multiple sites.
The proposed statistical methods are implemented as an R package MSstatsPTM available on Bioconductor and Github.

\section{Figure legends}

Figure 3.Data structure of a typical PTM experiment and goals of PTM characterization.(a)Schematic data representation, in a simplified case of two conditions and two replicate runs. Each PTM site is modeled and characterized separately, where a PTM is quantified with multiple spectral features (boxes), distinguished by different charge states of a peptide. The feature intensities are viewed as repeated measurements of the underlying abundance of the PTM, where the abundance in Condition i is denoted by i. Features corresponding to unmodified peptides are considered together to perform adjustment with respect to protein abundance, where the protein abundance in Condition i is denoted by *i. Peptides can be fully cleaved (solid lines) and/or partially cleaved (dashed lines). Some spectral features can be missing.(b)PTM relative quantification by statistical inference, which makes use of the feature intensities to infer the underlying PTM abundance and protein abundance with an estimate of associated uncertainty. (c) Model-based testing for differential PTM abundance, which corrects for the underlying protein abundance with a cost of increased uncertainty about the estimate of difference between conditions.(d) Statistical experimental design in terms of sample size calculations and power analysis.

Figure 4.False positive rate and statistical power in PTM significance analysis by the proposed approach, t-test with protein-level adjustment, and t-test without protein-level adjustment. (a) When there was no change in protein abundance, all considered methods well calibrated the Type I error rate. (b) When the PTM changes were entirely due to changes in protein abundance across conditions, analysis accounting for the protein-level changes resulted in off-target, high false positive rates. (c) Statistical powers by the proposed approach and t-test with protein-level adjustment, where the data consisted of 2 conditions and the SD corresponding unmodified feature intensities was 0.2. (d) Statistical powers by the proposed approach and t-test with protein-level adjustment, where the data consisted of 2 conditions, the SD corresponding unmodified feature intensities was 0.2, and the PTM was missing in Run 1 of Condition 1. (e) Same as in (d), but the data consisted of 4 conditions.

Figure 5.Results corresponding to estimation error and false positive rate and statistical power of the PTM significance analysis, where the data were acquired in two batches. The following parameters were considered to represent the batch effects: no batch-condition interaction, mean intensity level in Batch 2 was higher than Batch 1 by 2 on log scale, and the SDs in Batch 1 and Batch 2 were 0.2 and 0.3, respectively.(a)Estimation based on the most statistically significant batch with t-test was more variable than other methods and frequently biased. (b)The proposed approach better calibrated Type I error rate. (c)The proposed approach improved statistical power in all the considered cases with different number of conditions and replicates. This was achieved by properly characterizing batch effects and leveraging all available information. Ignoring batch effects by t-test (no batch) lost power dramatically. The negative impact was only partially reduced by increasing the sample size to 5.

Figure 6.Design of future PTM experiments in terms of sample size calculations and power analysis. (a) Protein-level adjustment relies on the inference of protein abundance, which introduces additional uncertainty in the estimate of PTM difference. Therefore, the required sample size to detect a systematic change is higher than as expected for standard differential analysis without adjustment. Sample size calculations without accounting for the uncertainty would lead to over-optimistic, under-powered studies. (b) In complex designs, simultaneously analyzing all the conditions effectively increases the degrees of freedom and requires fewer replicates. (c)Increasing the sample size and analyzing multiple conditions together both result in improved statistical power.

\end{document}n{document}

### 2021-01-25 supplementary/bcp/ptm_sm.tex
Latex supplementary

%\documentclass[letterpaper,12pt]{article}
\documentclass{mcp}

\usepackage{graphics}

\usepackage{geometry}
%\usepackage{amsmath}
%\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{bm}
\usepackage{algorithmic}
\usepackage{algorithm}
\usepackage{rotating}
\usepackage[caption=false,font=footnotesize]{subfig}
\usepackage{color}
\usepackage{tikz}
\usetikzlibrary{shapes,arrows,backgrounds,calc,positioning,fit,petri,plotmarks}
\usepackage{multirow}
\usepackage{graphicx}
\usepackage{subfig}
\usepackage{url}
\usepackage{booktabs}
\usepackage{enumitem}
\usepackage{authblk}

\def\todo#1{{\color{red}[TODO: #1]}}

%\usepackage{caption} 
%\captionsetup[table]{skip=10pt}
\captionsetup{belowskip=12pt,aboveskip=4pt}

% strikkeout \sout
\usepackage[normalem]{ulem}

\usepackage{fullpage}
\usepackage{amsfonts,amsthm}
\usepackage{amsmath}
%\usepackage[fleqn]{amsmath}
%\usepackage{setspace}
%\usepackage{url}

%\usepackage[table,dvipsnames]{xcolor}
%\usepackage{textcomp}
%\usepackage{gensymb}

\usepackage{fancyhdr}
\pagestyle{fancy}
\lhead{\footnotesize \parbox{11cm}{Supplementary }}
\headsep = 25pt

%\usepackage{tabulary,multirow,multicol,rotating}

%% big wide hat
\usepackage{scalerel,stackengine}
\stackMath
\newcommand\reallywidehat[1]{%
\savestack{\tmpbox}{\stretchto{%
  \scaleto{%
    \scalerel*[\widthof{\ensuremath{#1}}]{\kern-.6pt\bigwedge\kern-.6pt}%
    {\rule[-\textheight/2]{1ex}{\textheight}}%WIDTH-LIMITED BIG WEDGE
  }{\textheight}% 
}{0.5ex}}%
\stackon[1pt]{#1}{\tmpbox}%
}
\parskip 1ex

\renewcommand{\thetable}{S\arabic{table}}
\renewcommand{\thefigure}{S\arabic{figure}}
%\renewcommand{\tablename}{Supplementary Table}
%\renewcommand{\figurename}{Supplementary Figure}

\def\sfigref#1{{Figure~\ref{#1}}}
\def\secref#1{Section~\ref{#1}}
\def\stabref#1{{Table~\ref{#1}}}
\def\seqref#1{Eq.~(\ref{#1})}

%\floatname{algorithm}{Procedure}
%\renewcommand{\algorithmicrequire}{\textbf{Input:}}
%\renewcommand{\algorithmicensure}{\textbf{Output:}}

\title{Statistical methods for relative quantification of post-translational modifications in global proteomics experiments}
\author[1]{Tsung-Heng~Tsai}
%\author[1]{Olga~Vitek}
\affil[1]{Northeastern University, Boston, MA, USA}

\begin{document}
%\maketitle

%\newpage
%\tableofcontents

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{The motivating example}
\label{sec:motiv}

\todo{use a more representative plot}

\todo{change the order of the panels, so that the modified and unmodified peptides are in the first row, the modified peptides without a counterpart are in the second row left column (right column empty), and the unmodified peptides without a counterpart are in the third row right column (left column empty)}

\todo{the batches are not very clear from the plot}

\begin{figure}[h!]
\centering
\includegraphics[width=.9\textwidth]{fig/ex_profile}
\caption{Profile plot of peptide features categorized based on modification and matching status.}
\end{figure}

In a typical PTM experiment, modified and unmodified forms of the same peptide can have very different quantitative characteristics such as ionization efficiency and associated variability. Observed changes in signal intensities of modified peptides may result from changes in modification as well as changes in protein abundance. To account for this confounding factor, unmodified peptides that do not contain any modified sites are used for estimation of the underlying protein abundance. Data are often acquired in multiple batches, which is another factor affecting the quantitative characteristics. As shown in the profile plot, the number of peptide features and their intensities vary across batches.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\clearpage
\section{Overview of proposed approach and notation}
\label{sec:intro}

\todo{Make it more clear that this data structure reflects a particular site. Mention somewhere that we only consider peptides with one modifiable site}

\todo{The definition of a batch is not very clear. Can you define it in a way that is not very specific to these particular experiments?}

\todo{Why do you need stars in this table? Maybe simplify and say that this table is for unmodified peptides, and that a similar table is constructed for the modified peptides, with y*?}

The main objective is to detect modifications that are differentially expressed across experimental conditions.

%Unlike protein-level analysis, in which multiple quantified peptides are often observed, phospho-peptides are more frequently detected and quantified only once. There is a strong correlation between signal strength and reproducibility. When selecting sites for further study, investigators should pay close attention to the number of PSMs and the signal strength for peptides harboring each site.
%Simply calculating averages or medians of all peptides containing a given site might not reval the full complexity of cellular phosphorylation patterns. Singly and doubly phosphorylated forms of a peptide might be present at different levels. One must also not forget taht changes in total protein level are not refelcted in the phosphopeptide ratios. Whenever possible, separate protein-level measurements made from unmodified peptides should be performed and used to normalize phosphopeptide ratios.

\paragraph{Data structure of PTM quantification experiments.}
A set of fully-cleaved and/or partially-cleaved peptides with a same modification (e.g., ubiquitination) at one site are considered together. There are $I$ conditions and $J$ mass spectrometry runs (technical replicates) per condition, processed in $B$ batches in the experiment. In each batch, there are $K \times L$ runs. 
The modification is represented by multiple spectral features (peptide ions, distinguished by their cleavage sites, mass shifts and charge states), where the number of features in Batch $b$ is $K_b$. The log-intensity (base 2) of Feature $k$, in Run $j$ of Condition $i$ in Batch $b$ is denoted by $y_{b,ijk}^{\ast}$.

\begin{figure}[h!]
\centering
\begin{footnotesize}
\[
\begin{array}{l l c c c c c c c c c} 
\toprule
 &  & \multicolumn{4}{c}{\text{Condition }1} & \dots & \multicolumn{4}{c}{\text{Condition }I} \\ 
\cmidrule{3-6} 
\cmidrule{8-11} 
 &  &  \text{Run }1 & \text{Run }2 & \dots & \text{Run }J & \dots & \text{Run }1 & \text{Run }2 & \dots & \text{Run }J \\
\midrule
\multirow{4}{*}{Batch $1$} & \text{Feature }1 & y_{1,111}^{\ast} & y_{1,121}^{\ast} & \dots & y_{1,1J1}^{\ast} & \dots & y_{1,I11}^{\ast} & y_{1,I21}^{\ast} & \dots & y_{1,IJ1}^{\ast} \\
 & \text{Feature }2 & y_{1,112}^{\ast} & y_{1,122}^{\ast} & \dots & y_{1,1J2}^{\ast} & \dots & y_{1,I12}^{\ast} & y_{1,I22}^{\ast} & \dots & y_{1,IJ2}^{\ast}  \\
 & \dots & \dots & \dots & \dots & \dots & \dots & \dots & \dots & \dots & \dots \\
 & \text{Feature }K_1 & y_{1,11K_1}^{\ast} & y_{1,12K_1}^{\ast} & \dots & y_{1,1JK_1}^{\ast} & \dots & y_{1,I1K_1}^{\ast} & y_{1,I2K_1}^{\ast} & \dots & y_{1,IJK_1}^{\ast} \\
\midrule
\dots & \dots & \dots & \dots & \dots & \dots & \dots & \dots & \dots & \dots & \dots \\
\dots & \dots & \dots & \dots & \dots & \dots & \dots & \dots & \dots & \dots & \dots \\
\midrule
\multirow{4}{*}{Batch $b$} & \text{Feature }1 & y_{b,111}^{\ast} & y_{b,121}^{\ast} & \dots & y_{b,1J1}^{\ast} & \dots & y_{b,I11}^{\ast} & y_{b,I21}^{\ast} & \dots & y_{b,IJ1}^{\ast} \\
 & \text{Feature }2 & y_{b,112}^{\ast} & y_{b,122}^{\ast} & \dots & y_{b,1J2}^{\ast} & \dots & y_{b,I12}^{\ast} & y_{b,I22}^{\ast} & \dots & y_{b,IJ2}^{\ast}  \\
 & \dots & \dots & \dots & \dots & \dots & \dots & \dots & \dots & \dots & \dots \\
 & \text{Feature }K_b & y_{b,11K_b}^{\ast} & y_{b,12K_b}^{\ast} & \dots & y_{b,1JK_b}^{\ast} & \dots & y_{b,I1K_b}^{\ast} & y_{b,I2K_b}^{\ast} & \dots & y_{b,IJK_b}^{\ast} \\
\midrule
\dots & \dots & \dots & \dots & \dots & \dots & \dots & \dots & \dots & \dots & \dots \\
\dots & \dots & \dots & \dots & \dots & \dots & \dots & \dots & \dots & \dots & \dots \\
\midrule
\multirow{4}{*}{Batch $B$} & \text{Feature }1 & y_{B,111}^{\ast} & y_{B,121}^{\ast} & \dots & y_{B,1J1}^{\ast} & \dots & y_{B,I11}^{\ast} & y_{B,I21}^{\ast} & \dots & y_{B,IJ1}^{\ast} \\
 & \text{Feature }2 & y_{B,112}^{\ast} & y_{B,122}^{\ast} & \dots & y_{B,1J2}^{\ast} & \dots & y_{B,I12}^{\ast} & y_{B,I22}^{\ast} & \dots & y_{B,IJ2}^{\ast}  \\
 & \dots & \dots & \dots & \dots & \dots & \dots & \dots & \dots & \dots & \dots \\
 & \text{Feature }K_B & y_{B,11K_B}^{\ast} & y_{B,12K_B}^{\ast} & \dots & y_{B,1JK_B}^{\ast} & \dots & y_{B,I1K_B}^{\ast} & y_{B,I2K_B}^{\ast} & \dots & y_{B,IJK_B}^{\ast} \\
%\midrule
%\multirow{4}{*}{\text{Batch }1} & \text{Feature }1 & y_{b,111} & y_{b,121} & \dots & y_{b,1J1} & \dots & y_{b,I11} & y_{b,I21} & \dots & y_{b,IJ1} \\
% & \text{Feature }2 & y_{b,112} & y_{b,122} & \dots & y_{b,1J2} & \dots & y_{b,I12} & y_{b,I22} & \dots & y_{b,IJ2}  \\
% & \dots & \dots & \dots & \dots & \dots & \dots & \dots & \dots & \dots & \dots \\
% & \text{Feature }L & y_{b,11L} & y_{b,12L} & \dots & y_{b,1JL} & \dots & y_{b,I1L} & y_{b,I2L} & \dots & y_{b,IJL} \\
\bottomrule
\end{array}
\]
\end{footnotesize}
\caption{Representation of the data of modified peptides at one site, measured in $B$ batches with $I$ conditions and $J$ replicate runs. Modification level is quantified by multiple spectral features (peptide ions). Some spectral features can be missing (shown as $-$), either randomly in individual runs or completely from a particular form. In real practice, the number of runs can vary across conditions. \label{fig:dtable}}
\end{figure}

\sfigref{fig:dtable} shows an example data representation of modified peptide ions at one site. The modification measured in multiple batches is represented by different numbers of features. Unmodified peptides from the same protein are sometimes also measured, which may provide additional evidence on the underlying protein abundance. The features of unmodified peptides are denoted similarly. That is, the log-intensity (base 2) of Feature $l$, in Run $j$ of Condition $i$ in Batch $b$ is denoted by $y_{b,ijl}$, where $l \in \{1, \ldots , L_b \}$. To address the goal, statistical analysis needs to summarize values in this table using appropriate statistical models, translate the goal into a model-based quantity of interest, and draw inference (i.e., characterize the uncertainty) about the quantity.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Existing method: two-sample $t$-test}
\label{sec:ttest}

Two-sample $t$-test is based on the null hypothesis stating that there is no difference in mean modification abundance between Conditions $i$ and $i'$ against the alternative: 
\begin{align*}
H_{0}: \mu_{i}^{\ast} - \mu_{i'}^{\ast} &= 0\\
H_{a}: \mu_{i}^{\ast} - \mu_{i'}^{\ast} &\neq 0
\end{align*}
The test is not directly applicable to a problem with batches of data. Several \textit{ad-hoc} approaches may be used. Three commonly used approaches are 1) $t$-test (no batch): ignoring batch effects when applying $t$-test, 2) $t$-test (most significant batch): applying $t$-test in each batch and drawing conclusions based on the most significant batch, and 3) $t$-test (all batch): applying $t$-test in each batch and drawing conclusions based on all batches. Although simple, these \textit{ad-hoc} methods lack statistical justification. We reconstruct their meanings below and characterize their statistical properties under various forms of batch effects in \secref{sec:sim}. The abundance in each run is taken as input each of the three methods and is often estimated by sum of peak intensities, e.g., the abundance estimate for Run $j$ of Condition $i$ in Batch $b$ is given by
\[
y_{b, ij+} = \sum_{k=1}^{K_b} y_{b, ijk}
\]

%In our implementation of the approach using all batches, estimation of a change is the average over all batches and the readout for statistical significance (i.e., $p$-value) is based on the least significant batch.

\paragraph{Two-sample $t$-test (no batch).}
In this approach, the modification abundance is estimated as the average over runs in all batches. For example, the estimate for Group $i$ is given by
\[
\hat{\mu}_{i}^{\ast} = \frac{1}{BJ} \sum_{b=1}^{B} \sum_{j=1}^{J} y_{b, ij+}.
\]
This is equivalent to fitting a linear model with only an intercept, which does not account for different characteristics across batches.

\paragraph{Two-sample $t$-test (most significant batch).} 
The modification abundance is estimated in all batches, i.e., 
\[
\hat{\mu}_{b, i}^{\ast} = \frac{1}{J} \sum_{j=1}^{J} y_{b, ij+},
\]
where $\mathrm{SE} (\hat{\mu}_{b, i}^{\ast})$ is the standard error associated with the estimate. 
The test statistic in Batch $b$ is 
\[
t_{b} = \frac{\hat{\mu}_{b, i}^{\ast}}{\mathrm{SE} (\hat{\mu}_{b, i}^{\ast})}
\]
To summarize the evidence across batches, one strategy is to use the most significant batch. With balanced design, the batch with the largest test statistic across all batch is selected, i.e., 
\[
b_{\text{max}} = \{ b \mid t_b = \max_{b'=1, \ldots, B} | t_{b'} | \}.
\] 
The change in modification abundance is estimated as
\[
\hat{\mu}_{i}^{\ast} - \hat{\mu}_{i'}^{\ast} = \hat{\mu}_{b_{\text{max}}, i}^{\ast} - \hat{\mu}_{b_{\text{max}}, i'}^{\ast}
\]
and the test statistic for the $t$-test is given by
\[
\frac{\hat{\mu}_{i}^{\ast} - \hat{\mu}_{i'}^{\ast}}{\mathrm{SE}\left( \hat{\mu}_{i}^{\ast} - \hat{\mu}_{i'}^{\ast} \right)} 
= \frac{ \hat{\mu}_{b_{\text{max}}, i}^{\ast} - \hat{\mu}_{b_{\text{max}}, i'}^{\ast} }{\left[ \mathrm{SE}(\hat{\mu}_{b_{\text{max}}, i}^{\ast})^2 + \mathrm{SE}(\hat{\mu}_{b_{\text{max}}, i'}^{\ast})^2 \right]^{1/2}},
\]
%where $\hat{\sigma}_{\pi, ijk}^{2}$ and $\hat{\sigma}_{\pi, ijk'}^{2}$ are estimated as sample variances.

\paragraph{Two-sample $t$-test (all batch).}
A change in modification abundance is considered as statistically significant only if such change is significant in all batches. This is equivalent to take the smallest test statistic across batches for the $t$-test: 
\[
\frac{\hat{\mu}_{i}^{\ast} - \hat{\mu}_{i'}^{\ast}}{\mathrm{SE}\left( \hat{\mu}_{i}^{\ast} - \hat{\mu}_{i'}^{\ast} \right)} 
= \frac{ \hat{\mu}_{b_{\text{min}}, i}^{\ast} - \hat{\mu}_{b_{\text{min}}, i'}^{\ast} }{\left[ \mathrm{SE}(\hat{\mu}_{b_{\text{min}}, i}^{\ast})^2 + \mathrm{SE}(\hat{\mu}_{b_{\text{min}}, i'}^{\ast})^2 \right]^{1/2}},
\]
where 
\[
b_{\text{min}} = \{ b \mid t_b = \min_{b'=1, \ldots, B} | t_{b'} | \}.
\]

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Proposed statistical modeling and inference}

To characterize the observed feature intensities, different levels of variations are expressed in consideration of the following factors: modification, batch, condition, run, and feature. As the number of peptide features and their intensities are different across batches, variances of feature intensities are therefore expected to vary across batches. Also, different degrees of variability in the intensities of modified and unmodified features are present in the data; therefore, they are expressed by separate models.

\todo{identify and present the best model that we advocate for}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Linear models and parameter estimation}

The observed log-intensity of a modified peptide feature in batch $b$ is denoted by $y_{b,ijk}^{\ast}$ and represented as
\[
y_{b,ijk}^{\ast} = \psi_{b}^{\ast} + C_{b,i}^{\ast} + R_{b,j(i)}^{\ast} + F_{b,k}^{\ast} + (R \times F)_{b,ijk}^{\ast},
\]
where the effects of condition and feature are modeled as fixed effects: 
\[
\sum_{i=1}^{I} C_{b,i}^{\ast} = 0, \quad \sum_{k=1}^{K_b} F_{b,k}^{\ast} = 0,
\]
and the effects of run and its interaction with feature are considered as random effects arising from normal distribution with mean $0$: 
\[
R_{b,j(i)}^{\ast} = \gamma_{b,j(i)}^{\ast} \stackrel{\text{iid}}{\sim} \mathcal{N}(0, \sigma_{\gamma_b^{\ast}}^{2}), \qquad
(R \times F)_{b,ijk}^{\ast} = \epsilon_{b,ijk}^{\ast} \stackrel{\text{iid}}{\sim} \mathcal{N}(0, \sigma_{\epsilon_b^{\ast}}^{2}).
\]
Similarly, the observed log-intensity of an unmodified peptide feature is denoted by $y_{b,ijk}$ and represented as 
\[
y_{b,ijl} = \psi_{b} + C_{b,i} + R_{b,j(i)} + F_{b,l} + (R \times F)_{b,ijl}, 
\]
where the effects of condition and feature are modeled as fixed effects: 
\[
\sum_{i=1}^{I} C_{b,i} = 0, \quad \sum_{l=1}^{L_b} F_{b,l} = 0,
\]
and 
\[
R_{b,j(i)} = \gamma_{b,j(i)} \stackrel{\text{iid}}{\sim} \mathcal{N}(0, \sigma_{\gamma_b}^{2}), \qquad
(R \times F)_{b,ijl} = \epsilon_{b,ijl} \stackrel{\text{iid}}{\sim} \mathcal{N}(0, \sigma_{\epsilon_b}^{2}).
\]

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Run-level summarization of feature intensities}

The summarization of feature intensities at each site is carried out as in the sub-plot model of MSstats, which involves
\begin{enumerate}
\item Imputation of censored missing values
\item Summarization of feature intensities using Tukey's median polish (TMP), where the summarized log-intensity is denoted by $\hat{y}_{b, ij}$
\end{enumerate}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Model-based inference of the underlying abundance}

We distinguish two candidate models accounting for batch effects: 1) \emph{per-batch model}, which assumes changes between conditions and variability in the run-level summaries are not identical across batches, and each batch needs to be modeled separately, and 2) \emph{all-batch model}, which assumes differences between conditions are consistent across batches (i.e., no interaction effect between condition and batch) and identical variability in the run-level summarization is shared across batches. The details for each model are given below, with a summary in Table XX.

\paragraph{Per-batch model.}
Within Batch $b$, the abundance of the modification in each run is represented as 
\[
\hat{y}_{b, ij}^{\ast} = \psi_{b}^{\ast} + C_{b, i}^{\ast} + R_{b, j(i)}^{\ast},
\]
where $\sum_{i=1}^{I} C_{b, i}^{\ast} = 0$, $R_{b, j(i)}^{\ast} = \gamma_{b, j(i)}^{\ast} \stackrel{\text{iid}}{\sim} \mathcal{N}(0, \sigma_{\gamma_{b}^{\ast}}^{2})$.
Similarly, the model for protein abundance in each run is given by
\[
\hat{y}_{b, ij} = \psi_{b} + C_{b, i} + R_{b, j(i)},
\]
where $\sum_{i=1}^{I} C_{b, i} = 0$, $R_{b, j(i)} = \gamma_{b, j(i)} \stackrel{\text{iid}}{\sim} \mathcal{N}(0, \sigma_{\gamma_{b}}^{2})$. The expected values of log-abundances of the modification and protein in group $i$ are denoted by $\mu_{i}^{\ast}$ and $\mu_{i}$, respectively, and the values are defined as the average over the estimates in all batches:
\begin{align*}
\mu_{i}^{\ast} &= \psi^{\ast} + C_{i}^{\ast} = \frac{1}{B} \sum_{b=1}^{B} (\psi_{b}^{\ast} + C_{b, i}^{\ast})\\
\mu_{i} &= \psi + C_{i} = \frac{1}{B} \sum_{b=1}^{B} (\psi_{b} + C_{b, i})
%\mu_{i} &= \psi + C_{i}.
\end{align*}

\paragraph{All-batch model.} The whole-plot model for the modification abundance in each run is represented as 
\[
\hat{y}_{b, ij}^{\ast} = \psi^{\ast} + \delta_{b}^{\ast} + C_{i}^{\ast} + R_{b, j(i)}^{\ast},
\]
where $\sum_{b=1}^{B} \delta_{b}^{\ast} = 0$, $\sum_{i=1}^{I} C_{i}^{\ast} = 0$, $R_{b, j(i)}^{\ast} = \gamma_{b, j(i)}^{\ast} \stackrel{\text{iid}}{\sim} \mathcal{N}(0, \sigma_{\gamma^{\ast}}^{2})$.
Similarly, the model for protein abundance in each run is given by
\[
\hat{y}_{b, ij} = \psi + \delta_{b} + C_{i} + R_{b, j(i)},
\]
where $\sum_{b=1}^{B} \delta_{b} = 0$, $\sum_{i=1}^{I} C_{i} = 0$, $R_{b, j(i)} = \gamma_{b, j(i)} \stackrel{\text{iid}}{\sim} \mathcal{N}(0, \sigma_{\gamma}^{2})$. The expected values of log-abundances of the modification and protein in group $i$ are defined as $\mu_{i}^{\ast}$ and $\mu_{i}$, respectively. The values are represented as follows
\begin{align*}
\mu_{i}^{\ast} &= \psi^{\ast} + C_{i}^{\ast}\\
\mu_{i} &= \psi + C_{i}
\end{align*}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Model-based testing}
Hypothesis testing on the log-abundances of the modifications in Conditions $i$ and $i'$ is performed to detect systematic changes in modification abundance. The hypothesis states that there is no difference in log-abundance of the modification between groups $i$ and $i'$
\begin{align*}
H_{0}: \Delta = \mu_{i}^{\ast} - \mu_{i'}^{\ast} &= 0 \\
H_{a}: \Delta = \mu_{i}^{\ast} - \mu_{i'}^{\ast} &\neq 0
\end{align*}
The difference in modification abundance, $\Delta = \mu_{i}^{\ast} - \mu_{i'}^{\ast}$, is estimated with the per-batch model and all-batch model as follows.

\paragraph{Per-batch model.} 
The log of fold change in modification abundance is estimated as the average over batches
\[
\hat{\Delta} = \frac{1}{B} \sum_{b=1}^{B} \left[ \frac{1}{J} \left( \hat{y}_{b,i+}^{\ast} - \hat{y}_{b,i'+}^{\ast} \right) \right],
\]
and the standard error of the estimate is 
\[
\mathrm{SE}(\hat{\Delta}) = \left[ \left(\frac{1}{B}\right)^2 \cdot \frac{2}{J} \sum_{b=1}^{B} \hat{\sigma}_{\gamma_b^{\ast}}^{2} \right]^{1/2}
\]
The test statistic $\hat{\Delta} / \mathrm{SE}(\hat{\Delta})$ is compared against the $t$ distribution. The degrees of freedom of the $t$ distribution are approximated as
%\[
%\frac{ \left( \sum_{b=1}^{B} \hat{\sigma}_{\gamma_{b}^{\ast}}^{2} \right)^2 }
%{ \sum_{b=1}^{B} \left[  \frac{\hat{\sigma}_{\gamma_{b}^{\ast}}^{4}}{ \mathrm{df}(\gamma_b^{\ast}) }  \right] }.
%\]
\[
\left( \sum_{b=1}^{B} \hat{\sigma}_{\gamma_{b}^{\ast}}^{2} \right)^2 \bigg/
\sum_{b=1}^{B} \left[  \frac{\hat{\sigma}_{\gamma_{b}^{\ast}}^{4}}{ \mathrm{df}(\gamma_b^{\ast}) } \right].
\]

\paragraph{All-batch model.}
The log of fold change in modification abundance is estimated as the average over runs (and batches)
\[
\hat{\Delta} = \frac{1}{B} \sum_{b=1}^{B} \left[ \frac{1}{J} \left( \hat{y}_{b,i+}^{\ast} - \hat{y}_{b,i'+}^{\ast} \right) \right]
= \frac{1}{BJ} \left( \hat{y}_{+,i+}^{\ast} - \hat{y}_{+,i'+}^{\ast} \right),
\]
and the standard error of the estimate is 
\[
\mathrm{SE}(\hat{\Delta}) = \left( \frac{2}{BJ} \cdot \hat{\sigma}_{\gamma^{\ast}}^{2} \right)^{1/2}
\]
The test statistic $\hat{\Delta} / \mathrm{SE}(\hat{\Delta})$ is compared against the $t$ distribution with degrees of freedom $\mathrm{df}(\gamma^{\ast})$.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Adjustment with protein abundance}

Adjusted by protein abundance, the hypothesis states that there is no difference in log-abundance of the modification between groups $i$ and $i'$
\begin{align*}
H_{0}: \Delta = \left( \mu_{i}^{\ast} - \mu_{i'}^{\ast} \right) - \left( \mu_{i} - \mu_{i'} \right) &= 0 \\
H_{a}: \Delta = \left( \mu_{i}^{\ast} - \mu_{i'}^{\ast} \right) - \left( \mu_{i} - \mu_{i'} \right) &\neq 0
\end{align*}
Details of the testing with the per-batch model and all-batch model are provided below.

\paragraph{Per-batch model.}
The log of fold change in normalized modification abundance, $\Delta$, is estimated by 
\[
\hat{\Delta} = \frac{1}{B} \sum_{b=1}^{B} \left[ \frac{1}{J} \left( \hat{y}_{b,i+}^{\ast} - \hat{y}_{b,i'+}^{\ast} \right) \right] - \frac{1}{B} \sum_{b=1}^{B} \left[ \frac{1}{J} \left( \hat{y}_{b,i+} - \hat{y}_{b,i'+} \right) \right],
\]
%\begin{align*}
%\hat{\Delta} &= \left( \hat{\mu}_{i}^{\ast} - \hat{\mu}_{i'}^{\ast} \right) - \left( \hat{\mu}_{i} - \hat{\mu}_{i'} \right) \\
%&= \frac{1}{B} \sum_{b=1}^{B} \left[ \frac{1}{J} \left( \hat{y}_{b,i+}^{\ast} - \hat{y}_{b,i'+}^{\ast} \right) \right] - \frac{1}{B} \sum_{b=1}^{B} \left[ \frac{1}{J} \left( \hat{y}_{b,i+} - \hat{y}_{b,i'+} \right) \right],
%\end{align*}
and the standard error of the estimate $\mathrm{SE}(\hat{\Delta})$ is 
\[
\left[ \left(\frac{1}{B}\right)^2 \cdot \left( \frac{2}{J} \right) \cdot \sum_{b=1}^{B} \left( \hat{\sigma}_{\gamma_b^{\ast}}^{2} + \hat{\sigma}_{\gamma_b}^{2} \right) \right]^{1/2}.
\]
The test statistic $\hat{\Delta} / \mathrm{SE}(\hat{\Delta})$ is compared against the $t$ distribution, with degrees of freedom approximated by
%\[
%\frac{ \left[ \sum_{b=1}^{B} \left( \hat{\sigma}_{\gamma_{b}^{\ast}}^{2} + \hat{\sigma}_{\gamma_{b}}^{2} \right) \right]^2 }
%{ \sum_{b=1}^{B} \left[ \frac{\hat{\sigma}_{\gamma_{b}^{\ast}}^{4}}{\mathrm{df}(\gamma_b^{\ast})} + \frac{\hat{\sigma}_{\gamma_{b}}^{4}}{  \mathrm{df}(\gamma_b)} \right] }.
%\]
\[
\left[ \sum_{b=1}^{B} \left( \hat{\sigma}_{\gamma_{b}^{\ast}}^{2} + \hat{\sigma}_{\gamma_{b}}^{2} \right) \right]^2 \bigg/
\sum_{b=1}^{B} \left[ \frac{\hat{\sigma}_{\gamma_{b}^{\ast}}^{4}}{\mathrm{df}(\gamma_b^{\ast})} + \frac{\hat{\sigma}_{\gamma_{b}}^{4}}{  \mathrm{df}(\gamma_b)} \right].
\]

\paragraph{All-batch model.} 
The log of fold change in normalized modification abundance $\Delta$ which is estimated as 
\[
\hat{\Delta} = \frac{1}{BJ} \left( \hat{y}_{+,i+}^{\ast} - \hat{y}_{+,i'+}^{\ast} \right) - \frac{1}{BJ} \left( \hat{y}_{+,i+} - \hat{y}_{+,i'+} \right),
\]
and its standard error $\mathrm{SE}(\hat{\Delta})$ is 
\[
\left[ \frac{2}{BJ} \left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right) \right] ^{1/2}.
\]
The test statistic $\hat{\Delta} / \mathrm{SE}(\hat{\Delta})$ is compared against the $t$ distribution, with degrees of freedom approximated by
%\[
%\frac{ \left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right)^2 }
%{ \frac{\hat{\sigma}_{\gamma^{\ast}}^{4}}{\mathrm{df}(\gamma^{\ast})} + \frac{\hat{\sigma}_{\gamma}^{4}}{  \mathrm{df}(\gamma)} }.
%\]
\[
\left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right)^2 \bigg/
\left( \frac{\hat{\sigma}_{\gamma^{\ast}}^{4}}{\mathrm{df}(\gamma^{\ast})} + \frac{\hat{\sigma}_{\gamma}^{4}}{  \mathrm{df}(\gamma)} \right).
\]

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Computer simulation}
\label{sec:sim}

The proposed statistical approaches were evaluated using computer simulation. In particular, their properties under batch effects and adjustment with respect to unmodified peptides were evaluated.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Batch effects}

The proposed statistical approaches were evaluated and compared with those approaches based on two-sample $t$-test (\secref{sec:ttest}) using computer simulation. Two batches of data were generated, where various forms of batch effects were simulated, including difference in signal intensities across batches, difference in variability across batches, and interaction between batch and condition (i.e., change between condition affected by batch). The following interaction scenarios were considered in the simulation: 1) no interaction between condition and batch, i.e., same change across conditions in both batches, 2) positive interaction as $25$\% greater change in the batch of higher level, and 3) negative interaction as $25$\% lower change in the batch of higher level. In real experiment, multiple conditions are often compared together. Whereas $t$-test uses measurements from the two conditions being compared, the proposed models leverage measurements in all conditions for the inference of the underlying abundance. To highlight this distinction, multiple conditions of data were generated, where only one condition was simulated with a systematic change. Below are the parameters used in the simulation: 
\begin{itemize}
\item Mean of log-intensity: $25$
\item Increased intensity level of Batch 2 versus Batch 1: $0$, $1$, $2$
\item Standard deviations of log-intensity: 
\begin{itemize}
\item Identical variances across batches: $0.2$ in both Batch 1 and Batch 2
\item Different variances across batches: $0.2$ and $0.3$ in Batch 1 and Batch 2, respectively
\end{itemize}
\item Difference between conditions in Batch 1: $0.5$, $0.75$, $1$
\item Number of replicates: $2$, $3$, $5$
\item Number of conditions: $2$, $3$, $4$
\item Number of realizations: $500$
\end{itemize}

\todo{comment on which combination of parameters was observed most often in the experimental dataset}

\begin{figure}[h!]
\centering
\includegraphics[width=.825\textwidth]{sim/synnull_est_2}
\caption{Estiamtion based on the most statistically significant batch with $t$-test was highly variable and frequently biased. The observation is consistent in all the simulated scenarios. \label{fig:synnull_est}}
\end{figure}

\begin{figure}[h!]
\centering
\includegraphics[width=.825\textwidth]{sim/synnull_fpr}
\caption{The proposed per-batch model better calibrated Type I error rate. Similar results were observed in the cases with positive and negative interactions. \label{fig:synnull_fpr}}
\end{figure}

\clearpage
\begin{figure}[h!]
\centering
\includegraphics[width=.85\textwidth]{sim/synnull_pwr_2}\\
\includegraphics[width=.85\textwidth]{sim/synnull_pwr_3}
\caption{The proposed model improved power with small sample sizes in almost all the scenarios, under various forms of batch effects including difference in intensity level (higher on the right) and difference in variability. Using $t$-test while ignoring batch effect gave similar performance in special cases with no difference in intensity level between batches, but its performance dramatically decreased in general cases. The proposed methods gave consistently improved performance compared with other methods by properly characterizing batch effects and leveraging all available information. \label{fig:synnull_pwr}}
\end{figure}

\clearpage
\begin{figure}[h!]
\centering
\includegraphics[width=.85\textwidth]{sim/synpos_pwr_3}\\
\includegraphics[width=.85\textwidth]{sim/synneg_pwr_3}
\caption{With 3 replicates and under the scenarios with either positive or negative interaction, the proposed models consistently improved power over other methods. Similar results were observed with 2 replicates. As in those cases with no interaction, the approaches using two-sample $t$-test did not benefit from data in additional conditions. Also, ignoring batch effects as applying $t$-test (no batch) lost power dramatically. \label{fig:syn_pwr_3}}
\end{figure}

\begin{figure}[h!]
\centering
\includegraphics[width=.85\textwidth]{sim/synnull_pwr_5}
\caption{Ignoring batch effects lost power. The negative impact was only partially reduced by increasing the sample size to 5. Similar patterns were observed in other simulated scenarios. \label{fig:synnull_pwr_5}}
\end{figure}

\clearpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Protein-level adjustment}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{One site per protein}

Differential levels of modified peptides may be due to changes in modification or change in protein abundance. Adjustment with respect to unmodified peptides can correct for the confounding factor, with a cost of increased uncertainty. We evaluated such adjustment by computer simulation. Below are the parameters used in the simulation: 
\begin{itemize}
\item Mean of log-intensity: $25$
\item Standard deviations of log-intensity of modified peptides: 
\begin{itemize}
\item Identical variances across batches: $0.2$ in both Batch 1 and Batch 2
\item Different variances across batches: $0.2$ and $0.3$ in Batch 1 and Batch 2, respectively
\end{itemize}
\item Standard deviations of log-intensity of unmodified peptides: $0.2$ in both batches
\item Difference in modification level between conditions: $0$, $0.5$
\item Difference in protein level between conditions: $0$, $0.5$
\item Number of replicates: $2$, $3$, $5$
\item Number of conditions: $2$, $3$, $4$
\item Number of realizations: $500$
\end{itemize}

\begin{figure}[h!]
\centering
\includegraphics[width=.85\textwidth]{sim/prot_fpr}
\caption{
All the proposed methods well calibrated the Type I error rate when there was no change in protein abundance (leftmost two columns). When the modification changes were entirely due to the changes in protein abundance across conditions, analysis without accounting for the protein-level changes resulted in off-target, high false positive rates. \label{fig:prot_fpr}}
\end{figure}

\begin{figure}[h!]
\centering
\includegraphics[width=.85\textwidth]{sim/prot_pwr}
\caption{Adjustment with respect to unmodified peptides increase uncertainty in the estimation and testing. When there was no change in protein abundance (leftmost two columns), the approaches with no adjustment resulted in slightly higher power. However, their performances decreased significantly when such confounding indeed existed (rightmost two columns). \label{fig:prot_pwr}}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Multiple sites per protein}

Expression levels of modification sites are adjusted based on the abundance of their originating protein. Since the same reference is used for all the sites, it introduces correlation between estimates and test statistics for those sites. This may cause issues in controlling false discovery rate (FDR). We investigated the property by simulating data of two conditions and 1000 proteins with different numbers of modification sites. A fraction (50\% or 80\%) of the 1000 proteins had no changes between conditions while systematic changes were simulated for the rest of the proteins. Multiple testing correction was performed using the Benjamini and Hochberg's method. Performances of the considered approaches were assessed by their power and FDR. The power was calculated as the fraction of proteins with adjusted $p$-values $<0.05$ among those proteins with true differences between conditions, while the FDR was calculated as the fraction of proteins with adjusted $p$-values $<0.05$ among the proteins with true differences.

\begin{figure}[h!]
\centering
\includegraphics[width=.85\textwidth]{sim/prot_fdr}
\caption{Correlation between test statistics across sites due to protein-level adjustment did not affect the expected level of FDR under the considered scenarios. The Benjamini-Hochberg procedure was used to control the FDR. \label{fig:prot_fdr}}
\end{figure}

\todo{Storey's method?}
\todo{Proportion of null doesn't seem to be a big factor}

\clearpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Appendix}

This section contains full details of some plots that we are trying to simplify. It is only for the reference during the preparation and will be removed later.

\end{document}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\section{Modeling and inference: one batch}
%
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\subsection{Linear models and parameter estimation}
%\paragraph{Model.}
%For a protein, the observed log-intensity of a modified peptide feature is denoted by $y_{ijk}^{\ast}$ and represented using a linear mixed model
%\[
%y_{ijk}^{\ast} = \psi^{\ast} + \text{Group}_{i}^{\ast} + \text{Run}_{j(i)}^{\ast} + \text{Feature}_{k}^{\ast} + (\text{Run} \times \text{Feature})_{ijk}^{\ast},
%\]
%\[
%i = 1, \ldots, I, \quad j = 1, \ldots, J, \quad k = 1, \ldots, K
%\]
%where $\psi^{\ast}$ is the expected value of log-abundance of the modification, and the effects of condition, run, and feature characterize how the log-intensity of each feature deviates from this reference. The interaction between run and feature $(\text{Run} \times \text{Feature})_{ijk}^{\ast}$ is essentially viewed as a random noise. The effects of condition and feature are modeled as fixed effects: 
%\[
%\sum_{i=1}^{I} \text{Group}_{i}^{\ast} = 0, \quad \sum_{k=1}^{K} \text{Feature}_{k}^{\ast} = 0,
%\]
%and the effects of run and its interaction with feature are considered as random effects arising from normal distribution with mean $0$: 
%\[
%\text{Run}_{j(i)}^{\ast} = \gamma_{j(i)}^{\ast} \stackrel{\text{iid}}{\sim} \mathcal{N}(0, \sigma_{\gamma^{\ast}}^{2}), \qquad 
%(\text{Run} \times \text{Feature})_{ijk}^{\ast} = \epsilon_{ijk}^{\ast} \stackrel{\text{iid}}{\sim} \mathcal{N}(0, \sigma_{\epsilon^{\ast}}^{2}).
%\]
%%\[
%%\text{Run}_{j(i)}^{\ast} = \gamma_{j(i)}^{\ast} \stackrel{\text{iid}}{\sim} \mathcal{N}(0, \sigma_{\gamma^{\ast}}^{2}),
%%\]
%%\[
%%(\text{Run} \times \text{Feature})_{ijk}^{\ast} = \epsilon_{ijk}^{\ast} \stackrel{\text{iid}}{\sim} \mathcal{N}(0, \sigma_{\epsilon^{\ast}}^{2}).
%%\]
%Similarly, the observed log-intensity of an unmodified peptide feature is denoted by $y_{ijk}$ and represented as 
%\[
%y_{ijl} = \psi + \text{Group}_{i} + \text{Run}_{j(i)} + \text{Feature}_{l} + (\text{Run} \times \text{Feature})_{ijl}, 
%\]
%where $\psi$ the expected value of log-abundance of the protein, the effects of condition and feature are modeled as fixed effects: 
%\[
%\sum_{i=1}^{I} \text{Group}_{i} = 0, \quad \sum_{l=1}^{L} \text{Feature}_{l} = 0,
%\]
%and 
%\[
%\text{Run}_{j(i)} = \gamma_{j(i)} \stackrel{\text{iid}}{\sim} \mathcal{N}(0, \sigma_{\gamma}^{2}), \qquad 
%(\text{Run} \times \text{Feature})_{ijl} = \epsilon_{ijl} \stackrel{\text{iid}}{\sim} \mathcal{N}(0, \sigma_{\epsilon}^{2}).
%\]
%
%\paragraph{Parameter estimation.} The parameter estimation is performed by using a sub-plot model to summarize feature intensities per run, followed by a whole-plot model to carry out the model-based inference of the underlying abundance. 
%
%\paragraph{Step 1 of parameter estimation: run-level summarization.} The sub-plot model is used for the summarization of feature intensities. This step involves 
%\begin{enumerate}
%\item Imputation of censored missing values
%\item Summarization of feature intensities using Tukey's median polish (TMP), where the summarized log-intensity is denoted by $\hat{y}_{ij}$
%\end{enumerate}
%
%\paragraph{Step 2 of parameter estimation: model-based inference of the underlying abundance.}
%The whole-plot model for the modification abundance in each run is represented as 
%\[
%\hat{y}_{ij}^{\ast} = \psi^{\ast} + \text{Group}_{i}^{\ast} + \text{Run}_{j(i)}^{\ast},
%\]
%where $\sum_{i=1}^{I} \text{Group}_{i}^{\ast} = 0$, $\text{Run}_{j(i)}^{\ast} = \gamma_{j(i)}^{\ast} \stackrel{\text{iid}}{\sim} \mathcal{N}(0, \sigma_{\gamma^{\ast}}^{2})$.
%Similarly, the model for protein abundance in each run is given by
%\[
%\hat{y}_{ij} = \psi + \text{Group}_{i} + \text{Run}_{j(i)},
%\]
%where $\sum_{i=1}^{I} \text{Group}_{i} = 0$, $\text{Run}_{j(i)} = \gamma_{j(i)} \stackrel{\text{iid}}{\sim} \mathcal{N}(0, \sigma_{\gamma}^{2})$. The expected values of log-abundances of the modification and protein in group $i$ are defined as $\mu_{i}^{\ast}$ and $\mu_{i}$, respectively. The values are represented as follows
%\begin{align*}
%\mu_{i}^{\ast} &= \psi^{\ast} + \text{Group}_{i}^{\ast} \\
%\mu_{i} &= \psi + \text{Group}_{i}.
%\end{align*}
%
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\subsection{Model-based inference}
%Hypothesis testing for differential modification.
%
%\paragraph{Hypothesis $H_{0}^{(1)}$.} The hypothesis states that there is no difference in log-abundance of the modification between groups $i$ and $i'$
%\begin{align*}
%H_{0}^{(1)}: \mu_{i}^{\ast} - \mu_{i'}^{\ast} &= 0 \\
%H_{a}^{(1)}: \mu_{i}^{\ast} - \mu_{i'}^{\ast} &\neq 0
%\end{align*}
%The estimated log of fold change in modification abundance is given by
%\[
%\hat{\mu}_{i}^{\ast} - \hat{\mu}_{i'}^{\ast} = \frac{1}{J} \left( \hat{y}_{i+}^{\ast} - \hat{y}_{i'+}^{\ast} \right),
%\]
%and the standard error of the estimate is 
%\[
%\mathrm{SE}\left( \hat{\mu}_{i}^{\ast} - \hat{\mu}_{i'}^{\ast} \right) = \sqrt{\frac{2}{J} \hat{\sigma}_{\gamma^{\ast}}^{2}}
%\]
%The test statistic $\left( \hat{\mu}_{i}^{\ast} - \hat{\mu}_{i'}^{\ast} \right) / \mathrm{SE}\left( \hat{\mu}_{i}^{\ast} - \hat{\mu}_{i'}^{\ast} \right)$ is compared against the $t$ distribution. The degrees of freedom of the $t$ distribution are $2J-2$.
%
%\paragraph{Hypothesis $H_{0}^{(2)}$.} The hypothesis states that there is no difference in log-abundance of the modification between groups $i$ and $i'$, normalized by the protein abundance 
%\begin{align*}
%H_{0}^{(2)}: \left( \mu_{i}^{\ast} - \mu_{i'}^{\ast} \right) - \left( \mu_{i} - \mu_{i'} \right) &= 0 \\
%H_{a}^{(2)}: \left( \mu_{i}^{\ast} - \mu_{i'}^{\ast} \right) - \left( \mu_{i} - \mu_{i'} \right) &\neq 0
%\end{align*}
%The estimated log of fold change in normalized modification abundance is given by
%\[
%\left( \hat{\mu}_{i}^{\ast} - \hat{\mu}_{i'}^{\ast} \right) - \left( \hat{\mu}_{i} - \hat{\mu}_{i'} \right) 
%= \frac{1}{J} \left( \hat{y}_{i+}^{\ast} - \hat{y}_{i'+}^{\ast} \right) - \frac{1}{J} \left( \hat{y}_{i+} - \hat{y}_{i'+} \right),
%\]
%and the standard error of the estimate is 
%\[
%\sqrt{\frac{2}{J} \left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right) }
%\]
%The test statistic is compared against the $t$ distribution with degrees of freedom approximated as:
%\[
%df = 
%\frac{ 2 (J-1) \left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right)^2 }{ \left( \hat{\sigma}_{\gamma^{\ast}}^{4} + \hat{\sigma}_{\gamma}^{4} \right) }.
%\]

%\paragraph{Hypothesis $H_{0}^{(1)}$.} The hypothesis states that there is no difference in log-abundance of the modification between groups $i$ and $i'$
%\begin{align*}
%H_{0}^{(1)}: \mu_{i}^{\ast} - \mu_{i'}^{\ast} &= 0 \\
%H_{a}^{(1)}: \mu_{i}^{\ast} - \mu_{i'}^{\ast} &\neq 0
%\end{align*}
%The log of fold change in modification abundance is estimated as the average over batches
%\[
%\hat{\mu}_{i}^{\ast} - \hat{\mu}_{i'}^{\ast} = 
%\frac{1}{B} \sum_{b=1}^{B} \left[ \frac{1}{J} \left( \hat{y}_{b,i+}^{\ast} - \hat{y}_{b,i'+}^{\ast} \right) \right],
%\]
%and the standard error of the estimate is 
%\[
%\mathrm{SE}\left( \hat{\mu}_{i}^{\ast} - \hat{\mu}_{i'}^{\ast} \right) = 
%\left[ \left(\frac{1}{B}\right)^2 \cdot \frac{2}{J} \sum_{b=1}^{B} \hat{\sigma}_{\gamma_b^{\ast}}^{2} \right]^{1/2}
%\]
%The test statistic $\left( \hat{\mu}_{i}^{\ast} - \hat{\mu}_{i'}^{\ast} \right) / \mathrm{SE}\left( \hat{\mu}_{i}^{\ast} - \hat{\mu}_{i'}^{\ast} \right)$ is compared against the $t$ distribution. The degrees of freedom of the $t$ distribution are approximated as
%\[
%\frac{ \left( \sum_{b=1}^{B} \hat{\sigma}_{\gamma_{b}^{\ast}}^{2} \right)^2 }
%%{ \sum_{b=1}^{B} \left( \hat{\sigma}_{\gamma_{b}^{\ast}}^{4} /  \mathrm{df}(\gamma_b^{\ast}) \right) }
%{ \sum_{b=1}^{B} \left[  \frac{\hat{\sigma}_{\gamma_{b}^{\ast}}^{4}}{ \mathrm{df}(\gamma_b^{\ast}) }  \right] }
%\]
%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\subsection{All-batch model}
%
%\paragraph{Step 1 of parameter estimation: run-level summarization.} As in the per-batch model.
%
%\paragraph{Step 2 of parameter estimation: model-based inference of the underlying abundance.}
%The whole-plot model for the modification abundance in each run is represented as 
%\[
%\hat{y}_{b, ij}^{\ast} = \psi^{\ast} + \delta_{b}^{\ast} + C_{i}^{\ast} + R_{b, j(i)}^{\ast},
%\]
%where $\sum_{b=1}^{B} \delta_{b}^{\ast} = 0$, $\sum_{i=1}^{I} C_{i}^{\ast} = 0$, $R_{b, j(i)}^{\ast} = \gamma_{b, j(i)}^{\ast} \stackrel{\text{iid}}{\sim} \mathcal{N}(0, \sigma_{\gamma^{\ast}}^{2})$.
%Similarly, the model for protein abundance in each run is given by
%\[
%\hat{y}_{b, ij} = \psi + \delta_{b} + C_{i} + R_{b, j(i)},
%\]
%where $\sum_{b=1}^{B} \delta_{b} = 0$, $\sum_{i=1}^{I} C_{i} = 0$, $R_{b, j(i)} = \gamma_{b, j(i)} \stackrel{\text{iid}}{\sim} \mathcal{N}(0, \sigma_{\gamma}^{2})$. The expected values of log-abundances of the modification and protein in group $i$ are defined as $\mu_{i}^{\ast}$ and $\mu_{i}$, respectively. The values are represented as follows
%\begin{align*}
%\mu_{i}^{\ast} &= \psi^{\ast} + C_{i}^{\ast}\\
%\mu_{i} &= \psi + C_{i}
%\end{align*}

### 2021-01-25 supplementary/ptm_sm.tex
Latex supplementary

%\documentclass[letterpaper,12pt]{article}
\documentclass{mcp}

\usepackage{graphics}

\usepackage{geometry}
%\usepackage{amsmath}
%\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{bm}
\usepackage{algorithmic}
\usepackage{algorithm}
\usepackage{rotating}
\usepackage[caption=false,font=footnotesize]{subfig}
\usepackage{color}
\usepackage{tikz}
\usetikzlibrary{shapes,arrows,backgrounds,calc,positioning,fit,petri,plotmarks}
\usepackage{multirow}
\usepackage{graphicx}
\usepackage{subfig}
\usepackage{url}
\usepackage{booktabs}
\usepackage{enumitem}
\usepackage{authblk}

\def\todo#1{{\color{red}[TODO: #1]}}

%\usepackage{caption} 
%\captionsetup[table]{skip=10pt}
\captionsetup{belowskip=12pt,aboveskip=4pt}

% strikkeout \sout
\usepackage[normalem]{ulem}

\usepackage{fullpage}
\usepackage{amsfonts,amsthm}
\usepackage{amsmath}
%\usepackage[fleqn]{amsmath}
%\usepackage{setspace}
%\usepackage{url}

%\usepackage[table,dvipsnames]{xcolor}
%\usepackage{textcomp}
%\usepackage{gensymb}

\usepackage{fancyhdr}
\pagestyle{fancy}
\lhead{\footnotesize \parbox{11cm}{Supplementary }}
\headsep = 25pt

%\usepackage{tabulary,multirow,multicol,rotating}

%% big wide hat
\usepackage{scalerel,stackengine}
\stackMath
\newcommand\reallywidehat[1]{%
\savestack{\tmpbox}{\stretchto{%
  \scaleto{%
    \scalerel*[\widthof{\ensuremath{#1}}]{\kern-.6pt\bigwedge\kern-.6pt}%
    {\rule[-\textheight/2]{1ex}{\textheight}}%WIDTH-LIMITED BIG WEDGE
  }{\textheight}% 
}{0.5ex}}%
\stackon[1pt]{#1}{\tmpbox}%
}
\parskip 1ex

\renewcommand{\thetable}{S\arabic{table}}
\renewcommand{\thefigure}{S\arabic{figure}}
%\renewcommand{\tablename}{Supplementary Table}
%\renewcommand{\figurename}{Supplementary Figure}

\def\sfigref#1{{Figure~\ref{#1}}}
\def\secref#1{Section~\ref{#1}}
\def\stabref#1{{Table~\ref{#1}}}
\def\seqref#1{Eq.~(\ref{#1})}

%\floatname{algorithm}{Procedure}
%\renewcommand{\algorithmicrequire}{\textbf{Input:}}
%\renewcommand{\algorithmicensure}{\textbf{Output:}}

\title{Statistical methods for relative quantification of post-translational modifications in global proteomics experiments}
\author[1]{Tsung-Heng~Tsai}
\author[2]{Erik~Verschueren}
\author[1]{Olga~Vitek}
\affil[1]{Northeastern University, Boston, MA, USA}
\affil[2]{Protein Chemistry, Genentech, South San Francisco, CA, USA}

\begin{document}
%\maketitle

%\newpage
\tableofcontents

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\clearpage
\section{Overview of proposed approach and notation}
\label{sec:intro}

This study addresses three major goals in the characterization of post-translational modifications (PTMs): a) relative PTM quantification, b) PTM significance analysis, i.e., to detect PTM sites that are differentially modified across experimental conditions, and c) statistical design of PTM experiments.

%Unlike protein-level analysis, in which multiple quantified peptides are often observed, phospho-peptides are more frequently detected and quantified only once. There is a strong correlation between signal strength and reproducibility. When selecting sites for further study, investigators should pay close attention to the number of PSMs and the signal strength for peptides harboring each site.
%Simply calculating averages or medians of all peptides containing a given site might not reval the full complexity of cellular phosphorylation patterns. Singly and doubly phosphorylated forms of a peptide might be present at different levels. One must also not forget that changes in total protein level are not refelcted in the phosphopeptide ratios. Whenever possible, separate protein-level measurements made from unmodified peptides should be performed and used to normalize phosphopeptide ratios.

\paragraph{Data structure of PTM quantification experiments.}
A set of fully-cleaved and/or partially-cleaved peptides containing a same PTM (e.g., ubiquitination) at one site are considered together. There are $I$ conditions and $J$ mass spectrometry runs (technical replicates) per condition in the experiment. The PTM site is represented by $K$ spectral features (peptide ions, distinguished by their cleavage residues and charge states). The log-intensity (base 2) of Feature $k$, in Run $j$ of Condition $i$ is denoted by $y_{ijk}$. To account for the underlying protein abundance, features corresponding to the unmodified peptides from the same protein are considered together, except those unmodified peptides containing a modified site to avoid the confounding effect due to the PTM. The log-intensity of Feature $l$ from the unmodified peptides in the same run is denoted by $y_{ijl}^{\ast}$. \sfigref{fig:dtable} shows an example data representation of modified peptide ions at one site and unmodified peptide ions of the same protein. Unmodified peptides from the same protein provide additional evidence on the underlying protein abundance, which needs to be integrated for PTM characterization. To address the goals of PTM characterization, statistical analysis needs to summarize values in this table using appropriate statistical models, translate the goal into a model-based quantity of interest, and draw inference (i.e., characterize the uncertainty) about the quantity.

\begin{figure}[h!]
\centering
\begin{footnotesize}
\[
\begin{array}{l l c c c c c c c c c} 
\toprule
 &  & \multicolumn{4}{c}{\text{Condition }1} & \dots & \multicolumn{4}{c}{\text{Condition }I} \\ 
\cmidrule{3-6} 
\cmidrule{8-11} 
 &  &  \text{Run }1 & \text{Run }2 & \dots & \text{Run }J & \dots & \text{Run }1 & \text{Run }2 & \dots & \text{Run }J \\
\midrule
\multirow{4}{*}{Modified} & \text{Feature }1 & y_{111} & y_{121} & \dots & y_{1J1} & \dots & y_{I11} & y_{I21} & \dots & y_{IJ1} \\
 & \text{Feature }2 & y_{112} & - & \dots & y_{1J2} & \dots & y_{I12} & y_{I22} & \dots & y_{IJ2}  \\
 & \dots & \dots & \dots & \dots & \dots & \dots & \dots & \dots & \dots & \dots \\
 & \text{Feature }K & y_{11K} & y_{12K} & \dots & y_{1JK} & \dots & y_{I1K} & y_{I2K} & \dots & y_{IJK} \\
\midrule
\multirow{4}{*}{Unmodified} & \text{Feature }1 & y_{111}^{\ast} & y_{121}^{\ast} & \dots & y_{1J1}^{\ast} & \dots & y_{I11}^{\ast} & y_{I21}^{\ast} & \dots & y_{IJ1}^{\ast} \\
 & \text{Feature }2 & y_{112}^{\ast} & y_{122}^{\ast} & \dots & y_{1J2}^{\ast} & \dots & - & - & \dots & -  \\
% & \text{Feature }2 & y_{112}^{\ast} & y_{122}^{\ast} & \dots & y_{1J2}^{\ast} & \dots & y_{I12}^{\ast} & y_{I22}^{\ast} & \dots & y_{IJ2}^{\ast}  \\
 & \dots & \dots & \dots & \dots & \dots & \dots & \dots & \dots & \dots & \dots \\
 & \text{Feature }L & y_{11L}^{\ast} & y_{12L}^{\ast} & \dots & y_{1JL}^{\ast} & \dots & y_{I1L}^{\ast} & y_{I2L}^{\ast} & \dots & y_{IJL}^{\ast} \\
\bottomrule
\end{array}
\]
%\[
%\begin{array}{l l c c c c c} 
%\toprule
% &  & \multicolumn{2}{c}{\text{Condition }1} &  & \multicolumn{2}{c}{\text{Condition }2} \\ 
%\cmidrule{3-4} 
%\cmidrule{6-7} 
% &  &  \text{Run }1 & \text{Run }2 &  & \text{Run }1 & \text{Run }2 \\
%\midrule
%\multirow{4}{*}{\begin{turn}{90} {\text{Modified}} \end{turn}} & \text{Feature }1 & y_{111} & y_{121} &  & y_{211} & y_{221} \\
%%\multirow{4}{*}{Modified} & \text{Feature }1 & y_{111} & y_{121} &  & y_{211} & y_{221} \\
% & \text{Feature }2 & y_{112} & - &  & y_{212} & y_{222} \\
% & \dots & \dots & \dots &  & \dots & \dots \\
% & \text{Feature }5 & y_{115} & y_{125} &  & y_{215} & y_{225} \\
%\midrule
%\multirow{7}{*}{\begin{turn}{90} {\text{Unmodified}} \end{turn}} & \text{Feature }1 & y_{111}^{\ast} & y_{121}^{\ast} &  & y_{211}^{\ast} & y_{221}^{\ast} \\
% & \text{Feature }2 & y_{112}^{\ast} & y_{122}^{\ast} &  & - & y_{222}^{\ast} \\
% & \dots & \dots & \dots &  & \dots & \dots \\
% & \text{Feature }7 & y_{117}^{\ast} & y_{127}^{\ast} &  & y_{217}^{\ast} & y_{227}^{\ast} \\
% & \text{Feature }8 & y_{118}^{\ast} & y_{128}^{\ast} &  & y_{218}^{\ast} & y_{228}^{\ast} \\
% & \dots & \dots & \dots &  & \dots & \dots \\
% & \text{Feature }11 & y_{1,1,11}^{\ast} & y_{1,2,11}^{\ast} &  & y_{2,1,11}^{\ast} & y_{2,2,11}^{\ast} \\
%\bottomrule
%\end{array}
%\]
\end{footnotesize}
\caption{Representation of the data of modified peptides at one site and unmodified peptides of the same protein, with $I$ conditions and $J$ replicate runs. Abundances of the PTM and protein are quantified by multiple spectral features (peptide ions, $K$ for modified peptides and $L$ for unmodified peptides). Some spectral features can be missing (shown as $-$), either randomly in individual runs or completely in certain conditions. In real practice, the number of runs can vary across conditions. \label{fig:dtable}}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Existing method: two-sample $t$-test}
\label{sec:ttest}

Two-sample $t$-test is based on the null hypothesis that there is no difference in mean PTM abundance between Conditions $i$ and $i'$. The abundance in each run is taken as input and is often estimated by sum of peak intensities. The $t$-test is typically performed based on the log of summarized value. For example, the log-abundance estimate for the PTM in Run $j$ of Condition $i$ is given by
\[
\log \left( \sum_{k=1}^{K} 2^{y_{ijk}} \right).
\]
For adjustment with respect to unmodified peptides, the estimate of PTM abundance is divided by the protein abundance estimate, 
and the $t$-test for the adjusted PTM abundance on log scale takes as input the difference of their log-estimates.
%which is equivalent to take the difference of their log abundances. 
The quantity is denoted by $d_{ij}$ and is given by
\[
d_{ij} = \log \left( \sum_{k=1}^{K} 2^{y_{ijk}} \right) - \log \left( \sum_{l=1}^{L} 2^{y_{ijl}^{\ast}} \right).
\]
Alternatively, run-level summary to be described in \secref{sec:sum} can also be used. The difference between the means of PTM abundance in Conditions $i$ and $i'$ are estimated as
\[
\hat{\Delta} = \frac{1}{J} d_{i+} - \frac{1}{J} d_{i'+},
\]
where $d_{i+} = \sum_{j=1}^{J} d_{ij}$ and the test statistic for the $t$-test is given by $\hat{\Delta} / \mathrm{SE}(\hat{\Delta})$. The statistical significance of the difference is determined by comparing the test statistic against the $t$ distribution, with degrees of freedom $df=2J-2$ in balanced designs.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\subsection{$t$-test for data in batches}
%
%The test is not directly applicable to a problem with batches of data. Several \textit{ad-hoc} approaches may be used. Three commonly used approaches are 1) $t$-test (no batch): ignoring batch effects when applying $t$-test, 2) $t$-test (most significant batch): applying $t$-test in each batch and drawing conclusions based on the most significant batch, and 3) $t$-test (all batch): applying $t$-test in each batch and drawing conclusions based on all batches. Although simple, these \textit{ad-hoc} methods lack statistical justification. We reconstruct their meanings below and characterize their statistical properties under various forms of batch effects in \secref{sec:sim}. The abundance in each run is taken as input each of the three methods and is often estimated by sum of peak intensities, e.g., the abundance estimate for Run $j$ of Condition $i$ in Batch $b$ is given by
%\[
%y_{b, ij+} = \sum_{k=1}^{K_b} y_{b, ijk}
%\]
%
%\paragraph{Two-sample $t$-test (no batch).}
%In this approach, the modification abundance is estimated as the average over runs in all batches. For example, the estimate for Group $i$ is given by
%\[
%\hat{\mu}_{i}^{\ast} = \frac{1}{BJ} \sum_{b=1}^{B} \sum_{j=1}^{J} y_{b, ij+}.
%\]
%This is equivalent to fitting a linear model with only an intercept, which does not account for different characteristics across batches. 
%
%
%\paragraph{Two-sample $t$-test (most significant batch).} 
%The modification abundance is estimated in all batches, i.e., 
%\[
%\hat{\mu}_{b, i}^{\ast} = \frac{1}{J} \sum_{j=1}^{J} y_{b, ij+},
%\]
%where $\mathrm{SE} (\hat{\mu}_{b, i}^{\ast})$ is the standard error associated with the estimate. 
%The test statistic in Batch $b$ is 
%\[
%t_{b} = \frac{\hat{\mu}_{b, i}^{\ast}}{\mathrm{SE} (\hat{\mu}_{b, i}^{\ast})}
%\]
%To summarize the evidence across batches, one strategy is to use the most significant batch. With balanced design, the batch with the largest test statistic across all batch is selected, i.e., 
%\[
%b_{\text{max}} = \{ b \mid t_b = \max_{b'=1, \ldots, B} | t_{b'} | \}.
%\] 
%The change in modification abundance is estimated as
%\[
%\hat{\mu}_{i}^{\ast} - \hat{\mu}_{i'}^{\ast} = \hat{\mu}_{b_{\text{max}}, i}^{\ast} - \hat{\mu}_{b_{\text{max}}, i'}^{\ast}
%\]
%and the test statistic for the $t$-test is given by
%\[
%\frac{\hat{\mu}_{i}^{\ast} - \hat{\mu}_{i'}^{\ast}}{\mathrm{SE}\left( \hat{\mu}_{i}^{\ast} - \hat{\mu}_{i'}^{\ast} \right)} 
%= \frac{ \hat{\mu}_{b_{\text{max}}, i}^{\ast} - \hat{\mu}_{b_{\text{max}}, i'}^{\ast} }{\left[ \mathrm{SE}(\hat{\mu}_{b_{\text{max}}, i}^{\ast})^2 + \mathrm{SE}(\hat{\mu}_{b_{\text{max}}, i'}^{\ast})^2 \right]^{1/2}},
%\]
%%where $\hat{\sigma}_{\pi, ijk}^{2}$ and $\hat{\sigma}_{\pi, ijk'}^{2}$ are estimated as sample variances. 
%
%
%\paragraph{Two-sample $t$-test (all batch).}
%A change in modification abundance is considered as statistically significant only if such change is significant in all batches. This is equivalent to take the smallest test statistic across batches for the $t$-test: 
%\[
%\frac{\hat{\mu}_{i}^{\ast} - \hat{\mu}_{i'}^{\ast}}{\mathrm{SE}\left( \hat{\mu}_{i}^{\ast} - \hat{\mu}_{i'}^{\ast} \right)} 
%= \frac{ \hat{\mu}_{b_{\text{min}}, i}^{\ast} - \hat{\mu}_{b_{\text{min}}, i'}^{\ast} }{\left[ \mathrm{SE}(\hat{\mu}_{b_{\text{min}}, i}^{\ast})^2 + \mathrm{SE}(\hat{\mu}_{b_{\text{min}}, i'}^{\ast})^2 \right]^{1/2}},
%\]
%where 
%\[
%b_{\text{min}} = \{ b \mid t_b = \min_{b'=1, \ldots, B} | t_{b'} | \}.
%\]

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Proposed approach}
\label{sec:prop}

To characterize the observed feature intensities, different levels of variations are expressed using linear mixed models in consideration of the following factors: modification, condition, run, and feature. As different degrees of variability are present in the feature intensities of modified and unmodified peptides, they are expressed by separate models.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Statistical modeling and inference}
\label{sec:model}

The observed log-intensity of a modified peptide feature is denoted by $y_{ijk}$ and represented as
\[
y_{ijk} = \psi + C_{i} + R_{j(i)} + F_{k} + (R \times F)_{ijk},
\]
where the effects of condition and feature are modeled as fixed effects: 
\[
\sum_{i=1}^{I} C_{i} = 0, \quad \sum_{k=1}^{K} F_{k} = 0,
\]
and the effects of run and its interaction with feature are considered as random effects arising from normal distribution with mean $0$: 
\[
R_{j(i)} = \gamma_{j(i)} \stackrel{\text{iid}}{\sim} \mathcal{N}(0, \sigma_{\gamma}^{2}), \qquad
(R \times F)_{ijk} = \epsilon_{ijk} \stackrel{\text{iid}}{\sim} \mathcal{N}(0, \sigma_{\epsilon}^{2}).
\]
Similarly, the observed log-intensity of an unmodified peptide feature is denoted by $y_{ijl}^{\ast}$ and represented as 
\[
y_{ijl}^{\ast} = \psi^{\ast} + C_{i}^{\ast} + R_{j(i)}^{\ast} + F_{l}^{\ast} + (R \times F)_{ijl}^{\ast}, 
\]
where the effects of condition and feature are modeled as fixed effects: 
\[
\sum_{i=1}^{I} C_{i}^{\ast} = 0, \quad \sum_{l=1}^{L} F_{l}^{\ast} = 0,
\]
and 
\[
R_{j(i)}^{\ast} = \gamma_{j(i)}^{\ast} \stackrel{\text{iid}}{\sim} \mathcal{N}(0, \sigma_{\gamma^{\ast}}^{2}), \qquad
(R \times F)_{ijl}^{\ast} = \epsilon_{ijl}^{\ast} \stackrel{\text{iid}}{\sim} \mathcal{N}(0, \sigma_{\epsilon^{\ast}}^{2}).
\]

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Run-level summarization of feature intensities}
\label{sec:sum}

Run-level summarization of feature intensities for each PTM site is carried out as in the sub-plot model of MSstats~\cite{choi_etal_14a}, which involves a) imputation of censored missing values, and b) summarization of feature intensities using Tukey's median polish. The run-level summary for the PTM in Run $j$ of Condition $i$ is denoted by $\hat{y}_{ij}$.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Model-based inference of the underlying abundance}
\label{sec:infer}

The PTM abundance in each run is represented as 
\[
\hat{y}_{ij} = \psi + C_{i} + R_{j(i)},
\]
where $\sum_{i=1}^{I} C_{i} = 0$, $R_{j(i)} = \gamma_{j(i)} \stackrel{\text{iid}}{\sim} \mathcal{N}(0, \sigma_{\gamma}^{2})$. 
Similarly, the protein abundance in each run is expressed as
\[
\hat{y}_{ij} = \psi^{\ast} + C_{i}^{\ast} + R_{j(i)}^{\ast},
\]
where $\sum_{i=1}^{I} C_{i}^{\ast} = 0$, $R_{j(i)}^{\ast} = \gamma_{j(i)}^{\ast} \stackrel{\text{iid}}{\sim} \mathcal{N}(0, \sigma_{\gamma^{\ast}}^{2})$. The expected values of log-abundances of the PTM and protein in Condition $i$ are denoted by $\mu_{i}$ and $\mu_{i}^{\ast}$, respectively, and the values are estimated as: 
\begin{align*}
\hat{\mu}_{i} &= \hat{\psi} + \hat{C}_{i} = \frac{1}{J} \hat{y}_{i+} \\
\hat{\mu}_{i}^{\ast} &= \hat{\psi}^{\ast} + \hat{C}_{i}^{\ast} = \frac{1}{J} \hat{y}_{i+}^{\ast},
\end{align*}
where the standard errors of the estimates are $\mathrm{SE}(\hat{\mu}_{i}) = (\hat{\sigma}_{\gamma}^{2} / J)^{1/2}$ and $\mathrm{SE}(\hat{\mu}_{i}^{\ast}) = (\hat{\sigma}_{\gamma^{\ast}}^{2} / J)^{1/2}$.
%\[
%\left( \frac{\hat{\sigma}_{\gamma}^{2}}{J} \right)^{1/2}, \quad \left( \frac{\hat{\sigma}_{\gamma^{\ast}}^{2}}{J} \right) ^ {1/2}.
%\]
%\[
%\left( \frac{1}{J} \hat{\sigma}_{\gamma}^{2} \right)^{1/2}, \quad \left( \frac{\hat{\sigma}_{\gamma^{\ast}}^{2}}{J} \right) ^ {1/2}.
%\]
Based on the estimates $\hat{\mu}_{i}$ and $\hat{\mu}_{i}^{\ast}$, the adjusted log-abundance of the PTM is given by $(\hat{\mu}_{i} - \hat{\mu}_{i}^{\ast})$ and the standard error of the estimate is 
\[
\left[ \frac{1}{J} \left( \hat{\sigma}_{\gamma}^{2} + \hat{\sigma}_{\gamma^{\ast}}^{2} \right) \right]^{1/2}.
\]

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{PTM significance analysis}
\label{sec:test}

With protein-level adjustment, the model-based testing is based on the hypothesis that there is no difference in adjusted PTM abundance between Conditions $i$ and $i'$
%Adjusted by protein abundance, the hypothesis states that there is no difference in log-abundance of the modification between groups $i$ and $i'$
\begin{align*}
H_{0}: \Delta = \left( \mu_{i} - \mu_{i'} \right) - \left( \mu_{i}^{\ast} - \mu_{i'}^{\ast} \right) &= 0 \\
H_{a}: \Delta = \left( \mu_{i} - \mu_{i'} \right) - \left( \mu_{i}^{\ast} - \mu_{i'}^{\ast} \right) &\neq 0
\end{align*}
The log-fold change in the adjusted PTM abundance, $\Delta$, is estimated by 
\[
\hat{\Delta} = \left[ \frac{1}{J} \left( \hat{y}_{i+} - \hat{y}_{i'+} \right) \right] - \left[ \frac{1}{J} \left( \hat{y}_{i+}^{\ast} - \hat{y}_{i'+}^{\ast} \right) \right],
\]
and the standard error of the estimate $\mathrm{SE}(\hat{\Delta})$ is 
\[
\left[ \frac{2}{J} \left( \hat{\sigma}_{\gamma}^{2} + \hat{\sigma}_{\gamma^{\ast}}^{2} \right) \right]^{1/2}.
\]
%\[
%\left[ \left( \frac{2}{J} \right) \cdot \left( \hat{\sigma}_{\gamma}^{2} + \hat{\sigma}_{\gamma^{\ast}}^{2} \right) \right]^{1/2}.
%\]
The test statistic $\hat{\Delta} / \mathrm{SE}(\hat{\Delta})$ is compared against the $t$ distribution, with degrees of freedom approximated by
\[
\left( \hat{\sigma}_{\gamma}^{2} + \hat{\sigma}_{\gamma^{\ast}}^{2} \right)^2 \bigg/
\left( \frac{\hat{\sigma}_{\gamma}^{4}}{\mathrm{df}(\gamma)} + \frac{\hat{\sigma}_{\gamma^{\ast}}^{4}}{ \mathrm{df}(\gamma^{\ast})} \right).
\]
A distinctive property of the proposed model-based testing to the two-sample $t$-test is that even only the PTM abundances in Conditions $i$ and $i'$ are compared, measurements from all conditions are used for the modeling and inference.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Design of PTM experiments}
\label{sec:design}

The proposed statistical framework allows for design of PTM experiments in terms of sample size calculation and power analysis. 
%Sample size calculation based on linear models has been described for general applications~\cite{kutner_etal_04a} and specifically for protein significance analysis~\cite{oberg_vitek_09a}. 
Sample size calculation takes as input a) $q$, the desired false discovery rate, b) $\beta$, the average Type II error rate, c) $\Delta$, the minimal log-fold change in adjusted PTM abundance that we would like to detect, d) $m_0 / (m_0 + m_1)$, the fraction of truly differentially modified PTM sites in the comparison, and e) $\sigma_{\gamma}^{2}$ and $\sigma_{\gamma^{\ast}}^{2}$, the anticipated variances associated to modified and unmodified peptide features, respectively. The variances can be derived based on the dataset being analyzed, assuming similar quantitative properties and variations. With these values and a user-specified number of conditions, the corresponding number of technical replicates per condition can then be derived, as described in~\cite{kutner_etal_04a}. Given the above quantities, the minimal number of replicates $J$ is determined by the variance of the estimated log-fold change $\mathrm{SE}^{2}(\hat{\Delta})$ as
\[
\mathrm{SE}^{2}(\hat{\Delta}) = \left[ \frac{2}{J} \left( \hat{\sigma}_{\gamma}^{2} + \hat{\sigma}_{\gamma^{\ast}}^{2} \right) \right]
\leq \left( \frac{\Delta}{t_{1-\beta, df} + t_{1-\alpha /2, df}} \right)^{2},
\]
where 
\[
\alpha = (1 - \beta) \cdot \frac{q}{1 + (1-q) \cdot m_0 / m_1},
\]
and $t_{1-\beta, df}$ and $t_{1-\alpha /2, df}$ are the $100(1-\beta)^{\text{th}}$ and the $100(1-\alpha /2)^{\text{th}}$ percentiles of the $t$ distribution, with $df = I(J-1)$ degrees of freedom in balanced designs. More details can be found in~\cite{oberg_vitek_09a}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Extension to complex designs}
\label{sec:complex}

The proposed statistical framework can be extended to analyze data from experiments of complex designs, such as factorial design and time series. We discuss below a specific design commonly applied in PTM experiments, in which data are acquired in multiple batches.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Batch effects}
\label{sec:batch}

As in \secref{sec:test}, hypothesis testing on the adjusted log-abundances of the PTM in Conditions $i$ and $i'$ is performed to detect differentially modified PTM sites. The hypothesis is that there is no difference in adjusted log-abundance of the PTM between Conditions $i$ and $i'$
\begin{align*}
H_{0}: \Delta = (\mu_{i} - \mu_{i'}) - (\mu_{i}^{\ast} - \mu_{i'}^{\ast}) &= 0 \\
H_{a}: \Delta = (\mu_{i} - \mu_{i'}) - (\mu_{i}^{\ast} - \mu_{i'}^{\ast}) &\neq 0
\end{align*}
For batch-wise data, we consider two ways to estimate the difference in adjusted PTM abundance, $\Delta = (\mu_{i} - \mu_{i'}) - (\mu_{i}^{\ast} - \mu_{i'}^{\ast})$ based on different assumptions about the properties of batch effects, namely per-batch model (proposed approach) and all-batch model. For the following discussion, we denote the log-intensity of a modified peptide feature in Run $j$ of Condition $i$ and Batch $b$ by $y_{b, ijk}$, where $b = 1, \ldots, B$. Similarly, $y_{b, ijl}^{\ast}$ is denoted for the unmodified peptide feature in the same run.

\paragraph{Per-batch model (proposed approach).} 
The model assumes different levels of variability are present in different batches and the differences between conditions vary across batches (i.e., there is an interaction effect between condition and batch). Difference between conditions is estimated in each batch, and the overall log-fold change in adjusted PTM abundance is estimated as the average over batches
\[
\hat{\Delta} = \frac{1}{B} \sum_{b=1}^{B} \left[ \frac{1}{J} \left( \hat{y}_{b,i+} - \hat{y}_{b,i'+} \right) \right]
- \frac{1}{B} \sum_{b=1}^{B} \left[ \frac{1}{J} \left( \hat{y}_{b,i+}^{\ast} - \hat{y}_{b,i'+}^{\ast} \right) \right].
\]
The standard error associated to the estimate is 
\[
\left[ \left(\frac{1}{B}\right)^2 \cdot \left( \frac{2}{J} \right) \cdot \sum_{b=1}^{B} \left( \hat{\sigma}_{\gamma_b}^{2} + \hat{\sigma}_{\gamma_b^{\ast}}^{2} \right) \right]^{1/2}.
\]
The test statistic $\hat{\Delta} / \mathrm{SE}(\hat{\Delta})$ is compared against the $t$ distribution, where the degrees of freedom are approximated as
\[
\left[ \sum_{b=1}^{B} \left( \hat{\sigma}_{\gamma_{b}}^{2} + \hat{\sigma}_{\gamma_{b}^{\ast}}^{2} \right) \right]^2 \bigg/
\sum_{b=1}^{B} \left[ \frac{\hat{\sigma}_{\gamma_{b}}^{4}}{  \mathrm{df}(\gamma_b)} + \frac{\hat{\sigma}_{\gamma_{b}^{\ast}}^{4}}{\mathrm{df}(\gamma_b^{\ast})} \right].
\]

\paragraph{All-batch model.}
In contrast to the per-batch model, the all-batch model assumes identical variance and difference between conditions in all batches. The log-fold change in adjusted PTM abundance is estimated as the average over runs and batches
\[
\hat{\Delta} = \frac{1}{BJ} \left( \hat{y}_{+,i+} - \hat{y}_{+,i'+} \right) - \frac{1}{BJ} \left( \hat{y}_{+,i+}^{\ast} - \hat{y}_{+,i'+}^{\ast} \right),
\]
and the standard error of the estimate is 
\[
\left[ \frac{2}{BJ} \left( \hat{\sigma}_{\gamma}^{2} + \hat{\sigma}_{\gamma^{\ast}}^{2} \right) \right] ^{1/2}.
\]
The test statistic $\hat{\Delta} / \mathrm{SE}(\hat{\Delta})$ is compared against the $t$ distribution, with degrees of freedom approximated by
\[
\left( \hat{\sigma}_{\gamma}^{2} + \hat{\sigma}_{\gamma^{\ast}}^{2} \right)^2 \bigg/
\left( \frac{\hat{\sigma}_{\gamma}^{4}}{  \mathrm{df}(\gamma)} + \frac{\hat{\sigma}_{\gamma^{\ast}}^{4}}{\mathrm{df}(\gamma^{\ast})} \right).
\]

%\clearpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Computer simulation}
\label{sec:sim}

The proposed statistical approach was evaluated and compared to the $t$-test using computer simulation. In particular, their statistical properties with respect to protein-level adjustment and batch effects were evaluated.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Protein-level adjustment}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{One site per protein}

Differential levels of modified peptides may be due to differential modifications and/or changes in protein abundance. The proposed approach adjusts the PTM abundance with respect to protein abundance as introduced in \secref{sec:prop}. Alternatively, two-sample $t$-test taking as input the ratio between feature intensities of modified and unmodified peptides (difference on log scale) is commonly applied for the same purpose (\secref{sec:ttest}). 
Approaches without considering protein-level adjustment lose track of an important aspect in interpreting observed changes in PTM abundance, which may result in misleading conclusions. To highlight the necessity of the adjustment, we compared the following approaches: a) proposed approach, b) linear model (no adjustment), i.e., same as the proposed approach but ignoring the measurements with respect to unmodified peptides, c) $t$-test (with adjustment), and d) $t$-test (no adjustment).
In experiments of complex designs, multiple inter-related conditions are often compared together. Whereas $t$-test uses measurements from the two conditions being compared, the proposed approach leverages measurements in all conditions for the inference of the underlying abundance. To highlight this distinction, multiple conditions of data were generated, where only one condition was simulated with a systematic change. The simulation was based on the following parameters: 
\begin{itemize}
\item Mean of log-intensity: $25$
\item Standard deviations of log-intensities for modified and unmodified peptides: $0.2$, $0.3$
\item Difference in PTM abundance between conditions: $0$, $0.5$, $0.75$, $1$
\item Difference in protein abundance between conditions: $0$, $0.5$
\item Number of replicates: $2$, $3$, $5$
\item Number of conditions: $2$, $3$, $4$
\item Number of realizations: $500$
\item Missing data: no missing value, or PTM missing in Run 1, Condition 1
\end{itemize}
The results are summarized from \sfigref{fig:prot_fpr} to \sfigref{fig:protm1_pwr}, including false positive rate with or without changes in protein abundance by the considered methods (\sfigref{fig:prot_fpr}), power by the approaches with protein-level adjustment when there is no missing data (\sfigref{fig:prot_pwr}), and power by the approaches with protein-level adjustment when the PTM is missing in one run (\sfigref{fig:protm1_pwr}).

\begin{figure}[h!]
\centering
\includegraphics[width=.85\textwidth]{sim/protnull_fpr}\\
\includegraphics[width=.85\textwidth]{sim/protalt_fpr}
\caption{All the considered methods well calibrated the Type I error rate when there was no change in protein abundance (upper plot). When the
changes in PTM abundance were entirely due to changes in protein abundance across conditions (bottom plot), analysis without accounting for the protein-level changes resulted in off-target, high false positive rates. \label{fig:prot_fpr}}
\end{figure}

\begin{figure}[h!]
\centering
\includegraphics[width=\textwidth]{sim/prot_pwr}
%\includegraphics[width=.85\textwidth]{sim/prot_pwr}
\caption{In comparison between the proposed approach and $t$-test with protein-level adjustment, the proposed approach improved power with small sample sizes. Two-sample $t$-test only used data within the groups of interest while ignoring the rest the of data. Consequently, it gave similar performance across cases with different number of conditions. In contrast, the proposed approach leveraged all available information, which resulted in improved power with increased number of conditions. \label{fig:prot_pwr}}
\end{figure}

\begin{figure}[h!]
\centering
\includegraphics[width=\textwidth]{sim/protm1_pwr}
%\includegraphics[width=.85\textwidth]{sim/protm1_pwr}
\caption{The advantage by using the proposed approach over two-sample $t$-test became more profound in the presence of missing value. In cases with small sample sizes, performance by the $t$-test decreased dramatically with one missing value. As a specific example, the statistical power by using the $t$-test with two replicates was below $0.25$, and increasing the fold change to $2$ did not effectively reduce the negative impact. \label{fig:protm1_pwr}}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Multiple sites per protein}

Expression levels of PTM sites are adjusted based on the abundance of their originating protein. Since the same reference is used for all sites, it introduces correlation among estimates and test statistics for those sites. This may cause issues in controlling false discovery rate (FDR).
We investigated the property by simulating data of two conditions and 1000 proteins with different numbers of PTM sites and comparing the results of the proposed approach and the linear model with no adjustment. 
%We investigated the property by simulating data of two conditions and 1000 proteins with different numbers of PTM sites. 
A fraction (50\%) of the 1000 proteins had no changes between conditions while systematic changes were simulated for the rest of the proteins. Multiple testing correction was performed using the Benjamini and Hochberg's method. 
Performances of the considered approaches were assessed by their actual FDR, calculated as the fraction of proteins with adjusted $p$-values $<0.05$ among the proteins with true differences. The results are summarized in \sfigref{fig:prot_fdr}.
%Performances of the considered approaches were assessed by their power and FDR. The power was calculated as the fraction of proteins with adjusted $p$-values $<0.05$ among those proteins with true differences between conditions, while the FDR was calculated as the fraction of proteins with adjusted $p$-values $<0.05$ among the proteins with true differences. The results are summarized in \sfigref{fig:prot_fdr}.

\begin{figure}[h!]
\centering
\includegraphics[width=.85\textwidth]{sim/prot_fdr}
\caption{Correlation between test statistics across sites due to protein-level adjustment did not affect the expected level of FDR under the considered scenarios. The Benjamini-Hochberg procedure was used to control the FDR. \label{fig:prot_fdr}}
\end{figure}

\clearpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Batch effects}

We evaluated the statistical properties of the proposed approach under batch effects, in comparison to the two-sample $t$-test (\secref{sec:ttest}). While the $t$-test is not directly applicable to a problem with batches of data, several \textit{ad-hoc} approaches may be used. Two commonly used approaches are a) $t$-test (no batch): ignoring batch effects when applying $t$-test, and b) $t$-test (most significant batch): applying $t$-test in each batch and drawing conclusions based on the most significant batch. Although simple, these \textit{ad-hoc} methods lack statistical justification. We characterized their statistical properties under various forms of batch effects. In this part of the simulation, two batches of data were generated, with the following forms of batch effects: difference in signal intensities across batches, difference in variability across batches, and interaction effect between batch and condition (i.e., change between condition affected by batch). The following interaction scenarios were considered in the simulation: a) no interaction between condition and batch, i.e., same change across conditions in both batches, b) positive interaction as $25$\% greater change in the batch of higher level, and c) negative interaction as $25$\% lower change in the batch of higher level. Below are the parameters used in the simulation: 
\begin{itemize}
\item Mean of log-intensity: $25$
\item Increased intensity level of Batch 2 versus Batch 1: $0$, $1$, $2$
\item Standard deviations of log-intensity: $0.2$ in Batch 1 and $0.3$ in Batch 2
\item Difference between conditions in Batch 1: $0.5$, $0.75$, $1$
\item Number of replicates: $2$, $3$, $5$
\item Number of conditions: $2$, $3$, $4$
\item Number of realizations: $500$
\end{itemize}
The following four approaches were compared: a) proposed approach, i.e., per-batch model, b) all-batch model, c) $t$-test (no batch), and d) $t$-test (most significant batch). The results are summarized from \sfigref{fig:synnull_est} to \sfigref{fig:synnull_pwr_5}, on the aspects of estimation error (\sfigref{fig:synnull_est}), false positive rate (\sfigref{fig:synnull_fpr}), and power (\sfigref{fig:synnull_pwr}, \sfigref{fig:synnull_pwr_5}).

\begin{figure}[h!]
\centering
\includegraphics[width=.825\textwidth]{sim/synnull_est_2}
\caption{Estiamtion based on the most statistically significant batch with $t$-test was highly variable and frequently biased. The observation is consistent in all the simulated scenarios. \label{fig:synnull_est}}
\end{figure}

\begin{figure}[h!]
\centering
\includegraphics[width=.825\textwidth]{sim/synnull_fpr}
\caption{The proposed approach better calibrated Type I error rate. Similar results were observed in the cases with positive and negative batch-condition interactions. \label{fig:synnull_fpr}}
\end{figure}

\clearpage
\begin{figure}[h!]
\centering
\includegraphics[width=.825\textwidth]{sim/synnull_pwr_2}\\
\includegraphics[width=.825\textwidth]{sim/synnull_pwr_3}
\caption{The proposed approach improved power with small sample sizes in almost all the scenarios, under various forms of batch effects including difference in intensity level (higher on the right) and difference in variability. Using $t$-test while ignoring batch effect gave similar performance in special cases with no difference in intensity level between batches, but its performance dramatically decreased in general cases. The proposed methods gave consistently improved performance compared with other methods by properly characterizing batch effects and leveraging all available information. Similar patterns were observed in other simulated scenarios with positive and negative interactions.\label{fig:synnull_pwr}}
\end{figure}

%\clearpage
%\begin{figure}[h!]
%\centering
%\includegraphics[width=.85\textwidth]{sim/synpos_pwr_3}\\
%\includegraphics[width=.85\textwidth]{sim/synneg_pwr_3}
%\caption{With 3 replicates and under the scenarios with either positive or negative interaction, the proposed approach consistently improved power over other methods. Similar results were observed with 2 replicates. As in those cases with no interaction, the approaches using two-sample $t$-test did not benefit from data in additional conditions. Also, ignoring batch effects as applying $t$-test (no batch) lost power dramatically. \label{fig:syn_pwr_3}}
%\end{figure}

\begin{figure}[h!]
\centering
\includegraphics[width=.825\textwidth]{sim/synnull_pwr_5}
\caption{Ignoring batch effects lost power. The negative impact was only partially reduced by increasing the sample size to 5. Similar patterns were observed in other simulated scenarios. \label{fig:synnull_pwr_5}}
\end{figure}

\clearpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Sample size calculation and power analysis}

The proposed approach adjusts for the underlying protein abundance in the PTM significance analysis, which corrects the confounding factor with a cost of increased uncertainty. We compared the required sample size with or without the adjustment to highlight the property (\sfigref{fig:size_prot}), based on a design of two conditions, in consideration of three pairs of standard deviations of log-intensities for modified and unmodified peptides: (0.2, 0.1), (0.2, 0.2), and (0.2, 0.4). 
%We considered a design with two conditions, where standard deviations of log-intensities of both modified and unmodified peptides were both set as $0.2$. 
We then characterized the advantage of general statistical modeling in complex designs in terms of sample size calculation (\sfigref{fig:size_grp}) and power analysis (\sfigref{fig:pwr_grp}).

\begin{figure}[h!]
\centering
\includegraphics[width=\textwidth]{sim/size_prot}
\caption{Protein-level adjustment relies on the inference of protein abundance, which introduces additional uncertainty in the estimate of PTM difference. Therefore, the required sample size to detect a systematic change is higher than as expected for standard differential analysis without adjustment. The discrepancy can be profound if the uncertainty associated with the protein abundance estimate is greater than that of PTM abundance estimate. Sample size calculation without accounting for the uncertainty would lead to under-powered studies. \label{fig:size_prot}}
\end{figure}

\begin{figure}[h!]
\centering
\includegraphics[width=\textwidth]{sim/size_grp}
%\includegraphics[width=.725\textwidth]{sim/size_grp}
\caption{In complex designs, simultaneously analyzing all the conditions effectively increases the degrees of freedom and requires fewer replicates. \label{fig:size_grp}}
\end{figure}

\begin{figure}[h!]
\centering
\includegraphics[width=\textwidth]{sim/pwr_grp}
%\includegraphics[width=.725\textwidth]{sim/pwr_grp}
\caption{Statistical power can be improved by increasing the sample size and analyzing multiple conditions together. \label{fig:pwr_grp}}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\clearpage
\section*{References}

\bibliographystyle{plain}
\bibliography{ptm_ref}

\end{document}

### 2021-02-01 main/MSstatsPTM_Main.tex
Updated main latex

@@
 \documentclass[11pt]{article}
+%\documentclass[11pt, twocolumn]{article}
+
+\usepackage[colorinlistoftodos]{todonotes}
+\usepackage{graphicx}
+\graphicspath{ {./images/} }
 
@@
 
-\author{
-  LastName1, FirstName1\\
-  \texttt{first1.last1@xxxxx.com}
-  \and
-  LastName2, FirstName2\\
-  \texttt{first2.last2@xxxxx.com}
-}
-
+%\author{
+%  LastName1, FirstName1\\
+%  \texttt{first1.last1@xxxxx.com}
+%  \and
+%  LastName2, FirstName2\\
+%  \texttt{first2.last2@xxxxx.com}
+%}
+\date{}
 \begin{document}
@@
 \maketitle
+\todo{Add Author list}
 
 \section{Abstract}
+
+The scientific community widely utilizes mass spectrometry (MS)-based proteomics to quantify the abundance of proteins and their post-translational modifications (PTMs).
+Experiments targeting PTMs face several specific challenges.
+These include the low abundance of modified proteo-forms, few representative peptides that span modification sites, and convolution with abundance changes in the overall protein expression.
+Due to these challenges, a robust approach to estimate relative systematic changes in PTMs should combine information pertaining to PTM sites over several peptides, replicates in multiple conditions, and consider sources of confounding and variation present in the experiment.
+We propose a general statistical model and workflow that is both reproducible and comprehensive.
+The model measures PTM and protein abundance by summarizing intensities through Tukey’s median polish method.
+Then a model based on the family of linear mixed-effects models is fit.
+This model is automatically adjusted to the specific experimental design.
+Finally, the PTM abundances are adjusted to remove variance from changes in the overall protein.
+We implement this model in the free and open-source R package MSstatsPTM.
+\todo{Used overview from USHUPO poster}
 
@@
 The modeling framework, however, is not applicable for either label-free workflows or experiments with complex designs.
-[Source]
+\todo{Add Source. Compare this to the TMT part of MSstatsPTM}
 
@@
 \section{Proposed Approach}
-
+\begin{center}
+\includegraphics[scale=.8]{fig3}
+\end{center}
 Figure 3 schematically illustrates a simplified version of the data structure resulting from a typical bottom-up experiment for quantitative analysis of PTMs, in which there are multiple layers of variation present.
@@
 \subsection*{Statistical modeling and parameter estimation}
-
+\todo{Add in TMT approach}
 The proposed approach takes as input a list of log-transformed intensities of spectral features, identified and quantified across LC-MS runs.
@@
 Statistical modeling and quantification for global proteomics data are performed by the same procedure as for PTM data.
-
+\todo{Add Sources}
 \subsection*{Detection of changes in PTMs}
@@
 
-\paragraph*{Computer simulation: protein-level adjustment.} Differential intensity levels of modified peptides may be due to changes in modification, change in protein abundance, or both.
+\paragraph*{Computer simulation: protein-level adjustment.}
+
+Differential intensity levels of modified peptides may be due to changes in modification, change in protein abundance, or both.
 The proposed approach adjusts the abundance with respect to unmodified peptides by combining the inference of modified and unmodified peptide abundances.
@@
 
-\subsubsection*{Spike-in experiment [Need to write results of this experiment]}
+\subsubsection*{Spike-in experiment}
+\todo{Need to write results of this experiment}
+\todo{Devon: I don't know what these Label-Free experiments are or how to analyze them. Meeting with Tsung-Heng on 2/5 to review}
 
@@
 
+\todo{Add Shigella for TMT side}
+
 \section{Conclusion and discussion}
@@
 \section{Figure legends}
+\todo{Unclear where Tsung-Heng meant these plots to go. Need to add them into correct sections}
 
@@
 
+\todo{Add sources section}
+
 \end{document}n{document}

### 2021-02-05 main/MSstatsPTM_Main.tex
MC

@@
-\documentclass[11pt]{article}
+\documentclass[mcp]{article}
 %\documentclass[11pt, twocolumn]{article}
@@
 \graphicspath{ {./images/} }
+
+\usepackage{fullpage}
+\usepackage{amsfonts,amsthm}
+\usepackage{amsmath}
+\numberwithin{figure}{section} % numbering in each subsection
+\numberwithin{table}{section}
+\usepackage{setspace}
+\usepackage{url}
+\usepackage{lscape} %% rotating table
+\usepackage{rotating}
+\usepackage{hyperref}
+\usepackage[table,dvipsnames]{xcolor}
+%\usepackage{makecell}
+\usepackage{array}
+
+% remove number of section in the title
+\makeatletter
+\def\@seccntformat#1{%
+  \expandafter\ifx\csname c@#1\endcsname\c@section\else
+  \csname the#1\endcsname\quad
+  \fi}
+\makeatother
+
+\usepackage{caption}
+
+%% to remove zero preceding section number.
+\renewcommand\thesection{\arabic{section}}
+\renewcommand\thesubsection{\thesection.\arabic{subsection}}
+
+\usepackage{tabulary,multirow,multicol,rotating}
+
+\definecolor{Red}{rgb}{0.60,0.00,0.00}
+\definecolor{Blue}{rgb}{0.00,0.00,0.75}
+\definecolor{LightYellow}{rgb}{1.00,0.97,0.68}
+\definecolor{Green}{rgb}{0.30, 0.60, 0.30}
+\definecolor{MyLightMagenta}{cmyk}{0.1,0.8,0,0.1} 
+\definecolor{cornflowerblue}{rgb}{0.39, 0.58, 0.92}
+\definecolor{darkorange}{rgb}{0.8, 0.4, 0}
+\definecolor{LightPurple}{rgb}{1, 0.51, 0.98}
+\definecolor{DarkPurple}{rgb}{0.54, 0.27, 0.53}
+\definecolor{Purple1}{rgb}{0.83, 0.29, 0.95}
+\definecolor{Purple2}{rgb}{0.97, 0.12, 0.59}
+\definecolor{change}{rgb}{0.39, 0.58, 0.92}
+
+\definecolor{function}{RGB}{0, 112, 192}
+\definecolor{group}{rgb}{0.39, 0.58, 0.92}
+\definecolor{ratio}{rgb}{1, 0.6, 0.16}
+\definecolor{feature}{rgb}{0.83, 0.32, 0.48}
+\definecolor{subject}{rgb}{0.6, 0.3, 0.09}
+\definecolor{run}{rgb}{0, 0.68, 0.17}
+
+\usepackage{array}
+\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
+\setlength{\tabcolsep}{2pt}
+
+%% for comments
+\newcommand{\ignore}[1]{}
+\def\todo#1{{\color{red}[#1]}}
+\def\change#1{{\color{cornflowerblue}#1}}
+\newenvironment{todolong}{\color{red}[TODO:}{]}
+\def\note#1{{\color{OliveGreen}[NOTE: #1]}}
+\def\added#1{{\color{blue}[ADDED: #1]}}
+\def\devon#1{{\color{green}[4Devon: #1]}}
+%\def\maria#1{{\color{purple}[For Maria to update: #1]}}
+%\def\eduard#1{{\color{OliveGreen}[For Eduard to update: #1]}}
+
+\def\eqref#1{Eq.~(\ref{eq:#1})}
+\def\figshortref#1{{\bf Fig.~\ref{fig:#1}}}
+\def\figref#1{{\bf Figure~\ref{fig:#1}}}
+\def\secref#1{{\bf Section~\ref{sec:#1}}}
+\def\tabref#1{{\bf Table~\ref{tab:#1}}}
+
+%\usepackage{xr}
+%\externaldocument{supplementary}
+%\def\sfigref#1{{\bf Supplementary Fig.~\ref{fig:#1}}}
+%\def\snoteref#1{{\bf Supplementary Note~\ref{sec:#1}}}
+%\def\snoteshortref#1{{\bf Suppl. Note~\ref{sec:#1}}}
+%\def\stabref#1{{\bf Supplementary Table~\ref{tab:#1}}}
+
+%%% should be change =2
+\linespread{1}
+
+%\date{\vspace{-5ex}} % to remove date in the title
+
+\renewcommand{\deg}{\ensuremath{^{\circ}}\xspace}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 \todo{Add Author list}
+%\noindent\author{Tsung-Heng Tsai$^{1}$, Olga Vitek$^{1,\ast}$}
+%
+%\vspace{0.2in}
+%\noindent $^1$   Khoury College of Computer Sciences, Northeastern University, Boston, MA, USA\\
+%\noindent $\ast$   Corresponding author \\
 
+\devon{maybe good to have supplementary.txt in the same folder, to link figure, table number between main and supplementary.}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Abstract}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Introduction}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Proposed Approach}
-\begin{center}
-\includegraphics[scale=.8]{fig3}
-\end{center}
-Figure 3 schematically illustrates a simplified version of the data structure resulting from a typical bottom-up experiment for quantitative analysis of PTMs, in which there are multiple layers of variation present.
+
+\devon{Please change Fig.4a and so on using figref or tabref in the text. Put all figures and legends in figure section.}
+\figref{data-structure} schematically illustrates a simplified version of the data structure resulting from a typical bottom-up experiment for quantitative analysis of PTMs, in which there are multiple layers of variation present.
 A PTM site is quantified with multiple spectral features, which vary in sequence (e.g., fully or partially cleaved peptides), ionization efficiency, charge states, etc.
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection*{Statistical modeling and parameter estimation}
-\todo{Add in TMT approach}
+
 The proposed approach takes as input a list of log-transformed intensities of spectral features, identified and quantified across LC-MS runs.
@@
 Statistical modeling and quantification for global proteomics data are performed by the same procedure as for PTM data.
+
 \todo{Add Sources}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection*{Detection of changes in PTMs}
@@
 The details are provided in Supplementary Sec.
-S3.2.
+S3.2 \devon{link to supplementary using sfigref}.
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection*{Analysis with multiple batches}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection*{Extension to TMT experiment}
+
+\todo{Add in TMT approach}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Results}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsubsection*{Spike-in experiment}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsubsection*{Dilution experiment [What was this experiment?]}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsubsection*{Ubiquitin changes in bone marrow derived macrophages from Atg16L1 deficient mice Re-analysis of published dataset: USP30}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsubsection*{Re-analysis of published dataset: RIP1 [Need this dataset and original findings]}
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsubsection*{Re-analysis of published dataset: TMT experiments}
+
+\todo{need the result of TMT experiment}
 \todo{Add Shigella for TMT side}
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Conclusion and discussion}
@@
 
-\section{Figure legends}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Figure}
 \todo{Unclear where Tsung-Heng meant these plots to go. Need to add them into correct sections}
 
-Figure 3.Data structure of a typical PTM experiment and goals of PTM characterization.(a)Schematic data representation, in a simplified case of two conditions and two replicate runs.
+\begin{figure}[ht]
+\centering
+\includegraphics[scale=.8]{fig3}
+\caption{Data structure of a typical PTM experiment and goals of PTM characterization.(a)Schematic data representation, in a simplified case of two conditions and two replicate runs.
 Each PTM site is modeled and characterized separately, where a PTM is quantified with multiple spectral features (boxes), distinguished by different charge states of a peptide.
@@
 Some spectral features can be missing.(b)PTM relative quantification by statistical inference, which makes use of the feature intensities to infer the underlying PTM abundance and protein abundance with an estimate of associated uncertainty.
-(c) Model-based testing for differential PTM abundance, which corrects for the underlying protein abundance with a cost of increased uncertainty about the estimate of difference between conditions.(d) Statistical experimental design in terms of sample size calculations and power analysis.
+(c) Model-based testing for differential PTM abundance, which corrects for the underlying protein abundance with a cost of increased uncertainty about the estimate of difference between conditions.(d) Statistical experimental design in terms of sample size calculations and power analysis.}
+\label{fig:data-structure}
+\end{figure}
 
@@
 
-\todo{Add sources section}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%    REFERENCES   %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\clearpage
+\section{Reference}
+\bibliographystyle{mcp}
+\bibliography{bibliography}
 

### 2021-04-15 main/MSstatsPTM_Main.tex
Added bib

@@
 \usepackage{hyperref}
-\usepackage[table,dvipsnames]{xcolor}
+\usepackage{xcolor} %\[table,dvipsnames]
 %\usepackage{makecell}
-\usepackage{array}
+%\usepackage{array}
 
@@
 \usepackage{tabulary,multirow,multicol,rotating}
+\usepackage[backend=biber]{biblatex}
+\addbibresource{bibliography.bib}
 
@@
 \def\devon#1{{\color{green}[4Devon: #1]}}
-%\def\maria#1{{\color{purple}[For Maria to update: #1]}}
-%\def\eduard#1{{\color{OliveGreen}[For Eduard to update: #1]}}
 
@@
 We implement this model in the free and open-source R package MSstatsPTM.
-\todo{Used overview from USHUPO poster}
 
@@
 The signaling mechanisms that allow cells to mount a dynamic and fast response to a multitude of events are primarily facilitated by the modification of proteins at specific residues, acting as molecular on/off switches.
-Mass spectrometry-based label-free proteomics is broadly established as the tool-of-choice for unbiased and large-scale identification and quantification of proteins and their post-translational modifications (PTMs) using liquid chromatography coupled with mass spectrometry (LC-MS).
+Mass spectrometry-based label-free proteomics is broadly established as the tool-of-choice for unbiased and large-scale identification and quantification of proteins and their post-translational modifications (PTMs) using liquid chromatography coupled with mass spectrometry (LC-MS)\cite{Kall:2011ub}.
 Studies targeting the post-translationally modified proteome focus either on the accurate localization of modification sites on proteins, relative or absolute quantification of a modification site’s occupancy repertoire, or relative changes in occupancy across experimental conditions.
@@
 Therefore, a robust approach to estimate systematic relative changes in post-translational modifications, at scale, should not only combine the quantitative information pertaining to a PTM site over peptides and replicates in multiple conditions, but take into account various sources of variations and confounding factors present in the experiments.
+\cite{Olsen:2013}
 
@@
 Many investigations performed differential expression analysis of PTMs using two-sample t-test or its extensions.
-The approach takes as input intensities of individual features from modified peptides, or intensity ratios of modified and unmodified peptide features, and compares the mean abundance of a PTM site from one condition to another.
+\todo{Add Ref about current PTM approaches} The approach takes as input intensities of individual features from modified peptides, or intensity ratios of modified and unmodified peptide features, and compares the mean abundance of a PTM site from one condition to another.
 Modifications of the t-test such as moderated t-test with limma were also proposed.
 While simple, the approach does not fully account for the sources of variations, and it is not directly applicable to experiments with complex designs, e.g., comparisons of multiple conditions, acquisition in multiple batches, etc.
-Isobar-PTM was developed for experiments with MS/MS quantitative strategies that employ isobaric labels such as tandem mass tags (TMT) and isobaric tag for relative and absolute quantification (iTRAQ).[Source] Isobar-PTM expresses MS measurements with a linear model and performs adjustment with respect to protein abundance using the difference between log-ratio of modified peptides in two channels and log-ratio of protein level.
+Isobar-PTM was developed for experiments with MS/MS quantitative strategies that employ isobaric labels such as tandem mass tags (TMT) and isobaric tag for relative and absolute quantification (iTRAQ).\cite{Breitwieser:2013} Isobar-PTM expresses MS measurements with a linear model and performs adjustment with respect to protein abundance using the difference between log-ratio of modified peptides in two channels and log-ratio of protein level.
 The modeling framework, however, is not applicable for either label-free workflows or experiments with complex designs.
-\todo{Add Source. Compare this to the TMT part of MSstatsPTM}
 
@@
 The proposed approach was evaluated using datasets from computer simulations, benchmark controlled mixtures and biological investigations, in comparison with the commonly applied t-test.
-The results demonstrated that by appropriately leveraging the information from the entire dataset the proposed approach improves the reproducibility and accuracy of the estimates of PTM fold changes, resultsin a better calibrated type I error rate, and improves the statistical power of detecting changes in PTMs.
+The results demonstrated that by appropriately leveraging the information from the entire dataset the proposed approach improves the reproducibility and accuracy of the estimates of PTM fold changes, results in a better calibrated type I error rate, and improves the statistical power of detecting changes in PTMs.
 For analyses of datasets with multiple batches, the proposed approach also improves the overall performance under various forms of batch effects.
-The proposed approach is implemented as an open source R package MSstatsPTM, which employs similar input format as in MSstats.
+The proposed approach is implemented as an open source R package MSstatsPTM, which employs similar input format as in MSstats and MSstatsTMT {Source to MSstats/MSstatsTMT}.
 
@@
 The features, which are precursor ions of modified or unmodified peptides, are used to characterize the identified PTM sites and proteins.
-For each PTM site, the feature log-intensities of the modified peptides spanning the site are expressed using a linear mixed model in consideration of the effects of condition, run, feature and interaction between run and feature.The model parameters are estimated using the split-plot approach as in MSstats[Source], where the feature log-intensities are first summarized into a single value per site per run in the subplot model, and the site-level summaries are then used for the inference of the PTM site abundance.
-In the site-level summarization, Tukey's median polish (TMP), a simple and robust procedure is applied to iteratively fit a two-way additive model with the effects of run and feature, which in turn summarizes the log-intensities for each site.
-[Source]  After the summarization, the inference of the PTM site abundance in each condition is carried out in account for the experimental design (Fig.
-3a).
-Statistical modeling and quantification for global proteomics data are performed by the same procedure as for PTM data.
-
-\todo{Add Sources}
+For each PTM site, the feature log-intensities of the modified peptides spanning the site are expressed using a linear mixed model in consideration of the effects of condition, run, feature and interaction between run and feature.
+The model parameters are estimated using the split-plot approach as in MSstats, where the feature log-intensities are first summarized into a single value per site per run in the subplot model, and the site-level summaries are then used for the inference of the PTM site abundance.\cite{Choi:2014} In the site-level summarization, Tukey's median polish (TMP), a simple and robust procedure is applied to iteratively fit a two-way additive model with the effects of run and feature, which in turn summarizes the log-intensities for each site.\cite{Tukey:1977} After summarization, the inference of the PTM site abundance in each condition is carried through fitting a model based on the family of linear mixed-effects models, taking into account the specific experimental design \figref{(Fig.
+3a)}.\cite{Bolker2009} \cite{Faraway:2006} Statistical modeling and quantification for global proteomics data are performed by the same procedure as for PTM data.
 
@@
 Detection of differentially modified PTM sites is performed through testing the null hypothesis of ‘no change’ against the alternative.
-The null hypothesis states that there is no difference in log-abundance of the PTM site between conditions, adjusted with respect to protein abundance (Fig.
-3b).
+The null hypothesis states that there is no difference in log-abundance of the PTM site between conditions, adjusted with respect to protein abundance \figref{Fig.
+3b}.
 Specifically, the adjusted difference is given by the difference in log-abundance of the PTM site, subtracted by the difference in log-abundance of the underlying protein, which is equivalent to the log of the ratio of PTM abundance difference to protein abundance difference.
-The estimate of the adjusted difference and the standard error (SE) of the estimate are obtained by combining the difference estimates and the associated SEs from both counterparts (for details, see Supplementary Sec.
-S3.1).
+The estimate of the adjusted difference and the standard error (SE) of the estimate are obtained by combining the difference estimates and the associated SEs from both counterparts \todo{link to Supplementary Sec.
+S3.1}.
 The test statistic for the hypothesis testing is the ratio of the estimate of the adjusted difference to its SE.
-To determine the statistical significance of the difference in terms of p-value, the test statistic is compared against the t distribution with degrees of freedom approximated by the Satterthwaite method.Adjustment for multiple comparisons is performed using the Benjamini-Hochberg procedure to control the false discovery rate at a desired level, e.g., 0.05.
-The details are provided in Supplementary Sec.
+To determine the statistical significance of the difference in terms of p-value, the test statistic is compared against the t distribution with degrees of freedom approximated by the Satterthwaite method.
+\cite{satterthwaite:1946} Adjustment for multiple comparisons is performed using the Benjamini-Hochberg procedure to control the false discovery rate at a desired level, e.g., 0.05.\cite{Benjamini:1995} The details are provided in Supplementary Sec.
 S3.2 \devon{link to supplementary using sfigref}.
@@
 Alternatively, assuming identical variability in the site-level summaries and identical differences between conditions across batches (i.e., no interaction effect between condition and batch), the statistical inference and testing can be performed based on one single model with an additional fixed effect of batch.
-Details about these modeling considerations are discussed in Supplementary Sec.
-S3.4.
+Details about these modeling considerations are discussed in Supplementary \todo{link to Sec.
+S3.4}.
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection*{Extension to TMT experiment}
+\subsection*{Missing Value imputation}
 
-\todo{Add in TMT approach}
+MSstatsPTM gives the ability to impute missing feature intensities if desired.
+When values are imputes, it is assumed they are missing for reasons of low abundance.
+To impute missing features in each MS run the Accelerated Time Failure model is used.\cite{Tukey:1977} In order to impute as feature's missing values, the feature must be present in at least one MS Run.
+If the feature is not present across all MS Runs, the value will be left missing.
+Missing value imputation is done before summarization with Tukey's Median Polish in order to correct for the impact of outliers.
+Missing value imputation is done separately for both the PTM and global protein datasets.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection*{Extension to TMT experiments}
+
+The statistical modeling approaches discussed above can also be extended to Tandem Mass Tag (TMT) labeling methods.
+TMT experiments introduce an additional source of variation in the form of different mixtures.
+To account for the mixture variable, first feature intensities are summarized as described previously, then a new linear mixed effects model with an added term for the mixture is included.
+This workflow and model follow the methods used for MSstatsTMT.
+\cite{Huang:2020} These methods are repeated to quantify and model global protein abundance.
+Once both PTM and global protein are modeled, the PTM model is adjusted for changes in global protein abundance using the same methods are before.
+
+\todo{add figref for TMT model}
+\todo{add figref for TMT experiment table}
 
@@
 We evaluated the impact with such adjustment by computer simulation, considering the following factors: with/without protein-level changes, effect size, number of replicates and number of conditions.
-Details of the simulation can be found in Supplementary Sec.
-S4.1.
-In this simulation, when there was no change in protein abundance, all the considered approaches well calibrated the Type I error rate (Fig.
-4a).
-However, when the modification changes were entirely due to the changes in protein abundance across conditions, analysis without accounting for the protein-level changes resulted in off-target, high false positive rates (Fig.
-4b).
-As shown in Fig.
-4c-e, in detection of systematic changes in PTM abundance, the proposed approach improved statistical power with small sample sizes in almost all the considered scenarios.
-The advantage by using the proposed approach over t-test became more profound in the presence of missing data (Fig.
-4d-e).
-In cases with small sample sizes (e.g., 2 replicates) as shown in Fig.
-4d, performance by t-test decreased dramatically with one missing value, and increasing the log-fold change to 2 did not effectively reduce the negative impact.
+Details of the simulation can be found in Supplementary \todo{link to Sec.
+S4.1}.
+In this simulation, when there was no change in protein abundance, all the considered approaches well calibrated the Type I error rate \todo{figref Fig.
+4a}.
+However, when the modification changes were entirely due to the changes in protein abundance across conditions, analysis without accounting for the protein-level changes resulted in off-target, high false positive rates \todo{figref (Fig.
+4b)}.
+As shown in \todo{figref Fig.
+4c-e}, in detection of systematic changes in PTM abundance, the proposed approach improved statistical power with small sample sizes in almost all the considered scenarios.
+The advantage by using the proposed approach over t-test became more profound in the presence of missing data \todo{figref (Fig.
+4d-e)}.
+In cases with small sample sizes (e.g., 2 replicates) as shown in \todo{figref Fig.
+4d}, performance by t-test decreased dramatically with one missing value, and increasing the log-fold change to 2 did not effectively reduce the negative impact.
 Two-sample t-test only used data within the groups of interest while ignoring the rest of the data.
 Consequently, it gave similar performance across cases with different number of conditions.
-In contrast, the proposed approach leveraged all available information, which resulted in improved power with increased number of conditions (see for example, the improved performance in Fig.
-4e [with 4 conditions] over Fig.
-4d [with 2 conditions]).
-Further results are provided in Supplementary Sec.
-S4.1.
+In contrast, the proposed approach leveraged all available information, which resulted in improved power with increased number of conditions (see for example, the improved performance in \todo{figref Fig.
+4e} [with 4 conditions] over \todo{figref Fig.
+4d} [with 2 conditions]).
+Further results are provided in Supplementary \todo{link to Sec.
+S4.1}.
 
-\paragraph*{Computer simulation: batch effects.} PTM experiments often create batches during data acquisition.Whereas the proposed approach can be naturally extended to handle such scenarios by modeling various forms of batch effects; t-test is not directly applicableto a problem with batches of data.
-Two commonly used ad-hoc approaches are a) t-test (no batch): ignoring batch effects when applying t-test, and b) t-test (most significant batch): applying t-test in each batch and drawing conclusions based on themost significant batch.
+\paragraph*{Computer simulation: batch effects.}
+
+PTM experiments often create batches during data acquisition.
+Whereas the proposed approach can be naturally extended to handle such scenarios by modeling various forms of batch effects; t-test is not directly applicable to a problem with batches of data.
+Two commonly used ad-hoc approaches are a) t-test (no batch): ignoring batch effects when applying t-test, and b) t-test (most significant batch): applying t-test in each batch and drawing conclusions based on the most significant batch.
 Although simple, these ad-hoc methods lack statistical justification.
-We characterized their statistical properties under various forms of batch effects in Supplementary Sec.
-S4.2.
-Twobatches of data were generated, where various forms of batch effects were simulated, includingdifference in signal intensities across batches, difference in variability across batches, and interactionbetween batch and condition (i.e., change between conditions affected by batch).
-The major results are summarized in Fig.
-5 with additional details provided in Supplementary Sec.
-S4.2.In terms of estimation of fold change, the most statistically significant batch with t-test was highly variable and frequently biased (Fig.
+We characterized their statistical properties under various forms of batch effects in Supplementary \todo{link to Sec.
+S4.2}.
+Two batches of data were generated, where various forms of batch effects were simulated, including difference in signal intensities across batches, difference in variability across batches, and interaction between batch and condition (i.e., change between conditions affected by batch).
+The major results are summarized in \todo{figref Fig.
+5} with additional details provided in Supplementary \todo{link to Sec.
+S4.2}.
+In terms of estimation of fold change, the most statistically significant batch with t-test was highly variable and frequently biased (Fig.
 5a).
@@
 As shown Fig.
-5c, the proposed approach improved power with small sample sizes in almost all thescenarios, under various forms of batch effects including difference in intensity level and difference in variability.
-Using t-test while ignoring batch effect gave similar performancein special cases with no difference in intensity level between batches (Supplementary Fig.
-SXX), but its performance dramatically decreased in general cases (Supplementary Figs.
-S8-S9).
+5c, the proposed approach improved power with small sample sizes in almost all the scenarios, under various forms of batch effects including difference in intensity level and difference in variability.
+Using t-test while ignoring batch effect gave similar performance in special cases with no difference in intensity level between batches (Supplementary Fig.
+\todo{find and link this section SXX}), but its performance dramatically decreased in general cases (Supplementary \todo{link to Figs.
+S8-S9}).
 The negative impact was only partially reduced by increasing the sample size to 5.
-The proposed approach gave consistently improved performancecompared with other methods by properly characterizing batch effects and leveraging all availableinformation.
+The proposed approach gave consistently improved performance compared with other methods by properly characterizing batch effects and leveraging all available information.
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection*{Spike-in experiment}
+\subsection*{Spike-in experiment}
 \todo{Need to write results of this experiment}
-\todo{Devon: I don't know what these Label-Free experiments are or how to analyze them. Meeting with Tsung-Heng on 2/5 to review}
 
-We evaluated our approach using a custom designed benchmark experiment, where
+We evaluated our approach using a custom designed benchmark experiment, where 50 heavy-labeled KGG motif peptides from 20 human proteins were used as spike-in peptides.
+Quantitative changes in protein and site abundance changes of these 20 proteins were the target of the benchmark.
+Unmodified peptides from Human Lysate were used as the estimate of global protein abundance changes.
+Additionally, E coli Lysate was used to normalize total protein levels prior to enrichment or global protein profiling.
+Four mixes of spike in peptides and Human Lysate were created to create conditions with known fold change.
+\todo{Add figure of mixtures} Two sets of data were acquired for each mixture: modifed peptide data, including the spike in KGG enriched peptides, and unmodified peptides from Human Lysate.
+These datasets were used as the PTM and global protein data.
+
+\todo{figref volcano plot of spike in peptides}
+
+In *FigXX* we can clearly see the red labeled spike-in peptides do not follow the expected fold change before adjusting for changes in global protein level.
+After adjustment they follow the expected fold change much closer.
+Additionally, background peptides that should not show a significant change, colored grey, show many false positives before adjustment is made.
+Again after adjustment the results improve, and the number of false positives decrease significantly.
+In this experiment adjusting to remove confounding with changes in global protein abundance leads to a more accurate conclusion about the effect of modified peptides between conditions.
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection*{Dilution experiment [What was this experiment?]}
+\subsection*{Dilution experiment [What was this experiment?]}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection*{Ubiquitin changes in bone marrow derived macrophages from Atg16L1 deficient mice Re-analysis of published dataset: USP30}
+\subsection*{Ubiquitin changes in bone marrow derived macrophages from Atg16L1 deficient mice Re-analysis of published dataset: USP30}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection*{Re-analysis of published dataset: RIP1 [Need this dataset and original findings]}
+\subsection*{Re-analysis of published dataset: RIP1 [Need this dataset and original findings]}
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection*{Re-analysis of published dataset: TMT experiments}
+\subsection*{Re-analysis of published dataset: TMT experiment}
+
+To evaluate our approach on TMT experiments we reanalyzed an experiment targeting primary murine macrophages infected with Shigella flexneri (S.flexneri).
+\todo{Add Shigella paper as source} Tandem mass tagging was used to quantify changes in total proteins, phosphorylation, and ubiquitination in Wild type (WT) controls and ATG16L1-deficient (cKO) bone marrow macrophages (BMDMs) either infected or uninfected with Shigella flexneri.
+The TMT acquisition was run with two mixtures consisting of 11 samples over an extended time period.
+Samples were taken at three seperate time periods, for a total of three test conditions.
+A global profiling run was also conducted, which was used to correct for changes in overall protein abundance.
 
@@
 The framework is general and is applicable to a variety of experimental designs.
-It outperforms the ad-hoc methods underlying the t-test, and yields accurate results in the broad type of experimental circumstances, including the presence of missing values, changes in protein abundance, and batch effects.
+It outperforms the ad-hoc methods underlying the t-test, and yields accurate results in the broad type of experimental circumstances, including the presence of missing values, changes in protein abundance, batch effects, and different acquisition methods.
 The framework allows us to plan for subsequent experiments, and choose the appropriate number of replicates in consideration of adjustment with respect to protein abundance.
@@
 A potential limitation of the proposed framework is the assumption that all the peptides are correctly mapped to the underlying proteins and PTM sites, and the features are informative of the abundances of underlying protein and PTM.
-Also, characterizing PTMs with current data-dependent acquisition workflows is prone to being undersampled, leading to a sparse dataset with a large number of missing values for the analysis.
+Also, characterizing PTMs with current data-dependent acquisition workflows is prone to being under sampled, leading to a sparse dataset with a large number of missing values for the analysis.
 Statistical methods accounting for effects due to experimental units and missing values introduced in this manuscript help interpret the data in a more objective manner.
@@
 \centering
-\includegraphics[scale=.8]{fig3}
+\includegraphics[scale=.8]{images/fig3}
 \caption{Data structure of a typical PTM experiment and goals of PTM characterization.(a)Schematic data representation, in a simplified case of two conditions and two replicate runs.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\clearpage
-\section{Reference}
-\bibliographystyle{mcp}
-\bibliography{bibliography}
 
-\end{document}n{document}
+\printbibliography
+
+\end{document}

### 2021-05-07 main/MSstatsPTM_Main.tex
Updates to main manuscript

@@
 
-%\usepackage{xr}
-%\externaldocument{supplementary}
+\usepackage{xr}
+\externaldocument[supp-]{ptm_sm}
 %\def\sfigref#1{{\bf Supplementary Fig.~\ref{fig:#1}}}
@@
 
-The signaling mechanisms that allow cells to mount a dynamic and fast response to a multitude of events are primarily facilitated by the modification of proteins at specific residues, acting as molecular on/off switches.
-Mass spectrometry-based label-free proteomics is broadly established as the tool-of-choice for unbiased and large-scale identification and quantification of proteins and their post-translational modifications (PTMs) using liquid chromatography coupled with mass spectrometry (LC-MS)\cite{Kall:2011ub}.
-Studies targeting the post-translationally modified proteome focus either on the accurate localization of modification sites on proteins, relative or absolute quantification of a modification site’s occupancy repertoire, or relative changes in occupancy across experimental conditions.
-Regardless of the question at hand, interrogating the modified proteome is challenging due to a number of reasons.
+The signaling mechanisms that allow cells to mount a dynamic and fast response to a multitude of events are primarily facilitated by the modification of proteins at specific residues, acting as molecular on/off switches.\cite{Deribe} \cite{Cohen}  Mass spectrometry-based label-free proteomics is broadly established as the tool-of-choice for unbiased and large-scale identification and quantification of proteins and their post-translational modifications (PTMs) using liquid chromatography coupled with mass spectrometry (LC-MS)\cite{Kall:2011ub} \cite{Roepstorff}.
+Studies targeting the post-translationally modified proteome focus either on the accurate localization of modification sites on proteins, relative or absolute quantification of a modification site’s occupancy repertoire, or relative changes in occupancy across experimental conditions.\cite{Mann} Regardless of the question at hand, interrogating the modified proteome is challenging due to a number of reasons.
 First, the relatively lower abundance of modified proteo-forms dictates that a global interrogation can only be achieved through large-scale enrichment protocols with modification-specific antibodies or beads.
@@
 Third, unless early signaling events are interrogated, the interpretation of the relative changes in modification occupancy are inherently convoluted with changes in the overall protein expression, making the interpretation of the results not straightforward.
-Therefore, a robust approach to estimate systematic relative changes in post-translational modifications, at scale, should not only combine the quantitative information pertaining to a PTM site over peptides and replicates in multiple conditions, but take into account various sources of variations and confounding factors present in the experiments.
-\cite{Olsen:2013}
+\cite{Olsen:2013} Therefore, a robust approach to estimate systematic relative changes in post-translational modifications, at scale, should not only combine the quantitative information pertaining to a PTM site over peptides and replicates in multiple conditions, but take into account various sources of variations and confounding factors present in the experiments.
 
@@
 \todo{Add Ref about current PTM approaches} The approach takes as input intensities of individual features from modified peptides, or intensity ratios of modified and unmodified peptide features, and compares the mean abundance of a PTM site from one condition to another.
-Modifications of the t-test such as moderated t-test with limma were also proposed.
-While simple, the approach does not fully account for the sources of variations, and it is not directly applicable to experiments with complex designs, e.g., comparisons of multiple conditions, acquisition in multiple batches, etc.
+Modifications of the t-test such as moderated t-test with limma were also proposed.\cite{Zhu} While simple, the approach does not fully account for the sources of variations, and it is not directly applicable to experiments with complex designs, e.g., comparisons of multiple conditions, acquisition in multiple batches, etc.
 Isobar-PTM was developed for experiments with MS/MS quantitative strategies that employ isobaric labels such as tandem mass tags (TMT) and isobaric tag for relative and absolute quantification (iTRAQ).\cite{Breitwieser:2013} Isobar-PTM expresses MS measurements with a linear model and performs adjustment with respect to protein abundance using the difference between log-ratio of modified peptides in two channels and log-ratio of protein level.
@@
 We propose a general statistical approach, which explicitly characterizes the variations and confounding factors present in bottom-up PTM experiments.
-With analysis procedure developed for summarization of LC-MS data, quantitative characterization of site-specific PTMs, and adjustment with respect to protein abundance, the proposed approach is aimed at the detection of quantitative changes in PTMs between conditions.
+The proposed approach is aimed at the detection of quantitative changes in PTMs between conditions utilizing procedures developed for summarization of LC-MS data, quantitative characterization of site-specific PTMs, and adjustment with respect to protein abundance.
 Quantitative analyses of PTMs often involve comparisons between multiple inter-related conditions of the same biological system.
-The general statistical framework underlying the proposed approach allows for analyzing experiments with complex designs, including those with multiple conditions and/or multiple batches.
+The general statistical framework underlying the proposed approach allows for analyzing experiments with complex designs, including those with multiple conditions, multiple batches, and measured at different points in time.
 
-The proposed approach was evaluated using datasets from computer simulations, benchmark controlled mixtures and biological investigations, in comparison with the commonly applied t-test.
-The results demonstrated that by appropriately leveraging the information from the entire dataset the proposed approach improves the reproducibility and accuracy of the estimates of PTM fold changes, results in a better calibrated type I error rate, and improves the statistical power of detecting changes in PTMs.
-For analyses of datasets with multiple batches, the proposed approach also improves the overall performance under various forms of batch effects.
-The proposed approach is implemented as an open source R package MSstatsPTM, which employs similar input format as in MSstats and MSstatsTMT {Source to MSstats/MSstatsTMT}.
+The proposed approach was evaluated using datasets from computer simulations, benchmark controlled mixtures, and biological investigations.
+The approach was then compared against the commonly applied t-test.
+The results demonstrate that by appropriately leveraging the information from the entire dataset the proposed approach improves the reproducibility and accuracy of the estimates of PTM fold changes, results in a better calibrated type I error rate, and improves the statistical power of detecting changes in PTMs.
+The proposed approach is implemented as an open source R package MSstatsPTM, which employs similar input format as in MSstats and MSstatsTMT \cite{Choi:2014} \cite{Huang:2020}.
 
@@
 
-\devon{Please change Fig.4a and so on using figref or tabref in the text. Put all figures and legends in figure section.}
-\figref{data-structure} schematically illustrates a simplified version of the data structure resulting from a typical bottom-up experiment for quantitative analysis of PTMs, in which there are multiple layers of variation present.
+\todo{Not sure what to do with this part and figure. How would this look in a paper?}
+Figure \ref{fig:data-structure} schematically illustrates a simplified version of the data structure resulting from a typical bottom-up experiment for quantitative analysis of PTMs, in which there are multiple layers of variation present.
 A PTM site is quantified with multiple spectral features, which vary in sequence (e.g., fully or partially cleaved peptides), ionization efficiency, charge states, etc.
@@
 For each PTM site, the feature log-intensities of the modified peptides spanning the site are expressed using a linear mixed model in consideration of the effects of condition, run, feature and interaction between run and feature.
-The model parameters are estimated using the split-plot approach as in MSstats, where the feature log-intensities are first summarized into a single value per site per run in the subplot model, and the site-level summaries are then used for the inference of the PTM site abundance.\cite{Choi:2014} In the site-level summarization, Tukey's median polish (TMP), a simple and robust procedure is applied to iteratively fit a two-way additive model with the effects of run and feature, which in turn summarizes the log-intensities for each site.\cite{Tukey:1977} After summarization, the inference of the PTM site abundance in each condition is carried through fitting a model based on the family of linear mixed-effects models, taking into account the specific experimental design \figref{(Fig.
-3a)}.\cite{Bolker2009} \cite{Faraway:2006} Statistical modeling and quantification for global proteomics data are performed by the same procedure as for PTM data.
+The model parameters are estimated using the split-plot approach as in MSstats, where the feature log-intensities are first summarized into a single value per site per run in the subplot model, and the site-level summaries are then used for the inference of the PTM site abundance.\cite{Choi:2014} In the site-level summarization, Tukey's median polish (TMP), a simple and robust procedure is applied to iteratively fit a two-way additive model with the effects of run and feature, which in turn summarizes the log-intensities for each site.\cite{Tukey:1977} After summarization, the inference of the PTM site abundance in each condition is carried through fitting a model based on the family of linear mixed-effects models, taking into account the specific experimental design \ref{fig:data-structure}.\cite{Bolker2009} \cite{Faraway:2006} Statistical modeling and quantification for global proteomics data are performed by the same procedure as for PTM data.
 
@@
 Detection of differentially modified PTM sites is performed through testing the null hypothesis of ‘no change’ against the alternative.
-The null hypothesis states that there is no difference in log-abundance of the PTM site between conditions, adjusted with respect to protein abundance \figref{Fig.
-3b}.
+The null hypothesis states that there is no difference in log-abundance of the PTM site between conditions, adjusted with respect to protein abundance \ref{fig:data-structure}.
 Specifically, the adjusted difference is given by the difference in log-abundance of the PTM site, subtracted by the difference in log-abundance of the underlying protein, which is equivalent to the log of the ratio of PTM abundance difference to protein abundance difference.
-The estimate of the adjusted difference and the standard error (SE) of the estimate are obtained by combining the difference estimates and the associated SEs from both counterparts \todo{link to Supplementary Sec.
-S3.1}.
+The estimate of the adjusted difference and the standard error (SE) of the estimate are obtained by combining the difference estimates and the associated SEs from both counterparts.
+The standard error of the estimate $\mathrm{SE}(\hat{\Delta})$ is
+
+\[
+\left[ \frac{2}{J} \left( \hat{\sigma}_{\gamma}^{2} + \hat{\sigma}_{\gamma^{\ast}}^{2} \right) \right]^{1/2}.
+\]
+
 The test statistic for the hypothesis testing is the ratio of the estimate of the adjusted difference to its SE.
@@
 MSstatsPTM gives the ability to impute missing feature intensities if desired.
-When values are imputes, it is assumed they are missing for reasons of low abundance.
-To impute missing features in each MS run the Accelerated Time Failure model is used.\cite{Tukey:1977} In order to impute as feature's missing values, the feature must be present in at least one MS Run.
+When values are imputed, it is assumed they are missing for reasons of low abundance.
+Missing features are imputed in each MS run using the Accelerated Time Failure model.\cite{Tukey:1977} In order to impute a feature's missing values, the feature must be present in at least one MS Run.
 If the feature is not present across all MS Runs, the value will be left missing.
@@
 
-\todo{need the result of TMT experiment}
-\todo{Add Shigella for TMT side}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \section{Figure}
-\todo{Unclear where Tsung-Heng meant these plots to go. Need to add them into correct sections}
 
@@
 \centering
-\includegraphics[scale=.8]{images/fig3}
-\caption{Data structure of a typical PTM experiment and goals of PTM characterization.(a)Schematic data representation, in a simplified case of two conditions and two replicate runs.
+\includegraphics[scale=.8]{images/fig1.png}
+\caption{}
+\label{fig:workflow}
+\end{figure}
+
+\begin{figure}[ht]
+\centering
+\includegraphics[scale=.8]{images/fig2.png}
+\caption{}
+\label{fig:batch}
+\end{figure}
+
+\begin{figure}[ht]
+\centering
+\includegraphics[scale=.8]{images/fig3.png}
+\caption{
+Data structure of a typical PTM experiment and goals of PTM characterization.(a)Schematic data representation, in a simplified case of two conditions and two replicate runs.
 Each PTM site is modeled and characterized separately, where a PTM is quantified with multiple spectral features (boxes), distinguished by different charge states of a peptide.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
+\newpage
 \printbibliography

### 2021-05-17 main/MSstatsPTM_Main.tex
Updated main manuscript

@@
 S4.1}.
+
 In this simulation, when there was no change in protein abundance, all the considered approaches well calibrated the Type I error rate \todo{figref Fig.
@@
 PTM experiments often create batches during data acquisition.
-Whereas the proposed approach can be naturally extended to handle such scenarios by modeling various forms of batch effects; t-test is not directly applicable to a problem with batches of data.
-Two commonly used ad-hoc approaches are a) t-test (no batch): ignoring batch effects when applying t-test, and b) t-test (most significant batch): applying t-test in each batch and drawing conclusions based on the most significant batch.
-Although simple, these ad-hoc methods lack statistical justification.
+The proposed approach can be naturally extended to handle such scenarios by modeling various forms of batch effects, as a t-test is not directly applicable to a problem with batches of data.
+Two commonly used ad-hoc approaches are a) t-test (no batch): ignoring batch effects when applying t-test, and b) t-test (most significant batch): applying t-test in each batch and drawing conclusions based on the most significant batch.\todo{Find source for batch analysis} Although simple, these ad-hoc methods lack statistical justification.
 We characterized their statistical properties under various forms of batch effects in Supplementary \todo{link to Sec.
@@
 Two batches of data were generated, where various forms of batch effects were simulated, including difference in signal intensities across batches, difference in variability across batches, and interaction between batch and condition (i.e., change between conditions affected by batch).
+
 The major results are summarized in \todo{figref Fig.
@@
 
-We evaluated our approach using a custom designed benchmark experiment, where 50 heavy-labeled KGG motif peptides from 20 human proteins were used as spike-in peptides.
+We evaluated our approach using a custom designed spike-in benchmark experiment where 50 heavy-labeled KGG motif peptides from 20 human proteins were used as spike-in peptides.
 Quantitative changes in protein and site abundance changes of these 20 proteins were the target of the benchmark.
 Unmodified peptides from Human Lysate were used as the estimate of global protein abundance changes.
+All comparisons with respect to human lysate serve as a null model, there was no change in global protein nor PTM abundance between conditions.
 Additionally, E coli Lysate was used to normalize total protein levels prior to enrichment or global protein profiling.
@@
 In *FigXX* we can clearly see the red labeled spike-in peptides do not follow the expected fold change before adjusting for changes in global protein level.
-After adjustment they follow the expected fold change much closer.
+After adjustment the estimated fold change is much close to expectation.
 Additionally, background peptides that should not show a significant change, colored grey, show many false positives before adjustment is made.
@@
 
-We evaluated our approach using a custom designed benchmark experiment, where...
-Two sets of liver samples from five Atg16L1 deficient mice (one with KGG enrichment followed by trypsin digestion, one with trypsin digestion alone) were prepared.
+We evaluated our approach using a custom designed dilution benchmark experiment where two sets of liver samples from five Atg16L1 deficient mice (one with KGG enrichment followed by trypsin digestion, one with trypsin digestion alone) were prepared.
 Following a randomized order, each sample was injected at three different concentrations 4ul, 2ul, and 1ul.
@@
 K1P1, as shown in Table XX.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection*{Ubiquitin changes in bone marrow derived macrophages from Atg16L1 deficient mice Re-analysis of published dataset: USP30}
-
-Compare results with/without adjustment\\
-Very few measurements of unmodified peptides are available\\
-Compare results with original findings] [Need original findings\\
-Include USP30OE vs. Ctrl in addition to Combo vs. CCCP?\\
-Comment on features, degrees of freedom\\
-Different changes across batches – P54652, Q5VYK3
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection*{Re-analysis of published dataset: RIP1 [Need this dataset and original findings]}
+\todo{Create or find this table}
 
@@
 To evaluate our approach on TMT experiments we reanalyzed an experiment targeting primary murine macrophages infected with Shigella flexneri (S.flexneri).
-\todo{Add Shigella paper as source} Tandem mass tagging was used to quantify changes in total proteins, phosphorylation, and ubiquitination in Wild type (WT) controls and ATG16L1-deficient (cKO) bone marrow macrophages (BMDMs) either infected or uninfected with Shigella flexneri.
-The TMT acquisition was run with two mixtures consisting of 11 samples over an extended time period.
-Samples were taken at three seperate time periods, for a total of three test conditions.
-A global profiling run was also conducted, which was used to correct for changes in overall protein abundance.
+\cite{Maculins} Tandem mass tagging was used to quantify changes in total proteins, phosphorylation, and ubiquitination in Wild type (WT) controls and ATG16L1-deficient (cKO) bone marrow macrophages (BMDMs) either infected or uninfected with Shigella flexneri.
+11-plex isobaric multiplex was used with two separate mixtures.
+Cell lysates were prepared over three different time periods, either uninfected, infected at early (45-60 min) or infected at late (3-3.5 hour) time points.
+There were 103,700 peptides mapping to 9,430 proteins in the global profiling run.
+Additionally, there were 25,600 unique phosphorylation sites and 12,400 KGG ubiquitination sites.
+About ~90\% of the identified modified peptides derived from proteins that were also quantified in the global profiling run.
+
+The major results are summarized below.
 

### 2021-06-01 main/MSstatsPTM_Main.tex
add comments 2021/06/01

@@
 %\noindent $\ast$   Corresponding author \\
-
-\devon{maybe good to have supplementary.txt in the same folder, to link figure, table number between main and supplementary.}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%    REFERENCES   %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\newpage
+\printbibliography
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\newpage
 \section{Figure}
@@
 \includegraphics[scale=.8]{images/fig1.png}
-\caption{}
+\caption{\todo{todo}}
 \label{fig:workflow}
@@
 \includegraphics[scale=.8]{images/fig2.png}
-\caption{}
+\caption{\todo{todo}}
 \label{fig:batch}
@@
 Some spectral features can be missing.(b)PTM relative quantification by statistical inference, which makes use of the feature intensities to infer the underlying PTM abundance and protein abundance with an estimate of associated uncertainty.
-(c) Model-based testing for differential PTM abundance, which corrects for the underlying protein abundance with a cost of increased uncertainty about the estimate of difference between conditions.(d) Statistical experimental design in terms of sample size calculations and power analysis.}
+(c) Model-based testing for differential PTM abundance, which corrects for the underlying protein abundance with a cost of increased uncertainty about the estimate of difference between conditions.
+\todo{there is no d..}(d) Statistical experimental design in terms of sample size calculations and power analysis.}
 \label{fig:data-structure}
@@
 
-Figure 4.False positive rate and statistical power in PTM significance analysis by the proposed approach, t-test with protein-level adjustment, and t-test without protein-level adjustment.
+{\bf Figure 4} False positive rate and statistical power in PTM significance analysis by the proposed approach, t-test with protein-level adjustment, and t-test without protein-level adjustment.
 (a) When there was no change in protein abundance, all considered methods well calibrated the Type I error rate.
@@
 
-Figure 5.Results corresponding to estimation error and false positive rate and statistical power of the PTM significance analysis, where the data were acquired in two batches.
+{\bf Figure 5} Results corresponding to estimation error and false positive rate and statistical power of the PTM significance analysis, where the data were acquired in two batches.
 The following parameters were considered to represent the batch effects: no batch-condition interaction, mean intensity level in Batch 2 was higher than Batch 1 by 2 on log scale, and the SDs in Batch 1 and Batch 2 were 0.2 and 0.3, respectively.(a)Estimation based on the most statistically significant batch with t-test was more variable than other methods and frequently biased.
@@
 
-Figure 6.Design of future PTM experiments in terms of sample size calculations and power analysis.
+{\bf Figure 6} Design of future PTM experiments in terms of sample size calculations and power analysis.
 (a) Protein-level adjustment relies on the inference of protein abundance, which introduces additional uncertainty in the estimate of PTM difference.
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%    REFERENCES   %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\newpage
-\printbibliography
-
 \end{document}

### 2021-06-01 supplementary/ptm_sm.tex
add comments 2021/06/01

@@
 The statistical significance of the difference is determined by comparing the test statistic against the $t$ distribution, with degrees of freedom $df=2J-2$ in balanced designs.
+
+\todo{If we use MSstats protein summarization + t-test, how much would they be different, methodologically and performance?}
 
@@
 
-%\clearpage
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsubsection{TMT experiment}
+\label{sec:tmtmethod}
+
+\todo{add the method for TMT experiment, batch vs plex?}
+
+\clearpage
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \caption{Correlation between test statistics across sites due to protein-level adjustment did not affect the expected level of FDR under the considered scenarios.
-The Benjamini-Hochberg procedure was used to control the FDR.
-\label{fig:prot_fdr}}
+The Benjamini-Hochberg procedure was used to control the FDR.\todo{make same color for each method} \label{fig:prot_fdr}}
 \end{figure}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{TMT experiment}
+\label{sec:tmtsim}
+
+\todo{add the simulation for TMT experiment, batch vs plex?}
+
+\clearpage
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Sample size calculation and power analysis}
@@
 
+\clearpage
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Datasets : Biological investigation}
+
+\todo{description and results}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2021-06-02 supplementary/ptm_sm.tex
some comments 06/02

@@
 \[
-\left[ \frac{2}{J} \left( \hat{\sigma}_{\gamma}^{2} + \hat{\sigma}_{\gamma^{\ast}}^{2} \right) \right]^{1/2}.
+SE(\hat{\Delta}) = \left[ \frac{2}{J} \left( \hat{\sigma}_{\gamma}^{2} + \hat{\sigma}_{\gamma^{\ast}}^{2} \right) \right]^{1/2}.
 \]
@@
 Consequently, it gave similar performance across cases with different number of conditions.
-In contrast, the proposed approach leveraged all available information, which resulted in improved power with increased number of conditions.
-\label{fig:prot_pwr}}
+In contrast, the proposed approach leveraged all available information, which resulted in improved power with increased number of conditions.\todo{how about protein abundance change?} \todo{this conclusion seems that overfitting..
+more number of conditions increase the power..
+Then how about limma?} \todo{how to calculate estimated sigma for 4 conditions and t-test?} \label{fig:prot_pwr}}
 \end{figure}
@@
 The results are summarized in \sfigref{fig:prot_fdr}.
+
+\todo{should address number of replicates, standard deviations, missing data?, same as one site per protein.}
+\todo{not strong evidence. need to revisit this section.}
 
@@
 The proposed approach adjusts for the underlying protein abundance in the PTM significance analysis, which corrects the confounding factor with a cost of increased uncertainty.
-We compared the required sample size with or without the adjustment to highlight the property (\sfigref{fig:size_prot}), based on a design of two conditions, in consideration of three pairs of standard deviations of log-intensities for modified and unmodified peptides: (0.2, 0.1), (0.2, 0.2), and (0.2, 0.4).
+\todo{uncertainty?
+means variation?} We compared the required sample size with or without the adjustment to highlight the property (\sfigref{fig:size_prot}), based on a design of two conditions, in consideration of three pairs of standard deviations of log-intensities for modified and unmodified peptides: (0.2, 0.1), (0.2, 0.2), and (0.2, 0.4).
 %We considered a design with two conditions, where standard deviations of log-intensities of both modified and unmodified peptides were both set as $0.2$. 

### 2021-07-13 supplementary/ptm_sm.tex
small update

--- 2021-06-01 supplementary/ptm_sm.tex
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Existing method: two-sample $t$-test}
-\label{sec:ttest}
+\section{Existing methods} \label{sec:ttest}
+\subsection{two-sample $t$-test}
 
@@
 \todo{If we use MSstats protein summarization + t-test, how much would they be different, methodologically and performance?}
+
+\subsection{Limma}
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Extension to complex designs}
-\label{sec:complex}
+\todo{not ready in MSstatsPTM v2 version.}
 
-The proposed statistical framework can be extended to analyze data from experiments of complex designs, such as factorial design and time series.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%\subsection{Extension to complex designs}
+%\label{sec:complex}
+%
+%The proposed statistical framework can be extended to analyze data from experiments of complex designs, such as factorial design and time series.
 We discuss below a specific design commonly applied in PTM experiments, in which data are acquired in multiple batches.
+%
+%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%\subsubsection{Batch effects}
+%\label{sec:batch}
+%
+%As in \secref{sec:test}, hypothesis testing on the adjusted log-abundances of the PTM in Conditions $i$ and $i'$ is performed to detect differentially modified PTM sites.
+The hypothesis is that there is no difference in adjusted log-abundance of the PTM between Conditions $i$ and $i'$
+%\begin{align*}
+%H_{0}: \Delta = (\mu_{i} - \mu_{i'}) - (\mu_{i}^{\ast} - \mu_{i'}^{\ast}) &= 0 \\
+%H_{a}: \Delta = (\mu_{i} - \mu_{i'}) - (\mu_{i}^{\ast} - \mu_{i'}^{\ast}) &\neq 0
+%\end{align*}
+%For batch-wise data, we consider two ways to estimate the difference in adjusted PTM abundance, $\Delta = (\mu_{i} - \mu_{i'}) - (\mu_{i}^{\ast} - \mu_{i'}^{\ast})$ based on different assumptions about the properties of batch effects, namely per-batch model (proposed approach) and all-batch model.
+For the following discussion, we denote the log-intensity of a modified peptide feature in Run $j$ of Condition $i$ and Batch $b$ by $y_{b, ijk}$, where $b = 1, \ldots, B$.
+Similarly, $y_{b, ijl}^{\ast}$ is denoted for the unmodified peptide feature in the same run.
+%
+%\paragraph{Per-batch model (proposed approach).} 
+%The model assumes different levels of variability are present in different batches and the differences between conditions vary across batches (i.e., there is an interaction effect between condition and batch).
+Difference between conditions is estimated in each batch, and the overall log-fold change in adjusted PTM abundance is estimated as the average over batches
+%\[
+%\hat{\Delta} = \frac{1}{B} \sum_{b=1}^{B} \left[ \frac{1}{J} \left( \hat{y}_{b,i+} - \hat{y}_{b,i'+} \right) \right]
+%- \frac{1}{B} \sum_{b=1}^{B} \left[ \frac{1}{J} \left( \hat{y}_{b,i+}^{\ast} - \hat{y}_{b,i'+}^{\ast} \right) \right].
+%\]
+%The standard error associated to the estimate is 
+%\[
+%\left[ \left(\frac{1}{B}\right)^2 \cdot \left( \frac{2}{J} \right) \cdot \sum_{b=1}^{B} \left( \hat{\sigma}_{\gamma_b}^{2} + \hat{\sigma}_{\gamma_b^{\ast}}^{2} \right) \right]^{1/2}.
+%\]
+%The test statistic $\hat{\Delta} / \mathrm{SE}(\hat{\Delta})$ is compared against the $t$ distribution, where the degrees of freedom are approximated as
+%\[
+%\left[ \sum_{b=1}^{B} \left( \hat{\sigma}_{\gamma_{b}}^{2} + \hat{\sigma}_{\gamma_{b}^{\ast}}^{2} \right) \right]^2 \bigg/
+%\sum_{b=1}^{B} \left[ \frac{\hat{\sigma}_{\gamma_{b}}^{4}}{  \mathrm{df}(\gamma_b)} + \frac{\hat{\sigma}_{\gamma_{b}^{\ast}}^{4}}{\mathrm{df}(\gamma_b^{\ast})} \right].
+%\]
+%
+%\paragraph{All-batch model.}
+%In contrast to the per-batch model, the all-batch model assumes identical variance and difference between conditions in all batches.
+The log-fold change in adjusted PTM abundance is estimated as the average over runs and batches
+%\[
+%\hat{\Delta} = \frac{1}{BJ} \left( \hat{y}_{+,i+} - \hat{y}_{+,i'+} \right) - \frac{1}{BJ} \left( \hat{y}_{+,i+}^{\ast} - \hat{y}_{+,i'+}^{\ast} \right),
+%\]
+%and the standard error of the estimate is 
+%\[
+%\left[ \frac{2}{BJ} \left( \hat{\sigma}_{\gamma}^{2} + \hat{\sigma}_{\gamma^{\ast}}^{2} \right) \right] ^{1/2}.
+%\]
+%The test statistic $\hat{\Delta} / \mathrm{SE}(\hat{\Delta})$ is compared against the $t$ distribution, with degrees of freedom approximated by
+%\[
+%\left( \hat{\sigma}_{\gamma}^{2} + \hat{\sigma}_{\gamma^{\ast}}^{2} \right)^2 \bigg/
+%\left( \frac{\hat{\sigma}_{\gamma}^{4}}{  \mathrm{df}(\gamma)} + \frac{\hat{\sigma}_{\gamma^{\ast}}^{4}}{\mathrm{df}(\gamma^{\ast})} \right).
+%\]
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection{Batch effects}
-\label{sec:batch}
+\subsubsection{TMT experiment}
+\label{sec:tmtmethod}
 
-As in \secref{sec:test}, hypothesis testing on the adjusted log-abundances of the PTM in Conditions $i$ and $i'$ is performed to detect differentially modified PTM sites.
-The hypothesis is that there is no difference in adjusted log-abundance of the PTM between Conditions $i$ and $i'$
+The methods described in this paper can be applied to Tandem Mass Tag (TMT) experiments targeting PTMs, in addition to batch effects.
+The same hypothesis testing is performed to detect differences in adjusted log-abundance of modifications in Condition $i$ and $i'$, where the null hypothesis is that there is no difference in PTM abundance between $i$ and $i'$.
 \begin{align*}
@@
 \end{align*}
-For batch-wise data, we consider two ways to estimate the difference in adjusted PTM abundance, $\Delta = (\mu_{i} - \mu_{i'}) - (\mu_{i}^{\ast} - \mu_{i'}^{\ast})$ based on different assumptions about the properties of batch effects, namely per-batch model (proposed approach) and all-batch model.
-For the following discussion, we denote the log-intensity of a modified peptide feature in Run $j$ of Condition $i$ and Batch $b$ by $y_{b, ijk}$, where $b = 1, \ldots, B$.
-Similarly, $y_{b, ijl}^{\ast}$ is denoted for the unmodified peptide feature in the same run.
 
-\paragraph{Per-batch model (proposed approach).} 
-The model assumes different levels of variability are present in different batches and the differences between conditions vary across batches (i.e., there is an interaction effect between condition and batch).
-Difference between conditions is estimated in each batch, and the overall log-fold change in adjusted PTM abundance is estimated as the average over batches
+For TMT experiments there is an additional source of variation coming from the mixtures used.
+This increases the complexity as there are many options for allocating samples to mixtures, including multiple mixtures, biological replicates, and technical replicates.
+Going forward we will denote the log-intensity of a modified peptide for Technical Replicate j of Mixture m for Condition i by $y_{ijm}$, where $m = 1,....,M$.
+This is done in the same way for the unmodified peptide $y^*_{ijm}$.
+
+The model assumes the variability of the mixture is a random effect and is not convoluted with the other variables in the model.
+The model assumes that the difference between conditions is the same between mixtures.
+For a given modified peptide and corresponding global protein a mixed effects model is fit as follows.
+
+\begin{align*}
+Y_{mtcb} = \mu + Mixture_m + Techrep(Mixture)_{jm} + Condition_c + Subject_{mcb} + \epsilon_{mtcb}\\
+Y^*_{mtcb} = \mu^* + Mixture^*_m + Techrep(Mixture)^*_{jm} + Condition^*_c + Subject^*_{mcb} + \epsilon^*_{mtcb}
+\end{align*}
+\begin{gather*}
+\text{Where }  Mixture_m \sim N(0, \sigma^2_m), Techrep(Mixture)_{jm} \sim N(0, \sigma^2_{jm}), Subject \sim N(0, \sigma^2_{s}), \\
+\epsilon_{mtcb} \sim N(0, sigma^2)
+\end{gather*}
+
+The log-fold change for adjusted PTM abundance is estimated as the average over mixtures and technical replicates
+
+$$\hat{\Delta} = \frac{1}{MJ} \left( \hat{y}_{+,+,i+} - \hat{y}_{+,+,i'+} \right) - \frac{1}{BJ} \left( \hat{y}_{+,+,i+}^{\ast} - \hat{y}_{+,+,i'+}^{\ast} \right)$$
+
+and the standard error of the estimate is
+
 \[
-\hat{\Delta} = \frac{1}{B} \sum_{b=1}^{B} \left[ \frac{1}{J} \left( \hat{y}_{b,i+} - \hat{y}_{b,i'+} \right) \right]
-- \frac{1}{B} \sum_{b=1}^{B} \left[ \frac{1}{J} \left( \hat{y}_{b,i+}^{\ast} - \hat{y}_{b,i'+}^{\ast} \right) \right].
-\]
-The standard error associated to the estimate is 
-\[
-\left[ \left(\frac{1}{B}\right)^2 \cdot \left( \frac{2}{J} \right) \cdot \sum_{b=1}^{B} \left( \hat{\sigma}_{\gamma_b}^{2} + \hat{\sigma}_{\gamma_b^{\ast}}^{2} \right) \right]^{1/2}.
-\]
-The test statistic $\hat{\Delta} / \mathrm{SE}(\hat{\Delta})$ is compared against the $t$ distribution, where the degrees of freedom are approximated as
-\[
-\left[ \sum_{b=1}^{B} \left( \hat{\sigma}_{\gamma_{b}}^{2} + \hat{\sigma}_{\gamma_{b}^{\ast}}^{2} \right) \right]^2 \bigg/
-\sum_{b=1}^{B} \left[ \frac{\hat{\sigma}_{\gamma_{b}}^{4}}{  \mathrm{df}(\gamma_b)} + \frac{\hat{\sigma}_{\gamma_{b}^{\ast}}^{4}}{\mathrm{df}(\gamma_b^{\ast})} \right].
-\]
-
-\paragraph{All-batch model.}
-In contrast to the per-batch model, the all-batch model assumes identical variance and difference between conditions in all batches.
-The log-fold change in adjusted PTM abundance is estimated as the average over runs and batches
-\[
-\hat{\Delta} = \frac{1}{BJ} \left( \hat{y}_{+,i+} - \hat{y}_{+,i'+} \right) - \frac{1}{BJ} \left( \hat{y}_{+,i+}^{\ast} - \hat{y}_{+,i'+}^{\ast} \right),
-\]
-and the standard error of the estimate is 
-\[
-\left[ \frac{2}{BJ} \left( \hat{\sigma}_{\gamma}^{2} + \hat{\sigma}_{\gamma^{\ast}}^{2} \right) \right] ^{1/2}.
+\left[ \frac{2}{MJ} \left( \hat{\sigma}_{\gamma}^{2} + \hat{\sigma}_{\gamma^{\ast}}^{2} \right) \right] ^{1/2}.
 \]
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection{TMT experiment}
-\label{sec:tmtmethod}
-
-\todo{add the method for TMT experiment, batch vs plex?}
+\todo{anything related to time-course?}
 
@@
 
-The proposed statistical approach was evaluated and compared to the $t$-test using computer simulation.
-In particular, their statistical properties with respect to protein-level adjustment and batch effects were evaluated.
+The proposed statistical approach was evaluated and compared to the $t$-test and Limma using computer simulation.
+In particular, their statistical properties with respect to protein-level adjustment and real world experimental conditions were evaluated.
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection{One site per protein}
-
 Differential levels of modified peptides may be due to differential modifications and/or changes in protein abundance.
 The proposed approach adjusts the PTM abundance with respect to protein abundance as introduced in \secref{sec:prop}.
-Alternatively, two-sample $t$-test taking as input the ratio between feature intensities of modified and unmodified peptides (difference on log scale) is commonly applied for the same purpose (\secref{sec:ttest}).
+Alternatively, two-sample $t$-test or Limma taking as input the ratio between feature intensities of modified and unmodified peptides is commonly applied for the same purpose (\secref{sec:ttest}).
 Approaches without considering protein-level adjustment lose track of an important aspect in interpreting observed changes in PTM abundance, which may result in misleading conclusions.
-To highlight the necessity of the adjustment, we compared the following approaches: a) proposed approach, b) linear model (no adjustment), i.e., same as the proposed approach but ignoring the measurements with respect to unmodified peptides, c) $t$-test (with adjustment), and d) $t$-test (no adjustment).
+To highlight the necessity of the adjustment, we compared the following approaches: a) proposed approach, b) proposed approach without adjusting for unmodified peptides c) $t$-test (with adjustment), d) $t$-test (no adjustment), e) Limma (with adjustment), and f) Limme (no adjustment).
+
 In experiments of complex designs, multiple inter-related conditions are often compared together.
-Whereas $t$-test uses measurements from the two conditions being compared, the proposed approach leverages measurements in all conditions for the inference of the underlying abundance.
-To highlight this distinction, multiple conditions of data were generated, where only one condition was simulated with a systematic change.
+The proposed approach and Limma leverage measurements in all conditions for the inference of the underlying abundance, whereas $t$-test uses measurements from the two conditions being compared.
+To highlight this distinction, multiple conditions of data were generated.
 The simulation was based on the following parameters:
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsubsection{Simulation 1}
+
+In the first simulation an experiment with many observations per protein/PTM and no missing data was created.
+This is not representative of a real experiment, but provides a baseline for model performance.
+
 \begin{itemize}
@@
 \item Standard deviations of log-intensities for modified and unmodified peptides: $0.2$, $0.3$
-\item Difference in PTM abundance between conditions: $0$, $0.5$, $0.75$, $1$
-\item Difference in protein abundance between conditions: $0$, $0.5$
-\item Number of replicates: $2$, $3$, $5$
+\item Difference in PTM abundance between conditions: $0$, $0.75$, $1.5$, $2.25$
+\item Difference in protein abundance between conditions: $0$, $0.75$, $1.5$, $2.25$
+\item Number of replicates: $2$, $3$, $5$, $10$
 \item Number of conditions: $2$, $3$, $4$
-\item Number of realizations: $500$
-\item Missing data: no missing value, or PTM missing in Run 1, Condition 1
+\item Number of realizations: $1000$
+\item Number of features per PTM: $10$
+\item Number of features per protein: $10$
+\item Missing data: no missing value
 \end{itemize}
-The results are summarized from \sfigref{fig:prot_fpr} to \sfigref{fig:protm1_pwr}, including false positive rate with or without changes in protein abundance by the considered methods (\sfigref{fig:prot_fpr}), power by the approaches with protein-level adjustment when there is no missing data (\sfigref{fig:prot_pwr}), and power by the approaches with protein-level adjustment when the PTM is missing in one run (\sfigref{fig:protm1_pwr}).
+
+The results are summarized from \sfigref{fig:prot_fpr} to \sfigref{fig:fc_boxplot}, including false positive rate with or without changes in protein abundance by the considered methods (\sfigref{fig:prot_fpr}), false positive rate, and overall accuracy.
+Additionally the results are compared to simulation 2 described in the next section.
+
+When simulating the data, 500 of the 1000 peptides were simulated with a differencial fold change between conditons, while the other 500 were not.
+In terms of the 500 peptides that were not differencial, 250 were simulated with a change in both PTM and overall protein abundance, such that the change in PTM abundance is entirely due to the change in the overall protein.
+The other 250 were simulated with no significant change in either PTM or overall protein abundance.
+In this way there were 500 true positives and 500 true negative PTMs.
+These were used to calculate the summary statistics seen in the following plots.
+
+\subsubsection{Simulation 2}
+
+In the second simulation we introduced limited feature observations per PTM as well as masking a portion of the observation to simulate missing values.
+This is more in line with what we would expect with a real life experiment and provides a more realistic expectation of model performance.
+
+\begin{itemize}
+\item Mean of log-intensity: $25$
+\item Standard deviations of log-intensities for modified and unmodified peptides: $0.2$, $0.3$
+\item Difference in PTM abundance between conditions: $0$, $0.75$, $1.5$, $2.25$
+\item Difference in protein abundance between conditions: $0$, $0.75$, $1.5$, $2.25$
+\item Number of replicates: $2$, $3$, $5$, $10$
+\item Number of conditions: $2$, $3$, $4$
+\item Number of realizations: $1000$
+\item Number of features per PTM: $2$
+\item Number of features per protein: $10$
+\item Missing data: 20\% of the observations for PTMs and Proteins were masked with NA at random
+\end{itemize}
+
+The results are summarized from \sfigref{fig:sim1_fpr} to \sfigref{fig:fc_boxplot}, including false positive rate with or without changes in protein abundance by the considered methods (\sfigref{fig:sim1_fpr}), false positive rate, and overall accuracy.
+Again the results are compared to simulation 1.
+
+The portion of significant peptides were done in the same was as described in simulation 1.
 
@@
 \centering
-\includegraphics[width=.85\textwidth]{sim/protnull_fpr}\\
-\includegraphics[width=.85\textwidth]{sim/protalt_fpr}
-\caption{All the considered methods well calibrated the Type I error rate when there was no change in protein abundance (upper plot). When the
-changes in PTM abundance were entirely due to changes in protein abundance across conditions (bottom plot), analysis without accounting for the protein-level changes resulted in off-target, high false positive rates.
-\label{fig:prot_fpr}}
+\includegraphics[width=.85\textwidth]{sim_new/sim1_FDR_all_models}\\
+\includegraphics[width=.85\textwidth]{sim_new/sim1_FDR}
+\caption{(upper plot)All the considered methods in simulation 1 correctly calibrated the Type I error rate when adjusting for changes in protein abundance.
+In comparison, the methods without accounting for the protein-level changes resulted in off-target, high false positive rates.
+(lower plot) The considered methods with protein adjustment are compared in detail.
+The proposed method generally performed better than Limma, but had a higher FDR than T-test.
+All methods performed strongly, with all FDR rates near or below .03.
+\label{fig:sim1_fpr}}
 \end{figure}
@@
 \centering
-\includegraphics[width=\textwidth]{sim/prot_pwr}
-%\includegraphics[width=.85\textwidth]{sim/prot_pwr}
-\caption{In comparison between the proposed approach and $t$-test with protein-level adjustment, the proposed approach improved power with small sample sizes.
-Two-sample $t$-test only used data within the groups of interest while ignoring the rest the of data.
-Consequently, it gave similar performance across cases with different number of conditions.
-In contrast, the proposed approach leveraged all available information, which resulted in improved power with increased number of conditions.
-\label{fig:prot_pwr}}
+\includegraphics[width=.85\textwidth]{sim_new/sim1_Recall}\\
+\includegraphics[width=.85\textwidth]{sim_new/sim1_accuracy}
+\caption{(upper plot) The recall for simulation 1 methods with adjustment were compared.
+It is clear that Limma performs the strongest here when the number of replicates were low.
+At higher replicates the performance of the proposed methods and Limma are comparable.
+T-test clearly performs worse across all methods.
+(lower plot) The overall accuracy plot mimics the observations in the recall plot.
+Limma performs stronger than the proposed method at lower replicates, while at higher they are comparable.
+\label{fig:sim1_recall}}
 \end{figure}
@@
 \centering
-\includegraphics[width=\textwidth]{sim/protm1_pwr}
-%\includegraphics[width=.85\textwidth]{sim/protm1_pwr}
-\caption{The advantage by using the proposed approach over two-sample $t$-test became more profound in the presence of missing value.
-In cases with small sample sizes, performance by the $t$-test decreased dramatically with one missing value.
-As a specific example, the statistical power by using the $t$-test with two replicates was below $0.25$, and increasing the fold change to $2$ did not effectively reduce the negative impact.
-\label{fig:protm1_pwr}}
+\includegraphics[width=.85\textwidth]{sim_new/sim3_Recall}
+\includegraphics[width=.85\textwidth]{sim_new/sim3_Accuracy}
+\caption{The advantage of using the proposed approach over Limma and $t$-test is apparent when looking at simulation 2, which includes limited observations and the presence of missing values.
+(upper plot) In the case of recall the proposed method performs stronger than limma and t-test in nearly every model.
+Even at lower replicates the proposed method still outperformed limma, especially when there were more conditions in the simulation.
+Again the lowest performance was observed using the $t$-test method.
+(lower plot) Again in the overall accuracy plots, the results are similar to recall.
+The propsed method performed the strongest across all methods, even when replicates were low.
+While Limma shows strong performance in a clean experiment, when real world data problems are introduced it is clear the proposed method is more robust.
+\label{fig:sim2_recall}}
 \end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection{Multiple sites per protein}
-
-Expression levels of PTM sites are adjusted based on the abundance of their originating protein.
-Since the same reference is used for all sites, it introduces correlation among estimates and test statistics for those sites.
-This may cause issues in controlling false discovery rate (FDR).
-We investigated the property by simulating data of two conditions and 1000 proteins with different numbers of PTM sites and comparing the results of the proposed approach and the linear model with no adjustment.
-%We investigated the property by simulating data of two conditions and 1000 proteins with different numbers of PTM sites. 
-A fraction (50\%) of the 1000 proteins had no changes between conditions while systematic changes were simulated for the rest of the proteins.
-Multiple testing correction was performed using the Benjamini and Hochberg's method.
-Performances of the considered approaches were assessed by their actual FDR, calculated as the fraction of proteins with adjusted $p$-values $<0.05$ among the proteins with true differences.
-The results are summarized in \sfigref{fig:prot_fdr}.
-%Performances of the considered approaches were assessed by their power and FDR.
-The power was calculated as the fraction of proteins with adjusted $p$-values $<0.05$ among those proteins with true differences between conditions, while the FDR was calculated as the fraction of proteins with adjusted $p$-values $<0.05$ among the proteins with true differences.
-The results are summarized in \sfigref{fig:prot_fdr}.
 
@@
 \centering
-\includegraphics[width=.85\textwidth]{sim/prot_fdr}
-\caption{Correlation between test statistics across sites due to protein-level adjustment did not affect the expected level of FDR under the considered scenarios.
-The Benjamini-Hochberg procedure was used to control the FDR.\todo{make same color for each method} \label{fig:prot_fdr}}
+\includegraphics[width=.85\textwidth]{sim_new/sim1_FC_boxplot}
+\includegraphics[width=.85\textwidth]{sim_new/sim3_FC_boxplot}
+\caption{(upper plot) In simulation 1 all considered methods correctly estimated the fold change between conditions, with a median fold change estimation of .75 in all methods.
+The distributions around the median were also consistent acorss all methods.
+Predictably the distribution quantiles decrease as the number of replicates increases.
+(lower plot) In simulation 2 again all methods correctly estimated the fold change with a median log change of .75.
+The proposed method in this simulation had a visibly tighter distribution around the median.
+Both Limma and $t$-test showed a wider range around the fold change.
+In this case the proposed method showed a stronger performed in correctly estimating the fold change for all peptides.
+\label{fig:fc_boxplot}}
 \end{figure}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Batch effects}
+\subsection{Benchmark experiment}
+\label{sec:benchmark}
 
-We evaluated the statistical properties of the proposed approach under batch effects, in comparison to the two-sample $t$-test (\secref{sec:ttest}).
-While the $t$-test is not directly applicable to a problem with batches of data, several \textit{ad-hoc} approaches may be used.
-Two commonly used approaches are a) $t$-test (no batch): ignoring batch effects when applying $t$-test, and b) $t$-test (most significant batch): applying $t$-test in each batch and drawing conclusions based on the most significant batch.
-Although simple, these \textit{ad-hoc} methods lack statistical justification.
-We characterized their statistical properties under various forms of batch effects.
-In this part of the simulation, two batches of data were generated, with the following forms of batch effects: difference in signal intensities across batches, difference in variability across batches, and interaction effect between batch and condition (i.e., change between condition affected by batch).
-The following interaction scenarios were considered in the simulation: a) no interaction between condition and batch, i.e., same change across conditions in both batches, b) positive interaction as $25$\% greater change in the batch of higher level, and c) negative interaction as $25$\% lower change in the batch of higher level.
-Below are the parameters used in the simulation:
-\begin{itemize}
-\item Mean of log-intensity: $25$
-\item Increased intensity level of Batch 2 versus Batch 1: $0$, $1$, $2$
-\item Standard deviations of log-intensity: $0.2$ in Batch 1 and $0.3$ in Batch 2
-\item Difference between conditions in Batch 1: $0.5$, $0.75$, $1$
-\item Number of replicates: $2$, $3$, $5$
-\item Number of conditions: $2$, $3$, $4$
-\item Number of realizations: $500$
-\end{itemize}
-The following four approaches were compared: a) proposed approach, i.e., per-batch model, b) all-batch model, c) $t$-test (no batch), and d) $t$-test (most significant batch).
-The results are summarized from \sfigref{fig:synnull_est} to \sfigref{fig:synnull_pwr_5}, on the aspects of estimation error (\sfigref{fig:synnull_est}), false positive rate (\sfigref{fig:synnull_fpr}), and power (\sfigref{fig:synnull_pwr}, \sfigref{fig:synnull_pwr_5}).
+A custom designed experiment with labeling was used to assess the performance of the proposed method in a real experimental setting.
+Heavy-labeld KGG modified peptides were used as spike-in peptides.
+The spike-in peptides were mixed with human lysate to create four mixture conditions.
+Two sets of data were acquired for each mixture: KGG enriched + LC-MS, and LC-MS only.
+The KGG enriched dataset included the spike-in peptides, as well as modified and unmodified human lysate.
+The LC-MS dataset included only unmodified peptides.
+The spike-in peptides where a significant fold change between conditions are expected to be differencial, whereas none of the human lysate peptides in any comparison are expected to be differencial.
+
+Again we consider three different methods and assess their performance: the method proposed in this paper, Limma, and two sample $t$-test.
+All methods are analyzed after adjusting for changes in overall protein level.
+The propsed method summarizes feature intensities up to the run level using Tukey's Median Polish, while the other methods use the log of summed feature intensities.
+The results are summarized from \sfigref{fig:spike_volcano} to \sfigref{fig:spike_volcano}, including volcano plots, model summary statsitics, and fold change analysis.
 
@@
 \centering
-\includegraphics[width=.825\textwidth]{sim/synnull_est_2}
-\caption{Estiamtion based on the most statistically significant batch with $t$-test was highly variable and frequently biased.
-The observation is consistent in all the simulated scenarios.
-\label{fig:synnull_est}}
+\includegraphics[width=.85\textwidth]{sim_new/spike_in_msstatsptm_volcano}
+\caption{Using the proposed method to model the benchmark experiment the spike in peptides (colored red) do not follow the expected log fold change before adjustment.
+After adjusting for changes in overall protein abundance the spike in peptides are more in line with expectation.
+Additionally the background grey colored peptides showed many false positives before adjustment.
+After adjustment the false positives were decreased considerably.
+\label{fig:spike_volcano_msstats}}
 \end{figure}
@@
 \centering
-\includegraphics[width=.825\textwidth]{sim/synnull_fpr}
-\caption{The proposed approach better calibrated Type I error rate. Similar results were observed in the cases with positive and negative batch-condition interactions. \label{fig:synnull_fpr}}
+\includegraphics[width=.85\textwidth]{sim_new/spike_in_limma_volcano}
+\caption{Using the limma method the spike in peptides follow the expected log fold change better after adjusting for changes in protein level.
+However, while the fold change is much more accurate, the majority of spike in peptides do not have a significante adjusted pvalue.
+In terms of false positives, the results are very similar to the propsed method, with many false positives before adjustment and much fewer afterward.
+\label{fig:spike_volcano_limma}}
 \end{figure}
-
-\clearpage
-\begin{figure}[h!]
-\centering
-\includegraphics[width=.825\textwidth]{sim/synnull_pwr_2}\\
-\includegraphics[width=.825\textwidth]{sim/synnull_pwr_3}
-\caption{The proposed approach improved power with small sample sizes in almost all the scenarios, under various forms of batch effects including difference in intensity level (higher on the right) and difference in variability.
-Using $t$-test while ignoring batch effect gave similar performance in special cases with no difference in intensity level between batches, but its performance dramatically decreased in general cases.
-The proposed methods gave consistently improved performance compared with other methods by properly characterizing batch effects and leveraging all available information.
-Similar patterns were observed in other simulated scenarios with positive and negative interactions.\label{fig:synnull_pwr}}
-\end{figure}
-
-%\clearpage
-%\begin{figure}[h!]
-%\centering
-%\includegraphics[width=.85\textwidth]{sim/synpos_pwr_3}\\
-%\includegraphics[width=.85\textwidth]{sim/synneg_pwr_3}
-%\caption{With 3 replicates and under the scenarios with either positive or negative interaction, the proposed approach consistently improved power over other methods.
-Similar results were observed with 2 replicates.
-As in those cases with no interaction, the approaches using two-sample $t$-test did not benefit from data in additional conditions.
-Also, ignoring batch effects as applying $t$-test (no batch) lost power dramatically.
-\label{fig:syn_pwr_3}}
-%\end{figure}
 
@@
 \centering
-\includegraphics[width=.825\textwidth]{sim/synnull_pwr_5}
-\caption{Ignoring batch effects lost power.
-The negative impact was only partially reduced by increasing the sample size to 5.
-Similar patterns were observed in other simulated scenarios.
-\label{fig:synnull_pwr_5}}
+\includegraphics[width=.85\textwidth]{sim_new/spike_in_ttest_volcano}
+\caption{Using the two sample $t$-test, none of the comparisons either before or after adjustment do not show any significant peptides.
+With that being said, the fold change of the spike in peptides is much closer to expectation after adjusting for global protein abundance.
+\label{fig:spike_volcano_ttest}}
+\end{figure}
+
+\begin{figure}[h!]
+\centering
+\includegraphics[width=.85\textwidth]{sim_new/spike_in_statistic_comparison}
+\caption{Comparing the summary statistics between methods it is clear that the proposed method performs the strongest.
+In terms of accuracy and specificity the three methods are close, with Limma and $t$-test showing slightly higher values.
+Accuracy and specificity are dominated by the large number of true negatives (background peptides) compared to the true positives (spike in peptides).
+In terms of recall, the proposed approach far out performed the other two methods, showing that it correctly labeled the most spike in peptides.
+\label{fig:spike_stat}}
+\end{figure}
+
+\begin{figure}[h!]
+\centering
+\includegraphics[width=.85\textwidth]{sim_new/spike_in_fc}
+\caption{The fold change misalignment before adjustment is illustrated in the boxplots.
+The true log fold change is indicated by the red 'X' marks.
+Before adjustment the boxplots are not close to the true fold change.
+After adjustment all three models generally have median near the true fold change.
+In terms of model comparison, the proposed method shows a much tighter distribution around the true fold change, whole both Limma and $t$-test are much wider.
+\label{fig:spike_fc}}
 \end{figure}
@@
 
-The proposed approach adjusts for the underlying protein abundance in the PTM significance analysis, which corrects the confounding factor with a cost of increased uncertainty.
+The proposed approach adjusts for the underlying protein abundance in the PTM significance analysis, which corrects the confounding factor with a cost of increased variation.
 We compared the required sample size with or without the adjustment to highlight the property (\sfigref{fig:size_prot}), based on a design of two conditions, in consideration of three pairs of standard deviations of log-intensities for modified and unmodified peptides: (0.2, 0.1), (0.2, 0.2), and (0.2, 0.4).
@@
 \clearpage
+
+\section{Further Considerations}
+
+\subsubsection{Multiple sites per protein}
+
+Expression levels of PTM sites are adjusted based on the abundance of their originating protein.
+Since the same reference is used for all sites, it introduces correlation among estimates and test statistics for those sites.
+This may cause issues in controlling false discovery rate (FDR).
+We investigated the property by simulating data of two conditions and 1000 proteins with different numbers of PTM sites and comparing the results of the proposed approach and the linear model with no adjustment.
+%We investigated the property by simulating data of two conditions and 1000 proteins with different numbers of PTM sites. 
+A fraction (50\%) of the 1000 proteins had no changes between conditions while systematic changes were simulated for the rest of the proteins.
+Multiple testing correction was performed using the Benjamini and Hochberg's method.
+Performances of the considered approaches were assessed by their actual FDR, calculated as the fraction of proteins with adjusted $p$-values $<0.05$ among the proteins with true differences.
+The results are summarized in \sfigref{fig:prot_fdr}.
+%Performances of the considered approaches were assessed by their power and FDR.
+The power was calculated as the fraction of proteins with adjusted $p$-values $<0.05$ among those proteins with true differences between conditions, while the FDR was calculated as the fraction of proteins with adjusted $p$-values $<0.05$ among the proteins with true differences.
+The results are summarized in \sfigref{fig:prot_fdr}.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\clearpage
 \section*{References}

### 2021-07-20 supplementary/ptm_sm.tex
updated USP30 supplementary

@@
 \usepackage{rotating}
-\usepackage[caption=false,font=footnotesize]{subfig}
 \usepackage{color}
@@
 \usepackage{graphicx}
-\usepackage{subfig}
+\usepackage{caption}
+\usepackage{subcaption}
 \usepackage{url}
@@
 \def\todo#1{{\color{red}[TODO: #1]}}
+
+\DeclareCaptionFormat{subfig}{\figurename~#1#2#3}
+\DeclareCaptionSubType*{figure}
+\captionsetup[subfigure]{format=subfig,labelsep=colon,labelformat=simple}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Existing methods} \label{sec:ttest}
-\subsection{two-sample $t$-test}
+\section{Existing method: two-sample $t$-test}
+\label{sec:ttest}
 
@@
 \todo{If we use MSstats protein summarization + t-test, how much would they be different, methodologically and performance?}
-
-\subsection{Limma}
 
@@
 
-\todo{not ready in MSstatsPTM v2 version.}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Extension to complex designs}
+\label{sec:complex}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\subsection{Extension to complex designs}
-%\label{sec:complex}
-%
-%The proposed statistical framework can be extended to analyze data from experiments of complex designs, such as factorial design and time series.
+The proposed statistical framework can be extended to analyze data from experiments of complex designs, such as factorial design and time series.
 We discuss below a specific design commonly applied in PTM experiments, in which data are acquired in multiple batches.
-%
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\subsubsection{Batch effects}
-%\label{sec:batch}
-%
-%As in \secref{sec:test}, hypothesis testing on the adjusted log-abundances of the PTM in Conditions $i$ and $i'$ is performed to detect differentially modified PTM sites.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsubsection{Batch effects}
+\label{sec:batch}
+
+As in \secref{sec:test}, hypothesis testing on the adjusted log-abundances of the PTM in Conditions $i$ and $i'$ is performed to detect differentially modified PTM sites.
 The hypothesis is that there is no difference in adjusted log-abundance of the PTM between Conditions $i$ and $i'$
-%\begin{align*}
-%H_{0}: \Delta = (\mu_{i} - \mu_{i'}) - (\mu_{i}^{\ast} - \mu_{i'}^{\ast}) &= 0 \\
-%H_{a}: \Delta = (\mu_{i} - \mu_{i'}) - (\mu_{i}^{\ast} - \mu_{i'}^{\ast}) &\neq 0
-%\end{align*}
-%For batch-wise data, we consider two ways to estimate the difference in adjusted PTM abundance, $\Delta = (\mu_{i} - \mu_{i'}) - (\mu_{i}^{\ast} - \mu_{i'}^{\ast})$ based on different assumptions about the properties of batch effects, namely per-batch model (proposed approach) and all-batch model.
+\begin{align*}
+H_{0}: \Delta = (\mu_{i} - \mu_{i'}) - (\mu_{i}^{\ast} - \mu_{i'}^{\ast}) &= 0 \\
+H_{a}: \Delta = (\mu_{i} - \mu_{i'}) - (\mu_{i}^{\ast} - \mu_{i'}^{\ast}) &\neq 0
+\end{align*}
+For batch-wise data, we consider two ways to estimate the difference in adjusted PTM abundance, $\Delta = (\mu_{i} - \mu_{i'}) - (\mu_{i}^{\ast} - \mu_{i'}^{\ast})$ based on different assumptions about the properties of batch effects, namely per-batch model (proposed approach) and all-batch model.
 For the following discussion, we denote the log-intensity of a modified peptide feature in Run $j$ of Condition $i$ and Batch $b$ by $y_{b, ijk}$, where $b = 1, \ldots, B$.
 Similarly, $y_{b, ijl}^{\ast}$ is denoted for the unmodified peptide feature in the same run.
-%
-%\paragraph{Per-batch model (proposed approach).} 
-%The model assumes different levels of variability are present in different batches and the differences between conditions vary across batches (i.e., there is an interaction effect between condition and batch).
+
+\paragraph{Per-batch model (proposed approach).} 
+The model assumes different levels of variability are present in different batches and the differences between conditions vary across batches (i.e., there is an interaction effect between condition and batch).
 Difference between conditions is estimated in each batch, and the overall log-fold change in adjusted PTM abundance is estimated as the average over batches
-%\[
-%\hat{\Delta} = \frac{1}{B} \sum_{b=1}^{B} \left[ \frac{1}{J} \left( \hat{y}_{b,i+} - \hat{y}_{b,i'+} \right) \right]
-%- \frac{1}{B} \sum_{b=1}^{B} \left[ \frac{1}{J} \left( \hat{y}_{b,i+}^{\ast} - \hat{y}_{b,i'+}^{\ast} \right) \right].
-%\]
-%The standard error associated to the estimate is 
-%\[
-%\left[ \left(\frac{1}{B}\right)^2 \cdot \left( \frac{2}{J} \right) \cdot \sum_{b=1}^{B} \left( \hat{\sigma}_{\gamma_b}^{2} + \hat{\sigma}_{\gamma_b^{\ast}}^{2} \right) \right]^{1/2}.
-%\]
-%The test statistic $\hat{\Delta} / \mathrm{SE}(\hat{\Delta})$ is compared against the $t$ distribution, where the degrees of freedom are approximated as
-%\[
-%\left[ \sum_{b=1}^{B} \left( \hat{\sigma}_{\gamma_{b}}^{2} + \hat{\sigma}_{\gamma_{b}^{\ast}}^{2} \right) \right]^2 \bigg/
-%\sum_{b=1}^{B} \left[ \frac{\hat{\sigma}_{\gamma_{b}}^{4}}{  \mathrm{df}(\gamma_b)} + \frac{\hat{\sigma}_{\gamma_{b}^{\ast}}^{4}}{\mathrm{df}(\gamma_b^{\ast})} \right].
-%\]
-%
-%\paragraph{All-batch model.}
-%In contrast to the per-batch model, the all-batch model assumes identical variance and difference between conditions in all batches.
+\[
+\hat{\Delta} = \frac{1}{B} \sum_{b=1}^{B} \left[ \frac{1}{J} \left( \hat{y}_{b,i+} - \hat{y}_{b,i'+} \right) \right]
+- \frac{1}{B} \sum_{b=1}^{B} \left[ \frac{1}{J} \left( \hat{y}_{b,i+}^{\ast} - \hat{y}_{b,i'+}^{\ast} \right) \right].
+\]
+The standard error associated to the estimate is 
+\[
+\left[ \left(\frac{1}{B}\right)^2 \cdot \left( \frac{2}{J} \right) \cdot \sum_{b=1}^{B} \left( \hat{\sigma}_{\gamma_b}^{2} + \hat{\sigma}_{\gamma_b^{\ast}}^{2} \right) \right]^{1/2}.
+\]
+The test statistic $\hat{\Delta} / \mathrm{SE}(\hat{\Delta})$ is compared against the $t$ distribution, where the degrees of freedom are approximated as
+\[
+\left[ \sum_{b=1}^{B} \left( \hat{\sigma}_{\gamma_{b}}^{2} + \hat{\sigma}_{\gamma_{b}^{\ast}}^{2} \right) \right]^2 \bigg/
+\sum_{b=1}^{B} \left[ \frac{\hat{\sigma}_{\gamma_{b}}^{4}}{  \mathrm{df}(\gamma_b)} + \frac{\hat{\sigma}_{\gamma_{b}^{\ast}}^{4}}{\mathrm{df}(\gamma_b^{\ast})} \right].
+\]
+
+\paragraph{All-batch model.}
+In contrast to the per-batch model, the all-batch model assumes identical variance and difference between conditions in all batches.
 The log-fold change in adjusted PTM abundance is estimated as the average over runs and batches
-%\[
-%\hat{\Delta} = \frac{1}{BJ} \left( \hat{y}_{+,i+} - \hat{y}_{+,i'+} \right) - \frac{1}{BJ} \left( \hat{y}_{+,i+}^{\ast} - \hat{y}_{+,i'+}^{\ast} \right),
-%\]
-%and the standard error of the estimate is 
-%\[
-%\left[ \frac{2}{BJ} \left( \hat{\sigma}_{\gamma}^{2} + \hat{\sigma}_{\gamma^{\ast}}^{2} \right) \right] ^{1/2}.
-%\]
-%The test statistic $\hat{\Delta} / \mathrm{SE}(\hat{\Delta})$ is compared against the $t$ distribution, with degrees of freedom approximated by
-%\[
-%\left( \hat{\sigma}_{\gamma}^{2} + \hat{\sigma}_{\gamma^{\ast}}^{2} \right)^2 \bigg/
-%\left( \frac{\hat{\sigma}_{\gamma}^{4}}{  \mathrm{df}(\gamma)} + \frac{\hat{\sigma}_{\gamma^{\ast}}^{4}}{\mathrm{df}(\gamma^{\ast})} \right).
-%\]
+\[
+\hat{\Delta} = \frac{1}{BJ} \left( \hat{y}_{+,i+} - \hat{y}_{+,i'+} \right) - \frac{1}{BJ} \left( \hat{y}_{+,i+}^{\ast} - \hat{y}_{+,i'+}^{\ast} \right),
+\]
+and the standard error of the estimate is 
+\[
+\left[ \frac{2}{BJ} \left( \hat{\sigma}_{\gamma}^{2} + \hat{\sigma}_{\gamma^{\ast}}^{2} \right) \right] ^{1/2}.
+\]
+The test statistic $\hat{\Delta} / \mathrm{SE}(\hat{\Delta})$ is compared against the $t$ distribution, with degrees of freedom approximated by
+\[
+\left( \hat{\sigma}_{\gamma}^{2} + \hat{\sigma}_{\gamma^{\ast}}^{2} \right)^2 \bigg/
+\left( \frac{\hat{\sigma}_{\gamma}^{4}}{  \mathrm{df}(\gamma)} + \frac{\hat{\sigma}_{\gamma^{\ast}}^{4}}{\mathrm{df}(\gamma^{\ast})} \right).
+\]
 
@@
 \]
-
-\todo{anything related to time-course?}
 
@@
 The propsed method summarizes feature intensities up to the run level using Tukey's Median Polish, while the other methods use the log of summed feature intensities.
-The results are summarized from \sfigref{fig:spike_volcano} to \sfigref{fig:spike_volcano}, including volcano plots, model summary statsitics, and fold change analysis.
+The results are summarized from \sfigref{fig:spike_volcano_msstats} to \sfigref{fig:spike_fc}, including volcano plots, model summary statsitics, and fold change analysis.
 
@@
 
-\todo{description and results}
+\subsection{USP30}
+
+This experiment was analyzed using the proposed method, limma, and two sample $t$-test.
+Becuase this is a real biological experiment we do not know the true postives and negatives as with the spike-in and simulation evaluations.
+The results of each method can be seen in \sfigref{fig:results_table}
+
+\begin{figure}[!ht]
+\begin{subfigure}[b]{1\linewidth}
+\centering
+\resizebox{10cm}{!}{
+ \begin{tabular}{|c|c|c|}
+
+ Model & Before Adjustment & After Adjustment \\
+\hline
+ Proposed & 4502 & 1625 \\ 
+ Limma & 1211 & 361\\
+ $t$-test & 28 & 7
+
+\end{tabular}}
+\caption{Significant modified peptides per model}
+\label{fig:results_table1a}
+\end{subfigure}
+ \vspace{1cm}
+\begin{subfigure}[b]{1\linewidth}
+\centering
+\includegraphics[width=.6\textwidth]{sim_new/usp30_ven}
+\caption{Shared significant modified peptides between models}
+\label{fig:results_table1b}
+\end{subfigure}
+\caption{A.
+The proposed method identifies far more signficant modified peptides compared to Limma and $t$-test.
+All methods identify substantially fewer peptides after adjusting for changes in global protein level.
+B.
+The proposed method and Limma share 16\% of all peptides identified as significant.
+The proposed method is the only model that identified 79\% of the peptides, while Limma identified 5\% that no other model did.
+$t$-test only identified 7 peptides as signififcant, with all but 1 being shared with the other two models.}
+\label{fig:1asdfa}
+\end{figure}
 

### 2021-09-09 code/MSstatsPTM_Simulation_Analysis.Rmd
Added analysis code and experiment papers

---
title: "MSstatsPTM_SImulation_Analysis"
author: "Devon Kohler"
date: "6/16/2021"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load_packages}
library(MSstatsPTMold)
library(MSstatsPTM)
library(data.table)
library(tidyverse)
library(limma)
```

```{r function, include=FALSE}
## Define Modeling Functions ---------------------------------------------------
convert_to_msstats_format <- function(df){
  ## Format into MSstatsPTM input
  df$PTM$protein <- paste(df$PTM$protein, df$PTM$site, sep = "_")
  df$PTM$PrecursorCharge <- NA
  df$PTM$FragmentIon <- NA
  df$PTM$ProductCharge <- NA
  df$PTM$IsotopeLabelType <- "L"
  df$PTM$BioReplicate <- paste0(df$PTM$run, "rep")
  df$PTM$log2inty <- 2**df$PTM$log2inty
  df$PTM$PeptideSequence <- paste(df$PTM$protein, df$PTM$feature, sep = "_")

  setnames(df$PTM, c("protein", "group", "run", "log2inty"),
           c("ProteinName", "Condition", "Run", "Intensity"))

  df$PROTEIN$PrecursorCharge <- NA
  df$PROTEIN$FragmentIon <- NA
  df$PROTEIN$ProductCharge <- NA
  df$PROTEIN$IsotopeLabelType <- "L"
  df$PROTEIN$log2inty <- 2**df$PROTEIN$log2inty
  df$PROTEIN$BioReplicate <- paste0(df$PROTEIN$run, "rep")
  df$PROTEIN$PeptideSequence <- paste(df$PROTEIN$protein, df$PROTEIN$feature, sep = "_")
  setnames(df$PROTEIN, c("protein", "group", "run", "log2inty"),
           c("ProteinName", "Condition", "Run", "Intensity"))

  return(df)
}

## Run ttest using run level summared data
run_ttest <- function(summarized_data){

  ptms <- unique(summarized_data$PTM)

  ttest_temp <- data.table()
  ttest_adj_temp <- data.table()
  ## Run ttest for each ptm
  for (p in seq_along(ptms)){
    temp_joined <- summarized_data %>% filter(PTM == ptms[[p]])
    groups <- (temp_joined %>% distinct(Condition))[[1]]
    t <- 2

    ## Loop over groups
    for (g in 1:(length(groups)-1)){
      for (g2 in 2:length(groups)){
        tryCatch({ttest_ptm <- t.test((temp_joined %>% filter(Condition == groups[g]) %>% select(Abundance.x))[[1]], 
                                      (temp_joined %>% filter(Condition == groups[g2]) %>% select(Abundance.x))[[1]])},
                 error=function(e){cat("ERROR :", as.character(p), "\n")})
        tryCatch({ttest_adj_ptm <- t.test((temp_joined %>% filter(Condition == groups[g]) %>% select(Adj_Abundance))[[1]], 
                                          (temp_joined %>% filter(Condition == groups[g2]) %>% select(Adj_Abundance))[[1]])},
                 error=function(e){cat("ERROR :", as.character(p), "\n")})
        ttest_temp <- rbindlist(list(ttest_temp, data.table(ptm = ptms[[p]], 
                                                            label = paste(groups[g], "vs", groups[g2], sep = " "), 
                                                            pval = ttest_ptm$p.value, 
                                                            tstat = ttest_ptm$statistic[[1]],
                                                            SE = ttest_ptm$stderr,
                                                            df = ttest_ptm$parameter[[1]],
                                                            estimate = ttest_ptm$estimate[[2]] - ttest_ptm$estimate[[1]])))
        ttest_adj_temp <- rbindlist(list(ttest_adj_temp, data.table(ptm = ptms[[p]], 
                                                                    label = paste(groups[g], "vs", groups[g2], sep = " "), 
                                                                    pval = ttest_adj_ptm$p.value, 
                                                                    tstat = ttest_adj_ptm$statistic[[1]],
                                                                    SE = ttest_adj_ptm$stderr,
                                                                    df = ttest_adj_ptm$parameter[[1]],
                                                                    estimate = ttest_adj_ptm$estimate[[2]] - ttest_adj_ptm$estimate[[1]])))
      }
    }
  }
  return(list(ttest_temp = ttest_temp, ttest_adj_temp = ttest_adj_temp))
}

## Limma pairwise function
design.pairs <- function(levels) {
    n <- length(levels)
    design <- matrix(0,n,choose(n,2))
    rownames(design) <- levels
    colnames(design) <- 1:choose(n,2)
    k <- 0
    for (i in 1:(n-1))
      for (j in (i+1):n) {
        k <- k+1
        design[i,k] <- 1
        design[j,k] <- -1
        colnames(design)[k] <- paste(levels[i],"-",levels[j],sep="")
      }
    design
}

## Fit limma model (both not and with adjust) for given dataset
fit_limma <- function(summarized_data, conditions, runs){

  ## Convert into format required for limma
  input <- data.frame(summarized_data %>% select(PTM, Run, Condition, Abundance.x) %>% 
                        pivot_wider(names_from = c(Condition, Run), values_from = Abundance.x,
                                    names_sort = TRUE))
  rownames(input) <- input$PTM
  input <- input %>% select(-PTM)
  input_adj <- data.frame(summarized_data %>% select(PTM, Run, Condition, Adj_Abundance) %>% 
                            pivot_wider(names_from = c(Condition, Run), values_from = Adj_Abundance,
                                        names_sort = TRUE))
  rownames(input_adj) <- input_adj$PTM
  input_adj <- input_adj %>% select(-PTM)

  ## Create contrast matrix
  class <- c()
  for (x in seq_len(conditions)){
    cond <- rep(paste0("Condition_", as.character(x)), runs)
    class <- c(class, cond)
  }
  class <- as.factor(class)
  design <- model.matrix(~0+class)

  input.matrix <- as.matrix(input)
  input_adj.matrix <- as.matrix(input_adj)

  ## Run models
  fit <- lmFit(input.matrix, design=design)
  fit_no_adj <- lmFit(input_adj.matrix, design=design)

  contrast.matrix <- design.pairs(colnames(design))
  fit2 <- contrasts.fit(fit, contrast.matrix)
  fit2 <- eBayes(fit2)

  fit2_no_adj <- contrasts.fit(fit_no_adj, contrast.matrix)
  fit2_no_adj <- eBayes(fit2_no_adj)

  ## Output to data.table
  comparisons <- data.table()
  comparisons_no_adj <- data.table()

  for (g in seq_along(colnames(fit2$coefficients))){
    comparisons <- rbindlist(list(comparisons, 
                                   data.table(PTM = rownames(fit2$coefficients),
                              Label = colnames(fit2$coefficients)[g],
                              Log2FC = as.vector(fit2$coefficients[,g]),
                              pvalue = as.vector(fit2$p.value[,g]),
                              df = as.vector(fit2$df.residual),
                              se = as.vector(fit2$sigma))))
    comparisons_no_adj <- rbindlist(list(comparisons_no_adj, 
                                          data.table(PTM = rownames(fit2_no_adj$coefficients),
                                     Label = colnames(fit2_no_adj$coefficients)[g],
                                     Log2FC = as.vector(fit2_no_adj$coefficients[,g]),
                                     pvalue = as.vector(fit2_no_adj$p.value[,g]),
                                     df = as.vector(fit2_no_adj$df.residual),
                                     se = as.vector(fit2_no_adj$sigma))))
  }

  return(list(limma_test <- comparisons, limma_no_adj_test = comparisons_no_adj))
}

## Define metrics
calculate_summary_stats <- function(df, protein_col_name, pval_col_name, sd, rep, conditions){
  fpr <- df %>% filter(get(pval_col_name) < .05 & grepl("NoChange", get(protein_col_name))) %>% 
    nrow() / nrow(df %>% filter(grepl("NoChange", get(protein_col_name))))
  sensitivity <- df %>% filter(get(pval_col_name) < .05 & !grepl("NoChange", get(protein_col_name))) %>% 
    nrow() / nrow(df %>% filter(!grepl("NoChange", get(protein_col_name))))
  specificity <- df %>% filter(get(pval_col_name) >= .05 & grepl("NoChange", get(protein_col_name))) %>% 
    nrow() / nrow(df %>% filter(grepl("NoChange", get(protein_col_name))))
  precision <- df %>% filter(get(pval_col_name) < .05 & !grepl("NoChange", get(protein_col_name))) %>% 
    nrow() / nrow(df %>% filter(get(pval_col_name) < .05))
  accuracy <- df %>% filter((get(pval_col_name) < .05 & !grepl("NoChange", get(protein_col_name))) | 
                              (get(pval_col_name) >= .05 & grepl("NoChange", get(protein_col_name)))) %>% 
    nrow() / nrow(df)
  recall <- df %>% filter(get(pval_col_name) < .05 & !grepl("NoChange", get(protein_col_name))) %>% 
    nrow() / df %>% filter(!grepl("NoChange", get(protein_col_name))) %>% nrow()
  results_temp <- data.table(fpr = fpr, sensitivity = sensitivity, 
                             specificity = specificity, precision = precision, 
                             sd = sd, rep = rep, conditions = conditions,
                             accuracy = accuracy, recall = recall)
  return(results_temp)
}
```

## Computer Simulation

### Simulation Methods

To simulate data the `PTMsimulateExperiment()` function from Tsung-Heng's original MSstatsPTM was used. This function allows us to vary the number of conditions, replicates, number of proteins, number of sites per protein, number of spectral features per site/protein, mean log2-abundance of PTM and PROTEIN, deviation from the mean log2-abundance in each group, standard deviation among replicates, and standard deviation among log2-intensities.

Three different statistical modeling methods were applied to the simulated data: MSstatsPTM, limma, and two-sample t-test. These methods were tested both with and without applying protein level adjustment. MSstatsPTM uses TMP for summarization and post modeling calculations for the adjustment. To adjust limma and t-test, the run-level data was averaged for both ptm and protein datasets and then combined. The resulting dataset was then used for limma and t-test.

Graphs can help us vizualize the components of the simulation.

```{r, echo=FALSE}
data.frame(ConditionA = rnorm(1000000, 22,2), ConditionB = rnorm(1000000, 17,2)) %>% ggplot() + 
  geom_density(aes(ConditionA), fill = "coral1") + geom_density(aes(ConditionB), fill = "steelblue1") +
  geom_vline(aes(xintercept=24), size = 1.25, color = "red") + 
  labs(title = "PTM1 between Condition A and B", x = "Abundance")
```

Here we can see a sample PTM with two conditions A and B. The red line represents a same biological replicate. We can simulate different variance's for the conditions.

```{r, echo=FALSE}
data.frame(ConditionA = rnorm(1000000, 0,2)) %>% ggplot() + 
  geom_density(aes(ConditionA), fill = "coral1") +
  geom_vline(aes(xintercept=.5), size = 1.25, color = "red") + 
  labs(title = "BioRep1 for PTM1 and Condition A", x = "Abundance")
```

Here we see the biorep from before, where the red line represents a single feature.

**All charts are made using adjusted pvalue <.05 to designate a significant hit.**

### First simulation

All simulations were ran with half the PTMs being differencial, while the other half the difference was due to changes in global protein level. The first simulation is run with the following paramters:

* Mean of log-intensity: $25$
* Number of Features: $10$ (PTM), $10$ (Protein)
* Standard deviations of log-intensities for modified and unmodified peptides: $0.2$, $0.3$
* Difference in PTM abundance between conditions: $0$, $0.5$, $0.75$, $1$
* Difference in protein abundance between conditions: $0$, $0.5$
* Number of replicates: $3$, $4$, $5$
* Number of conditions: $2$, $3$, $4$
* Number of realizations: $500$

``` {r first_sim_data, include = FALSE}

load(file = "model_data/ptm_models_sim1.rda")#ptm_models_sim1
load(file = "model_data/adjusted_models_sim1.rda")#adjusted_models_sim1 
load(file = "model_data/ttest_models_sim1.rda")#ttest_sim1
load(file = "model_data/adj_ttest_models_sim1.rda")#adj_ttest_sim1
load(file = "model_data/limma_models_sim1.rda")#limma_results_sim1
load(file = "model_data/adj_limma_models_sim1.rda")#adj_limma_sim1

s <- c(.2,.3)
reps <- c(2,3,5,10)
cond <- c(2,3,4)
param_combos <- expand.grid(s, reps, cond)

no_adj_results <- data.table()
adj_results <- data.table()
ttest_results <- data.table()
ttest_adj_results <- data.table()
limma_model_results <- data.table()
limma_adj_model_results <- data.table()

## Calculate results
## Calculate metrics for each model and dataset
for (i in seq_along(ptm_models_sim1)){

  no_adj_results_temp <- calculate_summary_stats(ptm_models_sim1[[i]], c("Protein"),  c("pvalue"), param_combos[i, 1], 
                                                 param_combos[i, 2], param_combos[i, 3])
  no_adj_results <- rbindlist(list(no_adj_results, no_adj_results_temp))

  adj_results_temp <- calculate_summary_stats(adjusted_models_sim1[[i]], c("Protein"),  c("adj.pvalue"), 
                                              param_combos[i, 1], param_combos[i, 2], param_combos[i, 3])
  adj_results <- rbindlist(list(adj_results, adj_results_temp))

  ttest_results_temp <- calculate_summary_stats(ttest_sim1[[i]], c("ptm"), c("pval"), param_combos[i, 1], 
                                              param_combos[i, 2], param_combos[i, 3])
  ttest_results <- rbindlist(list(ttest_results, ttest_results_temp))

  temp <- adj_ttest_sim1[[i]]
  temp_ttest <- data.table()
  for (c in seq_along(unique(temp$label))){
    temp_ttest <- rbindlist(list(temp_ttest, 
                                 temp %>% filter(label == unique(temp$label)[c]) %>% 
                                   mutate(adj.pvalue = p.adjust(pval, method = "BH"))))
  }
  ttest_adj_results_temp <- calculate_summary_stats(temp_ttest, c("ptm"), c("adj.pvalue"), param_combos[i, 1], 
                                                param_combos[i, 2], param_combos[i, 3])
  ttest_adj_results <- rbindlist(list(ttest_adj_results, ttest_adj_results_temp))

  limma_results_temp <- calculate_summary_stats(limma_results_sim1[[i]], c("PTM"), c("pvalue"), param_combos[i, 1], 
                                                    param_combos[i, 2], param_combos[i, 3])
  limma_model_results <- rbindlist(list(limma_model_results, limma_results_temp))

  temp <- adj_limma_sim1[[i]]
  temp_limma <- data.table()
  for (c in seq_along(unique(temp$Label))){
    temp_limma <- rbindlist(list(temp_limma, 
                                 temp %>% filter(Label == unique(temp$Label)[c]) %>% 
                                   mutate(adj.pvalue = p.adjust(pvalue, method = "BH"))))
  }

  limma_results_temp <- calculate_summary_stats(temp_limma, c("PTM"), c("adj.pvalue"), param_combos[i, 1], 
                                                param_combos[i, 2], param_combos[i, 3])
  limma_adj_model_results <- rbindlist(list(limma_adj_model_results, limma_results_temp))
}

## Label Models
adj_results$Model <- "Proposed"
no_adj_results$Model <- "Proposed no Adjustment"
ttest_results$Model <- "T-Test no Adjustment"
ttest_adj_results$Model <- "T-Test"
limma_model_results$Model <- "Limma no Adjustment"
limma_adj_model_results$Model <- "Limma"

## Combine to plot
models <- rbindlist(list(adj_results, no_adj_results, ttest_results, 
                         ttest_adj_results, limma_model_results, limma_adj_model_results), fill=TRUE)

```

#### False Discovery Rate (FDR)

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width = 7, fig.height = 6}
png("supplementary/sim_new/sim1_FDR_all_models.png", width = 750, height = 500)

models %>% mutate(Model = factor(Model, levels=c("Proposed", "Proposed no Adjustment", "Limma", 
                                               "Limma no Adjustment", "T-Test", "T-Test no Adjustment")),
                  Replicates = as.factor(rep)) %>%
  ggplot() + geom_point(aes(x = Model, y = fpr, size = Replicates, color = Model)) + 
  facet_grid(vars(sd), vars(conditions)) + 
  theme_bw() +
  theme(axis.text.x = element_blank(), 
                                                 axis.text.y = element_text(size = 12), 
                                                 legend.text=element_text(size=12),
                                                 axis.title.y = element_text(size = 14),
                                                 axis.title.x = element_text(size = 14),
                                                 title = element_text(size = 16),
                                                 strip.text = element_text(size = 12)) + 
  labs(title = "Simulation 1: FDR all models", x = "Model", y = "FDR")

dev.off()

```

We can clearly see that the non-adjusted models cannot adequately capture the change in global protein abundance. The FDR for all non-adjusted models is very high, while the adjusted models are all near zero. Thus the adjustment is obviously necessary to capture differential PTMs when they are convoluted with changes in protein level.

Because we can clearly see the adjustment has a large effect, we will only look at the adjusted models going forward.

``` {r, echo = FALSE, message=FALSE, warning=FALSE}
png("supplementary/sim_new/sim1_FDR.png", width = 750, height = 500)

models %>% mutate(Model = factor(Model, levels=c("Proposed", "Proposed no Adjustment", "Limma", 
                                               "Limma no Adjustment", "T-Test", "T-Test no Adjustment")),
                  Replicates = as.factor(rep)) %>%
  filter(Model %in% c("Proposed", "T-Test", "Limma")) %>% 
  ggplot() + geom_point(aes(x = Model, y = fpr, size = Replicates, color = Model)) + 
  facet_grid(vars(sd), vars(conditions)) + 
  theme_bw() +
  theme(axis.text.x = element_blank(), 
                                                 axis.text.y = element_text(size = 12), 
                                                 legend.text=element_text(size=12),
                                                 axis.title.y = element_text(size = 14),
                                                 axis.title.x = element_text(size = 14),
                                                 title = element_text(size = 16),
                                                 strip.text = element_text(size = 12)) + 
  labs(title = "Simulation 1: FDR adjusted models", x = "Model", y = "FDR")
dev.off()
```

When looking closer at the adjusted models only we can see that all the models we can see that they are indeed all very close to zero. The proposed solution has a slightly lower FDR than limma, while the ttest has the lowest value. This may be because limma uses empirical bayes to lower the variance in the models, allowing more values to show as significant.

#### Recall (TPR)

Next we will look at the true positive rate, to see how the models perform when identifying differential ptms.

```{r echo=FALSE, message=FALSE, warning=FALSE}
png("supplementary/sim_new/sim1_Recall.png", width = 750, height = 500)
models %>% mutate(Model = factor(Model, levels=c("Proposed", "Proposed no Adjustment", "Limma", 
                                               "Limma no Adjustment", "T-Test", "T-Test no Adjustment")),
                  Replicates = as.factor(rep)) %>%
  filter(Model %in% c("Proposed", "T-Test", "Limma")) %>% 
  ggplot() + geom_point(aes(x = Model, y = recall, size = Replicates, color = Model)) + 
  facet_grid(vars(sd), vars(conditions)) + 
  theme_bw() +
  theme(axis.text.x = element_blank(), 
                                                 axis.text.y = element_text(size = 12), 
                                                 legend.text=element_text(size=12),
                                                 axis.title.y = element_text(size = 14),
                                                 axis.title.x = element_text(size = 14),
                                                 title = element_text(size = 16),
                                                 strip.text = element_text(size = 12)) + 
  labs(title = "Simulation 1: Recall adjusted models", x = "Model", y = "Recall")
dev.off()
```

Recall is clearly best in Limma, followed by the proposed method, and then ttest. All the models performed poorly with high variance and only one condition. Additionally the lower number of replicates performed much worse. Limma has the advantage here of adjusting the variance, allowing more PTMs to show significant with lower replicates/sample size/ This can be seen in the bigger differences between the lower replicates and less conditions. With more conditions the proposed method (and limma to an extent) can leverage the information across conditions to lower the variance. T-test does not have this advantage and thus shows the worst results.

#### Accuracy

Next we will look at the overall accuracy, to see how the models perform in general.

```{r echo=FALSE, message=FALSE, warning=FALSE}
png("supplementary/sim_new/sim1_accuracy.png", width = 750, height = 500)
models %>% mutate(Model = factor(Model, levels=c("Proposed", "Proposed no Adjustment", "Limma", 
                                               "Limma no Adjustment", "T-Test", "T-Test no Adjustment")),
                  Replicates = as.factor(rep)) %>%
  filter(Model %in% c("Proposed", "T-Test", "Limma")) %>% 
  ggplot() + geom_point(aes(x = Model, y = accuracy, size = Replicates, color = Model)) + 
  facet_grid(vars(sd), vars(conditions)) + 
  theme_bw() +
  theme(axis.text.x = element_blank(), 
                                                 axis.text.y = element_text(size = 12), 
                                                 legend.text=element_text(size=12),
                                                 axis.title.y = element_text(size = 14),
                                                 axis.title.x = element_text(size = 14),
                                                 title = element_text(size = 16),
                                                 strip.text = element_text(size = 12)) + 
  labs(title = "Simulation 1: Accuracy adjusted models", x = "Model", y = "Accuracy")

dev.off()
```

In the accuracy chart we see much the same results as in recall. This is mainly due to the recall affecting the accuracy much more than FDR (all our models performed well in identifying the non-differencial PTMs after adjustment).

``` {r, echo = FALSE, fig.width = 10, fig.height = 7}

conds <- c("G_1 vs G_2", "G_2 vs G_3", "G_3 vs G_4")
cond2 <- c("classCondition_1-classCondition_2",
"classCondition_2-classCondition_3",
"classCondition_3-classCondition_4")

boxplot_data <- data.table()
for (i in seq_len(nrow(param_combos))){
  temp1 <- adjusted_models_sim1[[i]] %>% filter(!grepl("NoChange", Protein) & Label %in% conds)
  temp1$Model <- "Proposed"
  temp1$log2FC <- temp1$log2FC*-1
  temp2 <- adj_ttest_sim1[[i]] %>% filter(!grepl("NoChange", ptm) & label %in% conds) %>% rename(log2FC = estimate, Label = label)
  temp2$Model <- "T-test"
  temp3 <- adj_limma_sim1[[i]] %>% filter(!grepl("NoChange", PTM) & Label %in% cond2) %>% rename(log2FC = Log2FC)
  temp3$Model <- "Limma"
  temp3$log2FC <- temp3$log2FC*-1

  temp <- rbindlist(list(temp1, temp2, temp3), fill = TRUE)
  temp$sd <- param_combos[i, 1]
  temp$reps <- param_combos[i, 2]
  temp$cond <- param_combos[i, 3]

  boxplot_data <- rbindlist(list(boxplot_data, temp))
}

png("supplementary/sim_new/sim1_FC_boxplot.png", width = 750, height = 500)
boxplot_data %>% mutate(Model = factor(Model, levels=c("Proposed", "Limma", "T-test")),
                        reps = factor(reps, levels = c(2,3,5,10))) %>% 
  ggplot()  + geom_hline(yintercept = .75, color = "red", size = 1.1) + 
  geom_boxplot(aes(y = log2FC, x = reps, fill = Model)) + facet_grid(vars(sd), vars(cond)) + 
  theme_bw() +
  theme(axis.text.x = element_text(size = 12), 
                                                 axis.text.y = element_text(size = 12), 
                                                 legend.text=element_text(size=12),
                                                 axis.title.y = element_text(size = 14),
                                                 axis.title.x = element_text(size = 14),
                                                 title = element_text(size = 16),
                                                 strip.text = element_text(size = 12)) + 
  labs(title = "Simulation 1: Fold Change Distribution", x = "Replicates", y = "Log2FC")
dev.off()
```

Here we can see that all the models generally correctly estimate the expected Fold Change. Additionally the distributions of estimated FC are very similar across models. Because the FC between models is so similar, Limma's increased performance must be due to empirical bayes reducing the variance and allowing for better identification of the truly differential peptides.

In this simulation, we created a very clean experiment with a large amount of features and no missing values. In this environment Limma clearly performs well. With that being said, in the real world our experiments will not be this clean. In the following experiments we will adjust the simulation to better mimic real world experiments.

### Simulation 2

In this simulation we reduce the number of features for each PTM to 2, while keeping the global protein features at 10. In real world experiments the PTMs have much fewer experiments, as the peptide we use must span the modification site. This leads to an average of only a couple features per PTM.

* Mean of log-intensity: $25$
* Number of Features: $2$ (PTM), $10$ (Protein)
* Standard deviations of log-intensities for modified and unmodified peptides: $0.2$, $0.3$
* Difference in PTM abundance between conditions: $0$, $0.75$, $1.5$, $2.25$
* Difference in protein abundance between conditions: $0$, $0.75$, $1.5$, $2.25$
* Number of replicates: $2$, $4$, $6$
* Number of conditions: $2$, $3$, $4$
* Number of realizations: $1000$

```{r, include=FALSE}

load(file = "model_data/ptm_models_sim2.rda")#ptm_models_sim1
load(file = "model_data/adjusted_models_sim2.rda")#adjusted_models_sim1 
load(file = "model_data/ttest_models_sim2.rda")#ttest_sim1
load(file = "model_data/adj_ttest_models_sim2.rda")#adj_ttest_sim1
load(file = "model_data/limma_models_sim2.rda")#limma_results_sim1
load(file = "model_data/adj_limma_models_sim2.rda")#adj_limma_sim1

s <- c(.2,.3)
reps <- c(2,4,6)
cond <- c(2,3,4)
param_combos <- expand.grid(s, reps, cond)

no_adj_results <- data.table()
adj_results <- data.table()
ttest_results <- data.table()
ttest_adj_results <- data.table()
limma_model_results <- data.table()
limma_adj_model_results <- data.table()

## Calculate results
## Calculate metrics for each model and dataset
for (i in seq_along(ptm_models_sim2)){

  no_adj_results_temp <- calculate_summary_stats(ptm_models_sim2[[i]], c("Protein"),  c("pvalue"), param_combos[i, 1], 
                                                 param_combos[i, 2], param_combos[i, 3])
  no_adj_results <- rbindlist(list(no_adj_results, no_adj_results_temp))

  adj_results_temp <- calculate_summary_stats(adjusted_models_sim2[[i]], c("Protein"),  c("adj.pvalue"), 
                                              param_combos[i, 1], param_combos[i, 2], param_combos[i, 3])
  adj_results <- rbindlist(list(adj_results, adj_results_temp))

  ttest_results_temp <- calculate_summary_stats(ttest_sim2[[i]], c("ptm"), c("pval"), param_combos[i, 1], 
                                              param_combos[i, 2], param_combos[i, 3])
  ttest_results <- rbindlist(list(ttest_results, ttest_results_temp))

  temp <- adj_ttest_sim2[[i]]
  temp_ttest <- data.table()
  for (c in seq_along(unique(temp$label))){
    temp_ttest <- rbindlist(list(temp_ttest, 
                                 temp %>% filter(label == unique(temp$label)[c]) %>% 
                                   mutate(adj.pvalue = p.adjust(pval, method = "BH"))))
  }
  ttest_adj_results_temp <- calculate_summary_stats(temp_ttest, c("ptm"), c("adj.pvalue"), param_combos[i, 1], 
                                                param_combos[i, 2], param_combos[i, 3])
  ttest_adj_results <- rbindlist(list(ttest_adj_results, ttest_adj_results_temp))

  limma_results_temp <- calculate_summary_stats(limma_results_sim2[[i]], c("PTM"), c("pvalue"), param_combos[i, 1], 
                                                    param_combos[i, 2], param_combos[i, 3])
  limma_model_results <- rbindlist(list(limma_model_results, limma_results_temp))

  temp <- adj_limma_sim2[[i]]
  temp_limma <- data.table()
  for (c in seq_along(unique(temp$Label))){
    temp_limma <- rbindlist(list(temp_limma, 
                                 temp %>% filter(Label == unique(temp$Label)[c]) %>% 
                                   mutate(adj.pvalue = p.adjust(pvalue, method = "BH"))))
  }

  limma_results_temp <- calculate_summary_stats(temp_limma, c("PTM"), c("adj.pvalue"), param_combos[i, 1], 
                                                param_combos[i, 2], param_combos[i, 3])
  limma_adj_model_results <- rbindlist(list(limma_adj_model_results, limma_results_temp))
}

## Label Models
adj_results$Model <- "Proposed"
no_adj_results$Model <- "Proposed_no_adj"
ttest_results$Model <- "ttest_no_adj"
ttest_adj_results$Model <- "T-Test"
limma_model_results$Model <- "limma_no_adj"
limma_adj_model_results$Model <- "Limma"

## Combine to plot
models <- rbindlist(list(adj_results, no_adj_results, ttest_results, 
                         ttest_adj_results, limma_model_results, limma_adj_model_results), fill=TRUE)

```

In the following plots we will show the adjusted models only.

#### FDR

```{r, echo = FALSE, message=FALSE, warning=FALSE}
models %>% mutate(Model = factor(Model, levels=c("Proposed", "Proposed_no_adj", "limma_no_adj", 
                                               "Limma", "ttest_no_adj", "T-Test")),
                  Replicates = as.factor(rep)) %>%
  filter(Model %in% c("Proposed", "T-Test", "Limma")) %>% 
  ggplot() + geom_point(aes(x = Model, y = fpr, size = Replicates, color = Model)) + 
  facet_grid(vars(sd), vars(conditions)) + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + 
  labs(title = "FDR adjusted models", x = "Model", y = "FDR")
```

As in the last simulation, the t-test method performs the strongest in terms of FDR, with the proposed method and then limma following.

#### Recall (TPR)

Next we will look at the true positive rate, to see how the models perform when identifying differential ptms.

```{r, echo = FALSE, message=FALSE, warning=FALSE}
models %>% mutate(Model = factor(Model, levels=c("Proposed", "Proposed_no_adj", "limma_no_adj", 
                                               "Limma", "ttest_no_adj", "T-Test")),
                  Replicates = as.factor(rep)) %>%
  filter(Model %in% c("Proposed", "T-Test", "Limma")) %>% 
  ggplot() + geom_point(aes(x = Model, y = recall, size = Replicates, color = Model)) + 
  facet_grid(vars(sd), vars(conditions)) + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + 
  labs(title = "Recall adjusted models", x = "Model", y = "Recall")

```

In terms of Recall, Limma still performs the best, with the proposed method and t-test following. However we do see the proposed method is much closer to Limma in this simulation.

#### Accuracy

Next we will look at the overall accuracy, to see how the models perform in general.

```{r, echo = FALSE, message=FALSE, warning=FALSE}
models %>% mutate(Model = factor(Model, levels=c("Proposed", "Proposed_no_adj", "limma_no_adj", 
                                               "Limma", "ttest_no_adj", "T-Test")),
                  Replicates = as.factor(rep)) %>%
  filter(Model %in% c("Proposed", "T-Test", "Limma")) %>% 
  ggplot() + geom_point(aes(x = Model, y = accuracy, size = Replicates, color = Model)) + 
  facet_grid(vars(sd), vars(conditions)) + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + 
  labs(title = "Accuracy adjusted models", x = "Model", y = "Accuracy")
```

In terms of overall accuracy, again Limma performs the best, with the proposed method and t-test following. However, as with recall, we do see the proposed method being much closer to Limma in this simulation. It performs worse in lower number of conditions and replicates.

``` {r, echo = FALSE, fig.width = 10, fig.height = 7, message=FALSE, warning=FALSE}

conds <- c("G_1 vs G_2", "G_2 vs G_3", "G_3 vs G_4")
cond2 <- c("classCondition_1-classCondition_2",
"classCondition_2-classCondition_3",
"classCondition_3-classCondition_4")

boxplot_data <- data.table()
for (i in seq_len(nrow(param_combos))){
  temp1 <- adjusted_models_sim2[[i]] %>% filter(!grepl("NoChange", Protein) & Label %in% conds)
  temp1$Model <- "Proposed"
  temp1$log2FC <- temp1$log2FC*-1
  temp2 <- adj_ttest_sim2[[i]] %>% filter(!grepl("NoChange", ptm) & label %in% conds) %>% rename(log2FC = estimate, Label = label)
  temp2$Model <- "T-test"
  temp3 <- adj_limma_sim2[[i]] %>% filter(!grepl("NoChange", PTM) & Label %in% cond2) %>% rename(log2FC = Log2FC)
  temp3$Model <- "Limma"
  temp3$log2FC <- temp3$log2FC*-1

  temp <- rbindlist(list(temp1, temp2, temp3), fill = TRUE)
  temp$sd <- param_combos[i, 1]
  temp$reps <- param_combos[i, 2]
  temp$cond <- param_combos[i, 3]

  boxplot_data <- rbindlist(list(boxplot_data, temp))
}

boxplot_data %>% mutate(Model = factor(Model, levels=c("Proposed", "Limma", "T-test"))) %>% 
  ggplot()  + geom_hline(yintercept = .75, color = "red", size = 1.1) + 
  geom_boxplot(aes(y = log2FC, x = as.character(reps), fill = Model)) + facet_grid(vars(sd), vars(cond)) + 
  labs(title = "Fold Change Distribution", x = "Model", y = "Log2FC")

```

Again we see all the models following the expected FC, and their distributions are similar.

### Simulation 3

In this simulation we introduce missing values into the experiment. 10% of the features simulated were selected at random and masked with an NA value. No missing value imputation was used.

* Mean of log-intensity: $25$
* Number of Features: $2$ (PTM), $10$ (Protein)
* Standard deviations of log-intensities for modified and unmodified peptides: $0.2$, $0.3$
* Difference in PTM abundance between conditions: $0$, $0.75$, $1.5$, $2.25$
* Difference in protein abundance between conditions: $0$, $0.75$, $1.5$, $2.25$
* Number of replicates: $2$, $4$, $6$
* Number of conditions: $2$, $3$, $4$
* Number of realizations: $1000$
* 10% of features missing at random in PTM and Protein datasets

```{r, include=FALSE}

load(file = "model_data/ptm_models_sim3.rda")#ptm_models_sim1
load(file = "model_data/adjusted_models_sim3.rda")#adjusted_models_sim1 
load(file = "model_data/ttest_models_sim3.rda")#ttest_sim1
load(file = "model_data/adj_ttest_models_sim3.rda")#adj_ttest_sim1
load(file = "model_data/limma_models_sim3.rda")#limma_results_sim1
load(file = "model_data/adj_limma_models_sim3.rda")#adj_limma_sim1

s <- c(.2,.3)
reps <- c(2,3,5,10)
cond <- c(2,3,4)
param_combos <- expand.grid(s, reps, cond)

no_adj_results <- data.table()
adj_results <- data.table()
ttest_results <- data.table()
ttest_adj_results <- data.table()
limma_model_results <- data.table()
limma_adj_model_results <- data.table()

## Calculate results
## Calculate metrics for each model and dataset
for (i in seq_along(ptm_models_sim3)){

  no_adj_results_temp <- calculate_summary_stats(ptm_models_sim3[[i]], c("Protein"),  c("pvalue"), param_combos[i, 1], 
                                                 param_combos[i, 2], param_combos[i, 3])
  no_adj_results <- rbindlist(list(no_adj_results, no_adj_results_temp))

  adj_results_temp <- calculate_summary_stats(adjusted_models_sim3[[i]], c("Protein"),  c("adj.pvalue"), 
                                              param_combos[i, 1], param_combos[i, 2], param_combos[i, 3])
  adj_results <- rbindlist(list(adj_results, adj_results_temp))

  ttest_results_temp <- calculate_summary_stats(ttest_sim3[[i]], c("ptm"), c("pval"), param_combos[i, 1], 
                                              param_combos[i, 2], param_combos[i, 3])
  ttest_results <- rbindlist(list(ttest_results, ttest_results_temp))

  temp <- adj_ttest_sim3[[i]]
  temp_ttest <- data.table()
  for (c in seq_along(unique(temp$label))){
    temp_ttest <- rbindlist(list(temp_ttest, 
                                 temp %>% filter(label == unique(temp$label)[c]) %>% 
                                   mutate(adj.pvalue = p.adjust(pval, method = "BH"))))
  }
  ttest_adj_results_temp <- calculate_summary_stats(temp_ttest, c("ptm"), c("adj.pvalue"), param_combos[i, 1], 
                                                param_combos[i, 2], param_combos[i, 3])
  ttest_adj_results <- rbindlist(list(ttest_adj_results, ttest_adj_results_temp))

  limma_results_temp <- calculate_summary_stats(limma_results_sim3[[i]], c("PTM"), c("pvalue"), param_combos[i, 1], 
                                                    param_combos[i, 2], param_combos[i, 3])
  limma_model_results <- rbindlist(list(limma_model_results, limma_results_temp))

  temp <- adj_limma_sim3[[i]]
  temp_limma <- data.table()
  for (c in seq_along(unique(temp$Label))){
    temp_limma <- rbindlist(list(temp_limma, 
                                 temp %>% filter(Label == unique(temp$Label)[c]) %>% 
                                   mutate(adj.pvalue = p.adjust(pvalue, method = "BH"))))
  }

  limma_results_temp <- calculate_summary_stats(temp_limma, c("PTM"), c("adj.pvalue"), param_combos[i, 1], 
                                                param_combos[i, 2], param_combos[i, 3])
  limma_adj_model_results <- rbindlist(list(limma_adj_model_results, limma_results_temp))
}

## Label Models
adj_results$Model <- "Proposed"
no_adj_results$Model <- "Proposed_no_adj"
ttest_results$Model <- "ttest_no_adj"
ttest_adj_results$Model <- "T-Test"
limma_model_results$Model <- "limma_no_adj"
limma_adj_model_results$Model <- "Limma"

## Combine to plot
models <- rbindlist(list(adj_results, no_adj_results, ttest_results, 
                         ttest_adj_results, limma_model_results, limma_adj_model_results), fill=TRUE)

```

In the following plots we will show the adjusted models only.

#### FDR

``` {r, echo = FALSE, message=FALSE, warning=FALSE}
png("supplementary/sim_new/sim3_FDR.png", width = 350, height = 350)
models %>% mutate(Model = factor(Model, levels=c("Proposed", "Proposed_no_adj", "limma_no_adj", 
                                               "Limma", "ttest_no_adj", "T-Test")),
                  Replicates = as.factor(rep)) %>%
  filter(Model %in% c("Proposed", "T-Test", "Limma")) %>% 
  ggplot() + geom_point(aes(x = Model, y = fpr, size = Replicates, color = Model)) + 
  facet_grid(vars(sd), vars(conditions)) + 
  theme_bw() +
  theme(axis.text.x = element_blank(), 
                                                 axis.text.y = element_text(size = 12), 
                                                 legend.text=element_text(size=12),
                                                 axis.title.y = element_text(size = 14),
                                                 axis.title.x = element_text(size = 14),
                                                 title = element_text(size = 16),
                                                 strip.text = element_text(size = 12)) + 
  labs(title = "FDR adjusted models", x = "Model", y = "FDR")
dev.off()
```

The FDR still generally shows the proposed method better than limma, although they are closer than what we saw previously.

#### Recall (TPR)

Next we will look at the true positive rate, to see how the models perform when identifying differential ptms.

```{r echo=FALSE, message=FALSE, warning=FALSE}
png("supplementary/sim_new/sim3_Recall.png", width = 750, height = 500)
models %>% mutate(Model = factor(Model, levels=c("Proposed", "Proposed_no_adj", "limma_no_adj", 
                                               "Limma", "ttest_no_adj", "T-Test")),
                  Replicates = as.factor(rep)) %>%
  filter(Model %in% c("Proposed", "T-Test", "Limma")) %>% 
  ggplot() + geom_point(aes(x = Model, y = recall, size = Replicates, color = Model)) + 
  facet_grid(vars(sd), vars(conditions)) + 
  theme_bw() +
  theme(axis.text.x = element_blank(), 
                                                 axis.text.y = element_text(size = 12), 
                                                 legend.text=element_text(size=12),
                                                 axis.title.y = element_text(size = 14),
                                                 axis.title.x = element_text(size = 14),
                                                 title = element_text(size = 16),
                                                 strip.text = element_text(size = 12)) + 
  labs(title = "Simulation 2: Recall adjusted models", x = "Model", y = "Recall")
dev.off()
```

In terms of recall we can see that the proposed method performs much better in the presence of missing values compared to Limma. There are some exceptions to this, notably at lower replicates with higher variance, however in general the proposed performs better.

#### Accuracy

Next we will look at the overall accuracy, to see how the models perform in general.

```{r echo=FALSE, message=FALSE, warning=FALSE}
png("supplementary/sim_new/sim3_Accuracy.png", width = 750, height = 500)
models %>% mutate(Model = factor(Model, levels=c("Proposed", "Proposed_no_adj", "limma_no_adj", 
                                               "Limma", "ttest_no_adj", "T-Test")),
                  Replicates = as.factor(rep)) %>%
  filter(Model %in% c("Proposed", "T-Test", "Limma")) %>% 
  ggplot() + geom_point(aes(x = Model, y = accuracy, size = Replicates, color = Model)) + 
  facet_grid(vars(sd), vars(conditions)) + 
  theme_bw() +
  theme(axis.text.x = element_blank(), 
                                                 axis.text.y = element_text(size = 12), 
                                                 legend.text=element_text(size=12),
                                                 axis.title.y = element_text(size = 14),
                                                 axis.title.x = element_text(size = 14),
                                                 title = element_text(size = 16),
                                                 strip.text = element_text(size = 12)) + 
  labs(title = "Simulation 2: Accuracy adjusted models", x = "Model", y = "Accuracy")
dev.off()
```

Overall accuracy is better in the proposed method than Limma for every model except two.

``` {r, echo = FALSE, fig.width = 10, fig.height = 7, message=FALSE, warning=FALSE}

conds <- c("G_1 vs G_2", "G_2 vs G_3", "G_3 vs G_4")
cond2 <- c("classCondition_1-classCondition_2",
"classCondition_2-classCondition_3",
"classCondition_3-classCondition_4")

boxplot_data <- data.table()
for (i in seq_len(nrow(param_combos))){
  temp1 <- adjusted_models_sim3[[i]] %>% filter(!grepl("NoChange", Protein) & Label %in% conds)
  temp1$Model <- "Proposed"
  temp1$log2FC <- temp1$log2FC*-1
  temp2 <- adj_ttest_sim3[[i]] %>% filter(!grepl("NoChange", ptm) & label %in% conds) %>% rename(log2FC = estimate, Label = label)
  temp2$Model <- "T-test"
  temp3 <- adj_limma_sim3[[i]] %>% filter(!grepl("NoChange", PTM) & Label %in% cond2) %>% rename(log2FC = Log2FC)
  temp3$Model <- "Limma"
  temp3$log2FC <- temp3$log2FC*-1

  temp <- rbindlist(list(temp1, temp2, temp3), fill = TRUE)
  temp$sd <- param_combos[i, 1]
  temp$reps <- param_combos[i, 2]
  temp$cond <- param_combos[i, 3]

  boxplot_data <- rbindlist(list(boxplot_data, temp))
}

png("supplementary/sim_new/sim3_FC_boxplot.png", width = 750, height = 500)
boxplot_data %>% mutate(Model = factor(Model, levels=c("Proposed", "Limma", "T-test")),
                        reps = factor(reps, levels = c(2,3,5,10))) %>% 
ggplot()  + geom_hline(yintercept = .75, color = "red", size = 1.1) + 
geom_boxplot(aes(y = log2FC, x = reps, fill = Model)) + facet_grid(vars(sd), vars(cond)) + 
  theme_bw() +
  theme(axis.text.x = element_text(size = 12), 
                                               axis.text.y = element_text(size = 12), 
                                               legend.text=element_text(size=12),
                                               axis.title.y = element_text(size = 14),
                                               axis.title.x = element_text(size = 14),
                                               title = element_text(size = 16),
                                               strip.text = element_text(size = 12)) + 
labs(title = "Simulation 2: Fold Change Distribution", x = "Replicates", y = "Log2FC")
dev.off()
```

Interesting, when introducing missing values we see a difference in this plot compared to the other simulations. The median FC still is in line with the expected FC across all models. However the proposed method has a tighter distribution compared to the other models, especially Limma.

### 2021-09-09 code/Spike-in_Analysis.Rmd
Added analysis code and experiment papers

---
title: "Spike-in_MSstatsPTM_Analysis"
author: "Devon Kohler"
date: "6/18/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load_packages}
library(MSstatsPTMold)
library(MSstatsPTM)
library(data.table)
library(tidyverse)
library(limma)
```

## Spike-in Dataset Method Comparison

This markdown file examines the custom made spike-in analysis targeting PTMs. This allows us to know which PTMs are actually differential and which are not. The models used in this analysis are: MSstatsPTM, limma, and two condition t-test. For MSstatsPTM the full workflow is used for run-level summarization and modeling. Limma and t-test use feature averaging for run summarization and then their coresponding models are fit. The model results are shown first and then a comparison is made.

```{r include=FALSE}
## Some general function definitions -------------------------------------------

## Define metrics
calculate_summary_stats <- function(df, pval_col_name, comparison_col_name){

  ## FP/N (FPR)
  fpr <- df %>% filter((get(pval_col_name) < .05 & labeling == "light") | 
                (get(comparison_col_name) == "mix1-mix4" & get(pval_col_name) < .05)) %>% 
    nrow() / nrow(df %>% filter(labeling == "light" | get(comparison_col_name) == "mix1-mix4"))

  ## TP/P (TPR)
  recall <- df %>% filter(get(pval_col_name) < .05 & labeling == "heavy" & 
                                 get(comparison_col_name) != "mix1-mix4") %>% 
    nrow() / df %>% filter(labeling == "heavy" & 
                             get(comparison_col_name) != "mix1-mix4") %>% nrow()

  ## TN/N (TNR)
  specificity <- df %>% filter(get(pval_col_name) >= .05 & (labeling == "light" |
                                 get(comparison_col_name) == "mix1-mix4")) %>% 
    nrow() / nrow(df %>% filter(labeling == "light" | get(comparison_col_name) == "mix1-mix4"))

  ## TP/TP+FP (PPV)
  precision <- df %>% filter(get(pval_col_name) < .05 & labeling == "heavy" & 
                               get(comparison_col_name) != "mix1-mix4") %>% 
    nrow() / nrow(df %>% filter(get(pval_col_name) < .05))

  ## TP + TN / N + P
  accuracy <- df %>% filter((get(pval_col_name) < .05 & labeling == "heavy" & 
                               get(comparison_col_name) != "mix1-mix4") | 
                              (get(pval_col_name) >= .05 & (labeling == "light" | 
                               get(comparison_col_name) == "mix1-mix4"))) %>% 
    nrow() / nrow(df)

  results_temp <- data.table(metric = c("FPR", "Recall", "Specificity", "Precision", "Accuracy"),
                             result = c(fpr, recall, specificity, precision, accuracy))
  return(results_temp)
}

```

### Data Stats

##### Missing Values

PTM: 20%
Protein: 30%

##### Average Features

PTM: 1.4
Protein: 10.2

### MSstatsPTM

#### Volcano Plot

First a comparison between the adjusted and unadjusted models are shown. The spike-in peptides are colored red and their expected log Fold Change is indicated in the subplot title. The background peptides are colored grey and should all be insignificant.

```{r pressure, echo=FALSE, fig.width = 10, fig.height = 7}
load(file = "model_data/Spike_in_MSstatsPTM_Model.rda")#model_df

model_df$ADJUSTED.Model <- model_df$ADJUSTED.Model %>% filter(is.finite(pvalue))
model_df$PTM.Model <- model_df$PTM.Model %>% filter(is.finite(pvalue))

plot_df <- rbind(model_df$ADJUSTED.Model %>% select(-GlobalProtein), 
              model_df$PTM.Model %>% select(-c(MissingPercentage,ImputationPercentage, issue)))
plot_df$protadj <- factor(plot_df$protadj, levels = c('no adjustment','protein adjustment'))

# Volcano plot
png("supplementary/sim_new/spike_in_msstatsptm_volcano.png", width = 750, height = 500)
plot_df %>% arrange(desc(labeling)) %>% 
  filter(abs(log2FC) < Inf & Label %in% c("mix1-mix3", "mix1-mix4", "mix2-mix3", "mix2-mix4")) %>%
  ggplot(aes(log2FC, -log10(adj.pvalue))) + 
  geom_point(aes(color = labeling, alpha = labeling)) + 
  scale_color_manual(values = c("#D55E00", "#999999")) + 
  scale_alpha_manual(values = c(1, 0.2)) + 
  scale_size_manual(values = c(2, 1)) + 
  geom_hline(yintercept = 0) + 
  geom_vline(xintercept = 0) + 
  geom_hline(yintercept = -log10(0.05), color = "darkred") + 
  geom_vline(xintercept = c(1, -1), color = "darkred", linetype = "dashed") + 
  facet_grid(protadj ~ contrast_truth) + 
  theme_bw() + 
  theme(axis.text.x = element_text(size = 12), 
       axis.text.y = element_text(size = 12), 
       legend.text=element_text(size=12),
       axis.title.y = element_text(size = 14),
       axis.title.x = element_text(size = 14),
       title = element_text(size = 16),
       strip.text = element_text(size = 12),
       legend.position = "bottom") + 
  labs(title = "Proposed Method VolcanoPlot")
dev.off()
```

We can clearly see the effect of the adjustment in the volcano plots. Before adjustment, the red spike-in peptides did not follow the expected fold change. Additionally there are a large amount of false positives, shown as grey dots above the significance threshold. After applying protein level adjustment the red spike-in peptides are much more in line with expectation and there are many less false positives.

#### Summary Statistics

Summary statistics for the models are given below.

```{r, echo = FALSE, fig.width = 8}

unadj_stats <- calculate_summary_stats(model_df$PTM.Model, 'adj.pvalue', "Label")
unadj_stats$Adjustment <- "no adjustment"
adj_stats <- calculate_summary_stats(model_df$ADJUSTED.Model, 'adj.pvalue', "Label")
adj_stats$Adjustment <- "protein adjustment"
msstatsptm_stats <- rbindlist(list(unadj_stats, adj_stats))

msstatsptm_stats %>% ggplot() + geom_col(aes(x = metric, y = result, fill = Adjustment), position = "dodge") + 
  scale_fill_manual(values = c("salmon1", "steelblue1")) + 
  labs(title = "MSstatsPTM - Summary Statistics", x = "Metric", y = "Value")

```

In the summary statistics we can clearly see the advantage of the protein adjustment. Overall accuracy increases to near 100% (although we need to keep in mind it is dominated by the large number of true negatives). The False Positive Rate decreases substantially. Precision increases but is still quiet low. This is mainly due to the very large number of true negatives. While we were able to correctly categorize a large portion of the true negatives, there are still a large number that were missed and dominate this statistic. Recall only saw a slight increase in the adjusted model. This is because, while the fold change does now follow what we expected, a large number of the spike-in peptides still had an adjusted p value over .05. This is most likely due to increases in the variance when combined with the protein model. Finally specificity also saw a large increase when adding in the protein adjustment, indicating we are correctly identifying most of the background peptides as negatives.

#### Log Fold Change - Spike-in Peptides

Lastly, the plot below looks at the distribution spike-in peptides log fold changes. The expected log Fold Change is indicated by the red `X` marks.

```{r, echo = FALSE, fig.width = 10, fig.height = 5}
unadj_spike <- model_df$PTM.Model %>% filter(labeling == "heavy")
unadj_spike$Adjustment <- "no adjustment"
adj_spike <- model_df$ADJUSTED.Model %>% filter(labeling == "heavy")
adj_spike$Adjustment <- "protein adjustment"
spike <- rbindlist(list(unadj_spike, adj_spike), fill = TRUE)

spike %>% ggplot() + geom_boxplot(aes(x = contrast_truth, y = log2FC), fill = "lightsteelblue4") + 
  geom_point(aes(y = trueLog2FC, x = contrast_truth), color = "red", shape = 4, size = 3) + facet_wrap(.~Adjustment) + 
  labs(title = "MSstatsPTM Fold Change of Spike-in Peptides", x = "Comparison with Expected FC") + theme(axis.text.x = element_text(angle = 30, vjust = .5))

```

We can see in the plot above that the fold change of the spike-in peptides falls in line with expectation much better after adjustment. The expected change falls very close to the median value, and is at least within the 1st and 3rd quartile boxes in every comparison. In the unadjusted plot, the expected falls outside the quartile boxes in the majority of comparisons.

### Limma

We will repeat the above analysis for Limma.

#### Volcano Plot

First a comparison between the adjusted and unadjusted models are shown. Again the spike-in peptides are colored red and their expected log Fold Change is indicated in the subplot title. The background peptides are colored grey and should all be insignificant.

```{r, echo=FALSE, fig.width = 10, fig.height = 7}
load(file = "model_data/Spike_in_Limma_Model.rda")#limma_model

limma_model <- list(PTM.Model = limma_model %>% filter(protadj == "no adjustment"),
                    ADJUSTED.Model = limma_model %>% filter(protadj == "protein adjustment"))

limma_model$ADJUSTED.Model <- limma_model$ADJUSTED.Model %>% filter(is.finite(pvalue))
limma_model$PTM.Model <- limma_model$PTM.Model %>% filter(is.finite(pvalue))

plot_df <- rbind(limma_model$ADJUSTED.Model, limma_model$PTM.Model, fill = TRUE)
plot_df$protadj <- factor(plot_df$protadj, levels = c('no adjustment','protein adjustment'))

# Volcano plot
# png("supplementary/sim_new/spike_in_limma_volcano.png", width = 750, height = 500)
plot_df %>% arrange(desc(labeling)) %>% 
  filter(abs(Log2FC) < Inf & Label %in% c("mix1-mix3", "mix1-mix4", "mix2-mix3", "mix2-mix4")) %>%
  ggplot(aes(Log2FC, -log10(adj.pvalue))) + 
  geom_point(aes(color = labeling, alpha = labeling)) + 
  scale_color_manual(values = c("#D55E00", "#999999")) + 
  scale_alpha_manual(values = c(1, 0.2)) + 
  scale_size_manual(values = c(2, 1)) + 
  geom_hline(yintercept = 0) + 
  geom_vline(xintercept = 0) + 
  geom_hline(yintercept = -log10(0.05), color = "darkred") + 
  geom_vline(xintercept = c(1, -1), color = "darkred", linetype = "dashed") + 
  facet_grid(protadj ~ contrast_truth) + 
  theme_bw() + 
  theme(axis.text.x = element_text(size = 12), 
       axis.text.y = element_text(size = 12), 
       legend.text=element_text(size=12),
       axis.title.y = element_text(size = 14),
       axis.title.x = element_text(size = 14),
       title = element_text(size = 16),
       strip.text = element_text(size = 12),
       legend.position = "bottom") + 
  labs(title = "Limma VolcanoPlot")
# dev.off()
```

The limma model shows an improvement in terms of the false positives, but loses insight into the majority of spike-in peptides. After adjustment, the spike-in peptides are much more inline with the expected log2FC, however they are mostly below the significant pvalue threshold. This is most likely due to an increase in standard error or lower degrees of freedom compared to MSstatsPTM.

#### Summary Statistics

Summary statistics for the models are given below.

```{r, echo = FALSE, fig.width = 8}

unadj_stats <- calculate_summary_stats(limma_model$PTM.Model, 'adj.pvalue', "Label")
unadj_stats$Adjustment <- "no adjustment"
adj_stats <- calculate_summary_stats(limma_model$ADJUSTED.Model, 'adj.pvalue', "Label")
adj_stats$Adjustment <- "protein adjustment"
limma_stats <- rbindlist(list(unadj_stats, adj_stats))

limma_stats %>% ggplot() + geom_col(aes(x = metric, y = result, fill = Adjustment), position = "dodge") + 
  scale_fill_manual(values = c("salmon1", "steelblue1")) + 
  labs(title = "Limma - Summary Statistics", x = "Metric", y = "Value")

```

The summary statistics repeat what we saw in the volcano plots. The accuracy and specificity are both much higher after adjustment due to the decrease in false positives. Of course this is also seen in the lower FPR. Recall after adjustment is much lower. We are correctly labeling very few of the true positives after adjustment. This is in contrast to MSstatsPTM which actually saw an increase in recall after adjustment.

#### Log Fold Change - Spike-in Peptides

Again the last thing we will look at is the distribution spike-in peptides log fold changes. The expected log Fold Change is indicated by the red `X` marks.

```{r, echo = FALSE, fig.width = 10, fig.height = 5}
unadj_spike <- limma_model$PTM.Model %>% filter(labeling == "heavy")
unadj_spike$Adjustment <- "no adjustment"
adj_spike <- limma_model$ADJUSTED.Model %>% filter(labeling == "heavy")
adj_spike$Adjustment <- "protein adjustment"
spike <- rbindlist(list(unadj_spike, adj_spike), fill = TRUE)

spike %>% ggplot() + geom_boxplot(aes(x = contrast_truth, y = Log2FC), fill = "lightsteelblue4") + 
  geom_point(aes(y = trueLog2FC, x = contrast_truth), color = "red", shape = 4, size = 3) + facet_wrap(.~Adjustment) + 
  labs(title = "Fold Change of Spike-in Peptides", x = "Comparison with Expected FC") + theme(axis.text.x = element_text(angle = 30, vjust = .5))

```

Once again we see a strong improvement in the Log2FC of spike-in peptides after adjustment. In contrast to MSstatsPTM the distribution after adjustment is much wider, with a median close to the true FC. This larger variance around the median indicates that while we are generally seeing the true FC, the results are not as strong as MSstatsPTM.

### T-Test

Finally we will look at the t-test results.

#### Volcano Plot

Again we look at the volcano plots. The spike-in peptides are colored red and their expected log Fold Change is indicated in the subplot title. The background peptides are colored grey and should all be insignificant.

```{r, echo=FALSE, fig.width = 10, fig.height = 7}
load(file = "model_data/Spike_in_ttest_model.rda")#ttest_final_model

# Volcano plot
png("supplementary/sim_new/spike_in_ttest_volcano.png", width = 750, height = 500)
ttest_final_model %>% arrange(desc(labeling)) %>% 
  filter(abs(estimate) < Inf & label %in% c("mix1-mix3", "mix1-mix4", "mix2-mix3", "mix2-mix4")) %>%
  ggplot(aes(estimate, -log10(adj.pvalue))) + 
  geom_point(aes(color = labeling, alpha = labeling)) + 
  scale_color_manual(values = c("#D55E00", "#999999")) + 
  scale_alpha_manual(values = c(1, 0.1)) + 
  scale_size_manual(values = c(2, 1)) + 
  geom_hline(yintercept = 0) + 
  geom_vline(xintercept = 0) + 
  geom_hline(yintercept = -log10(0.05), color = "darkred") + 
  geom_vline(xintercept = c(1, -1), color = "darkred", linetype = "dashed") + 
  facet_grid(protadj ~ contrast_truth) + 
  theme_bw() + 
  theme(axis.text.x = element_text(size = 12), 
       axis.text.y = element_text(size = 12), 
       legend.text=element_text(size=12),
       axis.title.y = element_text(size = 14),
       axis.title.x = element_text(size = 14),
       title = element_text(size = 16),
       strip.text = element_text(size = 12),
       legend.position = "bottom") + 
  labs(title = "T-test VolcanoPlot")
dev.off()
```

The results of the t-test are the weakest of the models we have observed. Both before and after adjustment we see no significant results. This may be due to an increased variance in the models, as each t-test model must be fit between two conditions and cannot share information. Additionally we only have two observations per condition, thus if one observation is missing the t-test model breaks down and the model cannot be fit.

To get a closer look at the effect of the adjustment we will look at the volcano plots using the unadjusted pvalue.

``` {r, echo=FALSE, fig.width = 10, fig.height = 7}
# Volcano plot
ttest_final_model %>% arrange(desc(labeling)) %>% 
  filter(abs(estimate) < Inf & label %in% c("mix1-mix3", "mix1-mix4", "mix2-mix3", "mix2-mix4")) %>%
  ggplot(aes(estimate, -log10(pval))) + 
  geom_point(aes(color = labeling, alpha = labeling)) + 
  scale_color_manual(values = c("#D55E00", "#999999")) + 
  scale_alpha_manual(values = c(1, 0.1)) + 
  scale_size_manual(values = c(2, 1)) + 
  geom_hline(yintercept = 0) + 
  geom_vline(xintercept = 0) + 
  geom_hline(yintercept = -log10(0.05), color = "darkred") + 
  geom_vline(xintercept = c(1, -1), color = "darkred", linetype = "dashed") + 
  facet_grid(protadj ~ contrast_truth) + 
  theme_bw() + 
  theme(legend.position = "bottom") + 
  labs(title = "T-test VolcanoPlot: Unadjusted Pvalue")
```

The unadjusted pvalue volcano plots look very similar to what we have seen in the previous two models. The false positives decrease and the spike-in peptides follow the expected FC closer after adjustment.

#### Summary Statistics

The summary statistics for this model are skipped because the adjusted pvalue shows no positive results.

#### Log Fold Change - Spike-in Peptides

Lastly, we will again look at the distribution of the spike-in peptides log fold changes. The expected log Fold Change is indicated by the red `X` marks.

```{r, echo = FALSE, fig.width = 10, fig.height = 5}
unadj_spike <- ttest_final_model %>% filter(protadj == "no adjustment" & labeling == "heavy")
unadj_spike$Adjustment <- "no adjustment"
adj_spike <- ttest_final_model %>% filter(protadj == "protein adjustment" & labeling == "heavy")
adj_spike$Adjustment <- "protein adjustment"
spike <- rbindlist(list(unadj_spike, adj_spike), fill = TRUE)

spike %>% ggplot() + geom_boxplot(aes(x = contrast_truth, y = estimate), fill = "lightsteelblue4") + 
  geom_point(aes(y = trueLog2FC, x = contrast_truth), color = "red", shape = 4, size = 3) + facet_wrap(.~Adjustment) + 
  labs(title = "T-test Fold Change of Spike-in Peptides", x = "Comparison with Expected FC", y = "log2FC") + theme(axis.text.x = element_text(angle = 30, vjust = .5))

```

This plots is very similar to the limma plot. The spike-in peptides follow the expected FC much better after adjustment. However the effect is much wider around the true FC than MSstatsPTM, meaning the effect is less accurate.

### Summary

From the analysis above it is clear MSstatsPTM is the most robust modeling method of the three tested. In the computer simulations we saw MSstatsPTM and Limma having similar results, however in the real world experiment Limma performanced significantly worse. This could be in part due to the robust TMP summary methods used by MSstatsPTM compared the log sum of abundance used for Limma. An extension to this analysis could look into the performance of Limma using the same summarization methods as used in MSstatsPTM. In either case it is clear that the full workflow of MSstatsPTM performs the best of the observed models.

In the following plot we compare the summary statsitics of the adjusted models. Note T-test is shown but all values are zero.
```{r, echo = FALSE, fig.width = 8}

limma_sum <- limma_stats %>% filter(Adjustment == "protein adjustment") 
limma_sum$Model <- "Limma"
msstatsptm_sum <- msstatsptm_stats  %>% filter(Adjustment == "protein adjustment") 
msstatsptm_sum$Model <- "MSstatsPTM"

ttest_sum <- calculate_summary_stats(ttest_final_model %>% filter(protadj == "protein adjustment"), 'adj.pvalue', "label")

# ttest_sum <- data.frame(metric = c('FPR', 'Recall','Specificity','Precision','Accuracy'),
#            result = c(0,0,0,0,0))
ttest_sum$Adjustment <- "protein adjustment"
ttest_sum$Model <- "T-test"

summary <- rbindlist(list(msstatsptm_sum, limma_sum, ttest_sum))
summary[Model == "MSstatsPTM"]$Model <- "Proposed"

png("supplementary/sim_new/spike_in_statistic_comparison.png", width = 750, height = 500)
summary %>% mutate(Model = factor(Model, levels=c("Proposed", "Limma", "T-test"))) %>% 
  ggplot() + geom_col(aes(x = metric, y = result, fill = Model), position = "dodge") + 
  #scale_fill_manual(values = c("salmon1", "steelblue1", "green")) + 
  labs(title = "Summary Statistic Comparison", x = "Metric", y = "Result") + theme_bw()
dev.off()
```

This plot confirms that MSstatsPTM and limma are comparable in terms of FDR, however MSstatsPTM performs much better when identifying the true positives, spike-in, peptides.

In the following plot we compare the log2FC of the spike-in peptides across models.

```{r, echo = FALSE, fig.width = 8}

unadj_spike_msstats <- model_df$PTM.Model %>% filter(labeling == "heavy")
unadj_spike_msstats$Adjustment <- "no adjustment"
unadj_spike_msstats$Model <- "MSstatsPTM"
adj_spike_msstats <- model_df$ADJUSTED.Model %>% filter(labeling == "heavy")
adj_spike_msstats$Adjustment <- "protein adjustment"
adj_spike_msstats$Model <- "MSstatsPTM"

unadj_spike_ttest <- ttest_final_model %>% filter(labeling == "heavy" & protadj == "no adjustment") %>% rename(log2FC = estimate)
unadj_spike_ttest$Adjustment <- "no adjustment"
unadj_spike_ttest$Model <- "T-test"
adj_spike_ttest <- ttest_final_model %>% filter(labeling == "heavy" & protadj != "no adjustment") %>% rename(log2FC = estimate)
adj_spike_ttest$Adjustment <- "protein adjustment"
adj_spike_ttest$Model <- "T-test"

unadj_spike_limma <- limma_model$PTM.Model %>% filter(labeling == "heavy") %>% rename(log2FC = Log2FC)
unadj_spike_limma$Adjustment <- "no adjustment"
unadj_spike_limma$Model <- "Limma"
adj_spike_limma <- limma_model$ADJUSTED.Model %>% filter(labeling == "heavy") %>% rename(log2FC = Log2FC)
adj_spike_limma$Adjustment <- "protein adjustment"
adj_spike_limma$Model <- "Limma"

spike <- rbindlist(list(unadj_spike_msstats, adj_spike_msstats, unadj_spike_ttest,
                        adj_spike_ttest, unadj_spike_limma, adj_spike_limma), fill = TRUE)
spike[Model == "MSstatsPTM"]$Model <- "Proposed"

png("supplementary/sim_new/spike_in_fc.png", width = 750, height = 500)
spike %>% mutate(Comparison = str_split(contrast_truth, " ", simplify = TRUE)[,1],
                 Model = factor(Model, levels=c("Proposed", "Limma", "T-test"))) %>%  ggplot() + 
  geom_boxplot(aes(x = Comparison, y = log2FC, fill = Model)) + 
  geom_point(aes(y = trueLog2FC, x = Comparison), color = "red", shape = 4, size = 4, stroke = 2) + 
  scale_fill_manual(values = c("salmon1", "steelblue1", "green")) + facet_wrap(.~Adjustment) + 
    theme_bw() + 
  theme(axis.text.x = element_text(size = 11, angle = 0, vjust = 0), 
       axis.text.y = element_text(size = 11), 
       legend.text=element_text(size=12),
       axis.title.y = element_text(size = 14),
       axis.title.x = element_text(size = 14),
       title = element_text(size = 16),
       strip.text = element_text(size = 12),
       legend.position = "bottom") + 
  labs(title = "Fold Change of Spike-in Peptides", x = "Comparison with Expected FC")
dev.off()
```

In this plot we can see that the unadjusted models are very similar and do not follow the expected FC. In comparison the adjusted models all generally follow the expectation, however MSstatsPTM has a tighter distribution around the expected FC. In contrast the other two models have a wider range around the true FC.

### 2021-09-09 supplementary/ptm_sm.tex
Added analysis code and experiment papers

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection{TMT experiment}
+\subsection{TMT experiment}
 \label{sec:tmtmethod}
@@
 \centering
-\includegraphics[width=.85\textwidth]{sim_new/sim1_FDR_all_models}\\
-\includegraphics[width=.85\textwidth]{sim_new/sim1_FDR}
-\caption{(upper plot)All the considered methods in simulation 1 correctly calibrated the Type I error rate when adjusting for changes in protein abundance.
-In comparison, the methods without accounting for the protein-level changes resulted in off-target, high false positive rates.
-(lower plot) The considered methods with protein adjustment are compared in detail.
+ \begin{subfigure}{.9\textwidth}
+	\includegraphics[width=.9\textwidth]{sim_new/sim1_FDR_all_models}\\
+	\caption{All the considered methods in simulation 1 correctly calibrated the Type I error rate when adjusting for changes in protein abundance.
+In comparison, the methods without accounting for the protein-level changes resulted in off-target, high false positive rates.}
+ \end{subfigure}
+ \begin{subfigure}{.9\textwidth}
+	\includegraphics[width=.9\textwidth]{sim_new/sim1_FDR}
+	\caption{The considered methods with protein adjustment are compared in detail.
 The proposed method generally performed better than Limma, but had a higher FDR than T-test.
-All methods performed strongly, with all FDR rates near or below .03.
-\label{fig:sim1_fpr}}
+All methods performed strongly, with all FDR rates near or below .03.}
+ \end{subfigure}
+\label{fig:sim1_fpr}
 \end{figure}
@@
 \centering
-\includegraphics[width=.85\textwidth]{sim_new/sim1_Recall}\\
-\includegraphics[width=.85\textwidth]{sim_new/sim1_accuracy}
-\caption{(upper plot) The recall for simulation 1 methods with adjustment were compared.
+ \begin{subfigure}{.9\textwidth}
+	\includegraphics[width=.9\textwidth]{sim_new/sim1_Recall}\\
+	\caption{The recall for simulation 1 methods with adjustment were compared.
 It is clear that Limma performs the strongest here when the number of replicates were low.
 At higher replicates the performance of the proposed methods and Limma are comparable.
-T-test clearly performs worse across all methods.
-(lower plot) The overall accuracy plot mimics the observations in the recall plot.
-Limma performs stronger than the proposed method at lower replicates, while at higher they are comparable.
-\label{fig:sim1_recall}}
+T-test clearly performs worse across all methods.}
+ \end{subfigure}
+ \begin{subfigure}{.9\textwidth}
+	\includegraphics[width=.9\textwidth]{sim_new/sim1_accuracy}
+	\caption{The overall accuracy plot mimics the observations in the recall plot. Limma performs stronger than the proposed method at lower replicates, while at higher they are comparable.}
+ \end{subfigure}
+\label{fig:sim1_recall}
 \end{figure}
@@
 \centering
-\includegraphics[width=.85\textwidth]{sim_new/sim3_Recall}
-\includegraphics[width=.85\textwidth]{sim_new/sim3_Accuracy}
-\caption{The advantage of using the proposed approach over Limma and $t$-test is apparent when looking at simulation 2, which includes limited observations and the presence of missing values.
-(upper plot) In the case of recall the proposed method performs stronger than limma and t-test in nearly every model.
+ \begin{subfigure}{.9\textwidth}
+	\includegraphics[width=.9\textwidth]{sim_new/sim3_Recall}
+	\caption{The advantage of using the proposed approach is apparent when looking at simulation 2, which includes limited observations and the presence of missing values.
+In the case of recall the proposed method performs stronger than limma and t-test in nearly every model.
 Even at lower replicates the proposed method still outperformed limma, especially when there were more conditions in the simulation.
-Again the lowest performance was observed using the $t$-test method.
-(lower plot) Again in the overall accuracy plots, the results are similar to recall.
-The propsed method performed the strongest across all methods, even when replicates were low.
-While Limma shows strong performance in a clean experiment, when real world data problems are introduced it is clear the proposed method is more robust.
-\label{fig:sim2_recall}}
+Again the lowest performance was observed using the $t$-test method.}
+ \end{subfigure}
+ \begin{subfigure}{.9\textwidth}
+	\includegraphics[width=.9\textwidth]{sim_new/sim3_Accuracy}
+	\caption{The overall accuracy plots show similar results to recall.
+The proposed method performed the strongest across all methods, even when replicates were low.
+Limma shows strong performance in a clean experiment, when real world data problems are introduced it is clear the proposed method is more robust.}
+ \end{subfigure}
+\label{fig:sim2_recall}
 \end{figure}
@@
 \centering
-\includegraphics[width=.85\textwidth]{sim_new/sim1_FC_boxplot}
+ \begin{subfigure}{.9\textwidth}
+	\includegraphics[width=.9\textwidth]{sim_new/sim1_FC_boxplot}
+	\caption{In simulation 1 all considered methods correctly estimated the fold change between conditions, with a median fold change estimation of .75 in all methods.
+The distributions around the median were also consistent acorss all methods.
+Predictably the distribution quantiles decrease as the number of replicates increases.}
+ \end{subfigure}
+ \begin{subfigure}{.9\textwidth}
 	\includegraphics[width=.85\textwidth]{sim_new/sim3_FC_boxplot}
-\caption{(upper plot) In simulation 1 all considered methods correctly estimated the fold change between conditions, with a median fold change estimation of .75 in all methods.
-The distributions around the median were also consistent acorss all methods.
-Predictably the distribution quantiles decrease as the number of replicates increases.
-(lower plot) In simulation 2 again all methods correctly estimated the fold change with a median log change of .75.
+	\caption{In simulation 2 again all methods correctly estimated the fold change with a median log change of .75.
 The proposed method in this simulation had a visibly tighter distribution around the median.
 Both Limma and $t$-test showed a wider range around the fold change.
-In this case the proposed method showed a stronger performed in correctly estimating the fold change for all peptides.
-\label{fig:fc_boxplot}}
+In this case the proposed method showed a stronger performed in correctly estimating the fold change for all peptides.}
+ \end{subfigure}
+\label{fig:fc_boxplot}
 \end{figure}
@@
 \end{figure}
-
-\clearpage
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{TMT experiment}
-\label{sec:tmtsim}
-
-\todo{add the simulation for TMT experiment, batch vs plex?}
 
@@
 
+\clearpage
+
+\subsection{Shigella flexneri}
+
+In this study, the correlation between the autophagy gene Atg16L1 and killing of virulent Shigella flexneri (S.flexneri) was assessed.
+It was discovered that the loss of Atg16L1 in macrophages enhanced the killing of S.flexneri.
+Multiplex proteomics was used to quantify the abundance of total protein, phosphorylation, and ubiquitination in wild type (WT) and ATG16L1-deficient (cKO) bone marrow macrophages, uninfected and uninfected with S.flexneri.
+The abundance of total protein and post-translation modifications were quantified at three time points, uninfected, early infections, and late infection.
+Quantifying the total protein along with the post-translational modifications allows us to adjust for changes in total protein and see the true impact of the site specific modifications.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2021-09-09 code/MSstatsPTM2_spike.R

@@ -1,0 +1,883 @@
+
+library(data.table)
+library(MSstatsPTM)
+library(tidyverse)
+
+setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
+
+# Prepare data set --------------------------------------------------------
+
+# Design for global profiling data
+design_prot1 <- read.table("000547_ptm_benchmark/ptm_tft_global_ip1.txt",
+  header = TRUE, sep = "\t")
+design_prot1 <- as.data.table(design_prot1)
+design_prot2 <- read.table("000547_ptm_benchmark/ptm_tft_global_ip2.txt",
+  header = TRUE, sep = "\t")
+design_prot2 <- as.data.table(design_prot2)
+design_prot2[,id_injectionset := 2]
+design_prot <- rbindlist(list(design_prot1, design_prot2))
+
+# Design for KGG enriched data 
+design_kgg1 <- read.table("000547_ptm_benchmark/ptm_tft_kgg_ip1.txt",
+  header = TRUE, sep = "\t")
+design_kgg1 <- as.data.table(design_kgg1)
+design_kgg1[, batch := "BCH1"]
+design_kgg2 <- read.table("000547_ptm_benchmark/ptm_tft_kgg_ip2.txt",
+  header = TRUE, sep = "\t")
+design_kgg2 <- as.data.table(design_kgg2)
+design_kgg2[, batch := "BCH2"]
+
+design_kgg <- rbindlist(list(design_kgg1, design_kgg2))
+
+# Prepare data set with columns of interest for later use
+fn_kgg <- paste0("vistaga_", design_kgg$run_id, "_kgg_with_unmodified_peptides.tsv")
+fn_prot <- paste0("vistaga_", design_prot$run_id, ".tsv")
+
+sub_cols <- c(
+  "Reference", "run_id", "peptide_id", "vista_peak_area_light", 
+  "vista_confidence_score", "peptide_m_z", "peptide_sequence", 
+  "peptide_trypticity", "peptide_miscleavages", "peptide_validity",
+  "label1", "ms2_rt", "ms2_charge", "is_xq"
+)
+
+dta_kgg <- vector("list", length(fn_kgg))
+for (i in seq_along(fn_kgg)) {
+  path_file <- paste0("000547_ptm_benchmark/", fn_kgg[i])
+  dta_kgg[[i]] <- read.table(
[…]

### 2021-09-09 code/Simulated_Analysis.R

@@ -1,0 +1,569 @@
+
+library(MSstatsPTMold)
+library(MSstatsPTM)
+library(data.table)
+library(tidyverse)
+library(limma)
+
+## Define Modeling Functions ---------------------------------------------------
+convert_to_msstats_format <- function(df){
+  ## Format into MSstatsPTM input
+  df$PTM$protein <- paste(df$PTM$protein, df$PTM$site, sep = "_")
+  df$PTM$PrecursorCharge <- NA
+  df$PTM$FragmentIon <- NA
+  df$PTM$ProductCharge <- NA
+  df$PTM$IsotopeLabelType <- "L"
+  df$PTM$BioReplicate <- paste0(df$PTM$run, "rep")
+  df$PTM$log2inty <- 2**df$PTM$log2inty
+  df$PTM$PeptideSequence <- paste(df$PTM$protein, df$PTM$feature, sep = "_")
+  
+  setnames(df$PTM, c("protein", "group", "run", "log2inty"),
+           c("ProteinName", "Condition", "Run", "Intensity"))
+  
+  df$PROTEIN$PrecursorCharge <- NA
+  df$PROTEIN$FragmentIon <- NA
+  df$PROTEIN$ProductCharge <- NA
+  df$PROTEIN$IsotopeLabelType <- "L"
+  df$PROTEIN$log2inty <- 2**df$PROTEIN$log2inty
+  df$PROTEIN$BioReplicate <- paste0(df$PROTEIN$run, "rep")
+  df$PROTEIN$PeptideSequence <- paste(df$PROTEIN$protein, df$PROTEIN$feature, sep = "_")
+  setnames(df$PROTEIN, c("protein", "group", "run", "log2inty"),
+           c("ProteinName", "Condition", "Run", "Intensity"))
+  
+  return(df)
+}
+
+## Run ttest using run level summared data
+run_ttest <- function(summarized_data){
+  
+  ptms <- unique(summarized_data$PTM)
+  
+  ttest_temp <- data.table()
+  ttest_adj_temp <- data.table()
+  ## Run ttest for each ptm
+  for (p in seq_along(ptms)){
+    temp_joined <- summarized_data %>% filter(PTM == ptms[[p]])
+    groups <- (temp_joined %>% distinct(Condition))[[1]]
+    t <- 2
+    
+    ## Loop over groups
+    for (g in 1:(length(groups)-1)){
+      for (g2 in 2:length(groups)){
[…]

### 2021-09-09 code/USP30_Method_Comparison.R

@@ -1,0 +1,283 @@
+
+library(tidyverse)
+library(MSstatsPTM)
+
+setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
+
+load("dta_usp30.RData")
+head(dta_usp30)
+
+processed_df <- readRDS("usp30-site.RDS")
+head(processed_df)
+
+processed_df$FragmentIon <- NA
+processed_df$ProductCharge <- NA
+processed_df$IsotopeLabelType <- 'L'
+processed_df <- processed_df %>% mutate(PrecursorCharge = stringr::str_split(processed_df$feature, "_", simplify = TRUE)[,2])
+
+colnames(processed_df) <- c('ProteinName', 'PeptideSequence', 'is_mod', 'Feature', 
+                            'Condition', "Batch", "Run", 'Intensity', 'site', 'FragmentIon',
+                       'ProductCharge', 'IsotopeLabelType', 'PrecursorCharge')
+
+#processed_df <- processed_df %>% mutate(BioReplicate = stringr::str_split(processed_df$Run, "-", simplify = TRUE)[,2])
+processed_df$BioReplicate <- processed_df$Batch
+processed_df$Intensity <- 2^processed_df$Intensity
+
+## Combine proteins and sites so that we can estimate PTMs (like MSstatsTMTPTM)
+PTM_df <- processed_df %>% filter(is_mod == TRUE)
+PTM_df$ProteinName <- paste(PTM_df$ProteinName, PTM_df$site, sep = '_')
+protein_df <- processed_df %>% filter(is_mod == FALSE)
+
+# mean((PTM_df %>% group_by(ProteinName) %>% summarize(num_feat = n_distinct(Feature)))$num_feat)
+# mean((protein_df %>% group_by(ProteinName) %>% summarize(num_feat = n_distinct(Feature)))$num_feat)
+
+drops <- c("is_mod", "site", "Batch", "Feature")
+PTM_df <- PTM_df[ , !(names(PTM_df) %in% drops)]
+protein_df <- protein_df[ , !(names(protein_df) %in% drops)]
+
+processed_df <- list('PTM' = PTM_df, 'PROTEIN' = protein_df)
+
+## Fit MSstatsPTM Model --------------------------------------------------------
+summarized_ptm <- dataSummarizationPTM(processed_df, min_feature_count.PTM = 1)
+
+model_ptm <- groupComparisonPTM(summarized_ptm, data.type = "LabelFree")
+
[…]

### 2021-09-09 supplementary/ptm_sm.tex
Updated sections and graphs of supplementary

@@
 Multiplex proteomics was used to quantify the abundance of total protein, phosphorylation, and ubiquitination in wild type (WT) and ATG16L1-deficient (cKO) bone marrow macrophages, uninfected and uninfected with S.flexneri.
-The abundance of total protein and post-translation modifications were quantified at three time points, uninfected, early infections, and late infection.
+The abundance of total protein and post-translation modifications were quantified at three time points, uninfected, early infection (45-60 minutes), and late infection (3-3.5 hours).
 Quantifying the total protein along with the post-translational modifications allows us to adjust for changes in total protein and see the true impact of the site specific modifications.

### 2021-09-16 code/MSstatsPTM_Simulation_Analysis.Rmd
Added Shigella and USP30 details to supplmentary

@@
   fpr <- df %>% filter(get(pval_col_name) < .05 & grepl("NoChange", get(protein_col_name))) %>% 
-    nrow() / nrow(df %>% filter(grepl("NoChange", get(protein_col_name))))
+    nrow() / nrow(df %>% filter(get(pval_col_name) < .05))
   sensitivity <- df %>% filter(get(pval_col_name) < .05 & !grepl("NoChange", get(protein_col_name))) %>% 
@@
 
-load(file = "model_data/ptm_models_sim1.rda")#ptm_models_sim1
-load(file = "model_data/adjusted_models_sim1.rda")#adjusted_models_sim1 
-load(file = "model_data/ttest_models_sim1.rda")#ttest_sim1
-load(file = "model_data/adj_ttest_models_sim1.rda")#adj_ttest_sim1
-load(file = "model_data/limma_models_sim1.rda")#limma_results_sim1
-load(file = "model_data/adj_limma_models_sim1.rda")#adj_limma_sim1
+load(file = "../data/ptm_models_sim1.rda")#ptm_models_sim1
+load(file = "../data/adjusted_models_sim1.rda")#adjusted_models_sim1 
+load(file = "../data/ttest_models_sim1.rda")#ttest_sim1
+load(file = "../data/adj_ttest_models_sim1.rda")#adj_ttest_sim1
+load(file = "../data/limma_models_sim1.rda")#limma_results_sim1
+load(file = "../data/adj_limma_models_sim1.rda")#adj_limma_sim1
 
@@
 
+cbPalette <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#999999")
+cbPalette_first <- c("#E69F00", "#F0E442", "#56B4E9", "#0072B2", "#009E73", "#D55E00", "#CC79A7", "#999999")
 ```
@@
 ```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width = 7, fig.height = 6}
-png("supplementary/sim_new/sim1_FDR_all_models.png", width = 750, height = 500)
+png("../supplementary/sim_new/sim1_FDR_all_models.png", width = 750, height = 500)
 
@@
   ggplot() + geom_point(aes(x = Model, y = fpr, size = Replicates, color = Model)) + 
+  scale_colour_manual(values=cbPalette_first) +
   facet_grid(vars(sd), vars(conditions)) + 
@@
 ``` {r, echo = FALSE, message=FALSE, warning=FALSE}
-png("supplementary/sim_new/sim1_FDR.png", width = 750, height = 500)
+png("../supplementary/sim_new/sim1_FDR.png", width = 750, height = 500)
 
@@
   ggplot() + geom_point(aes(x = Model, y = fpr, size = Replicates, color = Model)) + 
+  scale_colour_manual(values=cbPalette) +
   facet_grid(vars(sd), vars(conditions)) + 
@@
 ```{r echo=FALSE, message=FALSE, warning=FALSE}
-png("supplementary/sim_new/sim1_Recall.png", width = 750, height = 500)
+png("../supplementary/sim_new/sim1_Recall.png", width = 750, height = 500)
 models %>% mutate(Model = factor(Model, levels=c("Proposed", "Proposed no Adjustment", "Limma", 
@@
   ggplot() + geom_point(aes(x = Model, y = recall, size = Replicates, color = Model)) + 
+  scale_colour_manual(values=cbPalette) +
   facet_grid(vars(sd), vars(conditions)) + 
@@
 ```{r echo=FALSE, message=FALSE, warning=FALSE}
-png("supplementary/sim_new/sim1_accuracy.png", width = 750, height = 500)
+png("../supplementary/sim_new/sim1_accuracy.png", width = 750, height = 500)
 models %>% mutate(Model = factor(Model, levels=c("Proposed", "Proposed no Adjustment", "Limma", 
@@
   ggplot() + geom_point(aes(x = Model, y = accuracy, size = Replicates, color = Model)) + 
+  scale_colour_manual(values=cbPalette) +
   facet_grid(vars(sd), vars(conditions)) + 
@@
   temp1$log2FC <- temp1$log2FC*-1
-  temp2 <- adj_ttest_sim1[[i]] %>% filter(!grepl("NoChange", ptm) & label %in% conds) %>% rename(log2FC = estimate, Label = label)
+  temp2 <- adj_ttest_sim1[[i]] %>% filter(!grepl("NoChange", ptm) & label %in% conds)
+  temp2 <- setNames(temp2, c("ptm", "Label", "pval", "tstat", "SE", "df", "log2FC"))
   temp2$Model <- "T-test"
-  temp3 <- adj_limma_sim1[[i]] %>% filter(!grepl("NoChange", PTM) & Label %in% cond2) %>% rename(log2FC = Log2FC)
+  temp3 <- adj_limma_sim1[[i]] %>% filter(!grepl("NoChange", PTM) & Label %in% cond2) 
+  temp3 <- setNames(temp3, c('PTM', 'Label', 'log2FC', 'pvalue', 'df', 'se'))
   temp3$Model <- "Limma"
@@
 
-png("supplementary/sim_new/sim1_FC_boxplot.png", width = 750, height = 500)
+png("../supplementary/sim_new/sim1_FC_boxplot.png", width = 750, height = 500)
 boxplot_data %>% mutate(Model = factor(Model, levels=c("Proposed", "Limma", "T-test")),
@@
   geom_boxplot(aes(y = log2FC, x = reps, fill = Model)) + facet_grid(vars(sd), vars(cond)) + 
+  scale_fill_manual(values=cbPalette) +
   theme_bw() +
@@
 
-load(file = "model_data/ptm_models_sim3.rda")#ptm_models_sim1
-load(file = "model_data/adjusted_models_sim3.rda")#adjusted_models_sim1 
-load(file = "model_data/ttest_models_sim3.rda")#ttest_sim1
-load(file = "model_data/adj_ttest_models_sim3.rda")#adj_ttest_sim1
-load(file = "model_data/limma_models_sim3.rda")#limma_results_sim1
-load(file = "model_data/adj_limma_models_sim3.rda")#adj_limma_sim1
+load(file = "../data/ptm_models_sim3.rda")#ptm_models_sim1
+load(file = "../data/adjusted_models_sim3.rda")#adjusted_models_sim1 
+load(file = "../data/ttest_models_sim3.rda")#ttest_sim1
+load(file = "../data/adj_ttest_models_sim3.rda")#adj_ttest_sim1
+load(file = "../data/limma_models_sim3.rda")#limma_results_sim1
+load(file = "../data/adj_limma_models_sim3.rda")#adj_limma_sim1
 
@@
 ``` {r, echo = FALSE, message=FALSE, warning=FALSE}
-png("supplementary/sim_new/sim3_FDR.png", width = 350, height = 350)
+png("../supplementary/sim_new/sim3_FDR.png", width = 350, height = 350)
 models %>% mutate(Model = factor(Model, levels=c("Proposed", "Proposed_no_adj", "limma_no_adj", 
@@
 ```{r echo=FALSE, message=FALSE, warning=FALSE}
-png("supplementary/sim_new/sim3_Recall.png", width = 750, height = 500)
+png("../supplementary/sim_new/sim3_Recall.png", width = 750, height = 500)
 models %>% mutate(Model = factor(Model, levels=c("Proposed", "Proposed_no_adj", "limma_no_adj", 
@@
 ```{r echo=FALSE, message=FALSE, warning=FALSE}
-png("supplementary/sim_new/sim3_Accuracy.png", width = 750, height = 500)
+png("../supplementary/sim_new/sim3_Accuracy.png", width = 750, height = 500)
 models %>% mutate(Model = factor(Model, levels=c("Proposed", "Proposed_no_adj", "limma_no_adj", 
@@
   temp1$log2FC <- temp1$log2FC*-1
-  temp2 <- adj_ttest_sim3[[i]] %>% filter(!grepl("NoChange", ptm) & label %in% conds) %>% rename(log2FC = estimate, Label = label)
+  temp2 <- adj_ttest_sim3[[i]] %>% filter(!grepl("NoChange", ptm) & label %in% conds)
+  temp2 <- setNames(temp2, c("ptm", "Label", "pval", "tstat", "SE", "df", "log2FC"))
   temp2$Model <- "T-test"
-  temp3 <- adj_limma_sim3[[i]] %>% filter(!grepl("NoChange", PTM) & Label %in% cond2) %>% rename(log2FC = Log2FC)
+  temp3 <- adj_limma_sim3[[i]] %>% filter(!grepl("NoChange", PTM) & Label %in% cond2) 
+  temp3 <- setNames(temp3, c('PTM', 'Label', 'log2FC', 'pvalue', 'df', 'se'))
   temp3$Model <- "Limma"
@@
 
-png("supplementary/sim_new/sim3_FC_boxplot.png", width = 750, height = 500)
+png("../supplementary/sim_new/sim3_FC_boxplot.png", width = 750, height = 500)
 boxplot_data %>% mutate(Model = factor(Model, levels=c("Proposed", "Limma", "T-test")),
@@
 geom_boxplot(aes(y = log2FC, x = reps, fill = Model)) + facet_grid(vars(sd), vars(cond)) + 
+    scale_fill_manual(values=cbPalette) +
   theme_bw() +

### 2021-09-16 code/Spike-in_Analysis.Rmd
Added Shigella and USP30 details to supplmentary

@@
                 (get(comparison_col_name) == "mix1-mix4" & get(pval_col_name) < .05)) %>% 
-    nrow() / nrow(df %>% filter(labeling == "light" | get(comparison_col_name) == "mix1-mix4"))
+    nrow() / nrow(df %>% filter(labeling == "light"))
 
@@
 ```{r pressure, echo=FALSE, fig.width = 10, fig.height = 7}
-load(file = "model_data/Spike_in_MSstatsPTM_Model.rda")#model_df
+load(file = "../data/Spike_in_MSstatsPTM_Model.rda")#model_df
 
@@
 ```{r, echo=FALSE, fig.width = 10, fig.height = 7}
-load(file = "model_data/Spike_in_Limma_Model.rda")#limma_model
+load(file = "../data/Spike_in_Limma_Model.rda")#limma_model
 
@@
 ```{r, echo=FALSE, fig.width = 10, fig.height = 7}
-load(file = "model_data/Spike_in_ttest_model.rda")#ttest_final_model
+load(file = "../data/Spike_in_ttest_model.rda")#ttest_final_model
 
@@
 
-png("supplementary/sim_new/spike_in_statistic_comparison.png", width = 750, height = 500)
+png("../supplementary/sim_new/spike_in_statistic_comparison.png", width = 750, height = 500)
 summary %>% mutate(Model = factor(Model, levels=c("Proposed", "Limma", "T-test"))) %>% 
   ggplot() + geom_col(aes(x = metric, y = result, fill = Model), position = "dodge") + 
+    scale_fill_manual(values=cbPalette) +
   #scale_fill_manual(values = c("salmon1", "steelblue1", "green")) + 
@@
 
-unadj_spike_ttest <- ttest_final_model %>% filter(labeling == "heavy" & protadj == "no adjustment") %>% rename(log2FC = estimate)
+unadj_spike_ttest <- ttest_final_model %>% filter(labeling == "heavy" & protadj == "no adjustment")
+unadj_spike_ttest <- setNames(unadj_spike_ttest, c("ptm", "label", "pval", "tstat", "SE", 
+                                                   "df", "log2FC", "labeling", "adj.pvalue", 
+                                                   "trueLog2FC", "contrast_truth", "protadj"))
+
 unadj_spike_ttest$Adjustment <- "no adjustment"
 unadj_spike_ttest$Model <- "T-test"
-adj_spike_ttest <- ttest_final_model %>% filter(labeling == "heavy" & protadj != "no adjustment") %>% rename(log2FC = estimate)
+
+adj_spike_ttest <- ttest_final_model %>% filter(labeling == "heavy" & protadj != "no adjustment")
+adj_spike_ttest <- setNames(adj_spike_ttest, c("ptm", "label", "pval", "tstat", "SE", 
+                                                   "df", "log2FC", "labeling", "adj.pvalue", 
+                                                   "trueLog2FC", "contrast_truth", "protadj"))
 adj_spike_ttest$Adjustment <- "protein adjustment"
@@
 
-unadj_spike_limma <- limma_model$PTM.Model %>% filter(labeling == "heavy") %>% rename(log2FC = Log2FC)
+unadj_spike_limma <- limma_model$PTM.Model %>% filter(labeling == "heavy")
+unadj_spike_limma <- setNames(unadj_spike_limma, c("PTM", "Label", "log2FC", "pvalue", "df", 
+                                                   "se", "labeling", "adj.pvalue", 
+                                                   "trueLog2FC", "contrast_truth", "protadj"))
 unadj_spike_limma$Adjustment <- "no adjustment"
 unadj_spike_limma$Model <- "Limma"
-adj_spike_limma <- limma_model$ADJUSTED.Model %>% filter(labeling == "heavy") %>% rename(log2FC = Log2FC)
+adj_spike_limma <- limma_model$ADJUSTED.Model %>% filter(labeling == "heavy")
+adj_spike_limma <- setNames(adj_spike_limma, c("PTM", "Label", "log2FC", "pvalue", "df", 
+                                                   "se", "labeling", "adj.pvalue", 
+                                                   "trueLog2FC", "contrast_truth", "protadj"))
 adj_spike_limma$Adjustment <- "protein adjustment"
@@
 
-png("supplementary/sim_new/spike_in_fc.png", width = 750, height = 500)
+png("../supplementary/sim_new/spike_in_fc.png", width = 750, height = 500)
 spike %>% mutate(Comparison = str_split(contrast_truth, " ", simplify = TRUE)[,1],
@@
   geom_point(aes(y = trueLog2FC, x = Comparison), color = "red", shape = 4, size = 4, stroke = 2) + 
-  scale_fill_manual(values = c("salmon1", "steelblue1", "green")) + facet_wrap(.~Adjustment) + 
+      scale_fill_manual(values=cbPalette) +
+ facet_wrap(.~Adjustment) + 
     theme_bw() + 

### 2021-09-16 supplementary/ptm_sm.tex
Added Shigella and USP30 details to supplmentary

@@
  \begin{subfigure}{.9\textwidth}
-	\includegraphics[width=.85\textwidth]{sim_new/sim3_FC_boxplot}
+	\includegraphics[width=.9\textwidth]{sim_new/sim3_FC_boxplot}
 	\caption{In simulation 2 again all methods correctly estimated the fold change with a median log change of .75.
@@
 
-This experiment was analyzed using the proposed method, limma, and two sample $t$-test.
-Becuase this is a real biological experiment we do not know the true postives and negatives as with the spike-in and simulation evaluations.
-The results of each method can be seen in \sfigref{fig:results_table}
+This experiment looked into the relationship between USP30 and protein kinase PINK1, and their association with Parkinson's Disease.
+Ubiquitination site profiling was performed and the modified site abundance was analyzed.
+Four conditions were tested with two biological replicates per condition.
+The conditions were as follows: CCCP, USP30 overexpression (USP30\_OE), Combo, and Control.
+Label-free mass spectrometry quantification was used to quantify the abundance of modified peptides.
+A corresponding mixed effects model was fit per modification and global protein as described previously in this supplementary.
 
-\begin{figure}[!ht]
-\begin{subfigure}[b]{1\linewidth}
+In contrast to other experiments analyzed in this paper, there was no unmodified global protein profiling run performed in this experiment.
+Once identification and quantification of the Ubiquitinated profiling was performed, peptides which were unmodified were extracted and used in place of a global profiling run.
+This resulted in a significant lack of overlap between modified and unmodified peptides.
+Any modified peptide without a corresponding unmodified protein could not be adjusted.
+Of the 10,799 modified peptides identified, only 4526 had a corresponding unmodified run and could be adjusted.
+Finally, using this method also resulted in a low feature count for the unmodified protein model.
+
+An example profile plot for this experiment can be seen in \sfigref{fig:USP30_profile_plot}.
+
+\begin{figure}[h!]
 \centering
-\resizebox{10cm}{!}{
- \begin{tabular}{|c|c|c|}
-
- Model & Before Adjustment & After Adjustment \\
-\hline
- Proposed & 4502 & 1625 \\ 
- Limma & 1211 & 361\\
- $t$-test & 28 & 7
-
-\end{tabular}}
-\caption{Significant modified peptides per model}
-\label{fig:results_table1a}
-\end{subfigure}
- \vspace{1cm}
-\begin{subfigure}[b]{1\linewidth}
-\centering
-\includegraphics[width=.6\textwidth]{sim_new/usp30_ven}
-\caption{Shared significant modified peptides between models}
-\label{fig:results_table1b}
-\end{subfigure}
-\caption{A.
-The proposed method identifies far more signficant modified peptides compared to Limma and $t$-test.
-All methods identify substantially fewer peptides after adjusting for changes in global protein level.
-B.
-The proposed method and Limma share 16\% of all peptides identified as significant.
-The proposed method is the only model that identified 79\% of the peptides, while Limma identified 5\% that no other model did.
-$t$-test only identified 7 peptides as signififcant, with all but 1 being shared with the other two models.}
-\label{fig:1asdfa}
+\includegraphics[width=\textwidth]{sim_new/USP30_profile_plot}
+\caption{Comparing the global profiling of protein $P52209$ with the modification of the protein at site $K059$.
+The modification appears generally unchanged between conditions, whereas the global profiling run shows the CCCP and Combo conditions at a higher relative abundance compared to the Control and USP30\_OE.
+This indicates that the modification actually had a major effect when comparing CCCP and Combo to Control and USP30\_OE, which would have been missed without adjusting for global protein changes.}
+\label{fig:USP30_profile_plot}
 \end{figure}
-
 \clearpage
@@
 
-In this study, the correlation between the autophagy gene Atg16L1 and killing of virulent Shigella flexneri (S.flexneri) was assessed.
-It was discovered that the loss of Atg16L1 in macrophages enhanced the killing of S.flexneri.
-Multiplex proteomics was used to quantify the abundance of total protein, phosphorylation, and ubiquitination in wild type (WT) and ATG16L1-deficient (cKO) bone marrow macrophages, uninfected and uninfected with S.flexneri.
+In this study, the correlation between the gene Atg16L1 and killing of Shigella flexneri (S.flexneri) was assessed.
+Multiplex proteomics was used to quantify the abundance of total protein, phosphorylation, and ubiquitination in wild type (WT) and ATG16L1-deficient (cKO) samples, uninfected and uninfected with S.flexneri.
 The abundance of total protein and post-translation modifications were quantified at three time points, uninfected, early infection (45-60 minutes), and late infection (3-3.5 hours).
-Quantifying the total protein along with the post-translational modifications allows us to adjust for changes in total protein and see the true impact of the site specific modifications.
+Quantifying the total protein along with the post-translational modifications allowed us to adjust for changes in total protein and see the true impact of the site specific modifications.
+Two mixtures using 11-plex were ran over the six conditions.
+The six conditions were split between 11 channels leading to the experimental design being unbalanced.
+Each mixture contained two replicates per early and late WT and KO conditions.
+Mixture one contained one replicate of uninfected WT and two replicates of uninfected KO.
+Mixture two contained one replicate of uninfected KO and two uninfected WT.
+The experimental design can be seen in Table ~\ref{table:shigella_design}.
+
+\begin{table}[h!]
+\centering
+\begin{tabular}{|c | c c | c c | c|}
+\hline
+ & Mixture 1 & & Mixture 2 & & Condition \\ [0.5ex]
+ \hline\hline
+ Uninfected & 128C & & 128C & 131C & \\
+ \hline
+Early (1 Hour) & 126C & 129C & 126C & 129C & WT \\
+\hline
+Late (3 Hour) & 127C & 130C & 127C & 130C & \\
+\hline
+Uninfected & 129N & 131C & 129N & & \\
+\hline
+Early (1 Hour) & 127N & 130N & 127N & 130N & KO \\
+\hline
+Late (3 Hour) & 128N & 131N & 128N & 131N & \\
+\hline
+
+\end{tabular}
+\caption{The experimental design of the Shigella flexneri experiment}
+\label{table:shigella_design}
+\end{table}
+
+A model was fit for the total protein, phosphorylation, and ubiquitination seperately, as described previously for TMT experiments. The model formula can be seen below.
+
+$$Y_{mcb} = \mu + Mixture_m + Condition_c + Subject_{mcb} + \epsilon_{mcb}$$
+
+$$Mixture_m ~ \sim N(0, \sigma^2_M) ,\: \sum_{c=1}^C{Condition_c} = 0 ,\: Subject_{mcb} ~ \sim N(0, \sigma^2_S) ,\: \epsilon_{mcb} ~ \sim N(0, \sigma^2)$$
+
+The results of the proposed method to this experiment can be seen in \sfigref{fig:No_Diff_Shigella_PTM} and \sfigref{fig:Diff_Shigella_PTM}.
+
+\begin{figure}[h!]
+\centering
+ \begin{subfigure}{\textwidth}
+	\includegraphics[width=\textwidth]{sim_new/No_Difference_Shigella_Profile_Plot}
+	\caption{Comparing the global profiling of protein $TTP\_MOUSE|P22893$ with the modification of the protein at site $S178$.
+The individual PSM features are shown in grey, while the feature summarization is shown in red.
+When looking at the summary of the modification and global protein it is clear the difference between conditions follow the same trend.
+Specifically, there is a positive adjustment in abundance when comparing WT\_Uninfect to WT\_Late in both the modification and global profiling run.
+This indicates the movement is driven by changes in global protein.}
+ \end{subfigure}
+ \begin{subfigure}{\textwidth}
+	\includegraphics[width=\textwidth]{sim_new/No_Difference_Shigella_Volcano}
+	\caption{Volcano plots of WT\_Late vs WT\_Uninfect both before and after protein adjustment.
+The $TTP\_MOUSE|P22893\_S178$ modification is highlighted.
+Before adjustment the modification has a large fold change and significant adjusted pvalue.
+After adjustment the fold change is much smaller and the adjusted pvalue is insignificant.}
+ \end{subfigure}
+\label{fig:No_Diff_Shigella_PTM}
+\end{figure}
+
+\begin{figure}[h!]
+\centering
+ \begin{subfigure}{\textwidth}
+	\includegraphics[width=\textwidth]{sim_new/Difference_Shigella_Profile_Plot}
+	\caption{Comparing the global profiling of protein $KI67\_MOUSE|E9PVX6$ with the modification of the protein at site $T215$.
+In this case the modification and global protein trend in different directions.
+Specifically, comparing WT\_Uninfect and WT\_Early there is a slightly positive change in abundance, however in the global profiling there was a negative change.
+In this case the profile plot indicates the effect of the modification is masked by the change in global protein abundance.
+Additionally this profile plot shows the large difference in available features between modifications and global protein.}
+ \end{subfigure}
+ \begin{subfigure}{\textwidth}
+	\includegraphics[width=\textwidth]{sim_new/Difference_Shigella_Volcano}
+	\caption{Volcano plots of WT\_Early vs WT\_Uninfect both before and after protein adjustment.
+The $KI67\_MOUSE|E9PVX6$ modification is highlighted.
+Before adjustment the modification has a small positive fold change and insignificant adjusted pvalue.
+After adjustment the fold change increases and the adjusted pvalue is now significant.}
+ \end{subfigure}
+\label{fig:Diff_Shigella_PTM}
+\end{figure}
 
@@
 \clearpage
+
+\todo{move to main manuscript}
 

### 2021-09-16 code/USP30_MSstatsPTM_Analysis.R

@@ -1,0 +1,148 @@
+
+library(tidyverse)
+library(MSstatsPTM)
+
+setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
+
+load(file = "../data/ups30_msstatsptm_model.rda")#model_ptm
+load(file = "../data/ups30_msstatsptm_summary.rda")#summarized_ptm
+
+
+original_ptms <- unique(model_ptm$PTM.Model$Protein)
+adj_ptms <- unique(model_ptm$ADJUSTED.Model$Protein)
+global_prot <- unique(model_ptm$PROTEIN.Model$Protein)
+
+match_list <- c()
+for (i in seq_along(global_prot)){
+  for (y in seq_along(original_ptms))
+    if (grepl(global_prot[[i]], original_ptms[[y]]) & 
+        !global_prot[[i]] %in% match_list) {
+      match_list <- c(match_list, global_prot[[i]])
+    }
+}
+
+data.frame("Number" = c(original_ptms, adj_ptms, global_prot), Label = c("Total PTMs", "Matching PTMs", "Global Proteins")) %>% 
+  ggplot() + geom_col(aes(y = Number, x = Label))
+
+prot <- sapply(original_ptms, function(x){substring(str_extract_all(x, "^(.*?)_")[[1]], 
+                                                    1, nchar(str_extract_all(x, "^(.*?)_")[[1]]) - 1)})
+length(intersect(unique(prot), global_prot))
+
+venn.diagram(
+  x = list(prot, global_prot),
+  category.names = c("PTMs" , "Global Proteins"),
+  filename = "test.png",
+  output=FALSE
+)
+
+
+## Profile Plots
+combo <- merge(model_ptm$ADJUSTED.Model, model_ptm$PTM.Model, all.x = TRUE, by = c("Protein", "Label"))
+
+combo$abs_diff <- abs(combo$log2FC.x - combo$log2FC.y)
+pot <- (combo %>% arrange(desc(abs_diff)) %>% head(100) %>% top_n(100) %>% select(Protein))[[1]]
+
+
+for (i in seq_along(pot)){
+  
+  temp <- summarized_ptm$PTM$FeatureLevelData %>% filter(PROTEIN == pot[[i]])
+  num <- length(unique(temp$FEATURE))
+  if (num >= 2){
+    print(num)
+    print(pot[[i]])
+  }
+}
+
+
+ptm_list1 <- c("Q16186_K034")
+
+# pSTY_features$BioReplicate <- factor(pSTY_features$BioReplicate, 
[…]

### 2021-09-16 code/USP30_Method_Comparison.R

@@ -4,10 +4,10 @@ library(MSstatsPTM)
 
 setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
 
-load("dta_usp30.RData")
+load("../../PTM_Refractor/Label_Free_PTM_Data/USP30/dta_usp30.RData")
 head(dta_usp30)
 
-processed_df <- readRDS("usp30-site.RDS")
+processed_df <- readRDS("../../PTM_Refractor/Label_Free_PTM_Data/USP30/usp30-site.RDS")
 head(processed_df)
 
 processed_df$FragmentIon <- NA
@@ -38,7 +38,7 @@ protein_df <- protein_df[ , !(names(protein_df) %in% drops)]
 processed_df <- list('PTM' = PTM_df, 'PROTEIN' = protein_df)
 
 ## Fit MSstatsPTM Model --------------------------------------------------------
-summarized_ptm <- dataSummarizationPTM(processed_df, min_feature_count.PTM = 1)
+summarized_ptm <- dataSummarizationPTM(processed_df, min_feature_count.PTM = 3)
 
 model_ptm <- groupComparisonPTM(summarized_ptm, data.type = "LabelFree")
 

### 2021-09-20 supplementary/ptm_sm.tex
Updated supplementary

@@
 \paragraph{Data structure of PTM quantification experiments.}
-A set of fully-cleaved and/or partially-cleaved peptides containing a same PTM (e.g., ubiquitination) at one site are considered together.
+A set of fully-cleaved and/or partially-cleaved peptides containing a same PTM (e.g., ubiquitination/phosphorylation) at one site are considered together.
 There are $I$ conditions and $J$ mass spectrometry runs (technical replicates) per condition in the experiment.
@@
 
-\todo{If we use MSstats protein summarization + t-test, how much would they be different, methodologically and performance?}
+\subsection{Limma}
 
-\subsection{Limma}
+Limma uses linear models to test the null hypothesis that there is no difference in mean PTM abundances between conditions.
+Additionally it leverages Empirical Bayes moderation to share pooled variance information across individual modification models and moderate the individual residual variances.
+Using a linear model allows Limma to share variance information across conditions, providing a more accurate estimate.
+
+With respect to PTM analysis, feature level summarization is done in the same way as $t$-test. The log-abundace estimate for PTM in Run $j$ of Condition $i$ is given by
+
+\[
+\log \left( \sum_{k=1}^{K} 2^{y_{ijk}} \right).
+\]
+
+To adjust for unmodified peptides, the unmodified peptide is summarized in the same way and the result is combined with the PTM given by
+
+\[
+d_{ij} = \log \left( \sum_{k=1}^{K} 2^{y_{ijk}} \right) - \log \left( \sum_{l=1}^{L} 2^{y_{ijl}^{\ast}} \right).
+\]
+
+Alternatively run level summarization can be performed given by
+
+\[
+\hat{\Delta} = \frac{1}{J} d_{i+} - \frac{1}{J} d_{i'+},
+\]
 
@@
 
-\todo{not ready in MSstatsPTM v2 version.}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-The methods described in this paper can be applied to Tandem Mass Tag (TMT) experiments targeting PTMs, in addition to batch effects.
-The same hypothesis testing is performed to detect differences in adjusted log-abundance of modifications in Condition $i$ and $i'$, where the null hypothesis is that there is no difference in PTM abundance between $i$ and $i'$.
+The methods described previously can be extrapolated to Tandem Mass Tag (TMT) experiments targeting PTMs.
+The null hypothesis is that there is no difference in mean PTM abundance between Conditions $i$ and $i'$ after adjusting for changes in mean unmodified peptide abundance between Conditions $i$ and $i'$.
 \begin{align*}
@@
 The model assumes the variability of the mixture is a random effect and is not convoluted with the other variables in the model.
-The model assumes that the difference between conditions is the same between mixtures.
-For a given modified peptide and corresponding global protein a mixed effects model is fit as follows.
+The model also assumes that the difference between conditions is the same between mixtures.
+For a given modified peptide and corresponding unmodified protein a mixed effects model is fit as follows.
 
@@
 \end{align*}
-\begin{gather*}
-\text{Where }  Mixture_m \sim N(0, \sigma^2_m), Techrep(Mixture)_{jm} \sim N(0, \sigma^2_{jm}), Subject \sim N(0, \sigma^2_{s}), \\
-\epsilon_{mtcb} \sim N(0, sigma^2)
-\end{gather*}
+
+Where $Mixture_m \sim N(0, \sigma^2_m)$, $Techrep(Mixture)_{jm} \sim N(0, \sigma^2_{jm})$, $Subject \sim N(0, \sigma^2_{s})$ are random effects, and $\sum_{c=1}^{I} Condition_{c}^{\ast} = 0$ is a fixed effect.
+Finally the error is expressed as $\epsilon_{mtcb} \sim N(0, \sigma^2)$.
 
@@
 
-\todo{anything related to time-course?}
-
 \clearpage
@@
 Approaches without considering protein-level adjustment lose track of an important aspect in interpreting observed changes in PTM abundance, which may result in misleading conclusions.
-To highlight the necessity of the adjustment, we compared the following approaches: a) proposed approach, b) proposed approach without adjusting for unmodified peptides c) $t$-test (with adjustment), d) $t$-test (no adjustment), e) Limma (with adjustment), and f) Limme (no adjustment).
+To highlight the necessity of the adjustment, we compared the following approaches: a) proposed approach, b) proposed approach without adjusting for unmodified peptides c) $t$-test (with adjustment), d) $t$-test (no adjustment), e) Limma (with adjustment), and f) Limma (no adjustment).
 
@@
 To highlight this distinction, multiple conditions of data were generated.
-The simulation was based on the following parameters:
+Two simulations were generated based on the following parameters:
 
@@
 
-In the first simulation an experiment with many observations per protein/PTM and no missing data was created.
-This is not representative of a real experiment, but provides a baseline for model performance.
+In the first simulation an experiment with many features per PTM and unmodified protein was created.
+Additionally this simulation contained no missing data.
+These attributes are not representative of a real experiment, but provide a baseline for model performance.
 
@@
 
-The results are summarized from \sfigref{fig:prot_fpr} to \sfigref{fig:fc_boxplot}, including false positive rate with or without changes in protein abundance by the considered methods (\sfigref{fig:prot_fpr}), false positive rate, and overall accuracy.
+The results are summarized from \sfigref{fig:prot_fpr} to \sfigref{fig:fc_boxplot}, including false discovery rate with or without changes in protein abundance by the considered methods (\sfigref{fig:prot_fpr}), false positive rate, and overall accuracy.
 Additionally the results are compared to simulation 2 described in the next section.
 
-When simulating the data, 500 of the 1000 peptides were simulated with a differencial fold change between conditons, while the other 500 were not.
-In terms of the 500 peptides that were not differencial, 250 were simulated with a change in both PTM and overall protein abundance, such that the change in PTM abundance is entirely due to the change in the overall protein.
+When simulating the data, 500 of the 1000 peptides were simulated with a differential fold change between conditions, while the other 500 were not.
+In terms of the peptides that were not differential, 250 were simulated with a change in both PTM and overall protein abundance, such that the change in PTM abundance is entirely due to the change in the overall protein.
 The other 250 were simulated with no significant change in either PTM or overall protein abundance.
 In this way there were 500 true positives and 500 true negative PTMs.
-These were used to calculate the summary statistics seen in the following plots.
 
@@
 	\includegraphics[width=.9\textwidth]{sim_new/sim1_FDR_all_models}\\
-	\caption{All the considered methods in simulation 1 correctly calibrated the Type I error rate when adjusting for changes in protein abundance.
+	\caption{All the considered methods in simulation 1 correctly calibrated FDR rate when adjusting for changes in protein abundance.
 In comparison, the methods without accounting for the protein-level changes resulted in off-target, high false positive rates.}
@@
 	\includegraphics[width=.9\textwidth]{sim_new/sim1_FDR}
-	\caption{The considered methods with protein adjustment are compared in detail.
-The proposed method generally performed better than Limma, but had a higher FDR than T-test.
-All methods performed strongly, with all FDR rates near or below .03.}
+	\caption{The considered methods with protein adjustment are compared in detail. All three methods generally performed similarly in terms of FDR.}
  \end{subfigure}
@@
 	\includegraphics[width=.9\textwidth]{sim_new/sim1_Recall}\\
-	\caption{The recall for simulation 1 methods with adjustment were compared.
+	\caption{The recall(false positive rate) for simulation 1 methods with adjustment were compared.
 It is clear that Limma performs the strongest here when the number of replicates were low.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Sample size calculation and power analysis}
-
-The proposed approach adjusts for the underlying protein abundance in the PTM significance analysis, which corrects the confounding factor with a cost of increased variation.
-We compared the required sample size with or without the adjustment to highlight the property (\sfigref{fig:size_prot}), based on a design of two conditions, in consideration of three pairs of standard deviations of log-intensities for modified and unmodified peptides: (0.2, 0.1), (0.2, 0.2), and (0.2, 0.4).
-%We considered a design with two conditions, where standard deviations of log-intensities of both modified and unmodified peptides were both set as $0.2$. 
-We then characterized the advantage of general statistical modeling in complex designs in terms of sample size calculation (\sfigref{fig:size_grp}) and power analysis (\sfigref{fig:pwr_grp}).
-
-\begin{figure}[h!]
-\centering
-\includegraphics[width=\textwidth]{sim/size_prot}
-\caption{Protein-level adjustment relies on the inference of protein abundance, which introduces additional uncertainty in the estimate of PTM difference.
-Therefore, the required sample size to detect a systematic change is higher than as expected for standard differential analysis without adjustment.
-The discrepancy can be profound if the uncertainty associated with the protein abundance estimate is greater than that of PTM abundance estimate.
-Sample size calculation without accounting for the uncertainty would lead to under-powered studies.
-\label{fig:size_prot}}
-\end{figure}
-
-\begin{figure}[h!]
-\centering
-\includegraphics[width=\textwidth]{sim/size_grp}
-%\includegraphics[width=.725\textwidth]{sim/size_grp}
-\caption{In complex designs, simultaneously analyzing all the conditions effectively increases the degrees of freedom and requires fewer replicates. \label{fig:size_grp}}
-\end{figure}
-
-\begin{figure}[h!]
-\centering
-\includegraphics[width=\textwidth]{sim/pwr_grp}
-%\includegraphics[width=.725\textwidth]{sim/pwr_grp}
-\caption{Statistical power can be improved by increasing the sample size and analyzing multiple conditions together. \label{fig:pwr_grp}}
-\end{figure}
-
-\clearpage
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Datasets : Biological investigation}
-
-\subsection{USP30}
-
-This experiment looked into the relationship between USP30 and protein kinase PINK1, and their association with Parkinson's Disease.
-Ubiquitination site profiling was performed and the modified site abundance was analyzed.
-Four conditions were tested with two biological replicates per condition.
-The conditions were as follows: CCCP, USP30 overexpression (USP30\_OE), Combo, and Control.
-Label-free mass spectrometry quantification was used to quantify the abundance of modified peptides.
-A corresponding mixed effects model was fit per modification and global protein as described previously in this supplementary.
-
-In contrast to other experiments analyzed in this paper, there was no unmodified global protein profiling run performed in this experiment.
-Once identification and quantification of the Ubiquitinated profiling was performed, peptides which were unmodified were extracted and used in place of a global profiling run.
-This resulted in a significant lack of overlap between modified and unmodified peptides.
-Any modified peptide without a corresponding unmodified protein could not be adjusted.
-Of the 10,799 modified peptides identified, only 4526 had a corresponding unmodified run and could be adjusted.
-Finally, using this method also resulted in a low feature count for the unmodified protein model.
-
-An example profile plot for this experiment can be seen in \sfigref{fig:USP30_profile_plot}.
-
-\begin{figure}[h!]
-\centering
-\includegraphics[width=\textwidth]{sim_new/USP30_profile_plot}
-\caption{Comparing the global profiling of protein $P52209$ with the modification of the protein at site $K059$.
-The modification appears generally unchanged between conditions, whereas the global profiling run shows the CCCP and Combo conditions at a higher relative abundance compared to the Control and USP30\_OE.
-This indicates that the modification actually had a major effect when comparing CCCP and Combo to Control and USP30\_OE, which would have been missed without adjusting for global protein changes.}
-\label{fig:USP30_profile_plot}
-\end{figure}
-\clearpage
 
@@
 
+\clearpage
+
+\subsection{USP30}
+
+This experiment looked into the relationship between USP30 and protein kinase PINK1, and their association with Parkinson's Disease.
+Ubiquitination site profiling was performed and the modified site abundance was analyzed.
+Four conditions were tested with two biological replicates per condition.
+The conditions were as follows: CCCP, USP30 overexpression (USP30\_OE), Combo, and Control.
+Label-free mass spectrometry quantification was used to quantify the abundance of modified peptides.
+A corresponding mixed effects model was fit per modification and global protein as described previously in this supplementary.
+
+In contrast to other experiments analyzed in this paper, there was no unmodified global protein profiling run performed in this experiment.
+Once identification and quantification of the Ubiquitinated profiling was performed, peptides which were unmodified were extracted and used in place of a global profiling run.
+This resulted in a significant lack of overlap between modified and unmodified peptides.
+Any modified peptide without a corresponding unmodified protein could not be adjusted.
+Of the 10,799 modified peptides identified, only 4526 had a corresponding unmodified run and could be adjusted.
+Finally, using this method also resulted in a low feature count for the unmodified protein model.
+
+An example profile plot for this experiment can be seen in \sfigref{fig:USP30_profile_plot}.
+
+\begin{figure}[h!]
+\centering
+\includegraphics[width=\textwidth]{sim_new/USP30_profile_plot}
+\caption{Comparing the global profiling of protein $P52209$ with the modification of the protein at site $K059$.
+The modification appears generally unchanged between conditions, whereas the global profiling run shows the CCCP and Combo conditions at a higher relative abundance compared to the Control and USP30\_OE.
+This indicates that the modification actually had a major effect when comparing CCCP and Combo to Control and USP30\_OE, which would have been missed without adjusting for global protein changes.}
+\label{fig:USP30_profile_plot}
+\end{figure}
+\clearpage
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Sample size calculation and power analysis}
+
+The proposed approach adjusts for the underlying protein abundance in the PTM significance analysis, which corrects the confounding factor with a cost of increased variation.
+We compared the required sample size with or without the adjustment to highlight the property (\sfigref{fig:size_prot}), based on a design of two conditions, in consideration of three pairs of standard deviations of log-intensities for modified and unmodified peptides: (0.2, 0.1), (0.2, 0.2), and (0.2, 0.4).
+%We considered a design with two conditions, where standard deviations of log-intensities of both modified and unmodified peptides were both set as $0.2$. 
+We then characterized the advantage of general statistical modeling in complex designs in terms of sample size calculation (\sfigref{fig:size_grp}) and power analysis (\sfigref{fig:pwr_grp}).
+
+\begin{figure}[h!]
+\centering
+\includegraphics[width=\textwidth]{sim/size_prot}
+\caption{Protein-level adjustment relies on the inference of protein abundance, which introduces additional uncertainty in the estimate of PTM difference.
+Therefore, the required sample size to detect a systematic change is higher than as expected for standard differential analysis without adjustment.
+The discrepancy can be profound if the uncertainty associated with the protein abundance estimate is greater than that of PTM abundance estimate.
+Sample size calculation without accounting for the uncertainty would lead to under-powered studies.
+\label{fig:size_prot}}
+\end{figure}
+
+\begin{figure}[h!]
+\centering
+\includegraphics[width=\textwidth]{sim/size_grp}
+%\includegraphics[width=.725\textwidth]{sim/size_grp}
+\caption{In complex designs, simultaneously analyzing all the conditions effectively increases the degrees of freedom and requires fewer replicates. \label{fig:size_grp}}
+\end{figure}
+
+\begin{figure}[h!]
+\centering
+\includegraphics[width=\textwidth]{sim/pwr_grp}
+%\includegraphics[width=.725\textwidth]{sim/pwr_grp}
+\caption{Statistical power can be improved by increasing the sample size and analyzing multiple conditions together. \label{fig:pwr_grp}}
+\end{figure}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2021-09-20 code/USP30_MSstatsPTM_Analysis.R

@@ -1,6 +1,8 @@
 
 library(tidyverse)
 library(MSstatsPTM)
+library(data.table)
+library(gridExtra)
 
 setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
 
@@ -40,21 +42,30 @@ venn.diagram(
 combo <- merge(model_ptm$ADJUSTED.Model, model_ptm$PTM.Model, all.x = TRUE, by = c("Protein", "Label"))
 
 combo$abs_diff <- abs(combo$log2FC.x - combo$log2FC.y)
-pot <- (combo %>% arrange(desc(abs_diff)) %>% head(100) %>% top_n(100) %>% select(Protein))[[1]]
+pot <- combo %>% arrange(desc(abs_diff)) %>% head(1000) %>% top_n(1000)
 
 
-for (i in seq_along(pot)){
-  
-  temp <- summarized_ptm$PTM$FeatureLevelData %>% filter(PROTEIN == pot[[i]])
+for (i in seq_len(nrow(pot))){
+
+  temp <- summarized_ptm$PTM$FeatureLevelData %>% filter(PROTEIN == pot[i, Protein])
   num <- length(unique(temp$FEATURE))
+  
   if (num >= 2){
-    print(num)
-    print(pot[[i]])
+    temp_p <- summarized_ptm$PROTEIN$FeatureLevelData %>% 
+      filter(PROTEIN == pot[i, GlobalProtein])
+    
+    num2 <- length(unique(temp_p$FEATURE))
+    
+    if (num2 >= 2){
+      print(num)
+      print(num2)
+      print(pot[i, Protein])
+    }
   }
 }
 
 
-ptm_list1 <- c("Q16186_K034")
+ptm_list1 <- c("P52209_K059")
 
 # pSTY_features$BioReplicate <- factor(pSTY_features$BioReplicate, 
 #                                      levels = c("WT_Uninfect_1", "WT_Uninfect_2", "WT_Uninfect_3" , "WT_Uninfect_4", 
@@ -76,7 +87,7 @@ summarized$Feature <- "Summary"
 test <- rbindlist(list(features, summarized), fill = TRUE)
 
 p1 <- test %>% ggplot() + 
-  geom_line(aes(x = originalRUN, y = Abundance , group = FEATURE, color = Feature)) + #,  size = Feature)
+  geom_line(aes(x = originalRUN, y = Abundance , group = FEATURE, color = Feature,  size = Feature)) +
[…]

### 2021-09-20 code/sample_size_analysis.R

@@ -1,0 +1,50 @@
+
+
+m0_m1 <- 99
+FDR <- .05
+power <- .8
+alpha <- power * FDR / (1 + (1 - FDR) * m0_m1)
+
+z_alpha <- qnorm(1 - alpha / 2)
+z_beta <- qnorm(power)
+fc <- seq(.5, 1.5, 0.0013)
+delta <- sqrt(fc)
+aa <- (delta / (z_alpha + z_beta)) ^2
+
+#median_sigma_error + median_sigma_subject
+sigma_error_mod <- .2
+sigma_error_unmod <- .1
+
+numSample_nomod <- round(2 * (sigma_error_mod) / aa, 0)
+numSample_mod <- round(2 * (sqrt(sigma_error_mod^2 + sigma_error_unmod^2)) / aa, 0)
+
+p1 <- data.frame("type" = c(rep("With adjustment", length(numSample_mod)), 
+                      rep("Without adjustment", length(numSample_nomod))),
+           "Log2FC" = c(fc, fc),
+           "num_samples" = c(numSample_mod, numSample_nomod)) %>% 
+  ggplot() + geom_line(aes(x = Log2FC, y = num_samples, group = type, color = type))
+
+sigma_error_mod <- .2
+sigma_error_unmod <- .2
+
+numSample_nomod <- round(2 * (sigma_error_mod + .3) / aa, 0)
+numSample_mod <- round(2 * (sqrt(sigma_error_mod^2 + sigma_error_unmod^2) + .3) / aa, 0)
+
+p2 <- data.frame("type" = c(rep("With adjustment", length(numSample_mod)), 
+                      rep("Without adjustment", length(numSample_nomod))),
+           "Log2FC" = c(delta, delta),
+           "num_samples" = c(numSample_mod, numSample_nomod)) %>% 
+  ggplot() + geom_line(aes(x = Log2FC, y = num_samples, group = type, color = type))
+
+sigma_error_mod <- .2
+sigma_error_unmod <- .4
+
+numSample_nomod <- round(2 * (sigma_error_mod + .05) / aa, 0)
+numSample_mod <- round(2 * (sqrt(sigma_error_mod^2 + sigma_error_unmod^2) + .05) / aa, 0)
+
+p3 <- data.frame("type" = c(rep("With adjustment", length(numSample_mod)), 
+                      rep("Without adjustment", length(numSample_nomod))),
+           "Log2FC" = c(delta, delta),
[…]

### 2021-09-21 main/MSstatsPTM_Main.tex
Minor updates to main manuscript

@@
 
-%\author{
-%  LastName1, FirstName1\\
-%  \texttt{first1.last1@xxxxx.com}
-%  \and
-%  LastName2, FirstName2\\
-%  \texttt{first2.last2@xxxxx.com}
-%}
+\author{
+ Devon Kohler\\
+ \and
+ Tsung-Heng Tsai\\
+ \and
+ Ting Huang\\
+ \and
+ Erik Verschueren
+ \and
+ Trent Hinkle
+ \and
+  Meena Choi\\
+ \and
+ Olga Vitek
+}
 \date{}
@@
 The proposed approach was evaluated using datasets from computer simulations, benchmark controlled mixtures, and biological investigations.
-The approach was then compared against the commonly applied t-test.
+The approach was then compared against the commonly applied t-test and limma package.
 The results demonstrate that by appropriately leveraging the information from the entire dataset the proposed approach improves the reproducibility and accuracy of the estimates of PTM fold changes, results in a better calibrated type I error rate, and improves the statistical power of detecting changes in PTMs.
@@
 
-\todo{Not sure what to do with this part and figure. How would this look in a paper?}
 Figure \ref{fig:data-structure} schematically illustrates a simplified version of the data structure resulting from a typical bottom-up experiment for quantitative analysis of PTMs, in which there are multiple layers of variation present.
@@
 For more accurate estimation of protein abundance, separate global proteomics data of unenriched samples are often acquired.
+If unmodified features are unavailable for any given modified feature, unmodified intensity adjustment cannot be performed.
 As different levels of variability are present in the data, the log-intensities of the features for modified and unmodified peptides are modeled separately using two linear mixed models.
@@
 For each PTM site, the feature log-intensities of the modified peptides spanning the site are expressed using a linear mixed model in consideration of the effects of condition, run, feature and interaction between run and feature.
+In addition mixture may be included for features acquired via tandem mass tag (TMT) methods.
 The model parameters are estimated using the split-plot approach as in MSstats, where the feature log-intensities are first summarized into a single value per site per run in the subplot model, and the site-level summaries are then used for the inference of the PTM site abundance.\cite{Choi:2014} In the site-level summarization, Tukey's median polish (TMP), a simple and robust procedure is applied to iteratively fit a two-way additive model with the effects of run and feature, which in turn summarizes the log-intensities for each site.\cite{Tukey:1977} After summarization, the inference of the PTM site abundance in each condition is carried through fitting a model based on the family of linear mixed-effects models, taking into account the specific experimental design \ref{fig:data-structure}.\cite{Bolker2009} \cite{Faraway:2006} Statistical modeling and quantification for global proteomics data are performed by the same procedure as for PTM data.
@@
 The estimate of the adjusted difference and the standard error (SE) of the estimate are obtained by combining the difference estimates and the associated SEs from both counterparts.
-The standard error of the estimate $\mathrm{SE}(\hat{\Delta})$ is
-
-\[
-\left[ \frac{2}{J} \left( \hat{\sigma}_{\gamma}^{2} + \hat{\sigma}_{\gamma^{\ast}}^{2} \right) \right]^{1/2}.
-\]
 
@@
 
-The proposed statistical approach methods were evaluated using computer simulation. Specifically, their properties under adjustment with respect to protein abundance and batch effects were evaluated.
+The proposed statistical approach methods were evaluated using computer simulation.
+Specifically, their properties under adjustment with respect to protein abundance and batch effects were evaluated.
+Two simulations were ran, one including a high number of features and no missing values, and one with few modified features and including missing values.
 
@@
 The proposed approach adjusts the abundance with respect to unmodified peptides by combining the inference of modified and unmodified peptide abundances.
-Alternatively, two-sample t-test that takes as input the ratio between modified and unmodified peptide intensities (difference on log scale) is commonly applied for the same purpose.
+Alternatively, two-sample t-test or limma that takes as input the ratio between modified and unmodified peptide intensities (difference on log scale) is commonly applied for the same purpose.
 In real experiments, multiple inter-related conditions are often compared together.
-Whereas t-test uses measurements from the two conditions being compared, the proposed approach leverages measurements in all conditions for the inference of underlying abundance.
+Whereas t-test uses measurements from the two conditions being compared, the proposed approach and limma leverages measurements in all conditions for the inference of underlying abundance.
 We evaluated the impact with such adjustment by computer simulation, considering the following factors: with/without protein-level changes, effect size, number of replicates and number of conditions.
-Details of the simulation can be found in Supplementary \todo{link to Sec.
-S4.1}.
+Details of the simulation can be found in Supplementary %\cref{supp-sec:sim}.
 
-In this simulation, when there was no change in protein abundance, all the considered approaches well calibrated the Type I error rate \todo{figref Fig.
-4a}.
-However, when the modification changes were entirely due to the changes in protein abundance across conditions, analysis without accounting for the protein-level changes resulted in off-target, high false positive rates \todo{figref (Fig.
+In this simulation, when the modification changes were entirely due to the changes in protein abundance across conditions, analysis without accounting for the protein-level changes resulted in off-target, high false positive rates \todo{figref (Fig.
 4b)}.
@@
 
-\paragraph*{Computer simulation: batch effects.}
+%\paragraph*{Computer simulation: batch effects.}
 
-PTM experiments often create batches during data acquisition.
+%PTM experiments often create batches during data acquisition.
 The proposed approach can be naturally extended to handle such scenarios by modeling various forms of batch effects, as a t-test is not directly applicable to a problem with batches of data.
@@
 
-The major results are summarized in \todo{figref Fig.
+%The major results are summarized in \todo{figref Fig.
 5} with additional details provided in Supplementary \todo{link to Sec.

### 2021-09-21 code/sample_size_analysis.R
added sample size code

@@ -1,27 +1,39 @@
 
 
-m0_m1 <- 99
+## Code directly from MSstats designSampleSize function
+## Range of desired FC
+delta <- log2(seq(.5, 1.5, 0.025))
+desiredFC <- 2 ^ delta
+
+# delta <- seq(.5, 1.5, 0.025)
+# desiredFC <- delta
+
+## Define base metrics
 FDR <- .05
 power <- .8
+
+## function calculations
+m0_m1 <- 99 ## it means m0/m1=99, m0/(m0+m1)=0.99
 alpha <- power * FDR / (1 + (1 - FDR) * m0_m1)
 
 z_alpha <- qnorm(1 - alpha / 2)
 z_beta <- qnorm(power)
-fc <- seq(.5, 1.5, 0.0013)
-delta <- sqrt(fc)
 aa <- (delta / (z_alpha + z_beta)) ^2
 
-#median_sigma_error + median_sigma_subject
+#Specify sigma for ptm and unmod pep
 sigma_error_mod <- .2
 sigma_error_unmod <- .1
+
+## Original code containing sigma for subject.. can include or assume 0 for now
+## numSample = round(2 * (median_sigma_error + median_sigma_subject) / aa, 0)
 
 numSample_nomod <- round(2 * (sigma_error_mod) / aa, 0)
 numSample_mod <- round(2 * (sqrt(sigma_error_mod^2 + sigma_error_unmod^2)) / aa, 0)
 
-p1 <- data.frame("type" = c(rep("With adjustment", length(numSample_mod)), 
+p1 <- data.frame("type" = c(rep("With adjustment", length(numSample_mod)),
                       rep("Without adjustment", length(numSample_nomod))),
-           "Log2FC" = c(fc, fc),
-           "num_samples" = c(numSample_mod, numSample_nomod)) %>% 
+           "Log2FC" = c(desiredFC, desiredFC),
+           "num_samples" = c(numSample_mod, numSample_nomod)) %>%
   ggplot() + geom_line(aes(x = Log2FC, y = num_samples, group = type, color = type))
 
 sigma_error_mod <- .2
@@ -30,10 +42,10 @@ sigma_error_unmod <- .2
 numSample_nomod <- round(2 * (sigma_error_mod + .3) / aa, 0)
 numSample_mod <- round(2 * (sqrt(sigma_error_mod^2 + sigma_error_unmod^2) + .3) / aa, 0)
 
-p2 <- data.frame("type" = c(rep("With adjustment", length(numSample_mod)), 
[…]

### 2021-09-22 supplementary/ptm_sm.tex
Added sample size plots to supplementary

@@
 
-The proposed approach adjusts for the underlying protein abundance in the PTM significance analysis, which corrects the confounding factor with a cost of increased variation.
-We compared the required sample size with or without the adjustment to highlight the property (\sfigref{fig:size_prot}), based on a design of two conditions, in consideration of three pairs of standard deviations of log-intensities for modified and unmodified peptides: (0.2, 0.1), (0.2, 0.2), and (0.2, 0.4).
-%We considered a design with two conditions, where standard deviations of log-intensities of both modified and unmodified peptides were both set as $0.2$. 
-We then characterized the advantage of general statistical modeling in complex designs in terms of sample size calculation (\sfigref{fig:size_grp}) and power analysis (\sfigref{fig:pwr_grp}).
+The proposed approach allows us to calculate the sample size needed to achieve a desired statistical power as described in Section \ref{sec:design}.
+Given a desired power of .9, varying standard deviations, and an expected fold change, the number of replicates per condition can be seen in \sfigref{fig:sample_size}
 
@@
 \centering
-\includegraphics[width=\textwidth]{sim/size_prot}
-\caption{Protein-level adjustment relies on the inference of protein abundance, which introduces additional uncertainty in the estimate of PTM difference.
-Therefore, the required sample size to detect a systematic change is higher than as expected for standard differential analysis without adjustment.
-The discrepancy can be profound if the uncertainty associated with the protein abundance estimate is greater than that of PTM abundance estimate.
-Sample size calculation without accounting for the uncertainty would lead to under-powered studies.
-\label{fig:size_prot}}
+\includegraphics[width=\textwidth]{sim_new/simple_power_analysis}
+\caption{The number of replicates per condition required to reach a power of .9 can be seen in the plot.
+At a high standard deviation and low expected fold change the number of replicates is unrealistically high.
+As the expected fold change increases the number of replicates needed decreases dramatically.
+The standard deviation of .1 expectantly shows the least number of replicates needed regardless of fold change.}
+\label{fig:sample_size}
 \end{figure}
+
+As described in Section \ref{sec:test} the proposed method adjusts for the underlying protein abundance in the PTM significance analysis, which corrects the confounding factor with a cost of increased variation.
+When the variation is increased, the experiment requires a larger number of replicates to reach the same power.
+
+We compared the required sample size with varying standard deviations for the modified and unmodified peptide models.
+Three pairs of standard deviations of log-intensities for modified and unmodified peptides were used: (0.2, 0.1), (0.2, 0.2), and (0.2, 0.4).
+In the first pair the standard deviation for the unmodified peptide was less than the modified, in the second the standard deviations are the same, and in the third the standard deviation for the unmodified peptide was more than for the modified peptide.
+Again the desired power was .9 and the expected fold change was varied.
+The results can be seen in \sfigref{fig:power_sd_combo}
 
@@
 \centering
-\includegraphics[width=\textwidth]{sim/size_grp}
-%\includegraphics[width=.725\textwidth]{sim/size_grp}
-\caption{In complex designs, simultaneously analyzing all the conditions effectively increases the degrees of freedom and requires fewer replicates. \label{fig:size_grp}}
+\includegraphics[width=\textwidth]{sim_new/power_analysis_sd_combined}
+\caption{When the unmodified peptide has a low standard deviation the number of replicates is very similar to the baseline.
+As the standard deviation increases the number of replicates increases greatly.
+When the unmodified stand deviation is much higher than the modified, we can see that many more replicates are needed to reach the same power.}
+\label{fig:power_sd_combo}
 \end{figure}
 
-\begin{figure}[h!]
-\centering
-\includegraphics[width=\textwidth]{sim/pwr_grp}
+%\begin{figure}[h!]
+%\centering
+%\includegraphics[width=\textwidth]{sim/size_prot}
+%\caption{Protein-level adjustment relies on the inference of protein abundance, %which introduces additional uncertainty in the estimate of PTM difference.
+%Therefore, the required sample size to detect a systematic change is higher than as %expected for standard differential analysis without adjustment.
+The discrepancy can %be profound if the uncertainty associated with the protein abundance estimate is %greater than that of PTM abundance estimate.
+Sample size calculation without %accounting for the uncertainty would lead to under-powered studies.
+%\label{fig:size_prot}}
+%\end{figure}
+
+%\begin{figure}[h!]
+%\centering
+%\includegraphics[width=\textwidth]{sim/size_grp}
+%\includegraphics[width=.725\textwidth]{sim/size_grp}
+%\caption{In complex designs, simultaneously analyzing all the conditions% effectively increases the degrees of freedom and requires fewer replicates. %\label{fig:size_grp}}
+%\end{figure}
+
+%\begin{figure}[h!]
+%\centering
+%\includegraphics[width=\textwidth]{sim/pwr_grp}
 %\includegraphics[width=.725\textwidth]{sim/pwr_grp}
-\caption{Statistical power can be improved by increasing the sample size and analyzing multiple conditions together. \label{fig:pwr_grp}}
-\end{figure}
+%\caption{Statistical power can be improved by increasing the sample size and %analyzing multiple conditions together. \label{fig:pwr_grp}}
+%\end{figure}
 

### 2021-09-22 code/sample_size_analysis.R

@@ -2,7 +2,7 @@
 
 ## Code directly from MSstats designSampleSize function
 ## Range of desired FC
-delta <- log2(seq(.5, 1.5, 0.025))
+delta <- log2(seq(1.5, 2.5, 0.025))
 desiredFC <- 2 ^ delta
 
 # delta <- seq(.5, 1.5, 0.025)
@@ -21,42 +21,58 @@ z_beta <- qnorm(power)
 aa <- (delta / (z_alpha + z_beta)) ^2
 
 #Specify sigma for ptm and unmod pep
-sigma_error_mod <- .2
-sigma_error_unmod <- .1
+
 
 ## Original code containing sigma for subject.. can include or assume 0 for now
 ## numSample = round(2 * (median_sigma_error + median_sigma_subject) / aa, 0)
 
-numSample_nomod <- round(2 * (sigma_error_mod) / aa, 0)
-numSample_mod <- round(2 * (sqrt(sigma_error_mod^2 + sigma_error_unmod^2)) / aa, 0)
+cbPalette <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#999999")
 
-p1 <- data.frame("type" = c(rep("With adjustment", length(numSample_mod)),
-                      rep("Without adjustment", length(numSample_nomod))),
-           "Log2FC" = c(desiredFC, desiredFC),
-           "num_samples" = c(numSample_mod, numSample_nomod)) %>%
-  ggplot() + geom_line(aes(x = Log2FC, y = num_samples, group = type, color = type))
+numSample_nomod_1 <- round(2 * (.1) / aa, 0)
+numSample_nomod_2 <- round(2 * (.2) / aa, 0)
+numSample_nomod_3 <- round(2 * (.4) / aa, 0)
 
-sigma_error_mod <- .2
-sigma_error_unmod <- .2
+p1 <- data.frame("SD" = c(rep(".1", length(numSample_nomod_1)),
+                      rep(".2", length(numSample_nomod_1)),
+                      rep(".4", length(numSample_nomod_1))),
+           "Log2FC" = c(desiredFC, desiredFC, desiredFC),
+           "num_samples" = c(numSample_nomod_1, numSample_nomod_2, numSample_nomod_3)) %>%
+  ggplot() + geom_line(aes(x = Log2FC, y = num_samples, group = SD, color = SD), size = 1.2) +
+  scale_color_discrete(cbPalette) +
[…]

### 2021-09-22 code/sample_size_analysis.R
updated sample size script

@@ -26,7 +26,7 @@ aa <- (delta / (z_alpha + z_beta)) ^2
 ## Original code containing sigma for subject.. can include or assume 0 for now
 ## numSample = round(2 * (median_sigma_error + median_sigma_subject) / aa, 0)
 
-cbPalette <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#999999")
+cbPalette <- c("#E69F00", "#56B4E9", "#009E73", "#999999", "#0072B2", "#D55E00", "#CC79A7", "#999999")
 
 numSample_nomod_1 <- round(2 * (.1) / aa, 0)
 numSample_nomod_2 <- round(2 * (.2) / aa, 0)
@@ -38,7 +38,7 @@ p1 <- data.frame("SD" = c(rep(".1", length(numSample_nomod_1)),
            "Log2FC" = c(desiredFC, desiredFC, desiredFC),
            "num_samples" = c(numSample_nomod_1, numSample_nomod_2, numSample_nomod_3)) %>%
   ggplot() + geom_line(aes(x = Log2FC, y = num_samples, group = SD, color = SD), size = 1.2) +
-  scale_color_discrete(cbPalette) +
+  scale_colour_manual(values=cbPalette) +
   labs(title = "Replicates Needed to Reach Power of .9", y = "# of replicates per condition") +
   theme_bw() +
   theme(axis.text.x = element_text(size = 12),
@@ -60,12 +60,13 @@ numSample_nomod_4 <- round((2 * .2) / aa, 0)
 p2 <- data.frame("SD" = c(rep(".2, .1", length(numSample_nomod_1)),
                           rep(".2, .2", length(numSample_nomod_1)),
                           rep(".2, .4", length(numSample_nomod_1)),
-                          rep("Baseline - .2", length(numSample_nomod_1))),
+                          rep("Baseline: .2", length(numSample_nomod_1))),
                  "Log2FC" = c(desiredFC, desiredFC, desiredFC, desiredFC),
                  "num_samples" = c(numSample_nomod_1, numSample_nomod_2, numSample_nomod_3,
                                    numSample_nomod_4)) %>%
-  ggplot() + geom_line(aes(x = Log2FC, y = num_samples, group = SD, color = SD)) +
-  labs(title = "Replicates Needed to Reach Power of .9", y = "# of replicates per condition") +
[…]

### 2021-09-23 supplementary/ptm_sm.tex
added IpaH section to supplementary

@@
 
+\subsection{IpaH7.8}
+
+In this experiment Shigella ubiquitin ligase IpaH7.8 was shown to inhibit the protein gasdermin D (GSDMD).
+Multiplex proteomics was used to quantify the abundance of total protein, and ubiquitination in human epithelial cells.
+Cells were either infected or uninfected with IpaH7.8-deficient Shigella flexneri and measurements were taken at different time periods.
+Uninfected cells were measured at 0 and 6 hours, while infected cells were measured at 1, 2, 4, and 6 hour increments, resulting in six total conditions.
+The experiment was inbalanced with two bioreplicates per condition for all conditions except for infected 1 hour.
+The experimental design can be seen in Table~\ref{table:ipah_design}.
+
+\begin{table}[h!]
+\centering
+\begin{tabular}{| c | c | c |}
+\hline
+ Condition & BioReplicate & Channel \\ [0.5ex]
+ \hline\hline
+ Dox1hr & Dox1hr\_1 & 127C\\
+ \hline
+ Dox2hr & Dox2hr\_1 & 128N\\
+\hline
+ Dox2hr & Dox2hr\_2 & 130C\\
+\hline
+ Dox4hr & Dox4hr\_1 & 128C\\
+\hline
+ Dox4hr & Dox4hr\_2 & 131C\\
+\hline
+ Dox6hr & Dox6hr\_1 & 129N\\
+\hline
+ Dox6hr & Dox6hr\_2 & 131N\\
+\hline
+ NoDox0hr & NoDox0hr\_1 & 126C\\
+\hline
+ NoDox0hr & NoDox0hr\_2 & 129C\\
+\hline
+ NoDox6hr & NoDox6hr\_1 & 127N\\
+\hline
+ NoDox6hr & NoDox6hr\_2 & 130N\\
+\hline
+
+\end{tabular}
+\caption{The experimental design of the IpaH7.8 experiment}
+\label{table:ipah_design}
+\end{table}
+
+A model was fit for the total protein, and ubiquitination separately, as described previously for TMT experiments in Section~\ref{sec:tmtmethod}. The model formula can be seen below.
+
+\todo{only one mixture so we probably wouldn't include the mixture as a term in the model?}
+
+$$Y_{mcb} = \mu + Condition_c + Subject_{mcb} + \epsilon_{mcb}$$
+
+$$\sum_{c=1}^C{Condition_c} = 0 ,\: Subject_{mcb} ~ \sim N(0, \sigma^2_S) ,\: \epsilon_{mcb} ~ \sim N(0, \sigma^2)$$
+
+The results of the proposed method to this experiment can be seen in \sfigref{fig:ipah_figures}.
+
+\begin{figure}[h!]
+\centering
+ \begin{subfigure}{\textwidth}
+	\includegraphics[width=\textwidth]{sim_new/IpaH_prof_plot}
+	\caption{Comparing the global profiling of protein $GSDMD\_HUMAN|P57764$ with the ubiquitination of the protein at site $K62$.
+The individual PSM features are shown in grey, while the feature summarization is shown in red.
+When looking at the summary of the modification and global protein it is clear the conditions follow different trends.
+Specifically, there appears to be no change in abundance between Dox1hr and Dox4hr in the modified plot, however there is a large negative change when looking at the unmodified plot.
+This indicates the modification is confounded with changes in the unmodified protein.}
+ \end{subfigure}
+ \begin{subfigure}{\textwidth}
+	\includegraphics[width=\textwidth]{sim_new/IpaH_volcano_plot}
+	\caption{Volcano plots of Dox4hr vs Dox1hr both before and after protein adjustment.
+The $GSDMD\_HUMAN|P57764\_K62$ modification is highlighted.
+Before adjustment the modification has a small fold change and insignificant adjusted pvalue.
+After adjustment the fold change is much larger and the adjusted pvalue is very significant.
+In this case the proposed method allows us to identify differential modified peptides that could have otherwise been missed.}
+ \end{subfigure}
+\label{fig:ipah_figures}
+\end{figure}
+
+\clearpage
 \subsection{Shigella flexneri}
@@
 Mixture two contained one replicate of uninfected KO and two uninfected WT.
-The experimental design can be seen in Table ~\ref{table:shigella_design}.
+The experimental design can be seen in Table~\ref{table:shigella_design}.
 
@@
 
-A model was fit for the total protein, phosphorylation, and ubiquitination seperately, as described previously for TMT experiments. The model formula can be seen below.
+A model was fit for the total protein, phosphorylation, and ubiquitination separately, as described previously for TMT experiments. The model formula can be seen below.
 

### 2021-09-23 code/IpaH_analysis.R

@@ -1,0 +1,181 @@
+
+library(MSstatsPTM)
+library(data.table)
+library(tidyverse)
+library(gridExtra)
+library(ggrepel)
+
+setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
+
+ptm_input <- read.csv("../data/IpaH_PTM_Input.txt")
+protein_input <- read.csv("../data/Ipah_Global_Input.txt", sep = "\t")
+protein_input$FeatureType <- "Peptide"
+protein_input$Abundance <- log2(protein_input$Intensity)
+
+ptm_summarization <- read.csv("../data/Ipah_KGG_TMT11_corrected_msstats_quant_combined_results.txt",
+                              sep = "\t")
+summarized_model <- ptm_summarization %>% filter(PeptideSequence == "Model")
+protein_summarization <- read.csv("../data/Ipah_Global_TMT11_msstats_quant_results.txt",
+                                  sep = "\t")
+
+protein_summarization$PeptideSequence <- "Model"
+protein_summarization$FeatureType <- "Model"
+protein_summarization_combined <- rbindlist(list(protein_input, protein_summarization), fill = TRUE)
+
+
+unadj_ptm_model <- read.csv("../data/Ipah_PTM_test_results.txt", sep = "\t")
+adj_ptm_model <- read.csv("../data/Ipah_PTM_Adjusted_test_results.txt", sep = "\t")
+
+combined_models <- merge(adj_ptm_model, unadj_ptm_model, all.x=TRUE, all.y=TRUE, by = c("Protein", "Label"))
+
+combined_models$FC_Diff <- abs(combined_models$log2FC.x - combined_models$log2FC.y)
+combined_models %>% arrange(desc(FC_Diff))
+
+
+## Profile Plots
+"GSDMD_HUMAN|P57764_K204"
+"GSDMD_HUMAN|P57764_K62"
+
+gsmd_sum <- protein_summarization_combined %>% filter(ProteinName == "GSDMD_HUMAN|P57764")
+gsmd_204_sum <- ptm_summarization %>% filter(ProteinName == "GSDMD_HUMAN|P57764_K204")
+gsmd_62_sum <- ptm_summarization %>% filter(ProteinName == "GSDMD_HUMAN|P57764_K62")
+
+gsmd_62_sum$BioReplicate <- factor(gsmd_62_sum$BioReplicate,
+                                     levels = c("NoDox0hr_1", "NoDox0hr_2", "NoDox6hr_1",
[…]

### 2021-09-23 main/MSstatsPTM_Main.tex
Updated computer sim and benchmark results sections of main

@@
 \usepackage{xr}
-\externaldocument[supp-]{ptm_sm}
+\externaldocument[supp-]{D:\Northeastern\Research\MSstats\MSstatsPTM-manuscript\supplementary\ptm_sm}
 %\def\sfigref#1{{\bf Supplementary Fig.~\ref{fig:#1}}}
@@
 \subsection*{Computer simulations}
+\label{sec:comp_sim}
 
 The proposed statistical approach methods were evaluated using computer simulation.
-Specifically, their properties under adjustment with respect to protein abundance and batch effects were evaluated.
+Specifically, their properties under adjustment with respect to protein abundance were evaluated.
 Two simulations were ran, one including a high number of features and no missing values, and one with few modified features and including missing values.
-
-\paragraph*{Computer simulation: protein-level adjustment.}
 
@@
 In real experiments, multiple inter-related conditions are often compared together.
-Whereas t-test uses measurements from the two conditions being compared, the proposed approach and limma leverages measurements in all conditions for the inference of underlying abundance.
+Whereas t-test uses measurements from the two conditions being compared, the proposed approach and limma leverage measurements in all conditions for the inference of underlying abundance.
 We evaluated the impact with such adjustment by computer simulation, considering the following factors: with/without protein-level changes, effect size, number of replicates and number of conditions.
-Details of the simulation can be found in Supplementary %\cref{supp-sec:sim}.
+Details of the simulations can be found in Supplementary %\cref{supp-sec:sim}.
 
-In this simulation, when the modification changes were entirely due to the changes in protein abundance across conditions, analysis without accounting for the protein-level changes resulted in off-target, high false positive rates \todo{figref (Fig.
-4b)}.
-As shown in \todo{figref Fig.
-4c-e}, in detection of systematic changes in PTM abundance, the proposed approach improved statistical power with small sample sizes in almost all the considered scenarios.
-The advantage by using the proposed approach over t-test became more profound in the presence of missing data \todo{figref (Fig.
-4d-e)}.
-In cases with small sample sizes (e.g., 2 replicates) as shown in \todo{figref Fig.
-4d}, performance by t-test decreased dramatically with one missing value, and increasing the log-fold change to 2 did not effectively reduce the negative impact.
+\paragraph*{Computer simulation 1}
+
+In this simulation, a clean dataset was generated with a high number of modified and unmodified peptide features and no missing values.
+Not accounting for changes in unmodified peptide level resulted in a high false discovery rate for all methods\todo{Reference supplementary figure}.
+When adjusting for changes in the unmodified peptide, all methods performed similarly in terms of FDR.
+When looking at recall and accuracy the proposed approach and limma performed similarly, while t-test lagged behind.
 Two-sample t-test only used data within the groups of interest while ignoring the rest of the data.
 Consequently, it gave similar performance across cases with different number of conditions.
-In contrast, the proposed approach leveraged all available information, which resulted in improved power with increased number of conditions (see for example, the improved performance in \todo{figref Fig.
-4e} [with 4 conditions] over \todo{figref Fig.
-4d} [with 2 conditions]).
+In contrast, the proposed approach and limma leveraged all available information, which resulted in improved power with increased number of conditions.
 Further results are provided in Supplementary \todo{link to Sec.
-S4.1}.
+S4.1.1}.
+
+\paragraph*{Computer simulation 2}
+
+In this simulation real world experimental conditions were introduced, including missing values and low number of modified features.
+In this case, changes in unmodified peptide level still needed to be accounted for to control the FDR in all methods.
+Once controlling for changes in unmodified peptide, it is clear that the proposed method performed stronger across all methods when compared to limma and t-test.
+The proposed method well calibrated the Recall rate and outperformed the other methods.
+Additionally, in overall accuracy the proposed method showed the strongest performance, even when the number of biological replicates were lower.
+Finally, the distribution of estimated fold changes is shown in \todo{reference sim2 boxplot}.
+The proposed method shows a tighter distribution of estimated fold changes around the true fold change, while the other methods showed much wider distributions.
+Further results are provided in Supplementary \todo{link to Sec.
+S4.1.2}.
 
@@
 \subsection*{Spike-in experiment}
-\todo{Need to write results of this experiment}
 
@@
 In *FigXX* we can clearly see the red labeled spike-in peptides do not follow the expected fold change before adjusting for changes in global protein level.
-After adjustment the estimated fold change is much close to expectation.
-Additionally, background peptides that should not show a significant change, colored grey, show many false positives before adjustment is made.
-Again after adjustment the results improve, and the number of false positives decrease significantly.
-In this experiment adjusting to remove confounding with changes in global protein abundance leads to a more accurate conclusion about the effect of modified peptides between conditions.
+After adjustment the estimated fold change is more in line with expectation.
+Additionally, background peptides serving as the null model, colored grey, show many false positives before adjustment is made.
+Again after adjustment the results improve and the number of false positives decrease significantly.
+
+\todo{Add limma plot}
+
+In *FigXX* we can see the results of the same experiment modeled using limma with protein adjustment discussed in Section~\ref{sec:comp_sim}.
+Again the red labeled spike-in peptides are more inline with expected fold change after adjustment is applied.
+However, using limma we can see that the majority of the differential spike in peptides are not showing a significant adjusted pvalue.
+In this case, usin limma would have resulted in missing the majority of differential modified peptides.
+In terms of false positives, the results are very similar to the propsed method, with many false positives before adjustment and much fewer afterward.
+
+Further results for this experiment are provided in Supplementary \todo{link to Sec. S4.2}.
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection*{Dilution experiment [What was this experiment?]}
+%\subsection*{Dilution experiment [What was this experiment?]}
 
-We evaluated our approach using a custom designed dilution benchmark experiment where two sets of liver samples from five Atg16L1 deficient mice (one with KGG enrichment followed by trypsin digestion, one with trypsin digestion alone) were prepared.
+%We evaluated our approach using a custom designed dilution benchmark experiment where two sets of liver samples from five Atg16L1 deficient mice (one with KGG enrichment followed by trypsin digestion, one with trypsin digestion alone) were prepared.
 Following a randomized order, each sample was injected at three different concentrations 4ul, 2ul, and 1ul.
@@
 Among these, 3173 have corresponding measurements from global data.
-To simulate the effects of changes in protein abundance, we created three groups of data by matching one concentration for the KGG data with one for the global data: K1P1, K2P2, and K4P4.
+%To simulate the effects of changes in protein abundance, we created three groups of data by matching one concentration for the KGG data with one for the global data: K1P1, K2P2, and K4P4.
 Systematic changes in the abundance of ubiquitinated sites between the groups are present in the KGG data.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection*{Re-analysis of published dataset: TMT experiment}
+\subsection*{Re-analysis of published datasets}
+\subsubsection*{IpaH7.8}
 
-To evaluate our approach on TMT experiments we reanalyzed an experiment targeting primary murine macrophages infected with Shigella flexneri (S.flexneri).
+\cite{LUCHETTI2021}
+
+\subsubsection{Shigella flexneri}
+To evaluate our approach we reanalyzed an experiment targeting primary murine macrophages infected with Shigella flexneri (S.flexneri).
 \cite{Maculins} Tandem mass tagging was used to quantify changes in total proteins, phosphorylation, and ubiquitination in Wild type (WT) controls and ATG16L1-deficient (cKO) bone marrow macrophages (BMDMs) either infected or uninfected with Shigella flexneri.
@@
 The major results are summarized below.
+
+\subsubsection{USP30}
 

### 2021-09-23 main/MSstatsPTM_Main.tex
Updated results section and moved additional considerations to main

@@
 
-Further results for this experiment are provided in Supplementary \todo{link to Sec. S4.2}.
+Further results of this experiment are provided in Supplementary \todo{link to Sec. S4.2}.
 
@@
 \subsection*{Re-analysis of published datasets}
+
 \subsubsection*{IpaH7.8}
 
-\cite{LUCHETTI2021}
+The proposed approach was evaluated on an experiment where Shigella ubiquitin ligase IpaH7.8 was shown to inhibit the protein gasdermin D (GSDMD).\cite{LUCHETTI2021} Multiplex proteomics was used to quantify the abundance of total protein, and ubiquitination in human epithelial cells.
+Cells were either infected or uninfected with IpaH7.8-deficient Shigella flexneri and measurements were taken at different time periods.
+Uninfected cells were measured at 0 and 6 hours, while infected cells were measured at 1, 2, 4, and 6 hour increments, resulting in six total conditions.
+The experiment was inbalanced with two bioreplicates per condition for all conditions except for infected 1 hour.
+About ~95\% of the identified modified peptides derived from proteins that were quantified in the global profiling run.
 
-\subsubsection{Shigella flexneri}
+Looking at modified peptide $GSDMD\_HUMAN|P57764$ we can see the advantage of the proposed method.
+The modified peptide shows a flat abundance change between the infected 1 hour, 4 hour, and 6 hour conditions.
+This is contrasted with a strong negative change in the global profiling run between the same conditions.
+Looking at the Dox4hr vs Dox1hr conditions and modeling the modified peptide without adjusting for changes in the global profiling, the fold change was $-.501$ and the adjusted pvalue was insignificant at $.0644$.
+After adjusting for changes in the global protein abundance, the fold change is much higher, $2.79$, and the adjusted pvalue became very significant, $5.25e^{-8}$.
+In this case the effect of the modified peptide is strongly confounded with changes in the global protein.
+The proposed method allows us to remove this confounding and estimate the true effect.
+The results of this can be seen in Figure \todo{add profile plot}.
+
+Further details and results of this experiment are provided in Supplementary \todo{link to Sec. S5.1}.
+
+\subsubsection*{Shigella flexneri}
 To evaluate our approach we reanalyzed an experiment targeting primary murine macrophages infected with Shigella flexneri (S.flexneri).
-\cite{Maculins} Tandem mass tagging was used to quantify changes in total proteins, phosphorylation, and ubiquitination in Wild type (WT) controls and ATG16L1-deficient (cKO) bone marrow macrophages (BMDMs) either infected or uninfected with Shigella flexneri.
-11-plex isobaric multiplex was used with two separate mixtures.
-Cell lysates were prepared over three different time periods, either uninfected, infected at early (45-60 min) or infected at late (3-3.5 hour) time points.
-There were 103,700 peptides mapping to 9,430 proteins in the global profiling run.
-Additionally, there were 25,600 unique phosphorylation sites and 12,400 KGG ubiquitination sites.
+\cite{Maculins} Multiplex proteomics was used to quantify the abundance of total protein, phosphorylation, and ubiquitination in wild type (WT) and ATG16L1-deficient (cKO) samples, uninfected and uninfected with S.flexneri.
+The abundance of total protein and post-translation modifications were quantified at three time points, uninfected, early infection (45-60 minutes), and late infection (3-3.5 hours).
+Quantifying the total protein along with the post-translational modifications allowed us to adjust for changes in total protein and see the true impact of the site specific modifications.
+Two mixtures using 11-plex were ran over the six conditions.
+The six conditions were split between 11 channels leading to the experimental design being unbalanced.
+Each mixture contained two replicates per early and late WT and KO conditions.
+Mixture one contained one replicate of uninfected WT and two replicates of uninfected KO.
+Mixture two contained one replicate of uninfected KO and two uninfected WT.
 About ~90\% of the identified modified peptides derived from proteins that were also quantified in the global profiling run.
 
-The major results are summarized below.
+Two specific modified peptides, $TTP\_MOUSE|P22893_S178$ and $KI67\_MOUSE|E9PVX6_T215$, show the power of the proposed method.
+The profile plot of modified peptide $TTP\_MOUSE|P22893_S178$ shows a large positive log fold change of $2.9$ between the WT\_Late and WT\_Uninfected conditions.
+However, the global profiling run shows a similar log fold change of $2.015$ between  the same conditions.
+This indicates that the abundance change in the modified peptide is nearly entirely due to changes in the global protein.
+When adjusting for the global protein the modified peptide's adjusted pvalue becomes insignificant, going from $.0009$ to $.275$.
+In contrast $KI67\_MOUSE|E9PVX6_T215$ originally shows a small positive log fold change of $.212$ between conditions WT\_Early and WT\_Uninfected.
+This is contrast with a negative fold change in the global profiling run of $-.616$.
+In adjusting the modified peptide for changes in the global protein, the log fold change becomes .828 and the adjusted pvalue goes from $.452$ to $.037$.
 
-\subsubsection{USP30}
+Further details and results of this experiment are provided in Supplementary \todo{link to Sec. S5.2}.
+
+\subsubsection*{USP30}
+
+This experiment looked into the relationship between USP30 and protein kinase PINK1, and their association with Parkinson’s Disease.
+Ubiquitination site profiling was performed and the modified site abundance was analyzed.
+Four conditions were tested with two biological replicates per condition.
+The conditions were as follows: CCCP, USP30 overexpression (USP30 OE), Combo, and Control.
+Label-free mass spectrometry quantification was used to quantify the abundance of modified peptides.
+A corresponding mixed effects model was fit per modification and global protein as described previously in this supplementary.
+
+In contrast to other experiments analyzed in this paper, there was no unmodified global protein profiling run performed in this experiment.
+Once identification and quantification of the Ubiquitinated profiling was performed, peptides which were unmodified were extracted and used in place of a global profiling run.
+This resulted in a significant lack of overlap between modified and unmodified peptides.
+Any modified peptide without a corresponding unmodified protein could not be adjusted.
+Of the 10,799 modified peptides identified, only 4526 had a corresponding unmodified run and could be adjusted.
+Finally, using this method also resulted in a low feature count for the unmodified protein model.
+
+Given the lack of matches in the global profiling run, the proposed method was unable to be applied to as many modified peptides as in the previous examples.
+However we can still see the power of the method when looking at peptide $P52209\_K059$.
+The abundance between conditions appears flat in the modified peptide profile plot, however this is contrast with a large negative change in the corresponding global protein.
+Using the proposed method we get increased information that would otherwise have been missed.
+
+Further details and results of this experiment are provided in Supplementary \todo{link to Sec. S5.3}.
 
@@
 The framework is general and is applicable to a variety of experimental designs.
-It outperforms the ad-hoc methods underlying the t-test, and yields accurate results in the broad type of experimental circumstances, including the presence of missing values, changes in protein abundance, batch effects, and different acquisition methods.
+It outperforms the ad-hoc methods underlying t-test and limma, and yields accurate results in the broad type of experimental circumstances, including the presence of missing values, changes in protein abundance, few representative peptides, and different acquisition methods.
 The framework allows us to plan for subsequent experiments, and choose the appropriate number of replicates in consideration of adjustment with respect to protein abundance.
@@
 The gain is due to a more efficient use of the data, and to a more accurate understanding of the systematic and random variations.
-The proposed framework can be extended beyond the experimental designs with multiple batches and conditions discussed above.
+The proposed framework can be extended beyond the experimental designs with variation from multiple sources discussed above.
 For example, it can represent experimental designs with even more complex structures, such as time series or factorial investigations.
@@
 The latest development of targeted acquisition and data-independent acquisition methods are expected to further alleviate these issues.
+
+Additionally, expression levels of PTM sites can be convoluted with each other if there are two or more modification sites per peptide.
+In the current implementation the effect of a specific modification in a peptide with multiple modifications cannot be quantified.
+One potential solution to this is to measure the abundance of peptides with one modification and use this to adjust the peptide with multiple sites to remove the convolution.
+However, this method would most likely run into major challenges due to sparsity of features for both modified peptides with a single and multiple modification sites, as well as a potential lack of matching single site modified peptides.
+A more complex approach to addressing this problem is most likely necessary.
 

### 2021-09-23 supplementary/ptm_sm.tex
Updated results section and moved additional considerations to main

@@
 \clearpage
-
-\todo{move to main manuscript}
-
-\section{Further Considerations}
-
-\subsubsection{Multiple sites per protein}
-
-Expression levels of PTM sites are adjusted based on the abundance of their originating protein.
-Since the same reference is used for all sites, it introduces correlation among estimates and test statistics for those sites.
-This may cause issues in controlling false discovery rate (FDR).
-We investigated the property by simulating data of two conditions and 1000 proteins with different numbers of PTM sites and comparing the results of the proposed approach and the linear model with no adjustment.
-%We investigated the property by simulating data of two conditions and 1000 proteins with different numbers of PTM sites. 
-A fraction (50\%) of the 1000 proteins had no changes between conditions while systematic changes were simulated for the rest of the proteins.
-Multiple testing correction was performed using the Benjamini and Hochberg's method.
-Performances of the considered approaches were assessed by their actual FDR, calculated as the fraction of proteins with adjusted $p$-values $<0.05$ among the proteins with true differences.
-The results are summarized in \sfigref{fig:prot_fdr}.
-%Performances of the considered approaches were assessed by their power and FDR.
-The power was calculated as the fraction of proteins with adjusted $p$-values $<0.05$ among those proteins with true differences between conditions, while the FDR was calculated as the fraction of proteins with adjusted $p$-values $<0.05$ among the proteins with true differences.
-The results are summarized in \sfigref{fig:prot_fdr}.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\clearpage
 \section*{References}

### 2021-09-23 main/MSstatsPTM_Main.tex
fixed typo in shigella results section

@@
 
-Two specific modified peptides, $TTP\_MOUSE|P22893_S178$ and $KI67\_MOUSE|E9PVX6_T215$, show the power of the proposed method.
-The profile plot of modified peptide $TTP\_MOUSE|P22893_S178$ shows a large positive log fold change of $2.9$ between the WT\_Late and WT\_Uninfected conditions.
+Two specific modified peptides, $TTP\_MOUSE|P22893\_S178$ and $KI67\_MOUSE|E9PVX6\_T215$, show the power of the proposed method.
+The profile plot of modified peptide $TTP\_MOUSE|P22893\_S178$ shows a large positive log fold change of $2.9$ between the WT\_Late and WT\_Uninfected conditions.
 However, the global profiling run shows a similar log fold change of $2.015$ between  the same conditions.
@@
 When adjusting for the global protein the modified peptide's adjusted pvalue becomes insignificant, going from $.0009$ to $.275$.
-In contrast $KI67\_MOUSE|E9PVX6_T215$ originally shows a small positive log fold change of $.212$ between conditions WT\_Early and WT\_Uninfected.
+In contrast $KI67\_MOUSE|E9PVX6\_T215$ originally shows a small positive log fold change of $.212$ between conditions WT\_Early and WT\_Uninfected.
 This is contrast with a negative fold change in the global profiling run of $-.616$.

### 2021-09-24 main/MSstatsPTM_Main.tex
cleaned up supplementary, added minor updates to main

@@
 \usepackage{hyperref}
+\usepackage{authblk}
 \usepackage{xcolor} %\[table,dvipsnames]
@@
 
-\author{
- Devon Kohler\\
- \and
- Tsung-Heng Tsai\\
- \and
- Ting Huang\\
- \and
- Erik Verschueren
- \and
- Trent Hinkle
- \and
-  Meena Choi\\
- \and
- Olga Vitek
-}
+\author[1]{Devon~Kohler}
+\author[2]{Tsung-Heng~Tsai}
+\author[1]{Ting~Huang}
+\author[4]{Erik~Verschueren}
+\author[3]{Trent~Hinkle}
+\author[3]{Meena~Choi}
+\author[1]{Olga~Vitek}
+\affil[1]{Khoury College of Computer Science, Northeastern University, Boston, MA, USA}
+\affil[2]{Kent State University, Kent, OH, USA}
+\affil[3]{MPL, Genentech, South San Francisco, CA, USA}
+\affil[4]{Galapagos, Mechelen, Antwerp, Belgium}
+
 \date{}
+
 \begin{document}
@@
 \maketitle
-\todo{Add Author list}
 %\noindent\author{Tsung-Heng Tsai$^{1}$, Olga Vitek$^{1,\ast}$}
@@
 We propose a general statistical model and workflow that is both reproducible and comprehensive.
-The model measures PTM and protein abundance by summarizing intensities through Tukey’s median polish method.
+The model measures modified and unmodified peptide abundance by summarizing intensities through Tukey’s median polish method.
 Then a model based on the family of linear mixed-effects models is fit.
@@
 
-The signaling mechanisms that allow cells to mount a dynamic and fast response to a multitude of events are primarily facilitated by the modification of proteins at specific residues, acting as molecular on/off switches.\cite{Deribe} \cite{Cohen}  Mass spectrometry-based label-free proteomics is broadly established as the tool-of-choice for unbiased and large-scale identification and quantification of proteins and their post-translational modifications (PTMs) using liquid chromatography coupled with mass spectrometry (LC-MS)\cite{Kall:2011ub} \cite{Roepstorff}.
-Studies targeting the post-translationally modified proteome focus either on the accurate localization of modification sites on proteins, relative or absolute quantification of a modification site’s occupancy repertoire, or relative changes in occupancy across experimental conditions.\cite{Mann} Regardless of the question at hand, interrogating the modified proteome is challenging due to a number of reasons.
+The signaling mechanisms that allow cells to mount a dynamic and fast response to a multitude of events are primarily facilitated by the modification of proteins at specific residues, acting as molecular on/off switches.\cite{Deribe} \cite{Cohen}  Mass spectrometry-based label-free proteomics is broadly established as the tool-of-choice for unbiased and large-scale identification and quantification of proteins and their post-translational modifications (PTMs) using liquid chromatography coupled with mass spectrometry (LC-MS).\cite{Kall:2011ub} \cite{Roepstorff} Studies targeting the post-translationally modified proteome focus either on the accurate localization of modification sites on proteins, relative or absolute quantification of a modification site’s occupancy repertoire, or relative changes in occupancy across experimental conditions.\cite{Mann} Regardless of the question at hand, interrogating the modified proteome is challenging due to a number of reasons.
 First, the relatively lower abundance of modified proteo-forms dictates that a global interrogation can only be achieved through large-scale enrichment protocols with modification-specific antibodies or beads.
@@
 Many investigations performed differential expression analysis of PTMs using two-sample t-test or its extensions.
-\todo{Add Ref about current PTM approaches} The approach takes as input intensities of individual features from modified peptides, or intensity ratios of modified and unmodified peptide features, and compares the mean abundance of a PTM site from one condition to another.
+The approach takes as input intensities of individual features from modified peptides, or intensity ratios of modified and unmodified peptide features, and compares the mean abundance of a PTM site from one condition to another.
 Modifications of the t-test such as moderated t-test with limma were also proposed.\cite{Zhu} While simple, the approach does not fully account for the sources of variations, and it is not directly applicable to experiments with complex designs, e.g., comparisons of multiple conditions, acquisition in multiple batches, etc.
@@
 The proposed approach was evaluated using datasets from computer simulations, benchmark controlled mixtures, and biological investigations.
-The approach was then compared against the commonly applied t-test and limma package.
-The results demonstrate that by appropriately leveraging the information from the entire dataset the proposed approach improves the reproducibility and accuracy of the estimates of PTM fold changes, results in a better calibrated type I error rate, and improves the statistical power of detecting changes in PTMs.
-The proposed approach is implemented as an open source R package MSstatsPTM, which employs similar input format as in MSstats and MSstatsTMT \cite{Choi:2014} \cite{Huang:2020}.
+The approach was compared against the commonly applied t-test and limma methods.
+The results demonstrate that by appropriately leveraging the information from the entire dataset, the proposed approach improves the reproducibility and accuracy of the estimates of PTM fold changes, results in a better calibrated type I error rate, and improves the statistical power of detecting changes in PTMs.
+The proposed approach is implemented as an open source R package MSstatsPTM, which employs similar input format as in MSstats and MSstatsTMT.
+\cite{Choi:2014} \cite{Huang:2020}
 
@@
 \newpage
-\section{Figure}
+\section{Figures}
 

### 2021-09-24 supplementary/ptm_sm.tex
cleaned up supplementary, added minor updates to main

@@
 \usepackage{authblk}
+\usepackage[backend=biber]{biblatex}
+\bibliography{ptm_ref}
 
@@
 
-\paragraph{Data structure of PTM quantification experiments.}
+\subsection{Data structure of PTM quantification experiments.}
 A set of fully-cleaved and/or partially-cleaved peptides containing a same PTM (e.g., ubiquitination/phosphorylation) at one site are considered together.
@@
 Limma uses linear models to test the null hypothesis that there is no difference in mean PTM abundances between conditions.
-Additionally it leverages Empirical Bayes moderation to share pooled variance information across individual modification models and moderate the individual residual variances.
-Using a linear model allows Limma to share variance information across conditions, providing a more accurate estimate.
+Additionally, it leverages Empirical Bayes moderation to share pooled variance information across individual modification models and moderate the individual residual variances.
+Using a linear model allows Limma to share variance information across conditions, providing a more accurate estimate.\cite{Ritchie_15a}
 
@@
 \]
+
+After fitting a linear model, the final variance is moderated using Empirical Bayes and the global variance across all modified peptides.
+This is done as $s_{i} = P(v_i| v_\Theta)$ where $v_i$ is the resulting variance when modeling peptide $i$ and $v_\Theta$ is the global variance.
 
@@
 
-The methods described previously can be extrapolated to Tandem Mass Tag (TMT) experiments targeting PTMs.
-The null hypothesis is that there is no difference in mean PTM abundance between Conditions $i$ and $i'$ after adjusting for changes in mean unmodified peptide abundance between Conditions $i$ and $i'$.
+The methods described previously can be extrapolated to Tandem Mass Tag (TMT) experiments targeting PTMs, similarly to how MSstatsTMT targets the global protein.\cite{Huang:2020} The null hypothesis is still that there is no difference in mean PTM abundance between Conditions $i$ and $i'$ after adjusting for changes in mean unmodified peptide abundance between Conditions $i$ and $i'$.
 \begin{align*}
@@
 For TMT experiments there is an additional source of variation coming from the mixtures used.
-This increases the complexity as there are many options for allocating samples to mixtures, including multiple mixtures, biological replicates, and technical replicates.
+This increases the model complexity as variation can now come from multiple mixtures, biological replicates, and technical replicates.
 Going forward we will denote the log-intensity of a modified peptide for Technical Replicate j of Mixture m for Condition i by $y_{ijm}$, where $m = 1,....,M$.
@@
 \begin{align*}
-Y_{mtcb} = \mu + Mixture_m + Techrep(Mixture)_{jm} + Condition_c + Subject_{mcb} + \epsilon_{mtcb}\\
-Y^*_{mtcb} = \mu^* + Mixture^*_m + Techrep(Mixture)^*_{jm} + Condition^*_c + Subject^*_{mcb} + \epsilon^*_{mtcb}
+Y_{mtcb} = \mu + Mixture_m + Techrep(Mixture)_{tm} + Condition_c + Subject_{mcb} + \epsilon_{mtcb}\\
+Y^*_{mtcb} = \mu^* + Mixture^*_m + Techrep(Mixture)^*_{tm} + Condition^*_c + Subject^*_{mcb} + \epsilon^*_{mtcb}
 \end{align*}
 
-Where $Mixture_m \sim N(0, \sigma^2_m)$, $Techrep(Mixture)_{jm} \sim N(0, \sigma^2_{jm})$, $Subject \sim N(0, \sigma^2_{s})$ are random effects, and $\sum_{c=1}^{I} Condition_{c}^{\ast} = 0$ is a fixed effect.
+Where $Mixture_m \sim N(0, \sigma^2_m)$, $Techrep(Mixture)_{tm} \sim N(0, \sigma^2_{tm})$, $Subject \sim N(0, \sigma^2_{s})$ are random effects, and $\sum_{c=1}^{I} Condition_{c}^{\ast} = 0$ is a fixed effect.
 Finally the error is expressed as $\epsilon_{mtcb} \sim N(0, \sigma^2)$.
@@
 
-The results are summarized from \sfigref{fig:prot_fpr} to \sfigref{fig:fc_boxplot}, including false discovery rate with or without changes in protein abundance by the considered methods (\sfigref{fig:prot_fpr}), false positive rate, and overall accuracy.
+The results are summarized from \sfigref{fig:sim1_fpr} to \sfigref{fig:sim1_recall}, including false discovery rate with or without changes in protein abundance by the considered methods, false positive rate, and overall accuracy.
 Additionally the results are compared to simulation 2 described in the next section.
@@
 
-The results are summarized from \sfigref{fig:sim1_fpr} to \sfigref{fig:fc_boxplot}, including false positive rate with or without changes in protein abundance by the considered methods (\sfigref{fig:sim1_fpr}), false positive rate, and overall accuracy.
-Again the results are compared to simulation 1.
+The results are summarized from \sfigref{fig:sim2_recall} to \sfigref{fig:fc_boxplot}, including false positive rate with or without changes in protein abundance by the considered methods, false positive rate, and overall accuracy.
 
@@
 \centering
- \begin{subfigure}{.9\textwidth}
-	\includegraphics[width=.9\textwidth]{sim_new/sim1_FDR_all_models}\\
+ \begin{subfigure}{\textwidth}
+  \centering
+	\includegraphics[width=.8\textwidth]{sim_new/sim1_FDR_all_models}\\
 	\caption{All the considered methods in simulation 1 correctly calibrated FDR rate when adjusting for changes in protein abundance.
@@
  \end{subfigure}
- \begin{subfigure}{.9\textwidth}
-	\includegraphics[width=.9\textwidth]{sim_new/sim1_FDR}
+ \begin{subfigure}{\textwidth}
+  \centering
+	\includegraphics[width=.8\textwidth]{sim_new/sim1_FDR}
 	\caption{The considered methods with protein adjustment are compared in detail. All three methods generally performed similarly in terms of FDR.}
  \end{subfigure}
+ \caption{FDR of Simulation 1.}
 \label{fig:sim1_fpr}
@@
 \centering
- \begin{subfigure}{.9\textwidth}
-	\includegraphics[width=.9\textwidth]{sim_new/sim1_Recall}\\
+ \begin{subfigure}{\textwidth}
+  \centering
+	\includegraphics[width=.8\textwidth]{sim_new/sim1_Recall}\\
 	\caption{The recall(false positive rate) for simulation 1 methods with adjustment were compared.
@@
  \end{subfigure}
- \begin{subfigure}{.9\textwidth}
-	\includegraphics[width=.9\textwidth]{sim_new/sim1_accuracy}
+ \begin{subfigure}{\textwidth}
+  \centering
+	\includegraphics[width=.8\textwidth]{sim_new/sim1_accuracy}
 	\caption{The overall accuracy plot mimics the observations in the recall plot. Limma performs stronger than the proposed method at lower replicates, while at higher they are comparable.}
  \end{subfigure}
+\caption{Recall and Accuracy results of Simulation 1.}
 \label{fig:sim1_recall}
@@
 \centering
- \begin{subfigure}{.9\textwidth}
-	\includegraphics[width=.9\textwidth]{sim_new/sim3_Recall}
+ \begin{subfigure}{\textwidth}
+  \centering
+	\includegraphics[width=.8\textwidth]{sim_new/sim3_Recall}
 	\caption{The advantage of using the proposed approach is apparent when looking at simulation 2, which includes limited observations and the presence of missing values.
 In the case of recall the proposed method performs stronger than limma and t-test in nearly every model.
-Even at lower replicates the proposed method still outperformed limma, especially when there were more conditions in the simulation.
-Again the lowest performance was observed using the $t$-test method.}
+Even at lower replicates the proposed method still outperformed limma.
+Again the lowest performaning method was $t$-test.}
  \end{subfigure}
- \begin{subfigure}{.9\textwidth}
-	\includegraphics[width=.9\textwidth]{sim_new/sim3_Accuracy}
+ \begin{subfigure}{\textwidth}
+  \centering
+	\includegraphics[width=.8\textwidth]{sim_new/sim3_Accuracy}
 	\caption{The overall accuracy plots show similar results to recall.
@@
  \end{subfigure}
+\caption{Recall and Accuracy results of Simulation 2.}
 \label{fig:sim2_recall}
@@
 \centering
- \begin{subfigure}{.9\textwidth}
-	\includegraphics[width=.9\textwidth]{sim_new/sim1_FC_boxplot}
-	\caption{In simulation 1 all considered methods correctly estimated the fold change between conditions, with a median fold change estimation of .75 in all methods.
+ \begin{subfigure}{\textwidth}
+ \centering
+	\includegraphics[width=.8\textwidth]{sim_new/sim1_FC_boxplot}
+	\caption{In simulation 1 all considered methods correctly estimated the fold change between conditions, with a median fold change estimation of .75.
 The distributions around the median were also consistent acorss all methods.
@@
  \end{subfigure}
- \begin{subfigure}{.9\textwidth}
-	\includegraphics[width=.9\textwidth]{sim_new/sim3_FC_boxplot}
+ \begin{subfigure}{\textwidth}
+  \centering
+	\includegraphics[width=.8\textwidth]{sim_new/sim3_FC_boxplot}
 	\caption{In simulation 2 again all methods correctly estimated the fold change with a median log change of .75.
@@
  \end{subfigure}
+\caption{Fold change distribution comparison between Simulations 1 and 2.}
 \label{fig:fc_boxplot}
@@
  \begin{subfigure}{\textwidth}
-	\includegraphics[width=\textwidth]{sim_new/IpaH_prof_plot}
+ \centering
+	\includegraphics[width=.9\textwidth]{sim_new/IpaH_prof_plot}
 	\caption{Comparing the global profiling of protein $GSDMD\_HUMAN|P57764$ with the ubiquitination of the protein at site $K62$.
@@
  \begin{subfigure}{\textwidth}
-	\includegraphics[width=\textwidth]{sim_new/IpaH_volcano_plot}
+ \centering
+	\includegraphics[width=.9\textwidth]{sim_new/IpaH_volcano_plot}
 	\caption{Volcano plots of Dox4hr vs Dox1hr both before and after protein adjustment.
@@
  \end{subfigure}
+\caption{Summary plots for modification of protein GSDMD at site K62.}
 \label{fig:ipah_figures}
@@
  \begin{subfigure}{\textwidth}
-	\includegraphics[width=\textwidth]{sim_new/No_Difference_Shigella_Profile_Plot}
+ \centering
+	\includegraphics[width=.9\textwidth]{sim_new/No_Difference_Shigella_Profile_Plot}
 	\caption{Comparing the global profiling of protein $TTP\_MOUSE|P22893$ with the modification of the protein at site $S178$.
@@
  \begin{subfigure}{\textwidth}
-	\includegraphics[width=\textwidth]{sim_new/No_Difference_Shigella_Volcano}
+ \centering
+	\includegraphics[width=.9\textwidth]{sim_new/No_Difference_Shigella_Volcano}
 	\caption{Volcano plots of WT\_Late vs WT\_Uninfect both before and after protein adjustment.
@@
  \end{subfigure}
+\caption{Summary plots for modification of protein TTP at site S178.}
 \label{fig:No_Diff_Shigella_PTM}
@@
  \begin{subfigure}{\textwidth}
-	\includegraphics[width=\textwidth]{sim_new/Difference_Shigella_Profile_Plot}
+ \centering
+	\includegraphics[width=.9\textwidth]{sim_new/Difference_Shigella_Profile_Plot}
 	\caption{Comparing the global profiling of protein $KI67\_MOUSE|E9PVX6$ with the modification of the protein at site $T215$.
@@
  \begin{subfigure}{\textwidth}
-	\includegraphics[width=\textwidth]{sim_new/Difference_Shigella_Volcano}
+ \centering
+	\includegraphics[width=.9\textwidth]{sim_new/Difference_Shigella_Volcano}
 	\caption{Volcano plots of WT\_Early vs WT\_Uninfect both before and after protein adjustment.
@@
  \end{subfigure}
+ \caption{Summary plots for modification of protein KI67 at site T215.}
 \label{fig:Diff_Shigella_PTM}
@@
 \centering
-\includegraphics[width=\textwidth]{sim_new/simple_power_analysis}
+\includegraphics[width=.9\textwidth]{sim_new/simple_power_analysis}
 \caption{The number of replicates per condition required to reach a power of .9 can be seen in the plot.
@@
 \centering
-\includegraphics[width=\textwidth]{sim_new/power_analysis_sd_combined}
+\includegraphics[width=.9\textwidth]{sim_new/power_analysis_sd_combined}
 \caption{When the unmodified peptide has a low standard deviation the number of replicates is very similar to the baseline.
@@
 
-\bibliographystyle{plain}
-\bibliography{ptm_ref}
+\printbibliography
 

### 2021-09-28 code/MSstatsPTM_Simulation_Analysis.Rmd
Updated main manuscript

@@
   return(results_temp)
+}
+
+calc_roc <- function(df, protein_col_name, pval_col_name){
+
+  temp_df <- df %>% select(get(protein_col_name), get(pval_col_name))
+
+  temp_df$True <- ifelse(grepl("NoChange", get(protein_col_name)), 0, 1)
+  temp_df$Pred <- ifelse(get(pval_col_name) < .05, 1, 0)
+
 }

### 2021-09-28 main/MSstatsPTM_Main.tex
Updated main manuscript

@@
 \usepackage{rotating}
-\usepackage{hyperref}
 \usepackage{authblk}
@@
 \usepackage{xr}
-\externaldocument[supp-]{D:\Northeastern\Research\MSstats\MSstatsPTM-manuscript\supplementary\ptm_sm}
+\externaldocument[supp-]{../supplementary/ptm_sm}
 %\def\sfigref#1{{\bf Supplementary Fig.~\ref{fig:#1}}}
@@
 To determine the statistical significance of the difference in terms of p-value, the test statistic is compared against the t distribution with degrees of freedom approximated by the Satterthwaite method.
-\cite{satterthwaite:1946} Adjustment for multiple comparisons is performed using the Benjamini-Hochberg procedure to control the false discovery rate at a desired level, e.g., 0.05.\cite{Benjamini:1995} The details are provided in Supplementary Sec.
-S3.2 \devon{link to supplementary using sfigref}.
+\cite{satterthwaite:1946} Adjustment for multiple comparisons is performed using the Benjamini-Hochberg procedure to control the false discovery rate at a desired level, e.g., $0.05$.\cite{Benjamini:1995} More details are provided in Supplementary Sec \ref{supp-sec:test}
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection*{Analysis with multiple batches}
+%\subsection*{Analysis with multiple batches}
 
-The proposed statistical framework allows to analyze data from experiments of complex designs such as factorial design.
+%The proposed statistical framework allows to analyze data from experiments of complex designs such as factorial design.
 We discuss below a specific design commonly considered in PTM experiments, in which data are acquired in multiple batches, often as a result of multiple repeats for the antibody enrichment step.
@@
 Alternatively, assuming identical variability in the site-level summaries and identical differences between conditions across batches (i.e., no interaction effect between condition and batch), the statistical inference and testing can be performed based on one single model with an additional fixed effect of batch.
-Details about these modeling considerations are discussed in Supplementary \todo{link to Sec.
-S3.4}.
+Details about these modeling considerations are discussed in Supplementary Sec \ref{supp-sec:test}.
 
@@
 TMT experiments introduce an additional source of variation in the form of different mixtures.
-To account for the mixture variable, first feature intensities are summarized as described previously, then a new linear mixed effects model with an added term for the mixture is included.
-This workflow and model follow the methods used for MSstatsTMT.
-\cite{Huang:2020} These methods are repeated to quantify and model global protein abundance.
-Once both PTM and global protein are modeled, the PTM model is adjusted for changes in global protein abundance using the same methods are before.
+To account for mixture variance, first feature intensities are summarized again using Tukey's Median Polish and missing values are imputed, if desired, then a new linear mixed effects model with an added term for the mixture is included.
+The workflow and statisical model described generally follow the methods used for modeling proteins in MSstatsTMT.
+\cite{Huang:2020} These methods are repeated to quantify both modified and unmodified peptide abundance.
+Once both modified and unmodified peptides are modeled, the PTM model is adjusted for changes in global protein abundance using the same methods described previously.
 
-\todo{add figref for TMT model}
-\todo{add figref for TMT experiment table}
+For more information about the extension to Tandem Mass Tag experiments see Supplementary Sec. \ref{supp-sec:tmtmethod}.
 
@@
 We evaluated the impact with such adjustment by computer simulation, considering the following factors: with/without protein-level changes, effect size, number of replicates and number of conditions.
-Details of the simulations can be found in Supplementary %\cref{supp-sec:sim}.
+Details of the results of this simulation can be found in Supplementary Sec.
+\ref{supp-sec:sim}.
 
@@
 In this simulation, a clean dataset was generated with a high number of modified and unmodified peptide features and no missing values.
-Not accounting for changes in unmodified peptide level resulted in a high false discovery rate for all methods\todo{Reference supplementary figure}.
+Not accounting for changes in unmodified peptide level resulted in a high false discovery rate for all methods, seen in Figure \ref{fig:fdr_all_models}.
 When adjusting for changes in the unmodified peptide, all methods performed similarly in terms of FDR.
@@
 In contrast, the proposed approach and limma leveraged all available information, which resulted in improved power with increased number of conditions.
-Further results are provided in Supplementary \todo{link to Sec.
-S4.1.1}.
+Further results are provided in Supplementary Figures \ref{supp-fig:sim1_fpr} to \ref{supp-fig:sim1_recall}.
 
@@
 In this case, changes in unmodified peptide level still needed to be accounted for to control the FDR in all methods.
-Once controlling for changes in unmodified peptide, it is clear that the proposed method performed stronger across all methods when compared to limma and t-test.
+Once controlling for changes in unmodified peptide, it is clear that the proposed method performed stronger across all methods when compared to limma and t-test, as seen in Figure \ref{fig:sim_accuracy}.
 The proposed method well calibrated the Recall rate and outperformed the other methods.
 Additionally, in overall accuracy the proposed method showed the strongest performance, even when the number of biological replicates were lower.
-Finally, the distribution of estimated fold changes is shown in \todo{reference sim2 boxplot}.
-The proposed method shows a tighter distribution of estimated fold changes around the true fold change, while the other methods showed much wider distributions.
-Further results are provided in Supplementary \todo{link to Sec.
-S4.1.2}.
+Finally, the proposed method showed a tighter distribution of estimated fold changes around the true fold change, while the other methods showed much wider distributions.
+Further results are provided in Supplementary Figures \ref{supp-fig:sim2_recall} to \ref{supp-fig:fc_boxplot}.
 
@@
 Four mixes of spike in peptides and Human Lysate were created to create conditions with known fold change.
-\todo{Add figure of mixtures} Two sets of data were acquired for each mixture: modifed peptide data, including the spike in KGG enriched peptides, and unmodified peptides from Human Lysate.
+Two sets of data were acquired for each mixture: modified peptide data, including the spike in KGG enriched peptides, and unmodified peptides from Human Lysate.
 These datasets were used as the PTM and global protein data.
 
-\todo{figref volcano plot of spike in peptides}
-
-In *FigXX* we can clearly see the red labeled spike-in peptides do not follow the expected fold change before adjusting for changes in global protein level.
+In Figure \ref{fig:spikein_prop_volcano} we can clearly see the red labeled spike-in peptides do not follow the expected fold change before adjusting for changes in global protein level.
 After adjustment the estimated fold change is more in line with expectation.
@@
 
-\todo{Add limma plot}
-
-In *FigXX* we can see the results of the same experiment modeled using limma with protein adjustment discussed in Section~\ref{sec:comp_sim}.
+In Figure \ref{fig:spikein_limma_volcano} we can see the results of the same experiment modeled using limma with protein adjustment discussed in Section~\ref{sec:comp_sim}.
 Again the red labeled spike-in peptides are more inline with expected fold change after adjustment is applied.
 However, using limma we can see that the majority of the differential spike in peptides are not showing a significant adjusted pvalue.
-In this case, usin limma would have resulted in missing the majority of differential modified peptides.
-In terms of false positives, the results are very similar to the propsed method, with many false positives before adjustment and much fewer afterward.
+In this case, using limma would have resulted in missing the majority of differential modified peptides.
+In terms of false positives, the results are very similar to the proposed method, with many false positives before adjustment and much fewer afterward.
 
-Further results of this experiment are provided in Supplementary \todo{link to Sec. S4.2}.
+Further results of this experiment are provided in Supplementary Sec. \ref{supp-sec:benchmark}.
 
@@
 Uninfected cells were measured at 0 and 6 hours, while infected cells were measured at 1, 2, 4, and 6 hour increments, resulting in six total conditions.
-The experiment was inbalanced with two bioreplicates per condition for all conditions except for infected 1 hour.
+The experiment was unbalanced with two bioreplicates per condition for all conditions except for infected 1 hour.
 About ~95\% of the identified modified peptides derived from proteins that were quantified in the global profiling run.
@@
 The proposed method allows us to remove this confounding and estimate the true effect.
-The results of this can be seen in Figure \todo{add profile plot}.
+The results of this can be seen in Figure \ref{fig:ipah_profile}.
 
-Further details and results of this experiment are provided in Supplementary \todo{link to Sec. S5.1}.
+Further details and results of this experiment are provided in Supplementary Sec. \ref{supp-sec:ipah}.
 
@@
 
-Further details and results of this experiment are provided in Supplementary \todo{link to Sec. S5.2}.
+Further details and results of this experiment are provided in Supplementary Sec. \ref{supp-sec:shigella}.
 
@@
 Four conditions were tested with two biological replicates per condition.
-The conditions were as follows: CCCP, USP30 overexpression (USP30 OE), Combo, and Control.
+The conditions were as follows: CCCP, USP30 over expression (USP30 OE), Combo, and Control.
 Label-free mass spectrometry quantification was used to quantify the abundance of modified peptides.
@@
 
-Further details and results of this experiment are provided in Supplementary \todo{link to Sec. S5.3}.
+Further details and results of this experiment are provided in Supplementary Sec. \ref{supp-sec:usp30}.
 
@@
 \centering
-\includegraphics[scale=.8]{images/fig2.png}
-\caption{\todo{todo}}
-\label{fig:batch}
+\includegraphics[scale=.8]{images/fig3.png}
+\caption{
+Data structure of a typical PTM experiment and goals of PTM characterization.
+(a) Schematic data representation, in a simplified case of two conditions and two replicate runs.
+Each PTM site is modeled and characterized separately, where a PTM is quantified with multiple spectral features (boxes), distinguished by different charge states of a peptide.
+The feature intensities are viewed as repeated measurements of the underlying abundance of the PTM, where the abundance in Condition i is denoted by i.
+Features corresponding to unmodified peptides are considered together to perform adjustment with respect to protein abundance, where the protein abundance in Condition i is denoted by i*.
+Peptides can be fully cleaved (solid lines) and/or partially cleaved (dashed lines).
+Some spectral features can be missing.
+(b) PTM relative quantification by statistical inference, which makes use of the feature intensities to infer the underlying PTM abundance and protein abundance with an estimate of associated uncertainty.
+(c) Model-based testing for differential PTM abundance, which corrects for the underlying protein abundance with a cost of increased uncertainty about the estimate of difference between conditions.}
+\label{fig:data-structure}
 \end{figure}
@@
 \centering
-\includegraphics[scale=.8]{images/fig3.png}
-\caption{
-Data structure of a typical PTM experiment and goals of PTM characterization.(a)Schematic data representation, in a simplified case of two conditions and two replicate runs.
-Each PTM site is modeled and characterized separately, where a PTM is quantified with multiple spectral features (boxes), distinguished by different charge states of a peptide.
-The feature intensities are viewed as repeated measurements of the underlying abundance of the PTM, where the abundance in Condition i is denoted by i.
-Features corresponding to unmodified peptides are considered together to perform adjustment with respect to protein abundance, where the protein abundance in Condition i is denoted by *i.
-Peptides can be fully cleaved (solid lines) and/or partially cleaved (dashed lines).
-Some spectral features can be missing.(b)PTM relative quantification by statistical inference, which makes use of the feature intensities to infer the underlying PTM abundance and protein abundance with an estimate of associated uncertainty.
-(c) Model-based testing for differential PTM abundance, which corrects for the underlying protein abundance with a cost of increased uncertainty about the estimate of difference between conditions.
-\todo{there is no d..}(d) Statistical experimental design in terms of sample size calculations and power analysis.}
-\label{fig:data-structure}
+\includegraphics[width=1\textwidth]{images/sim1_FDR_all_models.png}
+\caption{All the considered methods in simulation 1 correctly calibrated FDR when adjusting for changes in protein abundance.
+In comparison, the methods without accounting for the protein-level changes resulted in off-target, high false positive rates.
+}\label{fig:fdr_all_models}
 \end{figure}
 
-{\bf Figure 4} False positive rate and statistical power in PTM significance analysis by the proposed approach, t-test with protein-level adjustment, and t-test without protein-level adjustment.
+\begin{figure}[ht]
+\centering
+\includegraphics[width=1\textwidth]{images/sim3_Accuracy.png}
+\caption{The advantage of using the proposed approach is apparent when looking at simulation 2, which includes limited observations and the presence of missing values.
+In overall accuracy, the proposed method performs stronger than limma and t-test in nearly every model.
+Even at lower replicates the proposed method still outperformed limma.
+The lowest performaning method was $t$-test.
+Limma shows comparable performance to the proposed method in a clean experiment, however when real world data problems are introduced it is clear the proposed method is more robust.}
+\label{fig:sim_accuracy}
+\end{figure}
+
+\begin{figure}[ht]
+\centering
+\includegraphics[width=1\textwidth]{images/spike_in_msstatsptm_volcano.png}
+\caption{Using the proposed method to model the benchmark experiment the spike in peptides (colored red) do not follow the expected log fold change before adjustment.
+After adjusting for changes in overall protein abundance the spike in peptides are more in line with expectation.
+Additionally the background grey colored peptides showed many false positives before adjustment.
+After adjustment the false positives were decreased considerably.
+}\label{fig:spikein_prop_volcano}
+\end{figure}
+
+\begin{figure}[ht]
+\centering
+\includegraphics[width=1\textwidth]{images/spike_in_limma_volcano.png}
+\caption{Limma method used to model the spike in experiment.
+Using limma the spike in peptides follow the expected log fold change better after adjusting for changes in protein level.
+However, while the fold change is much more accurate, the majority of spike in peptides do not have a significant adjusted pvalue.
+In terms of false positives, the results are very similar to the proposed method, with many false positives before adjustment and much fewer after.
+}\label{fig:spikein_limma_volcano}
+\end{figure}
+
+\begin{figure}[ht]
+\centering
+\includegraphics[width=1\textwidth]{images/IpaH_prof_plot.png}
+\caption{Comparing the global profiling of protein $GSDMD\_HUMAN|P57764$ with the ubiquitination of the protein at site $K62$.
+The individual PSM features are shown in grey, while the feature summarization is shown in red.
+When looking at the summary of the modification and global protein it is clear the conditions follow different trends.
+Specifically, there appears to be no change in abundance between Dox1hr and Dox4hr in the modified plot, however there is a large negative change when looking at the unmodified plot.
+This indicates the modification is confounded with changes in the unmodified protein.
+}\label{fig:ipah_profile}
+\end{figure}
+
+%{\bf Figure 4} False positive rate and statistical power in PTM significance analysis by the proposed approach, t-test with protein-level adjustment, and t-test without protein-level adjustment.
 (a) When there was no change in protein abundance, all considered methods well calibrated the Type I error rate.
@@
 
-{\bf Figure 5} Results corresponding to estimation error and false positive rate and statistical power of the PTM significance analysis, where the data were acquired in two batches.
+%{\bf Figure 5} Results corresponding to estimation error and false positive rate and statistical power of the PTM significance analysis, where the data were acquired in two batches.
 The following parameters were considered to represent the batch effects: no batch-condition interaction, mean intensity level in Batch 2 was higher than Batch 1 by 2 on log scale, and the SDs in Batch 1 and Batch 2 were 0.2 and 0.3, respectively.(a)Estimation based on the most statistically significant batch with t-test was more variable than other methods and frequently biased.
@@
 
-{\bf Figure 6} Design of future PTM experiments in terms of sample size calculations and power analysis.
+%{\bf Figure 6} Design of future PTM experiments in terms of sample size calculations and power analysis.
 (a) Protein-level adjustment relies on the inference of protein abundance, which introduces additional uncertainty in the estimate of PTM difference.

### 2021-09-28 supplementary/ptm_sm.tex
Updated main manuscript

@@
 \usepackage{authblk}
+\usepackage{xr}
+\externaldocument[main-]{../main/MSstatsPTM_Main}
+
 \usepackage[backend=biber]{biblatex}
@@
 	\includegraphics[width=.8\textwidth]{sim_new/sim1_FDR_all_models}\\
-	\caption{All the considered methods in simulation 1 correctly calibrated FDR rate when adjusting for changes in protein abundance.
+	\caption{All the considered methods in simulation 1 correctly calibrated FDR  when adjusting for changes in protein abundance.
 In comparison, the methods without accounting for the protein-level changes resulted in off-target, high false positive rates.}
@@
 \includegraphics[width=.85\textwidth]{sim_new/spike_in_limma_volcano}
-\caption{Using the limma method the spike in peptides follow the expected log fold change better after adjusting for changes in protein level.
-However, while the fold change is much more accurate, the majority of spike in peptides do not have a significante adjusted pvalue.
-In terms of false positives, the results are very similar to the propsed method, with many false positives before adjustment and much fewer afterward.
+\caption{Limma method used to model the spike in experiment.
+Using limma the spike in peptides follow the expected log fold change better after adjusting for changes in protein level.
+However, while the fold change is much more accurate, the majority of spike in peptides do not have a significant adjusted pvalue.
+In terms of false positives, the results are very similar to the proposed method, with many false positives before adjustment and much fewer after.
 \label{fig:spike_volcano_limma}}
@@
 \subsection{IpaH7.8}
+\label{sec:ipah}
 
@@
 \subsection{Shigella flexneri}
-
+\label{sec:shigella}
 In this study, the correlation between the gene Atg16L1 and killing of Shigella flexneri (S.flexneri) was assessed.
-Multiplex proteomics was used to quantify the abundance of total protein, phosphorylation, and ubiquitination in wild type (WT) and ATG16L1-deficient (cKO) samples, uninfected and uninfected with S.flexneri.
+\cite{Maculins} Multiplex proteomics was used to quantify the abundance of total protein, phosphorylation, and ubiquitination in wild type (WT) and ATG16L1-deficient (cKO) samples, uninfected and uninfected with S.flexneri.
 The abundance of total protein and post-translation modifications were quantified at three time points, uninfected, early infection (45-60 minutes), and late infection (3-3.5 hours).
@@
 \subsection{USP30}
+\label{sec:usp30}
 

### 2021-09-28 supplementary/sim_old/ptm_sm_original.tex
Updated main manuscript

--- 2021-01-25 supplementary/ptm_sm.tex
@@
 
+\todo{If we use MSstats protein summarization + t-test, how much would they be different, methodologically and performance?}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
+\todo{should address number of replicates, standard deviations, missing data?, same as one site per protein.}
+\todo{not strong evidence. need to revisit this section.}
+
 \begin{figure}[h!]

### 2021-09-28 main/MSstatsPTM_Main.tex
Edited text and phrasing errors

@@
 \usepackage{authblk}
+\usepackage{subcaption}
 \usepackage{xcolor} %\[table,dvipsnames]
@@
 
-The signaling mechanisms that allow cells to mount a dynamic and fast response to a multitude of events are primarily facilitated by the modification of proteins at specific residues, acting as molecular on/off switches.\cite{Deribe} \cite{Cohen}  Mass spectrometry-based label-free proteomics is broadly established as the tool-of-choice for unbiased and large-scale identification and quantification of proteins and their post-translational modifications (PTMs) using liquid chromatography coupled with mass spectrometry (LC-MS).\cite{Kall:2011ub} \cite{Roepstorff} Studies targeting the post-translationally modified proteome focus either on the accurate localization of modification sites on proteins, relative or absolute quantification of a modification site’s occupancy repertoire, or relative changes in occupancy across experimental conditions.\cite{Mann} Regardless of the question at hand, interrogating the modified proteome is challenging due to a number of reasons.
+The signaling mechanisms that allow cells to mount a dynamic and fast response to a multitude of events are primarily facilitated by the modification of proteins at specific residues, acting as molecular on/off switches.\cite{Deribe} \cite{Cohen}  Mass spectrometry-based label-free proteomics is broadly established as the tool-of-choice for unbiased and large-scale identification and quantification of proteins and their post-translational modifications (PTMs) using liquid chromatography coupled with mass spectrometry (LC-MS).\cite{Kall:2011ub} \cite{Roepstorff} Studies targeting the post-translationally modified proteome focus either on the accurate localization of modification sites on proteins, relative or absolute quantification of a modification site's occupancy repertoire, or relative changes in occupancy across experimental conditions.\cite{Mann} Regardless of the question at hand, interrogating the modified proteome is challenging due to a number of reasons.
 First, the relatively lower abundance of modified proteo-forms dictates that a global interrogation can only be achieved through large-scale enrichment protocols with modification-specific antibodies or beads.
@@
 Quantitative analyses of PTMs often involve comparisons between multiple inter-related conditions of the same biological system.
-The general statistical framework underlying the proposed approach allows for analyzing experiments with complex designs, including those with multiple conditions, multiple batches, and measured at different points in time.
+The general statistical framework underlying the proposed approach allows for analyzing experiments with complex designs and different acquisition methods, including label-free, DDA, DIA, and TMT.
 
@@
 The approach was compared against the commonly applied t-test and limma methods.
-The results demonstrate that by appropriately leveraging the information from the entire dataset, the proposed approach improves the reproducibility and accuracy of the estimates of PTM fold changes, results in a better calibrated type I error rate, and improves the statistical power of detecting changes in PTMs.
+The results demonstrate that by appropriately leveraging the information from the entire dataset, the proposed approach improves the reproducibility and accuracy of the estimates of PTM fold changes and results in a better calibrated type I error rate.
 The proposed approach is implemented as an open source R package MSstatsPTM, which employs similar input format as in MSstats and MSstatsTMT.
@@
 We evaluated the impact with such adjustment by computer simulation, considering the following factors: with/without protein-level changes, effect size, number of replicates and number of conditions.
-Details of the results of this simulation can be found in Supplementary Sec.
+Further details of how the simulations were generated and their results can be found in Supplementary Sec.
 \ref{supp-sec:sim}.
@@
 When adjusting for changes in the unmodified peptide, all methods performed similarly in terms of FDR.
-When looking at recall and accuracy the proposed approach and limma performed similarly, while t-test lagged behind.
-Two-sample t-test only used data within the groups of interest while ignoring the rest of the data.
+Recall and accuracy of the proposed approach and limma performed similarly, while t-test lagged behind.
+Two-sample t-test only uses data within the groups of interest while ignoring the remainder of the data.
 Consequently, it gave similar performance across cases with different number of conditions.
@@
 Once controlling for changes in unmodified peptide, it is clear that the proposed method performed stronger across all methods when compared to limma and t-test, as seen in Figure \ref{fig:sim_accuracy}.
-The proposed method well calibrated the Recall rate and outperformed the other methods.
-Additionally, in overall accuracy the proposed method showed the strongest performance, even when the number of biological replicates were lower.
+The proposed method well calibrated model accuracy and outperformed the other methods.
+Additionally, in the recall rate the proposed method showed the strongest performance, even when the number of biological replicates was low.
 Finally, the proposed method showed a tighter distribution of estimated fold changes around the true fold change, while the other methods showed much wider distributions.
@@
 \subsubsection*{Shigella flexneri}
+
 To evaluate our approach we reanalyzed an experiment targeting primary murine macrophages infected with Shigella flexneri (S.flexneri).
@@
 In adjusting the modified peptide for changes in the global protein, the log fold change becomes .828 and the adjusted pvalue goes from $.452$ to $.037$.
+The profile plots of these two PTMs can be seen in Figure \ref{fig:shigella_profile}
 
@@
 Label-free mass spectrometry quantification was used to quantify the abundance of modified peptides.
-A corresponding mixed effects model was fit per modification and global protein as described previously in this supplementary.
+A corresponding mixed effects model was fit per modification and global protein as described previously.
 
@@
 Any modified peptide without a corresponding unmodified protein could not be adjusted.
-Of the 10,799 modified peptides identified, only 4526 had a corresponding unmodified run and could be adjusted.
-Finally, using this method also resulted in a low feature count for the unmodified protein model.
+Of the 10,799 modified peptides identified, only 4526 had a corresponding unmodified protein and could be adjusted.
+Not having a separate global profiling run resulted in very low feature counts for the unmodified protein model.
 
@@
 The abundance between conditions appears flat in the modified peptide profile plot, however this is contrast with a large negative change in the corresponding global protein.
+Specifically looking at the comparison between the CCCP and Control conditions, the unadjusted model showed a small log fold change of $-.255$ and insignificant adjusted pvalue of $.104$.
+However when adjusting for changes in the undmodified protein, the log fold change increased to $-2.73$ and adjusted pvalue became significant at $1.067e-04$.
+The profile plot for this PTM can be seen in Figure \ref{fig:usp30_profile}.
 Using the proposed method we get increased information that would otherwise have been missed.
@@
 One potential solution to this is to measure the abundance of peptides with one modification and use this to adjust the peptide with multiple sites to remove the convolution.
-However, this method would most likely run into major challenges due to sparsity of features for both modified peptides with a single and multiple modification sites, as well as a potential lack of matching single site modified peptides.
+However, this method would most likely run into major challenges due to sparsity of features for modified peptides with both a single and multiple modification sites.
 A more complex approach to addressing this problem is most likely necessary.
@@
 Specifically, there appears to be no change in abundance between Dox1hr and Dox4hr in the modified plot, however there is a large negative change when looking at the unmodified plot.
-This indicates the modification is confounded with changes in the unmodified protein.
-}\label{fig:ipah_profile}
+This indicates the modification is confounded with changes in the unmodified protein.}
+\label{fig:ipah_profile}
+\end{figure}
+
+\begin{figure}[h!]
+\centering
+\begin{subfigure}{\textwidth}
+ \centering
+	\includegraphics[width=.95\textwidth]{images/No_Difference_Shigella_Profile_Plot}
+	\caption{Comparing the global profiling of protein $TTP\_MOUSE|P22893$ with the modification of the protein at site $S178$.
+The individual PSM features are shown in grey, while the feature summarization is shown in red.
+When looking at the summary of the modification and global protein it is clear the difference between conditions follow the same trend.
+Specifically, there is a positive adjustment in abundance when comparing WT\_Uninfect to WT\_Late in both the modification and global profiling run.
+This indicates the movement is driven by changes in global protein.}
+ \end{subfigure}\vspace{5mm}
+ \begin{subfigure}{\textwidth}
+ \centering
+	\includegraphics[width=.95\textwidth]{images/Difference_Shigella_Profile_Plot}
+	\caption{Comparing the global profiling of protein $KI67\_MOUSE|E9PVX6$ with the modification of the protein at site $T215$.
+In this case the modification and global protein trend in different directions.
+Specifically, comparing WT\_Uninfect and WT\_Early there is a slightly positive change in abundance, however in the global profiling there was a negative change.
+In this case the profile plot indicates the effect of the modification is masked by the change in global protein abundance.
+Additionally this profile plot shows the large difference in available features between modifications and global protein.}
+ \end{subfigure}
+\caption{Profile plots for PTMs P22893\_S178 and E9PVX6\_T215 in the Shigella flexneri experiment.}
+\label{fig:shigella_profile}
+\end{figure}
+
+\begin{figure}[ht]
+\centering
+\includegraphics[width=1\textwidth]{images/USP30_profile_plot.png}
+\caption{Comparing the global profiling of protein $P52209$ with the modification of the protein at site $K059$.
+The modification appears generally unchanged between conditions, whereas the global profiling run shows the CCCP and Combo conditions at a higher relative abundance compared to the Control and USP30\_OE.
+This indicates that the modification actually had a major effect when comparing CCCP and Combo to Control and USP30\_OE, which would have been missed without adjusting for global protein changes.}
+\label{fig:usp30_profile}
 \end{figure}

### 2021-09-28 supplementary/ptm_sm.tex
Edited text and phrasing errors

@@
 	\caption{All the considered methods in simulation 1 correctly calibrated FDR  when adjusting for changes in protein abundance.
-In comparison, the methods without accounting for the protein-level changes resulted in off-target, high false positive rates.}
+In comparison, the methods without accounting for the protein-level changes resulted in off-target, high false discovery rates.}
  \end{subfigure}
@@
 	\includegraphics[width=.8\textwidth]{sim_new/sim1_Recall}\\
-	\caption{The recall(false positive rate) for simulation 1 methods with adjustment were compared.
-It is clear that Limma performs the strongest here when the number of replicates were low.
-At higher replicates the performance of the proposed methods and Limma are comparable.
-T-test clearly performs worse across all methods.}
+	\caption{The recall for simulation 1 methods with adjustment were compared.
+It is clear that limma performs the strongest here when the number of replicates were low.
+At higher replicates the performance of the proposed methods and limma are comparable.
+$t$-test clearly performs worse across all methods.}
  \end{subfigure}
@@
 Even at lower replicates the proposed method still outperformed limma.
-Again the lowest performaning method was $t$-test.}
+Again the lowest performing method was $t$-test.}
  \end{subfigure}
@@
 The proposed method performed the strongest across all methods, even when replicates were low.
-Limma shows strong performance in a clean experiment, when real world data problems are introduced it is clear the proposed method is more robust.}
+Limma shows strong performance in a clean experiment, however when real world data problems are introduced it is clear the proposed method is more robust.}
  \end{subfigure}
@@
 	\caption{In simulation 1 all considered methods correctly estimated the fold change between conditions, with a median fold change estimation of .75.
-The distributions around the median were also consistent acorss all methods.
+The distributions around the median were also consistent across all methods.
 Predictably the distribution quantiles decrease as the number of replicates increases.}
@@
 The proposed method in this simulation had a visibly tighter distribution around the median.
-Both Limma and $t$-test showed a wider range around the fold change.
-In this case the proposed method showed a stronger performed in correctly estimating the fold change for all peptides.}
+Both limma and $t$-test showed a wider range around the fold change.
+In this case the proposed method showed a stronger performance in correctly estimating the fold change for all peptides.}
  \end{subfigure}
@@
 
-Again we consider three different methods and assess their performance: the method proposed in this paper, Limma, and two sample $t$-test.
+Again we consider three different methods and assess their performance: the method proposed in this paper, limma, and two sample $t$-test.
 All methods are analyzed after adjusting for changes in overall protein level.
-The propsed method summarizes feature intensities up to the run level using Tukey's Median Polish, while the other methods use the log of summed feature intensities.
-The results are summarized from \sfigref{fig:spike_volcano_msstats} to \sfigref{fig:spike_fc}, including volcano plots, model summary statsitics, and fold change analysis.
+The proposed method summarizes feature intensities up to the run level using Tukey's Median Polish, while the other methods use the log of summed feature intensities.
+The results are summarized from \sfigref{fig:spike_volcano_msstats} to \sfigref{fig:spike_fc}, including volcano plots, model summary statistics, and fold change analysis.
 
@@
 \includegraphics[width=.85\textwidth]{sim_new/spike_in_msstatsptm_volcano}
-\caption{Using the proposed method to model the benchmark experiment the spike in peptides (colored red) do not follow the expected log fold change before adjustment.
+\caption{Using the proposed method to model the benchmark experiment, the spike in peptides (colored red) do not follow the expected log fold change before adjustment.
 After adjusting for changes in overall protein abundance the spike in peptides are more in line with expectation.
-Additionally the background grey colored peptides showed many false positives before adjustment.
+Additionally the background grey colored peptides show many false positives before adjustment.
 After adjustment the false positives were decreased considerably.
@@
 \includegraphics[width=.85\textwidth]{sim_new/spike_in_limma_volcano}
-\caption{Limma method used to model the spike in experiment.
-Using limma the spike in peptides follow the expected log fold change better after adjusting for changes in protein level.
+\caption{When modeling the experiment with the limma method, the spike in peptides again follow the expected log fold change better after adjusting for changes in protein level.
 However, while the fold change is much more accurate, the majority of spike in peptides do not have a significant adjusted pvalue.
+In this case, the known differencial peptides are missed by the model.
 In terms of false positives, the results are very similar to the proposed method, with many false positives before adjustment and much fewer after.
@@
 \includegraphics[width=.85\textwidth]{sim_new/spike_in_ttest_volcano}
-\caption{Using the two sample $t$-test, none of the comparisons either before or after adjustment do not show any significant peptides.
+\caption{Using the two sample $t$-test, none of the comparisons either before or after adjustment show any significant peptides.
 With that being said, the fold change of the spike in peptides is much closer to expectation after adjusting for global protein abundance.
@@
 \caption{Comparing the summary statistics between methods it is clear that the proposed method performs the strongest.
-In terms of accuracy and specificity the three methods are close, with Limma and $t$-test showing slightly higher values.
+In terms of accuracy and specificity the three methods are close, with limma and $t$-test showing slightly higher values.
 Accuracy and specificity are dominated by the large number of true negatives (background peptides) compared to the true positives (spike in peptides).
@@
 Before adjustment the boxplots are not close to the true fold change.
-After adjustment all three models generally have median near the true fold change.
+After adjustment all three models generally have a median near the true fold change.
 In terms of model comparison, the proposed method shows a much tighter distribution around the true fold change, whole both Limma and $t$-test are much wider.
@@
 In this experiment Shigella ubiquitin ligase IpaH7.8 was shown to inhibit the protein gasdermin D (GSDMD).
-Multiplex proteomics was used to quantify the abundance of total protein, and ubiquitination in human epithelial cells.
+\cite{LUCHETTI2021} Multiplex proteomics was used to quantify the abundance of total protein, and ubiquitination in human epithelial cells.
 Cells were either infected or uninfected with IpaH7.8-deficient Shigella flexneri and measurements were taken at different time periods.
@@
 After adjustment the fold change is much larger and the adjusted pvalue is very significant.
-In this case the proposed method allows us to identify differential modified peptides that could have otherwise been missed.}
+In this case the proposed method allows us to identify a differential modified peptide that could have otherwise been missed.}
  \end{subfigure}
@@
 \label{sec:shigella}
+
 In this study, the correlation between the gene Atg16L1 and killing of Shigella flexneri (S.flexneri) was assessed.
@@
 In this case the profile plot indicates the effect of the modification is masked by the change in global protein abundance.
-Additionally this profile plot shows the large difference in available features between modifications and global protein.}
+Additionally this profile plot shows the large difference in available features between modified peptides and global proteins.}
  \end{subfigure}
@@
 Of the 10,799 modified peptides identified, only 4526 had a corresponding unmodified run and could be adjusted.
-Finally, using this method also resulted in a low feature count for the unmodified protein model.
+Not having a separate global profiling run resulted in very low feature counts for the unmodified protein model.
 
@@
 \caption{Comparing the global profiling of protein $P52209$ with the modification of the protein at site $K059$.
-The modification appears generally unchanged between conditions, whereas the global profiling run shows the CCCP and Combo conditions at a higher relative abundance compared to the Control and USP30\_OE.
-This indicates that the modification actually had a major effect when comparing CCCP and Combo to Control and USP30\_OE, which would have been missed without adjusting for global protein changes.}
+The modification appears generally unchanged between all conditions, whereas the global profiling run shows the CCCP and Combo conditions have a higher relative abundance compared to the Control and USP30\_OE.
+This indicates that the modification actually had a major effect when comparing CCCP and Combo to Control and USP30\_OE.
+This would have been missed without adjusting for global protein changes.}
 \label{fig:USP30_profile_plot}
@@
 As the standard deviation increases the number of replicates increases greatly.
-When the unmodified stand deviation is much higher than the modified, we can see that many more replicates are needed to reach the same power.}
+When the unmodified standard deviation is much higher than the modified, we can see that many more replicates are needed to reach the same power.}
 \label{fig:power_sd_combo}

### 2021-09-29 main/MSstatsPTM_Main.tex
Made updates to both main and supp

@@
 
-\title{Statistical methods for relative quantification of post-translational modifications in global proteomics experiments}
+\title{MSstatsPTM statistical relative quantification of post-translational modifications in global proteomics experiments}
 
@@
 Finally, the PTM abundances are adjusted to remove variance from changes in the overall protein.
-We implement this model in the free and open-source R package MSstatsPTM.
+We implement this model in the free and open-source R package $MSstatsPTM$.
 
@@
 
+We propose a general statistical approach, which explicitly characterizes the variations and confounding factors present in bottom-up PTM experiments.
+The proposed approach is aimed at the detection of quantitative changes in PTMs between conditions utilizing procedures developed for summarization of LC-MS data, quantitative characterization of site-specific PTMs, and adjustment with respect to protein abundance.
+Quantitative analyses of PTMs often involve comparisons between multiple inter-related conditions of the same biological system.
+The general statistical framework underlying the proposed approach allows for analyzing experiments with complex designs and different acquisition methods.
+
+The proposed approach was evaluated using datasets from computer simulations, benchmark controlled mixtures, and biological investigations.
+The approach was compared against the commonly applied t-test and limma methods.
+The results demonstrate that by appropriately leveraging the information from the entire dataset, the proposed approach improves the reproducibility and accuracy of the estimates of PTM fold changes and results in a better calibrated type I error rate.
+Finally, the proposed approach is implemented as a freely available open source R package $MSstatsPTM$, available on Bioconductor, which employs similar input format as in $MSstats$ and $MSstatsTMT$, and is compatible with many acquisition methods, such as label free, DIA, DDA, and TMT.
+\cite{Choi:2014} \cite{Huang:2020}
+
+\section{Experimental Procedure}
+
+This section describes the datasets and experiments used to evaluate the proposed method. Further details can be seen in Supplemental Sections \ref{supp-sec:sim} and \ref{supp-sec:bio_investigations}
+
+\subsection*{Computer simulations}
+\label{sec:comp_sim_procedure}
+
+The proposed statistical approach methods were evaluated using two computer simulations, with known ground truth.
+Specifically, their properties under adjustment with respect to protein abundance were evaluated.
+Two simulations were ran, one including a high number of features and no missing values, and one with few modified features and including missing values.
+
+Differential intensity levels of modified peptides may be due to changes in modification, change in protein abundance, or both.
+The proposed approach adjusts the abundance with respect to unmodified peptides by combining the inference of modified and unmodified peptide abundances.
+Alternatively, two-sample t-test or limma that takes as input the ratio between modified and unmodified peptide intensities (difference on log scale) is commonly applied for the same purpose.
+In real experiments, multiple inter-related conditions are often compared together.
+Whereas t-test uses measurements from the two conditions being compared, the proposed approach and limma leverage measurements in all conditions for the inference of underlying abundance.
+We evaluated the impact with such adjustment by computer simulation, considering the following factors: with/without protein-level changes, effect size, number of replicates and number of conditions.
+Further details of how the simulations were generated can be found in Supplementary Sec.
+\ref{supp-sec:sim}.
+
+\paragraph*{Computer simulation 1 - label free}
+
+In this simulation, a clean dataset was generated with a high number of modified and unmodified peptide features and no missing values.
+Multiple simulations were generated with these parameters, with different numbers of replicates and conditions.
+1000 modified peptides were generated, half of which were true positives and the other half true negatives.
+Of true positives simulated, half of them had their abundance's masked by changes in global protein.
+Modified peptides and unmodified proteins were both simulated with 10 features.
+
+\paragraph*{Computer simulation 2 - missing values and low replicates}
+
+In this simulation real world experimental conditions were introduced, including missing values and low number of modified features.
+Missing values and feature count values for this simulation were determined by looking at the averages from the biological experiments in this paper.
+Modified peptides were simulated with 2 features, while unmodified proteins were still simulated with 10 features.
+Additionally, 20\% of observations for both modified and unmodified peptides were masked completely at random.
+Adding missing values and few representative modified features provides a more realistic expectation of model performance.
+
+\subsection*{SpikeIn benchmark - KGG enriched}
+
+We evaluated our approach using a custom designed spike-in benchmark experiment where 50 heavy-labeled KGG motif peptides from 20 human proteins were used as spike-in peptides.
+Quantitative changes in protein and site abundance changes of these 20 proteins were the target of the benchmark.
+Unmodified peptides from Human Lysate were used as the estimate of global protein abundance changes.
+All comparisons with respect to human lysate serve as a null model, there was no change in global protein nor PTM abundance between conditions.
+Additionally, E coli Lysate was used to normalize total protein levels prior to enrichment or global protein profiling.
+Four mixes of spike in peptides and Human Lysate were created to create conditions with known fold change.
+Two sets of data were acquired for each mixture: modified peptide data, including the spike in KGG enriched peptides, and unmodified peptides from Human Lysate.
+
+\subsection*{Human-1mix-TMT - ubiquitination}
+
+The proposed approach was evaluated on an experiment where Shigella ubiquitin ligase IpaH7.8 was shown to inhibit the protein gasdermin D (GSDMD).\cite{LUCHETTI2021} Multiplex proteomics was used to quantify the abundance of total protein, and ubiquitination in human epithelial cells.
+Cells were either infected or uninfected with IpaH7.8-deficient Shigella flexneri and measurements were taken at different time periods.
+Uninfected cells were measured at 0 and 6 hours, while infected cells were measured at 1, 2, 4, and 6 hour increments, resulting in six total conditions.
+The experiment was unbalanced with two bioreplicates per condition for all conditions except for infected 1 hour.
+About ~95\% of the identified modified peptides derived from proteins that were quantified in the global profiling run.
+
+\subsection*{Mouse-2mix-TMT - phosphorylation}
+
+Here an experiment targeting primary murine macrophages infected with Shigella flexneri (S.flexneri).
+\cite{Maculins} Multiplex proteomics was used to quantify the abundance of total protein, and phosphorylation in wild type (WT) and ATG16L1-deficient (cKO) samples, uninfected and uninfected with S.flexneri.
+The abundance of total protein and post-translation modifications were quantified at three time points, uninfected, early infection (45-60 minutes), and late infection (3-3.5 hours).
+Quantifying the total protein along with the post-translational modifications allowed us to adjust for changes in total protein and see the true impact of the site specific modifications.
+Two mixtures using 11-plex were ran over the six conditions.
+The six conditions were split between 11 channels leading to the experimental design being unbalanced.
+Each mixture contained two replicates per early and late WT and KO conditions.
+Mixture one contained one replicate of uninfected WT and two replicates of uninfected KO.
+Mixture two contained one replicate of uninfected KO and two uninfected WT.
+About ~90\% of the identified modified peptides derived from proteins that were also quantified in the global profiling run.
+
+\subsection*{Human label free quantification - no global profiling run}
+
+This experiment looked into the relationship between USP30 and protein kinase PINK1, and their association with Parkinson’s Disease.
+Ubiquitination site profiling was performed and the modified site abundance was analyzed.
+Four conditions were tested with two biological replicates per condition.
+The conditions were as follows: CCCP, USP30 over expression (USP30 OE), Combo, and Control.
+Label-free mass spectrometry quantification was used to quantify the abundance of modified peptides.
+A corresponding mixed effects model was fit per modification and global protein as described previously.
+In contrast to the other biological experiments assessed in this paper, this experiment did not not have a separate global profiling run for the unmodified protein.
+In this case $MSstatsPTM$ can still be used by extracting unmodified peptides from the modified run, however this generally leads to substantial less matches between modified and unmodified peptides, in addition to low feature counts for unmodified peptides.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section*{Results}
+
+\subsection*{Existing statistical methods for experiments targeting PTMs}
+
 Despite the important implications of PTMs in biological functions, there is a lack of general framework to summarize the available quantitative information from LC-MS data, to perform statistical inference, and to draw conclusions to characterize the quantitative properties of PTM in a statistically rigorous manner.
@@
 The modeling framework, however, is not applicable for either label-free workflows or experiments with complex designs.
+Further details on existing methods can be seen in Supplementary \ref{supp-sec:ttest}.
 
-We propose a general statistical approach, which explicitly characterizes the variations and confounding factors present in bottom-up PTM experiments.
-The proposed approach is aimed at the detection of quantitative changes in PTMs between conditions utilizing procedures developed for summarization of LC-MS data, quantitative characterization of site-specific PTMs, and adjustment with respect to protein abundance.
-Quantitative analyses of PTMs often involve comparisons between multiple inter-related conditions of the same biological system.
-The general statistical framework underlying the proposed approach allows for analyzing experiments with complex designs and different acquisition methods, including label-free, DDA, DIA, and TMT.
-
-The proposed approach was evaluated using datasets from computer simulations, benchmark controlled mixtures, and biological investigations.
-The approach was compared against the commonly applied t-test and limma methods.
-The results demonstrate that by appropriately leveraging the information from the entire dataset, the proposed approach improves the reproducibility and accuracy of the estimates of PTM fold changes and results in a better calibrated type I error rate.
-The proposed approach is implemented as an open source R package MSstatsPTM, which employs similar input format as in MSstats and MSstatsTMT.
-\cite{Choi:2014} \cite{Huang:2020}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Proposed Approach}
+\subsection*{Proposed approach}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection*{Statistical modeling and parameter estimation}
+\subsubsection*{Statistical modeling and parameter estimation}
 
@@
 In addition mixture may be included for features acquired via tandem mass tag (TMT) methods.
-The model parameters are estimated using the split-plot approach as in MSstats, where the feature log-intensities are first summarized into a single value per site per run in the subplot model, and the site-level summaries are then used for the inference of the PTM site abundance.\cite{Choi:2014} In the site-level summarization, Tukey's median polish (TMP), a simple and robust procedure is applied to iteratively fit a two-way additive model with the effects of run and feature, which in turn summarizes the log-intensities for each site.\cite{Tukey:1977} After summarization, the inference of the PTM site abundance in each condition is carried through fitting a model based on the family of linear mixed-effects models, taking into account the specific experimental design \ref{fig:data-structure}.\cite{Bolker2009} \cite{Faraway:2006} Statistical modeling and quantification for global proteomics data are performed by the same procedure as for PTM data.
+The model parameters are estimated using the split-plot approach as in $MSstats$, where the feature log-intensities are first summarized into a single value per site per run in the subplot model, and the site-level summaries are then used for the inference of the PTM site abundance.\cite{Choi:2014} In the site-level summarization, Tukey's median polish (TMP), a simple and robust procedure is applied to iteratively fit a two-way additive model with the effects of run and feature, which in turn summarizes the log-intensities for each site.\cite{Tukey:1977} After summarization, the inference of the PTM site abundance in each condition is carried through fitting a model based on the family of linear mixed-effects models, taking into account the specific experimental design \ref{fig:data-structure}.\cite{Bolker2009} \cite{Faraway:2006} Statistical modeling and quantification for global proteomics data are performed by the same procedure as for PTM data.
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection*{Detection of changes in PTMs}
+\subsubsection*{Detection of changes in PTMs}
+\label{sec:adjustment_summary}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection*{Missing Value imputation}
+\subsubsection*{Missing value imputation}
+\label{sec:missing_imp}
 
-MSstatsPTM gives the ability to impute missing feature intensities if desired.
+$MSstatsPTM$ gives the ability to impute missing feature intensities if desired.
 When values are imputed, it is assumed they are missing for reasons of low abundance.
-Missing features are imputed in each MS run using the Accelerated Time Failure model.\cite{Tukey:1977} In order to impute a feature's missing values, the feature must be present in at least one MS Run.
+Missing features are imputed in each MS run using the Accelerated Time Failure (AFT) model.\cite{Tukey:1977} In order to impute a feature's missing values, the feature must be present in at least one MS Run.
 If the feature is not present across all MS Runs, the value will be left missing.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection*{Extension to TMT experiments}
+\subsubsection*{Extension to TMT experiments}
 
@@
 To account for mixture variance, first feature intensities are summarized again using Tukey's Median Polish and missing values are imputed, if desired, then a new linear mixed effects model with an added term for the mixture is included.
-The workflow and statisical model described generally follow the methods used for modeling proteins in MSstatsTMT.
+The workflow and statisical model described generally follow the methods used for modeling proteins in $MSstatsTMT$.
 \cite{Huang:2020} These methods are repeated to quantify both modified and unmodified peptide abundance.
@@
 
+\subsection*{Implementation}
+
+The proposed methods are implemented in the open source R package $MSstatsPTM$, available on Bioconductor.
+$MSstatsPTM$ includes converters for multiple spectral processing tools, including MaxQuant, Progenesis, and Spectronaut.
+The converters take as input the raw data from the tool, identify the modification site for modified peptide, and put the data into the correct format for analysis in $MSstatsPTM$.
+Conversion is done separately for the modified and global profiling runs and combined together for input into the rest of the package.
+If the global profiling run is not available, the package can still analyze the modified run, but will do so without adjusting for changes in unmodified protein abundance.
+
+After using the converters, the next step is peptide/protein summarization and missing value imputation.
+For the modified run, the package will only summarize features, PSMs, which include the same modification at the same site.
+Features with multiple modification sites are not included with single site features and are summarized separately.
+For the global profiling run all unmodified features from the same protein are summarized together, up to the protein level.
+Additionally, the summarization includes global median normalization and normalizes between MS runs.
+As mentioned in Section \ref{sec:missing_imp}, the package uses an AFT model to impute missing values in this step, although this step is optional.
+
+The final step of the package is modeling the summarized dataset.
+A linear mixed effects model is automatically fit for both the summarized modified and global profiling runs.
+This model is automatically adjusted depending on the experimental design and acquisition method.
+The comparisons of interest can either be predefined or a full pairwise comparison will be tested.
+After modeling the modified and unmodified data, the modified model is adjusted for changes in unmodified protein abundance, using the methods described in Section \ref{sec:adjustment_summary}.
+
+Beyond the core conversion, summary, and modeling functionality, the package also includes two plotting functions.
+The first creates plots for the summarization data to help with quality assurance analysis and identifying sources of variation.
+This function plots a quality control plot, summarizing the peptide abundance per run in the form of a boxplot, and profile plot, plotting each feature and the overall feature summarization as a line plot over each run.
+The second function plots the modeling results and includes a volcano plot, showing all peptides adjusted pvalues and fold changes, as well as a heatmap, which evaluates the fold change between conditions and peptides.
+
+The package relies on functionalities from the R packages $MSstats$ \cite{Choi:2014} and $MSstatsTMT$ \cite{Huang:2020}, depending on the data acquisition type.
+The statistical modeling relies on the functionality from the R packages $lme4$ \cite{Bates2015} and $lmerTest$ \cite{Kuznetsova2017}.
+All datasets in this paper were evaluated using the $MSstatsPTM$ package.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Results}
+\subsection*{Evaluation}
 
-\subsection*{Computer simulations}
-\label{sec:comp_sim}
+The performance of of the proposed method was evaluated on simulated and spike in datasets with known ground truth, as well as biological experimental data where the ground truth was not known.
+For the experiments where ground truth is known, we calculated summary statistics using the number of true positives (TP), false positives (FP), true negatives (TN), and false negatives (FN).
+To evaluate the performance of different methods, we considered the false discovery rate (FDR = FP/(TP+FP)), the recall rate (TP/(TP + FN), and total accuracy ((TP + TN) / (TP + TN + FP + FN)).
+For biological experiments where ground truth was not known, we evaluated significantly differential peptides with an $\alpha$ of .05.
+Comparisons were made before and after adjusting for changes in unmodified protein abundance and the results are presented in the form of venn diagrams and profile plots.
 
-The proposed statistical approach methods were evaluated using computer simulation.
-Specifically, their properties under adjustment with respect to protein abundance were evaluated.
-Two simulations were ran, one including a high number of features and no missing values, and one with few modified features and including missing values.
+\subsubsection*{Computer simulation 1 - label free}
 
-Differential intensity levels of modified peptides may be due to changes in modification, change in protein abundance, or both.
-The proposed approach adjusts the abundance with respect to unmodified peptides by combining the inference of modified and unmodified peptide abundances.
-Alternatively, two-sample t-test or limma that takes as input the ratio between modified and unmodified peptide intensities (difference on log scale) is commonly applied for the same purpose.
-In real experiments, multiple inter-related conditions are often compared together.
-Whereas t-test uses measurements from the two conditions being compared, the proposed approach and limma leverage measurements in all conditions for the inference of underlying abundance.
-We evaluated the impact with such adjustment by computer simulation, considering the following factors: with/without protein-level changes, effect size, number of replicates and number of conditions.
-Further details of how the simulations were generated and their results can be found in Supplementary Sec.
-\ref{supp-sec:sim}.
-
-\paragraph*{Computer simulation 1}
-
-In this simulation, a clean dataset was generated with a high number of modified and unmodified peptide features and no missing values.
-Not accounting for changes in unmodified peptide level resulted in a high false discovery rate for all methods, seen in Figure \ref{fig:fdr_all_models}.
+In this simulation, not accounting for changes in unmodified peptide level resulted in a high false discovery rate for all methods, seen in Figure \ref{fig:fdr_all_models}.
 When adjusting for changes in the unmodified peptide, all methods performed similarly in terms of FDR.
@@
 
-\paragraph*{Computer simulation 2}
+\subsubsection*{Computer simulation 2 - missing values and low replicates}
 
-In this simulation real world experimental conditions were introduced, including missing values and low number of modified features.
-In this case, changes in unmodified peptide level still needed to be accounted for to control the FDR in all methods.
+Changes in unmodified peptide level still needed to be accounted for to control the FDR in all methods.
 Once controlling for changes in unmodified peptide, it is clear that the proposed method performed stronger across all methods when compared to limma and t-test, as seen in Figure \ref{fig:sim_accuracy}.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection*{Spike-in experiment}
-
-We evaluated our approach using a custom designed spike-in benchmark experiment where 50 heavy-labeled KGG motif peptides from 20 human proteins were used as spike-in peptides.
-Quantitative changes in protein and site abundance changes of these 20 proteins were the target of the benchmark.
-Unmodified peptides from Human Lysate were used as the estimate of global protein abundance changes.
-All comparisons with respect to human lysate serve as a null model, there was no change in global protein nor PTM abundance between conditions.
-Additionally, E coli Lysate was used to normalize total protein levels prior to enrichment or global protein profiling.
-Four mixes of spike in peptides and Human Lysate were created to create conditions with known fold change.
-Two sets of data were acquired for each mixture: modified peptide data, including the spike in KGG enriched peptides, and unmodified peptides from Human Lysate.
-These datasets were used as the PTM and global protein data.
+\subsubsection*{SpikeIn benchmark - KGG enriched}
 
@@
 
-In Figure \ref{fig:spikein_limma_volcano} we can see the results of the same experiment modeled using limma with protein adjustment discussed in Section~\ref{sec:comp_sim}.
+In Figure \ref{fig:spikein_limma_volcano} we can see the results of the same experiment modeled using limma with protein adjustment.
 Again the red labeled spike-in peptides are more inline with expected fold change after adjustment is applied.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection*{Re-analysis of published datasets}
-
-\subsubsection*{IpaH7.8}
-
-The proposed approach was evaluated on an experiment where Shigella ubiquitin ligase IpaH7.8 was shown to inhibit the protein gasdermin D (GSDMD).\cite{LUCHETTI2021} Multiplex proteomics was used to quantify the abundance of total protein, and ubiquitination in human epithelial cells.
-Cells were either infected or uninfected with IpaH7.8-deficient Shigella flexneri and measurements were taken at different time periods.
-Uninfected cells were measured at 0 and 6 hours, while infected cells were measured at 1, 2, 4, and 6 hour increments, resulting in six total conditions.
-The experiment was unbalanced with two bioreplicates per condition for all conditions except for infected 1 hour.
-About ~95\% of the identified modified peptides derived from proteins that were quantified in the global profiling run.
+\subsubsection*{Human-1mix-TMT - ubiquitination}
 
@@
 
-\subsubsection*{Shigella flexneri}
-
-To evaluate our approach we reanalyzed an experiment targeting primary murine macrophages infected with Shigella flexneri (S.flexneri).
-\cite{Maculins} Multiplex proteomics was used to quantify the abundance of total protein, phosphorylation, and ubiquitination in wild type (WT) and ATG16L1-deficient (cKO) samples, uninfected and uninfected with S.flexneri.
-The abundance of total protein and post-translation modifications were quantified at three time points, uninfected, early infection (45-60 minutes), and late infection (3-3.5 hours).
-Quantifying the total protein along with the post-translational modifications allowed us to adjust for changes in total protein and see the true impact of the site specific modifications.
-Two mixtures using 11-plex were ran over the six conditions.
-The six conditions were split between 11 channels leading to the experimental design being unbalanced.
-Each mixture contained two replicates per early and late WT and KO conditions.
-Mixture one contained one replicate of uninfected WT and two replicates of uninfected KO.
-Mixture two contained one replicate of uninfected KO and two uninfected WT.
-About ~90\% of the identified modified peptides derived from proteins that were also quantified in the global profiling run.
+\subsubsection*{Mouse-2mix-TMT - phosphorylation}
 
@@
 
-\subsubsection*{USP30}
+\subsubsection*{Human label free quantification - no global profiling run}
 
-This experiment looked into the relationship between USP30 and protein kinase PINK1, and their association with Parkinson’s Disease.
-Ubiquitination site profiling was performed and the modified site abundance was analyzed.
-Four conditions were tested with two biological replicates per condition.
-The conditions were as follows: CCCP, USP30 over expression (USP30 OE), Combo, and Control.
-Label-free mass spectrometry quantification was used to quantify the abundance of modified peptides.
-A corresponding mixed effects model was fit per modification and global protein as described previously.
-
-In contrast to other experiments analyzed in this paper, there was no unmodified global protein profiling run performed in this experiment.
+As discussed previously, there was no unmodified global profiling run performed in this experiment.
 Once identification and quantification of the Ubiquitinated profiling was performed, peptides which were unmodified were extracted and used in place of a global profiling run.
@@
 
+\subsubsection*{Sample size calculation}
+
+\todo{Devon: Update this section in main and supp}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Conclusion and discussion}
+\section{Discussion}
 
@@
 We are also interested in characterizing the interplay of PTMs at multiple sites.
-The proposed statistical methods are implemented as an R package MSstatsPTM available on Bioconductor and Github.
+The proposed statistical methods are implemented as an R package $MSstatsPTM$ available on Bioconductor and Github.
 

### 2021-09-29 supplementary/ptm_sm.tex
Made updates to both main and supp

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection{Simulation 1}
+\subsubsection{Computer simulation 1 - label free}
 
@@
 
-\subsubsection{Simulation 2}
+\subsubsection{Computer simulation 2 - missing values and low replicates}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Benchmark experiment}
+\subsection{SpikeIn benchmark - KGG enriched}
 \label{sec:benchmark}
@@
 \section{Datasets : Biological investigation}
-
-\subsection{IpaH7.8}
+\label{sec:bio_investigations}
+
+\subsection{Human-1mix-TMT - ubiquitination}
 \label{sec:ipah}
@@
 
-The results of the proposed method to this experiment can be seen in \sfigref{fig:ipah_figures}.
+The results of the proposed method to this experiment can be seen in \sfigref{fig:ipha_venn_d} and \sfigref{fig:ipah_figures}.
+
+\begin{figure}[h!]
+\centering
+\includegraphics[height=.7\textwidth]{sim_new/ipah_venn_diagramm.png}
+\caption{The overlap of differencial modified peptides between the PTM model with and without global protein level adjustment.
+More PTMs became insignificant after adjustment then became significant.
+This indicates that for the peptides that became insignificant in the adjusted model, their change in abundance was mainly due to changes in global protein abundance.
+An additional question that must be addressed is if the decrease in significant peptides is due to the increased variance that comes from adjustment.
+This was tested by looking for modified peptides who's adjusted log fold change was within 10\% of the unadjusted log fold change but became insignificant after adjustment.
+When this test was applied on this experiment, only one peptide became insignificant due to an increase in variance.
+Thus we can conclude that the drop off in significant peptides was truly due to changes in global protein abundance.}
+\label{fig:ipha_venn_d}
+\end{figure}
 
@@
 \clearpage
-\subsection{Shigella flexneri}
+\subsection{Mouse-2mix-TMT - phosphorylation}
 \label{sec:shigella}
@@
 
-The results of the proposed method to this experiment can be seen in \sfigref{fig:No_Diff_Shigella_PTM} and \sfigref{fig:Diff_Shigella_PTM}.
+The results of the proposed method to this experiment can be seen in \sfigref{fig:shig_venn_d},  \sfigref{fig:No_Diff_Shigella_PTM}, and \sfigref{fig:Diff_Shigella_PTM}.
+
+\begin{figure}[h!]
+\centering
+\includegraphics[height=.7\textwidth]{sim_new/shig_venn_diagramm.png}
+\caption{The overlap of differencial modified peptides between the PTM model with and without global protein level adjustment.
+Much like in \sfigref{fig:ipha_venn_d}, more PTMs became insignificant after adjustment then became significant.
+This indicates that for the peptides that became insignificant in the adjusted model, their change in abundance was mainly due to changes in global protein abundance.
+Again we tested if the decrease in significant peptides was due to the increased variance that comes from adjustment.
+This was tested by looking for modified peptides who's adjusted log fold change was within 10\% of the unadjusted log fold change but became insignificant after adjustment.
+When this test was applied on this experiment, 776 peptides became insignificant due to an increase in variance.
+This is higher than in \sfigref{fig:ipha_venn_d}, however it is still a very small portion (4\%) of the peptides that became insignificant after adjustment.
+Thus we can conclude that the drop off in significant peptides was mainly due to changes in global protein abundance.}
+\label{fig:shig_venn_d}
+\end{figure}
 
@@
 
-\subsection{USP30}
+\subsection{Human label free quantification - no global profiling run}
 \label{sec:usp30}
@@
 \section{Sample size calculation and power analysis}
+\todo{Devon: Update this section per Olga's commentsv}
 

### 2021-09-29 code/IpaH_analysis.R

@@ -4,6 +4,7 @@ library(data.table)
 library(tidyverse)
 library(gridExtra)
 library(ggrepel)
+library(VennDiagram)
 
 setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
 
@@ -179,3 +180,46 @@ v2 <- adj_plot_df %>% ggplot() +
     fill="#69b3a2")
 
 grid.arrange(v1, v2, nrow = 1)
+
+
+## Venn Diagram
+colors <- c("#E69F00", "#56B4E9")
+
+unadj_ptm_model <- read.csv("../data/Ipah_PTM_test_results.txt", sep = "\t")
+adj_ptm_model <- read.csv("../data/Ipah_PTM_Adjusted_test_results.txt", sep = "\t")
+
+unadj_ptm_model$Protein_Label <- paste(unadj_ptm_model$Protein, unadj_ptm_model$Label)
+adj_ptm_model$Protein_Label <- paste(adj_ptm_model$Protein, adj_ptm_model$Label)
+
+sig_unadj_ptm_model <- unadj_ptm_model %>% filter(Protein_Label %in% adj_ptm_model$Protein_Label)
+
+sig_unadj_ptm_model <- sig_unadj_ptm_model %>% filter(adj.pvalue < .05)
+sig_adj_ptm_model <- adj_ptm_model %>% filter(adj.pvalue < .05)
+
+
+venn.diagram(
+  x = list(sig_unadj_ptm_model$Protein_Label, sig_adj_ptm_model$Protein_Label),
+  category.names = c("Unadjusted" , "Adjusted"),
+  filename = "ipah_venn_diagramm.png",
+  output=TRUE,
+  imagetype="png" ,
+  height = 1200,
+  width = 1200,
+  resolution = 100,
+  lwd = 2,
+  fill = colors,
+  main.fontface = "bold",
+  main.pos = c(.5,.965),
+  fontface = "bold",
+  cat.fontface = "bold",
+  cex = 3,
+  cat.cex = 3,
+  cat.pos = c(-40, 30),
+  cat.dist = c(.037, .03),
+  main.cex = 2.8,
+  main = "Overlap between signficant adjusted and unadjusted PTMs"
+)
+
+combined_models <- merge(adj_ptm_model, unadj_ptm_model, all.x=TRUE, all.y=TRUE, by = c("Protein_Label"))
[…]

### 2021-09-30 supplementary/ptm_sm.tex
added venn diagrams to supplementary

@@
 \centering
-\includegraphics[height=.7\textwidth]{sim_new/ipah_venn_diagramm.png}
+\includegraphics[height=.65\textwidth]{sim_new/ipah_venn_diagramm.png}
 \caption{The overlap of differencial modified peptides between the PTM model with and without global protein level adjustment.
@@
 \centering
-\includegraphics[height=.7\textwidth]{sim_new/shig_venn_diagramm.png}
+\includegraphics[height=.65\textwidth]{sim_new/shig_venn_diagramm.png}
 \caption{The overlap of differencial modified peptides between the PTM model with and without global protein level adjustment.
@@
 
-An example profile plot for this experiment can be seen in \sfigref{fig:USP30_profile_plot}.
+An example profile plot for this experiment can be seen in \sfigref{fig:usp30_ven_d} and \sfigref{fig:USP30_profile_plot}.
+
+\begin{figure}[h!]
+\centering
+ \begin{subfigure}{\textwidth}
+ \centering
+	\includegraphics[height=.5\textwidth]{sim_new/usp30_venn_diagramm}
+	\caption{The overlap of differencial modified peptides between the PTM model with and without global protein level adjustment.In this experiment, many more PTMs became insignificant after adjustment then became significant.
+In this case this is mainly due to the large lack of overlap between the modified peptides and unmodified proteins, stemming from not having a global profiling run.}
+ \end{subfigure}\vspace{-5mm}
+ \begin{subfigure}{\textwidth}
+ \centering
+	\includegraphics[height=.5\textwidth]{sim_new/usp30_venn_diagramm_matching_only}
+	\caption{The same overlap as above, but looking only at modified peptides with a matching unmodified protein so adjustment can be performed.
+In this case we see significantly less peptides become insignificant after adjustment.
+Again we can check if peptides are becoming insignificant due to an increase in variance.
+To check this we look at modified peptides whose adjusted fold change is within 10\% of their unadjusted fold change and they become insignificant after adjustment.
+In this experiment there are only 25 peptides that meet this criteria, a very small percentage of all modified peptides.}
+ \end{subfigure}
+ \caption{Overlap of significant PTMs before and after protein adjustment.}
+\label{fig:usp30_ven_d}
+\end{figure}
 
@@
 The proposed approach allows us to calculate the sample size needed to achieve a desired statistical power as described in Section \ref{sec:design}.
-Given a desired power of .9, varying standard deviations, and an expected fold change, the number of replicates per condition can be seen in \sfigref{fig:sample_size}
-
-\begin{figure}[h!]
-\centering
-\includegraphics[width=.9\textwidth]{sim_new/simple_power_analysis}
-\caption{The number of replicates per condition required to reach a power of .9 can be seen in the plot.
-At a high standard deviation and low expected fold change the number of replicates is unrealistically high.
-As the expected fold change increases the number of replicates needed decreases dramatically.
-The standard deviation of .1 expectantly shows the least number of replicates needed regardless of fold change.}
-\label{fig:sample_size}
-\end{figure}
-
 As described in Section \ref{sec:test} the proposed method adjusts for the underlying protein abundance in the PTM significance analysis, which corrects the confounding factor with a cost of increased variation.
 When the variation is increased, the experiment requires a larger number of replicates to reach the same power.
-
-We compared the required sample size with varying standard deviations for the modified and unmodified peptide models.
-Three pairs of standard deviations of log-intensities for modified and unmodified peptides were used: (0.2, 0.1), (0.2, 0.2), and (0.2, 0.4).
-In the first pair the standard deviation for the unmodified peptide was less than the modified, in the second the standard deviations are the same, and in the third the standard deviation for the unmodified peptide was more than for the modified peptide.
-Again the desired power was .9 and the expected fold change was varied.
+Thus the statistical power is dependent on the variance from both the modified peptide and unmodifed protein as well as the number of replicates in both the modified and unmodified runs.
+
+We compared the statistical power in experiments with different numbers of replicates, variance, and fold change for both the modified and unmodified runs.
 The results can be seen in \sfigref{fig:power_sd_combo}
@@
 \includegraphics[width=.9\textwidth]{sim_new/power_analysis_sd_combined}
-\caption{When the unmodified peptide has a low standard deviation the number of replicates is very similar to the baseline.
-As the standard deviation increases the number of replicates increases greatly.
-When the unmodified standard deviation is much higher than the modified, we can see that many more replicates are needed to reach the same power.}
+\caption{}
 \label{fig:power_sd_combo}

### 2021-09-30 code/USP30_MSstatsPTM_Analysis.R

@@ -17,16 +17,16 @@ global_prot <- unique(model_ptm$PROTEIN.Model$Protein)
 match_list <- c()
 for (i in seq_along(global_prot)){
   for (y in seq_along(original_ptms))
-    if (grepl(global_prot[[i]], original_ptms[[y]]) & 
+    if (grepl(global_prot[[i]], original_ptms[[y]]) &
         !global_prot[[i]] %in% match_list) {
       match_list <- c(match_list, global_prot[[i]])
     }
 }
 
-data.frame("Number" = c(original_ptms, adj_ptms, global_prot), Label = c("Total PTMs", "Matching PTMs", "Global Proteins")) %>% 
+data.frame("Number" = c(original_ptms, adj_ptms, global_prot), Label = c("Total PTMs", "Matching PTMs", "Global Proteins")) %>%
   ggplot() + geom_col(aes(y = Number, x = Label))
 
-prot <- sapply(original_ptms, function(x){substring(str_extract_all(x, "^(.*?)_")[[1]], 
+prot <- sapply(original_ptms, function(x){substring(str_extract_all(x, "^(.*?)_")[[1]],
                                                     1, nchar(str_extract_all(x, "^(.*?)_")[[1]]) - 1)})
 length(intersect(unique(prot), global_prot))
 
@@ -49,13 +49,13 @@ for (i in seq_len(nrow(pot))){
 
   temp <- summarized_ptm$PTM$FeatureLevelData %>% filter(PROTEIN == pot[i, Protein])
   num <- length(unique(temp$FEATURE))
-  
+
   if (num >= 2){
-    temp_p <- summarized_ptm$PROTEIN$FeatureLevelData %>% 
+    temp_p <- summarized_ptm$PROTEIN$FeatureLevelData %>%
       filter(PROTEIN == pot[i, GlobalProtein])
-    
+
     num2 <- length(unique(temp_p$FEATURE))
-    
+
     if (num2 >= 2){
       print(num)
       print(num2)
@@ -67,46 +67,46 @@ for (i in seq_len(nrow(pot))){
 
 ptm_list1 <- c("P52209_K059")
 
-# pSTY_features$BioReplicate <- factor(pSTY_features$BioReplicate, 
-#                                      levels = c("WT_Uninfect_1", "WT_Uninfect_2", "WT_Uninfect_3" , "WT_Uninfect_4", 
+# pSTY_features$BioReplicate <- factor(pSTY_features$BioReplicate,
[…]

### 2021-09-30 code/sample_size_analysis.R

@@ -9,18 +9,52 @@ desiredFC <- 2 ^ delta
 # desiredFC <- delta
 
 ## Define base metrics
-FDR <- .05
-power <- .8
 
-## function calculations
-m0_m1 <- 99 ## it means m0/m1=99, m0/(m0+m1)=0.99
-alpha <- power * FDR / (1 + (1 - FDR) * m0_m1)
+low <- 1
+high <- 3
 
-z_alpha <- qnorm(1 - alpha / 2)
-z_beta <- qnorm(power)
-aa <- (delta / (z_alpha + z_beta)) ^2
+mod_pep_sigma <- low
+unmod_pep_sigma <- low
+
+n_sample_list <- c(3,5,7,10)
+
+mod_num_sample <- n_sample_list[[1]]
+unmod_num_sample <- n_sample_list[[1]]
 
 #Specify sigma for ptm and unmod pep
+
+samples_mod <- rnorm(n_sample_list[[1]], mean = 0, low)
+std_dev_mod <- sum((samples_mod - mean(samples_mod)^2)) / (mod_num_sample)
+samples_unmod <- rnorm(n_sample_list[[1]], mean = 0, low)
+std_dev_unmod <- sum((samples_unmod - mean(samples_unmod)^2)) / (unmod_num_sample)
+
+t <- delta / sqrt(std_dev_mod^2 + std_dev_unmod^2)
+
+FDR <- 0.05
+m0_m1 <- 99
+powerTemp <- seq(0, 1, 0.01)
+power <- numeric(length(t))
+
+for (i in seq_along(t)) {
+  diff <- qnorm(powerTemp) + qnorm(1 - powerTemp * FDR / (1 + (1 - FDR) * m0_m1) / 2) - t[i]
+  min(abs(diff), na.rm = TRUE)
+  power[i] = powerTemp[order(abs(diff))][1]
+}
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
 
 
 ## Original code containing sigma for subject.. can include or assume 0 for now
@@ -32,7 +66,7 @@ numSample_nomod_1 <- round(2 * (.1) / aa, 0)
 numSample_nomod_2 <- round(2 * (.2) / aa, 0)
 numSample_nomod_3 <- round(2 * (.4) / aa, 0)
 
-p1 <- data.frame("SD" = c(rep(".1", length(numSample_nomod_1)),
+p1 <- data.frame("SD" = c(rep("`.1", length(numSample_nomod_1)),
[…]

### 2021-10-04 main/MSstatsPTM_Main.tex
made adjustments to supp and main

@@
 We propose a general statistical model and workflow that is both reproducible and comprehensive.
-The model measures modified and unmodified peptide abundance by summarizing intensities through Tukey’s median polish method.
+The method measures modified and unmodified peptide abundance by summarizing intensities through Tukey’s median polish method.
 Then a model based on the family of linear mixed-effects models is fit.
@@
 Second, contrary to the often large number of identified peptides that can be used as features to model protein abundance changes, there are relatively few representative peptides that span a modification sites, which often results in sparse, and sometimes, inherently convoluted models (i.e., single versus multiple modified sites on a single peptide).
-Third, unless early signaling events are interrogated, the interpretation of the relative changes in modification occupancy are inherently convoluted with changes in the overall protein expression, making the interpretation of the results not straightforward.
-\cite{Olsen:2013} Therefore, a robust approach to estimate systematic relative changes in post-translational modifications, at scale, should not only combine the quantitative information pertaining to a PTM site over peptides and replicates in multiple conditions, but take into account various sources of variations and confounding factors present in the experiments.
+Third, unless early signaling events are interrogated, the interpretation of the relative changes in modification occupancy are inherently convoluted with changes in the overall protein expression, making the interpretation of the results not straightforward.\cite{Olsen:2013} Therefore, a robust approach to estimate systematic relative changes in post-translational modifications, at scale, should not only combine the quantitative information pertaining to a PTM site over peptides and replicates in multiple conditions, but take into account various sources of variations and confounding factors present in the experiments.
 
@@
 
-The proposed approach was evaluated using datasets from computer simulations, benchmark controlled mixtures, and biological investigations.
-The approach was compared against the commonly applied t-test and limma methods.
-The results demonstrate that by appropriately leveraging the information from the entire dataset, the proposed approach improves the reproducibility and accuracy of the estimates of PTM fold changes and results in a better calibrated type I error rate.
-Finally, the proposed approach is implemented as a freely available open source R package $MSstatsPTM$, available on Bioconductor, which employs similar input format as in $MSstats$ and $MSstatsTMT$, and is compatible with many acquisition methods, such as label free, DIA, DDA, and TMT.
-\cite{Choi:2014} \cite{Huang:2020}
+The proposed approach was implemented using the R coding language and evaluated using datasets from computer simulations, benchmark controlled mixtures, and biological investigations.
+The approach was compared against the commonly applied t-test and Limma methods.\cite{Ritchie_15a} The results demonstrate that by appropriately leveraging the information from the entire dataset, the proposed approach improves the reproducibility and accuracy of the estimates of PTM fold changes and results in a better calibrated type I error rate.
+Finally, the proposed approach is implemented as a freely available open source R package $MSstatsPTM$, available on Bioconductor, which employs similar input format as in $MSstats$ and $MSstatsTMT$, and is compatible with many acquisition methods, such as label free, DIA, DDA, and TMT.\cite{Choi:2014} \cite{Huang:2020}
 
@@
 Specifically, their properties under adjustment with respect to protein abundance were evaluated.
-Two simulations were ran, one including a high number of features and no missing values, and one with few modified features and including missing values.
+Two simulations were ran, mimicking a label free experiment, one including a high number of features and no missing values, and one with few modified features and including missing values.
 
@@
 The proposed approach adjusts the abundance with respect to unmodified peptides by combining the inference of modified and unmodified peptide abundances.
-Alternatively, two-sample t-test or limma that takes as input the ratio between modified and unmodified peptide intensities (difference on log scale) is commonly applied for the same purpose.
+Alternatively, two-sample t-test or Limma that takes as input the ratio between modified and unmodified peptide intensities (difference on log scale) is commonly applied for the same purpose.
 In real experiments, multiple inter-related conditions are often compared together.
-Whereas t-test uses measurements from the two conditions being compared, the proposed approach and limma leverage measurements in all conditions for the inference of underlying abundance.
+Whereas t-test uses measurements from the two conditions being compared, the proposed approach and Limma leverage measurements in all conditions for the inference of underlying abundance.
 We evaluated the impact with such adjustment by computer simulation, considering the following factors: with/without protein-level changes, effect size, number of replicates and number of conditions.
@@
 
-In this simulation, a clean dataset was generated with a high number of modified and unmodified peptide features and no missing values.
-Multiple simulations were generated with these parameters, with different numbers of replicates and conditions.
-1000 modified peptides were generated, half of which were true positives and the other half true negatives.
-Of true positives simulated, half of them had their abundance's masked by changes in global protein.
+Here simulations were generated with a high number of modified and unmodified peptide features and no missing values.
+Multiple simulations were generated with different numbers of replicates and conditions.
+In all simulations 1000 modified peptides were generated, half of which were true positives and the other half true negatives.
+Half of true positives and true negatives were generated with convolution with global protein abundance.
+Before adjustment, the convoluted true positives should appear to be negatives, while the convoluted true negatives would appear differential.
 Modified peptides and unmodified proteins were both simulated with 10 features.
@@
 In this simulation real world experimental conditions were introduced, including missing values and low number of modified features.
-Missing values and feature count values for this simulation were determined by looking at the averages from the biological experiments in this paper.
+The percentage of missing values and feature counts were determined by looking at the averages from the biological experiments in this paper.
 Modified peptides were simulated with 2 features, while unmodified proteins were still simulated with 10 features.
@@
 Adding missing values and few representative modified features provides a more realistic expectation of model performance.
+Otherwise this simulation used the same parameters as in simulation 1, with 1000 modified peptides half of which were positives and the other half negative.
 
@@
 Additionally, E coli Lysate was used to normalize total protein levels prior to enrichment or global protein profiling.
-Four mixes of spike in peptides and Human Lysate were created to create conditions with known fold change.
-Two sets of data were acquired for each mixture: modified peptide data, including the spike in KGG enriched peptides, and unmodified peptides from Human Lysate.
+The spike-in peptides were mixed with human lysate to create four mixture conditions.
+Two sets of data were acquired for each mixture: KGG enriched + LC-MS, and LC-MS only.
+The four mixtures were compared with known fold changes for the spike-in peptides.
 
@@
 
-The proposed approach was evaluated on an experiment where Shigella ubiquitin ligase IpaH7.8 was shown to inhibit the protein gasdermin D (GSDMD).\cite{LUCHETTI2021} Multiplex proteomics was used to quantify the abundance of total protein, and ubiquitination in human epithelial cells.
+The proposed approach was evaluated on an experiment where Shigella ubiquitin ligase IpaH7.8 was shown to inhibit the protein gasdermin D (GSDMD).\cite{LUCHETTI2021} Multiplex proteomics was used to quantify the abundance of total protein and ubiquitination in human epithelial cells.
 Cells were either infected or uninfected with IpaH7.8-deficient Shigella flexneri and measurements were taken at different time periods.
@@
 In this case $MSstatsPTM$ can still be used by extracting unmodified peptides from the modified run, however this generally leads to substantial less matches between modified and unmodified peptides, in addition to low feature counts for unmodified peptides.
+In this case there only 41.9\% of the modified peptides had a corresponding unmodified protein to perform adjustment.
+In cases where there is no estimate for the matching unmodified protein, adjustment cannot be performed.
 
@@
 The approach takes as input intensities of individual features from modified peptides, or intensity ratios of modified and unmodified peptide features, and compares the mean abundance of a PTM site from one condition to another.
-Modifications of the t-test such as moderated t-test with limma were also proposed.\cite{Zhu} While simple, the approach does not fully account for the sources of variations, and it is not directly applicable to experiments with complex designs, e.g., comparisons of multiple conditions, acquisition in multiple batches, etc.
+Modifications of the t-test such as moderated t-test with Limma were also proposed.\cite{Zhu} While simple, the approach does not fully account for the sources of variations, and it is not directly applicable to experiments with complex designs, e.g., comparisons of multiple conditions, acquisition in multiple batches, etc.
+Additionally, while this approach can be applied to experiments targeting PTMs, there is not a self contained, straight forward implementation of the methods, making application challenging.
+
 Isobar-PTM was developed for experiments with MS/MS quantitative strategies that employ isobaric labels such as tandem mass tags (TMT) and isobaric tag for relative and absolute quantification (iTRAQ).\cite{Breitwieser:2013} Isobar-PTM expresses MS measurements with a linear model and performs adjustment with respect to protein abundance using the difference between log-ratio of modified peptides in two channels and log-ratio of protein level.
 The modeling framework, however, is not applicable for either label-free workflows or experiments with complex designs.
+
 Further details on existing methods can be seen in Supplementary \ref{supp-sec:ttest}.
@@
 After using the converters, the next step is peptide/protein summarization and missing value imputation.
-For the modified run, the package will only summarize features, PSMs, which include the same modification at the same site.
+For the modified run, the package summarizes features, PSMs, which include the same modification together.
 Features with multiple modification sites are not included with single site features and are summarized separately.
@@
 Additionally, the summarization includes global median normalization and normalizes between MS runs.
-As mentioned in Section \ref{sec:missing_imp}, the package uses an AFT model to impute missing values in this step, although this step is optional.
+The package uses an AFT model, as mentioned in Section \ref{sec:missing_imp}, to impute missing values, although this step is optional.
 
@@
 The comparisons of interest can either be predefined or a full pairwise comparison will be tested.
-After modeling the modified and unmodified data, the modified model is adjusted for changes in unmodified protein abundance, using the methods described in Section \ref{sec:adjustment_summary}.
+After fitting a model to both the modified and unmodified data, the modified model is adjusted for changes in unmodified protein abundance, using the methods described in Section \ref{sec:adjustment_summary}.
 
-Beyond the core conversion, summary, and modeling functionality, the package also includes two plotting functions.
-The first creates plots for the summarization data to help with quality assurance analysis and identifying sources of variation.
-This function plots a quality control plot, summarizing the peptide abundance per run in the form of a boxplot, and profile plot, plotting each feature and the overall feature summarization as a line plot over each run.
-The second function plots the modeling results and includes a volcano plot, showing all peptides adjusted pvalues and fold changes, as well as a heatmap, which evaluates the fold change between conditions and peptides.
+Beyond the core functionalities of conversion, summary, and modeling, the package also includes functions for plotting the results.
+These include plots for the summarized and modeled data to assist in the analysis of the experiment.
+The summarized plots help with quality assurance analysis and identifying sources of variation.
+This includes a quality control plot, summarizing the peptide abundance per run in the form of a boxplot, and a profile plot, plotting each feature and the overall feature summarization as a line plot over each run.
+Additionally, the model plots include a volcano plot, showing all peptides adjusted pvalues and fold changes, as well as a heatmap, which evaluates the fold change between conditions and peptides.
 
@@
 The statistical modeling relies on the functionality from the R packages $lme4$ \cite{Bates2015} and $lmerTest$ \cite{Kuznetsova2017}.
-All datasets in this paper were evaluated using the $MSstatsPTM$ package.
 
@@
 
-The performance of of the proposed method was evaluated on simulated and spike in datasets with known ground truth, as well as biological experimental data where the ground truth was not known.
+The performance of the proposed method was evaluated on simulated and spike in datasets with known ground truth, as well as biological experimental data where the ground truth was not known.
 For the experiments where ground truth is known, we calculated summary statistics using the number of true positives (TP), false positives (FP), true negatives (TN), and false negatives (FN).
@@
 Comparisons were made before and after adjusting for changes in unmodified protein abundance and the results are presented in the form of venn diagrams and profile plots.
+All datasets in this paper were evaluated using the $MSstatsPTM$ package.
 
@@
 
-In this simulation, not accounting for changes in unmodified peptide level resulted in a high false discovery rate for all methods, seen in Figure \ref{fig:fdr_all_models}.
+Here we see that not accounting for changes in unmodified peptide level results in a high false discovery rate, seen in Figure \ref{fig:fdr_all_models}.
 When adjusting for changes in the unmodified peptide, all methods performed similarly in terms of FDR.
-Recall and accuracy of the proposed approach and limma performed similarly, while t-test lagged behind.
+Recall and accuracy of the proposed approach and Limma performed similarly, with Limma performing slightly better with low replicates, while t-test lagged behind.
 Two-sample t-test only uses data within the groups of interest while ignoring the remainder of the data.
 Consequently, it gave similar performance across cases with different number of conditions.
-In contrast, the proposed approach and limma leveraged all available information, which resulted in improved power with increased number of conditions.
+In contrast, the proposed approach and Limma leverage all available information, which resulted in improved power with increased number of conditions.
 Further results are provided in Supplementary Figures \ref{supp-fig:sim1_fpr} to \ref{supp-fig:sim1_recall}.
@@
 
-Changes in unmodified peptide level still needed to be accounted for to control the FDR in all methods.
-Once controlling for changes in unmodified peptide, it is clear that the proposed method performed stronger across all methods when compared to limma and t-test, as seen in Figure \ref{fig:sim_accuracy}.
-The proposed method well calibrated model accuracy and outperformed the other methods.
-Additionally, in the recall rate the proposed method showed the strongest performance, even when the number of biological replicates was low.
-Finally, the proposed method showed a tighter distribution of estimated fold changes around the true fold change, while the other methods showed much wider distributions.
+In this simulation, changes in unmodified peptide level still needed to be accounted for to control the FDR.
+Once controlling for changes in unmodified peptide, it is clear that the proposed method performed strongest across all methods, as seen in Figure \ref{fig:sim_accuracy}.
+The proposed method well calibrated model accuracy and the recall rate, even when the number of replicates were low.
+Finally, when comparing the fold change estimation across all peptides, the proposed method showed a tighter distribution of estimated fold changes around the true fold change.
+In comparison, Limma and t-test showed much wider distributions.
+This fold change comparison can be seen in Figure \ref{fig:sim_boxplot}.
 Further results are provided in Supplementary Figures \ref{supp-fig:sim2_recall} to \ref{supp-fig:fc_boxplot}.
@@
 After adjustment the estimated fold change is more in line with expectation.
-Additionally, background peptides serving as the null model, colored grey, show many false positives before adjustment is made.
+Additionally, the background peptides serving as the null model, colored grey, show many false positives before adjustment is made.
 Again after adjustment the results improve and the number of false positives decrease significantly.
 
-In Figure \ref{fig:spikein_limma_volcano} we can see the results of the same experiment modeled using limma with protein adjustment.
+In Figure \ref{fig:spikein_limma_volcano} we can see the results of the same experiment modeled using Limma with protein adjustment.
 Again the red labeled spike-in peptides are more inline with expected fold change after adjustment is applied.
-However, using limma we can see that the majority of the differential spike in peptides are not showing a significant adjusted pvalue.
-In this case, using limma would have resulted in missing the majority of differential modified peptides.
+However, using Limma we can see that the majority of the differential spike in peptides are not showing a significant adjusted pvalue.
+In this case, using Limma would have resulted in missing the majority of differential modified peptides.
 In terms of false positives, the results are very similar to the proposed method, with many false positives before adjustment and much fewer afterward.
@@
 
-Looking at modified peptide $GSDMD\_HUMAN|P57764$ we can see the advantage of the proposed method.
+The results of this experiment are summarized in Figures \ref{fig:ipah_venn_diagramm} and \ref{fig:ipah_profile}.
+In Figure \ref{fig:ipah_venn_diagramm} the number of significant modified peptides before and after adjustment is shown.
+We can see that more peptides become insignificant after adjustment than become significant.
+3,236 peptides become insignificant, 1352 peptides become significant, while 4,282 peptides are significant in both models.
+For the peptides that became insignificant in the adjusted model, their change in abundance was mainly due to changes in global protein abundance.
+In contrast, for peptides that became significant after adjustment, their true abundance change was masked by underlying changes in the unmodified protein.
+Both of these issues are correct in adjustment, and the true abundance change is shown.
+An additional question that must be addressed is if the decrease in significant peptides is due to the increased variance that comes from adjustment.
+This was tested by looking for modified peptides who's adjusted log fold change was within 10\% of the unadjusted log fold change but became insignificant after adjustment.
+In other words, the fold change was the same between models but variance increased.
+When this test was applied on this experiment, only one peptide became insignificant due to an increase in variance.
+Thus we can conclude that the drop off in significant peptides was truly due to changes in global protein abundance.
+
+In Figure \ref{fig:ipah_profile}, looking at modified peptide $GSDMD\_HUMAN|P57764$ we can see the advantage of the proposed method.
 The modified peptide shows a flat abundance change between the infected 1 hour, 4 hour, and 6 hour conditions.
@@
 The proposed method allows us to remove this confounding and estimate the true effect.
-The results of this can be seen in Figure \ref{fig:ipah_profile}.
 
@@
 
-Two specific modified peptides, $TTP\_MOUSE|P22893\_S178$ and $KI67\_MOUSE|E9PVX6\_T215$, show the power of the proposed method.
+The results of this experiment are summarized in Figures \ref{fig:shig_venn_d} and \ref{fig:shigella_profile}.
+In Figure \ref{fig:shig_venn_d} the number of significant modified peptides before and after adjustment is shown.
+Again more PTMs became insignificant after adjustment than became significant.
+19,286 peptides become insignificant, 4,947 became significant, while 39,550 peptides are significant in both models.
+Again we tested if the decrease in significant peptides was due to the increased variance that comes from adjustment, or if it was mainly due to removing convolution with change in the umodified protein.
+This was tested by looking for modified peptides who's adjusted log fold change was within 10\% of the unadjusted log fold change but became insignificant after adjustment.
+When this test was applied on this experiment, 776 peptides became insignificant due to an increase in variance.
+This is still a very small portion (4.02\%) of the peptides that became insignificant after adjustment.
+Thus we can conclude that the drop off in significant peptides was mainly due to changes in global protein abundance.
+
+In Figure \ref{fig:shigella_profile} the profile plot of two specific modified peptides, $TTP\_MOUSE|P22893\_S178$ and $KI67\_MOUSE|E9PVX6\_T215$, show the power of the proposed method.
 The profile plot of modified peptide $TTP\_MOUSE|P22893\_S178$ shows a large positive log fold change of $2.9$ between the WT\_Late and WT\_Uninfected conditions.
@@
 In adjusting the modified peptide for changes in the global protein, the log fold change becomes .828 and the adjusted pvalue goes from $.452$ to $.037$.
-The profile plots of these two PTMs can be seen in Figure \ref{fig:shigella_profile}
 
@@
 Of the 10,799 modified peptides identified, only 4526 had a corresponding unmodified protein and could be adjusted.
-Not having a separate global profiling run resulted in very low feature counts for the unmodified protein model.
+Additionally, not having a separate global profiling run resulted in very low feature counts for the unmodified protein model.
 
-Given the lack of matches in the global profiling run, the proposed method was unable to be applied to as many modified peptides as in the previous examples.
-However we can still see the power of the method when looking at peptide $P52209\_K059$.
+The results of this experiment are summarized in Figures \ref{fig:usp30_ven_d} and \ref{fig:usp30_profile}.
+In \ref{fig:usp30_ven_d} the number of significant modified peptides before and after protein adjustment is shown.
+In the top plot we can see there are a very large number of peptides that become insignificant after adjustment.
+This is mainly due to the lack of overlap between modified and unmodified peptides, stemming from not having a global profiling run.
+In the bottom plot only modified peptides that could be adjusted are shown.
+Here there are much less peptides that become insignificant after adjustment.
+In this plot we see that 726 peptides became insignificant, 547 became significant, and 1,078 are significant in both models.
+As in the previous experiments, we can check if these peptides are becoming insignificant due to an increase in variance.
+To check this we look at modified peptides whose adjusted fold change is within 10\% of their unadjusted fold change and they become insignificant after adjustment.
+In this experiment there are only 25 peptides that meet this criteria, a very small percentage of all modified peptides.
+
+In Figure \ref{fig:usp30_profile} the profile plot of peptide $P52209\_K059$ is shown.
 The abundance between conditions appears flat in the modified peptide profile plot, however this is contrast with a large negative change in the corresponding global protein.
@@
 However when adjusting for changes in the undmodified protein, the log fold change increased to $-2.73$ and adjusted pvalue became significant at $1.067e-04$.
-The profile plot for this PTM can be seen in Figure \ref{fig:usp30_profile}.
 Using the proposed method we get increased information that would otherwise have been missed.
@@
 
-We proposed a general statistical modeling framework for PTM characterization.
+We proposed a general statistical modeling framework and implementation for PTM characterization.
 The framework is designed for bottom-up MS workflows, which are characterized with variations from multiple convoluted sources, frequent missing data, and associated uncertainty in the conclusions.
 The framework is general and is applicable to a variety of experimental designs.
-It outperforms the ad-hoc methods underlying t-test and limma, and yields accurate results in the broad type of experimental circumstances, including the presence of missing values, changes in protein abundance, few representative peptides, and different acquisition methods.
+It outperforms the ad-hoc methods underlying t-test and Limma, and yields accurate results in the broad type of experimental circumstances, including the presence of missing values, changes in protein abundance, few representative peptides, and different acquisition methods.
 The framework allows us to plan for subsequent experiments, and choose the appropriate number of replicates in consideration of adjustment with respect to protein abundance.
+The implementation allows for straightforward application of the methods discussed and allows for reproducible experimental analysis.
+Additionally, the implementation allows the methods to be applied many different acquisition methods.
 
-Our results show that when measurements from multiple related conditions are available, the proposed approach for joint modeling and summarization of all the LC-MS/MS runs leads to more sensitive PTM significance analysis and more accurate and precise quantification than when separately analyzing conditions of runs.
+Our results show that the proposed approach for modeling and summarization  leads to more sensitive PTM significance analysis and more accurate and precise quantification.
 The gain is due to a more efficient use of the data, and to a more accurate understanding of the systematic and random variations.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Code availability}
+
+\bf{Bioconductor:} http://www.bioconductor.org/packages/release/bioc/html/MSstatsPTM.html
+
+\noindent{\bf{Github:} https://github.com/Vitek-Lab/MSstatsPTM}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%    REFERENCES   %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \includegraphics[scale=.8]{images/fig1.png}
-\caption{\todo{todo}}
+\caption{\todo{Meena?: Write about how MS for PTMs is run?}}
 \label{fig:workflow}
@@
 \caption{The advantage of using the proposed approach is apparent when looking at simulation 2, which includes limited observations and the presence of missing values.
-In overall accuracy, the proposed method performs stronger than limma and t-test in nearly every model.
-Even at lower replicates the proposed method still outperformed limma.
+In overall accuracy, the proposed method performs stronger than Limma and t-test in nearly every model.
+Even at lower replicates the proposed method still outperformed Limma.
 The lowest performaning method was $t$-test.
@@
 \label{fig:sim_accuracy}
+\end{figure}
+
+\begin{figure}[ht]
+\centering
+\includegraphics[width=1\textwidth]{images/sim3_FC_boxplot.png}
+\caption{In the second simulation with missing values and low PTM features all methods correctly estimated the fold change with a median log change of .75.
+The proposed method in this simulation had a visibly tighter distribution around the median when compared to Limma and $t$-test.
+While the mean estimated fold changes were similar, the proposed method exhibited a stronger performance in correctly estimating the fold change for all peptides.}
+\label{fig:sim_boxplot}
 \end{figure}
@@
 \caption{Limma method used to model the spike in experiment.
-Using limma the spike in peptides follow the expected log fold change better after adjusting for changes in protein level.
+Using Limma the spike in peptides follow the expected log fold change better after adjusting for changes in protein level.
 However, while the fold change is much more accurate, the majority of spike in peptides do not have a significant adjusted pvalue.
@@
 }\label{fig:spikein_limma_volcano}
+\end{figure}
+
+\begin{figure}[ht]
+\centering
+\includegraphics[width=1\textwidth]{images/IpaH_prof_plot.png}
+\caption{The overlap of differencial modified peptides for the PTM model with and without global protein level adjustment.
+More PTMs became insignificant after adjustment then became significant.
+For the peptides that became insignificant in the adjusted model, their change in abundance was mainly due to changes in global protein abundance.
+In contrast, for peptides that became significant after adjustment, their true abundance change was masked by underlying changes in the unmodified protein.}
+\label{fig:ipah_venn_diagramm}
 \end{figure}
@@
 \label{fig:ipah_profile}
+\end{figure}
+
+\begin{figure}[h!]
+\centering
+\includegraphics[height=.65\textwidth]{images/shig_venn_diagramm.png}
+\caption{The overlap of differencial modified peptides between the PTM model with and without global protein level adjustment.
+Again more PTMs became insignificant after adjustment then became significant.}
+\label{fig:shig_venn_d}
 \end{figure}
@@
 \label{fig:shigella_profile}
+\end{figure}
+
+\begin{figure}[h!]
+\centering
+ \begin{subfigure}{\textwidth}
+ \centering
+	\includegraphics[height=.5\textwidth]{images/usp30_venn_diagramm}
+	\caption{The overlap of differencial modified peptides for the PTM model with and without global protein level adjustment.
+Here many more PTMs became insignificant then became significant after adjustment.
+This is due to not having a global profiling run, resulting in a large lack of overlap between the modified peptides and unmodified proteins.}
+ \end{subfigure}\vspace{-5mm}
+ \begin{subfigure}{\textwidth}
+ \centering
+	\includegraphics[height=.5\textwidth]{images/usp30_venn_diagramm_matching_only}
+	\caption{Here we make the same comparison but only for modified peptides with a matching unmodified protein, so adjustment can be performed.
+In this case we see significantly less peptides become insignificant after adjustment.
+This highlights the need for a global profiling run if protein adjustment is going to performed.}
+ \end{subfigure}
+ \caption{Overlap of significant PTMs before and after protein adjustment.}
+\label{fig:usp30_ven_d}
 \end{figure}

### 2021-10-04 supplementary/ptm_sm.tex
made adjustments to supp and main

@@
 
-To adjust for unmodified peptides, the unmodified peptide is summarized in the same way and the result is combined with the PTM given by
+Adjustment for changes in the unmodified peptides is performed by taking the difference of log estimates, as follows
 
@@
 Run-level summarization of feature intensities for each PTM site is carried out as in the sub-plot model of MSstats~\cite{choi_etal_14a}, which involves a) imputation of censored missing values, and b) summarization of feature intensities using Tukey's median polish.
-The run-level summary for the PTM in Run $j$ of Condition $i$ is denoted by $\hat{y}_{ij}$.
+\cite{tukey_77a} The run-level summary for the PTM in Run $j$ of Condition $i$ is denoted by $\hat{y}_{ij}$.
 
@@
 
-The methods described previously can be extrapolated to Tandem Mass Tag (TMT) experiments targeting PTMs, similarly to how MSstatsTMT targets the global protein.\cite{Huang:2020} The null hypothesis is still that there is no difference in mean PTM abundance between Conditions $i$ and $i'$ after adjusting for changes in mean unmodified peptide abundance between Conditions $i$ and $i'$.
+The methods described previously can be extrapolated to Tandem Mass Tag (TMT) experiments targeting PTMs.
+This is done in the same way to how MSstatsTMT targets the global protein.\cite{Huang:2020} The null hypothesis is that there is no difference in mean PTM abundance between Conditions $i$ and $i'$ after adjusting for changes in mean unmodified peptide abundance between Conditions $i$ and $i'$.
 \begin{align*}
@@
 For a given modified peptide and corresponding unmodified protein a mixed effects model is fit as follows.
-
 \begin{align*}
-Y_{mtcb} = \mu + Mixture_m + Techrep(Mixture)_{tm} + Condition_c + Subject_{mcb} + \epsilon_{mtcb}\\
-Y^*_{mtcb} = \mu^* + Mixture^*_m + Techrep(Mixture)^*_{tm} + Condition^*_c + Subject^*_{mcb} + \epsilon^*_{mtcb}
+Y_{mtcb} = \mu + M_m + T(M)_{tm} + C_c + S_{mcb} + \epsilon_{mtcb}\\
+Y^*_{mtcb} = \mu^* + M^*_m + T(M)^*_{tm} + C^*_c + S^*_{mcb} + \epsilon^*_{mtcb}
 \end{align*}
 
-Where $Mixture_m \sim N(0, \sigma^2_m)$, $Techrep(Mixture)_{tm} \sim N(0, \sigma^2_{tm})$, $Subject \sim N(0, \sigma^2_{s})$ are random effects, and $\sum_{c=1}^{I} Condition_{c}^{\ast} = 0$ is a fixed effect.
+Where $M_m \sim N(0, \sigma^2_m)$, $T(M)_{tm} \sim N(0, \sigma^2_{tm})$, $S\sim N(0, \sigma^2_{s})$ are random effects, and $\sum_{c=1}^{I} C_{c}^{\ast} = 0$ is a fixed effect.
 Finally the error is expressed as $\epsilon_{mtcb} \sim N(0, \sigma^2)$.
@@
 
-$$\hat{\Delta} = \frac{1}{MJ} \left( \hat{y}_{+,+,i+} - \hat{y}_{+,+,i'+} \right) - \frac{1}{BJ} \left( \hat{y}_{+,+,i+}^{\ast} - \hat{y}_{+,+,i'+}^{\ast} \right)$$
+$$\hat{\Delta} = \frac{1}{MJ} \left( \hat{y}_{+,+,i+} - \hat{y}_{+,+,i'+} \right) - \frac{1}{MJ} \left( \hat{y}_{+,+,i+}^{\ast} - \hat{y}_{+,+,i'+}^{\ast} \right)$$
 
@@
 
-The proposed statistical approach was evaluated and compared to the $t$-test and Limma using computer simulation.
-In particular, their statistical properties with respect to protein-level adjustment and real world experimental conditions were evaluated.
+The proposed statistical approach was evaluated and compared to the $t$-test and Limma methods using computer simulation.
+Specifically, their statistical properties with respect to protein-level adjustment and real world experimental conditions were evaluated.
 
@@
 Approaches without considering protein-level adjustment lose track of an important aspect in interpreting observed changes in PTM abundance, which may result in misleading conclusions.
-To highlight the necessity of the adjustment, we compared the following approaches: a) proposed approach, b) proposed approach without adjusting for unmodified peptides c) $t$-test (with adjustment), d) $t$-test (no adjustment), e) Limma (with adjustment), and f) Limma (no adjustment).
+To highlight the necessity of the adjustment, we compared six different approaches as follows: a) proposed approach, b) proposed approach without adjusting for unmodified peptides c) $t$-test (with adjustment), d) $t$-test (no adjustment), e) Limma (with adjustment), and f) Limma (no adjustment).
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection{Computer simulation 1 - label free}
+\subsubsection{Computer simulation 1 - label free experiment}
 
@@
 \item Number of features per PTM: $10$
-\item Number of features per protein: $10$
+\item Number of features per unmodified protein: $10$
 \item Missing data: no missing value
@@
 
-The results are summarized from \sfigref{fig:sim1_fpr} to \sfigref{fig:sim1_recall}, including false discovery rate with or without changes in protein abundance by the considered methods, false positive rate, and overall accuracy.
-Additionally the results are compared to simulation 2 described in the next section.
-
-When simulating the data, 500 of the 1000 peptides were simulated with a differential fold change between conditions, while the other 500 were not.
-In terms of the peptides that were not differential, 250 were simulated with a change in both PTM and overall protein abundance, such that the change in PTM abundance is entirely due to the change in the overall protein.
+The results are summarized from \sfigref{fig:sim1_fpr} to \sfigref{fig:sim1_recall}.
+These results include the false discovery rate with and without adjusting for changes in unmodified protein abundance, false positive rate, and overall accuracy.
+Additionally, the results are compared to simulation 2 described in the next section.
+
+When simulating the data, 500 of the 1000 peptides were simulated as true positives with a differential fold change between conditions, while the other 500 were not differential and treated as true negatives.
+In terms of the true negatives, 250 were simulated such that any abundance changes were entirely driven by changes in the unmodified peptide.
+We would expect any model without performing protein adjustment to treat these as false positives.
 The other 250 were simulated with no significant change in either PTM or overall protein abundance.
-In this way there were 500 true positives and 500 true negative PTMs.
 
@@
 In the second simulation we introduced limited feature observations per PTM as well as masking a portion of the observation to simulate missing values.
-This is more in line with what we would expect with a real life experiment and provides a more realistic expectation of model performance.
+This is more in line with what we would expect in a biological experiment and provides a more realistic expectation of model performance.
+The number of PTM features and missing data percentage where determined by looking at the features and missing data in the biological experiments in this paper.
 
@@
 \item Number of features per PTM: $2$
-\item Number of features per protein: $10$
+\item Number of features per unmodified protein: $10$
 \item Missing data: 20\% of the observations for PTMs and Proteins were masked with NA at random
@@
 
-The results are summarized from \sfigref{fig:sim2_recall} to \sfigref{fig:fc_boxplot}, including false positive rate with or without changes in protein abundance by the considered methods, false positive rate, and overall accuracy.
+The results are summarized from \sfigref{fig:sim2_recall} to \sfigref{fig:fc_boxplot}.
 
@@
 	\caption{All the considered methods in simulation 1 correctly calibrated FDR  when adjusting for changes in protein abundance.
-In comparison, the methods without accounting for the protein-level changes resulted in off-target, high false discovery rates.}
+In comparison, the methods without accounting for the protein-level changes resulted in off-target, high false discovery rates.
+The performance of the models without adjustment was much lower than those with adjustment, thus only models with adjustment are compared going forward.}
  \end{subfigure}
@@
 	\includegraphics[width=.8\textwidth]{sim_new/sim1_FDR}
-	\caption{The considered methods with protein adjustment are compared in detail. All three methods generally performed similarly in terms of FDR.}
+	\caption{The considered methods with protein adjustment are compared in detail. All three methods with adjustment generally performed similarly in terms of FDR.}
  \end{subfigure}
@@
 	\caption{The recall for simulation 1 methods with adjustment were compared.
-It is clear that limma performs the strongest here when the number of replicates were low.
-At higher replicates the performance of the proposed methods and limma are comparable.
+It is clear that Limma performs the strongest here when the number of replicates were low.
+At higher replicates the performance of the proposed methods and Limma are comparable.
 $t$-test clearly performs worse across all methods.}
@@
 	\caption{The advantage of using the proposed approach is apparent when looking at simulation 2, which includes limited observations and the presence of missing values.
-In the case of recall the proposed method performs stronger than limma and t-test in nearly every model.
-Even at lower replicates the proposed method still outperformed limma.
+In the case of recall the proposed method performs stronger than Limma and t-test in nearly every model.
+Even at lower replicates the proposed method still outperformed Limma.
 Again the lowest performing method was $t$-test.}
@@
 	\includegraphics[width=.8\textwidth]{sim_new/sim3_FC_boxplot}
-	\caption{In simulation 2 again all methods correctly estimated the fold change with a median log change of .75.
+	\caption{In simulation 2 all methods correctly estimated the fold change with a median log change of .75.
 The proposed method in this simulation had a visibly tighter distribution around the median.
-Both limma and $t$-test showed a wider range around the fold change.
+Both Limma and $t$-test showed a wider range around the fold change.
 In this case the proposed method showed a stronger performance in correctly estimating the fold change for all peptides.}
@@
 A custom designed experiment with labeling was used to assess the performance of the proposed method in a real experimental setting.
-Heavy-labeld KGG modified peptides were used as spike-in peptides.
+Heavy-labeled KGG modified peptides were used as spike-in peptides.
 The spike-in peptides were mixed with human lysate to create four mixture conditions.
@@
 The LC-MS dataset included only unmodified peptides.
-The spike-in peptides where a significant fold change between conditions are expected to be differencial, whereas none of the human lysate peptides in any comparison are expected to be differencial.
-
-Again we consider three different methods and assess their performance: the method proposed in this paper, limma, and two sample $t$-test.
+The spike-in peptides have a known fold change between conditions and are expected to be differential in most cases, whereas the human lysate modified peptides were treated as controls and are not expected to be differential in any comparison.
+
+Again we consider three different methods and assess their performance: the method proposed in this paper, Limma, and two sample $t$-test.
 All methods are analyzed after adjusting for changes in overall protein level.
@@
 \includegraphics[width=.85\textwidth]{sim_new/spike_in_msstatsptm_volcano}
-\caption{Using the proposed method to model the benchmark experiment, the spike in peptides (colored red) do not follow the expected log fold change before adjustment.
+\caption{The modeling results of the proposed method both before and after adjustment.
+The spike in peptides are colored red and the background peptides are colored grey.
+All grey peptides are expected to be insignificant.
+Using the proposed method to model the benchmark experiment, the spike in peptides (colored red) do not follow the expected log fold change before adjustment.
 After adjusting for changes in overall protein abundance the spike in peptides are more in line with expectation.
 Additionally the background grey colored peptides show many false positives before adjustment.
-After adjustment the false positives were decreased considerably.
+After adjustment these false positives were decreased considerably.
 \label{fig:spike_volcano_msstats}}
@@
 \includegraphics[width=.85\textwidth]{sim_new/spike_in_limma_volcano}
-\caption{When modeling the experiment with the limma method, the spike in peptides again follow the expected log fold change better after adjusting for changes in protein level.
+\caption{When modeling the experiment with the Limma method, the spike in peptides again follow the expected log fold change better after adjusting for changes in protein level.
 However, while the fold change is much more accurate, the majority of spike in peptides do not have a significant adjusted pvalue.
-In this case, the known differencial peptides are missed by the model.
+In this case, the known differential peptides are missed by the model.
 In terms of false positives, the results are very similar to the proposed method, with many false positives before adjustment and much fewer after.
@@
 \caption{Comparing the summary statistics between methods it is clear that the proposed method performs the strongest.
-In terms of accuracy and specificity the three methods are close, with limma and $t$-test showing slightly higher values.
+In terms of accuracy and specificity the three methods are close, with Limma and $t$-test showing slightly higher values.
 Accuracy and specificity are dominated by the large number of true negatives (background peptides) compared to the true positives (spike in peptides).
@@
 
-A model was fit for the total protein, and ubiquitination separately, as described previously for TMT experiments in Section~\ref{sec:tmtmethod}. The model formula can be seen below.
+A model was fit for the total protein and ubiquitination separately, as described previously for TMT experiments in Section~\ref{sec:tmtmethod}. The model formula can be seen below.
 
@@
 \includegraphics[height=.65\textwidth]{sim_new/ipah_venn_diagramm.png}
-\caption{The overlap of differencial modified peptides between the PTM model with and without global protein level adjustment.
+\caption{The overlap of differential modified peptides for the PTM model with and without global protein level adjustment.
 More PTMs became insignificant after adjustment then became significant.
-This indicates that for the peptides that became insignificant in the adjusted model, their change in abundance was mainly due to changes in global protein abundance.
+For the peptides that became insignificant in the adjusted model, their change in abundance was mainly due to changes in global protein abundance.
+In contrast, for peptides that became significant after adjustment, their true abundance change was masked by underlying changes in the unmodified protein.
 An additional question that must be addressed is if the decrease in significant peptides is due to the increased variance that comes from adjustment.
 This was tested by looking for modified peptides who's adjusted log fold change was within 10\% of the unadjusted log fold change but became insignificant after adjustment.
+In other words, the fold change was the same between models but variance increased.
 When this test was applied on this experiment, only one peptide became insignificant due to an increase in variance.
@@
 \hline
- & Mixture 1 & & Mixture 2 & & Condition \\ [0.5ex]
+ & \multicolumn{2}{c}{Mixture 1} & \multicolumn{2}{c}{Mixture 2} & Condition \\ [0.5ex]
  \hline\hline
@@
 \includegraphics[height=.65\textwidth]{sim_new/shig_venn_diagramm.png}
-\caption{The overlap of differencial modified peptides between the PTM model with and without global protein level adjustment.
+\caption{The overlap of differential modified peptides between the PTM model with and without global protein level adjustment.
 Much like in \sfigref{fig:ipha_venn_d}, more PTMs became insignificant after adjustment then became significant.
@@
 	\includegraphics[height=.5\textwidth]{sim_new/usp30_venn_diagramm}
-	\caption{The overlap of differencial modified peptides between the PTM model with and without global protein level adjustment.In this experiment, many more PTMs became insignificant after adjustment then became significant.
-In this case this is mainly due to the large lack of overlap between the modified peptides and unmodified proteins, stemming from not having a global profiling run.}
+	\caption{The overlap of differential modified peptides for the PTM model with and without global protein level adjustment.
+Here many more PTMs became insignificant then became significant after adjustment.
+This is due to not having a global profiling run, resulting in a large lack of overlap between the modified peptides and unmodified proteins.}
  \end{subfigure}\vspace{-5mm}
@@
 	\includegraphics[height=.5\textwidth]{sim_new/usp30_venn_diagramm_matching_only}
-	\caption{The same overlap as above, but looking only at modified peptides with a matching unmodified protein so adjustment can be performed.
+	\caption{Here we make the same comparison but only for modified peptides with a matching unmodified protein, so adjustment can be performed.
 In this case we see significantly less peptides become insignificant after adjustment.
-Again we can check if peptides are becoming insignificant due to an increase in variance.
+This highlights the need for a global profiling run if protein adjustment is going to performed.
+As in the previous experiments, we can check if peptides are becoming insignificant due to an increase in variance.
 To check this we look at modified peptides whose adjusted fold change is within 10\% of their unadjusted fold change and they become insignificant after adjustment.

### 2021-10-04 code/sample_size_analysis.R

@@ -1,37 +1,71 @@
 
+library(tidyverse)
+
+## Define standard deviation sampling
+sample_variance <- function(mod_samples, unmod_samples, mod_features,
+                            unmod_features, mod_variance, unmod_variance){
+
+  final_sd <- numeric(100000)
+
+  for (i in seq_len(100000)){
+
+    replicate_means_ptms <- rnorm(mod_samples, mean = 0, mod_variance)
+    replicate_observations_ptms <- numeric(length(mod_samples))
+
+    # for (y in seq_along(replicate_means_ptms)){
+    #   feature_samples <- rnorm(mod_features, mean = replicate_means_ptms[[y]], 2)
+    #   replicate_observations_ptms[[y]] <- mean(feature_samples)
+    # }
+
+    replicate_means_prot <- rnorm(unmod_samples, mean = 0, unmod_variance)
+    replicate_observations_prot <- numeric(length(unmod_samples))
+
+    # for (y in seq_along(replicate_means_prot)){
+    #   feature_samples_prot <- rnorm(unmod_features, mean = replicate_means_prot[[y]], 2)
+    #   replicate_observations_prot[[y]] <- mean(feature_samples_prot)
+    # }
+
+    std_ptm <- mod_variance^2 / sum((replicate_means_ptms - mean(replicate_means_ptms))^2)
+    std_prot <- unmod_variance^2 / sum((replicate_means_prot - mean(replicate_means_prot))^2)
+
+    final_sd[[i]] <- sqrt(std_ptm^2 + std_prot^2)
+
+  }
+
+  return(mean(final_sd))
+}
+
+
 
 ## Code directly from MSstats designSampleSize function
-## Range of desired FC
-delta <- log2(seq(1.5, 2.5, 0.025))
-desiredFC <- 2 ^ delta
-
-# delta <- seq(.5, 1.5, 0.025)
-# desiredFC <- delta
-
-## Define base metrics
-
-low <- 1
-high <- 3
-
-mod_pep_sigma <- low
-unmod_pep_sigma <- low
-
-n_sample_list <- c(3,5,7,10)
-
-mod_num_sample <- n_sample_list[[1]]
-unmod_num_sample <- n_sample_list[[1]]
-
-#Specify sigma for ptm and unmod pep
-
[…]

### 2021-10-05 main/MSstatsPTM_Main.tex
Notes

@@
 
-\section{Experimental Procedure}
+\section{Experimental Procedures}
 
@@
 
-\subsection*{SpikeIn benchmark - KGG enriched}
+\subsection*{SpikeIn benchmark - KGG enriched \todo{label-free?}}
 
-We evaluated our approach using a custom designed spike-in benchmark experiment where 50 heavy-labeled KGG motif peptides from 20 human proteins were used as spike-in peptides.
+We evaluated our approach using a custom designed spike-in benchmark experiment with known ground truth, where 50 heavy-labeled KGG motif peptides from 20 human proteins were used as spike-in peptides.
 Quantitative changes in protein and site abundance changes of these 20 proteins were the target of the benchmark.
@@
 
-Further details on existing methods can be seen in Supplementary \ref{supp-sec:ttest}.
+Further details \todo{clarify which details}  on existing methods can be seen in Supplementary \ref{supp-sec:ttest}.
 
@@
 In addition mixture may be included for features acquired via tandem mass tag (TMT) methods.
-The model parameters are estimated using the split-plot approach as in $MSstats$, where the feature log-intensities are first summarized into a single value per site per run in the subplot model, and the site-level summaries are then used for the inference of the PTM site abundance.\cite{Choi:2014} In the site-level summarization, Tukey's median polish (TMP), a simple and robust procedure is applied to iteratively fit a two-way additive model with the effects of run and feature, which in turn summarizes the log-intensities for each site.\cite{Tukey:1977} After summarization, the inference of the PTM site abundance in each condition is carried through fitting a model based on the family of linear mixed-effects models, taking into account the specific experimental design \ref{fig:data-structure}.\cite{Bolker2009} \cite{Faraway:2006} Statistical modeling and quantification for global proteomics data are performed by the same procedure as for PTM data.
+The model parameters are estimated using the split-plot approach as in $MSstats$, where the feature log-intensities are first summarized into a single value per site per run in the subplot model, and the site-level summaries are then used for the inference of the PTM site abundance.\cite{Choi:2014} In the site-level summarization, Tukey's median polish (TMP), a simple and robust procedure is applied to iteratively fit a two-way additive model with the effects of run and feature, which in turn summarizes the log-intensities for each site.\cite{Tukey:1977} After summarization, the inference of the PTM site abundance in each condition is carried through fitting a model based on the family of linear mixed-effects models, taking into account the specific experimental design \todo{weird reference?} \ref{fig:data-structure}.\cite{Bolker2009} \cite{Faraway:2006} Statistical modeling and quantification for global proteomics data are performed by the same procedure as for PTM data.
 
@@
 \subsection*{Evaluation}
+
+\todo{In the caption of each figure, could you clarify which dataset the figure refers to?}
 
@@
 
-Further details and results of this experiment are provided in Supplementary Sec. \ref{supp-sec:shigella}.
+Further details and results of this experiment are provided in Supplementary Sec. \ref{supp-sec:shigella}.  \todo{Maybe clarify what kind of results}
 
@@
 
-\bf{Bioconductor:} http://www.bioconductor.org/packages/release/bioc/html/MSstatsPTM.html
+\todo{This belongs to the 'implementation' section}
+
+{ \bf Bioconductor:} http://www.bioconductor.org/packages/release/bioc/html/MSstatsPTM.html
 

### 2021-10-05 main/MSstatsPTM_Main.tex
small note

@@
 
+\todo{Meena should be a co-corresponding author}
 The scientific community widely utilizes mass spectrometry (MS)-based proteomics to quantify the abundance of proteins and their post-translational modifications (PTMs).

### 2021-10-05 main/MSstatsPTM_Main.tex
More comments

@@
 The scientific community widely utilizes mass spectrometry (MS)-based proteomics to quantify the abundance of proteins and their post-translational modifications (PTMs).
-Experiments targeting PTMs face several specific challenges.
+Experiments targeting PTMs face several specific challenges \todo{challenges of statistical nature?}.
 These include the low abundance of modified proteo-forms, few representative peptides that span modification sites, and convolution with abundance changes in the overall protein expression.
-Due to these challenges, a robust approach to estimate relative systematic changes in PTMs should combine information pertaining to PTM sites over several peptides, replicates in multiple conditions, and consider sources of confounding and variation present in the experiment.
+Due to these challenges, a robust approach to estimating relative systematic changes in PTMs should combine information pertaining to PTM sites over several peptides, replicates in multiple conditions, and consider sources of confounding and variation present in the experiment.
 We propose a general statistical model and workflow that is both reproducible and comprehensive.
@@
 This model is automatically adjusted to the specific experimental design.
-Finally, the PTM abundances are adjusted to remove variance from changes in the overall protein.
+Finally, the PTM abundances are adjusted to remove variance \todo{bias?} from changes in the overall protein.
 We implement this model in the free and open-source R package $MSstatsPTM$.
+\todo{Evaluations show that XXX} \todo{Unlike in CS, usually the text is in the past tense}
 
@@
 
-The signaling mechanisms that allow cells to mount a dynamic and fast response to a multitude of events are primarily facilitated by the modification of proteins at specific residues, acting as molecular on/off switches.\cite{Deribe} \cite{Cohen}  Mass spectrometry-based label-free proteomics is broadly established as the tool-of-choice for unbiased and large-scale identification and quantification of proteins and their post-translational modifications (PTMs) using liquid chromatography coupled with mass spectrometry (LC-MS).\cite{Kall:2011ub} \cite{Roepstorff} Studies targeting the post-translationally modified proteome focus either on the accurate localization of modification sites on proteins, relative or absolute quantification of a modification site's occupancy repertoire, or relative changes in occupancy across experimental conditions.\cite{Mann} Regardless of the question at hand, interrogating the modified proteome is challenging due to a number of reasons.
+The signaling mechanisms that allow cells to mount a dynamic and fast response to a multitude of events are primarily facilitated by the modification of proteins at specific residues, acting as molecular on/off switches.\cite{Deribe} \cite{Cohen} \todo{Check that the references appear before the end of the sentence} Mass spectrometry-based label-free proteomics is broadly established as the tool-of-choice for unbiased and large-scale identification and quantification of proteins and their post-translational modifications (PTMs) using liquid chromatography coupled with mass spectrometry (LC-MS).\cite{Kall:2011ub} \cite{Roepstorff} Studies targeting the post-translationally modified proteome focus either on the accurate localization of modification sites on proteins, relative or absolute quantification of a modification site's occupancy repertoire, or relative changes in occupancy across experimental conditions.\cite{Mann} Regardless of the question at hand, interrogating the modified proteome is challenging due to a number of reasons.
 First, the relatively lower abundance of modified proteo-forms dictates that a global interrogation can only be achieved through large-scale enrichment protocols with modification-specific antibodies or beads.
@@
 
-This section describes the datasets and experiments used to evaluate the proposed method. Further details can be seen in Supplemental Sections \ref{supp-sec:sim} and \ref{supp-sec:bio_investigations}
+This section describes the datasets and experiments used to evaluate the proposed method.
+Further details can be seen in Supplemental Sections \ref{supp-sec:sim} and \ref{supp-sec:bio_investigations} \todo{Either specify which details exactly, or remove the sentence} \todo{A better use of this paragraph is to argue why we chose these particular datasets - some with known ground truth, various data acquisition strategies, various organisms}
+
+\todo{I am wondering if we should number the datasets}
 
@@
 \label{sec:comp_sim_procedure}
+
+\todo{These two paragraphs are a bit out of place - we have not yet introduced the proposed approach, Limma etc.
+This section should focus specifically on the data, not on the analyses.
+Much of the text below belongs to results}
 

### 2021-10-05 code/sample_size_analysis.R
added sample size analysis

@@ -1,70 +1,33 @@
 
 library(tidyverse)
 
-## Define standard deviation sampling
-sample_variance <- function(mod_samples, unmod_samples, mod_features,
-                            unmod_features, mod_variance, unmod_variance){
+## Calculations w/o feature variations -----------------------------------------
 
-  final_sd <- numeric(100000)
-
-  for (i in seq_len(100000)){
-
-    replicate_means_ptms <- rnorm(mod_samples, mean = 0, mod_variance)
-    replicate_observations_ptms <- numeric(length(mod_samples))
-
-    # for (y in seq_along(replicate_means_ptms)){
-    #   feature_samples <- rnorm(mod_features, mean = replicate_means_ptms[[y]], 2)
-    #   replicate_observations_ptms[[y]] <- mean(feature_samples)
-    # }
-
-    replicate_means_prot <- rnorm(unmod_samples, mean = 0, unmod_variance)
-    replicate_observations_prot <- numeric(length(unmod_samples))
-
-    # for (y in seq_along(replicate_means_prot)){
-    #   feature_samples_prot <- rnorm(unmod_features, mean = replicate_means_prot[[y]], 2)
-    #   replicate_observations_prot[[y]] <- mean(feature_samples_prot)
-    # }
-
-    std_ptm <- mod_variance^2 / sum((replicate_means_ptms - mean(replicate_means_ptms))^2)
-    std_prot <- unmod_variance^2 / sum((replicate_means_prot - mean(replicate_means_prot))^2)
-
-    final_sd[[i]] <- sqrt(std_ptm^2 + std_prot^2)
-
-  }
-
-  return(mean(final_sd))
+sigma_model <- function(ptm_var, ptm_n, prot_var, prot_n){
+  simga <- sqrt((ptm_var/(.25*ptm_n)) + (prot_var/(.25*prot_n)))
+  return(simga)
 }
 
+## Define adjustable metrics
+low <- .1
+high <- .3
+n_sample_list <- c(3,5,10)
 
-
-## Code directly from MSstats designSampleSize function
-## Power analysis
[…]

### 2021-10-07 main/MSstatsPTM_Main.tex
Updated main per Olgas comments, added sample size analysis

--- 2021-10-05 main/MSstatsPTM_Main.tex
@@
 
-\author[1]{Devon~Kohler}
-\author[2]{Tsung-Heng~Tsai}
+\author[1]{Devon~Kohler*}
+\author[2]{Tsung-Heng~Tsai*}
+\author[3]{Meena~Choi*}
 \author[1]{Ting~Huang}
@@
 \author[3]{Trent~Hinkle}
-\author[3]{Meena~Choi}
 \author[1]{Olga~Vitek}
@@
 \affil[4]{Galapagos, Mechelen, Antwerp, Belgium}
+\affil[*]{Corresponding Authors}
 
@@
 The scientific community widely utilizes mass spectrometry (MS)-based proteomics to quantify the abundance of proteins and their post-translational modifications (PTMs).
-Experiments targeting PTMs face several specific challenges.
+Experiments targeting PTMs face several specific challenges which are statistical in nature.
 These include the low abundance of modified proteo-forms, few representative peptides that span modification sites, and convolution with abundance changes in the overall protein expression.
-Due to these challenges, a robust approach to estimate relative systematic changes in PTMs should combine information pertaining to PTM sites over several peptides, replicates in multiple conditions, and consider sources of confounding and variation present in the experiment.
+Due to these challenges, a robust approach to estimating relative systematic changes in PTMs should combine information pertaining to PTM sites over several peptides, replicates in multiple conditions, and consider sources of confounding and variation present in the experiment.
 We propose a general statistical model and workflow that is both reproducible and comprehensive.
@@
 This model is automatically adjusted to the specific experimental design.
-Finally, the PTM abundances are adjusted to remove variance from changes in the overall protein.
-We implement this model in the free and open-source R package $MSstatsPTM$.
+Finally, the PTM abundances are adjusted to remove bias from changes in the overall protein.
+We implemented this model in the free and open-source R package $MSstatsPTM$.
+The approach is evaluated on computer simulations, a spike-in experiment with known ground truth, and three biological experiments with varying acquisition methods.
+Finally, the approach is compared against several existing methods and is shown to outperform all other methods.
 
@@
 
-The signaling mechanisms that allow cells to mount a dynamic and fast response to a multitude of events are primarily facilitated by the modification of proteins at specific residues, acting as molecular on/off switches.\cite{Deribe} \cite{Cohen}  Mass spectrometry-based label-free proteomics is broadly established as the tool-of-choice for unbiased and large-scale identification and quantification of proteins and their post-translational modifications (PTMs) using liquid chromatography coupled with mass spectrometry (LC-MS).\cite{Kall:2011ub} \cite{Roepstorff} Studies targeting the post-translationally modified proteome focus either on the accurate localization of modification sites on proteins, relative or absolute quantification of a modification site's occupancy repertoire, or relative changes in occupancy across experimental conditions.\cite{Mann} Regardless of the question at hand, interrogating the modified proteome is challenging due to a number of reasons.
+The signaling mechanisms that allow cells to mount a dynamic and fast response to a multitude of events are primarily facilitated by the modification of proteins at specific residues, acting as molecular on/off switches\cite{Deribe} \cite{Cohen}.
+Mass spectrometry-based label-free proteomics is broadly established as the tool-of-choice for unbiased and large-scale identification and quantification of proteins and their post-translational modifications (PTMs) using liquid chromatography coupled with mass spectrometry (LC-MS)\cite{Kall:2011ub} \cite{Roepstorff}.
+Studies targeting the post-translationally modified proteome focus either on the accurate localization of modification sites on proteins, relative or absolute quantification of a modification site's occupancy repertoire, or relative changes in occupancy across experimental conditions\cite{Mann}.
+Regardless of the question at hand, interrogating the modified proteome is challenging due to a number of reasons.
 First, the relatively lower abundance of modified proteo-forms dictates that a global interrogation can only be achieved through large-scale enrichment protocols with modification-specific antibodies or beads.
@@
 Second, contrary to the often large number of identified peptides that can be used as features to model protein abundance changes, there are relatively few representative peptides that span a modification sites, which often results in sparse, and sometimes, inherently convoluted models (i.e., single versus multiple modified sites on a single peptide).
-Third, unless early signaling events are interrogated, the interpretation of the relative changes in modification occupancy are inherently convoluted with changes in the overall protein expression, making the interpretation of the results not straightforward.\cite{Olsen:2013} Therefore, a robust approach to estimate systematic relative changes in post-translational modifications, at scale, should not only combine the quantitative information pertaining to a PTM site over peptides and replicates in multiple conditions, but take into account various sources of variations and confounding factors present in the experiments.
+Third, unless early signaling events are interrogated, the interpretation of the relative changes in modification occupancy are inherently convoluted with changes in the overall protein expression, making the interpretation of the results not straightforward\cite{Olsen:2013}.
+Therefore, a robust approach to estimate systematic relative changes in post-translational modifications, at scale, should not only combine the quantitative information pertaining to a PTM site over peptides and replicates in multiple conditions, but take into account various sources of variations and confounding factors present in the experiments.
 
@@
 The proposed approach was implemented using the R coding language and evaluated using datasets from computer simulations, benchmark controlled mixtures, and biological investigations.
-The approach was compared against the commonly applied t-test and Limma methods.\cite{Ritchie_15a} The results demonstrate that by appropriately leveraging the information from the entire dataset, the proposed approach improves the reproducibility and accuracy of the estimates of PTM fold changes and results in a better calibrated type I error rate.
-Finally, the proposed approach is implemented as a freely available open source R package $MSstatsPTM$, available on Bioconductor, which employs similar input format as in $MSstats$ and $MSstatsTMT$, and is compatible with many acquisition methods, such as label free, DIA, DDA, and TMT.\cite{Choi:2014} \cite{Huang:2020}
+The approach was compared against the commonly applied t-test and Limma methods\cite{Ritchie_15a}.
+The results demonstrate that by appropriately leveraging the information from the entire dataset, the proposed approach improves the reproducibility and accuracy of the estimates of PTM fold changes and results in a better calibrated type I error rate.
+Finally, the proposed approach is implemented as a freely available open source R package $MSstatsPTM$, available on Bioconductor, which employs similar input format as in $MSstats$ and $MSstatsTMT$, and is compatible with many acquisition methods, such as label free, DIA, DDA, and TMT\cite{Choi:2014} \cite{Huang:2020}.
 
@@
 
-This section describes the datasets and experiments used to evaluate the proposed method. Further details can be seen in Supplemental Sections \ref{supp-sec:sim} and \ref{supp-sec:bio_investigations}
+This section describes the datasets and experiments used to evaluate the proposed method.
+The datasets were chosen to provide a performance benchmark of the proposed approach against existing methods when ground truth is known, as well as represent a variety of experimental designs and acquisition methods where the approach can be applied.
+To benchmark the dataset, computer simulations and a spike-in experiment were used.
+The computer simulations vary in experimental realism; the first simulation was perfectly clean, with many replicates and no missing values, while the second simulation introduced real world qualities, such as limited modified features and missing values.
+The spike-in experiment took the real world qualities a step further and allowed us to compare the methods in an real experiment where ground truth is known.
+Finally the proposed approach was used to model three biological experiments which differed in experimental design, acquisition strategy, and use various organisms.
+These experiments were chosen to show the proposed method's applicability to many different types of experiments.
 
@@
 
-The proposed statistical approach methods were evaluated using two computer simulations, with known ground truth.
-Specifically, their properties under adjustment with respect to protein abundance were evaluated.
-Two simulations were ran, mimicking a label free experiment, one including a high number of features and no missing values, and one with few modified features and including missing values.
-
-Differential intensity levels of modified peptides may be due to changes in modification, change in protein abundance, or both.
-The proposed approach adjusts the abundance with respect to unmodified peptides by combining the inference of modified and unmodified peptide abundances.
-Alternatively, two-sample t-test or Limma that takes as input the ratio between modified and unmodified peptide intensities (difference on log scale) is commonly applied for the same purpose.
-In real experiments, multiple inter-related conditions are often compared together.
-Whereas t-test uses measurements from the two conditions being compared, the proposed approach and Limma leverage measurements in all conditions for the inference of underlying abundance.
-We evaluated the impact with such adjustment by computer simulation, considering the following factors: with/without protein-level changes, effect size, number of replicates and number of conditions.
-Further details of how the simulations were generated can be found in Supplementary Sec.
-\ref{supp-sec:sim}.
-
-\paragraph*{Computer simulation 1 - label free}
+\paragraph*{Dataset 1 : Computer simulation 1 - label free}
 
@@
 
-\paragraph*{Computer simulation 2 - missing values and low replicates}
+\paragraph*{Dataset 2 : Computer simulation 2 - missing values and low replicates}
 
@@
 
-\subsection*{SpikeIn benchmark - KGG enriched \todo{label-free?}}
+\medskip
+
+\noindent{Further details of how the simulations were generated can be found in Supplementary Sec. \ref{supp-sec:sim}. }
+
+\subsection*{Dataset 3 : SpikeIn benchmark - label-free - KGG enriched}
 
@@
 
-\subsection*{Human-1mix-TMT - ubiquitination}
+\subsection*{Dataset 4 : Human-1mix-TMT - ubiquitination}
 
-The proposed approach was evaluated on an experiment where Shigella ubiquitin ligase IpaH7.8 was shown to inhibit the protein gasdermin D (GSDMD).\cite{LUCHETTI2021} Multiplex proteomics was used to quantify the abundance of total protein and ubiquitination in human epithelial cells.
+The proposed approach was evaluated on an experiment where Shigella ubiquitin ligase IpaH7.8 was shown to inhibit the protein gasdermin D (GSDMD)\cite{LUCHETTI2021}.
+Multiplex proteomics was used to quantify the abundance of total protein and ubiquitination in human epithelial cells.
 Cells were either infected or uninfected with IpaH7.8-deficient Shigella flexneri and measurements were taken at different time periods.
@@
 
-\subsection*{Mouse-2mix-TMT - phosphorylation}
+Further details on the design of this experiment are provided in Supplementary Section \ref{supp-sec:ipah}.
 
-Here an experiment targeting primary murine macrophages infected with Shigella flexneri (S.flexneri).
-\cite{Maculins} Multiplex proteomics was used to quantify the abundance of total protein, and phosphorylation in wild type (WT) and ATG16L1-deficient (cKO) samples, uninfected and uninfected with S.flexneri.
+\subsection*{Dataset 5 : Mouse-2mix-TMT - phosphorylation}
+
+Here an experiment targeting primary murine macrophages infected with Shigella flexneri (S.flexneri)\cite{Maculins}.
+Multiplex proteomics was used to quantify the abundance of total protein, and phosphorylation in wild type (WT) and ATG16L1-deficient (cKO) samples, uninfected and uninfected with S.flexneri.
 The abundance of total protein and post-translation modifications were quantified at three time points, uninfected, early infection (45-60 minutes), and late infection (3-3.5 hours).
@@
 
-\subsection*{Human label free quantification - no global profiling run}
+Further details on the design of this experiment are provided in Supplementary Sec. \ref{supp-sec:shigella}.
+
+\subsection*{Dataset 6 : Human label free quantification - no global profiling run}
 
@@
 The approach takes as input intensities of individual features from modified peptides, or intensity ratios of modified and unmodified peptide features, and compares the mean abundance of a PTM site from one condition to another.
-Modifications of the t-test such as moderated t-test with Limma were also proposed.\cite{Zhu} While simple, the approach does not fully account for the sources of variations, and it is not directly applicable to experiments with complex designs, e.g., comparisons of multiple conditions, acquisition in multiple batches, etc.
+Modifications of the t-test such as moderated t-test with Limma were also proposed\cite{Zhu}.
+While simple, the approach does not fully account for the sources of variations, and it is not directly applicable to experiments with complex designs, e.g., comparisons of multiple conditions, acquisition in multiple batches, etc.
 Additionally, while this approach can be applied to experiments targeting PTMs, there is not a self contained, straight forward implementation of the methods, making application challenging.
 
-Isobar-PTM was developed for experiments with MS/MS quantitative strategies that employ isobaric labels such as tandem mass tags (TMT) and isobaric tag for relative and absolute quantification (iTRAQ).\cite{Breitwieser:2013} Isobar-PTM expresses MS measurements with a linear model and performs adjustment with respect to protein abundance using the difference between log-ratio of modified peptides in two channels and log-ratio of protein level.
+Isobar-PTM was developed for experiments with MS/MS quantitative strategies that employ isobaric labels such as tandem mass tags (TMT) and isobaric tag for relative and absolute quantification (iTRAQ)\cite{Breitwieser:2013}.
+Isobar-PTM expresses MS measurements with a linear model and performs adjustment with respect to protein abundance using the difference between log-ratio of modified peptides in two channels and log-ratio of protein level.
 The modeling framework, however, is not applicable for either label-free workflows or experiments with complex designs.
 
-Further details \todo{clarify which details}  on existing methods can be seen in Supplementary \ref{supp-sec:ttest}.
+Further details on the specific equations and workflow of existing methods can be seen in Supplementary \ref{supp-sec:ttest}.
 
@@
 In addition mixture may be included for features acquired via tandem mass tag (TMT) methods.
-The model parameters are estimated using the split-plot approach as in $MSstats$, where the feature log-intensities are first summarized into a single value per site per run in the subplot model, and the site-level summaries are then used for the inference of the PTM site abundance.\cite{Choi:2014} In the site-level summarization, Tukey's median polish (TMP), a simple and robust procedure is applied to iteratively fit a two-way additive model with the effects of run and feature, which in turn summarizes the log-intensities for each site.\cite{Tukey:1977} After summarization, the inference of the PTM site abundance in each condition is carried through fitting a model based on the family of linear mixed-effects models, taking into account the specific experimental design \todo{weird reference?} \ref{fig:data-structure}.\cite{Bolker2009} \cite{Faraway:2006} Statistical modeling and quantification for global proteomics data are performed by the same procedure as for PTM data.
+The model parameters are estimated using the split-plot approach as in $MSstats$, where the feature log-intensities are first summarized into a single value per site per run in the subplot model, and the site-level summaries are then used for the inference of the PTM site abundance\cite{Choi:2014}.
+In the site-level summarization, Tukey's median polish (TMP), a simple and robust procedure is applied to iteratively fit a two-way additive model with the effects of run and feature, which in turn summarizes the log-intensities for each site\cite{Tukey:1977}.
+After summarization, the inference of the PTM site abundance in each condition is carried through fitting a model based on the family of linear mixed-effects models\cite{Bolker2009} \cite{Faraway:2006}.
+Statistical modeling and quantification for global proteomics data are performed by the same procedure as for PTM data.
 
@@
 The test statistic for the hypothesis testing is the ratio of the estimate of the adjusted difference to its SE.
-To determine the statistical significance of the difference in terms of p-value, the test statistic is compared against the t distribution with degrees of freedom approximated by the Satterthwaite method.
-\cite{satterthwaite:1946} Adjustment for multiple comparisons is performed using the Benjamini-Hochberg procedure to control the false discovery rate at a desired level, e.g., $0.05$.\cite{Benjamini:1995} More details are provided in Supplementary Sec \ref{supp-sec:test}
+To determine the statistical significance of the difference in terms of p-value, the test statistic is compared against the t distribution with degrees of freedom approximated by the Satterthwaite method\cite{satterthwaite:1946}.
+Adjustment for multiple comparisons is performed using the Benjamini-Hochberg procedure to control the false discovery rate at a desired level, e.g., $0.05$\cite{Benjamini:1995}.
+More details are provided in Supplementary Sec \ref{supp-sec:test}
 
@@
 When values are imputed, it is assumed they are missing for reasons of low abundance.
-Missing features are imputed in each MS run using the Accelerated Time Failure (AFT) model.\cite{Tukey:1977} In order to impute a feature's missing values, the feature must be present in at least one MS Run.
+Missing features are imputed in each MS run using the Accelerated Time Failure (AFT) model\cite{Tukey:1977}.
+In order to impute a feature's missing values, the feature must be present in at least one MS Run.
 If the feature is not present across all MS Runs, the value will be left missing.
@@
 To account for mixture variance, first feature intensities are summarized again using Tukey's Median Polish and missing values are imputed, if desired, then a new linear mixed effects model with an added term for the mixture is included.
-The workflow and statisical model described generally follow the methods used for modeling proteins in $MSstatsTMT$.
-\cite{Huang:2020} These methods are repeated to quantify both modified and unmodified peptide abundance.
+The workflow and statisical model described generally follow the methods used for modeling proteins in $MSstatsTMT$\cite{Huang:2020}.
+These methods are repeated to quantify both modified and unmodified peptide abundance.
 Once both modified and unmodified peptides are modeled, the PTM model is adjusted for changes in global protein abundance using the same methods described previously.
@@
 
+\paragraph{Code availability}\mbox{}\\
+
+\noindent{{\bf Bioconductor:} http://www.bioconductor.org/packages/release/bioc/html/MSstatsPTM.html}
+
+\noindent{{\bf Github:} https://github.com/Vitek-Lab/MSstatsPTM}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \subsection*{Evaluation}
-
-\todo{In the caption of each figure, could you clarify which dataset the figure refers to?}
 
@@
 
-\subsubsection*{Computer simulation 1 - label free}
+\subsubsection*{Dataset 1 : Computer simulation 1 - label free}
 
+The proposed approach was evaluated against two-sample t-test and Limma. 
 Here we see that not accounting for changes in unmodified peptide level results in a high false discovery rate, seen in Figure \ref{fig:fdr_all_models}.
@@
 In contrast, the proposed approach and Limma leverage all available information, which resulted in improved power with increased number of conditions.
-Further results are provided in Supplementary Figures \ref{supp-fig:sim1_fpr} to \ref{supp-fig:sim1_recall}.
 
-\subsubsection*{Computer simulation 2 - missing values and low replicates}
+\subsubsection*{Dataset 2 : Computer simulation 2 - missing values and low replicates}
 
@@
 This fold change comparison can be seen in Figure \ref{fig:sim_boxplot}.
-Further results are provided in Supplementary Figures \ref{supp-fig:sim2_recall} to \ref{supp-fig:fc_boxplot}.
+
+Further details reviewing the results of each simulation and their comparison with respect to method performance can be seen in Supplementary Section \ref{supp-sec:sim}.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection*{SpikeIn benchmark - KGG enriched}
+\subsubsection*{Dataset 3 : SpikeIn benchmark - label-free - KGG enriched}
 
@@
 
-Further results of this experiment are provided in Supplementary Sec. \ref{supp-sec:benchmark}.
+Further results of this experiment, including a comparison to t-test and fold change distribution of each method before and after adjustment, are provided in Supplementary Section \ref{supp-sec:benchmark}.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection*{Human-1mix-TMT - ubiquitination}
+\subsubsection*{Dataset 4 : Human-1mix-TMT - ubiquitination}
 
@@
 
-Further details and results of this experiment are provided in Supplementary Sec. \ref{supp-sec:ipah}.
-
-\subsubsection*{Mouse-2mix-TMT - phosphorylation}
+\subsubsection*{Dataset 5 : Mouse-2mix-TMT - phosphorylation}
 
@@
 
-Further details and results of this experiment are provided in Supplementary Sec. \ref{supp-sec:shigella}.  \todo{Maybe clarify what kind of results}
-
-\subsubsection*{Human label free quantification - no global profiling run}
+\subsubsection*{Dataset 6 : Human label free quantification - no global profiling run}
 
@@
 Using the proposed method we get increased information that would otherwise have been missed.
-
-Further details and results of this experiment are provided in Supplementary Sec. \ref{supp-sec:usp30}.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Code availability}
-
-\todo{This belongs to the 'implementation' section}
-
-{ \bf Bioconductor:} http://www.bioconductor.org/packages/release/bioc/html/MSstatsPTM.html
-
-\noindent{\bf{Github:} https://github.com/Vitek-Lab/MSstatsPTM}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%    REFERENCES   %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \includegraphics[scale=.8]{images/fig1.png}
-\caption{\todo{Meena?: Write about how MS for PTMs is run?}}
+\caption{\todo{Meena?: Can you write about how MS for PTMs is run?}}
 \label{fig:workflow}
@@
 \includegraphics[width=1\textwidth]{images/sim1_FDR_all_models.png}
-\caption{All the considered methods in simulation 1 correctly calibrated FDR when adjusting for changes in protein abundance.
+\caption{Dataset 1 : Computer simulation 1 - label free.
+All the considered methods in the first computer simulation correctly calibrated FDR when adjusting for changes in protein abundance.
 In comparison, the methods without accounting for the protein-level changes resulted in off-target, high false positive rates.
@@
 \includegraphics[width=1\textwidth]{images/sim3_Accuracy.png}
-\caption{The advantage of using the proposed approach is apparent when looking at simulation 2, which includes limited observations and the presence of missing values.
+\caption{Dataset 2 : Computer simulation 2 - missing values and low replicates.
+The advantage of using the proposed approach is apparent when looking at the second computer simulation, which includes limited observations and the presence of missing values.
 In overall accuracy, the proposed method performs stronger than Limma and t-test in nearly every model.
@@
 \includegraphics[width=1\textwidth]{images/sim3_FC_boxplot.png}
-\caption{In the second simulation with missing values and low PTM features all methods correctly estimated the fold change with a median log change of .75.
+\caption{Dataset 2 : Computer simulation 2 - missing values and low replicates.
+In the second computer simulation with missing values and low PTM features all methods correctly estimated the fold change with a median log change of .75.
 The proposed method in this simulation had a visibly tighter distribution around the median when compared to Limma and $t$-test.
@@
 \includegraphics[width=1\textwidth]{images/spike_in_msstatsptm_volcano.png}
-\caption{Using the proposed method to model the benchmark experiment the spike in peptides (colored red) do not follow the expected log fold change before adjustment.
+\caption{Dataset 3 : SpikeIn benchmark - label-free - KGG enriched.
+Using the proposed method to model the benchmark experiment the spike in peptides (colored red) do not follow the expected log fold change before adjustment.
 After adjusting for changes in overall protein abundance the spike in peptides are more in line with expectation.
@@
 \includegraphics[width=1\textwidth]{images/spike_in_limma_volcano.png}
-\caption{Limma method used to model the spike in experiment.
+\caption{Dataset 3 : SpikeIn benchmark - label-free - KGG enriched.
+Here the Limma method was used to model the spike in experiment.
 Using Limma the spike in peptides follow the expected log fold change better after adjusting for changes in protein level.
@@
 \includegraphics[width=1\textwidth]{images/IpaH_prof_plot.png}
-\caption{The overlap of differencial modified peptides for the PTM model with and without global protein level adjustment.
+\caption{Dataset 4 : Human-1mix-TMT - ubiquitination.
+The overlap of differencial modified peptides for the PTM model with and without global protein level adjustment.
 More PTMs became insignificant after adjustment then became significant.
@@
 \includegraphics[width=1\textwidth]{images/IpaH_prof_plot.png}
-\caption{Comparing the global profiling of protein $GSDMD\_HUMAN|P57764$ with the ubiquitination of the protein at site $K62$.
+\caption{Dataset 4 : Human-1mix-TMT - ubiquitination.
+Comparing the global profiling of protein $GSDMD\_HUMAN|P57764$ with the ubiquitination of the protein at site $K62$.
 The individual PSM features are shown in grey, while the feature summarization is shown in red.
@@
 \includegraphics[height=.65\textwidth]{images/shig_venn_diagramm.png}
-\caption{The overlap of differencial modified peptides between the PTM model with and without global protein level adjustment.
+\caption{Dataset 5 : Mouse-2mix-TMT - phosphorylation.
+The overlap of differencial modified peptides between the PTM model with and without global protein level adjustment.
 Again more PTMs became insignificant after adjustment then became significant.}
@@
  \end{subfigure}
-\caption{Profile plots for PTMs P22893\_S178 and E9PVX6\_T215 in the Shigella flexneri experiment.}
+\caption{Dataset 5 : Mouse-2mix-TMT - phosphorylation. Profile plots for PTMs P22893\_S178 and E9PVX6\_T215 in the Shigella flexneri experiment.}
 \label{fig:shigella_profile}
@@
  \end{subfigure}
- \caption{Overlap of significant PTMs before and after protein adjustment.}
+ \caption{Dataset 6 : Human label free quantification - no global profiling run. Overlap of significant PTMs before and after protein adjustment.}
 \label{fig:usp30_ven_d}
@@
 \includegraphics[width=1\textwidth]{images/USP30_profile_plot.png}
-\caption{Comparing the global profiling of protein $P52209$ with the modification of the protein at site $K059$.
+\caption{Dataset 6 : Human label free quantification - no global profiling run.
+Comparing the global profiling of protein $P52209$ with the modification of the protein at site $K059$.
 The modification appears generally unchanged between conditions, whereas the global profiling run shows the CCCP and Combo conditions at a higher relative abundance compared to the Control and USP30\_OE.

### 2021-10-07 supplementary/ptm_sm.tex
Updated main per Olgas comments, added sample size analysis

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection{Computer simulation 1 - label free experiment}
+\subsubsection{Dataset 1 : Computer simulation 1 - label free experiment}
 
@@
 
-\subsubsection{Computer simulation 2 - missing values and low replicates}
+\subsubsection{Dataset 2 : Computer simulation 2 - missing values and low replicates}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{SpikeIn benchmark - KGG enriched}
+\subsection{Dataset 3 : SpikeIn benchmark - label-free - KGG enriched}
 \label{sec:benchmark}
@@
 
-\subsection{Human-1mix-TMT - ubiquitination}
+\subsection{Dataset 4 : Human-1mix-TMT - ubiquitination}
 \label{sec:ipah}
@@
 \clearpage
-\subsection{Mouse-2mix-TMT - phosphorylation}
+\subsection{Dataset 5 : Mouse-2mix-TMT - phosphorylation}
 \label{sec:shigella}
@@
 
-\subsection{Human label free quantification - no global profiling run}
+\subsection{Dataset 6 : Human label free quantification - no global profiling run}
 \label{sec:usp30}
@@
 \section{Sample size calculation and power analysis}
-\todo{Devon: Update this section per Olga's commentsv}
 
 The proposed approach allows us to calculate the sample size needed to achieve a desired statistical power as described in Section \ref{sec:design}.
-As described in Section \ref{sec:test} the proposed method adjusts for the underlying protein abundance in the PTM significance analysis, which corrects the confounding factor with a cost of increased variation.
+The proposed method adjusts for the underlying protein abundance in the PTM significance analysis, which corrects the confounding factor with a cost of increased variation.
 When the variation is increased, the experiment requires a larger number of replicates to reach the same power.
-Thus the statistical power is dependent on the variance from both the modified peptide and unmodifed protein as well as the number of replicates in both the modified and unmodified runs.
-
-We compared the statistical power in experiments with different numbers of replicates, variance, and fold change for both the modified and unmodified runs.
+Thus the statistical power is dependent on the variance from both the modified peptide and unmodifed protein as well as the number of replicates in both runs.
+
+We compared the statistical power in experiments with differing numbers of replicates, variance, and fold change for both the modified and unmodified runs.
+For variance we used the variance from the biological experiments to determine what values to set.
+In Dataset 4 and 5 the variance of the PTM was generally higher than the global protein.
+In contrast, in dataset 6 the variance of the PTM and Protein was generally the same.
+Because of this we will analyze the power of our experiments when the PTM variance is higher than the protein, and also when they are the same.
+For the variance numbers, in the case when the PTM and protein are the same we chose a variance of .15.
+When the PTM is higher than the protein we chose a PTM variance of .2 and a protein variance of .1.
+These numbers were derived from the biological experiments in this paper.
+
 The results can be seen in \sfigref{fig:power_sd_combo}
@@
 \centering
-\includegraphics[width=.9\textwidth]{sim_new/power_analysis_sd_combined}
-\caption{}
+ \begin{subfigure}{\textwidth}
+ \centering
+	\includegraphics[width=.9\textwidth]{sim_new/same_var_power}
+	\caption{blah}
+ \end{subfigure}\vspace{-5mm}
+ \begin{subfigure}{\textwidth}
+ \centering
+	\includegraphics[width=.9\textwidth]{sim_new/high_ptm_var_power}
+	\caption{blah}
+ \end{subfigure}
+ \caption{Power analysis of experiments with differing variances.}
 \label{fig:power_sd_combo}

### 2021-10-07 code/IpaH_analysis.R

@@ -25,6 +25,7 @@ protein_summarization_combined <- rbindlist(list(protein_input, protein_summariz
 
 
 unadj_ptm_model <- read.csv("../data/Ipah_PTM_test_results.txt", sep = "\t")
+protein_model <- read.csv("../data/Ipah_Global_TMT11_msstats_test_results.txt", sep = "\t")
 adj_ptm_model <- read.csv("../data/Ipah_PTM_Adjusted_test_results.txt", sep = "\t")
 
 combined_models <- merge(adj_ptm_model, unadj_ptm_model, all.x=TRUE, all.y=TRUE, by = c("Protein", "Label"))
@@ -32,6 +33,9 @@ combined_models <- merge(adj_ptm_model, unadj_ptm_model, all.x=TRUE, all.y=TRUE,
 combined_models$FC_Diff <- abs(combined_models$log2FC.x - combined_models$log2FC.y)
 combined_models %>% arrange(desc(FC_Diff))
 
+## SE analysis
+unadj_ptm_model %>% ggplot + geom_boxplot(aes(y = SE))
+protein_model %>% ggplot + geom_boxplot(aes(y = SE))
 
 ## Profile Plots
 "GSDMD_HUMAN|P57764_K204"

### 2021-10-07 code/sample_size_analysis.R

@@ -1,102 +1,68 @@
 
 library(tidyverse)
+library(ggthemes)
+library(data.table)
 
 ## Calculations w/o feature variations -----------------------------------------
 
 sigma_model <- function(ptm_var, ptm_n, prot_var, prot_n){
-  simga <- sqrt((ptm_var/(.25*ptm_n)) + (prot_var/(.25*prot_n)))
+  simga <- sqrt((ptm_var/ptm_n) + (prot_var/prot_n))
   return(simga)
 }
 
 ## Define adjustable metrics
 low <- .1
+mid <- .2
 high <- .3
-n_sample_list <- c(3,5,10)
+n_sample_list <- c(2,5)
 
 ## Hard metrics
 FDR <- 0.05
 m0_m1 <- 99
 delta <- seq(1, 2, 0.025)
 
-var_list <- numeric(9)
+bind_list <- c()
+
 i <- 1
 for (x in seq_along(n_sample_list)){
   for (y in seq_along(n_sample_list)){
-    var_list[i] <- sigma_model(low, n_sample_list[[x]], low, n_sample_list[[y]])
-    i <- i+1
-  }
-}
-
-t <- delta / var_list[[9]]
-
-powerTemp <- seq(0, 1, 0.01)
-power <- numeric(length(t))
-
-for (i in seq_along(t)) {
-  diff <- qnorm(powerTemp) + qnorm(1 - powerTemp * FDR / (1 + (1 - FDR) * m0_m1) / 2) - t[i]
-  min(abs(diff), na.rm = TRUE)
-  power[i] = powerTemp[order(abs(diff))][1]
-}
-power
-
-
-## Simulate feature effect -----------------------------------------------------
-
-n_feature_list <- 1:10
-
-final_var <- c()
-
-for (n in seq_along(n_feature_list)){
-  test <- c()
-  for (i in 1:1000) {
-    r1 <- c()
-    r2 <- c()
-
-    for (x in seq_len(3)){
-
-      t1 <- rnorm(1, mean = 0, .5)
-      t2 <- rnorm(1, mean = 2, .5)
-
-      for (y in seq_len(n_feature_list[[n]])){
-        f1 <- rnorm(y, mean = t1, .001)
-        f2 <- rnorm(y, mean = t2, .001)
-      }
-
-      r1 <- c(r1, log(sum(f1^2)))
-      r2 <- c(r2, log(sum(f2^2)))
-
[…]

### 2021-10-11 main/MSstatsPTM_Main.tex
Added sample size and updated model section

@@
 Typically, because of the enrichment step for PTMs, very few of those features are present in original LC-MS runs.
-For more accurate estimation of protein abundance, separate global proteomics data of unenriched samples are often acquired.
+For more accurate estimation of protein abundance, separate global proteomics data of unenriched samples are recommended.
 If unmodified features are unavailable for any given modified feature, unmodified intensity adjustment cannot be performed.
 As different levels of variability are present in the data, the log-intensities of the features for modified and unmodified peptides are modeled separately using two linear mixed models.
+An overview of the steps of the proposed approach are illustrated in Figure \ref{fig:msstatsptm_design}.
 
@@
 For each PTM site, the feature log-intensities of the modified peptides spanning the site are expressed using a linear mixed model in consideration of the effects of condition, run, feature and interaction between run and feature.
-In addition mixture may be included for features acquired via tandem mass tag (TMT) methods.
 The model parameters are estimated using the split-plot approach as in $MSstats$, where the feature log-intensities are first summarized into a single value per site per run in the subplot model, and the site-level summaries are then used for the inference of the PTM site abundance\cite{Choi:2014}.
+This approach allows the method to be extended to cases where the experimental design in inbalanced, and where additional sources of variation are present.
 In the site-level summarization, Tukey's median polish (TMP), a simple and robust procedure is applied to iteratively fit a two-way additive model with the effects of run and feature, which in turn summarizes the log-intensities for each site\cite{Tukey:1977}.
@@
 Detection of differentially modified PTM sites is performed through testing the null hypothesis of ‘no change’ against the alternative.
-The null hypothesis states that there is no difference in log-abundance of the PTM site between conditions, adjusted with respect to protein abundance \ref{fig:data-structure}.
+The null hypothesis states that there is no difference in log-abundance of the PTM site between conditions, adjusted with respect to protein abundance.
+Notationally, the hypothesis test is formulated as follows
+
+\begin{align*}
+H_{0}: \Delta = \left( \mu_{i}^{\ast} - \mu_{i'}^{\ast} \right) - \left( \mu_{i} - \mu_{i'} \right) &= 0 \\
+H_{a}: \Delta = \left( \mu_{i}^{\ast} - \mu_{i'}^{\ast} \right) - \left( \mu_{i} - \mu_{i'} \right) &\neq 0
+\end{align*}
+
 Specifically, the adjusted difference is given by the difference in log-abundance of the PTM site, subtracted by the difference in log-abundance of the underlying protein, which is equivalent to the log of the ratio of PTM abundance difference to protein abundance difference.
 The estimate of the adjusted difference and the standard error (SE) of the estimate are obtained by combining the difference estimates and the associated SEs from both counterparts.
+
+\[
+SE(\hat{\Delta}) = \left[ \left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right) \right]^{1/2}.
+\]
+
+Finally the degrees of freedom are given by
+
+\[
+\left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right)^2 \bigg/
+\left( \frac{\hat{\sigma}_{\gamma^{\ast}}^{4}}{\mathrm{df}(\gamma^{\ast})} + \frac{\hat{\sigma}_{\gamma}^{4}}{ \mathrm{df}(\gamma)} \right).
+\]
 
@@
 Adjustment for multiple comparisons is performed using the Benjamini-Hochberg procedure to control the false discovery rate at a desired level, e.g., $0.05$\cite{Benjamini:1995}.
-More details are provided in Supplementary Sec \ref{supp-sec:test}
+More details on how the PTM model is adjusted are provided in Supplementary Sec \ref{supp-sec:test}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection*{Extension to TMT experiments}
+\subsubsection*{Extension to complex designs}
 
-The statistical modeling approaches discussed above can also be extended to Tandem Mass Tag (TMT) labeling methods.
+The statistical modeling approaches discussed above can also be extended to complex designs with additional sources of variation and inbalanced designs.
+One specific complex design the method can be applied to is Tandem Mass Tag (TMT) labeling methods.
 TMT experiments introduce an additional source of variation in the form of different mixtures.
@@
 
-For more information about the extension to Tandem Mass Tag experiments see Supplementary Sec. \ref{supp-sec:tmtmethod}.
+For more information about the extension to complex designs and Tandem Mass Tag experiments see Supplementary Sec. \ref{supp-sec:complex_methods}.
 
@@
 
-\todo{Devon: Update this section in main and supp}
+Here we analyze the sample size needed to achieve a desired statistical power.
+The proposed approach corrects for confounding between the modified and unmodified peptides by adjusting the modified peptide for changes in the unmodified.
+This adjustment comes at a cost of increased variation.
+This can be seen in the calculation for variance in Section \ref{sec:adjustment_summary}.
+When the variation is increased, the experiment requires a larger number of replicates to reach the same power.
+Thus the statistical power is dependent on the variance from both the modified peptide and unmodifed protein as well as the number of replicates in both runs.
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+We compared the statistical power in experiments with differing numbers of replicates, variance, and fold change for both the modified and unmodified runs.
+In terms of the number of replicates, we tested scenarios with equal replicates in both the modified and unmodified runs, as well as scenarios where the replicates differed between runs.
+For variance we used the biological experiments to determine what scenarios and values to test.
+In dataset 4 and 5 the variance of the PTM was higher than the global protein.
+In dataset 6 the variance of the PTM and Protein was generally the same due to not using a separate global profiling run.
+We mimicked these scenarios and analyzed the power of experiments when the PTM variance was higher than the protein and when they were equal.
+When the PTM and protein were the same we chose a variance of .15, whereas when the PTM was higher than the protein we chose a PTM variance of .2 and a protein variance of .1.
+
+The results show that when the variance and replicates were equal, higher replicates predictably lead to higher power.
+In cases where the replicates were unbalanced, but the variance was still the same, it did not matter if there were more replicates in the modified or unmodified runs.
+As long as there were the same overall number of replicates the results were the same.
+In comparison, with differing variance and equal replicates, higher replicates still lead to higher power.
+When the replicates were unbalanced and the variance was higher in the PTM, there was more power when there were more replicates in the PTM than the protein.
+It is clearly more important to have high replicates for the PTM run than the unmodified protein, when the PTM variance is higher.
+In cases where the number of replicates has to be limited, it is better to weight them on the PTM side.
+
+The results of the power and sample size analysis can be seen in Figure \ref{fig:power_sd_combo}.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \label{fig:data-structure}
+\end{figure}
+
+\begin{figure}[ht]
+\centering
+\includegraphics[scale=.8]{images/MSstatsPTM_design.png}
+\caption{blah.}
+\label{fig:msstatsptm_design}
 \end{figure}
@@
 
+\begin{figure}[h!]
+\centering
+ \begin{subfigure}{\textwidth}
+ \centering
+	\includegraphics[width=.725\textwidth]{images/same_var_power}
+	\caption{Here we can see the power of an experiment targeting PTMs with the same variance, .15, for the modified and unmodified peptides.
+Predictably when the replicates are high for both modified and unmodified peptides the power is much higher.
+Conversely at low replicates for each the power is much lower.
+The interesting part of this chart is when the replicates are different between runs.
+With equal variance, it does not matter if the PTM replicates or protein replicates are higher.}
+ \end{subfigure}\vspace{-5mm}
+ \begin{subfigure}{\textwidth}
+ \centering
+	\includegraphics[width=.725\textwidth]{images/high_ptm_var_power}
+	\caption{In this chart the variance for the PTM is higher than the unmodified protein.
+The PTM variance is .2, while the unmodified protein variance is .1.
+With equal replicates the results are the same as above, more replicates equals more power.
+When the replicates are not the same we can clearly see that having more replicates for the PTM leads to higher power.}
+ \end{subfigure}
+ \caption{Power analysis of experiments with differing variances.}
+\label{fig:power_sd_combo}
+\end{figure}
+
 %{\bf Figure 4} False positive rate and statistical power in PTM significance analysis by the proposed approach, t-test with protein-level adjustment, and t-test without protein-level adjustment.

### 2021-10-11 supplementary/ptm_sm.tex
Added sample size and updated model section

@@
 The PTM site is represented by $K$ spectral features (peptide ions, distinguished by their cleavage residues and charge states).
-The log-intensity (base 2) of Feature $k$, in Run $j$ of Condition $i$ is denoted by $y_{ijk}$.
+The log-intensity (base 2) of Feature $k$, in Run $j$ of Condition $i$ is denoted by $y_{ijk}^{\ast}$.
 To account for the underlying protein abundance, features corresponding to the unmodified peptides from the same protein are considered together, except those unmodified peptides containing a modified site to avoid the confounding effect due to the PTM.
-The log-intensity of Feature $l$ from the unmodified peptides in the same run is denoted by $y_{ijl}^{\ast}$.
+The log-intensity of Feature $l$ from the unmodified peptides in the same run is denoted by $y_{ijl}$.
 \sfigref{fig:dtable} shows an example data representation of modified peptide ions at one site and unmodified peptide ions of the same protein.
@@
 \midrule
-\multirow{4}{*}{Modified} & \text{Feature }1 & y_{111} & y_{121} & \dots & y_{1J1} & \dots & y_{I11} & y_{I21} & \dots & y_{IJ1} \\
- & \text{Feature }2 & y_{112} & - & \dots & y_{1J2} & \dots & y_{I12} & y_{I22} & \dots & y_{IJ2}  \\
+\multirow{4}{*}{Modified} & \text{Feature }1 & y_{111}^{\ast} & y_{121}^{\ast} & \dots & y_{1J1}^{\ast} & \dots & y_{I11}^{\ast} & y_{I21}^{\ast} & \dots & y_{IJ1}^{\ast} \\
+ & \text{Feature }2 & y_{112}^{\ast} & - & \dots & y_{1J2}^{\ast} & \dots & y_{I12}^{\ast} & y_{I22}^{\ast} & \dots & y_{IJ2}^{\ast}  \\
  & \dots & \dots & \dots & \dots & \dots & \dots & \dots & \dots & \dots & \dots \\
- & \text{Feature }K & y_{11K} & y_{12K} & \dots & y_{1JK} & \dots & y_{I1K} & y_{I2K} & \dots & y_{IJK} \\
+ & \text{Feature }K & y_{11K}^{\ast} & y_{12K}^{\ast} & \dots & y_{1JK}^{\ast} & \dots & y_{I1K}^{\ast} & y_{I2K}^{\ast} & \dots & y_{IJK}^{\ast} \\
 \midrule
-\multirow{4}{*}{Unmodified} & \text{Feature }1 & y_{111}^{\ast} & y_{121}^{\ast} & \dots & y_{1J1}^{\ast} & \dots & y_{I11}^{\ast} & y_{I21}^{\ast} & \dots & y_{IJ1}^{\ast} \\
- & \text{Feature }2 & y_{112}^{\ast} & y_{122}^{\ast} & \dots & y_{1J2}^{\ast} & \dots & - & - & \dots & -  \\
-% & \text{Feature }2 & y_{112}^{\ast} & y_{122}^{\ast} & \dots & y_{1J2}^{\ast} & \dots & y_{I12}^{\ast} & y_{I22}^{\ast} & \dots & y_{IJ2}^{\ast}  \\
+\multirow{4}{*}{Unmodified} & \text{Feature }1 & y_{111} & y_{121} & \dots & y_{1J1} & \dots & y_{I11} & y_{I21} & \dots & y_{IJ1} \\
+ & \text{Feature }2 & y_{112} & y_{122} & \dots & y_{1J2} & \dots & - & - & \dots & -  \\
+% & \text{Feature }2 & y_{112} & y_{122} & \dots & y_{1J2} & \dots & y_{I12} & y_{I22} & \dots & y_{IJ2}  \\
  & \dots & \dots & \dots & \dots & \dots & \dots & \dots & \dots & \dots & \dots \\
- & \text{Feature }L & y_{11L}^{\ast} & y_{12L}^{\ast} & \dots & y_{1JL}^{\ast} & \dots & y_{I1L}^{\ast} & y_{I2L}^{\ast} & \dots & y_{IJL}^{\ast} \\
+ & \text{Feature }L & y_{11L} & y_{12L} & \dots & y_{1JL} & \dots & y_{I1L} & y_{I2L} & \dots & y_{IJL} \\
 \bottomrule
@@
 \[
-\log \left( \sum_{k=1}^{K} 2^{y_{ijk}} \right).
+\log \left( \sum_{k=1}^{K} 2^{y_{ijk}^{\ast}} \right).
 \]
@@
 \[
-d_{ij} = \log \left( \sum_{k=1}^{K} 2^{y_{ijk}} \right) - \log \left( \sum_{l=1}^{L} 2^{y_{ijl}^{\ast}} \right).
+d_{ij} = \log \left( \sum_{k=1}^{K} 2^{y_{ijk}^{\ast}} \right) - \log \left( \sum_{l=1}^{L} 2^{y_{ijl}} \right).
 \]
@@
 \[
-\hat{\Delta} = \frac{1}{J} d_{i+} - \frac{1}{J} d_{i'+},
-\]
-where $d_{i+} = \sum_{j=1}^{J} d_{ij}$ and the test statistic for the $t$-test is given by $\hat{\Delta} / \mathrm{SE}(\hat{\Delta})$.
+\hat{\Delta} = \frac{1}{J} d_{i+}^{\ast} - \frac{1}{J} d_{i'+},
+\]
+where $d_{i+}^{\ast} = \sum_{j=1}^{J} d_{ij}^{\ast}$ and the test statistic for the $t$-test is given by $\hat{\Delta} / \mathrm{SE}(\hat{\Delta})$.
 The statistical significance of the difference is determined by comparing the test statistic against the $t$ distribution, with degrees of freedom $df=2J-2$ in balanced designs.
@@
 \[
-\hat{\Delta} = \frac{1}{J} d_{i+} - \frac{1}{J} d_{i'+},
+\hat{\Delta} = \frac{1}{J} d_{i+}^{\ast} - \frac{1}{J} d_{i'+},
 \]
@@
 
-To characterize the observed feature intensities, different levels of variations are expressed using linear mixed models in consideration of the following factors: modification, condition, run, and feature.
+To characterize the observed feature intensities, different levels of variations are expressed using linear mixed models in consideration of the following factors in a simple experimental design: modification, condition, run, and feature.
 As different degrees of variability are present in the feature intensities of modified and unmodified peptides, they are expressed by separate models.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Statistical modeling and inference}
-\label{sec:model}
-
-The observed log-intensity of a modified peptide feature is denoted by $y_{ijk}$ and represented as
-\[
-y_{ijk} = \psi + C_{i} + R_{j(i)} + F_{k} + (R \times F)_{ijk},
-\]
-where the effects of condition and feature are modeled as fixed effects: 
-\[
-\sum_{i=1}^{I} C_{i} = 0, \quad \sum_{k=1}^{K} F_{k} = 0,
-\]
-and the effects of run and its interaction with feature are considered as random effects arising from normal distribution with mean $0$: 
-\[
-R_{j(i)} = \gamma_{j(i)} \stackrel{\text{iid}}{\sim} \mathcal{N}(0, \sigma_{\gamma}^{2}), \qquad
-(R \times F)_{ijk} = \epsilon_{ijk} \stackrel{\text{iid}}{\sim} \mathcal{N}(0, \sigma_{\epsilon}^{2}).
-\]
-Similarly, the observed log-intensity of an unmodified peptide feature is denoted by $y_{ijl}^{\ast}$ and represented as 
-\[
-y_{ijl}^{\ast} = \psi^{\ast} + C_{i}^{\ast} + R_{j(i)}^{\ast} + F_{l}^{\ast} + (R \times F)_{ijl}^{\ast}, 
-\]
-where the effects of condition and feature are modeled as fixed effects: 
-\[
-\sum_{i=1}^{I} C_{i}^{\ast} = 0, \quad \sum_{l=1}^{L} F_{l}^{\ast} = 0,
-\]
-and 
-\[
-R_{j(i)}^{\ast} = \gamma_{j(i)}^{\ast} \stackrel{\text{iid}}{\sim} \mathcal{N}(0, \sigma_{\gamma^{\ast}}^{2}), \qquad
-(R \times F)_{ijl}^{\ast} = \epsilon_{ijl}^{\ast} \stackrel{\text{iid}}{\sim} \mathcal{N}(0, \sigma_{\epsilon^{\ast}}^{2}).
-\]
+%\subsection{Statistical modeling and inference}
+%\label{sec:model}
+
+%The observed log-intensity of a modified peptide feature is denoted by $y_{ijk}$ and represented as
+%\[
+%y_{ijk} = \psi + C_{i} + R_{j(i)} + F_{k} + (R \times F)_{ijk},
+%\]
+%where the effects of condition and feature are modeled as fixed effects: 
+%\[
+%\sum_{i=1}^{I} C_{i} = 0, \quad \sum_{k=1}^{K} F_{k} = 0,
+%\]
+%and the effects of run and its interaction with feature are considered as random effects arising from normal distribution with mean $0$: 
+%\[
+%R_{j(i)} = \gamma_{j(i)} \stackrel{\text{iid}}{\sim} \mathcal{N}(0, \sigma_{\gamma}^{2}), \qquad
+%(R \times F)_{ijk} = \epsilon_{ijk} \stackrel{\text{iid}}{\sim} \mathcal{N}(0, \sigma_{\epsilon}^{2}).
+%\]
+%Similarly, the observed log-intensity of an unmodified peptide feature is denoted by $y_{ijl}^{\ast}$ and represented as 
+%\[
+%y_{ijl}^{\ast} = \psi^{\ast} + C_{i}^{\ast} + R_{j(i)}^{\ast} + F_{l}^{\ast} + (R \times F)_{ijl}^{\ast}, 
+%\]
+%where the effects of condition and feature are modeled as fixed effects: 
+%\[
+%\sum_{i=1}^{I} C_{i}^{\ast} = 0, \quad \sum_{l=1}^{L} F_{l}^{\ast} = 0,
+%\]
+%and 
+%\[
+%R_{j(i)}^{\ast} = \gamma_{j(i)}^{\ast} \stackrel{\text{iid}}{\sim} \mathcal{N}(0, \sigma_{\gamma^{\ast}}^{2}), \qquad
+%(R \times F)_{ijl}^{\ast} = \epsilon_{ijl}^{\ast} \stackrel{\text{iid}}{\sim} \mathcal{N}(0, \sigma_{\epsilon^{\ast}}^{2}).
+%\]
 
@@
 Run-level summarization of feature intensities for each PTM site is carried out as in the sub-plot model of MSstats~\cite{choi_etal_14a}, which involves a) imputation of censored missing values, and b) summarization of feature intensities using Tukey's median polish.
-\cite{tukey_77a} The run-level summary for the PTM in Run $j$ of Condition $i$ is denoted by $\hat{y}_{ij}$.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection{Model-based inference of the underlying abundance}
+\cite{tukey_77a} The run-level summary for the PTM in Run $j$ of Condition $i$ is denoted by $\hat{y}_{ij}^{\ast}$.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsubsection{Model-based inference of the underlying abundance in a simple balanced experiment}
 \label{sec:infer}
 
-The PTM abundance in each run is represented as 
+In the case of a balanced design with multiple conditions and without technical replicates, the PTM abundance in each run is represented as 
+\[
+\hat{y}_{ij}^{\ast} = \psi^{\ast} + C_{i}^{\ast} + R_{j(i)}^{\ast},
+\]
+where $\sum_{i=1}^{I} C_{i}^{\ast} = 0$, $R_{j(i)}^{\ast} = \gamma_{j(i)}^{\ast} \stackrel{\text{iid}}{\sim} \mathcal{N}(0, \sigma_{\gamma^{\ast}}^{2})$. 
+Similarly, the protein abundance in each run is expressed as
 \[
@@
 where $\sum_{i=1}^{I} C_{i} = 0$, $R_{j(i)} = \gamma_{j(i)} \stackrel{\text{iid}}{\sim} \mathcal{N}(0, \sigma_{\gamma}^{2})$.
-Similarly, the protein abundance in each run is expressed as
-\[
-\hat{y}_{ij} = \psi^{\ast} + C_{i}^{\ast} + R_{j(i)}^{\ast},
-\]
-where $\sum_{i=1}^{I} C_{i}^{\ast} = 0$, $R_{j(i)}^{\ast} = \gamma_{j(i)}^{\ast} \stackrel{\text{iid}}{\sim} \mathcal{N}(0, \sigma_{\gamma^{\ast}}^{2})$.
-The expected values of log-abundances of the PTM and protein in Condition $i$ are denoted by $\mu_{i}$ and $\mu_{i}^{\ast}$, respectively, and the values are estimated as:
+The expected values of log-abundances of the PTM and protein in Condition $i$ are denoted by $\mu_{i}^{\ast}$ and $\mu_{i}$, respectively, and the values are estimated as:
 \begin{align*}
-\hat{\mu}_{i} &= \hat{\psi} + \hat{C}_{i} = \frac{1}{J} \hat{y}_{i+} \\
-\hat{\mu}_{i}^{\ast} &= \hat{\psi}^{\ast} + \hat{C}_{i}^{\ast} = \frac{1}{J} \hat{y}_{i+}^{\ast},
+\hat{\mu}_{i}^{\ast} &= \hat{\psi}^{\ast} + \hat{C}_{i}^{\ast} = \frac{1}{J} \hat{y}_{i+}^{\ast} \\
+\hat{\mu}_{i} &= \hat{\psi} + \hat{C}_{i} = \frac{1}{J} \hat{y}_{i+},
 \end{align*}
-where the standard errors of the estimates are $\mathrm{SE}(\hat{\mu}_{i}) = (\hat{\sigma}_{\gamma}^{2} / J)^{1/2}$ and $\mathrm{SE}(\hat{\mu}_{i}^{\ast}) = (\hat{\sigma}_{\gamma^{\ast}}^{2} / J)^{1/2}$.
+where the standard errors of the estimates are $\mathrm{SE^{\ast}}(\hat{\mu}_{i}) = (\hat{\sigma}_{\gamma^{\ast}}^{2} / J)^{1/2}$ and $\mathrm{SE}(\hat{\mu}_{i}) = (\hat{\sigma}_{\gamma^{\ast}}^{2} / J)^{1/2}$.
 %\[
@@
 %\]
-Based on the estimates $\hat{\mu}_{i}$ and $\hat{\mu}_{i}^{\ast}$, the adjusted log-abundance of the PTM is given by $(\hat{\mu}_{i} - \hat{\mu}_{i}^{\ast})$ and the standard error of the estimate is 
-\[
-\left[ \frac{1}{J} \left( \hat{\sigma}_{\gamma}^{2} + \hat{\sigma}_{\gamma^{\ast}}^{2} \right) \right]^{1/2}.
+Based on the estimates $\hat{\mu}_{i}^{\ast}$ and $\hat{\mu}_{i}$, the adjusted log-abundance of the PTM is given by $(\hat{\mu}_{i}^{\ast} - \hat{\mu}_{i})$ and the standard error of the estimate is 
+\[
+\left[ \frac{1}{J} \left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right) \right]^{1/2}.
 \]
@@
 \begin{align*}
-H_{0}: \Delta = \left( \mu_{i} - \mu_{i'} \right) - \left( \mu_{i}^{\ast} - \mu_{i'}^{\ast} \right) &= 0 \\
-H_{a}: \Delta = \left( \mu_{i} - \mu_{i'} \right) - \left( \mu_{i}^{\ast} - \mu_{i'}^{\ast} \right) &\neq 0
+H_{0}: \Delta = \left( \mu_{i}^{\ast} - \mu_{i'}^{\ast} \right) - \left( \mu_{i} - \mu_{i'} \right) &= 0 \\
+H_{a}: \Delta = \left( \mu_{i}^{\ast} - \mu_{i'}^{\ast} \right) - \left( \mu_{i} - \mu_{i'} \right) &\neq 0
 \end{align*}
-The log-fold change in the adjusted PTM abundance, $\Delta$, is estimated by 
-\[
-\hat{\Delta} = \left[ \frac{1}{J} \left( \hat{y}_{i+} - \hat{y}_{i'+} \right) \right] - \left[ \frac{1}{J} \left( \hat{y}_{i+}^{\ast} - \hat{y}_{i'+}^{\ast} \right) \right],
+In a balanced design, the log-fold change in the adjusted PTM abundance, $\Delta$, is estimated by 
+\[
+\hat{\Delta} = \left[ \frac{1}{J} \left( \hat{y}_{i+}^{\ast} - \hat{y}_{i'+}^{\ast} \right) \right] - \left[ \frac{1}{J} \left( \hat{y}_{i+} - \hat{y}_{i'+} \right) \right],
 \]
@@
 \[
-SE(\hat{\Delta}) = \left[ \frac{2}{J} \left( \hat{\sigma}_{\gamma}^{2} + \hat{\sigma}_{\gamma^{\ast}}^{2} \right) \right]^{1/2}.
+SE(\hat{\Delta}) = \left[ \frac{2}{J} \left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right) \right]^{1/2}.
 \]
@@
 \[
-\left( \hat{\sigma}_{\gamma}^{2} + \hat{\sigma}_{\gamma^{\ast}}^{2} \right)^2 \bigg/
-\left( \frac{\hat{\sigma}_{\gamma}^{4}}{\mathrm{df}(\gamma)} + \frac{\hat{\sigma}_{\gamma^{\ast}}^{4}}{ \mathrm{df}(\gamma^{\ast})} \right).
+\left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right)^2 \bigg/
+\left( \frac{\hat{\sigma}_{\gamma^{\ast}}^{4}}{\mathrm{df}(\gamma^{\ast})} + \frac{\hat{\sigma}_{\gamma}^{4}}{ \mathrm{df}(\gamma)} \right).
 \]
 A distinctive property of the proposed model-based testing to the two-sample $t$-test is that even only the PTM abundances in Conditions $i$ and $i'$ are compared, measurements from all conditions are used for the modeling and inference.
+
+The formulations for $\hat{\Delta}$, $SE(\hat{\Delta})$, and degrees of freedom shown above are for a simple balanced experiment.
+They can also be extended to more realistic experiments with complex designs.
+This is discussed further in Section \ref{sec:complex_methods}
 
@@
 %Sample size calculation based on linear models has been described for general applications~\cite{kutner_etal_04a} and specifically for protein significance analysis~\cite{oberg_vitek_09a}. 
-Sample size calculation takes as input a) $q$, the desired false discovery rate, b) $\beta$, the average Type II error rate, c) $\Delta$, the minimal log-fold change in adjusted PTM abundance that we would like to detect, d) $m_0 / (m_0 + m_1)$, the fraction of truly differentially modified PTM sites in the comparison, and e) $\sigma_{\gamma}^{2}$ and $\sigma_{\gamma^{\ast}}^{2}$, the anticipated variances associated to modified and unmodified peptide features, respectively.
+Sample size calculation takes as input a) $q$, the desired false discovery rate, b) $\beta$, the average Type II error rate, c) $\Delta$, the minimal log-fold change in adjusted PTM abundance that we would like to detect, d) $m_0 / (m_0 + m_1)$, the fraction of truly differentially modified PTM sites in the comparison, and e) $\sigma_{\gamma^{\ast}}^{2}$ and $\sigma_{\gamma}^{2}$, the anticipated variances associated to modified and unmodified peptide features, respectively.
 The variances can be derived based on the dataset being analyzed, assuming similar quantitative properties and variations.
@@
 \[
-\mathrm{SE}^{2}(\hat{\Delta}) = \left[ \frac{2}{J} \left( \hat{\sigma}_{\gamma}^{2} + \hat{\sigma}_{\gamma^{\ast}}^{2} \right) \right]
+\mathrm{SE}^{2}(\hat{\Delta}) = \left[ \frac{2}{J} \left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right) \right]
 \leq \left( \frac{\Delta}{t_{1-\beta, df} + t_{1-\alpha /2, df}} \right)^{2},
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{TMT experiment}
-\label{sec:tmtmethod}
-
-The methods described previously can be extrapolated to Tandem Mass Tag (TMT) experiments targeting PTMs.
-This is done in the same way to how MSstatsTMT targets the global protein.\cite{Huang:2020} The null hypothesis is that there is no difference in mean PTM abundance between Conditions $i$ and $i'$ after adjusting for changes in mean unmodified peptide abundance between Conditions $i$ and $i'$.
+\subsection{Extension to complex designs}
+\label{sec:complex_methods}
+
+The methods described previously in Section \ref{sec:test} can be extrapolated to experiments with complex designs targeting PTMs.
+This includes experiments with additional sources of variation and unbalanced designs.
+This is done in the same way to how MSstats~\cite{choi_etal_14a} and MSstatsTMT~\cite{Huang:2020} target the global protein.
+The null hypothesis is still that there is no difference in mean PTM abundance between Conditions $i$ and $i'$ after adjusting for changes in mean unmodified peptide abundance between Conditions $i$ and $i'$.
 \begin{align*}
@@
 
-For TMT experiments there is an additional source of variation coming from the mixtures used.
-This increases the model complexity as variation can now come from multiple mixtures, biological replicates, and technical replicates.
+In cases where the design in unbalanced, restricted maximum likelihood is used to estimate the parameters of each model.
+Once the parameters are estimated, we can combine the models using modified versions of the formulas in Section \ref{sec:test}.
+
+The log-fold change in the adjusted PTM abundance, $\Delta$, is estimated by 
+\[
+\hat{\Delta} = \left[ \left( \hat{y}_{i+}^{\ast} - \hat{y}_{i'+}^{\ast} \right) \right] - \left[ \left( \hat{y}_{i+} - \hat{y}_{i'+} \right) \right],
+\]
+and the standard error of the estimate $\mathrm{SE}(\hat{\Delta})$ is 
+\[
+SE(\hat{\Delta}) = \left[ \left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right) \right]^{1/2}.
+\]
+%\[
+%\left[ \left( \frac{2}{J} \right) \cdot \left( \hat{\sigma}_{\gamma}^{2} + \hat{\sigma}_{\gamma^{\ast}}^{2} \right) \right]^{1/2}.
+%\]
+The test statistic $\hat{\Delta} / \mathrm{SE}(\hat{\Delta})$ is compared against the $t$ distribution, with degrees of freedom approximated by
+\[
+\left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right)^2 \bigg/
+\left( \frac{\hat{\sigma}_{\gamma^{\ast}}^{4}}{\mathrm{df}(\gamma^{\ast})} + \frac{\hat{\sigma}_{\gamma}^{4}}{ \mathrm{df}(\gamma)} \right).
+\]
+
+The model can be expanded to account for multiple sources of variation.
+An example of this is in TMT experiments wheere additional variation comes from the mixtures used.
 Going forward we will denote the log-intensity of a modified peptide for Technical Replicate j of Mixture m for Condition i by $y_{ijm}$, where $m = 1,....,M$.
@@
 \begin{align*}
-Y_{mtcb} = \mu + M_m + T(M)_{tm} + C_c + S_{mcb} + \epsilon_{mtcb}\\
-Y^*_{mtcb} = \mu^* + M^*_m + T(M)^*_{tm} + C^*_c + S^*_{mcb} + \epsilon^*_{mtcb}
+y^*_{mtcb} = \mu^* + M^*_m + T(M)^*_{tm} + C^*_c + S^*_{mcb} + \epsilon^*_{mtcb}\\
+y_{mtcb} = \mu + M_m + T(M)_{tm} + C_c + S_{mcb} + \epsilon_{mtcb}
 \end{align*}
@@
 Finally the error is expressed as $\epsilon_{mtcb} \sim N(0, \sigma^2)$.
-
-The log-fold change for adjusted PTM abundance is estimated as the average over mixtures and technical replicates
-
-$$\hat{\Delta} = \frac{1}{MJ} \left( \hat{y}_{+,+,i+} - \hat{y}_{+,+,i'+} \right) - \frac{1}{MJ} \left( \hat{y}_{+,+,i+}^{\ast} - \hat{y}_{+,+,i'+}^{\ast} \right)$$
-
-and the standard error of the estimate is
-
-\[
-\left[ \frac{2}{MJ} \left( \hat{\sigma}_{\gamma}^{2} + \hat{\sigma}_{\gamma^{\ast}}^{2} \right) \right] ^{1/2}.
-\]
-The test statistic $\hat{\Delta} / \mathrm{SE}(\hat{\Delta})$ is compared against the $t$ distribution, with degrees of freedom approximated by
-\[
-\left( \hat{\sigma}_{\gamma}^{2} + \hat{\sigma}_{\gamma^{\ast}}^{2} \right)^2 \bigg/
-\left( \frac{\hat{\sigma}_{\gamma}^{4}}{  \mathrm{df}(\gamma)} + \frac{\hat{\sigma}_{\gamma^{\ast}}^{4}}{\mathrm{df}(\gamma^{\ast})} \right).
-\]
 
@@
 For variance we used the variance from the biological experiments to determine what values to set.
-In Dataset 4 and 5 the variance of the PTM was generally higher than the global protein.
+In dataset 4 and 5 the variance of the PTM was generally higher than the global protein.
 In contrast, in dataset 6 the variance of the PTM and Protein was generally the same.
@@
 
-The results can be seen in \sfigref{fig:power_sd_combo}
-
-\begin{figure}[h!]
-\centering
- \begin{subfigure}{\textwidth}
- \centering
-	\includegraphics[width=.9\textwidth]{sim_new/same_var_power}
-	\caption{blah}
+The results of the power and sample size analysis can be seen in \sfigref{fig:power_sd_combo}.
+
+\begin{figure}[h!]
+\centering
+ \begin{subfigure}{\textwidth}
+ \centering
+	\includegraphics[width=.725\textwidth]{sim_new/same_var_power}
+	\caption{Here we can see the power of an experiment targeting PTMs with the same variance, .15, for the modified and unmodified peptides.
+Predictably when the replicates are high for both modified and unmodified peptides the power is much higher.
+Conversely at low replicates for each the power is much lower.
+The interesting part of this chart is when the replicates are different between runs.
+With equal variance, it does not matter if the PTM replicates or protein replicates are higher.
+Both cases result in the same power with equal variance.}
  \end{subfigure}\vspace{-5mm}
@@
  \centering
-	\includegraphics[width=.9\textwidth]{sim_new/high_ptm_var_power}
-	\caption{blah}
+	\includegraphics[width=.725\textwidth]{sim_new/high_ptm_var_power}
+	\caption{In this chart the variance for the PTM is higher than the unmodified protein.
+The PTM variance is .2, while the unmodified protein variance is .1.
+With equal replicates the results are the same as above, more replicates equals more power.
+When the replicates are not the same we can clearly see that having more replicates for the PTM leads to more power.
+It is clearly more important to have high replicates for the PTM run than the unmodified protein, when the PTM variance is higher.}
  \end{subfigure}

### 2021-10-11 code/USP30_MSstatsPTM_Analysis.R

@@ -9,6 +9,16 @@ setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
 load(file = "../data/ups30_msstatsptm_model.rda")#model_ptm
 load(file = "../data/ups30_msstatsptm_summary.rda")#summarized_ptm
 
+## SE analyis
+pyst_model
+
+model_ptm$PTM.Model %>% ggplot + geom_boxplot(aes(y = SE))
+model_ptm$PROTEIN.Model %>% ggplot + geom_boxplot(aes(y = SE))
+
+model_ptm$PTM.Model %>% summarize(mean(SE, na.rm= TRUE))
+model_ptm$PROTEIN.Model %>% summarize(mean(SE, na.rm= TRUE))
+
+##
 
 original_ptms <- unique(model_ptm$PTM.Model$Protein)
 adj_ptms <- unique(model_ptm$ADJUSTED.Model$Protein)

### 2021-10-11 code/sample_size_analysis.R

@@ -3,6 +3,8 @@ library(tidyverse)
 library(ggthemes)
 library(data.table)
 
+setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
+
 ## Calculations w/o feature variations -----------------------------------------
 
 sigma_model <- function(ptm_var, ptm_n, prot_var, prot_n){
@@ -12,8 +14,8 @@ sigma_model <- function(ptm_var, ptm_n, prot_var, prot_n){
 
 ## Define adjustable metrics
 low <- .1
-mid <- .2
-high <- .3
+mid <- .15
+high <- .2
 n_sample_list <- c(2,5)
 
 ## Hard metrics
@@ -26,7 +28,7 @@ bind_list <- c()
 i <- 1
 for (x in seq_along(n_sample_list)){
   for (y in seq_along(n_sample_list)){
-    var <- sigma_model(mid, n_sample_list[[x]], mid, n_sample_list[[y]])
+    var <- sigma_model(high, n_sample_list[[x]], low, n_sample_list[[y]])
     
     
     t <- delta / var
@@ -87,32 +89,12 @@ p2 <- high_ptm_var_df %>%
         strip.text = element_text(size = 12))
 p2
 
-
-## Original code containing sigma for subject.. can include or assume 0 for now
-## numSample = round(2 * (median_sigma_error + median_sigma_subject) / aa, 0)
-
-
-
-numSample_nomod_1 <- round(2 * (.1) / aa, 0)
-numSample_nomod_2 <- round(2 * (.2) / aa, 0)
-numSample_nomod_3 <- round(2 * (.4) / aa, 0)
-
-p1 <- data.frame("SD" = c(rep("`.1", length(numSample_nomod_1)),
-                      rep(".2", length(numSample_nomod_1)),
-                      rep(".4", length(numSample_nomod_1))),
-           "Log2FC" = c(desiredFC, desiredFC, desiredFC),
-           "num_samples" = c(numSample_nomod_1, numSample_nomod_2, numSample_nomod_3)) %>%
-  ggplot() + geom_line(aes(x = Log2FC, y = num_samples, group = SD, color = SD), size = 1.2) +
-  scale_colour_manual(values=cbPalette) +
-  labs(title = "Replicates Needed to Reach Power of .9", y = "# of replicates per condition") +
-  theme_bw() +
-  theme(axis.text.x = element_text(size = 12),
-        axis.text.y = element_text(size = 12),
-        legend.text=element_text(size=12),
[…]

### 2021-10-12 main/MSstatsPTM_Main.tex
Made updates to formulas

@@
 The null hypothesis states that there is no difference in log-abundance of the PTM site between conditions, adjusted with respect to protein abundance.
-Notationally, the hypothesis test is formulated as follows
+Notationally, given two conditions $i$ and $i'$, the hypothesis test is formulated as follows
 
@@
 \includegraphics[scale=.8]{images/MSstatsPTM_design.png}
-\caption{blah.}
+\caption{The workflow of MSstatsPTM and how it fits into the experimental pipeline.
+MSstatsPTM's workflow starts after modified and unmodified peptide quantification.
+First tool specific pre-processing is done, this includes modification site identification, general data cleaning, and formatting the data into the format needed for the package.
+The next step is feature level summarization, which summarizes features up to the modification level for the PTM data, and the protein level for the protein data.
+In the final step a model is fit to identify differential PTMs across condtions.}
 \label{fig:msstatsptm_design}

### 2021-10-13 main/MSstatsPTM_Main.tex
Updated shigella to time series model

@@
 
-\subsection*{Computer simulations}
-\label{sec:comp_sim_procedure}
-
-\paragraph*{Dataset 1 : Computer simulation 1 - label free}
+\subsection*{Dataset 1 : Computer simulation 1 - Label-free} 
+\label{sec:comp_sim_procedure1}
 
@@
 
-\paragraph*{Dataset 2 : Computer simulation 2 - missing values and low replicates}
+\subsection*{Dataset 2 : Computer simulation 2 - Label-free missing values and low replicates} 
+\label{sec:comp_sim_procedure2}
 
@@
 
-\subsection*{Dataset 3 : SpikeIn benchmark - label-free - KGG enriched}
+\subsection*{Dataset 3 : SpikeIn benchmark - Ubiquitination - Label-free}
 
@@
 
-\subsection*{Dataset 4 : Human-1mix-TMT - ubiquitination}
+\subsection*{Dataset 4 : Human - Ubiquitination - 1mix-TMT}
 
@@
 
-\subsection*{Dataset 5 : Mouse-2mix-TMT - phosphorylation}
+\subsection*{Dataset 5 : Mouse - Phosphorylation - 2mix-TMT time series}
 
@@
 
-\subsection*{Dataset 6 : Human label free quantification - no global profiling run}
+\subsection*{Dataset 6 : Human - Ubiquitination - Label-free no global profiling run}
 
@@
 
-\subsubsection*{Dataset 1 : Computer simulation 1 - label free}
+\subsubsection*{Dataset 1 : Computer simulation 1 - Label-free}
 
@@
 
-\subsubsection*{Dataset 2 : Computer simulation 2 - missing values and low replicates}
+\subsubsection*{Dataset 2 : Computer simulation 2 - Label-free missing values and low replicates}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection*{Dataset 3 : SpikeIn benchmark - label-free - KGG enriched}
+\subsubsection*{Dataset 3 : SpikeIn benchmark - Ubiquitination - Label-free}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection*{Dataset 4 : Human-1mix-TMT - ubiquitination}
+\subsubsection*{Dataset 4 : Human - Ubiquitination - 1mix-TMT}
 
 The results of this experiment are summarized in Figures \ref{fig:ipah_venn_diagramm} and \ref{fig:ipah_profile}.
+The experiment was modeled as a group comparison.
 In Figure \ref{fig:ipah_venn_diagramm} the number of significant modified peptides before and after adjustment is shown.
@@
 
-\subsubsection*{Dataset 5 : Mouse-2mix-TMT - phosphorylation}
+\subsubsection*{Dataset 5 : Mouse - Phosphorylation - 2mix-TMT time series}
 
 The results of this experiment are summarized in Figures \ref{fig:shig_venn_d} and \ref{fig:shigella_profile}.
+The experiment was modeled as a time series, with the same bioreplicate being measured at each time point.
 In Figure \ref{fig:shig_venn_d} the number of significant modified peptides before and after adjustment is shown.
@@
 This was tested by looking for modified peptides who's adjusted log fold change was within 10\% of the unadjusted log fold change but became insignificant after adjustment.
-When this test was applied on this experiment, 776 peptides became insignificant due to an increase in variance.
-This is still a very small portion (4.02\%) of the peptides that became insignificant after adjustment.
+When this test was applied on this experiment, 548 peptides became insignificant due to an increase in variance.
+This is still a very small portion (3.4\%) of the peptides that became insignificant after adjustment.
 Thus we can conclude that the drop off in significant peptides was mainly due to changes in global protein abundance.
@@
 The profile plot of modified peptide $TTP\_MOUSE|P22893\_S178$ shows a large positive log fold change of $2.9$ between the WT\_Late and WT\_Uninfected conditions.
-However, the global profiling run shows a similar log fold change of $2.015$ between the same conditions.
+However, the global profiling run shows a similar log fold change of $2.014$ between the same conditions.
 This indicates that the abundance change in the modified peptide is nearly entirely due to changes in the global protein.
-When adjusting for the global protein the modified peptide's adjusted pvalue becomes insignificant, going from $.0009$ to $.275$.
-In contrast $KI67\_MOUSE|E9PVX6\_T215$ originally shows a small positive log fold change of $.212$ between conditions WT\_Early and WT\_Uninfected.
-This is contrast with a negative fold change in the global profiling run of $-.616$.
-In adjusting the modified peptide for changes in the global protein, the log fold change becomes .828 and the adjusted pvalue goes from $.452$ to $.037$.
+When adjusting for the global protein the modified peptide's adjusted pvalue becomes insignificant, going from $.0009$ to $.248$.
+In contrast $KI67\_MOUSE|E9PVX6\_T215$ originally shows a small positive log fold change of $.185$ between conditions WT\_Early and WT\_Uninfected.
+This is contrast with a negative fold change in the global profiling run of $-.619$.
+In adjusting the modified peptide for changes in the global protein, the log fold change becomes $.804$ and the adjusted pvalue goes from $.338$ to $.0156$.
 
-\subsubsection*{Dataset 6 : Human label free quantification - no global profiling run}
+\subsubsection*{Dataset 6 : Human - Ubiquitination - Label-free no global profiling run}
 
@@
 \includegraphics[width=1\textwidth]{images/sim1_FDR_all_models.png}
-\caption{Dataset 1 : Computer simulation 1 - label free.
+\caption{Dataset 1 : Computer simulation 1 - Label-free.
 All the considered methods in the first computer simulation correctly calibrated FDR when adjusting for changes in protein abundance.
@@
 \includegraphics[width=1\textwidth]{images/sim3_Accuracy.png}
-\caption{Dataset 2 : Computer simulation 2 - missing values and low replicates.
+\caption{Dataset 2 : Computer simulation 2 - Label-free missing values and low replicates.
 The advantage of using the proposed approach is apparent when looking at the second computer simulation, which includes limited observations and the presence of missing values.
@@
 \includegraphics[width=1\textwidth]{images/sim3_FC_boxplot.png}
-\caption{Dataset 2 : Computer simulation 2 - missing values and low replicates.
+\caption{Dataset 2 : Computer simulation 2 - Label-free missing values and low replicates.
 In the second computer simulation with missing values and low PTM features all methods correctly estimated the fold change with a median log change of .75.
@@
 \includegraphics[width=1\textwidth]{images/spike_in_msstatsptm_volcano.png}
-\caption{Dataset 3 : SpikeIn benchmark - label-free - KGG enriched.
+\caption{Dataset 3 : SpikeIn benchmark - Ubiquitination - Label-free.
 Using the proposed method to model the benchmark experiment the spike in peptides (colored red) do not follow the expected log fold change before adjustment.
@@
 \includegraphics[width=1\textwidth]{images/spike_in_limma_volcano.png}
-\caption{Dataset 3 : SpikeIn benchmark - label-free - KGG enriched.
+\caption{Dataset 3 : SpikeIn benchmark - Ubiquitination - Label-free.
 Here the Limma method was used to model the spike in experiment.
@@
 \centering
-\includegraphics[width=1\textwidth]{images/IpaH_prof_plot.png}
-\caption{Dataset 4 : Human-1mix-TMT - ubiquitination.
+\includegraphics[height=.65\textwidth]{images/ipah_venn_diagramm.png}
+\caption{Dataset 4 : Human - Ubiquitination - 1mix-TMT.
 The overlap of differencial modified peptides for the PTM model with and without global protein level adjustment.
@@
 \includegraphics[width=1\textwidth]{images/IpaH_prof_plot.png}
-\caption{Dataset 4 : Human-1mix-TMT - ubiquitination.
+\caption{Dataset 4 : Human - Ubiquitination - 1mix-TMT.
 Comparing the global profiling of protein $GSDMD\_HUMAN|P57764$ with the ubiquitination of the protein at site $K62$.
@@
 \includegraphics[height=.65\textwidth]{images/shig_venn_diagramm.png}
-\caption{Dataset 5 : Mouse-2mix-TMT - phosphorylation.
+\caption{Dataset 5 : Mouse - Phosphorylation - 2mix-TMT time series.
 The overlap of differencial modified peptides between the PTM model with and without global protein level adjustment.
@@
  \end{subfigure}
-\caption{Dataset 5 : Mouse-2mix-TMT - phosphorylation. Profile plots for PTMs P22893\_S178 and E9PVX6\_T215 in the Shigella flexneri experiment.}
+\caption{Dataset 5 : Mouse - Phosphorylation - 2mix-TMT time series. Profile plots for PTMs P22893\_S178 and E9PVX6\_T215 in the Shigella flexneri experiment.}
 \label{fig:shigella_profile}
@@
  \end{subfigure}
- \caption{Dataset 6 : Human label free quantification - no global profiling run. Overlap of significant PTMs before and after protein adjustment.}
+ \caption{Dataset 6 : Human - Ubiquitination - Label-free no global profiling run. Overlap of significant PTMs before and after protein adjustment.}
 \label{fig:usp30_ven_d}
@@
 \includegraphics[width=1\textwidth]{images/USP30_profile_plot.png}
-\caption{Dataset 6 : Human label free quantification - no global profiling run.
+\caption{Dataset 6 : Human - Ubiquitination - Label-free no global profiling run.
 Comparing the global profiling of protein $P52209$ with the modification of the protein at site $K059$.

### 2021-10-13 supplementary/ptm_sm.tex
Made updates to formulas
Updated shigella to time series model

@@
 \begin{align*}
-H_{0}: \Delta = (\mu_{i} - \mu_{i'}) - (\mu_{i}^{\ast} - \mu_{i'}^{\ast}) &= 0 \\
-H_{a}: \Delta = (\mu_{i} - \mu_{i'}) - (\mu_{i}^{\ast} - \mu_{i'}^{\ast}) &\neq 0
+H_{0}: \Delta = (\mu_{i}^{\ast} - \mu_{i'}^{\ast}) - (\mu_{i} - \mu_{i'}) &= 0 \\
+H_{a}: \Delta = (\mu_{i}^{\ast} - \mu_{i'}^{\ast}) - (\mu_{i} - \mu_{i'}) &\neq 0
 \end{align*}
@@
 \[
-\hat{\Delta} = \left[ \left( \hat{y}_{i+}^{\ast} - \hat{y}_{i'+}^{\ast} \right) \right] - \left[ \left( \hat{y}_{i+} - \hat{y}_{i'+} \right) \right],
+\hat{\Delta} = (\hat{\mu}_{RML_i}^{\ast} - \hat{\mu}_{RML_{i'}}^{\ast}) - (\hat{\mu}_{RML_i} - \hat{\mu}_{RML_{i'}})
 \]
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection{Dataset 1 : Computer simulation 1 - label free experiment}
+\subsubsection{Dataset 1 : Computer simulation 1 - Label-free}
 
@@
 
-\subsubsection{Dataset 2 : Computer simulation 2 - missing values and low replicates}
+\subsubsection{Dataset 2 : Computer simulation 2 - Label-free missing values and low replicates}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Dataset 3 : SpikeIn benchmark - label-free - KGG enriched}
+\subsection{Dataset 3 : SpikeIn benchmark - Ubiquitination - Label-free}
 \label{sec:benchmark}
@@
 
-\subsection{Dataset 4 : Human-1mix-TMT - ubiquitination}
+\subsection{Dataset 4 : Human - Ubiquitination - 1mix-TMT}
 \label{sec:ipah}
@@
 The experiment was inbalanced with two bioreplicates per condition for all conditions except for infected 1 hour.
+The experiment was modeled as a group comparison, with different replicates measured at each time point.
 The experimental design can be seen in Table~\ref{table:ipah_design}.
@@
 \clearpage
-\subsection{Dataset 5 : Mouse-2mix-TMT - phosphorylation}
+\subsection{Dataset 5 : Mouse - Phosphorylation - 2mix-TMT time series}
 \label{sec:shigella}
@@
 The abundance of total protein and post-translation modifications were quantified at three time points, uninfected, early infection (45-60 minutes), and late infection (3-3.5 hours).
+The experiment was modeled as a time series, with the same bioreplicate being measured at each time point.
 Quantifying the total protein along with the post-translational modifications allowed us to adjust for changes in total protein and see the true impact of the site specific modifications.
@@
 This was tested by looking for modified peptides who's adjusted log fold change was within 10\% of the unadjusted log fold change but became insignificant after adjustment.
-When this test was applied on this experiment, 776 peptides became insignificant due to an increase in variance.
-This is higher than in \sfigref{fig:ipha_venn_d}, however it is still a very small portion (4\%) of the peptides that became insignificant after adjustment.
+When this test was applied on this experiment, 548 peptides became insignificant due to an increase in variance.
+This is higher than in \sfigref{fig:ipha_venn_d}, however it is still a very small portion (3.4\%) of the peptides that became insignificant after adjustment.
 Thus we can conclude that the drop off in significant peptides was mainly due to changes in global protein abundance.}
@@
 
-\subsection{Dataset 6 : Human label free quantification - no global profiling run}
+\subsection{Dataset 6 : Human - Ubiquitination - Label-free no global profiling run}
 \label{sec:usp30}
@@
 A corresponding mixed effects model was fit per modification and global protein as described previously in this supplementary.
+The experiment was modeled as a group comparison.
 

### 2021-10-14 main/MSstatsPTM_Main.tex
Updated modeling sections

@@
 
-\author[1]{Devon~Kohler*}
-\author[2]{Tsung-Heng~Tsai*}
-\author[3]{Meena~Choi*}
+\author[1]{Devon~Kohler}
+\author[2]{Tsung-Heng~Tsai}
 \author[1]{Ting~Huang}
@@
 \author[3]{Trent~Hinkle}
-\author[1]{Olga~Vitek}
+\author[3]{Meena~Choi*}
+\author[1]{Olga~Vitek*}
 \affil[1]{Khoury College of Computer Science, Northeastern University, Boston, MA, USA}
@@
 The computer simulations vary in experimental realism; the first simulation was perfectly clean, with many replicates and no missing values, while the second simulation introduced real world qualities, such as limited modified features and missing values.
-The spike-in experiment took the real world qualities a step further and allowed us to compare the methods in an real experiment where ground truth is known.
-Finally the proposed approach was used to model three biological experiments which differed in experimental design, acquisition strategy, and use various organisms.
+The spike-in experiment took the real world qualities a step further and allowed us to compare the methods in a real experiment where ground truth is known.
+Finally the proposed approach was used to model three biological experiments which differed in experimental design, acquisition strategy, and organism.
 These experiments were chosen to show the proposed method's applicability to many different types of experiments.
@@
 
-Here simulations were generated with a high number of modified and unmodified peptide features and no missing values.
+Here simulations were generated mimicking a Label-free experiment, with a high number of modified and unmodified peptide features and no missing values.
 Multiple simulations were generated with different numbers of replicates and conditions.
 In all simulations 1000 modified peptides were generated, half of which were true positives and the other half true negatives.
-Half of true positives and true negatives were generated with convolution with global protein abundance.
+Half of the true positives and true negatives were generated convoluted with global protein abundance.
 Before adjustment, the convoluted true positives should appear to be negatives, while the convoluted true negatives would appear differential.
@@
 
-\medskip
-
-\noindent{Further details of how the simulations were generated can be found in Supplementary Sec. \ref{supp-sec:sim}. }
+Further details of how both simulations were generated, including exact parameters used, can be found in Supplementary Sec. \ref{supp-sec:sim}.
 
@@
 We evaluated our approach using a custom designed spike-in benchmark experiment with known ground truth, where 50 heavy-labeled KGG motif peptides from 20 human proteins were used as spike-in peptides.
-Quantitative changes in protein and site abundance changes of these 20 proteins were the target of the benchmark.
+Quantitative changes in protein and site abundance of these 20 proteins were the target of the benchmark.
 Unmodified peptides from Human Lysate were used as the estimate of global protein abundance changes.
@@
 
-Further details on the design of this experiment are provided in Supplementary Section \ref{supp-sec:ipah}.
+Further details on the design of this experiment are provided in Supplementary Sec. \ref{supp-sec:ipah}.
 
@@
 Label-free mass spectrometry quantification was used to quantify the abundance of modified peptides.
-A corresponding mixed effects model was fit per modification and global protein as described previously.
 In contrast to the other biological experiments assessed in this paper, this experiment did not not have a separate global profiling run for the unmodified protein.
-In this case $MSstatsPTM$ can still be used by extracting unmodified peptides from the modified run, however this generally leads to substantial less matches between modified and unmodified peptides, in addition to low feature counts for unmodified peptides.
+In this case $MSstatsPTM$ can still be used by extracting unmodified peptides from the modified run, however this generally leads to substantially less matches between modified and unmodified peptides, in addition to low feature counts for unmodified peptides.
 In this case there only 41.9\% of the modified peptides had a corresponding unmodified protein to perform adjustment.
@@
 
-Further details on the specific equations and workflow of existing methods can be seen in Supplementary \ref{supp-sec:ttest}.
+Further details on the specific equations and workflow of existing methods can be seen in Supplementary Sec. \ref{supp-sec:ttest}.
 
@@
 As different levels of variability are present in the data, the log-intensities of the features for modified and unmodified peptides are modeled separately using two linear mixed models.
-An overview of the steps of the proposed approach are illustrated in Figure \ref{fig:msstatsptm_design}.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection*{Detection of changes in PTMs}
+\subsubsection*{Missing value imputation}
+\label{sec:missing_imp}
+
+$MSstatsPTM$ gives the ability to impute missing feature intensities if desired.
+When values are imputed, it is assumed they are missing for reasons of low abundance.
+Missing features are imputed in each MS run using the Accelerated Time Failure (AFT) model\cite{Tukey:1977}.
+In order to impute a feature's missing values, the feature must be present in at least one MS Run.
+If the feature is not present across all MS Runs, the value will be left missing.
+Missing value imputation is done before summarization with Tukey's Median Polish in order to correct for the impact of outliers.
+Missing value imputation is done separately for both the PTM and global protein datasets.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsubsection*{Detection of changes in PTMs - Balanced design with one source of variation}
 \label{sec:adjustment_summary}
+
+The methods when applied to a balanced design with multiple conditions and one technical replicate.
+In practice the experimental design can be unbalanced and variation can come from multiple sources.
+We show how the methods can be applied to more complex experimental designs in the next section.
 
@@
 The null hypothesis states that there is no difference in log-abundance of the PTM site between conditions, adjusted with respect to protein abundance.
-Notationally, given two conditions $i$ and $i'$, the hypothesis test is formulated as follows
+Notationally, given two conditions $i$ and $i'$, the hypothesis test is formulated as follows:
 
@@
 Specifically, the adjusted difference is given by the difference in log-abundance of the PTM site, subtracted by the difference in log-abundance of the underlying protein, which is equivalent to the log of the ratio of PTM abundance difference to protein abundance difference.
+
+\[
+\hat{\Delta} = \left[ \frac{1}{J} \left( \hat{y}_{i+}^{\ast} - \hat{y}_{i'+}^{\ast} \right) \right] - \left[ \frac{1}{J} \left( \hat{y}_{i+} - \hat{y}_{i'+} \right) \right],
+\]
+
 The estimate of the adjusted difference and the standard error (SE) of the estimate are obtained by combining the difference estimates and the associated SEs from both counterparts.
@@
 
-Finally the degrees of freedom are given by
+Finally the degrees of freedom are given by:
 
@@
 Adjustment for multiple comparisons is performed using the Benjamini-Hochberg procedure to control the false discovery rate at a desired level, e.g., $0.05$\cite{Benjamini:1995}.
-More details on how the PTM model is adjusted are provided in Supplementary Sec \ref{supp-sec:test}
+More details on how the PTM model is adjusted are provided in Supplementary Sec.
+\ref{supp-sec:test}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection*{Missing value imputation}
-\label{sec:missing_imp}
+\subsubsection*{Detection of changes in PTMs - Extension to complex designs}
 
-$MSstatsPTM$ gives the ability to impute missing feature intensities if desired.
-When values are imputed, it is assumed they are missing for reasons of low abundance.
-Missing features are imputed in each MS run using the Accelerated Time Failure (AFT) model\cite{Tukey:1977}.
-In order to impute a feature's missing values, the feature must be present in at least one MS Run.
-If the feature is not present across all MS Runs, the value will be left missing.
-Missing value imputation is done before summarization with Tukey's Median Polish in order to correct for the impact of outliers.
-Missing value imputation is done separately for both the PTM and global protein datasets.
+The statistical modeling approaches discussed above can also be extended to complex designs including experiments with additional sources of variation and unbalanced designs.
+The null hypothesis is still that there is no difference in mean PTM abundance between Conditions $i$ and $i'$ after adjusting for changes in mean unmodified peptide abundance between Conditions $i$ and $i'$.
+\begin{align*}
+H_{0}: \Delta = (\mu_{i}^{\ast} - \mu_{i'}^{\ast}) - (\mu_{i} - \mu_{i'}) &= 0 \\
+H_{a}: \Delta = (\mu_{i}^{\ast} - \mu_{i'}^{\ast}) - (\mu_{i} - \mu_{i'}) &\neq 0
+\end{align*}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection*{Extension to complex designs}
+Model inference is done in the same way to how MSstats~\cite{choi_etal_14a} and MSstatsTMT~\cite{Huang:2020} target the global protein.
+In cases where the design in unbalanced, a linear mixed effects model is fit, which takes into account all potential sources of variation, and restricted maximum likelihood is used to estimate the parameters of each model.
+Once the parameters are estimated, we can combine the models using modified versions of the formulas in Section \ref{sec:adjustment_summary}.
 
-The statistical modeling approaches discussed above can also be extended to complex designs with additional sources of variation and inbalanced designs.
-One specific complex design the method can be applied to is Tandem Mass Tag (TMT) labeling methods.
-TMT experiments introduce an additional source of variation in the form of different mixtures.
-To account for mixture variance, first feature intensities are summarized again using Tukey's Median Polish and missing values are imputed, if desired, then a new linear mixed effects model with an added term for the mixture is included.
-The workflow and statisical model described generally follow the methods used for modeling proteins in $MSstatsTMT$\cite{Huang:2020}.
-These methods are repeated to quantify both modified and unmodified peptide abundance.
-Once both modified and unmodified peptides are modeled, the PTM model is adjusted for changes in global protein abundance using the same methods described previously.
-
-For more information about the extension to complex designs and Tandem Mass Tag experiments see Supplementary Sec. \ref{supp-sec:complex_methods}.
+The log-fold change in the adjusted PTM abundance, $\Delta$, is estimated by 
+\[
+\hat{\Delta} = (\hat{\mu}_{RML_i}^{\ast} - \hat{\mu}_{RML_{i'}}^{\ast}) - (\hat{\mu}_{RML_i} - \hat{\mu}_{RML_{i'}})
+\]
+and the standard error of the estimate $\mathrm{SE}(\hat{\Delta})$ is 
+\[
+SE(\hat{\Delta}) = \left[ \left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right) \right]^{1/2}.
+\]
+%\[
+%\left[ \left( \frac{2}{J} \right) \cdot \left( \hat{\sigma}_{\gamma}^{2} + \hat{\sigma}_{\gamma^{\ast}}^{2} \right) \right]^{1/2}.
+%\]
+The test statistic $\hat{\Delta} / \mathrm{SE}(\hat{\Delta})$ is compared against the $t$ distribution, with degrees of freedom approximated by
+\[
+\left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right)^2 \bigg/
+\left( \frac{\hat{\sigma}_{\gamma^{\ast}}^{4}}{\mathrm{df}(\gamma^{\ast})} + \frac{\hat{\sigma}_{\gamma}^{4}}{ \mathrm{df}(\gamma)} \right).
+\]
 
@@
 The converters take as input the raw data from the tool, identify the modification site for modified peptide, and put the data into the correct format for analysis in $MSstatsPTM$.
-Conversion is done separately for the modified and global profiling runs and combined together for input into the rest of the package.
+Conversion is done separately for the modified and global profiling runs.
 If the global profiling run is not available, the package can still analyze the modified run, but will do so without adjusting for changes in unmodified protein abundance.
@@
 After using the converters, the next step is peptide/protein summarization and missing value imputation.
+In the methods sections it was assumed that all the input data was summarized; in the implementation we need to run a specific summarization step to make this happen.
 For the modified run, the package summarizes features, PSMs, which include the same modification together.
@@
 The final step of the package is modeling the summarized dataset.
-A linear mixed effects model is automatically fit for both the summarized modified and global profiling runs.
+A linear mixed effects model is fit for both the summarized modified and global profiling runs.
 This model is automatically adjusted depending on the experimental design and acquisition method.
 The comparisons of interest can either be predefined or a full pairwise comparison will be tested.
-After fitting a model to both the modified and unmodified data, the modified model is adjusted for changes in unmodified protein abundance, using the methods described in Section \ref{sec:adjustment_summary}.
+After fitting a model to both the modified and unmodified data, the modified model is adjusted for changes in unmodified protein abundance, using the methods described in this paper.
 
@@
 The statistical modeling relies on the functionality from the R packages $lme4$ \cite{Bates2015} and $lmerTest$ \cite{Kuznetsova2017}.
+An overview of the steps of the package are illustrated in Figure \ref{fig:msstatsptm_design}.
 
@@
 To evaluate the performance of different methods, we considered the false discovery rate (FDR = FP/(TP+FP)), the recall rate (TP/(TP + FN), and total accuracy ((TP + TN) / (TP + TN + FP + FN)).
-For biological experiments where ground truth was not known, we evaluated significantly differential peptides with an $\alpha$ of .05.
+For biological experiments where ground truth was not known, we evaluated significantly differential peptides with $\alpha=.05$.
 Comparisons were made before and after adjusting for changes in unmodified protein abundance and the results are presented in the form of venn diagrams and profile plots.
@@
 The proposed approach was evaluated against two-sample t-test and Limma. 
-Here we see that not accounting for changes in unmodified peptide level results in a high false discovery rate, seen in Figure \ref{fig:fdr_all_models}.
+In Figure \ref{fig:fdr_all_models} we see that not accounting for changes in unmodified peptide level results in a high false discovery rate.
 When adjusting for changes in the unmodified peptide, all methods performed similarly in terms of FDR.
@@
 
-Further details reviewing the results of each simulation and their comparison with respect to method performance can be seen in Supplementary Section \ref{supp-sec:sim}.
+Further details reviewing the results of each simulation and their comparison with respect to method performance can be seen in Supplementary Sec. \ref{supp-sec:sim}.
 
@@
 After adjustment the estimated fold change is more in line with expectation.
-Additionally, the background peptides serving as the null model, colored grey, show many false positives before adjustment is made.
-Again after adjustment the results improve and the number of false positives decrease significantly.
+Additionally, the background peptides serving as the null model, show many false positives before adjustment is made.
+After adjustment the results improve and the number of false positives decrease significantly.
 
@@
 
-Further results of this experiment, including a comparison to t-test and fold change distribution of each method before and after adjustment, are provided in Supplementary Section \ref{supp-sec:benchmark}.
+Further results of this experiment, including a comparison to t-test and fold change distribution of each method before and after adjustment, are provided in Supplementary Sec.
+\ref{supp-sec:benchmark}.
 
@@
 The results of this experiment are summarized in Figures \ref{fig:shig_venn_d} and \ref{fig:shigella_profile}.
-The experiment was modeled as a time series, with the same bioreplicate being measured at each time point.
+The experiment was modeled as a time series, with the same bioreplicate measured at each time point.
 In Figure \ref{fig:shig_venn_d} the number of significant modified peptides before and after adjustment is shown.
 Again more PTMs became insignificant after adjustment than became significant.
-19,286 peptides become insignificant, 4,947 became significant, while 39,550 peptides are significant in both models.
+19,286 peptides become insignificant, 4,947 became significant, while 41,552 peptides are significant in both models.
 Again we tested if the decrease in significant peptides was due to the increased variance that comes from adjustment, or if it was mainly due to removing convolution with change in the umodified protein.
@@
 Here we analyze the sample size needed to achieve a desired statistical power.
-The proposed approach corrects for confounding between the modified and unmodified peptides by adjusting the modified peptide for changes in the unmodified.
-This adjustment comes at a cost of increased variation.
+The proposed approach corrects for confounding between the modified peptide and unmodified protein at a cost of increased variation.
 This can be seen in the calculation for variance in Section \ref{sec:adjustment_summary}.
@@
 For variance we used the biological experiments to determine what scenarios and values to test.
-In dataset 4 and 5 the variance of the PTM was higher than the global protein.
+In datasets 4 and 5 the variance of the PTM was higher than the global protein.
 In dataset 6 the variance of the PTM and Protein was generally the same due to not using a separate global profiling run.
@@
 
-The results show that when the variance and replicates were equal, higher replicates predictably lead to higher power.
+The results of the power and sample size analysis can be seen in Figure \ref{fig:power_sd_combo}.
+We can see that when the variance and replicates were equal, higher replicates predictably lead to higher power.
 In cases where the replicates were unbalanced, but the variance was still the same, it did not matter if there were more replicates in the modified or unmodified runs.
@@
 It is clearly more important to have high replicates for the PTM run than the unmodified protein, when the PTM variance is higher.
-In cases where the number of replicates has to be limited, it is better to weight them on the PTM side.
-
-The results of the power and sample size analysis can be seen in Figure \ref{fig:power_sd_combo}.
+In cases where the number of replicates has to be limited, it is better to weigh them on the PTM side.
 
@@
 The implementation allows for straightforward application of the methods discussed and allows for reproducible experimental analysis.
-Additionally, the implementation allows the methods to be applied many different acquisition methods.
 
@@
 We are also interested in characterizing the interplay of PTMs at multiple sites.
-The proposed statistical methods are implemented as an R package $MSstatsPTM$ available on Bioconductor and Github.
 
@@
 This is due to not having a global profiling run, resulting in a large lack of overlap between the modified peptides and unmodified proteins.}
- \end{subfigure}\vspace{-5mm}
+ \end{subfigure}\vspace{5mm}
  \begin{subfigure}{\textwidth}
@@
 With equal variance, it does not matter if the PTM replicates or protein replicates are higher.}
- \end{subfigure}\vspace{-5mm}
+ \end{subfigure}\vspace{5mm}
  \begin{subfigure}{\textwidth}

### 2021-10-14 supplementary/ptm_sm.tex
Updated modeling sections

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection{Run-level summarization of feature intensities}
+\subsection{Run-level summarization of feature intensities}
 \label{sec:sum}
 
-Run-level summarization of feature intensities for each PTM site is carried out as in the sub-plot model of MSstats~\cite{choi_etal_14a}, which involves a) imputation of censored missing values, and b) summarization of feature intensities using Tukey's median polish.
+Run-level summarization of feature intensities for each PTM site is carried out as in the sub-plot model of MSstats~\cite{choi_etal_14a} or MSstatsTMT~\cite{Huang:2020} depending on acquisition strategy, which involves a) imputation of censored missing values, and b) summarization of feature intensities using Tukey's median polish.
 \cite{tukey_77a} The run-level summary for the PTM in Run $j$ of Condition $i$ is denoted by $\hat{y}_{ij}^{\ast}$.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection{Model-based inference of the underlying abundance in a simple balanced experiment}
+From here forward the input to the method assumes that all features are summarized.
+
+\subsection{Balanced design with one source of variation}
+\label{sec:model}
+The methods when applied to a balanced design with multiple conditions and one technical replicate are shown in this section.
+In practice the experimental design can be unbalanced and variation can come from multiple sources.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsubsection{Model-based inference of the underlying abundance}
 \label{sec:infer}
 
-In the case of a balanced design with multiple conditions and without technical replicates, the PTM abundance in each run is represented as 
+The PTM abundance in each run is represented as 
 \[
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{PTM significance analysis}
+\subsubsection{PTM significance analysis}
 \label{sec:test}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Design of PTM experiments}
+\subsubsection{Design of PTM experiments}
 \label{sec:design}
@@
 
-The methods described previously in Section \ref{sec:test} can be extrapolated to experiments with complex designs targeting PTMs.
+The methods described previously in Section \ref{sec:model} can be extrapolated to experiments with complex designs targeting PTMs.
 This includes experiments with additional sources of variation and unbalanced designs.
-This is done in the same way to how MSstats~\cite{choi_etal_14a} and MSstatsTMT~\cite{Huang:2020} target the global protein.
+
+\subsubsection{Model-based inference and significance analysis}
+
+Model inference is done in the same way to how MSstats~\cite{choi_etal_14a} and MSstatsTMT~\cite{Huang:2020} target the global protein.
 The null hypothesis is still that there is no difference in mean PTM abundance between Conditions $i$ and $i'$ after adjusting for changes in mean unmodified peptide abundance between Conditions $i$ and $i'$.
@@
 
-In cases where the design in unbalanced, restricted maximum likelihood is used to estimate the parameters of each model.
+In cases where the design in unbalanced, a linear mixed effects model is fit, which takes into account all potential sources of variation, and restricted maximum likelihood is used to estimate the parameters of each model.
 Once the parameters are estimated, we can combine the models using modified versions of the formulas in Section \ref{sec:test}.
@@
 
-The model can be expanded to account for multiple sources of variation.
-An example of this is in TMT experiments wheere additional variation comes from the mixtures used.
+%An example of this is in TMT experiments where additional variation comes from the mixtures used.
 Going forward we will denote the log-intensity of a modified peptide for Technical Replicate j of Mixture m for Condition i by $y_{ijm}$, where $m = 1,....,M$.
@@
 
-The model assumes the variability of the mixture is a random effect and is not convoluted with the other variables in the model.
+%The model assumes the variability of the mixture is a random effect and is not convoluted with the other variables in the model.
 The model also assumes that the difference between conditions is the same between mixtures.
 For a given modified peptide and corresponding unmodified protein a mixed effects model is fit as follows.
-\begin{align*}
-y^*_{mtcb} = \mu^* + M^*_m + T(M)^*_{tm} + C^*_c + S^*_{mcb} + \epsilon^*_{mtcb}\\
-y_{mtcb} = \mu + M_m + T(M)_{tm} + C_c + S_{mcb} + \epsilon_{mtcb}
-\end{align*}
-
-Where $M_m \sim N(0, \sigma^2_m)$, $T(M)_{tm} \sim N(0, \sigma^2_{tm})$, $S\sim N(0, \sigma^2_{s})$ are random effects, and $\sum_{c=1}^{I} C_{c}^{\ast} = 0$ is a fixed effect.
+%\begin{align*}
+%y^*_{mtcb} = \mu^* + M^*_m + T(M)^*_{tm} + C^*_c + S^*_{mcb} + \epsilon^*_{mtcb}\\
+%y_{mtcb} = \mu + M_m + T(M)_{tm} + C_c + S_{mcb} + \epsilon_{mtcb}
+%\end{align*}
+
+%Where $M_m \sim N(0, \sigma^2_m)$, $T(M)_{tm} \sim N(0, \sigma^2_{tm})$, $S\sim N(0, \sigma^2_{s})$ are random effects, and $\sum_{c=1}^{I} C_{c}^{\ast} = 0$ is a fixed effect.
 Finally the error is expressed as $\epsilon_{mtcb} \sim N(0, \sigma^2)$.
 
+\subsubsection{Design of PTM experiments}
+
+\todo{Devon: Add Ting's contrast calculations.}
+
 \clearpage
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Computer simulation}
+\section{Experiments with known ground truth}
 \label{sec:sim}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Protein-level adjustment}
+\subsection{Computer simulation}
 
@@
 
-\todo{only one mixture so we probably wouldn't include the mixture as a term in the model?}
-
 $$Y_{mcb} = \mu + Condition_c + Subject_{mcb} + \epsilon_{mcb}$$

### 2021-10-19 supplementary/ptm_sm.tex
Added sample size for complex designs

@@
 \subsubsection{Design of PTM experiments}
-
-\todo{Devon: Add Ting's contrast calculations.}
+\label{sec:complex_design}
+
+The proposed approach applied to experiments with complex designs still allows for sample size calculation and power analysis.
+To make the calculation the standard deviation and degrees of freedom are needed.
+The estimation of these quantities is dependent on the experimental design.
+In \sfigref{fig:statistical_inference} different experimental designs that the proposed approach can be applied to, along with their estimation of standard error and degrees of freedom is shown.
+
+\begin{figure}[h!]
+\makebox[\textwidth][c]{\includegraphics[width=1.2\textwidth]{sim_new/Statistical_inference}}%
+\caption{Different models that are fit depending on the experimental design.
+Models are fit for Label-free and TMT acquisition methods, as well as group comparison and time course experimental designs.
+The estimation of standard error and degrees of freedom for each model is shown.
+\label{fig:statistical_inference}}
+\end{figure}
+
+The standard error and degrees of freedom are estimated separately for both the PTM and unmodified protein models.
+Once this is done we can use these these values to calculate the sample size and perform power analysis.
+This is done in a similar way to Section \ref{sec:design}.
+Given $q$, the desired false discovery rate, $\beta$, the average Type II error rate, $m_0 / (m_0 + m_1)$, the fraction of truly differentially modified PTM sites in the comparison.
+The minimal number of replicates is determined by the variance of the estimated log-fold change $\mathrm{SE}^{2}(\hat{\Delta})$ as
+\[
+\mathrm{SE}^{2}(\hat{\Delta}) = \left[ \left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right) \right]
+\leq \left( \frac{\Delta}{t_{1-\beta, df} + t_{1-\alpha /2, df}} \right)^{2},
+\]
+where 
+\[
+\alpha = (1 - \beta) \cdot \frac{q}{1 + (1-q) \cdot m_0 / m_1},
+\]
+and $t_{1-\beta, df}$ and $t_{1-\alpha /2, df}$ are the $100(1-\beta)^{\text{th}}$ and the $100(1-\alpha /2)^{\text{th}}$ percentiles of the $t$ distribution.
+The degrees of freedom are estimated for the appropriate model using restricted maximum likelihood.
 

### 2021-10-26 main/MSstatsPTM_Main.tex
Removed duplicate info in supp, changed headers in main

@@
 \renewcommand{\deg}{\ensuremath{^{\circ}}\xspace}
+\newcommand{\dd}[1]{\mathrm{d}#1}
 
@@
 \affil[3]{MPL, Genentech, South San Francisco, CA, USA}
-\affil[4]{Galapagos, Mechelen, Antwerp, Belgium}
+\affil[4]{ULUA BV, Arendstraat 29, 2018 Antwerp, Belgium}
 \affil[*]{Corresponding Authors}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection*{Statistical modeling and parameter estimation}
+\subsubsection*{Existing work in statistical modeling and parameter estimation}
 
-The proposed approach takes as input a list of log-transformed intensities of spectral features, identified and quantified across LC-MS runs.
+The proposed approach first leverages the large amount of work done to create MSstats~\cite{Choi:2014}.
+It takes as input a list of log-transformed intensities of spectral features, identified and quantified across LC-MS runs.
 The features, which are precursor ions of modified or unmodified peptides, are used to characterize the identified PTM sites and proteins.
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection*{Missing value imputation}
-\label{sec:missing_imp}
-
-$MSstatsPTM$ gives the ability to impute missing feature intensities if desired.
+Additionally, the proposed method can include imputing missing feature intensities if desired.
 When values are imputed, it is assumed they are missing for reasons of low abundance.
@@
 
-The methods when applied to a balanced design with multiple conditions and one technical replicate.
+Here we propose a statistical method for detecting changes in PTMs and removing confounding with changes in global protein abundance.
+The method is shown when applied to a balanced design with multiple conditions and one technical replicate.
 In practice the experimental design can be unbalanced and variation can come from multiple sources.
@@
 The null hypothesis states that there is no difference in log-abundance of the PTM site between conditions, adjusted with respect to protein abundance.
-Notationally, given two conditions $i$ and $i'$, the hypothesis test is formulated as follows:
+There are $I$ conditions and $J$ mass spectrometry runs (technical replicates) per condition in the experiment.
+The PTM site is represented by $K$ spectral features (peptide ions, distinguished by their cleavage residues and charge states).
+The log-intensity (base 2) of Feature $k$, in Run $j$ of Condition $i$ is denoted by $y_{ijk}^{\ast}$.
+Conversely, the log-intensity of Feature $l$ from the unmodified peptides in the same run is denoted by $y_{ijl}$.
+The expected values of log-abundances of the PTM and protein in Condition $i$ are denoted by $\mu_{i}^{\ast}$ and $\mu_{i}$.
+The feature and run level summarization for condition $i$ is denoted as $\hat{y}_{i+}^{\ast}$ for the modified peptide and $\hat{y}_{i+}$ for the unmodified peptide.
+For two conditions $i$ and $i'$ the hypothesis test is formulated as follows:
 
-\begin{align*}
-H_{0}: \Delta = \left( \mu_{i}^{\ast} - \mu_{i'}^{\ast} \right) - \left( \mu_{i} - \mu_{i'} \right) &= 0 \\
-H_{a}: \Delta = \left( \mu_{i}^{\ast} - \mu_{i'}^{\ast} \right) - \left( \mu_{i} - \mu_{i'} \right) &\neq 0
-\end{align*}
+\begin{equation}
+\begin{aligned}
+H_{0}: \Delta = ( \mu_{i}^{\ast} - \mu_{i'}^{\ast} ) - ( \mu_{i} - \mu_{i'} ) = 0 \\
+H_{a}: \Delta = ( \mu_{i}^{\ast} - \mu_{i'}^{\ast} ) - ( \mu_{i} - \mu_{i'} ) \neq 0
+\end{aligned}
+\end{equation}
 
@@
 
-\[
-\hat{\Delta} = \left[ \frac{1}{J} \left( \hat{y}_{i+}^{\ast} - \hat{y}_{i'+}^{\ast} \right) \right] - \left[ \frac{1}{J} \left( \hat{y}_{i+} - \hat{y}_{i'+} \right) \right],
-\]
+\begin{equation}
+\hat{\Delta} = \left[ \frac{1}{J} \left( \hat{y}_{i+}^{\ast} - \hat{y}_{i'+}^{\ast} \right) \right] - \left[ \frac{1}{J} \left( \hat{y}_{i+} - \hat{y}_{i'+} \right) \right]
+\end{equation}
 
@@
 
-\[
-SE(\hat{\Delta}) = \left[ \left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right) \right]^{1/2}.
-\]
+\begin{equation}
+SE(\hat{\Delta}) = \left[ \left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right) \right]^{1/2}
+\end{equation}
 
-Finally the degrees of freedom are given by:
+Finally the degrees of freedom are calculated using the following formula.
 
-\[
+\begin{equation}
 \left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right)^2 \bigg/
-\left( \frac{\hat{\sigma}_{\gamma^{\ast}}^{4}}{\mathrm{df}(\gamma^{\ast})} + \frac{\hat{\sigma}_{\gamma}^{4}}{ \mathrm{df}(\gamma)} \right).
-\]
+\left( \frac{\hat{\sigma}_{\gamma^{\ast}}^{4}}{\mathrm{df}(\gamma^{\ast})} + \frac{\hat{\sigma}_{\gamma}^{4}}{ \mathrm{df}(\gamma)} \right)
+\end{equation}
 
@@
 Adjustment for multiple comparisons is performed using the Benjamini-Hochberg procedure to control the false discovery rate at a desired level, e.g., $0.05$\cite{Benjamini:1995}.
-More details on how the PTM model is adjusted are provided in Supplementary Sec.
-\ref{supp-sec:test}
+
+\subsubsection*{Design of PTM experiments in the balanced case}
+\label{sec:design}
+
+The proposed statistical framework allows for design of PTM experiments in terms of sample size calculation and power analysis. 
+%Sample size calculation based on linear models has been described for general applications~\cite{kutner_etal_04a} and specifically for protein significance analysis~\cite{oberg_vitek_09a}. 
+Sample size calculation takes as input a) $q$, the desired false discovery rate, b) $\beta$, the average Type II error rate, c) $\Delta$, the minimal log-fold change in adjusted PTM abundance that we would like to detect, d) $m_0 / (m_0 + m_1)$, the fraction of truly differentially modified PTM sites in the comparison, and e) $\sigma_{\gamma^{\ast}}^{2}$ and $\sigma_{\gamma}^{2}$, the anticipated variances associated to modified and unmodified peptide features, respectively.
+The variances can be derived based on the dataset being analyzed, assuming similar quantitative properties and variations.
+With these values and a user-specified number of conditions, the corresponding number of technical replicates per condition can then be derived, as described in~\cite{kutner_etal_04a}.
+Given the above quantities, the minimal number of replicates $J$ is determined by the variance of the estimated log-fold change $\mathrm{SE}^{2}(\hat{\Delta})$ as
+\[
+\mathrm{SE}^{2}(\hat{\Delta}) = \left[ \frac{2}{J} \left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right) \right]
+\leq \left( \frac{\Delta}{t_{1-\beta, df} + t_{1-\alpha /2, df}} \right)^{2},
+\]
+where 
+\[
+\alpha = (1 - \beta) \cdot \frac{q}{1 + (1-q) \cdot m_0 / m_1},
+\]
+and $t_{1-\beta, df}$ and $t_{1-\alpha /2, df}$ are the $100(1-\beta)^{\text{th}}$ and the $100(1-\alpha /2)^{\text{th}}$ percentiles of the $t$ distribution, with $df = I(J-1)$ degrees of freedom in balanced designs.
+More details can be found in~\cite{oberg_vitek_09a}.
 
@@
 The statistical modeling approaches discussed above can also be extended to complex designs including experiments with additional sources of variation and unbalanced designs.
-The null hypothesis is still that there is no difference in mean PTM abundance between Conditions $i$ and $i'$ after adjusting for changes in mean unmodified peptide abundance between Conditions $i$ and $i'$.
+Equation (1) is the same in this case and the null hypothesis is still that there is no difference in mean PTM abundance between Conditions $i$ and $i'$ after adjusting for changes in mean unmodified peptide abundance between Conditions $i$ and $i'$.
+
 \begin{align*}
@@
 Model inference is done in the same way to how MSstats~\cite{choi_etal_14a} and MSstatsTMT~\cite{Huang:2020} target the global protein.
-In cases where the design in unbalanced, a linear mixed effects model is fit, which takes into account all potential sources of variation, and restricted maximum likelihood is used to estimate the parameters of each model.
+In cases where the design in unbalanced, a linear mixed effects model is fit, which takes into account all potential sources of variation, and restricted maximum likelihood (RML) is used to estimate the parameters of each model.
 Once the parameters are estimated, we can combine the models using modified versions of the formulas in Section \ref{sec:adjustment_summary}.
 
-The log-fold change in the adjusted PTM abundance, $\Delta$, is estimated by 
+The log-fold change in the adjusted PTM abundance, $\Delta$, is now estimated by RML, and Equation (2) becomes 
 \[
@@
 \]
-and the standard error of the estimate $\mathrm{SE}(\hat{\Delta})$ is 
-\[
-SE(\hat{\Delta}) = \left[ \left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right) \right]^{1/2}.
-\]
-%\[
-%\left[ \left( \frac{2}{J} \right) \cdot \left( \hat{\sigma}_{\gamma}^{2} + \hat{\sigma}_{\gamma^{\ast}}^{2} \right) \right]^{1/2}.
-%\]
-The test statistic $\hat{\Delta} / \mathrm{SE}(\hat{\Delta})$ is compared against the $t$ distribution, with degrees of freedom approximated by
-\[
-\left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right)^2 \bigg/
-\left( \frac{\hat{\sigma}_{\gamma^{\ast}}^{4}}{\mathrm{df}(\gamma^{\ast})} + \frac{\hat{\sigma}_{\gamma}^{4}}{ \mathrm{df}(\gamma)} \right).
-\]
+
+The standard error of the estimate $\mathrm{SE}(\hat{\Delta})$ and degrees of freedom in Equations (3) and (4) are unchanged from the balanced case, although $\hat{\sigma}_{\gamma^z\ast}$ and $\hat{\sigma}_{\gamma}$ are now estimated using RML.
+
+Details on how the proposed method can be used to run sample size calculations and power analysis on experiments with complex design can be seen in Supplementary Sec. \ref{supp-sec:complex_design}
 
@@
 
-\paragraph{Code availability}\mbox{}\\
-
-\noindent{{\bf Bioconductor:} http://www.bioconductor.org/packages/release/bioc/html/MSstatsPTM.html}
-
-\noindent{{\bf Github:} https://github.com/Vitek-Lab/MSstatsPTM}
+The code is available on Bioconductor, http://www.bioconductor.org/packages/release/bioc/html/MSstatsPTM.html, and Github, https://github.com/Vitek-Lab/MSstatsPTM.
 
@@
 
-\subsubsection*{Dataset 1 : Computer simulation 1 - Label-free}
+\subsubsection*{The proposed approach corrects for high FDR in computer simulations}
 
@@
 
-\subsubsection*{Dataset 2 : Computer simulation 2 - Label-free missing values and low replicates}
+\subsubsection*{The proposed approach outperforms existing methods in simulations with missing values and low replicates}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection*{Dataset 3 : SpikeIn benchmark - Ubiquitination - Label-free}
+\subsubsection*{In a Label-free benchmark experiment the proposed approach corrects for bias and outperforms existing methods}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection*{Dataset 4 : Human - Ubiquitination - 1mix-TMT}
+\subsubsection*{MSstatsPTM corrects for bias in group comparison TMT experiment}
 
@@
 
-\subsubsection*{Dataset 5 : Mouse - Phosphorylation - 2mix-TMT time series}
+\subsubsection*{Proposed approach removes confounding in time series TMT experiment}
 
@@
 
-\subsubsection*{Dataset 6 : Human - Ubiquitination - Label-free no global profiling run}
+\subsubsection*{Proposed approach is less effective without a separate global profiling run but still fixes bias in Label-free experiment}
 
@@
 
-\subsubsection*{Sample size calculation}
+\subsubsection*{Power analysis shows more replicates are needed in the PTM profiling run in cases of unequal variance.}
 
 Here we analyze the sample size needed to achieve a desired statistical power.
-The proposed approach corrects for confounding between the modified peptide and unmodified protein at a cost of increased variation.
+The proposed approach corrects for confounding between the modified peptide and unmodified protein at the cost of increased variation.
 This can be seen in the calculation for variance in Section \ref{sec:adjustment_summary}.
 When the variation is increased, the experiment requires a larger number of replicates to reach the same power.
-Thus the statistical power is dependent on the variance from both the modified peptide and unmodifed protein as well as the number of replicates in both runs.
+Thus the statistical power is dependent on the variance from both the modified peptide and unmodifed protein as well as the number of replicates in each run.
 

### 2021-10-26 supplementary/ptm_sm.tex
Removed duplicate info in supp, changed headers in main

@@
 \]
-Alternatively, run-level summary to be described in \secref{sec:sum} can also be used. The difference between the means of PTM abundance in Conditions $i$ and $i'$ are estimated as
+Alternatively, run-level summary to be described can also be used. The difference between the means of PTM abundance in Conditions $i$ and $i'$ are estimated as
 \[
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Run-level summarization of feature intensities}
-\label{sec:sum}
-
-Run-level summarization of feature intensities for each PTM site is carried out as in the sub-plot model of MSstats~\cite{choi_etal_14a} or MSstatsTMT~\cite{Huang:2020} depending on acquisition strategy, which involves a) imputation of censored missing values, and b) summarization of feature intensities using Tukey's median polish.
+%\subsection{Run-level summarization of feature intensities}
+%\label{sec:sum}
+
+%Run-level summarization of feature intensities for each PTM site is carried out as in the sub-plot model of MSstats~\cite{choi_etal_14a} or MSstatsTMT~\cite{Huang:2020} depending on acquisition strategy, which involves a) imputation of censored missing values, and b) summarization of feature intensities using Tukey's median polish.
 \cite{tukey_77a} The run-level summary for the PTM in Run $j$ of Condition $i$ is denoted by $\hat{y}_{ij}^{\ast}$.
@@
 
-\subsection{Balanced design with one source of variation}
-\label{sec:model}
-The methods when applied to a balanced design with multiple conditions and one technical replicate are shown in this section.
+%\subsection{Balanced design with one source of variation}
+%\label{sec:model}
+%The methods when applied to a balanced design with multiple conditions and one technical replicate are shown in this section.
 In practice the experimental design can be unbalanced and variation can come from multiple sources.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection{Model-based inference of the underlying abundance}
-\label{sec:infer}
-
-The PTM abundance in each run is represented as 
-\[
-\hat{y}_{ij}^{\ast} = \psi^{\ast} + C_{i}^{\ast} + R_{j(i)}^{\ast},
-\]
-where $\sum_{i=1}^{I} C_{i}^{\ast} = 0$, $R_{j(i)}^{\ast} = \gamma_{j(i)}^{\ast} \stackrel{\text{iid}}{\sim} \mathcal{N}(0, \sigma_{\gamma^{\ast}}^{2})$. 
-Similarly, the protein abundance in each run is expressed as
-\[
-\hat{y}_{ij} = \psi + C_{i} + R_{j(i)},
-\]
-where $\sum_{i=1}^{I} C_{i} = 0$, $R_{j(i)} = \gamma_{j(i)} \stackrel{\text{iid}}{\sim} \mathcal{N}(0, \sigma_{\gamma}^{2})$.
+%\subsubsection{Model-based inference of the underlying abundance}
+%\label{sec:infer}
+
+%The PTM abundance in each run is represented as 
+%\[
+%\hat{y}_{ij}^{\ast} = \psi^{\ast} + C_{i}^{\ast} + R_{j(i)}^{\ast},
+%\]
+%where $\sum_{i=1}^{I} C_{i}^{\ast} = 0$, $R_{j(i)}^{\ast} = \gamma_{j(i)}^{\ast} \stackrel{\text{iid}}{\sim} \mathcal{N}(0, \sigma_{\gamma^{\ast}}^{2})$. 
+%Similarly, the protein abundance in each run is expressed as
+%\[
+%\hat{y}_{ij} = \psi + C_{i} + R_{j(i)},
+%\]
+%where $\sum_{i=1}^{I} C_{i} = 0$, $R_{j(i)} = \gamma_{j(i)} \stackrel{\text{iid}}{\sim} \mathcal{N}(0, \sigma_{\gamma}^{2})$.
 The expected values of log-abundances of the PTM and protein in Condition $i$ are denoted by $\mu_{i}^{\ast}$ and $\mu_{i}$, respectively, and the values are estimated as:
-\begin{align*}
-\hat{\mu}_{i}^{\ast} &= \hat{\psi}^{\ast} + \hat{C}_{i}^{\ast} = \frac{1}{J} \hat{y}_{i+}^{\ast} \\
-\hat{\mu}_{i} &= \hat{\psi} + \hat{C}_{i} = \frac{1}{J} \hat{y}_{i+},
-\end{align*}
-where the standard errors of the estimates are $\mathrm{SE^{\ast}}(\hat{\mu}_{i}) = (\hat{\sigma}_{\gamma^{\ast}}^{2} / J)^{1/2}$ and $\mathrm{SE}(\hat{\mu}_{i}) = (\hat{\sigma}_{\gamma^{\ast}}^{2} / J)^{1/2}$.
+%\begin{align*}
+%\hat{\mu}_{i}^{\ast} &= \hat{\psi}^{\ast} + \hat{C}_{i}^{\ast} = \frac{1}{J} \hat{y}_{i+}^{\ast} \\
+%\hat{\mu}_{i} &= \hat{\psi} + \hat{C}_{i} = \frac{1}{J} \hat{y}_{i+},
+%\end{align*}
+%where the standard errors of the estimates are $\mathrm{SE^{\ast}}(\hat{\mu}_{i}) = (\hat{\sigma}_{\gamma^{\ast}}^{2} / J)^{1/2}$ and $\mathrm{SE}(\hat{\mu}_{i}) = (\hat{\sigma}_{\gamma^{\ast}}^{2} / J)^{1/2}$.
 %\[
@@
 %\]
-Based on the estimates $\hat{\mu}_{i}^{\ast}$ and $\hat{\mu}_{i}$, the adjusted log-abundance of the PTM is given by $(\hat{\mu}_{i}^{\ast} - \hat{\mu}_{i})$ and the standard error of the estimate is 
-\[
-\left[ \frac{1}{J} \left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right) \right]^{1/2}.
-\]
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection{PTM significance analysis}
-\label{sec:test}
-
-With protein-level adjustment, the model-based testing is based on the hypothesis that there is no difference in adjusted PTM abundance between Conditions $i$ and $i'$
+%Based on the estimates $\hat{\mu}_{i}^{\ast}$ and $\hat{\mu}_{i}$, the adjusted log-abundance of the PTM is given by $(\hat{\mu}_{i}^{\ast} - \hat{\mu}_{i})$ and the standard error of the estimate is
+%\[
+%\left[ \frac{1}{J} \left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right) \right]^{1/2}.
+%\]
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%\subsubsection{PTM significance analysis}
+%\label{sec:test}
+
+%With protein-level adjustment, the model-based testing is based on the hypothesis that there is no difference in adjusted PTM abundance between Conditions $i$ and $i'$
 %Adjusted by protein abundance, the hypothesis states that there is no difference in log-abundance of the modification between groups $i$ and $i'$
-\begin{align*}
-H_{0}: \Delta = \left( \mu_{i}^{\ast} - \mu_{i'}^{\ast} \right) - \left( \mu_{i} - \mu_{i'} \right) &= 0 \\
-H_{a}: \Delta = \left( \mu_{i}^{\ast} - \mu_{i'}^{\ast} \right) - \left( \mu_{i} - \mu_{i'} \right) &\neq 0
-\end{align*}
-In a balanced design, the log-fold change in the adjusted PTM abundance, $\Delta$, is estimated by 
-\[
-\hat{\Delta} = \left[ \frac{1}{J} \left( \hat{y}_{i+}^{\ast} - \hat{y}_{i'+}^{\ast} \right) \right] - \left[ \frac{1}{J} \left( \hat{y}_{i+} - \hat{y}_{i'+} \right) \right],
-\]
-and the standard error of the estimate $\mathrm{SE}(\hat{\Delta})$ is 
-\[
-SE(\hat{\Delta}) = \left[ \frac{2}{J} \left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right) \right]^{1/2}.
-\]
+%\begin{align*}
+%H_{0}: \Delta = \left( \mu_{i}^{\ast} - \mu_{i'}^{\ast} \right) - \left( \mu_{i} - \mu_{i'} \right) &= 0 \\
+%H_{a}: \Delta = \left( \mu_{i}^{\ast} - \mu_{i'}^{\ast} \right) - \left( \mu_{i} - \mu_{i'} \right) &\neq 0
+%\end{align*}
+%In a balanced design, the log-fold change in the adjusted PTM abundance, $\Delta$, is estimated by 
+%\[
+%\hat{\Delta} = \left[ \frac{1}{J} \left( \hat{y}_{i+}^{\ast} - \hat{y}_{i'+}^{\ast} \right) \right] - \left[ \frac{1}{J} \left( \hat{y}_{i+} - \hat{y}_{i'+} \right) \right],
+%\]
+%and the standard error of the estimate $\mathrm{SE}(\hat{\Delta})$ is 
+%\[
+%SE(\hat{\Delta}) = \left[ \frac{2}{J} \left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right) \right]^{1/2}.
+%\]
 %\[
@@
 %\]
-The test statistic $\hat{\Delta} / \mathrm{SE}(\hat{\Delta})$ is compared against the $t$ distribution, with degrees of freedom approximated by
-\[
-\left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right)^2 \bigg/
-\left( \frac{\hat{\sigma}_{\gamma^{\ast}}^{4}}{\mathrm{df}(\gamma^{\ast})} + \frac{\hat{\sigma}_{\gamma}^{4}}{ \mathrm{df}(\gamma)} \right).
-\]
-A distinctive property of the proposed model-based testing to the two-sample $t$-test is that even only the PTM abundances in Conditions $i$ and $i'$ are compared, measurements from all conditions are used for the modeling and inference.
-
-The formulations for $\hat{\Delta}$, $SE(\hat{\Delta})$, and degrees of freedom shown above are for a simple balanced experiment.
+%The test statistic $\hat{\Delta} / \mathrm{SE}(\hat{\Delta})$ is compared against the $t$ distribution, with degrees of freedom approximated by
+%\[
+%\left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right)^2 \bigg/
+%\left( \frac{\hat{\sigma}_{\gamma^{\ast}}^{4}}{\mathrm{df}(\gamma^{\ast})} + \frac{\hat{\sigma}_{\gamma}^{4}}{ \mathrm{df}(\gamma)} \right).
+%\]
+%A distinctive property of the proposed model-based testing to the two-sample $t$-test is that even only the PTM abundances in Conditions $i$ and $i'$ are compared, measurements from all conditions are used for the modeling and inference.
+
+%The formulations for $\hat{\Delta}$, $SE(\hat{\Delta})$, and degrees of freedom shown above are for a simple balanced experiment.
 They can also be extended to more realistic experiments with complex designs.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection{Design of PTM experiments}
-\label{sec:design}
-
-The proposed statistical framework allows for design of PTM experiments in terms of sample size calculation and power analysis. 
+%\subsubsection{Design of PTM experiments}
+%\label{sec:design}
+
+%The proposed statistical framework allows for design of PTM experiments in terms of sample size calculation and power analysis. 
 %Sample size calculation based on linear models has been described for general applications~\cite{kutner_etal_04a} and specifically for protein significance analysis~\cite{oberg_vitek_09a}. 
-Sample size calculation takes as input a) $q$, the desired false discovery rate, b) $\beta$, the average Type II error rate, c) $\Delta$, the minimal log-fold change in adjusted PTM abundance that we would like to detect, d) $m_0 / (m_0 + m_1)$, the fraction of truly differentially modified PTM sites in the comparison, and e) $\sigma_{\gamma^{\ast}}^{2}$ and $\sigma_{\gamma}^{2}$, the anticipated variances associated to modified and unmodified peptide features, respectively.
+%Sample size calculation takes as input a) $q$, the desired false discovery rate, b) $\beta$, the average Type II error rate, c) $\Delta$, the minimal log-fold change in adjusted PTM abundance that we would like to detect, d) $m_0 / (m_0 + m_1)$, the fraction of truly differentially modified PTM sites in the comparison, and e) $\sigma_{\gamma^{\ast}}^{2}$ and $\sigma_{\gamma}^{2}$, the anticipated variances associated to modified and unmodified peptide features, respectively.
 The variances can be derived based on the dataset being analyzed, assuming similar quantitative properties and variations.
@@
 Given the above quantities, the minimal number of replicates $J$ is determined by the variance of the estimated log-fold change $\mathrm{SE}^{2}(\hat{\Delta})$ as
-\[
-\mathrm{SE}^{2}(\hat{\Delta}) = \left[ \frac{2}{J} \left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right) \right]
-\leq \left( \frac{\Delta}{t_{1-\beta, df} + t_{1-\alpha /2, df}} \right)^{2},
-\]
-where 
-\[
-\alpha = (1 - \beta) \cdot \frac{q}{1 + (1-q) \cdot m_0 / m_1},
-\]
-and $t_{1-\beta, df}$ and $t_{1-\alpha /2, df}$ are the $100(1-\beta)^{\text{th}}$ and the $100(1-\alpha /2)^{\text{th}}$ percentiles of the $t$ distribution, with $df = I(J-1)$ degrees of freedom in balanced designs.
+%\[
+%\mathrm{SE}^{2}(\hat{\Delta}) = \left[ \frac{2}{J} \left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right) \right]
+%\leq \left( \frac{\Delta}{t_{1-\beta, df} + t_{1-\alpha /2, df}} \right)^{2},
+%\]
+%where 
+%\[
+%\alpha = (1 - \beta) \cdot \frac{q}{1 + (1-q) \cdot m_0 / m_1},
+%\]
+%and $t_{1-\beta, df}$ and $t_{1-\alpha /2, df}$ are the $100(1-\beta)^{\text{th}}$ and the $100(1-\alpha /2)^{\text{th}}$ percentiles of the $t$ distribution, with $df = I(J-1)$ degrees of freedom in balanced designs.
 More details can be found in~\cite{oberg_vitek_09a}.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Extension to complex designs}
-\label{sec:complex_methods}
-
-The methods described previously in Section \ref{sec:model} can be extrapolated to experiments with complex designs targeting PTMs.
+%\subsection{Extension to complex designs}
+%\label{sec:complex_methods}
+
+%The methods described previously in Section \ref{sec:model} can be extrapolated to experiments with complex designs targeting PTMs.
 This includes experiments with additional sources of variation and unbalanced designs.
 
-\subsubsection{Model-based inference and significance analysis}
-
-Model inference is done in the same way to how MSstats~\cite{choi_etal_14a} and MSstatsTMT~\cite{Huang:2020} target the global protein.
+%\subsubsection{Model-based inference and significance analysis}
+
+%Model inference is done in the same way to how MSstats~\cite{choi_etal_14a} and MSstatsTMT~\cite{Huang:2020} target the global protein.
 The null hypothesis is still that there is no difference in mean PTM abundance between Conditions $i$ and $i'$ after adjusting for changes in mean unmodified peptide abundance between Conditions $i$ and $i'$.
-\begin{align*}
-H_{0}: \Delta = (\mu_{i}^{\ast} - \mu_{i'}^{\ast}) - (\mu_{i} - \mu_{i'}) &= 0 \\
-H_{a}: \Delta = (\mu_{i}^{\ast} - \mu_{i'}^{\ast}) - (\mu_{i} - \mu_{i'}) &\neq 0
-\end{align*}
-
-In cases where the design in unbalanced, a linear mixed effects model is fit, which takes into account all potential sources of variation, and restricted maximum likelihood is used to estimate the parameters of each model.
+%\begin{align*}
+%H_{0}: \Delta = (\mu_{i}^{\ast} - \mu_{i'}^{\ast}) - (\mu_{i} - \mu_{i'}) &= 0 \\
+%H_{a}: \Delta = (\mu_{i}^{\ast} - \mu_{i'}^{\ast}) - (\mu_{i} - \mu_{i'}) &\neq 0
+%\end{align*}
+
+%In cases where the design in unbalanced, a linear mixed effects model is fit, which takes into account all potential sources of variation, and restricted maximum likelihood is used to estimate the parameters of each model.
 Once the parameters are estimated, we can combine the models using modified versions of the formulas in Section \ref{sec:test}.
 
-The log-fold change in the adjusted PTM abundance, $\Delta$, is estimated by 
-\[
-\hat{\Delta} = (\hat{\mu}_{RML_i}^{\ast} - \hat{\mu}_{RML_{i'}}^{\ast}) - (\hat{\mu}_{RML_i} - \hat{\mu}_{RML_{i'}})
-\]
-and the standard error of the estimate $\mathrm{SE}(\hat{\Delta})$ is 
-\[
-SE(\hat{\Delta}) = \left[ \left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right) \right]^{1/2}.
-\]
+%The log-fold change in the adjusted PTM abundance, $\Delta$, is estimated by 
+%\[
+%\hat{\Delta} = (\hat{\mu}_{RML_i}^{\ast} - \hat{\mu}_{RML_{i'}}^{\ast}) - (\hat{\mu}_{RML_i} - \hat{\mu}_{RML_{i'}})
+%\]
+%and the standard error of the estimate $\mathrm{SE}(\hat{\Delta})$ is 
+%\[
+%SE(\hat{\Delta}) = \left[ \left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right) \right]^{1/2}.
+%\]
 %\[
@@
 %\]
-The test statistic $\hat{\Delta} / \mathrm{SE}(\hat{\Delta})$ is compared against the $t$ distribution, with degrees of freedom approximated by
-\[
-\left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right)^2 \bigg/
-\left( \frac{\hat{\sigma}_{\gamma^{\ast}}^{4}}{\mathrm{df}(\gamma^{\ast})} + \frac{\hat{\sigma}_{\gamma}^{4}}{ \mathrm{df}(\gamma)} \right).
-\]
+%The test statistic $\hat{\Delta} / \mathrm{SE}(\hat{\Delta})$ is compared against the $t$ distribution, with degrees of freedom approximated by
+%\[
+%\left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right)^2 \bigg/
+%\left( \frac{\hat{\sigma}_{\gamma^{\ast}}^{4}}{\mathrm{df}(\gamma^{\ast})} + \frac{\hat{\sigma}_{\gamma}^{4}}{ \mathrm{df}(\gamma)} \right).
+%\]
+%
 
@@
 Once this is done we can use these these values to calculate the sample size and perform power analysis.
-This is done in a similar way to Section \ref{sec:design}.
 Given $q$, the desired false discovery rate, $\beta$, the average Type II error rate, $m_0 / (m_0 + m_1)$, the fraction of truly differentially modified PTM sites in the comparison.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Datasets : Biological investigation}
+\section{Biological investigations}
 \label{sec:bio_investigations}
@@
 
-A model was fit for the total protein and ubiquitination separately, as described previously for TMT experiments in Section~\ref{sec:tmtmethod}. The model formula can be seen below.
+A model was fit for the total protein and ubiquitination separately. The model formula can be seen below.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Sample size calculation and power analysis}
-
-The proposed approach allows us to calculate the sample size needed to achieve a desired statistical power as described in Section \ref{sec:design}.
+%\section{Sample size calculation and power analysis}
+
+%The proposed approach allows us to calculate the sample size needed to achieve a desired statistical power as described in Section \ref{sec:design}.
 The proposed method adjusts for the underlying protein abundance in the PTM significance analysis, which corrects the confounding factor with a cost of increased variation.
@@
 
-We compared the statistical power in experiments with differing numbers of replicates, variance, and fold change for both the modified and unmodified runs.
+%We compared the statistical power in experiments with differing numbers of replicates, variance, and fold change for both the modified and unmodified runs.
 For variance we used the variance from the biological experiments to determine what values to set.
@@
 
-The results of the power and sample size analysis can be seen in \sfigref{fig:power_sd_combo}.
-
-\begin{figure}[h!]
-\centering
- \begin{subfigure}{\textwidth}
- \centering
-	\includegraphics[width=.725\textwidth]{sim_new/same_var_power}
-	\caption{Here we can see the power of an experiment targeting PTMs with the same variance, .15, for the modified and unmodified peptides.
+%The results of the power and sample size analysis can be seen in \sfigref{fig:power_sd_combo}.
+
+%\begin{figure}[h!]
+%\centering
+% \begin{subfigure}{\textwidth}
+% \centering
+%	\includegraphics[width=.725\textwidth]{sim_new/same_var_power}
+%	\caption{Here we can see the power of an experiment targeting PTMs with the same variance, .15, for the modified and unmodified peptides.
 Predictably when the replicates are high for both modified and unmodified peptides the power is much higher.
@@
 Both cases result in the same power with equal variance.}
- \end{subfigure}\vspace{-5mm}
- \begin{subfigure}{\textwidth}
- \centering
-	\includegraphics[width=.725\textwidth]{sim_new/high_ptm_var_power}
-	\caption{In this chart the variance for the PTM is higher than the unmodified protein.
+% \end{subfigure}\vspace{-5mm}
+% \begin{subfigure}{\textwidth}
+% \centering
+%	\includegraphics[width=.725\textwidth]{sim_new/high_ptm_var_power}
+%	\caption{In this chart the variance for the PTM is higher than the unmodified protein.
 The PTM variance is .2, while the unmodified protein variance is .1.
@@
 It is clearly more important to have high replicates for the PTM run than the unmodified protein, when the PTM variance is higher.}
- \end{subfigure}
- \caption{Power analysis of experiments with differing variances.}
-\label{fig:power_sd_combo}
-\end{figure}
+% \end{subfigure}
+% \caption{Power analysis of experiments with differing variances.}
+%\label{fig:power_sd_combo}
+%\end{figure}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\clearpage
-\section*{References}
+%\clearpage
 

### 2021-10-26 main/MSstatsPTM_Main.tex
Meeting notes

@@
 Additionally, the summarization includes global median normalization and normalizes between MS runs.
-The package uses an AFT model, as mentioned in Section \ref{sec:missing_imp}, to impute missing values, although this step is optional.
+The package uses an AFT model, as mentioned in Section\todo{ \ref{sec:missing_imp}}, to impute missing values, although this step is optional.
 
@@
 \subsubsection*{Power analysis shows more replicates are needed in the PTM profiling run in cases of unequal variance.}
+
+\todo{Noisy PTM measurements benefit from additional biological replicates }
 

### 2021-10-27 main/MSstatsPTM_Main.tex
Updated figures and sample size sections

--- 2021-10-26 main/MSstatsPTM_Main.tex
@@
 Given the above quantities, the minimal number of replicates $J$ is determined by the variance of the estimated log-fold change $\mathrm{SE}^{2}(\hat{\Delta})$ as
-\[
+
+\begin{equation}
 \mathrm{SE}^{2}(\hat{\Delta}) = \left[ \frac{2}{J} \left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right) \right]
-\leq \left( \frac{\Delta}{t_{1-\beta, df} + t_{1-\alpha /2, df}} \right)^{2},
-\]
+\leq \left( \frac{\Delta}{t_{1-\beta, df} + t_{1-\alpha /2, df}} \right)^{2}
+\end{equation}
 where 
-\[
-\alpha = (1 - \beta) \cdot \frac{q}{1 + (1-q) \cdot m_0 / m_1},
-\]
+\begin{equation}
+\alpha = (1 - \beta) \cdot \frac{q}{1 + (1-q) \cdot m_0 / m_1}
+\end{equation}
+
 and $t_{1-\beta, df}$ and $t_{1-\alpha /2, df}$ are the $100(1-\beta)^{\text{th}}$ and the $100(1-\alpha /2)^{\text{th}}$ percentiles of the $t$ distribution, with $df = I(J-1)$ degrees of freedom in balanced designs.
-More details can be found in~\cite{oberg_vitek_09a}.
+
+Solving for J, the number of biological replicates per group is calculated as
+
+\begin{equation}
+J \geq \frac{(2\hat{\sigma}_{\gamma^{\ast}}^{2} + 2\hat{\sigma}_{\gamma}^{2})(t_{1-\beta, df} + t_{1-\alpha /2, df})^2}{\Delta^2}
+\end{equation}
+
+More details on sample size calculation can be found in~\cite{oberg_vitek_09a}.
 
@@
 
-\begin{align*}
-H_{0}: \Delta = (\mu_{i}^{\ast} - \mu_{i'}^{\ast}) - (\mu_{i} - \mu_{i'}) &= 0 \\
-H_{a}: \Delta = (\mu_{i}^{\ast} - \mu_{i'}^{\ast}) - (\mu_{i} - \mu_{i'}) &\neq 0
-\end{align*}
-
-Model inference is done in the same way to how MSstats~\cite{choi_etal_14a} and MSstatsTMT~\cite{Huang:2020} target the global protein.
+Model inference is done in the same way to how MSstats~\cite{Choi:2014} and MSstatsTMT~\cite{Huang:2020} target the global protein.
 In cases where the design in unbalanced, a linear mixed effects model is fit, which takes into account all potential sources of variation, and restricted maximum likelihood (RML) is used to estimate the parameters of each model.
@@
 The log-fold change in the adjusted PTM abundance, $\Delta$, is now estimated by RML, and Equation (2) becomes 
-\[
+\begin{equation}
 \hat{\Delta} = (\hat{\mu}_{RML_i}^{\ast} - \hat{\mu}_{RML_{i'}}^{\ast}) - (\hat{\mu}_{RML_i} - \hat{\mu}_{RML_{i'}})
-\]
+\end{equation}
 
@@
 Additionally, the summarization includes global median normalization and normalizes between MS runs.
-The package uses an AFT model, as mentioned in Section \ref{sec:missing_imp}, to impute missing values, although this step is optional.
+The package uses an AFT model to impute missing values, although this step is optional.
 
@@
 
-\subsubsection*{Power analysis shows more replicates are needed in the PTM profiling run in cases of unequal variance.}
+\subsubsection*{Noisy PTM measurements benefit from additional biological replicates}
 
@@
 \centering
-\includegraphics[scale=.8]{images/fig1.png}
-\caption{\todo{Meena?: Can you write about how MS for PTMs is run?}}
-\label{fig:workflow}
-\end{figure}
-
-\begin{figure}[ht]
-\centering
-\includegraphics[scale=.8]{images/fig3.png}
+\includegraphics[scale=.65]{images/fig3.png}
 \caption{
@@
 The next step is feature level summarization, which summarizes features up to the modification level for the PTM data, and the protein level for the protein data.
-In the final step a model is fit to identify differential PTMs across condtions.}
+In the final step a model is fit to identify differential PTMs and unmodified proteins across conditions and the PTM model is adjusted for changes in the unmodified protein.}
 \label{fig:msstatsptm_design}

### 2021-11-08 main/MSstatsPTM_Main.tex
Updated main and supp for author review

@@
 \usepackage{amsmath}
-\numberwithin{figure}{section} % numbering in each subsection
+%\numberwithin{figure}{section} % numbering in each subsection
 \numberwithin{table}{section}
@@
 For example, it can represent experimental designs with even more complex structures, such as time series or factorial investigations.
+Additionally, the approach can handle experiments with modified peptides processed using label-free methods and unmodified peptides processed using TMT labeling, or vice versa.
+In this case summarization and modeling is still done separately for both the modified and unmodified data, and then combined after modeling.
 

### 2021-11-15 main/MSstatsPTM_Main.tex
Updated per author comments

@@
 The datasets were chosen to provide a performance benchmark of the proposed approach against existing methods when ground truth is known, as well as represent a variety of experimental designs and acquisition methods where the approach can be applied.
-To benchmark the dataset, computer simulations and a spike-in experiment were used.
+To benchmark the approach, computer simulations and a spike-in experiment were used.
 The computer simulations vary in experimental realism; the first simulation was perfectly clean, with many replicates and no missing values, while the second simulation introduced real world qualities, such as limited modified features and missing values.
@@
 Modified peptides were simulated with 2 features, while unmodified proteins were still simulated with 10 features.
-Additionally, 20\% of observations for both modified and unmodified peptides were masked completely at random.
+Additionally, 20\% of observations for both modified and unmodified peptides were missing completely at random.
 Adding missing values and few representative modified features provides a more realistic expectation of model performance.
@@
 For more accurate estimation of protein abundance, separate global proteomics data of unenriched samples are recommended.
-If unmodified features are unavailable for any given modified feature, unmodified intensity adjustment cannot be performed.
+If unmodified peptides are unavailable for any given modification, unmodified intensity adjustment cannot be performed.
 As different levels of variability are present in the data, the log-intensities of the features for modified and unmodified peptides are modeled separately using two linear mixed models.

### 2022-01-27 code/Spike-in_Analysis.Rmd
updated plots per review

@@
 
+## Update some naming conventions
+plot_df[plot_df$Label == 'mix1-mix3'][['contrast_truth']] <- "LogFC: 1 (mix1-mix3)"
+plot_df[plot_df$Label == 'mix1-mix4'][['contrast_truth']] <- "LogFC: 0 (mix1-mix4)"
+plot_df[plot_df$Label == 'mix2-mix3'][['contrast_truth']] <- "LogFC: -2 (mix2-mix3)"
+plot_df[plot_df$Label == 'mix2-mix4'][['contrast_truth']] <- "LogFC: -1 (mix2-mix4)"
+
+plot_df[plot_df$labeling == 'light'][['labeling']] <- "Background Peptides"
+plot_df[plot_df$labeling == 'heavy'][['labeling']] <- "SpikeIn Peptides"
+
 # Volcano plot
-png("supplementary/sim_new/spike_in_msstatsptm_volcano.png", width = 750, height = 500)
-plot_df %>% arrange(desc(labeling)) %>% 
+png("../supplementary/sim_new/spike_in_msstatsptm_volcano.png", width = 750, height = 500)
+plot_df %>% arrange(labeling) %>% 
   filter(abs(log2FC) < Inf & Label %in% c("mix1-mix3", "mix1-mix4", "mix2-mix3", "mix2-mix4")) %>%
   ggplot(aes(log2FC, -log10(adj.pvalue))) + 
-  geom_point(aes(color = labeling, alpha = labeling)) + 
-  scale_color_manual(values = c("#D55E00", "#999999")) + 
-  scale_alpha_manual(values = c(1, 0.2)) + 
-  scale_size_manual(values = c(2, 1)) + 
+  geom_point(aes(color = labeling, alpha = labeling), size = 2.2) + 
+  scale_color_manual(values = c("#999999", "#D55E00")) + 
+  scale_alpha_manual(values = c(0.2, 1)) + 
+  #scale_size_manual(values = c(2, 1)) + 
   geom_hline(yintercept = 0) + 
-  geom_vline(xintercept = 0) + 
+  geom_vline(xintercept = 0, linetype = "dashed") + 
   geom_hline(yintercept = -log10(0.05), color = "darkred") + 
-  geom_vline(xintercept = c(1, -1), color = "darkred", linetype = "dashed") + 
   facet_grid(protadj ~ contrast_truth) + 
   theme_bw() + 
-  theme(axis.text.x = element_text(size = 12), 
-       axis.text.y = element_text(size = 12), 
-       legend.text=element_text(size=12),
-       axis.title.y = element_text(size = 14),
-       axis.title.x = element_text(size = 14),
-       title = element_text(size = 16),
-       strip.text = element_text(size = 12),
-       legend.position = "bottom") + 
+  theme(axis.text.x = element_text(size = 16), 
+       axis.text.y = element_text(size = 16), 
+       legend.text=element_text(size=16),
+       axis.title.y = element_text(size = 18, vjust=2),
+       axis.title.x = element_text(size = 18),
+       title = element_text(size = 20),
+       strip.text = element_text(size = 15),
+       legend.title =  element_blank(),
+       legend.position = "bottom",
+       legend.box.margin=margin(-10,-10,-10,-10)) + 
+  scale_y_continuous(limits = c(0,5)) + 
+  guides(colour = guide_legend(override.aes = list(size=5))) +
   labs(title = "Proposed Method VolcanoPlot")
@@
 
+## Update some naming conventions
+plot_df[plot_df$Label == 'mix1-mix3'][['contrast_truth']] <- "LogFC: 1 (mix1-mix3)"
+plot_df[plot_df$Label == 'mix1-mix4'][['contrast_truth']] <- "LogFC: 0 (mix1-mix4)"
+plot_df[plot_df$Label == 'mix2-mix3'][['contrast_truth']] <- "LogFC: -2 (mix2-mix3)"
+plot_df[plot_df$Label == 'mix2-mix4'][['contrast_truth']] <- "LogFC: -1 (mix2-mix4)"
+
+plot_df[plot_df$labeling == 'light'][['labeling']] <- "Background Peptides"
+plot_df[plot_df$labeling == 'heavy'][['labeling']] <- "SpikeIn Peptides"
+
 # Volcano plot
-# png("supplementary/sim_new/spike_in_limma_volcano.png", width = 750, height = 500)
-plot_df %>% arrange(desc(labeling)) %>% 
+png("../supplementary/sim_new/spike_in_limma_volcano.png", width = 750, height = 500)
+plot_df %>% arrange(labeling) %>% 
   filter(abs(Log2FC) < Inf & Label %in% c("mix1-mix3", "mix1-mix4", "mix2-mix3", "mix2-mix4")) %>%
   ggplot(aes(Log2FC, -log10(adj.pvalue))) + 
-  geom_point(aes(color = labeling, alpha = labeling)) + 
-  scale_color_manual(values = c("#D55E00", "#999999")) + 
-  scale_alpha_manual(values = c(1, 0.2)) + 
-  scale_size_manual(values = c(2, 1)) + 
+  geom_point(aes(color = labeling, alpha = labeling), size = 2.2) + 
+  scale_color_manual(values = c("#999999", "#D55E00")) + 
+  scale_alpha_manual(values = c(0.2, 1)) + 
+  #scale_size_manual(values = c(2, 1)) + 
   geom_hline(yintercept = 0) + 
-  geom_vline(xintercept = 0) + 
+  geom_vline(xintercept = 0, linetype = "dashed") + 
   geom_hline(yintercept = -log10(0.05), color = "darkred") + 
-  geom_vline(xintercept = c(1, -1), color = "darkred", linetype = "dashed") + 
   facet_grid(protadj ~ contrast_truth) + 
   theme_bw() + 
-  theme(axis.text.x = element_text(size = 12), 
-       axis.text.y = element_text(size = 12), 
-       legend.text=element_text(size=12),
-       axis.title.y = element_text(size = 14),
-       axis.title.x = element_text(size = 14),
-       title = element_text(size = 16),
-       strip.text = element_text(size = 12),
-       legend.position = "bottom") + 
+  theme(axis.text.x = element_text(size = 16), 
+       axis.text.y = element_text(size = 16), 
+       legend.text=element_text(size=16),
+       axis.title.y = element_text(size = 18, vjust=2),
+       axis.title.x = element_text(size = 18),
+       title = element_text(size = 20),
+       strip.text = element_text(size = 15),
+       legend.title =  element_blank(),
+       legend.position = "bottom",
+       legend.box.margin=margin(-10,-10,-10,-10)) + 
+  scale_y_continuous(limits = c(0,3)) + 
+  guides(colour = guide_legend(override.aes = list(size=5))) +
   labs(title = "Limma VolcanoPlot")
-# dev.off()
+dev.off()
 ```
@@
 
+## Update some naming conventions
+ttest_final_model[ttest_final_model$label == 'mix1-mix3'][['contrast_truth']] <- "LogFC: 1 (mix1-mix3)"
+ttest_final_model[ttest_final_model$label == 'mix1-mix4'][['contrast_truth']] <- "LogFC: 0 (mix1-mix4)"
+ttest_final_model[ttest_final_model$label == 'mix2-mix3'][['contrast_truth']] <- "LogFC: -2 (mix2-mix3)"
+ttest_final_model[ttest_final_model$label == 'mix2-mix4'][['contrast_truth']] <- "LogFC: -1 (mix2-mix4)"
+
+ttest_final_model[ttest_final_model$labeling == 'light'][['labeling']] <- "Background Peptides"
+ttest_final_model[ttest_final_model$labeling == 'heavy'][['labeling']] <- "SpikeIn Peptides"
+
+ttest_final_model$log2FC <- ttest_final_model$estimate
+
 # Volcano plot
-png("supplementary/sim_new/spike_in_ttest_volcano.png", width = 750, height = 500)
-ttest_final_model %>% arrange(desc(labeling)) %>% 
-  filter(abs(estimate) < Inf & label %in% c("mix1-mix3", "mix1-mix4", "mix2-mix3", "mix2-mix4")) %>%
-  ggplot(aes(estimate, -log10(adj.pvalue))) + 
-  geom_point(aes(color = labeling, alpha = labeling)) + 
-  scale_color_manual(values = c("#D55E00", "#999999")) + 
-  scale_alpha_manual(values = c(1, 0.1)) + 
-  scale_size_manual(values = c(2, 1)) + 
-  geom_hline(yintercept = 0) + 
+png("../supplementary/sim_new/spike_in_ttest_volcano.png", width = 750, height = 500)
+ttest_final_model %>% arrange(labeling) %>% 
+  filter(abs(log2FC) < Inf & label %in% c("mix1-mix3", "mix1-mix4", "mix2-mix3", "mix2-mix4")) %>%
+  ggplot(aes(log2FC, -log10(adj.pvalue))) + 
+  geom_point(aes(color = labeling, alpha = labeling), size = 2.2) + 
+  scale_color_manual(values = c("#999999", "#D55E00")) + 
+  scale_alpha_manual(values = c(0.1, 1)) + 
+  #scale_size_manual(values = c(2, 1)) + 
+  geom_hline(yintercept = 0, linetype = "dashed") + 
   geom_vline(xintercept = 0) + 
   geom_hline(yintercept = -log10(0.05), color = "darkred") + 
-  geom_vline(xintercept = c(1, -1), color = "darkred", linetype = "dashed") + 
   facet_grid(protadj ~ contrast_truth) + 
   theme_bw() + 
-  theme(axis.text.x = element_text(size = 12), 
-       axis.text.y = element_text(size = 12), 
-       legend.text=element_text(size=12),
-       axis.title.y = element_text(size = 14),
-       axis.title.x = element_text(size = 14),
-       title = element_text(size = 16),
-       strip.text = element_text(size = 12),
-       legend.position = "bottom") + 
+  theme(axis.text.x = element_text(size = 14), 
+       axis.text.y = element_text(size = 14), 
+       legend.text=element_text(size=16),
+       axis.title.y = element_text(size = 18, vjust=2),
+       axis.title.x = element_text(size = 18),
+       title = element_text(size = 20),
+       strip.text = element_text(size = 15),
+       legend.title =  element_blank(),
+       legend.position = "bottom",
+       legend.box.margin=margin(-10,-10,-10,-10)) + 
+  scale_y_continuous(limits = c(0,1.5)) + 
+  guides(colour = guide_legend(override.aes = list(size=5))) + 
   labs(title = "T-test VolcanoPlot")
@@
 ```{r, echo = FALSE, fig.width = 8}
+cbPalette <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#999999")
 
@@
   #scale_fill_manual(values = c("salmon1", "steelblue1", "green")) + 
-  labs(title = "Summary Statistic Comparison", x = "Metric", y = "Result") + theme_bw()
+  labs(title = "Summary Statistic Comparison", x = "Metric", y = "Result") + 
+  theme_bw() + 
+    theme(axis.text.x = element_text(size = 18), 
+       axis.text.y = element_text(size = 18), 
+       legend.text=element_text(size=18),
+       axis.title.y = element_text(size = 20, vjust=2),
+       axis.title.x = element_text(size = 20),
+       title = element_text(size = 20),
+       #strip.text = element_text(size = 15),
+       legend.position = "right")
 dev.off()
@@
 
-unadj_spike_ttest <- ttest_final_model %>% filter(labeling == "heavy" & protadj == "no adjustment")
-unadj_spike_ttest <- setNames(unadj_spike_ttest, c("ptm", "label", "pval", "tstat", "SE", 
+ttest_final_model[, log2FC:=NULL]
+unadj_spike_ttest <- ttest_final_model %>% filter(labeling == "SpikeIn Peptides" & protadj == "no adjustment")
+unadj_spike_ttest <- setNames(unadj_spike_ttest, c("ptm", "Label", "pval", "tstat", "SE", 
                                                    "df", "log2FC", "labeling", "adj.pvalue", 
@@
 
-adj_spike_ttest <- ttest_final_model %>% filter(labeling == "heavy" & protadj != "no adjustment")
-adj_spike_ttest <- setNames(adj_spike_ttest, c("ptm", "label", "pval", "tstat", "SE", 
+adj_spike_ttest <- ttest_final_model %>% filter(labeling == "SpikeIn Peptides" & protadj != "no adjustment")
+adj_spike_ttest <- setNames(adj_spike_ttest, c("ptm", "Label", "pval", "tstat", "SE", 
                                                    "df", "log2FC", "labeling", "adj.pvalue", 
@@
                         adj_spike_ttest, unadj_spike_limma, adj_spike_limma), fill = TRUE)
-spike[Model == "MSstatsPTM"]$Model <- "Proposed"
+spike <- as.data.table(spike)
+spike[spike$Model == "MSstatsPTM", Model := "Proposed"]
+#plot_df[plot_df$Label == 'mix1-mix3'][['contrast_truth']]
 
@@
                  Model = factor(Model, levels=c("Proposed", "Limma", "T-test"))) %>%  ggplot() + 
-  geom_boxplot(aes(x = Comparison, y = log2FC, fill = Model)) + 
-  geom_point(aes(y = trueLog2FC, x = Comparison), color = "red", shape = 4, size = 4, stroke = 2) + 
+  geom_boxplot(aes(x = Label, y = log2FC, fill = Model)) + 
+  geom_point(aes(y = trueLog2FC, x = Label), color = "red", shape = 4, size = 4, stroke = 2) + 
       scale_fill_manual(values=cbPalette) +
@@
     theme_bw() + 
-  theme(axis.text.x = element_text(size = 11, angle = 0, vjust = 0), 
-       axis.text.y = element_text(size = 11), 
-       legend.text=element_text(size=12),
-       axis.title.y = element_text(size = 14),
-       axis.title.x = element_text(size = 14),
-       title = element_text(size = 16),
-       strip.text = element_text(size = 12),
+  theme(axis.text.x = element_text(angle = 60, hjust=1, size = 16), 
+       axis.text.y = element_text(size = 16), 
+       legend.text=element_text(size=18),
+       axis.title.y = element_text(size = 18),
+       axis.title.x = element_text(size = 18),
+       title = element_text(size = 20),
+       strip.text = element_text(size = 18),
        legend.position = "bottom") + 

### 2022-01-27 main/MSstatsPTM_Main.tex
updated plots per review

@@
 \usepackage{tabulary,multirow,multicol,rotating}
-\usepackage[backend=biber]{biblatex}
+\usepackage[backend=biber,sorting=none]{biblatex}
 \addbibresource{bibliography.bib}
@@
 The signaling mechanisms that allow cells to mount a dynamic and fast response to a multitude of events are primarily facilitated by the modification of proteins at specific residues, acting as molecular on/off switches\cite{Deribe} \cite{Cohen}.
-Mass spectrometry-based label-free proteomics is broadly established as the tool-of-choice for unbiased and large-scale identification and quantification of proteins and their post-translational modifications (PTMs) using liquid chromatography coupled with mass spectrometry (LC-MS)\cite{Kall:2011ub} \cite{Roepstorff}.
+Mass spectrometry-based proteomics is broadly established as the tool-of-choice for unbiased and large-scale identification and quantification of proteins and their post-translational modifications (PTMs) using liquid chromatography coupled with mass spectrometry (LC-MS)\cite{Kall:2011ub} \cite{Roepstorff}.
 Studies targeting the post-translationally modified proteome focus either on the accurate localization of modification sites on proteins, relative or absolute quantification of a modification site's occupancy repertoire, or relative changes in occupancy across experimental conditions\cite{Mann}.
@@
 \centering
+\begin{subfigure}[c]{0.825\linewidth}
 \includegraphics[width=1\textwidth]{images/spike_in_msstatsptm_volcano.png}
+\caption{}
+\label{fig:spikein_prop_volcano}
+\end{subfigure}
+\begin{subfigure}[c]{0.825\linewidth}
+\includegraphics[width=1\textwidth]{images/spike_in_limma_volcano.png}
+\caption{}
+\label{fig:spikein_limma_volcano}
+\end{subfigure}
 \caption{Dataset 3 : SpikeIn benchmark - Ubiquitination - Label-free.
-Using the proposed method to model the benchmark experiment the spike in peptides (colored red) do not follow the expected log fold change before adjustment.
-After adjusting for changes in overall protein abundance the spike in peptides are more in line with expectation.
-Additionally the background grey colored peptides showed many false positives before adjustment.
-After adjustment the false positives were decreased considerably.
-}\label{fig:spikein_prop_volcano}
+A) Using the proposed method the spike in peptides (colored red) do not follow the expected log fold change before adjustment.
+After adjustment the spike in peptides are more in line with expectation.
+Additionally, the background peptides showed many false positives before adjustment, but decrease considerably after adjustment.
+B) Using Limma the spike in peptides again follow the expected log fold change after adjusting for changes in protein level.
+However, while the fold change is more accurate, the majority of spike in peptides do not have a significant adjusted pvalue.}
+\label{fig:spikein_volcano}
 \end{figure}
@@
 \centering
-\includegraphics[width=1\textwidth]{images/spike_in_limma_volcano.png}
-\caption{Dataset 3 : SpikeIn benchmark - Ubiquitination - Label-free.
-Here the Limma method was used to model the spike in experiment.
-Using Limma the spike in peptides follow the expected log fold change better after adjusting for changes in protein level.
-However, while the fold change is much more accurate, the majority of spike in peptides do not have a significant adjusted pvalue.
-In terms of false positives, the results are very similar to the proposed method, with many false positives before adjustment and much fewer after.
-}\label{fig:spikein_limma_volcano}
-\end{figure}
-
-\begin{figure}[ht]
-\centering
-\includegraphics[height=.65\textwidth]{images/ipah_venn_diagramm.png}
+\begin{subfigure}[c]{0.5\linewidth}
+\includegraphics[width=1\textwidth]{images/ipah_venn_diagramm.png}
+\caption{}
+\label{fig:data4_venn_diagram}
+\end{subfigure}
+\begin{subfigure}[c]{0.95\linewidth}
+\includegraphics[width=1\textwidth]{images/IpaH_prof_plot.png}
+\caption{}
+\label{fig:data4_profile_plot}
+\end{subfigure}
 \caption{Dataset 4 : Human - Ubiquitination - 1mix-TMT.
-The overlap of differencial modified peptides for the PTM model with and without global protein level adjustment.
+A) The overlap of differential modified peptides for the PTM model with and without global protein level adjustment.
 More PTMs became insignificant after adjustment then became significant.
-For the peptides that became insignificant in the adjusted model, their change in abundance was mainly due to changes in global protein abundance.
-In contrast, for peptides that became significant after adjustment, their true abundance change was masked by underlying changes in the unmodified protein.}
-\label{fig:ipah_venn_diagramm}
-\end{figure}
-
-\begin{figure}[ht]
-\centering
-\includegraphics[width=1\textwidth]{images/IpaH_prof_plot.png}
-\caption{Dataset 4 : Human - Ubiquitination - 1mix-TMT.
-Comparing the global profiling of protein $GSDMD\_HUMAN|P57764$ with the ubiquitination of the protein at site $K62$.
-The individual PSM features are shown in grey, while the feature summarization is shown in red.
+For the peptides that became insignificant in the adjusted model, their change in abundance was driven by changes in the global protein.
+In contrast, peptides that became significant after adjustment saw their true abundance change masked by underlying changes in the unmodified protein.
+B) Comparing the global profiling of protein $GSDMD$ with the ubiquitination of the protein at site $K62$.
 When looking at the summary of the modification and global protein it is clear the conditions follow different trends.
@@
 This indicates the modification is confounded with changes in the unmodified protein.}
-\label{fig:ipah_profile}
+\label{fig:data4_plots}
 \end{figure}
@@
  \centering
-	\includegraphics[width=.725\textwidth]{images/high_ptm_var_power}
+	\includegraphics[width=.9\textwidth]{images/high_ptm_var_power}
 	\caption{In this chart the variance for the PTM is higher than the unmodified protein.
@@
 
-%{\bf Figure 4} False positive rate and statistical power in PTM significance analysis by the proposed approach, t-test with protein-level adjustment, and t-test without protein-level adjustment.
-(a) When there was no change in protein abundance, all considered methods well calibrated the Type I error rate.
-(b) When the PTM changes were entirely due to changes in protein abundance across conditions, analysis accounting for the protein-level changes resulted in off-target, high false positive rates.
-(c) Statistical powers by the proposed approach and t-test with protein-level adjustment, where the data consisted of 2 conditions and the SD corresponding unmodified feature intensities was 0.2.
-(d) Statistical powers by the proposed approach and t-test with protein-level adjustment, where the data consisted of 2 conditions, the SD corresponding unmodified feature intensities was 0.2, and the PTM was missing in Run 1 of Condition 1.
-(e) Same as in (d), but the data consisted of 4 conditions.
-
-%{\bf Figure 5} Results corresponding to estimation error and false positive rate and statistical power of the PTM significance analysis, where the data were acquired in two batches.
-The following parameters were considered to represent the batch effects: no batch-condition interaction, mean intensity level in Batch 2 was higher than Batch 1 by 2 on log scale, and the SDs in Batch 1 and Batch 2 were 0.2 and 0.3, respectively.(a)Estimation based on the most statistically significant batch with t-test was more variable than other methods and frequently biased.
-(b)The proposed approach better calibrated Type I error rate.
-(c)The proposed approach improved statistical power in all the considered cases with different number of conditions and replicates.
-This was achieved by properly characterizing batch effects and leveraging all available information.
-Ignoring batch effects by t-test (no batch) lost power dramatically.
-The negative impact was only partially reduced by increasing the sample size to 5.
-
-%{\bf Figure 6} Design of future PTM experiments in terms of sample size calculations and power analysis.
-(a) Protein-level adjustment relies on the inference of protein abundance, which introduces additional uncertainty in the estimate of PTM difference.
-Therefore, the required sample size to detect a systematic change is higher than as expected for standard differential analysis without adjustment.
-Sample size calculations without accounting for the uncertainty would lead to over-optimistic, under-powered studies.
-(b) In complex designs, simultaneously analyzing all the conditions effectively increases the degrees of freedom and requires fewer replicates.
-(c)Increasing the sample size and analyzing multiple conditions together both result in improved statistical power.
-
 \end{document}

### 2022-01-27 supplementary/ptm_sm.tex
Meeting notes
updated plots per review

@@
 
-\usepackage[backend=biber]{biblatex}
+\usepackage[backend=biber,sorting=none]{biblatex}
 \bibliography{ptm_ref}
@@
 \section{Existing methods} \label{sec:ttest}
-\subsection{two-sample $t$-test}
+\subsection{Two-sample $t$-test}
 
@@
 
-\subsubsection{Design of PTM experiments}
+\subsection{Design of PTM experiments}
 \label{sec:complex_design}
@@
 \centering
-\includegraphics[height=.65\textwidth]{sim_new/ipah_venn_diagramm.png}
-\caption{The overlap of differential modified peptides for the PTM model with and without global protein level adjustment.
-More PTMs became insignificant after adjustment then became significant.
-For the peptides that became insignificant in the adjusted model, their change in abundance was mainly due to changes in global protein abundance.
-In contrast, for peptides that became significant after adjustment, their true abundance change was masked by underlying changes in the unmodified protein.
-An additional question that must be addressed is if the decrease in significant peptides is due to the increased variance that comes from adjustment.
-This was tested by looking for modified peptides who's adjusted log fold change was within 10\% of the unadjusted log fold change but became insignificant after adjustment.
-In other words, the fold change was the same between models but variance increased.
-When this test was applied on this experiment, only one peptide became insignificant due to an increase in variance.
-Thus we can conclude that the drop off in significant peptides was truly due to changes in global protein abundance.}
-\label{fig:ipha_venn_d}
-\end{figure}
-
-\begin{figure}[h!]
-\centering
  \begin{subfigure}{\textwidth}
@@
  \centering
-	\includegraphics[width=.9\textwidth]{sim_new/No_Difference_Shigella_Profile_Plot}
+	\includegraphics[width=.95\textwidth]{sim_new/No_Difference_Shigella_Profile_Plot}
 	\caption{Comparing the global profiling of protein $TTP\_MOUSE|P22893$ with the modification of the protein at site $S178$.

### 2022-01-27 code/IpaH_analysis.R

@@ -56,56 +56,69 @@ gsmd_sum$BioReplicate <- factor(gsmd_sum$BioReplicate,
                                            "NoDox6hr_2", "Dox1hr_1", "Dox2hr_1",
                                            "Dox2hr_2", "Dox4hr_1",
                                            "Dox4hr_2", "Dox6hr_1", "Dox6hr_2"))
+gsmd_62_sum <- as.data.table(gsmd_62_sum)
+gsmd_62_sum[gsmd_62_sum$FeatureType == 'Model'][['FeatureType']] <- "PTM Summarized"
+gsmd_62_sum[gsmd_62_sum$FeatureType == 'Peptide'][['FeatureType']] <- "PTM Feature"
 
-p1 <- gsmd_62_sum %>% ggplot() + geom_line(aes(x = BioReplicate, y = Abundance, color = FeatureType,
+gsmd_sum[gsmd_sum$FeatureType == 'Model'][['FeatureType']] <- "Protein Summarized"
+gsmd_sum[gsmd_sum$FeatureType == 'Peptide'][['FeatureType']] <- "Protein Feature"
+
+p1 <- gsmd_62_sum %>% ggplot() + geom_line(aes(x = BioReplicate, y = Abundance, 
+                                               color = FeatureType,
                                                group = PSM, size = FeatureType)) +
-  geom_point(aes(x = BioReplicate, y = Abundance, color = FeatureType), size = 3) +
+  geom_point(aes(x = BioReplicate, y = Abundance, color = FeatureType), size = 5) +
   geom_vline(data=data.frame(x = c(2.5, 4.5, 5.5, 7.5, 9.5)),
              aes(xintercept=as.numeric(x)), linetype = "dashed") +
-  scale_colour_manual(values = c("#D55E00", "#C3C3C3")) +
+  scale_colour_manual(values = c("#C3C3C3", "#D55E00")) +
   scale_size_manual(values = c(1.5,1)) +
-  labs(title = "GSDMD_HUMAN|P57764_K62", x = "BioReplicate", y = "Abundance") +
+  labs(title = "GSDMD with ubiquitination at site K62", x = "BioReplicate", 
+       y = "Abundance") +
   theme_bw() +
-  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size = 12),
-        axis.text.y = element_text(size = 12),
-        legend.text=element_text(size=12),
-        axis.title.y = element_text(size = 14),
-        axis.title.x = element_text(size = 14),
[…]

### 2022-01-27 code/Shigella_Analysis.R

@@ -1,0 +1,474 @@
+
+library(MSstatsPTM)
+library(data.table)
+library(tidyverse)
+library(gridExtra)
+library(ggrepel)
+library(VennDiagram)
+
+## Load msstats input data
+load(file = "shigella_pST_TMTFormat.rda")#shigella_peptide_pST
+load(file = "shigella_global_TMTFormat.rda")#shigella_global
+load(file = "shigella_pY_TMTFormat.rda")#shigella_peptide_pY
+
+## Setup for summarization
+shigella_global <- shigella_global %>% group_by(ProteinName, PeptideSequence, 
+  Charge, PSM, Mixture, TechRepMixture, Run, Channel, Condition, BioReplicate) %>%
+  summarize(Intensity = max(Intensity))
+
+pY_input <- list(PTM = data.frame(shigella_peptide_pY), 
+                 PROTEIN = data.frame(shigella_global))
+pST_input <- list(PTM = data.frame(shigella_peptide_pST), 
+                  PROTEIN = data.frame(shigella_global))
+
+## Data summarization ----------------------------------------------------------
+py_summarization <- dataSummarizationPTM_TMT(pY_input)
+pST_summarization <- dataSummarizationPTM_TMT(pST_input)
+
+save(py_summarization, file = "Shigella_pY_Summarization.rda")
+save(pST_summarization, file = "Shigella_pST_Summarization.rda")
+
+py_summarized_data <- py_summarization$PTM$ProteinLevelData
+py_feature_data <- py_summarization$PTM$FeatureLevelData
+
+pst_summarized_data <- pST_summarization$PTM$ProteinLevelData
+pst_feature_data <- pST_summarization$PTM$FeatureLevelData
+
+global_summarized_data <- pST_summarization$PROTEIN$ProteinLevelData
+global_feature_data <- pST_summarization$PROTEIN$FeatureLevelData
+
+
+## Create pYST dataset
+## Create pY/pST Dataset
+## check the overlapped sites between pST and pY
+shared_sites <- intersect(unique(py_summarized_data$Protein), 
+                          unique(pst_summarized_data$Protein))
+shared_ST <- shared_sites[grepl("_S", shared_sites)|grepl("_T", shared_sites)]
+shared_Y <- shared_sites[grepl("_Y", shared_sites)]
+
+## remove the overlapped Y sites from data_ST
[…]

### 2022-01-27 code/Simulated_Analysis.R

@@ -16,10 +16,10 @@ convert_to_msstats_format <- function(df){
   df$PTM$BioReplicate <- paste0(df$PTM$run, "rep")
   df$PTM$log2inty <- 2**df$PTM$log2inty
   df$PTM$PeptideSequence <- paste(df$PTM$protein, df$PTM$feature, sep = "_")
-  
+
   setnames(df$PTM, c("protein", "group", "run", "log2inty"),
            c("ProteinName", "Condition", "Run", "Intensity"))
-  
+
   df$PROTEIN$PrecursorCharge <- NA
   df$PROTEIN$FragmentIon <- NA
   df$PROTEIN$ProductCharge <- NA
@@ -29,15 +29,15 @@ convert_to_msstats_format <- function(df){
   df$PROTEIN$PeptideSequence <- paste(df$PROTEIN$protein, df$PROTEIN$feature, sep = "_")
   setnames(df$PROTEIN, c("protein", "group", "run", "log2inty"),
            c("ProteinName", "Condition", "Run", "Intensity"))
-  
+
   return(df)
 }
 
 ## Run ttest using run level summared data
 run_ttest <- function(summarized_data){
-  
+
   ptms <- unique(summarized_data$PTM)
-  
+
   ttest_temp <- data.table()
   ttest_adj_temp <- data.table()
   ## Run ttest for each ptm
@@ -45,32 +45,32 @@ run_ttest <- function(summarized_data){
     temp_joined <- summarized_data %>% filter(PTM == ptms[[p]])
     groups <- (temp_joined %>% distinct(Condition))[[1]]
     t <- 2
-    
+
     ## Loop over groups
     for (g in 1:(length(groups)-1)){
       for (g2 in 2:length(groups)){
-        tryCatch({ttest_ptm <- t.test((temp_joined %>% filter(Condition == groups[g]) %>% select(Abundance.x))[[1]], 
+        tryCatch({ttest_ptm <- t.test((temp_joined %>% filter(Condition == groups[g]) %>% select(Abundance.x))[[1]],
                                       (temp_joined %>% filter(Condition == groups[g2]) %>% select(Abundance.x))[[1]])
-        ttest_temp <- rbindlist(list(ttest_temp, data.table(ptm = ptms[[p]], 
-                                                            label = paste(groups[g], "vs", groups[g2], sep = " "), 
-                                                            pval = ttest_ptm$p.value, 
+        ttest_temp <- rbindlist(list(ttest_temp, data.table(ptm = ptms[[p]],
[…]

### 2022-01-27 code/USP30_MSstatsPTM_Analysis.R

@@ -86,45 +86,50 @@ ptm_list1 <- c("P52209_K059")
 #                                                 "KO_Late_1", "KO_Late_2", "KO_Late_3", "KO_Late_4"))
 
 features <- summarized_ptm$PTM$FeatureLevelData %>% filter(PROTEIN == ptm_list1[[1]])
-features$Feature <- "PSM"
+features$FeatureType <- "Peptide"
 features$Protein <- features$PROTEIN
 features$Abundance <- features$ABUNDANCE
 
 summarized <- summarized_ptm$PTM$ProteinLevelData %>% filter(Protein == ptm_list1[[1]])
 summarized$Abundance <- summarized$LogIntensities
-summarized$Feature <- "Summary"
+summarized$FeatureType <- "Model"
 
 test <- rbindlist(list(features, summarized), fill = TRUE)
 
+test[test$FeatureType == 'Model'][['FeatureType']] <- "PTM Summarized"
+test[test$FeatureType == 'Peptide'][['FeatureType']] <- "PTM Feature"
+
 p1 <- test %>% ggplot() +
-  geom_line(aes(x = originalRUN, y = Abundance , group = FEATURE, color = Feature,  size = Feature)) +
-  geom_point(aes(x = originalRUN, y = Abundance , group = FEATURE, color = Feature), size = 3) +
+  geom_line(aes(x = originalRUN, y = Abundance , group = FEATURE, color = FeatureType), size =2) +
+  geom_point(aes(x = originalRUN, y = Abundance , group = FEATURE, color = FeatureType), size = 5) +
   geom_vline(data=data.frame(x = c(4.5, 8.5, 12.5)),
              aes(xintercept=as.numeric(x)), linetype = "dashed") +
   scale_colour_manual(values = c("#C3C3C3", "#D55E00")) +
-  scale_size_manual(values = c(1, 2)) +
-  labs(title = ptm_list1[[1]], x = "BioReplicate", y = "Abundance") +
+  #scale_size_manual(values = c(1, 2)) +
+  labs(title = "P52209 with ubiquitination at site K059", x = "BioReplicate", y = "Abundance") +
   theme_bw() +
-  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size = 12),
-        axis.text.y = element_text(size = 12),
-        legend.text=element_text(size=12),
-        axis.title.y = element_text(size = 14),
[…]

### 2022-01-28 code/MSstatsPTM_Simulation_Analysis.Rmd
Changes for internal review

@@
 adj_results$Model <- "Proposed"
-no_adj_results$Model <- "Proposed no Adjustment"
-ttest_results$Model <- "T-Test no Adjustment"
+no_adj_results$Model <- "Proposed no adjustment"
+ttest_results$Model <- "T-Test no adjustment"
 ttest_adj_results$Model <- "T-Test"
-limma_model_results$Model <- "Limma no Adjustment"
+limma_model_results$Model <- "Limma no adjustment"
 limma_adj_model_results$Model <- "Limma"
@@
 ```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width = 7, fig.height = 6}
+
+models$plot_conditions = paste0(models$conditions, " Conditions")
+models$plot_sd = paste0("SD ", models$sd)
+
 png("../supplementary/sim_new/sim1_FDR_all_models.png", width = 750, height = 500)
 
-models %>% mutate(Model = factor(Model, levels=c("Proposed", "Proposed no Adjustment", "Limma", 
-                                               "Limma no Adjustment", "T-Test", "T-Test no Adjustment")),
+models %>% mutate(Model = factor(Model, levels=c("Proposed", "Proposed no adjustment", "Limma", 
+                                               "Limma no adjustment", "T-Test", "T-Test no adjustment")),
                   Replicates = as.factor(rep)) %>%
@@
   scale_colour_manual(values=cbPalette_first) +
-  facet_grid(vars(sd), vars(conditions)) + 
+  scale_size_manual(values = c(3,5,7,9)) + 
+  facet_grid(vars(plot_sd), vars(plot_conditions)) + 
   theme_bw() +
-  theme(axis.text.x = element_blank(), 
-                                                 axis.text.y = element_text(size = 12), 
-                                                 legend.text=element_text(size=12),
-                                                 axis.title.y = element_text(size = 14),
-                                                 axis.title.x = element_text(size = 14),
-                                                 title = element_text(size = 16),
-                                                 strip.text = element_text(size = 12)) + 
-  labs(title = "Simulation 1: FDR all models", x = "Model", y = "FDR")
+  theme(axis.text.x = element_text(angle = 60, hjust=1, size = 12), 
+         axis.text.y = element_text(size = 14), 
+         legend.text=element_text(size=14),
+         axis.title.y = element_text(size = 22),
+         axis.title.x = element_text(size = 22),
+         title = element_text(size = 22),
+         strip.text = element_text(size = 16)) + 
+  labs(title = "Simulation 1: FDR all models", x = "Model", y = "FDR") + 
+     guides(color = guide_legend(order = 1, override.aes = list(size=5)), 
+              size = guide_legend(order = 2, reverse = TRUE))
 
@@
 ``` {r, echo = FALSE, message=FALSE, warning=FALSE}
+
+models$plot_conditions = paste0(models$conditions, " Conditions")
+models$plot_sd = paste0("SD ", models$sd)
+
 png("../supplementary/sim_new/sim1_FDR.png", width = 750, height = 500)
-
-models %>% mutate(Model = factor(Model, levels=c("Proposed", "Proposed no Adjustment", "Limma", 
-                                               "Limma no Adjustment", "T-Test", "T-Test no Adjustment")),
+models %>% mutate(Model = factor(Model, levels=c("Proposed", "Proposed no adjustment", "Limma", 
+                                               "Limma no adjustment", "T-Test", "T-Test no adjustment")),
                   Replicates = as.factor(rep)) %>%
@@
   scale_colour_manual(values=cbPalette) +
-  facet_grid(vars(sd), vars(conditions)) + 
+  scale_size_manual(values = c(3,5,7,9)) + 
+  facet_grid(vars(plot_sd), vars(plot_conditions)) + 
   theme_bw() +
-  theme(axis.text.x = element_blank(), 
-                                                 axis.text.y = element_text(size = 12), 
-                                                 legend.text=element_text(size=12),
-                                                 axis.title.y = element_text(size = 14),
-                                                 axis.title.x = element_text(size = 14),
-                                                 title = element_text(size = 16),
-                                                 strip.text = element_text(size = 12)) + 
-  labs(title = "Simulation 1: FDR adjusted models", x = "Model", y = "FDR")
+  theme(axis.text.x = element_text(angle = 60, hjust=1, size = 16), 
+         axis.text.y = element_text(size = 16), 
+         legend.text=element_text(size=18),
+         axis.title.y = element_text(size = 22),
+         axis.title.x = element_text(size = 22),
+         title = element_text(size = 22),
+         strip.text = element_text(size = 16)) + 
+  labs(title = "Simulation 1: FDR adjusted models", x = "Model", y = "FDR") + 
+   guides(color = guide_legend(order = 1, override.aes = list(size=5)), 
+              size = guide_legend(order = 2, reverse = TRUE))
 dev.off()
+
+# axis.text.x = element_text(angle = 60, hjust=1, size = 16),
+#         axis.text.y = element_text(size = 16),
+#         legend.text=element_text(size=18),
+#         axis.title.y = element_text(size = 22),
+#         axis.title.x = element_text(size = 22),
+#         title = element_text(size = 22),
+#         strip.text = element_text(size = 16),
+
 ```
@@
 ```{r echo=FALSE, message=FALSE, warning=FALSE}
+
+models$plot_conditions = paste0(models$conditions, " Conditions")
+models$plot_sd = paste0("SD ", models$sd)
+
 png("../supplementary/sim_new/sim1_Recall.png", width = 750, height = 500)
-models %>% mutate(Model = factor(Model, levels=c("Proposed", "Proposed no Adjustment", "Limma", 
-                                               "Limma no Adjustment", "T-Test", "T-Test no Adjustment")),
+models %>% mutate(Model = factor(Model, levels=c("Proposed", "Proposed no adjustment", "Limma", 
+                                               "Limma no adjustment", "T-Test", "T-Test no adjustment")),
                   Replicates = as.factor(rep)) %>%
@@
   scale_colour_manual(values=cbPalette) +
-  facet_grid(vars(sd), vars(conditions)) + 
+  scale_size_manual(values = c(3,5,7,9)) + 
+  facet_grid(vars(plot_sd), vars(plot_conditions)) + 
   theme_bw() +
-  theme(axis.text.x = element_blank(), 
-                                                 axis.text.y = element_text(size = 12), 
-                                                 legend.text=element_text(size=12),
-                                                 axis.title.y = element_text(size = 14),
-                                                 axis.title.x = element_text(size = 14),
-                                                 title = element_text(size = 16),
-                                                 strip.text = element_text(size = 12)) + 
-  labs(title = "Simulation 1: Recall adjusted models", x = "Model", y = "Recall")
+  theme(axis.text.x = element_text(angle = 60, hjust=1, size = 16), 
+         axis.text.y = element_text(size = 16), 
+         legend.text=element_text(size=18),
+         axis.title.y = element_text(size = 22),
+         axis.title.x = element_text(size = 22),
+         title = element_text(size = 22),
+         strip.text = element_text(size = 16)) + 
+  labs(title = "Simulation 1: Recall adjusted models", x = "Model", y = "Recall") + 
+   guides(color = guide_legend(order = 1, override.aes = list(size=5)), 
+              size = guide_legend(order = 2, reverse = TRUE))
 dev.off()
@@
   scale_colour_manual(values=cbPalette) +
-  facet_grid(vars(sd), vars(conditions)) + 
+  scale_size_manual(values = c(3,5,7,9)) + 
+  facet_grid(vars(plot_sd), vars(plot_conditions)) + 
   theme_bw() +
-  theme(axis.text.x = element_blank(), 
-                                                 axis.text.y = element_text(size = 12), 
-                                                 legend.text=element_text(size=12),
-                                                 axis.title.y = element_text(size = 14),
-                                                 axis.title.x = element_text(size = 14),
-                                                 title = element_text(size = 16),
-                                                 strip.text = element_text(size = 12)) + 
-  labs(title = "Simulation 1: Accuracy adjusted models", x = "Model", y = "Accuracy")
+  theme(axis.text.x = element_text(angle = 60, hjust=1, size = 16), 
+         axis.text.y = element_text(size = 16), 
+         legend.text=element_text(size=18),
+         axis.title.y = element_text(size = 22),
+         axis.title.x = element_text(size = 22),
+         title = element_text(size = 22),
+         strip.text = element_text(size = 16)) + 
+  labs(title = "Simulation 1: Accuracy adjusted models", x = "Model", y = "Accuracy") + 
+   guides(color = guide_legend(order = 1, override.aes = list(size=5)), 
+              size = guide_legend(order = 2, reverse = TRUE))
 
@@
 
+boxplot_data$plot_conditions = paste0(boxplot_data$cond, " Conditions")
+boxplot_data$plot_sd = paste0("SD ", boxplot_data$sd)
+
 png("../supplementary/sim_new/sim1_FC_boxplot.png", width = 750, height = 500)
@@
   ggplot()  + geom_hline(yintercept = .75, color = "red", size = 1.1) + 
-  geom_boxplot(aes(y = log2FC, x = reps, fill = Model)) + facet_grid(vars(sd), vars(cond)) + 
+  geom_boxplot(aes(y = log2FC, x = reps, fill = Model)) + 
+  facet_grid(vars(plot_sd), vars(plot_conditions)) + 
   scale_fill_manual(values=cbPalette) +
   theme_bw() +
-  theme(axis.text.x = element_text(size = 12), 
-                                                 axis.text.y = element_text(size = 12), 
-                                                 legend.text=element_text(size=12),
-                                                 axis.title.y = element_text(size = 14),
-                                                 axis.title.x = element_text(size = 14),
-                                                 title = element_text(size = 16),
-                                                 strip.text = element_text(size = 12)) + 
+  theme(axis.text.x = element_text(size = 16), 
+         axis.text.y = element_text(size = 16), 
+         legend.text=element_text(size=18),
+         axis.title.y = element_text(size = 22),
+         axis.title.x = element_text(size = 22),
+         title = element_text(size = 22),
+         strip.text = element_text(size = 16)) + 
   labs(title = "Simulation 1: Fold Change Distribution", x = "Replicates", y = "Log2FC")
@@
 
-load(file = "model_data/ptm_models_sim2.rda")#ptm_models_sim1
-load(file = "model_data/adjusted_models_sim2.rda")#adjusted_models_sim1 
-load(file = "model_data/ttest_models_sim2.rda")#ttest_sim1
-load(file = "model_data/adj_ttest_models_sim2.rda")#adj_ttest_sim1
-load(file = "model_data/limma_models_sim2.rda")#limma_results_sim1
-load(file = "model_data/adj_limma_models_sim2.rda")#adj_limma_sim1
+load(file = "../data/ptm_models_sim2.rda")#ptm_models_sim1
+load(file = "../data/adjusted_models_sim2.rda")#adjusted_models_sim1 
+load(file = "../data/ttest_models_sim2.rda")#ttest_sim1
+load(file = "../data/adj_ttest_models_sim2.rda")#adj_ttest_sim1
+load(file = "../data/limma_models_sim2.rda")#limma_results_sim1
+load(file = "../data/adj_limma_models_sim2.rda")#adj_limma_sim1
 
@@
 ``` {r, echo = FALSE, message=FALSE, warning=FALSE}
+models$plot_conditions = paste0(models$conditions, " Conditions")
+models$plot_sd = paste0("SD ", models$sd)
+
 png("../supplementary/sim_new/sim3_FDR.png", width = 350, height = 350)
@@
 ```{r echo=FALSE, message=FALSE, warning=FALSE}
+models$plot_conditions = paste0(models$conditions, " Conditions")
+models$plot_sd = paste0("SD ", models$sd)
+
 png("../supplementary/sim_new/sim3_Recall.png", width = 750, height = 500)
@@
   scale_colour_manual(values=cbPalette) +
-  facet_grid(vars(sd), vars(conditions)) + 
+  scale_size_manual(values = c(3,5,7,9)) + 
+  facet_grid(vars(plot_sd), vars(plot_conditions)) + 
   theme_bw() +
-  theme(axis.text.x = element_blank(), 
-                                                 axis.text.y = element_text(size = 12), 
-                                                 legend.text=element_text(size=12),
-                                                 axis.title.y = element_text(size = 14),
-                                                 axis.title.x = element_text(size = 14),
-                                                 title = element_text(size = 16),
-                                                 strip.text = element_text(size = 12)) + 
-  labs(title = "Simulation 2: Recall adjusted models", x = "Model", y = "Recall")
+  theme(axis.text.x = element_text(angle = 60, hjust=1, size = 16), 
+         axis.text.y = element_text(size = 16), 
+         legend.text=element_text(size=18),
+         axis.title.y = element_text(size = 22),
+         axis.title.x = element_text(size = 22),
+         title = element_text(size = 22),
+         strip.text = element_text(size = 16)) + 
+  labs(title = "Simulation 2: Recall adjusted models", x = "Model", y = "Recall") + 
+   guides(color = guide_legend(order = 1, override.aes = list(size=5)), 
+              size = guide_legend(order = 2, reverse = TRUE))
 dev.off()
+
 ```
@@
   scale_colour_manual(values=cbPalette) +
-  facet_grid(vars(sd), vars(conditions)) + 
+  scale_size_manual(values = c(3,5,7,9)) + 
+  facet_grid(vars(plot_sd), vars(plot_conditions)) + 
   theme_bw() +
-  theme(axis.text.x = element_blank(), 
-                                                 axis.text.y = element_text(size = 12), 
-                                                 legend.text=element_text(size=12),
-                                                 axis.title.y = element_text(size = 14),
-                                                 axis.title.x = element_text(size = 14),
-                                                 title = element_text(size = 16),
-                                                 strip.text = element_text(size = 12)) + 
-  labs(title = "Simulation 2: Accuracy adjusted models", x = "Model", y = "Accuracy")
+  theme(axis.text.x = element_text(angle = 60, hjust=1, size = 16), 
+         axis.text.y = element_text(size = 16), 
+         legend.text=element_text(size=18),
+         axis.title.y = element_text(size = 22),
+         axis.title.x = element_text(size = 22),
+         title = element_text(size = 22),
+         strip.text = element_text(size = 16)) + 
+  labs(title = "Simulation 2: Accuracy adjusted models", x = "Model", y = "Accuracy") + 
+   guides(color = guide_legend(order = 1, override.aes = list(size=5)), 
+              size = guide_legend(order = 2, reverse = TRUE))
 dev.off()
@@
 
+boxplot_data$plot_conditions = paste0(boxplot_data$cond, " Conditions")
+boxplot_data$plot_sd = paste0("SD ", boxplot_data$sd)
+
 png("../supplementary/sim_new/sim3_FC_boxplot.png", width = 750, height = 500)
@@
   theme_bw() +
-  theme(axis.text.x = element_text(size = 12), 
-                                               axis.text.y = element_text(size = 12), 
-                                               legend.text=element_text(size=12),
-                                               axis.title.y = element_text(size = 14),
-                                               axis.title.x = element_text(size = 14),
-                                               title = element_text(size = 16),
-                                               strip.text = element_text(size = 12)) + 
+  theme(axis.text.x = element_text(size = 16), 
+         axis.text.y = element_text(size = 16), 
+         legend.text=element_text(size=18),
+         axis.title.y = element_text(size = 22),
+         axis.title.x = element_text(size = 22),
+         title = element_text(size = 22),
+         strip.text = element_text(size = 16)) + 
 labs(title = "Simulation 2: Fold Change Distribution", x = "Replicates", y = "Log2FC")

### 2022-01-28 main/MSstatsPTM_Main.tex
Changes for internal review

@@
 
+There are two main biological questions that are generally relevant in PTM analysis.
+First, some researchers run mass-spectrometry based proteomic experiments targeting PTMs when looking for biomarkers\cite{Petushkova_2017}.
+In this case the researcher may not necessarily want to correct for confounding with the total protein.
+The interest is in detecting modified peptides which change in abundance in a meaningful way, such as between disease and control, fast or slow-progressing, or responder and non-responder groups.
+Conversely, researchers may be interested in whether the fraction of peptides modified at a given site is changing between conditions.
+This is an important question for thinking about biological and physiological models and very short timescale experiments.
+In this case, correcting for confounding with the total protein is relevant.
+These two questions are both important areas of study, and require different statistical methods to correctly address.
+In this paper we will mainly address a statistical solution to the second question, which requires more rigorous methods as we must address the convolution with the total protein.
+
 We propose a general statistical approach, which explicitly characterizes the variations and confounding factors present in bottom-up PTM experiments.
@@
 
-The proposed approach was evaluated on an experiment where Shigella ubiquitin ligase IpaH7.8 was shown to inhibit the protein gasdermin D (GSDMD)\cite{LUCHETTI2021}.
+The proposed approach was evaluated on an experiment where Shigella ubiquitin ligase IpaH7.8 was shown to degrade the protein gasdermin D (GSDMD)\cite{LUCHETTI2021}.
 Multiplex proteomics was used to quantify the abundance of total protein and ubiquitination in human epithelial cells.
-Cells were either infected or uninfected with IpaH7.8-deficient Shigella flexneri and measurements were taken at different time periods.
+Cells were engineered to express IpaH7.8 under a dox inducible promoter and measurements were taken at different time periods.
+GSDMD was actively degraded when IpaH7.8 expression was induced by dox treatment.
 Uninfected cells were measured at 0 and 6 hours, while infected cells were measured at 1, 2, 4, and 6 hour increments, resulting in six total conditions.
@@
 The model parameters are estimated using the split-plot approach as in $MSstats$, where the feature log-intensities are first summarized into a single value per site per run in the subplot model, and the site-level summaries are then used for the inference of the PTM site abundance\cite{Choi:2014}.
-This approach allows the method to be extended to cases where the experimental design in inbalanced, and where additional sources of variation are present.
+This approach allows the method to be extended to cases where the experimental design is unbalanced, and where additional sources of variation are present.
 In the site-level summarization, Tukey's median polish (TMP), a simple and robust procedure is applied to iteratively fit a two-way additive model with the effects of run and feature, which in turn summarizes the log-intensities for each site\cite{Tukey:1977}.
@@
 \end{aligned}
+\label{eq_1}
 \end{equation}
@@
 Model inference is done in the same way to how MSstats~\cite{Choi:2014} and MSstatsTMT~\cite{Huang:2020} target the global protein.
-In cases where the design in unbalanced, a linear mixed effects model is fit, which takes into account all potential sources of variation, and restricted maximum likelihood (RML) is used to estimate the parameters of each model.
+In cases where the design is unbalanced, a linear mixed effects model is fit, which takes into account all potential sources of variation, and restricted maximum likelihood (RML) is used to estimate the parameters of each model.
 Once the parameters are estimated, we can combine the models using modified versions of the formulas in Section \ref{sec:adjustment_summary}.
@@
 If the global profiling run is not available, the package can still analyze the modified run, but will do so without adjusting for changes in unmodified protein abundance.
+Specifically, if a global profiling run is not available the null and alternative hypothesis in Equation \ref{eq_1} will reduce to:
+
+\begin{equation}
+\begin{aligned}
+H_{0}: \Delta = \mu_{i}^{\ast} - \mu_{i} = 0 \\
+H_{a}: \Delta = \mu_{i}^{\ast} - \mu_{i} \neq 0
+\end{aligned}
+\label{eq_9}
+\end{equation}
+
+Which amounts to running a the process without accounting for protein adjustment, as seen in MSstats and MSstatsTMT \cite{Choi:2014} \cite{Huang:2020}.
 
@@
 For the experiments where ground truth is known, we calculated summary statistics using the number of true positives (TP), false positives (FP), true negatives (TN), and false negatives (FN).
-To evaluate the performance of different methods, we considered the false discovery rate (FDR = FP/(TP+FP)), the recall rate (TP/(TP + FN), and total accuracy ((TP + TN) / (TP + TN + FP + FN)).
+To evaluate the performance of different methods, we considered the false discovery rate (FDR = FP/(TP+FP)), the recall rate (TP/(TP + FN), and accuracy ((TP + TN) / (TP + TN + FP + FN)).
 For biological experiments where ground truth was not known, we evaluated significantly differential peptides with $\alpha=.05$.
@@
 Once controlling for changes in unmodified peptide, it is clear that the proposed method performed strongest across all methods, as seen in Figure \ref{fig:sim_accuracy}.
-The proposed method well calibrated model accuracy and the recall rate, even when the number of replicates were low.
+The proposed method calibrated model accuracy and recall rate well, even when the number of replicates were low.
 Finally, when comparing the fold change estimation across all peptides, the proposed method showed a tighter distribution of estimated fold changes around the true fold change.
-In comparison, Limma and t-test showed much wider distributions.
+In comparison, Limma and t-test showed wider distributions.
 This fold change comparison can be seen in Figure \ref{fig:sim_boxplot}.
@@
 An additional question that must be addressed is if the decrease in significant peptides is due to the increased variance that comes from adjustment.
-This was tested by looking for modified peptides who's adjusted log fold change was within 10\% of the unadjusted log fold change but became insignificant after adjustment.
+This was tested by looking for modified peptides whose adjusted log fold change was within 10\% of the unadjusted log fold change but became insignificant after adjustment.
 In other words, the fold change was the same between models but variance increased.
@@
 Looking at the Dox4hr vs Dox1hr conditions and modeling the modified peptide without adjusting for changes in the global profiling, the fold change was $-.501$ and the adjusted pvalue was insignificant at $.0644$.
-After adjusting for changes in the global protein abundance, the fold change is much higher, $2.79$, and the adjusted pvalue became very significant, $5.25e^{-8}$.
+After adjusting for changes in the global protein abundance, the fold change is much higher, $2.79$, and the adjusted pvalue became very significant, $5.25e-8$.
 In this case the effect of the modified peptide is strongly confounded with changes in the global protein.
@@
 Again we tested if the decrease in significant peptides was due to the increased variance that comes from adjustment, or if it was mainly due to removing convolution with change in the umodified protein.
-This was tested by looking for modified peptides who's adjusted log fold change was within 10\% of the unadjusted log fold change but became insignificant after adjustment.
+This was tested by looking for modified peptides whose adjusted log fold change was within 10\% of the unadjusted log fold change but became insignificant after adjustment.
 When this test was applied on this experiment, 548 peptides became insignificant due to an increase in variance.
@@
 Each PTM site is modeled and characterized separately, where a PTM is quantified with multiple spectral features (boxes), distinguished by different charge states of a peptide.
-The feature intensities are viewed as repeated measurements of the underlying abundance of the PTM, where the abundance in Condition i is denoted by i.
-Features corresponding to unmodified peptides are considered together to perform adjustment with respect to protein abundance, where the protein abundance in Condition i is denoted by i*.
+The feature intensities are viewed as repeated measurements of the underlying abundance of the PTM, where the abundance in Condition i is denoted by i*.
+Features corresponding to unmodified peptides are considered together to perform adjustment with respect to protein abundance, where the protein abundance in Condition i is denoted by i.
 Peptides can be fully cleaved (solid lines) and/or partially cleaved (dashed lines).
@@
 Even at lower replicates the proposed method still outperformed Limma.
-The lowest performaning method was $t$-test.
+The lowest performing method was $t$-test.
 Limma shows comparable performance to the proposed method in a clean experiment, however when real world data problems are introduced it is clear the proposed method is more robust.}

### 2022-01-28 supplementary/ptm_sm.tex
Changes for internal review

@@
 After adjustment all three models generally have a median near the true fold change.
-In terms of model comparison, the proposed method shows a much tighter distribution around the true fold change, whole both Limma and $t$-test are much wider.
+In terms of model comparison, the proposed method shows a much tighter distribution around the true fold change, whole both Limma and $t$-test are wider.
 \label{fig:spike_fc}}
@@
 Uninfected cells were measured at 0 and 6 hours, while infected cells were measured at 1, 2, 4, and 6 hour increments, resulting in six total conditions.
-The experiment was inbalanced with two bioreplicates per condition for all conditions except for infected 1 hour.
+The experiment was unbalanced with two bioreplicates per condition for all conditions except for infected 1 hour.
 The experiment was modeled as a group comparison, with different replicates measured at each time point.
@@
 Again we tested if the decrease in significant peptides was due to the increased variance that comes from adjustment.
-This was tested by looking for modified peptides who's adjusted log fold change was within 10\% of the unadjusted log fold change but became insignificant after adjustment.
+This was tested by looking for modified peptides whose adjusted log fold change was within 10\% of the unadjusted log fold change but became insignificant after adjustment.
 When this test was applied on this experiment, 548 peptides became insignificant due to an increase in variance.

### 2022-01-28 code/Simulated_Analysis.R

@@ -347,12 +347,20 @@ adj_limma_sim1 <- adj_limma
 # save(limma_results, file = "model_data/limma_models_5_rep_75change.rda")
 # save(adj_limma, file = "model_data/adj_limma_models_5_rep_75change.rda")
 
-load(file = "model_data/ptm_models_sim1.rda")
-load(file = "model_data/adjusted_models_sim1.rda")
-load(file = "model_data/ttest_models_sim1.rda")
-load(file = "model_data/adj_ttest_models_sim1.rda")
-load(file = "model_data/limma_models_sim1.rda")
-load(file = "model_data/adj_limma_models_sim1.rda")
+
+load(file = "../data/ptm_models_sim1.rda")
+load(file = "../data/adjusted_models_sim1.rda")
+load(file = "../data/ttest_models_sim1.rda")
+load(file = "../data/adj_ttest_models_sim1.rda")
+load(file = "../data/limma_models_sim1.rda")
+load(file = "../data/adj_limma_models_sim1.rda")
+
+ptm_models <- ptm_models_sim1
+adjusted_models <- adjusted_models_sim1
+ttest <- ttest_sim1
+adj_ttest <- adj_ttest_sim1
+limma_results <- limma_results_sim1
+adj_limma <- adj_limma_sim1
 
 ## Calculate Summary Stats -----------------------------------------------------
 no_adj_results <- data.table()

### 2022-02-01 code/MSstatsPTM_Simulation_Analysis.Rmd
Fixed all plots

@@
 png("../supplementary/sim_new/sim1_FDR_all_models.png", width = 750, height = 500)
-
 models %>% mutate(Model = factor(Model, levels=c("Proposed", "Proposed no adjustment", "Limma", 
                                                "Limma no adjustment", "T-Test", "T-Test no adjustment")),
-                  Replicates = as.factor(rep)) %>%
-  ggplot() + geom_point(aes(x = Model, y = fpr, size = Replicates, color = Model)) + 
+                  ) %>%
+  filter(rep == 5) %>% 
+  ggplot() + geom_point(aes(x = Model, y = fpr, color = Model), size = 9) + #, size = Replicates
   scale_colour_manual(values=cbPalette_first) +
-  scale_size_manual(values = c(3,5,7,9)) + 
+  # scale_size_manual(values = c(3,5,7,9)) + 
   facet_grid(vars(plot_sd), vars(plot_conditions)) + 
   theme_bw() +
-  theme(axis.text.x = element_text(angle = 60, hjust=1, size = 12), 
+  ylim(0.0, 0.45) + 
+  theme(axis.text.x = element_text(angle = 45, hjust=1, vjust = 1, size = 12), 
          axis.text.y = element_text(size = 14), 

### 2022-02-01 code/Spike-in_Analysis.Rmd
Fixed all plots

@@
 
+proposed_plot_df <- plot_df
+
 # Volcano plot
@@
 
+#### Calculating background peptide percentages
+
+```{r, echo=FALSE}
+
+light_pep <- model_df$PTM.Model %>% filter(labeling == "light")
+
+unadj_sig_light <- model_df$PTM.Model %>% filter(labeling == "light" & adj.pvalue < .05)
+adj_sig_light <- model_df$ADJUSTED.Model %>% filter(labeling == "light" & adj.pvalue < .05)
+
+nrow(unadj_sig_light) / nrow(light_pep)
+nrow(adj_sig_light) / nrow(light_pep)
+
+```
+
 ### Limma
@@
 plot_df[plot_df$labeling == 'heavy'][['labeling']] <- "SpikeIn Peptides"
+
+limma_plot_df <- plot_df
 
@@
 This is most likely due to an increase in standard error or lower degrees of freedom compared to MSstatsPTM.
+
+#### Create combo plot of proposed and limma
+
+``` {r, echo = FALSE}
+
+proposed_plot_df$Model <- "Proposed Method"
+limma_plot_df$Model <- "Limma"
+limma_plot_df$Protein <- limma_plot_df$PTM
+limma_plot_df$log2FC <- limma_plot_df$Log2FC
+
+combo_plot_df <- rbind(proposed_plot_df[, c("Protein", "Label", "log2FC", 
+                                            "adj.pvalue", "labeling", 
+                                            "trueLog2FC", "contrast_truth", 
+                                            "protadj", "Model")], 
+                       limma_plot_df[, c("Protein", "Label", "log2FC", 
+                                            "adj.pvalue", "labeling", 
+                                            "trueLog2FC", "contrast_truth", 
+                                            "protadj", "Model")])
+combo_plot_df$Model <- factor(combo_plot_df$Model, levels=c('Proposed Method','Limma'))
+
+png("../main/images/spike_in_volcano.png", width = 750, height = 500)
+combo_plot_df %>% arrange(Model, labeling) %>% 
+  filter(abs(log2FC) < Inf & Label %in% c("mix2-mix3", "mix2-mix4")) %>%
+  ggplot(aes(log2FC, -log10(adj.pvalue))) + 
+  geom_point(aes(color = labeling, alpha = labeling), size = 2.2) + 
+  scale_color_manual(values = c("#999999", "#D55E00")) + 
+  scale_alpha_manual(values = c(0.2, 1)) + 
+  #scale_size_manual(values = c(2, 1)) + 
+  geom_hline(yintercept = 0) + 
+  geom_vline(xintercept = 0, linetype = "dashed") + 
+  geom_hline(yintercept = -log10(0.05), color = "darkred") + 
+  facet_grid(protadj ~ Model + contrast_truth) + 
+  theme_bw() + 
+  theme(axis.text.x = element_text(size = 16), 
+       axis.text.y = element_text(size = 16), 
+       legend.text=element_text(size=16),
+       axis.title.y = element_text(size = 18, vjust=2),
+       axis.title.x = element_text(size = 18),
+       title = element_text(size = 20),
+       strip.text = element_text(size = 15),
+       legend.title =  element_blank(),
+       legend.position = "bottom",
+       legend.box.margin=margin(-10,-10,-10,-10)) + 
+  scale_y_continuous(limits = c(0,3)) + 
+  guides(colour = guide_legend(override.aes = list(size=5))) +
+  labs(title = "SpikeIn VolcanoPlot", x= "Log2FC")
+dev.off()
+
+```
 
@@
 #plot_df[plot_df$Label == 'mix1-mix3'][['contrast_truth']]
+spike$`color_label` <- "True LogFC"
 
@@
 spike %>% mutate(Comparison = str_split(contrast_truth, " ", simplify = TRUE)[,1],
-                 Model = factor(Model, levels=c("Proposed", "Limma", "T-test"))) %>%  ggplot() + 
+      Model = factor(Model, levels=c("Proposed", "Limma", "T-test"))) %>%
+  filter(!Label %in% c("mix1-mix2", "mix3-mix4")) %>% 
+  ggplot() + 
   geom_boxplot(aes(x = Label, y = log2FC, fill = Model)) + 
-  geom_point(aes(y = trueLog2FC, x = Label), color = "red", shape = 4, size = 4, stroke = 2) + 
+  geom_point(aes(y = trueLog2FC, x = Label, color = color_label), 
+             shape = 4, size = 4, stroke = 2) + 
   scale_fill_manual(values=cbPalette) +
+  scale_color_manual(name = "", values="red", label="True LogFC") + 
  facet_wrap(.~Adjustment) + 
     theme_bw() + 
-  theme(axis.text.x = element_text(angle = 60, hjust=1, size = 16), 
+  theme(axis.text.x = element_text(angle = 45, hjust=1, vjust = 1, size = 16), 
        axis.text.y = element_text(size = 16), 

### 2022-02-01 main/MSstatsPTM_Main.tex
Fixed all plots

@@
 
-There are two main biological questions that are generally relevant in PTM analysis.
-First, some researchers run mass-spectrometry based proteomic experiments targeting PTMs when looking for biomarkers\cite{Petushkova_2017}.
+Generally, there are two main biological questions that are relevant in PTM analysis.
+First, some researchers run mass-spectrometry based experiments targeting PTMs when looking for biomarkers\cite{Petushkova_2017}.
 In this case the researcher may not necessarily want to correct for confounding with the total protein.
@@
 This is an important question for thinking about biological and physiological models and very short timescale experiments.
-In this case, correcting for confounding with the total protein is relevant.
-These two questions are both important areas of study, and require different statistical methods to correctly address.
-In this paper we will mainly address a statistical solution to the second question, which requires more rigorous methods as we must address the convolution with the total protein.
+In this case, we are interested in the relative abundance change solely from the modification and correcting for confounding with the total protein is relevant.
+These two questions are both important areas of study and require different statistical methods to correctly address.
+In this paper we will mainly address a statistical solution to the second question, which requires more rigorous methods as we must address confounding with the total protein.
 
@@
 
-Here simulations were generated mimicking a Label-free experiment, with a high number of modified and unmodified peptide features and no missing values.
+Here simulations were generated mimicking a Label-free experiment with a high number of modified and unmodified peptide features and no missing values.
 Multiple simulations were generated with different numbers of replicates and conditions.
-In all simulations 1000 modified peptides were generated, half of which were true positives and the other half true negatives.
+In all simulations 1000 modified peptides were generated, half of which were true positives and half true negatives.
 Half of the true positives and true negatives were generated convoluted with global protein abundance.
@@
 The performance of the proposed method was evaluated on simulated and spike in datasets with known ground truth, as well as biological experimental data where the ground truth was not known.
-For the experiments where ground truth is known, we calculated summary statistics using the number of true positives (TP), false positives (FP), true negatives (TN), and false negatives (FN).
-To evaluate the performance of different methods, we considered the false discovery rate (FDR = FP/(TP+FP)), the recall rate (TP/(TP + FN), and accuracy ((TP + TN) / (TP + TN + FP + FN)).
-For biological experiments where ground truth was not known, we evaluated significantly differential peptides with $\alpha=.05$.
+For the experiments where ground truth is known, we calculated the number of true positives (TP), false positives (FP), true negatives (TN), and false negatives (FN).
+To determine positives and negatives we adjusted the p-values for multiple testing and used an $\alpha=.05$ cutoff to determine differential sites.
+These results were then compared to the ground truth to determine TP and TN.
+We then used these results to calculate the false discovery rate (FDR = FP/(TP+FP)), the recall rate (TP/(TP + FN), and accuracy ((TP + TN) / (TP + TN + FP + FN)).
+Finally, the summary statistics were compared across methods to analyze method performance.
+
+For biological experiments where ground truth was not known, we again adjusted the p-values for multiple testing and used an $\alpha=.05$ cutoff to determine differential sites.
 Comparisons were made before and after adjusting for changes in unmodified protein abundance and the results are presented in the form of venn diagrams and profile plots.
+Each biological dataset presents a different experimental scenario, showing the applicability of the proposed method.
 All datasets in this paper were evaluated using the $MSstatsPTM$ package.
@@
 \centering
+\begin{subfigure}[c]{0.825\linewidth}
 \includegraphics[width=1\textwidth]{images/sim1_FDR_all_models.png}
-\caption{Dataset 1 : Computer simulation 1 - Label-free.
-All the considered methods in the first computer simulation correctly calibrated FDR when adjusting for changes in protein abundance.
+\caption{}
+\label{fig:spikein_prop_volcano}
+\end{subfigure}
+\begin{subfigure}[c]{0.825\linewidth}
+\includegraphics[width=1\textwidth]{images/sim3_Accuracy.png}
+\caption{}
+\end{subfigure}
+\caption{Dataset 1 \& 2: Computer simulation.
+a) All the considered methods in the first computer simulation correctly calibrated FDR when adjusting for changes in protein abundance.
 In comparison, the methods without accounting for the protein-level changes resulted in off-target, high false positive rates.
-}\label{fig:fdr_all_models}
-\end{figure}
-
-\begin{figure}[ht]
-\centering
-\includegraphics[width=1\textwidth]{images/sim3_Accuracy.png}
-\caption{Dataset 2 : Computer simulation 2 - Label-free missing values and low replicates.
-The advantage of using the proposed approach is apparent when looking at the second computer simulation, which includes limited observations and the presence of missing values.
-In overall accuracy, the proposed method performs stronger than Limma and t-test in nearly every model.
-Even at lower replicates the proposed method still outperformed Limma.
-The lowest performing method was $t$-test.
-Limma shows comparable performance to the proposed method in a clean experiment, however when real world data problems are introduced it is clear the proposed method is more robust.}
-\label{fig:sim_accuracy}
-\end{figure}
-
-\begin{figure}[ht]
-\centering
-\includegraphics[width=1\textwidth]{images/sim3_FC_boxplot.png}
-\caption{Dataset 2 : Computer simulation 2 - Label-free missing values and low replicates.
-In the second computer simulation with missing values and low PTM features all methods correctly estimated the fold change with a median log change of .75.
-The proposed method in this simulation had a visibly tighter distribution around the median when compared to Limma and $t$-test.
-While the mean estimated fold changes were similar, the proposed method exhibited a stronger performance in correctly estimating the fold change for all peptides.}
-\label{fig:sim_boxplot}
+b) The advantage of using the proposed approach is apparent when including limited observations and missing values.
+Looking at accuracy, the proposed method outperforms Limma and t-test in nearly every model.
+}
+\label{fig:computer_sim}
 \end{figure}
@@
 \begin{subfigure}[c]{0.825\linewidth}
-\includegraphics[width=1\textwidth]{images/spike_in_msstatsptm_volcano.png}
+\includegraphics[width=1\textwidth]{images/spike_in_volcano.png}
 \caption{}
@@
 \begin{subfigure}[c]{0.825\linewidth}
-\includegraphics[width=1\textwidth]{images/spike_in_limma_volcano.png}
+\includegraphics[width=1\textwidth]{images/spike_in_fc.png}
 \caption{}
@@
 \caption{Dataset 3 : SpikeIn benchmark - Ubiquitination - Label-free.
-A) Using the proposed method the spike in peptides (colored red) do not follow the expected log fold change before adjustment.
+a) Using the proposed method the spike in peptides (colored red) do not follow the expected log fold change before adjustment.
 After adjustment the spike in peptides are more in line with expectation.
-Additionally, the background peptides showed many false positives before adjustment, but decrease considerably after adjustment.
-B) Using Limma the spike in peptides again follow the expected log fold change after adjusting for changes in protein level.
-However, while the fold change is more accurate, the majority of spike in peptides do not have a significant adjusted pvalue.}
+Using Limma the spike in peptides again follow the expected log fold change after adjusting for changes in protein level.
+However, while the fold change is more accurate, the majority of spike in peptides do not have a significant adjusted pvalue.
+b) say something about this one.}
 \label{fig:spikein_volcano}
@@
 \caption{Dataset 4 : Human - Ubiquitination - 1mix-TMT.
-A) The overlap of differential modified peptides for the PTM model with and without global protein level adjustment.
+a) The overlap of differential modified peptides for the PTM model with and without global protein level adjustment.
 More PTMs became insignificant after adjustment then became significant.
@@
 In contrast, peptides that became significant after adjustment saw their true abundance change masked by underlying changes in the unmodified protein.
-B) Comparing the global profiling of protein $GSDMD$ with the ubiquitination of the protein at site $K62$.
+b) Comparing the global profiling of protein $GSDMD$ with the ubiquitination of the protein at site $K62$.
 When looking at the summary of the modification and global protein it is clear the conditions follow different trends.
@@
 \centering
-\includegraphics[height=.65\textwidth]{images/shig_venn_diagramm.png}
+\begin{subfigure}{\textwidth}
+ \centering
+\includegraphics[height=.5\textwidth]{images/shig_venn_diagramm.png}
+\caption{}
+ \end{subfigure}
+ \begin{subfigure}{\textwidth}
+ \centering
+	\includegraphics[width=1.0\textwidth]{images/No_Difference_Shigella_Profile_Plot}
+	\caption{}
+	 \end{subfigure}
 \caption{Dataset 5 : Mouse - Phosphorylation - 2mix-TMT time series.
-The overlap of differencial modified peptides between the PTM model with and without global protein level adjustment.
-Again more PTMs became insignificant after adjustment then became significant.}
-\label{fig:shig_venn_d}
+a) The overlap of differentially modified peptides between the PTM model with and without global protein level adjustment.
+Again more PTMs became insignificant after adjustment then became significant.
+Comparing the global profiling of protein $TTP$ with the modification of the protein at site $S178$.
+When looking at the summary of the modification and global protein it is clear the difference between conditions follow the same trend.
+Specifically, there is a positive adjustment in abundance when comparing WT\_Uninfect to WT\_Late in both the modification and global profiling run.
+This indicates the movement is driven by changes in global protein that is only accounted for in the model when adjusting for global protein abundance change.}
+\label{fig:shigella_plots}
 \end{figure}
@@
  \centering
-	\includegraphics[width=.95\textwidth]{images/No_Difference_Shigella_Profile_Plot}
-	\caption{Comparing the global profiling of protein $TTP\_MOUSE|P22893$ with the modification of the protein at site $S178$.
-The individual PSM features are shown in grey, while the feature summarization is shown in red.
-When looking at the summary of the modification and global protein it is clear the difference between conditions follow the same trend.
-Specifically, there is a positive adjustment in abundance when comparing WT\_Uninfect to WT\_Late in both the modification and global profiling run.
-This indicates the movement is driven by changes in global protein.}
- \end{subfigure}\vspace{5mm}
+	\includegraphics[height=.525\textwidth]{images/usp30_venn_diagramm}
+	\caption{}
+ \end{subfigure}
  \begin{subfigure}{\textwidth}
  \centering
-	\includegraphics[width=.95\textwidth]{images/Difference_Shigella_Profile_Plot}
-	\caption{Comparing the global profiling of protein $KI67\_MOUSE|E9PVX6$ with the modification of the protein at site $T215$.
-In this case the modification and global protein trend in different directions.
-Specifically, comparing WT\_Uninfect and WT\_Early there is a slightly positive change in abundance, however in the global profiling there was a negative change.
-In this case the profile plot indicates the effect of the modification is masked by the change in global protein abundance.
-Additionally this profile plot shows the large difference in available features between modifications and global protein.}
+	\includegraphics[height=.525\textwidth]{images/usp30_venn_diagramm_matching_only}
+	\caption{}
  \end{subfigure}
-\caption{Dataset 5 : Mouse - Phosphorylation - 2mix-TMT time series. Profile plots for PTMs P22893\_S178 and E9PVX6\_T215 in the Shigella flexneri experiment.}
-\label{fig:shigella_profile}
+ \caption{Dataset 6 : Human - Ubiquitination - Label-free no global profiling run.
+a) The overlap of differencial modified peptides for the PTM model with and without global protein level adjustment.
+Here many more PTMs became insignificant then became significant after adjustment.
+This is due to not having a global profiling run, resulting in a large lack of overlap between the modified peptides and unmodified proteins.
+b) Here we make the same comparison but only for modified peptides with a matching unmodified protein, so adjustment can be performed.
+In this case we see significantly less peptides become insignificant after adjustment.
+This highlights the need for a global profiling run if protein adjustment is going to performed.}
+\label{fig:usp30_ven_d}
 \end{figure}
@@
  \centering
-	\includegraphics[height=.5\textwidth]{images/usp30_venn_diagramm}
-	\caption{The overlap of differencial modified peptides for the PTM model with and without global protein level adjustment.
-Here many more PTMs became insignificant then became significant after adjustment.
-This is due to not having a global profiling run, resulting in a large lack of overlap between the modified peptides and unmodified proteins.}
+	\includegraphics[width=.75\textwidth]{images/same_var_power}
+	\caption{}
  \end{subfigure}\vspace{5mm}
@@
  \centering
-	\includegraphics[height=.5\textwidth]{images/usp30_venn_diagramm_matching_only}
-	\caption{Here we make the same comparison but only for modified peptides with a matching unmodified protein, so adjustment can be performed.
-In this case we see significantly less peptides become insignificant after adjustment.
-This highlights the need for a global profiling run if protein adjustment is going to performed.}
+	\includegraphics[width=.75\textwidth]{images/high_ptm_var_power}
+	\caption{}
  \end{subfigure}
- \caption{Dataset 6 : Human - Ubiquitination - Label-free no global profiling run. Overlap of significant PTMs before and after protein adjustment.}
-\label{fig:usp30_ven_d}
-\end{figure}
-
-\begin{figure}[ht]
-\centering
-\includegraphics[width=1\textwidth]{images/USP30_profile_plot.png}
-\caption{Dataset 6 : Human - Ubiquitination - Label-free no global profiling run.
-Comparing the global profiling of protein $P52209$ with the modification of the protein at site $K059$.
-The modification appears generally unchanged between conditions, whereas the global profiling run shows the CCCP and Combo conditions at a higher relative abundance compared to the Control and USP30\_OE.
-This indicates that the modification actually had a major effect when comparing CCCP and Combo to Control and USP30\_OE, which would have been missed without adjusting for global protein changes.}
-\label{fig:usp30_profile}
-\end{figure}
-
-\begin{figure}[h!]
-\centering
- \begin{subfigure}{\textwidth}
- \centering
-	\includegraphics[width=.725\textwidth]{images/same_var_power}
-	\caption{Here we can see the power of an experiment targeting PTMs with the same variance, .15, for the modified and unmodified peptides.
+ \caption{Power analysis of experiments with differing variances.
+a) The power of an experiment targeting PTMs with the same variance, .15, for the modified and unmodified peptides.
 Predictably when the replicates are high for both modified and unmodified peptides the power is much higher.
@@
 The interesting part of this chart is when the replicates are different between runs.
-With equal variance, it does not matter if the PTM replicates or protein replicates are higher.}
- \end{subfigure}\vspace{5mm}
- \begin{subfigure}{\textwidth}
- \centering
-	\includegraphics[width=.9\textwidth]{images/high_ptm_var_power}
-	\caption{In this chart the variance for the PTM is higher than the unmodified protein.
+With equal variance, it does not matter if the PTM replicates or protein replicates are higher.
+b) In this chart the variance for the PTM is higher than the unmodified protein.
 The PTM variance is .2, while the unmodified protein variance is .1.
@@
 When the replicates are not the same we can clearly see that having more replicates for the PTM leads to higher power.}
- \end{subfigure}
- \caption{Power analysis of experiments with differing variances.}
 \label{fig:power_sd_combo}

### 2022-02-01 supplementary/ptm_sm.tex
Fixed all plots

@@
 
+To compare performance across methods, we first calculated the number of true positives (TP), false positives (FP), true negatives (TN), and false negatives (FN).
+To determine TP and TN we adjusted the p-values for multiple testing and used an $\alpha=.05$ cutoff to determine differential sites.
+These results were then compared to the ground truth to determine TP and TN.
+We then used these results to calculate the false discovery rate (FDR = FP/(TP+FP)), the recall rate (TP/(TP + FN), and accuracy ((TP + TN) / (TP + TN + FP + FN)).
+Finally, the summary statistics were compared across methods to analyze method performance.
+
 In experiments of complex designs, multiple inter-related conditions are often compared together.
@@
 To highlight this distinction, multiple conditions of data were generated.
-Two simulations were generated based on the following parameters:
+While multiple conditions were generated, all comparisons were still made with the same fold change between conditions ($0.75$).
+This ensured that all peptides were differentially abundant by the same amount and the only difference in TP/FP/TN/FN calculations was using more conditions in the modeling.
 
@@
 The LC-MS dataset included only unmodified peptides.
-The spike-in peptides have a known fold change between conditions and are expected to be differential in most cases, whereas the human lysate modified peptides were treated as controls and are not expected to be differential in any comparison.
+The spike-in peptides have a known fold change between conditions and are expected to be differential in most cases.
+Additionally, background E coli lysate was used to normalize total protein levels prior to the enrichment and global protein profiling.
+The background lysate was treated as controls and are not expected to be differential in any comparison.
+
+\begin{figure}[h!]
+\centering
+\includegraphics[width=.8\textwidth]{sim_new/SpikeIn_Overview}
+\caption{An overview of the SpikeIn benchmark experimental design.
+The corresponding levels of SpikeIn peptide, human lysate, and background lysate for each mixture are shown.
+The SpikeIn peptide levels are adjusted for the human lysate levels.
+For example, the true SpikeIn abundance of mixture 1 is entirely accounted for by human lysate, whereas mix 2 has double the amount human lysate compared to SpikeIn abundance.
+When comparing mix 1 vs mix 2 we would expect to see a log fold change of -1.
+In this way we know the expected log fold change between mixture comparisons.
+\label{fig:spikein_overview}}
+\end{figure}
 
@@
 	\includegraphics[width=.9\textwidth]{sim_new/Difference_Shigella_Profile_Plot}
-	\caption{Comparing the global profiling of protein $KI67\_MOUSE|E9PVX6$ with the modification of the protein at site $T215$.
-In this case the modification and global protein trend in different directions.
-Specifically, comparing WT\_Uninfect and WT\_Early there is a slightly positive change in abundance, however in the global profiling there was a negative change.
-In this case the profile plot indicates the effect of the modification is masked by the change in global protein abundance.
-Additionally this profile plot shows the large difference in available features between modified peptides and global proteins.}
+	\caption{Comparing the global profiling of protein $KI67\_MOUSE|E9PVX6$ with the modification at site $T215$.
+In this case the modification and global protein appear to show very small to no differences between conditions, however after adjusting for change in global protein abundance, the modification is statistically significant.
+Additionally, this profile plot shows the large difference in available features between modified peptides and global proteins.}
  \end{subfigure}
@@
 	\includegraphics[width=.9\textwidth]{sim_new/Difference_Shigella_Volcano}
-	\caption{Volcano plots of WT\_Early vs WT\_Uninfect both before and after protein adjustment.
-The $KI67\_MOUSE|E9PVX6$ modification is highlighted.
-Before adjustment the modification has a small positive fold change and insignificant adjusted pvalue.
-After adjustment the fold change increases and the adjusted pvalue is now significant.}
+	\caption{The volcano plot of the WT\_Uninfect and WT\_Early comparison show the specifics of the adjustment.
+The modification looks to be flat, with a log fold change of $.185$, while the global profiling showed a small negative fold change of $-.619$.
+While both exhibit small changes, when combined we can see a log fold change of $.804$ and adjusted p-value of $.0156$.
+In this case we can see the strength of the proposed method, showing noteworthy modifications which would be missed otherwise.}
  \end{subfigure}

### 2022-02-01 code/Shigella_Analysis.R

@@ -468,6 +468,55 @@ v2 <- adj_model_df %>% ggplot() +
 
 grid.arrange(v1, v2, nrow = 1)
 
+## Venndiagram -----------------------------------------------------------------
+
+colors <- c("#E69F00", "#56B4E9")
+
+unadj_ptm_model <- pSTY_model$PTM.Model
+adj_ptm_model <- pSTY_model$ADJUSTED.Model
+
+unadj_ptm_model$Protein_Label <- paste(unadj_ptm_model$Protein, unadj_ptm_model$Label)
+adj_ptm_model$Protein_Label <- paste(adj_ptm_model$Protein, adj_ptm_model$Label)
+
+sig_unadj_ptm_model <- unadj_ptm_model %>% filter(Protein_Label %in% adj_ptm_model$Protein_Label)
+
+sig_unadj_ptm_model <- sig_unadj_ptm_model %>% filter(adj.pvalue < .05 & is.finite(log2FC))
+sig_adj_ptm_model <- adj_ptm_model %>% filter(adj.pvalue < .05 & is.finite(log2FC))
+
+
+venn.diagram(
+  x = list(sig_unadj_ptm_model$Protein_Label, sig_adj_ptm_model$Protein_Label),
+  category.names = c("Unadjusted" , "Adjusted"),
+  filename = "shig_venn_diagramm_time_series.png",
+  output=TRUE,
+  imagetype="png" ,
+  height = 1400,
+  width = 1400,
+  resolution = 100,
+  lwd = 2,
+  fill = colors,
+  main.fontface = "bold",
+  main.fontfamily="sans",
+  main.pos = c(.5,1),
+  fontface = "bold",
+  cat.fontface = "bold",
+  cex = 3,
+  cat.cex = 3,
+  cat.pos = c(-40, 30),
+  cat.dist = c(.037, .03),
+  main.cex = 2.8,
+  main = "Overlap between signficant adjusted and unadjusted PTMs"
+)
+
+combo <- adj_ptm_model %>% merge(unadj_ptm_model, by = "Protein_Label")
+
+combo %>% filter(log2FC.x < log2FC.y*1.09 & log2FC.x > log2FC.y*.91 &
+                   adj.pvalue.x >= .05 & adj.pvalue.y < .05)
+
+
+
+
+
 
 
 

### 2022-02-01 code/sample_size_analysis.R

@@ -50,7 +50,6 @@ for (x in seq_along(n_sample_list)){
   }
 }
 
-
 low_var_df <- rbindlist(bind_list)
 same_var_df <- rbindlist(bind_list)
 high_ptm_var_df <- rbindlist(bind_list)
@@ -66,13 +65,13 @@ p1 <- same_var_df %>%
   scale_colour_manual(values=cbPalette) +
   labs(title = "Power analysis with same variance") +
   theme_bw() +
-  theme(axis.text.x = element_text(size = 12),
-        axis.text.y = element_text(size = 12),
-        legend.text=element_text(size=12),
-        axis.title.y = element_text(size = 14),
-        axis.title.x = element_text(size = 14),
-        title = element_text(size = 16),
-        strip.text = element_text(size = 12))
+  theme(axis.text.x = element_text(size = 16),
+        axis.text.y = element_text(size = 16),
+        legend.text=element_text(size=16),
+        axis.title.y = element_text(size = 22),
+        axis.title.x = element_text(size = 22),
+        title = element_text(size = 22),
+        strip.text = element_text(size = 16))
 p1
 
 p2 <- high_ptm_var_df %>%
@@ -80,13 +79,13 @@ p2 <- high_ptm_var_df %>%
   scale_colour_manual(values=cbPalette) +
   labs(title = "Power analysis with high PTM variance") +
   theme_bw() +
-  theme(axis.text.x = element_text(size = 12),
-        axis.text.y = element_text(size = 12),
-        legend.text=element_text(size=12),
-        axis.title.y = element_text(size = 14),
-        axis.title.x = element_text(size = 14),
-        title = element_text(size = 16),
-        strip.text = element_text(size = 12))
+  theme(axis.text.x = element_text(size = 16),
+        axis.text.y = element_text(size = 16),
+        legend.text=element_text(size=16),
+        axis.title.y = element_text(size = 22),
+        axis.title.x = element_text(size = 22),
+        title = element_text(size = 22),
+        strip.text = element_text(size = 16))
 p2
 
 png("../supplementary/sim_new/same_var_power.png", width = 750, height = 500)

### 2022-02-12 code/MSstatsPTM_Simulation_Analysis.Rmd
Updates to text

--- 2022-01-28 code/MSstatsPTM_Simulation_Analysis.Rmd
@@
 models$plot_sd = paste0("SD ", models$sd)
+models$Replicates = factor(models$rep)
 
-png("../supplementary/sim_new/sim1_FDR_all_models.png", width = 750, height = 500)
-
+png("../main/images/sim1_FDR_all_models.png", width = 750, height = 500)
 models %>% mutate(Model = factor(Model, levels=c("Proposed", "Proposed no adjustment", "Limma", 
                                                "Limma no adjustment", "T-Test", "T-Test no adjustment")),
-                  Replicates = as.factor(rep)) %>%
-  ggplot() + geom_point(aes(x = Model, y = fpr, size = Replicates, color = Model)) + 
+                  ) %>%
+  filter(conditions == 4 & sd == .3 & rep == 5) %>% 
+  ggplot() + geom_col(aes(x = Model, y = fpr, fill = Model)) + #, size = Replicates
   scale_colour_manual(values=cbPalette_first) +
-  scale_size_manual(values = c(3,5,7,9)) + 
-  facet_grid(vars(plot_sd), vars(plot_conditions)) + 
+  # scale_size_manual(values = c(3,5,7,9)) +
+  # facet_grid(vars(plot_sd), vars(plot_conditions)) + 
   theme_bw() +
-  theme(axis.text.x = element_text(angle = 60, hjust=1, size = 12), 
-         axis.text.y = element_text(size = 14), 
-         legend.text=element_text(size=14),
+  ylim(0.0, 0.4) + 
+  theme(axis.text.x = element_text(angle = 45, hjust=1, vjust = 1, size = 16), 
+         axis.text.y = element_text(size = 16), 
+         legend.text=element_text(size=16),
          axis.title.y = element_text(size = 22),
@@
 dev.off()
+
+## Caluclate #s of IQR
+setDT(boxplot_data)[,list(IQR=IQR(log2FC, na.rm = TRUE)),by=list(Model)]
+
+(0.3787906 - 0.4228233) / 0.4228233
+(0.3787906 - 0.4843184) / 0.4843184
+
 ```

### 2022-02-12 code/Spike-in_Analysis.Rmd
Updates to text

@@
 dev.off()
+
+spike
+setDT(spike)[,list(IQR=IQR(log2FC, na.rm = TRUE)),by=list(Model, Label, protadj)] %>% 
+  filter(Label %in% c('mix2-mix3', 'mix1-mix3', 'mix1-mix4', 'mix2-mix4') & 
+           protadj == 'protein adjustment') %>% 
+  group_by(Model) %>% summarize(mean(IQR))
+
+(0.7557581 - 1.1954719) / 1.1954719
+(0.7557581 - 1.1277441) / 1.1277441
+
 ```

### 2022-02-12 main/MSstatsPTM_Main.tex
Updates to text

@@
 
-Generally, there are two main biological questions that are relevant in PTM analysis.
-First, some researchers run mass-spectrometry based experiments targeting PTMs when looking for biomarkers\cite{Petushkova_2017}.
+There are two main biological questions that may be relevant in PTM analysis.
+First, some researchers run mass-spectrometry based experiments targeting PTMs when looking for biomarkers.\cite{Petushkova_2017}.
 In this case the researcher may not necessarily want to correct for confounding with the total protein.
@@
 These two questions are both important areas of study and require different statistical methods to correctly address.
-In this paper we will mainly address a statistical solution to the second question, which requires more rigorous methods as we must address confounding with the total protein.
+In this paper we will mainly address a statistical solution to the second question, which requires more rigorous methods that account for confounding with the total protein.
 
@@
 Unmodified peptides from Human Lysate were used as the estimate of global protein abundance changes.
-All comparisons with respect to human lysate serve as a null model, there was no change in global protein nor PTM abundance between conditions.
-Additionally, E coli Lysate was used to normalize total protein levels prior to enrichment or global protein profiling.
+Background E coli Lysate was used to normalize total protein levels prior to enrichment or global protein profiling.
+The background lysate was treated as controls and are not expected to be differential in any comparison.
 The spike-in peptides were mixed with human lysate to create four mixture conditions.
@@
 The four mixtures were compared with known fold changes for the spike-in peptides.
+
+More information on the experimental design of the benchmark experiment can be found in Supplementary Sec. \ref{supp-sec:benchmark}.
 
@@
 
-The proposed approach first leverages the large amount of work done to create MSstats~\cite{Choi:2014}.
+The proposed approach leverages the large amount of work done to create $MSstats$~\cite{Choi:2014}.
 It takes as input a list of log-transformed intensities of spectral features, identified and quantified across LC-MS runs.
@@
 
-Which amounts to running a the process without accounting for protein adjustment, as seen in MSstats and MSstatsTMT \cite{Choi:2014} \cite{Huang:2020}.
+Which amounts to running the methods seen in MSstats and MSstatsTMT on peptide level data \cite{Choi:2014} \cite{Huang:2020}.
 
@@
 
-The code is available on Bioconductor, http://www.bioconductor.org/packages/release/bioc/html/MSstatsPTM.html, and Github, https://github.com/Vitek-Lab/MSstatsPTM.
+The code is available on Bioconductor, \url{http://www.bioconductor.org/packages/release/bioc/html/MSstatsPTM.html}, and Github, \url{https://github.com/Vitek-Lab/MSstatsPTM}.
 
@@
 For the experiments where ground truth is known, we calculated the number of true positives (TP), false positives (FP), true negatives (TN), and false negatives (FN).
+True positives are modified peptides that are differential after accounting for changes in global protein abundance.
+True negatives were modified peptides that were not differentially abundant after accounting for global protein abundance.
 To determine positives and negatives we adjusted the p-values for multiple testing and used an $\alpha=.05$ cutoff to determine differential sites.
 These results were then compared to the ground truth to determine TP and TN.
-We then used these results to calculate the false discovery rate (FDR = FP/(TP+FP)), the recall rate (TP/(TP + FN), and accuracy ((TP + TN) / (TP + TN + FP + FN)).
+We used these results to calculate the false discovery rate (FDR = FP/(TP+FP)), the recall rate (TP/(TP + FN), and accuracy ((TP + TN) / (TP + TN + FP + FN)).
 Finally, the summary statistics were compared across methods to analyze method performance.
 
-For biological experiments where ground truth was not known, we again adjusted the p-values for multiple testing and used an $\alpha=.05$ cutoff to determine differential sites.
-Comparisons were made before and after adjusting for changes in unmodified protein abundance and the results are presented in the form of venn diagrams and profile plots.
-Each biological dataset presents a different experimental scenario, showing the applicability of the proposed method.
+For biological experiments where ground truth was not known, we adjusted the p-values for multiple testing and used an $\alpha=.05$ cutoff to determine differential sites.
+Comparisons were made before and after adjusting for changes in unmodified protein abundance and differences between the comparisons were noted.
+Each biological dataset presents a different experimental scenario, showing the applicability of the proposed method on a variety of experimental designs.
 All datasets in this paper were evaluated using the $MSstatsPTM$ package.
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsubsection*{The proposed approach corrects for high FDR in computer simulations}
 
-The proposed approach was evaluated against two-sample t-test and Limma. 
-In Figure \ref{fig:fdr_all_models} we see that not accounting for changes in unmodified peptide level results in a high false discovery rate.
+The proposed approach was evaluated against two-sample t-test and Limma in computer simulations. 
+In Figure \ref{fig:computer_sim} we see that not accounting for changes in unmodified peptide level results in a high false discovery rate.
 When adjusting for changes in the unmodified peptide, all methods performed similarly in terms of FDR.
-Recall and accuracy of the proposed approach and Limma performed similarly, with Limma performing slightly better with low replicates, while t-test lagged behind.
-Two-sample t-test only uses data within the groups of interest while ignoring the remainder of the data.
-Consequently, it gave similar performance across cases with different number of conditions.
-In contrast, the proposed approach and Limma leverage all available information, which resulted in improved power with increased number of conditions.
+In a clean simulation, recall and accuracy of the proposed approach and Limma performed similarly, with Limma performing slightly better with low replicates.
+In contrast, the performance of the t-test method lagged behind the others.
+Two-sample t-test only uses data within the groups of interest while ignoring the remainder of the data, leading to similar performance across all simulations.
+In contrast, the proposed approach and Limma leverage all available information, which resulted in improved power in simulations with more conditions.
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsubsection*{The proposed approach outperforms existing methods in simulations with missing values and low replicates}
 
-In this simulation, changes in unmodified peptide level still needed to be accounted for to control the FDR.
-Once controlling for changes in unmodified peptide, it is clear that the proposed method performed strongest across all methods, as seen in Figure \ref{fig:sim_accuracy}.
+In this simulation the proposed method outperformed t-test and limma when real world experimental conditions were introduced.
+Changes in unmodified peptide level still needed to be accounted for to control the FDR.
+Once controlling for changes in the unmodified peptide, the proposed method out performed the other methods, as seen in Figure \ref{fig:computer_sim}.
 The proposed method calibrated model accuracy and recall rate well, even when the number of replicates were low.
-Finally, when comparing the fold change estimation across all peptides, the proposed method showed a tighter distribution of estimated fold changes around the true fold change.
-In comparison, Limma and t-test showed wider distributions.
-This fold change comparison can be seen in Figure \ref{fig:sim_boxplot}.
+Additionally, when comparing the fold change estimation across all peptides, the proposed method showed a tighter distribution of estimated fold changes around the true fold change than t-test and Limma.
+Specifically, the inner quartile range of the estimated fold change for the proposed method was on average 10.4\% smaller than t-test and 21.8\% smaller than Limma.
+This fold change comparison can be seen in Supplementary Figure \ref{supp-fig:fc_boxplot}.
 
@@
 
-%\paragraph*{Computer simulation: batch effects.}
-
-%PTM experiments often create batches during data acquisition.
-The proposed approach can be naturally extended to handle such scenarios by modeling various forms of batch effects, as a t-test is not directly applicable to a problem with batches of data.
-Two commonly used ad-hoc approaches are a) t-test (no batch): ignoring batch effects when applying t-test, and b) t-test (most significant batch): applying t-test in each batch and drawing conclusions based on the most significant batch.\todo{Find source for batch analysis} Although simple, these ad-hoc methods lack statistical justification.
-We characterized their statistical properties under various forms of batch effects in Supplementary \todo{link to Sec.
-S4.2}.
-Two batches of data were generated, where various forms of batch effects were simulated, including difference in signal intensities across batches, difference in variability across batches, and interaction between batch and condition (i.e., change between conditions affected by batch).
-
-%The major results are summarized in \todo{figref Fig.
-5} with additional details provided in Supplementary \todo{link to Sec.
-S4.2}.
-In terms of estimation of fold change, the most statistically significant batch with t-test was highly variable and frequently biased (Fig.
-5a).
-In the detection of systematic changes in PTM abundance, it also frequently resulted in higher false positive rates than the expected value (Fig.
-5b).
-As shown Fig.
-5c, the proposed approach improved power with small sample sizes in almost all the scenarios, under various forms of batch effects including difference in intensity level and difference in variability.
-Using t-test while ignoring batch effect gave similar performance in special cases with no difference in intensity level between batches (Supplementary Fig.
-\todo{find and link this section SXX}), but its performance dramatically decreased in general cases (Supplementary \todo{link to Figs.
-S8-S9}).
-The negative impact was only partially reduced by increasing the sample size to 5.
-The proposed approach gave consistently improved performance compared with other methods by properly characterizing batch effects and leveraging all available information.
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsubsection*{In a Label-free benchmark experiment the proposed approach corrects for bias and outperforms existing methods}
+
+In this experiment all models incorrectly estimated the fold change of the modified spike-in peptides before adjusting for changes in global protein abundance.
+After adjustment, the spike-in peptides' fold change was in line with expectation.
+This can be seen in Figure \ref{fig:spikein_boxplot}.
+While the median fold change for the spike-in peptides was generally in line with the true fold change after adjustment, we can see that the distribution of fold changes for the peptides was visibly wider after adjustment.
+However, the proposed method a tighter distribution around the true log fold change.
+Comparing the inner quartile range (IQR) of the spike-in peptide's log fold change, the proposed method's IQR was 36.78\% smaller than Limma's and 32.98\% smaller than t-test's.
+This means that the proposed method's fold change estimate was more consistently correct, when compared to Limma and t-test.
 
@@
 After adjustment the results improve and the number of false positives decrease significantly.
+Specifically, for the proposed method, the number of false positives goes from 20.88\% to 1.84\% of all true negatives after adjustment is applied.
+Using both the proposed method and Limma, the red labeled spike-in peptides are more inline with expected fold change after the adjustment is applied.
+However, when using Limma we can see that the majority of the differential spike-in peptides are not showing a significant adjusted pvalue.
+In this case, using Limma would have resulted in missing the majority of differentially modified peptides.
 
-In Figure \ref{fig:spikein_limma_volcano} we can see the results of the same experiment modeled using Limma with protein adjustment.
-Again the red labeled spike-in peptides are more inline with expected fold change after adjustment is applied.
-However, using Limma we can see that the majority of the differential spike in peptides are not showing a significant adjusted pvalue.
-In this case, using Limma would have resulted in missing the majority of differential modified peptides.
-In terms of false positives, the results are very similar to the proposed method, with many false positives before adjustment and much fewer afterward.
+Further results of this experiment are provided in Supplementary Sec. \ref{supp-sec:benchmark}.
 
-Further results of this experiment, including a comparison to t-test and fold change distribution of each method before and after adjustment, are provided in Supplementary Sec.
-\ref{supp-sec:benchmark}.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\subsection*{Dilution experiment [What was this experiment?]}
-
-%We evaluated our approach using a custom designed dilution benchmark experiment where two sets of liver samples from five Atg16L1 deficient mice (one with KGG enrichment followed by trypsin digestion, one with trypsin digestion alone) were prepared.
-Following a randomized order, each sample was injected at three different concentrations 4ul, 2ul, and 1ul.
-The experiment resulted in 15 KGG and 15 global profiling runs.
-To correct for run-to-run variation, the AQUA peptide mixture was spiked into each sample.
-There are 4696 ubiquitinated proteins identified in this dataset.
-Among these, 3173 have corresponding measurements from global data.
-%To simulate the effects of changes in protein abundance, we created three groups of data by matching one concentration for the KGG data with one for the global data: K1P1, K2P2, and K4P4.
-Systematic changes in the abundance of ubiquitinated sites between the groups are present in the KGG data.
-These changes are considered as artifacts driven by the changes in the abundance of their corresponding proteins.
-Using the comparison of K2P2 vs.
-K1P1 as example, there are 19355 possible comparisons with KGG data, and 3860 changes are detected.
-Out of the 19355 sites, 15036 have corresponding measurements from global data and are eligible for protein-level correction.
-The correction significantly reduced the number of false positives to 79.
-Similar observations were made in the other two comparisons K4P4 vs.
-K2P2 and K4P4 vs.
-K1P1, as shown in Table XX.
-\todo{Create or find this table}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsubsection*{MSstatsPTM corrects for bias in group comparison TMT experiment}
 
-The results of this experiment are summarized in Figures \ref{fig:ipah_venn_diagramm} and \ref{fig:ipah_profile}.
+The results of this experiment are summarized in Figure \ref{fig:data4_plots}.
 The experiment was modeled as a group comparison.
-In Figure \ref{fig:ipah_venn_diagramm} the number of significant modified peptides before and after adjustment is shown.
+In Figure \ref{fig:data4_venn_diagram} the number of significant modified peptides before and after adjustment is shown.
 We can see that more peptides become insignificant after adjustment than become significant.
@@
 In contrast, for peptides that became significant after adjustment, their true abundance change was masked by underlying changes in the unmodified protein.
-Both of these issues are correct in adjustment, and the true abundance change is shown.
+Both of these issues are corrected in adjustment, and the true abundance change is shown.
 An additional question that must be addressed is if the decrease in significant peptides is due to the increased variance that comes from adjustment.
@@
 
-In Figure \ref{fig:ipah_profile}, looking at modified peptide $GSDMD\_HUMAN|P57764$ we can see the advantage of the proposed method.
+In Figure \ref{fig:data4_profile_plot}, looking at the modification of protein $GSDMD$ at $K62$ we can see the advantage of the proposed method.
 The modified peptide shows a flat abundance change between the infected 1 hour, 4 hour, and 6 hour conditions.
@@
 Looking at the Dox4hr vs Dox1hr conditions and modeling the modified peptide without adjusting for changes in the global profiling, the fold change was $-.501$ and the adjusted pvalue was insignificant at $.0644$.
-After adjusting for changes in the global protein abundance, the fold change is much higher, $2.79$, and the adjusted pvalue became very significant, $5.25e-8$.
+After adjusting for changes in the global protein abundance, the fold change is much higher, $2.79$, and the adjusted pvalue became significant, $5.25e-8$.
 In this case the effect of the modified peptide is strongly confounded with changes in the global protein.
@@
 
-The results of this experiment are summarized in Figures \ref{fig:shig_venn_d} and \ref{fig:shigella_profile}.
-The experiment was modeled as a time series, with the same bioreplicate measured at each time point.
-In Figure \ref{fig:shig_venn_d} the number of significant modified peptides before and after adjustment is shown.
+The results of this experiment are summarized in Figure \ref{fig:data5_plots}.
+The experiment was modeled as a time series, with the same biological replicate measured at each time point.
+Modeling the experiment as a time series lowered the model variance as compared to modeling the experiment as a group comparison.
+In Figure \ref{fig:data5_venn_diagram} the number of significant modified peptides before and after adjustment is shown.
 Again more PTMs became insignificant after adjustment than became significant.
@@
 When this test was applied on this experiment, 548 peptides became insignificant due to an increase in variance.
-This is still a very small portion (3.4\%) of the peptides that became insignificant after adjustment.
+This is still a small portion (3.4\%) of the peptides that became insignificant after adjustment.
 Thus we can conclude that the drop off in significant peptides was mainly due to changes in global protein abundance.
 
-In Figure \ref{fig:shigella_profile} the profile plot of two specific modified peptides, $TTP\_MOUSE|P22893\_S178$ and $KI67\_MOUSE|E9PVX6\_T215$, show the power of the proposed method.
-The profile plot of modified peptide $TTP\_MOUSE|P22893\_S178$ shows a large positive log fold change of $2.9$ between the WT\_Late and WT\_Uninfected conditions.
+In Figure \ref{fig:data5_profile_plot} the profile plot of protein $TTP$ modified at site $S178$ shows the power of the proposed method.
+The profile plot shows a large positive log fold change of $2.9$ between the WT\_Late and WT\_Uninfected conditions.
 However, the global profiling run shows a similar log fold change of $2.014$ between the same conditions.
@@
 When adjusting for the global protein the modified peptide's adjusted pvalue becomes insignificant, going from $.0009$ to $.248$.
-In contrast $KI67\_MOUSE|E9PVX6\_T215$ originally shows a small positive log fold change of $.185$ between conditions WT\_Early and WT\_Uninfected.
-This is contrast with a negative fold change in the global profiling run of $-.619$.
-In adjusting the modified peptide for changes in the global protein, the log fold change becomes $.804$ and the adjusted pvalue goes from $.338$ to $.0156$.
+Correcting for changes in global protein abundance allowed us to see the true impact of the modification at $S178$ which would have otherwise been challenging to perceive.
+
+Analysis of other modifications in this experiment can be seen in Supplementary Sec. \ref{supp-sec:shigella}.
 
@@
 As discussed previously, there was no unmodified global profiling run performed in this experiment.
-Once identification and quantification of the Ubiquitinated profiling was performed, peptides which were unmodified were extracted and used in place of a global profiling run.
+Once identification and quantification of the Ubiquitination profiling was performed, peptides which were unmodified were extracted and used in place of a global profiling run.
 This resulted in a significant lack of overlap between modified and unmodified peptides.
@@
 Of the 10,799 modified peptides identified, only 4526 had a corresponding unmodified protein and could be adjusted.
-Additionally, not having a separate global profiling run resulted in very low feature counts for the unmodified protein model.
+Additionally, not having a separate global profiling run resulted in low feature counts for the unmodified protein model.
 
-The results of this experiment are summarized in Figures \ref{fig:usp30_ven_d} and \ref{fig:usp30_profile}.
-In \ref{fig:usp30_ven_d} the number of significant modified peptides before and after protein adjustment is shown.
-In the top plot we can see there are a very large number of peptides that become insignificant after adjustment.
+The results of this experiment are summarized in Figure \ref{fig:data6_plots}.
+In \ref{fig:data6_vd1} the number of significant modified peptides before and after protein adjustment is shown.
+In the top plot we can see there are a large number of peptides that become insignificant after adjustment.
 This is mainly due to the lack of overlap between modified and unmodified peptides, stemming from not having a global profiling run.
@@
 To check this we look at modified peptides whose adjusted fold change is within 10\% of their unadjusted fold change and they become insignificant after adjustment.
-In this experiment there are only 25 peptides that meet this criteria, a very small percentage of all modified peptides.
+In this experiment there are only 25 peptides that meet this criteria, a small percentage of all modified peptides.
 
-In Figure \ref{fig:usp30_profile} the profile plot of peptide $P52209\_K059$ is shown.
+Analysis of modifications before and after protein adjustment can be seen in Supplementary Sec. \ref{supp-sec:usp30}.
+
+%In Figure \ref{fig:usp30_profile} the profile plot of peptide $P52209\_K059$ is shown.
 The abundance between conditions appears flat in the modified peptide profile plot, however this is contrast with a large negative change in the corresponding global protein.
@@
 We mimicked these scenarios and analyzed the power of experiments when the PTM variance was higher than the protein and when they were equal.
-When the PTM and protein were the same we chose a variance of .15, whereas when the PTM was higher than the protein we chose a PTM variance of .2 and a protein variance of .1.
+When the PTM and protein were the same we chose a variance of $.15$, whereas when the PTM was higher than the protein we chose a PTM variance of $.2$ and a protein variance of $.1$.
 
-The results of the power and sample size analysis can be seen in Figure \ref{fig:power_sd_combo}.
-We can see that when the variance and replicates were equal, higher replicates predictably lead to higher power.
+The results of the power and sample size analysis can be seen in Supplementary Figure \ref{supp-fig:power_sd_combo}.
+When the variance and replicates were equal, higher replicates predictably lead to higher power.
 In cases where the replicates were unbalanced, but the variance was still the same, it did not matter if there were more replicates in the modified or unmodified runs.
@@
 One potential solution to this is to measure the abundance of peptides with one modification and use this to adjust the peptide with multiple sites to remove the convolution.
-However, this method would most likely run into major challenges due to sparsity of features for modified peptides with both a single and multiple modification sites.
+However, this method would likely run into challenges due to sparsity of features for modified peptides with both a single and multiple modification sites.
 A more complex approach to addressing this problem is most likely necessary.
@@
 Overall, the proposed approach balances accuracy and practicality, and enables the analysis of complex experiments in high throughput.
-Future work is to carry out the inference and testing for not only the relative change of PTM abundance, but also the fraction of the protein that is modified at the particular site (site occupancy, or stoichiometry).
-We are also interested in characterizing the interplay of PTMs at multiple sites.
+Future work is to carry out the inference and testing for not only the relative change of PTM abundance, but also the fraction of the protein that is modified at the particular site (site occupancy, or stoichiometry), and attempt to deconvolute the effect of individual PTMs in peptides with multiple modifications.
+
+\section{Data Availability}
+
+Shigella: https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=4878d777c6b34cf8aaf8477e93140c4d
 
@@
 \centering
-\includegraphics[scale=.65]{images/fig3.png}
+\includegraphics[scale=.6]{images/fig3.png}
 \caption{
@@
 \caption{}
-\label{fig:spikein_prop_volcano}
 \end{subfigure}
@@
 \begin{subfigure}[c]{0.825\linewidth}
+\includegraphics[width=1\textwidth]{images/spike_in_fc.png}
+\caption{}
+\label{fig:spikein_boxplot}
+\end{subfigure}
+\begin{subfigure}[c]{0.825\linewidth}
 \includegraphics[width=1\textwidth]{images/spike_in_volcano.png}
@@
 \end{subfigure}
-\begin{subfigure}[c]{0.825\linewidth}
-\includegraphics[width=1\textwidth]{images/spike_in_fc.png}
-\caption{}
-\label{fig:spikein_limma_volcano}
-\end{subfigure}
 \caption{Dataset 3 : SpikeIn benchmark - Ubiquitination - Label-free.
-a) Using the proposed method the spike in peptides (colored red) do not follow the expected log fold change before adjustment.
-After adjustment the spike in peptides are more in line with expectation.
-Using Limma the spike in peptides again follow the expected log fold change after adjusting for changes in protein level.
-However, while the fold change is more accurate, the majority of spike in peptides do not have a significant adjusted pvalue.
-b) say something about this one.}
+a) Before adjustment all models show the fold change of the spike-in peptides is systematically different from the expected fold change.
+After adjustment, this systemic difference is removed, however the inner quartile range of the Limma and t-test models is wider than the proposed method.
+b) The spike in peptides (colored red) do not follow the expected log fold change before adjustment.
+However, after adjustment, the spike in peptides are more in line with expectation.
+Using Limma the spike in peptides follow the expected log fold change after adjustment, however the majority of spike in peptides do not have a significant adjusted pvalue.}
 \label{fig:spikein_volcano}
@@
 \caption{}
+\label{fig:data5_venn_diagram}
  \end{subfigure}
@@
 	\caption{}
+	\label{fig:data5_profile_plot}
 	 \end{subfigure}
@@
 This indicates the movement is driven by changes in global protein that is only accounted for in the model when adjusting for global protein abundance change.}
-\label{fig:shigella_plots}
+\label{fig:data5_plots}
 \end{figure}
@@
 	\caption{}
+	\label{fig:data6_vd1}
  \end{subfigure}
@@
 	\caption{}
+	\label{fig:data6_vd2}
  \end{subfigure}
@@
 This highlights the need for a global profiling run if protein adjustment is going to performed.}
-\label{fig:usp30_ven_d}
+\label{fig:data6_plots}
 \end{figure}
 
-\begin{figure}[h!]
-\centering
- \begin{subfigure}{\textwidth}
- \centering
-	\includegraphics[width=.75\textwidth]{images/same_var_power}
-	\caption{}
- \end{subfigure}\vspace{5mm}
- \begin{subfigure}{\textwidth}
- \centering
-	\includegraphics[width=.75\textwidth]{images/high_ptm_var_power}
-	\caption{}
- \end{subfigure}
- \caption{Power analysis of experiments with differing variances.
+%\begin{figure}[h!]
+%\centering
+% \begin{subfigure}{\textwidth}
+% \centering
+%	\includegraphics[width=.75\textwidth]{images/same_var_power}
+%	\caption{}
+% \end{subfigure}\vspace{5mm}
+% \begin{subfigure}{\textwidth}
+% \centering
+%	\includegraphics[width=.75\textwidth]{images/high_ptm_var_power}
+%	\caption{}
+% \end{subfigure}
+% \caption{Power analysis of experiments with differing variances.
 a) The power of an experiment targeting PTMs with the same variance, .15, for the modified and unmodified peptides.
@@
 When the replicates are not the same we can clearly see that having more replicates for the PTM leads to higher power.}
-\label{fig:power_sd_combo}
-\end{figure}
+%\label{fig:power_sd_combo}
+%\end{figure}
 

### 2022-02-12 supplementary/ptm_sm.tex
Updates to text

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Experiments with known ground truth}
+\section{Experimental datasets in this manuscript}
+
+%%%% Summary of controlled mixtures
+\begin{table}
+\begin{tiny}
+\begin{tabular}{|c|c|ccccc|c|}
+\hline
+%%%% colnames of table
+&  & \multicolumn{5}{c|}{Experimental design} & Data availability \\
+& Dataset & \multirow{2}{*}{Comparison} & No. of & No. of Bio.& No. of Tech.& \multirow{2}{*}{Details in}   & (MassIVE.quant  \\
+&  & & Conditions & Replicates & Replicates &  & or R package) \\
+\hline
+\hline
+% &&&&& \\[-0.05in]
+%%%% DDA:ControlledMix
+\multirow{3}{*}{\rotatebox[origin=c]{90}{Known}}  \multirow{3}{*}{\rotatebox[origin=c]{90}{ Ground}} \multirow{3}{*}{\rotatebox[origin=c]{90}{ Truth}} & DDA:ControlledMix \cite{Chiva2014} & Group & 5 & 1 & 3 & Suppl.
+Sec.
+& \\
+ &&&&&&& \\[-0.05in]
+%\hline 
+%&&&&&&&& \\[-0.05in]
+% \cline{2-9}
+% &&&&&&&& \\[-0.05in]
+%%% DDA iPRG2015
+& DDA:Choi2017 \cite{Choi:2017jan}  & Group & 4 & 1 & 3 &  & \\
+ &&&&&&& \\[-0.05in]
+%\hline 
+% &&&&&&& \\[-0.05in]
+%%% DDA Cox2014
+& DDA:Cox2014 \cite{Cox:2014fi} & Group & 2 & 1 & 4 &  & \\
+ &&&&&&& \\[-0.05in]
+\hline
+\multicolumn{8}{c}{ } \\ [0.02in]
+\hline 
+ &&&&&&& \\[-0.05in]
+%%%%%%%%%%%%%%%%%%
+%%%% biological study
+%%%%%%%%%%%%%%%%%%
+%% DDA Meierhofer
+\multirow{3}{*}{\rotatebox[origin=c]{90}{Biological}} \multirow{3}{*}{\rotatebox[origin=c]{90}{investigation}} & DDA:Meierhofer2016 \cite{Meierhofer2016} & Paired & 2 & 6 & 2 &  &  \\
+&&&&&&& \\[-0.05in]
+% \cline{2-9}
+% &&&&&&& \\[-0.05in]
+%%%% SRM Picotti 2009
+& SRM:Picotti2009  \cite{Picotti09} & Time course & 10 & 3 & 1 &  & \\
+&&&&&&& \\[-0.05in]
+% \cline{2-9}
+% &&&&&&& \\[-0.05in]
+%%%% SRM Surinova 2015, training
+& SRM:Surinova2015:Training \cite{Surinova:2015jn} & Group & 2 & 100 & 1 & &\ [0.02in]
+\end{tabular}\\
+\end{tiny}
+\caption{
+}
+\label{tab:dataDescriptionSpikein}
+\end{table}
+
+\subsection{Experiments with known ground truth}
 \label{sec:sim}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Computer simulation}
+\subsubsection{Computer simulation}
 
@@
 
-To compare performance across methods, we first calculated the number of true positives (TP), false positives (FP), true negatives (TN), and false negatives (FN).
-To determine TP and TN we adjusted the p-values for multiple testing and used an $\alpha=.05$ cutoff to determine differential sites.
-These results were then compared to the ground truth to determine TP and TN.
-We then used these results to calculate the false discovery rate (FDR = FP/(TP+FP)), the recall rate (TP/(TP + FN), and accuracy ((TP + TN) / (TP + TN + FP + FN)).
-Finally, the summary statistics were compared across methods to analyze method performance.
-
 In experiments of complex designs, multiple inter-related conditions are often compared together.
@@
 \subsubsection{Dataset 1 : Computer simulation 1 - Label-free}
-
+\label{dataset1}
 In the first simulation an experiment with many features per PTM and unmodified protein was created.
@@
 \centering
- \begin{subfigure}{\textwidth}
-  \centering
 	\includegraphics[width=.8\textwidth]{sim_new/sim3_Recall}
-	\caption{The advantage of using the proposed approach is apparent when looking at simulation 2, which includes limited observations and the presence of missing values.
+	\caption{}
+\caption{Recall results of Simulation 2.
+The advantage of using the proposed approach is apparent when looking at simulation 2, which includes limited observations and the presence of missing values.
 In the case of recall the proposed method performs stronger than Limma and t-test in nearly every model.
@@
 Again the lowest performing method was $t$-test.}
- \end{subfigure}
- \begin{subfigure}{\textwidth}
-  \centering
-	\includegraphics[width=.8\textwidth]{sim_new/sim3_Accuracy}
-	\caption{The overall accuracy plots show similar results to recall.
-The proposed method performed the strongest across all methods, even when replicates were low.
-Limma shows strong performance in a clean experiment, however when real world data problems are introduced it is clear the proposed method is more robust.}
- \end{subfigure}
-\caption{Recall and Accuracy results of Simulation 2.}
 \label{fig:sim2_recall}
@@
 Both Limma and $t$-test showed a wider range around the fold change.
+The inner quartile range of the Proposed method was on average 10.4\% smaller than t-test and 21.8\% smaller than Limma.
 In this case the proposed method showed a stronger performance in correctly estimating the fold change for all peptides.}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Dataset 3 : SpikeIn benchmark - Ubiquitination - Label-free}
+\subsubsection{Dataset 3 : SpikeIn benchmark - Ubiquitination - Label-free}
 \label{sec:benchmark}
@@
 \includegraphics[width=.8\textwidth]{sim_new/SpikeIn_Overview}
-\caption{An overview of the SpikeIn benchmark experimental design.
-The corresponding levels of SpikeIn peptide, human lysate, and background lysate for each mixture are shown.
-The SpikeIn peptide levels are adjusted for the human lysate levels.
-For example, the true SpikeIn abundance of mixture 1 is entirely accounted for by human lysate, whereas mix 2 has double the amount human lysate compared to SpikeIn abundance.
+\caption{An overview of the spike-in benchmark experimental design.
+The corresponding levels of spike-in peptide, human lysate, and background lysate for each mixture are shown.
+The spike-in peptide levels are adjusted for the human lysate levels.
+For example, the true spike-in abundance of mixture 1 is entirely accounted for by human lysate, whereas mix 2 has double the amount human lysate compared to spike-in abundance.
 When comparing mix 1 vs mix 2 we would expect to see a log fold change of -1.
@@
 \caption{The modeling results of the proposed method both before and after adjustment.
-The spike in peptides are colored red and the background peptides are colored grey.
+The spike-in peptides are colored red and the background peptides are colored grey.
 All grey peptides are expected to be insignificant.
-Using the proposed method to model the benchmark experiment, the spike in peptides (colored red) do not follow the expected log fold change before adjustment.
-After adjusting for changes in overall protein abundance the spike in peptides are more in line with expectation.
+Using the proposed method to model the benchmark experiment, the spike-in peptides (colored red) do not follow the expected log fold change before adjustment.
+After adjusting for changes in overall protein abundance the spike-in peptides are more in line with expectation.
 Additionally the background grey colored peptides show many false positives before adjustment.
@@
 \includegraphics[width=.85\textwidth]{sim_new/spike_in_limma_volcano}
-\caption{When modeling the experiment with the Limma method, the spike in peptides again follow the expected log fold change better after adjusting for changes in protein level.
-However, while the fold change is much more accurate, the majority of spike in peptides do not have a significant adjusted pvalue.
+\caption{When modeling the experiment with the Limma method, the spike-in peptides again follow the expected log fold change better after adjusting for changes in protein level.
+However, while the fold change is much more accurate, the majority of spike-in peptides do not have a significant adjusted pvalue.
 In this case, the known differential peptides are missed by the model.
@@
 \caption{Using the two sample $t$-test, none of the comparisons either before or after adjustment show any significant peptides.
-With that being said, the fold change of the spike in peptides is much closer to expectation after adjusting for global protein abundance.
+With that being said, the fold change of the spike-in peptides is much closer to expectation after adjusting for global protein abundance.
 \label{fig:spike_volcano_ttest}}
@@
 In terms of accuracy and specificity the three methods are close, with Limma and $t$-test showing slightly higher values.
-Accuracy and specificity are dominated by the large number of true negatives (background peptides) compared to the true positives (spike in peptides).
-In terms of recall, the proposed approach far out performed the other two methods, showing that it correctly labeled the most spike in peptides.
+Accuracy and specificity are dominated by the large number of true negatives (background peptides) compared to the true positives (spike-in peptides).
+In terms of recall, the proposed approach far out performed the other two methods, showing that it correctly labeled the most spike-in peptides.
 \label{fig:spike_stat}}
@@
 
-\begin{figure}[h!]
-\centering
-\includegraphics[width=.85\textwidth]{sim_new/spike_in_fc}
-\caption{The fold change misalignment before adjustment is illustrated in the boxplots.
-The true log fold change is indicated by the red 'X' marks.
-Before adjustment the boxplots are not close to the true fold change.
-After adjustment all three models generally have a median near the true fold change.
-In terms of model comparison, the proposed method shows a much tighter distribution around the true fold change, whole both Limma and $t$-test are wider.
-\label{fig:spike_fc}}
-\end{figure}
-
 \clearpage
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Biological investigations}
+\subsection{Biological investigations}
 \label{sec:bio_investigations}
 
-\subsection{Dataset 4 : Human - Ubiquitination - 1mix-TMT}
+\subsubsection{Dataset 4 : Human - Ubiquitination - 1mix-TMT}
 \label{sec:ipah}
@@
 Uninfected cells were measured at 0 and 6 hours, while infected cells were measured at 1, 2, 4, and 6 hour increments, resulting in six total conditions.
-The experiment was unbalanced with two bioreplicates per condition for all conditions except for infected 1 hour.
+The experiment was unbalanced with two biological replicates per condition for all conditions except for infected 1 hour.
 The experiment was modeled as a group comparison, with different replicates measured at each time point.
@@
 \end{tabular}
-\caption{The experimental design of the IpaH7.8 experiment}
+\caption{The experimental design of Dataset 4}
 \label{table:ipah_design}
@@
 
-The results of the proposed method to this experiment can be seen in \sfigref{fig:ipha_venn_d} and \sfigref{fig:ipah_figures}.
+The results of the proposed method to this experiment can be seen in \sfigref{fig:ipah_figures}.
 
@@
 \clearpage
-\subsection{Dataset 5 : Mouse - Phosphorylation - 2mix-TMT time series}
+\subsubsection{Dataset 5 : Mouse - Phosphorylation - 2mix-TMT time series}
 \label{sec:shigella}
@@
 The abundance of total protein and post-translation modifications were quantified at three time points, uninfected, early infection (45-60 minutes), and late infection (3-3.5 hours).
-The experiment was modeled as a time series, with the same bioreplicate being measured at each time point.
+The experiment was modeled as a time series, with the same biological replicate being measured at each time point.
 Quantifying the total protein along with the post-translational modifications allowed us to adjust for changes in total protein and see the true impact of the site specific modifications.
@@
 \end{tabular}
-\caption{The experimental design of the Shigella flexneri experiment}
+\caption{The experimental design of Dataset 5}
 \label{table:shigella_design}
@@
 
-The results of the proposed method to this experiment can be seen in \sfigref{fig:shig_venn_d},  \sfigref{fig:No_Diff_Shigella_PTM}, and \sfigref{fig:Diff_Shigella_PTM}.
-
-\begin{figure}[h!]
-\centering
-\includegraphics[height=.65\textwidth]{sim_new/shig_venn_diagramm.png}
-\caption{The overlap of differential modified peptides between the PTM model with and without global protein level adjustment.
-Much like in \sfigref{fig:ipha_venn_d}, more PTMs became insignificant after adjustment then became significant.
-This indicates that for the peptides that became insignificant in the adjusted model, their change in abundance was mainly due to changes in global protein abundance.
-Again we tested if the decrease in significant peptides was due to the increased variance that comes from adjustment.
-This was tested by looking for modified peptides whose adjusted log fold change was within 10\% of the unadjusted log fold change but became insignificant after adjustment.
-When this test was applied on this experiment, 548 peptides became insignificant due to an increase in variance.
-This is higher than in \sfigref{fig:ipha_venn_d}, however it is still a very small portion (3.4\%) of the peptides that became insignificant after adjustment.
-Thus we can conclude that the drop off in significant peptides was mainly due to changes in global protein abundance.}
-\label{fig:shig_venn_d}
-\end{figure}
+The results of the proposed method to this experiment can be seen in  \sfigref{fig:No_Diff_Shigella_PTM} and \sfigref{fig:Diff_Shigella_PTM}.
 
@@
 
-\subsection{Dataset 6 : Human - Ubiquitination - Label-free no global profiling run}
+\subsubsection{Dataset 6 : Human - Ubiquitination - Label-free no global profiling run}
 \label{sec:usp30}
@@
 Once identification and quantification of the Ubiquitinated profiling was performed, peptides which were unmodified were extracted and used in place of a global profiling run.
-This resulted in a significant lack of overlap between modified and unmodified peptides.
-Any modified peptide without a corresponding unmodified protein could not be adjusted.
-Of the 10,799 modified peptides identified, only 4526 had a corresponding unmodified run and could be adjusted.
-Not having a separate global profiling run resulted in very low feature counts for the unmodified protein model.
-
-An example profile plot for this experiment can be seen in \sfigref{fig:usp30_ven_d} and \sfigref{fig:USP30_profile_plot}.
-
-\begin{figure}[h!]
-\centering
- \begin{subfigure}{\textwidth}
- \centering
-	\includegraphics[height=.5\textwidth]{sim_new/usp30_venn_diagramm}
-	\caption{The overlap of differential modified peptides for the PTM model with and without global protein level adjustment.
-Here many more PTMs became insignificant then became significant after adjustment.
-This is due to not having a global profiling run, resulting in a large lack of overlap between the modified peptides and unmodified proteins.}
- \end{subfigure}\vspace{-5mm}
- \begin{subfigure}{\textwidth}
- \centering
-	\includegraphics[height=.5\textwidth]{sim_new/usp30_venn_diagramm_matching_only}
-	\caption{Here we make the same comparison but only for modified peptides with a matching unmodified protein, so adjustment can be performed.
-In this case we see significantly less peptides become insignificant after adjustment.
-This highlights the need for a global profiling run if protein adjustment is going to performed.
-As in the previous experiments, we can check if peptides are becoming insignificant due to an increase in variance.
-To check this we look at modified peptides whose adjusted fold change is within 10\% of their unadjusted fold change and they become insignificant after adjustment.
-In this experiment there are only 25 peptides that meet this criteria, a very small percentage of all modified peptides.}
- \end{subfigure}
- \caption{Overlap of significant PTMs before and after protein adjustment.}
-\label{fig:usp30_ven_d}
-\end{figure}
+This resulted in a significant lack of overlap between modified and unmodified peptides and low feature counts for the unmodified protein model.
+
+An example profile plot for this experiment can be seen in  \sfigref{fig:USP30_profile_plot}.
 
@@
 The modification appears generally unchanged between all conditions, whereas the global profiling run shows the CCCP and Combo conditions have a higher relative abundance compared to the Control and USP30\_OE.
-This indicates that the modification actually had a major effect when comparing CCCP and Combo to Control and USP30\_OE.
-This would have been missed without adjusting for global protein changes.}
+This indicates that the modification actually had an effect when comparing CCCP and Combo to Control and USP30\_OE.
+However it is not entirely clear as one unmodified peptide feature appears to be changed, while the other does not.
+This uncertainty is another result of not running a separate global profiling run.
+With a global profiling run, many unmodified features are generally quantified, removing the uncertainty that comes with low feature counts.}
 \label{fig:USP30_profile_plot}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\section{Sample size calculation and power analysis}
+\section{Sample size calculation and power analysis}
+
+\begin{figure}[h!]
+\centering
+ \begin{subfigure}{\textwidth}
+ \centering
+	\includegraphics[width=.75\textwidth]{sim_new/same_var_power}
+	\caption{The power of an experiment targeting PTMs with the same variance, .15, for the modified and unmodified peptides.
+Predictably when the replicates are high for both modified and unmodified peptides the power is much higher.
+Conversely at low replicates for each the power is much lower.
+The interesting part of this chart is when the replicates are different between runs.
+With equal variance, it does not matter if the PTM replicates or protein replicates are higher.}
+ \end{subfigure}
+ \begin{subfigure}{\textwidth}
+ \centering
+	\includegraphics[width=.75\textwidth]{sim_new/high_ptm_var_power}
+	\caption{In this chart the variance for the PTM is higher than the unmodified protein.
+The PTM variance is .2, while the unmodified protein variance is .1.
+With equal replicates the results are the same as above, more replicates equals more power.
+When the replicates are not the same we can clearly see that having more replicates for the PTM leads to higher power.}
+ \end{subfigure}
+ \caption{Power analysis of experiments with differing variances.}
+\label{fig:power_sd_combo}
+\end{figure}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\clearpage
+\clearpage
 

### 2022-02-13 main/MSstatsPTM_Main.tex
Finalized main and supp text

@@
 
-\subsection*{Dataset 2 : Computer simulation 2 - Label-free missing values and low replicates}
+\subsection*{Dataset 2 : Computer simulation 2 - Label-free missing values and low features}
 \label{sec:comp_sim_procedure2}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection*{The proposed approach outperforms existing methods in simulations with missing values and low replicates}
+\subsubsection*{The proposed approach outperforms existing methods in simulations with missing values and low features}
 

### 2022-02-13 supplementary/ptm_sm.tex
Finalized main and supp text

@@
 \usepackage{xr}
+\usepackage{hyperref}
 \externaldocument[main-]{../main/MSstatsPTM_Main}
@@
 The log-intensity of Feature $l$ from the unmodified peptides in the same run is denoted by $y_{ijl}$.
-\sfigref{fig:dtable} shows an example data representation of modified peptide ions at one site and unmodified peptide ions of the same protein.
+Figure \ref{fig:dtable} shows an example data representation of modified peptide ions at one site and unmodified peptide ions of the same protein.
 Unmodified peptides from the same protein provide additional evidence on the underlying protein abundance, which needs to be integrated for PTM characterization.
@@
 The estimation of these quantities is dependent on the experimental design.
-In \sfigref{fig:statistical_inference} different experimental designs that the proposed approach can be applied to, along with their estimation of standard error and degrees of freedom is shown.
+In Figure \ref{fig:statistical_inference} different experimental designs that the proposed approach can be applied to, along with their estimation of standard error and degrees of freedom is shown.
 
@@
 
+Table \ref{tab:dataDescription} summarizes the datasets used in this manuscript and their experimental designs.
+There were 6 total experiments, 3 for which ground truth was known and 3 biological investigations.
+For the ground truth experiments, 2 were generated via computer simulation with differing parameters and the other was an experiment with heavy labeled spike-in peptides.
+The biological investigations represent a variety of experimental designs and acquisition methods.
+Two were acquired via Tandem Mass Tag (TMT) labeling, while the other was a label free experiment.
+Of the two TMT experiments, one was analyzed as a time series and the other a group comparison.
+
+Details of raw data, R scripts with MSstatsPTM analyses, and result of statistical analyses for the datasets are available in either MassIVE.quant or Github (with direct links available in the table).
+
 %%%% Summary of controlled mixtures
-\begin{table}
+\begin{table}[h!]
 \begin{tiny}
-\begin{tabular}{|c|c|ccccc|c|}
+\begin{tabular}{|c|c|cccc|c|}
 \hline
 %%%% colnames of table
-&  & \multicolumn{5}{c|}{Experimental design} & Data availability \\
-& Dataset & \multirow{2}{*}{Comparison} & No. of & No. of Bio.& No. of Tech.& \multirow{2}{*}{Details in}   & (MassIVE.quant  \\
-&  & & Conditions & Replicates & Replicates &  & or R package) \\
-\hline
-\hline
-% &&&&& \\[-0.05in]
+&  & \multicolumn{4}{c|}{Experimental design} & Data availability \\
+& Dataset & \multirow{2}{*}{Comparison} & No. of & No. of Bio.& \multirow{2}{*}{Details in}   & (MassIVE.quant  \\
+&  & & Conditions & Replicates &  & or R package) \\
+\hline
+\hline
+ &&&&&& \\[-0.05in]
 %%%% DDA:ControlledMix
-\multirow{3}{*}{\rotatebox[origin=c]{90}{Known}}  \multirow{3}{*}{\rotatebox[origin=c]{90}{ Ground}} \multirow{3}{*}{\rotatebox[origin=c]{90}{ Truth}} & DDA:ControlledMix \cite{Chiva2014} & Group & 5 & 1 & 3 & Suppl.
+\multirow{5}{*}{\rotatebox[origin=c]{90}{Known}}  \multirow{5}{*}{\rotatebox[origin=c]{90}{Ground}} \multirow{5}{*}{\rotatebox[origin=c]{90}{Truth}} & Computer simulation 1 - Label-free & Group & 2/3/4 & 2/3/5/10 & Suppl.
 Sec.
-& \\
- &&&&&&& \\[-0.05in]
+\ref{sec:dataset1} & \href{https://github.com/devonjkohler/MSstatsPTM_simulations}{Github}\\
+ &&&&&& \\%[-0.05in]
 %\hline 
@@
 %%% DDA iPRG2015
-& DDA:Choi2017 \cite{Choi:2017jan}  & Group & 4 & 1 & 3 &  & \\
- &&&&&&& \\[-0.05in]
+& Computer simulation 2 - missing and low features & Group & 2/3/4 & 2/3/5/10 & Suppl. Sec. \ref{sec:dataset2} & \href{https://github.com/devonjkohler/MSstatsPTM_simulations}{Github} \\
+ &&&&&& \\%[0.05in]
 %\hline 
-% &&&&&&& \\[-0.05in]
+% &&&&&& \\[-0.05in]
 %%% DDA Cox2014
-& DDA:Cox2014 \cite{Cox:2014fi} & Group & 2 & 1 & 4 &  & \\
- &&&&&&& \\[-0.05in]
-\hline
-\multicolumn{8}{c}{ } \\ [0.02in]
-\hline 
- &&&&&&& \\[-0.05in]
+& SpikeIn benchmark & Group & 4 & 2 & Suppl. Sec. \ref{sec:benchmark} & \\
+ &&&&&& \\%[-0.05in]
+\hline
+\multicolumn{7}{c}{ } \\ [0.02in]
+\hline 
+ &&&&&& \\%[-0.05in]
 %%%%%%%%%%%%%%%%%%
@@
 %% DDA Meierhofer
-\multirow{3}{*}{\rotatebox[origin=c]{90}{Biological}} \multirow{3}{*}{\rotatebox[origin=c]{90}{investigation}} & DDA:Meierhofer2016 \cite{Meierhofer2016} & Paired & 2 & 6 & 2 &  &  \\
-&&&&&&& \\[-0.05in]
+\multirow{3}{*}{\rotatebox[origin=c]{90}{Biological}} \multirow{3}{*}{\rotatebox[origin=c]{90}{Investigation}} & Human - Ubiquitination - TMT & Group & 6 & 2 & Suppl. Sec. \ref{sec:ipah} &  \\
+&&&&&& \\%[-0.05in]
 % \cline{2-9}
@@
 %%%% SRM Picotti 2009
-& SRM:Picotti2009  \cite{Picotti09} & Time course & 10 & 3 & 1 &  & \\
-&&&&&&& \\[-0.05in]
+& Mouse - Phosphorylation - TMT & Time course & 6 & 2 & Suppl. Sec. \ref{sec:shigella} & \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=4878d777c6b34cf8aaf8477e93140c4d}{MSV000085565} \\
+&&&&&& \\%[-0.05in]
 % \cline{2-9}
@@
 %%%% SRM Surinova 2015, training
-& SRM:Surinova2015:Training \cite{Surinova:2015jn} & Group & 2 & 100 & 1 & &\ [0.02in]
+& Human - Ubiquitination - Label-free & Group & 4 & 2 & Suppl. Sec. \ref{sec:usp30} &\\ [0.02in]
+ &&&&&& \\
+ \hline
 \end{tabular}\\
 \end{tiny}
-\caption{
+\caption{ \small {\bf Experimental datasets in this manuscript} 
+```Dataset'' is the code name of the dataset used in this manuscript.
+`Comparison'' is the type of experimental design (group comparison, or time course).
+``No. of Conditions'' states the number of conditions in the dataset.
+``No. of Bio. Replicates'' states the number of biological replicates per condition.
+Details regarding the experimental design for each dataset are available in the section listed in the ``Details in'' column.
+``Data availability'' column shows the MassIVE.quant ID or location on github. 
 }
-\label{tab:dataDescriptionSpikein}
+\label{tab:dataDescription}
 \end{table}
@@
 \subsubsection{Dataset 1 : Computer simulation 1 - Label-free}
-\label{dataset1}
+\label{sec:dataset1}
 In the first simulation an experiment with many features per PTM and unmodified protein was created.
@@
 
-The results are summarized from \sfigref{fig:sim1_fpr} to \sfigref{fig:sim1_recall}.
+The results are summarized from Figure \ref{fig:sim1_fpr} to Figure \ref{fig:sim1_recall}.
 These results include the false discovery rate with and without adjusting for changes in unmodified protein abundance, false positive rate, and overall accuracy.
@@
 
-\subsubsection{Dataset 2 : Computer simulation 2 - Label-free missing values and low replicates}
+\subsubsection{Dataset 2 : Computer simulation 2 - Label-free missing values and low features}
+\label{sec:dataset2}
 
@@
 
-The results are summarized from \sfigref{fig:sim2_recall} to \sfigref{fig:fc_boxplot}.
+The results are summarized from Figure \ref{fig:sim2_recall} to Figure \ref{fig:fc_boxplot}.
 
@@
 The proposed method summarizes feature intensities up to the run level using Tukey's Median Polish, while the other methods use the log of summed feature intensities.
-The results are summarized from \sfigref{fig:spike_volcano_msstats} to \sfigref{fig:spike_fc}, including volcano plots, model summary statistics, and fold change analysis.
+The results are summarized from Figure \ref{fig:spike_volcano_msstats} to Figure \ref{fig:spike_fc}, including volcano plots, model summary statistics, and fold change analysis.
 
@@
 
-The results of the proposed method to this experiment can be seen in \sfigref{fig:ipah_figures}.
+The results of the proposed method to this experiment can be seen in Figure \ref{fig:ipah_figures}.
 
@@
 
-The results of the proposed method to this experiment can be seen in  \sfigref{fig:No_Diff_Shigella_PTM} and \sfigref{fig:Diff_Shigella_PTM}.
+The results of the proposed method to this experiment can be seen in Figure \ref{fig:No_Diff_Shigella_PTM} and Figure \ref{fig:Diff_Shigella_PTM}.
 
@@
 
-An example profile plot for this experiment can be seen in  \sfigref{fig:USP30_profile_plot}.
+An example profile plot for this experiment can be seen in  Figure \ref{fig:USP30_profile_plot}.
 
@@
 With equal variance, it does not matter if the PTM replicates or protein replicates are higher.}
- \end{subfigure}
+ \end{subfigure}\vspace{-1mm}
  \begin{subfigure}{\textwidth}
@@
 When the replicates are not the same we can clearly see that having more replicates for the PTM leads to higher power.}
- \end{subfigure}
+ \end{subfigure}\vspace{-1mm}
  \caption{Power analysis of experiments with differing variances.}

### 2022-02-13 code/MSstatsPTM2_spike.R

@@ -367,6 +367,9 @@ summary_df$PTM$ProteinLevelData <- summary_df$PTM$ProteinLevelData %>%
                                  LogIntensities -1, LogIntensities))
 
 save(summary_df, file = "v2_summarized.rda")
+load(file = "D:\\Northeastern\\Research\\MSstats\\PTM_Refractor\\Label_Free_PTM_Data\\benchmark-spike\\saved_data\\v2_summarized.rda")
+
+
 
 ## Model
 model_df <- groupComparisonPTM(summary_df, data.type = "LabelFree", use_log_file=FALSE)

### 2022-02-13 code/Shigella_Analysis.R

@@ -25,8 +25,9 @@ pST_input <- list(PTM = data.frame(shigella_peptide_pST),
 py_summarization <- dataSummarizationPTM_TMT(pY_input)
 pST_summarization <- dataSummarizationPTM_TMT(pST_input)
 
-save(py_summarization, file = "Shigella_pY_Summarization.rda")
-save(pST_summarization, file = "Shigella_pST_Summarization.rda")
+save(py_summarization, file = "../data/Shigella_pY_Summarization.rda")
+load(file = "../data/Shigella_pY_Summarization.rda")
+save(pST_summarization, file = "../data/Shigella_pST_Summarization.rda")
 
 py_summarized_data <- py_summarization$PTM$ProteinLevelData
 py_feature_data <- py_summarization$PTM$FeatureLevelData

### 2022-02-14 main/MSstatsPTM_Main.tex
Abstract+intro

@@
 
-The scientific community widely utilizes mass spectrometry (MS)-based proteomics to quantify the abundance of proteins and their post-translational modifications (PTMs).
-Experiments targeting PTMs face several specific challenges which are statistical in nature.
-These include the low abundance of modified proteo-forms, few representative peptides that span modification sites, and convolution with abundance changes in the overall protein expression.
-Due to these challenges, a robust approach to estimating relative systematic changes in PTMs should combine information pertaining to PTM sites over several peptides, replicates in multiple conditions, and consider sources of confounding and variation present in the experiment.
-We propose a general statistical model and workflow that is both reproducible and comprehensive.
-The method measures modified and unmodified peptide abundance by summarizing intensities through Tukey’s median polish method.
-Then a model based on the family of linear mixed-effects models is fit.
-This model is automatically adjusted to the specific experimental design.
-Finally, the PTM abundances are adjusted to remove bias from changes in the overall protein.
-We implemented this model in the free and open-source R package $MSstatsPTM$.
-The approach is evaluated on computer simulations, a spike-in experiment with known ground truth, and three biological experiments with varying acquisition methods.
-Finally, the approach is compared against several existing methods and is shown to outperform all other methods.
+The scientific community increasingly utilizes mass spectrometry (MS)-based proteomics to detect changes in post-translational modifications (PTMs) between samples from various conditions.
+Experiments targeting PTMs face numerous challenges.
+These include the low abundance of modified proteoforms, few representative peptides that span modification sites, and confounding between changes in the abundance of PTM and the overall changes in the protein abundance.
+Therefore, statistical approaches to estimating relative changes in PTMs, and to detecting differential PTM abundance, should combine information pertaining to PTM sites over several peptides, replicates in multiple conditions, and consider relevant sources of confounding and variation.
+In this manuscript we propose such a general statistical framework, which is comprehensive, accurate, and leads to reproducible results.
+The framework requires an experimental design\todo{that, for each sample, quantifies the peptides with post-translational modifications, as well as  peptides from the same proteins with no modification sites}, \todo{and is applicable to both label-free and tandem mass tag (TMT)-based acquisitions}.
+The proposed framework summarizes the abundances of \todo{peptides with and without the modification sites}, and fits a linear mixed effects model that automatically adjusts to the biological and technological aspects of the experimental design.
+Finally, changes in the PTM and the protein-level abundances are combined to remove bias due to the confounding of these two sources.
+Evaluations on computer simulations, a spike-in experiment with known ground truth, and three biological experiments with \todo{which?} acquisition methods \todo{demonstrate the improved - what?
+- as compared to the currently used approaches for relative PTM quantification}.
+The proposed framework is implemented in the free and open-source R package $MSstatsPTM$.
 
@@
 
-The signaling mechanisms that allow cells to mount a dynamic and fast response to a multitude of events are primarily facilitated by the modification of proteins at specific residues, acting as molecular on/off switches\cite{Deribe} \cite{Cohen}.
-Mass spectrometry-based proteomics is broadly established as the tool-of-choice for unbiased and large-scale identification and quantification of proteins and their post-translational modifications (PTMs) using liquid chromatography coupled with mass spectrometry (LC-MS)\cite{Kall:2011ub} \cite{Roepstorff}.
+The signaling mechanisms that allow cells to mount a dynamic and fast response to a multitude of events are facilitated by the modification of proteins at specific residues, acting as molecular on/off switches\cite{Deribe,Cohen}.
+Bottom-up liquid chromatography coupled with mass spectrometry (LC-MS) is broadly established as the tool-of-choice for unbiased and large-scale identification and quantification of proteins and their post-translational modifications (PTMs) \cite{Kall:2011ub,Roepstorff}.
 Studies targeting the post-translationally modified proteome focus either on the accurate localization of modification sites on proteins, relative or absolute quantification of a modification site's occupancy repertoire, or relative changes in occupancy across experimental conditions\cite{Mann}.
+\todo{The previous sentence is great, but I am wondering whether it can lead the reviewers in a wrong direction, since we only address one of these problems } Some scientific questions,  such as discovery of protein biomarkers of a disease, consider meaningful patterns of changes in PTM abundance alone \cite{Petushkova_2017}.
+Other questions, e.g.
+in biological and physiological studies based on very short timescales, require us to distinguish the confounding of quantitative changes in PTM and of overall changes of the protein abundance.
+
 Regardless of the question at hand, interrogating the modified proteome is challenging due to a number of reasons.
-First, the relatively lower abundance of modified proteo-forms dictates that a global interrogation can only be achieved through large-scale enrichment protocols with modification-specific antibodies or beads.
+First, the relatively lower abundance of modified proteoforms dictates that a global interrogation can only be achieved through large-scale enrichment protocols with modification-specific antibodies or beads \todo{refs for these points?}.
 Variability in the enrichment efficiency inevitably affects the reproducibility of the number of spectral features (e.g., peptide precursor ions or their fragments) and their intensities, which imposes challenges in both quantification and statistical modeling.
 Second, contrary to the often large number of identified peptides that can be used as features to model protein abundance changes, there are relatively few representative peptides that span a modification sites, which often results in sparse, and sometimes, inherently convoluted models (i.e., single versus multiple modified sites on a single peptide).
-Third, unless early signaling events are interrogated, the interpretation of the relative changes in modification occupancy are inherently convoluted with changes in the overall protein expression, making the interpretation of the results not straightforward\cite{Olsen:2013}.
-Therefore, a robust approach to estimate systematic relative changes in post-translational modifications, at scale, should not only combine the quantitative information pertaining to a PTM site over peptides and replicates in multiple conditions, but take into account various sources of variations and confounding factors present in the experiments.
+Third, unless early signaling events are interrogated, the interpretation of the relative changes in modification occupancy are inherently confounded with changes in the overall protein abundance, complication the interpretation of the results \cite{Olsen:2013}.
+\todo{Finally, technological aspects of bottom-up MS experiments, and presence of labeling, e.g.
+by tandem mass tag (TMT), introduce additional sources of uncertainty and variation.}
 
-There are two main biological questions that may be relevant in PTM analysis.
-First, some researchers run mass-spectrometry based experiments targeting PTMs when looking for biomarkers.\cite{Petushkova_2017}.
-In this case the researcher may not necessarily want to correct for confounding with the total protein.
-The interest is in detecting modified peptides which change in abundance in a meaningful way, such as between disease and control, fast or slow-progressing, or responder and non-responder groups.
-Conversely, researchers may be interested in whether the fraction of peptides modified at a given site is changing between conditions.
-This is an important question for thinking about biological and physiological models and very short timescale experiments.
-In this case, we are interested in the relative abundance change solely from the modification and correcting for confounding with the total protein is relevant.
-These two questions are both important areas of study and require different statistical methods to correctly address.
-In this paper we will mainly address a statistical solution to the second question, which requires more rigorous methods that account for confounding with the total protein.
+\todo{Currently statistical analyses of these experiments are performed with methods such as t-test and Limma methods\cite{Ritchie_15a}. They have drawbacks XXX}
 
-We propose a general statistical approach, which explicitly characterizes the variations and confounding factors present in bottom-up PTM experiments.
-The proposed approach is aimed at the detection of quantitative changes in PTMs between conditions utilizing procedures developed for summarization of LC-MS data, quantitative characterization of site-specific PTMs, and adjustment with respect to protein abundance.
-Quantitative analyses of PTMs often involve comparisons between multiple inter-related conditions of the same biological system.
-The general statistical framework underlying the proposed approach allows for analyzing experiments with complex designs and different acquisition methods.
+This manuscript proposes a general statistical analysis framework that detects relative changes in post-translational modifications.
+The proposed framework combines the quantitative information pertaining to a PTM site over peptides and replicates in multiple conditions, and takes into account various sources of variation and confounding.
+The framework requires an experimental design\todo{that, for each sample, quantifies the peptides with post-translational modifications, as well as  peptides from the same proteins with no modification sites}.
+\todo{It is applicable to both label-free and tandem mass tag (TMT)-based acquisitions}.
+The proposed framework summarizes the abundances of \todo{peptides with and without the modification sites}, and fits a linear mixed effects model that automatically adjusts to the biological and technological aspects of the experimental design.
+\todo{For questions focusing on PTM, can analyze changes in the modified peptides.
+For eliminating the confounding, combine...} Finally, changes in the PTM and the protein-level abundances are combined to remove bias due to the confounding of these two sources.
 
-The proposed approach was implemented using the R coding language and evaluated using datasets from computer simulations, benchmark controlled mixtures, and biological investigations.
-The approach was compared against the commonly applied t-test and Limma methods\cite{Ritchie_15a}.
-The results demonstrate that by appropriately leveraging the information from the entire dataset, the proposed approach improves the reproducibility and accuracy of the estimates of PTM fold changes and results in a better calibrated type I error rate.
-Finally, the proposed approach is implemented as a freely available open source R package $MSstatsPTM$, available on Bioconductor, which employs similar input format as in $MSstats$ and $MSstatsTMT$, and is compatible with many acquisition methods, such as label free, DIA, DDA, and TMT\cite{Choi:2014} \cite{Huang:2020}.
+We evaluated the proposed framework on datasets from computer simulations, benchmark controlled mixtures, and biological investigations \todo{expand on the relevance and the comprehensive nature of the datasets - diverse organisms, acquisitions}.
+The evaluations demonstrated that by appropriately leveraging the information from \todo{which sources}, the proposed approach improves the accuracy of the estimates of PTM fold changes and results in a better calibrated proportion of false positive \todo{what?}.
+\todo{In particular, accounting for the confounding from protein abundance...}
+
+The proposed approach is implemented as a freely available open source R package $MSstatsPTM$, available on Bioconductor, which employs similar input format as in $MSstats$ and $MSstatsTMT$, and is compatible with many acquisition methods, such as label free, DIA, DDA, and TMT\cite{Choi:2014,Huang:2020}.
+\todo{we claim that the results are reproducible - what does this mean?}
 
@@
 
-This section describes the datasets and experiments used to evaluate the proposed method.
-The datasets were chosen to provide a performance benchmark of the proposed approach against existing methods when ground truth is known, as well as represent a variety of experimental designs and acquisition methods where the approach can be applied.
-To benchmark the approach, computer simulations and a spike-in experiment were used.
-The computer simulations vary in experimental realism; the first simulation was perfectly clean, with many replicates and no missing values, while the second simulation introduced real world qualities, such as limited modified features and missing values.
-The spike-in experiment took the real world qualities a step further and allowed us to compare the methods in a real experiment where ground truth is known.
-Finally the proposed approach was used to model three biological experiments which differed in experimental design, acquisition strategy, and organism.
-These experiments were chosen to show the proposed method's applicability to many different types of experiments.
+The datasets were chosen to provide a performance benchmark of the proposed approach in situations with known ground truth, and to represent a variety of experimental designs and acquisition methods.
+The computer simulations varied in experimental realism.
+The first simulation was perfectly clean, with many replicates and no missing values, while the second simulation introduced real world qualities, such as limited modified features and missing values.
+The spike-in experiment took the real world qualities a step further and allowed us to compare the methods in a real experiment with known \todo{changes in what?}.
+Finally, three biological experiments demonstrated the applicability of the proposed approach across different biological organisms, experimental designs, and data acquisition strategies.
 
@@
 
-Here simulations were generated mimicking a Label-free experiment with a high number of modified and unmodified peptide features and no missing values.
-Multiple simulations were generated with different numbers of replicates and conditions.
+The simulation mimicked a label-free experiment with a high number of modified and unmodified peptide features and no missing values.
+Multiple datasets were generated with different numbers of replicates and conditions.
 In all simulations 1000 modified peptides were generated, half of which were true positives and half true negatives.
@@
 Despite the important implications of PTMs in biological functions, there is a lack of general framework to summarize the available quantitative information from LC-MS data, to perform statistical inference, and to draw conclusions to characterize the quantitative properties of PTM in a statistically rigorous manner.
-Many investigations performed differential expression analysis of PTMs using two-sample t-test or its extensions.
+Many investigations performed differential abundance analysis of PTMs using two-sample t-test or its extensions.
 The approach takes as input intensities of individual features from modified peptides, or intensity ratios of modified and unmodified peptide features, and compares the mean abundance of a PTM site from one condition to another.
@@
 
+\todo{Check that this section uses past tense; look out for informal expressions}
+
 The performance of the proposed method was evaluated on simulated and spike in datasets with known ground truth, as well as biological experimental data where the ground truth was not known.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection*{In a Label-free benchmark experiment the proposed approach corrects for bias and outperforms existing methods}
+\subsubsection*{In a label-free benchmark experiment the proposed approach corrects for bias and outperforms existing methods}
 
@@
 
-\subsubsection*{Proposed approach is less effective without a separate global profiling run but still fixes bias in Label-free experiment}
+\subsubsection*{In label-free experiment without a separate global profiling run, the proposed approach could eliminate the bias \todo{due to what?}, albeit less effectively than \todo{in presence of a global profiling run?}}
 
@@
 
-Additionally, expression levels of PTM sites can be convoluted with each other if there are two or more modification sites per peptide.
+Additionally, abundance levels of PTM sites can be convoluted with each other if there are two or more modification sites per peptide.
 In the current implementation the effect of a specific modification in a peptide with multiple modifications cannot be quantified.

### 2022-02-17 main/MSstatsPTM_Main.tex
main intro updates

@@
 In this manuscript we propose such a general statistical framework, which is comprehensive, accurate, and leads to reproducible results.
-The framework requires an experimental design\todo{that, for each sample, quantifies the peptides with post-translational modifications, as well as  peptides from the same proteins with no modification sites}, \todo{and is applicable to both label-free and tandem mass tag (TMT)-based acquisitions}.
-The proposed framework summarizes the abundances of \todo{peptides with and without the modification sites}, and fits a linear mixed effects model that automatically adjusts to the biological and technological aspects of the experimental design.
+The framework requires an experimental design that, for each sample, quantifies the peptides with post-translational modifications, as well as  peptides from the same proteins with no modification sites, and is applicable to both label-free and tandem mass tag (TMT)-based acquisitions.
+The proposed framework summarizes the abundances of peptides with and without the modification sites, and fits a linear mixed effects model that automatically adjusts to the biological and technological aspects of the experimental design.
 Finally, changes in the PTM and the protein-level abundances are combined to remove bias due to the confounding of these two sources.
-Evaluations on computer simulations, a spike-in experiment with known ground truth, and three biological experiments with \todo{which?} acquisition methods \todo{demonstrate the improved - what?
-- as compared to the currently used approaches for relative PTM quantification}.
+Evaluations on computer simulations, a spike-in experiment with known ground truth, and three biological experiments which demonstrate the improved fold change estimation as compared to the currently used approaches for relative PTM quantification.
 The proposed framework is implemented in the free and open-source R package $MSstatsPTM$.
@@
 Studies targeting the post-translationally modified proteome focus either on the accurate localization of modification sites on proteins, relative or absolute quantification of a modification site's occupancy repertoire, or relative changes in occupancy across experimental conditions\cite{Mann}.
-\todo{The previous sentence is great, but I am wondering whether it can lead the reviewers in a wrong direction, since we only address one of these problems } Some scientific questions,  such as discovery of protein biomarkers of a disease, consider meaningful patterns of changes in PTM abundance alone \cite{Petushkova_2017}.
+Some scientific questions,  such as discovery of protein biomarkers of a disease, consider meaningful patterns of changes in PTM abundance alone \cite{Petushkova_2017}.
 Other questions, e.g.
@@
 Regardless of the question at hand, interrogating the modified proteome is challenging due to a number of reasons.
-First, the relatively lower abundance of modified proteoforms dictates that a global interrogation can only be achieved through large-scale enrichment protocols with modification-specific antibodies or beads \todo{refs for these points?}.
+First, the relatively lower abundance of modified proteoforms dictates that a global interrogation can only be achieved through large-scale enrichment protocols with modification-specific antibodies or beads \cite{Huang:2014}.
 Variability in the enrichment efficiency inevitably affects the reproducibility of the number of spectral features (e.g., peptide precursor ions or their fragments) and their intensities, which imposes challenges in both quantification and statistical modeling.
-Second, contrary to the often large number of identified peptides that can be used as features to model protein abundance changes, there are relatively few representative peptides that span a modification sites, which often results in sparse, and sometimes, inherently convoluted models (i.e., single versus multiple modified sites on a single peptide).
+Second, contrary to the often large number of identified peptides that can be used as features to model protein abundance changes, there are relatively few representative peptides that span a modification sites, which often results in sparse, and sometimes, inherently convoluted models (i.e., single versus multiple modified sites on a single peptide) \cite{Mann}.
 Third, unless early signaling events are interrogated, the interpretation of the relative changes in modification occupancy are inherently confounded with changes in the overall protein abundance, complication the interpretation of the results \cite{Olsen:2013}.
-\todo{Finally, technological aspects of bottom-up MS experiments, and presence of labeling, e.g.
-by tandem mass tag (TMT), introduce additional sources of uncertainty and variation.}
+Finally, technological aspects of bottom-up MS experiments, and presence of labeling, e.g.
+by tandem mass tag (TMT), introduce additional sources of uncertainty and variation.
 
-\todo{Currently statistical analyses of these experiments are performed with methods such as t-test and Limma methods\cite{Ritchie_15a}. They have drawbacks XXX}
+Currently statistical analyses of these experiments are performed with methods such as t-test and Limma\cite{Ritchie_15a}.
+These approaches generally use the intensity ratios of modified and unmodified peptide features and compare the mean abundance of different PTM sites.
+These approaches do not fully account for different sources of variations and as such, are not directly applicable to experiments with complex designs, such as multiple conditions and mixtures.
+Additionally, there is no self contained implementation of these methods, requiring manual application of the methods by the user and increasing complexity.
 
@@
 The proposed framework combines the quantitative information pertaining to a PTM site over peptides and replicates in multiple conditions, and takes into account various sources of variation and confounding.
-The framework requires an experimental design\todo{that, for each sample, quantifies the peptides with post-translational modifications, as well as  peptides from the same proteins with no modification sites}.
-\todo{It is applicable to both label-free and tandem mass tag (TMT)-based acquisitions}.
-The proposed framework summarizes the abundances of \todo{peptides with and without the modification sites}, and fits a linear mixed effects model that automatically adjusts to the biological and technological aspects of the experimental design.
+The framework requires an experimental design that, for each sample, quantifies the peptides with post-translational modifications, as well as  peptides from the same proteins with no modification sites.
+It is applicable to both label-free and tandem mass tag (TMT)-based acquisitions.
+The proposed framework summarizes the abundances of peptides with and without the modification sites, and fits a linear mixed effects model that automatically adjusts to the biological and technological aspects of the experimental design.
 \todo{For questions focusing on PTM, can analyze changes in the modified peptides.

### 2022-02-17 main/MSstatsPTM_Main.tex
intro changes

@@
 The proposed framework summarizes the abundances of peptides with and without the modification sites, and fits a linear mixed effects model that automatically adjusts to the biological and technological aspects of the experimental design.
-\todo{For questions focusing on PTM, can analyze changes in the modified peptides.
-For eliminating the confounding, combine...} Finally, changes in the PTM and the protein-level abundances are combined to remove bias due to the confounding of these two sources.
+Finally changes in the PTM and protein-level abundances are combined to remove bias due to the confounding of these two sources.
 
-We evaluated the proposed framework on datasets from computer simulations, benchmark controlled mixtures, and biological investigations \todo{expand on the relevance and the comprehensive nature of the datasets - diverse organisms, acquisitions}.
-The evaluations demonstrated that by appropriately leveraging the information from \todo{which sources}, the proposed approach improves the accuracy of the estimates of PTM fold changes and results in a better calibrated proportion of false positive \todo{what?}.
-\todo{In particular, accounting for the confounding from protein abundance...}
+We evaluated the proposed framework on datasets from computer simulations, benchmark controlled mixtures, and biological investigations, which illustrate a diverse set of organisms, acquisition methods, and experimental designs, showing the applicability of the methods to a variety of situations.
+The evaluations demonstrated that by appropriately leveraging the information from the unmodified peptide, the proposed approach improves the accuracy of the estimates of PTM fold changes and results in a better calibrated false positive rate.
+In particular, accounting for the confounding from protein abundance allows us to see the true effect of the modification, avoiding the need for more manual and time intensive methods.
 
 The proposed approach is implemented as a freely available open source R package $MSstatsPTM$, available on Bioconductor, which employs similar input format as in $MSstats$ and $MSstatsTMT$, and is compatible with many acquisition methods, such as label free, DIA, DDA, and TMT\cite{Choi:2014,Huang:2020}.
-\todo{we claim that the results are reproducible - what does this mean?}
 
@@
 The first simulation was perfectly clean, with many replicates and no missing values, while the second simulation introduced real world qualities, such as limited modified features and missing values.
-The spike-in experiment took the real world qualities a step further and allowed us to compare the methods in a real experiment with known \todo{changes in what?}.
+The spike-in experiment took the real world qualities a step further and allowed us to compare the methods in a real experiment with known changes in modified spike-in peptides.
 Finally, three biological experiments demonstrated the applicability of the proposed approach across different biological organisms, experimental designs, and data acquisition strategies.
+A summary of the experiments used in this manuscript can be seen in \todo{add and reference table}.
 

### 2022-02-21 main/MSstatsPTM_Main.tex
changes to intro

@@
 \usepackage{subcaption}
+\usepackage{hyperref}
 \usepackage{xcolor} %\[table,dvipsnames]
@@
 
+This section summarizes the datasets in this manuscript.
 The datasets were chosen to provide a performance benchmark of the proposed approach in situations with known ground truth, and to represent a variety of experimental designs and acquisition methods.
@@
 Finally, three biological experiments demonstrated the applicability of the proposed approach across different biological organisms, experimental designs, and data acquisition strategies.
-A summary of the experiments used in this manuscript can be seen in \todo{add and reference table}.
+A summary of the experiments used in this manuscript can be seen in Table \ref{tab:dataDescription}.
+Details of raw data, R scripts with MSstatsPTM analyses, and result of statistical analyses for the datasets are available in either MassIVE.quant or Github (with direct links available in the table).
+
+%%%% Summary of controlled mixtures
+\begin{table}[h!]
+\centering
+\begin{tiny}
+\begin{tabular}{|c|c|cccc|c|}
+\hline
+%%%% colnames of table
+&  & \multicolumn{4}{c|}{Experimental design} & Data availability \\
+& Dataset & \multirow{2}{*}{Comparison} & No. of & No. of Bio.& \multirow{2}{*}{Details in} & (MassIVE.quant  \\
+&  & & Conditions & Replicates &  & or R package) \\
+\hline
+\hline
+ &&&&&& \\[-0.05in]
+%%%% DDA:ControlledMix
+\multirow{5}{*}{\rotatebox[origin=c]{90}{Known}}  \multirow{5}{*}{\rotatebox[origin=c]{90}{Ground}} \multirow{5}{*}{\rotatebox[origin=c]{90}{Truth}} & Computer simulation 1 - Label-free & Group & 2/3/4 & 2/3/5/10 & Suppl.
+Sec.
+\ref{sec:dataset1} & \href{https://github.com/devonjkohler/MSstatsPTM_simulations}{Github}\\
+ &&&&&& \\%[-0.05in]
+%\hline 
+%&&&&&&&& \\[-0.05in]
+% \cline{2-9}
+% &&&&&&&& \\[-0.05in]
+%%% DDA iPRG2015
+& Computer simulation 2 - missing and low features & Group & 2/3/4 & 2/3/5/10 & Suppl. Sec. \ref{sec:dataset2} & \href{https://github.com/devonjkohler/MSstatsPTM_simulations}{Github} \\
+ &&&&&& \\%[0.05in]
+%\hline 
+% &&&&&& \\[-0.05in]
+%%% DDA Cox2014
+& SpikeIn benchmark & Group & 4 & 2 & Suppl. Sec. \ref{sec:benchmark} & \\
+ &&&&&& \\%[-0.05in]
+\hline
+\multicolumn{7}{c}{ } \\ [0.02in]
+\hline 
+ &&&&&& \\%[-0.05in]
+%%%%%%%%%%%%%%%%%%
+%%%% biological study
+%%%%%%%%%%%%%%%%%%
+%% DDA Meierhofer
+\multirow{3}{*}{\rotatebox[origin=c]{90}{Biological}} \multirow{3}{*}{\rotatebox[origin=c]{90}{Investigation}} & Human - Ubiquitination - TMT & Group & 6 & 2 & Suppl. Sec. \ref{sec:ipah} &  \\
+&&&&&& \\%[-0.05in]
+% \cline{2-9}
+% &&&&&&& \\[-0.05in]
+%%%% SRM Picotti 2009
+& Mouse - Phosphorylation - TMT & Time course & 6 & 2 & Suppl. Sec. \ref{sec:shigella} & \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=4878d777c6b34cf8aaf8477e93140c4d}{MSV000085565} \\
+&&&&&& \\%[-0.05in]
+% \cline{2-9}
+% &&&&&&& \\[-0.05in]
+%%%% SRM Surinova 2015, training
+& Human - Ubiquitination - Label-free & Group & 4 & 2 & Suppl. Sec. \ref{sec:usp30} &\\ [0.02in]
+ &&&&&& \\
+ \hline
+\end{tabular}\\
+\end{tiny}
+\caption{ \small {\bf Experimental datasets in this manuscript} 
+```Dataset'' is the code name of the dataset used in this manuscript.
+`Comparison'' is the type of experimental design (group comparison, or time course).
+``No. of Conditions'' states the number of conditions in the dataset.
+``No. of Bio. Replicates'' states the number of biological replicates per condition.
+Details regarding the experimental design for each dataset are available in the section listed in the ``Details in'' column.
+``Data availability'' column shows the MassIVE.quant ID or location on github. 
+}
+\label{tab:dataDescription}
+\end{table}
 
@@
 
-The simulation mimicked a label-free experiment with a high number of modified and unmodified peptide features and no missing values.
-Multiple datasets were generated with different numbers of replicates and conditions.
-In all simulations 1000 modified peptides were generated, half of which were true positives and half true negatives.
-Half of the true positives and true negatives were generated convoluted with global protein abundance.
-Before adjustment, the convoluted true positives should appear to be negatives, while the convoluted true negatives would appear differential.
-Modified peptides and unmodified proteins were both simulated with 10 features.
+To evaluate the proposed method in situations with biological variation and known ground truth we created synthetic datasets which mimicked a label-free experiment with a high number of modified and unmodified peptide features and no missing values.
+Multiple simulations were ran with differing numbers of replicates and conditions per simulation.
+This was done to see how the method performed in different experimental design scenarios.
+The exact numbers of replicates and conditions can be seen in Table \ref{tab:dataDescription}.
+
+In each simulation, 1000 PTMs and 1000 unmodified proteins were generated.
+Each PTM and protein was created with 10 corresponding peptide features.
+500 of the PTMs were generated with a differential fold change between conditions, while the other half were generated with no difference between conditions.
+Of the 500 differential PTMs, the fold change of 250 were fully masked by changes in the unmodified protein.
+Of the 500 non-differential PTMs, 250 were generated with a differential fold change that was entirely due to the unmodified protein.
+All differential PTMs were generated with an expected log fold change of .75.
+
+Finally, each simulation was generated with random biological variation.
+Specifically, let $Y_{ibc}$ denote the abundance of peptide $i$ in biological replicate $b$ of condition $c$.
+The simulated peptide abundance  $Z_{ibc}$ was generated as $Z_{ibc} = Y_{ibc} + \epsilon_{ibc}$ where $\epsilon_{ibc}\sim^{iid}N(0,\sigma^2)$.
+The simulations were generated with two value $\sigma^2 = {.2, .3}$ motivated by the biological experiments in this manuscript.
+
+Further details of how the first simulation was generated, can be found in Supplementary Sec. \ref{supp-sec:sim}.
 

### 2022-02-21 supplementary/ptm_sm.tex
changes to intro

@@
 \section{Experimental datasets in this manuscript}
-
-Table \ref{tab:dataDescription} summarizes the datasets used in this manuscript and their experimental designs.
-There were 6 total experiments, 3 for which ground truth was known and 3 biological investigations.
-For the ground truth experiments, 2 were generated via computer simulation with differing parameters and the other was an experiment with heavy labeled spike-in peptides.
-The biological investigations represent a variety of experimental designs and acquisition methods.
-Two were acquired via Tandem Mass Tag (TMT) labeling, while the other was a label free experiment.
-Of the two TMT experiments, one was analyzed as a time series and the other a group comparison.
-
-Details of raw data, R scripts with MSstatsPTM analyses, and result of statistical analyses for the datasets are available in either MassIVE.quant or Github (with direct links available in the table).
-
-%%%% Summary of controlled mixtures
-\begin{table}[h!]
-\begin{tiny}
-\begin{tabular}{|c|c|cccc|c|}
-\hline
-%%%% colnames of table
-&  & \multicolumn{4}{c|}{Experimental design} & Data availability \\
-& Dataset & \multirow{2}{*}{Comparison} & No. of & No. of Bio.& \multirow{2}{*}{Details in}   & (MassIVE.quant  \\
-&  & & Conditions & Replicates &  & or R package) \\
-\hline
-\hline
- &&&&&& \\[-0.05in]
-%%%% DDA:ControlledMix
-\multirow{5}{*}{\rotatebox[origin=c]{90}{Known}}  \multirow{5}{*}{\rotatebox[origin=c]{90}{Ground}} \multirow{5}{*}{\rotatebox[origin=c]{90}{Truth}} & Computer simulation 1 - Label-free & Group & 2/3/4 & 2/3/5/10 & Suppl.
-Sec.
-\ref{sec:dataset1} & \href{https://github.com/devonjkohler/MSstatsPTM_simulations}{Github}\\
- &&&&&& \\%[-0.05in]
-%\hline 
-%&&&&&&&& \\[-0.05in]
-% \cline{2-9}
-% &&&&&&&& \\[-0.05in]
-%%% DDA iPRG2015
-& Computer simulation 2 - missing and low features & Group & 2/3/4 & 2/3/5/10 & Suppl. Sec. \ref{sec:dataset2} & \href{https://github.com/devonjkohler/MSstatsPTM_simulations}{Github} \\
- &&&&&& \\%[0.05in]
-%\hline 
-% &&&&&& \\[-0.05in]
-%%% DDA Cox2014
-& SpikeIn benchmark & Group & 4 & 2 & Suppl. Sec. \ref{sec:benchmark} & \\
- &&&&&& \\%[-0.05in]
-\hline
-\multicolumn{7}{c}{ } \\ [0.02in]
-\hline 
- &&&&&& \\%[-0.05in]
-%%%%%%%%%%%%%%%%%%
-%%%% biological study
-%%%%%%%%%%%%%%%%%%
-%% DDA Meierhofer
-\multirow{3}{*}{\rotatebox[origin=c]{90}{Biological}} \multirow{3}{*}{\rotatebox[origin=c]{90}{Investigation}} & Human - Ubiquitination - TMT & Group & 6 & 2 & Suppl. Sec. \ref{sec:ipah} &  \\
-&&&&&& \\%[-0.05in]
-% \cline{2-9}
-% &&&&&&& \\[-0.05in]
-%%%% SRM Picotti 2009
-& Mouse - Phosphorylation - TMT & Time course & 6 & 2 & Suppl. Sec. \ref{sec:shigella} & \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=4878d777c6b34cf8aaf8477e93140c4d}{MSV000085565} \\
-&&&&&& \\%[-0.05in]
-% \cline{2-9}
-% &&&&&&& \\[-0.05in]
-%%%% SRM Surinova 2015, training
-& Human - Ubiquitination - Label-free & Group & 4 & 2 & Suppl. Sec. \ref{sec:usp30} &\\ [0.02in]
- &&&&&& \\
- \hline
-\end{tabular}\\
-\end{tiny}
-\caption{ \small {\bf Experimental datasets in this manuscript} 
-```Dataset'' is the code name of the dataset used in this manuscript.
-`Comparison'' is the type of experimental design (group comparison, or time course).
-``No. of Conditions'' states the number of conditions in the dataset.
-``No. of Bio. Replicates'' states the number of biological replicates per condition.
-Details regarding the experimental design for each dataset are available in the section listed in the ``Details in'' column.
-``Data availability'' column shows the MassIVE.quant ID or location on github. 
-}
-\label{tab:dataDescription}
-\end{table}
 

### 2022-02-21 main/MSstatsPTM_Main.tex
main edits

@@
 The datasets were chosen to provide a performance benchmark of the proposed approach in situations with known ground truth, and to represent a variety of experimental designs and acquisition methods.
-The computer simulations varied in experimental realism.
+Two computer simulations were run which varied in experimental realism.
 The first simulation was perfectly clean, with many replicates and no missing values, while the second simulation introduced real world qualities, such as limited modified features and missing values.
@@
 A summary of the experiments used in this manuscript can be seen in Table \ref{tab:dataDescription}.
-Details of raw data, R scripts with MSstatsPTM analyses, and result of statistical analyses for the datasets are available in either MassIVE.quant or Github (with direct links available in the table).
+Details of raw data, R scripts with $MSstatsPTM$ analyses, and result of statistical analyses for the datasets are available in either MassIVE.quant or Github (with direct links available in the table).
 
@@
 Details regarding the experimental design for each dataset are available in the section listed in the ``Details in'' column.
-``Data availability'' column shows the MassIVE.quant ID or location on github. 
+``Data availability'' column shows the MassIVE.quant ID or github repository. 
 }
@@
 
-To evaluate the proposed method in situations with biological variation and known ground truth we created synthetic datasets which mimicked a label-free experiment with a high number of modified and unmodified peptide features and no missing values.
-Multiple simulations were ran with differing numbers of replicates and conditions per simulation.
+To evaluate the proposed method in situations with biological variation and known ground truth we created synthetic datasets which mimicked a label-free experiment; with a high number of modified peptide features and no missing values.
+Multiple simulations were generated with differing numbers of replicates and conditions per simulation.
 This was done to see how the method performed in different experimental design scenarios.
@@
 All differential PTMs were generated with an expected log fold change of .75.
-
-Finally, each simulation was generated with random biological variation.
+Each simulation was generated with random biological variation.
 Specifically, let $Y_{ibc}$ denote the abundance of peptide $i$ in biological replicate $b$ of condition $c$.
 The simulated peptide abundance  $Z_{ibc}$ was generated as $Z_{ibc} = Y_{ibc} + \epsilon_{ibc}$ where $\epsilon_{ibc}\sim^{iid}N(0,\sigma^2)$.
-The simulations were generated with two value $\sigma^2 = {.2, .3}$ motivated by the biological experiments in this manuscript.
+The simulations were generated with one of two values $\sigma^2 = \{.2, .3\}$ motivated by the biological experiments in this manuscript.
 
-Further details of how the first simulation was generated, can be found in Supplementary Sec. \ref{supp-sec:sim}.
+Further details of how the first simulation was generated, can be found in Supplementary Sec. \ref{supp-sec:dataset1}.
 
@@
 
-In this simulation real world experimental conditions were introduced, including missing values and low number of modified features.
-The percentage of missing values and feature counts were determined by looking at the averages from the biological experiments in this paper.
-Modified peptides were simulated with 2 features, while unmodified proteins were still simulated with 10 features.
+In this simulation, data was generated using the same parameters as described in Section \ref{sec:comp_sim_procedure1} and including real world experimental conditions.
+Specifically these simulations included missing values and a low number of modified features.
+The percentage of missing values and feature counts were determined by looking at the biological experiments in this paper.
+PTMs were simulated with 2 modified peptide features, while unmodified proteins were still simulated with 10 features.
 Additionally, 20\% of observations for both modified and unmodified peptides were missing completely at random.
 Adding missing values and few representative modified features provides a more realistic expectation of model performance.
-Otherwise this simulation used the same parameters as in simulation 1, with 1000 modified peptides half of which were positives and the other half negative.
 
-Further details of how both simulations were generated, including exact parameters used, can be found in Supplementary Sec. \ref{supp-sec:sim}.
+Further details of how this simulation was generated, including exact parameters used, can be found in Supplementary Sec. \ref{supp-sec:dataset2}.
 
@@
 
-We evaluated our approach using a custom designed spike-in benchmark experiment with known ground truth, where 50 heavy-labeled KGG motif peptides from 20 human proteins were used as spike-in peptides.
+This custom designed experiment had 4 conditions with two biological replicates per condition and a balanced design.
+
+Experimental Design: We evaluated our approach using a custom designed spike-in benchmark experiment with known ground truth, where 50 heavy-labeled KGG motif peptides from 20 human proteins were used as spike-in peptides.
 Quantitative changes in protein and site abundance of these 20 proteins were the target of the benchmark.

### 2022-03-05 code/Spike-in_Analysis.Rmd
experimental design changes

@@
 ```{r pressure, echo=FALSE, fig.width = 10, fig.height = 7}
+load(file = "D:\\Northeastern\\Research\\MSstats\\PTM_Refractor\\Label_Free_PTM_Data\\benchmark-spike\\MSstatsPTM_Summarized.rda")
+
+summary_df$PTM$RunlevelData
+
 load(file = "../data/Spike_in_MSstatsPTM_Model.rda")#model_df

### 2022-03-05 main/MSstatsPTM_Main.tex
experimental design changes

@@
 
-The signaling mechanisms that allow cells to mount a dynamic and fast response to a multitude of events are facilitated by the modification of proteins at specific residues, acting as molecular on/off switches\cite{Deribe,Cohen}.
+The signaling mechanisms that allow cells to mount a dynamic and fast response to a multitude of events are facilitated by the modification of proteins at specific residues, acting as molecular on/off switches \cite{Deribe,Cohen}.
 Bottom-up liquid chromatography coupled with mass spectrometry (LC-MS) is broadly established as the tool-of-choice for unbiased and large-scale identification and quantification of proteins and their post-translational modifications (PTMs) \cite{Kall:2011ub,Roepstorff}.
-Studies targeting the post-translationally modified proteome focus either on the accurate localization of modification sites on proteins, relative or absolute quantification of a modification site's occupancy repertoire, or relative changes in occupancy across experimental conditions\cite{Mann}.
+Studies targeting the post-translationally modified proteome focus either on the accurate localization of modification sites on proteins, relative or absolute quantification of a modification site's occupancy repertoire, or relative changes in occupancy across experimental conditions \cite{Mann}.
 Some scientific questions,  such as discovery of protein biomarkers of a disease, consider meaningful patterns of changes in PTM abundance alone \cite{Petushkova_2017}.
@@
 
-Currently statistical analyses of these experiments are performed with methods such as t-test and Limma\cite{Ritchie_15a}.
+Currently statistical analyses of these experiments are performed with methods such as t-test and Limma \cite{Ritchie_15a}.
 These approaches generally use the intensity ratios of modified and unmodified peptide features and compare the mean abundance of different PTM sites.
@@
 The proposed framework summarizes the abundances of peptides with and without the modification sites, and fits a linear mixed effects model that automatically adjusts to the biological and technological aspects of the experimental design.
-Finally changes in the PTM and protein-level abundances are combined to remove bias due to the confounding of these two sources.
+Finally, changes in the PTM and protein-level abundances are combined to remove bias due to the confounding of these two sources.
 
@@
 
-The proposed approach is implemented as a freely available open source R package $MSstatsPTM$, available on Bioconductor, which employs similar input format as in $MSstats$ and $MSstatsTMT$, and is compatible with many acquisition methods, such as label free, DIA, DDA, and TMT\cite{Choi:2014,Huang:2020}.
+The proposed approach is implemented as a freely available open source R package $MSstatsPTM$, available on Bioconductor, which employs similar input format as in $MSstats$ and $MSstatsTMT$, and is compatible with many acquisition methods, such as label free, DIA, DDA, and TMT \cite{Choi:2014,Huang:2020}.
 
@@
 A summary of the experiments used in this manuscript can be seen in Table \ref{tab:dataDescription}.
-Details of raw data, R scripts with $MSstatsPTM$ analyses, and result of statistical analyses for the datasets are available in either MassIVE.quant or Github (with direct links available in the table).
+Details of the raw data, R scripts with $MSstatsPTM$ analysis, and the results of the statistical analysis are available in either MassIVE.quant or Github (with direct links available in the table).
 
@@
 \begin{tiny}
-\begin{tabular}{|c|c|cccc|c|}
+\begin{tabular}{|c|c|ccc|c|}
 \hline
 %%%% colnames of table
-&  & \multicolumn{4}{c|}{Experimental design} & Data availability \\
-& Dataset & \multirow{2}{*}{Comparison} & No. of & No. of Bio.& \multirow{2}{*}{Details in} & (MassIVE.quant  \\
-&  & & Conditions & Replicates &  & or R package) \\
+&  & \multicolumn{3}{c|}{Experimental design} & Data availability \\
+& Dataset & \multirow{2}{*}{Comparison} & No. of & No. of Bio. & (MassIVE.quant  \\
+&  & & Conditions & Replicates & or R package) \\
 \hline
 \hline
- &&&&&& \\[-0.05in]
+ &&&&& \\[-0.05in]
 %%%% DDA:ControlledMix
-\multirow{5}{*}{\rotatebox[origin=c]{90}{Known}}  \multirow{5}{*}{\rotatebox[origin=c]{90}{Ground}} \multirow{5}{*}{\rotatebox[origin=c]{90}{Truth}} & Computer simulation 1 - Label-free & Group & 2/3/4 & 2/3/5/10 & Suppl.
-Sec.
-\ref{sec:dataset1} & \href{https://github.com/devonjkohler/MSstatsPTM_simulations}{Github}\\
- &&&&&& \\%[-0.05in]
+\multirow{5}{*}{\rotatebox[origin=c]{90}{Known}}  \multirow{5}{*}{\rotatebox[origin=c]{90}{Ground}} \multirow{5}{*}{\rotatebox[origin=c]{90}{Truth}} & Computer simulation 1 - Label-free & Group & 2/3/4 & 2/3/5/10 & \href{https://github.com/devonjkohler/MSstatsPTM_simulations}{Github}\\
+ &&&&& \\%[-0.05in]
 %\hline 
@@
 %%% DDA iPRG2015
-& Computer simulation 2 - missing and low features & Group & 2/3/4 & 2/3/5/10 & Suppl. Sec. \ref{sec:dataset2} & \href{https://github.com/devonjkohler/MSstatsPTM_simulations}{Github} \\
- &&&&&& \\%[0.05in]
+& Computer simulation 2 - missing and low features & Group & 2/3/4 & 2/3/5/10 & \href{https://github.com/devonjkohler/MSstatsPTM_simulations}{Github} \\
+ &&&&& \\%[0.05in]
 %\hline 
@@
 %%% DDA Cox2014
-& SpikeIn benchmark & Group & 4 & 2 & Suppl. Sec. \ref{sec:benchmark} & \\
- &&&&&& \\%[-0.05in]
+& SpikeIn benchmark - Ubiquitination - Label-free & Group & 4 & 2 & \href{https://massive.ucsd.edu/ProteoSAFe/private-dataset.jsp?task=c4c583ecf7f941cdac87f7a4f872517b}{MSV000088971}\\
+ &&&&& \\%[-0.05in]
 \hline
-\multicolumn{7}{c}{ } \\ [0.02in]
+\multicolumn{6lm}{c}{ } \\ [0.02in]
 \hline 
- &&&&&& \\%[-0.05in]
+ &&&&& \\%[-0.05in]
 %%%%%%%%%%%%%%%%%%
@@
 %% DDA Meierhofer
-\multirow{3}{*}{\rotatebox[origin=c]{90}{Biological}} \multirow{3}{*}{\rotatebox[origin=c]{90}{Investigation}} & Human - Ubiquitination - TMT & Group & 6 & 2 & Suppl. Sec. \ref{sec:ipah} &  \\
-&&&&&& \\%[-0.05in]
+\multirow{3}{*}{\rotatebox[origin=c]{90}{Biological}} \multirow{3}{*}{\rotatebox[origin=c]{90}{Investigation}} & Human - Ubiquitination - 1mix-TMT & Group & 6 & 2 & \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=b6f0c74c234247678fb0888c6df1f225}{MSV000088966} \\
+&&&&& \\%[-0.05in]
 % \cline{2-9}
@@
 %%%% SRM Picotti 2009
-& Mouse - Phosphorylation - TMT & Time course & 6 & 2 & Suppl. Sec. \ref{sec:shigella} & \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=4878d777c6b34cf8aaf8477e93140c4d}{MSV000085565} \\
-&&&&&& \\%[-0.05in]
+& Mouse - Phosphorylation - 2mix-TMT & Time course & 6 & 4 & \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=4878d777c6b34cf8aaf8477e93140c4d}{MSV000085565} \\
+&&&&& \\%[-0.05in]
 % \cline{2-9}
@@
 %%%% SRM Surinova 2015, training
-& Human - Ubiquitination - Label-free & Group & 4 & 2 & Suppl. Sec. \ref{sec:usp30} &\\ [0.02in]
- &&&&&& \\
+& Human - Ubiquitination - Label-free & Group & 4 & 2 & \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=1b516164de5345108b40b75147dd58b5}{MSV000078977}\\ [0.02in]
+ &&&&& \\
  \hline
@@
 ``No. of Bio. Replicates'' states the number of biological replicates per condition.
-Details regarding the experimental design for each dataset are available in the section listed in the ``Details in'' column.
 ``Data availability'' column shows the MassIVE.quant ID or github repository. 
@@
 
-To evaluate the proposed method in situations with biological variation and known ground truth we created synthetic datasets which mimicked a label-free experiment; with a high number of modified peptide features and no missing values.
+To evaluate the proposed method in situations with biological variation and known ground truth we created synthetic datasets which mimicked a label-free experiment.
 Multiple simulations were generated with differing numbers of replicates and conditions per simulation.
+All simulations were created with a high number of modified peptide features and no missing values.
 This was done to see how the method performed in different experimental design scenarios.
@@
 In each simulation, 1000 PTMs and 1000 unmodified proteins were generated.
-Each PTM and protein was created with 10 corresponding peptide features.
+Each PTM and protein was created with 10 corresponding features.
 500 of the PTMs were generated with a differential fold change between conditions, while the other half were generated with no difference between conditions.
@@
 Of the 500 non-differential PTMs, 250 were generated with a differential fold change that was entirely due to the unmodified protein.
-All differential PTMs were generated with an expected log fold change of .75.
+All differential PTMs were generated with an expected log fold change of .75 between conditions.
 Each simulation was generated with random biological variation.
@@
 
-In this simulation, data was generated using the same parameters as described in Section \ref{sec:comp_sim_procedure1} and including real world experimental conditions.
+In this simulation, data was generated using the same parameters as described in the previous section and including real world experimental conditions.
 Specifically these simulations included missing values and a low number of modified features.
@@
 \subsection*{Dataset 3 : SpikeIn benchmark - Ubiquitination - Label-free}
+\label{sec:exp_proc_dataset3}
+We evaluated our approach using a custom designed spike-in benchmark experiment with known ground truth.
+This experiment had 4 conditions with two biological replicates per condition and a balanced design.
 
-This custom designed experiment had 4 conditions with two biological replicates per condition and a balanced design.
-
-Experimental Design: We evaluated our approach using a custom designed spike-in benchmark experiment with known ground truth, where 50 heavy-labeled KGG motif peptides from 20 human proteins were used as spike-in peptides.
+Experimental Design: In this experiment 50 heavy-labeled KGG motif peptides from 20 human proteins were used as spike-in peptides.
 Quantitative changes in protein and site abundance of these 20 proteins were the target of the benchmark.
@@
 Two sets of data were acquired for each mixture: KGG enriched + LC-MS, and LC-MS only.
-The four mixtures were compared with known fold changes for the spike-in peptides.
 
-More information on the experimental design of the benchmark experiment can be found in Supplementary Sec. \ref{supp-sec:benchmark}.
+Global Profiling Run: This experiment included a separate global profiling run to measure unmodified peptides.
+There was a 90.2\% overlap between the identified background modified peptides and proteins that were quantified in the global profiling run.
+
+Pairwise Comparisons: We evaluated the ability of the methods to detect known changes in abundance between pairs of conditions.
+The four mixtures were treated as individual conditions and labeled mix1, mix2, mix3, and mix4.
+The pairwise comparisons were mix1-mix2, mix1-mix3, mix1-mix4, mix2-mix3, mix2-mix4, and mix3-mix4.
+The true log fold changes of spike-in modified peptides in these comparisons can be seen in Table \ref{table:spikein_fold_change}.
+
+\begin{table}[h!]
+\centering
+\begin{tabular}{| c | c |}
+\hline
+ Comparison & Expected log fold change\\ [0.5ex]
+ \hline\hline
+ mix1-mix2 & -1\\
+ \hline
+ mix1-mix3 & 1\\
+\hline
+ mix1-mix4 & 0\\
+\hline
+ mix2-mix3 & -2\\
+\hline
+ mix2-mix4 & -1\\
+\hline
+ mix3-mix4 & 1\\
+\hline
+\end{tabular}
+\caption{The expected log fold change of modified spike-in peptides in Dataset 3.}
+\label{table:spikein_fold_change}
+\end{table}
+
+More information on the experimental design of the benchmark experiment and how the expected fold change was calculated can be found in Supplementary Sec. \ref{supp-sec:benchmark}.
 
 \subsection*{Dataset 4 : Human - Ubiquitination - 1mix-TMT}
+\label{sec:exp_proc_dataset4}
+The proposed approach was evaluated on an experiment with six conditions and an unbalanced design.
 
-The proposed approach was evaluated on an experiment where Shigella ubiquitin ligase IpaH7.8 was shown to degrade the protein gasdermin D (GSDMD)\cite{LUCHETTI2021}.
-Multiplex proteomics was used to quantify the abundance of total protein and ubiquitination in human epithelial cells.
+Experimental Design: In this study Luchetti et al.
+\cite{LUCHETTI2021} used multiplex proteomics to quantify the abundance of total protein and ubiquitination in human epithelial cells.
 Cells were engineered to express IpaH7.8 under a dox inducible promoter and measurements were taken at different time periods.
@@
 Uninfected cells were measured at 0 and 6 hours, while infected cells were measured at 1, 2, 4, and 6 hour increments, resulting in six total conditions.
-The experiment was unbalanced with two bioreplicates per condition for all conditions except for infected 1 hour.
-About ~95\% of the identified modified peptides derived from proteins that were quantified in the global profiling run.
+
+Global Profiling Run: This experiment included a separate global profiling run to measure unmodified peptides.
+There was a 95\% overlap between the identified modified peptides and proteins that were quantified in the global profiling run.
+
+Pairwise comparisons: We evaluated the statistical approach by its ability to detect changes in the abundance of modified peptides both before and after adjusting for changes in global protein abundance.
+The six condition were labeled Dox1hr, Dox2hr, Dox4hr, Dox6hr, NoDox0hr, and NoDox6hr.
+All conditions were compared with each other in a full pairwise comparison, resulting in 15 comparisons.
+Because the dataset is a biological investigation, the true positive modifications were unknown.
 
@@
 \subsection*{Dataset 5 : Mouse - Phosphorylation - 2mix-TMT time series}
+\label{sec:exp_proc_dataset5}
+In this experiment 8 biological replicates were allocated to 2 TMT mixtures in an unbalanced design.
 
-Here an experiment targeting primary murine macrophages infected with Shigella flexneri (S.flexneri)\cite{Maculins}.
+Experimental Design: Maculins et al.
+\cite{Maculins} designed an experiment targeting primary murine macrophages infected with Shigella flexneri (S.flexneri).
 Multiplex proteomics was used to quantify the abundance of total protein, and phosphorylation in wild type (WT) and ATG16L1-deficient (cKO) samples, uninfected and uninfected with S.flexneri.
 The abundance of total protein and post-translation modifications were quantified at three time points, uninfected, early infection (45-60 minutes), and late infection (3-3.5 hours).
-Quantifying the total protein along with the post-translational modifications allowed us to adjust for changes in total protein and see the true impact of the site specific modifications.
-Two mixtures using 11-plex were ran over the six conditions.
-The six conditions were split between 11 channels leading to the experimental design being unbalanced.
+11-plex isobaric multiplexing via tandem mass tagging (TMT) in combination with liquid-chromatography and tandem mass spectrometry (LC-MS/MS) was used.
+Six conditions were split between 11 channels leading to the experimental design being unbalanced.
 Each mixture contained two replicates per early and late WT and KO conditions.
@@
 Mixture two contained one replicate of uninfected KO and two uninfected WT.
-About ~90\% of the identified modified peptides derived from proteins that were also quantified in the global profiling run.
+
+Global Profiling Run: This experiment included a separate global profiling run to measure unmodified peptides.
+There was a 90\% overlap between the identified background modified peptides and proteins that were quantified in the global profiling run.
+
+Time series comparison: We evaluated the statistical approach by its ability to detect changes in the abundance of modified peptides both before and after adjusting for changes in global protein abundance.
+The six condition were labeled KO\_Uninfect, KO\_Early, KO\_Late, WT\_Uninfect, WT\_Early, and WT\_Late.
+9 total comparisons were made named KO\_Early-WT\_Early, KO\_Late-WT\_Late, KO\_Uninfected-WT\_Uninfected, KO\_Early-KO\_Uninfected, KO\_Late-KO\_Uninfected, WT\_Early-WT\_Uninfected, WT\_Late-WT\_Uninfected, Infected-Uninfected, and KO-WT.
+Because the dataset is a biological investigation, the true positive modifications were unknown.
 
@@
 \subsection*{Dataset 6 : Human - Ubiquitination - Label-free no global profiling run}
+\label{sec:exp_proc_dataset6}
 
-This experiment looked into the relationship between USP30 and protein kinase PINK1, and their association with Parkinson’s Disease.
+In this experiment 8 biological replicates were split between 4 conditions in a balanced design.
+
+Experimental Design: Cunningham et al.
+\cite{Cunningham2015} investigated the relationship between USP30 and protein kinase PINK1, and their association with Parkinson’s Disease.
 Ubiquitination site profiling was performed and the modified site abundance was analyzed.
@@
 Label-free mass spectrometry quantification was used to quantify the abundance of modified peptides.
-In contrast to the other biological experiments assessed in this paper, this experiment did not not have a separate global profiling run for the unmodified protein.
-In this case $MSstatsPTM$ can still be used by extracting unmodified peptides from the modified run, however this generally leads to substantially less matches between modified and unmodified peptides, in addition to low feature counts for unmodified peptides.
-In this case there only 41.9\% of the modified peptides had a corresponding unmodified protein to perform adjustment.
-In cases where there is no estimate for the matching unmodified protein, adjustment cannot be performed.
+
+Global Profiling Run: This experiment did not included a separate global profiling run to measure unmodified peptides.
+In this case $MSstatsPTM$ can still be used by extracting unmodified peptides from the modified run, however this leads to substantially less matches between modified and unmodified peptides, in addition to low feature counts for unmodified peptides.
+There was a 41.9\% overlap between the identified background modified peptides and proteins that were quantified in the global profiling run.
+
+Pairwise comparisons: We evaluated the statistical approach by its ability to detect changes in the abundance of modified peptides both before and after adjusting for changes in global protein abundance.
+The four condition were labeled CCCP, Combo, Ctrl, and USP30\_OE.
+All conditions were compared with each other in a full pairwise comparison, resulting in 6 comparisons.
+Because the dataset is a biological investigation, the true positive modifications were unknown.
 
@@
 The approach takes as input intensities of individual features from modified peptides, or intensity ratios of modified and unmodified peptide features, and compares the mean abundance of a PTM site from one condition to another.
-Modifications of the t-test such as moderated t-test with Limma were also proposed\cite{Zhu}.
+Modifications of the t-test such as moderated t-test with Limma were also proposed \cite{Zhu}.
 While simple, the approach does not fully account for the sources of variations, and it is not directly applicable to experiments with complex designs, e.g., comparisons of multiple conditions, acquisition in multiple batches, etc.
@@
 For each PTM site, the feature log-intensities of the modified peptides spanning the site are expressed using a linear mixed model in consideration of the effects of condition, run, feature and interaction between run and feature.
-The model parameters are estimated using the split-plot approach as in $MSstats$, where the feature log-intensities are first summarized into a single value per site per run in the subplot model, and the site-level summaries are then used for the inference of the PTM site abundance\cite{Choi:2014}.
+The model parameters are estimated using the split-plot approach as in $MSstats$, where the feature log-intensities are first summarized into a single value per site per run in the subplot model, and the site-level summaries are then used for the inference of the PTM site abundance \cite{Choi:2014}.
 This approach allows the method to be extended to cases where the experimental design is unbalanced, and where additional sources of variation are present.
-In the site-level summarization, Tukey's median polish (TMP), a simple and robust procedure is applied to iteratively fit a two-way additive model with the effects of run and feature, which in turn summarizes the log-intensities for each site\cite{Tukey:1977}.
-After summarization, the inference of the PTM site abundance in each condition is carried through fitting a model based on the family of linear mixed-effects models\cite{Bolker2009} \cite{Faraway:2006}.
+In the site-level summarization, Tukey's median polish (TMP), a simple and robust procedure is applied to iteratively fit a two-way additive model with the effects of run and feature, which in turn summarizes the log-intensities for each site \cite{Tukey:1977}.
+After summarization, the inference of the PTM site abundance in each condition is carried through fitting a model based on the family of linear mixed-effects models \cite{Bolker2009} \cite{Faraway:2006}.
 Statistical modeling and quantification for global proteomics data are performed by the same procedure as for PTM data.
@@
 When values are imputed, it is assumed they are missing for reasons of low abundance.
-Missing features are imputed in each MS run using the Accelerated Time Failure (AFT) model\cite{Tukey:1977}.
+Missing features are imputed in each MS run using the Accelerated Time Failure (AFT) model \cite{Tukey:1977}.
 In order to impute a feature's missing values, the feature must be present in at least one MS Run.
@@
 The statistical modeling approaches discussed above can also be extended to complex designs including experiments with additional sources of variation and unbalanced designs.
-Equation (1) is the same in this case and the null hypothesis is still that there is no difference in mean PTM abundance between Conditions $i$ and $i'$ after adjusting for changes in mean unmodified peptide abundance between Conditions $i$ and $i'$.
+Equation 1 is the same in this case and the null hypothesis is still that there is no difference in mean PTM abundance between Conditions $i$ and $i'$ after adjusting for changes in mean unmodified peptide abundance between Conditions $i$ and $i'$.
 
@@
 In cases where the design is unbalanced, a linear mixed effects model is fit, which takes into account all potential sources of variation, and restricted maximum likelihood (RML) is used to estimate the parameters of each model.
-Once the parameters are estimated, we can combine the models using modified versions of the formulas in Section \ref{sec:adjustment_summary}.
+Once the parameters are estimated, we can combine the models using modified versions of Equations 2, 3, and 4.
 
-The log-fold change in the adjusted PTM abundance, $\Delta$, is now estimated by RML, and Equation (2) becomes 
+The log-fold change in the adjusted PTM abundance, $\Delta$, is now estimated by RML, and Equation 2 becomes 
 \begin{equation}
@@
 
-The standard error of the estimate $\mathrm{SE}(\hat{\Delta})$ and degrees of freedom in Equations (3) and (4) are unchanged from the balanced case, although $\hat{\sigma}_{\gamma^z\ast}$ and $\hat{\sigma}_{\gamma}$ are now estimated using RML.
+The standard error of the estimate $\mathrm{SE}(\hat{\Delta})$ and degrees of freedom in Equations 3 and 4 are unchanged from the balanced case, although $\hat{\sigma}_{\gamma^z\ast}$ and $\hat{\sigma}_{\gamma}$ are now estimated using RML.
 
@@
 
-The package relies on functionalities from the R packages $MSstats$ \cite{Choi:2014} and $MSstatsTMT$ \cite{Huang:2020}, depending on the data acquisition type.
+The package relies on functionalities from the R packages $MSstats$ \cite{Choi:2014} and $MSstatsTMT$ \cite{Huang:2020}.
 The statistical modeling relies on the functionality from the R packages $lme4$ \cite{Bates2015} and $lmerTest$ \cite{Kuznetsova2017}.
@@
 
-\todo{Check that this section uses past tense; look out for informal expressions}
-
 The performance of the proposed method was evaluated on simulated and spike in datasets with known ground truth, as well as biological experimental data where the ground truth was not known.
-For the experiments where ground truth is known, we calculated the number of true positives (TP), false positives (FP), true negatives (TN), and false negatives (FN).
-True positives are modified peptides that are differential after accounting for changes in global protein abundance.
+For the experiments where ground truth was known, we calculated the number of true positives (TP), false positives (FP), true negatives (TN), and false negatives (FN).
+True positives were modified peptides that are differential after accounting for changes in global protein abundance.
 True negatives were modified peptides that were not differentially abundant after accounting for global protein abundance.
@@
 Comparisons were made before and after adjusting for changes in unmodified protein abundance and differences between the comparisons were noted.
-Each biological dataset presents a different experimental scenario, showing the applicability of the proposed method on a variety of experimental designs.
+Each biological dataset presented a different experimental scenario, showing the applicability of the proposed method on a variety of experimental designs.
 All datasets in this paper were evaluated using the $MSstatsPTM$ package.
@@
 The proposed approach was evaluated against two-sample t-test and Limma in computer simulations. 
-In Figure \ref{fig:computer_sim} we see that not accounting for changes in unmodified peptide level results in a high false discovery rate.
+In Figure \ref{fig:sim1_fdr} we see that not accounting for changes in unmodified peptide level results in a high false discovery rate.
 When adjusting for changes in the unmodified peptide, all methods performed similarly in terms of FDR.
@@
 Changes in unmodified peptide level still needed to be accounted for to control the FDR.
-Once controlling for changes in the unmodified peptide, the proposed method out performed the other methods, as seen in Figure \ref{fig:computer_sim}.
+Once controlling for changes in the unmodified peptide, the proposed method out performed the other methods, as seen in Figure \ref{fig:sim2_acc}.
 The proposed method calibrated model accuracy and recall rate well, even when the number of replicates were low.
@@
 In this experiment all models incorrectly estimated the fold change of the modified spike-in peptides before adjusting for changes in global protein abundance.
-After adjustment, the spike-in peptides' fold change was in line with expectation.
-This can be seen in Figure \ref{fig:spikein_boxplot}.
+After adjustment, the spike-in peptides' fold change was in line with expectation (Figure \ref{fig:spikein_boxplot}).
 While the median fold change for the spike-in peptides was generally in line with the true fold change after adjustment, we can see that the distribution of fold changes for the peptides was visibly wider after adjustment.
@@
 In Figure \ref{fig:spikein_prop_volcano} we can clearly see the red labeled spike-in peptides do not follow the expected fold change before adjusting for changes in global protein level.
-After adjustment the estimated fold change is more in line with expectation.
-Additionally, the background peptides serving as the null model, show many false positives before adjustment is made.
-After adjustment the results improve and the number of false positives decrease significantly.
-Specifically, for the proposed method, the number of false positives goes from 20.88\% to 1.84\% of all true negatives after adjustment is applied.
-Using both the proposed method and Limma, the red labeled spike-in peptides are more inline with expected fold change after the adjustment is applied.
-However, when using Limma we can see that the majority of the differential spike-in peptides are not showing a significant adjusted pvalue.
+After adjustment the estimated fold change was more in line with expectation.
+Additionally, the background peptides serving as the null model, show many false positives before adjustment was made.
+After adjustment the results improved and the number of false positives decreased significantly.
+Specifically, for the proposed method, the number of false positives went from 20.88\% to 1.84\% of all true negatives after adjustment was applied.
+Using both the proposed method and Limma, the red labeled spike-in peptides were more inline with expected fold change after the adjustment was applied.
+However, when using Limma we can see that the majority of the differential spike-in peptides were not showing a significant adjusted pvalue.
 In this case, using Limma would have resulted in missing the majority of differentially modified peptides.
@@
 In Figure \ref{fig:data4_venn_diagram} the number of significant modified peptides before and after adjustment is shown.
-We can see that more peptides become insignificant after adjustment than become significant.
-3,236 peptides become insignificant, 1352 peptides become significant, while 4,282 peptides are significant in both models.
+We can see that more peptides became insignificant after adjustment than became significant.
+3,236 peptides became insignificant, 1352 peptides became significant, while 4,282 peptides were significant in both models.
 For the peptides that became insignificant in the adjusted model, their change in abundance was mainly due to changes in global protein abundance.
@@
 Looking at the Dox4hr vs Dox1hr conditions and modeling the modified peptide without adjusting for changes in the global profiling, the fold change was $-.501$ and the adjusted pvalue was insignificant at $.0644$.
-After adjusting for changes in the global protein abundance, the fold change is much higher, $2.79$, and the adjusted pvalue became significant, $5.25e-8$.
-In this case the effect of the modified peptide is strongly confounded with changes in the global protein.
-The proposed method allows us to remove this confounding and estimate the true effect.
+After adjusting for changes in the global protein abundance, the fold change was much higher, $2.79$, and the adjusted pvalue became significant, $5.25e-8$.
+In this case the effect of the modified peptide was strongly confounded with changes in the global protein.
+The proposed method allowed us to remove this confounding and estimate the true effect.
 
@@
 This indicates that the abundance change in the modified peptide is nearly entirely due to changes in the global protein.
-When adjusting for the global protein the modified peptide's adjusted pvalue becomes insignificant, going from $.0009$ to $.248$.
+When adjusting for the global protein the modified peptide's adjusted pvalue became insignificant, going from $.0009$ to $.248$.
 Correcting for changes in global protein abundance allowed us to see the true impact of the modification at $S178$ which would have otherwise been challenging to perceive.
@@
 
-\subsubsection*{In label-free experiment without a separate global profiling run, the proposed approach could eliminate the bias \todo{due to what?}, albeit less effectively than \todo{in presence of a global profiling run?}}
+\subsubsection*{In label-free experiment without a separate global profiling run, the proposed approach could eliminate the bias due to changes in the unmodified protein, albeit less effectively than in the presence of a global profiling run}
 
-As discussed previously, there was no unmodified global profiling run performed in this experiment.
+As discussed in Section \ref{sec:exp_proc_dataset6}, there was no unmodified global profiling run performed in this experiment.
 Once identification and quantification of the Ubiquitination profiling was performed, peptides which were unmodified were extracted and used in place of a global profiling run.
@@
 The results of this experiment are summarized in Figure \ref{fig:data6_plots}.
-In \ref{fig:data6_vd1} the number of significant modified peptides before and after protein adjustment is shown.
-In the top plot we can see there are a large number of peptides that become insignificant after adjustment.
-This is mainly due to the lack of overlap between modified and unmodified peptides, stemming from not having a global profiling run.
+In \ref{fig:data6_vd1} the number of significant modified peptides before and after protein adjustment are shown.
+In the top plot we can see there were a large number of peptides that become insignificant after adjustment.
+This was mainly due to the lack of overlap between modified and unmodified peptides, stemming from not having a global profiling run.
 In the bottom plot only modified peptides that could be adjusted are shown.
-Here there are much less peptides that become insignificant after adjustment.
+Here there were much less peptides that become insignificant after adjustment.
 In this plot we see that 726 peptides became insignificant, 547 became significant, and 1,078 are significant in both models.
 As in the previous experiments, we can check if these peptides are becoming insignificant due to an increase in variance.
-To check this we look at modified peptides whose adjusted fold change is within 10\% of their unadjusted fold change and they become insignificant after adjustment.
-In this experiment there are only 25 peptides that meet this criteria, a small percentage of all modified peptides.
+To check this we look at modified peptides whose adjusted fold change was within 10\% of their unadjusted fold change and they became insignificant after adjustment.
+In this experiment there were only 25 peptides that meet this criteria, a small percentage of all modified peptides.
 
@@
 
+\todo{Add massive and github links}
+
 Shigella: https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=4878d777c6b34cf8aaf8477e93140c4d
@@
 \caption{}
+\label{fig:sim1_fdr}
 \end{subfigure}
@@
 \caption{}
+\label{fig:sim2_acc}
 \end{subfigure}

### 2022-03-05 supplementary/ptm_sm.tex
experimental design changes

@@
 
-\title{Statistical methods for relative quantification of post-translational modifications in global proteomics experiments}
-\author[1]{Tsung-Heng~Tsai}
-\author[2]{Erik~Verschueren}
-\author[1]{Olga~Vitek}
-\affil[1]{Northeastern University, Boston, MA, USA}
-\affil[2]{Protein Chemistry, Genentech, South San Francisco, CA, USA}
+\title{MSstatsPTM statistical relative quantification of post-translational modifications in global proteomics experiments
+
+\textbf{Supplementary Information}
+}
+
+\author[1]{Devon~Kohler}
+\author[2]{Tsung-Heng~Tsai}
+\author[1]{Ting~Huang}
+\author[4]{Erik~Verschueren}
+\author[3]{Trent~Hinkle}
+\author[3]{Meena~Choi*}
+\author[1]{Olga~Vitek*}
+\affil[1]{Khoury College of Computer Science, Northeastern University, Boston, MA, USA}
+\affil[2]{Kent State University, Kent, OH, USA}
+\affil[3]{MPL, Genentech, South San Francisco, CA, USA}
+\affil[4]{ULUA BV, Arendstraat 29, 2018 Antwerp, Belgium}
+\affil[*]{Corresponding Authors}
 
 \begin{document}
-%\maketitle
-
-%\newpage
+\maketitle
+
+\newpage
 \tableofcontents
@@
 
-This study addresses three major goals in the characterization of post-translational modifications (PTMs): a) relative PTM quantification, b) PTM significance analysis, i.e., to detect PTM sites that are differentially modified across experimental conditions, and c) statistical design of PTM experiments.
+This study addresses three major goals in the characterization of post-translational modifications (PTMs): a) relative PTM quantification, b) PTM significance analysis, i.e., to detect PTM sites that are differentially modified across experimental conditions, and c) statistical design and analysis of PTM experiments.
 
@@
 
-The proposed statistical approach was evaluated and compared to the $t$-test and Limma methods using computer simulation.
+The proposed statistical approach was evaluated and compared to the $t$-test and Limma methods using computer simulations.
 Specifically, their statistical properties with respect to protein-level adjustment and real world experimental conditions were evaluated.
@@
 The proposed method summarizes feature intensities up to the run level using Tukey's Median Polish, while the other methods use the log of summed feature intensities.
-The results are summarized from Figure \ref{fig:spike_volcano_msstats} to Figure \ref{fig:spike_fc}, including volcano plots, model summary statistics, and fold change analysis.
+The results are summarized from Figure \ref{fig:spike_volcano_msstats} to Figure \ref{fig:spike_stat}, including volcano plots, model summary statistics, and fold change analysis.
 

### 2022-03-05 code/Shigella_Analysis.R

@@ -28,6 +28,7 @@ pST_summarization <- dataSummarizationPTM_TMT(pST_input)
 save(py_summarization, file = "../data/Shigella_pY_Summarization.rda")
 load(file = "../data/Shigella_pY_Summarization.rda")
 save(pST_summarization, file = "../data/Shigella_pST_Summarization.rda")
+load(file = "../data/Shigella_pST_Summarization.rda")
 
 py_summarized_data <- py_summarization$PTM$ProteinLevelData
 py_feature_data <- py_summarization$PTM$FeatureLevelData

### 2022-03-06 main/MSstatsPTM_Main.tex
Intro

@@
 
-\title{MSstatsPTM statistical relative quantification of post-translational modifications in global proteomics experiments}
+\title{MSstatsPTM statistical relative quantification of post-translational modifications in bottom-up proteomic experiments}
 
@@
 \maketitle
-%\noindent\author{Tsung-Heng Tsai$^{1}$, Olga Vitek$^{1,\ast}$}
-%
-%\vspace{0.2in}
-%\noindent $^1$   Khoury College of Computer Sciences, Northeastern University, Boston, MA, USA\\
-%\noindent $\ast$   Corresponding author \\
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Abstract}
+\begin{abstract}
 
-The scientific community increasingly utilizes mass spectrometry (MS)-based proteomics to detect changes in post-translational modifications (PTMs) between samples from various conditions.
-Experiments targeting PTMs face numerous challenges.
-These include the low abundance of modified proteoforms, few representative peptides that span modification sites, and confounding between changes in the abundance of PTM and the overall changes in the protein abundance.
-Therefore, statistical approaches to estimating relative changes in PTMs, and to detecting differential PTM abundance, should combine information pertaining to PTM sites over several peptides, replicates in multiple conditions, and consider relevant sources of confounding and variation.
-In this manuscript we propose such a general statistical framework, which is comprehensive, accurate, and leads to reproducible results.
-The framework requires an experimental design that, for each sample, quantifies the peptides with post-translational modifications, as well as  peptides from the same proteins with no modification sites, and is applicable to both label-free and tandem mass tag (TMT)-based acquisitions.
-The proposed framework summarizes the abundances of peptides with and without the modification sites, and fits a linear mixed effects model that automatically adjusts to the biological and technological aspects of the experimental design.
-Finally, changes in the PTM and the protein-level abundances are combined to remove bias due to the confounding of these two sources.
-Evaluations on computer simulations, a spike-in experiment with known ground truth, and three biological experiments which demonstrate the improved fold change estimation as compared to the currently used approaches for relative PTM quantification.
-The proposed framework is implemented in the free and open-source R package $MSstatsPTM$.
+Mass spectrometry (MS)-based proteomics is increasingly used to detect changes in post-translational modifications (PTMs) between samples from various conditions.
+Analysis of data from such experiments faces numerous statistical challenges.
+These include the low abundance of modified proteoforms, the small number of representative peptides that span modification sites, and confounding between changes in the abundance of PTM and the overall changes in the protein abundance.
+Therefore, statistical approaches for detecting differential PTM abundance must integrate all the available information pertaining to a PTM site, and consider all the relevant sources of confounding and variation.
+In this manuscript we propose such a statistical framework, which is comprehensive, accurate, and leads to reproducible results.
+The framework requires an experimental design, which for each sample quantifies both the peptides with post-translational modifications, and peptides from the same proteins with no modification sites.
+The framework supports both label-free and tandem mass tag (TMT)-based acquisitions.
+The statistical methodology separately summarizes the abundances of peptides with and without the modification sites, and fits separate linear mixed effects models that reflect the biological and technological aspects of the experimental design.
+Next, model-based inferences regarding the PTM and the protein-level abundances are combined to account for the confounding between these two sources.
+Evaluations on computer simulations, a spike-in experiment with known ground truth, and three biological experiments demonstrate the improved fold change estimation and detection of differential PTM abundance, as compared to currently used approaches.
+The proposed framework is implemented in the free and open-source R/Bioconductor package $MSstatsPTM$.
+
+\end{abstract}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\clearpage
 \section{Introduction}
 
-The signaling mechanisms that allow cells to mount a dynamic and fast response to a multitude of events are facilitated by the modification of proteins at specific residues, acting as molecular on/off switches \cite{Deribe,Cohen}.
-Bottom-up liquid chromatography coupled with mass spectrometry (LC-MS) is broadly established as the tool-of-choice for unbiased and large-scale identification and quantification of proteins and their post-translational modifications (PTMs) \cite{Kall:2011ub,Roepstorff}.
-Studies targeting the post-translationally modified proteome focus either on the accurate localization of modification sites on proteins, relative or absolute quantification of a modification site's occupancy repertoire, or relative changes in occupancy across experimental conditions \cite{Mann}.
-Some scientific questions,  such as discovery of protein biomarkers of a disease, consider meaningful patterns of changes in PTM abundance alone \cite{Petushkova_2017}.
-Other questions, e.g.
-in biological and physiological studies based on very short timescales, require us to distinguish the confounding of quantitative changes in PTM and of overall changes of the protein abundance.
+\todo{This section would benefit from more biology/technology background, and more extensive references}
+Signaling mechanisms allow cells to mount a fast and dynamic response to a multitude of bimolecular events.
+Signaling is facilitated by the modification of proteins at specific residues, acting as molecular on/off switches~\cite{Deribe,Cohen}.
+Localizing the modification sites, or characterizing relative abundance of a modification site's occupancy repertoire across experimental conditions, provides important insights~\cite{Mann}.
+For example, meaningful patterns of changes in post-translational modifications (PTMs) abundance can serve as biomarkers of a disease~\cite{Petushkova_2017}.
+Alternatively, distinguishing the quantitative changes in a PTM from the overall changes of the protein abundance helps gain insight into biological and physiological processes operating on a very short timescale \todo{would be good to have a reference}.
 
-Regardless of the question at hand, interrogating the modified proteome is challenging due to a number of reasons.
+Bottom-up liquid chromatography coupled with tandem mass spectrometry (LC-MS/MS) is a tool-of-choice for unbiased and large-scale identification and quantification of proteins and their PTMs~\cite{Kall:2011ub,Roepstorff}.
+However, LC-MS-based interrogation of the modified proteome is challenging, for a number of reasons.
 First, the relatively lower abundance of modified proteoforms dictates that a global interrogation can only be achieved through large-scale enrichment protocols with modification-specific antibodies or beads \cite{Huang:2014}.
-Variability in the enrichment efficiency inevitably affects the reproducibility of the number of spectral features (e.g., peptide precursor ions or their fragments) and their intensities, which imposes challenges in both quantification and statistical modeling.
+Variability in the enrichment efficiency inevitably affects the reproducibility of the number of spectral features (e.g., peptide precursor ions or their fragments) and their intensities.
 Second, contrary to the often large number of identified peptides that can be used as features to model protein abundance changes, there are relatively few representative peptides that span a modification sites, which often results in sparse, and sometimes, inherently convoluted models (i.e., single versus multiple modified sites on a single peptide) \cite{Mann}.
@@
 
-Currently statistical analyses of these experiments are performed with methods such as t-test and Limma \cite{Ritchie_15a}.
-These approaches generally use the intensity ratios of modified and unmodified peptide features and compare the mean abundance of different PTM sites.
-These approaches do not fully account for different sources of variations and as such, are not directly applicable to experiments with complex designs, such as multiple conditions and mixtures.
-Additionally, there is no self contained implementation of these methods, requiring manual application of the methods by the user and increasing complexity.
+The technological challenges in PTM identification and quantification translate themselves into challenges of statistical data analyses.
+Yet frequently data from these experiments are analyzed using statistical methods that were not originally designed for this task.
+In particular, methods such as t-test, Analysis of Variance, or Limma take as input the intensity ratios of modified and unmodified peptide features, and compare the mean abundance of different PTM sites~ \cite{Ritchie_15a}.
+Such methods do not fully account for all the sources of uncertainty and variation in the experiment.
+As the result, they are either not directly applicable to experiments with non-trivial designs (such as experiments with multiple conditions, paired and time course designs, and experiments with labeling), or require the analysts to exercise non-trivial statistical expertise.
 
 This manuscript proposes a general statistical analysis framework that detects relative changes in post-translational modifications.
-The proposed framework combines the quantitative information pertaining to a PTM site over peptides and replicates in multiple conditions, and takes into account various sources of variation and confounding.
-The framework requires an experimental design that, for each sample, quantifies the peptides with post-translational modifications, as well as  peptides from the same proteins with no modification sites.
-It is applicable to both label-free and tandem mass tag (TMT)-based acquisitions.
-The proposed framework summarizes the abundances of peptides with and without the modification sites, and fits a linear mixed effects model that automatically adjusts to the biological and technological aspects of the experimental design.
-Finally, changes in the PTM and protein-level abundances are combined to remove bias due to the confounding of these two sources.
+The framework requires an experimental design, which for each sample quantifies both the peptides with post-translational modifications, and peptides from the same proteins with no modification sites.
+The framework supports both label-free (DIA, DDA) and tandem mass tag (DDA-TMT)-based acquisitions.
+The statistical methodology separately summarizes the abundances of peptides with and without the modification sites, and fits separate linear mixed effects models that reflect the biological and technological aspects of the experimental design.
+Next, model-based inferences regarding the PTM and the protein-level abundances are combined to account for the confounding between these two sources.
 
-We evaluated the proposed framework on datasets from computer simulations, benchmark controlled mixtures, and biological investigations, which illustrate a diverse set of organisms, acquisition methods, and experimental designs, showing the applicability of the methods to a variety of situations.
-The evaluations demonstrated that by appropriately leveraging the information from the unmodified peptide, the proposed approach improves the accuracy of the estimates of PTM fold changes and results in a better calibrated false positive rate.
-In particular, accounting for the confounding from protein abundance allows us to see the true effect of the modification, avoiding the need for more manual and time intensive methods.
+We evaluated the proposed framework on \todo{how many} datasets from computer simulations, \todo{how many} benchmark controlled mixtures, and \todo{how many} biological investigations.
+The datasets illustrate a diverse set of organisms, acquisition methods, and experimental designs, showing the applicability of the framework to a variety of situations.
+The evaluations demonstrated that by appropriately leveraging the information from the unmodified peptide, the proposed approach improves the accuracy of the estimates of PTM fold changes and results in a better calibrated false positive rate of detecting differentially abundant PTMs.
+In particular, accounting for the confounding from protein abundance allows us to characterize the true effect of the modification, avoiding the need for more manual and time intensive methods.
 
-The proposed approach is implemented as a freely available open source R package $MSstatsPTM$, available on Bioconductor, which employs similar input format as in $MSstats$ and $MSstatsTMT$, and is compatible with many acquisition methods, such as label free, DIA, DDA, and TMT \cite{Choi:2014,Huang:2020}.
+The proposed approach is implemented as a freely available open source R package $MSstatsPTM$, available on Bioconductor, which employs similar input format as in $MSstats$ and $MSstatsTMT$ \cite{Choi:2014,Huang:2020}.
 

### 2022-03-07 code/MSstatsPTM_Simulation_Analysis.Rmd
Added data availability info

@@
 ---
-title: "MSstatsPTM_SImulation_Analysis"
+title: "MSstatsPTM Simulation Analysis"
 author: "Devon Kohler"
-date: "6/16/2021"
+date: "3/7/2022"
 output:
@@
 ```{r load_packages}
-library(MSstatsPTMold)
 library(MSstatsPTM)
@@
 library(tidyverse)
-library(limma)
 ```

### 2022-03-07 code/SpikeIn_Method_Comparison.Rmd
Added data availability info

--- 2022-02-01 code/Spike-in_Analysis.Rmd
@@
 ---
-title: "Spike-in_MSstatsPTM_Analysis"
+title: "SpikeIn MSstatsPTM Method Comparison"
 author: "Devon Kohler"
-date: "6/18/2021"
-output: html_document
+date: "3/7/2022"
+output:
+  pdf_document: default
+  html_document: default
 ---
@@
 
-### Data Stats
-
-##### Missing Values
-
-PTM: 20%
-Protein: 30%
-
-##### Average Features
-
-PTM: 1.4
-Protein: 10.2
-
 ### MSstatsPTM
@@
 ```{r pressure, echo=FALSE, fig.width = 10, fig.height = 7}
-load(file = "../data/Spike_in_MSstatsPTM_Model.rda")#model_df
+load(file = "MSstatsPTM_Summarized.rda")
+load(file = "Spike_in_MSstatsPTM_Model.rda")#model_df
 
@@
 # Volcano plot
-png("../supplementary/sim_new/spike_in_msstatsptm_volcano.png", width = 750, height = 500)
 plot_df %>% arrange(labeling) %>% 
@@
   labs(title = "Proposed Method VolcanoPlot")
-dev.off()
 ```
@@
 This is because, while the fold change does now follow what we expected, a large number of the spike-in peptides still had an adjusted p value over .05.
-This is most likely due to increases in the variance when combined with the protein model.
 Finally specificity also saw a large increase when adding in the protein adjustment, indicating we are correctly identifying most of the background peptides as negatives.
@@
 
-spike %>% ggplot() + geom_boxplot(aes(x = contrast_truth, y = log2FC), fill = "lightsteelblue4") + 
+spike %>% filter(!Label %in% c("mix1-mix2", "mix3-mix4")) %>% 
+  ggplot() + geom_boxplot(aes(x = contrast_truth, y = log2FC), fill = "lightsteelblue4") + 
   geom_point(aes(y = trueLog2FC, x = contrast_truth), color = "red", shape = 4, size = 3) + facet_wrap(.~Adjustment) + 
@@
 In the unadjusted plot, the expected falls outside the quartile boxes in the majority of comparisons.
-
-#### Calculating background peptide percentages
-
-```{r, echo=FALSE}
-
-light_pep <- model_df$PTM.Model %>% filter(labeling == "light")
-
-unadj_sig_light <- model_df$PTM.Model %>% filter(labeling == "light" & adj.pvalue < .05)
-adj_sig_light <- model_df$ADJUSTED.Model %>% filter(labeling == "light" & adj.pvalue < .05)
-
-nrow(unadj_sig_light) / nrow(light_pep)
-nrow(adj_sig_light) / nrow(light_pep)
-
-```
 
@@
 ```{r, echo=FALSE, fig.width = 10, fig.height = 7}
-load(file = "../data/Spike_in_Limma_Model.rda")#limma_model
+load(file = "Spike_in_Limma_Model.rda")#limma_model
 
@@
 # Volcano plot
-png("../supplementary/sim_new/spike_in_limma_volcano.png", width = 750, height = 500)
 plot_df %>% arrange(labeling) %>% 
@@
   labs(title = "Limma VolcanoPlot")
-dev.off()
 ```
@@
 After adjustment, the spike-in peptides are much more inline with the expected log2FC, however they are mostly below the significant pvalue threshold.
-This is most likely due to an increase in standard error or lower degrees of freedom compared to MSstatsPTM.
 
@@
 
-png("../main/images/spike_in_volcano.png", width = 750, height = 500)
 combo_plot_df %>% arrange(Model, labeling) %>% 
@@
   labs(title = "SpikeIn VolcanoPlot", x= "Log2FC")
-dev.off()
 
@@
 
-spike %>% ggplot() + geom_boxplot(aes(x = contrast_truth, y = Log2FC), fill = "lightsteelblue4") + 
+spike %>% 
+  filter(!Label %in% c("mix1-mix2", "mix3-mix4")) %>% 
+  ggplot() + geom_boxplot(aes(x = contrast_truth, y = Log2FC), fill = "lightsteelblue4") + 
   geom_point(aes(y = trueLog2FC, x = contrast_truth), color = "red", shape = 4, size = 3) + facet_wrap(.~Adjustment) + 
@@
 ```{r, echo=FALSE, fig.width = 10, fig.height = 7}
-load(file = "../data/Spike_in_ttest_model.rda")#ttest_final_model
+load(file = "Spike_in_ttest_model.rda")#ttest_final_model
 
@@
 # Volcano plot
-png("../supplementary/sim_new/spike_in_ttest_volcano.png", width = 750, height = 500)
 ttest_final_model %>% arrange(labeling) %>% 
@@
   labs(title = "T-test VolcanoPlot")
-dev.off()
 ```
@@
 ```{r, echo = FALSE, fig.width = 10, fig.height = 5}
-unadj_spike <- ttest_final_model %>% filter(protadj == "no adjustment" & labeling == "heavy")
+unadj_spike <- ttest_final_model %>% filter(protadj == "no adjustment" & labeling == "SpikeIn Peptides")
 unadj_spike$Adjustment <- "no adjustment"
-adj_spike <- ttest_final_model %>% filter(protadj == "protein adjustment" & labeling == "heavy")
+adj_spike <- ttest_final_model %>% filter(protadj == "protein adjustment" & labeling == "SpikeIn Peptides")
 adj_spike$Adjustment <- "protein adjustment"
@@
 
-spike %>% ggplot() + geom_boxplot(aes(x = contrast_truth, y = estimate), fill = "lightsteelblue4") + 
+spike %>% 
+  filter(!label %in% c("mix1-mix2", "mix3-mix4")) %>% 
+  ggplot() + geom_boxplot(aes(x = contrast_truth, y = estimate), fill = "lightsteelblue4") + 
   geom_point(aes(y = trueLog2FC, x = contrast_truth), color = "red", shape = 4, size = 3) + facet_wrap(.~Adjustment) + 
@@
 
-From the analysis above it is clear MSstatsPTM is the most robust modeling method of the three tested.
-In the computer simulations we saw MSstatsPTM and Limma having similar results, however in the real world experiment Limma performanced significantly worse.
-This could be in part due to the robust TMP summary methods used by MSstatsPTM compared the log sum of abundance used for Limma.
-An extension to this analysis could look into the performance of Limma using the same summarization methods as used in MSstatsPTM.
-In either case it is clear that the full workflow of MSstatsPTM performs the best of the observed models.
+In the following plot we compare the summary statsitics of the adjusted models. Note T-test is shown but all values are zero.
 
-In the following plot we compare the summary statsitics of the adjusted models. Note T-test is shown but all values are zero.
 ```{r, echo = FALSE, fig.width = 8}
@@
 
-png("../supplementary/sim_new/spike_in_statistic_comparison.png", width = 750, height = 500)
 summary %>% mutate(Model = factor(Model, levels=c("Proposed", "Limma", "T-test"))) %>% 
@@
        legend.position = "right")
-dev.off()
 ```
@@
 
-png("../supplementary/sim_new/spike_in_fc.png", width = 750, height = 500)
 spike %>% mutate(Comparison = str_split(contrast_truth, " ", simplify = TRUE)[,1],
@@
   labs(title = "Fold Change of Spike-in Peptides", x = "Comparison with Expected FC")
-dev.off()
+
 ```

### 2022-03-07 main/MSstatsPTM_Main.tex
Added data availability info

--- 2022-03-05 main/MSstatsPTM_Main.tex
@@
 
-\todo{Add massive and github links}
+All datasets for the four experiments and two simulations, including raw files and analysis results, are available online.
 
-Shigella: https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=4878d777c6b34cf8aaf8477e93140c4d
+The details of the four experiments, including raw data, and intermediate data processing outputs are located in individual MassIVE repositories: SpikeIn benchmark - Ubiquitination - Label-free (MassIVE repository identifier: \href{https://massive.ucsd.edu/ProteoSAFe/private-dataset.jsp?task=c4c583ecf7f941cdac87f7a4f872517b}{MSV000088971}), Human - Ubiquitination - 1mix-TMT (MassIVE repository identifier: \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=b6f0c74c234247678fb0888c6df1f225}{MSV000088966}), Mouse - Phosphorylation - 2mix-TMT time series (MassIVE repository identifier: \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=4878d777c6b34cf8aaf8477e93140c4d}{MSV000085565}), Human - Ubiquitination - Label-free no global profiling run (MassIVE repository identifier: \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=1b516164de5345108b40b75147dd58b5}{MSV000078977}).
+
+The R scripts and results of the statistical analysis for the four experiments are available in MassIVE.quant reanalysis containers: SpikeIn benchmark - Ubiquitination - Label-free (MassIVE reanalysis container identifier: \href{https://massive.ucsd.edu/ProteoSAFe/private-dataset.jsp?task=c4c583ecf7f941cdac87f7a4f872517b}{TBD}), Human - Ubiquitination - 1mix-TMT (MassIVE reanalysis container identifier: \href{https://massive.ucsd.edu/ProteoSAFe/reanalysis_container.jsp?task=b639fa5d0ff24ce78abdca13f4b18c0a}{RMSV000000356}), Mouse - Phosphorylation - 2mix-TMT time series (MassIVE reanalysis container identifier: \href{https://massive.ucsd.edu/ProteoSAFe/reanalysis_container.jsp?task=79406d32c1f9413daf7756e6eeb24968}{RMSV000000357}), Human - Ubiquitination - Label-free no global profiling run (MassIVE reanalysis container identifier: \href{https://massive.ucsd.edu/ProteoSAFe/reanalysis_container.jsp?task=f9e6117e00794ce5b2e8802601be16bc}{RMSV000000358}).
+
+The code, data, and analysis results of the two simulations, Computer simulation 1 - Label-free and Computer simulation 2 - Label-free missing values and low features, are available on Github: \href{https://github.com/devonjkohler/MSstatsPTM_simulations}{MSstatsPTM\_simulations}.
 

### 2022-03-07 code/MSstatsPTM_Analysis.R

@@ -1,0 +1,515 @@
+
+library(MSstatsPTM)
+library(data.table)
+library(tidyverse)
+library(gridExtra)
+library(ggrepel)
+library(VennDiagram)
+
+## Load msstats input data
+load(file = "pST_MSstatsPTM_Format.rda")#shigella_peptide_pST
+load(file = "global_MSstatsPTM_Format.rda")#shigella_global
+load(file = "pY_MSstatsPTM_Format.rda")#shigella_peptide_pY
+
+## Setup for summarization
+shigella_global <- shigella_global %>% group_by(ProteinName, PeptideSequence, 
+  Charge, PSM, Mixture, TechRepMixture, Run, Channel, Condition, BioReplicate) %>%
+  summarize(Intensity = max(Intensity))
+
+pY_input <- list(PTM = data.frame(shigella_peptide_pY), 
+                 PROTEIN = data.frame(shigella_global))
+pST_input <- list(PTM = data.frame(shigella_peptide_pST), 
+                  PROTEIN = data.frame(shigella_global))
+
+## Data summarization ----------------------------------------------------------
+py_summarization <- dataSummarizationPTM_TMT(pY_input)
+pST_summarization <- dataSummarizationPTM_TMT(pST_input)
+
+# save(py_summarization, file = "pY_Summarization.rda")
+# save(pST_summarization, file = "pST_Summarization.rda")
+
+py_summarized_data <- py_summarization$PTM$ProteinLevelData
+py_feature_data <- py_summarization$PTM$FeatureLevelData
+
+pst_summarized_data <- pST_summarization$PTM$ProteinLevelData
+pst_feature_data <- pST_summarization$PTM$FeatureLevelData
+
+global_summarized_data <- pST_summarization$PROTEIN$ProteinLevelData
+global_feature_data <- pST_summarization$PROTEIN$FeatureLevelData
+
+
+## Create pYST dataset
+## Create pY/pST Dataset
+## check the overlapped sites between pST and pY
+shared_sites <- intersect(unique(py_summarized_data$Protein), 
+                          unique(pst_summarized_data$Protein))
+shared_ST <- shared_sites[grepl("_S", shared_sites)|grepl("_T", shared_sites)]
+shared_Y <- shared_sites[grepl("_Y", shared_sites)]
+
[…]

### 2022-03-07 code/MSstatsPTM_SpikeIn_Analysis.R

@@ -1,0 +1,884 @@
+
+library(data.table)
+library(MSstatsPTM)
+library(tidyverse)
+
+setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
+
+# Prepare data set -------------------------------------------------------------
+
+# Design for global profiling data
+design_prot1 <- read.table("000547_ptm_benchmark/ptm_tft_global_ip1.txt",
+  header = TRUE, sep = "\t")
+design_prot1 <- as.data.table(design_prot1)
+design_prot2 <- read.table("000547_ptm_benchmark/ptm_tft_global_ip2.txt",
+  header = TRUE, sep = "\t")
+design_prot2 <- as.data.table(design_prot2)
+design_prot2[,id_injectionset := 2]
+design_prot <- rbindlist(list(design_prot1, design_prot2))
+
+# Design for KGG enriched data 
+design_kgg1 <- read.table("000547_ptm_benchmark/ptm_tft_kgg_ip1.txt",
+  header = TRUE, sep = "\t")
+design_kgg1 <- as.data.table(design_kgg1)
+design_kgg1[, batch := "BCH1"]
+design_kgg2 <- read.table("000547_ptm_benchmark/ptm_tft_kgg_ip2.txt",
+  header = TRUE, sep = "\t")
+design_kgg2 <- as.data.table(design_kgg2)
+design_kgg2[, batch := "BCH2"]
+
+design_kgg <- rbindlist(list(design_kgg1, design_kgg2))
+
+# Prepare data set with columns of interest for later use
+fn_kgg <- paste0("vistaga_", design_kgg$run_id, 
+                 "_kgg_with_unmodified_peptides.tsv")
+fn_prot <- paste0("vistaga_", design_prot$run_id, ".tsv")
+
+sub_cols <- c(
+  "Reference", "run_id", "peptide_id", "vista_peak_area_light", 
+  "vista_confidence_score", "peptide_m_z", "peptide_sequence", 
+  "peptide_trypticity", "peptide_miscleavages", "peptide_validity",
+  "label1", "ms2_rt", "ms2_charge", "is_xq"
+)
+
+dta_kgg <- vector("list", length(fn_kgg))
+for (i in seq_along(fn_kgg)) {
+  path_file <- paste0("000547_ptm_benchmark/", fn_kgg[i])
+  dta_kgg[[i]] <- read.table(
[…]

### 2022-03-07 code/MSstatsPTM_USP30_Analysis.R

@@ -1,0 +1,238 @@
+
+library(MSstatsPTM)
+library(tidyverse)
+
+setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
+
+## Load input data--------------------------------------------------------------
+raw_input_data <- readRDS("usp30_input_data.RDS")
+head(raw_input_data)
+
+## Convert to MSstatsPTM input format
+raw_input_data$FragmentIon <- NA
+raw_input_data$ProductCharge <- NA
+raw_input_data$IsotopeLabelType <- 'L'
+raw_input_data <- raw_input_data %>% 
+  mutate(PrecursorCharge = stringr::str_split(
+    raw_input_data$feature, "_", simplify = TRUE)[,2])
+
+colnames(raw_input_data) <- c('ProteinName', 'PeptideSequence', 'is_mod', 'Feature', 
+                            'Condition', "Batch", "Run", 'Intensity', 'site', 'FragmentIon',
+                            'ProductCharge', 'IsotopeLabelType', 'PrecursorCharge')
+
+raw_input_data$BioReplicate <- raw_input_data$Batch
+raw_input_data$Intensity <- 2^raw_input_data$Intensity
+
+## Split modified and unmodified peptides
+PTM_df <- raw_input_data %>% filter(is_mod == TRUE)
+PTM_df$ProteinName <- paste(PTM_df$ProteinName, PTM_df$site, sep = '_')
+protein_df <- raw_input_data %>% filter(is_mod == FALSE)
+
+## Remove unneeded columns
+drops <- c("is_mod", "site", "Batch", "Feature")
+PTM_df <- PTM_df[ , !(names(PTM_df) %in% drops)]
+protein_df <- protein_df[ , !(names(protein_df) %in% drops)]
+
+input_df <- list('PTM' = PTM_df, 'PROTEIN' = protein_df)
+
+## Summarize features-----------------------------------------------------------
+summarized_ptm <- dataSummarizationPTM(input_df)
+
+## Model summarized results-----------------------------------------------------
+## Full pairwise comparison used
+model_ptm <- groupComparisonPTM(summarized_ptm, data.type = "LabelFree")
+
+## Plot results-----------------------------------------------------------------
+## Look at overlap between modified and unmodified peptides
[…]

### 2022-03-07 code/Simulated_Analysis.R

@@ -1,9 +1,12 @@
 
-library(MSstatsPTMold)
+## Load packages
 library(MSstatsPTM)
 library(data.table)
 library(tidyverse)
 library(limma)
+
+## Source simulation code
+source("PTMsimulateExperiment.R")
 
 ## Define Modeling Functions ---------------------------------------------------
 convert_to_msstats_format <- function(df){

### 2022-03-08 main/MSstatsPTM_Main.tex
Small

--- 2022-03-06 main/MSstatsPTM_Main.tex
@@
 
-This section summarizes the datasets in this manuscript.
-The datasets were chosen to provide a performance benchmark of the proposed approach in situations with known ground truth, and to represent a variety of experimental designs and acquisition methods.
-Two computer simulations were run which varied in experimental realism.
-The first simulation was perfectly clean, with many replicates and no missing values, while the second simulation introduced real world qualities, such as limited modified features and missing values.
-The spike-in experiment took the real world qualities a step further and allowed us to compare the methods in a real experiment with known changes in modified spike-in peptides.
+The benchmark datasets in this manuscript provide a performance benchmark of the proposed approach in situations with known ground truth, and to represent a variety of experimental designs and acquisition methods.
+Two computer simulations varied in experimental realism.
+The first simulation produced a perfectly clean dataset, with many replicates and no missing values.
+The second simulation introduced real-world characteristics, such as limited modified features and missing values.
+The spike-in experiment took the real world characteristics a step further and allowed us to compare the methods in a real experiment with known changes in modified spike-in peptides.
 Finally, three biological experiments demonstrated the applicability of the proposed approach across different biological organisms, experimental designs, and data acquisition strategies.
-A summary of the experiments used in this manuscript can be seen in Table \ref{tab:dataDescription}.
-Details of the raw data, R scripts with $MSstatsPTM$ analysis, and the results of the statistical analysis are available in either MassIVE.quant or Github (with direct links available in the table).
+Table \ref{tab:dataDescription} summarizes the experiments.
+Details of the experimental data, R scripts with $MSstatsPTM$ analysis, and the results of the statistical analysis are available in either MassIVE.quant \todo{cite the paper, provide url}.
+Details of computer simulations are available on GitHub.
 
@@
 %%%% colnames of table
-&  & \multicolumn{3}{c|}{Experimental design} & Data availability \\
-& Dataset & \multirow{2}{*}{Comparison} & No. of & No. of Bio. & (MassIVE.quant  \\
-&  & & Conditions & Replicates & or R package) \\
+& \multirow{2}{*}{Dataset} & Experimental & No. of & No. of Bio. &  Data \\
+&  & Design & Conditions & Replicates & availability \\
 \hline
@@
 %%%% DDA:ControlledMix
-\multirow{5}{*}{\rotatebox[origin=c]{90}{Known}}  \multirow{5}{*}{\rotatebox[origin=c]{90}{Ground}} \multirow{5}{*}{\rotatebox[origin=c]{90}{Truth}} & Computer simulation 1 - Label-free & Group & 2/3/4 & 2/3/5/10 & \href{https://github.com/devonjkohler/MSstatsPTM_simulations}{Github}\\
+\multirow{5}{*}{\rotatebox[origin=c]{90}{Known}}  \multirow{5}{*}{\rotatebox[origin=c]{90}{Ground}} \multirow{5}{*}{\rotatebox[origin=c]{90}{Truth}} & Computer simulation 1 - Label-free & Group comparison & 2/3/4 & 2/3/5/10 & \href{https://github.com/devonjkohler/MSstatsPTM_simulations}{Github \todo{url?}}\\
  &&&&& \\%[-0.05in]
@@
 %%% DDA iPRG2015
-& Computer simulation 2 - missing and low features & Group & 2/3/4 & 2/3/5/10 & \href{https://github.com/devonjkohler/MSstatsPTM_simulations}{Github} \\
+& Computer simulation 2 - missing and low features & Group comparison & 2/3/4 & 2/3/5/10 & \href{https://github.com/devonjkohler/MSstatsPTM_simulations}{Github \todo{url?}} \\
  &&&&& \\%[0.05in]
@@
 %%% DDA Cox2014
-& SpikeIn benchmark - Ubiquitination - Label-free & Group & 4 & 2 & \href{https://massive.ucsd.edu/ProteoSAFe/private-dataset.jsp?task=c4c583ecf7f941cdac87f7a4f872517b}{MSV000088971}\\
+& SpikeIn benchmark - Ubiquitination - Label-free & Group comparison & 4 & 2 & \href{https://massive.ucsd.edu/ProteoSAFe/private-dataset.jsp?task=c4c583ecf7f941cdac87f7a4f872517b}{MSV000088971}\\
  &&&&& \\%[-0.05in]
 \hline
-\multicolumn{6lm}{c}{ } \\ [0.02in]
+\multicolumn{6}{c}{ } \\ [0.02in]
 \hline 
@@
 %% DDA Meierhofer
-\multirow{3}{*}{\rotatebox[origin=c]{90}{Biological}} \multirow{3}{*}{\rotatebox[origin=c]{90}{Investigation}} & Human - Ubiquitination - 1mix-TMT & Group & 6 & 2 & \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=b6f0c74c234247678fb0888c6df1f225}{MSV000088966} \\
+\multirow{3}{*}{\rotatebox[origin=c]{90}{Biological}} \multirow{3}{*}{\rotatebox[origin=c]{90}{Investigation}} & Human - Ubiquitination - 1mix-TMT & Group comparison & 6 & 2 & \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=b6f0c74c234247678fb0888c6df1f225}{MSV000088966} \\
 &&&&& \\%[-0.05in]
@@
 %%%% SRM Surinova 2015, training
-& Human - Ubiquitination - Label-free & Group & 4 & 2 & \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=1b516164de5345108b40b75147dd58b5}{MSV000078977}\\ [0.02in]
+& Human - Ubiquitination - Label-free & Group comparison & 4 & 2 & \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=1b516164de5345108b40b75147dd58b5}{MSV000078977}\\ [0.02in]
  &&&&& \\
@@
 \end{tiny}
-\caption{ \small {\bf Experimental datasets in this manuscript} 
-```Dataset'' is the code name of the dataset used in this manuscript.
-`Comparison'' is the type of experimental design (group comparison, or time course).
-``No. of Conditions'' states the number of conditions in the dataset.
-``No. of Bio. Replicates'' states the number of biological replicates per condition.
-``Data availability'' column shows the MassIVE.quant ID or github repository. 
+\caption{ \small {\bf Simulated and experimental datasets in this manuscript} 
+```Dataset'' is the code name of the dataset in this manuscript.
+``Data availability'' shows the ID of the MassIVE.quant repository \todo{url} or the GitHub repository. 
 }
@@
 
-To evaluate the proposed method in situations with biological variation and known ground truth we created synthetic datasets which mimicked a label-free experiment.
-Multiple simulations were generated with differing numbers of replicates and conditions per simulation.
-All simulations were created with a high number of modified peptide features and no missing values.
-This was done to see how the method performed in different experimental design scenarios.
-The exact numbers of replicates and conditions can be seen in Table \ref{tab:dataDescription}.
-
-In each simulation, 1000 PTMs and 1000 unmodified proteins were generated.
-Each PTM and protein was created with 10 corresponding features.
+The simulation mimicked a label-free experiment.
+Synthetic datasets had a high number of modified peptide features and no missing values, while varying numbers of replicates and conditions.
+Specifically, each dataset had 1000 PTMs and 1000 unmodified proteins, each represented by 10 LC-MS features observed across all the mass spectrometry runs.
 500 of the PTMs were generated with a differential fold change between conditions, while the other half were generated with no difference between conditions.

### 2022-03-15 main/MSstatsPTM_Main.tex
cleaned up ref

@@
 
-We evaluated the proposed framework on \todo{how many} datasets from computer simulations, \todo{how many} benchmark controlled mixtures, and \todo{how many} biological investigations.
+We evaluated the proposed framework on two datasets from computer simulations, one benchmark controlled mixture, and three biological investigations.
 The datasets illustrate a diverse set of organisms, acquisition methods, and experimental designs, showing the applicability of the framework to a variety of situations.
@@
 Table \ref{tab:dataDescription} summarizes the experiments.
-Details of the experimental data, R scripts with $MSstatsPTM$ analysis, and the results of the statistical analysis are available in either MassIVE.quant \todo{cite the paper, provide url}.
-Details of computer simulations are available on GitHub.
+Details of the experimental data, R scripts with $MSstatsPTM$ analysis, and the results of the statistical analysis are available in MassIVE.quant (\url{https://massive.ucsd.edu/ProteoSAFe/static/massive-quant.jsp}) \cite{Choi:2020}.
+Details of computer simulations are available on GitHub (\url{https://github.com/devonjkohler/MSstatsPTM_simulations}).
 
@@
 %%%% DDA:ControlledMix
-\multirow{5}{*}{\rotatebox[origin=c]{90}{Known}}  \multirow{5}{*}{\rotatebox[origin=c]{90}{Ground}} \multirow{5}{*}{\rotatebox[origin=c]{90}{Truth}} & Computer simulation 1 - Label-free & Group comparison & 2/3/4 & 2/3/5/10 & \href{https://github.com/devonjkohler/MSstatsPTM_simulations}{Github \todo{url?}}\\
+\multirow{5}{*}{\rotatebox[origin=c]{90}{Known}}  \multirow{5}{*}{\rotatebox[origin=c]{90}{Ground}} \multirow{5}{*}{\rotatebox[origin=c]{90}{Truth}} & Computer simulation 1 - Label-free & Group comparison & 2/3/4 & 2/3/5/10 & \href{https://github.com/devonjkohler/MSstatsPTM_simulations}{Github}\\
  &&&&& \\%[-0.05in]
@@
 %%% DDA iPRG2015
-& Computer simulation 2 - missing and low features & Group comparison & 2/3/4 & 2/3/5/10 & \href{https://github.com/devonjkohler/MSstatsPTM_simulations}{Github \todo{url?}} \\
+& Computer simulation 2 - missing and low features & Group comparison & 2/3/4 & 2/3/5/10 & \href{https://github.com/devonjkohler/MSstatsPTM_simulations}{Github} \\
  &&&&& \\%[0.05in]
@@
 ```Dataset'' is the code name of the dataset in this manuscript.
-``Data availability'' shows the ID of the MassIVE.quant repository \todo{url} or the GitHub repository. 
+``Data availability'' shows the ID of the MassIVE.quant repository or the GitHub repository. 
 }

### 2022-03-16 main/MSstatsPTM_Main.tex
Datasets

@@
 Signaling is facilitated by the modification of proteins at specific residues, acting as molecular on/off switches~\cite{Deribe,Cohen}.
-Localizing the modification sites, or characterizing relative abundance of a modification site's occupancy repertoire across experimental conditions, provides important insights~\cite{Mann}.
+Localizing the modification sites, or characterizing relative abundance of a modification site's occupancy repertoire across experimental conditions provides important insights~\cite{Mann}.
 For example, meaningful patterns of changes in post-translational modifications (PTMs) abundance can serve as biomarkers of a disease~\cite{Petushkova_2017}.
@@
 
-Bottom-up liquid chromatography coupled with tandem mass spectrometry (LC-MS/MS) is a tool-of-choice for unbiased and large-scale identification and quantification of proteins and their PTMs~\cite{Kall:2011ub,Roepstorff}.
+Bottom-up liquid chromatography coupled with tandem mass spectrometry (LC-MS/MS) is a tool of choice for unbiased and large-scale identification and quantification of proteins and their PTMs~\cite{Kall:2011ub,Roepstorff}.
 However, LC-MS-based interrogation of the modified proteome is challenging, for a number of reasons.
@@
 Variability in the enrichment efficiency inevitably affects the reproducibility of the number of spectral features (e.g., peptide precursor ions or their fragments) and their intensities.
-Second, contrary to the often large number of identified peptides that can be used as features to model protein abundance changes, there are relatively few representative peptides that span a modification sites, which often results in sparse, and sometimes, inherently convoluted models (i.e., single versus multiple modified sites on a single peptide) \cite{Mann}.
-Third, unless early signaling events are interrogated, the interpretation of the relative changes in modification occupancy are inherently confounded with changes in the overall protein abundance, complication the interpretation of the results \cite{Olsen:2013}.
-Finally, technological aspects of bottom-up MS experiments, and presence of labeling, e.g.
-by tandem mass tag (TMT), introduce additional sources of uncertainty and variation.
+Second, contrary to the often large number of identified peptides that can be used to quantify protein abundance, there are relatively few representative peptides that span a modification site, and there may be multiple modified sites on a single peptide~\cite{Mann}.
+Third, unless early signaling events are interrogated, the interpretation of the relative changes in modification occupancy are inherently confounded with changes in the overall protein abundance, complicating the interpretation of the results \cite{Olsen:2013}.
+Finally, technological aspects of bottom-up MS experiments, such as presence of labeling by tandem mass tag (TMT), introduce additional sources of uncertainty and variation.
 
-The technological challenges in PTM identification and quantification translate themselves into challenges of statistical data analyses.
+The technological challenges in PTM identification and quantification challenge the downstream statistical analyses.
 Yet frequently data from these experiments are analyzed using statistical methods that were not originally designed for this task.
@@
 
-The proposed approach is implemented as a freely available open source R package $MSstatsPTM$, available on Bioconductor, which employs similar input format as in $MSstats$ and $MSstatsTMT$ \cite{Choi:2014,Huang:2020}.
+The proposed approach is implemented as a freely available open source R package $MSstatsPTM$, available on Bioconductor, which employs similar input format as $MSstats$ and $MSstatsTMT$ \cite{Choi:2014,Huang:2020}.
 
@@
 
-The benchmark datasets in this manuscript provide a performance benchmark of the proposed approach in situations with known ground truth, and to represent a variety of experimental designs and acquisition methods.
+The datasets in this manuscript  benchmarked the proposed approach in situations with known ground truth, and represented a variety of experimental designs and acquisition methods.
 Two computer simulations varied in experimental realism.
@@
 The spike-in experiment took the real world characteristics a step further and allowed us to compare the methods in a real experiment with known changes in modified spike-in peptides.
-Finally, three biological experiments demonstrated the applicability of the proposed approach across different biological organisms, experimental designs, and data acquisition strategies.
+Finally, three biological experiments demonstrated the applicability of the proposed approach across different biological organisms, PTM types, experimental designs and acquisition strategies.
 Table \ref{tab:dataDescription} summarizes the experiments.
@@
 ```Dataset'' is the code name of the dataset in this manuscript.
-``Data availability'' shows the ID of the MassIVE.quant repository or the GitHub repository. 
+``Data availability'' shows the ID of the MassIVE.quant repository or the GitHub repository.  \todo{All the experiments were conducted in data-dependent acquisition (DDA) mode.}
 }
@@
 
-The simulation mimicked a label-free experiment.
-Synthetic datasets had a high number of modified peptide features and no missing values, while varying numbers of replicates and conditions.
-Specifically, each dataset had 1000 PTMs and 1000 unmodified proteins, each represented by 10 LC-MS features observed across all the mass spectrometry runs.
+The simulation provides an idealistic case.
+\todo{generated (how many?)  repeated synthetic datasets?}.
+\todo{I have a hard time following this section.
+Would it be possible to rewrite the description, by explaining (1) how many repeated datasets, (2) how many proteins per dataset; (3) how many PTMs per protein, (4) how many modified/unmodified features per protein?
+Then talk about the quantitative aspects: systematic fold change; random variation; where the parameters of the simulation were inspired from.
+It will help relate this simulation to real data} had 1000 PTMs and 1000 unmodified proteins, each represented by 10 LC-MS features observed across all the mass spectrometry runs.
 500 of the PTMs were generated with a differential fold change between conditions, while the other half were generated with no difference between conditions.
@@
 The simulations were generated with one of two values $\sigma^2 = \{.2, .3\}$ motivated by the biological experiments in this manuscript.
-
-Further details of how the first simulation was generated, can be found in Supplementary Sec. \ref{supp-sec:dataset1}.
+Further details can be found in Supplementary Sec. \ref{supp-sec:dataset1}.
 
@@
 
-In this simulation, data was generated using the same parameters as described in the previous section and including real world experimental conditions.
-Specifically these simulations included missing values and a low number of modified features.
-The percentage of missing values and feature counts were determined by looking at the biological experiments in this paper.
-PTMs were simulated with 2 modified peptide features, while unmodified proteins were still simulated with 10 features.
+The data were simulated as above, while including missing values and a low number of modified features to provide a more realistic representation of the experiments.
+In this simulation, the feature counts and the proportion of missing values were as observed \todo{on average over all the experimental datasets in this manuscript?}.
+PTMs were simulated with 2 modified peptide features, while unmodified proteins were simulated with 10 features.
 Additionally, 20\% of observations for both modified and unmodified peptides were missing completely at random.
-Adding missing values and few representative modified features provides a more realistic expectation of model performance.
-
-Further details of how this simulation was generated, including exact parameters used, can be found in Supplementary Sec. \ref{supp-sec:dataset2}.
+\todo{There will likely be a comment that missing values are more common among low-abundance features.
+Did the simulation account for that?} Further details can be found in Supplementary Sec.
+\ref{supp-sec:dataset2}.
 
@@
 \label{sec:exp_proc_dataset3}
-We evaluated our approach using a custom designed spike-in benchmark experiment with known ground truth.
-This experiment had 4 conditions with two biological replicates per condition and a balanced design.
 
-Experimental Design: In this experiment 50 heavy-labeled KGG motif peptides from 20 human proteins were used as spike-in peptides.
+The custom spike-in experiment had 4 conditions with two biological replicates per condition, and a balanced design. Additional information can be found in Supplementary Sec. \ref{supp-sec:benchmark}.
+
+\medskip \noindent {\bf Experimental Design:} 50 heavy-labeled KGG motif peptides from 20 human proteins were used as spike-in peptides.
 Quantitative changes in protein and site abundance of these 20 proteins were the target of the benchmark.
-Unmodified peptides from Human Lysate were used as the estimate of global protein abundance changes.
-Background E coli Lysate was used to normalize total protein levels prior to enrichment or global protein profiling.
-The background lysate was treated as controls and are not expected to be differential in any comparison.
+Unmodified peptides from Human lysate were used to estimate changes in global protein abundance.
+Background {\it E.
+coli} lysate was used to normalize total protein levels prior to enrichment or global protein profiling.
+The background lysate was treated as control, i.e.
+it had no expected changes in any comparison.
+
 The spike-in peptides were mixed with human lysate to create four mixture conditions.
-Two sets of data were acquired for each mixture: KGG enriched + LC-MS, and LC-MS only.
+The true log fold changes of spike-in modified peptides in these comparisons can be seen in Table \ref{table:spikein_fold_change}.
+Two sets of data were acquired for each mixture in label-free LC-MS/MS mode: KGG enriched + LC-MS/MS, and LC-MS/MS only.
+\todo{How the biological replicates in each condition were obtained?}
 
-Global Profiling Run: This experiment included a separate global profiling run to measure unmodified peptides.
+\medskip \noindent {\bf Global Profiling Run:} This experiment included a separate global profiling run to measure unmodified peptides.
 There was a 90.2\% overlap between the identified background modified peptides and proteins that were quantified in the global profiling run.
 
-Pairwise Comparisons: We evaluated the ability of the methods to detect known changes in abundance between pairs of conditions.
+\medskip \noindent {\bf Pairwise Comparisons:} We evaluated the ability of the statistical methods to detect known changes in abundance between pairs of conditions.
 The four mixtures were treated as individual conditions and labeled mix1, mix2, mix3, and mix4.
 The pairwise comparisons were mix1-mix2, mix1-mix3, mix1-mix4, mix2-mix3, mix2-mix4, and mix3-mix4.
-The true log fold changes of spike-in modified peptides in these comparisons can be seen in Table \ref{table:spikein_fold_change}.
 
@@
 
-More information on the experimental design of the benchmark experiment and how the expected fold change was calculated can be found in Supplementary Sec. \ref{supp-sec:benchmark}.
-
 \subsection*{Dataset 4 : Human - Ubiquitination - 1mix-TMT}
 \label{sec:exp_proc_dataset4}
-The proposed approach was evaluated on an experiment with six conditions and an unbalanced design.
+\todo{In this experiment XXX biological replicates were allocated to 1 TMT mixture in an unbalanced design.} Further details are in Supplementary Sec. \ref{supp-sec:ipah}.
 
-Experimental Design: In this study Luchetti et al.
-\cite{LUCHETTI2021} used multiplex proteomics to quantify the abundance of total protein and ubiquitination in human epithelial cells.
+\medskip \noindent {\bf Experimental Design:} Luchetti et al.
+\cite{LUCHETTI2021} quantified the abundance of total protein and ubiquitination in human epithelial cells.
 Cells were engineered to express IpaH7.8 under a dox inducible promoter and measurements were taken at different time periods.
@@
 Uninfected cells were measured at 0 and 6 hours, while infected cells were measured at 1, 2, 4, and 6 hour increments, resulting in six total conditions.
+The proteome was quantified in 11-plex \todo{?} isobaric multiplexing via tandem mass tagging (TMT) in combination with LC-MS/MS.
+\todo{One MS run total?}
 
-Global Profiling Run: This experiment included a separate global profiling run to measure unmodified peptides.
+\medskip \noindent {\bf Global Profiling Run:} This experiment included a separate global profiling run \todo{with the same design?} to measure unmodified peptides.
 There was a 95\% overlap between the identified modified peptides and proteins that were quantified in the global profiling run.
 
-Pairwise comparisons: We evaluated the statistical approach by its ability to detect changes in the abundance of modified peptides both before and after adjusting for changes in global protein abundance.
+\medskip \noindent {\bf Pairwise comparisons:} We evaluated the ability of the statistical methods to detect changes in the abundance of modified peptides both before and after adjusting for changes in global protein abundance.
 The six condition were labeled Dox1hr, Dox2hr, Dox4hr, Dox6hr, NoDox0hr, and NoDox6hr.
-All conditions were compared with each other in a full pairwise comparison, resulting in 15 comparisons.
-Because the dataset is a biological investigation, the true positive modifications were unknown.
-
-Further details on the design of this experiment are provided in Supplementary Sec. \ref{supp-sec:ipah}.
+All conditions were compared with each other, resulting in 15 pairwise comparisons.
+Since the dataset was a biological investigation, the true positive modifications were unknown.
 
@@
 \label{sec:exp_proc_dataset5}
-In this experiment 8 biological replicates were allocated to 2 TMT mixtures in an unbalanced design.
+In this experiment 8 biological replicates were allocated to 2 TMT mixtures in an unbalanced design. More details are in Supplementary Sec. \ref{supp-sec:shigella}.
 
-Experimental Design: Maculins et al.
-\cite{Maculins} designed an experiment targeting primary murine macrophages infected with Shigella flexneri (S.flexneri).
-Multiplex proteomics was used to quantify the abundance of total protein, and phosphorylation in wild type (WT) and ATG16L1-deficient (cKO) samples, uninfected and uninfected with S.flexneri.
+\medskip \noindent {\bf Experimental Design:} Maculins et al.
+\cite{Maculins} studied primary murine macrophages infected with {\it Shigella flexneri} ({\it S.
+flexneri}).
+The experiment quantified the abundance of total protein and of phosphorylation in wild type (WT), and in ATG16L1-deficient (cKO) samples, uninfected and uninfected with {\it S.
+flexneri}.
 The abundance of total protein and post-translation modifications were quantified at three time points, uninfected, early infection (45-60 minutes), and late infection (3-3.5 hours).
-11-plex isobaric multiplexing via tandem mass tagging (TMT) in combination with liquid-chromatography and tandem mass spectrometry (LC-MS/MS) was used.
-Six conditions were split between 11 channels leading to the experimental design being unbalanced.
+
+The proteome was quantified in 11-plex isobaric multiplexing via tandem mass tagging (TMT) in combination with LC-MS/MS.
+The six conditions were split between 11 channels, and as the result the experiment was unbalanced.
 Each mixture contained two replicates per early and late WT and KO conditions.
@@
 
-Global Profiling Run: This experiment included a separate global profiling run to measure unmodified peptides.
+\medskip \noindent {\bf Global Profiling Run:} This experiment included a separate global profiling run \todo{with the same design?} to measure unmodified peptides.
 There was a 90\% overlap between the identified background modified peptides and proteins that were quantified in the global profiling run.
 
-Time series comparison: We evaluated the statistical approach by its ability to detect changes in the abundance of modified peptides both before and after adjusting for changes in global protein abundance.
+\medskip \noindent {\bf Time series comparison:} We evaluated the ability of the statistical methods to detect changes in the abundance of modified peptides both before and after adjusting for changes in global protein abundance.
 The six condition were labeled KO\_Uninfect, KO\_Early, KO\_Late, WT\_Uninfect, WT\_Early, and WT\_Late.
-9 total comparisons were made named KO\_Early-WT\_Early, KO\_Late-WT\_Late, KO\_Uninfected-WT\_Uninfected, KO\_Early-KO\_Uninfected, KO\_Late-KO\_Uninfected, WT\_Early-WT\_Uninfected, WT\_Late-WT\_Uninfected, Infected-Uninfected, and KO-WT.
-Because the dataset is a biological investigation, the true positive modifications were unknown.
-
-Further details on the design of this experiment are provided in Supplementary Sec. \ref{supp-sec:shigella}.
+9 total comparisons were made, namely KO\_Early-WT\_Early, KO\_Late-WT\_Late, KO\_Uninfected-WT\_Uninfected, KO\_Early-KO\_Uninfected, KO\_Late-KO\_Uninfected, WT\_Early-WT\_Uninfected, WT\_Late-WT\_Uninfected, Infected-Uninfected, and KO-WT.
+Since the dataset was a biological investigation, the true positive modifications were unknown.
 
@@
 
-In this experiment 8 biological replicates were split between 4 conditions in a balanced design.
+In this experiment 8 biological replicates were split between 4 conditions in a balanced design. \todo{Further details in Suppler Sec XXX?}
 
-Experimental Design: Cunningham et al.
+\medskip \noindent {\bf Experimental Design:} Cunningham et al.
 \cite{Cunningham2015} investigated the relationship between USP30 and protein kinase PINK1, and their association with Parkinson’s Disease.
-Ubiquitination site profiling was performed and the modified site abundance was analyzed.
-Four conditions were tested with two biological replicates per condition.
-The conditions were as follows: CCCP, USP30 over expression (USP30 OE), Combo, and Control.
-Label-free mass spectrometry quantification was used to quantify the abundance of modified peptides.
+The experiment profiled ubiquitination sites, and analyzed changes in the modified site abundance.
+The experiment had four conditions, CCCP, USP30 over expression (USP30 OE), Combo, and Control.
+Each condition had two biological replicates.
+The abundance of modified peptides was quantified with label-free LC-MS/MS.
 
-Global Profiling Run: This experiment did not included a separate global profiling run to measure unmodified peptides.
-In this case $MSstatsPTM$ can still be used by extracting unmodified peptides from the modified run, however this leads to substantially less matches between modified and unmodified peptides, in addition to low feature counts for unmodified peptides.
+\medskip \noindent {\bf Global Profiling Run:} This experiment did not include a separate global profiling run to measure unmodified peptides.
+In this case $MSstatsPTM$ can still be used by extracting unmodified peptides from the modified run.
+However, in addition to low feature counts for unmodified peptides,  this leads to substantially fewer matches between modified and unmodified peptides.
 There was a 41.9\% overlap between the identified background modified peptides and proteins that were quantified in the global profiling run.
 
-Pairwise comparisons: We evaluated the statistical approach by its ability to detect changes in the abundance of modified peptides both before and after adjusting for changes in global protein abundance.
-The four condition were labeled CCCP, Combo, Ctrl, and USP30\_OE.
-All conditions were compared with each other in a full pairwise comparison, resulting in 6 comparisons.
-Because the dataset is a biological investigation, the true positive modifications were unknown.
+\medskip \noindent {\bf Pairwise comparisons:} We evaluated the ability of the statistical methods to detect  changes in the abundance of modified peptides both before and after adjusting for changes in global protein abundance.
+All the conditions were compared with each other in a full pairwise comparison, resulting in 6 comparisons.
+Since the dataset is a biological investigation, the true positive modifications were unknown.
 

### 2022-03-16 main/MSstatsPTM_Main.tex
Suggestion for the results

@@
 
+\todo{Would it be possible to re-organize this section?
+(1) Overview, including the data structures and the figure; (2) Existing methods: limma and t-test, including the description in the notation of the figure; also Isobar-PTM - can express in the same notation?
+Ok if not; (3) Existing methods: MSstats.
+Not applicable for PTM but useful background; (4) Proposed approach in the notation of the figure.
+I feel that the experimental design in Suppler is redundant.
+Could we fold it into the main text as well?} \todo{In this way, the entire Suppler Sec 2 will be merged into main \; the only methods in supplementary will be the table }
+
 \subsection*{Existing statistical methods for experiments targeting PTMs}
 
-Despite the important implications of PTMs in biological functions, there is a lack of general framework to summarize the available quantitative information from LC-MS data, to perform statistical inference, and to draw conclusions to characterize the quantitative properties of PTM in a statistically rigorous manner.
+Despite the importance of PTMs for biological function, there is a lack of general statistical analysis framework that summarizes the available quantitative information from LC-MS/MS experiments, performs statistical inference, and characterizes the quantitative properties of PTM in a statistically rigorous manner.
+
 Many investigations performed differential abundance analysis of PTMs using two-sample t-test or its extensions.
-The approach takes as input intensities of individual features from modified peptides, or intensity ratios of modified and unmodified peptide features, and compares the mean abundance of a PTM site from one condition to another.
+The approach takes as input intensities of individual features from modified peptides, or intensity ratios of modified and unmodified peptide features, and compares the mean abundance of a PTM site from one condition to another \todo{add references that substantiate that claim}.
 Modifications of the t-test such as moderated t-test with Limma were also proposed \cite{Zhu}.
-While simple, the approach does not fully account for the sources of variations, and it is not directly applicable to experiments with complex designs, e.g., comparisons of multiple conditions, acquisition in multiple batches, etc.
-Additionally, while this approach can be applied to experiments targeting PTMs, there is not a self contained, straight forward implementation of the methods, making application challenging.
+While simple, these approaches do not fully account for the sources of variations, and are not directly applicable to experiments with complex designs, e.g., comparisons of multiple conditions, acquisition in multiple batches, etc.
+Additionally, while these approach can be applied to experiments studying PTMs, there is not a self contained, straight forward implementation of the methods, making application challenging.
 
-Isobar-PTM was developed for experiments with MS/MS quantitative strategies that employ isobaric labels such as tandem mass tags (TMT) and isobaric tag for relative and absolute quantification (iTRAQ)\cite{Breitwieser:2013}.
+More recently, Isobar-PTM was proposed for experiments with LC-MS/MS quantitative strategies that employ isobaric labels such as tandem mass tags (TMT) or isobaric tag for relative and absolute quantification (iTRAQ)\cite{Breitwieser:2013}.
 Isobar-PTM expresses MS measurements with a linear model and performs adjustment with respect to protein abundance using the difference between log-ratio of modified peptides in two channels and log-ratio of protein level.
-The modeling framework, however, is not applicable for either label-free workflows or experiments with complex designs.
+Unfortunately, this statistical modeling framework is not applicable to either label-free workflows to experiments with complex designs.
 
-Further details on the specific equations and workflow of existing methods can be seen in Supplementary Sec. \ref{supp-sec:ttest}.
+Mathematical details regarding these existing methods is in Supplementary Sec. \ref{supp-sec:ttest}. \todo{Does the section describe Isobar-PTM as well?}
 
@@
 
+\todo{I do not think you need to repeat all the MassIVE.quant ids - just refer to the table in the beginning.
+Then add that in addition to the data the repository containers also include analysis scripts and results}
+
 The details of the four experiments, including raw data, and intermediate data processing outputs are located in individual MassIVE repositories: SpikeIn benchmark - Ubiquitination - Label-free (MassIVE repository identifier: \href{https://massive.ucsd.edu/ProteoSAFe/private-dataset.jsp?task=c4c583ecf7f941cdac87f7a4f872517b}{MSV000088971}), Human - Ubiquitination - 1mix-TMT (MassIVE repository identifier: \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=b6f0c74c234247678fb0888c6df1f225}{MSV000088966}), Mouse - Phosphorylation - 2mix-TMT time series (MassIVE repository identifier: \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=4878d777c6b34cf8aaf8477e93140c4d}{MSV000085565}), Human - Ubiquitination - Label-free no global profiling run (MassIVE repository identifier: \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=1b516164de5345108b40b75147dd58b5}{MSV000078977}).

### 2022-03-16 main/MSstatsPTM_Main.tex
Small

@@
 \todo{Would it be possible to re-organize this section?
-(1) Overview, including the data structures and the figure; (2) Existing methods: limma and t-test, including the description in the notation of the figure; also Isobar-PTM - can express in the same notation?
+(1) Overview, including the data structures and the figure - essentially the section that describes Fig 1; (2) Existing methods: limma and t-test, including the description in the notation of the figure; also Isobar-PTM - can express in the same notation?
 Ok if not; (3) Existing methods: MSstats.

### 2022-03-16 main/MSstatsPTM_Main.tex
Small

--- 2022-03-16 main/MSstatsPTM_Main.tex
@@
 \todo{Would it be possible to re-organize this section?
-(1) Overview, including the data structures and the figure; (2) Existing methods: limma and t-test, including the description in the notation of the figure; also Isobar-PTM - can express in the same notation?
+(1) Overview, including the data structures  the figure and the notation - essentially the section that describes Fig 1; (2) Existing methods: limma and t-test, including the description in the notation of the figure; also Isobar-PTM - can express in the same notation?
 Ok if not; (3) Existing methods: MSstats.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection*{The proposed approach corrects for high FDR in computer simulations}
+\subsubsection*{The proposed approach corrects \todo{starting here, put all the section titles in past tense} for high FDR in computer simulations}
 

### 2022-03-17 main/MSstatsPTM_Main.tex
main changes

@@
 ```Dataset'' is the code name of the dataset in this manuscript.
-``Data availability'' shows the ID of the MassIVE.quant repository or the GitHub repository.  \todo{All the experiments were conducted in data-dependent acquisition (DDA) mode.}
+``Data availability'' shows the ID of the MassIVE.quant repository or the GitHub repository. All the experiments were conducted in data-dependent acquisition (DDA) mode.
 }
@@
 The simulation provides an idealistic case.
-\todo{generated (how many?)  repeated synthetic datasets?}.
-\todo{I have a hard time following this section.
-Would it be possible to rewrite the description, by explaining (1) how many repeated datasets, (2) how many proteins per dataset; (3) how many PTMs per protein, (4) how many modified/unmodified features per protein?
-Then talk about the quantitative aspects: systematic fold change; random variation; where the parameters of the simulation were inspired from.
-It will help relate this simulation to real data} had 1000 PTMs and 1000 unmodified proteins, each represented by 10 LC-MS features observed across all the mass spectrometry runs.
+24 synthetic datasets were generated exhibiting scenarios with different experimental designs and biological variation.
+1000 proteins were generated per dataset with 10 unmodified features per protein.
+One PTM was created for each protein in the dataset, resulting in 1000 PTMs, each with 10 modified features.
 500 of the PTMs were generated with a differential fold change between conditions, while the other half were generated with no difference between conditions.
@@
 
+\todo{generated (how many?)  repeated synthetic datasets?}.
+\todo{I have a hard time following this section.
+Would it be possible to rewrite the description, by explaining (1) how many repeated datasets, (2) how many proteins per dataset; (3) how many PTMs per protein, (4) how many modified/unmodified features per protein?
+Then talk about the quantitative aspects: systematic fold change; random variation; where the parameters of the simulation were inspired from.
+It will help relate this simulation to real data} had 1000 PTMs and 1000 unmodified proteins, each represented by 10 LC-MS features observed across all the mass spectrometry runs.
+
 \subsection*{Dataset 2 : Computer simulation 2 - Label-free missing values and low features}
@@
 The data were simulated as above, while including missing values and a low number of modified features to provide a more realistic representation of the experiments.
-In this simulation, the feature counts and the proportion of missing values were as observed \todo{on average over all the experimental datasets in this manuscript?}.
+In this simulation, the feature counts and the proportion of missing values were as observed on average over all the the experimental datasets in this manuscript.
 PTMs were simulated with 2 modified peptide features, while unmodified proteins were simulated with 10 features.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection*{The proposed approach corrects \todo{starting here, put all the section titles in past tense} for high FDR in computer simulations}
+\subsubsection*{The proposed approach corrected for high FDR in computer simulations}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection*{The proposed approach outperforms existing methods in simulations with missing values and low features}
+\subsubsection*{The proposed approach outperformed existing methods in simulations with missing values and low features}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection*{In a label-free benchmark experiment the proposed approach corrects for bias and outperforms existing methods}
+\subsubsection*{In a label-free benchmark experiment the proposed approach corrected for bias and outperformed existing methods}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection*{MSstatsPTM corrects for bias in group comparison TMT experiment}
+\subsubsection*{MSstatsPTM corrected for bias in group comparison TMT experiment}
 
@@
 
-\subsubsection*{Proposed approach removes confounding in time series TMT experiment}
+\subsubsection*{Proposed approach removed confounding in time series TMT experiment}
 
@@
 
-\subsubsection*{Noisy PTM measurements benefit from additional biological replicates}
+\subsubsection*{Noisy PTM measurements benefited from additional biological replicates}
 

### 2022-03-19 main/MSstatsPTM_Main.tex
moved sections to main

@@
 Yet frequently data from these experiments are analyzed using statistical methods that were not originally designed for this task.
-In particular, methods such as t-test, Analysis of Variance, or Limma take as input the intensity ratios of modified and unmodified peptide features, and compare the mean abundance of different PTM sites~ \cite{Ritchie_15a}.
+In particular, methods such as $t$-test, Analysis of Variance, or Limma take as input the intensity ratios of modified and unmodified peptide features, and compare the mean abundance of different PTM sites~\cite{Ritchie_15a}.
 Such methods do not fully account for all the sources of uncertainty and variation in the experiment.
@@
 Table \ref{tab:dataDescription} summarizes the experiments.
-Details of the experimental data, R scripts with $MSstatsPTM$ analysis, and the results of the statistical analysis are available in MassIVE.quant (\url{https://massive.ucsd.edu/ProteoSAFe/static/massive-quant.jsp}) \cite{Choi:2020}.
+Details of the experimental data, R scripts with $MSstatsPTM$ analysis, and results of the statistical analysis are available in MassIVE.quant (\url{https://massive.ucsd.edu/ProteoSAFe/static/massive-quant.jsp}) \cite{Choi:2020}.
 Details of computer simulations are available on GitHub (\url{https://github.com/devonjkohler/MSstatsPTM_simulations}).
@@
 %% DDA Meierhofer
-\multirow{3}{*}{\rotatebox[origin=c]{90}{Biological}} \multirow{3}{*}{\rotatebox[origin=c]{90}{Investigation}} & Human - Ubiquitination - 1mix-TMT & Group comparison & 6 & 2 & \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=b6f0c74c234247678fb0888c6df1f225}{MSV000088966} \\
+\multirow{3}{*}{\rotatebox[origin=c]{90}{Biological}} \multirow{3}{*}{\rotatebox[origin=c]{90}{Experiment}} & Human - Ubiquitination - 1mix-TMT & Group comparison & 6 & 2 & \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=b6f0c74c234247678fb0888c6df1f225}{MSV000088966} \\
 &&&&& \\%[-0.05in]
@@
 500 of the PTMs were generated with a differential fold change between conditions, while the other half were generated with no difference between conditions.
-Of the 500 differential PTMs, the fold change of 250 were fully masked by changes in the unmodified protein.
-Of the 500 non-differential PTMs, 250 were generated with a differential fold change that was entirely due to the unmodified protein.
+The fold change of half the 500 differential PTMs were fully masked by changes in the unmodified protein.
+Additionally, the fold change of half the 500 non-differential PTMs was entirely due to changes in the unmodified protein.
 All differential PTMs were generated with an expected log fold change of .75 between conditions.
@@
 The simulated peptide abundance  $Z_{ibc}$ was generated as $Z_{ibc} = Y_{ibc} + \epsilon_{ibc}$ where $\epsilon_{ibc}\sim^{iid}N(0,\sigma^2)$.
-The simulations were generated with one of two values $\sigma^2 = \{.2, .3\}$ motivated by the biological experiments in this manuscript.
+The simulations were generated with one of two values $\sigma^2 = \{.2, .3\}$.
+The values for $\sigma^2$ were motivated by the biological experiments in this manuscript.
 Further details can be found in Supplementary Sec. \ref{supp-sec:dataset1}.
-
-\todo{generated (how many?)  repeated synthetic datasets?}.
-\todo{I have a hard time following this section.
-Would it be possible to rewrite the description, by explaining (1) how many repeated datasets, (2) how many proteins per dataset; (3) how many PTMs per protein, (4) how many modified/unmodified features per protein?
-Then talk about the quantitative aspects: systematic fold change; random variation; where the parameters of the simulation were inspired from.
-It will help relate this simulation to real data} had 1000 PTMs and 1000 unmodified proteins, each represented by 10 LC-MS features observed across all the mass spectrometry runs.
 
@@
 \todo{There will likely be a comment that missing values are more common among low-abundance features.
-Did the simulation account for that?} Further details can be found in Supplementary Sec.
+Did the simulation account for that?
+Devon: The simulation did not include any missing values due to low abundance.
+Should we mention this or just leave it as is?} Further details can be found in Supplementary Sec.
 \ref{supp-sec:dataset2}.
@@
 
-The custom spike-in experiment had 4 conditions with two biological replicates per condition, and a balanced design. Additional information can be found in Supplementary Sec. \ref{supp-sec:benchmark}.
+In the custom spike-in experiment 8 biological replicates were allocated to 4 conditions, resulting in 2 biological replicates per condition, and a balanced design.
+Additional information can be found in Supplementary Sec.
+\ref{supp-sec:benchmark}.
 
@@
 Two sets of data were acquired for each mixture in label-free LC-MS/MS mode: KGG enriched + LC-MS/MS, and LC-MS/MS only.
-\todo{How the biological replicates in each condition were obtained?}
 
@@
 \label{sec:exp_proc_dataset4}
-\todo{In this experiment XXX biological replicates were allocated to 1 TMT mixture in an unbalanced design.} Further details are in Supplementary Sec. \ref{supp-sec:ipah}.
+In this experiment 11 biological replicates were allocated to 1 TMT mixture in an unbalanced design. Further details are in Supplementary Sec. \ref{supp-sec:ipah}.
 
@@
 Uninfected cells were measured at 0 and 6 hours, while infected cells were measured at 1, 2, 4, and 6 hour increments, resulting in six total conditions.
-The proteome was quantified in 11-plex \todo{?} isobaric multiplexing via tandem mass tagging (TMT) in combination with LC-MS/MS.
-\todo{One MS run total?}
-
-\medskip \noindent {\bf Global Profiling Run:} This experiment included a separate global profiling run \todo{with the same design?} to measure unmodified peptides.
+The proteome was quantified using isobaric multiplexing via tandem mass tagging (TMT) in combination with LC-MS/MS.
+\todo{One MS run total?
+Only one MS run was used.
+Not sure the best way to say this.}
+
+\medskip \noindent {\bf Global Profiling Run:} This experiment included a separate global profiling run, with the same design, to measure unmodified peptides.
 There was a 95\% overlap between the identified modified peptides and proteins that were quantified in the global profiling run.
@@
 
-\medskip \noindent {\bf Global Profiling Run:} This experiment included a separate global profiling run \todo{with the same design?} to measure unmodified peptides.
+\medskip \noindent {\bf Global Profiling Run:} This experiment included a separate global profiling run with the same design to measure unmodified peptides.
 There was a 90\% overlap between the identified background modified peptides and proteins that were quantified in the global profiling run.
@@
 
-In this experiment 8 biological replicates were split between 4 conditions in a balanced design. \todo{Further details in Suppler Sec XXX?}
+In this experiment 8 biological replicates were split between 4 conditions in a balanced design. Further details in Suppler Sec \ref{supp-sec:usp30}.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section*{Results}
+\section*{Background Information}
+
+\subsection*{Experimental Design}
 
 \todo{Would it be possible to re-organize this section?
-(1) Overview, including the data structures  the figure and the notation - essentially the section that describes Fig 1; (2) Existing methods: limma and t-test, including the description in the notation of the figure; also Isobar-PTM - can express in the same notation?
+(1) Overview, including the data structures  the figure and the notation - essentially the section that describes Fig 1; (2) Existing methods: limma and $t$-test, including the description in the notation of the figure; also Isobar-PTM - can express in the same notation?
 Ok if not; (3) Existing methods: MSstats.
@@
 
-\subsection*{Existing statistical methods for experiments targeting PTMs}
-
-Despite the importance of PTMs for biological function, there is a lack of general statistical analysis framework that summarizes the available quantitative information from LC-MS/MS experiments, performs statistical inference, and characterizes the quantitative properties of PTM in a statistically rigorous manner.
-
-Many investigations performed differential abundance analysis of PTMs using two-sample t-test or its extensions.
-The approach takes as input intensities of individual features from modified peptides, or intensity ratios of modified and unmodified peptide features, and compares the mean abundance of a PTM site from one condition to another \todo{add references that substantiate that claim}.
-Modifications of the t-test such as moderated t-test with Limma were also proposed \cite{Zhu}.
-While simple, these approaches do not fully account for the sources of variations, and are not directly applicable to experiments with complex designs, e.g., comparisons of multiple conditions, acquisition in multiple batches, etc.
-Additionally, while these approach can be applied to experiments studying PTMs, there is not a self contained, straight forward implementation of the methods, making application challenging.
-
-More recently, Isobar-PTM was proposed for experiments with LC-MS/MS quantitative strategies that employ isobaric labels such as tandem mass tags (TMT) or isobaric tag for relative and absolute quantification (iTRAQ)\cite{Breitwieser:2013}.
-Isobar-PTM expresses MS measurements with a linear model and performs adjustment with respect to protein abundance using the difference between log-ratio of modified peptides in two channels and log-ratio of protein level.
-Unfortunately, this statistical modeling framework is not applicable to either label-free workflows to experiments with complex designs.
-
-Mathematical details regarding these existing methods is in Supplementary Sec. \ref{supp-sec:ttest}. \todo{Does the section describe Isobar-PTM as well?}
-
-\subsection*{Proposed approach}
+\todo{Devon: I moved most everything around as recommended.
+The notation is mainly described once we start getting into the equations (t-test section).
+We can move this into the design section but I thought it was clearer to describe it with the equations.
+The results section now starts with the proposed method in the same notation.
+The supplementary now only includes the table and extra results.}
 
@@
 A PTM site is quantified with multiple spectral features, which vary in sequence (e.g., fully or partially cleaved peptides), ionization efficiency, charge states, etc.
-The number quantified features vary across replicate LC-MS/MS runs of the same sample, and across conditions.
+Additionally, some spectral features can be missing.
+The feature intensities are viewed as repeated measurements of the underlying abundance of the PTM, where the abundance in Condition $i$ is denoted by $i^{\ast}$.
+The number of quantified features vary across replicate LC-MS/MS runs of the same sample, and across conditions.
 To perform adjustment with respect to protein abundance, features of unmodified peptides are used for the inference of underlying protein abundance.
+Features corresponding to unmodified peptides are considered together, where the protein abundance in Condition $i$ is denoted by $i$.
 Typically, because of the enrichment step for PTMs, very few of those features are present in original LC-MS runs.
@@
 If unmodified peptides are unavailable for any given modification, unmodified intensity adjustment cannot be performed.
-As different levels of variability are present in the data, the log-intensities of the features for modified and unmodified peptides are modeled separately using two linear mixed models.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection*{Existing work in statistical modeling and parameter estimation}
+
+\subsection*{Existing statistical methods for experiments targeting PTMs}
+
+Despite the importance of PTMs for biological function, there is a lack of general statistical analysis framework that summarizes the available quantitative information from LC-MS/MS experiments, performs statistical inference, and characterizes the quantitative properties of PTM in a statistically rigorous manner.
+
+Many investigations performed differential abundance analysis of PTMs using two-sample $t$-test or its extensions.
+The approach takes as input intensities of individual features from modified peptides, or intensity ratios of modified and unmodified peptide features, and compares the mean abundance of a PTM site from one condition to another~\cite{Schwammle2015}\cite{THOMAS2020}.
+Modifications of the $t$-test such as moderated $t$-test with Limma were also proposed~\cite{Ritchie_15a}\cite{Zhu}.
+While simple, these approaches do not fully account for the sources of variations, and are not directly applicable to experiments with complex designs, e.g., comparisons of multiple conditions, acquisition in multiple batches, etc.
+Additionally, while these approaches can be applied to experiments studying PTMs, there is not a self contained, straight forward implementation of the methods, making application challenging.
+
+Isobar-PTM was also proposed for experiments with LC-MS/MS quantitative strategies that employ isobaric labels such as tandem mass tags (TMT) or isobaric tag for relative and absolute quantification (iTRAQ)\cite{Breitwieser:2013}.
+Isobar-PTM expresses MS measurements with a linear model and performs adjustment with respect to protein abundance using the difference between log-ratio of modified peptides in two channels and log-ratio of protein level.
+Unfortunately, this statistical modeling framework is not applicable to either label-free workflows or experiments with complex designs.
+
+An overview of the existing approach using $t$-test and Limma are shown in the following sections.
+
+\subsubsection*{Two-sample $t$-test}
+
+Two-sample $t$-test is based on the null hypothesis that there is no difference in mean PTM abundance between Conditions $i$ and $i^*$.
+There are $I$ conditions and $J$ mass spectrometry runs (technical replicates) per condition in the experiment.
+The PTM site is represented by $K$ spectral features (peptide ions, distinguished by their cleavage residues and charge states).
+The log-intensity (base 2) of Feature $k$, in Run $j$ of Condition $i$ is denoted by $y_{ijk}^{\ast}$.
+Conversely, the unmodified protein is represented by $L$ spectral features, and the log-intensity of Feature $l$ from the unmodified peptides in the same run is denoted by $y_{ijl}$.
+The abundance in each run is taken as input and is often estimated by sum of peak intensities.
+For example, the log-abundance estimate for the PTM in Run $j$ of Condition $i$ is given by
+\[
+\log \left( \sum_{k=1}^{K} 2^{y_{ijk}^{\ast}} \right).
+\]
+For adjustment with respect to unmodified peptides, the estimate of PTM abundance is divided by the protein abundance estimate, 
+and the $t$-test for the adjusted PTM abundance on log scale takes as input the difference of their log-estimates.
+%which is equivalent to take the difference of their log abundances. 
+The quantity is denoted by $d_{ij}$ and is given by
+\[
+d_{ij} = \log \left( \sum_{k=1}^{K} 2^{y_{ijk}^{\ast}} \right) - \log \left( \sum_{l=1}^{L} 2^{y_{ijl}} \right).
+\]
+Alternatively, run-level summarization can also be used. The difference between the means of PTM abundance in Conditions $i$ and $i'$ are estimated as
+\[
+\hat{\Delta} = \frac{1}{J} d_{i+}^{\ast} - \frac{1}{J} d_{i'+},
+\]
+where $d_{i+}^{\ast} = \sum_{j=1}^{J} d_{ij}^{\ast}$ and the test statistic for the $t$-test is given by $\hat{\Delta} / \mathrm{SE}(\hat{\Delta})$.
+The statistical significance of the difference is determined by comparing the test statistic against the $t$ distribution, with degrees of freedom $df=2J-2$ in balanced designs.
+
+\subsubsection*{Limma}
+
+Limma uses linear models to test the null hypothesis that there is no difference in mean PTM abundances between conditions.
+It leverages Empirical Bayes moderation to share pooled variance information across individual modification models and moderate the individual residual variances.
+Using a linear model allows Limma to share variance information across conditions, providing a more accurate estimate~\cite{Ritchie_15a}.
+
+With respect to PTM analysis, feature level summarization and adjustment for changes in the unmodified peptides are performed in the same way as $t$-test.
+After fitting a linear model, the final variance is moderated using the global variance across all modified peptides.
+This is done as $s_{i} = P(v_i| v_\Theta)$ where $v_i$ is the resulting variance when modeling PTM $i$ and $v_\Theta$ is the global variance.
+
+\subsection*{Existing work in statistical modeling and parameter estimation}
 
 The proposed approach leverages the large amount of work done to create $MSstats$~\cite{Choi:2014}.
-It takes as input a list of log-transformed intensities of spectral features, identified and quantified across LC-MS runs.
-The features, which are precursor ions of modified or unmodified peptides, are used to characterize the identified PTM sites and proteins.
-For each PTM site, the feature log-intensities of the modified peptides spanning the site are expressed using a linear mixed model in consideration of the effects of condition, run, feature and interaction between run and feature.
-The model parameters are estimated using the split-plot approach as in $MSstats$, where the feature log-intensities are first summarized into a single value per site per run in the subplot model, and the site-level summaries are then used for the inference of the PTM site abundance \cite{Choi:2014}.
-This approach allows the method to be extended to cases where the experimental design is unbalanced, and where additional sources of variation are present.
-In the site-level summarization, Tukey's median polish (TMP), a simple and robust procedure is applied to iteratively fit a two-way additive model with the effects of run and feature, which in turn summarizes the log-intensities for each site \cite{Tukey:1977}.
-After summarization, the inference of the PTM site abundance in each condition is carried through fitting a model based on the family of linear mixed-effects models \cite{Bolker2009} \cite{Faraway:2006}.
-Statistical modeling and quantification for global proteomics data are performed by the same procedure as for PTM data.
-
-Additionally, the proposed method can include imputing missing feature intensities if desired.
+$MSstats$ takes as input a list of log-transformed intensities of spectral features which are used to characterize the identified protein.
+For each protein, the feature log-intensities are expressed using a linear mixed model in consideration of the effects of condition, run, feature and interaction between run and feature.
+The model parameters are estimated using a split-plot approach, where the feature log-intensities are first summarized into a single value per protein per run in the subplot model, and the protein-level summaries are then used for the inference of the protein abundance.
+This approach allows the method to be extended to cases where the experimental design is unbalanced and where additional sources of variation are present.
+In the protein-level summarization, Tukey's median polish (TMP), a simple and robust procedure, is applied to iteratively fit a two-way additive model with the effects of run and feature, which in turn summarizes the log-intensities for each protein~\cite{Tukey:1977}.
+After summarization, the inference of the protein abundance in each condition is carried through fitting a model based on the family of linear mixed-effects models \cite{Bolker2009} \cite{Faraway:2006}.
+We leverage this statistical modeling and quantification workflow and apply it to PTMs, in addition to unmodified proteins.
+
+Additionally, $MSstats$ provides a method for imputing missing feature intensities if desired.
 When values are imputed, it is assumed they are missing for reasons of low abundance.
-Missing features are imputed in each MS run using the Accelerated Time Failure (AFT) model \cite{Tukey:1977}.
+Missing features are imputed in each MS run using the Accelerated Time Failure (AFT) model~\cite{Tukey:1977}.
 In order to impute a feature's missing values, the feature must be present in at least one MS Run.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section*{Results}
+
 \subsubsection*{Detection of changes in PTMs - Balanced design with one source of variation}
@@
 
-Here we propose a statistical method for detecting changes in PTMs and removing confounding with changes in global protein abundance.
-The method is shown when applied to a balanced design with multiple conditions and one technical replicate.
-In practice the experimental design can be unbalanced and variation can come from multiple sources.
-We show how the methods can be applied to more complex experimental designs in the next section.
+We propose a statistical method for detecting changes in PTMs while removing confounding with changes in global protein abundance.
+Here the method is applied to a balanced design with multiple conditions and one technical replicate.
+In practice the experimental design can be unbalanced and variation can come from multiple sources, as shown in the next section.
 
@@
 The null hypothesis states that there is no difference in log-abundance of the PTM site between conditions, adjusted with respect to protein abundance.
-There are $I$ conditions and $J$ mass spectrometry runs (technical replicates) per condition in the experiment.
-The PTM site is represented by $K$ spectral features (peptide ions, distinguished by their cleavage residues and charge states).
-The log-intensity (base 2) of Feature $k$, in Run $j$ of Condition $i$ is denoted by $y_{ijk}^{\ast}$.
-Conversely, the log-intensity of Feature $l$ from the unmodified peptides in the same run is denoted by $y_{ijl}$.
-The expected values of log-abundances of the PTM and protein in Condition $i$ are denoted by $\mu_{i}^{\ast}$ and $\mu_{i}$.
+The expected values of log-abundances of the PTM and protein in Condition $i$ are denoted by $\mu_{i}^{\ast}$ and $\mu_{i}$ respectively.
 The feature and run level summarization for condition $i$ is denoted as $\hat{y}_{i+}^{\ast}$ for the modified peptide and $\hat{y}_{i+}$ for the unmodified peptide.
@@
 
-Finally the degrees of freedom are calculated using the following formula.
+Finally the degrees of freedom are calculated as follows.
 
@@
 The test statistic for the hypothesis testing is the ratio of the estimate of the adjusted difference to its SE.
-To determine the statistical significance of the difference in terms of p-value, the test statistic is compared against the t distribution with degrees of freedom approximated by the Satterthwaite method\cite{satterthwaite:1946}.
-Adjustment for multiple comparisons is performed using the Benjamini-Hochberg procedure to control the false discovery rate at a desired level, e.g., $0.05$\cite{Benjamini:1995}.
+To determine the statistical significance of the difference in terms of p-value, the test statistic is compared against the $t$ distribution with degrees of freedom approximated by the Satterthwaite method~\cite{satterthwaite:1946}.
+Adjustment for multiple comparisons is performed using the Benjamini-Hochberg procedure to control the false discovery rate at a desired level, e.g., $0.05$~\cite{Benjamini:1995}.
 
@@
 
-The statistical modeling approaches discussed above can also be extended to complex designs including experiments with additional sources of variation and unbalanced designs.
-Equation 1 is the same in this case and the null hypothesis is still that there is no difference in mean PTM abundance between Conditions $i$ and $i'$ after adjusting for changes in mean unmodified peptide abundance between Conditions $i$ and $i'$.
+The statistical modeling approaches discussed above can be extended to complex designs including experiments with additional sources of variation and unbalanced designs.
+In this case our null hypothesis and Equation 1 are unchanged.
 
@@
 
-The standard error of the estimate $\mathrm{SE}(\hat{\Delta})$ and degrees of freedom in Equations 3 and 4 are unchanged from the balanced case, although $\hat{\sigma}_{\gamma^z\ast}$ and $\hat{\sigma}_{\gamma}$ are now estimated using RML.
+The standard error of the estimate $\mathrm{SE}(\hat{\Delta})$ and degrees of freedom in Equations 3 and 4 are unchanged from the balanced case, although $\hat{\sigma}_{\gamma^\ast}$ and $\hat{\sigma}_{\gamma}$ are now estimated using RML.
 
@@
 This includes a quality control plot, summarizing the peptide abundance per run in the form of a boxplot, and a profile plot, plotting each feature and the overall feature summarization as a line plot over each run.
-Additionally, the model plots include a volcano plot, showing all peptides adjusted pvalues and fold changes, as well as a heatmap, which evaluates the fold change between conditions and peptides.
+Additionally, the model plots include a volcano plot, showing all peptides adjusted p-values and fold changes, as well as a heatmap, which evaluates the fold change between conditions and peptides.
 
@@
 For the experiments where ground truth was known, we calculated the number of true positives (TP), false positives (FP), true negatives (TN), and false negatives (FN).
-True positives were modified peptides that are differential after accounting for changes in global protein abundance.
-True negatives were modified peptides that were not differentially abundant after accounting for global protein abundance.
+True positives were modified peptides that were differential after accounting for changes in global protein abundance.
+True negatives were modified peptides that were not differential after accounting for changes in global protein abundance.
 To determine positives and negatives we adjusted the p-values for multiple testing and used an $\alpha=.05$ cutoff to determine differential sites.
 These results were then compared to the ground truth to determine TP and TN.
-We used these results to calculate the false discovery rate (FDR = FP/(TP+FP)), the recall rate (TP/(TP + FN), and accuracy ((TP + TN) / (TP + TN + FP + FN)).
+We used these results to calculate the false discovery rate (FDR = FP/(TP+FP)), recall rate (TP/(TP + FN), and accuracy ((TP + TN) / (TP + TN + FP + FN)).
 Finally, the summary statistics were compared across methods to analyze method performance.
@@
 
-The proposed approach was evaluated against two-sample t-test and Limma in computer simulations. 
-In Figure \ref{fig:sim1_fdr} we see that not accounting for changes in unmodified peptide level results in a high false discovery rate.
-When adjusting for changes in the unmodified peptide, all methods performed similarly in terms of FDR.
-In a clean simulation, recall and accuracy of the proposed approach and Limma performed similarly, with Limma performing slightly better with low replicates.
-In contrast, the performance of the t-test method lagged behind the others.
-Two-sample t-test only uses data within the groups of interest while ignoring the remainder of the data, leading to similar performance across all simulations.
+The proposed approach was evaluated against two-sample $t$-test and Limma in computer simulations. 
+In Figure \ref{fig:sim1_fdr} we see that not accounting for changes in the unmodified protein resulted in a high false discovery rate.
+After adjusting for changes in the unmodified protein, all methods performed similarly in terms of FDR.
+In a clean simulation, recall and accuracy of the proposed approach and Limma performed similarly, although Limma performed slightly better in simulations with fewer biological replicates.
+In contrast, the performance of the $t$-test method lagged behind the others.
+Two-sample $t$-test only uses data within the groups of interest while ignoring the remainder of the data, leading to similar performance across all simulations.
 In contrast, the proposed approach and Limma leverage all available information, which resulted in improved power in simulations with more conditions.
@@
 
-In this simulation the proposed method outperformed t-test and limma when real world experimental conditions were introduced.
-Changes in unmodified peptide level still needed to be accounted for to control the FDR.
-Once controlling for changes in the unmodified peptide, the proposed method out performed the other methods, as seen in Figure \ref{fig:sim2_acc}.
+In this simulation the proposed method outperformed $t$-test and Limma.
+Changes in unmodified protein level still needed to be accounted for to control the FDR.
+Once controlling for changes in the unmodified protein, the proposed method outperformed the other methods, as seen in Figure \ref{fig:sim2_acc}.
 The proposed method calibrated model accuracy and recall rate well, even when the number of replicates were low.
-Additionally, when comparing the fold change estimation across all peptides, the proposed method showed a tighter distribution of estimated fold changes around the true fold change than t-test and Limma.
-Specifically, the inner quartile range of the estimated fold change for the proposed method was on average 10.4\% smaller than t-test and 21.8\% smaller than Limma.
+Additionally, when comparing the fold change estimation across all modified peptides, the proposed method showed a tighter distribution of estimated fold changes around the true fold change, as compared to $t$-test and Limma.
+Specifically, the inner quartile range of the estimated fold change for the proposed method was on average 10.4\% smaller than $t$-test and 21.8\% smaller than Limma.
 This fold change comparison can be seen in Supplementary Figure \ref{supp-fig:fc_boxplot}.
 
-Further details reviewing the results of each simulation and their comparison with respect to method performance can be seen in Supplementary Sec. \ref{supp-sec:sim}.
+Further details reviewing the results of each simulation and  method comparison can be seen in Supplementary Sec. \ref{supp-sec:sim}.
 
@@
 In this experiment all models incorrectly estimated the fold change of the modified spike-in peptides before adjusting for changes in global protein abundance.
-After adjustment, the spike-in peptides' fold change was in line with expectation (Figure \ref{fig:spikein_boxplot}).
-While the median fold change for the spike-in peptides was generally in line with the true fold change after adjustment, we can see that the distribution of fold changes for the peptides was visibly wider after adjustment.
-However, the proposed method a tighter distribution around the true log fold change.
-Comparing the inner quartile range (IQR) of the spike-in peptide's log fold change, the proposed method's IQR was 36.78\% smaller than Limma's and 32.98\% smaller than t-test's.
-This means that the proposed method's fold change estimate was more consistently correct, when compared to Limma and t-test.
-
-In Figure \ref{fig:spikein_prop_volcano} we can clearly see the red labeled spike-in peptides do not follow the expected fold change before adjusting for changes in global protein level.
-After adjustment the estimated fold change was more in line with expectation.
+After adjustment, the spike-in peptides' fold change was generally in line with expectation (Figure \ref{fig:spikein_boxplot}), however the distribution of estimated fold changes was visibly wider in all methods.
+Of the three approaches, the proposed method showed the tightest distribution around the true log fold change.
+Comparing the inner quartile range (IQR) of the spike-in peptide's log fold change, the proposed method's IQR was 36.78\% smaller than Limma's and 32.98\% smaller than $t$-test's.
+This means that the proposed method's fold change estimate was more consistently correct, when compared to Limma and $t$-test.
+
+In Figure \ref{fig:spikein_prop_volcano} we can clearly see the red labeled spike-in peptides do not follow the expected fold change before adjusting for changes in global protein abundance.
+After adjustment, the estimated fold change was more in line with expectation.
 Additionally, the background peptides serving as the null model, show many false positives before adjustment was made.
@@
 Using both the proposed method and Limma, the red labeled spike-in peptides were more inline with expected fold change after the adjustment was applied.
-However, when using Limma we can see that the majority of the differential spike-in peptides were not showing a significant adjusted pvalue.
+However, using Limma resulted in the majority of the differential spike-in peptides not showing a significant adjusted p-value.
 In this case, using Limma would have resulted in missing the majority of differentially modified peptides.
@@
 We can see that more peptides became insignificant after adjustment than became significant.
-3,236 peptides became insignificant, 1352 peptides became significant, while 4,282 peptides were significant in both models.
+3,236 modified peptides became insignificant, 1352 became significant, while 4,282 were significant in both models.
 For the peptides that became insignificant in the adjusted model, their change in abundance was mainly due to changes in global protein abundance.
@@
 When this test was applied on this experiment, only one peptide became insignificant due to an increase in variance.
-Thus we can conclude that the drop off in significant peptides was truly due to changes in global protein abundance.
-
-In Figure \ref{fig:data4_profile_plot}, looking at the modification of protein $GSDMD$ at $K62$ we can see the advantage of the proposed method.
-The modified peptide shows a flat abundance change between the infected 1 hour, 4 hour, and 6 hour conditions.
-This is contrasted with a strong negative change in the global profiling run between the same conditions.
-Looking at the Dox4hr vs Dox1hr conditions and modeling the modified peptide without adjusting for changes in the global profiling, the fold change was $-.501$ and the adjusted pvalue was insignificant at $.0644$.
-After adjusting for changes in the global protein abundance, the fold change was much higher, $2.79$, and the adjusted pvalue became significant, $5.25e-8$.
+Thus we can conclude that the drop off in significant peptides was due to changes in global protein abundance.
+
+In Figure \ref{fig:data4_profile_plot}, the modification of protein $GSDMD$ at $K62$ showed the advantage of the proposed method.
+The modified peptide originally showed no abundance change between the infected 1 hour, 4 hour, and 6 hour conditions.
+This was contrasted with a strong negative change in the global profiling run between the same conditions.
+Looking at the Dox4hr vs Dox1hr conditions and modeling the modified peptide without adjusting for changes in the global profiling, the fold change was $-.501$ and the adjusted p-value was insignificant at $.0644$.
+After adjusting for changes in the global protein abundance, the fold change was much higher, $2.79$, and the adjusted p-value became significant, $5.25e-8$.
 In this case the effect of the modified peptide was strongly confounded with changes in the global protein.
@@
 Again more PTMs became insignificant after adjustment than became significant.
-19,286 peptides become insignificant, 4,947 became significant, while 41,552 peptides are significant in both models.
-Again we tested if the decrease in significant peptides was due to the increased variance that comes from adjustment, or if it was mainly due to removing convolution with change in the umodified protein.
+19,286 peptides became insignificant, 4,947 became significant, and 41,552 were significant in both models.
+Again we tested if the decrease in significant peptides was due to increased variance or if it was due to removing convolution with changes in the umodified protein.
 This was tested by looking for modified peptides whose adjusted log fold change was within 10\% of the unadjusted log fold change but became insignificant after adjustment.
-When this test was applied on this experiment, 548 peptides became insignificant due to an increase in variance.
-This is still a small portion (3.4\%) of the peptides that became insignificant after adjustment.
+When this test was applied, 548 peptides became insignificant due to an increase in variance.
+This is only 3.4\% of the total peptides that became insignificant after adjustment.
 Thus we can conclude that the drop off in significant peptides was mainly due to changes in global protein abundance.
 
-In Figure \ref{fig:data5_profile_plot} the profile plot of protein $TTP$ modified at site $S178$ shows the power of the proposed method.
-The profile plot shows a large positive log fold change of $2.9$ between the WT\_Late and WT\_Uninfected conditions.
-However, the global profiling run shows a similar log fold change of $2.014$ between the same conditions.
-This indicates that the abundance change in the modified peptide is nearly entirely due to changes in the global protein.
-When adjusting for the global protein the modified peptide's adjusted pvalue became insignificant, going from $.0009$ to $.248$.
+In Figure \ref{fig:data5_profile_plot} the profile plot of protein $TTP$ modified at site $S178$ showed the power of the proposed method.
+Without adjustment, there was a large positive log fold change of $2.9$ between the WT\_Late and WT\_Uninfected conditions.
+However, the global profiling run showed a similar log fold change of $2.014$ between the same conditions.
+This indicated that the abundance change in the modified peptide is nearly entirely due to changes in the global protein.
+When adjusting for the global protein the modified peptide's adjusted p-value became insignificant, going from $.0009$ to $.248$.
 Correcting for changes in global protein abundance allowed us to see the true impact of the modification at $S178$ which would have otherwise been challenging to perceive.
@@
 Once identification and quantification of the Ubiquitination profiling was performed, peptides which were unmodified were extracted and used in place of a global profiling run.
-This resulted in a significant lack of overlap between modified and unmodified peptides.
+This resulted in a lack of overlap between modified and unmodified peptides.
 Any modified peptide without a corresponding unmodified protein could not be adjusted.
 Of the 10,799 modified peptides identified, only 4526 had a corresponding unmodified protein and could be adjusted.
-Additionally, not having a separate global profiling run resulted in low feature counts for the unmodified protein model.
+Additionally, the lack of a separate global profiling run resulted in low feature counts for the unmodified protein model.
 
 The results of this experiment are summarized in Figure \ref{fig:data6_plots}.
-In \ref{fig:data6_vd1} the number of significant modified peptides before and after protein adjustment are shown.
-In the top plot we can see there were a large number of peptides that become insignificant after adjustment.
-This was mainly due to the lack of overlap between modified and unmodified peptides, stemming from not having a global profiling run.
-In the bottom plot only modified peptides that could be adjusted are shown.
-Here there were much less peptides that become insignificant after adjustment.
-In this plot we see that 726 peptides became insignificant, 547 became significant, and 1,078 are significant in both models.
-As in the previous experiments, we can check if these peptides are becoming insignificant due to an increase in variance.
-To check this we look at modified peptides whose adjusted fold change was within 10\% of their unadjusted fold change and they became insignificant after adjustment.
-In this experiment there were only 25 peptides that meet this criteria, a small percentage of all modified peptides.
-
-Analysis of modifications before and after protein adjustment can be seen in Supplementary Sec. \ref{supp-sec:usp30}.
+After adjusting for changes in the unmodified protein, there were fewer significant modified peptides.
+In total, 2709 modified peptides became insignificant, while only 547 became significant.
+However, this was mainly due to not having a global profiling run, resulting in a lack of overlap between modified and unmodified peptides.
+In the bottom plot, only modified peptides that could be adjusted are shown.
+Here there were much fewer peptides that become insignificant after adjustment.
+726 modified peptides became insignificant, 547 became significant, and 1,078 were significant in both models.
+As in the previous experiments, we can check if these peptides became insignificant due to an increase in variance.
+To check this we looked at modified peptides whose adjusted fold change was within 10\% of their unadjusted fold change and which became insignificant after adjustment.
+In this experiment there were only 25 modified peptides that meet this criteria.
+
+Further analysis of modifications before and after protein adjustment can be seen in Supplementary Sec. \ref{supp-sec:usp30}.
 
@@
 
-Here we analyze the sample size needed to achieve a desired statistical power.
-The proposed approach corrects for confounding between the modified peptide and unmodified protein at the cost of increased variation.
+\todo{I would like to move this section to the supplementary and just referencing it in the main.}
+
+Here we analyzed the sample size needed to achieve a desired statistical power.
+The proposed approach corrected for confounding between the modified peptide and unmodified protein at the cost of increased variation.
 This can be seen in the calculation for variance in Section \ref{sec:adjustment_summary}.
-When the variation is increased, the experiment requires a larger number of replicates to reach the same power.
-Thus the statistical power is dependent on the variance from both the modified peptide and unmodifed protein as well as the number of replicates in each run.
+Increased variation required a larger number of replicates to reach the same power.
+Thus the statistical power was dependent on the variance from both the modified peptide and unmodifed protein.
 
@@
 In terms of the number of replicates, we tested scenarios with equal replicates in both the modified and unmodified runs, as well as scenarios where the replicates differed between runs.
-For variance we used the biological experiments to determine what scenarios and values to test.
+We used the biological experiments to determine what variance values to test.
 In datasets 4 and 5 the variance of the PTM was higher than the global protein.
-In dataset 6 the variance of the PTM and Protein was generally the same due to not using a separate global profiling run.
+In dataset 6 the variance of the PTM and Protein was generally the same.
 We mimicked these scenarios and analyzed the power of experiments when the PTM variance was higher than the protein and when they were equal.
@@
 In cases where the replicates were unbalanced, but the variance was still the same, it did not matter if there were more replicates in the modified or unmodified runs.
-As long as there were the same overall number of replicates the results were the same.
 In comparison, with differing variance and equal replicates, higher replicates still lead to higher power.
 When the replicates were unbalanced and the variance was higher in the PTM, there was more power when there were more replicates in the PTM than the protein.
-It is clearly more important to have high replicates for the PTM run than the unmodified protein, when the PTM variance is higher.
-In cases where the number of replicates has to be limited, it is better to weigh them on the PTM side.
+In this case it was clearly more important to have high replicates for the PTM run than the unmodified protein.
+In cases where the number of replicates has to be limited, it was better to add more to the PTM side.
 
@@
 The framework is general and is applicable to a variety of experimental designs.
-It outperforms the ad-hoc methods underlying t-test and Limma, and yields accurate results in the broad type of experimental circumstances, including the presence of missing values, changes in protein abundance, few representative peptides, and different acquisition methods.
+It outperforms the ad-hoc methods underlying $t$-test and Limma, and yields accurate results in the broad type of experimental circumstances, including the presence of missing values, changes in protein abundance, few representative peptides, and different acquisition methods.
 The framework allows us to plan for subsequent experiments, and choose the appropriate number of replicates in consideration of adjustment with respect to protein abundance.
@@
 
-\todo{I do not think you need to repeat all the MassIVE.quant ids - just refer to the table in the beginning.
-Then add that in addition to the data the repository containers also include analysis scripts and results}
-
-The details of the four experiments, including raw data, and intermediate data processing outputs are located in individual MassIVE repositories: SpikeIn benchmark - Ubiquitination - Label-free (MassIVE repository identifier: \href{https://massive.ucsd.edu/ProteoSAFe/private-dataset.jsp?task=c4c583ecf7f941cdac87f7a4f872517b}{MSV000088971}), Human - Ubiquitination - 1mix-TMT (MassIVE repository identifier: \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=b6f0c74c234247678fb0888c6df1f225}{MSV000088966}), Mouse - Phosphorylation - 2mix-TMT time series (MassIVE repository identifier: \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=4878d777c6b34cf8aaf8477e93140c4d}{MSV000085565}), Human - Ubiquitination - Label-free no global profiling run (MassIVE repository identifier: \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=1b516164de5345108b40b75147dd58b5}{MSV000078977}).
+The details of the four experiments, including raw data, and intermediate data processing outputs are located in individual MassIVE repositories, referenced in \ref{tab:dataDescription}.
 
@@
 Data structure of a typical PTM experiment and goals of PTM characterization.
+(a) Schematic data representation, in a simplified case of two conditions and two replicate runs.
+Each PTM site is modeled and characterized separately, where a PTM is quantified with multiple spectral features (boxes), distinguished by different charge states of a peptide.
+The feature intensities the PTM in Condition i are denoted by i*.
+Features corresponding to unmodified peptides in Condition i is denoted by i.
+Peptides can be fully cleaved (solid lines) and/or partially cleaved (dashed lines).
+(b) PTM relative quantification by statistical inference, which makes use of the feature intensities to infer the underlying PTM abundance and protein abundance with an estimate of associated uncertainty.
+(c) Model-based testing for differential PTM abundance, which corrects for the underlying protein abundance with a cost of increased uncertainty about the estimate of difference between conditions.}
+\label{fig:data-structure}
+\end{figure}
+
+%Data structure of a typical PTM experiment and goals of PTM characterization.
 (a) Schematic data representation, in a simplified case of two conditions and two replicate runs.
@@
 (b) PTM relative quantification by statistical inference, which makes use of the feature intensities to infer the underlying PTM abundance and protein abundance with an estimate of associated uncertainty.
-(c) Model-based testing for differential PTM abundance, which corrects for the underlying protein abundance with a cost of increased uncertainty about the estimate of difference between conditions.}
-\label{fig:data-structure}
-\end{figure}
+(c) Model-based testing for differential PTM abundance, which corrects for the underlying protein abundance with a cost of increased uncertainty about the estimate of difference between conditions.
 
@@
 In comparison, the methods without accounting for the protein-level changes resulted in off-target, high false positive rates.
-b) The advantage of using the proposed approach is apparent when including limited observations and missing values.
-Looking at accuracy, the proposed method outperforms Limma and t-test in nearly every model.
+b) The advantage of using the proposed approach was apparent when including limited observations and missing values.
+Looking at accuracy, the proposed method outperformed Limma and $t$-test in nearly every model.
 }
@@
 \caption{Dataset 3 : SpikeIn benchmark - Ubiquitination - Label-free.
-a) Before adjustment all models show the fold change of the spike-in peptides is systematically different from the expected fold change.
-After adjustment, this systemic difference is removed, however the inner quartile range of the Limma and t-test models is wider than the proposed method.
-b) The spike in peptides (colored red) do not follow the expected log fold change before adjustment.
-However, after adjustment, the spike in peptides are more in line with expectation.
-Using Limma the spike in peptides follow the expected log fold change after adjustment, however the majority of spike in peptides do not have a significant adjusted pvalue.}
+a) Before adjustment the fold change of the spike-in peptides' were systematically different from the expected fold change in all models.
+After adjustment, this systemic difference was removed, however the inner quartile range of the Limma and $t$-test models was wider than the proposed method.
+b) The spike in peptides (colored red) did not follow the expected log fold change before adjustment.
+However, after adjustment, the spike in peptides were more in line with expectation.
+Using Limma the spike in peptides followed the expected log fold change after adjustment, however the majority of spike in peptides did not have a significant adjusted p-value.}
 \label{fig:spikein_volcano}
@@
 b) Comparing the global profiling of protein $GSDMD$ with the ubiquitination of the protein at site $K62$.
-When looking at the summary of the modification and global protein it is clear the conditions follow different trends.
-Specifically, there appears to be no change in abundance between Dox1hr and Dox4hr in the modified plot, however there is a large negative change when looking at the unmodified plot.
-This indicates the modification is confounded with changes in the unmodified protein.}
+When looking at the summary of the modification and global protein it was clear the conditions follow different trends.
+Specifically, there appeared to be no change in abundance between Dox1hr and Dox4hr in the modified plot, however there was a large negative change when looking at the unmodified plot.
+This indicated the modification was confounded with changes in the unmodified protein.}
 \label{fig:data4_plots}
@@
 Again more PTMs became insignificant after adjustment then became significant.
-Comparing the global profiling of protein $TTP$ with the modification of the protein at site $S178$.
-When looking at the summary of the modification and global protein it is clear the difference between conditions follow the same trend.
-Specifically, there is a positive adjustment in abundance when comparing WT\_Uninfect to WT\_Late in both the modification and global profiling run.
-This indicates the movement is driven by changes in global protein that is only accounted for in the model when adjusting for global protein abundance change.}
+b) Comparing the global profiling of protein $TTP$ with the modification of the protein at site $S178$.
+When looking at the summary of the modification and global protein it was clear the difference between conditions followed the same trend.
+Specifically, there was a positive adjustment in abundance when comparing WT\_Uninfect to WT\_Late in both the modification and global profiling run.
+This indicated the movement was driven by changes in global protein that was only accounted for in the model after adjusting for global protein abundance change.}
 \label{fig:data5_plots}
@@
  \caption{Dataset 6 : Human - Ubiquitination - Label-free no global profiling run.
-a) The overlap of differencial modified peptides for the PTM model with and without global protein level adjustment.
-Here many more PTMs became insignificant then became significant after adjustment.
-This is due to not having a global profiling run, resulting in a large lack of overlap between the modified peptides and unmodified proteins.
-b) Here we make the same comparison but only for modified peptides with a matching unmodified protein, so adjustment can be performed.
-In this case we see significantly less peptides become insignificant after adjustment.
-This highlights the need for a global profiling run if protein adjustment is going to performed.}
+a) The overlap of differential modified peptides for the PTM model with and without global protein level adjustment.
+More PTMs became insignificant than became significant after adjustment.
+This was due to not having a global profiling run, resulting in a lack of overlap between modified peptides and unmodified proteins.
+b) Here we made the same comparison but only looked at modified peptides where adjustment could be performed, ie they had a matching unmodified protein.
+In this case there were significantly less peptides that became insignificant after adjustment.
+This highlighted the need for a global profiling run.}
 \label{fig:data6_plots}

### 2022-03-19 supplementary/ptm_sm.tex
moved sections to main

@@
 \clearpage
-\section{Overview of proposed approach and notation}
-\label{sec:intro}
-
-This study addresses three major goals in the characterization of post-translational modifications (PTMs): a) relative PTM quantification, b) PTM significance analysis, i.e., to detect PTM sites that are differentially modified across experimental conditions, and c) statistical design and analysis of PTM experiments.
+%\section{Overview of proposed approach and notation}
+%\label{sec:intro}
+
+%This study addresses three major goals in the characterization of post-translational modifications (PTMs): a) relative PTM quantification, b) PTM significance analysis, i.e., to detect PTM sites that are differentially modified across experimental conditions, and c) statistical design and analysis of PTM experiments.
 
@@
 
-\subsection{Data structure of PTM quantification experiments.}
-A set of fully-cleaved and/or partially-cleaved peptides containing a same PTM (e.g., ubiquitination/phosphorylation) at one site are considered together.
+%\subsection{Data structure of PTM quantification experiments.}
+%A set of fully-cleaved and/or partially-cleaved peptides containing a same PTM (e.g., ubiquitination/phosphorylation) at one site are considered together.
 There are $I$ conditions and $J$ mass spectrometry runs (technical replicates) per condition in the experiment.
@@
 
-\begin{figure}[h!]
-\centering
-\begin{footnotesize}
-\[
-\begin{array}{l l c c c c c c c c c} 
-\toprule
- &  & \multicolumn{4}{c}{\text{Condition }1} & \dots & \multicolumn{4}{c}{\text{Condition }I} \\ 
-\cmidrule{3-6} 
-\cmidrule{8-11} 
- &  &  \text{Run }1 & \text{Run }2 & \dots & \text{Run }J & \dots & \text{Run }1 & \text{Run }2 & \dots & \text{Run }J \\
-\midrule
-\multirow{4}{*}{Modified} & \text{Feature }1 & y_{111}^{\ast} & y_{121}^{\ast} & \dots & y_{1J1}^{\ast} & \dots & y_{I11}^{\ast} & y_{I21}^{\ast} & \dots & y_{IJ1}^{\ast} \\
- & \text{Feature }2 & y_{112}^{\ast} & - & \dots & y_{1J2}^{\ast} & \dots & y_{I12}^{\ast} & y_{I22}^{\ast} & \dots & y_{IJ2}^{\ast}  \\
- & \dots & \dots & \dots & \dots & \dots & \dots & \dots & \dots & \dots & \dots \\
- & \text{Feature }K & y_{11K}^{\ast} & y_{12K}^{\ast} & \dots & y_{1JK}^{\ast} & \dots & y_{I1K}^{\ast} & y_{I2K}^{\ast} & \dots & y_{IJK}^{\ast} \\
-\midrule
-\multirow{4}{*}{Unmodified} & \text{Feature }1 & y_{111} & y_{121} & \dots & y_{1J1} & \dots & y_{I11} & y_{I21} & \dots & y_{IJ1} \\
- & \text{Feature }2 & y_{112} & y_{122} & \dots & y_{1J2} & \dots & - & - & \dots & -  \\
-% & \text{Feature }2 & y_{112} & y_{122} & \dots & y_{1J2} & \dots & y_{I12} & y_{I22} & \dots & y_{IJ2}  \\
- & \dots & \dots & \dots & \dots & \dots & \dots & \dots & \dots & \dots & \dots \\
- & \text{Feature }L & y_{11L} & y_{12L} & \dots & y_{1JL} & \dots & y_{I1L} & y_{I2L} & \dots & y_{IJL} \\
-\bottomrule
-\end{array}
-\]
+%\begin{figure}[h!]
+%\centering
+%\begin{footnotesize}
+%\[
+%\begin{array}{l l c c c c c c c c c} 
+%\toprule
+% &  & \multicolumn{4}{c}{\text{Condition }1} & \dots & \multicolumn{4}{c}{\text{Condition }I} \\ 
+%\cmidrule{3-6} 
+%\cmidrule{8-11} 
+% &  &  \text{Run }1 & \text{Run }2 & \dots & \text{Run }J & \dots & \text{Run }1 & \text{Run }2 & \dots & \text{Run }J \\
+%\midrule
+%\multirow{4}{*}{Modified} & \text{Feature }1 & y_{111}^{\ast} & y_{121}^{\ast} & \dots & y_{1J1}^{\ast} & \dots & y_{I11}^{\ast} & y_{I21}^{\ast} & \dots & y_{IJ1}^{\ast} \\
+% & \text{Feature }2 & y_{112}^{\ast} & - & \dots & y_{1J2}^{\ast} & \dots & y_{I12}^{\ast} & y_{I22}^{\ast} & \dots & y_{IJ2}^{\ast}  \\
+% & \dots & \dots & \dots & \dots & \dots & \dots & \dots & \dots & \dots & \dots \\
+% & \text{Feature }K & y_{11K}^{\ast} & y_{12K}^{\ast} & \dots & y_{1JK}^{\ast} & \dots & y_{I1K}^{\ast} & y_{I2K}^{\ast} & \dots & y_{IJK}^{\ast} \\
+%\midrule
+%\multirow{4}{*}{Unmodified} & \text{Feature }1 & y_{111} & y_{121} & \dots & y_{1J1} & \dots & y_{I11} & y_{I21} & \dots & y_{IJ1} \\
+% & \text{Feature }2 & y_{112} & y_{122} & \dots & y_{1J2} & \dots & - & - & \dots & -  \\
+%% & \text{Feature }2 & y_{112} & y_{122} & \dots & y_{1J2} & \dots & y_{I12} & y_{I22} & \dots & y_{IJ2}  \\
+% & \dots & \dots & \dots & \dots & \dots & \dots & \dots & \dots & \dots & \dots \\
+% & \text{Feature }L & y_{11L} & y_{12L} & \dots & y_{1JL} & \dots & y_{I1L} & y_{I2L} & \dots & y_{IJL} \\
+%\bottomrule
+%\end{array}
+%\]
 %\[
@@
 %\]
-\end{footnotesize}
-\caption{Representation of the data of modified peptides at one site and unmodified peptides of the same protein, with $I$ conditions and $J$ replicate runs.
+%\end{footnotesize}
+%\caption{Representation of the data of modified peptides at one site and unmodified peptides of the same protein, with $I$ conditions and $J$ replicate runs.
 Abundances of the PTM and protein are quantified by multiple spectral features (peptide ions, $K$ for modified peptides and $L$ for unmodified peptides).
@@
 \label{fig:dtable}}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Existing methods} \label{sec:ttest}
-\subsection{Two-sample $t$-test}
-
-Two-sample $t$-test is based on the null hypothesis that there is no difference in mean PTM abundance between Conditions $i$ and $i'$.
-The abundance in each run is taken as input and is often estimated by sum of peak intensities.
-The $t$-test is typically performed based on the log of summarized value.
-For example, the log-abundance estimate for the PTM in Run $j$ of Condition $i$ is given by
-\[
-\log \left( \sum_{k=1}^{K} 2^{y_{ijk}^{\ast}} \right).
-\]
-For adjustment with respect to unmodified peptides, the estimate of PTM abundance is divided by the protein abundance estimate, 
-and the $t$-test for the adjusted PTM abundance on log scale takes as input the difference of their log-estimates.
-%which is equivalent to take the difference of their log abundances. 
-The quantity is denoted by $d_{ij}$ and is given by
-\[
-d_{ij} = \log \left( \sum_{k=1}^{K} 2^{y_{ijk}^{\ast}} \right) - \log \left( \sum_{l=1}^{L} 2^{y_{ijl}} \right).
-\]
-Alternatively, run-level summary to be described can also be used. The difference between the means of PTM abundance in Conditions $i$ and $i'$ are estimated as
-\[
-\hat{\Delta} = \frac{1}{J} d_{i+}^{\ast} - \frac{1}{J} d_{i'+},
-\]
-where $d_{i+}^{\ast} = \sum_{j=1}^{J} d_{ij}^{\ast}$ and the test statistic for the $t$-test is given by $\hat{\Delta} / \mathrm{SE}(\hat{\Delta})$.
-The statistical significance of the difference is determined by comparing the test statistic against the $t$ distribution, with degrees of freedom $df=2J-2$ in balanced designs.
-
-\subsection{Limma}
-
-Limma uses linear models to test the null hypothesis that there is no difference in mean PTM abundances between conditions.
-Additionally, it leverages Empirical Bayes moderation to share pooled variance information across individual modification models and moderate the individual residual variances.
-Using a linear model allows Limma to share variance information across conditions, providing a more accurate estimate.\cite{Ritchie_15a}
-
-With respect to PTM analysis, feature level summarization is done in the same way as $t$-test. The log-abundace estimate for PTM in Run $j$ of Condition $i$ is given by
-
-\[
-\log \left( \sum_{k=1}^{K} 2^{y_{ijk}^{\ast}} \right).
-\]
-
-Adjustment for changes in the unmodified peptides is performed by taking the difference of log estimates, as follows
-
-\[
-d_{ij} = \log \left( \sum_{k=1}^{K} 2^{y_{ijk}^{\ast}} \right) - \log \left( \sum_{l=1}^{L} 2^{y_{ijl}} \right).
-\]
-
-Alternatively run level summarization can be performed given by
-
-\[
-\hat{\Delta} = \frac{1}{J} d_{i+}^{\ast} - \frac{1}{J} d_{i'+},
-\]
-
-After fitting a linear model, the final variance is moderated using Empirical Bayes and the global variance across all modified peptides.
-This is done as $s_{i} = P(v_i| v_\Theta)$ where $v_i$ is the resulting variance when modeling peptide $i$ and $v_\Theta$ is the global variance.
+%\end{figure}
+%
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%\section{Existing methods} \label{sec:ttest}
 
@@
 
-To characterize the observed feature intensities, different levels of variations are expressed using linear mixed models in consideration of the following factors in a simple experimental design: modification, condition, run, and feature.
+%To characterize the observed feature intensities, different levels of variations are expressed using linear mixed models in consideration of the following factors in a simple experimental design: modification, condition, run, and feature.
 As different degrees of variability are present in the feature intensities of modified and unmodified peptides, they are expressed by separate models.
@@
 
-\subsection{Design of PTM experiments}
+\subsection{Design of PTM experiments - Extension to complex designs}
 \label{sec:complex_design}

### 2022-03-21 main/MSstatsPTM_Main.tex
Background

@@
 
-Mass spectrometry (MS)-based proteomics is increasingly used to detect changes in post-translational modifications (PTMs) between samples from various conditions.
+Liquid chromatography coupled with mass spectrometry (LC-MS/MS)-based proteomics is increasingly used to detect changes in post-translational modifications (PTMs) between samples from various conditions.
 Analysis of data from such experiments faces numerous statistical challenges.
@@
 Therefore, statistical approaches for detecting differential PTM abundance must integrate all the available information pertaining to a PTM site, and consider all the relevant sources of confounding and variation.
-In this manuscript we propose such a statistical framework, which is comprehensive, accurate, and leads to reproducible results.
+In this manuscript we propose such a statistical framework, which is versatile, accurate, and leads to reproducible results.
 The framework requires an experimental design, which for each sample quantifies both the peptides with post-translational modifications, and peptides from the same proteins with no modification sites.
-The framework supports both label-free and tandem mass tag (TMT)-based acquisitions.
+The framework supports both label-free and tandem mass tag (TMT)-based LC-MS/MS acquisitions.
 The statistical methodology separately summarizes the abundances of peptides with and without the modification sites, and fits separate linear mixed effects models that reflect the biological and technological aspects of the experimental design.
@@
 The framework requires an experimental design, which for each sample quantifies both the peptides with post-translational modifications, and peptides from the same proteins with no modification sites.
-The framework supports both label-free (DIA, DDA) and tandem mass tag (DDA-TMT)-based acquisitions.
+The framework supports data-dependent acquisitions (DDA) that are label-free or tandem mass tag (TMT)-based.
 The statistical methodology separately summarizes the abundances of peptides with and without the modification sites, and fits separate linear mixed effects models that reflect the biological and technological aspects of the experimental design.
@@
 The datasets illustrate a diverse set of organisms, acquisition methods, and experimental designs, showing the applicability of the framework to a variety of situations.
-The evaluations demonstrated that by appropriately leveraging the information from the unmodified peptide, the proposed approach improves the accuracy of the estimates of PTM fold changes and results in a better calibrated false positive rate of detecting differentially abundant PTMs.
+The evaluations demonstrated that by appropriately leveraging the information from the unmodified peptides, the proposed approach improves the accuracy of the estimates of PTM fold changes and results in a better calibrated false positive rate of detecting differentially abundant PTMs.
 In particular, accounting for the confounding from protein abundance allows us to characterize the true effect of the modification, avoiding the need for more manual and time intensive methods.
@@
 The second simulation introduced real-world characteristics, such as limited modified features and missing values.
-The spike-in experiment took the real world characteristics a step further and allowed us to compare the methods in a real experiment with known changes in modified spike-in peptides.
+The spike-in experiment took the real world characteristics a step further and allowed us to compare the methods in a real-world experimental settings with known changes in modified spike-in peptides.
 Finally, three biological experiments demonstrated the applicability of the proposed approach across different biological organisms, PTM types, experimental designs and acquisition strategies.
@@
 
-The simulation provides an idealistic case.
+The simulation represented an idealistic case.
 24 synthetic datasets were generated exhibiting scenarios with different experimental designs and biological variation.
-1000 proteins were generated per dataset with 10 unmodified features per protein.
-One PTM was created for each protein in the dataset, resulting in 1000 PTMs, each with 10 modified features.
-500 of the PTMs were generated with a differential fold change between conditions, while the other half were generated with no difference between conditions.
-The fold change of half the 500 differential PTMs were fully masked by changes in the unmodified protein.
+In each dataset, 1000 proteins had 10 unmodified features per protein.
+Each of the 1000 proteins had one PTM, each PTM represented by 10 modified features.
+The PTMs of 500 proteins had a differential fold change between conditions, while the other 500 proteins were generated with no difference between conditions.
+Furthermore, the fold changes of half of the 500 differential PTMs were fully masked by changes in the unmodified protein.
 Additionally, the fold change of half the 500 non-differential PTMs was entirely due to changes in the unmodified protein.
-All differential PTMs were generated with an expected log fold change of .75 between conditions.
+All differential PTMs were generated with an expected log fold change of 0.75 between conditions.
+
 Each simulation was generated with random biological variation.
@@
 The simulations were generated with one of two values $\sigma^2 = \{.2, .3\}$.
-The values for $\sigma^2$ were motivated by the biological experiments in this manuscript.
-Further details can be found in Supplementary Sec. \ref{supp-sec:dataset1}.
+The values for $\sigma^2$ were motivated \todo{motivated how specifically?
+corresponded to median variances?}  by the biological experiments in this manuscript.
+Further details are in Supplementary Sec. \ref{supp-sec:dataset1}.
 
@@
 In this simulation, the feature counts and the proportion of missing values were as observed on average over all the the experimental datasets in this manuscript.
-PTMs were simulated with 2 modified peptide features, while unmodified proteins were simulated with 10 features.
+Specifically, PTMs were simulated with 2 modified peptide features, while unmodified proteins were simulated with 10 features.
 Additionally, 20\% of observations for both modified and unmodified peptides were missing completely at random.
@@
 Devon: The simulation did not include any missing values due to low abundance.
-Should we mention this or just leave it as is?} Further details can be found in Supplementary Sec.
+Should we mention this or just leave it as is?} Further details are in Supplementary Sec.
 \ref{supp-sec:dataset2}.
@@
 
-In the custom spike-in experiment 8 biological replicates were allocated to 4 conditions, resulting in 2 biological replicates per condition, and a balanced design.
+In the custom spike-in experiment 8 biological \todo{I am not sure these are bio reps.
+If this is the same lysate, then these are in fact technical reps (there is no between-subject variation)} replicates were allocated to 4 conditions, resulting in 2 biological replicates per condition, and a balanced design.
 Additional information can be found in Supplementary Sec.
@@
 it had no expected changes in any comparison.
+\todo{I am a bit confused.
+Fig S7 shows that there is a difference in Background between Mix 1,2 and Mix 3,4.
+Could you explain better?
+In fact, would it be possible to move the figure here for clarity, and have the figure as panel (a), and the table as panel (b)?
+And in the text above refer to the figure.
+This will help the understanding} \todo{Also, why is it a balanced design - as stated in the paragraph above - when the background is not present everywhere?}
 
@@
 \label{sec:exp_proc_dataset4}
-In this experiment 11 biological replicates were allocated to 1 TMT mixture in an unbalanced design. Further details are in Supplementary Sec. \ref{supp-sec:ipah}.
+In this experiment 11 biological replicates were allocated to 1 TMT mixture in an \todo{group comparison or time course?} unbalanced design.
+Further details are in Supplementary Sec.
+\ref{supp-sec:ipah}.
 
@@
 Uninfected cells were measured at 0 and 6 hours, while infected cells were measured at 1, 2, 4, and 6 hour increments, resulting in six total conditions.
-The proteome was quantified using isobaric multiplexing via tandem mass tagging (TMT) in combination with LC-MS/MS.
+\todo{If this is a group comparison, explain why, even though the conditions are times.
+The table above says it is group comparison} The proteome was quantified using isobaric multiplexing via tandem mass tagging (TMT) in combination with LC-MS/MS.
 \todo{One MS run total?
@@
 
-\medskip \noindent {\bf Pairwise comparisons:} We evaluated the ability of the statistical methods to detect changes in the abundance of modified peptides both before and after adjusting for changes in global protein abundance.
+\medskip \noindent {\bf Pairwise Comparisons:} We evaluated the ability of the statistical methods to detect changes in the abundance of modified peptides both before and after adjusting for changes in global protein abundance.
 The six condition were labeled Dox1hr, Dox2hr, Dox4hr, Dox6hr, NoDox0hr, and NoDox6hr.
@@
 \label{sec:exp_proc_dataset5}
-In this experiment 8 biological replicates were allocated to 2 TMT mixtures in an unbalanced design. More details are in Supplementary Sec. \ref{supp-sec:shigella}.
+In this experiment 8 biological replicates were allocated to 2 TMT mixtures in an unbalanced \todo{time course} design. More details are in Supplementary Sec. \ref{supp-sec:shigella}.
 
@@
 Mixture two contained one replicate of uninfected KO and two uninfected WT.
+\todo{was there a reference channel?}
 
@@
 
-In this experiment 8 biological replicates were split between 4 conditions in a balanced design. Further details in Suppler Sec \ref{supp-sec:usp30}.
+In this experiment 8 biological replicates were split between 4 conditions in a balanced \todo{group comparison} design. Further details in Suppler Sec \ref{supp-sec:usp30}.
 
@@
 \medskip \noindent {\bf Global Profiling Run:} This experiment did not include a separate global profiling run to measure unmodified peptides.
-In this case $MSstatsPTM$ can still be used by extracting unmodified peptides from the modified run.
-However, in addition to low feature counts for unmodified peptides,  this leads to substantially fewer matches between modified and unmodified peptides.
+In this case $MSstatsPTM$ could still be used by extracting unmodified peptides from the modified run.
+However, in addition to low feature counts for unmodified peptides,  this lead to substantially fewer matches between modified and unmodified peptides.
 There was a 41.9\% overlap between the identified background modified peptides and proteins that were quantified in the global profiling run.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section*{Background Information}
-
-\subsection*{Experimental Design}
-
-\todo{Would it be possible to re-organize this section?
-(1) Overview, including the data structures  the figure and the notation - essentially the section that describes Fig 1; (2) Existing methods: limma and $t$-test, including the description in the notation of the figure; also Isobar-PTM - can express in the same notation?
-Ok if not; (3) Existing methods: MSstats.
-Not applicable for PTM but useful background; (4) Proposed approach in the notation of the figure.
-I feel that the experimental design in Suppler is redundant.
-Could we fold it into the main text as well?} \todo{In this way, the entire Suppler Sec 2 will be merged into main \; the only methods in supplementary will be the table }
-
-\todo{Devon: I moved most everything around as recommended.
-The notation is mainly described once we start getting into the equations (t-test section).
-We can move this into the design section but I thought it was clearer to describe it with the equations.
-The results section now starts with the proposed method in the same notation.
-The supplementary now only includes the table and extra results.}
-
-Figure \ref{fig:data-structure} schematically illustrates a simplified version of the data structure resulting from a typical bottom-up experiment for quantitative analysis of PTMs, in which there are multiple layers of variation present.
-A PTM site is quantified with multiple spectral features, which vary in sequence (e.g., fully or partially cleaved peptides), ionization efficiency, charge states, etc.
-Additionally, some spectral features can be missing.
-The feature intensities are viewed as repeated measurements of the underlying abundance of the PTM, where the abundance in Condition $i$ is denoted by $i^{\ast}$.
-The number of quantified features vary across replicate LC-MS/MS runs of the same sample, and across conditions.
-To perform adjustment with respect to protein abundance, features of unmodified peptides are used for the inference of underlying protein abundance.
-Features corresponding to unmodified peptides are considered together, where the protein abundance in Condition $i$ is denoted by $i$.
-Typically, because of the enrichment step for PTMs, very few of those features are present in original LC-MS runs.
-For more accurate estimation of protein abundance, separate global proteomics data of unenriched samples are recommended.
-If unmodified peptides are unavailable for any given modification, unmodified intensity adjustment cannot be performed.
-
-\subsection*{Existing statistical methods for experiments targeting PTMs}
-
-Despite the importance of PTMs for biological function, there is a lack of general statistical analysis framework that summarizes the available quantitative information from LC-MS/MS experiments, performs statistical inference, and characterizes the quantitative properties of PTM in a statistically rigorous manner.
-
-Many investigations performed differential abundance analysis of PTMs using two-sample $t$-test or its extensions.
-The approach takes as input intensities of individual features from modified peptides, or intensity ratios of modified and unmodified peptide features, and compares the mean abundance of a PTM site from one condition to another~\cite{Schwammle2015}\cite{THOMAS2020}.
-Modifications of the $t$-test such as moderated $t$-test with Limma were also proposed~\cite{Ritchie_15a}\cite{Zhu}.
-While simple, these approaches do not fully account for the sources of variations, and are not directly applicable to experiments with complex designs, e.g., comparisons of multiple conditions, acquisition in multiple batches, etc.
-Additionally, while these approaches can be applied to experiments studying PTMs, there is not a self contained, straight forward implementation of the methods, making application challenging.
-
-Isobar-PTM was also proposed for experiments with LC-MS/MS quantitative strategies that employ isobaric labels such as tandem mass tags (TMT) or isobaric tag for relative and absolute quantification (iTRAQ)\cite{Breitwieser:2013}.
-Isobar-PTM expresses MS measurements with a linear model and performs adjustment with respect to protein abundance using the difference between log-ratio of modified peptides in two channels and log-ratio of protein level.
-Unfortunately, this statistical modeling framework is not applicable to either label-free workflows or experiments with complex designs.
-
-An overview of the existing approach using $t$-test and Limma are shown in the following sections.
-
-\subsubsection*{Two-sample $t$-test}
-
-Two-sample $t$-test is based on the null hypothesis that there is no difference in mean PTM abundance between Conditions $i$ and $i^*$.
-There are $I$ conditions and $J$ mass spectrometry runs (technical replicates) per condition in the experiment.
-The PTM site is represented by $K$ spectral features (peptide ions, distinguished by their cleavage residues and charge states).
+\section*{Background}
+
+\subsection*{Goals of PTM characterization, input to statistical analyses, and notation}
+
+Consider an LC-MS/MS experiment with $I$ conditions and $J$ mass spectrometry runs (technical \todo{of biological?} replicates) per condition in the experiment.
+For one protein, the PTM site is represented by $K$ spectral features (peptide ions, distinguished by their cleavage residues and charge states).
 The log-intensity (base 2) of Feature $k$, in Run $j$ of Condition $i$ is denoted by $y_{ijk}^{\ast}$.
 Conversely, the unmodified protein is represented by $L$ spectral features, and the log-intensity of Feature $l$ from the unmodified peptides in the same run is denoted by $y_{ijl}$.
+
+Figure \ref{fig:data-structure}(a) schematically illustrates this data structure for one protein in a simplified experiment with $I=2$ conditions.
+\todo{Describe the population parameters in the figure (quantities of interest}.
+The data are collected in $J=2$ \todo{biological?} replicate runs \todo{describe how the data are representative of the population parameters of interest}
+The number of quantified features vary across replicate LC-MS/MS runs of the same sample, and across conditions. \todo{There may be outliers}. Some spectral features can be missing.
+
+\todo{in Figure 2 b, there are two quantities of interest, stated at the population level. Describe the quantities, what is of interest, why relevant}
+
+%---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------
+
+\subsection*{Existing statistical methods for experiments targeting PTMs}
+
+\noindent {\bf Two-sample $t$-test} \todo{When describing the existing methods, could you refer to Fig 1?
+Here is the suggested order: (1) What is the null hypothesis?
+Do these methods test the same null hypothesis?
+(If not, it may be a good idea to add this null hypothesis to the figure for contrast) (2) What are the data (how summarized prior to the testing); (3) what is the model assumptions?
+(4) how inference is performed} Many investigations perform differential abundance analysis of PTMs using two-sample $t$-test or its extensions.
+
+\todo{These sentences were combined from different places.
+Could you edit?} The approach takes as input intensities of individual features from modified peptides, or intensity ratios of modified and unmodified peptide features, and compares the mean abundance of a PTM site from one condition to another~\cite{Schwammle2015}\cite{THOMAS2020}.
+Two-sample $t$-test is based on the null hypothesis that there is no difference in mean PTM abundance between Conditions $i$ and $i^*$.
 The abundance in each run is taken as input and is often estimated by sum of peak intensities.
@@
 
-\subsubsection*{Limma}
+\medskip \noindent {\bf Limma} \todo{Could you fine-tune, using the same order as above for Limma?
+I.e., null hypothesis, the data, the model, the inference} Modifications of the $t$-test such as moderated $t$-test with Limma were also proposed~\cite{Ritchie_15a}\cite{Zhu}.
+While simple, these approaches do not fully account for the sources of variations, and are not directly applicable to experiments with complex designs, e.g., comparisons of multiple conditions, acquisition in multiple batches, etc.
+Additionally, while these approaches can be applied to experiments studying PTMs, there is not a self contained, straight forward implementation of the methods, making application challenging.
 
@@
 
-\subsection*{Existing work in statistical modeling and parameter estimation}
-
-The proposed approach leverages the large amount of work done to create $MSstats$~\cite{Choi:2014}.
+\medskip \noindent {\bf Isobar-PTM} was also proposed for experiments with LC-MS/MS quantitative strategies that employ isobaric labels such as tandem mass tags (TMT) or isobaric tag for relative and absolute quantification (iTRAQ)\cite{Breitwieser:2013}.
+Isobar-PTM expresses MS measurements with a linear model and performs adjustment with respect to protein abundance using the difference between log-ratio of modified peptides in two channels and log-ratio of protein level.
+Unfortunately, this statistical modeling framework is not applicable to either label-free workflows or experiments with complex designs.
+
+\subsection*{Statistical modeling and parameter estimation in MSstats}
+
+\todo{Could you also refer to Figure 1?
+Essentially, MSstats only analyzes a part of the figure.
+Refer to the null hypothesis.
+Then describe the summarization (including AFT etc), and the model for he summarized data (the fact that different designs require different models - this is your argument for the limitation of the methods above.)} The proposed approach leverages the large amount of work done to create $MSstats$~\cite{Choi:2014}.
 $MSstats$ takes as input a list of log-transformed intensities of spectral features which are used to characterize the identified protein.
@@
 
+\todo{Although demonstrated here on DDA, is also applicable to DIA, SRP and PRM acquisitions}
+
 We proposed a general statistical modeling framework and implementation for PTM characterization.
@@
 
-All datasets for the four experiments and two simulations, including raw files and analysis results, are available online.
+\todo{This overlaps with the Datasets section.
+I would merge all of the below to the experimental data section} All datasets for the four experiments and two simulations, including raw files and analysis results, are available online.
 
@@
 \caption{
-Data structure of a typical PTM experiment and goals of PTM characterization.
-(a) Schematic data representation, in a simplified case of two conditions and two replicate runs.
+Goals of PTM characterization, input to statistical analyses, and notation.
+\todo{Could you update the caption?
+The figure now has only 2 parts.
+It doesn't have subscript $i$ etc.
+On the other hand, it distinguishes the population parameters, and the observed data.
+Could you mention this here briefly, and describe in more detail in the text?}  (a) Schematic data representation, in a simplified case of two conditions and two \todo{biological?} replicate runs.
 Each PTM site is modeled and characterized separately, where a PTM is quantified with multiple spectral features (boxes), distinguished by different charge states of a peptide.

### 2022-03-21 main/MSstatsPTM_Main.tex
Results - beginning

@@
 We propose a statistical method for detecting changes in PTMs while removing confounding with changes in global protein abundance.
-Here the method is applied to a balanced design with multiple conditions and one technical replicate.
-In practice the experimental design can be unbalanced and variation can come from multiple sources, as shown in the next section.
-
-Detection of differentially modified PTM sites is performed through testing the null hypothesis of ‘no change’ against the alternative.
-The null hypothesis states that there is no difference in log-abundance of the PTM site between conditions, adjusted with respect to protein abundance.
+For simplicity, we discuss the method in the special case of a balanced design with multiple conditions and one technical replicate.
+However, in  practice the experimental design can be unbalanced and variation can come from multiple sources, as shown in the next section.
+
+\todo{Put it in the context of the figure again, and also in the context of the existing methods in Background - how is it similar to the existing methods, and how different} The null hypothesis states that there is no difference in log-abundance of the PTM site between conditions, adjusted with respect to protein abundance.
 The expected values of log-abundances of the PTM and protein in Condition $i$ are denoted by $\mu_{i}^{\ast}$ and $\mu_{i}$ respectively.
-The feature and run level summarization for condition $i$ is denoted as $\hat{y}_{i+}^{\ast}$ for the modified peptide and $\hat{y}_{i+}$ for the unmodified peptide.
+The feature and run level summarization \todo{Summarized how?} for condition $i$ is denoted as $\hat{y}_{i+}^{\ast}$ for the modified peptide and $\hat{y}_{i+}$ for the unmodified peptide.
 For two conditions $i$ and $i'$ the hypothesis test is formulated as follows:

### 2022-03-21 main/MSstatsPTM_Main.tex
Small

@@
 
-\todo{in Figure 2 b, there are two quantities of interest, stated at the population level. Describe the quantities, what is of interest, why relevant}
+\todo{in Figure 2 b, there are two quantities of interest, stated at the population level.
+Describe the quantities, what is of interest, why relevant.
+It may help to introduce multiple versions of the null hypotheses, to contrast the methods below} \todo{Also please edit the figure caption}
 
@@
 \medskip \noindent {\bf Limma} \todo{Could you fine-tune, using the same order as above for Limma?
-I.e., null hypothesis, the data, the model, the inference} Modifications of the $t$-test such as moderated $t$-test with Limma were also proposed~\cite{Ritchie_15a}\cite{Zhu}.
+I.e., null hypothesis, the data, the model, the inference.
+I think the text can be made more compact, and a bit more technical, by directly referring to either Fig 1 or the equations above} Modifications of the $t$-test such as moderated $t$-test with Limma were also proposed~\cite{Ritchie_15a}\cite{Zhu}.
 While simple, these approaches do not fully account for the sources of variations, and are not directly applicable to experiments with complex designs, e.g., comparisons of multiple conditions, acquisition in multiple batches, etc.

### 2022-03-28 main/MSstatsPTM_Main.tex
rewrote background section

@@
 The simulated peptide abundance  $Z_{ibc}$ was generated as $Z_{ibc} = Y_{ibc} + \epsilon_{ibc}$ where $\epsilon_{ibc}\sim^{iid}N(0,\sigma^2)$.
-The simulations were generated with one of two values $\sigma^2 = \{.2, .3\}$.
-The values for $\sigma^2$ were motivated \todo{motivated how specifically?
-corresponded to median variances?}  by the biological experiments in this manuscript.
+The simulations were generated with one of two values $\sigma^2 = \{.2, .3\}$, motivated by the biological experiments in this manuscript.
 Further details are in Supplementary Sec. \ref{supp-sec:dataset1}.
@@
 \label{sec:exp_proc_dataset5}
-In this experiment 8 biological replicates were allocated to 2 TMT mixtures in an unbalanced \todo{time course} design. More details are in Supplementary Sec. \ref{supp-sec:shigella}.
+In this experiment 8 biological replicates were allocated to 2 TMT mixtures in an unbalanced time course design. More details are in Supplementary Sec. \ref{supp-sec:shigella}.
 
@@
 Mixture two contained one replicate of uninfected KO and two uninfected WT.
-\todo{was there a reference channel?}
 
@@
 
-In this experiment 8 biological replicates were split between 4 conditions in a balanced \todo{group comparison} design. Further details in Suppler Sec \ref{supp-sec:usp30}.
+In this experiment 8 biological replicates were split between 4 conditions in a balanced group comparison design. Further details in Suppler Sec \ref{supp-sec:usp30}.
 
@@
 Figure \ref{fig:data-structure}(a) schematically illustrates this data structure for one protein in a simplified experiment with $I=2$ conditions.
-\todo{Describe the population parameters in the figure (quantities of interest}.
-The data are collected in $J=2$ \todo{biological?} replicate runs \todo{describe how the data are representative of the population parameters of interest}
-The number of quantified features vary across replicate LC-MS/MS runs of the same sample, and across conditions. \todo{There may be outliers}. Some spectral features can be missing.
+The data are collected in $J=2$ \todo{biological?} replicate runs.
+Each run represents an observation used to estimate the relative abundance of the PTM in condition $i$, $u_i^{\ast}$.
+The number of quantified features vary across replicate LC-MS/MS runs of the same sample, and across conditions.
+There may be feature outliers which could be caused by random variation or measurement error.
+Additionally, some spectral features can be missing.
+
+The population parameter we are interested in estimating is the difference between the PTM in condition $1$ and condition $2$, denoted by $u_1^{\ast}$ and $u_2^{\ast}$ respectively.
+To calculate the quantity of interest, statistical analysis needs to summarize values in the table in Figure \ref{fig:data-structure}(a) using appropriate models.
+For condition $i$, the models provide estimates for the mean abundance, $\hat{u}_i^{\ast}$, and standard error, $SE(\hat{u}_i^{\ast})$.
+Once summarized, we can formulate the Null and Alternative Hypotheses as seen in Equation \ref{eq:conv_null_hyp}.
+
+\begin{equation}
+\begin{aligned}
+H_{0}: \Delta = \mu_{1}^{\ast} - \mu_{2}^{\ast} = 0 \\
+H_{a}: \Delta = \mu_{1}^{\ast} - \mu_{2}^{\ast} \neq 0
+\end{aligned}
+\label{eq:conv_null_hyp}
+\end{equation}
+
+However, these population parameters are inherently convoluted with changes in the unmodified protein.
+To account for this, we must estimate the unmodified protein in conditions $1$ and $2$, denoted by $\mu_1$ and $\mu_2$.
+We estimate these quantities in the same way as for the PTM, using their correspond features and observations, resulting in an estimated mean abundance, $\hat{u}_i$, and standard error, $SE(\hat{u}_i)$.
+These estimates are used to adjust the quantities estimated for the PTM, resulting in the new Null and Alternative Hypotheses seen in Equation \ref{eq:null_hyp}.
+
+\begin{equation}
+\begin{aligned}
+H_{0}: \Delta = ( \mu_{1}^{\ast} - \mu_{1}) - ( \mu_{2}^{\ast} - \mu_{2} ) = 0 \\
+H_{a}: \Delta = ( \mu_{1}^{\ast} - \mu_{1}) - ( \mu_{2}^{\ast} - \mu_{2} ) \neq 0
+\end{aligned}
+\label{eq:null_hyp}
+\end{equation}
 
@@
 (If not, it may be a good idea to add this null hypothesis to the figure for contrast) (2) What are the data (how summarized prior to the testing); (3) what is the model assumptions?
-(4) how inference is performed} Many investigations perform differential abundance analysis of PTMs using two-sample $t$-test or its extensions.
-
-\todo{These sentences were combined from different places.
-Could you edit?} The approach takes as input intensities of individual features from modified peptides, or intensity ratios of modified and unmodified peptide features, and compares the mean abundance of a PTM site from one condition to another~\cite{Schwammle2015}\cite{THOMAS2020}.
-Two-sample $t$-test is based on the null hypothesis that there is no difference in mean PTM abundance between Conditions $i$ and $i^*$.
+(4) how inference is performed}
+
+\todo{These sentences were combined from different places. Could you edit?}
+
+Many investigations perform differential abundance analysis of PTMs using two-sample $t$-test or its extensions.
+While simple, these approaches do not fully account for the sources of variations, and are not directly applicable to experiments with complex designs, e.g., comparisons of multiple conditions, acquisition in multiple batches, etc.
+Additionally, while these approaches can be applied to experiments studying PTMs, there is not a self contained, straightforward implementation of the methods, making application challenging.
+
+The approach generally takes as input intensities of individual features from modified peptides, or intensity ratios of modified and unmodified peptide features, and compares the mean abundance of a PTM site from one condition to another~\cite{Schwammle2015}\cite{THOMAS2020}.
+Two-sample $t$-test is based on the null hypothesis that there is no difference in mean PTM abundance between Conditions $1$ and $2$.
 The abundance in each run is taken as input and is often estimated by sum of peak intensities.
-For example, the log-abundance estimate for the PTM in Run $j$ of Condition $i$ is given by
-\[
-\log \left( \sum_{k=1}^{K} 2^{y_{ijk}^{\ast}} \right).
-\]
-For adjustment with respect to unmodified peptides, the estimate of PTM abundance is divided by the protein abundance estimate, 
-and the $t$-test for the adjusted PTM abundance on log scale takes as input the difference of their log-estimates.
+For example, the log-abundance estimate for the PTM in Run $j$ of Condition $1$ is given by
+
+\begin{equation}
+\begin{aligned}
+\log \left( \sum_{k=1}^{K} 2^{y_{1jk}^{\ast}} \right).
+\end{aligned}
+\label{eq:feature_estimation}
+\end{equation}
+
+For adjustment with respect to unmodified peptides, the estimate of PTM abundance is divided by the protein abundance estimate.
 %which is equivalent to take the difference of their log abundances. 
-The quantity is denoted by $d_{ij}$ and is given by
-\[
-d_{ij} = \log \left( \sum_{k=1}^{K} 2^{y_{ijk}^{\ast}} \right) - \log \left( \sum_{l=1}^{L} 2^{y_{ijl}} \right).
-\]
-Alternatively, run-level summarization can also be used. The difference between the means of PTM abundance in Conditions $i$ and $i'$ are estimated as
-\[
-\hat{\Delta} = \frac{1}{J} d_{i+}^{\ast} - \frac{1}{J} d_{i'+},
-\]
-where $d_{i+}^{\ast} = \sum_{j=1}^{J} d_{ij}^{\ast}$ and the test statistic for the $t$-test is given by $\hat{\Delta} / \mathrm{SE}(\hat{\Delta})$.
+For condition $1$ the quantity is denoted by $d_{1j}$ and is given by
+
+\begin{equation}
+\begin{aligned}
+d_{1j}^{\ast} = \log \left( \sum_{k=1}^{K} 2^{y_{1jk}^{\ast}} \right) - \log \left( \sum_{l=1}^{L} 2^{y_{1jl}} \right).
+\end{aligned}
+\label{eq:adj_estimation}
+\end{equation}
+
+$t$-test then takes the adjusted PTM abundance of condition $1$, $d_{1j}$, and condition $2$, $d_{2j}$, for each run in $J$ as input to estimate the quantity of interest $\Delta$, as well as the standard error, $SE(\Delta)$.
+This results in the Null and Alternative Hypotheses show in Equation \ref{eq:ttest_null_hyp}, where $\hat{d}_{1+}^{\ast}$ represents all $J$ runs for condition $1$ and is the estimate of $(\mu_{1}^{\ast} - \mu_{1})$ seen in Equation \ref{eq:null_hyp}.
+
+\begin{equation}
+\begin{aligned}
+H_{0}: \Delta = d_{1+}^{\ast} - d_{2+}^{\ast} = 0 \\
+H_{a}: \Delta = d_{1+}^{\ast} - d_{2+}^{\ast} \neq 0
+\end{aligned}
+\label{eq:ttest_null_hyp}
+\end{equation}
+
+%Alternatively, run-level summarization can also be used. The difference between the means of PTM abundance in Conditions $i$ and $i'$ are estimated as
+%\[
+%\hat{\Delta} = \frac{1}{J} d_{i+}^{\ast} - \frac{1}{J} d_{i'+},
+%\]
+%where $d_{i+}^{\ast} = \sum_{j=1}^{J} d_{ij}^{\ast}$ and the test statistic for the $t$-test is given by $\hat{\Delta} / \mathrm{SE}(\hat{\Delta})$.
 The statistical significance of the difference is determined by comparing the test statistic against the $t$ distribution, with degrees of freedom $df=2J-2$ in balanced designs.
 
-\medskip \noindent {\bf Limma} \todo{Could you fine-tune, using the same order as above for Limma?
-I.e., null hypothesis, the data, the model, the inference.
-I think the text can be made more compact, and a bit more technical, by directly referring to either Fig 1 or the equations above} Modifications of the $t$-test such as moderated $t$-test with Limma were also proposed~\cite{Ritchie_15a}\cite{Zhu}.
-While simple, these approaches do not fully account for the sources of variations, and are not directly applicable to experiments with complex designs, e.g., comparisons of multiple conditions, acquisition in multiple batches, etc.
-Additionally, while these approaches can be applied to experiments studying PTMs, there is not a self contained, straight forward implementation of the methods, making application challenging.
-
+\medskip \noindent {\bf Limma} Modifications of $t$-test such as moderated $t$-test with Limma were also proposed~\cite{Ritchie_15a}\cite{Zhu}.
 Limma uses linear models to test the null hypothesis that there is no difference in mean PTM abundances between conditions.
@@
 
-With respect to PTM analysis, feature level summarization and adjustment for changes in the unmodified peptides are performed in the same way as $t$-test.
-After fitting a linear model, the final variance is moderated using the global variance across all modified peptides.
-This is done as $s_{i} = P(v_i| v_\Theta)$ where $v_i$ is the resulting variance when modeling PTM $i$ and $v_\Theta$ is the global variance.
+With respect to PTM analysis, feature level summarization and unmodified protein adjustment are performed in the same way as $t$-test, shown in Equations \ref{eq:feature_estimation} and \ref{eq:adj_estimation}.
+Limma fits a linear model using the adjusted summarized features, $d_{1+}^{\ast}$ and $d_{2+}^{\ast}$, to estimate the difference between condition $1$ and $2$, $\Delta$.
+This results in the same hypothesis as seen in Equation \ref{eq:ttest_null_hyp}.
+The variance for the estimation is moderated using information from the variance across all PTMs modeled.
+For the estimate of a single PTM $m$, $\Delta_m$, the variance is estimated as $s_{\Delta_m} = P(v_{\Delta_m}| v_\Theta)$ where $v_{\Delta_m}$ is the resulting variance when modeling PTM $m$ and $v_\Theta$ is the global variance across all PTMs.
+
+\todo{Could you fine-tune, using the same order as above for Limma?
+I.e., null hypothesis, the data, the model, the inference.
+I think the text can be made more compact, and a bit more technical, by directly referring to either Fig 1 or the equations above}
 
@@
 
-\todo{Could you also refer to Figure 1?
-Essentially, MSstats only analyzes a part of the figure.
-Refer to the null hypothesis.
-Then describe the summarization (including AFT etc), and the model for he summarized data (the fact that different designs require different models - this is your argument for the limitation of the methods above.)} The proposed approach leverages the large amount of work done to create $MSstats$~\cite{Choi:2014}.
+The proposed approach leverages the large amount of work done to create $MSstats$~\cite{Choi:2014}.
+$MSstats$ can analyze the PTM without accounting for convolution with the unmodified protein.
+In terms of Figure \ref{fig:data-structure}, this means we can either model the modified or unmodified features, but we cannot combine them as seen in part Figure \ref{fig:data-structure}(c).
+The only Null and Alternative Hypotheses we can address are those in Equation \ref{eq:conv_null_hyp}, without adjusting for unmodified protein abundance.
+
 $MSstats$ takes as input a list of log-transformed intensities of spectral features which are used to characterize the identified protein.
@@
 The model parameters are estimated using a split-plot approach, where the feature log-intensities are first summarized into a single value per protein per run in the subplot model, and the protein-level summaries are then used for the inference of the protein abundance.
-This approach allows the method to be extended to cases where the experimental design is unbalanced and where additional sources of variation are present.
+This approach allows the method to be extended to cases where the experimental design is unbalanced and additional sources of variation are present, overcoming the limitations in the previous sections.
 In the protein-level summarization, Tukey's median polish (TMP), a simple and robust procedure, is applied to iteratively fit a two-way additive model with the effects of run and feature, which in turn summarizes the log-intensities for each protein~\cite{Tukey:1977}.
@@
 
+\todo{Could you also refer to Figure 1?
+Essentially, MSstats only analyzes a part of the figure.
+Refer to the null hypothesis.
+Then describe the summarization (including AFT etc), and the model for he summarized data (the fact that different designs require different models - this is your argument for the limitation of the methods above.)}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-We propose a statistical method for detecting changes in PTMs while removing confounding with changes in global protein abundance.
+We propose a statistical method for detecting changes in PTMs.
+The approach removes confounding with the unmodified protein, while being applicable to experiments with complex designs.
 For simplicity, we discuss the method in the special case of a balanced design with multiple conditions and one technical replicate.
@@
 
-\todo{Put it in the context of the figure again, and also in the context of the existing methods in Background - how is it similar to the existing methods, and how different} The null hypothesis states that there is no difference in log-abundance of the PTM site between conditions, adjusted with respect to protein abundance.
-The expected values of log-abundances of the PTM and protein in Condition $i$ are denoted by $\mu_{i}^{\ast}$ and $\mu_{i}$ respectively.
-The feature and run level summarization \todo{Summarized how?} for condition $i$ is denoted as $\hat{y}_{i+}^{\ast}$ for the modified peptide and $\hat{y}_{i+}$ for the unmodified peptide.
-For two conditions $i$ and $i'$ the hypothesis test is formulated as follows:
+The proposed approach first applies the methods used in $MSstats$ to both the PTM and unmodified protein separately.
+This includes run-level feature summarization using Tukey's Median Polish and fitting a model based on the family of linear mixed-effects models.
+After a model is fit for the both the PTM and unmodified protein, the modeling results are then combined to estimate the quantity of interest.
+
+The null hypothesis states that there is no difference in log-abundance of the PTM site between conditions, adjusted with respect to protein abundance.
+Relating to the example in Figure \ref{fig:data-structure}, the expected value of log-abundance of the PTM and unmodified protein in Condition $1$ are denoted by $\mu_{1}^{\ast}$ and $\mu_{1}$ respectively.
+The feature and run level summarization for Condition $1$ is denoted as $\hat{y}_{1+}^{\ast}$ for the modified peptide and $\hat{y}_{1+}$ for the unmodified peptide.
+The Null and Alternative Hypotheses for Conditions $1$ and $2$ are shown in Equation \ref{eq:prop_null_hyp}.
+This is in contrast to the hypotheses show in Equation \ref{eq:null_hyp}, where the PTM was adjusted before modeling.
 
@@
 \begin{aligned}
-H_{0}: \Delta = ( \mu_{i}^{\ast} - \mu_{i'}^{\ast} ) - ( \mu_{i} - \mu_{i'} ) = 0 \\
-H_{a}: \Delta = ( \mu_{i}^{\ast} - \mu_{i'}^{\ast} ) - ( \mu_{i} - \mu_{i'} ) \neq 0
+H_{0}: \Delta = ( \mu_{1}^{\ast} - \mu_{2}^{\ast} ) - ( \mu_{1} - \mu_{2} ) = 0 \\
+H_{a}: \Delta = ( \mu_{1}^{\ast} - \mu_{2}^{\ast} ) - ( \mu_{1} - \mu_{2} ) \neq 0
 \end{aligned}
-\label{eq_1}
-\end{equation}
-
-Specifically, the adjusted difference is given by the difference in log-abundance of the PTM site, subtracted by the difference in log-abundance of the underlying protein, which is equivalent to the log of the ratio of PTM abundance difference to protein abundance difference.
-
-\begin{equation}
-\hat{\Delta} = \left[ \frac{1}{J} \left( \hat{y}_{i+}^{\ast} - \hat{y}_{i'+}^{\ast} \right) \right] - \left[ \frac{1}{J} \left( \hat{y}_{i+} - \hat{y}_{i'+} \right) \right]
+\label{eq:prop_null_hyp}
+\end{equation}
+
+Specifically, the adjusted difference is given by the difference in log-abundance of the PTM site, subtracted by the difference in log-abundance of the underlying protein, which is equivalent to the log of the ratio of PTM abundance difference to protein abundance difference, in a balanced case.
+
+\begin{equation}
+\hat{\Delta} = \left[ \frac{1}{J} \left( \hat{y}_{1+}^{\ast} - \hat{y}_{2+}^{\ast} \right) \right] - \left[ \frac{1}{J} \left( \hat{y}_{1+} - \hat{y}_{2+} \right) \right]
+\label{eq:prop_fc}
 \end{equation}
@@
 The estimate of the adjusted difference and the standard error (SE) of the estimate are obtained by combining the difference estimates and the associated SEs from both counterparts.
+The equation accounts for the increased uncertainty when incorporating more information from the unmodified protein.
+This may lead to an increased SE, when compared to modeling the PTM alone.
 
@@
 SE(\hat{\Delta}) = \left[ \left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right) \right]^{1/2}
+\label{eq:prop_se}
 \end{equation}
@@
 \left( \frac{\hat{\sigma}_{\gamma^{\ast}}^{4}}{\mathrm{df}(\gamma^{\ast})} + \frac{\hat{\sigma}_{\gamma}^{4}}{ \mathrm{df}(\gamma)} \right)
+\label{eq:prop_df}
 \end{equation}
@@
 Adjustment for multiple comparisons is performed using the Benjamini-Hochberg procedure to control the false discovery rate at a desired level, e.g., $0.05$~\cite{Benjamini:1995}.
+
+\todo{Put it in the context of the figure again, and also in the context of the existing methods in Background - how is it similar to the existing methods, and how different}
 
@@
 
-The statistical modeling approaches discussed above can be extended to complex designs including experiments with additional sources of variation and unbalanced designs.
-In this case our null hypothesis and Equation 1 are unchanged.
+The proposed approach can be extended to complex designs, including experiments with additional sources of variation and unbalanced designs.
+In this case our null hypothesis in Equation \ref{eq:prop_null_hyp} is unchanged.
 
@@
 In cases where the design is unbalanced, a linear mixed effects model is fit, which takes into account all potential sources of variation, and restricted maximum likelihood (RML) is used to estimate the parameters of each model.
-Once the parameters are estimated, we can combine the models using modified versions of Equations 2, 3, and 4.
-
-The log-fold change in the adjusted PTM abundance, $\Delta$, is now estimated by RML, and Equation 2 becomes 
+Once the parameters are estimated, we can combine the models using modified versions of Equations \ref{eq:prop_fc}, \ref{eq:prop_se}, and \ref{eq:prop_df}.
+
+The log-fold change in the adjusted PTM abundance, $\Delta$, is now estimated by RML, and Equation \ref{eq:prop_fc} becomes 
 \begin{equation}
@@
 
-The proposed methods are implemented in the open source R package $MSstatsPTM$, available on Bioconductor.
+The proposed methods are implemented in the open source R package $MSstatsPTM$, available on Bioconductor and Github.
+An overview of the steps of the package are illustrated in Figure \ref{fig:msstatsptm_design}.
 $MSstatsPTM$ includes converters for multiple spectral processing tools, including MaxQuant, Progenesis, and Spectronaut.
@@
 If the global profiling run is not available, the package can still analyze the modified run, but will do so without adjusting for changes in unmodified protein abundance.
-Specifically, if a global profiling run is not available the null and alternative hypothesis in Equation \ref{eq_1} will reduce to:
-
-\begin{equation}
-\begin{aligned}
-H_{0}: \Delta = \mu_{i}^{\ast} - \mu_{i} = 0 \\
-H_{a}: \Delta = \mu_{i}^{\ast} - \mu_{i} \neq 0
-\end{aligned}
-\label{eq_9}
-\end{equation}
-
-Which amounts to running the methods seen in MSstats and MSstatsTMT on peptide level data \cite{Choi:2014} \cite{Huang:2020}.
+Specifically, if a global profiling run is not available the hypotheses in Equation \ref{eq:conv_null_hyp} will be used; which amounts to running the methods seen in $MSstats$ on modified peptide level data \cite{Choi:2014} \cite{Huang:2020}.
 
 After using the converters, the next step is peptide/protein summarization and missing value imputation.
-In the methods sections it was assumed that all the input data was summarized; in the implementation we need to run a specific summarization step to make this happen.
-For the modified run, the package summarizes features, PSMs, which include the same modification together.
+For the modified run, the package summarizes features which include the same modification together.
 Features with multiple modification sites are not included with single site features and are summarized separately.
@@
 Additionally, the summarization includes global median normalization and normalizes between MS runs.
-The package uses an AFT model to impute missing values, although this step is optional.
+The package can optionally impute missing values using an AFT model.
 
@@
 The statistical modeling relies on the functionality from the R packages $lme4$ \cite{Bates2015} and $lmerTest$ \cite{Kuznetsova2017}.
-An overview of the steps of the package are illustrated in Figure \ref{fig:msstatsptm_design}.
 
@@
 Additionally, when comparing the fold change estimation across all modified peptides, the proposed method showed a tighter distribution of estimated fold changes around the true fold change, as compared to $t$-test and Limma.
-Specifically, the inner quartile range of the estimated fold change for the proposed method was on average 10.4\% smaller than $t$-test and 21.8\% smaller than Limma.
+Specifically, the inner quartile range (IQR) of the estimated fold change for the proposed method was on average 10.4\% smaller than $t$-test and 21.8\% smaller than Limma.
 This fold change comparison can be seen in Supplementary Figure \ref{supp-fig:fc_boxplot}.
@@
 In this experiment all models incorrectly estimated the fold change of the modified spike-in peptides before adjusting for changes in global protein abundance.
-After adjustment, the spike-in peptides' fold change was generally in line with expectation (Figure \ref{fig:spikein_boxplot}), however the distribution of estimated fold changes was visibly wider in all methods.
+After adjustment, the spike-in peptides' fold change was generally in line with expectation in all methods, however the distribution of estimated fold changes was visibly wider (Figure \ref{fig:spikein_boxplot}).
 Of the three approaches, the proposed method showed the tightest distribution around the true log fold change.
-Comparing the inner quartile range (IQR) of the spike-in peptide's log fold change, the proposed method's IQR was 36.78\% smaller than Limma's and 32.98\% smaller than $t$-test's.
-This means that the proposed method's fold change estimate was more consistently correct, when compared to Limma and $t$-test.
+Comparing the IQR of the spike-in peptide's log fold change, the proposed method's IQR was 36.78\% smaller than Limma's and 32.98\% smaller than $t$-test's.
+This means that the proposed method's fold change estimate was more consistently correct when compared to Limma and $t$-test.
 
@@
 After adjustment, the estimated fold change was more in line with expectation.
-Additionally, the background peptides serving as the null model, show many false positives before adjustment was made.
-After adjustment the results improved and the number of false positives decreased significantly.
-Specifically, for the proposed method, the number of false positives went from 20.88\% to 1.84\% of all true negatives after adjustment was applied.
+Additionally, the background peptides serving as the null model show many false positives before adjustment was made, however after adjustment the number of false positives decreased significantly.
+Specifically, for the proposed method, the number of false positives went from 20.88\% to 1.84\% after adjustment was applied.
 Using both the proposed method and Limma, the red labeled spike-in peptides were more inline with expected fold change after the adjustment was applied.
-However, using Limma resulted in the majority of the differential spike-in peptides not showing a significant adjusted p-value.
+However, Limma resulted in the majority of the differential spike-in peptides having an insignificant adjusted p-value.
 In this case, using Limma would have resulted in missing the majority of differentially modified peptides.
@@
 The results of this experiment are summarized in Figure \ref{fig:data4_plots}.
-The experiment was modeled as a group comparison.
 In Figure \ref{fig:data4_venn_diagram} the number of significant modified peptides before and after adjustment is shown.
@@
 
-In Figure \ref{fig:data4_profile_plot}, the modification of protein $GSDMD$ at $K62$ showed the advantage of the proposed method.
+In Figure \ref{fig:data4_profile_plot}, the modification of protein $GSDMD$ at site $K62$ showed the advantage of the proposed method.
 The modified peptide originally showed no abundance change between the infected 1 hour, 4 hour, and 6 hour conditions.
@@
 Any modified peptide without a corresponding unmodified protein could not be adjusted.
-Of the 10,799 modified peptides identified, only 4526 had a corresponding unmodified protein and could be adjusted.
+Of the 10,799 ubiquitin sites identified, only 4526 had a corresponding unmodified protein and could be adjusted.
 Additionally, the lack of a separate global profiling run resulted in low feature counts for the unmodified protein model.
@@
 
-\subsubsection*{Noisy PTM measurements benefited from additional biological replicates}
-
-\todo{I would like to move this section to the supplementary and just referencing it in the main.}
-
-Here we analyzed the sample size needed to achieve a desired statistical power.
-The proposed approach corrected for confounding between the modified peptide and unmodified protein at the cost of increased variation.
-This can be seen in the calculation for variance in Section \ref{sec:adjustment_summary}.
-Increased variation required a larger number of replicates to reach the same power.
-Thus the statistical power was dependent on the variance from both the modified peptide and unmodifed protein.
-
-We compared the statistical power in experiments with differing numbers of replicates, variance, and fold change for both the modified and unmodified runs.
-In terms of the number of replicates, we tested scenarios with equal replicates in both the modified and unmodified runs, as well as scenarios where the replicates differed between runs.
-We used the biological experiments to determine what variance values to test.
-In datasets 4 and 5 the variance of the PTM was higher than the global protein.
-In dataset 6 the variance of the PTM and Protein was generally the same.
-We mimicked these scenarios and analyzed the power of experiments when the PTM variance was higher than the protein and when they were equal.
-When the PTM and protein were the same we chose a variance of $.15$, whereas when the PTM was higher than the protein we chose a PTM variance of $.2$ and a protein variance of $.1$.
-
-The results of the power and sample size analysis can be seen in Supplementary Figure \ref{supp-fig:power_sd_combo}.
-When the variance and replicates were equal, higher replicates predictably lead to higher power.
-In cases where the replicates were unbalanced, but the variance was still the same, it did not matter if there were more replicates in the modified or unmodified runs.
-In comparison, with differing variance and equal replicates, higher replicates still lead to higher power.
-When the replicates were unbalanced and the variance was higher in the PTM, there was more power when there were more replicates in the PTM than the protein.
-In this case it was clearly more important to have high replicates for the PTM run than the unmodified protein.
-In cases where the number of replicates has to be limited, it was better to add more to the PTM side.
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Discussion}
-
-\todo{Although demonstrated here on DDA, is also applicable to DIA, SRP and PRM acquisitions}
 
@@
 The proposed framework can be extended beyond the experimental designs with variation from multiple sources discussed above.
-For example, it can represent experimental designs with even more complex structures, such as time series or factorial investigations.
+Although demonstrated here on DDA, is also applicable to DIA, SRP and PRM acquisitions.
 Additionally, the approach can handle experiments with modified peptides processed using label-free methods and unmodified peptides processed using TMT labeling, or vice versa.
@@
 Goals of PTM characterization, input to statistical analyses, and notation.
-\todo{Could you update the caption?
-The figure now has only 2 parts.
-It doesn't have subscript $i$ etc.
-On the other hand, it distinguishes the population parameters, and the observed data.
-Could you mention this here briefly, and describe in more detail in the text?}  (a) Schematic data representation, in a simplified case of two conditions and two \todo{biological?} replicate runs.
-Each PTM site is modeled and characterized separately, where a PTM is quantified with multiple spectral features (boxes), distinguished by different charge states of a peptide.
-The feature intensities the PTM in Condition i are denoted by i*.
-Features corresponding to unmodified peptides in Condition i is denoted by i.
+(a) Schematic data representation, in a simplified case of two conditions and two \todo{biological?} replicate runs.
+Each PTM site is modeled and characterized separately, where a PTM is quantified with observed spectral features (boxes), distinguished by different charge states of a peptide.
+The feature intensities of the PTM in Condition 1 are denoted by $y_1^\ast$.
+Features corresponding to unmodified peptides in Condition 1 are denoted by $y_1$.
 Peptides can be fully cleaved (solid lines) and/or partially cleaved (dashed lines).
-(b) PTM relative quantification by statistical inference, which makes use of the feature intensities to infer the underlying PTM abundance and protein abundance with an estimate of associated uncertainty.
+(b) PTM relative quantification of the population parameters by statistical inference, which makes use of the feature intensities to infer the underlying PTM abundance and protein abundance with an estimate of associated uncertainty.
 (c) Model-based testing for differential PTM abundance, which corrects for the underlying protein abundance with a cost of increased uncertainty about the estimate of difference between conditions.}
@@
 \includegraphics[scale=.8]{images/MSstatsPTM_design.png}
-\caption{The workflow of MSstatsPTM and how it fits into the experimental pipeline.
-MSstatsPTM's workflow starts after modified and unmodified peptide quantification.
-First tool specific pre-processing is done, this includes modification site identification, general data cleaning, and formatting the data into the format needed for the package.
+\caption{The workflow of $MSstatsPTM$ and how it fits into the experimental pipeline.
+The package's workflow starts after modified and unmodified peptide quantification.
+First tool specific pre-processing is performed, this includes modification site identification, general data cleaning, and formatting the data into the format needed for the package.
 The next step is feature level summarization, which summarizes features up to the modification level for the PTM data, and the protein level for the protein data.
-In the final step a model is fit to identify differential PTMs and unmodified proteins across conditions and the PTM model is adjusted for changes in the unmodified protein.}
+In the final step a model is fit to identify differential PTMs and unmodified proteins across conditions and the PTM model is adjusted for changes in the unmodified protein.
+After both the summarization and group comparison steps, plots can be created to summarize the results.}
 \label{fig:msstatsptm_design}
@@
 After adjustment, this systemic difference was removed, however the inner quartile range of the Limma and $t$-test models was wider than the proposed method.
-b) The spike in peptides (colored red) did not follow the expected log fold change before adjustment.
-However, after adjustment, the spike in peptides were more in line with expectation.
-Using Limma the spike in peptides followed the expected log fold change after adjustment, however the majority of spike in peptides did not have a significant adjusted p-value.}
+b) Before adjustment the spike in peptides (colored red) did not follow the expected log fold change; after adjustment, the spike in peptides were more in line with expectation.
+Using Limma, the spike in peptides followed the expected log fold change after adjustment, however the majority of spike in peptides did not have a significant adjusted p-value.}
 \label{fig:spikein_volcano}
@@
 a) The overlap of differential modified peptides for the PTM model with and without global protein level adjustment.
-More PTMs became insignificant after adjustment then became significant.
+More PTMs became insignificant after adjustment than became significant.
 For the peptides that became insignificant in the adjusted model, their change in abundance was driven by changes in the global protein.
-In contrast, peptides that became significant after adjustment saw their true abundance change masked by underlying changes in the unmodified protein.
+In contrast, peptides that became significant after adjustment had their true abundance change masked by underlying changes in the unmodified protein.
 b) Comparing the global profiling of protein $GSDMD$ with the ubiquitination of the protein at site $K62$.

### 2022-03-28 supplementary/ptm_sm.tex
rewrote background section

@@
 Differential levels of modified peptides may be due to differential modifications and/or changes in protein abundance.
-The proposed approach adjusts the PTM abundance with respect to protein abundance as introduced in \secref{sec:prop}.
-Alternatively, two-sample $t$-test or Limma taking as input the ratio between feature intensities of modified and unmodified peptides is commonly applied for the same purpose (\secref{sec:ttest}).
 Approaches without considering protein-level adjustment lose track of an important aspect in interpreting observed changes in PTM abundance, which may result in misleading conclusions.
@@
 While multiple conditions were generated, all comparisons were still made with the same fold change between conditions ($0.75$).
-This ensured that all peptides were differentially abundant by the same amount and the only difference in TP/FP/TN/FN calculations was using more conditions in the modeling.
+This ensured that all peptides were differentially abundant by the same amount and the only difference in calculations was using more conditions in the modeling.
 
@@
 
-When simulating the data, 500 of the 1000 peptides were simulated as true positives with a differential fold change between conditions, while the other 500 were not differential and treated as true negatives.
-In terms of the true negatives, 250 were simulated such that any abundance changes were entirely driven by changes in the unmodified peptide.
-We would expect any model without performing protein adjustment to treat these as false positives.
-The other 250 were simulated with no significant change in either PTM or overall protein abundance.
-
 \subsubsection{Dataset 2 : Computer simulation 2 - Label-free missing values and low features}
@@
 
-The portion of significant peptides were done in the same was as described in simulation 1.
-
 \begin{figure}[h!]
@@
 	\includegraphics[width=.8\textwidth]{sim_new/sim1_FDR}
-	\caption{The considered methods with protein adjustment are compared in detail. All three methods with adjustment generally performed similarly in terms of FDR.}
+	\caption{The considered methods with protein adjustment were compared in detail. All three methods with adjustment generally performed similarly in terms of FDR.}
  \end{subfigure}
@@
 	\includegraphics[width=.8\textwidth]{sim_new/sim1_Recall}\\
-	\caption{The recall for simulation 1 methods with adjustment were compared.
-It is clear that Limma performs the strongest here when the number of replicates were low.
-At higher replicates the performance of the proposed methods and Limma are comparable.
-$t$-test clearly performs worse across all methods.}
+	\caption{All methods with adjustment were compared by comparing recall in simulation 1.
+Limma performed the strongest here when the number of replicates were low.
+At higher replicates the performance of the proposed methods and Limma were comparable.
+$t$-test clearly performed worse across all methods.}
  \end{subfigure}
@@
 	\includegraphics[width=.8\textwidth]{sim_new/sim1_accuracy}
-	\caption{The overall accuracy plot mimics the observations in the recall plot. Limma performs stronger than the proposed method at lower replicates, while at higher they are comparable.}
+	\caption{The overall accuracy plot mimicked the observations in the recall plot. Limma performed strongest when replicates were low, but was comparable to the proposed method with more replicates.}
  \end{subfigure}
@@
 \caption{Recall results of Simulation 2.
-The advantage of using the proposed approach is apparent when looking at simulation 2, which includes limited observations and the presence of missing values.
-In the case of recall the proposed method performs stronger than Limma and t-test in nearly every model.
+The advantage of using the proposed approach were apparent when looking at simulation 2, which included limited observations and the presence of missing values.
+In the case of recall the proposed method performed stronger than Limma and t-test in nearly every model.
 Even at lower replicates the proposed method still outperformed Limma.
@@
 	\caption{In simulation 1 all considered methods correctly estimated the fold change between conditions, with a median fold change estimation of .75.
-The distributions around the median were also consistent across all methods.
-Predictably the distribution quantiles decrease as the number of replicates increases.}
+The distributions around the median were also consistent across all methods.}
  \end{subfigure}
@@
 Both Limma and $t$-test showed a wider range around the fold change.
-The inner quartile range of the Proposed method was on average 10.4\% smaller than t-test and 21.8\% smaller than Limma.
-In this case the proposed method showed a stronger performance in correctly estimating the fold change for all peptides.}
+The inner quartile range of the Proposed method was on average 10.4\% smaller than t-test and 21.8\% smaller than Limma.}
  \end{subfigure}
@@
 All methods are analyzed after adjusting for changes in overall protein level.
-The proposed method summarizes feature intensities up to the run level using Tukey's Median Polish, while the other methods use the log of summed feature intensities.
 The results are summarized from Figure \ref{fig:spike_volcano_msstats} to Figure \ref{fig:spike_stat}, including volcano plots, model summary statistics, and fold change analysis.
@@
 All grey peptides are expected to be insignificant.
-Using the proposed method to model the benchmark experiment, the spike-in peptides (colored red) do not follow the expected log fold change before adjustment.
-After adjusting for changes in overall protein abundance the spike-in peptides are more in line with expectation.
-Additionally the background grey colored peptides show many false positives before adjustment.
+Using the proposed method to model the benchmark experiment, the spike-in peptides (colored red) did not follow the expected log fold change before adjustment.
+After adjusting for changes in overall protein abundance the spike-in peptides were more in line with expectation.
+Additionally the background grey colored peptides showed many false positives before adjustment.
 After adjustment these false positives were decreased considerably.
@@
 \caption{When modeling the experiment with the Limma method, the spike-in peptides again follow the expected log fold change better after adjusting for changes in protein level.
-However, while the fold change is much more accurate, the majority of spike-in peptides do not have a significant adjusted pvalue.
-In this case, the known differential peptides are missed by the model.
-In terms of false positives, the results are very similar to the proposed method, with many false positives before adjustment and much fewer after.
+However, while the fold change was more accurate, the majority of spike-in peptides did not have a significant adjusted pvalue.
+In this case, the known differential peptides were missed by the model.
+In terms of false positives, the results were very similar to the proposed method, with many false positives before adjustment and fewer after.
 \label{fig:spike_volcano_limma}}
@@
 \caption{Using the two sample $t$-test, none of the comparisons either before or after adjustment show any significant peptides.
-With that being said, the fold change of the spike-in peptides is much closer to expectation after adjusting for global protein abundance.
+With that being said, the fold change of the spike-in peptides was much closer to expectation after adjusting for global protein abundance.
 \label{fig:spike_volcano_ttest}}
@@
 \includegraphics[width=.85\textwidth]{sim_new/spike_in_statistic_comparison}
-\caption{Comparing the summary statistics between methods it is clear that the proposed method performs the strongest.
-In terms of accuracy and specificity the three methods are close, with Limma and $t$-test showing slightly higher values.
-Accuracy and specificity are dominated by the large number of true negatives (background peptides) compared to the true positives (spike-in peptides).
-In terms of recall, the proposed approach far out performed the other two methods, showing that it correctly labeled the most spike-in peptides.
+\caption{Comparing the summary statistics between methods, the proposed method performed the strongest.
+In terms of accuracy and specificity the three methods were close, with Limma and $t$-test showing slightly higher values.
+Accuracy and specificity were dominated by the large number of true negatives (background peptides) compared to the true positives (spike-in peptides).
+In terms of recall, the proposed approach out performed the other two methods, showing that it correctly labeled the most spike-in peptides.
 \label{fig:spike_stat}}
@@
 
-In this experiment Shigella ubiquitin ligase IpaH7.8 was shown to inhibit the protein gasdermin D (GSDMD).
-\cite{LUCHETTI2021} Multiplex proteomics was used to quantify the abundance of total protein, and ubiquitination in human epithelial cells.
+In this experiment Shigella ubiquitin ligase IpaH7.8 was shown to inhibit the protein gasdermin D (GSDMD) ~\cite{LUCHETTI2021}.
+Multiplex proteomics was used to quantify the abundance of total protein, and ubiquitination in human epithelial cells.
 Cells were either infected or uninfected with IpaH7.8-deficient Shigella flexneri and measurements were taken at different time periods.
@@
 	\includegraphics[width=.9\textwidth]{sim_new/IpaH_prof_plot}
-	\caption{Comparing the global profiling of protein $GSDMD\_HUMAN|P57764$ with the ubiquitination of the protein at site $K62$.
-The individual PSM features are shown in grey, while the feature summarization is shown in red.
-When looking at the summary of the modification and global protein it is clear the conditions follow different trends.
-Specifically, there appears to be no change in abundance between Dox1hr and Dox4hr in the modified plot, however there is a large negative change when looking at the unmodified plot.
-This indicates the modification is confounded with changes in the unmodified protein.}
+	\caption{Here the global profiling of protein $GSDMD\_HUMAN|P57764$ with the ubiquitination of the protein at site $K62$ was compared.
+The individual PSM features areshown in grey, while the feature summarization is shown in red.
+The summary of the modification and global protein showed that the conditions followed different trends.
+Specifically, there appeared to be no change in abundance between Dox1hr and Dox4hr in the modified plot, however there was a large negative change in the unmodified plot.
+This indicates the modification was confounded with changes in the unmodified protein.}
  \end{subfigure}
@@
 The $GSDMD\_HUMAN|P57764\_K62$ modification is highlighted.
-Before adjustment the modification has a small fold change and insignificant adjusted pvalue.
-After adjustment the fold change is much larger and the adjusted pvalue is very significant.
-In this case the proposed method allows us to identify a differential modified peptide that could have otherwise been missed.}
+Before adjustment the modification had a small fold change and insignificant adjusted pvalue.
+After adjustment the fold change was much larger and the adjusted pvalue was significant.
+In this case the proposed method allowed us to identify a differential modified peptide that could have otherwise been missed.}
  \end{subfigure}
@@
 
-In this study, the correlation between the gene Atg16L1 and killing of Shigella flexneri (S.flexneri) was assessed.
-\cite{Maculins} Multiplex proteomics was used to quantify the abundance of total protein, phosphorylation, and ubiquitination in wild type (WT) and ATG16L1-deficient (cKO) samples, uninfected and uninfected with S.flexneri.
+In this study, the correlation between the gene Atg16L1 and killing of Shigella flexneri (S.flexneri) was assessed~\cite{Maculins}.
+Multiplex proteomics was used to quantify the abundance of total protein, phosphorylation, and ubiquitination in wild type (WT) and ATG16L1-deficient (cKO) samples, uninfected and uninfected with S.flexneri.
 The abundance of total protein and post-translation modifications were quantified at three time points, uninfected, early infection (45-60 minutes), and late infection (3-3.5 hours).
@@
 	\includegraphics[width=.95\textwidth]{sim_new/No_Difference_Shigella_Profile_Plot}
-	\caption{Comparing the global profiling of protein $TTP\_MOUSE|P22893$ with the modification of the protein at site $S178$.
+	\caption{The global profiling of protein $TTP\_MOUSE|P22893$ with the modification of the protein at site $S178$ was compared.
 The individual PSM features are shown in grey, while the feature summarization is shown in red.
-When looking at the summary of the modification and global protein it is clear the difference between conditions follow the same trend.
-Specifically, there is a positive adjustment in abundance when comparing WT\_Uninfect to WT\_Late in both the modification and global profiling run.
-This indicates the movement is driven by changes in global protein.}
+The summary of the modification and global protein showed that the conditions followed the same trend.
+Specifically, there was a positive adjustment in abundance when comparing WT\_Uninfect to WT\_Late in both the modification and global profiling run.
+This indicated the movement is driven by changes in global protein.}
  \end{subfigure}
@@
 The $TTP\_MOUSE|P22893\_S178$ modification is highlighted.
-Before adjustment the modification has a large fold change and significant adjusted pvalue.
-After adjustment the fold change is much smaller and the adjusted pvalue is insignificant.}
+Before adjustment the modification had a large fold change and significant adjusted pvalue.
+After adjustment the fold change was much smaller and the adjusted pvalue was insignificant.}
  \end{subfigure}
@@
 	\includegraphics[width=.9\textwidth]{sim_new/Difference_Shigella_Profile_Plot}
-	\caption{Comparing the global profiling of protein $KI67\_MOUSE|E9PVX6$ with the modification at site $T215$.
-In this case the modification and global protein appear to show very small to no differences between conditions, however after adjusting for change in global protein abundance, the modification is statistically significant.
-Additionally, this profile plot shows the large difference in available features between modified peptides and global proteins.}
+	\caption{The global profiling of protein $KI67\_MOUSE|E9PVX6$ with the modification at site $T215$ was compared.
+In this case the modification and global protein appeared to show small or no difference between conditions, however after adjusting for change in global protein abundance, the modification was statistically significant.
+Additionally, this profile plot showed the large difference in available features between modified peptides and global proteins.}
  \end{subfigure}
@@
 	\includegraphics[width=.9\textwidth]{sim_new/Difference_Shigella_Volcano}
-	\caption{The volcano plot of the WT\_Uninfect and WT\_Early comparison show the specifics of the adjustment.
-The modification looks to be flat, with a log fold change of $.185$, while the global profiling showed a small negative fold change of $-.619$.
-While both exhibit small changes, when combined we can see a log fold change of $.804$ and adjusted p-value of $.0156$.
-In this case we can see the strength of the proposed method, showing noteworthy modifications which would be missed otherwise.}
+	\caption{The volcano plot of the WT\_Uninfect and WT\_Early comparison showed the specifics of the adjustment.
+The modification looked to be flat, with a log fold change of $.185$, while the global profiling showed a small negative fold change of $-.619$.
+While both exhibit small changes, when combined we saw a log fold change of $.804$ and adjusted p-value of $.0156$.}
  \end{subfigure}
@@
 \includegraphics[width=\textwidth]{sim_new/USP30_profile_plot}
-\caption{Comparing the global profiling of protein $P52209$ with the modification of the protein at site $K059$.
-The modification appears generally unchanged between all conditions, whereas the global profiling run shows the CCCP and Combo conditions have a higher relative abundance compared to the Control and USP30\_OE.
-This indicates that the modification actually had an effect when comparing CCCP and Combo to Control and USP30\_OE.
-However it is not entirely clear as one unmodified peptide feature appears to be changed, while the other does not.
-This uncertainty is another result of not running a separate global profiling run.
+\caption{The global profiling of protein $P52209$ with the modification of the protein at site $K059$ was compared.
+The modification appeared generally unchanged between all conditions, whereas the global profiling run showed the CCCP and Combo conditions had a higher relative abundance compared to the Control and USP30\_OE.
+This indicated that the modification actually had an effect when comparing CCCP and Combo to Control and USP30\_OE.
+However it was not entirely clear as one unmodified peptide feature appeared to be changed, while the other did not.
+This uncertainty was another result of not running a separate global profiling run.
 With a global profiling run, many unmodified features are generally quantified, removing the uncertainty that comes with low feature counts.}
@@
 
+\subsubsection*{Noisy PTM measurements benefited from additional biological replicates}
+
+Here we analyzed the sample size needed to achieve a desired statistical power.
+The proposed approach corrected for confounding between the modified peptide and unmodified protein at the cost of increased variation.
+This can be seen in the calculation for variance.
+Increased variation required a larger number of replicates to reach the same power.
+Thus the statistical power was dependent on the variance from both the modified peptide and unmodifed protein.
+
+We compared the statistical power in experiments with differing numbers of replicates, variance, and fold change for both the modified and unmodified runs.
+In terms of the number of replicates, we tested scenarios with equal replicates in both the modified and unmodified runs, as well as scenarios where the replicates differed between runs.
+We used the biological experiments to determine what variance values to test.
+In datasets 4 and 5 the variance of the PTM was higher than the global protein.
+In dataset 6 the variance of the PTM and Protein were generally the same.
+We mimicked these scenarios and analyzed the power of experiments when the PTM variance was higher than the protein and when they were equal.
+When the PTM and protein were the same we chose a variance of $.15$, whereas when the PTM was higher than the protein we chose a PTM variance of $.2$ and a protein variance of $.1$.
+
+The results of the power and sample size analysis can be seen in Figure \ref{fig:power_sd_combo}.
+When the variance and replicates were equal, higher replicates predictably lead to higher power.
+In cases where the replicates were unbalanced, but the variance was still the same, it did not matter if there were more replicates in the modified or unmodified runs.
+In comparison, with differing variance and equal replicates, higher replicates still lead to higher power.
+When the replicates were unbalanced and the variance was higher in the PTM, there was more power when more replicates were allocated to the PTM run than the unmodified protein run.
+The results lead to the takeaway that in cases where the number of replicates have to be limited, it is better to allocate more to the PTM run.
+
 \begin{figure}[h!]
@@
 	\caption{The power of an experiment targeting PTMs with the same variance, .15, for the modified and unmodified peptides.
-Predictably when the replicates are high for both modified and unmodified peptides the power is much higher.
-Conversely at low replicates for each the power is much lower.
-The interesting part of this chart is when the replicates are different between runs.
-With equal variance, it does not matter if the PTM replicates or protein replicates are higher.}
+Predictably when the replicates were high for both modified and unmodified peptides the power was much higher.
+Conversely at low replicates for each the power was much lower.
+With equal variance, it did not matter if the PTM replicates or protein replicates were higher.}
  \end{subfigure}\vspace{-1mm}
@@
 	\includegraphics[width=.75\textwidth]{sim_new/high_ptm_var_power}
-	\caption{In this chart the variance for the PTM is higher than the unmodified protein.
-The PTM variance is .2, while the unmodified protein variance is .1.
-With equal replicates the results are the same as above, more replicates equals more power.
-When the replicates are not the same we can clearly see that having more replicates for the PTM leads to higher power.}
+	\caption{In this chart the variance for the PTM was higher than the unmodified protein.
+The PTM variance was .2, while the unmodified protein variance was .1.
+With equal replicates the results were the same as above.
+When the replicates were not equal, more replicates allocated to the PTM runs lead to higher power.}
  \end{subfigure}\vspace{-1mm}

### 2022-04-03 main/MSstatsPTM_Main.tex
updated main background sec

@@
 Table \ref{tab:dataDescription} summarizes the experiments.
+
+\subsection*{Data Availability}
+
 Details of the experimental data, R scripts with $MSstatsPTM$ analysis, and results of the statistical analysis are available in MassIVE.quant (\url{https://massive.ucsd.edu/ProteoSAFe/static/massive-quant.jsp}) \cite{Choi:2020}.
@@
 
+The R scripts and results of the statistical analysis for the four experiments are available in MassIVE.quant reanalysis containers: SpikeIn benchmark - Ubiquitination - Label-free (MassIVE reanalysis container identifier: \href{https://massive.ucsd.edu/ProteoSAFe/private-dataset.jsp?task=c4c583ecf7f941cdac87f7a4f872517b}{TBD}), Human - Ubiquitination - 1mix-TMT (MassIVE reanalysis container identifier: \href{https://massive.ucsd.edu/ProteoSAFe/reanalysis_container.jsp?task=b639fa5d0ff24ce78abdca13f4b18c0a}{RMSV000000356}), Mouse - Phosphorylation - 2mix-TMT time series (MassIVE reanalysis container identifier: \href{https://massive.ucsd.edu/ProteoSAFe/reanalysis_container.jsp?task=79406d32c1f9413daf7756e6eeb24968}{RMSV000000357}), Human - Ubiquitination - Label-free no global profiling run (MassIVE reanalysis container identifier: \href{https://massive.ucsd.edu/ProteoSAFe/reanalysis_container.jsp?task=f9e6117e00794ce5b2e8802601be16bc}{RMSV000000358}).
+
 \subsection*{Dataset 1 : Computer simulation 1 - Label-free}
@@
 Additionally, 20\% of observations for both modified and unmodified peptides were missing completely at random.
-\todo{There will likely be a comment that missing values are more common among low-abundance features.
-Did the simulation account for that?
-Devon: The simulation did not include any missing values due to low abundance.
-Should we mention this or just leave it as is?} Further details are in Supplementary Sec.
+Further details are in Supplementary Sec.
 \ref{supp-sec:dataset2}.
@@
 
-In the custom spike-in experiment 8 biological \todo{I am not sure these are bio reps.
-If this is the same lysate, then these are in fact technical reps (there is no between-subject variation)} replicates were allocated to 4 conditions, resulting in 2 biological replicates per condition, and a balanced design.
-Additional information can be found in Supplementary Sec.
-\ref{supp-sec:benchmark}.
+\todo{Devon: I moved the figure to main and tried to adjust this section to be clearer. I still think this section could benefit from Erik and Tsung-Heng's insight. Will ask them to look in email.}
+
+In the custom spike-in experiment 8 replicates were allocated to 4 conditions, resulting in 2 replicates per condition.
 
@@
 Quantitative changes in protein and site abundance of these 20 proteins were the target of the benchmark.
-Unmodified peptides from Human lysate were used to estimate changes in global protein abundance.
+An overview of the experimental design is shown in Figure \ref{fig:benchmark-design}.
+Four mixture conditions were created with differing amounts of human lysate, background lysate, and human spike-in ub-peptide mixture.
+Unmodified peptides from human lysate were used to estimate changes in global protein abundance.
 Background {\it E.
 coli} lysate was used to normalize total protein levels prior to enrichment or global protein profiling.
-The background lysate was treated as control, i.e.
-it had no expected changes in any comparison.
-\todo{I am a bit confused.
-Fig S7 shows that there is a difference in Background between Mix 1,2 and Mix 3,4.
-Could you explain better?
-In fact, would it be possible to move the figure here for clarity, and have the figure as panel (a), and the table as panel (b)?
-And in the text above refer to the figure.
-This will help the understanding} \todo{Also, why is it a balanced design - as stated in the paragraph above - when the background is not present everywhere?}
-
-The spike-in peptides were mixed with human lysate to create four mixture conditions.
-The true log fold changes of spike-in modified peptides in these comparisons can be seen in Table \ref{table:spikein_fold_change}.
+The background lysate was treated as control and is expected to show no fold change in all comparisons except mix1-mix2 and mix3-mix4.
+The true log fold changes of spike-in modified peptides in each comparison can be seen in Figure \ref{fig:benchmark-design}(b).
 Two sets of data were acquired for each mixture in label-free LC-MS/MS mode: KGG enriched + LC-MS/MS, and LC-MS/MS only.
@@
 
-\begin{table}[h!]
-\centering
-\begin{tabular}{| c | c |}
-\hline
- Comparison & Expected log fold change\\ [0.5ex]
- \hline\hline
- mix1-mix2 & -1\\
- \hline
- mix1-mix3 & 1\\
-\hline
- mix1-mix4 & 0\\
-\hline
- mix2-mix3 & -2\\
-\hline
- mix2-mix4 & -1\\
-\hline
- mix3-mix4 & 1\\
-\hline
-\end{tabular}
-\caption{The expected log fold change of modified spike-in peptides in Dataset 3.}
-\label{table:spikein_fold_change}
-\end{table}
+%\begin{table}[h!]
+%\centering
+%\begin{tabular}{| c | c |}
+%\hline
+% Comparison & Expected log fold change\\ [0.5ex]
+% \hline\hline
+% mix1-mix2 & -1\\
+% \hline
+% mix1-mix3 & 1\\
+%\hline
+% mix1-mix4 & 0\\
+%\hline
+% mix2-mix3 & -2\\
+%\hline
+% mix2-mix4 & -1\\
+%\hline
+% mix3-mix4 & 1\\
+%\hline
+%\end{tabular}
+%\caption{The expected log fold change of modified spike-in peptides in Dataset 3.}
+%\label{table:spikein_fold_change}
+%\end{table}
 
@@
 \label{sec:exp_proc_dataset4}
-In this experiment 11 biological replicates were allocated to 1 TMT mixture in an \todo{group comparison or time course?} unbalanced design.
-Further details are in Supplementary Sec.
-\ref{supp-sec:ipah}.
+In this experiment 11 biological replicates were allocated to 1 TMT mixture in a group comparison unbalanced design. Further details are in Supplementary Sec. \ref{supp-sec:ipah}.
 
@@
 \todo{If this is a group comparison, explain why, even though the conditions are times.
-The table above says it is group comparison} The proteome was quantified using isobaric multiplexing via tandem mass tagging (TMT) in combination with LC-MS/MS.
-\todo{One MS run total?
-Only one MS run was used.
-Not sure the best way to say this.}
+The table above says it is group comparison} The proteome was quantified using isobaric multiplexing via tandem mass tagging (TMT) in combination with one LC-MS/MS run.
 
@@
 
-Consider an LC-MS/MS experiment with $I$ conditions and $J$ mass spectrometry runs (technical \todo{of biological?} replicates) per condition in the experiment.
+For simplicity, consider an LC-MS/MS experiment in the special case of a balanced design with $I=2$ conditions and $J=2$ biological replicates (mass spectrometry runs) per condition.
 For one protein, the PTM site is represented by $K$ spectral features (peptide ions, distinguished by their cleavage residues and charge states).
@@
 
-Figure \ref{fig:data-structure}(a) schematically illustrates this data structure for one protein in a simplified experiment with $I=2$ conditions.
-The data are collected in $J=2$ \todo{biological?} replicate runs.
-Each run represents an observation used to estimate the relative abundance of the PTM in condition $i$, $u_i^{\ast}$.
+Figure \ref{fig:data-structure}(a) schematically illustrates this data structure for one PTM and corresponding unmodified protein.
 The number of quantified features vary across replicate LC-MS/MS runs of the same sample, and across conditions.
@@
 Additionally, some spectral features can be missing.
+The features in each mass spectrometry run are generally summarized and represent a single observation used to estimate the relative abundance of the PTM in condition $i$, $u_i^{\ast}$.
 
@@
 To calculate the quantity of interest, statistical analysis needs to summarize values in the table in Figure \ref{fig:data-structure}(a) using appropriate models.
-For condition $i$, the models provide estimates for the mean abundance, $\hat{u}_i^{\ast}$, and standard error, $SE(\hat{u}_i^{\ast})$.
+For condition $i$, the models provide estimates for the mean abundance, $\hat{u}_i^{\ast}$, and standard error, $\hat{SE}(\hat{u}_i^{\ast})$.
 Once summarized, we can formulate the Null and Alternative Hypotheses as seen in Equation \ref{eq:conv_null_hyp}.
@@
 However, these population parameters are inherently convoluted with changes in the unmodified protein.
-To account for this, we must estimate the unmodified protein in conditions $1$ and $2$, denoted by $\mu_1$ and $\mu_2$.
-We estimate these quantities in the same way as for the PTM, using their correspond features and observations, resulting in an estimated mean abundance, $\hat{u}_i$, and standard error, $SE(\hat{u}_i)$.
+To account for this, we must estimate the relative abundance of the unmodified protein in conditions $1$ and $2$, denoted by $\mu_1$ and $\mu_2$.
+We estimate these quantities in the same way as for the PTM, using their correspond features and observations and resulting in an estimated mean abundance, $\hat{u}_i$, and standard error, $\hat{SE}(\hat{u}_i)$.
 These estimates are used to adjust the quantities estimated for the PTM, resulting in the new Null and Alternative Hypotheses seen in Equation \ref{eq:null_hyp}.
@@
 
-\todo{in Figure 2 b, there are two quantities of interest, stated at the population level.
-Describe the quantities, what is of interest, why relevant.
-It may help to introduce multiple versions of the null hypotheses, to contrast the methods below} \todo{Also please edit the figure caption}
-
 %---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------
@@
 
-\noindent {\bf Two-sample $t$-test} \todo{When describing the existing methods, could you refer to Fig 1?
-Here is the suggested order: (1) What is the null hypothesis?
-Do these methods test the same null hypothesis?
-(If not, it may be a good idea to add this null hypothesis to the figure for contrast) (2) What are the data (how summarized prior to the testing); (3) what is the model assumptions?
-(4) how inference is performed}
-
-\todo{These sentences were combined from different places. Could you edit?}
-
-Many investigations perform differential abundance analysis of PTMs using two-sample $t$-test or its extensions.
-While simple, these approaches do not fully account for the sources of variations, and are not directly applicable to experiments with complex designs, e.g., comparisons of multiple conditions, acquisition in multiple batches, etc.
+\noindent {\bf Two-sample $t$-test} Many investigations perform differential abundance analysis of PTMs using two-sample $t$-test or its extensions.
+While simple, these approaches do not fully account for all sources of variations, and are not directly applicable to experiments with complex designs.
+These include experiments with comparisons of multiple conditions, acquisition in multiple batches, and unbalanced designs.
 Additionally, while these approaches can be applied to experiments studying PTMs, there is not a self contained, straightforward implementation of the methods, making application challenging.
 
-The approach generally takes as input intensities of individual features from modified peptides, or intensity ratios of modified and unmodified peptide features, and compares the mean abundance of a PTM site from one condition to another~\cite{Schwammle2015}\cite{THOMAS2020}.
-Two-sample $t$-test is based on the null hypothesis that there is no difference in mean PTM abundance between Conditions $1$ and $2$.
+When modeling the simple experiment seen in Figure \ref{fig:data-structure}, the same Null Hypothesis is addressed as seen in Equation \ref{eq:null_hyp}.
+To estimate the parameters of interest in the trivial model, the approach takes as input intensities of individual features from modified peptides, or intensity ratios of modified and unmodified peptide features, and compares the mean abundance of a PTM site from one condition to another~\cite{Schwammle2015}\cite{THOMAS2020}.
 The abundance in each run is taken as input and is often estimated by sum of peak intensities.
-For example, the log-abundance estimate for the PTM in Run $j$ of Condition $1$ is given by
+For example, the log-abundance estimate for the PTM in Run $j$ of Condition $i$ is given by Equation \ref{eq:feature_estimation}.
 
@@
 \begin{aligned}
-\log \left( \sum_{k=1}^{K} 2^{y_{1jk}^{\ast}} \right).
+\log \left( \sum_{k=1}^{K} 2^{y_{ijk}^{\ast}} \right).
 \end{aligned}
@@
 
-For adjustment with respect to unmodified peptides, the estimate of PTM abundance is divided by the protein abundance estimate.
-%which is equivalent to take the difference of their log abundances. 
-For condition $1$ the quantity is denoted by $d_{1j}$ and is given by
+For adjustment with respect to unmodified peptides, the estimate of PTM abundance is divided by the protein abundance estimate, which is equivalent to take the difference of their log abundances.
+For condition $i$ and biological replicate $j$, the quantity is denoted by $\hat{u}^{\ast}_{ij}$ and is given by Equation \ref{eq:adj_estimation}.
 
@@
 \begin{aligned}
-d_{1j}^{\ast} = \log \left( \sum_{k=1}^{K} 2^{y_{1jk}^{\ast}} \right) - \log \left( \sum_{l=1}^{L} 2^{y_{1jl}} \right).
+\hat{u}^{\ast}_{ij} = \log \left( \sum_{k=1}^{K} 2^{y_{ijk}^{\ast}} \right) - \log \left( \sum_{l=1}^{L} 2^{y_{ijl}} \right).
 \end{aligned}
@@
 
-$t$-test then takes the adjusted PTM abundance of condition $1$, $d_{1j}$, and condition $2$, $d_{2j}$, for each run in $J$ as input to estimate the quantity of interest $\Delta$, as well as the standard error, $SE(\Delta)$.
-This results in the Null and Alternative Hypotheses show in Equation \ref{eq:ttest_null_hyp}, where $\hat{d}_{1+}^{\ast}$ represents all $J$ runs for condition $1$ and is the estimate of $(\mu_{1}^{\ast} - \mu_{1})$ seen in Equation \ref{eq:null_hyp}.
-
-\begin{equation}
-\begin{aligned}
-H_{0}: \Delta = d_{1+}^{\ast} - d_{2+}^{\ast} = 0 \\
-H_{a}: \Delta = d_{1+}^{\ast} - d_{2+}^{\ast} \neq 0
-\end{aligned}
-\label{eq:ttest_null_hyp}
-\end{equation}
-
-%Alternatively, run-level summarization can also be used. The difference between the means of PTM abundance in Conditions $i$ and $i'$ are estimated as
-%\[
-%\hat{\Delta} = \frac{1}{J} d_{i+}^{\ast} - \frac{1}{J} d_{i'+},
-%\]
-%where $d_{i+}^{\ast} = \sum_{j=1}^{J} d_{ij}^{\ast}$ and the test statistic for the $t$-test is given by $\hat{\Delta} / \mathrm{SE}(\hat{\Delta})$.
-The statistical significance of the difference is determined by comparing the test statistic against the $t$ distribution, with degrees of freedom $df=2J-2$ in balanced designs.
+$t$-test then takes as input the adjusted PTM abundance of each condition and biological replicate, $\hat{u}^{\ast}_{11}$, $\hat{u}^{\ast}_{12}$, $\hat{u}^{\ast}_{21}$, and $\hat{u}^{\ast}_{22}$, to estimate the quantity of interest $\hat{\Delta}$ as well as the standard error, $\hat{SE}(\hat{\Delta})$.
+$\hat{\Delta}$ is the estimate of the difference for the Null and Alternative Hypotheses show in Equation \ref{eq:null_hyp}.
+The statistical significance is determined by comparing the test statistic against the $t$ distribution, with degrees of freedom $df=2J-2$ in balanced designs.
 
 \medskip \noindent {\bf Limma} Modifications of $t$-test such as moderated $t$-test with Limma were also proposed~\cite{Ritchie_15a}\cite{Zhu}.
-Limma uses linear models to test the null hypothesis that there is no difference in mean PTM abundances between conditions.
+The Null Hypothesis we are interested in answering is the same as in Equation \ref{eq:null_hyp}.
+To estimate $\hat{\Delta}$ in Figure \ref{fig:data-structure}, feature level summarization and unmodified protein adjustment are performed in the same way as $t$-test, shown in Equations \ref{eq:feature_estimation} and \ref{eq:adj_estimation}.
+Limma then fits a linear model using the same adjusted summarized features as in $t$-test, $\hat{u}^{\ast}_{11}$, $\hat{u}^{\ast}_{12}$, $\hat{u}^{\ast}_{21}$ and $\hat{u}^{\ast}_{22}$, to estimate $\hat{\Delta}$.
+To estimate $\hat{SE}(\hat{\Delta})$, Limma leverages Empirical Bayes moderation to moderate the model variance of the individual PTM using the shared global variance of all PTMs modeled in the experiment.
+For a single PTM $m$ with an estimated fold change $\hat{\Delta}_m$, the estimated variance is calculated as $s^{adj}_{m} = P(s_{m}| s_\Theta)$ where $s_{m}$ is the variance resulting from the linear model of PTM $m$ and $s_\Theta$ is the global variance across all PTMs.
+The statistical significance is determined using $\hat{\Delta}_m$ and $\hat{SE}(\hat{\Delta})$ calculated using the adjusted variance $s^{adj}_{m}$.
+
+%Limma uses linear models to test the null hypothesis that there is no difference in mean PTM abundances between conditions.
 It leverages Empirical Bayes moderation to share pooled variance information across individual modification models and moderate the individual residual variances.
 Using a linear model allows Limma to share variance information across conditions, providing a more accurate estimate~\cite{Ritchie_15a}.
-
-With respect to PTM analysis, feature level summarization and unmodified protein adjustment are performed in the same way as $t$-test, shown in Equations \ref{eq:feature_estimation} and \ref{eq:adj_estimation}.
-Limma fits a linear model using the adjusted summarized features, $d_{1+}^{\ast}$ and $d_{2+}^{\ast}$, to estimate the difference between condition $1$ and $2$, $\Delta$.
+With respect to PTM analysis,  Limma fits a linear model using the adjusted summarized features, $d_{1+}^{\ast}$ and $d_{2+}^{\ast}$, to estimate , $\Delta$.
 This results in the same hypothesis as seen in Equation \ref{eq:ttest_null_hyp}.
 The variance for the estimation is moderated using information from the variance across all PTMs modeled.
-For the estimate of a single PTM $m$, $\Delta_m$, the variance is estimated as $s_{\Delta_m} = P(v_{\Delta_m}| v_\Theta)$ where $v_{\Delta_m}$ is the resulting variance when modeling PTM $m$ and $v_\Theta$ is the global variance across all PTMs.
-
-\todo{Could you fine-tune, using the same order as above for Limma?
-I.e., null hypothesis, the data, the model, the inference.
-I think the text can be made more compact, and a bit more technical, by directly referring to either Fig 1 or the equations above}
+For the estimate of a single PTM $m$, $\Delta_m$, the variance is estimated as
 
@@
 The proposed approach leverages the large amount of work done to create $MSstats$~\cite{Choi:2014}.
-$MSstats$ can analyze the PTM without accounting for convolution with the unmodified protein.
-In terms of Figure \ref{fig:data-structure}, this means we can either model the modified or unmodified features, but we cannot combine them as seen in part Figure \ref{fig:data-structure}(c).
-The only Null and Alternative Hypotheses we can address are those in Equation \ref{eq:conv_null_hyp}, without adjusting for unmodified protein abundance.
+$MSstats$ was designed for protein significance analysis and statistical relative quantification of proteins and peptides in global, targeted and data-independent proteomics.
+In terms of Figure \ref{fig:data-structure}, $MSstats$ was designed to estimate the difference in the unmodified protein in condition $1$ and condition $2$, denoted by $u_1$ and $u_2$ respectively.
+In other words it allows us to answer the new Null Hypothesis seen in Equation \ref{eq:msstats_null_hyp}.
+
+\begin{equation}
+\begin{aligned}
+H_{0}: \Delta = \mu_{1} - \mu_{2} = 0 \\
+H_{a}: \Delta = \mu_{1} - \mu_{2} \neq 0
+\end{aligned}
+\label{eq:msstats_null_hyp}
+\end{equation}
+
+The package can also be adjusted to analyze information on the peptide level, allowing us to estimate the convoluted Null Hypothesis in Equation \ref{eq:conv_null_hyp}.
+However, $MSstats$ currently includes no built in method to estimate the the deconvoluted Null Hypothesis in Equation \ref{eq:null_hyp}.
 
@@
 
-\todo{Could you also refer to Figure 1?
-Essentially, MSstats only analyzes a part of the figure.
-Refer to the null hypothesis.
-Then describe the summarization (including AFT etc), and the model for he summarized data (the fact that different designs require different models - this is your argument for the limitation of the methods above.)}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-We propose a statistical method for detecting changes in PTMs.
-The approach removes confounding with the unmodified protein, while being applicable to experiments with complex designs.
-For simplicity, we discuss the method in the special case of a balanced design with multiple conditions and one technical replicate.
-However, in  practice the experimental design can be unbalanced and variation can come from multiple sources, as shown in the next section.
-
-The proposed approach first applies the methods used in $MSstats$ to both the PTM and unmodified protein separately.
-This includes run-level feature summarization using Tukey's Median Polish and fitting a model based on the family of linear mixed-effects models.
-After a model is fit for the both the PTM and unmodified protein, the modeling results are then combined to estimate the quantity of interest.
-
-The null hypothesis states that there is no difference in log-abundance of the PTM site between conditions, adjusted with respect to protein abundance.
-Relating to the example in Figure \ref{fig:data-structure}, the expected value of log-abundance of the PTM and unmodified protein in Condition $1$ are denoted by $\mu_{1}^{\ast}$ and $\mu_{1}$ respectively.
-The feature and run level summarization for Condition $1$ is denoted as $\hat{y}_{1+}^{\ast}$ for the modified peptide and $\hat{y}_{1+}$ for the unmodified peptide.
-The Null and Alternative Hypotheses for Conditions $1$ and $2$ are shown in Equation \ref{eq:prop_null_hyp}.
-This is in contrast to the hypotheses show in Equation \ref{eq:null_hyp}, where the PTM was adjusted before modeling.
-
-\begin{equation}
-\begin{aligned}
-H_{0}: \Delta = ( \mu_{1}^{\ast} - \mu_{2}^{\ast} ) - ( \mu_{1} - \mu_{2} ) = 0 \\
-H_{a}: \Delta = ( \mu_{1}^{\ast} - \mu_{2}^{\ast} ) - ( \mu_{1} - \mu_{2} ) \neq 0
-\end{aligned}
-\label{eq:prop_null_hyp}
-\end{equation}
-
-Specifically, the adjusted difference is given by the difference in log-abundance of the PTM site, subtracted by the difference in log-abundance of the underlying protein, which is equivalent to the log of the ratio of PTM abundance difference to protein abundance difference, in a balanced case.
+We propose a statistical method for detecting changes in PTMs that allows us to answer the Null Hypothesis stated in Equation \ref{eq:null_hyp}; there is no difference in log-abundance of the PTM site between conditions, adjusted with respect to unmodified protein abundance.
+It leverages the methods used in $MSstats$ to answer this hypothesis, while being applicable to experiments with complex designs.
+Relating to the example in Figure \ref{fig:data-structure}, the expected value of the log-abundance of the PTM and unmodified protein in Condition $1$ are denoted by $\hat{\mu}_{1}^{\ast}$ and $\hat{\mu}_{1}$ respectively.
+The feature level summarization for Condition $1$ is denoted as $\hat{y}_{1+}^{\ast}$ for the modified peptide and $\hat{y}_{1+}$ for the unmodified peptide.
+
+To calculate $\hat{y}_{1+}^{\ast}$ and $\hat{y}_{1+}$, feature and run level summarization is carried out as in the sub-plot model of MSstats using Tukey's median polish.
+The summarized values for the PTM and unmodified protein are then used to fit two models based on the family of linear mixed-effects models.
+Finally, the adjusted difference, $\hat{\Delta}$, is given by the difference in log-abundance of the PTM site, subtracted by the difference in log-abundance of the underlying protein, which is equivalent to the log of the ratio of PTM abundance difference to protein abundance difference.
 
@@
 Adjustment for multiple comparisons is performed using the Benjamini-Hochberg procedure to control the false discovery rate at a desired level, e.g., $0.05$~\cite{Benjamini:1995}.
-
-\todo{Put it in the context of the figure again, and also in the context of the existing methods in Background - how is it similar to the existing methods, and how different}
 
@@
 The proposed approach can be extended to complex designs, including experiments with additional sources of variation and unbalanced designs.
-In this case our null hypothesis in Equation \ref{eq:prop_null_hyp} is unchanged.
+In this case our null hypothesis in Equation \ref{eq:null_hyp} is unchanged.
 
 Model inference is done in the same way to how MSstats~\cite{Choi:2014} and MSstatsTMT~\cite{Huang:2020} target the global protein.
-In cases where the design is unbalanced, a linear mixed effects model is fit, which takes into account all potential sources of variation, and restricted maximum likelihood (RML) is used to estimate the parameters of each model.
-Once the parameters are estimated, we can combine the models using modified versions of Equations \ref{eq:prop_fc}, \ref{eq:prop_se}, and \ref{eq:prop_df}.
-
-The log-fold change in the adjusted PTM abundance, $\Delta$, is now estimated by RML, and Equation \ref{eq:prop_fc} becomes 
+First, run and feature level summarization is performed separately for both the PTM and unmodified protein using Tukey's Median Polish.
+Then a separate linear mixed effects model is fit to both the PTM and unmodified protein.
+The model takes into account all potential sources of variation, and restricted maximum likelihood (RML) is used to estimate the parameters of each model.
+Finally, once the parameters are estimated, we can combine the models using modified versions of Equations \ref{eq:prop_fc}, \ref{eq:prop_se}, and \ref{eq:prop_df}.
+
+The log-fold change in the adjusted PTM abundance, $\hat{\Delta
+}$, is now estimated by RML, and Equation \ref{eq:prop_fc} becomes 
 \begin{equation}
@@
 
-The performance of the proposed method was evaluated on simulated and spike in datasets with known ground truth, as well as biological experimental data where the ground truth was not known.
+The performance of the proposed method was evaluated on simulated and spike-in datasets with known ground truth, as well as biological experimental data where the ground truth was not known.
 For the experiments where ground truth was known, we calculated the number of true positives (TP), false positives (FP), true negatives (TN), and false negatives (FN).
@@
 
-\section{Data Availability}
-
-\todo{This overlaps with the Datasets section.
-I would merge all of the below to the experimental data section} All datasets for the four experiments and two simulations, including raw files and analysis results, are available online.
-
-The details of the four experiments, including raw data, and intermediate data processing outputs are located in individual MassIVE repositories, referenced in \ref{tab:dataDescription}.
-
-The R scripts and results of the statistical analysis for the four experiments are available in MassIVE.quant reanalysis containers: SpikeIn benchmark - Ubiquitination - Label-free (MassIVE reanalysis container identifier: \href{https://massive.ucsd.edu/ProteoSAFe/private-dataset.jsp?task=c4c583ecf7f941cdac87f7a4f872517b}{TBD}), Human - Ubiquitination - 1mix-TMT (MassIVE reanalysis container identifier: \href{https://massive.ucsd.edu/ProteoSAFe/reanalysis_container.jsp?task=b639fa5d0ff24ce78abdca13f4b18c0a}{RMSV000000356}), Mouse - Phosphorylation - 2mix-TMT time series (MassIVE reanalysis container identifier: \href{https://massive.ucsd.edu/ProteoSAFe/reanalysis_container.jsp?task=79406d32c1f9413daf7756e6eeb24968}{RMSV000000357}), Human - Ubiquitination - Label-free no global profiling run (MassIVE reanalysis container identifier: \href{https://massive.ucsd.edu/ProteoSAFe/reanalysis_container.jsp?task=f9e6117e00794ce5b2e8802601be16bc}{RMSV000000358}).
-
-The code, data, and analysis results of the two simulations, Computer simulation 1 - Label-free and Computer simulation 2 - Label-free missing values and low features, are available on Github: \href{https://github.com/devonjkohler/MSstatsPTM_simulations}{MSstatsPTM\_simulations}.
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \section{Figures}
+
+\begin{figure}[ht]
+\centering
+\includegraphics[scale=.5]{images/benchmark_fig.png}
+\caption{An overview of the experimental design for Dataset 3 : SpikeIn benchmark - Ubiquitination - Label-free.
+(a) The corresponding levels of spike-in peptide, human lysate, and background lysate are shown.
+The spike-in peptides are adjusted by the human lysate.
+For example, the spike-in abundance of mixture 1 is entirely accounted for by human lysate.
+The background lysate serves as our null model.
+(b) The expected fold change of the spike-in peptides, after adjusting for changes in the unmodified peptide (human lysate).
+The expected fold change is calculated using the values of the spike-in and human lysate in part (a).}
+\label{fig:benchmark-design}
+\end{figure}
 
@@
 Goals of PTM characterization, input to statistical analyses, and notation.
-(a) Schematic data representation, in a simplified case of two conditions and two \todo{biological?} replicate runs.
+(a) Schematic data representation, in a simplified case of two conditions and two biological replicates.
 Each PTM site is modeled and characterized separately, where a PTM is quantified with observed spectral features (boxes), distinguished by different charge states of a peptide.
@@
 \end{figure}
-
-%Data structure of a typical PTM experiment and goals of PTM characterization.
-(a) Schematic data representation, in a simplified case of two conditions and two replicate runs.
-Each PTM site is modeled and characterized separately, where a PTM is quantified with multiple spectral features (boxes), distinguished by different charge states of a peptide.
-The feature intensities are viewed as repeated measurements of the underlying abundance of the PTM, where the abundance in Condition i is denoted by i*.
-Features corresponding to unmodified peptides are considered together to perform adjustment with respect to protein abundance, where the protein abundance in Condition i is denoted by i.
-Peptides can be fully cleaved (solid lines) and/or partially cleaved (dashed lines).
-Some spectral features can be missing.
-(b) PTM relative quantification by statistical inference, which makes use of the feature intensities to infer the underlying PTM abundance and protein abundance with an estimate of associated uncertainty.
-(c) Model-based testing for differential PTM abundance, which corrects for the underlying protein abundance with a cost of increased uncertainty about the estimate of difference between conditions.
 
@@
 After adjustment, this systemic difference was removed, however the inner quartile range of the Limma and $t$-test models was wider than the proposed method.
-b) Before adjustment the spike in peptides (colored red) did not follow the expected log fold change; after adjustment, the spike in peptides were more in line with expectation.
-Using Limma, the spike in peptides followed the expected log fold change after adjustment, however the majority of spike in peptides did not have a significant adjusted p-value.}
+b) Before adjustment the spike-in peptides (colored red) did not follow the expected log fold change; after adjustment, the spike-in peptides were more in line with expectation.
+Using Limma, the spike-in peptides followed the expected log fold change after adjustment, however the majority of spike-in peptides did not have a significant adjusted p-value.}
 \label{fig:spikein_volcano}

### 2022-04-03 supplementary/ptm_sm.tex
updated main background sec

@@
 
-A custom designed experiment with labeling was used to assess the performance of the proposed method in a real experimental setting.
+%A custom designed experiment with labeling was used to assess the performance of the proposed method in a real experimental setting.
 Heavy-labeled KGG modified peptides were used as spike-in peptides.
@@
 
-\begin{figure}[h!]
-\centering
-\includegraphics[width=.8\textwidth]{sim_new/SpikeIn_Overview}
-\caption{An overview of the spike-in benchmark experimental design.
+%\begin{figure}[h!]
+%\centering
+%\includegraphics[width=.8\textwidth]{sim_new/SpikeIn_Overview}
+%\caption{An overview of the spike-in benchmark experimental design.
 The corresponding levels of spike-in peptide, human lysate, and background lysate for each mixture are shown.
@@
 \label{fig:spikein_overview}}
-\end{figure}
+%\end{figure}
 

### 2022-04-06 main/MSstatsPTM_Main.tex
updated notation

@@
 Additionally, some spectral features can be missing.
-The features in each mass spectrometry run are generally summarized and represent a single observation used to estimate the relative abundance of the PTM in condition $i$, $u_i^{\ast}$.
-
-The population parameter we are interested in estimating is the difference between the PTM in condition $1$ and condition $2$, denoted by $u_1^{\ast}$ and $u_2^{\ast}$ respectively.
+The features in each mass spectrometry run are generally summarized and represent a single observation used to estimate the relative abundance of the PTM in condition $i$, $\mu_i^{\ast}$.
+
+The population parameter we are interested in estimating is the difference between the PTM in condition $1$ and condition $2$, denoted by $\mu_1^{\ast}$ and $\mu_2^{\ast}$ respectively.
 To calculate the quantity of interest, statistical analysis needs to summarize values in the table in Figure \ref{fig:data-structure}(a) using appropriate models.
-For condition $i$, the models provide estimates for the mean abundance, $\hat{u}_i^{\ast}$, and standard error, $\hat{SE}(\hat{u}_i^{\ast})$.
+For condition $i$, the models provide estimates for the mean abundance, $\hat{\mu}_i^{\ast}$, and standard error, $\widehat{SE(\mu_i^{\ast})}$.
 Once summarized, we can formulate the Null and Alternative Hypotheses as seen in Equation \ref{eq:conv_null_hyp}.
@@
 To account for this, we must estimate the relative abundance of the unmodified protein in conditions $1$ and $2$, denoted by $\mu_1$ and $\mu_2$.
-We estimate these quantities in the same way as for the PTM, using their correspond features and observations and resulting in an estimated mean abundance, $\hat{u}_i$, and standard error, $\hat{SE}(\hat{u}_i)$.
+We estimate these quantities in the same way as for the PTM, using their correspond features and observations and resulting in an estimated mean abundance, $\hat{\mu}_i$, and standard error, $\widehat{SE(\hat{\mu}_i)}$.
 These estimates are used to adjust the quantities estimated for the PTM, resulting in the new Null and Alternative Hypotheses seen in Equation \ref{eq:null_hyp}.
@@
 For adjustment with respect to unmodified peptides, the estimate of PTM abundance is divided by the protein abundance estimate, which is equivalent to take the difference of their log abundances.
-For condition $i$ and biological replicate $j$, the quantity is denoted by $\hat{u}^{\ast}_{ij}$ and is given by Equation \ref{eq:adj_estimation}.
+For condition $i$ and biological replicate $j$, the estimated quantity is denoted by $\hat{u}^{\ast}_{ij}$ and is given by Equation \ref{eq:adj_estimation}.
 
@@
 
-$t$-test then takes as input the adjusted PTM abundance of each condition and biological replicate, $\hat{u}^{\ast}_{11}$, $\hat{u}^{\ast}_{12}$, $\hat{u}^{\ast}_{21}$, and $\hat{u}^{\ast}_{22}$, to estimate the quantity of interest $\hat{\Delta}$ as well as the standard error, $\hat{SE}(\hat{\Delta})$.
+$t$-test then takes as input the adjusted PTM abundance of each condition and biological replicate, $\hat{u}^{\ast}_{11}$, $\hat{u}^{\ast}_{12}$, $\hat{u}^{\ast}_{21}$, and $\hat{u}^{\ast}_{22}$, to estimate the quantity of interest $\hat{\Delta}$ as well as the standard error, $\widehat{SE(\hat{\Delta})}$.
 $\hat{\Delta}$ is the estimate of the difference for the Null and Alternative Hypotheses show in Equation \ref{eq:null_hyp}.
@@
 Limma then fits a linear model using the same adjusted summarized features as in $t$-test, $\hat{u}^{\ast}_{11}$, $\hat{u}^{\ast}_{12}$, $\hat{u}^{\ast}_{21}$ and $\hat{u}^{\ast}_{22}$, to estimate $\hat{\Delta}$.
-To estimate $\hat{SE}(\hat{\Delta})$, Limma leverages Empirical Bayes moderation to moderate the model variance of the individual PTM using the shared global variance of all PTMs modeled in the experiment.
+To estimate $\widehat{SE(\hat{\Delta})}$, Limma leverages Empirical Bayes moderation to moderate the model variance of the individual PTM using the shared global variance of all PTMs modeled in the experiment.
 For a single PTM $m$ with an estimated fold change $\hat{\Delta}_m$, the estimated variance is calculated as $s^{adj}_{m} = P(s_{m}| s_\Theta)$ where $s_{m}$ is the variance resulting from the linear model of PTM $m$ and $s_\Theta$ is the global variance across all PTMs.
-The statistical significance is determined using $\hat{\Delta}_m$ and $\hat{SE}(\hat{\Delta})$ calculated using the adjusted variance $s^{adj}_{m}$.
+The statistical significance is determined using $\hat{\Delta}_m$ and $\widehat{SE(\hat{\Delta})}$ calculated using the adjusted variance $s^{adj}_{m}$.
 
@@
 $MSstats$ was designed for protein significance analysis and statistical relative quantification of proteins and peptides in global, targeted and data-independent proteomics.
-In terms of Figure \ref{fig:data-structure}, $MSstats$ was designed to estimate the difference in the unmodified protein in condition $1$ and condition $2$, denoted by $u_1$ and $u_2$ respectively.
+In terms of Figure \ref{fig:data-structure}, $MSstats$ was designed to estimate the difference in the unmodified protein in condition $1$ and condition $2$, denoted by $\mu_1$ and $\mu_2$ respectively.
 In other words it allows us to answer the new Null Hypothesis seen in Equation \ref{eq:msstats_null_hyp}.
@@
 Relating to the example in Figure \ref{fig:data-structure}, the expected value of the log-abundance of the PTM and unmodified protein in Condition $1$ are denoted by $\hat{\mu}_{1}^{\ast}$ and $\hat{\mu}_{1}$ respectively.
-The feature level summarization for Condition $1$ is denoted as $\hat{y}_{1+}^{\ast}$ for the modified peptide and $\hat{y}_{1+}$ for the unmodified peptide.
-
-To calculate $\hat{y}_{1+}^{\ast}$ and $\hat{y}_{1+}$, feature and run level summarization is carried out as in the sub-plot model of MSstats using Tukey's median polish.
+The feature and run level summarization for Condition $1$ is denoted as $\hat{y}_{1++}^{\ast}$ for the modified peptide and $\hat{y}_{1++}$ for the unmodified peptide.
+
+To calculate $\hat{y}_{1++}^{\ast}$ and $\hat{y}_{1++}$, feature and run level summarization is carried out as in the sub-plot model of MSstats using Tukey's median polish.
 The summarized values for the PTM and unmodified protein are then used to fit two models based on the family of linear mixed-effects models.
@@
 \begin{equation}
-\hat{\Delta} = \left[ \frac{1}{J} \left( \hat{y}_{1+}^{\ast} - \hat{y}_{2+}^{\ast} \right) \right] - \left[ \frac{1}{J} \left( \hat{y}_{1+} - \hat{y}_{2+} \right) \right]
+\hat{\Delta} = \left[ \frac{1}{J} \left( \hat{y}_{1++}^{\ast} - \hat{y}_{2++}^{\ast} \right) \right] - \left[ \frac{1}{J} \left( \hat{y}_{1++} - \hat{y}_{2++} \right) \right]
 \label{eq:prop_fc}
@@
 \begin{equation}
-SE(\hat{\Delta}) = \left[ \left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right) \right]^{1/2}
+\widehat{SE(\Delta)} = \left[ \left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right) \right]^{1/2}
 \label{eq:prop_se}
@@
 The package can optionally impute missing values using an AFT model.
+Missing value imputation should be performed with care when applied to PTM data.
+PTMs generally exhibit low feature counts and may be missing due to reasons other than low abundance.
+These issues can cause the missing value imputation to be unstable.
 

### 2022-05-19 main/MSstatsPTM_Main.tex
changes to spike in sections

@@
 
-\todo{This section would benefit from more biology/technology background, and more extensive references}
-Signaling mechanisms allow cells to mount a fast and dynamic response to a multitude of bimolecular events.
-Signaling is facilitated by the modification of proteins at specific residues, acting as molecular on/off switches~\cite{Deribe,Cohen}.
+Signaling mechanisms allow cells to mount a fast and dynamic response to a multitude of biomolecular events.
+Signaling is facilitated by the modification of proteins at specific residues, acting as molecular on/off switches~\cite{Deribe,Cohen,Bludau2022}.
 Localizing the modification sites, or characterizing relative abundance of a modification site's occupancy repertoire across experimental conditions provides important insights~\cite{Mann}.
 For example, meaningful patterns of changes in post-translational modifications (PTMs) abundance can serve as biomarkers of a disease~\cite{Petushkova_2017}.
-Alternatively, distinguishing the quantitative changes in a PTM from the overall changes of the protein abundance helps gain insight into biological and physiological processes operating on a very short timescale \todo{would be good to have a reference}.
+Alternatively, distinguishing the quantitative changes in a PTM from the overall changes of the protein abundance helps gain insight into biological and physiological processes operating on a very short timescale~\cite{Chandramouli:2009}.
 
@@
 
-The technological challenges in PTM identification and quantification challenge the downstream statistical analyses.
-Yet frequently data from these experiments are analyzed using statistical methods that were not originally designed for this task.
+The technological difficulties in PTM identification and quantification challenge the downstream statistical analyses.
+Frequently data from these experiments are analyzed using statistical methods that were not originally designed for this task.
 In particular, methods such as $t$-test, Analysis of Variance, or Limma take as input the intensity ratios of modified and unmodified peptide features, and compare the mean abundance of different PTM sites~\cite{Ritchie_15a}.
@@
 
-The datasets in this manuscript  benchmarked the proposed approach in situations with known ground truth, and represented a variety of experimental designs and acquisition methods.
+The datasets in this manuscript both benchmarked the proposed approach in situations with known ground truth and represented a variety of experimental designs and acquisition methods.
 Two computer simulations varied in experimental realism.
@@
 The spike-in experiment took the real world characteristics a step further and allowed us to compare the methods in a real-world experimental settings with known changes in modified spike-in peptides.
-Finally, three biological experiments demonstrated the applicability of the proposed approach across different biological organisms, PTM types, experimental designs and acquisition strategies.
+Finally, three biological experiments demonstrated the applicability of the proposed approach across different biological organisms, modifications, experimental designs and acquisition strategies.
 Table \ref{tab:dataDescription} summarizes the experiments.
@@
 In each dataset, 1000 proteins had 10 unmodified features per protein.
-Each of the 1000 proteins had one PTM, each PTM represented by 10 modified features.
+Each of the 1000 proteins had one PTM; each PTM was represented by 10 modified features.
 The PTMs of 500 proteins had a differential fold change between conditions, while the other 500 proteins were generated with no difference between conditions.
@@
 \label{sec:exp_proc_dataset3}
-
-\todo{Devon: I moved the figure to main and tried to adjust this section to be clearer. I still think this section could benefit from Erik and Tsung-Heng's insight. Will ask them to look in email.}
 
@@
 coli} lysate was used to normalize total protein levels prior to enrichment or global protein profiling.
-The background lysate was treated as control and is expected to show no fold change in all comparisons except mix1-mix2 and mix3-mix4.
+The background lysate is expected to change in the same way as the corresponding protein.
+Thus it is treated as control and is expected to show no fold change in all comparisons after adjusting for changes in the unmodified protein abundance.
 The true log fold changes of spike-in modified peptides in each comparison can be seen in Figure \ref{fig:benchmark-design}(b).
@@
 %\hline
-% Comparison & Expected log fold change\\ [0.5ex]
+% Comparison & Expected log fold change\\ %[0.5ex]
 % \hline\hline
@@
 %\hline
-% mix2-mix4 & -1\\
+% mix2-mix4 & 1\\
 %\hline
-% mix3-mix4 & 1\\
+% mix3-mix4 & -1\\
 %\hline
 %\end{tabular}
-%\caption{The expected log fold change of modified spike-in peptides in Dataset 3.}
+%\caption{The expected log fold change of %modified spike-in peptides in Dataset 3.}
 %\label{table:spikein_fold_change}
@@
 Uninfected cells were measured at 0 and 6 hours, while infected cells were measured at 1, 2, 4, and 6 hour increments, resulting in six total conditions.
-\todo{If this is a group comparison, explain why, even though the conditions are times.
-The table above says it is group comparison} The proteome was quantified using isobaric multiplexing via tandem mass tagging (TMT) in combination with one LC-MS/MS run.
+The proteome was quantified using isobaric multiplexing via tandem mass tagging (TMT) in combination with one LC-MS/MS run.
 

### 2022-05-29 main/MSstatsPTM_Main.tex
Small - Fig 4

@@
 \caption{Dataset 1 \& 2: Computer simulation.
-a) All the considered methods in the first computer simulation correctly calibrated FDR when adjusting for changes in protein abundance.
+\todo{OV: in (a) could you use the same color scheme for the methods as in the remaining figures?
+Use different fill/opacity/shade of a same color for methods with and without the adjustment} a) All the considered methods in the first computer simulation correctly calibrated FDR when adjusting for changes in protein abundance.
 In comparison, the methods without accounting for the protein-level changes resulted in off-target, high false positive rates.

### 2022-05-31 main/MSstatsPTM_Main.tex
Small

@@
 \label{sec:exp_proc_dataset5}
-In this experiment 8 biological replicates were allocated to 2 TMT mixtures in an unbalanced time course design. More details are in Supplementary Sec. \ref{supp-sec:shigella}.
+\todo{For each experimental dataset, could you say how many proteins were quantified, and the average number of {SM per protein (modified/unmodified)?
+I was looking everywhere and could not find the info} In this experiment 8 biological replicates were allocated to 2 TMT mixtures in an unbalanced time course design.
+More details are in Supplementary Sec.
+\ref{supp-sec:shigella}.
 
@@
 \caption{Dataset 5 : Mouse - Phosphorylation - 2mix-TMT time series.
-a) The overlap of differentially modified peptides between the PTM model with and without global protein level adjustment.
+\todo{OV: What are the vertical panels 1 and 2 in (b)??
+Are they bio reps?
+Could you clarify?
+I would remove the work 'BioReplicate' at the bottom, it is not clear what it refers to} a) The overlap of differentially modified peptides between the PTM model with and without global protein level adjustment \todo{OV: in which pairwise comparison?
+Across all the comparisons?}.
 Again more PTMs became insignificant after adjustment then became significant.

### 2022-06-03 main/MSstatsPTM_Main.tex
Intro and Data

@@
 \newenvironment{todolong}{\color{red}[TODO:}{]}
+\def\ov#1{{\color{red}#1}}
 \def\note#1{{\color{OliveGreen}[NOTE: #1]}}
@@
 
-\title{MSstatsPTM statistical relative quantification of post-translational modifications in bottom-up proteomic experiments}
+\title{MSstatsPTM: Statistical relative quantification of post-translational modifications in bottom-up mass spectrometry-based proteomics}
 
@@
 
-Liquid chromatography coupled with mass spectrometry (LC-MS/MS)-based proteomics is increasingly used to detect changes in post-translational modifications (PTMs) between samples from various conditions.
+Liquid chromatography coupled with bottom up mass spectrometry (LC-MS/MS)-based proteomics is increasingly used to detect changes in post-translational modifications (PTMs) in samples conditions.
 Analysis of data from such experiments faces numerous statistical challenges.
-These include the low abundance of modified proteoforms, the small number of representative peptides that span modification sites, and confounding between changes in the abundance of PTM and the overall changes in the protein abundance.
+These include the low abundance of modified proteoforms, the small number of observed peptides that span modification sites, and confounding between changes in the abundance of PTM and the overall changes in the protein abundance.
 Therefore, statistical approaches for detecting differential PTM abundance must integrate all the available information pertaining to a PTM site, and consider all the relevant sources of confounding and variation.
 In this manuscript we propose such a statistical framework, which is versatile, accurate, and leads to reproducible results.
-The framework requires an experimental design, which for each sample quantifies both the peptides with post-translational modifications, and peptides from the same proteins with no modification sites.
-The framework supports both label-free and tandem mass tag (TMT)-based LC-MS/MS acquisitions.
-The statistical methodology separately summarizes the abundances of peptides with and without the modification sites, and fits separate linear mixed effects models that reflect the biological and technological aspects of the experimental design.
+The framework requires an experimental design, which quantifies, for each sample, both peptides with post-translational modifications and peptides from the same proteins with no modification sites.
+The proposed framework supports both label-free and tandem mass tag (TMT)-based LC-MS/MS acquisitions.
+The statistical methodology separately summarizes the abundances of peptides with and without the modification sites, by fitting separate linear mixed effects models appropriate for the experimental design.
 Next, model-based inferences regarding the PTM and the protein-level abundances are combined to account for the confounding between these two sources.
-Evaluations on computer simulations, a spike-in experiment with known ground truth, and three biological experiments demonstrate the improved fold change estimation and detection of differential PTM abundance, as compared to currently used approaches.
+Evaluations on computer simulations, a spike-in experiment with known ground truth, and three biological experiments \ov{with different organisms, modification types and data acquisition types} demonstrate the improved fold change estimation and detection of differential PTM abundance, as compared to currently used approaches.
 The proposed framework is implemented in the free and open-source R/Bioconductor package $MSstatsPTM$.
@@
 Signaling is facilitated by the modification of proteins at specific residues, acting as molecular on/off switches~\cite{Deribe,Cohen,Bludau2022}.
-Localizing the modification sites, or characterizing relative abundance of a modification site's occupancy repertoire across experimental conditions provides important insights~\cite{Mann}.
+Characterizing relative abundance of a modification site's occupancy repertoire across experimental conditions provides important insights~\cite{Mann}.
 For example, meaningful patterns of changes in post-translational modifications (PTMs) abundance can serve as biomarkers of a disease~\cite{Petushkova_2017}.
@@
 
-The technological difficulties in PTM identification and quantification challenge the downstream statistical analyses.
+The technological difficulties in PTM identification and quantification increase the uncertainty and the variation in the data, and challenge the downstream statistical analyses.
 Frequently data from these experiments are analyzed using statistical methods that were not originally designed for this task.
-In particular, methods such as $t$-test, Analysis of Variance, or Limma take as input the intensity ratios of modified and unmodified peptide features, and compare the mean abundance of different PTM sites~\cite{Ritchie_15a}.
-Such methods do not fully account for all the sources of uncertainty and variation in the experiment.
-As the result, they are either not directly applicable to experiments with non-trivial designs (such as experiments with multiple conditions, paired and time course designs, and experiments with labeling), or require the analysts to exercise non-trivial statistical expertise.
-
-This manuscript proposes a general statistical analysis framework that detects relative changes in post-translational modifications.
-The framework requires an experimental design, which for each sample quantifies both the peptides with post-translational modifications, and peptides from the same proteins with no modification sites.
+Researchers use methods such as $t$-test \todo{ref}, Analysis of Variance\todo{ref}, or Limma\todo{ref}, by taking as input the intensity ratios of modified and unmodified peptide features, and comparing the mean abundance of different PTM sites~\cite{Ritchie_15a}.
+Such approaches do not fully account for all the sources of uncertainty.
+As the result, these approaches are either not directly applicable to experiments with non-trivial designs (such as experiments with multiple conditions, paired and time course designs, and experiments with labeling), or require the analysts to exercise non-trivial statistical expertise.
+
+This manuscript proposes a versatile statistical analysis framework that accurately detects relative changes in post-translational modifications.
+The framework requires an experimental design, which quantifies, for each sample, both the peptides with post-translational modifications, and peptides from the same proteins with no modification sites.
 The framework supports data-dependent acquisitions (DDA) that are label-free or tandem mass tag (TMT)-based.
@@
 We evaluated the proposed framework on two datasets from computer simulations, one benchmark controlled mixture, and three biological investigations.
-The datasets illustrate a diverse set of organisms, acquisition methods, and experimental designs, showing the applicability of the framework to a variety of situations.
-The evaluations demonstrated that by appropriately leveraging the information from the unmodified peptides, the proposed approach improves the accuracy of the estimates of PTM fold changes and results in a better calibrated false positive rate of detecting differentially abundant PTMs.
-In particular, accounting for the confounding from protein abundance allows us to characterize the true effect of the modification, avoiding the need for more manual and time intensive methods.
-
-The proposed approach is implemented as a freely available open source R package $MSstatsPTM$, available on Bioconductor, which employs similar input format as $MSstats$ and $MSstatsTMT$ \cite{Choi:2014,Huang:2020}.
-
-\section{Experimental Procedures}
-
-The datasets in this manuscript both benchmarked the proposed approach in situations with known ground truth and represented a variety of experimental designs and acquisition methods.
-Two computer simulations varied in experimental realism.
-The first simulation produced a perfectly clean dataset, with many replicates and no missing values.
-The second simulation introduced real-world characteristics, such as limited modified features and missing values.
-The spike-in experiment took the real world characteristics a step further and allowed us to compare the methods in a real-world experimental settings with known changes in modified spike-in peptides.
-Finally, three biological experiments demonstrated the applicability of the proposed approach across different biological organisms, modifications, experimental designs and acquisition strategies.
-Table \ref{tab:dataDescription} summarizes the experiments.
-
-\subsection*{Data Availability}
-
-Details of the experimental data, R scripts with $MSstatsPTM$ analysis, and results of the statistical analysis are available in MassIVE.quant (\url{https://massive.ucsd.edu/ProteoSAFe/static/massive-quant.jsp}) \cite{Choi:2020}.
-Details of computer simulations are available on GitHub (\url{https://github.com/devonjkohler/MSstatsPTM_simulations}).
+The datasets illustrate a diverse set of organisms, modification types, acquisition methods and experimental designs, showing the applicability of the framework to a variety of situations.
+By appropriately leveraging the information from the unmodified peptides, the proposed approach improved the accuracy of the estimates of PTM fold changes and produced a better calibrated false positive rate of detecting differentially abundant PTMs \todo{as compared to methods XXX}.
+In particular, accounting for the confounding from protein abundance allowed us to characterize the true effect of the modification, avoiding the need for more manual and time intensive follow-up investigation.
+
+The proposed approach is implemented as a freely available open source R package $MSstatsPTM$, as part of the $MSstats$ family of packages \cite{Choi:2014,Huang:2020}, and is available on Bioconductor.
+
+\section{Experimental procedures}
 
 %%%% Summary of controlled mixtures
-\begin{table}[h!]
+\begin{table}[t!]
 \centering
@@
 %%%% colnames of table
-& \multirow{2}{*}{Dataset} & Experimental & No. of & No. of Bio. &  Data \\
-&  & Design & Conditions & Replicates & availability \\
+& \multirow{2}{*}{Dataset} & Experimental & No. of & No. of bio. &  Data \\
+&  & design & Conditions & replicates & availability \\
 \hline
@@
 %%% DDA iPRG2015
-& Computer simulation 2 - missing and low features & Group comparison & 2/3/4 & 2/3/5/10 & \href{https://github.com/devonjkohler/MSstatsPTM_simulations}{Github} \\
+& Computer simulation 2 - Missing and low features & Group comparison & 2/3/4 & 2/3/5/10 & \href{https://github.com/devonjkohler/MSstatsPTM_simulations}{Github} \\
  &&&&& \\%[0.05in]
@@
 \end{tiny}
-\caption{ \small {\bf Simulated and experimental datasets in this manuscript} 
-```Dataset'' is the code name of the dataset in this manuscript.
-``Data availability'' shows the ID of the MassIVE.quant repository or the GitHub repository. All the experiments were conducted in data-dependent acquisition (DDA) mode.
+\caption{ \small {\bf Simulated and experimental datasets.} 
+```Dataset'' is the dataset code name.
+``Data availability'' shows the ID of the MassIVE.quant repository or the GitHub repository.
+All the experiments were conducted in data-dependent acquisition (DDA) mode.
+\todo{Could you add to the table the number of proteins, and the average number of modified/unmodified peptides per protein} \todo{If the figures are at the end, I think the table should also be at the end}
+\todo{Could you check the table and the entire section: when we say 'number of biological replicates', we typically mean per condition.
+But it seems that some descriptions imply the total number of biological samples across conditions.
+Could you clarify that we use number of biological replicates per condition throughout?
+(E.g, this helps calculate the degrees of freedom in balanced designs)}
 }
@@
 
-The R scripts and results of the statistical analysis for the four experiments are available in MassIVE.quant reanalysis containers: SpikeIn benchmark - Ubiquitination - Label-free (MassIVE reanalysis container identifier: \href{https://massive.ucsd.edu/ProteoSAFe/private-dataset.jsp?task=c4c583ecf7f941cdac87f7a4f872517b}{TBD}), Human - Ubiquitination - 1mix-TMT (MassIVE reanalysis container identifier: \href{https://massive.ucsd.edu/ProteoSAFe/reanalysis_container.jsp?task=b639fa5d0ff24ce78abdca13f4b18c0a}{RMSV000000356}), Mouse - Phosphorylation - 2mix-TMT time series (MassIVE reanalysis container identifier: \href{https://massive.ucsd.edu/ProteoSAFe/reanalysis_container.jsp?task=79406d32c1f9413daf7756e6eeb24968}{RMSV000000357}), Human - Ubiquitination - Label-free no global profiling run (MassIVE reanalysis container identifier: \href{https://massive.ucsd.edu/ProteoSAFe/reanalysis_container.jsp?task=f9e6117e00794ce5b2e8802601be16bc}{RMSV000000358}).
-
-\subsection*{Dataset 1 : Computer simulation 1 - Label-free}
+\subsection*{Data overview and availability}
+
+Table \ref{tab:dataDescription} summarizes the experiments.
+Two computer simulations had known ground truth, and varied in experimental realism.
+The first simulation produced a perfectly clean dataset, with many replicates and no missing values.
+The second simulation introduced real-world characteristics, such as limited modified features and missing values.
+Details of computer simulations are available on GitHub (\url{https://github.com/devonjkohler/MSstatsPTM_simulations}).
+
+One spike-in experiment also had known changes in modified spike-in peptides, but had real-world experimental characteristics.
+Finally, three biological experiments demonstrated the applicability of the proposed approach across different biological organisms, modifications, experimental designs and acquisition strategies.
+Details of the experimental data, R scripts with $MSstatsPTM$ analysis, and results of the statistical analysis are available in MassIVE.quant (\url{https://massive.ucsd.edu/ProteoSAFe/static/massive-quant.jsp}) \cite{Choi:2020}.
+
+\todo{Additional details on every dataset are available in Supplementary Section 2 (I think it's ok to only mention this once)}
+
+\todo{Could you delete this paragraph, and move the container identifies to a new column in table 1?} The R scripts and results of the statistical analysis for the four experiments are available in MassIVE.quant reanalysis containers: SpikeIn benchmark - Ubiquitination - Label-free (MassIVE reanalysis container identifier: \href{https://massive.ucsd.edu/ProteoSAFe/private-dataset.jsp?task=c4c583ecf7f941cdac87f7a4f872517b}{TBD}), Human - Ubiquitination - 1mix-TMT (MassIVE reanalysis container identifier: \href{https://massive.ucsd.edu/ProteoSAFe/reanalysis_container.jsp?task=b639fa5d0ff24ce78abdca13f4b18c0a}{RMSV000000356}), Mouse - Phosphorylation - 2mix-TMT time series (MassIVE reanalysis container identifier: \href{https://massive.ucsd.edu/ProteoSAFe/reanalysis_container.jsp?task=79406d32c1f9413daf7756e6eeb24968}{RMSV000000357}), Human - Ubiquitination - Label-free no global profiling run (MassIVE reanalysis container identifier: \href{https://massive.ucsd.edu/ProteoSAFe/reanalysis_container.jsp?task=f9e6117e00794ce5b2e8802601be16bc}{RMSV000000358}).
+
+\subsection*{Dataset 1: Computer simulation 1 - Label-free}
 \label{sec:comp_sim_procedure1}
 
-The simulation represented an idealistic case.
-24 synthetic datasets were generated exhibiting scenarios with different experimental designs and biological variation.
+{\bf Simulation design:} The simulation represented an idealistic case.
+24 synthetic label-free datasets were generated with different experimental designs and different biological variation.
 In each dataset, 1000 proteins had 10 unmodified features per protein.
-Each of the 1000 proteins had one PTM; each PTM was represented by 10 modified features.
-The PTMs of 500 proteins had a differential fold change between conditions, while the other 500 proteins were generated with no difference between conditions.
+Each of the 1000 proteins had one PTM.
+Each PTM was represented by 10 modified features.
+The PTMs of 500 proteins had a differential fold change between conditions, while the other 500 proteins were generated with no changes in abundance between conditions.
 Furthermore, the fold changes of half of the 500 differential PTMs were fully masked by changes in the unmodified protein.
 Additionally, the fold change of half the 500 non-differential PTMs was entirely due to changes in the unmodified protein.
-All differential PTMs were generated with an expected log fold change of 0.75 between conditions.
+All the differential PTMs were generated with an expected log fold change of 0.75 between conditions.
 
 Each simulation was generated with random biological variation.
-Specifically, let $Y_{ibc}$ denote the abundance of peptide $i$ in biological replicate $b$ of condition $c$.
-The simulated peptide abundance  $Z_{ibc}$ was generated as $Z_{ibc} = Y_{ibc} + \epsilon_{ibc}$ where $\epsilon_{ibc}\sim^{iid}N(0,\sigma^2)$.
-The simulations were generated with one of two values $\sigma^2 = \{.2, .3\}$, motivated by the biological experiments in this manuscript.
-Further details are in Supplementary Sec. \ref{supp-sec:dataset1}.
-
-\subsection*{Dataset 2 : Computer simulation 2 - Label-free missing values and low features}
+The observed peptide abundances were simulated by adding random noise $\mathcal{N}(0,\sigma^2)$ to the deterministic abundances described above.
+Two values $\sigma^2 = \{.2, .3\}$ were motivated by the experimental datasets in this manuscript.
+
+\medskip \noindent {\bf Evaluation:} We evaluated the ability of the statistical methods \todo{XXX - what and how}
+
+\subsection*{Dataset 2: Computer simulation 2 - Label-free missing values and low features}
 \label{sec:comp_sim_procedure2}
 
-The data were simulated as above, while including missing values and a low number of modified features to provide a more realistic representation of the experiments.
-In this simulation, the feature counts and the proportion of missing values were as observed on average over all the the experimental datasets in this manuscript.
-Specifically, PTMs were simulated with 2 modified peptide features, while unmodified proteins were simulated with 10 features.
+{\bf Simulation design:} The data were simulated as above, while providing a more realistic representation of the experiments.
+The feature counts and the proportion of missing values were as observed on average over all the the experimental datasets in this manuscript.
+Specifically, PTMs were simulated with 2 modified peptide features, and unmodified proteins were simulated with 10 features.
 Additionally, 20\% of observations for both modified and unmodified peptides were missing completely at random.
-Further details are in Supplementary Sec.
-\ref{supp-sec:dataset2}.
-
-\subsection*{Dataset 3 : SpikeIn benchmark - Ubiquitination - Label-free}
+
+\medskip \noindent {\bf Evaluation:} \todo{As above, we evaluated... what and how...}
+
+\subsection*{Dataset 3: SpikeIn benchmark - Ubiquitination - Label-free}
 \label{sec:exp_proc_dataset3}
 
-In the custom spike-in experiment 8 replicates were allocated to 4 conditions, resulting in 2 replicates per condition.
-
-\medskip \noindent {\bf Experimental Design:} 50 heavy-labeled KGG motif peptides from 20 human proteins were used as spike-in peptides.
+\medskip \noindent {\bf Experimental Design:} Figure \ref{fig:benchmark-design}(a) overviews the experimental design.
+Four mixtures (i.e., conditions) were created with varying amounts of human lysate, background {\it E.
+coli} lysate, and human spike-in ub-peptide mixture.
+Unmodified peptides from human lysate were viewed as the global proteome.
+Background {\it E.
+coli} lysate were used to equalize total protein levels prior to enrichment or global protein profiling \ov{between Mix 1 and 3, and between Mix 2 and 4}.
+50 heavy-labeled KGG motif peptides from 20 human proteins were used as spike-in peptides.
 Quantitative changes in protein and site abundance of these 20 proteins were the target of the benchmark.
-An overview of the experimental design is shown in Figure \ref{fig:benchmark-design}.
-Four mixture conditions were created with differing amounts of human lysate, background lysate, and human spike-in ub-peptide mixture.
-Unmodified peptides from human lysate were used to estimate changes in global protein abundance.
-Background {\it E.
-coli} lysate was used to normalize total protein levels prior to enrichment or global protein profiling.
-The background lysate is expected to change in the same way as the corresponding protein.
-Thus it is treated as control and is expected to show no fold change in all comparisons after adjusting for changes in the unmodified protein abundance.
-The true log fold changes of spike-in modified peptides in each comparison can be seen in Figure \ref{fig:benchmark-design}(b).
-Two sets of data were acquired for each mixture in label-free LC-MS/MS mode: KGG enriched + LC-MS/MS, and LC-MS/MS only.
-
-\medskip \noindent {\bf Global Profiling Run:} This experiment included a separate global profiling run to measure unmodified peptides.
+\todo{In particular, we distinguish the changes in the abundances of the modified peptides (i.e., unadjusted changes) and the changes of their abundance relative to the changes in the abundances of the human lysate (i.e., protein-level adjusted changed).
+} The true log-fold changes \ov{between the relevant components of the relevant mixtures} are summarized in Figure \ref{fig:benchmark-design}(b).
+Two replicates mixtures were created per condition.
+
+\medskip \noindent {\bf Data acquisition:} Each mixture was analyzed with KGG enrichment, and without KGG enrichment (i.e., in a global profiling run), with label-free LC-MS/MS.
 There was a 90.2\% overlap between the identified background modified peptides and proteins that were quantified in the global profiling run.
 
-\medskip \noindent {\bf Pairwise Comparisons:} We evaluated the ability of the statistical methods to detect known changes in abundance between pairs of conditions.
-The four mixtures were treated as individual conditions and labeled mix1, mix2, mix3, and mix4.
-The pairwise comparisons were mix1-mix2, mix1-mix3, mix1-mix4, mix2-mix3, mix2-mix4, and mix3-mix4.
-
-%\begin{table}[h!]
-%\centering
-%\begin{tabular}{| c | c |}
-%\hline
-% Comparison & Expected log fold change\\ %[0.5ex]
-% \hline\hline
-% mix1-mix2 & -1\\
-% \hline
-% mix1-mix3 & 1\\
-%\hline
-% mix1-mix4 & 0\\
-%\hline
-% mix2-mix3 & -2\\
-%\hline
-% mix2-mix4 & 1\\
-%\hline
-% mix3-mix4 & -1\\
-%\hline
-%\end{tabular}
-%\caption{The expected log fold change of %modified spike-in peptides in Dataset 3.}
-%\label{table:spikein_fold_change}
-%\end{table}
-
-\subsection*{Dataset 4 : Human - Ubiquitination - 1mix-TMT}
+\medskip \noindent {\bf Evaluation:} \todo{Here is the place to describe: 'we expect XXX to not change'.
+Therefore we evaluated the ability of statistical methods to avoid false positives (i.e., specificity).
+We expect XXX to have fold changes XXX between conditions XXX.
+We evaluated the accuracy of estimation of fold changes...
+and the sensitivity of detecting differentially abundant...}
+%
+\todo{I think some mixtures are not comparable because the total protein amount is not the same.
+I believe only in Mix 1 vs 2 and Mix 3 vs 4 are comparable.
+I do not think it is correct to compare Mix 1 vs 3 and 2 vs 4 directly because they differ in the total amount of the proteome.
+Or was it normalized based on what we know is the same?
+If yes, we need to describe in the beginning of 'Evaluation'}  \todo{The unadjusted proteins...
+estimation...
+The adjusted proteins...
+estimation...
+Positive controls...
+Negative controls...}
+
+\subsection*{Dataset 4: Human - Ubiquitination - 1mix-TMT}
 \label{sec:exp_proc_dataset4}
-In this experiment 11 biological replicates were allocated to 1 TMT mixture in a group comparison unbalanced design. Further details are in Supplementary Sec. \ref{supp-sec:ipah}.
 
 \medskip \noindent {\bf Experimental Design:}  Luchetti et al.
-\cite{LUCHETTI2021} quantified the abundance of total protein and ubiquitination in human epithelial cells.
-Cells were engineered to express IpaH7.8 under a dox inducible promoter and measurements were taken at different time periods.
+\cite{LUCHETTI2021} profiled human epithelial cells engineered to express IpaH7.8 under a dox inducible promoter.
+Uninfected cells were measured at 0 and 6 hours, while cells infected \todo{with what?
+Not clear about the infection} were measured at 1, 2, 4, and 6 hour increments, resulting in six total conditions.
 GSDMD was actively degraded when IpaH7.8 expression was induced by dox treatment.
-Uninfected cells were measured at 0 and 6 hours, while infected cells were measured at 1, 2, 4, and 6 hour increments, resulting in six total conditions.
-The proteome was quantified using isobaric multiplexing via tandem mass tagging (TMT) in combination with one LC-MS/MS run.
-
-\medskip \noindent {\bf Global Profiling Run:} This experiment included a separate global profiling run, with the same design, to measure unmodified peptides.
+\todo{Is GSDMD a known protein that will be used for validation?}
+
+\medskip \noindent {\bf Data acquisition:} 11 samples \todo{why 11 and not 12?
+Which condition didn't have a biological replicate?} were allocated to 1 TMT mixture in an unbalanced group comparison design.
+The ubiquitinated peptides, and the total proteome (i.e., global profiling) were each conducted in a single LC-MS/MS run.
 There was a 95\% overlap between the identified modified peptides and proteins that were quantified in the global profiling run.
 
-\medskip \noindent {\bf Pairwise Comparisons:} We evaluated the ability of the statistical methods to detect changes in the abundance of modified peptides both before and after adjusting for changes in global protein abundance.
+\medskip \noindent {\bf Evaluation:} We evaluated the ability of the statistical methods to detect changes in the abundance of modified peptides both before and after adjusting for changes in global protein abundance.
 The six condition were labeled Dox1hr, Dox2hr, Dox4hr, Dox6hr, NoDox0hr, and NoDox6hr.
@@
 
-\subsection*{Dataset 5 : Mouse - Phosphorylation - 2mix-TMT time series}
+\subsection*{Dataset 5: Mouse - Phosphorylation - 2mix-TMT time series}
 \label{sec:exp_proc_dataset5}
-\todo{For each experimental dataset, could you say how many proteins were quantified, and the average number of {SM per protein (modified/unmodified)?
-I was looking everywhere and could not find the info} In this experiment 8 biological replicates were allocated to 2 TMT mixtures in an unbalanced time course design.
-More details are in Supplementary Sec.
-\ref{supp-sec:shigella}.
 
@@
 flexneri}).
-The experiment quantified the abundance of total protein and of phosphorylation in wild type (WT), and in ATG16L1-deficient (cKO) samples, uninfected and uninfected with {\it S.
+The experiment quantified the abundance of total protein and of phosphorylation in wild type (WT), and in ATG16L1-deficient (cKO) samples, uninfected and infected with {\it S.
 flexneri}.
@@
 
-The proteome was quantified in 11-plex isobaric multiplexing via tandem mass tagging (TMT) in combination with LC-MS/MS.
+\medskip \noindent {\bf Data acquisition:} 8 biological samples \todo{total?
+Or per condition?
+I am having difficulties understanding the allocation from the description...} were allocated to 2 TMT mixtures in an unbalanced time course design.
+Each mixture contained two replicates per early and late WT and KO conditions.
 The six conditions were split between 11 channels, and as the result the experiment was unbalanced.
-Each mixture contained two replicates per early and late WT and KO conditions.
 Mixture one contained one replicate of uninfected WT and two replicates of uninfected KO.
 Mixture two contained one replicate of uninfected KO and two uninfected WT.
-
-\medskip \noindent {\bf Global Profiling Run:} This experiment included a separate global profiling run with the same design to measure unmodified peptides.
-There was a 90\% overlap between the identified background modified peptides and proteins that were quantified in the global profiling run.
+This experiment included a total proteome (i.e., a global profiling run) and a phosphopeptide enrichment run.
+There was a 90\% overlap between the identified modified peptides and proteins that were quantified in the global profiling run.
 
@@
 
-\subsection*{Dataset 6 : Human - Ubiquitination - Label-free no global profiling run}
+\subsection*{Dataset 6: Human - Ubiquitination - Label-free no global profiling run}
 \label{sec:exp_proc_dataset6}
-
-In this experiment 8 biological replicates were split between 4 conditions in a balanced group comparison design. Further details in Suppler Sec \ref{supp-sec:usp30}.
 
@@
 
-\medskip \noindent {\bf Global Profiling Run:} This experiment did not include a separate global profiling run to measure unmodified peptides.
-In this case $MSstatsPTM$ could still be used by extracting unmodified peptides from the modified run.
-However, in addition to low feature counts for unmodified peptides,  this lead to substantially fewer matches between modified and unmodified peptides.
+\medskip \noindent {\bf Data acquisition:} This experiment did not include a separate global profiling run to measure unmodified peptides.
+In addition to low feature counts for unmodified peptides,  this lead to substantially fewer matches between modified and unmodified peptides.
 There was a 41.9\% overlap between the identified background modified peptides and proteins that were quantified in the global profiling run.
 
-\medskip \noindent {\bf Pairwise comparisons:} We evaluated the ability of the statistical methods to detect  changes in the abundance of modified peptides both before and after adjusting for changes in global protein abundance.
+\medskip \noindent {\bf Evaluation:} We evaluated the ability of the statistical methods to detect  changes in the abundance of modified peptides both before and after adjusting for changes in global protein abundance.
 All the conditions were compared with each other in a full pairwise comparison, resulting in 6 comparisons.
@@
 
-For simplicity, consider an LC-MS/MS experiment in the special case of a balanced design with $I=2$ conditions and $J=2$ biological replicates (mass spectrometry runs) per condition.
+Consider a label-free LC-MS/MS experiment in the special case of a balanced design with $I$ conditions and $J$ biological replicates (mass spectrometry runs) per condition.
 For one protein, the PTM site is represented by $K$ spectral features (peptide ions, distinguished by their cleavage residues and charge states).
@@
 Conversely, the unmodified protein is represented by $L$ spectral features, and the log-intensity of Feature $l$ from the unmodified peptides in the same run is denoted by $y_{ijl}$.
-
-Figure \ref{fig:data-structure}(a) schematically illustrates this data structure for one PTM and corresponding unmodified protein.
-The number of quantified features vary across replicate LC-MS/MS runs of the same sample, and across conditions.
+\ov{The features can be quantified as part of a same mass spectrometry run, or in separate enrichment and global proteome profiling runs.}
+
+Figure \ref{fig:data-structure}(a) schematically illustrates this data structure for one PTM and the corresponding unmodified features of the protein, in a special case of 2 conditions and 2 biological replicates per condition.
+The number of quantified features varies across replicate LC-MS/MS runs of the same sample, and across conditions.
 There may be feature outliers which could be caused by random variation or measurement error.
@@
 \centering
-\includegraphics[scale=.5]{images/benchmark_fig.png}
-\caption{An overview of the experimental design for Dataset 3 : SpikeIn benchmark - Ubiquitination - Label-free.
-(a) The corresponding levels of spike-in peptide, human lysate, and background lysate are shown.
-The spike-in peptides are adjusted by the human lysate.
-For example, the spike-in abundance of mixture 1 is entirely accounted for by human lysate.
-The background lysate serves as our null model.
-(b) The expected fold change of the spike-in peptides, after adjusting for changes in the unmodified peptide (human lysate).
-The expected fold change is calculated using the values of the spike-in and human lysate in part (a).}
+\includegraphics[scale=.4]{images/benchmark_fig.png}
+\caption{{\bf Dataset 3: SpikeIn benchmark - Ubiquitination - Label-free}.
+(a) Relative amounts of human lysate, background {\it E.
+Coli} lysate, and spike-in peptides.
+\todo{Clarify what is changing how} (b) The expected fold change of the spike-in peptides, after adjusting for changes in the unmodified peptide (human lysate).
+The expected fold change is calculated using the values of the spike-in and human lysate in part (a).\todo{Could you make the columns in the table more complete?
+True fold-fold changes between the backgrounds; unadjusted true fold changes, adjusted true fold changes - essentially any value for any comparison that we may want to evaluate}
+ %
+ \todo{It's silly but...
+could you make the legend wide rather than long?
+I.e., have the colors horizontally, and in the same order as in the figure.
+And maybe use 2 lines for the text to make the widths of the bars in the legend more comparable with the widths of the bars in the text.
+This will help readability and save space}
+%
+\todo{Also, why is C between A and B? The letters have no meaning, it would help to have them in order.}
+}
 \label{fig:benchmark-design}
@@
 \end{subfigure}
-\caption{Dataset 3 : SpikeIn benchmark - Ubiquitination - Label-free.
+\caption{Dataset 3: SpikeIn benchmark - Ubiquitination - Label-free.
 a) Before adjustment the fold change of the spike-in peptides' were systematically different from the expected fold change in all models.
@@
 \end{subfigure}
-\caption{Dataset 4 : Human - Ubiquitination - 1mix-TMT.
+\caption{Dataset 4: Human - Ubiquitination - 1mix-TMT.
 a) The overlap of differential modified peptides for the PTM model with and without global protein level adjustment.
@@
 	 \end{subfigure}
-\caption{Dataset 5 : Mouse - Phosphorylation - 2mix-TMT time series.
+\caption{Dataset 5: Mouse - Phosphorylation - 2mix-TMT time series.
 \todo{OV: What are the vertical panels 1 and 2 in (b)??
@@
  \end{subfigure}
- \caption{Dataset 6 : Human - Ubiquitination - Label-free no global profiling run.
+ \caption{Dataset 6: Human - Ubiquitination - Label-free no global profiling run.
 a) The overlap of differential modified peptides for the PTM model with and without global protein level adjustment.

### 2022-06-12 main/MSstatsPTM_Main.tex
Small
Methods - small

@@
 The experiment had four conditions, CCCP, USP30 over expression (USP30 OE), Combo, and Control.
-Each condition had two biological replicates.
+Each condition had two biological replicates \todo{were these patients or cell lines?}.
 The abundance of modified peptides was quantified with label-free LC-MS/MS.
@@
 
-Consider a label-free LC-MS/MS experiment in the special case of a balanced design with $I$ conditions and $J$ biological replicates (mass spectrometry runs) per condition.
+Consider a label-free LC-MS/MS experiment in the special case of a balanced design with $I$ conditions and $J$ biological replicates per condition.
+For simplicity, we assume that the experiment has no technical replicates, and each biological replicate corresponds to an LC-MS/MS run.
 For one protein, the PTM site is represented by $K$ spectral features (peptide ions, distinguished by their cleavage residues and charge states).
@@
 
-Figure \ref{fig:data-structure}(a) schematically illustrates this data structure for one PTM and the corresponding unmodified features of the protein, in a special case of 2 conditions and 2 biological replicates per condition.
-The number of quantified features varies across replicate LC-MS/MS runs of the same sample, and across conditions.
-There may be feature outliers which could be caused by random variation or measurement error.
-Additionally, some spectral features can be missing.
-The features in each mass spectrometry run are generally summarized and represent a single observation used to estimate the relative abundance of the PTM in condition $i$, $\mu_i^{\ast}$.
-
-The population parameter we are interested in estimating is the difference between the PTM in condition $1$ and condition $2$, denoted by $\mu_1^{\ast}$ and $\mu_2^{\ast}$ respectively.
+Figure \ref{fig:data-structure}(a) schematically illustrates this data structure for one PTM site and the corresponding unmodified features of the protein, in a special case of 2 conditions and 2 biological replicates per condition.
+The number of quantified modified and unmodified features varies across proteins.
+Some log-intensities may be outliers.
+Some spectral features can be missing.
+
+\todo{OV: from here on, it is unclear if you introduce the general notation, or talk about our model specifically.
+If it is our model, it should be in the Results} The population parameter we are interested in estimating is the difference between the abundances of a PTM site in condition $1$ and condition $2$, denoted by $\mu_1^{\ast}$ and $\mu_2^{\ast}$ respectively.
 To calculate the quantity of interest, statistical analysis needs to summarize values in the table in Figure \ref{fig:data-structure}(a) using appropriate models.
@@
 \end{equation}
-
 However, these population parameters are inherently convoluted with changes in the unmodified protein.
 To account for this, we must estimate the relative abundance of the unmodified protein in conditions $1$ and $2$, denoted by $\mu_1$ and $\mu_2$.
-We estimate these quantities in the same way as for the PTM, using their correspond features and observations and resulting in an estimated mean abundance, $\hat{\mu}_i$, and standard error, $\widehat{SE(\hat{\mu}_i)}$.
+We estimate the quantities $\hat{\mu}_i$ and $\widehat{SE(\hat{\mu}_i)}$ in the same way as for the PTM site, using their corresponding features and observations and resulting in an estimated mean abundance.
 These estimates are used to adjust the quantities estimated for the PTM, resulting in the new Null and Alternative Hypotheses seen in Equation \ref{eq:null_hyp}.

### 2022-06-12 main/MSstatsPTM_Main.tex
small

@@
 To account for this, we must estimate the relative abundance of the unmodified protein in conditions $1$ and $2$, denoted by $\mu_1$ and $\mu_2$.
-We estimate the quantities $\hat{\mu}_i$ and $\widehat{SE(\hat{\mu}_i)}$ in the same way as for the PTM site, using their corresponding features and observations and resulting in an estimated mean abundance.
+We estimate \todo{OV: our method, or everyone?} the quantities $\hat{\mu}_i$ and $\widehat{SE(\hat{\mu}_i)}$ in the same way as for the PTM site, using their corresponding features and observations and resulting in an estimated mean abundance.
 These estimates are used to adjust the quantities estimated for the PTM, resulting in the new Null and Alternative Hypotheses seen in Equation \ref{eq:null_hyp}.

### 2022-06-13 main/MSstatsPTM_Main.tex
Notation

--- 2022-06-12 main/MSstatsPTM_Main.tex
@@
 Consider a label-free LC-MS/MS experiment in the special case of a balanced design with $I$ conditions and $J$ biological replicates per condition.
-For simplicity, we assume that the experiment has no technical replicates, and each biological replicate corresponds to an LC-MS/MS run.
+For simplicity, we assume that the experiment has no technical replicates, such that each biological replicate corresponds to an LC-MS/MS run.
 For one protein, the PTM site is represented by $K$ spectral features (peptide ions, distinguished by their cleavage residues and charge states).
@@
 
-Figure \ref{fig:data-structure}(a) schematically illustrates this data structure for one PTM site and the corresponding unmodified features of the protein, in a special case of 2 conditions and 2 biological replicates per condition.
-The number of quantified modified and unmodified features varies across proteins.
-Some log-intensities may be outliers.
-Some spectral features can be missing.
-
-\todo{OV: from here on, it is unclear if you introduce the general notation, or talk about our model specifically.
-If it is our model, it should be in the Results} The population parameter we are interested in estimating is the difference between the abundances of a PTM site in condition $1$ and condition $2$, denoted by $\mu_1^{\ast}$ and $\mu_2^{\ast}$ respectively.
-To calculate the quantity of interest, statistical analysis needs to summarize values in the table in Figure \ref{fig:data-structure}(a) using appropriate models.
-For condition $i$, the models provide estimates for the mean abundance, $\hat{\mu}_i^{\ast}$, and standard error, $\widehat{SE(\mu_i^{\ast})}$.
-Once summarized, we can formulate the Null and Alternative Hypotheses as seen in Equation \ref{eq:conv_null_hyp}.
-
+Figure \ref{fig:data-structure}(a) schematically illustrates this data structure for one PTM site and the corresponding unmodified features of the protein, in a special case of 2 conditions and 2 biological replicates per condition \todo{OV: replace 'run' with 'bio rep' in the figure}.
+The number of modified and unmodified features varies across proteins.
+Some log-intensities may be outliers, and some spectral features can be missing.
+The population parameter of interest is the difference between the abundances of a PTM site in condition $1$ and condition $2$, denoted by $\mu_1^{\ast}$ and $\mu_2^{\ast}$ respectively.
+We are interested in testing the null hypothesis
 \begin{equation}
@@
 \end{equation}
-However, these population parameters are inherently convoluted with changes in the unmodified protein.
-To account for this, we must estimate the relative abundance of the unmodified protein in conditions $1$ and $2$, denoted by $\mu_1$ and $\mu_2$.
-We estimate the quantities $\hat{\mu}_i$ and $\widehat{SE(\hat{\mu}_i)}$ in the same way as for the PTM site, using their corresponding features and observations and resulting in an estimated mean abundance.
-These estimates are used to adjust the quantities estimated for the PTM, resulting in the new Null and Alternative Hypotheses seen in Equation \ref{eq:null_hyp}.
-
+To test the hypothesis, a statistical model must represent the values in the table in Figure \ref{fig:data-structure}(a), and provide a model-based estimate of $\mu_{1}^{\ast} - \mu_{2}^{\ast}$, and its standard error.
+
+Unfortunately, the population parameter is inherently convoluted with the overall changes in protein abundance. To account for this, we can change the null hypothesis to
 \begin{equation}
@@
 \end{equation}
+where $\mu_1$ and $\mu_2$ reflect the overall protein abundances in condition $1$ and condition $2$. These quantities can be estimated using protein features without the modification site.
 
@@
 \caption{
-Goals of PTM characterization, input to statistical analyses, and notation.
+\todo{In 2a, it may be more meaningful to replace 'run' with 'bio rep'} Goals of PTM characterization, input to statistical analyses, and notation.
 (a) Schematic data representation, in a simplified case of two conditions and two biological replicates.

### 2022-06-13 main/MSstatsPTM_Main.tex
Suggestions for methods

@@
 
-\subsection*{Existing statistical methods for experiments targeting PTMs}
-
-\noindent {\bf Two-sample $t$-test} Many investigations perform differential abundance analysis of PTMs using two-sample $t$-test or its extensions.
-While simple, these approaches do not fully account for all sources of variations, and are not directly applicable to experiments with complex designs.
-These include experiments with comparisons of multiple conditions, acquisition in multiple batches, and unbalanced designs.
-Additionally, while these approaches can be applied to experiments studying PTMs, there is not a self contained, straightforward implementation of the methods, making application challenging.
-
-When modeling the simple experiment seen in Figure \ref{fig:data-structure}, the same Null Hypothesis is addressed as seen in Equation \ref{eq:null_hyp}.
-To estimate the parameters of interest in the trivial model, the approach takes as input intensities of individual features from modified peptides, or intensity ratios of modified and unmodified peptide features, and compares the mean abundance of a PTM site from one condition to another~\cite{Schwammle2015}\cite{THOMAS2020}.
+\subsection*{Existing statistical methods for detecting differentially abundant PTMs}
+
+\noindent {\bf Two-sample $t$-test} 
+\todo{OV: could you help me with this part a little bit?
+I tried to organize the text, but could not quite finish.
+We should have the following order: (1) input, (2) model for the input, (3) model-based inference, (4) implementation; (5) good things and limitations.
+I think it may help to separate two versions: based on the features, and based on the ratios.
+Otherwise it is not very clear}
+Many investigations perform differential abundance analysis of PTMs using two-sample $t$-test or its extensions.
+$t$-test takes as input intensities of individual features from modified peptides, or intensity ratios of modified and unmodified peptide features \todo{OV: let's separate the intensities and the ratios}, and compares the mean abundance of a PTM site from one condition to another~\cite{Schwammle2015}\cite{THOMAS2020}.
 The abundance in each run is taken as input and is often estimated by sum of peak intensities.
 For example, the log-abundance estimate for the PTM in Run $j$ of Condition $i$ is given by Equation \ref{eq:feature_estimation}.
+as input the adjusted PTM abundance of each condition and biological replicate, and estimates $\Delta$ in Equation \ref{eq:null_hyp}.
+\todo{OV: what is the model?}
+To estimate the parameters of interest in this trivial model, the approach takes as input
 
@@
 
-$t$-test then takes as input the adjusted PTM abundance of each condition and biological replicate, $\hat{u}^{\ast}_{11}$, $\hat{u}^{\ast}_{12}$, $\hat{u}^{\ast}_{21}$, and $\hat{u}^{\ast}_{22}$, to estimate the quantity of interest $\hat{\Delta}$ as well as the standard error, $\widehat{SE(\hat{\Delta})}$.
-$\hat{\Delta}$ is the estimate of the difference for the Null and Alternative Hypotheses show in Equation \ref{eq:null_hyp}.
+to estimate the quantity of interest $\hat{\Delta}$ as well as its standard error, $\widehat{SE(\hat{\Delta})}$.
 The statistical significance is determined by comparing the test statistic against the $t$ distribution, with degrees of freedom $df=2J-2$ in balanced designs.
 
+While simple, these approaches do not fully account for all sources of variations, and are not directly applicable to experiments with complex designs.
+These include experiments with comparisons of multiple conditions, acquisition in multiple batches, and unbalanced designs.
+Additionally, while these approaches can be applied to experiments studying PTMs, there is not a self contained, straightforward implementation of the methods, making application challenging.
+
 \medskip \noindent {\bf Limma} Modifications of $t$-test such as moderated $t$-test with Limma were also proposed~\cite{Ritchie_15a}\cite{Zhu}.
-The Null Hypothesis we are interested in answering is the same as in Equation \ref{eq:null_hyp}.
+
+\todo{OV: could you also re-organize: (1) input, (2) model for the input, (3) model-based inference, (4) implementation, (5) good things and limitations.
+Limitation: limited in experimental design (at most two variance components, with strong assumptions)}
+
 To estimate $\hat{\Delta}$ in Figure \ref{fig:data-structure}, feature level summarization and unmodified protein adjustment are performed in the same way as $t$-test, shown in Equations \ref{eq:feature_estimation} and \ref{eq:adj_estimation}.
@@
 
-The proposed approach leverages the large amount of work done to create $MSstats$~\cite{Choi:2014}.
-$MSstats$ was designed for protein significance analysis and statistical relative quantification of proteins and peptides in global, targeted and data-independent proteomics.
+\todo{OV: could you also re-organize: (1) input, (2) model for the input, (3) model-based inference, (4) implementation, (5) good things and limitations.
+Here good things are: complex designs, including multiple variance components; EB (for TMT); good summarization; implementation.
+Limitation: MSstats is not designed for protein-level adjustment of changes in PTM}
+
+$MSstats$~\cite{Choi:2014} is designed for protein significance analysis and statistical relative quantification of proteins and peptides in global, targeted and data-independent proteomics.
 In terms of Figure \ref{fig:data-structure}, $MSstats$ was designed to estimate the difference in the unmodified protein in condition $1$ and condition $2$, denoted by $\mu_1$ and $\mu_2$ respectively.
@@
 Each PTM site is modeled and characterized separately, where a PTM is quantified with observed spectral features (boxes), distinguished by different charge states of a peptide.
-The feature intensities of the PTM in Condition 1 are denoted by $y_1^\ast$.
-Features corresponding to unmodified peptides in Condition 1 are denoted by $y_1$.
+The feature \todo{log-}intensities of the PTM in Condition 1 are denoted by $y_1^\ast$ \todo{In the figure $y^*$ has 3 subscripts.
+Could you edit here as well?}.
+\todo{Log-intensities of } features corresponding to unmodified peptides in Condition 1 are denoted by $y_1$ \todo{In the figure $y$ has 3 subscripts.
+Could you edit here as well?}.
 Peptides can be fully cleaved (solid lines) and/or partially cleaved (dashed lines).
-(b) PTM relative quantification of the population parameters by statistical inference, which makes use of the feature intensities to infer the underlying PTM abundance and protein abundance with an estimate of associated uncertainty.
+\todo{OV: I am wondering whether we need Fig 2(b).
+The null hypothesis is specified in the text.
+I am not sure it is referenced anywhere.
+Maybe remove it for simplicity?} (b) PTM relative quantification of the population parameters by statistical inference, which makes use of the feature intensities to infer the underlying PTM abundance and protein abundance with an estimate of associated uncertainty.
 (c) Model-based testing for differential PTM abundance, which corrects for the underlying protein abundance with a cost of increased uncertainty about the estimate of difference between conditions.}

### 2022-06-15 code/MSstatsPTM_Simulation_Analysis.Rmd
addressed some comments

@@
 adj_results$Model <- "Proposed"
+adj_results$Color_Model <- "Proposed"
+adj_results$Adj = TRUE
 no_adj_results$Model <- "Proposed no adjustment"
+no_adj_results$Color_Model <- "Proposed"
+no_adj_results$Adj = FALSE
 ttest_results$Model <- "T-Test no adjustment"
+ttest_results$Color_Model <- "T-Test"
+ttest_results$Adj = FALSE
 ttest_adj_results$Model <- "T-Test"
+ttest_adj_results$Color_Model <- "T-Test"
+ttest_adj_results$Adj = TRUE
 limma_model_results$Model <- "Limma no adjustment"
+limma_model_results$Color_Model <- "Limma"
+limma_model_results$Adj = FALSE
 limma_adj_model_results$Model <- "Limma"
+limma_adj_model_results$Color_Model <- "Limma"
+limma_adj_model_results$Adj = TRUE
 
@@
 cbPalette <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#999999")
-cbPalette_first <- c("#E69F00", "#F0E442", "#56B4E9", "#0072B2", "#009E73", "#D55E00", "#CC79A7", "#999999")
+cbPalette_first <- c("#56B4E9", "#E69F00", "#009E73", "#56B4E9", "#009E73", "#009E73", "#CC79A7", "#999999")
 ```
@@
   filter(conditions == 4 & sd == .3 & rep == 5) %>% 
-  ggplot() + geom_col(aes(x = Model, y = fpr, fill = Model)) + #, size = Replicates
-  scale_colour_manual(values=cbPalette_first) +
+  ggplot() + geom_col(aes(x = Model, y = fpr, fill = Color_Model, alpha = Adj)) + #, size = Replicates
+  scale_fill_manual(values=cbPalette_first) +
+  scale_alpha_manual(values = c(.5, 1.0)) + 
   # scale_size_manual(values = c(3,5,7,9)) +
@@
   ylim(0.0, 0.4) + 
+  guides(alpha = "none") + 
+  guides(fill = "none") + 
   theme(axis.text.x = element_text(angle = 45, hjust=1, vjust = 1, size = 16), 
@@
          strip.text = element_text(size = 16)) + 
-  labs(title = "Simulation 1: FDR all models", x = "Model", y = "FDR") + 
+  labs(title = "Simulation 1: False Discovery Rate by Model", x = "Model", y = "FDR") + 
      guides(color = guide_legend(order = 1, override.aes = list(size=5)), 
@@
 ```{r echo=FALSE, message=FALSE, warning=FALSE}
-png("../supplementary/sim_new/sim3_Accuracy.png", width = 750, height = 500)
+png("../main/images/sim3_Accuracy.png", width = 750, height = 500)
 models %>% mutate(Model = factor(Model, levels=c("Proposed", "Proposed_no_adj", "limma_no_adj", 
@@
          strip.text = element_text(size = 16)) + 
-  labs(title = "Simulation 2: Accuracy adjusted models", x = "Model", y = "Accuracy") + 
+  labs(title = "Simulation 2: Model Accuracy", x = "Model", y = "Accuracy") + 
    guides(color = guide_legend(order = 1, override.aes = list(size=5)), 

### 2022-06-15 code/Spike-in_Analysis.Rmd
Added data availability info
addressed some comments

@@
 ---
-title: "Spike-in_MSstatsPTM_Analysis"
+title: "SpikeIn MSstatsPTM Analysis"
 author: "Devon Kohler"
-date: "6/18/2021"
+date: "3/7/2022"
 output: html_document
@@
 load(file = "D:\\Northeastern\\Research\\MSstats\\PTM_Refractor\\Label_Free_PTM_Data\\benchmark-spike\\MSstatsPTM_Summarized.rda")
+
+summary_df$PTM$ProcessedData
+summary_df$PTM$ProcessedData$global_prot = sapply(summary_df$PTM$ProcessedData$PROTEIN, function(x){str_split(x, "_")[[1]][1]})
+
+summary_df$PTM$ProcessedData$global_prot %>% n_distinct()
+summary_df$PTM$ProcessedData$PROTEIN %>% n_distinct()
+summary_df$PTM$ProcessedData$FEATURE %>% n_distinct()
+
+summary_df$PROTEIN$ProcessedData$PROTEIN %>% n_distinct()
+summary_df$PROTEIN$ProcessedData$PEPTIDE %>% n_distinct()
 

### 2022-06-15 main/MSstatsPTM_Main.tex
addressed some comments

--- 2022-06-12 main/MSstatsPTM_Main.tex
@@
 
-\usepackage{fullpage}
+%\usepackage{fullpage}
+\usepackage[margin=1in,headheight=13.6pt]{geometry}
 \usepackage{amsfonts,amsthm}
@@
 \usepackage{tabulary,multirow,multicol,rotating}
-\usepackage[backend=biber,sorting=none]{biblatex}
+\usepackage[backend=biber, citestyle=numeric, bibstyle=authoryear, sorting=none]{biblatex}
 \addbibresource{bibliography.bib}
+
+\makeatletter
+\input{numeric.bbx}
+\makeatother
 
@@
 
+\usepackage{fancyhdr}
+\pagestyle{fancy}
+\fancyhead[L]{MSstatsPTM: Statistical relative quantification of PTMs}
+
 %\date{\vspace{-5ex}} % to remove date in the title
@@
 \author[2]{Tsung-Heng~Tsai}
+\author[4]{Erik~Verschueren}
 \author[1]{Ting~Huang}
-\author[4]{Erik~Verschueren}
 \author[3]{Trent~Hinkle}
+\author[3]{Lilian~Phu}
 \author[3]{Meena~Choi*}
@@
 \affil[*]{Corresponding Authors}
+%\affil[]{MSstatsPTM: Statistical relative quantification of PTMs}
 
@@
 Next, model-based inferences regarding the PTM and the protein-level abundances are combined to account for the confounding between these two sources.
-Evaluations on computer simulations, a spike-in experiment with known ground truth, and three biological experiments \ov{with different organisms, modification types and data acquisition types} demonstrate the improved fold change estimation and detection of differential PTM abundance, as compared to currently used approaches.
+Evaluations on computer simulations, a spike-in experiment with known ground truth, and three biological experiments with different organisms, modification types and data acquisition types demonstrate the improved fold change estimation and detection of differential PTM abundance, as compared to currently used approaches.
 The proposed framework is implemented in the free and open-source R/Bioconductor package $MSstatsPTM$.
@@
 For example, meaningful patterns of changes in post-translational modifications (PTMs) abundance can serve as biomarkers of a disease~\cite{Petushkova_2017}.
-Alternatively, distinguishing the quantitative changes in a PTM from the overall changes of the protein abundance helps gain insight into biological and physiological processes operating on a very short timescale~\cite{Chandramouli:2009}.
+Alternatively, distinguishing the quantitative changes in a PTM from the overall changes of the protein abundance helps gain insight into biological and physiological processes operating on a very short timescale~\cite{Chandramouli:2009}\cite{Kim2016}.
+This helps to distinguish between relative site occupancy changes at steady-state protein levels, typical for short time-scale signaling events, and observed relative changes of PTMs as a result of underlying gene expression or protein abundance levels.
 
@@
 Frequently data from these experiments are analyzed using statistical methods that were not originally designed for this task.
-Researchers use methods such as $t$-test \todo{ref}, Analysis of Variance\todo{ref}, or Limma\todo{ref}, by taking as input the intensity ratios of modified and unmodified peptide features, and comparing the mean abundance of different PTM sites~\cite{Ritchie_15a}.
+Researchers use methods such as $t$-test\cite{Kalpic:2011}, Analysis of Variance\cite{girden:1992}, or Limma\cite{Ritchie_15a}, by taking as input the intensity ratios of modified and unmodified peptide features, and comparing the mean abundance of different PTM sites.
 Such approaches do not fully account for all the sources of uncertainty.
@@
 The datasets illustrate a diverse set of organisms, modification types, acquisition methods and experimental designs, showing the applicability of the framework to a variety of situations.
-By appropriately leveraging the information from the unmodified peptides, the proposed approach improved the accuracy of the estimates of PTM fold changes and produced a better calibrated false positive rate of detecting differentially abundant PTMs \todo{as compared to methods XXX}.
+By appropriately leveraging the information from the unmodified peptides, the proposed approach improved the accuracy of the estimates of PTM fold changes and produced a better calibrated false positive rate of detecting differentially abundant PTMs as compared to existing methods.
 In particular, accounting for the confounding from protein abundance allowed us to characterize the true effect of the modification, avoiding the need for more manual and time intensive follow-up investigation.
@@
 \section{Experimental procedures}
-
-%%%% Summary of controlled mixtures
-\begin{table}[t!]
-\centering
-\begin{tiny}
-\begin{tabular}{|c|c|ccc|c|}
-\hline
-%%%% colnames of table
-& \multirow{2}{*}{Dataset} & Experimental & No. of & No. of bio. &  Data \\
-&  & design & Conditions & replicates & availability \\
-\hline
-\hline
- &&&&& \\[-0.05in]
-%%%% DDA:ControlledMix
-\multirow{5}{*}{\rotatebox[origin=c]{90}{Known}}  \multirow{5}{*}{\rotatebox[origin=c]{90}{Ground}} \multirow{5}{*}{\rotatebox[origin=c]{90}{Truth}} & Computer simulation 1 - Label-free & Group comparison & 2/3/4 & 2/3/5/10 & \href{https://github.com/devonjkohler/MSstatsPTM_simulations}{Github}\\
- &&&&& \\%[-0.05in]
-%\hline 
-%&&&&&&&& \\[-0.05in]
-% \cline{2-9}
-% &&&&&&&& \\[-0.05in]
-%%% DDA iPRG2015
-& Computer simulation 2 - Missing and low features & Group comparison & 2/3/4 & 2/3/5/10 & \href{https://github.com/devonjkohler/MSstatsPTM_simulations}{Github} \\
- &&&&& \\%[0.05in]
-%\hline 
-% &&&&&& \\[-0.05in]
-%%% DDA Cox2014
-& SpikeIn benchmark - Ubiquitination - Label-free & Group comparison & 4 & 2 & \href{https://massive.ucsd.edu/ProteoSAFe/private-dataset.jsp?task=c4c583ecf7f941cdac87f7a4f872517b}{MSV000088971}\\
- &&&&& \\%[-0.05in]
-\hline
-\multicolumn{6}{c}{ } \\ [0.02in]
-\hline 
- &&&&& \\%[-0.05in]
-%%%%%%%%%%%%%%%%%%
-%%%% biological study
-%%%%%%%%%%%%%%%%%%
-%% DDA Meierhofer
-\multirow{3}{*}{\rotatebox[origin=c]{90}{Biological}} \multirow{3}{*}{\rotatebox[origin=c]{90}{Experiment}} & Human - Ubiquitination - 1mix-TMT & Group comparison & 6 & 2 & \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=b6f0c74c234247678fb0888c6df1f225}{MSV000088966} \\
-&&&&& \\%[-0.05in]
-% \cline{2-9}
-% &&&&&&& \\[-0.05in]
-%%%% SRM Picotti 2009
-& Mouse - Phosphorylation - 2mix-TMT & Time course & 6 & 4 & \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=4878d777c6b34cf8aaf8477e93140c4d}{MSV000085565} \\
-&&&&& \\%[-0.05in]
-% \cline{2-9}
-% &&&&&&& \\[-0.05in]
-%%%% SRM Surinova 2015, training
-& Human - Ubiquitination - Label-free & Group comparison & 4 & 2 & \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=1b516164de5345108b40b75147dd58b5}{MSV000078977}\\ [0.02in]
- &&&&& \\
- \hline
-\end{tabular}\\
-\end{tiny}
-\caption{ \small {\bf Simulated and experimental datasets.} 
-```Dataset'' is the dataset code name.
-``Data availability'' shows the ID of the MassIVE.quant repository or the GitHub repository.
-All the experiments were conducted in data-dependent acquisition (DDA) mode.
-\todo{Could you add to the table the number of proteins, and the average number of modified/unmodified peptides per protein} \todo{If the figures are at the end, I think the table should also be at the end}
-\todo{Could you check the table and the entire section: when we say 'number of biological replicates', we typically mean per condition.
-But it seems that some descriptions imply the total number of biological samples across conditions.
-Could you clarify that we use number of biological replicates per condition throughout?
-(E.g, this helps calculate the degrees of freedom in balanced designs)}
-}
-\label{tab:dataDescription}
-\end{table}
 
@@
 
-\todo{Additional details on every dataset are available in Supplementary Section 2 (I think it's ok to only mention this once)}
-
-\todo{Could you delete this paragraph, and move the container identifies to a new column in table 1?} The R scripts and results of the statistical analysis for the four experiments are available in MassIVE.quant reanalysis containers: SpikeIn benchmark - Ubiquitination - Label-free (MassIVE reanalysis container identifier: \href{https://massive.ucsd.edu/ProteoSAFe/private-dataset.jsp?task=c4c583ecf7f941cdac87f7a4f872517b}{TBD}), Human - Ubiquitination - 1mix-TMT (MassIVE reanalysis container identifier: \href{https://massive.ucsd.edu/ProteoSAFe/reanalysis_container.jsp?task=b639fa5d0ff24ce78abdca13f4b18c0a}{RMSV000000356}), Mouse - Phosphorylation - 2mix-TMT time series (MassIVE reanalysis container identifier: \href{https://massive.ucsd.edu/ProteoSAFe/reanalysis_container.jsp?task=79406d32c1f9413daf7756e6eeb24968}{RMSV000000357}), Human - Ubiquitination - Label-free no global profiling run (MassIVE reanalysis container identifier: \href{https://massive.ucsd.edu/ProteoSAFe/reanalysis_container.jsp?task=f9e6117e00794ce5b2e8802601be16bc}{RMSV000000358}).
+Additional details on every dataset are available in Supplementary Section 2.
 
@@
 24 synthetic label-free datasets were generated with different experimental designs and different biological variation.
-In each dataset, 1000 proteins had 10 unmodified features per protein.
-Each of the 1000 proteins had one PTM.
+In each dataset, 1,000 proteins had 10 unmodified features per protein.
+Each of the 1,000 proteins had one PTM.
 Each PTM was represented by 10 modified features.
@@
 
-\medskip \noindent {\bf Evaluation:} We evaluated the ability of the statistical methods \todo{XXX - what and how}
+\medskip \noindent {\bf Evaluation:} We evaluated the ability of the statistical methods to correctly detect which PTMs were simulated with changes between conditions and which were not.
+We evaluated the methods ability to avoid false positives (i.e.
+specificity), accurately estimate the fold change between conditions, and analyzed the sensitivity of detecting differentially abundant PTMs.
+Each method was used to analyze changes in PTM abundance in the presence of confounding with changes in the unmodified protein, and with this confounding removed.
 
@@
 
-\medskip \noindent {\bf Evaluation:} \todo{As above, we evaluated... what and how...}
+\medskip \noindent {\bf Evaluation:} The methods were evaluated in the same way as above.
+We evaluated their ability to correctly detect PTM's specificity, fold change estimation, and sensitivity.
+These statistics were analyzed both in the presence of, and without, confounding with changes in the unmodified protein.
 
@@
 Background {\it E.
-coli} lysate were used to equalize total protein levels prior to enrichment or global protein profiling \ov{between Mix 1 and 3, and between Mix 2 and 4}.
-50 heavy-labeled KGG motif peptides from 20 human proteins were used as spike-in peptides.
-Quantitative changes in protein and site abundance of these 20 proteins were the target of the benchmark.
-\todo{In particular, we distinguish the changes in the abundances of the modified peptides (i.e., unadjusted changes) and the changes of their abundance relative to the changes in the abundances of the human lysate (i.e., protein-level adjusted changed).
-} The true log-fold changes \ov{between the relevant components of the relevant mixtures} are summarized in Figure \ref{fig:benchmark-design}(b).
-Two replicates mixtures were created per condition.
+coli} lysate were used to equalize total protein levels prior to enrichment or global protein profiling between Mix 1 and 3, and between Mix 2 and 4.
+50 heavy-labeled KGG motif peptides from 20 human proteins were used as spike-in peptides in a mixed background of lysates from {\it E.
+coli} and a human cell line.
+Quantitative changes in protein and site abundance of these 20 human proteins were the target of the benchmark.
+In particular, we distinguish the changes in the abundances of the modified peptides (i.e., unadjusted changes) and the changes of their abundance relative to the changes in the abundances of the human lysate (i.e., protein-level adjusted changed).
+The true log-fold changes between the relevant components of the relevant mixtures are summarized in Figure \ref{fig:benchmark-design}(b).
+Two replicate mixtures were created per condition.
 
 \medskip \noindent {\bf Data acquisition:} Each mixture was analyzed with KGG enrichment, and without KGG enrichment (i.e., in a global profiling run), with label-free LC-MS/MS.
-There was a 90.2\% overlap between the identified background modified peptides and proteins that were quantified in the global profiling run.
-
-\medskip \noindent {\bf Evaluation:} \todo{Here is the place to describe: 'we expect XXX to not change'.
+There was a 90.2\% overlap of protein identifications between the identified background modified peptides and proteins that were quantified in the global profiling run.
+
+\medskip \noindent {\bf Evaluation:} We expect the spike-in peptides to change corresponding to the values in Figure \ref{fig:benchmark-design}(b).
+These peptides are treated as positive controls in all comparisons except Mix 1 vs Mix 4.
+Additionally, the background {\it E.
+coli} lysate peptides are treated as negative controls and are not expected to change in any comparison.
+Therefor we evaluated the statistical methods ability to avoid false positives, as well as their sensitivity in detecting the differentially abundant spike-in peptides and accurately estimate their expected fold change.
+
+\todo{Here is the place to describe: 'we expect XXX to not change'.
 Therefore we evaluated the ability of statistical methods to avoid false positives (i.e., specificity).
@@
 \cite{LUCHETTI2021} profiled human epithelial cells engineered to express IpaH7.8 under a dox inducible promoter.
-Uninfected cells were measured at 0 and 6 hours, while cells infected \todo{with what?
-Not clear about the infection} were measured at 1, 2, 4, and 6 hour increments, resulting in six total conditions.
-GSDMD was actively degraded when IpaH7.8 expression was induced by dox treatment.
-\todo{Is GSDMD a known protein that will be used for validation?}
-
-\medskip \noindent {\bf Data acquisition:} 11 samples \todo{why 11 and not 12?
-Which condition didn't have a biological replicate?} were allocated to 1 TMT mixture in an unbalanced group comparison design.
+Uninfected cells were measured at 0 and 6 hours, while cells infected with {\it Shigella flexneri} ({\it S.
+flexneri}) bacteria were measured at 1, 2, 4, and 6 hour increments, resulting in six total conditions.
+Shigella ubiquitin ligase IpaH7.8 was shown to function as an inhibitor of the protein gasdermin D (GSDMD).
+GSDMD was actively degraded when IpaH7.8 expression was induced by dox treatment in human cells.
+
+\medskip \noindent {\bf Data acquisition:} 11 samples were allocated to 1 TMT mixture in an unbalanced group comparison design.
+All conditions were allocated two biological replicates except for the Dox1hr condition, which was allocated one replicate.
 The ubiquitinated peptides, and the total proteome (i.e., global profiling) were each conducted in a single LC-MS/MS run.
@@
 
-\subsection*{Dataset 5: Mouse - Phosphorylation - 2mix-TMT time series}
+\subsection*{Dataset 5: Mouse - Phosphorylation - 2mix-TMT}
 \label{sec:exp_proc_dataset5}
@@
 \medskip \noindent {\bf Experimental Design:} Maculins et al.
-\cite{Maculins} studied primary murine macrophages infected with {\it Shigella flexneri} ({\it S.
-flexneri}).
+\cite{Maculins} studied primary murine macrophages infected with {\it S.
+flexneri}.
 The experiment quantified the abundance of total protein and of phosphorylation in wild type (WT), and in ATG16L1-deficient (cKO) samples, uninfected and infected with {\it S.
@@
 
-\medskip \noindent {\bf Data acquisition:} 8 biological samples \todo{total?
-Or per condition?
-I am having difficulties understanding the allocation from the description...} were allocated to 2 TMT mixtures in an unbalanced time course design.
-Each mixture contained two replicates per early and late WT and KO conditions.
-The six conditions were split between 11 channels, and as the result the experiment was unbalanced.
-Mixture one contained one replicate of uninfected WT and two replicates of uninfected KO.
-Mixture two contained one replicate of uninfected KO and two uninfected WT.
+\medskip \noindent {\bf Data acquisition:} 22 biological samples were allocated to 2 TMT mixtures in an unbalanced design, with 11 samples allocated to each mixture.
+16 replicates were spread equally between the early and late WT and cKO conditions, resulting in four replicates per condition.
+Both the uninfected WT and cKO contained 3 replicates, with mixture one allocating one replicate to uninfected WT and two replicates to uninfected cKO.
+Conversely, mixture two contained one replicate of uninfected cKO and two uninfected WT.
 This experiment included a total proteome (i.e., a global profiling run) and a phosphopeptide enrichment run.
@@
 
-\medskip \noindent {\bf Time series comparison:} We evaluated the ability of the statistical methods to detect changes in the abundance of modified peptides both before and after adjusting for changes in global protein abundance.
+\medskip \noindent {\bf Evaluation:} We evaluated the ability of the statistical methods to detect changes in the abundance of modified peptides both before and after adjusting for changes in global protein abundance.
 The six condition were labeled KO\_Uninfect, KO\_Early, KO\_Late, WT\_Uninfect, WT\_Early, and WT\_Late.
@@
 The experiment had four conditions, CCCP, USP30 over expression (USP30 OE), Combo, and Control.
-Each condition had two biological replicates \todo{were these patients or cell lines?}.
+Cell lines were used to create two biological replicates per condition.
 The abundance of modified peptides was quantified with label-free LC-MS/MS.
@@
 
-Consider a label-free LC-MS/MS experiment in the special case of a balanced design with $I$ conditions and $J$ biological replicates per condition.
-For simplicity, we assume that the experiment has no technical replicates, and each biological replicate corresponds to an LC-MS/MS run.
+Consider a label-free LC-MS/MS experiment in the special case of a balanced design with $I$ conditions and $J$ biological replicates (mass spectrometry runs) per condition.
 For one protein, the PTM site is represented by $K$ spectral features (peptide ions, distinguished by their cleavage residues and charge states).
@@
 Conversely, the unmodified protein is represented by $L$ spectral features, and the log-intensity of Feature $l$ from the unmodified peptides in the same run is denoted by $y_{ijl}$.
-\ov{The features can be quantified as part of a same mass spectrometry run, or in separate enrichment and global proteome profiling runs.}
-
-Figure \ref{fig:data-structure}(a) schematically illustrates this data structure for one PTM site and the corresponding unmodified features of the protein, in a special case of 2 conditions and 2 biological replicates per condition.
-The number of quantified modified and unmodified features varies across proteins.
-Some log-intensities may be outliers.
-Some spectral features can be missing.
-
-\todo{OV: from here on, it is unclear if you introduce the general notation, or talk about our model specifically.
-If it is our model, it should be in the Results} The population parameter we are interested in estimating is the difference between the abundances of a PTM site in condition $1$ and condition $2$, denoted by $\mu_1^{\ast}$ and $\mu_2^{\ast}$ respectively.
+The features can be quantified as part of a same mass spectrometry run, or in separate enrichment and global proteome profiling runs.
+
+Figure \ref{fig:data-structure}(a) schematically illustrates this data structure for one PTM and the corresponding unmodified features of the protein in a special case of 2 conditions and 2 biological replicates per condition.
+The number of quantified features varies across replicate LC-MS/MS runs of the same sample, and across conditions.
+There may be feature outliers which could be caused by random variation or measurement error.
+Additionally, some spectral features can be missing.
+The features in each mass spectrometry run are generally summarized and represent a single observation used to estimate the relative abundance of the PTM in condition $i$, $\mu_i^{\ast}$.
+
+The population parameter we are interested in estimating is the difference between the PTM in condition $1$ and condition $2$, denoted by $\mu_1^{\ast}$ and $\mu_2^{\ast}$ respectively.
 To calculate the quantity of interest, statistical analysis needs to summarize values in the table in Figure \ref{fig:data-structure}(a) using appropriate models.
@@
 \end{equation}
+
 However, these population parameters are inherently convoluted with changes in the unmodified protein.
 To account for this, we must estimate the relative abundance of the unmodified protein in conditions $1$ and $2$, denoted by $\mu_1$ and $\mu_2$.
-We estimate \todo{OV: our method, or everyone?} the quantities $\hat{\mu}_i$ and $\widehat{SE(\hat{\mu}_i)}$ in the same way as for the PTM site, using their corresponding features and observations and resulting in an estimated mean abundance.
+We estimate these quantities in the same way as for the PTM, using their corresponding features and observations, resulting in an estimated mean abundance, $\hat{\mu}_i$, and standard error, $\widehat{SE(\hat{\mu}_i)}$.
 These estimates are used to adjust the quantities estimated for the PTM, resulting in the new Null and Alternative Hypotheses seen in Equation \ref{eq:null_hyp}.
@@
 
-\noindent {\bf Two-sample $t$-test} Many investigations perform differential abundance analysis of PTMs using two-sample $t$-test or its extensions.
+\noindent {\bf Two-sample $t$-test} Many investigations perform differential abundance analysis of PTMs using two-sample $t$-test or its extensions~\cite{Schwammle2015}\cite{THOMAS2020}.
 While simple, these approaches do not fully account for all sources of variations, and are not directly applicable to experiments with complex designs.
@@
 
-When modeling the simple experiment seen in Figure \ref{fig:data-structure}, the same Null Hypothesis is addressed as seen in Equation \ref{eq:null_hyp}.
-To estimate the parameters of interest in the trivial model, the approach takes as input intensities of individual features from modified peptides, or intensity ratios of modified and unmodified peptide features, and compares the mean abundance of a PTM site from one condition to another~\cite{Schwammle2015}\cite{THOMAS2020}.
+When modeling the simple experiment seen in Figure \ref{fig:data-structure}, we are interested in answering the Null Hypothesis in Equation \ref{eq:null_hyp}.
+To estimate the parameters of interest in the trivial model, the approach takes as input intensities of individual features from modified peptides, or intensity ratios of modified and unmodified peptide features, and compares the mean abundance of a PTM site from one condition to another~\cite{Schwammle2015}.
 The abundance in each run is taken as input and is often estimated by sum of peak intensities.
@@
 
-\medskip \noindent {\bf Isobar-PTM} was also proposed for experiments with LC-MS/MS quantitative strategies that employ isobaric labels such as tandem mass tags (TMT) or isobaric tag for relative and absolute quantification (iTRAQ)\cite{Breitwieser:2013}.
+\medskip \noindent {\bf Isobar-PTM} was also proposed for experiments with LC-MS/MS quantitative strategies that employ isobaric labels such as TMT or isobaric tag for relative and absolute quantification (iTRAQ)\cite{Breitwieser:2013}.
 Isobar-PTM expresses MS measurements with a linear model and performs adjustment with respect to protein abundance using the difference between log-ratio of modified peptides in two channels and log-ratio of protein level.
@@
 In terms of Figure \ref{fig:data-structure}, $MSstats$ was designed to estimate the difference in the unmodified protein in condition $1$ and condition $2$, denoted by $\mu_1$ and $\mu_2$ respectively.
-In other words it allows us to answer the new Null Hypothesis seen in Equation \ref{eq:msstats_null_hyp}.
+In other words it allows us to answer the Null Hypothesis seen in Equation \ref{eq:msstats_null_hyp}.
 
@@
 
-The package can also be adjusted to analyze information on the peptide level, allowing us to estimate the convoluted Null Hypothesis in Equation \ref{eq:conv_null_hyp}.
-However, $MSstats$ currently includes no built in method to estimate the the deconvoluted Null Hypothesis in Equation \ref{eq:null_hyp}.
+The package can be adjusted to analyze information on the peptide level, allowing us to estimate the convoluted Null Hypothesis in Equation \ref{eq:conv_null_hyp}.
+However, $MSstats$ currently includes no built in method to estimate the deconvoluted Null Hypothesis in Equation \ref{eq:null_hyp}.
 
@@
 
-Additionally, $MSstats$ provides a method for imputing missing feature intensities if desired.
+%Additionally, $MSstats$ provides a method for imputing missing feature intensities if desired.
 When values are imputed, it is assumed they are missing for reasons of low abundance.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section*{Results}
-
-\subsubsection*{Detection of changes in PTMs - Balanced design with one source of variation}
+\section{Results}
+
+\subsection*{Detection of changes in PTMs - Balanced design with one source of variation}
 \label{sec:adjustment_summary}
 
-We propose a statistical method for detecting changes in PTMs that allows us to answer the Null Hypothesis stated in Equation \ref{eq:null_hyp}; there is no difference in log-abundance of the PTM site between conditions, adjusted with respect to unmodified protein abundance.
-It leverages the methods used in $MSstats$ to answer this hypothesis, while being applicable to experiments with complex designs.
+We propose a statistical method for detecting changes in PTMs that allows us to answer the Null Hypothesis stated in Equation \ref{eq:null_hyp}, i.e.
+that there is no difference in the log-abundance of the PTM site between conditions, adjusted with respect to unmodified protein abundance.
+We leverage the methods used in $MSstats$ to answer this hypothesis, allowing the methods to be directly applicable to experiments with complex designs.
 Relating to the example in Figure \ref{fig:data-structure}, the expected value of the log-abundance of the PTM and unmodified protein in Condition $1$ are denoted by $\hat{\mu}_{1}^{\ast}$ and $\hat{\mu}_{1}$ respectively.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection*{Detection of changes in PTMs - Extension to complex designs}
+\subsection*{Detection of changes in PTMs - Extension to complex designs}
 
@@
 
-Details on how the proposed method can be used to run sample size calculations and power analysis on experiments with complex design can be seen in Supplementary Sec. \ref{supp-sec:complex_design}
+Details on how the proposed method can be used to run sample size calculations and power analysis on experiments with complex design can be seen in Supplementary Sec. \ref{supp-sec:complex_design}.
 
@@
 The converters take as input the raw data from the tool, identify the modification site for modified peptide, and put the data into the correct format for analysis in $MSstatsPTM$.
-Conversion is done separately for the modified and global profiling runs.
-If the global profiling run is not available, the package can still analyze the modified run, but will do so without adjusting for changes in unmodified protein abundance.
-Specifically, if a global profiling run is not available the hypotheses in Equation \ref{eq:conv_null_hyp} will be used; which amounts to running the methods seen in $MSstats$ on modified peptide level data \cite{Choi:2014} \cite{Huang:2020}.
+Conversion is done separately for the modified peptide and global profiling runs.
+If the global profiling information is not available, the package can still analyze the modified run, but will do so without adjusting for changes in unmodified protein abundance.
+Specifically, this will reduce our null hypotheses to Equation \ref{eq:conv_null_hyp}; which amounts to running the methods seen in $MSstats$ on modified peptide level data \cite{Choi:2014} \cite{Huang:2020}.
 
@@
 Additionally, the summarization includes global median normalization and normalizes between MS runs.
-The package can optionally impute missing values using an AFT model.
+The package can optionally impute missing values using an accelerated failure time (AFT) model~\cite{Wei:1992}.
 Missing value imputation should be performed with care when applied to PTM data.
@@
 The comparisons of interest can either be predefined or a full pairwise comparison will be tested.
-After fitting a model to both the modified and unmodified data, the modified model is adjusted for changes in unmodified protein abundance, using the methods described in this paper.
-
-Beyond the core functionalities of conversion, summary, and modeling, the package also includes functions for plotting the results.
-These include plots for the summarized and modeled data to assist in the analysis of the experiment.
-The summarized plots help with quality assurance analysis and identifying sources of variation.
-This includes a quality control plot, summarizing the peptide abundance per run in the form of a boxplot, and a profile plot, plotting each feature and the overall feature summarization as a line plot over each run.
-Additionally, the model plots include a volcano plot, showing all peptides adjusted p-values and fold changes, as well as a heatmap, which evaluates the fold change between conditions and peptides.
+After fitting a model to both the modified and unmodified data, the modified model is adjusted for changes in unmodified protein abundance, using the methods described above.
+
+Beyond the core functionality, the package also includes functions for plotting the results.
+These include summarization plots to help with quality assurance and identifying sources of variation (such as quality control plots and profile plots), as well as modeling plots (such as volcano plots and heatmaps), which evaluates the fold change between conditions and peptides.
 
@@
 The statistical modeling relies on the functionality from the R packages $lme4$ \cite{Bates2015} and $lmerTest$ \cite{Kuznetsova2017}.
-
 The code is available on Bioconductor, \url{http://www.bioconductor.org/packages/release/bioc/html/MSstatsPTM.html}, and Github, \url{https://github.com/Vitek-Lab/MSstatsPTM}.
@@
 True negatives were modified peptides that were not differential after accounting for changes in global protein abundance.
-To determine positives and negatives we adjusted the p-values for multiple testing and used an $\alpha=.05$ cutoff to determine differential sites.
+To determine positives and negatives we adjusted the p-values for multiple testing and used an $\alpha=.05$ cutoff.
 These results were then compared to the ground truth to determine TP and TN.
@@
 All datasets in this paper were evaluated using the $MSstatsPTM$ package.
+
+Further details reviewing the results of each experimental analysis can be seen in Supplementary Sec. \ref{supp-sec:experiments}.
 
@@
 Additionally, when comparing the fold change estimation across all modified peptides, the proposed method showed a tighter distribution of estimated fold changes around the true fold change, as compared to $t$-test and Limma.
-Specifically, the inner quartile range (IQR) of the estimated fold change for the proposed method was on average 10.4\% smaller than $t$-test and 21.8\% smaller than Limma.
+Specifically, the inner quartile range (IQR) of the estimated fold change for the proposed method was on average 21.8\% smaller than Limma's IQR and 10.4\% smaller than $t$-test's IQR.
 This fold change comparison can be seen in Supplementary Figure \ref{supp-fig:fc_boxplot}.
-
-Further details reviewing the results of each simulation and  method comparison can be seen in Supplementary Sec. \ref{supp-sec:sim}.
 
@@
 Comparing the IQR of the spike-in peptide's log fold change, the proposed method's IQR was 36.78\% smaller than Limma's and 32.98\% smaller than $t$-test's.
-This means that the proposed method's fold change estimate was more consistently correct when compared to Limma and $t$-test.
-
-In Figure \ref{fig:spikein_prop_volcano} we can clearly see the red labeled spike-in peptides do not follow the expected fold change before adjusting for changes in global protein abundance.
-After adjustment, the estimated fold change was more in line with expectation.
-Additionally, the background peptides serving as the null model show many false positives before adjustment was made, however after adjustment the number of false positives decreased significantly.
+This means that the proposed method estimated the correct fold change more consistently, compared to Limma and $t$-test.
+
+In Figure \ref{fig:spikein_prop_volcano} we can clearly see the fold change of the red labeled spike-in peptides was only correctly estimated when accounting for changes in the unmodified protein abundance.
+Additionally, the background peptides, serving as the null model, show many false positives before adjustment.
+After adjustment the number of false positives decreased significantly.
 Specifically, for the proposed method, the number of false positives went from 20.88\% to 1.84\% after adjustment was applied.
-Using both the proposed method and Limma, the red labeled spike-in peptides were more inline with expected fold change after the adjustment was applied.
-However, Limma resulted in the majority of the differential spike-in peptides having an insignificant adjusted p-value.
+While the proposed method and Limma both correctly estimated the fold change of the spike-in peptides, Limma resulted in the majority having an insignificant adjusted p-value.
 In this case, using Limma would have resulted in missing the majority of differentially modified peptides.
-
-Further results of this experiment are provided in Supplementary Sec. \ref{supp-sec:benchmark}.
 
@@
 We can see that more peptides became insignificant after adjustment than became significant.
-3,236 modified peptides became insignificant, 1352 became significant, while 4,282 were significant in both models.
+3,236 modified peptides became insignificant, 1,352 became significant, while 4,282 were significant in both models.
 For the peptides that became insignificant in the adjusted model, their change in abundance was mainly due to changes in global protein abundance.
 In contrast, for peptides that became significant after adjustment, their true abundance change was masked by underlying changes in the unmodified protein.
-Both of these issues are corrected in adjustment, and the true abundance change is shown.
+Both of these issues were corrected in adjustment, and the deconfounded abundance change is shown.
 An additional question that must be addressed is if the decrease in significant peptides is due to the increased variance that comes from adjustment.
@@
 
-\subsubsection*{Proposed approach removed confounding in time series TMT experiment}
+\subsubsection*{Proposed approach removed confounding in two mixture TMT experiment}
 
 The results of this experiment are summarized in Figure \ref{fig:data5_plots}.
-The experiment was modeled as a time series, with the same biological replicate measured at each time point.
-Modeling the experiment as a time series lowered the model variance as compared to modeling the experiment as a group comparison.
 In Figure \ref{fig:data5_venn_diagram} the number of significant modified peptides before and after adjustment is shown.
@@
 
-Analysis of other modifications in this experiment can be seen in Supplementary Sec. \ref{supp-sec:shigella}.
-
 \subsubsection*{In label-free experiment without a separate global profiling run, the proposed approach could eliminate the bias due to changes in the unmodified protein, albeit less effectively than in the presence of a global profiling run}
@@
 Any modified peptide without a corresponding unmodified protein could not be adjusted.
-Of the 10,799 ubiquitin sites identified, only 4526 had a corresponding unmodified protein and could be adjusted.
+Of the 10,799 ubiquitin sites identified, only 4,526 had a corresponding unmodified protein and could be adjusted.
 Additionally, the lack of a separate global profiling run resulted in low feature counts for the unmodified protein model.
@@
 In the bottom plot, only modified peptides that could be adjusted are shown.
-Here there were much fewer peptides that become insignificant after adjustment.
+Here there were much fewer peptides that became insignificant after adjustment.
 726 modified peptides became insignificant, 547 became significant, and 1,078 were significant in both models.
@@
 In this experiment there were only 25 modified peptides that meet this criteria.
-
-Further analysis of modifications before and after protein adjustment can be seen in Supplementary Sec. \ref{supp-sec:usp30}.
 
@@
 
-Additionally, abundance levels of PTM sites can be convoluted with each other if there are two or more modification sites per peptide.
+Additionally, abundance levels of PTM sites can be convoluted with each other if there are multiple modification sites per peptide, or convoluted with changes in the unmodified peptide (as opposed to the unmodified protein).
 In the current implementation the effect of a specific modification in a peptide with multiple modifications cannot be quantified.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\newpage
+\section{Tables}
+
+%%%% Summary of controlled mixtures
+\begin{table}[ht]
+\centering
+\caption*{Experimental Datasets Overview}
+\begin{tiny}
+\begin{tabular}{|c|c|ccccc|cc|}
+\hline
+%%%% colnames of table
+& \multirow{2}{*}{Dataset} & No. of & No. of bio. & No. of & No. of mod. & No. of unmod.&  Data & \multirow{2}{*}{Analysis}\\
+&  & Conditions & replicates & proteins & features/site & peptides/protein & availability & \\
+\hline
+\hline
+ &&&&&&&& \\[-0.05in]
+%%%% DDA:ControlledMix
+\multirow{5}{*}{\rotatebox[origin=c]{90}{Known}}  \multirow{5}{*}{\rotatebox[origin=c]{90}{Ground}} \multirow{5}{*}{\rotatebox[origin=c]{90}{Truth}} & Computer simulation 1 - Label-free & 2/3/4 & 2/3/5/10 & 1000 & 10 & 10 & \multicolumn{2}{c|}{\href{https://github.com/devonjkohler/MSstatsPTM_simulations}{Github}} \\
+ &&&&&&&& \\%[-0.05in]
+%\hline 
+%&&&&&&&& \\[-0.05in]
+% \cline{2-9}
+% &&&&&&&& \\[-0.05in]
+%%% DDA iPRG2015
+& Computer simulation 2 - Missing and low features & 2/3/4 & 2/3/5/10 & 1000 & 2 & 10 & \multicolumn{2}{c|}{\href{https://github.com/devonjkohler/MSstatsPTM_simulations}{Github}} \\
+ &&&&&&&& \\%[0.05in]
+%\hline 
+% &&&&&& \\[-0.05in]
+%%% DDA Cox2014
+& SpikeIn benchmark - Ubiquitination - Label-free& 4 & 2 & 3198 & 1.37 & 10.17 & \href{https://massive.ucsd.edu/ProteoSAFe/private-dataset.jsp?task=c4c583ecf7f941cdac87f7a4f872517b}{MSV000088971} & \href{https://massive.ucsd.edu/ProteoSAFe/private-dataset.jsp?task=c4c583ecf7f941cdac87f7a4f872517b}{TBD} \\
+ &&&&&&&& \\%[-0.05in]
+\hline
+\multicolumn{6}{c}{ } \\ [0.02in]
+\hline 
+ &&&&&&&& \\%[-0.05in]
+%%%%%%%%%%%%%%%%%%
+%%%% biological study
+%%%%%%%%%%%%%%%%%%
+%% DDA Meierhofer
+\multirow{3}{*}{\rotatebox[origin=c]{90}{Biological}} \multirow{3}{*}{\rotatebox[origin=c]{90}{Experiment}} & Human - Ubiquitination - 1mix-TMT & 6 & 2 & 3550 & 1.21 & 10.75 & \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=b6f0c74c234247678fb0888c6df1f225}{MSV000088966} & \href{https://massive.ucsd.edu/ProteoSAFe/reanalysis_container.jsp?task=b639fa5d0ff24ce78abdca13f4b18c0a}{RMSV000000356}\\
+&&&&&&&& \\%[-0.05in]
+% \cline{2-9}
+% &&&&&&& \\[-0.05in]
+%%%% SRM Picotti 2009
+& Mouse - Phosphorylation - 2mix-TMT& 6 & 4 & 5002 & 1.67 & 10.95 & \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=4878d777c6b34cf8aaf8477e93140c4d}{MSV000085565} & \href{https://massive.ucsd.edu/ProteoSAFe/reanalysis_container.jsp?task=79406d32c1f9413daf7756e6eeb24968}{RMSV000000357}\\
+&&&&&&&& \\%[-0.05in]
+% \cline{2-9}
+% &&&&&&& \\[-0.05in]
+%%%% SRM Surinova 2015, training
+& Human - Ubiquitination - Label-free & 4 & 2 & 3863 & 1.40 & 1.65 & \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=1b516164de5345108b40b75147dd58b5}{MSV000078977} & \href{https://massive.ucsd.edu/ProteoSAFe/reanalysis_container.jsp?task=f9e6117e00794ce5b2e8802601be16bc}{RMSV000000358}\\ [0.02in]
+ &&&&&&&& \\
+ \hline
+\end{tabular}\\
+\end{tiny}
+\caption{ \small {\bf Simulated and experimental datasets.} 
+``Dataset'' is the dataset code name.
+``Data availability'' shows the ID of the MassIVE.quant repository or the GitHub repository.
+``Analysis'' shows the ID of the MassIVE.quant reanalysis container, containing analysis code and modeling results.
+All the experiments were conducted in data-dependent acquisition (DDA) mode.
+\todo{Could you add to the table the number of proteins, and the average number of modified/unmodified peptides per protein} \todo{If the figures are at the end, I think the table should also be at the end}
+\todo{Could you check the table and the entire section: when we say 'number of biological replicates', we typically mean per condition.
+But it seems that some descriptions imply the total number of biological samples across conditions.
+Could you clarify that we use number of biological replicates per condition throughout?
+(E.g, this helps calculate the degrees of freedom in balanced designs)}
+}
+\label{tab:dataDescription}
+\end{table}
+
 \newpage
@@
 The expected fold change is calculated using the values of the spike-in and human lysate in part (a).\todo{Could you make the columns in the table more complete?
-True fold-fold changes between the backgrounds; unadjusted true fold changes, adjusted true fold changes - essentially any value for any comparison that we may want to evaluate}
- %
- \todo{It's silly but...
-could you make the legend wide rather than long?
-I.e., have the colors horizontally, and in the same order as in the figure.
-And maybe use 2 lines for the text to make the widths of the bars in the legend more comparable with the widths of the bars in the text.
-This will help readability and save space}
-%
-\todo{Also, why is C between A and B? The letters have no meaning, it would help to have them in order.}
-}
+True fold-fold changes between the backgrounds; unadjusted true fold changes, adjusted true fold changes - essentially any value for any comparison that we may want to evaluate}}
+
 \label{fig:benchmark-design}
@@
 \caption{The workflow of $MSstatsPTM$ and how it fits into the experimental pipeline.
+The name of the R functions used for each step are highlighted in red.
 The package's workflow starts after modified and unmodified peptide quantification.
@@
 \caption{Dataset 1 \& 2: Computer simulation.
-\todo{OV: in (a) could you use the same color scheme for the methods as in the remaining figures?
-Use different fill/opacity/shade of a same color for methods with and without the adjustment} a) All the considered methods in the first computer simulation correctly calibrated FDR when adjusting for changes in protein abundance.
+a) All the considered methods in the first computer simulation correctly calibrated FDR when adjusting for changes in protein abundance.
 In comparison, the methods without accounting for the protein-level changes resulted in off-target, high false positive rates.
@@
 \caption{Dataset 4: Human - Ubiquitination - 1mix-TMT.
-a) The overlap of differential modified peptides for the PTM model with and without global protein level adjustment.
+a) The overlap of differential modified peptides for the PTM model with and without global protein level adjustment across all comparisons.
 More PTMs became insignificant after adjustment than became significant.
@@
 	 \end{subfigure}
-\caption{Dataset 5: Mouse - Phosphorylation - 2mix-TMT time series.
-\todo{OV: What are the vertical panels 1 and 2 in (b)??
-Are they bio reps?
-Could you clarify?
-I would remove the work 'BioReplicate' at the bottom, it is not clear what it refers to} a) The overlap of differentially modified peptides between the PTM model with and without global protein level adjustment \todo{OV: in which pairwise comparison?
-Across all the comparisons?}.
+\caption{Dataset 5: Mouse - Phosphorylation - 2mix-TMT.
+a) The overlap of differentially modified peptides between the PTM model with and without global protein level adjustment across all comparisons.
 Again more PTMs became insignificant after adjustment then became significant.
@@
  \caption{Dataset 6: Human - Ubiquitination - Label-free no global profiling run.
-a) The overlap of differential modified peptides for the PTM model with and without global protein level adjustment.
+a) The overlap of differential modified peptides for the PTM model with and without global protein level adjustment across all comparisons.
 More PTMs became insignificant than became significant after adjustment.

### 2022-06-15 supplementary/ptm_sm.tex
changes to spike in sections
addressed some comments

@@
 \pagestyle{fancy}
-\lhead{\footnotesize \parbox{11cm}{Supplementary }}
+\lhead{\footnotesize \parbox{11cm}{Supplementary}}
 \headsep = 25pt
@@
 \affil[*]{Corresponding Authors}
+
+\date{}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Experimental datasets in this manuscript}
-
+\section{Experimental datasets}
+\label{sec:experiments}
 \subsection{Experiments with known ground truth}
@@
 \clearpage
-\subsubsection{Dataset 5 : Mouse - Phosphorylation - 2mix-TMT time series}
+\subsubsection{Dataset 5 : Mouse - Phosphorylation - 2mix-TMT}
 \label{sec:shigella}
@@
 The abundance of total protein and post-translation modifications were quantified at three time points, uninfected, early infection (45-60 minutes), and late infection (3-3.5 hours).
-The experiment was modeled as a time series, with the same biological replicate being measured at each time point.
 Quantifying the total protein along with the post-translational modifications allowed us to adjust for changes in total protein and see the true impact of the site specific modifications.
 Two mixtures using 11-plex were ran over the six conditions.
-The six conditions were split between 11 channels leading to the experimental design being unbalanced.
+The six conditions were split between 11 channels leading to an unbalanced experimental design.
 Each mixture contained two replicates per early and late WT and KO conditions.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\clearpage
-
-\printbibliography
+%\clearpage
+
+%\printbibliography
 

### 2022-06-15 code/IpaH_analysis.R

@@ -16,8 +16,23 @@ protein_input$Abundance <- log2(protein_input$Intensity)
 ptm_summarization <- read.csv("../data/Ipah_KGG_TMT11_corrected_msstats_quant_combined_results.txt",
                               sep = "\t")
 summarized_model <- ptm_summarization %>% filter(PeptideSequence == "Model")
+ptm_summarization <- ptm_summarization %>% filter(PeptideSequence != "Model")
 protein_summarization <- read.csv("../data/Ipah_Global_TMT11_msstats_quant_results.txt",
                                   sep = "\t")
+
+protein_summarization = read.csv("../data/Ipah_Global_Input.txt",
+         sep = "\t")
+
+ptm_summarization %>% head()
+ptm_summarization$global_prot = sapply(ptm_summarization$ProteinName, function(x){str_split(x, "\\|")[[1]][1]})
+
+ptm_summarization$global_prot %>% n_distinct()
+ptm_input$ProteinName %>% n_distinct()
+ptm_input$PSM %>% n_distinct()
+
+protein_summarization$ProteinName %>% n_distinct()
+protein_summarization$PeptideSequence %>% n_distinct()
+
 
 protein_summarization$PeptideSequence <- "Model"
 protein_summarization$FeatureType <- "Model"

### 2022-06-15 code/PTMsimulateExperiment.R

@@ -1,0 +1,178 @@
+#' Simulate PTM quantification experiments
+#'
+#' \code{PTMsimulateExperiment} simulates a PTM quantification experiment with
+#' a list of log2-intensities of multiple spectral features, PTM sites and the
+#' corresponding proteins, in multiple MS runs and conditions.
+#'
+#' @param nGroup An integer to specify the number of conditions.
+#' @param nRep An integer to specify the number of replicates per condition.
+#' @param nProtein An integer to specify the number of protein.
+#' @param nSite An integer to specify the number of PTM sites per protein.
+#' @param nFeature An integer to specify the number of features per site.
+#' @param logAbundance A list of two lists named \code{PTM} and \code{PROTEIN}.
+#'   Each contains four elements: \code{mu} (a numeric representing the overall
+#'   mean log2-abundance), \code{delta} (a numeric vector for the deviation of
+#'   the mean log2-abundance for each group from the overall mean), \code{sRep}
+#'   (a numeric representing the standard deviation for run-to-run variation),
+#'   and \code{sPeak} (a numeric representing the standard deviation in peak
+#'   log2-intensities).
+#'
+#' @return A tibble with columns of \code{protein}, \code{site}, \code{group},
+#'   \code{run}, \code{feature}, \code{log2inty}.
+#'
+#' @examples
+#' PTMsimulateExperiment(
+#'     nGroup=2, nRep=2, nProtein=1, nSite=1, nFeature=5,
+#'     logAbundance=list(
+#'         PTM=list(mu=25, delta=c(0, 1), sRep=0.2, sPeak=0.05),
+#'         PROTEIN=list(mu=25, delta=c(0, 1), sRep=0.2, sPeak=0.05)
+#'     )
+#' )
+#'
+#' @export
+PTMsimulateExperiment <- function(nGroup, nRep, nProtein, nSite, nFeature, nFeature_prot,
+                                  logAbundance) {
+  
+  # PTM data
+  peaks <- vector("list", nProtein)
+  for (i in seq_len(nProtein)) {
+    peaks[[i]] <- simulateSites(
+      nGroup, nRep, nSite, nFeature,
+      logAbundance$PTM$mu, logAbundance$PTM$delta,
+      logAbundance$PTM$sRep, logAbundance$PTM$sPeak
+    )
+  }
+  sites <- bind_rows(peaks)
+  sites$protein <- rep(
[…]

### 2022-06-15 code/Shigella_Analysis.R

@@ -26,9 +26,9 @@ py_summarization <- dataSummarizationPTM_TMT(pY_input)
 pST_summarization <- dataSummarizationPTM_TMT(pST_input)
 
 save(py_summarization, file = "../data/Shigella_pY_Summarization.rda")
-load(file = "../data/Shigella_pY_Summarization.rda")
+load(file = "../data/pY_Summarization.rda")
 save(pST_summarization, file = "../data/Shigella_pST_Summarization.rda")
-load(file = "../data/Shigella_pST_Summarization.rda")
+load(file = "../data/pST_Summarization.rda")
 
 py_summarized_data <- py_summarization$PTM$ProteinLevelData
 py_feature_data <- py_summarization$PTM$FeatureLevelData
@@ -38,7 +38,6 @@ pst_feature_data <- pST_summarization$PTM$FeatureLevelData
 
 global_summarized_data <- pST_summarization$PROTEIN$ProteinLevelData
 global_feature_data <- pST_summarization$PROTEIN$FeatureLevelData
-
 
 ## Create pYST dataset
 ## Create pY/pST Dataset
@@ -63,6 +62,22 @@ data_Y_feature <- py_feature_data %>% filter(!ProteinName %in% shared_ST)
 
 ## combine pST and pY
 pSTY_features <- rbind(data_ST_feature, data_Y_feature)
+pyst_sum_input = list(PTM = list(ProteinLevelData = pSTY_sum_msstats,
+                                 FeatureLevelData = pSTY_features),
+                      PROTEIN = list(ProteinLevelData = global_summarized_data,
+                                     FeatureLevelData = global_feature_data)
+                      )
+
+
+## Calculate Peptide and Protein numbers
+pSTY_features$global_prot = sapply(pSTY_features$ProteinName, function(x){str_split(x, "\\|")[[1]][1]})
+
+pSTY_features$global_prot %>% n_distinct()
+pSTY_features$ProteinName %>% n_distinct()
+pSTY_features$PSM %>% n_distinct()
+
+global_feature_data$ProteinName %>% n_distinct()
+global_feature_data$PeptideSequence %>% n_distinct()
 
 ## Plot some interesting PTMs --------------------------------------------------
 ## TTP_MOUSE
@@ -98,7 +113,7 @@ test <- rbindlist(list(temp_plot1, pSTY_sum_msstats %>% filter(
 
[…]

### 2022-06-15 code/Simulated_Analysis.R

@@ -52,7 +52,8 @@ run_ttest <- function(summarized_data){
     ## Loop over groups
     for (g in 1:(length(groups)-1)){
       for (g2 in 2:length(groups)){
-        tryCatch({ttest_ptm <- t.test((temp_joined %>% filter(Condition == groups[g]) %>% select(Abundance.x))[[1]],
+        tryCatch({
+        ttest_ptm <- t.test((temp_joined %>% filter(Condition == groups[g]) %>% select(Abundance.x))[[1]],
                                       (temp_joined %>% filter(Condition == groups[g2]) %>% select(Abundance.x))[[1]])
         ttest_temp <- rbindlist(list(ttest_temp, data.table(ptm = ptms[[p]],
                                                             label = paste(groups[g], "vs", groups[g2], sep = " "),
@@ -194,19 +195,19 @@ for (row in seq_len(nrow(param_combos))){
 
   ## Sim
   sim <- PTMsimulateExperiment(
-    nGroup=param_combos[row, 3], nRep=param_combos[row, 2], nProtein=500, nSite=1, nFeature=10, nFeature_prot = 10,
+    nGroup=param_combos[row, 3], nRep=param_combos[row, 2], nProtein=500, nSite=1, nFeature=3, nFeature_prot = 10,
     logAbundance=list(
       PTM=list(mu=25, delta = del_arr, sRep=param_combos[row, 1], sPeak=.25),#0.05),
       PROTEIN=list(mu=25, delta = del_arr_no_change, sRep=param_combos[row, 1], sPeak=0.25))
   )
   sim_no_change1 <- PTMsimulateExperiment(
-    nGroup=param_combos[row, 3], nRep=param_combos[row, 2], nProtein=250, nSite=1, nFeature=10, nFeature_prot = 10,
+    nGroup=param_combos[row, 3], nRep=param_combos[row, 2], nProtein=250, nSite=1, nFeature=3, nFeature_prot = 10,
     logAbundance=list(
       PTM=list(mu=25, delta = del_arr, sRep=param_combos[row, 1], sPeak=0.25),
       PROTEIN=list(mu=25, delta = del_arr, sRep=param_combos[row, 1], sPeak=0.25))
   )
   sim_no_change2 <- PTMsimulateExperiment(
[…]

### 2022-06-15 code/USP30_MSstatsPTM_Analysis.R

@@ -6,8 +6,18 @@ library(gridExtra)
 
 setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
 
-load(file = "../data/ups30_msstatsptm_model.rda")#model_ptm
-load(file = "../data/ups30_msstatsptm_summary.rda")#summarized_ptm
+load(file = "../data/MSstatsPTM_USP30_model.rda")#model_ptm
+load(file = "../data/MSstatsPTM_USP30_summary.rda")#summarized_ptm
+
+
+## Calc # Prot and # pep
+summarized_ptm$PTM$FeatureLevelData$global_prot = sapply(summarized_ptm$PTM$FeatureLevelData$PROTEIN, function(x){str_split(x, "_")[[1]][1]})
+summarized_ptm$PTM$FeatureLevelData$global_prot %>% n_distinct()
+summarized_ptm$PTM$FeatureLevelData$PROTEIN %>% n_distinct()
+summarized_ptm$PTM$FeatureLevelData$PEPTIDE %>% n_distinct()
+
+summarized_ptm$PROTEIN$FeatureLevelData$PROTEIN %>% n_distinct()
+summarized_ptm$PROTEIN$FeatureLevelData$FEATURE  %>% n_distinct()
 
 ## SE analyis
 pyst_model

### 2022-06-15 main/MSstatsPTM_Main.tex

@@
 
-Consider a label-free LC-MS/MS experiment in the special case of a balanced design with $I$ conditions and $J$ biological replicates (mass spectrometry runs) per condition.
+Consider a label-free LC-MS/MS experiment in the special case of a balanced design with $I$ conditions and $J$ biological replicates per condition.
+For simplicity, we assume that the experiment has no technical replicates, such that each biological replicate corresponds to an LC-MS/MS run.
 For one protein, the PTM site is represented by $K$ spectral features (peptide ions, distinguished by their cleavage residues and charge states).
@@
 The features can be quantified as part of a same mass spectrometry run, or in separate enrichment and global proteome profiling runs.
-
-Figure \ref{fig:data-structure}(a) schematically illustrates this data structure for one PTM and the corresponding unmodified features of the protein in a special case of 2 conditions and 2 biological replicates per condition.
-The number of quantified features varies across replicate LC-MS/MS runs of the same sample, and across conditions.
-There may be feature outliers which could be caused by random variation or measurement error.
-Additionally, some spectral features can be missing.
-The features in each mass spectrometry run are generally summarized and represent a single observation used to estimate the relative abundance of the PTM in condition $i$, $\mu_i^{\ast}$.
-
-The population parameter we are interested in estimating is the difference between the PTM in condition $1$ and condition $2$, denoted by $\mu_1^{\ast}$ and $\mu_2^{\ast}$ respectively.
-To calculate the quantity of interest, statistical analysis needs to summarize values in the table in Figure \ref{fig:data-structure}(a) using appropriate models.
-For condition $i$, the models provide estimates for the mean abundance, $\hat{\mu}_i^{\ast}$, and standard error, $\widehat{SE(\mu_i^{\ast})}$.
-Once summarized, we can formulate the Null and Alternative Hypotheses as seen in Equation \ref{eq:conv_null_hyp}.
-
+Figure \ref{fig:data-structure}(a) schematically illustrates this data structure for one PTM site and the corresponding unmodified features of the protein, in a special case of 2 conditions and 2 biological replicates per condition \todo{OV: replace 'run' with 'bio rep' in the figure}.
+The number of modified and unmodified features varies across proteins.
+Some log-intensities may be outliers, and some spectral features can be missing.
+The population parameter of interest is the difference between the abundances of a PTM site in condition $1$ and condition $2$, denoted by $\mu_1^{\ast}$ and $\mu_2^{\ast}$ respectively.
+We are interested in testing the null hypothesis
 \begin{equation}
@@
 \end{equation}
-
-However, these population parameters are inherently convoluted with changes in the unmodified protein.
-To account for this, we must estimate the relative abundance of the unmodified protein in conditions $1$ and $2$, denoted by $\mu_1$ and $\mu_2$.
-We estimate these quantities in the same way as for the PTM, using their corresponding features and observations, resulting in an estimated mean abundance, $\hat{\mu}_i$, and standard error, $\widehat{SE(\hat{\mu}_i)}$.
-These estimates are used to adjust the quantities estimated for the PTM, resulting in the new Null and Alternative Hypotheses seen in Equation \ref{eq:null_hyp}.
+To test the hypothesis, a statistical model must represent the values in the table in Figure \ref{fig:data-structure}(a), and provide a model-based estimate of $\mu_{1}^{\ast} - \mu_{2}^{\ast}$, and its standard error.
+
+Unfortunately, the population parameter is inherently convoluted with the overall changes in protein abundance. To account for this, we can change the null hypothesis to
 
@@
 \end{equation}
+where $\mu_1$ and $\mu_2$ reflect the overall protein abundances in condition $1$ and condition $2$. These quantities can be estimated using protein features without the modification site.
 
 %---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------
-
-\subsection*{Existing statistical methods for experiments targeting PTMs}
-
-\noindent {\bf Two-sample $t$-test} Many investigations perform differential abundance analysis of PTMs using two-sample $t$-test or its extensions~\cite{Schwammle2015}\cite{THOMAS2020}.
-While simple, these approaches do not fully account for all sources of variations, and are not directly applicable to experiments with complex designs.
-These include experiments with comparisons of multiple conditions, acquisition in multiple batches, and unbalanced designs.
-Additionally, while these approaches can be applied to experiments studying PTMs, there is not a self contained, straightforward implementation of the methods, making application challenging.
-
-When modeling the simple experiment seen in Figure \ref{fig:data-structure}, we are interested in answering the Null Hypothesis in Equation \ref{eq:null_hyp}.
-To estimate the parameters of interest in the trivial model, the approach takes as input intensities of individual features from modified peptides, or intensity ratios of modified and unmodified peptide features, and compares the mean abundance of a PTM site from one condition to another~\cite{Schwammle2015}.
+\subsection*{Existing statistical methods for detecting differentially abundant PTMs}
+
+\noindent {\bf Two-sample $t$-test} 
+\todo{OV: could you help me with this part a little bit?
+I tried to organize the text, but could not quite finish.
+We should have the following order: (1) input, (2) model for the input, (3) model-based inference, (4) implementation; (5) good things and limitations.
+I think it may help to separate two versions: based on the features, and based on the ratios.
+Otherwise it is not very clear}
+Many investigations perform differential abundance analysis of PTMs using two-sample $t$-test or its extensions.
+$t$-test takes as input intensities of individual features from modified peptides, or intensity ratios of modified and unmodified peptide features \todo{OV: let's separate the intensities and the ratios}, and compares the mean abundance of a PTM site from one condition to another~\cite{Schwammle2015}\cite{THOMAS2020}.
 The abundance in each run is taken as input and is often estimated by sum of peak intensities.
 For example, the log-abundance estimate for the PTM in Run $j$ of Condition $i$ is given by Equation \ref{eq:feature_estimation}.
+as input the adjusted PTM abundance of each condition and biological replicate, and estimates $\Delta$ in Equation \ref{eq:null_hyp}.
+\todo{OV: what is the model?}
+To estimate the parameters of interest in this trivial model, the approach takes as input
 
@@
 
-The proposed approach leverages the large amount of work done to create $MSstats$~\cite{Choi:2014}.
-$MSstats$ was designed for protein significance analysis and statistical relative quantification of proteins and peptides in global, targeted and data-independent proteomics.
+\todo{OV: could you also re-organize: (1) input, (2) model for the input, (3) model-based inference, (4) implementation, (5) good things and limitations.
+Here good things are: complex designs, including multiple variance components; EB (for TMT); good summarization; implementation.
+Limitation: MSstats is not designed for protein-level adjustment of changes in PTM}
+
+$MSstats$~\cite{Choi:2014} is designed for protein significance analysis and statistical relative quantification of proteins and peptides in global, targeted and data-independent proteomics.
 In terms of Figure \ref{fig:data-structure}, $MSstats$ was designed to estimate the difference in the unmodified protein in condition $1$ and condition $2$, denoted by $\mu_1$ and $\mu_2$ respectively.
-In other words it allows us to answer the Null Hypothesis seen in Equation \ref{eq:msstats_null_hyp}.
+In other words it allows us to answer the new Null Hypothesis seen in Equation \ref{eq:msstats_null_hyp}.
 
@@
 \caption{
-Goals of PTM characterization, input to statistical analyses, and notation.
+\todo{In 2a, it may be more meaningful to replace 'run' with 'bio rep'} Goals of PTM characterization, input to statistical analyses, and notation.
 (a) Schematic data representation, in a simplified case of two conditions and two biological replicates.
 Each PTM site is modeled and characterized separately, where a PTM is quantified with observed spectral features (boxes), distinguished by different charge states of a peptide.
-The feature intensities of the PTM in Condition 1 are denoted by $y_1^\ast$.
-Features corresponding to unmodified peptides in Condition 1 are denoted by $y_1$.
+The feature \todo{log-}intensities of the PTM in Condition 1 are denoted by $y_1^\ast$ \todo{In the figure $y^*$ has 3 subscripts.
+Could you edit here as well?}.
+\todo{Log-intensities of } features corresponding to unmodified peptides in Condition 1 are denoted by $y_1$ \todo{In the figure $y$ has 3 subscripts.
+Could you edit here as well?}.
 Peptides can be fully cleaved (solid lines) and/or partially cleaved (dashed lines).
-(b) PTM relative quantification of the population parameters by statistical inference, which makes use of the feature intensities to infer the underlying PTM abundance and protein abundance with an estimate of associated uncertainty.
+\todo{OV: I am wondering whether we need Fig 2(b).
+The null hypothesis is specified in the text.
+I am not sure it is referenced anywhere.
+Maybe remove it for simplicity?} (b) PTM relative quantification of the population parameters by statistical inference, which makes use of the feature intensities to infer the underlying PTM abundance and protein abundance with an estimate of associated uncertainty.
 (c) Model-based testing for differential PTM abundance, which corrects for the underlying protein abundance with a cost of increased uncertainty about the estimate of difference between conditions.}

### 2022-06-15 main/MSstatsPTM_Main.tex
Rewrote statistical results sections

@@
 
-We propose a statistical method for detecting changes in PTMs that allows us to answer the Null Hypothesis stated in Equation \ref{eq:null_hyp}, i.e.
-that there is no difference in the log-abundance of the PTM site between conditions, adjusted with respect to unmodified protein abundance.
-We leverage the methods used in $MSstats$ to answer this hypothesis, allowing the methods to be directly applicable to experiments with complex designs.
-Relating to the example in Figure \ref{fig:data-structure}, the expected value of the log-abundance of the PTM and unmodified protein in Condition $1$ are denoted by $\hat{\mu}_{1}^{\ast}$ and $\hat{\mu}_{1}$ respectively.
-The feature and run level summarization for Condition $1$ is denoted as $\hat{y}_{1++}^{\ast}$ for the modified peptide and $\hat{y}_{1++}$ for the unmodified peptide.
-
-To calculate $\hat{y}_{1++}^{\ast}$ and $\hat{y}_{1++}$, feature and run level summarization is carried out as in the sub-plot model of MSstats using Tukey's median polish.
-The summarized values for the PTM and unmodified protein are then used to fit two models based on the family of linear mixed-effects models.
-Finally, the adjusted difference, $\hat{\Delta}$, is given by the difference in log-abundance of the PTM site, subtracted by the difference in log-abundance of the underlying protein, which is equivalent to the log of the ratio of PTM abundance difference to protein abundance difference.
+We propose a statistical method for detecting changes in PTMs that allows us to remove confounding with changes in the global protein, while being directly applicable to experiments with complex designs.
+To do this we leverage the existing methods in $MSstats$ and extend them to remove confounding between changes in the PTM and global protein.
+The method takes as input the spectral features of the PTM, $y_{ijk}^{\ast}$, and it's corresponding global protein, $y_{ijk}$.
+The features are summarized separately using the methods described previously for $MSstats$, resulting in run level values $\hat{y}_{1++}^{\ast}$ for the modified peptide and $\hat{y}_{1++}$ for the unmodified peptide.
+These summarized values are then used as the input to our models.
+Two separate linear mixed-effects models are fit, one using the summarized PTM values and another using the summarized global protein values.
+The models give estimates of the difference in log-abundance of the underlying PTM, $\hat{\Delta}_{PTM}$, and global protein, $\hat{\Delta}_{protein}$.
+Finally, the adjusted difference, $\hat{\Delta}$, is given by the difference in log-abundance of the PTM site, subtracted by the difference in log-abundance of the underlying protein.
+
+\begin{equation}
+\hat{\Delta} = \hat{\Delta}_{PTM} - \hat{\Delta}_{protein} 
+\end{equation}
+
+In the simple case of a balanced design with one source of variation, this is equivalent to the log of the ratio of PTM abundance difference to protein abundance difference.
 
@@
 
-Finally the degrees of freedom are calculated as follows.
+Finally the degrees of freedom are calculated as follows
 
@@
 
+The results of this model are used to answer the Null Hypothesis stated in Equation \ref{eq:null_hyp}.
 The test statistic for the hypothesis testing is the ratio of the estimate of the adjusted difference to its SE.
@@
 
-The proposed approach can be extended to complex designs, including experiments with additional sources of variation and unbalanced designs.
-In this case our null hypothesis in Equation \ref{eq:null_hyp} is unchanged.
-
-Model inference is done in the same way to how MSstats~\cite{Choi:2014} and MSstatsTMT~\cite{Huang:2020} target the global protein.
-First, run and feature level summarization is performed separately for both the PTM and unmodified protein using Tukey's Median Polish.
-Then a separate linear mixed effects model is fit to both the PTM and unmodified protein.
+As discussed above, the proposed approach can be extended to complex designs, including experiments with multiple sources of variation and unbalanced designs.
+Additionally, the approach can model experiments with separate designs for the PTM and global profiling run, i.e.
+differing number of runs and biological replicates, as long as the conditions of interest are the same.
+This extension is achieved through using the methods in $MSstats$, which can account for multiple sources of variation, and fully modeling both the PTM and global protein before adjustment.
+In terms of the workflow, first run level summarization is performed separately for both the PTM and unmodified protein using Tukey's Median Polish.
+Then a linear mixed effects model is fit to both the PTM and unmodified protein summarized features.
 The model takes into account all potential sources of variation, and restricted maximum likelihood (RML) is used to estimate the parameters of each model.
 Finally, once the parameters are estimated, we can combine the models using modified versions of Equations \ref{eq:prop_fc}, \ref{eq:prop_se}, and \ref{eq:prop_df}.
+Modeling the PTM and global protein separately and then combining them removes the need to match runs for adjustment, allowing us to model experiments with differing designs.
 
@@
 Comparing the IQR of the spike-in peptide's log fold change, the proposed method's IQR was 36.78\% smaller than Limma's and 32.98\% smaller than $t$-test's.
-This means that the proposed method estimated the correct fold change more consistently, compared to Limma and $t$-test.
+While Limma and $t$-test generally estimated the correct fold change, the proposed method's estimation was more consistent across all modifications.
 

### 2022-06-16 main/MSstatsPTM_Main.tex
Updated t-test and limma methods

@@
 \noindent {\bf Two-sample $t$-test} 
-\todo{OV: could you help me with this part a little bit?
-I tried to organize the text, but could not quite finish.
-We should have the following order: (1) input, (2) model for the input, (3) model-based inference, (4) implementation; (5) good things and limitations.
-I think it may help to separate two versions: based on the features, and based on the ratios.
-Otherwise it is not very clear}
+\todo{DK: I mainly explained the ratio version.
+I briefly mentioned the feature version, which is used to model ptms w/o adjustment.
+We could reference the versions back to hypotheses in Equation 1 and 2 but I didnt yet.}
+
 Many investigations perform differential abundance analysis of PTMs using two-sample $t$-test or its extensions.
-$t$-test takes as input intensities of individual features from modified peptides, or intensity ratios of modified and unmodified peptide features \todo{OV: let's separate the intensities and the ratios}, and compares the mean abundance of a PTM site from one condition to another~\cite{Schwammle2015}\cite{THOMAS2020}.
-The abundance in each run is taken as input and is often estimated by sum of peak intensities.
-For example, the log-abundance estimate for the PTM in Run $j$ of Condition $i$ is given by Equation \ref{eq:feature_estimation}.
-as input the adjusted PTM abundance of each condition and biological replicate, and estimates $\Delta$ in Equation \ref{eq:null_hyp}.
-\todo{OV: what is the model?}
-To estimate the parameters of interest in this trivial model, the approach takes as input
-
-\begin{equation}
-\begin{aligned}
-\log \left( \sum_{k=1}^{K} 2^{y_{ijk}^{\ast}} \right).
-\end{aligned}
-\label{eq:feature_estimation}
-\end{equation}
-
-For adjustment with respect to unmodified peptides, the estimate of PTM abundance is divided by the protein abundance estimate, which is equivalent to take the difference of their log abundances.
+$t$-test can be performed without accounting for confounding with global protein abundance by taking as input intensities of individual features from modified peptides, denoted by $y_{ijk}^{\ast}$.
+Additionally, it can be extended to remove confounding by taking as input intensity ratios of modified and unmodified peptide features and comparing the mean abundance of a PTM site from one condition to another~\cite{Schwammle2015}\cite{THOMAS2020}\cite{Mertins:2013}.
+This is done by taking the sum of peak intensities in each run for modified and unmodified features separately and dividing the estimate of PTM abundance by the protein abundance estimate.
+This is equivalent to taking the difference of their log abundances.
 For condition $i$ and biological replicate $j$, the estimated quantity is denoted by $\hat{u}^{\ast}_{ij}$ and is given by Equation \ref{eq:adj_estimation}.
@@
 
-to estimate the quantity of interest $\hat{\Delta}$ as well as its standard error, $\widehat{SE(\hat{\Delta})}$.
+These estimated ratios are then used as input to the two sample $t$-test, the output of which is the estimate of our quantity of interest $\hat{\Delta}$ as well as the standard error, $\widehat{SE(\hat{\Delta})}$.
 The statistical significance is determined by comparing the test statistic against the $t$ distribution, with degrees of freedom $df=2J-2$ in balanced designs.
+$t$-test assumes that the data follows a normal distribution, sampled from a representative portion of the total population, and that the conditions of interest have equal variance.
 
@@
 These include experiments with comparisons of multiple conditions, acquisition in multiple batches, and unbalanced designs.
-Additionally, while these approaches can be applied to experiments studying PTMs, there is not a self contained, straightforward implementation of the methods, making application challenging.
-
-\medskip \noindent {\bf Limma} Modifications of $t$-test such as moderated $t$-test with Limma were also proposed~\cite{Ritchie_15a}\cite{Zhu}.
+Additionally, because the adjustment is done on the replicate, or run, level, there must be a matching number of replicates in both the modified and unmodified experimental designs.
+If the designs are different, this method is no longer directly applicable.
+Finally, while these approaches can be applied to experiments studying PTMs, there is not a self contained, straightforward implementation of the methods, making application challenging.
+
+\medskip \noindent {\bf Limma} Modifications of $t$-test such as moderated $t$-test with Limma were also proposed~\cite{Ritchie_15a}\cite{Zhu}\cite{Chappell:2021}.
 
@@
 
-To estimate $\hat{\Delta}$ in Figure \ref{fig:data-structure}, feature level summarization and unmodified protein adjustment are performed in the same way as $t$-test, shown in Equations \ref{eq:feature_estimation} and \ref{eq:adj_estimation}.
-Limma then fits a linear model using the same adjusted summarized features as in $t$-test, $\hat{u}^{\ast}_{11}$, $\hat{u}^{\ast}_{12}$, $\hat{u}^{\ast}_{21}$ and $\hat{u}^{\ast}_{22}$, to estimate $\hat{\Delta}$.
-To estimate $\widehat{SE(\hat{\Delta})}$, Limma leverages Empirical Bayes moderation to moderate the model variance of the individual PTM using the shared global variance of all PTMs modeled in the experiment.
+The inputs to Limma are the same as $t$-test.
+The intensities of individual features from modified peptides, $y_{ijk}^{\ast}$, can be used directly (in the presence of confounding with the unmodified protein).
+To remove confounding we transform the input by taking the intensity ratios of modified and unmodified peptide features, $\hat{u}^{\ast}_{ij}$, calculated in the same way as Equation \ref{eq:adj_estimation}.
+Limma then fits a linear model using the transformed inputs to estimate $\hat{\Delta}$.
+To estimate $\widehat{SE(\hat{\Delta})}$, Limma leverages empirical bayes moderation~\cite{Robbins1992} to moderate the model variance of the individual PTM using the shared global variance of all PTMs modeled in the experiment.
 For a single PTM $m$ with an estimated fold change $\hat{\Delta}_m$, the estimated variance is calculated as $s^{adj}_{m} = P(s_{m}| s_\Theta)$ where $s_{m}$ is the variance resulting from the linear model of PTM $m$ and $s_\Theta$ is the global variance across all PTMs.
 The statistical significance is determined using $\hat{\Delta}_m$ and $\widehat{SE(\hat{\Delta})}$ calculated using the adjusted variance $s^{adj}_{m}$.
+
+\todo{DK: @Olga can you expand on the good things/limitations? I think I am missing some points here but not sure what exactly}
+
+Limma provides an alternative that overcomes some of the limitations of $t$-test.
+First, utilizing empirical bayes moderation can increase the sensitivity of the analysis and is especially useful in cases of low replicates.
+Additionally, using a linear model allows for the sharing of variance information across comparisons of interest.
+This means we can more accurately measure the variance as the number of comparisons increases.
+
+While a general improvement over $t$-test, there are still a number of limitations when using Limma.
+First, the method is limited in the types of experimental designs it can be applied to.
+The method is only directly applicable to experiments with at most two variance components, meaning that in more complex experiments it cannot account for all sources of variation.
+Additionally, it has the same issue as $t$-test with requiring the same experimental design for the modified and unmodified features, so that there are matching replicates in the ratio calculation.
+Finally, there is again no self contained implementation of the methods to PTMs, requiring manual transformation and application by the user at every step.
 

### 2022-06-16 main/MSstatsPTM_Main.tex
Updated MSstats and MSstatsPTM stats sections

@@
 The features can be quantified as part of a same mass spectrometry run, or in separate enrichment and global proteome profiling runs.
-Figure \ref{fig:data-structure}(a) schematically illustrates this data structure for one PTM site and the corresponding unmodified features of the protein, in a special case of 2 conditions and 2 biological replicates per condition \todo{OV: replace 'run' with 'bio rep' in the figure}.
+Figure \ref{fig:data-structure}(a) schematically illustrates this data structure for one PTM site and the corresponding unmodified features of the protein, in a special case of 2 conditions and 2 biological replicates per condition.
 The number of modified and unmodified features varies across proteins.
@@
 
-Many investigations perform differential abundance analysis of PTMs using two-sample $t$-test or its extensions.
-$t$-test can be performed without accounting for confounding with global protein abundance by taking as input intensities of individual features from modified peptides, denoted by $y_{ijk}^{\ast}$.
-Additionally, it can be extended to remove confounding by taking as input intensity ratios of modified and unmodified peptide features and comparing the mean abundance of a PTM site from one condition to another~\cite{Schwammle2015}\cite{THOMAS2020}\cite{Mertins:2013}.
+Many investigations perform differential abundance analysis of PTMs using two-sample $t$-test or its extensions~\cite{Schwammle2015}.
+These methods can be performed without accounting for convolution with global protein abundance by taking as input intensities of individual features from modified peptides, denoted by $y_{ijk}^{\ast}$.
+Additionally, they can be extended to remove convolution by taking as input intensity ratios of modified and unmodified peptide features and comparing the mean abundance of a PTM site from one condition to another~\cite{THOMAS2020}\cite{Mertins:2013}.
 This is done by taking the sum of peak intensities in each run for modified and unmodified features separately and dividing the estimate of PTM abundance by the protein abundance estimate.
@@
 
-These estimated ratios are then used as input to the two sample $t$-test, the output of which is the estimate of our quantity of interest $\hat{\Delta}$ as well as the standard error, $\widehat{SE(\hat{\Delta})}$.
+These estimated ratios are then used as input to the two sample $t$-test, the output of which is the estimate of our quantity of interest $\hat{\Delta}$ and the standard error, $\widehat{SE(\hat{\Delta})}$.
 The statistical significance is determined by comparing the test statistic against the $t$ distribution, with degrees of freedom $df=2J-2$ in balanced designs.
@@
 These include experiments with comparisons of multiple conditions, acquisition in multiple batches, and unbalanced designs.
-Additionally, because the adjustment is done on the replicate, or run, level, there must be a matching number of replicates in both the modified and unmodified experimental designs.
+Additionally, because the adjustment is done on the replicate level, there must be a matching number of replicates in both the modified and unmodified experimental designs.
 If the designs are different, this method is no longer directly applicable.
@@
 \medskip \noindent {\bf Limma} Modifications of $t$-test such as moderated $t$-test with Limma were also proposed~\cite{Ritchie_15a}\cite{Zhu}\cite{Chappell:2021}.
-
-\todo{OV: could you also re-organize: (1) input, (2) model for the input, (3) model-based inference, (4) implementation, (5) good things and limitations.
-Limitation: limited in experimental design (at most two variance components, with strong assumptions)}
-
-The inputs to Limma are the same as $t$-test.
+The inputs to the Limma based methods are the same as the inputs to $t$-test.
 The intensities of individual features from modified peptides, $y_{ijk}^{\ast}$, can be used directly (in the presence of confounding with the unmodified protein).
@@
 
-While a general improvement over $t$-test, there are still a number of limitations when using Limma.
+While Limma offers improvements over $t$-test, the method still has a number of limitations.
 First, the method is limited in the types of experimental designs it can be applied to.
 The method is only directly applicable to experiments with at most two variance components, meaning that in more complex experiments it cannot account for all sources of variation.
-Additionally, it has the same issue as $t$-test with requiring the same experimental design for the modified and unmodified features, so that there are matching replicates in the ratio calculation.
-Finally, there is again no self contained implementation of the methods to PTMs, requiring manual transformation and application by the user at every step.
+Additionally, it still requires the same experimental design for the modified and unmodified features so that the ratio calculation can be performed.
+Finally, there is no self contained implementation of the methods to PTMs, requiring manual transformation and application by the user.
 
@@
 
-\todo{OV: could you also re-organize: (1) input, (2) model for the input, (3) model-based inference, (4) implementation, (5) good things and limitations.
-Here good things are: complex designs, including multiple variance components; EB (for TMT); good summarization; implementation.
-Limitation: MSstats is not designed for protein-level adjustment of changes in PTM}
-
 $MSstats$~\cite{Choi:2014} is designed for protein significance analysis and statistical relative quantification of proteins and peptides in global, targeted and data-independent proteomics.
-In terms of Figure \ref{fig:data-structure}, $MSstats$ was designed to estimate the difference in the unmodified protein in condition $1$ and condition $2$, denoted by $\mu_1$ and $\mu_2$ respectively.
-In other words it allows us to answer the new Null Hypothesis seen in Equation \ref{eq:msstats_null_hyp}.
+The package takes as input a list of intensities of individual features, denoted by $y_{ijk}$, used to characterize the identified protein.
+For each protein, the feature log-intensities are first summarized into a single value per protein per run in the subplot model.
+This is done using Tukey's median polish (TMP)~\cite{Tukey:1977}, a simple and robust procedure, which iteratively fits a two-way additive model with the effects of run and feature, summarizing them into a single value per run.
+These summarized values are then used as input to fit a linear model based on the family of linear mixed-effects models \cite{Bolker2009} \cite{Faraway:2006}.
+These methods are implemented in the packages $MSstats$ and $MSstatsTMT$~\cite{Huang:2020} and include functionalities for end to end quantification of proteins and differential analysis between experimental conditions.
+
+The $MSstats$ analysis pipeline has a number advantages over simpler methods.
+First, the approach extends to many types of complex experimental design, including unbalanced designs, and additional sources of variation, such as fractionation.
+The methods include optional implementation of empirical bayes moderation, described previously, when applied to TMT experiments~\cite{Huang:2020}.
+Additionally, the summarization step includes options for missing value imputation, using an accelerated failure time (AFT) model~\cite{Wei:1992}, and has been shown to be more robust compared to using individual features~\cite{Choi:2014}.
+
+While $MSstats$ is a powerful tool for the quantification of the global protein, it does not have built in tools to analyze PTMs.
+It is designed to estimate the difference in a protein between conditions, answering the Null Hypothesis
 
@@
 
-The package can be adjusted to analyze information on the peptide level, allowing us to estimate the convoluted Null Hypothesis in Equation \ref{eq:conv_null_hyp}.
-However, $MSstats$ currently includes no built in method to estimate the deconvoluted Null Hypothesis in Equation \ref{eq:null_hyp}.
-
-$MSstats$ takes as input a list of log-transformed intensities of spectral features which are used to characterize the identified protein.
-For each protein, the feature log-intensities are expressed using a linear mixed model in consideration of the effects of condition, run, feature and interaction between run and feature.
-The model parameters are estimated using a split-plot approach, where the feature log-intensities are first summarized into a single value per protein per run in the subplot model, and the protein-level summaries are then used for the inference of the protein abundance.
-This approach allows the method to be extended to cases where the experimental design is unbalanced and additional sources of variation are present, overcoming the limitations in the previous sections.
-In the protein-level summarization, Tukey's median polish (TMP), a simple and robust procedure, is applied to iteratively fit a two-way additive model with the effects of run and feature, which in turn summarizes the log-intensities for each protein~\cite{Tukey:1977}.
-After summarization, the inference of the protein abundance in each condition is carried through fitting a model based on the family of linear mixed-effects models \cite{Bolker2009} \cite{Faraway:2006}.
-We leverage this statistical modeling and quantification workflow and apply it to PTMs, in addition to unmodified proteins.
+It is not designed to remove confounding with the global protein meaning that it cannot directly address the Null Hypothesis in Equation \ref{eq:null_hyp}.
 
@@
 
-We propose a statistical method for detecting changes in PTMs that allows us to remove confounding with changes in the global protein, while being directly applicable to experiments with complex designs.
+We propose a statistical method for detecting changes in PTMs that allows us to remove confounding with changes in the global protein and is applicable to experiments with complex designs.
 To do this we leverage the existing methods in $MSstats$ and extend them to remove confounding between changes in the PTM and global protein.
@@
 Adjustment for multiple comparisons is performed using the Benjamini-Hochberg procedure to control the false discovery rate at a desired level, e.g., $0.05$~\cite{Benjamini:1995}.
+
+\todo{DK: I'm not sure where this part should be. could move it to the first paragraph of this section, or in the extensions to complex design}
+Finally, this approach allows us to model experiments with separate designs for the PTM and global profiling runs.
+This includes differing number of runs, biological replicates, or acquisition methods.
+This is achieved through fully modeling the PTM and global protein separately before adjusting.
+Because we are not combining on the replicate level, the number of replicates for each group do not need to match for adjustment.
+This makes the proposed method more flexible and applicable to a wider variety of experimental designs.
 
@@
 As discussed above, the proposed approach can be extended to complex designs, including experiments with multiple sources of variation and unbalanced designs.
-Additionally, the approach can model experiments with separate designs for the PTM and global profiling run, i.e.
-differing number of runs and biological replicates, as long as the conditions of interest are the same.
-This extension is achieved through using the methods in $MSstats$, which can account for multiple sources of variation, and fully modeling both the PTM and global protein before adjustment.
-In terms of the workflow, first run level summarization is performed separately for both the PTM and unmodified protein using Tukey's Median Polish.
-Then a linear mixed effects model is fit to both the PTM and unmodified protein summarized features.
-The model takes into account all potential sources of variation, and restricted maximum likelihood (RML) is used to estimate the parameters of each model.
-Finally, once the parameters are estimated, we can combine the models using modified versions of Equations \ref{eq:prop_fc}, \ref{eq:prop_se}, and \ref{eq:prop_df}.
-Modeling the PTM and global protein separately and then combining them removes the need to match runs for adjustment, allowing us to model experiments with differing designs.
+This extension is achieved through utilizing the methods in $MSstats$ which, in complex cases, use restricted maximum likelihood (REML)~\cite{Kenward} to estimate the parameters of the models.
+Once the parameters are estimated, we can adjust the PTM model for changes in the global protein using modified versions of Equations \ref{eq:prop_fc}, \ref{eq:prop_se}, and \ref{eq:prop_df}.
 
 The log-fold change in the adjusted PTM abundance, $\hat{\Delta
-}$, is now estimated by RML, and Equation \ref{eq:prop_fc} becomes 
+}$, is now estimated by REML, and Equation \ref{eq:prop_fc} becomes 
 \begin{equation}
@@
 
-The standard error of the estimate $\mathrm{SE}(\hat{\Delta})$ and degrees of freedom in Equations 3 and 4 are unchanged from the balanced case, although $\hat{\sigma}_{\gamma^\ast}$ and $\hat{\sigma}_{\gamma}$ are now estimated using RML.
+The standard error of the estimate $\mathrm{SE}(\hat{\Delta})$ and degrees of freedom in Equations 3 and 4 are unchanged from the balanced case, although $\hat{\sigma}_{\gamma^\ast}$ and $\hat{\sigma}_{\gamma}$ are now estimated using REML.
 
@@
 \includegraphics[scale=.6]{images/fig3.png}
-\caption{
-\todo{In 2a, it may be more meaningful to replace 'run' with 'bio rep'} Goals of PTM characterization, input to statistical analyses, and notation.
+\caption{Goals of PTM characterization, input to statistical analyses, and notation.
 (a) Schematic data representation, in a simplified case of two conditions and two biological replicates.
 Each PTM site is modeled and characterized separately, where a PTM is quantified with observed spectral features (boxes), distinguished by different charge states of a peptide.
-The feature \todo{log-}intensities of the PTM in Condition 1 are denoted by $y_1^\ast$ \todo{In the figure $y^*$ has 3 subscripts.
-Could you edit here as well?}.
-\todo{Log-intensities of } features corresponding to unmodified peptides in Condition 1 are denoted by $y_1$ \todo{In the figure $y$ has 3 subscripts.
-Could you edit here as well?}.
+The log-intensity of the PTM in Condition i, Run j, and feature k are denoted by $y_{ijk}^{\ast}$.
+Log-intensity of feature corresponding to unmodified peptide in Condition i, Run j, and feature k are denoted by $y_{ijk}$.
 Peptides can be fully cleaved (solid lines) and/or partially cleaved (dashed lines).
-\todo{OV: I am wondering whether we need Fig 2(b).
-The null hypothesis is specified in the text.
-I am not sure it is referenced anywhere.
-Maybe remove it for simplicity?} (b) PTM relative quantification of the population parameters by statistical inference, which makes use of the feature intensities to infer the underlying PTM abundance and protein abundance with an estimate of associated uncertainty.
 (c) Model-based testing for differential PTM abundance, which corrects for the underlying protein abundance with a cost of increased uncertainty about the estimate of difference between conditions.}

### 2022-06-16 main/MSstatsPTM_Main.tex
fixed some typos

@@
 Finally, this approach allows us to model experiments with separate designs for the PTM and global profiling runs.
-This includes differing number of runs, biological replicates, or acquisition methods.
+This includes differing number of runs, biological replicates, or types of acquisition methods.
 This is achieved through fully modeling the PTM and global protein separately before adjusting.
@@
 
-The proposed methods are implemented in the open source R package $MSstatsPTM$, available on Bioconductor and Github.
+The proposed methods are implemented in the open source R package $MSstatsPTM$.
 An overview of the steps of the package are illustrated in Figure \ref{fig:msstatsptm_design}.
@@
 If the global profiling information is not available, the package can still analyze the modified run, but will do so without adjusting for changes in unmodified protein abundance.
-Specifically, this will reduce our null hypotheses to Equation \ref{eq:conv_null_hyp}; which amounts to running the methods seen in $MSstats$ on modified peptide level data \cite{Choi:2014} \cite{Huang:2020}.
+This will reduce our null hypotheses to Equation \ref{eq:conv_null_hyp}; which amounts to running the methods seen in $MSstats$ on modified peptide level data \cite{Choi:2014} \cite{Huang:2020}.
 
@@
 Additionally, the summarization includes global median normalization and normalizes between MS runs.
-The package can optionally impute missing values using an accelerated failure time (AFT) model~\cite{Wei:1992}.
+The package can optionally impute missing values using an AFT model~\cite{Wei:1992}.
 Missing value imputation should be performed with care when applied to PTM data.
@@
 Beyond the core functionality, the package also includes functions for plotting the results.
-These include summarization plots to help with quality assurance and identifying sources of variation (such as quality control plots and profile plots), as well as modeling plots (such as volcano plots and heatmaps), which evaluates the fold change between conditions and peptides.
+These include summarization plots to help with quality assurance and identifying sources of variation (such as quality control plots and profile plots), as well as modeling plots (such as volcano plots and heatmaps), which evaluate the fold change between conditions and peptides.
 

### 2022-06-20 main/MSstatsPTM_Main.tex
general rewording and updated table

@@
 
-Additional details on every dataset are available in Supplementary Section 2.
+Additional details on all experiments are available in Supplementary Section 2.
 
@@
 \medskip \noindent {\bf Evaluation:} We evaluated the ability of the statistical methods to correctly detect which PTMs were simulated with changes between conditions and which were not.
-We evaluated the methods ability to avoid false positives (i.e.
+We gauged the methods ability to avoid false positives (i.e.
 specificity), accurately estimate the fold change between conditions, and analyzed the sensitivity of detecting differentially abundant PTMs.
-Each method was used to analyze changes in PTM abundance in the presence of confounding with changes in the unmodified protein, and with this confounding removed.
+The evaluation was performed both in the presence of convolution with changes in the unmodified protein and after applying adjustment to correct for convolution.
 
@@
 
-\medskip \noindent {\bf Evaluation:} The methods were evaluated in the same way as above.
+\medskip \noindent {\bf Evaluation:} The methods were evaluated as above.
 We evaluated their ability to correctly detect PTM's specificity, fold change estimation, and sensitivity.
-These statistics were analyzed both in the presence of, and without, confounding with changes in the unmodified protein.
+These statistics were analyzed both in the presence of, and without, convolution with changes in the unmodified protein.
 
@@
 \medskip \noindent {\bf Limma} Modifications of $t$-test such as moderated $t$-test with Limma were also proposed~\cite{Ritchie_15a}\cite{Zhu}\cite{Chappell:2021}.
-The inputs to the Limma based methods are the same as the inputs to $t$-test.
+The inputs to  Limma are the same as the inputs to $t$-test.
 The intensities of individual features from modified peptides, $y_{ijk}^{\ast}$, can be used directly (in the presence of confounding with the unmodified protein).
@@
 
-Limma provides an alternative that overcomes some of the limitations of $t$-test.
+Limma provides an alternative that overcomes limitations of $t$-test.
 First, utilizing empirical bayes moderation can increase the sensitivity of the analysis and is especially useful in cases of low replicates.
@@
 
+\todo{DK: This section could use some rewording}
+
 $MSstats$~\cite{Choi:2014} is designed for protein significance analysis and statistical relative quantification of proteins and peptides in global, targeted and data-independent proteomics.
@@
 These summarized values are then used as input to fit a linear model based on the family of linear mixed-effects models \cite{Bolker2009} \cite{Faraway:2006}.
+The output of this model is the quantity of interest $\hat{\Delta}$ and the standard error, $\widehat{SE(\hat{\Delta})}$.
 These methods are implemented in the packages $MSstats$ and $MSstatsTMT$~\cite{Huang:2020} and include functionalities for end to end quantification of proteins and differential analysis between experimental conditions.
@@
 The $MSstats$ analysis pipeline has a number advantages over simpler methods.
-First, the approach extends to many types of complex experimental design, including unbalanced designs, and additional sources of variation, such as fractionation.
+First, the approach extends to many types of complex experimental design, including unbalanced designs and multiple sources of variation, such as fractionation.
 The methods include optional implementation of empirical bayes moderation, described previously, when applied to TMT experiments~\cite{Huang:2020}.
@@
 We propose a statistical method for detecting changes in PTMs that allows us to remove confounding with changes in the global protein and is applicable to experiments with complex designs.
-To do this we leverage the existing methods in $MSstats$ and extend them to remove confounding between changes in the PTM and global protein.
+To do this, we leverage the existing methods in $MSstats$ and extend them to remove confounding between changes in the PTM and global protein.
 The method takes as input the spectral features of the PTM, $y_{ijk}^{\ast}$, and it's corresponding global protein, $y_{ijk}$.
-The features are summarized separately using the methods described previously for $MSstats$, resulting in run level values $\hat{y}_{1++}^{\ast}$ for the modified peptide and $\hat{y}_{1++}$ for the unmodified peptide.
+The features are summarized separately using the methods described previously for $MSstats$, resulting in run level values for the PTM, $\hat{y}_{1++}^{\ast}$, and unmodified peptide, $\hat{y}_{1++}$.
 These summarized values are then used as the input to our models.
@@
 \todo{DK: I'm not sure where this part should be. could move it to the first paragraph of this section, or in the extensions to complex design}
-Finally, this approach allows us to model experiments with separate designs for the PTM and global profiling runs.
-This includes differing number of runs, biological replicates, or types of acquisition methods.
+This approach allows us to model experiments with separate designs for the PTM and global profiling runs.
+This includes differing number of runs, biological replicates, and/or types of acquisition methods.
 This is achieved through fully modeling the PTM and global protein separately before adjusting.
@@
 The performance of the proposed method was evaluated on simulated and spike-in datasets with known ground truth, as well as biological experimental data where the ground truth was not known.
-For the experiments where ground truth was known, we calculated the number of true positives (TP), false positives (FP), true negatives (TN), and false negatives (FN).
-True positives were modified peptides that were differential after accounting for changes in global protein abundance.
-True negatives were modified peptides that were not differential after accounting for changes in global protein abundance.
+When ground truth was known, we calculated the number of true positives (TP), false positives (FP), true negatives (TN), and false negatives (FN).
+True positives were PTMs that were expected to be differential after accounting for changes in global protein abundance.
+True negatives were PTMs that were not expected to be differential after accounting for changes in global protein abundance.
 To determine positives and negatives we adjusted the p-values for multiple testing and used an $\alpha=.05$ cutoff.
 These results were then compared to the ground truth to determine TP and TN.
-We used these results to calculate the false discovery rate (FDR = FP/(TP+FP)), recall rate (TP/(TP + FN), and accuracy ((TP + TN) / (TP + TN + FP + FN)).
+Summary statistics were calculated, including the false discovery rate (FDR) (FP/(TP+FP)), recall rate (TP/(TP + FN), and accuracy ((TP + TN) / (TP + TN + FP + FN)).
 Finally, the summary statistics were compared across methods to analyze method performance.
@@
 The proposed approach was evaluated against two-sample $t$-test and Limma in computer simulations. 
-In Figure \ref{fig:sim1_fdr} we see that not accounting for changes in the unmodified protein resulted in a high false discovery rate.
+In Figure \ref{fig:sim1_fdr} we see that not accounting for changes in the unmodified protein resulted in a high FDR.
 After adjusting for changes in the unmodified protein, all methods performed similarly in terms of FDR.
@@
 Two-sample $t$-test only uses data within the groups of interest while ignoring the remainder of the data, leading to similar performance across all simulations.
-In contrast, the proposed approach and Limma leverage all available information, which resulted in improved power in simulations with more conditions.
+In contrast, the proposed approach and Limma leverage information across comparisons, resulting in improved power in simulations with more comparisons.
 
@@
 
-In this simulation the proposed method outperformed $t$-test and Limma.
+In simulations with missing values and low features the proposed method outperformed $t$-test and Limma.
 Changes in unmodified protein level still needed to be accounted for to control the FDR.
@@
 Specifically, for the proposed method, the number of false positives went from 20.88\% to 1.84\% after adjustment was applied.
-While the proposed method and Limma both correctly estimated the fold change of the spike-in peptides, Limma resulted in the majority having an insignificant adjusted p-value.
-In this case, using Limma would have resulted in missing the majority of differentially modified peptides.
+While the proposed method and Limma both correctly estimated the fold change of the spike-in peptides, using Limma resulted in many insignificant adjusted p-values.
+Using Limma would have resulted in missing the majority of differential PTMs.
 
@@
 The results of this experiment are summarized in Figure \ref{fig:data4_plots}.
-In Figure \ref{fig:data4_venn_diagram} the number of significant modified peptides before and after adjustment is shown.
-We can see that more peptides became insignificant after adjustment than became significant.
+Figure \ref{fig:data4_venn_diagram} shows the number of significant modified peptides before and after adjustment.
+More peptides became insignificant after adjustment than became significant.
 3,236 modified peptides became insignificant, 1,352 became significant, while 4,282 were significant in both models.
@@
 In contrast, for peptides that became significant after adjustment, their true abundance change was masked by underlying changes in the unmodified protein.
-Both of these issues were corrected in adjustment, and the deconfounded abundance change is shown.
+Both of these issues were corrected in adjustment, and the unconvoluted abundance change is shown.
 An additional question that must be addressed is if the decrease in significant peptides is due to the increased variance that comes from adjustment.
@@
 The results of this experiment are summarized in Figure \ref{fig:data5_plots}.
-In Figure \ref{fig:data5_venn_diagram} the number of significant modified peptides before and after adjustment is shown.
+Figure \ref{fig:data5_venn_diagram} shows the number of significant modified peptides before and after adjustment.
 Again more PTMs became insignificant after adjustment than became significant.
 19,286 peptides became insignificant, 4,947 became significant, and 41,552 were significant in both models.
-Again we tested if the decrease in significant peptides was due to increased variance or if it was due to removing convolution with changes in the umodified protein.
+Again we tested if the decrease in significant peptides was due to removing convolution with changes in the umodified protein or increased variance after adjustment was applied.
 This was tested by looking for modified peptides whose adjusted log fold change was within 10\% of the unadjusted log fold change but became insignificant after adjustment.
@@
 Of the 10,799 ubiquitin sites identified, only 4,526 had a corresponding unmodified protein and could be adjusted.
-Additionally, the lack of a separate global profiling run resulted in low feature counts for the unmodified protein model.
+Additionally, the lack of a separate global profiling run resulted in low feature counts for the unmodified protein model compared to other experiments, seen in Table \ref{tab:dataDescription}.
 
@@
 726 modified peptides became insignificant, 547 became significant, and 1,078 were significant in both models.
-As in the previous experiments, we can check if these peptides became insignificant due to an increase in variance.
-To check this we looked at modified peptides whose adjusted fold change was within 10\% of their unadjusted fold change and which became insignificant after adjustment.
-In this experiment there were only 25 modified peptides that meet this criteria.
+Only 25 PTMs became insignificant due to increased variance from adjustment.
 
@@
 The proposed framework can be extended beyond the experimental designs with variation from multiple sources discussed above.
-Although demonstrated here on DDA, is also applicable to DIA, SRP and PRM acquisitions.
+Although demonstrated here on DDA, is also applicable to DIA, SRM and PRM acquisitions.
 Additionally, the approach can handle experiments with modified peptides processed using label-free methods and unmodified peptides processed using TMT labeling, or vice versa.
@@
 However, this method would likely run into challenges due to sparsity of features for modified peptides with both a single and multiple modification sites.
-A more complex approach to addressing this problem is most likely necessary.
+A more complex approach to addressing this problem is likely necessary.
 
@@
 %%%% colnames of table
-& \multirow{2}{*}{Dataset} & No. of & No. of bio. & No. of & No. of mod. & No. of unmod.&  Data & \multirow{2}{*}{Analysis}\\
-&  & Conditions & replicates & proteins & features/site & peptides/protein & availability & \\
+& \multirow{2}{*}{Dataset} & No. of & No. of bio. & No. of mod. & No. of mod. & No. of unmod.&  Data & \multirow{2}{*}{Analysis}\\
+&  & Conditions & replicates & peptides & features/site & peptides/protein & availability & \\
 \hline
@@
 %%%% DDA:ControlledMix
-\multirow{5}{*}{\rotatebox[origin=c]{90}{Known}}  \multirow{5}{*}{\rotatebox[origin=c]{90}{Ground}} \multirow{5}{*}{\rotatebox[origin=c]{90}{Truth}} & Computer simulation 1 - Label-free & 2/3/4 & 2/3/5/10 & 1000 & 10 & 10 & \multicolumn{2}{c|}{\href{https://github.com/devonjkohler/MSstatsPTM_simulations}{Github}} \\
+\multirow{5}{*}{\rotatebox[origin=c]{90}{Known}}  \multirow{5}{*}{\rotatebox[origin=c]{90}{Ground}} \multirow{5}{*}{\rotatebox[origin=c]{90}{Truth}} & Computer simulation 1 - Label-free & 2/3/4 & 2/3/5/10 & 1,000 & 10 & 10 & \multicolumn{2}{c|}{\href{https://github.com/devonjkohler/MSstatsPTM_simulations}{Github}} \\
  &&&&&&&& \\%[-0.05in]
@@
 %%% DDA iPRG2015
-& Computer simulation 2 - Missing and low features & 2/3/4 & 2/3/5/10 & 1000 & 2 & 10 & \multicolumn{2}{c|}{\href{https://github.com/devonjkohler/MSstatsPTM_simulations}{Github}} \\
+& Computer simulation 2 - Missing and low features & 2/3/4 & 2/3/5/10 & 1,000 & 2 & 10 & \multicolumn{2}{c|}{\href{https://github.com/devonjkohler/MSstatsPTM_simulations}{Github}} \\
  &&&&&&&& \\%[0.05in]
@@
 %%% DDA Cox2014
-& SpikeIn benchmark - Ubiquitination - Label-free& 4 & 2 & 3198 & 1.37 & 10.17 & \href{https://massive.ucsd.edu/ProteoSAFe/private-dataset.jsp?task=c4c583ecf7f941cdac87f7a4f872517b}{MSV000088971} & \href{https://massive.ucsd.edu/ProteoSAFe/private-dataset.jsp?task=c4c583ecf7f941cdac87f7a4f872517b}{TBD} \\
+& SpikeIn benchmark - Ubiquitination - Label-free& 4 & 2 & 12,137 & 1.37 & 10.17 & \href{https://massive.ucsd.edu/ProteoSAFe/private-dataset.jsp?task=c4c583ecf7f941cdac87f7a4f872517b}{MSV000088971} & \href{https://massive.ucsd.edu/ProteoSAFe/private-dataset.jsp?task=c4c583ecf7f941cdac87f7a4f872517b}{TBD} \\
  &&&&&&&& \\%[-0.05in]
@@
 %% DDA Meierhofer
-\multirow{3}{*}{\rotatebox[origin=c]{90}{Biological}} \multirow{3}{*}{\rotatebox[origin=c]{90}{Experiment}} & Human - Ubiquitination - 1mix-TMT & 6 & 2 & 3550 & 1.21 & 10.75 & \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=b6f0c74c234247678fb0888c6df1f225}{MSV000088966} & \href{https://massive.ucsd.edu/ProteoSAFe/reanalysis_container.jsp?task=b639fa5d0ff24ce78abdca13f4b18c0a}{RMSV000000356}\\
+\multirow{3}{*}{\rotatebox[origin=c]{90}{Biological}} \multirow{3}{*}{\rotatebox[origin=c]{90}{Experiment}} & Human - Ubiquitination - 1mix-TMT & 6 & 2/1 & 8,848 & 1.21 & 10.75 & \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=b6f0c74c234247678fb0888c6df1f225}{MSV000088966} & \href{https://massive.ucsd.edu/ProteoSAFe/reanalysis_container.jsp?task=b639fa5d0ff24ce78abdca13f4b18c0a}{RMSV000000356}\\
 &&&&&&&& \\%[-0.05in]
@@
 %%%% SRM Picotti 2009
-& Mouse - Phosphorylation - 2mix-TMT& 6 & 4 & 5002 & 1.67 & 10.95 & \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=4878d777c6b34cf8aaf8477e93140c4d}{MSV000085565} & \href{https://massive.ucsd.edu/ProteoSAFe/reanalysis_container.jsp?task=79406d32c1f9413daf7756e6eeb24968}{RMSV000000357}\\
+& Mouse - Phosphorylation - 2mix-TMT& 6 & 4/3 & 26,433 & 1.67 & 10.95 & \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=4878d777c6b34cf8aaf8477e93140c4d}{MSV000085565} & \href{https://massive.ucsd.edu/ProteoSAFe/reanalysis_container.jsp?task=79406d32c1f9413daf7756e6eeb24968}{RMSV000000357}\\
 &&&&&&&& \\%[-0.05in]
@@
 %%%% SRM Surinova 2015, training
-& Human - Ubiquitination - Label-free & 4 & 2 & 3863 & 1.40 & 1.65 & \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=1b516164de5345108b40b75147dd58b5}{MSV000078977} & \href{https://massive.ucsd.edu/ProteoSAFe/reanalysis_container.jsp?task=f9e6117e00794ce5b2e8802601be16bc}{RMSV000000358}\\ [0.02in]
+& Human - Ubiquitination - Label-free & 4 & 2 & 10,799 & 1.40 & 1.65 & \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=1b516164de5345108b40b75147dd58b5}{MSV000078977} & \href{https://massive.ucsd.edu/ProteoSAFe/reanalysis_container.jsp?task=f9e6117e00794ce5b2e8802601be16bc}{RMSV000000358}\\ [0.02in]
  &&&&&&&& \\
@@
 ``Dataset'' is the dataset code name.
+``No.
+of bio.
+replicates'' shows the number of biological replicates per condition.
+Simulations were generated with different numbers of replicates.
+The design of two Biological Experiments were unbalanced with unequal replicates per condition.
+``No.
+of mod.
+features/site'' is the number of features used to estimate the abundance of a single modification.
+A feature is a peptide and its corresponding charge.
+``No.
+of unmod.
+peptides/protein'' is the number of peptides without modifications that were used to estimate the global protein abundance.
 ``Data availability'' shows the ID of the MassIVE.quant repository or the GitHub repository.
@@
 All the experiments were conducted in data-dependent acquisition (DDA) mode.
-\todo{Could you add to the table the number of proteins, and the average number of modified/unmodified peptides per protein} \todo{If the figures are at the end, I think the table should also be at the end}
-\todo{Could you check the table and the entire section: when we say 'number of biological replicates', we typically mean per condition.
-But it seems that some descriptions imply the total number of biological samples across conditions.
-Could you clarify that we use number of biological replicates per condition throughout?
-(E.g, this helps calculate the degrees of freedom in balanced designs)}
 }
+
 \label{tab:dataDescription}

### 2022-06-21 main/MSstatsPTM_Main.tex
Up to methods

@@
 The datasets illustrate a diverse set of organisms, modification types, acquisition methods and experimental designs, showing the applicability of the framework to a variety of situations.
-By appropriately leveraging the information from the unmodified peptides, the proposed approach improved the accuracy of the estimates of PTM fold changes and produced a better calibrated false positive rate of detecting differentially abundant PTMs as compared to existing methods.
+By appropriately leveraging the information from the unmodified peptides, the proposed approach improved the accuracy of the estimates of PTM fold changes, and produced a better calibrated false positive rate of detecting differentially abundant PTMs as compared to existing methods.
 In particular, accounting for the confounding from protein abundance allowed us to characterize the true effect of the modification, avoiding the need for more manual and time intensive follow-up investigation.
@@
 The proposed approach is implemented as a freely available open source R package $MSstatsPTM$, as part of the $MSstats$ family of packages \cite{Choi:2014,Huang:2020}, and is available on Bioconductor.
+\todo{OV: could you clean up the references a little bit?
+E.g., remove DOI, ISBN}
 
@@
 Furthermore, the fold changes of half of the 500 differential PTMs were fully masked by changes in the unmodified protein.
-Additionally, the fold change of half the 500 non-differential PTMs was entirely due to changes in the unmodified protein.
-All the differential PTMs were generated with an expected log fold change of 0.75 between conditions.
+Finally, the fold change of half the 500 non-differential PTMs was entirely due to changes in the unmodified protein.
+All the differential PTMs were generated with an expected log \todo{OV: log base?} fold change of 0.75 between conditions.
 
@@
 
-\medskip \noindent {\bf Evaluation:} We evaluated the ability of the statistical methods to correctly detect which PTMs were simulated with changes between conditions and which were not.
+\medskip \noindent {\bf Evaluation:} We evaluated the ability of the statistical methods to correctly detect differentially abundant PTMs.
 We gauged the methods ability to avoid false positives (i.e.
@@
 Quantitative changes in protein and site abundance of these 20 human proteins were the target of the benchmark.
-In particular, we distinguish the changes in the abundances of the modified peptides (i.e., unadjusted changes) and the changes of their abundance relative to the changes in the abundances of the human lysate (i.e., protein-level adjusted changed).
+In particular, we distinguished the changes in the abundances of the modified peptides (i.e., unadjusted changes) and the changes of their abundance relative to the changes in the abundances of the human lysate (i.e., protein-level adjusted changed).
 The true log-fold changes between the relevant components of the relevant mixtures are summarized in Figure \ref{fig:benchmark-design}(b).
@@
 flexneri}) bacteria were measured at 1, 2, 4, and 6 hour increments, resulting in six total conditions.
-Shigella ubiquitin ligase IpaH7.8 was shown to function as an inhibitor of the protein gasdermin D (GSDMD).
+\todo{OV: is it a time course design or a repeated measures design?} \todo{OV: shouldn't the next sentence be part of the evaluation?
+Should we be evaluating the results wrt these criteria?}Shigella ubiquitin ligase IpaH7.8 was shown to function as an inhibitor of the protein gasdermin D (GSDMD).
 GSDMD was actively degraded when IpaH7.8 expression was induced by dox treatment in human cells.
@@
 \medskip \noindent {\bf Data acquisition:} 11 samples were allocated to 1 TMT mixture in an unbalanced group comparison design.
-All conditions were allocated two biological replicates except for the Dox1hr condition, which was allocated one replicate.
-The ubiquitinated peptides, and the total proteome (i.e., global profiling) were each conducted in a single LC-MS/MS run.
+All conditions had two biological replicates except for the Dox1hr condition, which was allocated one replicate.
+\todo{OV: the text up to here should be part of the experimental design} The ubiquitinated peptides, and the total proteome (i.e., global profiling) were each conducted in a single LC-MS/MS run.
 There was a 95\% overlap between the identified modified peptides and proteins that were quantified in the global profiling run.
@@
 The abundance of total protein and post-translation modifications were quantified at three time points, uninfected, early infection (45-60 minutes), and late infection (3-3.5 hours).
+\todo{OV: is it a time course design or a repeated measures design?}
 
@@
 Conversely, mixture two contained one replicate of uninfected cKO and two uninfected WT.
-This experiment included a total proteome (i.e., a global profiling run) and a phosphopeptide enrichment run.
+\todo{OV: the text up to here should be part of the experimental design}  This experiment included a total proteome (i.e., a global profiling run) and a phosphopeptide enrichment run.
 There was a 90\% overlap between the identified modified peptides and proteins that were quantified in the global profiling run.
@@
 The features can be quantified as part of a same mass spectrometry run, or in separate enrichment and global proteome profiling runs.
-Figure \ref{fig:data-structure}(a) schematically illustrates this data structure for one PTM site and the corresponding unmodified features of the protein, in a special case of 2 conditions and 2 biological replicates per condition.
+Figure \ref{fig:data-structure} schematically illustrates this data structure for one PTM site and the corresponding unmodified features of the protein, in a special case of $I=2$ conditions and $J=2$ biological replicates per condition.
 The number of modified and unmodified features varies across proteins.
 Some log-intensities may be outliers, and some spectral features can be missing.
+
 The population parameter of interest is the difference between the abundances of a PTM site in condition $1$ and condition $2$, denoted by $\mu_1^{\ast}$ and $\mu_2^{\ast}$ respectively.
@@
 \end{equation}
-To test the hypothesis, a statistical model must represent the values in the table in Figure \ref{fig:data-structure}(a), and provide a model-based estimate of $\mu_{1}^{\ast} - \mu_{2}^{\ast}$, and its standard error.
-
-Unfortunately, the population parameter is inherently convoluted with the overall changes in protein abundance. To account for this, we can change the null hypothesis to
+To test the hypothesis, a statistical model must represent the values in the table in Figure \ref{fig:data-structure}, and provide a model-based estimate of $\mu_{1}^{\ast} - \mu_{2}^{\ast}$, and its standard error.
+
+Unfortunately, the population parameter is inherently convoluted with the overall changes in protein abundance. To account for this, it is advantageous to change the null hypothesis to
 
@@
 Simulations were generated with different numbers of replicates.
-The design of two Biological Experiments were unbalanced with unequal replicates per condition.
+The designs of two biological experiments were unbalanced with unequal replicates per condition.
 ``No.
 of mod.
-features/site'' is the number of features used to estimate the abundance of a single modification.
-A feature is a peptide and its corresponding charge.
+features/site'' is the number of features (i.e., peptide ions) used to estimate the abundance of a single modification.
 ``No.
 of unmod.
-peptides/protein'' is the number of peptides without modifications that were used to estimate the global protein abundance.
-``Data availability'' shows the ID of the MassIVE.quant repository or the GitHub repository.
-``Analysis'' shows the ID of the MassIVE.quant reanalysis container, containing analysis code and modeling results.
+peptides/protein'' is the number of peptides \todo{OV: peptides or peptide ions?} without modifications that were used to estimate the global protein abundance.
+``Data availability'' is the ID of the MassIVE.quant repository or the GitHub repository.
+``Analysis'' is the ID of the MassIVE.quant reanalysis container, containing analysis code and modeling results.
 All the experiments were conducted in data-dependent acquisition (DDA) mode.
@@
 \includegraphics[scale=.6]{images/fig3.png}
-\caption{Goals of PTM characterization, input to statistical analyses, and notation.
+\caption{\todo{OV: remove (a) from the figure.
+Remove part (b)?
+Looks like an inset now...} Goals of PTM characterization, input to statistical analyses, and notation.
 (a) Schematic data representation, in a simplified case of two conditions and two biological replicates.

### 2022-06-23 main/MSstatsPTM_Main.tex
Added models to background

@@
 Finally, the fold change of half the 500 non-differential PTMs was entirely due to changes in the unmodified protein.
-All the differential PTMs were generated with an expected log \todo{OV: log base?} fold change of 0.75 between conditions.
+All the differential PTMs were generated with an expected log base 2 fold change of 0.75 between conditions.
 
@@
 flexneri}) bacteria were measured at 1, 2, 4, and 6 hour increments, resulting in six total conditions.
-\todo{OV: is it a time course design or a repeated measures design?} \todo{OV: shouldn't the next sentence be part of the evaluation?
-Should we be evaluating the results wrt these criteria?}Shigella ubiquitin ligase IpaH7.8 was shown to function as an inhibitor of the protein gasdermin D (GSDMD).
-GSDMD was actively degraded when IpaH7.8 expression was induced by dox treatment in human cells.
-
-\medskip \noindent {\bf Data acquisition:} 11 samples were allocated to 1 TMT mixture in an unbalanced group comparison design.
+11 samples were allocated to 1 TMT mixture in an unbalanced repeated measure design.
 All conditions had two biological replicates except for the Dox1hr condition, which was allocated one replicate.
-\todo{OV: the text up to here should be part of the experimental design} The ubiquitinated peptides, and the total proteome (i.e., global profiling) were each conducted in a single LC-MS/MS run.
+
+\medskip \noindent {\bf Data acquisition:} The ubiquitinated peptides, and the total proteome (i.e., global profiling) were each conducted in a single LC-MS/MS run.
 There was a 95\% overlap between the identified modified peptides and proteins that were quantified in the global profiling run.
@@
 Since the dataset was a biological investigation, the true positive modifications were unknown.
+Shigella ubiquitin ligase IpaH7.8 was shown to function as an inhibitor of the protein gasdermin D (GSDMD).
+GSDMD was actively degraded when IpaH7.8 expression was induced by dox treatment in human cells.
+We expect IpaH7.8 to function as an inhibitor of GSDMD in the global profiling run.
 
@@
 The abundance of total protein and post-translation modifications were quantified at three time points, uninfected, early infection (45-60 minutes), and late infection (3-3.5 hours).
-\todo{OV: is it a time course design or a repeated measures design?}
-
-\medskip \noindent {\bf Data acquisition:} 22 biological samples were allocated to 2 TMT mixtures in an unbalanced design, with 11 samples allocated to each mixture.
+22 biological samples were allocated to 2 TMT mixtures in an unbalanced repeated measure design, with 11 samples allocated to each mixture.
 16 replicates were spread equally between the early and late WT and cKO conditions, resulting in four replicates per condition.
@@
 Conversely, mixture two contained one replicate of uninfected cKO and two uninfected WT.
-\todo{OV: the text up to here should be part of the experimental design}  This experiment included a total proteome (i.e., a global profiling run) and a phosphopeptide enrichment run.
+
+\medskip \noindent {\bf Data acquisition:} This experiment included a total proteome (i.e., a global profiling run) and a phosphopeptide enrichment run.
 There was a 90\% overlap between the identified modified peptides and proteins that were quantified in the global profiling run.
@@
 For one protein, the PTM site is represented by $K$ spectral features (peptide ions, distinguished by their cleavage residues and charge states).
-The log-intensity (base 2) of Feature $k$, in Run $j$ of Condition $i$ is denoted by $y_{ijk}^{\ast}$.
+The log-intensity (base 2) of Feature $k$, in Replicate $j$ of Condition $i$ is denoted by $y_{ijk}^{\ast}$.
 Conversely, the unmodified protein is represented by $L$ spectral features, and the log-intensity of Feature $l$ from the unmodified peptides in the same run is denoted by $y_{ijl}$.
@@
 \begin{aligned}
-\hat{u}^{\ast}_{ij} = \log \left( \sum_{k=1}^{K} 2^{y_{ijk}^{\ast}} \right) - \log \left( \sum_{l=1}^{L} 2^{y_{ijl}} \right).
+\hat{u}^{\ast}_{ij} = \log \left( \sum_{k=1}^{K} 2^{y_{ijk}^{\ast}} \right) - \log \left( \sum_{l=1}^{L} 2^{y_{ijl}} \right)
 \end{aligned}
@@
 
-These estimated ratios are then used as input to the two sample $t$-test, the output of which is the estimate of our quantity of interest $\hat{\Delta}$ and the standard error, $\widehat{SE(\hat{\Delta})}$.
+These estimated ratios are then used as input to the two sample $t$-test.
+
+\begin{equation}
+\begin{aligned}
+\hat{u}^{\ast}_{ij} = \hat{\mu}^{\ast}_{i} + \epsilon^{\ast}_{ij}, \text{where } \epsilon^{\ast}_{ij} \mathop\sim^{iid} N(0, \sigma^2), i \in \{1, 2\}
+\end{aligned}
+\label{eq:ttest_model}
+\end{equation}
+
+$t$-test assumes that the data follows a normal distribution, sampled from a representative portion of the total population, and that the conditions of interest have equal variance.
+The method is only applicable comparisons between two conditions of interest.
+The output of this model is the estimate of our quantity of interest $\hat{\Delta}$ and the standard error, $\widehat{SE(\hat{\Delta})}$.
 The statistical significance is determined by comparing the test statistic against the $t$ distribution, with degrees of freedom $df=2J-2$ in balanced designs.
-$t$-test assumes that the data follows a normal distribution, sampled from a representative portion of the total population, and that the conditions of interest have equal variance.
 
@@
 To remove confounding we transform the input by taking the intensity ratios of modified and unmodified peptide features, $\hat{u}^{\ast}_{ij}$, calculated in the same way as Equation \ref{eq:adj_estimation}.
-Limma then fits a linear model using the transformed inputs to estimate $\hat{\Delta}$.
-To estimate $\widehat{SE(\hat{\Delta})}$, Limma leverages empirical bayes moderation~\cite{Robbins1992} to moderate the model variance of the individual PTM using the shared global variance of all PTMs modeled in the experiment.
-For a single PTM $m$ with an estimated fold change $\hat{\Delta}_m$, the estimated variance is calculated as $s^{adj}_{m} = P(s_{m}| s_\Theta)$ where $s_{m}$ is the variance resulting from the linear model of PTM $m$ and $s_\Theta$ is the global variance across all PTMs.
-The statistical significance is determined using $\hat{\Delta}_m$ and $\widehat{SE(\hat{\Delta})}$ calculated using the adjusted variance $s^{adj}_{m}$.
+Limma then fits a linear model using the transformed inputs.
+
+\begin{equation}
+\begin{aligned}
+\hat{u}^{\ast}_{ij} = \beta_0 + c_{i}\beta_1 + + r_j\beta_2 + \epsilon^{\ast}_{ij}, \text{where } \epsilon^{\ast}_{ij} \mathop\sim^{iid} N(0, \sigma^2/w_j), i \in \{1, 2\}, j \in \{1, 2\}
+\end{aligned}
+\label{eq:limma_model}
+\end{equation}
+
+Using a linear model allows Limma to incorporate the replicate, $r_j$, as an additional source of variation.
+Additionally, the model allows for the sharing of variance information across comparisons of interest.
+The model is used to estimate our condition of interest, $\hat{\Delta}$.
+To estimate $\widehat{SE(\hat{\Delta})}$, Limma leverages empirical bayes moderation~\cite{Robbins1992} to adjust the model variance of the individual PTM using the shared variance of all PTMs modeled in the experiment.
+\todo{DK: I introduced a new subscript here for the PTM p.
+It feel awkward to add it across the other methods so I only included it in the moderation describtion.} For a single PTM $p$ with an estimated fold change $\hat{\Delta}_p$, the estimated variance is calculated as $var(\hat{u}^{\ast}_{pij}) = \sigma^2_p / w_{pj}$ where $\sigma^2_p$ is the variance resulting from the linear model of PTM $p$ and $w_{pj}$ is the global variance across all PTMs in replicate $j$.
+The statistical significance is then determined using $\hat{\Delta}_p$ and $\widehat{SE(\hat{\Delta}_p)}$ calculated using the adjusted variance $var(\hat{u}^{\ast}_{pij})$.
 
@@
 First, utilizing empirical bayes moderation can increase the sensitivity of the analysis and is especially useful in cases of low replicates.
-Additionally, using a linear model allows for the sharing of variance information across comparisons of interest.
-This means we can more accurately measure the variance as the number of comparisons increases.
+Additionally, sharing variance information across comparisons means we can more accurately measure the variance as the number of comparisons increases.
 
@@
 These summarized values are then used as input to fit a linear model based on the family of linear mixed-effects models \cite{Bolker2009} \cite{Faraway:2006}.
-The output of this model is the quantity of interest $\hat{\Delta}$ and the standard error, $\widehat{SE(\hat{\Delta})}$.
-These methods are implemented in the packages $MSstats$ and $MSstatsTMT$~\cite{Huang:2020} and include functionalities for end to end quantification of proteins and differential analysis between experimental conditions.
-
-The $MSstats$ analysis pipeline has a number advantages over simpler methods.
-First, the approach extends to many types of complex experimental design, including unbalanced designs and multiple sources of variation, such as fractionation.
-The methods include optional implementation of empirical bayes moderation, described previously, when applied to TMT experiments~\cite{Huang:2020}.
-Additionally, the summarization step includes options for missing value imputation, using an accelerated failure time (AFT) model~\cite{Wei:1992}, and has been shown to be more robust compared to using individual features~\cite{Choi:2014}.
+For example, the model for a TMT experiment would be as follows
+
+\todo{need to fix the centering}
+\begin{equation}
+\begin{aligned}
+u_{ijm} = \beta_0 + c_{i}\beta_1 + + r_j\beta_2 + mix_m\beta_3 + \epsilon^{\ast}_{ij}, \text{where } \epsilon^{\ast}_{ijm} \mathop\sim^{iid} N(0, \sigma^2), \\ 
+mix_m \mathop\sim^{iid} N(0, \sigma^2_m), r_j \mathop\sim^{iid} N(0, \sigma^2_j), \sum_{i=0}^{I} c_i = 0
+\end{aligned}
+\end{equation}
+
+We then use this model to estimate the quantity of interest $\hat{\Delta}$ and the standard error, $\widehat{SE(\hat{\Delta})}$.
+These methods are implemented in the R packages $MSstats$ and $MSstatsTMT$~\cite{Huang:2020}.
+
+$MSstats$ has a number advantages over simpler methods.
+First, the approach extends to many types of complex experimental design, including those with multiple sources of variation, such as the TMT model above, and unbalanced designs.
+Run-level summarization has been shown to be more robust compared to directly using individual features as input to the statistical model~\cite{Choi:2014}.
+Additionally, missing value imputation can be performed before run-level summarizing using an accelerated failure time (AFT) model~\cite{Wei:1992}.
+Finally, the methods include optional implementation of empirical bayes moderation, described above, when applied to TMT experiments~\cite{Huang:2020}.
 
@@
 & \multirow{2}{*}{Dataset} & No. of & No. of bio. & No. of mod. & No. of mod. & No. of unmod.&  Data & \multirow{2}{*}{Analysis}\\
-&  & Conditions & replicates & peptides & features/site & peptides/protein & availability & \\
+&  & Conditions & replicates & peptides & features/site & features/protein & availability & \\
 \hline
@@
 %% DDA Meierhofer
-\multirow{3}{*}{\rotatebox[origin=c]{90}{Biological}} \multirow{3}{*}{\rotatebox[origin=c]{90}{Experiment}} & Human - Ubiquitination - 1mix-TMT & 6 & 2/1 & 8,848 & 1.21 & 10.75 & \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=b6f0c74c234247678fb0888c6df1f225}{MSV000088966} & \href{https://massive.ucsd.edu/ProteoSAFe/reanalysis_container.jsp?task=b639fa5d0ff24ce78abdca13f4b18c0a}{RMSV000000356}\\
+\multirow{3}{*}{\rotatebox[origin=c]{90}{Biological}} \multirow{3}{*}{\rotatebox[origin=c]{90}{Experiment}} & Human - Ubiquitination - 1mix-TMT & 6 & 2/1 & 8,848 & 1.21 & 11.01 & \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=b6f0c74c234247678fb0888c6df1f225}{MSV000088966} & \href{https://massive.ucsd.edu/ProteoSAFe/reanalysis_container.jsp?task=b639fa5d0ff24ce78abdca13f4b18c0a}{RMSV000000356}\\
 &&&&&&&& \\%[-0.05in]
@@
 %%%% SRM Picotti 2009
-& Mouse - Phosphorylation - 2mix-TMT& 6 & 4/3 & 26,433 & 1.67 & 10.95 & \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=4878d777c6b34cf8aaf8477e93140c4d}{MSV000085565} & \href{https://massive.ucsd.edu/ProteoSAFe/reanalysis_container.jsp?task=79406d32c1f9413daf7756e6eeb24968}{RMSV000000357}\\
+& Mouse - Phosphorylation - 2mix-TMT& 6 & 4/3 & 26,433 & 1.67 & 11.61 & \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=4878d777c6b34cf8aaf8477e93140c4d}{MSV000085565} & \href{https://massive.ucsd.edu/ProteoSAFe/reanalysis_container.jsp?task=79406d32c1f9413daf7756e6eeb24968}{RMSV000000357}\\
 &&&&&&&& \\%[-0.05in]
@@
 of unmod.
-peptides/protein'' is the number of peptides \todo{OV: peptides or peptide ions?} without modifications that were used to estimate the global protein abundance.
+peptides/protein'' is the number of peptide ions without modifications that were used to estimate the global protein abundance.
 ``Data availability'' is the ID of the MassIVE.quant repository or the GitHub repository.
@@
 \centering
-\includegraphics[scale=.6]{images/fig3.png}
-\caption{\todo{OV: remove (a) from the figure.
-Remove part (b)?
-Looks like an inset now...} Goals of PTM characterization, input to statistical analyses, and notation.
+\includegraphics[scale=.45]{images/fig3.png}
+\caption{Goals of PTM characterization, input to statistical analyses, and notation.
 (a) Schematic data representation, in a simplified case of two conditions and two biological replicates.

### 2022-06-26 main/MSstatsPTM_Main.tex
cleaned up ref
Existing methids

@@
 The proposed approach is implemented as a freely available open source R package $MSstatsPTM$, as part of the $MSstats$ family of packages \cite{Choi:2014,Huang:2020}, and is available on Bioconductor.
-\todo{OV: could you clean up the references a little bit?
-E.g., remove DOI, ISBN}
 
@@
 specificity), accurately estimate the fold change between conditions, and analyzed the sensitivity of detecting differentially abundant PTMs.
-The evaluation was performed both in the presence of convolution with changes in the unmodified protein and after applying adjustment to correct for convolution.
+The evaluation was performed both in the presence of confounding with changes in the unmodified protein and after applying adjustment to correct for the confounding.
 
@@
 We evaluated their ability to correctly detect PTM's specificity, fold change estimation, and sensitivity.
-These statistics were analyzed both in the presence of, and without, convolution with changes in the unmodified protein.
+These statistics were analyzed both in the presence of, and without, confounding with the overall changes in protein abundance.
 
@@
 For one protein, the PTM site is represented by $K$ spectral features (peptide ions, distinguished by their cleavage residues and charge states). 
-The log-intensity (base 2) of Feature $k$, in Replicate $j$ of Condition $i$ is denoted by $y_{ijk}^{\ast}$.
+The number of modified and unmodified features can vary across proteins. Some log-intensities may be outliers, and some spectral features can be missing. 
+The log$_2$-intensity of Feature $k$, in Replicate $j$ of Condition $i$ is denoted by $y_{ijk}^{\ast}$.
 Conversely, the unmodified protein is represented by $L$ spectral features, and the log-intensity of Feature $l$ from the unmodified peptides in the same run is denoted by $y_{ijl}$.
 The features can be quantified as part of a same mass spectrometry run, or in separate enrichment and global proteome profiling runs.
-Figure \ref{fig:data-structure} schematically illustrates this data structure for one PTM site and the corresponding unmodified features of the protein, in a special case of $I=2$ conditions and $J=2$ biological replicates per condition.
-The number of modified and unmodified features varies across proteins.
-Some log-intensities may be outliers, and some spectral features can be missing.
-
-The population parameter of interest is the difference between the abundances of a PTM site in condition $1$ and condition $2$, denoted by $\mu_1^{\ast}$ and $\mu_2^{\ast}$ respectively.
+Figure \ref{fig:data-structure} schematically illustrates this data structure for one PTM site, in a special case of $I=2$ conditions and $J=2$ biological replicates per condition.
+
+The population parameter of interest is the difference between the abundances of a PTM site in Condition $i$ and Condition $i^{\prime}$, denoted by $\mu_i^{\ast}$ and $\mu_{i^{\prime}}^{\ast}$ respectively.
 We are interested in testing the null hypothesis
+\begin{eqnarray}
+H_{0}: \Delta = \mu_{i}^{\ast} - \mu_{i^{\prime}}^{\ast} = 0 \ \text{vs} \ 
+H_{a}: \Delta = \mu_{i}^{\ast} - \mu_{i^{\prime}}^{\ast} \neq 0
+\label{eq:conv_null_hyp}
+\end{eqnarray}
+Unfortunately, the population parameter is inherently confounded with the overall changes in protein abundance. To account for this, it is advantageous to change the null hypothesis to
+\begin{eqnarray}
+H_{0}: \Delta_{adj} = ( \mu_{i}^{\ast} - \mu_{i}) - ( \mu_{i^{\prime}}^{\ast} - \mu_{i^{\prime}} ) = 0 \ \text{vs} \ 
+H_{a}: \Delta_{adj} = ( \mu_{i}^{\ast} - \mu_{i}) - ( \mu_{i^{\prime}}^{\ast} - \mu_{i^{\prime}} ) \neq 0
+\label{eq:null_hyp}
+\end{eqnarray}
+where $\mu_i$ and $\mu_{i^{\prime}}$ reflect the overall protein abundances in Condition $i$ and Condition $i^{\prime}$.
+These quantities can be estimated using protein features without the modification site.
+
+%---------------------------------------------------------------------------------------------------------------------------------------------------------------------------
+\subsection*{Existing statistical methods for detecting differentially abundant PTMs}
+
+\todo{OV: even though we say 't-test', the model that we describe is in fact an ANOVA (i.e., nothing limits us to 2 conditions. Should we change it to ANOVA directly?)}
+
+\noindent {\bf Two-sample $t$-test based on modified log$_2$-intensities}
+
+\medskip \noindent 
+\todo{OV: this doesn't look right, as it seems that all the features and replicates equally contribute to the error.
+Do people do some kind of summarization of the features first?
+Let's first describe the summarization}
+The basic two-sample $t$-test \todo{ref - ideally from this literature if you can find} is based on the model
+\begin{eqnarray}
+y^*_{ijk} = \hat{\mu}^{*}_{i} + \epsilon^{\ast}_{ij}, \ \epsilon^{\ast}_{ij} \mathop\sim^{iid} N(0, \sigma^{*2}),\ i=1,\ldots,I,\ j=1,\ldots,J,\ k=1,\ldots,K
+\label{eq:ttest_model}
+\end{eqnarray}
+The model allows us to estimate $\hat{\Delta}$ and its standard error.
+Based on the model, the approach tests the null hypothesis in \eqref{conv_null_hyp}, by comparing the test statistic against the Student distribution with $df=2J-2$ degrees of freedom in balanced designs.
+Unfortunately, this approach is fundamentally flawed as it does not account for the confounding between changes in the PTM abundance and the overall changes in the abundance of the protein.
+
+\medskip \noindent
+
+\noindent {\bf Two-sample $t$-test based on ratios of modified and unmodified log$_2$-intensities}
+
+\medskip \noindent 
+The basic $t$-test can be extended to account for the confounding of changes in PTM abundance and overall changes in protein abundance\todo{OV: not sure where to put this ref ~\cite{Schwammle2015}- could you check it belongs here?} ~\cite{THOMAS2020,Mertins:2013}.
+Typically this is done by first summarizing the feature intensities in each run, separately for modified and unmodified features, e.g.
+\todo{with the sum on the original scale}, dividing the summary of the PTM abundance by the summary of the protein abundance,  \todo{and then taking a log?}.
+Specifically, for condition $i$ and biological replicate $j$, the approach calculates
+\begin{eqnarray}
+u_{ij} = \log \left( \sum_{k=1}^{K} 2^{y_{ijk}^{\ast}} \right) - \log \left( \sum_{l=1}^{L} 2^{y_{ijl}} \right)
+\label{eq:adj_estimation}
+\end{eqnarray}
+The approach then models these values as
+\begin{eqnarray}
+u_{ij} = (\mu^{\ast}_{i}-\mu_i) + \epsilon^{\prime}_{ij}, \text{where } \epsilon^{\prime}_{ij} \mathop\sim^{iid} N(0, \sigma^{\prime 2}), i \in \{1, 2\}
+\label{eq:ttest_model}
+\end{eqnarray}
+The model allows us to estimate $\hat{\Delta}_{adj}$ and its standard error.
+Based on this model, we can test the more relevant null hypothesis in~\eqref{null_hyp}, by comparing the test statistic against the Student distribution with \todo{how many} degrees of freedom in balanced designs.
+
+Although effective, the approach is somewhat simplistic.
+It is not applicable to experimental designs with more complex sources of biological and technological variation, such as experiments with repeated measurements, experiments with multiple batches or experiments with multi-run multiplexing.
+Since \eqref{adj_estimation} performs the adjustment on the replicate level, the experiment must contain a matching number of replicates in both the modified and unmodified runs.
+Technological artifacts such as missing values further undermine the calculation of $u_{ij}$ in~\eqref{adj_estimation}.
+Finally, the lack of a self contained, straightforward implementation of the method challenges the applications \todo{OV: is the last statement true?
+Seems simple enough}.
+
+\medskip \noindent {\bf Limma}
+
+\medskip \noindent The two versions of $t$-test \todo{OV: ANOVA?} above can be expanded with Empirical Bayes moderation in Limma~\cite{Ritchie_15a,Zhu,Chappell:2021} \todo{cite the original limma paper}.
+In particular, the ratio-based version models the ratios $\hat{u}_{ij}$ in~\eqref{adj_estimation}.
+Limma then fits a linear model using the transformed inputs.
+
 \begin{equation}
 \begin{aligned}
-H_{0}: \Delta = \mu_{1}^{\ast} - \mu_{2}^{\ast} = 0 \\
-H_{a}: \Delta = \mu_{1}^{\ast} - \mu_{2}^{\ast} \neq 0
-\end{aligned}
-\label{eq:conv_null_hyp}
-\end{equation}
-To test the hypothesis, a statistical model must represent the values in the table in Figure \ref{fig:data-structure}, and provide a model-based estimate of $\mu_{1}^{\ast} - \mu_{2}^{\ast}$, and its standard error.
-
-Unfortunately, the population parameter is inherently convoluted with the overall changes in protein abundance. To account for this, it is advantageous to change the null hypothesis to
-
-\begin{equation}
-\begin{aligned}
-H_{0}: \Delta = ( \mu_{1}^{\ast} - \mu_{1}) - ( \mu_{2}^{\ast} - \mu_{2} ) = 0 \\
-H_{a}: \Delta = ( \mu_{1}^{\ast} - \mu_{1}) - ( \mu_{2}^{\ast} - \mu_{2} ) \neq 0
-\end{aligned}
-\label{eq:null_hyp}
-\end{equation}
-where $\mu_1$ and $\mu_2$ reflect the overall protein abundances in condition $1$ and condition $2$. These quantities can be estimated using protein features without the modification site.
-
-%---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------
-\subsection*{Existing statistical methods for detecting differentially abundant PTMs}
-
-\noindent {\bf Two-sample $t$-test} 
-\todo{DK: I mainly explained the ratio version.
-I briefly mentioned the feature version, which is used to model ptms w/o adjustment.
-We could reference the versions back to hypotheses in Equation 1 and 2 but I didnt yet.}
-
-Many investigations perform differential abundance analysis of PTMs using two-sample $t$-test or its extensions~\cite{Schwammle2015}.
-These methods can be performed without accounting for convolution with global protein abundance by taking as input intensities of individual features from modified peptides, denoted by $y_{ijk}^{\ast}$.
-Additionally, they can be extended to remove convolution by taking as input intensity ratios of modified and unmodified peptide features and comparing the mean abundance of a PTM site from one condition to another~\cite{THOMAS2020}\cite{Mertins:2013}.
-This is done by taking the sum of peak intensities in each run for modified and unmodified features separately and dividing the estimate of PTM abundance by the protein abundance estimate.
-This is equivalent to taking the difference of their log abundances.
-For condition $i$ and biological replicate $j$, the estimated quantity is denoted by $\hat{u}^{\ast}_{ij}$ and is given by Equation \ref{eq:adj_estimation}.
-
-\begin{equation}
-\begin{aligned}
-\hat{u}^{\ast}_{ij} = \log \left( \sum_{k=1}^{K} 2^{y_{ijk}^{\ast}} \right) - \log \left( \sum_{l=1}^{L} 2^{y_{ijl}} \right)
-\end{aligned}
-\label{eq:adj_estimation}
-\end{equation}
-
-These estimated ratios are then used as input to the two sample $t$-test.
-
-\begin{equation}
-\begin{aligned}
-\hat{u}^{\ast}_{ij} = \hat{\mu}^{\ast}_{i} + \epsilon^{\ast}_{ij}, \text{where } \epsilon^{\ast}_{ij} \mathop\sim^{iid} N(0, \sigma^2), i \in \{1, 2\}
-\end{aligned}
-\label{eq:ttest_model}
-\end{equation}
-
-$t$-test assumes that the data follows a normal distribution, sampled from a representative portion of the total population, and that the conditions of interest have equal variance.
-The method is only applicable comparisons between two conditions of interest.
-The output of this model is the estimate of our quantity of interest $\hat{\Delta}$ and the standard error, $\widehat{SE(\hat{\Delta})}$.
-The statistical significance is determined by comparing the test statistic against the $t$ distribution, with degrees of freedom $df=2J-2$ in balanced designs.
-
-While simple, these approaches do not fully account for all sources of variations, and are not directly applicable to experiments with complex designs.
-These include experiments with comparisons of multiple conditions, acquisition in multiple batches, and unbalanced designs.
-Additionally, because the adjustment is done on the replicate level, there must be a matching number of replicates in both the modified and unmodified experimental designs.
-If the designs are different, this method is no longer directly applicable.
-Finally, while these approaches can be applied to experiments studying PTMs, there is not a self contained, straightforward implementation of the methods, making application challenging.
-
-\medskip \noindent {\bf Limma} Modifications of $t$-test such as moderated $t$-test with Limma were also proposed~\cite{Ritchie_15a}\cite{Zhu}\cite{Chappell:2021}.
-The inputs to  Limma are the same as the inputs to $t$-test.
-The intensities of individual features from modified peptides, $y_{ijk}^{\ast}$, can be used directly (in the presence of confounding with the unmodified protein).
-To remove confounding we transform the input by taking the intensity ratios of modified and unmodified peptide features, $\hat{u}^{\ast}_{ij}$, calculated in the same way as Equation \ref{eq:adj_estimation}.
-Limma then fits a linear model using the transformed inputs.
-
-\begin{equation}
-\begin{aligned}
-\hat{u}^{\ast}_{ij} = \beta_0 + c_{i}\beta_1 + + r_j\beta_2 + \epsilon^{\ast}_{ij}, \text{where } \epsilon^{\ast}_{ij} \mathop\sim^{iid} N(0, \sigma^2/w_j), i \in \{1, 2\}, j \in \{1, 2\}
+\hat{u}_{ij} = \beta_0 + c_{i}\beta_1 + + r_j\beta_2 + \epsilon^{\prime\prime}_{ij}, \text{where } \epsilon^{\prime\prime}_{ij} \mathop\sim^{iid} N(0, \sigma^{\prime\prime 2}/w_j), i \in \{1, 2\}, j \in \{1, 2\}
 \end{aligned}
@@
 \end{equation}
-
-Using a linear model allows Limma to incorporate the replicate, $r_j$, as an additional source of variation.
-Additionally, the model allows for the sharing of variance information across comparisons of interest.
-The model is used to estimate our condition of interest, $\hat{\Delta}$.
+\todo{Essentially, Limma can fit all kinds of models - with or without the bio replicate, including the ones in the equations above, it depends on the user.
+The only difference from the approaches above is the EB moderation.
+I am confused by $r_j$ do you mean $I_{biorep=j}$?
+Since replicate id takes continuous values, the model above is a regression with replicate id as a single predictor / 1 df - not sure that's right;  I am confused by $w_j$ - this is not how Limma model is specified.
+Also what are the $\beta$ and why we only have one of them?
+Let's check which versions of the models the references use and how Ting describes Limma in her paper.
+}
+
+The goal of Limma is to increase the sensitivity of detection of differential abundance when the experiment has few biological replicates.
+This is done by means of Empirical Bayes moderation, i.e.
+by borrowing the information regarding the variation across all the modification sites.
+Specifically, \todo{OV: this should be another model equation, rather than the text}
+
+For each modification site, the model yields the same estimate $\hat{\Delta}_{adj}$ as \eqref{ttest_model}, but a different standard error.
 To estimate $\widehat{SE(\hat{\Delta})}$, Limma leverages empirical bayes moderation~\cite{Robbins1992} to adjust the model variance of the individual PTM using the shared variance of all PTMs modeled in the experiment.
@@
 
-\todo{DK: @Olga can you expand on the good things/limitations? I think I am missing some points here but not sure what exactly}
-
-Limma provides an alternative that overcomes limitations of $t$-test.
-First, utilizing empirical bayes moderation can increase the sensitivity of the analysis and is especially useful in cases of low replicates.
-Additionally, sharing variance information across comparisons means we can more accurately measure the variance as the number of comparisons increases.
-
-While Limma offers improvements over $t$-test, the method still has a number of limitations.
-First, the method is limited in the types of experimental designs it can be applied to.
+Since Limma only improves upon the methods above in estimation of variation, it has the same limitations in terms of the experimental designs.
 The method is only directly applicable to experiments with at most two variance components, meaning that in more complex experiments it cannot account for all sources of variation.
@@
 
-\medskip \noindent {\bf Isobar-PTM} was also proposed for experiments with LC-MS/MS quantitative strategies that employ isobaric labels such as TMT or isobaric tag for relative and absolute quantification (iTRAQ)\cite{Breitwieser:2013}.
+\medskip \noindent {\bf Isobar-PTM} \medskip \noindent Isobar-PTM was also proposed for experiments with LC-MS/MS quantitative strategies that employ isobaric labels such as TMT or isobaric tag for relative and absolute quantification (iTRAQ)\cite{Breitwieser:2013}.
 Isobar-PTM expresses MS measurements with a linear model and performs adjustment with respect to protein abundance using the difference between log-ratio of modified peptides in two channels and log-ratio of protein level.
@@
 
-\todo{DK: This section could use some rewording}
-
 $MSstats$~\cite{Choi:2014} is designed for protein significance analysis and statistical relative quantification of proteins and peptides in global, targeted and data-independent proteomics.
-The package takes as input a list of intensities of individual features, denoted by $y_{ijk}$, used to characterize the identified protein.
+The package takes as input a list of log$_2$-intensities $y_{ijk}$, used to characterize the identified protein.
 For each protein, the feature log-intensities are first summarized into a single value per protein per run in the subplot model.
@@
 For example, the model for a TMT experiment would be as follows
-
-\todo{need to fix the centering}
-\begin{equation}
-\begin{aligned}
-u_{ijm} = \beta_0 + c_{i}\beta_1 + + r_j\beta_2 + mix_m\beta_3 + \epsilon^{\ast}_{ij}, \text{where } \epsilon^{\ast}_{ijm} \mathop\sim^{iid} N(0, \sigma^2), \\ 
-mix_m \mathop\sim^{iid} N(0, \sigma^2_m), r_j \mathop\sim^{iid} N(0, \sigma^2_j), \sum_{i=0}^{I} c_i = 0
-\end{aligned}
-\end{equation}
-
-We then use this model to estimate the quantity of interest $\hat{\Delta}$ and the standard error, $\widehat{SE(\hat{\Delta})}$.
-These methods are implemented in the R packages $MSstats$ and $MSstatsTMT$~\cite{Huang:2020}.
+\todo{OV: this is very different from how we usually present MSstats.
+Check either the MSstatsTMT paper or the draft of Ting's most recent paper (or maybe her dissertation?) Could you try using the same presentation?}
+\begin{eqnarray}
+& u_{ijm} = \beta_0 + c_{i}\beta_1 + + r_j\beta_2 + mix_m\beta_3 + \epsilon^{\ast}_{ij}, \text{where } \epsilon^{\ast}_{ijm} \mathop\sim^{iid} N(0, \sigma^2), & \\ 
+& mix_m \mathop\sim^{iid} N(0, \sigma^2_m), r_j \mathop\sim^{iid} N(0, \sigma^2_j), \sum_{i=0}^{I} c_i = 0 &
+\end{eqnarray}
+
+The model enables the estimation of $\hat{\Delta}$ and the standard error. These methods are implemented in the R packages $MSstats$ and $MSstatsTMT$~\cite{Huang:2020}.
 
@@
 Finally, the methods include optional implementation of empirical bayes moderation, described above, when applied to TMT experiments~\cite{Huang:2020}.
-
-While $MSstats$ is a powerful tool for the quantification of the global protein, it does not have built in tools to analyze PTMs.
-It is designed to estimate the difference in a protein between conditions, answering the Null Hypothesis
-
-\begin{equation}
-\begin{aligned}
-H_{0}: \Delta = \mu_{1} - \mu_{2} = 0 \\
-H_{a}: \Delta = \mu_{1} - \mu_{2} \neq 0
-\end{aligned}
-\label{eq:msstats_null_hyp}
-\end{equation}
-
-It is not designed to remove confounding with the global protein meaning that it cannot directly address the Null Hypothesis in Equation \ref{eq:null_hyp}.
+While $MSstats$ is a powerful tool for the quantification of the global protein, it does not have the tools to detect differentially abundant PTMs.
+It is designed to estimate the difference in a protein between conditions in~\eqref{conv_null_hyp}, and is not designed to remove confounding with the changes in overall protein abundance, and cannot directly test the null hypothesis in~\eqref{null_hyp}.
 
@@
 
-\subsection*{Detection of changes in PTMs - Balanced design with one source of variation}
+\subsection*{Detection of changes in PTMs}
 \label{sec:adjustment_summary}
 
+\medskip \noindent {\bf Detection of changes in PTMs in balanced design with one source of variation}
+
+\medskip \noindent
 We propose a statistical method for detecting changes in PTMs that allows us to remove confounding with changes in the global protein and is applicable to experiments with complex designs.
@@
 
-\subsubsection*{Design of PTM experiments in the balanced case}
+\subsubsection*{Design of balanced PTM experiments with one source of variation}
 \label{sec:design}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection*{Detection of changes in PTMs - Extension to complex designs}
-
-As discussed above, the proposed approach can be extended to complex designs, including experiments with multiple sources of variation and unbalanced designs.
+\medskip \noindent {\bf Detection of changes in PTMs in experiments with complex designs}
+
+\medskip \noindent The proposed approach can be extended to complex designs, including experiments with multiple sources of variation and unbalanced designs.
 This extension is achieved through utilizing the methods in $MSstats$ which, in complex cases, use restricted maximum likelihood (REML)~\cite{Kenward} to estimate the parameters of the models.
@@
 
+%----------------------------------------------------------------------------------------------------------------
 \subsection*{Implementation}
@@
 In contrast, for peptides that became significant after adjustment, their true abundance change was masked by underlying changes in the unmodified protein.
-Both of these issues were corrected in adjustment, and the unconvoluted abundance change is shown.
+Both of these issues were corrected in adjustment, and the unconfounded abundance change is shown.
 An additional question that must be addressed is if the decrease in significant peptides is due to the increased variance that comes from adjustment.
@@
 19,286 peptides became insignificant, 4,947 became significant, and 41,552 were significant in both models.
-Again we tested if the decrease in significant peptides was due to removing convolution with changes in the umodified protein or increased variance after adjustment was applied.
+Again we tested if the decrease in significant peptides was due to removing confounding with changes in the umodified protein or increased variance after adjustment was applied.
 This was tested by looking for modified peptides whose adjusted log fold change was within 10\% of the unadjusted log fold change but became insignificant after adjustment.
@@
 We proposed a general statistical modeling framework and implementation for PTM characterization.
-The framework is designed for bottom-up MS workflows, which are characterized with variations from multiple convoluted sources, frequent missing data, and associated uncertainty in the conclusions.
+The framework is designed for bottom-up MS workflows, which are characterized with variations from multiple confounded sources, frequent missing data, and associated uncertainty in the conclusions.
 The framework is general and is applicable to a variety of experimental designs.
@@
 
-Additionally, abundance levels of PTM sites can be convoluted with each other if there are multiple modification sites per peptide, or convoluted with changes in the unmodified peptide (as opposed to the unmodified protein).
+Additionally, abundance levels of PTM sites can be confounded with each other if there are multiple modification sites per peptide, or confounded with changes in the unmodified peptide (as opposed to the unmodified protein).
 In the current implementation the effect of a specific modification in a peptide with multiple modifications cannot be quantified.
-One potential solution to this is to measure the abundance of peptides with one modification and use this to adjust the peptide with multiple sites to remove the convolution.
+One potential solution to this is to measure the abundance of peptides with one modification and use this to adjust the peptide with multiple sites to remove the confounding.
 However, this method would likely run into challenges due to sparsity of features for modified peptides with both a single and multiple modification sites.
@@
 Overall, the proposed approach balances accuracy and practicality, and enables the analysis of complex experiments in high throughput.
-Future work is to carry out the inference and testing for not only the relative change of PTM abundance, but also the fraction of the protein that is modified at the particular site (site occupancy, or stoichiometry), and attempt to deconvolute the effect of individual PTMs in peptides with multiple modifications.
+Future work is to carry out the inference and testing for not only the relative change of PTM abundance, but also the fraction of the protein that is modified at the particular site (site occupancy, or stoichiometry), and attempt to remove the confounding of individual PTMs in peptides with multiple modifications.
 
@@
 %% DDA Meierhofer
-\multirow{3}{*}{\rotatebox[origin=c]{90}{Biological}} \multirow{3}{*}{\rotatebox[origin=c]{90}{Experiment}} & Human - Ubiquitination - 1mix-TMT & 6 & 2/1 & 8,848 & 1.21 & 11.01 & \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=b6f0c74c234247678fb0888c6df1f225}{MSV000088966} & \href{https://massive.ucsd.edu/ProteoSAFe/reanalysis_container.jsp?task=b639fa5d0ff24ce78abdca13f4b18c0a}{RMSV000000356}\\
+\multirow{3}{*}{\rotatebox[origin=c]{90}{Biological}} \multirow{3}{*}{\rotatebox[origin=c]{90}{Experiment}} & Human - Ubiquitination - 1mix-TMT & 6 & 2 or 1 & 8,848 & 1.21 & 11.01 & \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=b6f0c74c234247678fb0888c6df1f225}{MSV000088966} & \href{https://massive.ucsd.edu/ProteoSAFe/reanalysis_container.jsp?task=b639fa5d0ff24ce78abdca13f4b18c0a}{RMSV000000356}\\
 &&&&&&&& \\%[-0.05in]
@@
 %%%% SRM Picotti 2009
-& Mouse - Phosphorylation - 2mix-TMT& 6 & 4/3 & 26,433 & 1.67 & 11.61 & \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=4878d777c6b34cf8aaf8477e93140c4d}{MSV000085565} & \href{https://massive.ucsd.edu/ProteoSAFe/reanalysis_container.jsp?task=79406d32c1f9413daf7756e6eeb24968}{RMSV000000357}\\
+& Mouse - Phosphorylation - 2mix-TMT& 6 & 4 or 3 & 26,433 & 1.67 & 11.61 & \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=4878d777c6b34cf8aaf8477e93140c4d}{MSV000085565} & \href{https://massive.ucsd.edu/ProteoSAFe/reanalysis_container.jsp?task=79406d32c1f9413daf7756e6eeb24968}{RMSV000000357}\\
 &&&&&&&& \\%[-0.05in]
@@
 \includegraphics[scale=.45]{images/fig3.png}
-\caption{Goals of PTM characterization, input to statistical analyses, and notation.
-(a) Schematic data representation, in a simplified case of two conditions and two biological replicates.
-Each PTM site is modeled and characterized separately, where a PTM is quantified with observed spectral features (boxes), distinguished by different charge states of a peptide.
-The log-intensity of the PTM in Condition i, Run j, and feature k are denoted by $y_{ijk}^{\ast}$.
-Log-intensity of feature corresponding to unmodified peptide in Condition i, Run j, and feature k are denoted by $y_{ijk}$.
-Peptides can be fully cleaved (solid lines) and/or partially cleaved (dashed lines).
-(c) Model-based testing for differential PTM abundance, which corrects for the underlying protein abundance with a cost of increased uncertainty about the estimate of difference between conditions.}
+\caption{Schematic representation of one PTM site, in a special case of $I=2$ conditions and $J=2$ biological replicates per condition.
+We are interested in estimating the difference between the population-level PTM abundance between Condition 1 and Condition 2 (i.e, $\mu^*_1-\mu^*_2$), relative to the population-level difference of the overall protein abundance (i.e., $\mu_1-\mu_2$).
+These quantities are characterized by the observed spectral Features (boxes), i.e.
+peptides of different charge states.
+The peptides can be fully cleaved (solid lines), or partially cleaved (dashed lines).
+The log$_2$-intensities of the modified peptides in Condition $i$, Run $j$, and Feature $k$ are denoted by $y_{ijk}^{\ast}$.
+The log$_2$-intensities of Feature $k$ corresponding to the unmodified peptide in Condition $i$ and Run $j$ are denoted by $y_{ijk}$.
+}
 \label{fig:data-structure}

### 2022-06-28 main/MSstatsPTM_Main.tex
added comments to methods

@@
 Second, contrary to the often large number of identified peptides that can be used to quantify protein abundance, there are relatively few representative peptides that span a modification site, and there may be multiple modified sites on a single peptide~\cite{Mann}.
-Third, unless early signaling events are interrogated, the interpretation of the relative changes in modification occupancy are inherently confounded with changes in the overall protein abundance, complicating the interpretation of the results \cite{Olsen:2013}.
+Third, unless early signaling events are interrogated, the interpretation of the relative changes in modification occupancy are inherently confounded with changes in the overall protein abundance, complicating the interpretation of the results \cite{Olsen:2013}\cite{Wu:2011}.
 Finally, technological aspects of bottom-up MS experiments, such as presence of labeling by tandem mass tag (TMT), introduce additional sources of uncertainty and variation.
@@
 These peptides are treated as positive controls in all comparisons except Mix 1 vs Mix 4.
+Mix 1 vs Mix 4 was treated as a negative control, with no expected fold change.
 Additionally, the background {\it E.
-coli} lysate peptides are treated as negative controls and are not expected to change in any comparison.
+coli} lysate peptides were also treated as negative controls and are not expected to change in any comparison.
 Therefor we evaluated the statistical methods ability to avoid false positives, as well as their sensitivity in detecting the differentially abundant spike-in peptides and accurately estimate their expected fold change.
@@
 Let's first describe the summarization}
+\todo{DK: I think we should just delete this part entirely. It isn't how we compared the method and is not really used in the literature from what I can tell.}
+
 The basic two-sample $t$-test \todo{ref - ideally from this literature if you can find} is based on the model
@@
 \medskip \noindent 
-The basic $t$-test can be extended to account for the confounding of changes in PTM abundance and overall changes in protein abundance\todo{OV: not sure where to put this ref ~\cite{Schwammle2015}- could you check it belongs here?} ~\cite{THOMAS2020,Mertins:2013}.
+The basic $t$-test can be extended to account for the confounding of changes in PTM abundance and overall changes in protein abundance~\cite{Schwammle2015,THOMAS2020,Mertins:2013}.
 Typically this is done by first summarizing the feature intensities in each run, separately for modified and unmodified features, e.g.
-\todo{with the sum on the original scale}, dividing the summary of the PTM abundance by the summary of the protein abundance,  \todo{and then taking a log?}.
+with the sum on the original scale, dividing the summary of the PTM abundance by the summary of the protein abundance and then taking a log.
 Specifically, for condition $i$ and biological replicate $j$, the approach calculates
@@
 Seems simple enough}.
+\todo{DK: I meant more that there was no simple package implementation.
+Researchers need to program the methods themselves.
+I cant find an implementation at least}
 
@@
 
-\medskip \noindent The two versions of $t$-test \todo{OV: ANOVA?} above can be expanded with Empirical Bayes moderation in Limma~\cite{Ritchie_15a,Zhu,Chappell:2021} \todo{cite the original limma paper}.
+\medskip \noindent The two versions of $t$-test \todo{OV: ANOVA?} above can be expanded with Empirical Bayes moderation in Limma~\cite{ Ritchie_15a, Schwammle2015, Smyth:2005, Zhu,Chappell:2021}.
 In particular, the ratio-based version models the ratios $\hat{u}_{ij}$ in~\eqref{adj_estimation}.
@@
 For example, the model for a TMT experiment would be as follows
-\todo{OV: this is very different from how we usually present MSstats.
-Check either the MSstatsTMT paper or the draft of Ting's most recent paper (or maybe her dissertation?) Could you try using the same presentation?}
+
 \begin{eqnarray}
-& u_{ijm} = \beta_0 + c_{i}\beta_1 + + r_j\beta_2 + mix_m\beta_3 + \epsilon^{\ast}_{ij}, \text{where } \epsilon^{\ast}_{ijm} \mathop\sim^{iid} N(0, \sigma^2), & \\ 
-& mix_m \mathop\sim^{iid} N(0, \sigma^2_m), r_j \mathop\sim^{iid} N(0, \sigma^2_j), \sum_{i=0}^{I} c_i = 0 &
+& Y_{mtcb} = \mu + \text{Mixture}_m + \text{TechRep(Mixture)}_{t(m)} + \text{Condition}_c + Subject_{mcb} + \epsilon_{mtcb}, & \\ 
+& \text{where }  \text{Mixture}_m  \mathop\sim^{iid} N(0, \sigma^2_M),  \text{TechRep(Mixture)} \mathop\sim^{iid} N(0, \sigma^2_T), & \\
+& \sum_{c=1}^{C} \text{Condition}_c = 0, \epsilon_{mtcb} \mathop\sim^{iid} N(0, \sigma^2) &
 \end{eqnarray}
@@
 
+%\begin{table}[]
+%\begin{tabular}{|c|c|c|}
+%\hline
+%Comparison & Unadjusted true fold change & Adjusted true fold change \\ \hline \hline
+%mix1-mix2  & 0                           & -1                        \\ \hline
+%mix1-mix3  & 0                           & 1                         \\ \hline
+%mix1-mix4  & -1                          & 0                         \\ \hline
+%mix2-mix3  & -1                          & -2                        \\ \hline
+%mix2-mix4  & 0                           & -1                        \\ \hline
+%mix3-mix4  & 0                           & 1                         \\ \hline
+%\end{tabular}
+%\end{table}
+
 \begin{figure}[ht]
 \centering
-\includegraphics[scale=.4]{images/benchmark_fig.png}
+\includegraphics[scale=.5]{images/benchmark_fig.png}
 \caption{{\bf Dataset 3: SpikeIn benchmark - Ubiquitination - Label-free}.
@@
 Coli} lysate, and spike-in peptides.
-\todo{Clarify what is changing how} (b) The expected fold change of the spike-in peptides, after adjusting for changes in the unmodified peptide (human lysate).
+\todo{TODO: Clarify what is changing how} (b) The expected fold change of the spike-in peptides, after adjusting for changes in the unmodified peptide (human lysate).
 The expected fold change is calculated using the values of the spike-in and human lysate in part (a).\todo{Could you make the columns in the table more complete?

### 2022-06-30 main/MSstatsPTM_Main.tex
Added refs

@@
 
-\medskip \noindent The two versions of $t$-test \todo{OV: ANOVA?} above can be expanded with Empirical Bayes moderation in Limma~\cite{ Ritchie_15a, Schwammle2015, Smyth:2005, Zhu,Chappell:2021}.
+\medskip \noindent The two versions of $t$-test \todo{OV: ANOVA?} above can be expanded with Empirical Bayes moderation in Limma~\cite{ Ritchie_15a, Schwammle2015, Smyth:2004, Smyth:2005, Zhu,Chappell:2021}.
 In particular, the ratio-based version models the ratios $\hat{u}_{ij}$ in~\eqref{adj_estimation}.
@@
 This is done using Tukey's median polish (TMP)~\cite{Tukey:1977}, a simple and robust procedure, which iteratively fits a two-way additive model with the effects of run and feature, summarizing them into a single value per run.
-These summarized values are then used as input to fit a linear model based on the family of linear mixed-effects models \cite{Bolker2009} \cite{Faraway:2006}.
+These summarized values are then used as input to fit a linear model based on the family of linear mixed-effects models \cite{McLean:1991, Faraway:2006, Bolker2009}.
 For example, the model for a TMT experiment would be as follows

### 2022-07-02 main/MSstatsPTM_Main.tex
Methods

@@
 The second simulation introduced real-world characteristics, such as limited modified features and missing values.
-Details of computer simulations are available on GitHub (\url{https://github.com/devonjkohler/MSstatsPTM_simulations}).
+Details of computer simulations are available in Supplementary Section 2 and on GitHub (\url{https://github.com/devonjkohler/MSstatsPTM_simulations}).
+\todo{OV: there is a way to cross-reference sections dynamically across files.
+Check Meena's MSstats paper, or just double-check that all the sections are referenced correctly in the end}
 
@@
 Finally, three biological experiments demonstrated the applicability of the proposed approach across different biological organisms, modifications, experimental designs and acquisition strategies.
-Details of the experimental data, R scripts with $MSstatsPTM$ analysis, and results of the statistical analysis are available in MassIVE.quant (\url{https://massive.ucsd.edu/ProteoSAFe/static/massive-quant.jsp}) \cite{Choi:2020}.
-
-Additional details on all experiments are available in Supplementary Section 2.
+The experimental data, R scripts with $MSstatsPTM$ analysis, and results of the statistical analysis are available in MassIVE.quant (\url{https://massive.ucsd.edu/ProteoSAFe/static/massive-quant.jsp}) \cite{Choi:2020}.
 
@@
 For simplicity, we assume that the experiment has no technical replicates, such that each biological replicate corresponds to an LC-MS/MS run.
-For one protein, the PTM site is represented by $K$ spectral features (peptide ions, distinguished by their cleavage residues and charge states). 
-The number of modified and unmodified features can vary across proteins. Some log-intensities may be outliers, and some spectral features can be missing. 
+Figure \ref{fig:data-structure} schematically illustrates this data structure for one PTM site, in a special case of $I=2$ conditions and $J=2$ biological replicates per condition.
+For one protein, the PTM site is represented by $K$ spectral features (i.e., peptide ions, distinguished by their cleavage residues and charge states). 
+The number of modified and unmodified features typically varies across proteins. Some log-intensities may be outliers, and some spectral features can be missing. 
 The log$_2$-intensity of Feature $k$, in Replicate $j$ of Condition $i$ is denoted by $y_{ijk}^{\ast}$.
@@
 The features can be quantified as part of a same mass spectrometry run, or in separate enrichment and global proteome profiling runs.
-Figure \ref{fig:data-structure} schematically illustrates this data structure for one PTM site, in a special case of $I=2$ conditions and $J=2$ biological replicates per condition.
 
@@
 \begin{eqnarray}
-H_{0}: \Delta = \mu_{i}^{\ast} - \mu_{i^{\prime}}^{\ast} = 0 \ \text{vs} \ 
-H_{a}: \Delta = \mu_{i}^{\ast} - \mu_{i^{\prime}}^{\ast} \neq 0
+H_{0}: \Delta_{PTM} = \mu_{i}^{\ast} - \mu_{i^{\prime}}^{\ast} = 0 \ \text{vs} \ 
+H_{a}: \Delta_{PTM} = \mu_{i}^{\ast} - \mu_{i^{\prime}}^{\ast} \neq 0
 \label{eq:conv_null_hyp}
 \end{eqnarray}
-Unfortunately, the population parameter is inherently confounded with the overall changes in protein abundance. To account for this, it is advantageous to change the null hypothesis to
+Unfortunately, the population parameter is inherently confounded with the overall changes in protein abundance. To account for this, it is advantageous to consider a different null hypothesis:
 \begin{eqnarray}
@@
 where $\mu_i$ and $\mu_{i^{\prime}}$ reflect the overall protein abundances in Condition $i$ and Condition $i^{\prime}$.
-These quantities can be estimated using protein features without the modification site.
+These quantities are estimated using protein features without with and without the modification site.
 
@@
 Let's first describe the summarization}
-\todo{DK: I think we should just delete this part entirely. It isn't how we compared the method and is not really used in the literature from what I can tell.}
+\todo{DK: I think we should just delete this part entirely.
+It isn't how we compared the method and is not really used in the literature from what I can tell.} \todo{OV: below in the results we use the notation $y_{i++}$ etc.
+Alternatively, since we put a hat on $\hat{y}_{ij}$, maybe we do not need '++', as it is clear that this is a summary.
+I removed '++' in the results section for now.
+Let's use the same notation in this section consistently}
 
@@
 \begin{eqnarray}
-y^*_{ijk} = \hat{\mu}^{*}_{i} + \epsilon^{\ast}_{ij}, \ \epsilon^{\ast}_{ij} \mathop\sim^{iid} N(0, \sigma^{*2}),\ i=1,\ldots,I,\ j=1,\ldots,J,\ k=1,\ldots,K
+y^*_{ijk} = \hat{\mu}^{*}_{i} + \epsilon^{\ast}_{ij}, \ \epsilon^{\ast}_{ij} \mathop\sim^{iid} \mathcal{N}(0, \sigma^{*2}),\ i=1,\ldots,I,\ j=1,\ldots,J,\ k=1,\ldots,K
 \label{eq:ttest_model}
@@
 \begin{eqnarray}
-u_{ij} = (\mu^{\ast}_{i}-\mu_i) + \epsilon^{\prime}_{ij}, \text{where } \epsilon^{\prime}_{ij} \mathop\sim^{iid} N(0, \sigma^{\prime 2}), i \in \{1, 2\}
+u_{ij} = (\mu^{\ast}_{i}-\mu_i) + \epsilon^{\prime}_{ij}, \text{where } \epsilon^{\prime}_{ij} \mathop\sim^{iid} \mathcal{N}(0, \sigma^{\prime 2}), i \in \{1, 2\}
 \label{eq:ttest_model}
@@
 \begin{aligned}
-\hat{u}_{ij} = \beta_0 + c_{i}\beta_1 + + r_j\beta_2 + \epsilon^{\prime\prime}_{ij}, \text{where } \epsilon^{\prime\prime}_{ij} \mathop\sim^{iid} N(0, \sigma^{\prime\prime 2}/w_j), i \in \{1, 2\}, j \in \{1, 2\}
+\hat{u}_{ij} = \beta_0 + c_{i}\beta_1 + + r_j\beta_2 + \epsilon^{\prime\prime}_{ij}, \text{where } \epsilon^{\prime\prime}_{ij} \mathop\sim^{iid} \mathcal{N}(0, \sigma^{\prime\prime 2}/w_j), i \in \{1, 2\}, j \in \{1, 2\}
 \end{aligned}
@@
 
-$MSstats$~\cite{Choi:2014} is designed for protein significance analysis and statistical relative quantification of proteins and peptides in global, targeted and data-independent proteomics.
-The package takes as input a list of log$_2$-intensities $y_{ijk}$, used to characterize the identified protein.
-For each protein, the feature log-intensities are first summarized into a single value per protein per run in the subplot model.
-This is done using Tukey's median polish (TMP)~\cite{Tukey:1977}, a simple and robust procedure, which iteratively fits a two-way additive model with the effects of run and feature, summarizing them into a single value per run.
-These summarized values are then used as input to fit a linear model based on the family of linear mixed-effects models \cite{McLean:1991, Faraway:2006, Bolker2009}.
-For example, the model for a TMT experiment would be as follows
-
+$MSstats$~\cite{Choi:2014} and $MSstatsTMT$~\cite{Huang:2020} are a series of R/Bioconductor packages designed for protein significance analysis and statistical relative quantification of proteins and peptides in global, targeted and data-independent proteomics.
+Models in these packages takes as input a list of log$_2$-intensities $y_{ijk}$.
+For each protein, the feature log$_2$-intensities are first summarized into a single value per protein per run using Tukey's median polish~\cite{Tukey:1977}.
+These summarized values are then used as input to fit a flexible family of linear mixed-effects models \cite{McLean:1991, Faraway:2006, Bolker2009}.
+The specific model instances depend on the experimental design and data acquisition.
+For example, for an experiment with \todo{group comparison or time course design?} $MSstatsTMT$ fits the following model \todo{OV: could we keep the notation as consistent as possible across this section?
+Above conditions had subscripts $i$ and bio reps had subscripts $j$.
+Could we keep a consistent notation below (or change the notation elsewhere in the manuscript?)} \todo{Should we maybe give a model for label-free, to stay consistent with the example in figure 2?
+Just use $I$ and $J$ etc for the indices...}
 \begin{eqnarray}
-& Y_{mtcb} = \mu + \text{Mixture}_m + \text{TechRep(Mixture)}_{t(m)} + \text{Condition}_c + Subject_{mcb} + \epsilon_{mtcb}, & \\ 
-& \text{where }  \text{Mixture}_m  \mathop\sim^{iid} N(0, \sigma^2_M),  \text{TechRep(Mixture)} \mathop\sim^{iid} N(0, \sigma^2_T), & \\
-& \sum_{c=1}^{C} \text{Condition}_c = 0, \epsilon_{mtcb} \mathop\sim^{iid} N(0, \sigma^2) &
+& y_{mtcb} = \mu + Mixture_m + TechRep(Mixture)_{t(m)} + Condition_c + Subject_{mcb} + \epsilon_{mtcb}, & \\ 
+& \text{where }  Mixture_m  \mathop\sim^{iid} \mathcal{N}(0, \sigma^2_M),  TechRep(Mixture) \mathop\sim^{iid} \mathcal{N}(0, \sigma^2_T), &  \nonumber \\
+& \sum\limits_{c=1}^{C} Condition_c = 0,\ \todo{Subject \stackrel{iid}{\sim}\mathcal{N}(0,?)}\  \epsilon_{mtcb} \mathop\sim^{iid} \mathcal{N}(0, \sigma^2) & \nonumber
 \end{eqnarray}
-
-The model enables the estimation of $\hat{\Delta}$ and the standard error. These methods are implemented in the R packages $MSstats$ and $MSstatsTMT$~\cite{Huang:2020}.
-
-$MSstats$ has a number advantages over simpler methods.
-First, the approach extends to many types of complex experimental design, including those with multiple sources of variation, such as the TMT model above, and unbalanced designs.
-Run-level summarization has been shown to be more robust compared to directly using individual features as input to the statistical model~\cite{Choi:2014}.
-Additionally, missing value imputation can be performed before run-level summarizing using an accelerated failure time (AFT) model~\cite{Wei:1992}.
-Finally, the methods include optional implementation of empirical bayes moderation, described above, when applied to TMT experiments~\cite{Huang:2020}.
-While $MSstats$ is a powerful tool for the quantification of the global protein, it does not have the tools to detect differentially abundant PTMs.
-It is designed to estimate the difference in a protein between conditions in~\eqref{conv_null_hyp}, and is not designed to remove confounding with the changes in overall protein abundance, and cannot directly test the null hypothesis in~\eqref{null_hyp}.
-
-%Additionally, $MSstats$ provides a method for imputing missing feature intensities if desired.
-When values are imputed, it is assumed they are missing for reasons of low abundance.
-Missing features are imputed in each MS run using the Accelerated Time Failure (AFT) model~\cite{Tukey:1977}.
-In order to impute a feature's missing values, the feature must be present in at least one MS Run.
-If the feature is not present across all MS Runs, the value will be left missing.
-Missing value imputation is done before summarization with Tukey's Median Polish in order to correct for the impact of outliers.
-Missing value imputation is done separately for both the PTM and global protein datasets.
+Parameters of the model are estimated using restricted maximum likelihood (REML)~\cite{Kenward}.
+In turn, the parameters allow us to estimate the relevant quantity $\hat{\Delta}_{Protein}$ \todo{this has not been defined, Define here, using the notation above?} and its standard error.
+Similarly to $Limma$, $MSstatsTMT$ includes an optional Empirical Bayes moderation of the standard error~\cite{Huang:2020}. 
+Based on the model, the approach tests the null hypothesis in \eqref{conv_null_hyp}, by comparing the test statistic against the Student distribution.
+
+The $MSstats$ framework has a number advantages over the methods above. 
+First, unlike \todo{t-test (or ANOVA?) and $Limma$}, the family of linear mixed effects models is applicable to many complex experimental designs, including designs with multiple sources of variation, and unbalanced designs.
+Second, the approach is applicable to various data acquisition types, including label-free DDA and DIA, and experiments with TMT labeling, and is compatible with various data processing tools such as \todo{XXX}
+Third, the run-level summarization has been shown to be more robust to outliers and missing values as compared to feature averages~\cite{Choi:2014}. 
+Finally, the custom $MSstats$ and $MSstatsTMT$ implementation accounts for potential data artifacts, and is numerically scalable and stable, and \todo{is available through both command line and a dedicated graphical user interface}.
+Unfortunately, MSstats estimates the difference in overall protein abundance and tests the null hypothesis in~\eqref{conv_null_hyp}.
+It does not account for the confounding between the changes in PTM abundance and the overall changes in protein abundance.
+In other words, it cannot directly test the null hypothesis in~\eqref{null_hyp}.
 
@@
 \medskip \noindent
-We propose a statistical method for detecting changes in PTMs that allows us to remove confounding with changes in the global protein and is applicable to experiments with complex designs.
-To do this, we leverage the existing methods in $MSstats$ and extend them to remove confounding between changes in the PTM and global protein.
-The method takes as input the spectral features of the PTM, $y_{ijk}^{\ast}$, and it's corresponding global protein, $y_{ijk}$.
-The features are summarized separately using the methods described previously for $MSstats$, resulting in run level values for the PTM, $\hat{y}_{1++}^{\ast}$, and unmodified peptide, $\hat{y}_{1++}$.
-These summarized values are then used as the input to our models.
-Two separate linear mixed-effects models are fit, one using the summarized PTM values and another using the summarized global protein values.
-The models give estimates of the difference in log-abundance of the underlying PTM, $\hat{\Delta}_{PTM}$, and global protein, $\hat{\Delta}_{protein}$.
-Finally, the adjusted difference, $\hat{\Delta}$, is given by the difference in log-abundance of the PTM site, subtracted by the difference in log-abundance of the underlying protein.
-
-\begin{equation}
-\hat{\Delta} = \hat{\Delta}_{PTM} - \hat{\Delta}_{protein} 
-\end{equation}
-
-In the simple case of a balanced design with one source of variation, this is equivalent to the log of the ratio of PTM abundance difference to protein abundance difference.
-
-\begin{equation}
-\hat{\Delta} = \left[ \frac{1}{J} \left( \hat{y}_{1++}^{\ast} - \hat{y}_{2++}^{\ast} \right) \right] - \left[ \frac{1}{J} \left( \hat{y}_{1++} - \hat{y}_{2++} \right) \right]
-\label{eq:prop_fc}
-\end{equation}
-
-The estimate of the adjusted difference and the standard error (SE) of the estimate are obtained by combining the difference estimates and the associated SEs from both counterparts.
-The equation accounts for the increased uncertainty when incorporating more information from the unmodified protein.
-This may lead to an increased SE, when compared to modeling the PTM alone.
-
-\begin{equation}
-\widehat{SE(\Delta)} = \left[ \left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right) \right]^{1/2}
+We propose a statistical approach for detecting changes in PTMs.
+The method takes as input the modified spectral features $y_{ijk}^{\ast}$, and the corresponding unmodified features $y_{ijk}$.
+Ideally, the modified features are acquired separately, after an enrichment to maximize the information content in the resulting dataset, and the unmodified features are acquired separately as part of a global proteome profiling.
+However the method can also take as input a combination of modified and unmodified features acquired within a same run.
+
+We leverage the existing methods in $MSstats$ and $MSstatsTMT$ and extend them to remove confounding between changes in the PTM and the overall changes in protein abundance.
+Each feature type is summarized separately using $MSstats$ or $MSstatsTMT$, resulting in run level values for the PTM, \todo{$\hat{y}_{ij}^{\ast}$} \todo{OV: I am confused: since we summarize per run, we need at least an index for a condition and an index for bio rep.
+Could we use this consistently across all the sections?}, and unmodified peptide.
+The estimated summaries of the modified features are then used as the input to the model \ov{such as in eq XXX} in $MSstats$ or $MSstatsTMT$.
+Similarly, the estimated summaries of the unmodified features are used as the input to the corresponding model in $MSstats$ or $MSstatsTMT$.
+
+For the modified features, the model-based estimates in $MSstats$ and $MSstatsTMT$ include $\hat{\Delta}_{PTM}=\hat{\mu}^{\ast}_{i}-\hat{\mu}_{i^{\prime}}^{\ast}$, and its standard error $\widehat{SE}(\hat{\Delta}_{PTM})$.
+Similarly, for the unmodified features, the model-based estimates in $MSstats$ and $MSstatsTMT$ incluse $\hat{\Delta}_{protein}=\hat{\mu}_{i}-\hat{\mu}_{i^{\prime}}$, and its standard error $\widehat{SE}(\hat{\Delta}_{protein})$.
+Therefore, the adjusted difference $\hat{\Delta}_{adj}$ in~\eqref{null_hyp} can be estimated as
+\begin{eqnarray}
+\hat{\Delta}_{adj} = (\hat{\mu}^{\ast}_{i}-\hat{\mu}_{i}) - (\hat{\mu}_{i^{\prime}}^{\ast}-\hat{\mu}_{i^{\prime}}) = (\hat{\mu}^{\ast}_{i}-\hat{\mu}_{i^{\prime}}^{\ast}) - (\hat{\mu}_{i}-\hat{\mu}_{i^{\prime}}) = \hat{\Delta}_{PTM} - \hat{\Delta}_{protein}
+\label{eq:Delta_hat_adj}
+\end{eqnarray}
+Assuming that the unexplained by the model  sources of variation in the modified features are independent from the corresponding sources of variation in the modified features, the standard error $\widehat{SE}(\hat{\Delta}_{adj})$ is obtained by combining the standard errors from the unmodified and modified model fits.
+\begin{eqnarray}
+\widehat{SE}(\Delta)_{adj} = \sqrt{ \widehat{SE}(\hat{\Delta}_{PTM})^{2} + \widehat{SE}(\hat{\Delta}_{protein})^{2}} 
 \label{eq:prop_se}
-\end{equation}
-
-Finally the degrees of freedom are calculated as follows
-
-\begin{equation}
-\left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right)^2 \bigg/
-\left( \frac{\hat{\sigma}_{\gamma^{\ast}}^{4}}{\mathrm{df}(\gamma^{\ast})} + \frac{\hat{\sigma}_{\gamma}^{4}}{ \mathrm{df}(\gamma)} \right)
+\end{eqnarray}
+The estimated standard error is larger than the standard errors associated with each distinct feature type, reflecting the increased uncertainty inherent to combining the uncertainty in two estimators.
+Finally, the degrees of freedom associated with~\eqref{prop_se} are obtained via the Satterthwaite approximation~\cite{satterthwaite:1946}\todo{OV: maybe site a more recent exp design book?}
+\begin{eqnarray}
+\left( \widehat{SE}(\hat{\Delta}_{PTM})^{2} + \widehat{SE}(\hat{\Delta}_{protein})^{2} \right)^2 \bigg/
+\left( \frac{\widehat{SE}(\hat{\Delta}_{PTM})^{4}}{\mathrm{df}(\widehat{SE}(\hat{\Delta}_{PTM}))} + \frac{\widehat{SE}(\hat{\Delta}_{protein})^{4}}{ \mathrm{df}(\widehat{SE}(\hat{\Delta}_{protein}))} \right)
 \label{eq:prop_df}
-\end{equation}
-
-The results of this model are used to answer the Null Hypothesis stated in Equation \ref{eq:null_hyp}.
-The test statistic for the hypothesis testing is the ratio of the estimate of the adjusted difference to its SE.
-To determine the statistical significance of the difference in terms of p-value, the test statistic is compared against the $t$ distribution with degrees of freedom approximated by the Satterthwaite method~\cite{satterthwaite:1946}.
-Adjustment for multiple comparisons is performed using the Benjamini-Hochberg procedure to control the false discovery rate at a desired level, e.g., $0.05$~\cite{Benjamini:1995}.
-
-\todo{DK: I'm not sure where this part should be. could move it to the first paragraph of this section, or in the extensions to complex design}
-This approach allows us to model experiments with separate designs for the PTM and global profiling runs.
-This includes differing number of runs, biological replicates, and/or types of acquisition methods.
-This is achieved through fully modeling the PTM and global protein separately before adjusting.
-Because we are not combining on the replicate level, the number of replicates for each group do not need to match for adjustment.
-This makes the proposed method more flexible and applicable to a wider variety of experimental designs.
+\end{eqnarray}
+To test the null hypothesis in~\eqref{null_hyp}, the test statistic $\hat{\Delta}_{adj}/\widehat{SE}(\Delta)_{adj}$ is compared with the Student distribution with the degrees of freedom in~\eqref{prop_df}.
+The p-values of the comparison are adjusted for multiple testing using the approach by Benjamin and Hochberg~\cite{Benjamini:1995}.
+
+By leveraging the implementations in $MSstats$ and $MSstatsTMT$, the proposed approach is very versatile.
+It is applicable to a wide variety of experimental designs, including group comparison, paired designs, time course designs and unbalanced designs.
+It is applicable to label-free data acquisition strategies such as DDA and DIA and experiments with TMT labeling.
+It is compatible with data processing tools \todo{XXX}.
+The proposed approach is also applicable to experiments where the experimental designs for PTM profiling and global proteome profiling vary in properties such as number of biological replicates, data acquisition strategies and runs.
+\todo{OV: We do have a requirement that the scale of measurements is comparable between TMT and PTM, correct?
+Otherwise the SE may not be comparable}
 
@@
 With these values and a user-specified number of conditions, the corresponding number of technical replicates per condition can then be derived, as described in~\cite{kutner_etal_04a}.
-Given the above quantities, the minimal number of replicates $J$ is determined by the variance of the estimated log-fold change $\mathrm{SE}^{2}(\hat{\Delta})$ as
+Given the above quantities, the minimal number of replicates $J$ is determined by the variance of the estimated log-fold change $\mathrm{SE}^{2}(\hat{\Delta})$ as \todo{OV: hmm...
+For what design and what model?
+It is unclear where eq 11 came from.
+This will be quite specific to the model and the design.
+Maybe refer to the model in Eq 7?
+(Which we should probably simplify to reflect the experiment in Figure2, while keeping the general indeces) }
 
 \begin{equation}
-\mathrm{SE}^{2}(\hat{\Delta}) = \left[ \frac{2}{J} \left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right) \right]
+\widehat{SE}(\Delta)_{adj}^2 = \left[ \frac{2}{J} \left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right) \right]
 \leq \left( \frac{\Delta}{t_{1-\beta, df} + t_{1-\alpha /2, df}} \right)^{2}
@@
 
+Details on how the proposed method can be used to run sample size calculations and power analysis on experiments with complex design can be seen in Supplementary Sec.
+\ref{supp-sec:complex_design}.
 More details on sample size calculation can be found in~\cite{oberg_vitek_09a}.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\subsection*{Analysis with multiple batches}
-
-%The proposed statistical framework allows to analyze data from experiments of complex designs such as factorial design.
-We discuss below a specific design commonly considered in PTM experiments, in which data are acquired in multiple batches, often as a result of multiple repeats for the antibody enrichment step.
-Typically, the number of spectral features and their intensities are different across batches, and the run-to-run variation sometimes varies across batches as well.
-The proposed approach summarizes log-intensities in each batch separately, and considers two ways to perform the statistical inference and testing.
-When variability in site-level summaries and in PTM changes is expected across batches, the proposed approach performs inference of the PTM site abundance and the adjusted difference for each batch separately.
-The evidence about differential PTM sites between conditions is averaged over batches.
-Alternatively, assuming identical variability in the site-level summaries and identical differences between conditions across batches (i.e., no interaction effect between condition and batch), the statistical inference and testing can be performed based on one single model with an additional fixed effect of batch.
-Details about these modeling considerations are discussed in Supplementary Sec \ref{supp-sec:test}.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\medskip \noindent {\bf Detection of changes in PTMs in experiments with complex designs}
-
-\medskip \noindent The proposed approach can be extended to complex designs, including experiments with multiple sources of variation and unbalanced designs.
-This extension is achieved through utilizing the methods in $MSstats$ which, in complex cases, use restricted maximum likelihood (REML)~\cite{Kenward} to estimate the parameters of the models.
-Once the parameters are estimated, we can adjust the PTM model for changes in the global protein using modified versions of Equations \ref{eq:prop_fc}, \ref{eq:prop_se}, and \ref{eq:prop_df}.
-
-The log-fold change in the adjusted PTM abundance, $\hat{\Delta
-}$, is now estimated by REML, and Equation \ref{eq:prop_fc} becomes 
-\begin{equation}
-\hat{\Delta} = (\hat{\mu}_{RML_i}^{\ast} - \hat{\mu}_{RML_{i'}}^{\ast}) - (\hat{\mu}_{RML_i} - \hat{\mu}_{RML_{i'}})
-\end{equation}
-
-The standard error of the estimate $\mathrm{SE}(\hat{\Delta})$ and degrees of freedom in Equations 3 and 4 are unchanged from the balanced case, although $\hat{\sigma}_{\gamma^\ast}$ and $\hat{\sigma}_{\gamma}$ are now estimated using REML.
-
-Details on how the proposed method can be used to run sample size calculations and power analysis on experiments with complex design can be seen in Supplementary Sec. \ref{supp-sec:complex_design}.
-
-%----------------------------------------------------------------------------------------------------------------
 \subsection*{Implementation}

### 2022-07-02 main/MSstatsPTM_Main.tex
Implementation

@@
 Similarly to $Limma$, $MSstatsTMT$ includes an optional Empirical Bayes moderation of the standard error~\cite{Huang:2020}. 
-Based on the model, the approach tests the null hypothesis in \eqref{conv_null_hyp}, by comparing the test statistic against the Student distribution.
+\todo{OV: maybe elaborate on that?} Based on the model, the approach tests the null hypothesis in \eqref{conv_null_hyp}, by comparing the test statistic against the Student distribution.
 
@@
 
-\subsection*{Detection of changes in PTMs}
+\subsection*{Statistical methods in MSstatsPTM}
 \label{sec:adjustment_summary}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection*{Implementation}
-
-The proposed methods are implemented in the open source R package $MSstatsPTM$.
+\subsection*{Implementation of MSstatsPTM}
+
+\todo{OV: I feel that some parts of this section are either specific to MSstats/MSstatsTMT, or a bit redundant with the methods.
+For the parts that we do not implement anew but reuse, could you move them to the 'Existing methods'/MSstats section?
+Here let us focus only on things that we had to develop from scratch for this project.
+If we implement the methods, we do not need to re-describe the methods}
+
+\todo{OV: could you make this section more specific to PTM converters?
+E.g., I remember that the input should be different, there are some parts relevant to the localization of PTMs.
+This was not part of the base MSstats.
+Could you emphasize that?} The proposed methods are implemented in the open source R package $MSstatsPTM$.
 An overview of the steps of the package are illustrated in Figure \ref{fig:msstatsptm_design}.
@@
 If the global profiling information is not available, the package can still analyze the modified run, but will do so without adjusting for changes in unmodified protein abundance.
-This will reduce our null hypotheses to Equation \ref{eq:conv_null_hyp}; which amounts to running the methods seen in $MSstats$ on modified peptide level data \cite{Choi:2014} \cite{Huang:2020}.
-
-After using the converters, the next step is peptide/protein summarization and missing value imputation.
+This amounts to testing the null hypothesis in~\eqref{conv_null_hyp}; which amounts to running the methods seen in $MSstats$ on modified peptide level data \cite{Choi:2014} \cite{Huang:2020}.
+
+\todo{OV: this is the part that looks redundant with the MSstats section.
+In the MSstats section, do not repeat the methods, just focus on the specifics of the implementation.
+Relavant parts would be that we fit simplified models if the base model is not applicable.
+In this section,  could you just say that we do the same as MSstats (including fitting simplified models).
+But there are some caveats, e.g.
+assumptions of missing values due to low abundance are not necessarily appropriate here so we recommend to not use that option?
+Also specialized sample size calculations} After using the converters, the next step is peptide/protein summarization and missing value imputation.
 For the modified run, the package summarizes features which include the same modification together.
@@
 
-Beyond the core functionality, the package also includes functions for plotting the results.
+\todo{OV: similarly, some of this belongs to MSstats.
+Could you focus on project-specific implementation.
+E.g., we separately visualize modified and unmodified peptides.
+(Btw, sequence-specific visualization such as the one you did for MSstatsLiP would be awesome if there is a chance)} Beyond the core functionality, the package also includes functions for plotting the results.
 These include summarization plots to help with quality assurance and identifying sources of variation (such as quality control plots and profile plots), as well as modeling plots (such as volcano plots and heatmaps), which evaluate the fold change between conditions and peptides.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection*{The proposed approach corrected for high FDR in computer simulations}
+\subsubsection*{MSstatsPTM corrected for high FDR in computer simulations}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection*{The proposed approach outperformed existing methods in simulations with missing values and low features}
+\subsubsection*{MSstatsPTM outperformed existing methods in simulations with missing values and low features}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection*{In a label-free benchmark experiment the proposed approach corrected for bias and outperformed existing methods}
+\subsubsection*{In a label-free benchmark experiment MSstatsPTM corrected for bias and outperformed existing methods}
 
@@
 
-\subsubsection*{Proposed approach removed confounding in two mixture TMT experiment}
+\subsubsection*{MSstatsPTM removed confounding in two mixture TMT experiment}
 
@@
 
-\subsubsection*{In label-free experiment without a separate global profiling run, the proposed approach could eliminate the bias due to changes in the unmodified protein, albeit less effectively than in the presence of a global profiling run}
+\subsubsection*{In label-free experiment without a separate global profiling run, MSstatsPTM eliminated the bias due to changes in the unmodified protein, albeit less effectively than in the presence of a global profiling run}
 

### 2022-07-02 main/MSstatsPTM_Main.tex
Comments on the evaluations

@@
 
-\medskip \noindent {\bf Experimental Design:} Figure \ref{fig:benchmark-design}(a) overviews the experimental design.
+\medskip \noindent {\bf Experimental Design:}~\figref{benchmark-design}(a) overviews the experimental design.
 Four mixtures (i.e., conditions) were created with varying amounts of human lysate, background {\it E.
@@
 In particular, we distinguished the changes in the abundances of the modified peptides (i.e., unadjusted changes) and the changes of their abundance relative to the changes in the abundances of the human lysate (i.e., protein-level adjusted changed).
-The true log-fold changes between the relevant components of the relevant mixtures are summarized in Figure \ref{fig:benchmark-design}(b).
+The true log-fold changes between the relevant components of the relevant mixtures are summarized in~\figref{benchmark-design}(b).
 Two replicate mixtures were created per condition.
@@
 
-The performance of the proposed method was evaluated on simulated and spike-in datasets with known ground truth, as well as biological experimental data where the ground truth was not known.
-When ground truth was known, we calculated the number of true positives (TP), false positives (FP), true negatives (TN), and false negatives (FN).
-True positives were PTMs that were expected to be differential after accounting for changes in global protein abundance.
-True negatives were PTMs that were not expected to be differential after accounting for changes in global protein abundance.
-To determine positives and negatives we adjusted the p-values for multiple testing and used an $\alpha=.05$ cutoff.
-These results were then compared to the ground truth to determine TP and TN.
-Summary statistics were calculated, including the false discovery rate (FDR) (FP/(TP+FP)), recall rate (TP/(TP + FN), and accuracy ((TP + TN) / (TP + TN + FP + FN)).
-Finally, the summary statistics were compared across methods to analyze method performance.
-
-For biological experiments where ground truth was not known, we adjusted the p-values for multiple testing and used an $\alpha=.05$ cutoff to determine differential sites.
-Comparisons were made before and after adjusting for changes in unmodified protein abundance and differences between the comparisons were noted.
-Each biological dataset presented a different experimental scenario, showing the applicability of the proposed method on a variety of experimental designs.
-All datasets in this paper were evaluated using the $MSstatsPTM$ package.
-
-Further details reviewing the results of each experimental analysis can be seen in Supplementary Sec. \ref{supp-sec:experiments}.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection*{MSstatsPTM corrected for high FDR in computer simulations}
-
-The proposed approach was evaluated against two-sample $t$-test and Limma in computer simulations. 
-In Figure \ref{fig:sim1_fdr} we see that not accounting for changes in the unmodified protein resulted in a high FDR.
-After adjusting for changes in the unmodified protein, all methods performed similarly in terms of FDR.
-In a clean simulation, recall and accuracy of the proposed approach and Limma performed similarly, although Limma performed slightly better in simulations with fewer biological replicates.
-In contrast, the performance of the $t$-test method lagged behind the others.
-Two-sample $t$-test only uses data within the groups of interest while ignoring the remainder of the data, leading to similar performance across all simulations.
-In contrast, the proposed approach and Limma leverage information across comparisons, resulting in improved power in simulations with more comparisons.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection*{MSstatsPTM outperformed existing methods in simulations with missing values and low features}
-
-In simulations with missing values and low features the proposed method outperformed $t$-test and Limma.
+We evaluated $MSstatsPTM$ on simulated and spike-in datasets with known ground truth in terms of true positives ($TP$), false positives ($FP$), true negatives ($TN$), and false negatives ($FN$) differentially abundant PTM.
+The true positives were defined as PTMs with changes distinct from the overall changes in protein abundance.
+The true negatives were defined as PTM which, after accounting for the changes in the overall protein abundance, were not differentially abundant.
+Additional summaries were the false discovery rate $FDR=FP/(TP+FP)$, $Recall=TP/(TP + FN)$, and $Accuracy=(TP + TN) / (TP + TN + FP + FN)$.
+For biological experiments with unknown ground truth, we compared the differentially abundant PTM  with and without adjusting for changes in unmodified protein abundance.
+
+The evaluations were based on the $MSstatsPTM$ \todo{any details of the options?
+E.g., was Empirical Bayes moderation used in MSstatsPTM?} $MSstatsPTM$ was compared \todo{with methods XXX - why chosen?
+We did not compare to IsobarPTM because XXX.
+Do we say t-test or ANOVA?} All the evaluations were done at the adjusted p-value cutoff of $\alpha=.05$.
+More details are in Supplementary Sec.
+\ref{supp-sec:experiments}.
+
+\todo{OV:Most of the following sections simply re-state the results that can already be read off the figures.
+The sections would have more value if in addition/instead of describing the figures, we could provide the reasons/insight into why we see this performance.
+Could you try?
+In general, no need to repeat what we see in the figure - this is what the figures are for.}
+
+\todo{OV: Also, maybe combine two sections that make a similar point?}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsubsection*{In clean simulations, MSstatsPTM corrected for high FDR of differentially abundant PTM}
+
+\todo{OV: maybe say something about the fact that this is a group comparisons design that doesn't require many variance components, and as such is favorable to Limma and t-test?
+Later when you have repeated measurements, clarify that these situations are more challenging for the existing methods and therefore the difference in the results is more pronounced.}
+
+\figref{sim1_fdr} illustrated the high FDR observed on the simulated datasets when not accounting for the overall changes in protein abundance.
+After the adjustment, the FDR of all methods improved.
+In a clean simulation, recall and accuracy of $MSstatsPTM$ and $Limma$ performed similarly.
+\todo{As expected when not using EB moderation in MSstatsPTM?,} Limma performed slightly better in simulations with fewer biological replicates.
+In contrast, the performance of the two-sample $t$-test method lagged behind.
+This is due to the fact that the $t$-test only uses data within the groups of interest while ignoring the remainder of the data.
+In contrast, $MSstatsPTM$ and Limma leverage information across comparisons, resulting in improved power in simulations with more comparisons.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsubsection*{In noisy simulations, MSstatsPTM \todo{had a higher metric XXX than} the existing methods}
+
+\todo{This section just describes the figure.
+Could you emphasize why we see these results?
+What are the reasons for differences in performance?} In simulations with missing values and few features, $MSstatsPTM$ outperformed $t$-test and $Limma$.
 Changes in unmodified protein level still needed to be accounted for to control the FDR.
-Once controlling for changes in the unmodified protein, the proposed method outperformed the other methods, as seen in Figure \ref{fig:sim2_acc}.
+Once controlling for changes in the unmodified protein, the proposed method outperformed the other methods, as seen in~\figref{sim2_acc}.
 The proposed method calibrated model accuracy and recall rate well, even when the number of replicates were low.
-Additionally, when comparing the fold change estimation across all modified peptides, the proposed method showed a tighter distribution of estimated fold changes around the true fold change, as compared to $t$-test and Limma.
-Specifically, the inner quartile range (IQR) of the estimated fold change for the proposed method was on average 21.8\% smaller than Limma's IQR and 10.4\% smaller than $t$-test's IQR.
-This fold change comparison can be seen in Supplementary Figure \ref{supp-fig:fc_boxplot}.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection*{In a label-free benchmark experiment MSstatsPTM corrected for bias and outperformed existing methods}
-
-In this experiment all models incorrectly estimated the fold change of the modified spike-in peptides before adjusting for changes in global protein abundance.
-After adjustment, the spike-in peptides' fold change was generally in line with expectation in all methods, however the distribution of estimated fold changes was visibly wider (Figure \ref{fig:spikein_boxplot}).
+Additionally, when comparing the fold change estimation across all modified peptides, the proposed method showed a tighter distribution of estimated fold changes around the true fold change, as compared to $t$-test and $Limma$.
+Specifically, the inner quartile range (IQR) of the estimated fold change for the proposed method was on average 21.8\% smaller than $Limma$ IQR and 10.4\% smaller than $t$-test's IQR.
+This fold change comparison can be seen in {\bf Supplementary Figure~\ref{supp-fig:fc_boxplot}}.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsubsection*{In the label-free benchmark experiment, MSstatsPTM \todo{had a higher metric XXX than} the existing methods}
+
+\todo{This section just describes the figure.
+Could you emphasize why we see these results?
+What are the reasons for differences in performance?} In this experiment all models incorrectly estimated the fold change of the modified spike-in peptides before adjusting for changes in global protein abundance.
+After adjustment, the spike-in peptides' fold change was generally in line with expectation in all methods, however the distribution of estimated fold changes was visibly wider (\figref{spikein_boxplot}).
 Of the three approaches, the proposed method showed the tightest distribution around the true log fold change.
-Comparing the IQR of the spike-in peptide's log fold change, the proposed method's IQR was 36.78\% smaller than Limma's and 32.98\% smaller than $t$-test's.
+Comparing the IQR of the spike-in peptide's log fold change, the proposed method's IQR was 36.78\% smaller than $Limma$ and 32.98\% smaller than $t$-test's.
 While Limma and $t$-test generally estimated the correct fold change, the proposed method's estimation was more consistent across all modifications.
 
-In Figure \ref{fig:spikein_prop_volcano} we can clearly see the fold change of the red labeled spike-in peptides was only correctly estimated when accounting for changes in the unmodified protein abundance.
+In~\figref{spikein_prop_volcano} we can clearly see the fold change of the red labeled spike-in peptides was only correctly estimated when accounting for changes in the unmodified protein abundance.
 Additionally, the background peptides, serving as the null model, show many false positives before adjustment.
@@
 Specifically, for the proposed method, the number of false positives went from 20.88\% to 1.84\% after adjustment was applied.
-While the proposed method and Limma both correctly estimated the fold change of the spike-in peptides, using Limma resulted in many insignificant adjusted p-values.
-Using Limma would have resulted in missing the majority of differential PTMs.
+While the proposed method and $Limma$ both correctly estimated the fold change of the spike-in peptides, using $Limma$ resulted in many insignificant adjusted p-values.
+Using $Limma$ would have resulted in missing the majority of differential PTMs.
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection*{MSstatsPTM corrected for bias in group comparison TMT experiment}
-
-The results of this experiment are summarized in Figure \ref{fig:data4_plots}.
-Figure \ref{fig:data4_venn_diagram} shows the number of significant modified peptides before and after adjustment.
-More peptides became insignificant after adjustment than became significant.
+\subsubsection*{In the group comparison TMT experiment, MSstatsPTM corrected for bias \todo{OV: 'corrected for bias' is unclear. We talk about confounding. Maybe use a more specific term?}}
+
+\todo{This section just describes the figure.
+Could you emphasize why we see these results?
+What are the reasons for differences in performance?} The results of this experiment are summarized in~\figref{data4_plots}.
+\figref{data4_venn_diagram} shows the number of significant modified peptides before and after adjustment.
+More peptides lost than gained differential abundance after the adjustment.
 3,236 modified peptides became insignificant, 1,352 became significant, while 4,282 were significant in both models.
@@
 
-In Figure \ref{fig:data4_profile_plot}, the modification of protein $GSDMD$ at site $K62$ showed the advantage of the proposed method.
+In~\figref{data4_profile_plot}, the modification of protein $GSDMD$ at site $K62$ showed the advantage of the proposed method.
 The modified peptide originally showed no abundance change between the infected 1 hour, 4 hour, and 6 hour conditions.
@@
 
-\subsubsection*{MSstatsPTM removed confounding in two mixture TMT experiment}
-
-The results of this experiment are summarized in Figure \ref{fig:data5_plots}.
+\subsubsection*{In the two-mixture TMT experiment, MSstatsPTM removed confounding \todo{OV: between what and what? Is it similar to the previous section?}}
+
+\todo{This section just describes the figure.
+Could you emphasize why we see these results?
+What are the reasons for differences in performance?} The results of this experiment are summarized in Figure \ref{fig:data5_plots}.
 Figure \ref{fig:data5_venn_diagram} shows the number of significant modified peptides before and after adjustment.
@@
 
-\subsubsection*{In label-free experiment without a separate global profiling run, MSstatsPTM eliminated the bias due to changes in the unmodified protein, albeit less effectively than in the presence of a global profiling run}
-
-As discussed in Section \ref{sec:exp_proc_dataset6}, there was no unmodified global profiling run performed in this experiment.
+\subsubsection*{In label-free experiment without a separate global profiling run, MSstatsPTM eliminated the confounding due to changes in the unmodified protein, albeit less effectively than in the presence of a global profiling run}
+
+\todo{This section just describes the figure.
+Could you emphasize why we see these results?
+What are the reasons for differences in performance?}  As discussed in \secref{exp_proc_dataset6}, there was no unmodified global profiling run performed in this experiment.
 Once identification and quantification of the Ubiquitination profiling was performed, peptides which were unmodified were extracted and used in place of a global profiling run.
@@
 
-The results of this experiment are summarized in Figure \ref{fig:data6_plots}.
+The results of this experiment are summarized in~\figref{data6_plots}.
 After adjusting for changes in the unmodified protein, there were fewer significant modified peptides.

### 2022-07-02 main/MSstatsPTM_Main.tex
Small typos
More small typos

@@
 \begin{eqnarray}
-\widehat{SE}(\Delta)_{adj} = \sqrt{ \widehat{SE}(\hat{\Delta}_{PTM})^{2} + \widehat{SE}(\hat{\Delta}_{protein})^{2}} 
+\widehat{SE}(\hat{\Delta})_{adj} = \sqrt{ \widehat{SE}(\hat{\Delta}_{PTM})^{2} + \widehat{SE}(\hat{\Delta}_{protein})^{2}} 
 \label{eq:prop_se}
@@
 \begin{equation}
-\widehat{SE}(\Delta)_{adj}^2 = \left[ \frac{2}{J} \left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right) \right]
+\widehat{SE}(\hat{\Delta})_{adj}^2 = \left[ \frac{2}{J} \left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right) \right]
 \leq \left( \frac{\Delta}{t_{1-\beta, df} + t_{1-\alpha /2, df}} \right)^{2}
@@
 
-Solving for J, the number of biological replicates per group is calculated as
+Solving for $J$, the number of biological replicates per group is calculated as
 
@@
 
-Details on how the proposed method can be used to run sample size calculations and power analysis on experiments with complex design can be seen in Supplementary Sec.
-\ref{supp-sec:complex_design}.
+Details on how the proposed method can be used to run sample size calculations and power analysis on experiments with complex design can be seen in {\bf Supplementary Sec.
+\ref{supp-sec:complex_design}}.
 More details on sample size calculation can be found in~\cite{oberg_vitek_09a}.
@@
 Do we say t-test or ANOVA?} All the evaluations were done at the adjusted p-value cutoff of $\alpha=.05$.
-More details are in Supplementary Sec.
-\ref{supp-sec:experiments}.
+More details are in {\bf Supplementary Sec.
+\ref{supp-sec:experiments}}.
 

### 2022-07-02 main/MSstatsPTM_Main.tex
More general advice in Evaluation

@@
 
+\todo{OV: also avoid the terms 'significant'. Replace it with 'differentially abundant'. Instead of 'bias' and 'convolution', use 'confounding'}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 Additionally, the background peptides, serving as the null model, show many false positives before adjustment.
-After adjustment the number of false positives decreased significantly.
+After adjustment the number of false positives substantially decreased.
 Specifically, for the proposed method, the number of false positives went from 20.88\% to 1.84\% after adjustment was applied.
-While the proposed method and $Limma$ both correctly estimated the fold change of the spike-in peptides, using $Limma$ resulted in many insignificant adjusted p-values.
+While the proposed method and $Limma$ both correctly estimated the fold change of the spike-in peptides, using $Limma$ resulted in many large adjusted p-values.
 Using $Limma$ would have resulted in missing the majority of differential PTMs.

### 2022-07-07 main/MSstatsPTM_Main.tex
updated spikein section

@@
 The second simulation introduced real-world characteristics, such as limited modified features and missing values.
-Details of computer simulations are available in Supplementary Section 2 and on GitHub (\url{https://github.com/devonjkohler/MSstatsPTM_simulations}).
-\todo{OV: there is a way to cross-reference sections dynamically across files.
-Check Meena's MSstats paper, or just double-check that all the sections are referenced correctly in the end}
+Details of computer simulations are available in Supplementary Section \ref{supp-sec:sim} and on GitHub (\url{https://github.com/devonjkohler/MSstatsPTM_simulations}).
 
@@
 \medskip \noindent {\bf Evaluation:} We expect the spike-in peptides to change corresponding to the values in Figure \ref{fig:benchmark-design}(b).
-These peptides are treated as positive controls in all comparisons except Mix 1 vs Mix 4.
-Mix 1 vs Mix 4 was treated as a negative control, with no expected fold change.
+These peptides are treated as positive controls in all comparisons except Mix 1 vs Mix 4, which was treated as a negative control.
 Additionally, the background {\it E.
-coli} lysate peptides were also treated as negative controls and are not expected to change in any comparison.
-Therefor we evaluated the statistical methods ability to avoid false positives, as well as their sensitivity in detecting the differentially abundant spike-in peptides and accurately estimate their expected fold change.
-
-\todo{Here is the place to describe: 'we expect XXX to not change'.
-Therefore we evaluated the ability of statistical methods to avoid false positives (i.e., specificity).
-We expect XXX to have fold changes XXX between conditions XXX.
-We evaluated the accuracy of estimation of fold changes...
-and the sensitivity of detecting differentially abundant...}
-%
-\todo{I think some mixtures are not comparable because the total protein amount is not the same.
-I believe only in Mix 1 vs 2 and Mix 3 vs 4 are comparable.
-I do not think it is correct to compare Mix 1 vs 3 and 2 vs 4 directly because they differ in the total amount of the proteome.
-Or was it normalized based on what we know is the same?
-If yes, we need to describe in the beginning of 'Evaluation'}  \todo{The unadjusted proteins...
-estimation...
-The adjusted proteins...
-estimation...
-Positive controls...
-Negative controls...}
+coli} lysate peptides were treated as negative controls and were not expected to change in any comparison.
+We evaluated the statistical methods ability to avoid false positives, as well as their sensitivity in detecting the differentially abundant spike-in peptides and accurately estimate their expected fold change.
 
@@
 %\hline
-%Comparison & Unadjusted true fold change & Adjusted true fold change \\ \hline \hline
-%mix1-mix2  & 0                           & -1                        \\ \hline
-%mix1-mix3  & 0                           & 1                         \\ \hline
+%Comparison & Unadj. true FC & Adj. true FC \\ \hline \hline
+%mix1-mix2  & -1                           & 1                        \\ \hline
+%mix1-mix3  & 0                           & -1                         \\ \hline
 %mix1-mix4  & -1                          & 0                         \\ \hline
-%mix2-mix3  & -1                          & -2                        \\ \hline
+%mix2-mix3  & 1                          & -2                        \\ \hline
 %mix2-mix4  & 0                           & -1                        \\ \hline
-%mix3-mix4  & 0                           & 1                         \\ \hline
+%mix3-mix4  & -1                           & 1                         \\ \hline
 %\end{tabular}
@@
 Coli} lysate, and spike-in peptides.
-\todo{TODO: Clarify what is changing how} (b) The expected fold change of the spike-in peptides, after adjusting for changes in the unmodified peptide (human lysate).
-The expected fold change is calculated using the values of the spike-in and human lysate in part (a).\todo{Could you make the columns in the table more complete?
-True fold-fold changes between the backgrounds; unadjusted true fold changes, adjusted true fold changes - essentially any value for any comparison that we may want to evaluate}}
+Volumes of each mixture totaled 20mg, with 10mg of background {\it E.
+Coli} lysate used to normalize mix 3 and mix 4 to make all mixtures directly comparable.
+2 pmol of spike-in peptides was added to mix 1 and mix 3, while 1 pmol was added to mix 2 and mix 4.
+(b) The expected fold change of the spike-in peptides in each comparison.
+``Unadj.
+true FC'' is the fold change of the spike-in peptides (red) without adjusting for changes in human lysate (blue).
+This was calculated by directly comparing the amounts of spike-in in each condition.
+``Adj.
+true FC'' is the fold change of the spike-in peptides when adjusting for changes in human lysate.
+This was calculated by determining the ratio of spike-in peptides to human lysate in each mixture and then using the ratio to calculate the true fold change across comparisons.}
 

### 2022-07-07 supplementary/ptm_sm.tex
updated spikein section

@@
 
-The proposed approach applied to experiments with complex designs still allows for sample size calculation and power analysis.
+%The proposed approach applied to experiments with complex designs still allows for sample size calculation and power analysis.
 To make the calculation the standard deviation and degrees of freedom are needed.
@@
 
-The standard error and degrees of freedom are estimated separately for both the PTM and unmodified protein models.
+%The standard error and degrees of freedom are estimated separately for both the PTM and unmodified protein models.
 Once this is done we can use these these values to calculate the sample size and perform power analysis.
@@
 The minimal number of replicates is determined by the variance of the estimated log-fold change $\mathrm{SE}^{2}(\hat{\Delta})$ as
-\[
-\mathrm{SE}^{2}(\hat{\Delta}) = \left[ \left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right) \right]
-\leq \left( \frac{\Delta}{t_{1-\beta, df} + t_{1-\alpha /2, df}} \right)^{2},
-\]
-where 
-\[
-\alpha = (1 - \beta) \cdot \frac{q}{1 + (1-q) \cdot m_0 / m_1},
-\]
-and $t_{1-\beta, df}$ and $t_{1-\alpha /2, df}$ are the $100(1-\beta)^{\text{th}}$ and the $100(1-\alpha /2)^{\text{th}}$ percentiles of the $t$ distribution.
+%\[
+%\mathrm{SE}^{2}(\hat{\Delta}) = \left[ \left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right) \right]
+%\leq \left( \frac{\Delta}{t_{1-\beta, df} + t_{1-\alpha /2, df}} \right)^{2},
+%\]
+%Where 
+%\[
+%\alpha = (1 - \beta) \cdot \frac{q}{1 + (1-q) \cdot m_0 / m_1},
+%\]
+%and $t_{1-\beta, df}$ and $t_{1-\alpha /2, df}$ are the $100(1-\beta)^{\text{th}}$ and the $100(1-\alpha /2)^{\text{th}}$ percentiles of the $t$ distribution.
 The degrees of freedom are estimated for the appropriate model using restricted maximum likelihood.
@@
 \subsubsection{Computer simulation}
-
+\label{sec:comp_sim}
 Differential levels of modified peptides may be due to differential modifications and/or changes in protein abundance.

### 2022-07-07 main/MSstatsPTM_Main.tex
Updated methods sections

@@
 \medskip \noindent
-\todo{OV: this doesn't look right, as it seems that all the features and replicates equally contribute to the error.
-Do people do some kind of summarization of the features first?
-Let's first describe the summarization}
-\todo{DK: I think we should just delete this part entirely.
-It isn't how we compared the method and is not really used in the literature from what I can tell.} \todo{OV: below in the results we use the notation $y_{i++}$ etc.
-Alternatively, since we put a hat on $\hat{y}_{ij}$, maybe we do not need '++', as it is clear that this is a summary.
-I removed '++' in the results section for now.
-Let's use the same notation in this section consistently}
 
@@
 The model allows us to estimate $\hat{\Delta}_{adj}$ and its standard error.
-Based on this model, we can test the more relevant null hypothesis in~\eqref{null_hyp}, by comparing the test statistic against the Student distribution with \todo{how many} degrees of freedom in balanced designs.
+Based on this model, we can test the more relevant null hypothesis in~\eqref{null_hyp}, by comparing the test statistic against the Student distribution with \todo{$df=2J-2$} \todo{this should be the same df as the previous section?} degrees of freedom in balanced designs.
 
@@
 Technological artifacts such as missing values further undermine the calculation of $u_{ij}$ in~\eqref{adj_estimation}.
-Finally, the lack of a self contained, straightforward implementation of the method challenges the applications \todo{OV: is the last statement true?
-Seems simple enough}.
-\todo{DK: I meant more that there was no simple package implementation.
-Researchers need to program the methods themselves.
-I cant find an implementation at least}
+Finally, their is no self contained, straightforward implementation of the method, such as in the form of a coding package, requiring a more challenging manual application.
 
@@
 The specific model instances depend on the experimental design and data acquisition.
-For example, for an experiment with \todo{group comparison or time course design?} $MSstatsTMT$ fits the following model \todo{OV: could we keep the notation as consistent as possible across this section?
-Above conditions had subscripts $i$ and bio reps had subscripts $j$.
-Could we keep a consistent notation below (or change the notation elsewhere in the manuscript?)} \todo{Should we maybe give a model for label-free, to stay consistent with the example in figure 2?
-Just use $I$ and $J$ etc for the indices...}
+For example, to model the unmodified protein in \figref{data-structure} $MSstats$ fits the following model
 \begin{eqnarray}
-& y_{mtcb} = \mu + Mixture_m + TechRep(Mixture)_{t(m)} + Condition_c + Subject_{mcb} + \epsilon_{mtcb}, & \\ 
-& \text{where }  Mixture_m  \mathop\sim^{iid} \mathcal{N}(0, \sigma^2_M),  TechRep(Mixture) \mathop\sim^{iid} \mathcal{N}(0, \sigma^2_T), &  \nonumber \\
-& \sum\limits_{c=1}^{C} Condition_c = 0,\ \todo{Subject \stackrel{iid}{\sim}\mathcal{N}(0,?)}\  \epsilon_{mtcb} \mathop\sim^{iid} \mathcal{N}(0, \sigma^2) & \nonumber
+& y_{ij} = \mu + Condition_i + Run_{j} + \epsilon_{ij}, & \\ 
+& \text{where } \sum\limits_{i=1}^{J} Condition_i = 0,\ Run \mathop\sim^{iid} \mathcal{N}(0, \sigma^2_j), \epsilon_{ij} \mathop\sim^{iid} \mathcal{N}(0, \sigma^2) & \nonumber
+\label{eq:msstats_model}
 \end{eqnarray}
 Parameters of the model are estimated using restricted maximum likelihood (REML)~\cite{Kenward}.
-In turn, the parameters allow us to estimate the relevant quantity $\hat{\Delta}_{Protein}$ \todo{this has not been defined, Define here, using the notation above?} and its standard error.
+In turn, the parameters allow us to estimate the relevant quantity $\hat{\Delta}_{Protein}$ and its standard error. Where $\hat{\Delta}_{Protein}$ is estimated as
+
+\begin{eqnarray}
+\Delta_{Protein} = \mu_{i} - \mu_{i^{\prime}}
+\end{eqnarray}
+
 Similarly to $Limma$, $MSstatsTMT$ includes an optional Empirical Bayes moderation of the standard error~\cite{Huang:2020}.
-\todo{OV: maybe elaborate on that?} Based on the model, the approach tests the null hypothesis in \eqref{conv_null_hyp}, by comparing the test statistic against the Student distribution.
+Using Empirical Bayes helps overcome cases where the number of biological replicates in each condition is small.
+Based on the model, the approach tests the null hypothesis in \eqref{conv_null_hyp}, by comparing the test statistic against the Student distribution.
 
@@
 First, unlike \todo{t-test (or ANOVA?) and $Limma$}, the family of linear mixed effects models is applicable to many complex experimental designs, including designs with multiple sources of variation, and unbalanced designs.
-Second, the approach is applicable to various data acquisition types, including label-free DDA and DIA, and experiments with TMT labeling, and is compatible with various data processing tools such as \todo{XXX}
+Second, the approach is applicable to various data acquisition types, including label-free DDA and DIA, and experiments with TMT labeling, and is compatible with various data processing tools such as Skyline, Spectronaut, MaxQuant, Progenesis, Proteome Discoverer, and OpenMS.
 Third, the run-level summarization has been shown to be more robust to outliers and missing values as compared to feature averages~\cite{Choi:2014}.
-Finally, the custom $MSstats$ and $MSstatsTMT$ implementation accounts for potential data artifacts, and is numerically scalable and stable, and \todo{is available through both command line and a dedicated graphical user interface}.
+Finally, the custom $MSstats$ and $MSstatsTMT$ implementation accounts for potential data artifacts, and is numerically scalable and stable, and is available through both command line and a dedicated graphical user interface.
+
 Unfortunately, MSstats estimates the difference in overall protein abundance and tests the null hypothesis in~\eqref{conv_null_hyp}.
@@
 We leverage the existing methods in $MSstats$ and $MSstatsTMT$ and extend them to remove confounding between changes in the PTM and the overall changes in protein abundance.
-Each feature type is summarized separately using $MSstats$ or $MSstatsTMT$, resulting in run level values for the PTM, \todo{$\hat{y}_{ij}^{\ast}$} \todo{OV: I am confused: since we summarize per run, we need at least an index for a condition and an index for bio rep.
-Could we use this consistently across all the sections?}, and unmodified peptide.
-The estimated summaries of the modified features are then used as the input to the model \ov{such as in eq XXX} in $MSstats$ or $MSstatsTMT$.
+Each feature type is summarized separately using $MSstats$ or $MSstatsTMT$, resulting in run level values for the PTM, $\hat{y}_{ij}^{\ast}$, and unmodified peptide.
+The estimated summaries of the modified features are then used as the input to the model, such as in \eqref{msstats_model}, in $MSstats$ or $MSstatsTMT$.
 Similarly, the estimated summaries of the unmodified features are used as the input to the corresponding model in $MSstats$ or $MSstatsTMT$.
@@
 The estimated standard error is larger than the standard errors associated with each distinct feature type, reflecting the increased uncertainty inherent to combining the uncertainty in two estimators.
-Finally, the degrees of freedom associated with~\eqref{prop_se} are obtained via the Satterthwaite approximation~\cite{satterthwaite:1946}\todo{OV: maybe site a more recent exp design book?}
+Finally, the degrees of freedom associated with~\eqref{prop_se} are obtained via the Satterthwaite approximation~\cite{kutner_etal_04a, satterthwaite:1946}
 \begin{eqnarray}
@@
 It is applicable to label-free data acquisition strategies such as DDA and DIA and experiments with TMT labeling.
-It is compatible with data processing tools \todo{XXX}.
 The proposed approach is also applicable to experiments where the experimental designs for PTM profiling and global proteome profiling vary in properties such as number of biological replicates, data acquisition strategies and runs.
-\todo{OV: We do have a requirement that the scale of measurements is comparable between TMT and PTM, correct?
-Otherwise the SE may not be comparable}
 
@@
 
-\todo{OV: I feel that some parts of this section are either specific to MSstats/MSstatsTMT, or a bit redundant with the methods.
-For the parts that we do not implement anew but reuse, could you move them to the 'Existing methods'/MSstats section?
-Here let us focus only on things that we had to develop from scratch for this project.
-If we implement the methods, we do not need to re-describe the methods}
-
-\todo{OV: could you make this section more specific to PTM converters?
-E.g., I remember that the input should be different, there are some parts relevant to the localization of PTMs.
-This was not part of the base MSstats.
-Could you emphasize that?} The proposed methods are implemented in the open source R package $MSstatsPTM$.
+The proposed methods are implemented in the open source R package $MSstatsPTM$.
 An overview of the steps of the package are illustrated in Figure \ref{fig:msstatsptm_design}.
 $MSstatsPTM$ includes converters for multiple spectral processing tools, including MaxQuant, Progenesis, and Spectronaut.
-The converters take as input the raw data from the tool, identify the modification site for modified peptide, and put the data into the correct format for analysis in $MSstatsPTM$.
 Conversion is done separately for the modified peptide and global profiling runs.
+To identify the local modification site we require the tool to identify where in the sequence the modification occurred.
 If the global profiling information is not available, the package can still analyze the modified run, but will do so without adjusting for changes in unmodified protein abundance.
-This amounts to testing the null hypothesis in~\eqref{conv_null_hyp}; which amounts to running the methods seen in $MSstats$ on modified peptide level data \cite{Choi:2014} \cite{Huang:2020}.
-
-\todo{OV: this is the part that looks redundant with the MSstats section.
-In the MSstats section, do not repeat the methods, just focus on the specifics of the implementation.
-Relavant parts would be that we fit simplified models if the base model is not applicable.
-In this section,  could you just say that we do the same as MSstats (including fitting simplified models).
-But there are some caveats, e.g.
-assumptions of missing values due to low abundance are not necessarily appropriate here so we recommend to not use that option?
-Also specialized sample size calculations} After using the converters, the next step is peptide/protein summarization and missing value imputation.
-For the modified run, the package summarizes features which include the same modification together.
-Features with multiple modification sites are not included with single site features and are summarized separately.
-For the global profiling run all unmodified features from the same protein are summarized together, up to the protein level.
-Additionally, the summarization includes global median normalization and normalizes between MS runs.
-The package can optionally impute missing values using an AFT model~\cite{Wei:1992}.
+This amounts to testing the null hypothesis in~\eqref{conv_null_hyp}; which amounts to running the methods seen in $MSstats$ on modified peptide level data.
+
+The next step is peptide/protein summarization.
+Summarization is performed separately for the PTM and global protein in the same way as described above for $MSstats$.
+The package can optionally impute missing values using an Accelerated Failure Time (AFT) model~\cite{Wei:1992}.
 Missing value imputation should be performed with care when applied to PTM data.
@@
 These issues can cause the missing value imputation to be unstable.
-
-The final step of the package is modeling the summarized dataset.
-A linear mixed effects model is fit for both the summarized modified and global profiling runs.
+Finally, a separate model is fit to both the modified and global profiling runs.
 This model is automatically adjusted depending on the experimental design and acquisition method.
-The comparisons of interest can either be predefined or a full pairwise comparison will be tested.
-After fitting a model to both the modified and unmodified data, the modified model is adjusted for changes in unmodified protein abundance, using the methods described above.
-
-\todo{OV: similarly, some of this belongs to MSstats.
-Could you focus on project-specific implementation.
-E.g., we separately visualize modified and unmodified peptides.
-(Btw, sequence-specific visualization such as the one you did for MSstatsLiP would be awesome if there is a chance)} Beyond the core functionality, the package also includes functions for plotting the results.
-These include summarization plots to help with quality assurance and identifying sources of variation (such as quality control plots and profile plots), as well as modeling plots (such as volcano plots and heatmaps), which evaluate the fold change between conditions and peptides.
+If the base model is not applicable, a simplified model is fit.
+After modeling, the modified model is adjusted for changes in unmodified protein abundance, using the methods described above.
+
+We separately visualize the PTM and global protein side by side, allowing for easier interpretation and comparison across the datasets.
+In the modeling stage this is further separated into PTM, global protein, and adjusted PTM visualizations.
+This allows users to see the affect of the adjustment.
 
@@
 
-The evaluations were based on the $MSstatsPTM$ \todo{any details of the options?
-E.g., was Empirical Bayes moderation used in MSstatsPTM?} $MSstatsPTM$ was compared \todo{with methods XXX - why chosen?
-We did not compare to IsobarPTM because XXX.
-Do we say t-test or ANOVA?} All the evaluations were done at the adjusted p-value cutoff of $\alpha=.05$.
+The evaluations were based on the $MSstatsPTM$, and were performed without imputing missing values and without using Empirical Bayes moderation.
+$MSstatsPTM$ was compared with \todo{$t$-test?} and Limma.
+We did not compare to IsobarPTM because it is only applicable to TMT experiments, meaning it could not be applied to our ground truth data.
+All the evaluations were done at the adjusted p-value cutoff of $\alpha=.05$.
 More details are in {\bf Supplementary Sec.
@@
 In a clean simulation, recall and accuracy of $MSstatsPTM$ and $Limma$ performed similarly.
-\todo{As expected when not using EB moderation in MSstatsPTM?,} Limma performed slightly better in simulations with fewer biological replicates.
+As expected when not using EB moderation in MSstatsPTM, Limma performed slightly better in simulations with fewer biological replicates.
 In contrast, the performance of the two-sample $t$-test method lagged behind.
@@
 In contrast, $MSstatsPTM$ and Limma leverage information across comparisons, resulting in improved power in simulations with more comparisons.
+Since this simulation is a group comparisons design that doesn't require many variance components, it is favorable to Limma and t-test
 
@@
 
-\todo{This section just describes the figure.
-Could you emphasize why we see these results?
-What are the reasons for differences in performance?} In simulations with missing values and few features, $MSstatsPTM$ outperformed $t$-test and $Limma$.
+\todo{This section just describes the figure. Could you emphasize why we see these results? What are the reasons for differences in performance?}
+
+In simulations with missing values and few features, $MSstatsPTM$ outperformed $t$-test and $Limma$.
 Changes in unmodified protein level still needed to be accounted for to control the FDR.
@@
 The proposed method calibrated model accuracy and recall rate well, even when the number of replicates were low.
+The proposed approach can handle low feature counts well due to using Tukey's Median Polish for summarization.
+The existing approaches summation approach require more observations to correctly summarize the intensities in the run, whereas TMP is more robust.
+\todo{more/better}
+
 Additionally, when comparing the fold change estimation across all modified peptides, the proposed method showed a tighter distribution of estimated fold changes around the true fold change, as compared to $t$-test and $Limma$.
 Specifically, the inner quartile range (IQR) of the estimated fold change for the proposed method was on average 21.8\% smaller than $Limma$ IQR and 10.4\% smaller than $t$-test's IQR.
+This means that while the mean of the estimated fold changes was generally correct for all methods, the proposed approach correctly estimated the fold change more often across all PTMs.
 This fold change comparison can be seen in {\bf Supplementary Figure~\ref{supp-fig:fc_boxplot}}.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection*{In the label-free benchmark experiment, MSstatsPTM \todo{had a higher metric XXX than} the existing methods}
-
-\todo{This section just describes the figure.
-Could you emphasize why we see these results?
-What are the reasons for differences in performance?} In this experiment all models incorrectly estimated the fold change of the modified spike-in peptides before adjusting for changes in global protein abundance.
+\subsubsection*{In the label-free benchmark experiment, MSstatsPTM had a higher sensitivity than the existing methods}
+
+In this experiment all models incorrectly estimated the fold change of the modified spike-in peptides before adjusting for changes in global protein abundance.
 After adjustment, the spike-in peptides' fold change was generally in line with expectation in all methods, however the distribution of estimated fold changes was visibly wider (\figref{spikein_boxplot}).
@@
 Using $Limma$ would have resulted in missing the majority of differential PTMs.
+
+This experiment introduced more variance components, which is challenging for the existing methods and therefore the difference in the results is more pronounced than in the simulations.
 
@@
 %Comparison & Unadj. true FC & Adj. true FC \\ \hline \hline
-%mix1-mix2  & -1                           & 1                        \\ \hline
+%mix1-mix2  & 1                           & 1                        \\ \hline
 %mix1-mix3  & 0                           & -1                         \\ \hline
-%mix1-mix4  & -1                          & 0                         \\ \hline
-%mix2-mix3  & 1                          & -2                        \\ \hline
+%mix1-mix4  & 1                          & 0                         \\ \hline
+%mix2-mix3  & -1                          & -2                        \\ \hline
 %mix2-mix4  & 0                           & -1                        \\ \hline
-%mix3-mix4  & -1                           & 1                         \\ \hline
+%mix3-mix4  & 1                           & 1                         \\ \hline
 %\end{tabular}

### 2022-07-07 main/MSstatsPTM_Main.tex
minor changes

@@
 \begin{eqnarray}
-\Delta_{Protein} = \mu_{i} - \mu_{i^{\prime}}
+\hat{\Delta}_{Protein} = \hat{\mu}_{i} - \hat{\mu}_{i^{\prime}}
 \end{eqnarray}
@@
 
-We separately visualize the PTM and global protein side by side, allowing for easier interpretation and comparison across the datasets.
-In the modeling stage this is further separated into PTM, global protein, and adjusted PTM visualizations.
-This allows users to see the affect of the adjustment.
+We include visualizations for QC and model assessment.
+In QC plots the PTM and global protein are shown side by side, allowing for easier interpretation and comparison across data.
+In the model plots this is further separated into PTM, global protein, and adjusted PTM visualizations, allowing users to see the affect of the adjustment.
 

### 2022-07-08 code/Spike-in_Analysis.Rmd
updated spikein sections, methods, and results

--- 2022-03-05 code/Spike-in_Analysis.Rmd
@@
 
-summary_df$PTM$RunlevelData
-
 load(file = "../data/Spike_in_MSstatsPTM_Model.rda")#model_df
@@
 ## Update some naming conventions
-plot_df[plot_df$Label == 'mix1-mix3'][['contrast_truth']] <- "LogFC: 1 (mix1-mix3)"
-plot_df[plot_df$Label == 'mix1-mix4'][['contrast_truth']] <- "LogFC: 0 (mix1-mix4)"
-plot_df[plot_df$Label == 'mix2-mix3'][['contrast_truth']] <- "LogFC: -2 (mix2-mix3)"
-plot_df[plot_df$Label == 'mix2-mix4'][['contrast_truth']] <- "LogFC: -1 (mix2-mix4)"
+plot_df[plot_df$Label == 'mix2 vs mix1'][['contrast_truth']] <- "LogFC: -1 (mix2-mix1)"
+plot_df[plot_df$Label == 'mix3 vs mix1'][['contrast_truth']] <- "LogFC: 1 (mix3-mix1)"
+plot_df[plot_df$Label == 'mix4 vs mix1'][['contrast_truth']] <- "LogFC: 0 (mix4-mix1)"
+plot_df[plot_df$Label == 'mix3 vs mix2'][['contrast_truth']] <- "LogFC: 2 (mix3-mix2)"
+plot_df[plot_df$Label == 'mix4 vs mix2'][['contrast_truth']] <- "LogFC: 1 (mix4-mix2)"
+plot_df[plot_df$Label == 'mix4 vs mix3'][['contrast_truth']] <- "LogFC: -1 (mix4-mix3)"
+plot_df$contrast_truth = factor(plot_df$contrast_truth, levels = 
+                                  c("LogFC: -1 (mix2-mix1)",
+                                    "LogFC: 1 (mix3-mix1)",
+                                    "LogFC: 0 (mix4-mix1)",
+                                    "LogFC: 2 (mix3-mix2)",
+                                    "LogFC: 1 (mix4-mix2)",
+                                    "LogFC: -1 (mix4-mix3)"))
 
 plot_df[plot_df$labeling == 'light'][['labeling']] <- "Background Peptides"
-plot_df[plot_df$labeling == 'heavy'][['labeling']] <- "SpikeIn Peptides"
+plot_df[plot_df$labeling == 'heavy'][['labeling']] <- "Spike-in Peptides"
 
@@
 # Volcano plot
-png("../supplementary/sim_new/spike_in_msstatsptm_volcano.png", width = 750, height = 500)
+png("../supplementary/sim_new/spike_in_msstatsptm_volcano.png", width = 1000, height = 500)
 plot_df %>% arrange(labeling) %>% 
-  filter(abs(log2FC) < Inf & Label %in% c("mix1-mix3", "mix1-mix4", "mix2-mix3", "mix2-mix4")) %>%
+  filter(abs(log2FC) < Inf) %>% # & Label %in% c("mix1-mix3", "mix1-mix4", "mix2-mix3", "mix2-mix4")
   ggplot(aes(log2FC, -log10(adj.pvalue))) + 
@@
   geom_hline(yintercept = 0) + 
-  geom_vline(xintercept = 0, linetype = "dashed") + 
+  geom_vline(xintercept = 0) + 
+  geom_vline(aes(xintercept = trueLog2FC),color = "darkred", linetype= "dashed") +
   geom_hline(yintercept = -log10(0.05), color = "darkred") + 

### 2022-07-08 main/MSstatsPTM_Main.tex
updated spikein sections, methods, and results

@@
 
-\subsection*{Dataset 3: SpikeIn benchmark - Ubiquitination - Label-free}
+\subsection*{Dataset 3: Spike-in benchmark - Ubiquitination - Label-free}
 \label{sec:exp_proc_dataset3}
@@
 
-\medskip \noindent {\bf Experimental Design:}  Luchetti et al.
-\cite{LUCHETTI2021} profiled human epithelial cells engineered to express IpaH7.8 under a dox inducible promoter.
+\medskip \noindent {\bf Experimental Design:}  Luchetti et al.~\cite{LUCHETTI2021} profiled human epithelial cells engineered to express IpaH7.8 under a dox inducible promoter.
 Uninfected cells were measured at 0 and 6 hours, while cells infected with {\it Shigella flexneri} ({\it S.
@@
 
-The basic two-sample $t$-test \todo{ref - ideally from this literature if you can find} is based on the model
+The basic two-sample $t$-test~\cite{kutner_etal_04a} is based on the model
 \begin{eqnarray}
@@
 
-\medskip \noindent The two versions of $t$-test \todo{OV: ANOVA?} above can be expanded with Empirical Bayes moderation in Limma~\cite{ Ritchie_15a, Schwammle2015, Smyth:2004, Smyth:2005, Zhu,Chappell:2021}.
-In particular, the ratio-based version models the ratios $\hat{u}_{ij}$ in~\eqref{adj_estimation}.
-Limma then fits a linear model using the transformed inputs.
-
-\begin{equation}
-\begin{aligned}
-\hat{u}_{ij} = \beta_0 + c_{i}\beta_1 + + r_j\beta_2 + \epsilon^{\prime\prime}_{ij}, \text{where } \epsilon^{\prime\prime}_{ij} \mathop\sim^{iid} \mathcal{N}(0, \sigma^{\prime\prime 2}/w_j), i \in \{1, 2\}, j \in \{1, 2\}
-\end{aligned}
+\todo{DK: Current plan is to follow what Ting did for MSstatsTMT and keep this part brief but expand on EB methods in the supplementary.}\medskip \noindent The two versions of $t$-test \todo{OV: ANOVA?} above can be expanded with Empirical Bayes moderation in Limma~\cite{ Ritchie_15a, Schwammle2015, Smyth:2004, Smyth:2005, Zhu,Chappell:2021}.
+In particular, the ratio-based version which models the ratios $\hat{u}_{ij}$ in~\eqref{adj_estimation} can be used to answer the null hypothesis in \eqref{null_hyp}.
+Limma uses the transformed inputs to fit a linear model with a fixed condition and run effect for each PTM.
+
+\begin{eqnarray}
+& u_{ij} = \mu + Condition_i + Run_{j} + \epsilon_{ij}, & \\ 
+& \text{where } \sum\limits_{i=1}^{I} Condition_i = 0,\ \sum\limits_{i=1}^{J} Run_j = 0, \epsilon_{ij} \mathop\sim^{iid} \mathcal{N}(0, \sigma^2) & \nonumber
 \label{eq:limma_model}
-\end{equation}
-\todo{Essentially, Limma can fit all kinds of models - with or without the bio replicate, including the ones in the equations above, it depends on the user.
-The only difference from the approaches above is the EB moderation.
-I am confused by $r_j$ do you mean $I_{biorep=j}$?
-Since replicate id takes continuous values, the model above is a regression with replicate id as a single predictor / 1 df - not sure that's right;  I am confused by $w_j$ - this is not how Limma model is specified.
-Also what are the $\beta$ and why we only have one of them?
-Let's check which versions of the models the references use and how Ting describes Limma in her paper.
-}
-
+\end{eqnarray}
+
+The method than uses Empirical Bayes moderation to combine the variance estimations across all PTMs modeled.
+This increases the degrees of freedom with which the individual PTM variances are estimated~\cite{Ritchie_15a}.
 The goal of Limma is to increase the sensitivity of detection of differential abundance when the experiment has few biological replicates.
-This is done by means of Empirical Bayes moderation, i.e.
-by borrowing the information regarding the variation across all the modification sites.
-Specifically, \todo{OV: this should be another model equation, rather than the text}
-
-For each modification site, the model yields the same estimate $\hat{\Delta}_{adj}$ as \eqref{ttest_model}, but a different standard error.
-To estimate $\widehat{SE(\hat{\Delta})}$, Limma leverages empirical bayes moderation~\cite{Robbins1992} to adjust the model variance of the individual PTM using the shared variance of all PTMs modeled in the experiment.
-\todo{DK: I introduced a new subscript here for the PTM p.
-It feel awkward to add it across the other methods so I only included it in the moderation describtion.} For a single PTM $p$ with an estimated fold change $\hat{\Delta}_p$, the estimated variance is calculated as $var(\hat{u}^{\ast}_{pij}) = \sigma^2_p / w_{pj}$ where $\sigma^2_p$ is the variance resulting from the linear model of PTM $p$ and $w_{pj}$ is the global variance across all PTMs in replicate $j$.
-The statistical significance is then determined using $\hat{\Delta}_p$ and $\widehat{SE(\hat{\Delta}_p)}$ calculated using the adjusted variance $var(\hat{u}^{\ast}_{pij})$.
 
@@
 Finally, there is no self contained implementation of the methods to PTMs, requiring manual transformation and application by the user.
-
-%Limma uses linear models to test the null hypothesis that there is no difference in mean PTM abundances between conditions.
-It leverages Empirical Bayes moderation to share pooled variance information across individual modification models and moderate the individual residual variances.
-Using a linear model allows Limma to share variance information across conditions, providing a more accurate estimate~\cite{Ritchie_15a}.
-With respect to PTM analysis,  Limma fits a linear model using the adjusted summarized features, $d_{1+}^{\ast}$ and $d_{2+}^{\ast}$, to estimate , $\Delta$.
-This results in the same hypothesis as seen in Equation \ref{eq:ttest_null_hyp}.
-The variance for the estimation is moderated using information from the variance across all PTMs modeled.
-For the estimate of a single PTM $m$, $\Delta_m$, the variance is estimated as
 
@@
 & y_{ij} = \mu + Condition_i + Run_{j} + \epsilon_{ij}, & \\ 
-& \text{where } \sum\limits_{i=1}^{J} Condition_i = 0,\ Run \mathop\sim^{iid} \mathcal{N}(0, \sigma^2_j), \epsilon_{ij} \mathop\sim^{iid} \mathcal{N}(0, \sigma^2) & \nonumber
+& \text{where } \sum\limits_{i=1}^{I} Condition_i = 0,\ Run \mathop\sim^{iid} \mathcal{N}(0, \sigma^2_j), \epsilon_{ij} \mathop\sim^{iid} \mathcal{N}(0, \sigma^2) & \nonumber
 \label{eq:msstats_model}
@@
 
-Similarly to $Limma$, $MSstatsTMT$ includes an optional Empirical Bayes moderation of the standard error~\cite{Huang:2020}.
-Using Empirical Bayes helps overcome cases where the number of biological replicates in each condition is small.
+Similarly to $Limma$, $MSstatsTMT$ includes an optional Empirical Bayes moderation of the standard error~\cite{Huang:2020}, which increases the sensitivity of the model when the number of biological replicates in each condition is small.
 Based on the model, the approach tests the null hypothesis in \eqref{conv_null_hyp}, by comparing the test statistic against the Student distribution.
@@
 In a clean simulation, recall and accuracy of $MSstatsPTM$ and $Limma$ performed similarly.
-As expected when not using EB moderation in MSstatsPTM, Limma performed slightly better in simulations with fewer biological replicates.
+As expected when not using Empricial Bayes moderation in MSstatsPTM, Limma performed slightly better in simulations with fewer biological replicates.
 In contrast, the performance of the two-sample $t$-test method lagged behind.
-This is due to the fact that the $t$-test only uses data within the groups of interest while ignoring the remainder of the data.
+This was due to the fact that the $t$-test only uses data within the groups of interest while ignoring the remainder of the data.
 In contrast, $MSstatsPTM$ and Limma leverage information across comparisons, resulting in improved power in simulations with more comparisons.
-Since this simulation is a group comparisons design that doesn't require many variance components, it is favorable to Limma and t-test
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection*{In noisy simulations, MSstatsPTM \todo{had a higher metric XXX than} the existing methods}
+Additionally, since this simulation is a group comparison design that doesn't require many variance components, it is favorable to Limma and t-test
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsubsection*{In noisy simulations, MSstatsPTM better calibrated the true fold change over the existing methods}
 
@@
 The proposed method calibrated model accuracy and recall rate well, even when the number of replicates were low.
-The proposed approach can handle low feature counts well due to using Tukey's Median Polish for summarization.
-The existing approaches summation approach require more observations to correctly summarize the intensities in the run, whereas TMP is more robust.
-\todo{more/better}
-
-Additionally, when comparing the fold change estimation across all modified peptides, the proposed method showed a tighter distribution of estimated fold changes around the true fold change, as compared to $t$-test and $Limma$.
+The proposed approach can handle low feature counts due to using Tukey's Median Polish for summarization.
+The existing methods summation approach requires more observations to correctly summarize the intensities in the run, whereas TMP is robust to few measurements.
+
+Additionally, when comparing the fold change estimation across all modified peptides, the proposed method showed a tighter distribution of estimated fold changes around the true fold change.
 Specifically, the inner quartile range (IQR) of the estimated fold change for the proposed method was on average 21.8\% smaller than $Limma$ IQR and 10.4\% smaller than $t$-test's IQR.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection*{In the group comparison TMT experiment, MSstatsPTM corrected for bias \todo{OV: 'corrected for bias' is unclear. We talk about confounding. Maybe use a more specific term?}}
-
-\todo{This section just describes the figure.
-Could you emphasize why we see these results?
-What are the reasons for differences in performance?} The results of this experiment are summarized in~\figref{data4_plots}.
-\figref{data4_venn_diagram} shows the number of significant modified peptides before and after adjustment.
-More peptides lost than gained differential abundance after the adjustment.
-3,236 modified peptides became insignificant, 1,352 became significant, while 4,282 were significant in both models.
-For the peptides that became insignificant in the adjusted model, their change in abundance was mainly due to changes in global protein abundance.
-In contrast, for peptides that became significant after adjustment, their true abundance change was masked by underlying changes in the unmodified protein.
-Both of these issues were corrected in adjustment, and the unconfounded abundance change is shown.
-An additional question that must be addressed is if the decrease in significant peptides is due to the increased variance that comes from adjustment.
-This was tested by looking for modified peptides whose adjusted log fold change was within 10\% of the unadjusted log fold change but became insignificant after adjustment.
+\subsubsection*{In two group comparison TMT experiments, MSstatsPTM corrected for confounding with the unmodified protein}
+
+\todo{This section just describes the figure. Could you emphasize why we see these results? What are the reasons for differences in performance?}
+
+The results of these experiments are summarized in~\figref{data4_plots} and~\figref{data5_plots}.
+\figref{data4_venn_diagram} and~\figref{data5_venn_diagram} show the number of differentially abundant modified peptides before and after adjustment.
+In both cases, more peptides lost than gained differential abundance after the adjustment.
+A question that must be addressed is if the decrease in differentially abundant peptides is due to the increased variance that comes from adjustment.
+This was tested by looking for modified peptides whose adjusted log fold change was within 10\% of the unadjusted log fold change but became statistically insignificant after adjustment.
 In other words, the fold change was the same between models but variance increased.
-When this test was applied on this experiment, only one peptide became insignificant due to an increase in variance.
-Thus we can conclude that the drop off in significant peptides was due to changes in global protein abundance.
-
-In~\figref{data4_profile_plot}, the modification of protein $GSDMD$ at site $K62$ showed the advantage of the proposed method.
-The modified peptide originally showed no abundance change between the infected 1 hour, 4 hour, and 6 hour conditions.
-This was contrasted with a strong negative change in the global profiling run between the same conditions.
-Looking at the Dox4hr vs Dox1hr conditions and modeling the modified peptide without adjusting for changes in the global profiling, the fold change was $-.501$ and the adjusted p-value was insignificant at $.0644$.
-After adjusting for changes in the global protein abundance, the fold change was much higher, $2.79$, and the adjusted p-value became significant, $5.25e-8$.
-In this case the effect of the modified peptide was strongly confounded with changes in the global protein.
-The proposed method allowed us to remove this confounding and estimate the true effect.
-
-\subsubsection*{In the two-mixture TMT experiment, MSstatsPTM removed confounding \todo{OV: between what and what? Is it similar to the previous section?}}
-
-\todo{This section just describes the figure.
+When this test was applied on this experiment, only one peptide became statistically insignificant due to an increase in variance.
+Thus we can conclude that the drop off in differentially abundant peptides was due to changes in global protein abundance.
+
+Interestingly, the global protein was more likely to drive changes in abundance rather than mask changes in the PTM, i.e.
+when the protein and PTM moved in opposite directions.
+This can be seen in \figref{data5_profile_plot}, where the global protein $TTP$ followed the same trend as the modification at $S178$.
+Originally the modificaiton was shown to be differentially abundant between KO\_Uninfect and KO\_Early, however this was due to confounding with the global protein.
+The proposed method corrects for these changes automatically, allowing us to see the true fold change, without having to do a side by side comparison.
+
+Conversely, cases where the PTM and global protein move in opposite directions are interesting and should be highlighted, as the modification and protein had contradictory abundance changes.
+One such case can be seen in~\figref{data4_profile_plot}.
+Luchetti et al.~\cite{LUCHETTI2021} showed that $GSDMD$ was actively degraded when IpaH7.8 expression was induced by dox treatment.
+Our reanalysis confirmed this, with the global $GSDMD$ protein being downregulated when Dox treatments reached the 4 and 6 hour marks.
+Conversely, the ubiquitination of $GSDMD$ at site $K62$ was actually upregulated between the same conditions.
+This upregulation was originally confounded by the downregulation of unmodified $GSDMD$, as can be seen in \figref{data4_profile_plot}, appearing to have little change between no dox and dox 4 and 6 hour.
+The proposed approach accounted for this confounding and the modification was shown to be differentially abundant, with an absolute fold change of $1.415$ between the Dox1hr and Dox4hr conditions.
+The proposed method allowed us to see modifications of interest, while filtering out those that would originally be false positives.
+
+%\subsubsection*{In the two-mixture TMT experiment, MSstatsPTM removed confounding \todo{OV: between what and what? Is it similar to the previous section?}}
+
+%\todo{This section just describes the figure.
 Could you emphasize why we see these results?
@@
 Again more PTMs became insignificant after adjustment than became significant.
-19,286 peptides became insignificant, 4,947 became significant, and 41,552 were significant in both models.
 Again we tested if the decrease in significant peptides was due to removing confounding with changes in the umodified protein or increased variance after adjustment was applied.
@@
 
-In Figure \ref{fig:data5_profile_plot} the profile plot of protein $TTP$ modified at site $S178$ showed the power of the proposed method.
+%In Figure \ref{fig:data5_profile_plot} the profile plot of protein $TTP$ modified at site $S178$ showed the power of the proposed method.
 Without adjustment, there was a large positive log fold change of $2.9$ between the WT\_Late and WT\_Uninfected conditions.
@@
 %\hline
-%Comparison & Unadj. true FC & Adj. true FC \\ \hline \hline
-%mix1-mix2  & 1                           & 1                        \\ \hline
-%mix1-mix3  & 0                           & -1                         \\ \hline
-%mix1-mix4  & 1                          & 0                         \\ \hline
-%mix2-mix3  & -1                          & -2                        \\ \hline
-%mix2-mix4  & 0                           & -1                        \\ \hline
-%mix3-mix4  & 1                           & 1                         \\ \hline
+%Comparison & Unadj. true FC & Adj. true FC \\ %\hline \hline
+%mix2 vs mix1  & -1 & -1 \\ \hline
+%mix3 vs mix1  & 0 & 1 \\ \hline
+%mix4 vs mix1  & -1 & 0 \\ \hline
+%mix3 vs mix2  & 1 & 2 \\ \hline
+%mix4 vs mix2  & 0 & 1 \\ \hline
+%mix4 vs mix3  & -1 & -1 \\ \hline
 %\end{tabular}
@@
 \includegraphics[scale=.5]{images/benchmark_fig.png}
-\caption{{\bf Dataset 3: SpikeIn benchmark - Ubiquitination - Label-free}.
+\caption{{\bf Dataset 3: Spike-in benchmark - Ubiquitination - Label-free}.
 (a) Relative amounts of human lysate, background {\it E.
@@
 2 pmol of spike-in peptides was added to mix 1 and mix 3, while 1 pmol was added to mix 2 and mix 4.
-(b) The expected fold change of the spike-in peptides in each comparison.
+(b) The expected log2 fold change of the spike-in peptides in each comparison.
 ``Unadj.
-true FC'' is the fold change of the spike-in peptides (red) without adjusting for changes in human lysate (blue).
+true FC'' is the log2 fold change of the spike-in peptides (red) without adjusting for changes in human lysate (blue).
 This was calculated by directly comparing the amounts of spike-in in each condition.
 ``Adj.
-true FC'' is the fold change of the spike-in peptides when adjusting for changes in human lysate.
+true FC'' is the log2 fold change of the spike-in peptides when adjusting for changes in human lysate.
 This was calculated by determining the ratio of spike-in peptides to human lysate in each mixture and then using the ratio to calculate the true fold change across comparisons.}
@@
 \end{subfigure}
-\caption{Dataset 3: SpikeIn benchmark - Ubiquitination - Label-free.
+\caption{Dataset 3: Spike-in benchmark - Ubiquitination - Label-free.
 a) Before adjustment the fold change of the spike-in peptides' were systematically different from the expected fold change in all models.
@@
 a) The overlap of differential modified peptides for the PTM model with and without global protein level adjustment across all comparisons.
-More PTMs became insignificant after adjustment than became significant.
-For the peptides that became insignificant in the adjusted model, their change in abundance was driven by changes in the global protein.
-In contrast, peptides that became significant after adjustment had their true abundance change masked by underlying changes in the unmodified protein.
+3,236 modified peptides became insignificant, 1,352 became significant, while 4,282 were significant in both models.
 b) Comparing the global profiling of protein $GSDMD$ with the ubiquitination of the protein at site $K62$.
@@
 a) The overlap of differentially modified peptides between the PTM model with and without global protein level adjustment across all comparisons.
-Again more PTMs became insignificant after adjustment then became significant.
+19,286 peptides became insignificant, 4,947 became significant, and 41,552 were significant in both models.
 b) Comparing the global profiling of protein $TTP$ with the modification of the protein at site $S178$.

### 2022-07-08 supplementary/ptm_sm.tex
updated spikein sections, methods, and results

@@
 \centering
-\includegraphics[width=.85\textwidth]{sim_new/spike_in_msstatsptm_volcano}
+\includegraphics[width=1\textwidth]{sim_new/spike_in_msstatsptm_volcano}
 \caption{The modeling results of the proposed method both before and after adjustment.
@@
 
-\subsubsection{Dataset 4 : Human - Ubiquitination - 1mix-TMT}
+\subsubsection{Dataset 4 : Human - Ubiquitination - 1mix-TMT~\cite{LUCHETTI2021}}
 \label{sec:ipah}
 
-In this experiment Shigella ubiquitin ligase IpaH7.8 was shown to inhibit the protein gasdermin D (GSDMD) ~\cite{LUCHETTI2021}.
-Multiplex proteomics was used to quantify the abundance of total protein, and ubiquitination in human epithelial cells.
-Cells were either infected or uninfected with IpaH7.8-deficient Shigella flexneri and measurements were taken at different time periods.
-Uninfected cells were measured at 0 and 6 hours, while infected cells were measured at 1, 2, 4, and 6 hour increments, resulting in six total conditions.
-The experiment was unbalanced with two biological replicates per condition for all conditions except for infected 1 hour.
-The experiment was modeled as a group comparison, with different replicates measured at each time point.
 The experimental design can be seen in Table~\ref{table:ipah_design}.

### 2022-07-08 code/MSstatsPTM_SpikeIn_Analysis_backup.R

@@ -173,6 +173,9 @@ aqua <- bind_rows(
   aqua2 %>% mutate(batch = "BCH2")
 )
 
+rm(list = ls()[!(ls() %in% c("aqua", "design_prot", "design_kgg",
+                             "dta_prot", "dta_kgg"))])
+
 # Load data set & housekeeping --------------------------------------------
 
 # Parameters
@@ -315,6 +318,16 @@ colnames(df_prot) <- c('ProteinName', 'PeptideSequence', 'PrecursorCharge',
                        'Intensity', 'FragmentIon', 'ProductCharge',
                        'IsotopeLabelType')
 
+## Clean Data ------------------------------------------------------------------
+df_site <- df_site %>% filter(is_mod == TRUE)
+df_prot <- df_prot %>% filter(!is_mod)
+
+df_site <- df_site[ , !(names(df_site) %in% c('is_mod'))]
+df_prot <- df_prot[ , !(names(df_prot) %in% c('is_mod'))]
+
+df_site <- .makeBalancedDesign(df_site, fill_missing = TRUE)
+df_prot <- .makeBalancedDesign(df_prot, fill_missing = TRUE)
+
 # Visualization ----------------------------------------------------------------
 
 # QC plot (boxplot)
@@ -403,21 +416,9 @@ Reference_Norm <- function(df, ref) {
   df_aug = df_aug %>% select(-log2inty, -adjLog2inty)
   return(df_aug)
 }
-saved_refs$PTM = saved_refs$PTM %>% rename(Run = run)
-saved_refs$PROTEIN = saved_refs$PROTEIN %>% rename(Run = run)
 
-df_site = Reference_Norm(df_site, saved_refs$PTM)
-df_prot = Reference_Norm(df_prot, saved_refs$PROTEIN)
-
-## Clean Data ------------------------------------------------------------------
-df_site <- df_site %>% filter(is_mod == TRUE)
-df_prot <- df_prot %>% filter(!is_mod)
-
-df_site <- df_site[ , !(names(df_site) %in% c('is_mod'))]
-df_prot <- df_prot[ , !(names(df_prot) %in% c('is_mod'))]
-
-df_site <- .makeBalancedDesign(df_site, fill_missing = TRUE)
-df_prot <- .makeBalancedDesign(df_prot, fill_missing = TRUE)
+df_site = Reference_Norm(df_site, refs$PTM)
[…]

### 2022-07-09 main/MSstatsPTM_Main.tex
Dataset 3

@@
 
-\medskip \noindent {\bf Experimental Design:}~\figref{benchmark-design}(a) overviews the experimental design.
-Four mixtures (i.e., conditions) were created with varying amounts of human lysate, background {\it E.
-coli} lysate, and human spike-in ub-peptide mixture.
+\medskip \noindent {\bf Experimental design:}~\figref{benchmark-design}(a) overviews the experimental design.
+Four mixtures (i.e., conditions) were created with varying amounts of human lysate \todo{which cell line?}, background {\it E.
+Coli} lysate, and human spike-in ub-peptide mixture.
 Unmodified peptides from human lysate were viewed as the global proteome.
 Background {\it E.
-coli} lysate were used to equalize total protein levels prior to enrichment or global protein profiling between Mix 1 and 3, and between Mix 2 and 4.
-50 heavy-labeled KGG motif peptides from 20 human proteins were used as spike-in peptides in a mixed background of lysates from {\it E.
-coli} and a human cell line.
+coli} lysate were used to equalize total protein levels.
+50 heavy-labeled KGG motif peptides from 20 human proteins were spiked into the mixed background of the lysates.
 Quantitative changes in protein and site abundance of these 20 human proteins were the target of the benchmark.
-In particular, we distinguished the changes in the abundances of the modified peptides (i.e., unadjusted changes) and the changes of their abundance relative to the changes in the abundances of the human lysate (i.e., protein-level adjusted changed).
+In particular, we distinguished the unadjusted changes (i.e.
+changes in the abundances of the modified peptides) and the protein-level adjusted changes of (i.e., changes in the abundances of the modified peptides relative to the changes in the abundances of the human lysate).
 The true log-fold changes between the relevant components of the relevant mixtures are summarized in~\figref{benchmark-design}(b).
@@
 \medskip \noindent {\bf Data acquisition:} Each mixture was analyzed with KGG enrichment, and without KGG enrichment (i.e., in a global profiling run), with label-free LC-MS/MS.
-There was a 90.2\% overlap of protein identifications between the identified background modified peptides and proteins that were quantified in the global profiling run.
-
-\medskip \noindent {\bf Evaluation:} We expect the spike-in peptides to change corresponding to the values in Figure \ref{fig:benchmark-design}(b).
-These peptides are treated as positive controls in all comparisons except Mix 1 vs Mix 4, which was treated as a negative control.
-Additionally, the background {\it E.
-coli} lysate peptides were treated as negative controls and were not expected to change in any comparison.
+There was a 90.2\% overlap of protein identifications between the identified background modified peptides and proteins quantified in the global profiling run.
+
+\medskip \noindent {\bf Evaluation:} We expect the relative abundances of the spike-in peptides to change as in~\figref{benchmark-design}(b).
+The changes in peptide abundances in all the comparisons except Mix 1 vs Mix 4 were distinct from changes in the global proteome abundances and distinct from zero, and were viewed as positive controls.
+In the comparison of Mix1 vs Mix 4 both the modified peptides and the global proteome background changed two-fold, and as the result the peptides in this comparison were viewed as a negative control.
+Since the background {\it E.
+Coli} lysate peptides were not expected to change in abundance in any comparison \todo{in any, or in 3 vs 4?}, they were viewed as additional (unadjusted) negative controls.
 We evaluated the statistical methods ability to avoid false positives, as well as their sensitivity in detecting the differentially abundant spike-in peptides and accurately estimate their expected fold change.
@@
 \medskip \noindent {\bf Experimental Design:}  Luchetti et al.~\cite{LUCHETTI2021} profiled human epithelial cells engineered to express IpaH7.8 under a dox inducible promoter.
-Uninfected cells were measured at 0 and 6 hours, while cells infected with {\it Shigella flexneri} ({\it S.
-flexneri}) bacteria were measured at 1, 2, 4, and 6 hour increments, resulting in six total conditions.
+Uninfected cells were measured at 0 and 6 hours, while cells infected with {\it Shigella Flexneri} ({\it S.
+Flexneri}) bacteria were measured at 1, 2, 4, and 6 hour increments, resulting in six total conditions.
 11 samples were allocated to 1 TMT mixture in an unbalanced repeated measure design.
@@
 Since the dataset was a biological investigation, the true positive modifications were unknown.
-Shigella ubiquitin ligase IpaH7.8 was shown to function as an inhibitor of the protein gasdermin D (GSDMD).
+Shigella ubiquitin ligase IpaH7.8 was shown to function as an inhibitor of the protein Gasdermin D (GSDMD).
 GSDMD was actively degraded when IpaH7.8 expression was induced by dox treatment in human cells.
@@
 \cite{Maculins} studied primary murine macrophages infected with {\it S.
-flexneri}.
+Flexneri}.
 The experiment quantified the abundance of total protein and of phosphorylation in wild type (WT), and in ATG16L1-deficient (cKO) samples, uninfected and infected with {\it S.
-flexneri}.
+Flexneri}.
 The abundance of total protein and post-translation modifications were quantified at three time points, uninfected, early infection (45-60 minutes), and late infection (3-3.5 hours).
@@
 \caption{{\bf Dataset 3: Spike-in benchmark - Ubiquitination - Label-free}.
-(a) Relative amounts of human lysate, background {\it E.
-Coli} lysate, and spike-in peptides.
-Volumes of each mixture totaled 20mg, with 10mg of background {\it E.
-Coli} lysate used to normalize mix 3 and mix 4 to make all mixtures directly comparable.
-2 pmol of spike-in peptides was added to mix 1 and mix 3, while 1 pmol was added to mix 2 and mix 4.
-(b) The expected log2 fold change of the spike-in peptides in each comparison.
+\todo{In the legend, could you mark the blue box "Human lysate (global proteome)", the grey box "E.
+Coli lysate (background proteome)" and mark the red box as "50 spiked human ub-peptides".
+Check the capitalization} \todo{In the table, could you change the headers to log$_2$FC}
+%
+(a) Four mixtures (i.e., conditions) were created with varying amounts of human lysate \todo{which cell line?}, background {\it E.
+Coli} lysate, and human spike-in ub-peptide mixture.
+Unmodified peptides from human lysate were viewed as the global proteome.
+Background {\it E.
+coli} lysate were used to equalize total protein levels.
+50 heavy-labeled KGG motif peptides from 20 human proteins were spiked into the mixed background of the lysates.
+Quantitative changes in protein and site abundance of these 20 human proteins were the target of the benchmark.
+%
+(b)  We distinguished the unadjusted changes (i.e.
+changes in the abundances of the modified peptides) and the protein-level adjusted changes of (i.e., changes in the abundances of the modified peptides relative to the changes in the abundances of the human lysate).
 ``Unadj.
-true FC'' is the log2 fold change of the spike-in peptides (red) without adjusting for changes in human lysate (blue).
-This was calculated by directly comparing the amounts of spike-in in each condition.
+true log$_2$FC'' are the log-ratios of the abundances of the spiked peptides between each condition.
 ``Adj.
-true FC'' is the log2 fold change of the spike-in peptides when adjusting for changes in human lysate.
-This was calculated by determining the ratio of spike-in peptides to human lysate in each mixture and then using the ratio to calculate the true fold change across comparisons.}
-
+true log$_2$FC'' \todo{Could you rephrase the calculation more precisely?
+Ratios of the ratios?
+I am wondering if a formula wold be helpful?} was calculated by determining the ratio of spike-in peptides to human lysate in each mixture and then using the ratio to calculate the true fold change across comparisons.}
 \label{fig:benchmark-design}

### 2022-07-11 main/MSstatsPTM_Main.tex
Implementation (has not checked the figure caption yet)

@@
 
-The proposed methods are implemented in the open source R package $MSstatsPTM$.
-An overview of the steps of the package are illustrated in Figure \ref{fig:msstatsptm_design}.
-$MSstatsPTM$ includes converters for multiple spectral processing tools, including MaxQuant, Progenesis, and Spectronaut.
-Conversion is done separately for the modified peptide and global profiling runs.
-To identify the local modification site we require the tool to identify where in the sequence the modification occurred.
-If the global profiling information is not available, the package can still analyze the modified run, but will do so without adjusting for changes in unmodified protein abundance.
-This amounts to testing the null hypothesis in~\eqref{conv_null_hyp}; which amounts to running the methods seen in $MSstats$ on modified peptide level data.
-
-The next step is peptide/protein summarization.
+\todo{Could we try to use a consistent vocabulary for 'global protein' (we sometimes use unmodified peptides, unmodified features, global profiling etc.) Could you check the entire manuscript and see that we use the same terminology for that}
+
+The proposed methods are implemented in the open source R package $MSstatsPTM$, overviewed in~\figref{msstatsptm_design}.
+\todo{$MSstatsPTM$ is applicable to most experimental designs, label-free data acquisitions such as XXX and label-based data acquisitions such as XXX} $MSstatsPTM$ takes as input lists of identified and quantified spectral features, produced by spectral processing tools such as MaxQuant, Progenesis, or Spectronaut.
+\todo{Could you rephrase the next sentence to be more specific.
+If someone is a tool developer, what exactly should they do?} To identify the local modification site we require the tool to identify where in the sequence the modification occurred.
+The package includes converters for seamless integration with multiple such tools.
+\todo{Could you clarify what the converters do, especially for PTM?
+Do we discard peptides with multiple modification sites?}.
+For experimental designs that include separate runs enriched in modified peptides and global profiling runs,  conversion is performed separately for the runs enriched in modified peptides and for the global profiling runs.
+For experimental designs that include modified and unmodified peptides in a same run, \todo{what happens for the conversion?}
+
+\todo{Could you rephrase the next sentence?
+What is summarized on the PTM side?
+The peptide sequence?
+Or any peptide sequence that contains the modification site?} The next step is peptide/protein summarization.
 Summarization is performed separately for the PTM and global protein in the same way as described above for $MSstats$.
-The package can optionally impute missing values using an Accelerated Failure Time (AFT) model~\cite{Wei:1992}.
-Missing value imputation should be performed with care when applied to PTM data.
+When summarizing the unmodified peptide features, the package optionally imputes missing values using an Accelerated Failure Time (AFT) model~\cite{Wei:1992}.
+When summarizing the modified features, missing value imputation is also possible but should be performed with care.
 PTMs generally exhibit low feature counts and may be missing due to reasons other than low abundance.
-These issues can cause the missing value imputation to be unstable.
-Finally, a separate model is fit to both the modified and global profiling runs.
-This model is automatically adjusted depending on the experimental design and acquisition method.
-If the base model is not applicable, a simplified model is fit.
+These issues can violate the assumptions underlying the imputation, and lead to numerically unstable  results.
+Next, separate statistical models are fit to both the summaries of the modified features and the to summaries of the unmodified features.
+The models are automatically selected to reflect the experimental design and the data acquisition.
+If the base model is not applicable for a particular PTM or protein, e.g.
+due to missing data, a simplified model is fit.
+
 After modeling, the modified model is adjusted for changes in unmodified protein abundance, using the methods described above.
-
-We include visualizations for QC and model assessment.
-In QC plots the PTM and global protein are shown side by side, allowing for easier interpretation and comparison across data.
-In the model plots this is further separated into PTM, global protein, and adjusted PTM visualizations, allowing users to see the affect of the adjustment.
-
-The package relies on functionalities from the R packages $MSstats$ \cite{Choi:2014} and $MSstatsTMT$ \cite{Huang:2020}.
-The statistical modeling relies on the functionality from the R packages $lme4$ \cite{Bates2015} and $lmerTest$ \cite{Kuznetsova2017}.
-The code is available on Bioconductor, \url{http://www.bioconductor.org/packages/release/bioc/html/MSstatsPTM.html}, and Github, \url{https://github.com/Vitek-Lab/MSstatsPTM}.
+Modification sites for which lacking the corresponding global profiling information cannot be adjusted for changes in protein abundance.
+In this case the implementation reverts to testing the null hypothesis in~\eqref{conv_null_hyp} using the statistical methods seen in $MSstats$, applied separately to each modified peptide.
+In addition to the statistical functionalities, the implementation includes visualizations for quality control and assessment of the quality of model fit.
+
+The implementation relies on functionalities from the R packages $MSstats$ \cite{Choi:2014} and $MSstatsTMT$ \cite{Huang:2020}, which in turn rely on the R packages $lme4$ \cite{Bates2015} and $lmerTest$ \cite{Kuznetsova2017}.
+$MSstatsPTM$ is available on Bioconductor, \url{http://www.bioconductor.org/packages/release/bioc/html/MSstatsPTM.html}, and Github, \url{https://github.com/Vitek-Lab/MSstatsPTM}.
 
@@
 In a clean simulation, recall and accuracy of $MSstatsPTM$ and $Limma$ performed similarly.
-As expected when not using Empricial Bayes moderation in MSstatsPTM, Limma performed slightly better in simulations with fewer biological replicates.
+As expected when not using Empiricial Bayes moderation in MSstatsPTM, Limma performed slightly better in simulations with fewer biological replicates.
 In contrast, the performance of the two-sample $t$-test method lagged behind.
@@
 \includegraphics[scale=.8]{images/MSstatsPTM_design.png}
-\caption{The workflow of $MSstatsPTM$ and how it fits into the experimental pipeline.
-The name of the R functions used for each step are highlighted in red.
+\caption{The $MSstatsPTM$ workflow.
+The names of the R functions used for each step are highlighted in red.
 The package's workflow starts after modified and unmodified peptide quantification.

### 2022-07-11 main/MSstatsPTM_Main.tex
finished spike in reanalysis, updated results

--- 2022-07-08 main/MSstatsPTM_Main.tex
@@
 
-We evaluated $MSstatsPTM$ on simulated and spike-in datasets with known ground truth in terms of true positives ($TP$), false positives ($FP$), true negatives ($TN$), and false negatives ($FN$) differentially abundant PTM.
+We evaluated $MSstatsPTM$ on simulated and spike-in datasets with known ground truth in terms of true positives ($TP$), false positives ($FP$), true negatives ($TN$), and false negatives ($FN$) differentially abundant PTMs.
 The true positives were defined as PTMs with changes distinct from the overall changes in protein abundance.
-The true negatives were defined as PTM which, after accounting for the changes in the overall protein abundance, were not differentially abundant.
+The true negatives were defined as PTMs which, after accounting for the changes in the overall protein abundance, were not differentially abundant.
 Additional summaries were the false discovery rate $FDR=FP/(TP+FP)$, $Recall=TP/(TP + FN)$, and $Accuracy=(TP + TN) / (TP + TN + FP + FN)$.
-For biological experiments with unknown ground truth, we compared the differentially abundant PTM  with and without adjusting for changes in unmodified protein abundance.
-
-The evaluations were based on the $MSstatsPTM$, and were performed without imputing missing values and without using Empirical Bayes moderation.
+For biological experiments with unknown ground truth, we compared the differentially abundant PTMs  with and without adjusting for changes in unmodified protein abundance.
+
+The evaluations were based on the $MSstatsPTM$, and were performed without imputing missing values or using Empirical Bayes moderation.
 $MSstatsPTM$ was compared with \todo{$t$-test?} and Limma.
@@
 In general, no need to repeat what we see in the figure - this is what the figures are for.}
-
-\todo{OV: Also, maybe combine two sections that make a similar point?}
-
-\todo{OV: also avoid the terms 'significant'. Replace it with 'differentially abundant'. Instead of 'bias' and 'convolution', use 'confounding'}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection*{In the label-free benchmark experiment, MSstatsPTM had a higher sensitivity than the existing methods}
+information \subsubsection*{In the label-free benchmark experiment, MSstatsPTM had a higher sensitivity than the existing methods}
 
@@
 Using $Limma$ would have resulted in missing the majority of differential PTMs.
-
 This experiment introduced more variance components, which is challenging for the existing methods and therefore the difference in the results is more pronounced than in the simulations.
@@
 
-\todo{This section just describes the figure. Could you emphasize why we see these results? What are the reasons for differences in performance?}
-
-The results of these experiments are summarized in~\figref{data4_plots} and~\figref{data5_plots}.
+The results of these experiments are summarized in~\figref{venn_diagrams} and~\figref{profile_plots}.
 \figref{data4_venn_diagram} and~\figref{data5_venn_diagram} show the number of differentially abundant modified peptides before and after adjustment.
-In both cases, more peptides lost than gained differential abundance after the adjustment.
+Adjusting for changes in the global protein caused more PTMs to change from differentially abundant to non-differentially abundant than vice versa.
 A question that must be addressed is if the decrease in differentially abundant peptides is due to the increased variance that comes from adjustment.
@@
 In other words, the fold change was the same between models but variance increased.
-When this test was applied on this experiment, only one peptide became statistically insignificant due to an increase in variance.
-Thus we can conclude that the drop off in differentially abundant peptides was due to changes in global protein abundance.
-
-Interestingly, the global protein was more likely to drive changes in abundance rather than mask changes in the PTM, i.e.
-when the protein and PTM moved in opposite directions.
-This can be seen in \figref{data5_profile_plot}, where the global protein $TTP$ followed the same trend as the modification at $S178$.
-Originally the modificaiton was shown to be differentially abundant between KO\_Uninfect and KO\_Early, however this was due to confounding with the global protein.
-The proposed method corrects for these changes automatically, allowing us to see the true fold change, without having to do a side by side comparison.
-
-Conversely, cases where the PTM and global protein move in opposite directions are interesting and should be highlighted, as the modification and protein had contradictory abundance changes.
+When the test was applied to Dataset 4, only one peptide became non-differentially abundant due to an increase in variance.
+When applied to Dataset 5, 548 peptides became non-differentially abundant due to an increase in variance (only 3.4\% of all PTMs that became became non-differentially abundant).
+Thus we can conclude that the drop off in differentially abundant peptides was due to changes in global protein abundance, not simply increased model variance.
+
+Interestingly, the test above showed that the global protein was more likely to cause observed changes in PTM abundance, rather than mask changes in the PTM.
+The later case occurs when the global protein exhibits a change in abundance, while the modification does not.
+An example of this can be seen in \figref{data5_profile_plot}.
+Originally the modification at $S178$ of protein $TTP$ was shown to be differentially abundant between KO\_Uninfect and KO\_Early, with a log fold change of $2.9$.
+However the global protein was shown to contribute $2.014$ of this change, meaning that the modification only accounted for $.886$ after adjustment.
+This caused the modification to go from differentially abundant to non-differentially abundant.
+As discussed previously, these cases occur frequently and can be thought of as false positives that would occur without accounting for confounding with the global protein.
+The proposed approach corrects for this confounding, allowing us to see the true fold change of PTMs, while excluding those that do not exhibit differential changes.
+
+Conversely, cases where the global protein masks changes in the PTM are more infrequent.
+In this case, the modification and protein exhibit contradictory changes in abundance, moving in opposite directions.
 One such case can be seen in~\figref{data4_profile_plot}.
@@
 Our reanalysis confirmed this, with the global $GSDMD$ protein being downregulated when Dox treatments reached the 4 and 6 hour marks.
-Conversely, the ubiquitination of $GSDMD$ at site $K62$ was actually upregulated between the same conditions.
-This upregulation was originally confounded by the downregulation of unmodified $GSDMD$, as can be seen in \figref{data4_profile_plot}, appearing to have little change between no dox and dox 4 and 6 hour.
+Conversely, ubiquitination of $GSDMD$ at site $K62$   upregulated abundance between the same conditions.
+This upregulation was originally confounded by the downregulation of unmodified $GSDMD$, appearing to have little change between no dox and dox 4 and 6 hour.
 The proposed approach accounted for this confounding and the modification was shown to be differentially abundant, with an absolute fold change of $1.415$ between the Dox1hr and Dox4hr conditions.
-The proposed method allowed us to see modifications of interest, while filtering out those that would originally be false positives.
-
-%\subsubsection*{In the two-mixture TMT experiment, MSstatsPTM removed confounding \todo{OV: between what and what? Is it similar to the previous section?}}
+This was an interesting result, as the modification contradicts what previous research showed for the global protein.
+This effect can be thought of as a false negative and would have been more challenging to observe without the proposed approach.
+There are many non-differentially abundant PTMs, and having to comb through each one individually to determine false negatives is generally an unfeasible task.
+The proposed approach removes the confounding on an experimental level, revealing interesting PTMs that are entirely masked due to changes in the global protein.
+
+%\subsubsection*{In the two-mi[xture TMT experiment, MSstatsPTM removed confounding \todo{OV: between what and what? Is it similar to the previous section?}}
 
@@
 
-\todo{This section just describes the figure.
-Could you emphasize why we see these results?
-What are the reasons for differences in performance?}  As discussed in \secref{exp_proc_dataset6}, there was no unmodified global profiling run performed in this experiment.
+As discussed in \secref{exp_proc_dataset6}, there was no unmodified global profiling run performed in this experiment.
 Once identification and quantification of the Ubiquitination profiling was performed, peptides which were unmodified were extracted and used in place of a global profiling run.
@@
 
-The results of this experiment are summarized in~\figref{data6_plots}.
+The number of differentially abundant PTMs before and after adjustment are shown in~\figref{data6_plots}.
 After adjusting for changes in the unmodified protein, there were fewer significant modified peptides.
-In total, 2709 modified peptides became insignificant, while only 547 became significant.
-However, this was mainly due to not having a global profiling run, resulting in a lack of overlap between modified and unmodified peptides.
-In the bottom plot, only modified peptides that could be adjusted are shown.
-Here there were much fewer peptides that became insignificant after adjustment.
-726 modified peptides became insignificant, 547 became significant, and 1,078 were significant in both models.
-Only 25 PTMs became insignificant due to increased variance from adjustment.
+However, this was mainly due to not having a global profiling run.
+When looking only at PTMs with a corresponding global protein, the number of differentially modified PTMs exhibited a small drop, but was more consistent.
+As above, we tested if this drop in differentially abundant PTMs was due to an increase in variance.
+Here only 25 PTMs became insignificant due to increased variance.
+When there was no global profiling run the methods were less effective at removing confounding with the global protein.
+This was entirely due to not being able to perform adjustment without a corresponding global protein.
+In this case adjustment caused us to toss out a large number of the PTMs in the original model.
+In cases without a global profiling run, the proposed method can still be used, but it should be used in conjunction with the unadjusted model so that information is not lost.
 
@@
 \end{subfigure}
-\begin{subfigure}[c]{0.95\linewidth}
-\includegraphics[width=1\textwidth]{images/IpaH_prof_plot.png}
+\begin{subfigure}[c]{0.5\linewidth}
+\includegraphics[width=1\textwidth]{images/shig_venn_diagramm.png}
+\caption{}
+\label{fig:data5_venn_diagram}
+\end{subfigure}
+\caption{a) Dataset 4: Human - Ubiquitination - 1mix-TMT.
+The overlap of differential modified peptides for the PTM model with and without global protein level adjustment across all comparisons.
+3,236 modified peptides became insignificant, 1,352 became significant, while 4,282 were significant in both models.
+b) Dataset 5: Mouse - Phosphorylation - 2mix-TMT.
+a) The overlap of differentially modified peptides between the PTM model with and without global protein level adjustment across all comparisons.
+19,286 peptides became insignificant, 4,947 became significant, and 41,552 were significant in both models.
+}
+\label{fig:venn_diagrams}
+\end{figure}
+
+\begin{figure}[h!]
+\centering
+\begin{subfigure}{\textwidth}
+ \centering
+\includegraphics[width=1.0\textwidth]{images/No_Difference_Shigella_Profile_Plot}
 \caption{}
@@
  \end{subfigure}
-\caption{Dataset 4: Human - Ubiquitination - 1mix-TMT.
-a) The overlap of differential modified peptides for the PTM model with and without global protein level adjustment across all comparisons.
-3,236 modified peptides became insignificant, 1,352 became significant, while 4,282 were significant in both models.
-b) Comparing the global profiling of protein $GSDMD$ with the ubiquitination of the protein at site $K62$.
+ \begin{subfigure}{\textwidth}
+ \centering
+	\includegraphics[width=1.0\textwidth]{images/IpaH_prof_plot.png}
+	\caption{}
+	\label{fig:data5_profile_plot}
+	 \end{subfigure}
+\caption{a) Dataset 5: Mouse - Phosphorylation - 2mix-TMT.
+Comparing the global profiling of protein $TTP$ with the modification of the protein at site $S178$.
+When looking at the summary of the modification and global protein it was clear the difference between conditions followed the same trend.
+Specifically, there was a positive adjustment in abundance when comparing WT\_Uninfect to WT\_Late in both the modification and global profiling run.
+This indicated the movement was driven by changes in global protein that was only accounted for in the model after adjusting for global protein abundance change.
+b) Dataset 4: Human - Ubiquitination - 1mix-TMT.
+Comparing the global profiling of protein $GSDMD$ with the ubiquitination of the protein at site $K62$.
 When looking at the summary of the modification and global protein it was clear the conditions follow different trends.
@@
 This indicated the modification was confounded with changes in the unmodified protein.}
-\label{fig:data4_plots}
-\end{figure}
-
-\begin{figure}[h!]
-\centering
-\begin{subfigure}{\textwidth}
- \centering
-\includegraphics[height=.5\textwidth]{images/shig_venn_diagramm.png}
-\caption{}
-\label{fig:data5_venn_diagram}
- \end{subfigure}
- \begin{subfigure}{\textwidth}
- \centering
-	\includegraphics[width=1.0\textwidth]{images/No_Difference_Shigella_Profile_Plot}
-	\caption{}
-	\label{fig:data5_profile_plot}
-	 \end{subfigure}
-\caption{Dataset 5: Mouse - Phosphorylation - 2mix-TMT.
-a) The overlap of differentially modified peptides between the PTM model with and without global protein level adjustment across all comparisons.
-19,286 peptides became insignificant, 4,947 became significant, and 41,552 were significant in both models.
-b) Comparing the global profiling of protein $TTP$ with the modification of the protein at site $S178$.
-When looking at the summary of the modification and global protein it was clear the difference between conditions followed the same trend.
-Specifically, there was a positive adjustment in abundance when comparing WT\_Uninfect to WT\_Late in both the modification and global profiling run.
-This indicated the movement was driven by changes in global protein that was only accounted for in the model after adjusting for global protein abundance change.}
-\label{fig:data5_plots}
+\label{fig:profile_plots}
 \end{figure}
@@
 More PTMs became insignificant than became significant after adjustment.
-This was due to not having a global profiling run, resulting in a lack of overlap between modified peptides and unmodified proteins.
+In total, 2709 modified peptides became insignificant, while only 547 became significant.
 b) Here we made the same comparison but only looked at modified peptides where adjustment could be performed, ie they had a matching unmodified protein.
 In this case there were significantly less peptides that became insignificant after adjustment.
-This highlighted the need for a global profiling run.}
+726 modified peptides became insignificant, 547 became significant, and 1,078 were significant in both models.}
 \label{fig:data6_plots}

### 2022-07-11 code/MSstatsPTM_SpikeIn_Analysis.R

@@ -889,7 +889,7 @@ run_ttest <- function(summarized_data){
 
 
 ttest_model <- run_ttest(joined)
-save(ttest_model, file = "Spike_in_ttest.rda")
+save(ttest_model, file = "D:\\Northeastern\\Research\\MSstats\\MSstatsPTM-manuscript\\data\\Spike_in_ttest.rda")
 
 ttest_model <- ttest_modelkeep
 ttest_model$ttest_adj_temp <- ttest_model$ttest_adj_temp %>% filter(pval != 1)
@@ -952,7 +952,7 @@ temp_test <- rbind(ttest_model$ttest_temp, ttest_model$ttest_adj_temp)
 temp_test$protadj <- factor(temp_test$protadj, levels = c('no adjustment','protein adjustment'))
 
 ttest_final_model <- temp_test
-save(ttest_final_model, file = "Spike_in_ttest_model.rda")
+save(ttest_final_model, file = "D:\\Northeastern\\Research\\MSstats\\MSstatsPTM-manuscript\\data\\Spike_in_ttest_model.rda")
 
 # Volcano plot
 temp_test %>% arrange(desc(labeling)) %>%

### 2022-07-11 main/MSstatsPTM_Main.tex

--- 2022-07-11 main/MSstatsPTM_Main.tex
@@
 \hat{\Delta}_{Protein} = \hat{\mu}_{i} - \hat{\mu}_{i^{\prime}}
+\label{eq:protein_delta}
 \end{eqnarray}

### 2022-07-11 main/MSstatsPTM_Main.tex
Suggestion for Fig 3 caption

--- 2022-07-11 main/MSstatsPTM_Main.tex
@@
 The names of the R functions used for each step are highlighted in red.
-The package's workflow starts after modified and unmodified peptide quantification.
+\todo{In the caption, could you try the following order: applicable to experimental designs XXX; takes as input XXX; separate parallel analyses for global proteome and PTM.
+(No need to read off the figure, just clarify the content)} \todo{Would it be possible to modify the figure to reflect the parallel nature of the processing/analyses?} The package's workflow starts after modified and unmodified peptide quantification.
 First tool specific pre-processing is performed, this includes modification site identification, general data cleaning, and formatting the data into the format needed for the package.

### 2022-07-11 code/Spike-in_Analysis.Rmd

@@
 ## Update some naming conventions
-plot_df[plot_df$Label == 'mix2 vs mix1'][['contrast_truth']] <- "LogFC: -1 (mix2-mix1)"
-plot_df[plot_df$Label == 'mix3 vs mix1'][['contrast_truth']] <- "LogFC: 1 (mix3-mix1)"
-plot_df[plot_df$Label == 'mix4 vs mix1'][['contrast_truth']] <- "LogFC: 0 (mix4-mix1)"
-plot_df[plot_df$Label == 'mix3 vs mix2'][['contrast_truth']] <- "LogFC: 2 (mix3-mix2)"
-plot_df[plot_df$Label == 'mix4 vs mix2'][['contrast_truth']] <- "LogFC: 1 (mix4-mix2)"
-plot_df[plot_df$Label == 'mix4 vs mix3'][['contrast_truth']] <- "LogFC: -1 (mix4-mix3)"
+plot_df[plot_df$Label == 'mix1-mix2'][['contrast_truth']] <- "LogFC: -1 (mix2-mix1)"
+plot_df[plot_df$Label == 'mix1-mix3'][['contrast_truth']] <- "LogFC: 1 (mix3-mix1)"
+plot_df[plot_df$Label == 'mix1-mix4'][['contrast_truth']] <- "LogFC: 0 (mix4-mix1)"
+plot_df[plot_df$Label == 'mix2-mix3'][['contrast_truth']] <- "LogFC: 2 (mix3-mix2)"
+plot_df[plot_df$Label == 'mix2-mix4'][['contrast_truth']] <- "LogFC: 1 (mix4-mix2)"
+plot_df[plot_df$Label == 'mix3-mix4'][['contrast_truth']] <- "LogFC: -1 (mix4-mix3)"
 plot_df$contrast_truth = factor(plot_df$contrast_truth, levels = 
@@
 plot_df[plot_df$labeling == 'heavy'][['labeling']] <- "Spike-in Peptides"
+plot_df$trueLog2FC=plot_df$trueLog2FC*-1
+plot_df[plot_df$Label == 'mix1-mix2'][['trueLog2FC']] <- plot_df[plot_df$Label == 'mix1-mix2'][['trueLog2FC']]*-1
+plot_df[plot_df$Label == 'mix1-mix3'][['trueLog2FC']] <- plot_df[plot_df$Label == 'mix1-mix3'][['trueLog2FC']]*-1
 
@@
 # Volcano plot
-png("../supplementary/sim_new/spike_in_limma_volcano.png", width = 750, height = 500)
+png("../supplementary/sim_new/spike_in_limma_volcano.png", width = 1000, height = 500)
 plot_df %>% arrange(labeling) %>% 
-  filter(abs(Log2FC) < Inf & Label %in% c("mix1-mix3", "mix1-mix4", "mix2-mix3", "mix2-mix4")) %>%
+  filter(abs(Log2FC) < Inf) %>%
   ggplot(aes(Log2FC, -log10(adj.pvalue))) + 
@@
 combo_plot_df %>% arrange(Model, labeling) %>% 
-  filter(abs(log2FC) < Inf & Label %in% c("mix2-mix3", "mix2-mix4")) %>%
+  filter(abs(log2FC) < Inf & Label %in% c("mix2-mix3", "mix1-mix4", "mix3 vs mix2",
+                                          "mix4 vs mix1")) %>%
   ggplot(aes(log2FC, -log10(adj.pvalue))) + 
@@
 ## Update some naming conventions
-ttest_final_model[ttest_final_model$label == 'mix1-mix3'][['contrast_truth']] <- "LogFC: 1 (mix1-mix3)"
-ttest_final_model[ttest_final_model$label == 'mix1-mix4'][['contrast_truth']] <- "LogFC: 0 (mix1-mix4)"
-ttest_final_model[ttest_final_model$label == 'mix2-mix3'][['contrast_truth']] <- "LogFC: -2 (mix2-mix3)"
-ttest_final_model[ttest_final_model$label == 'mix2-mix4'][['contrast_truth']] <- "LogFC: -1 (mix2-mix4)"
+ttest_final_model[ttest_final_model$label == 'mix1-mix2'][['contrast_truth']] <- "LogFC: -1 (mix2-mix1)"
+ttest_final_model[ttest_final_model$label == 'mix1-mix3'][['contrast_truth']] <- "LogFC: 1 (mix3-mix1)"
+ttest_final_model[ttest_final_model$label == 'mix1-mix4'][['contrast_truth']] <- "LogFC: 0 (mix4-mix1)"
+ttest_final_model[ttest_final_model$label == 'mix2-mix3'][['contrast_truth']] <- "LogFC: 2 (mix3-mix2)"
+ttest_final_model[ttest_final_model$label == 'mix2-mix4'][['contrast_truth']] <- "LogFC: 1 (mix4-mix2)"
+ttest_final_model[ttest_final_model$label == 'mix3-mix4'][['contrast_truth']] <- "LogFC: -1 (mix4-mix3)"
 
@@
 
+ttest_final_model$trueLog2FC=ttest_final_model$trueLog2FC*-1
+ttest_final_model[ttest_final_model$Label == 'mix1-mix2'][['trueLog2FC']] <- ttest_final_model[ttest_final_model$Label == 'mix1-mix2'][['trueLog2FC']]*-1
+ttest_final_model[ttest_final_model$Label == 'mix1-mix3'][['trueLog2FC']] <- ttest_final_model[ttest_final_model$Label == 'mix1-mix3'][['trueLog2FC']]*-1
+
 # Volcano plot
@@
 ttest_final_model %>% arrange(labeling) %>% 
-  filter(abs(log2FC) < Inf & label %in% c("mix1-mix3", "mix1-mix4", "mix2-mix3", "mix2-mix4")) %>%
+  filter(abs(log2FC) < Inf) %>%
   ggplot(aes(log2FC, -log10(adj.pvalue))) + 
@@
   geom_vline(xintercept = 0) + 
+  geom_vline(aes(xintercept = trueLog2FC),color = "darkred", linetype= "dashed") +
   geom_hline(yintercept = -log10(0.05), color = "darkred") + 

### 2022-07-11 main/MSstatsPTM_Main.tex
updated protein wording

--- 2022-07-11 main/MSstatsPTM_Main.tex
@@
 The statistical methodology separately summarizes the abundances of peptides with and without the modification sites, and fits separate linear mixed effects models that reflect the biological and technological aspects of the experimental design.
-Next, model-based inferences regarding the PTM and the protein-level abundances are combined to account for the confounding between these two sources.
+Next, model-based inferences regarding the PTM and the unmodified protein-level abundances are combined to account for the confounding between these two sources.
 
@@
 The datasets illustrate a diverse set of organisms, modification types, acquisition methods and experimental designs, showing the applicability of the framework to a variety of situations.
-By appropriately leveraging the information from the unmodified peptides, the proposed approach improved the accuracy of the estimates of PTM fold changes, and produced a better calibrated false positive rate of detecting differentially abundant PTMs as compared to existing methods.
-In particular, accounting for the confounding from protein abundance allowed us to characterize the true effect of the modification, avoiding the need for more manual and time intensive follow-up investigation.
+By appropriately leveraging the information from the unmodified proteins, the proposed approach improved the accuracy of the estimates of PTM fold changes, and produced a better calibrated false positive rate of detecting differentially abundant PTMs as compared to existing methods.
+In particular, accounting for the confounding from unmodified protein abundance allowed us to characterize the true effect of the modification, avoiding the need for more manual and time intensive follow-up investigation.
 
@@
 The log$_2$-intensity of Feature $k$, in Replicate $j$ of Condition $i$ is denoted by $y_{ijk}^{\ast}$.
-Conversely, the unmodified protein is represented by $L$ spectral features, and the log-intensity of Feature $l$ from the unmodified peptides in the same run is denoted by $y_{ijl}$.
+Conversely, the unmodified protein is represented by $L$ spectral features, and the log-intensity of Feature $l$ from the unmodified protein in the same run is denoted by $y_{ijl}$.
 The features can be quantified as part of a same mass spectrometry run, or in separate enrichment and global proteome profiling runs.
@@
 Based on the model, the approach tests the null hypothesis in \eqref{conv_null_hyp}, by comparing the test statistic against the Student distribution with $df=2J-2$ degrees of freedom in balanced designs.
-Unfortunately, this approach is fundamentally flawed as it does not account for the confounding between changes in the PTM abundance and the overall changes in the abundance of the protein.
+Unfortunately, this approach is fundamentally flawed as it does not account for the confounding between changes in the PTM abundance and the overall changes in the abundance of the unmodified protein.
 
@@
 
-\todo{DK: Current plan is to follow what Ting did for MSstatsTMT and keep this part brief but expand on EB methods in the supplementary.}\medskip \noindent The two versions of $t$-test \todo{OV: ANOVA?} above can be expanded with Empirical Bayes moderation in Limma~\cite{ Ritchie_15a, Schwammle2015, Smyth:2004, Smyth:2005, Zhu,Chappell:2021}.
+\medskip \noindent The two versions of $t$-test \todo{OV: ANOVA?} above can be expanded with Empirical Bayes moderation in Limma~\cite{ Ritchie_15a, Schwammle2015, Smyth:2004, Smyth:2005, Zhu,Chappell:2021}.
 In particular, the ratio-based version which models the ratios $\hat{u}_{ij}$ in~\eqref{adj_estimation} can be used to answer the null hypothesis in \eqref{null_hyp}.
@@
 We leverage the existing methods in $MSstats$ and $MSstatsTMT$ and extend them to remove confounding between changes in the PTM and the overall changes in protein abundance.
-Each feature type is summarized separately using $MSstats$ or $MSstatsTMT$, resulting in run level values for the PTM, $\hat{y}_{ij}^{\ast}$, and unmodified peptide.
+Each feature type is summarized separately using $MSstats$ or $MSstatsTMT$, resulting in run level values for the PTM, $\hat{y}_{ij}^{\ast}$, and unmodified protein.
 The estimated summaries of the modified features are then used as the input to the model, such as in \eqref{msstats_model}, in $MSstats$ or $MSstatsTMT$.
@@
 
-\todo{Could we try to use a consistent vocabulary for 'global protein' (we sometimes use unmodified peptides, unmodified features, global profiling etc.) Could you check the entire manuscript and see that we use the same terminology for that}
-
 The proposed methods are implemented in the open source R package $MSstatsPTM$, overviewed in~\figref{msstatsptm_design}.
-\todo{$MSstatsPTM$ is applicable to most experimental designs, label-free data acquisitions such as XXX and label-based data acquisitions such as XXX} $MSstatsPTM$ takes as input lists of identified and quantified spectral features, produced by spectral processing tools such as MaxQuant, Progenesis, or Spectronaut.
+$MSstatsPTM$ is applicable to most experimental designs, label-free data acquisitions such as DDA, DIA, SRM and label-based data acquisitions such as TMT.
+$MSstatsPTM$ takes as input lists of identified and quantified spectral features, produced by spectral processing tools such as MaxQuant, Progenesis, or Spectronaut.
 \todo{Could you rephrase the next sentence to be more specific.
@@
 
-\todo{Could you rephrase the next sentence?
-What is summarized on the PTM side?
-The peptide sequence?
-Or any peptide sequence that contains the modification site?} The next step is peptide/protein summarization.
-Summarization is performed separately for the PTM and global protein in the same way as described above for $MSstats$.
-When summarizing the unmodified peptide features, the package optionally imputes missing values using an Accelerated Failure Time (AFT) model~\cite{Wei:1992}.
+The next step is PTM/protein summarization.
+Summarization is performed separately for the PTM and unmodified protein.
+When summarizing the PTM, modified peptide features that span the same modification site are summarized together.
+Peptides that include multiple modifications are not included in the single modification summarization, and are grouped separately.
+The unmodified protein summarization is performed as discussed above for $MSstats$.
+When summarizing the unmodified protein features, the package optionally imputes missing values using an Accelerated Failure Time (AFT) model~\cite{Wei:1992}.
 When summarizing the modified features, missing value imputation is also possible but should be performed with care.
@@
 We evaluated $MSstatsPTM$ on simulated and spike-in datasets with known ground truth in terms of true positives ($TP$), false positives ($FP$), true negatives ($TN$), and false negatives ($FN$) differentially abundant PTMs.
-The true positives were defined as PTMs with changes distinct from the overall changes in protein abundance.
+The true positives were defined as PTMs with changes distinct from the overall changes in unmodified protein abundance.
 The true negatives were defined as PTMs which, after accounting for the changes in the overall protein abundance, were not differentially abundant.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-information \subsubsection*{In the label-free benchmark experiment, MSstatsPTM had a higher sensitivity than the existing methods}
-
-In this experiment all models incorrectly estimated the fold change of the modified spike-in peptides before adjusting for changes in global protein abundance.
+\subsubsection*{In the label-free benchmark experiment, MSstatsPTM had a higher sensitivity than the existing methods}
+
+In this experiment all models incorrectly estimated the fold change of the modified spike-in peptides before adjusting for changes in unmodified protein abundance.
 After adjustment, the spike-in peptides' fold change was generally in line with expectation in all methods, however the distribution of estimated fold changes was visibly wider (\figref{spikein_boxplot}).
@@
 \figref{data4_venn_diagram} and~\figref{data5_venn_diagram} show the number of differentially abundant modified peptides before and after adjustment.
-Adjusting for changes in the global protein caused more PTMs to change from differentially abundant to non-differentially abundant than vice versa.
+Adjusting for changes in the unmodified protein caused more PTMs to change from differentially abundant to non-differentially abundant than vice versa.
 A question that must be addressed is if the decrease in differentially abundant peptides is due to the increased variance that comes from adjustment.
@@
 In other words, the fold change was the same between models but variance increased.
-When the test was applied to Dataset 4, only one peptide became non-differentially abundant due to an increase in variance.
-When applied to Dataset 5, 548 peptides became non-differentially abundant due to an increase in variance (only 3.4\% of all PTMs that became became non-differentially abundant).
-Thus we can conclude that the drop off in differentially abundant peptides was due to changes in global protein abundance, not simply increased model variance.
-
-Interestingly, the test above showed that the global protein was more likely to cause observed changes in PTM abundance, rather than mask changes in the PTM.
-The later case occurs when the global protein exhibits a change in abundance, while the modification does not.
+When the test was applied to Dataset 4, only one PTM became non-differentially abundant due to an increase in variance.
+When applied to Dataset 5, 548 PTMs became non-differentially abundant due to an increase in variance (only 3.4\% of all PTMs that became became non-differentially abundant).
+Thus we can conclude that the drop off in differentially abundant PTMs was due to changes in global protein abundance, not simply increased model variance.
+
+Interestingly, the test above showed that the unmodified protein was more likely to cause observed changes in PTM abundance, rather than mask changes in the PTM.
+The later case occurs when the unmodified protein exhibits a change in abundance, while the modification does not.
 An example of this can be seen in \figref{data5_profile_plot}.
 Originally the modification at $S178$ of protein $TTP$ was shown to be differentially abundant between KO\_Uninfect and KO\_Early, with a log fold change of $2.9$.
-However the global protein was shown to contribute $2.014$ of this change, meaning that the modification only accounted for $.886$ after adjustment.
+However the unmodified protein was shown to contribute $2.014$ of this change, meaning that the modification only accounted for $.886$ after adjustment.
 This caused the modification to go from differentially abundant to non-differentially abundant.
-As discussed previously, these cases occur frequently and can be thought of as false positives that would occur without accounting for confounding with the global protein.
+As discussed previously, these cases occur frequently and can be thought of as false positives which are caused by confounding with the unmodified protein.
 The proposed approach corrects for this confounding, allowing us to see the true fold change of PTMs, while excluding those that do not exhibit differential changes.
 
-Conversely, cases where the global protein masks changes in the PTM are more infrequent.
-In this case, the modification and protein exhibit contradictory changes in abundance, moving in opposite directions.
+Conversely, cases where the unmodified protein masks changes in the PTM are more infrequent.
+In this case, the modification and unmodified protein exhibit contradictory changes in abundance, moving in opposite directions.
 One such case can be seen in~\figref{data4_profile_plot}.
 Luchetti et al.~\cite{LUCHETTI2021} showed that $GSDMD$ was actively degraded when IpaH7.8 expression was induced by dox treatment.
-Our reanalysis confirmed this, with the global $GSDMD$ protein being downregulated when Dox treatments reached the 4 and 6 hour marks.
+Our reanalysis confirmed this, with the $GSDMD$ protein being downregulated when Dox treatments reached the 4 and 6 hour marks.
 Conversely, ubiquitination of $GSDMD$ at site $K62$ upregulated abundance between the same conditions.
-This upregulation was originally confounded by the downregulation of unmodified $GSDMD$, appearing to have little change between no dox and dox 4 and 6 hour.
-The proposed approach accounted for this confounding and the modification was shown to be differentially abundant, with an absolute fold change of $1.415$ between the Dox1hr and Dox4hr conditions.
-This was an interesting result, as the modification contradicts what previous research showed for the global protein.
-This effect can be thought of as a false negative and would have been more challenging to observe without the proposed approach.
+This upregulation was originally confounded by the downregulation of unmodified $GSDMD$, appearing to have little change between no Dox and Dox 4 and 6 hour.
+The proposed approach accounted for this confounding and the modification was shown to be differentially abundant, with an absolute fold change of $1.415$ between the Dox 1 hour and Dox 4 hour conditions.
+This was an interesting result, as the modification contradicts what previous research showed for the unmodified protein.
+This effect can be thought of as a false negative and would have been challenging to observe without the proposed approach.
 There are many non-differentially abundant PTMs, and having to comb through each one individually to determine false negatives is generally an unfeasible task.
-The proposed approach removes the confounding on an experimental level, revealing interesting PTMs that are entirely masked due to changes in the global protein.
+The proposed approach removes the confounding on an experimental level, revealing interesting PTMs that are entirely masked due to changes in the unmodified protein.
 
@@
 Once identification and quantification of the Ubiquitination profiling was performed, peptides which were unmodified were extracted and used in place of a global profiling run.
-This resulted in a lack of overlap between modified and unmodified peptides.
-Any modified peptide without a corresponding unmodified protein could not be adjusted.
+This resulted in a lack of overlap between PTMs and unmodified proteins.
+Any PTM without a corresponding unmodified protein could not be adjusted.
 Of the 10,799 ubiquitin sites identified, only 4,526 had a corresponding unmodified protein and could be adjusted.
@@
 The number of differentially abundant PTMs before and after adjustment are shown in~\figref{data6_plots}.
-After adjusting for changes in the unmodified protein, there were fewer significant modified peptides.
+After adjusting for changes in the unmodified protein, there were fewer differentially modified peptides.
 However, this was mainly due to not having a global profiling run.
-When looking only at PTMs with a corresponding global protein, the number of differentially modified PTMs exhibited a small drop, but was more consistent.
+When looking only at PTMs with a corresponding unmodified protein, the number of differentially modified PTMs exhibited a small drop, but was more consistent.
 As above, we tested if this drop in differentially abundant PTMs was due to an increase in variance.
 Here only 25 PTMs became insignificant due to increased variance.
-When there was no global profiling run the methods were less effective at removing confounding with the global protein.
+When there was no global profiling run the methods were less effective at removing confounding with the unmodified protein.
 This was entirely due to not being able to perform adjustment without a corresponding global protein.

### 2022-07-12 code/Spike-in_Analysis.Rmd
small updates to results

@@
 
+combo_plot_df
+combo_plot_df[combo_plot_df$Label == 'mix2-mix3'][['trueLog2FC']] <- combo_plot_df[combo_plot_df$Label == 'mix2-mix3'][['trueLog2FC']]*-1
+
 png("../main/images/spike_in_volcano.png", width = 750, height = 500)
@@
 ttest_final_model$trueLog2FC=ttest_final_model$trueLog2FC*-1
-ttest_final_model[ttest_final_model$Label == 'mix1-mix2'][['trueLog2FC']] <- ttest_final_model[ttest_final_model$Label == 'mix1-mix2'][['trueLog2FC']]*-1
-ttest_final_model[ttest_final_model$Label == 'mix1-mix3'][['trueLog2FC']] <- ttest_final_model[ttest_final_model$Label == 'mix1-mix3'][['trueLog2FC']]*-1
+ttest_final_model[ttest_final_model$label == 'mix1-mix2'][['trueLog2FC']] <- ttest_final_model[ttest_final_model$label == 'mix1-mix2'][['trueLog2FC']]*-1
+ttest_final_model[ttest_final_model$label == 'mix3-mix4'][['trueLog2FC']] <- ttest_final_model[ttest_final_model$label == 'mix3-mix4'][['trueLog2FC']]*-1
+ttest_final_model[ttest_final_model$label == 'mix1-mix3'][['trueLog2FC']] <- ttest_final_model[ttest_final_model$label == 'mix1-mix3'][['trueLog2FC']]*-1
+
+ttest_final_model$contrast_truth = factor(ttest_final_model$contrast_truth, 
+                                          levels = 
+                                  c("LogFC: -1 (mix2-mix1)",
+                                    "LogFC: 1 (mix3-mix1)",
+                                    "LogFC: 0 (mix4-mix1)",
+                                    "LogFC: 2 (mix3-mix2)",
+                                    "LogFC: 1 (mix4-mix2)",
+                                    "LogFC: -1 (mix4-mix3)"))
 
@@
 
+spike[spike$label == 'mix1-mix2'][['contrast_truth']] <- "LogFC: -1 (mix2-mix1)"
+spike[spike$label == 'mix1-mix3'][['contrast_truth']] <- "LogFC: 1 (mix3-mix1)"
+spike[spike$label == 'mix1-mix4'][['contrast_truth']] <- "LogFC: 0 (mix4-mix1)"
+spike[spike$label == 'mix2-mix3'][['contrast_truth']] <- "LogFC: 2 (mix3-mix2)"
+spike[spike$label == 'mix2-mix4'][['contrast_truth']] <- "LogFC: 1 (mix4-mix2)"
+spike[spike$label == 'mix3-mix4'][['contrast_truth']] <- "LogFC: -1 (mix4-mix3)"
+
+spike[spike$Label == 'mix3-mix4'][["Label"]] = "mix4 vs mix3"
+spike[spike$Label == 'mix2-mix4'][["Label"]] = "mix4 vs mix2"
+spike[spike$Label == 'mix1-mix4'][["Label"]] = "mix4 vs mix1"
+spike[spike$Label == 'mix2-mix3'][["Label"]] = "mix3 vs mix2"
+spike[spike$Label == 'mix1-mix3'][["Label"]] = "mix3 vs mix1"
+spike[spike$Label == 'mix1-mix2'][["Label"]] = "mix2 vs mix1"
+
+spike[spike$Label == "mix3 vs mix2" & 
+        spike$Model == "T-test"][["trueLog2FC"]] = 2
+spike[spike$Label == "mix3 vs mix2" & 
+        spike$Model == "T-test"][["log2FC"]] = spike[
+          spike$Label == "mix3 vs mix2" & 
+        spike$Model == "T-test"][["log2FC"]] * -1
+spike[spike$Label == "mix4 vs mix2" & 
+        spike$Model == "T-test"][["trueLog2FC"]] = 1
+spike[spike$Label == "mix4 vs mix2" & 
+        spike$Model == "T-test"][["log2FC"]] = spike[
+          spike$Label == "mix4 vs mix2" & 
+        spike$Model == "T-test"][["log2FC"]] * -1
+spike[spike$Label == "mix4 vs mix3" & 
+        spike$Model == "T-test"][["trueLog2FC"]] = -1
+spike[spike$Label == "mix4 vs mix3" & 
+        spike$Model == "T-test"][["log2FC"]] = spike[
+          spike$Label == "mix4 vs mix3" & 
+        spike$Model == "T-test"][["log2FC"]] * -1
+
+spike[spike$Label == "mix3 vs mix2" & 
+        spike$Model == "Limma"][["trueLog2FC"]] = 2
+spike[spike$Label == "mix4 vs mix2" & 
+        spike$Model == "Limma"][["trueLog2FC"]] = 1
+spike[spike$Label == "mix4 vs mix3" & 
+        spike$Model == "Limma"][["trueLog2FC"]] = -1
+
 png("../supplementary/sim_new/spike_in_fc.png", width = 750, height = 500)

### 2022-07-12 main/MSstatsPTM_Main.tex
small updates to results

@@
 
-Interestingly, the test above showed that the unmodified protein was more likely to cause observed changes in PTM abundance, rather than mask changes in the PTM.
-The later case occurs when the unmodified protein exhibits a change in abundance, while the modification does not.
+\figref{venn_diagrams} demonstrates that the unmodified protein was more likely to cause the observed changes in PTM abundance, rather than mask changes in the PTM.
+The later case can occur when the unmodified protein exhibits a change in abundance, while the modification does not.
+Additionally, this case may come about due to the high variability of PTMs compared to the global protein.
 An example of this can be seen in \figref{data5_profile_plot}.
@@
 This caused the modification to go from differentially abundant to non-differentially abundant.
-As discussed previously, these cases occur frequently and can be thought of as false positives which are caused by confounding with the unmodified protein.
+These cases occur frequently and can be thought of as false positives.
 The proposed approach corrects for this confounding, allowing us to see the true fold change of PTMs, while excluding those that do not exhibit differential changes.
@@
 Luchetti et al.~\cite{LUCHETTI2021} showed that $GSDMD$ was actively degraded when IpaH7.8 expression was induced by dox treatment.
-Our reanalysis confirmed this, with the $GSDMD$ protein being downregulated when Dox treatments reached the 4 and 6 hour marks.
-Conversely, ubiquitination of $GSDMD$ at site $K62$ upregulated abundance between the same conditions.
-This upregulation was originally confounded by the downregulation of unmodified $GSDMD$, appearing to have little change between no Dox and Dox 4 and 6 hour.
+Our reanalysis confirmed this, with the $GSDMD$ protein being down-regulated when Dox treatments reached the 4 and 6 hour marks.
+Conversely, ubiquitination of $GSDMD$ at site $K62$ up-regulated abundance between the same conditions.
+This up-regulation was originally confounded by the down-regulation of unmodified $GSDMD$, and made the modification appear to have little change between no Dox and Dox 4 and 6 hour conditions.
 The proposed approach accounted for this confounding and the modification was shown to be differentially abundant, with an absolute fold change of $1.415$ between the Dox 1 hour and Dox 4 hour conditions.

### 2022-07-12 main/MSstatsPTM_Main.tex
spike in fig comments

@@
 \medskip \noindent {\bf Experimental design:}~\figref{benchmark-design}(a) overviews the experimental design.
-Four mixtures (i.e., conditions) were created with varying amounts of human lysate \todo{which cell line?}, background {\it E.
+Four mixtures (i.e., conditions) were created with varying amounts of human lysate, background {\it E.
 Coli} lysate, and human spike-in ub-peptide mixture.
@@
 \medskip \noindent {\bf Evaluation:} We expect the relative abundances of the spike-in peptides to change as in~\figref{benchmark-design}(b).
-The changes in peptide abundances in all the comparisons except Mix 1 vs Mix 4 were distinct from changes in the global proteome abundances and distinct from zero, and were viewed as positive controls.
-In the comparison of Mix1 vs Mix 4 both the modified peptides and the global proteome background changed two-fold, and as the result the peptides in this comparison were viewed as a negative control.
-Since the background {\it E.
-Coli} lysate peptides were not expected to change in abundance in any comparison \todo{in any, or in 3 vs 4?}, they were viewed as additional (unadjusted) negative controls.
+The changes in peptide abundances in all the comparisons except Mix 4 vs Mix 1 were distinct from changes in the global proteome abundances and distinct from zero, and were viewed as positive controls.
+In the comparison of Mix4 vs Mix 1 both the modified peptides and the global proteome background changed two-fold, and as the result the peptides in this comparison were viewed as a negative control.
+The background {\it E.
+Coli} lysate peptides were not expected to change in abundance in comparison, after accounting for adjustment, and were viewed as additional negative controls.
 We evaluated the statistical methods ability to avoid false positives, as well as their sensitivity in detecting the differentially abundant spike-in peptides and accurately estimate their expected fold change.
@@
 %\hline
-%Comparison & Unadj. true FC & Adj. true FC \\ %\hline \hline
+%Comparison & Unadj. true log$_{2}$FC & Adj. true log$_{2}$FC \\ %\hline \hline
 %mix2 vs mix1  & -1 & -1 \\ \hline
@@
 \caption{{\bf Dataset 3: Spike-in benchmark - Ubiquitination - Label-free}. 
-\todo{In the legend, could you mark the blue box "Human lysate (global proteome)", the grey box "E.
-Coli lysate (background proteome)" and mark the red box as "50 spiked human ub-peptides".
-Check the capitalization} \todo{In the table, could you change the headers to log$_2$FC}
 %
-(a) Four mixtures (i.e., conditions) were created with varying amounts of human lysate \todo{which cell line?}, background {\it E.
+(a) Four mixtures (i.e., conditions) were created with varying amounts of human lysate, background {\it E.
 Coli} lysate, and human spike-in ub-peptide mixture.
@@
 ``Adj.
-true log$_2$FC'' \todo{Could you rephrase the calculation more precisely?
-Ratios of the ratios?
-I am wondering if a formula wold be helpful?} was calculated by determining the ratio of spike-in peptides to human lysate in each mixture and then using the ratio to calculate the true fold change across comparisons.}
+true log$_2$FC'' was calculated by determining the ratios of the abundances of the spiked peptides and human lysate between each condition and then adjusting the ratio of the spiked peptides by the human lysate, similarly to \eqref{null_hyp}.}
 \label{fig:benchmark-design}

### 2022-07-12 main/MSstatsPTM_Main.tex
SE example formula and sample size notation

@@
 For simplicity, we assume that the experiment has no technical replicates, such that each biological replicate corresponds to an LC-MS/MS run.
-Figure \ref{fig:data-structure} schematically illustrates this data structure for one PTM site, in a special case of $I=2$ conditions and $J=2$ biological replicates per condition.
+\figref{data-structure} schematically illustrates this data structure for one PTM site, in a special case of $I=2$ conditions and $J=2$ biological replicates per condition.
 For one protein, the PTM site is represented by $K$ spectral features (i.e., peptide ions, distinguished by their cleavage residues and charge states). 
@@
 \begin{eqnarray}
-\widehat{SE}(\hat{\Delta})_{adj} = \sqrt{ \widehat{SE}(\hat{\Delta}_{PTM})^{2} + \widehat{SE}(\hat{\Delta}_{protein})^{2}} 
+\widehat{SE}(\hat{\Delta}_{adj}) = \sqrt{ \widehat{SE}(\hat{\Delta}_{PTM})^{2} + \widehat{SE}(\hat{\Delta}_{protein})^{2}} 
 \label{eq:prop_se}
 \end{eqnarray}
+
+For example, in the simple case of \figref{data-structure} with $J=2$ conditions, where $\hat{\sigma}^2_{PTM}$ and $\hat{\sigma}^2_{protein}$ are the estimates of the error variance for the PTM and protein model described in \eqref{msstats_model}, the standard error would be calculated as
+
+\begin{eqnarray}
+& \widehat{SE}(\hat{\Delta})_{adj} = \sqrt{\widehat{SE}(\hat{\Delta}_{PTM})^2 + \widehat{SE}(\hat{\Delta}_{protein})^2} = &\\ 
+& \sqrt{\frac{1}{J}\hat{\sigma}^2_{PTM} + \frac{1}{J}\hat{\sigma}^2_{protein}} = \sqrt{\frac{2}{J}(\hat{\sigma}^2_{PTM} + \hat{\sigma}^2_{protein})} = \sqrt{\hat{\sigma}^2_{PTM} + \hat{\sigma}^2_{protein}} & \nonumber
+\label{eq:se_calc}
+\end{eqnarray}
+
 The estimated standard error is larger than the standard errors associated with each distinct feature type, reflecting the increased uncertainty inherent to combining the uncertainty in two estimators.
+
 Finally, the degrees of freedom associated with~\eqref{prop_se} are obtained via the Satterthwaite approximation~\cite{kutner_etal_04a, satterthwaite:1946}
@@
 %Sample size calculation based on linear models has been described for general applications~\cite{kutner_etal_04a} and specifically for protein significance analysis~\cite{oberg_vitek_09a}. 
-Sample size calculation takes as input a) $q$, the desired false discovery rate, b) $\beta$, the average Type II error rate, c) $\Delta$, the minimal log-fold change in adjusted PTM abundance that we would like to detect, d) $m_0 / (m_0 + m_1)$, the fraction of truly differentially modified PTM sites in the comparison, and e) $\sigma_{\gamma^{\ast}}^{2}$ and $\sigma_{\gamma}^{2}$, the anticipated variances associated to modified and unmodified peptide features, respectively.
+Sample size calculation takes as input a) $q$, the desired false discovery rate, b) $\beta$, the average Type II error rate, c) $\Delta_{adj}$, the minimal log-fold change in adjusted PTM abundance that we would like to detect, d) $m_0 / (m_0 + m_1)$, the fraction of truly differentially modified PTM sites in the comparison, and e) $\sigma_{PTM}^{2}$ and $\sigma_{protein}^{2}$, the anticipated variances associated to modified and unmodified peptide features, respectively.
 The variances can be derived based on the dataset being analyzed, assuming similar quantitative properties and variations.
 With these values and a user-specified number of conditions, the corresponding number of technical replicates per condition can then be derived, as described in~\cite{kutner_etal_04a}.
-Given the above quantities, the minimal number of replicates $J$ is determined by the variance of the estimated log-fold change $\mathrm{SE}^{2}(\hat{\Delta})$ as \todo{OV: hmm...
-For what design and what model?
-It is unclear where eq 11 came from.
-This will be quite specific to the model and the design.
-Maybe refer to the model in Eq 7?
-(Which we should probably simplify to reflect the experiment in Figure2, while keeping the general indeces) }
+Given the above quantities, the minimal number of replicates $J$ is determined by the variance of the estimated log-fold change $\mathrm{SE}^{2}(\hat{\Delta}_{adj})$ as
 
 \begin{equation}
-\widehat{SE}(\hat{\Delta})_{adj}^2 = \left[ \frac{2}{J} \left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right) \right]
-\leq \left( \frac{\Delta}{t_{1-\beta, df} + t_{1-\alpha /2, df}} \right)^{2}
+\widehat{SE}(\hat{\Delta}_{adj})^2 = \left[ \frac{2}{J} \left( \hat{\sigma}_{PTM}^{2} + \hat{\sigma}_{protein}^{2} \right) \right]
+\leq \left( \frac{\Delta_{adj}}{t_{1-\beta, df} + t_{1-\alpha /2, df}} \right)^{2}
 \end{equation}
@@
 \begin{equation}
-J \geq \frac{(2\hat{\sigma}_{\gamma^{\ast}}^{2} + 2\hat{\sigma}_{\gamma}^{2})(t_{1-\beta, df} + t_{1-\alpha /2, df})^2}{\Delta^2}
+J \geq \frac{(2\hat{\sigma}_{PTM}^{2} + 2\hat{\sigma}_{protein}^{2})(t_{1-\beta, df} + t_{1-\alpha /2, df})^2}{\Delta_{adj}^2}
 \end{equation}

### 2022-07-12 main/MSstatsPTM_Main.tex
updated implementation figure

@@
 \centering
-\includegraphics[scale=.8]{images/MSstatsPTM_design.png}
+\includegraphics[scale=.6]{images/MSstatsPTM_design.png}
 \caption{The $MSstatsPTM$ workflow.
 The names of the R functions used for each step are highlighted in red.
-\todo{In the caption, could you try the following order: applicable to experimental designs XXX; takes as input XXX; separate parallel analyses for global proteome and PTM.
-(No need to read off the figure, just clarify the content)} \todo{Would it be possible to modify the figure to reflect the parallel nature of the processing/analyses?} The package's workflow starts after modified and unmodified peptide quantification.
-First tool specific pre-processing is performed, this includes modification site identification, general data cleaning, and formatting the data into the format needed for the package.
-The next step is feature level summarization, which summarizes features up to the modification level for the PTM data, and the protein level for the protein data.
-In the final step a model is fit to identify differential PTMs and unmodified proteins across conditions and the PTM model is adjusted for changes in the unmodified protein.
-After both the summarization and group comparison steps, plots can be created to summarize the results.}
+The package is applicable to label-free data acquisitions such as DDA, DIA, SRM and label-based data acquisitions such as TMT.
+The package's workflow starts after modified and unmodified peptide quantification.
+It takes as input lists of identified and quantified spectral features for the PTM and unmodified protein, produced by spectral processing tools such as MaxQuant, Progenesis, or Spectronaut.
+Separate parallel analysis for the PTM and unmodified protein is performed, including feature summarization and modeling.
+After a model is fit to both the PTM and unmodified protein, the PTM model is adjusted for changes in unmodified protein abundance.
+Plots can optionally be created at each step to summarize the results.}
 \label{fig:msstatsptm_design}

### 2022-07-12 main/MSstatsPTM_Main.tex
minor fig caption changes

@@
 \end{subfigure}
-\caption{a) Dataset 4: Human - Ubiquitination - 1mix-TMT.
+\caption{
+\todo{Update the figure titles}
+%
+a) Dataset 4: Human - Ubiquitination - 1mix-TMT.
 The overlap of differential modified peptides for the PTM model with and without global protein level adjustment across all comparisons.
@@
 b) Dataset 5: Mouse - Phosphorylation - 2mix-TMT.
-a) The overlap of differentially modified peptides between the PTM model with and without global protein level adjustment across all comparisons.
+The overlap of differentially modified peptides between the PTM model with and without global protein level adjustment across all comparisons.
 19,286 peptides became insignificant, 4,947 became significant, and 41,552 were significant in both models.
@@
 \caption{a) Dataset 5: Mouse - Phosphorylation - 2mix-TMT.
-Comparing the global profiling of protein $TTP$ with the modification of the protein at site $S178$.
-When looking at the summary of the modification and global protein it was clear the difference between conditions followed the same trend.
+Comparing the global profiling of protein $TTP$ with the modification at site $S178$.
+The modification and unmodified protein abundance followed the same trend across conditions.
 Specifically, there was a positive adjustment in abundance when comparing WT\_Uninfect to WT\_Late in both the modification and global profiling run.
-This indicated the movement was driven by changes in global protein that was only accounted for in the model after adjusting for global protein abundance change.
 b) Dataset 4: Human - Ubiquitination - 1mix-TMT.
 Comparing the global profiling of protein $GSDMD$ with the ubiquitination of the protein at site $K62$.
-When looking at the summary of the modification and global protein it was clear the conditions follow different trends.
-Specifically, there appeared to be no change in abundance between Dox1hr and Dox4hr in the modified plot, however there was a large negative change when looking at the unmodified plot.
+Here the modification and global protein follow different trends.
+There appeared to be no change in abundance between Dox1hr and Dox4hr in the modified plot, however there was a large negative change when looking at the unmodified plot.
 This indicated the modification was confounded with changes in the unmodified protein.}

### 2022-07-13 main/MSstatsPTM_Main.tex
implementation converter section

@@
 $MSstatsPTM$ takes as input lists of identified and quantified spectral features, produced by spectral processing tools such as MaxQuant, Progenesis, or Spectronaut.
-\todo{Could you rephrase the next sentence to be more specific.
-If someone is a tool developer, what exactly should they do?} To identify the local modification site we require the tool to identify where in the sequence the modification occurred.
+We require the processing tool to identify the amino acid in the protein sequence where the modification occurred (i.e.
+the modification site).
+This will generally include the amino acid abbreviation, plus it's number in the protein sequence.
+For example, a modification on a serine, which was the 70th amino acid in the sequence would be marked ``S70''.
+Occasionally tools will only include the peptide sequence with the modified amino acid highlighted.
+If there is no way to determine where in the protein sequence the modification occurred, $MSstatsPTM$ includes functionality to do this manually, given the modified peptide sequence and a FASTA file with the entire protein sequence.
 The package includes converters for seamless integration with multiple such tools.
-\todo{Could you clarify what the converters do, especially for PTM?
-Do we discard peptides with multiple modification sites?}.
+The converters can optionally keep peptides with both single and multiple modification sites, or filter out only those with multiple sites.
+If multiple sites are kept, these are grouped separately from the singly modified peptide so as not to confound the effect.
+Additionally, peptides with oxidation can also be removed if the effect is believed to confound the modifications of interest.
 For experimental designs that include separate runs enriched in modified peptides and global profiling runs, conversion is performed separately for the runs enriched in modified peptides and for the global profiling runs.
-For experimental designs that include modified and unmodified peptides in a same run, \todo{what happens for the conversion?}
 

### 2022-07-13 main/MSstatsPTM_Main.tex
removed comments in evaluation that were (maybe?) addressed

@@
 
-\todo{OV:Most of the following sections simply re-state the results that can already be read off the figures.
-The sections would have more value if in addition/instead of describing the figures, we could provide the reasons/insight into why we see this performance.
-Could you try?
-In general, no need to repeat what we see in the figure - this is what the figures are for.}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsubsection*{In clean simulations, MSstatsPTM corrected for high FDR of differentially abundant PTM}
-
-\todo{OV: maybe say something about the fact that this is a group comparisons design that doesn't require many variance components, and as such is favorable to Limma and t-test?
-Later when you have repeated measurements, clarify that these situations are more challenging for the existing methods and therefore the difference in the results is more pronounced.}
 
@@
 \subsubsection*{In noisy simulations, MSstatsPTM better calibrated the true fold change over the existing methods}
-
-\todo{This section just describes the figure. Could you emphasize why we see these results? What are the reasons for differences in performance?}
 

### 2022-07-14 code/Spike-in_Analysis.Rmd
updated limma model

@@
   fpr <- df %>% filter((get(pval_col_name) < .05 & labeling == "light") | 
-                (get(comparison_col_name) == "mix1-mix4" & get(pval_col_name) < .05)) %>% 
-    nrow() / nrow(df %>% filter(labeling == "light"))
+                (get(comparison_col_name) == "mix4 vs mix1" & get(pval_col_name) < .05)) %>% 
+    nrow() / nrow(df %>% filter(labeling == "light" | get(comparison_col_name) == "mix4 vs mix1"))
 
@@
   recall <- df %>% filter(get(pval_col_name) < .05 & labeling == "heavy" & 
-                                 get(comparison_col_name) != "mix1-mix4") %>% 
-    nrow() / df %>% filter(labeling == "heavy" & 
-                             get(comparison_col_name) != "mix1-mix4") %>% nrow()
+                                 get(comparison_col_name) != "mix4 vs mix1") %>% 
+    nrow() / (50*5)
 
@@
   specificity <- df %>% filter(get(pval_col_name) >= .05 & (labeling == "light" |
-                                 get(comparison_col_name) == "mix1-mix4")) %>% 
-    nrow() / nrow(df %>% filter(labeling == "light" | get(comparison_col_name) == "mix1-mix4"))
+                                 get(comparison_col_name) == "mix4 vs mix1")) %>% 
+    nrow() / nrow(df %>% filter(labeling == "light" | get(comparison_col_name) == "mix4 vs mix1"))
 
@@
   precision <- df %>% filter(get(pval_col_name) < .05 & labeling == "heavy" & 
-                               get(comparison_col_name) != "mix1-mix4") %>% 
+                               get(comparison_col_name) != "mix4 vs mix1") %>% 
     nrow() / nrow(df %>% filter(get(pval_col_name) < .05))
@@
   accuracy <- df %>% filter((get(pval_col_name) < .05 & labeling == "heavy" & 
-                               get(comparison_col_name) != "mix1-mix4") | 
+                               get(comparison_col_name) != "mix4 vs mix1") | 
                               (get(pval_col_name) >= .05 & (labeling == "light" | 
-                               get(comparison_col_name) == "mix1-mix4"))) %>% 
+                               get(comparison_col_name) == "mix4 vs mix1"))) %>% 
     nrow() / nrow(df)
@@
 ## Update some naming conventions
-plot_df[plot_df$Label == 'mix1-mix2'][['contrast_truth']] <- "LogFC: -1 (mix2-mix1)"
-plot_df[plot_df$Label == 'mix1-mix3'][['contrast_truth']] <- "LogFC: 1 (mix3-mix1)"
-plot_df[plot_df$Label == 'mix1-mix4'][['contrast_truth']] <- "LogFC: 0 (mix4-mix1)"
-plot_df[plot_df$Label == 'mix2-mix3'][['contrast_truth']] <- "LogFC: 2 (mix3-mix2)"
-plot_df[plot_df$Label == 'mix2-mix4'][['contrast_truth']] <- "LogFC: 1 (mix4-mix2)"
-plot_df[plot_df$Label == 'mix3-mix4'][['contrast_truth']] <- "LogFC: -1 (mix4-mix3)"
+plot_df[plot_df$Label == 'mix2 vs mix1'][['contrast_truth']] <- "LogFC: -1 (mix2-mix1)"
+plot_df[plot_df$Label == 'mix3 vs mix1'][['contrast_truth']] <- "LogFC: 1 (mix3-mix1)"
+plot_df[plot_df$Label == 'mix4 vs mix1'][['contrast_truth']] <- "LogFC: 0 (mix4-mix1)"
+plot_df[plot_df$Label == 'mix3 vs mix2'][['contrast_truth']] <- "LogFC: 2 (mix3-mix2)"
+plot_df[plot_df$Label == 'mix4 vs mix2'][['contrast_truth']] <- "LogFC: 1 (mix4-mix2)"
+plot_df[plot_df$Label == 'mix4 vs mix3'][['contrast_truth']] <- "LogFC: -1 (mix4-mix3)"
 plot_df$contrast_truth = factor(plot_df$contrast_truth, levels = 
@@
 plot_df[plot_df$labeling == 'heavy'][['labeling']] <- "Spike-in Peptides"
-plot_df$trueLog2FC=plot_df$trueLog2FC*-1
-plot_df[plot_df$Label == 'mix1-mix2'][['trueLog2FC']] <- plot_df[plot_df$Label == 'mix1-mix2'][['trueLog2FC']]*-1
-plot_df[plot_df$Label == 'mix1-mix3'][['trueLog2FC']] <- plot_df[plot_df$Label == 'mix1-mix3'][['trueLog2FC']]*-1
 
@@
 plot_df %>% arrange(labeling) %>% 
-  filter(abs(Log2FC) < Inf) %>%
+  filter(is.finite(Log2FC)) %>%
   ggplot(aes(Log2FC, -log10(adj.pvalue))) + 
@@
        legend.box.margin=margin(-10,-10,-10,-10)) + 
-  scale_y_continuous(limits = c(0,3)) + 
+  # scale_y_continuous(limits = c(0,3)) + 
   guides(colour = guide_legend(override.aes = list(size=5))) +
@@
 
+### Anova
+
+#### Volcano Plot
+
+```{r, echo=FALSE, fig.width = 10, fig.height = 7}
+load(file = "../data/anova_results.rda")#anova_results
+
+anova_results$contrast_truth = ""
+anova_results$trueLog2FC=0
+
+## Update some naming conventions
+anova_results[anova_results$label == 'mix2 vs mix1'][['contrast_truth']] <- "LogFC: -1 (mix2-mix1)"
+anova_results[anova_results$label == 'mix3 vs mix1'][['contrast_truth']] <- "LogFC: 1 (mix3-mix1)"
+anova_results[anova_results$label == 'mix4 vs mix1'][['contrast_truth']] <- "LogFC: 0 (mix4-mix1)"
+anova_results[anova_results$label == 'mix3 vs mix2'][['contrast_truth']] <- "LogFC: 2 (mix3-mix2)"
+anova_results[anova_results$label == 'mix4 vs mix2'][['contrast_truth']] <- "LogFC: 1 (mix4-mix2)"
+anova_results[anova_results$label == 'mix4 vs mix3'][['contrast_truth']] <- "LogFC: -1 (mix4-mix3)"
+
+anova_results[anova_results$label == 'mix2 vs mix1'][['trueLog2FC']] <- -1
+anova_results[anova_results$label == 'mix3 vs mix1'][['trueLog2FC']] <- 1
+anova_results[anova_results$label == 'mix4 vs mix1'][['trueLog2FC']] <- 0
+anova_results[anova_results$label == 'mix3 vs mix2'][['trueLog2FC']] <- 2
+anova_results[anova_results$label == 'mix4 vs mix2'][['trueLog2FC']] <- 1
+anova_results[anova_results$label == 'mix4 vs mix3'][['trueLog2FC']] <- -1
+
+anova_results[anova_results$labeling == 'light'][['labeling']] <- "Background Peptides"
+anova_results[anova_results$labeling == 'heavy'][['labeling']] <- "SpikeIn Peptides"
+
+anova_results$contrast_truth = factor(anova_results$contrast_truth, 
+                                          levels = 
+                                  c("LogFC: -1 (mix2-mix1)",
+                                    "LogFC: 1 (mix3-mix1)",
+                                    "LogFC: 0 (mix4-mix1)",
+                                    "LogFC: 2 (mix3-mix2)",
+                                    "LogFC: 1 (mix4-mix2)",
+                                    "LogFC: -1 (mix4-mix3)"))
+
+# Volcano plot
+png("../supplementary/sim_new/spike_in_ttest_volcano.png", width = 750, height = 500)
+anova_results %>% arrange(labeling) %>% 
+  filter(is.finite(log2FC)) %>%
+  ggplot(aes(log2FC, -log10(adj.pvalue))) + 
+  geom_point(aes(color = labeling, alpha = labeling), size = 2.2) + 
+  scale_color_manual(values = c("#999999", "#D55E00")) + 
+  scale_alpha_manual(values = c(0.1, 1)) + 
+  #scale_size_manual(values = c(2, 1)) + 
+  geom_hline(yintercept = 0, linetype = "dashed") + 
+  geom_vline(xintercept = 0) + 
+  geom_vline(aes(xintercept = trueLog2FC),color = "darkred", linetype= "dashed") +
+  geom_hline(yintercept = -log10(0.05), color = "darkred") + 
+  facet_grid(protadj ~ contrast_truth) + 
+  theme_bw() + 
+  theme(axis.text.x = element_text(size = 14), 
+       axis.text.y = element_text(size = 14), 
+       legend.text=element_text(size=16),
+       axis.title.y = element_text(size = 18, vjust=2),
+       axis.title.x = element_text(size = 18),
+       title = element_text(size = 20),
+       strip.text = element_text(size = 15),
+       legend.title =  element_blank(),
+       legend.position = "bottom",
+       legend.box.margin=margin(-10,-10,-10,-10)) + 
+  # scale_y_continuous(limits = c(0,1.5)) + 
+  guides(colour = guide_legend(override.aes = list(size=5))) + 
+  labs(title = "Anova VolcanoPlot")
+dev.off()
+```
+
+```{r}
+
+limma_comp = limma_plot_df %>% filter(protadj == "protein adjustment") %>% select(PTM, Label, Log2FC)
+anova_comp = anova_results %>% filter(protadj == "protein adjustment") %>% select(ptm, label, log2FC)
+
+comp = merge(limma_comp, anova_comp, by.x = c("PTM", "Label"), by.y = c("ptm", "label"))
+comp$diff = comp$Log2FC - comp$log2FC
+
+comp %>% ggplot() + geom_histogram(aes(diff)) + facet_grid(~Label)
+
+limma_comp = limma_plot_df %>% filter(protadj == "no adjustment") %>% select(PTM, Label, Log2FC)
+anova_comp = anova_results %>% filter(protadj == "no adjustment") %>% select(ptm, label, log2FC)
+
+comp = merge(limma_comp, anova_comp, by.x = c("PTM", "Label"), by.y = c("ptm", "label"))
+comp$diff = comp$Log2FC - comp$log2FC
+
+comp %>% ggplot() + geom_histogram(aes(diff)) + facet_grid(~Label)
+
+```
+
+#### Summary Statistics
+
+The summary statistics for this model are skipped because the adjusted pvalue shows no positive results.
+
+#### Log Fold Change - Spike-in Peptides
+
+Lastly, we will again look at the distribution of the spike-in peptides log fold changes. The expected log Fold Change is indicated by the red `X` marks.
+
+```{r, echo = FALSE, fig.width = 10, fig.height = 5}
+unadj_spike <- ttest_final_model %>% filter(protadj == "no adjustment" & labeling == "heavy")
+unadj_spike$Adjustment <- "no adjustment"
+adj_spike <- ttest_final_model %>% filter(protadj == "protein adjustment" & labeling == "heavy")
+adj_spike$Adjustment <- "protein adjustment"
+spike <- rbindlist(list(unadj_spike, adj_spike), fill = TRUE)
+
+spike %>% ggplot() + geom_boxplot(aes(x = contrast_truth, y = estimate), fill = "lightsteelblue4") + 
+  geom_point(aes(y = trueLog2FC, x = contrast_truth), color = "red", shape = 4, size = 3) + facet_wrap(.~Adjustment) + 
+  labs(title = "T-test Fold Change of Spike-in Peptides", x = "Comparison with Expected FC", y = "log2FC") + theme(axis.text.x = element_text(angle = 30, vjust = .5))
+
+```
+
+This plots is very similar to the limma plot.
+The spike-in peptides follow the expected FC much better after adjustment.
+However the effect is much wider around the true FC than MSstatsPTM, meaning the effect is less accurate.
+
 ### Summary
@@
 
-ttest_sum <- calculate_summary_stats(ttest_final_model %>% filter(protadj == "protein adjustment"), 'adj.pvalue', "label")
+anova_results[anova_results$labeling == 'Background Peptides'][['labeling']] <- "light"
+anova_results[anova_results$labeling == 'SpikeIn Peptides'][['labeling']] <- "heavy"
+anova_summary <- calculate_summary_stats(anova_results %>% filter(protadj == "protein adjustment"), 'adj.pvalue', "label")
+anova_summary$Adjustment <- "protein adjustment"
+anova_summary$Model <- "Anova"
 
-# ttest_sum <- data.frame(metric = c('FPR', 'Recall','Specificity','Precision','Accuracy'),
-#            result = c(0,0,0,0,0))
-ttest_sum$Adjustment <- "protein adjustment"
-ttest_sum$Model <- "T-test"
-
-summary <- rbindlist(list(msstatsptm_sum, limma_sum, ttest_sum))
+summary <- rbindlist(list(msstatsptm_sum, limma_sum, anova_summary))
 summary[Model == "MSstatsPTM"]$Model <- "Proposed"
@@
 png("../supplementary/sim_new/spike_in_statistic_comparison.png", width = 750, height = 500)
-summary %>% mutate(Model = factor(Model, levels=c("Proposed", "Limma", "T-test"))) %>% 
+summary %>% mutate(Model = factor(Model, levels=c("Proposed", "Limma", "Anova"))) %>% 
   ggplot() + geom_col(aes(x = metric, y = result, fill = Model), position = "dodge") + 

### 2022-07-14 main/MSstatsPTM_Main.tex
updated limma model

@@
 In particular, the ratio-based version which models the ratios $\hat{u}_{ij}$ in~\eqref{adj_estimation} can be used to answer the null hypothesis in \eqref{null_hyp}.
-Limma uses the transformed inputs to fit a linear model with a fixed condition and run effect for each PTM.
-
-\begin{eqnarray}
-& u_{ij} = \mu + Condition_i + Run_{j} + \epsilon_{ij}, & \\ 
-& \text{where } \sum\limits_{i=1}^{I} Condition_i = 0,\ \sum\limits_{i=1}^{J} Run_j = 0, \epsilon_{ij} \mathop\sim^{iid} \mathcal{N}(0, \sigma^2) & \nonumber
+Limma uses the transformed inputs to fit a linear model with a fixed condition effect for each PTM.
+
+\begin{eqnarray}
+& u_{ij} = \mu + Condition_i + \epsilon_{ij}, & \\ 
+& \text{where } \sum\limits_{i=1}^{I} Condition_i = 0, \epsilon_{ij} \mathop\sim^{iid} \mathcal{N}(0, \sigma^2) & \nonumber
 \label{eq:limma_model}

### 2022-07-14 code/MSstatsPTM_SpikeIn_Analysis.R

@@ -403,11 +403,9 @@ Reference_Norm <- function(df, ref) {
   df_aug = df_aug %>% select(-log2inty, -adjLog2inty)
   return(df_aug)
 }
-saved_refs$PTM = saved_refs$PTM %>% rename(Run = run)
-saved_refs$PROTEIN = saved_refs$PROTEIN %>% rename(Run = run)
 
-df_site = Reference_Norm(df_site, saved_refs$PTM)
-df_prot = Reference_Norm(df_prot, saved_refs$PROTEIN)
+df_site = Reference_Norm(df_site, refs$PTM)
+df_prot = Reference_Norm(df_prot, refs$PROTEIN)
 
 ## Clean Data ------------------------------------------------------------------
 df_site <- df_site %>% filter(is_mod == TRUE)
@@ -628,11 +626,12 @@ fit_limma <- function(summarized_data, conditions, runs){
   ## Create contrast matrix
   class <- c()
   for (x in seq_len(conditions)){
-    cond <- rep(paste0("mix_", as.character(x)), runs)
+    cond <- rep(paste0("mix", as.character(x)), runs)
     class <- c(class, cond)
   }
   class <- as.factor(class)
   design <- model.matrix(~0+class)
+  colnames(design) <- c("mix1", "mix2", "mix3", "mix4")
 
   input.matrix <- as.matrix(input)
   input_adj.matrix <- as.matrix(input_adj)
@@ -641,7 +640,10 @@ fit_limma <- function(summarized_data, conditions, runs){
   fit <- lmFit(input.matrix, design=design)
   fit_adj <- lmFit(input_adj.matrix, design=design)
 
-  contrast.matrix <- design.pairs(colnames(design))
+  # contrast.matrix <- design.pairs(colnames(design))
+  contrast.matrix <- makeContrasts(mix2-mix1, mix3-mix1,
+                                   mix4-mix1, mix3-mix2,
+                                   mix4-mix2, mix4-mix3, levels=design)
   fit2 <- contrasts.fit(fit, contrast.matrix)
   fit2 <- eBayes(fit2)
 
@@ -671,22 +673,6 @@ fit_limma <- function(summarized_data, conditions, runs){
 
   return(list(limma_test = comparisons, limma_adj_test = comparisons_adj))
 }
-## Limma pairwise function
-design.pairs <- function(levels) {
-  n <- length(levels)
[…]

### 2022-07-15 code/MSstatsPTM_Simulation_Analysis_new.Rmd
replaced ttest with anova

--- 2022-06-15 code/MSstatsPTM_Simulation_Analysis.Rmd
@@
 author: "Devon Kohler"
-date: "3/7/2022"
+date: "7/14/2022"
 output:
@@
 
-To simulate data the `PTMsimulateExperiment()` function from Tsung-Heng's original MSstatsPTM was used.
+To simulate data the `PTMsimulateExperiment()` function was used.
 This function allows us to vary the number of conditions, replicates, number of proteins, number of sites per protein, number of spectral features per site/protein, mean log2-abundance of PTM and PROTEIN, deviation from the mean log2-abundance in each group, standard deviation among replicates, and standard deviation among log2-intensities.

### 2022-07-15 code/Spike-in_Analysis.Rmd
replaced ttest with anova

@@
 
-combo_plot_df
-combo_plot_df[combo_plot_df$Label == 'mix2-mix3'][['trueLog2FC']] <- combo_plot_df[combo_plot_df$Label == 'mix2-mix3'][['trueLog2FC']]*-1
-
 png("../main/images/spike_in_volcano.png", width = 750, height = 500)
 combo_plot_df %>% arrange(Model, labeling) %>% 
-  filter(abs(log2FC) < Inf & Label %in% c("mix2-mix3", "mix1-mix4", "mix3 vs mix2",
-                                          "mix4 vs mix1")) %>%
+  filter(abs(log2FC) < Inf & Label %in% c("mix3 vs mix2",
+                                          "mix4 vs mix2")) %>%
   ggplot(aes(log2FC, -log10(adj.pvalue))) + 
@@
 
-### T-Test
-
-Finally we will look at the t-test results.
-
-#### Volcano Plot
-
-Again we look at the volcano plots.
-The spike-in peptides are colored red and their expected log Fold Change is indicated in the subplot title.
-The background peptides are colored grey and should all be insignificant.
-
-```{r, echo=FALSE, fig.width = 10, fig.height = 7}
-load(file = "../data/Spike_in_ttest_model.rda")#ttest_final_model
-
-## Update some naming conventions
-ttest_final_model[ttest_final_model$label == 'mix1-mix2'][['contrast_truth']] <- "LogFC: -1 (mix2-mix1)"
-ttest_final_model[ttest_final_model$label == 'mix1-mix3'][['contrast_truth']] <- "LogFC: 1 (mix3-mix1)"
-ttest_final_model[ttest_final_model$label == 'mix1-mix4'][['contrast_truth']] <- "LogFC: 0 (mix4-mix1)"
-ttest_final_model[ttest_final_model$label == 'mix2-mix3'][['contrast_truth']] <- "LogFC: 2 (mix3-mix2)"
-ttest_final_model[ttest_final_model$label == 'mix2-mix4'][['contrast_truth']] <- "LogFC: 1 (mix4-mix2)"
-ttest_final_model[ttest_final_model$label == 'mix3-mix4'][['contrast_truth']] <- "LogFC: -1 (mix4-mix3)"
-
-ttest_final_model[ttest_final_model$labeling == 'light'][['labeling']] <- "Background Peptides"
-ttest_final_model[ttest_final_model$labeling == 'heavy'][['labeling']] <- "SpikeIn Peptides"
-
-ttest_final_model$log2FC <- ttest_final_model$estimate
-
-ttest_final_model$trueLog2FC=ttest_final_model$trueLog2FC*-1
-ttest_final_model[ttest_final_model$label == 'mix1-mix2'][['trueLog2FC']] <- ttest_final_model[ttest_final_model$label == 'mix1-mix2'][['trueLog2FC']]*-1
-ttest_final_model[ttest_final_model$label == 'mix3-mix4'][['trueLog2FC']] <- ttest_final_model[ttest_final_model$label == 'mix3-mix4'][['trueLog2FC']]*-1
-ttest_final_model[ttest_final_model$label == 'mix1-mix3'][['trueLog2FC']] <- ttest_final_model[ttest_final_model$label == 'mix1-mix3'][['trueLog2FC']]*-1
-
-ttest_final_model$contrast_truth = factor(ttest_final_model$contrast_truth, 
-                                          levels = 
-                                  c("LogFC: -1 (mix2-mix1)",
-                                    "LogFC: 1 (mix3-mix1)",
-                                    "LogFC: 0 (mix4-mix1)",
-                                    "LogFC: 2 (mix3-mix2)",
-                                    "LogFC: 1 (mix4-mix2)",
-                                    "LogFC: -1 (mix4-mix3)"))
-
-# Volcano plot
-png("../supplementary/sim_new/spike_in_ttest_volcano.png", width = 750, height = 500)
-ttest_final_model %>% arrange(labeling) %>% 
-  filter(abs(log2FC) < Inf) %>%
-  ggplot(aes(log2FC, -log10(adj.pvalue))) + 
-  geom_point(aes(color = labeling, alpha = labeling), size = 2.2) + 
-  scale_color_manual(values = c("#999999", "#D55E00")) + 
-  scale_alpha_manual(values = c(0.1, 1)) + 
-  #scale_size_manual(values = c(2, 1)) + 
-  geom_hline(yintercept = 0, linetype = "dashed") + 
-  geom_vline(xintercept = 0) + 
-  geom_vline(aes(xintercept = trueLog2FC),color = "darkred", linetype= "dashed") +
-  geom_hline(yintercept = -log10(0.05), color = "darkred") + 
-  facet_grid(protadj ~ contrast_truth) + 
-  theme_bw() + 
-  theme(axis.text.x = element_text(size = 14), 
-       axis.text.y = element_text(size = 14), 
-       legend.text=element_text(size=16),
-       axis.title.y = element_text(size = 18, vjust=2),
-       axis.title.x = element_text(size = 18),
-       title = element_text(size = 20),
-       strip.text = element_text(size = 15),
-       legend.title =  element_blank(),
-       legend.position = "bottom",
-       legend.box.margin=margin(-10,-10,-10,-10)) + 
-  scale_y_continuous(limits = c(0,1.5)) + 
-  guides(colour = guide_legend(override.aes = list(size=5))) + 
-  labs(title = "T-test VolcanoPlot")
-dev.off()
-```
-
-The results of the t-test are the weakest of the models we have observed.
-Both before and after adjustment we see no significant results.
-This may be due to an increased variance in the models, as each t-test model must be fit between two conditions and cannot share information.
-Additionally we only have two observations per condition, thus if one observation is missing the t-test model breaks down and the model cannot be fit.
-
-To get a closer look at the effect of the adjustment we will look at the volcano plots using the unadjusted pvalue.
-
-``` {r, echo=FALSE, fig.width = 10, fig.height = 7}
-# Volcano plot
-ttest_final_model %>% arrange(desc(labeling)) %>% 
-  filter(abs(estimate) < Inf & label %in% c("mix1-mix3", "mix1-mix4", "mix2-mix3", "mix2-mix4")) %>%
-  ggplot(aes(estimate, -log10(pval))) + 
-  geom_point(aes(color = labeling, alpha = labeling)) + 
-  scale_color_manual(values = c("#D55E00", "#999999")) + 
-  scale_alpha_manual(values = c(1, 0.1)) + 
-  scale_size_manual(values = c(2, 1)) + 
-  geom_hline(yintercept = 0) + 
-  geom_vline(xintercept = 0) + 
-  geom_hline(yintercept = -log10(0.05), color = "darkred") + 
-  geom_vline(xintercept = c(1, -1), color = "darkred", linetype = "dashed") + 
-  facet_grid(protadj ~ contrast_truth) + 
-  theme_bw() + 
-  theme(legend.position = "bottom") + 
-  labs(title = "T-test VolcanoPlot: Unadjusted Pvalue")
-```
-
-The unadjusted pvalue volcano plots look very similar to what we have seen in the previous two models.
-The false positives decrease and the spike-in peptides follow the expected FC closer after adjustment.
-
-#### Summary Statistics
-
-The summary statistics for this model are skipped because the adjusted pvalue shows no positive results.
-
-#### Log Fold Change - Spike-in Peptides
-
-Lastly, we will again look at the distribution of the spike-in peptides log fold changes. The expected log Fold Change is indicated by the red `X` marks.
-
-```{r, echo = FALSE, fig.width = 10, fig.height = 5}
-unadj_spike <- ttest_final_model %>% filter(protadj == "no adjustment" & labeling == "heavy")
-unadj_spike$Adjustment <- "no adjustment"
-adj_spike <- ttest_final_model %>% filter(protadj == "protein adjustment" & labeling == "heavy")
-adj_spike$Adjustment <- "protein adjustment"
-spike <- rbindlist(list(unadj_spike, adj_spike), fill = TRUE)
-
-spike %>% ggplot() + geom_boxplot(aes(x = contrast_truth, y = estimate), fill = "lightsteelblue4") + 
-  geom_point(aes(y = trueLog2FC, x = contrast_truth), color = "red", shape = 4, size = 3) + facet_wrap(.~Adjustment) + 
-  labs(title = "T-test Fold Change of Spike-in Peptides", x = "Comparison with Expected FC", y = "log2FC") + theme(axis.text.x = element_text(angle = 30, vjust = .5))
-
-```
-
-This plots is very similar to the limma plot.
-The spike-in peptides follow the expected FC much better after adjustment.
-However the effect is much wider around the true FC than MSstatsPTM, meaning the effect is less accurate.
-
 ### Anova
@@
 # Volcano plot
-png("../supplementary/sim_new/spike_in_ttest_volcano.png", width = 750, height = 500)
+png("../supplementary/sim_new/spike_in_anova_volcano.png", width = 1000, height = 500)
 anova_results %>% arrange(labeling) %>% 
@@
 dev.off()
-```
-
-```{r}
-
-limma_comp = limma_plot_df %>% filter(protadj == "protein adjustment") %>% select(PTM, Label, Log2FC)
-anova_comp = anova_results %>% filter(protadj == "protein adjustment") %>% select(ptm, label, log2FC)
-
-comp = merge(limma_comp, anova_comp, by.x = c("PTM", "Label"), by.y = c("ptm", "label"))
-comp$diff = comp$Log2FC - comp$log2FC
-
-comp %>% ggplot() + geom_histogram(aes(diff)) + facet_grid(~Label)
-
-limma_comp = limma_plot_df %>% filter(protadj == "no adjustment") %>% select(PTM, Label, Log2FC)
-anova_comp = anova_results %>% filter(protadj == "no adjustment") %>% select(ptm, label, log2FC)
-
-comp = merge(limma_comp, anova_comp, by.x = c("PTM", "Label"), by.y = c("ptm", "label"))
-comp$diff = comp$Log2FC - comp$log2FC
-
-comp %>% ggplot() + geom_histogram(aes(diff)) + facet_grid(~Label)
-
 ```
@@
 
-ttest_final_model[, log2FC:=NULL]
-unadj_spike_ttest <- ttest_final_model %>% filter(labeling == "SpikeIn Peptides" & protadj == "no adjustment")
-unadj_spike_ttest <- setNames(unadj_spike_ttest, c("ptm", "Label", "pval", "tstat", "SE", 
-                                                   "df", "log2FC", "labeling", "adj.pvalue", 
-                                                   "trueLog2FC", "contrast_truth", "protadj"))
+unadj_spike_ttest <- anova_results %>% filter(labeling == "SpikeIn Peptides" & protadj == "no adjustment")
+unadj_spike_ttest <- setNames(unadj_spike_ttest, c("Protein", "Label", "log2FC", "SE", "Tvalue", 
+                                                   "pvalue", "protadj", "adj.pvalue", "labeling", 
+                                                   "contrast_truth", "trueLog2FC"))
 
 unadj_spike_ttest$Adjustment <- "no adjustment"
-unadj_spike_ttest$Model <- "T-test"
+unadj_spike_ttest$Model <- "Anova"
 
-adj_spike_ttest <- ttest_final_model %>% filter(labeling == "SpikeIn Peptides" & protadj != "no adjustment")
-adj_spike_ttest <- setNames(adj_spike_ttest, c("ptm", "Label", "pval", "tstat", "SE", 
-                                                   "df", "log2FC", "labeling", "adj.pvalue", 
-                                                   "trueLog2FC", "contrast_truth", "protadj"))
+adj_spike_ttest <- anova_results %>% filter(labeling == "SpikeIn Peptides" & protadj != "no adjustment")
+adj_spike_ttest <- setNames(adj_spike_ttest, c("Protein", "Label", "log2FC", "SE", "Tvalue", 
+                                                   "pvalue", "protadj", "adj.pvalue", "labeling", 
+                                                   "contrast_truth", "trueLog2FC"))
 adj_spike_ttest$Adjustment <- "protein adjustment"
-adj_spike_ttest$Model <- "T-test"
+adj_spike_ttest$Model <- "Anova"
 
 unadj_spike_limma <- limma_model$PTM.Model %>% filter(labeling == "heavy")
-unadj_spike_limma <- setNames(unadj_spike_limma, c("PTM", "Label", "log2FC", "pvalue", "df", 
-                                                   "se", "labeling", "adj.pvalue", 
+unadj_spike_limma <- setNames(unadj_spike_limma, c("Protein", "Label", "log2FC", "pvalue", "DF", 
+                                                   "SE", "labeling", "adj.pvalue", 
                                                    "trueLog2FC", "contrast_truth", "protadj"))
@@
 adj_spike_limma <- limma_model$ADJUSTED.Model %>% filter(labeling == "heavy")
-adj_spike_limma <- setNames(adj_spike_limma, c("PTM", "Label", "log2FC", "pvalue", "df", 
-                                                   "se", "labeling", "adj.pvalue", 
+adj_spike_limma <- setNames(adj_spike_limma, c("Protein", "Label", "log2FC", "pvalue", "df", 
+                                                   "SE", "labeling", "adj.pvalue", 
                                                    "trueLog2FC", "contrast_truth", "protadj"))
@@
 
-spike[spike$label == 'mix1-mix2'][['contrast_truth']] <- "LogFC: -1 (mix2-mix1)"
-spike[spike$label == 'mix1-mix3'][['contrast_truth']] <- "LogFC: 1 (mix3-mix1)"
-spike[spike$label == 'mix1-mix4'][['contrast_truth']] <- "LogFC: 0 (mix4-mix1)"
-spike[spike$label == 'mix2-mix3'][['contrast_truth']] <- "LogFC: 2 (mix3-mix2)"
-spike[spike$label == 'mix2-mix4'][['contrast_truth']] <- "LogFC: 1 (mix4-mix2)"
-spike[spike$label == 'mix3-mix4'][['contrast_truth']] <- "LogFC: -1 (mix4-mix3)"
-
-spike[spike$Label == 'mix3-mix4'][["Label"]] = "mix4 vs mix3"
-spike[spike$Label == 'mix2-mix4'][["Label"]] = "mix4 vs mix2"
-spike[spike$Label == 'mix1-mix4'][["Label"]] = "mix4 vs mix1"
-spike[spike$Label == 'mix2-mix3'][["Label"]] = "mix3 vs mix2"
-spike[spike$Label == 'mix1-mix3'][["Label"]] = "mix3 vs mix1"
-spike[spike$Label == 'mix1-mix2'][["Label"]] = "mix2 vs mix1"
-
-spike[spike$Label == "mix3 vs mix2" & 
-        spike$Model == "T-test"][["trueLog2FC"]] = 2
-spike[spike$Label == "mix3 vs mix2" & 
-        spike$Model == "T-test"][["log2FC"]] = spike[
-          spike$Label == "mix3 vs mix2" & 
-        spike$Model == "T-test"][["log2FC"]] * -1
-spike[spike$Label == "mix4 vs mix2" & 
-        spike$Model == "T-test"][["trueLog2FC"]] = 1
-spike[spike$Label == "mix4 vs mix2" & 
-        spike$Model == "T-test"][["log2FC"]] = spike[
-          spike$Label == "mix4 vs mix2" & 
-        spike$Model == "T-test"][["log2FC"]] * -1
-spike[spike$Label == "mix4 vs mix3" & 
-        spike$Model == "T-test"][["trueLog2FC"]] = -1
-spike[spike$Label == "mix4 vs mix3" & 
-        spike$Model == "T-test"][["log2FC"]] = spike[
-          spike$Label == "mix4 vs mix3" & 
-        spike$Model == "T-test"][["log2FC"]] * -1
-
-spike[spike$Label == "mix3 vs mix2" & 
-        spike$Model == "Limma"][["trueLog2FC"]] = 2
-spike[spike$Label == "mix4 vs mix2" & 
-        spike$Model == "Limma"][["trueLog2FC"]] = 1
-spike[spike$Label == "mix4 vs mix3" & 
-        spike$Model == "Limma"][["trueLog2FC"]] = -1
-
 png("../supplementary/sim_new/spike_in_fc.png", width = 750, height = 500)
 spike %>% mutate(Comparison = str_split(contrast_truth, " ", simplify = TRUE)[,1],
-      Model = factor(Model, levels=c("Proposed", "Limma", "T-test"))) %>%
+      Model = factor(Model, levels=c("Proposed", "Limma", "Anova"))) %>%
   filter(!Label %in% c("mix1-mix2", "mix3-mix4")) %>% 
@@
 setDT(spike)[,list(IQR=IQR(log2FC, na.rm = TRUE)),by=list(Model, Label, protadj)] %>% 
-  filter(Label %in% c('mix2-mix3', 'mix1-mix3', 'mix1-mix4', 'mix2-mix4') & 
-           protadj == 'protein adjustment') %>% 
+  filter(protadj == 'protein adjustment') %>% 
   group_by(Model) %>% summarize(mean(IQR))
 
-(0.7557581 - 1.1954719) / 1.1954719
-(0.7557581 - 1.1277441) / 1.1277441
+spike_se = setDT(spike)[,list(SE=mean(SE, na.rm = TRUE)),by=list(Model, Label, protadj)] %>% 
+  filter(protadj == 'protein adjustment') %>% 
+  group_by(Model) %>% summarize(mean(SE))
+
+all_se = setDT(spike)[,list(SE=mean(SE, na.rm = TRUE)),by=list(Model, Label, protadj)] %>% 
+  filter(protadj == 'protein adjustment') %>% 
+  group_by(Model) %>% summarize(mean(SE))
+
+(0.5984246 - 0.9309134) / 0.9309134
 

### 2022-07-15 main/MSstatsPTM_Main.tex
replaced ttest with anova

@@
 Frequently data from these experiments are analyzed using statistical methods that were not originally designed for this task.
-Researchers use methods such as $t$-test\cite{Kalpic:2011}, Analysis of Variance\cite{girden:1992}, or Limma\cite{Ritchie_15a}, by taking as input the intensity ratios of modified and unmodified peptide features, and comparing the mean abundance of different PTM sites.
+Researchers use methods such as $t$-test\cite{Kalpic:2011}, Analysis of Variance (Anova)\cite{girden:1992}, or Limma\cite{Ritchie_15a}, by taking as input the intensity ratios of modified and unmodified peptide features, and comparing the mean abundance of different PTM sites.
 Such approaches do not fully account for all the sources of uncertainty.
@@
 
-\todo{OV: even though we say 't-test', the model that we describe is in fact an ANOVA (i.e., nothing limits us to 2 conditions. Should we change it to ANOVA directly?)}
-
-\noindent {\bf Two-sample $t$-test based on modified log$_2$-intensities}
+\noindent {\bf Anova based on modified log$_2$-intensities}
 
@@
 
-The basic two-sample $t$-test~\cite{kutner_etal_04a} is based on the model
+The basic Anova~\cite{kutner_etal_04a} is based on the model
 \begin{eqnarray}
@@
 The model allows us to estimate $\hat{\Delta}$ and its standard error.
-Based on the model, the approach tests the null hypothesis in \eqref{conv_null_hyp}, by comparing the test statistic against the Student distribution with $df=2J-2$ degrees of freedom in balanced designs.
+Based on the model, the approach tests the null hypothesis in \eqref{conv_null_hyp}, by comparing the test statistic against the Student distribution with $df=IJ-I$ degrees of freedom in balanced designs.
 Unfortunately, this approach is fundamentally flawed as it does not account for the confounding between changes in the PTM abundance and the overall changes in the abundance of the unmodified protein.
@@
 
-\noindent {\bf Two-sample $t$-test based on ratios of modified and unmodified log$_2$-intensities}
+\noindent {\bf Anova based on ratios of modified and unmodified log$_2$-intensities}
 
 \medskip \noindent 
-The basic $t$-test can be extended to account for the confounding of changes in PTM abundance and overall changes in protein abundance~\cite{Schwammle2015,THOMAS2020,Mertins:2013}.
+The basic Anova can be extended to account for the confounding of changes in PTM abundance and overall changes in protein abundance~\cite{Schwammle2015,THOMAS2020,Mertins:2013}.
 Typically this is done by first summarizing the feature intensities in each run, separately for modified and unmodified features, e.g.
@@
 \begin{eqnarray}
-u_{ij} = (\mu^{\ast}_{i}-\mu_i) + \epsilon^{\prime}_{ij}, \text{where } \epsilon^{\prime}_{ij} \mathop\sim^{iid} \mathcal{N}(0, \sigma^{\prime 2}), i \in \{1, 2\}
+u_{ij} = (\mu^{\ast}_{i}-\mu_i) + \epsilon^{\prime}_{ij}, \text{where } \epsilon^{\prime}_{ij} \mathop\sim^{iid} \mathcal{N}(0, \sigma^{\prime 2}),\ i=1,\ldots,I,\ j=1,\ldots,J
 \label{eq:ttest_model}
@@
 The model allows us to estimate $\hat{\Delta}_{adj}$ and its standard error.
-Based on this model, we can test the more relevant null hypothesis in~\eqref{null_hyp}, by comparing the test statistic against the Student distribution with \todo{$df=2J-2$} \todo{this should be the same df as the previous section?} degrees of freedom in balanced designs.
+Based on this model, we can test the more relevant null hypothesis in~\eqref{null_hyp}, by comparing the test statistic against the Student distribution with $df=IJ-I$ degrees of freedom in balanced designs.
 
@@
 Technological artifacts such as missing values further undermine the calculation of $u_{ij}$ in~\eqref{adj_estimation}.
-Finally, their is no self contained, straightforward implementation of the method, such as in the form of a coding package, requiring a more challenging manual application.
+Finally, there is no self contained, straightforward implementation of the method, such as in the form of a coding package, requiring a manual application.
 
@@
 
-\medskip \noindent The two versions of $t$-test \todo{OV: ANOVA?} above can be expanded with Empirical Bayes moderation in Limma~\cite{ Ritchie_15a, Schwammle2015, Smyth:2004, Smyth:2005, Zhu,Chappell:2021}.
+\medskip \noindent The two versions of Anova above can be expanded with Empirical Bayes moderation in Limma~\cite{ Ritchie_15a, Schwammle2015, Smyth:2004, Smyth:2005, Zhu,Chappell:2021}.
 In particular, the ratio-based version which models the ratios $\hat{u}_{ij}$ in~\eqref{adj_estimation} can be used to answer the null hypothesis in \eqref{null_hyp}.
@@
 \begin{eqnarray}
-& u_{ij} = \mu + Condition_i + \epsilon_{ij}, & \\ 
-& \text{where } \sum\limits_{i=1}^{I} Condition_i = 0, \epsilon_{ij} \mathop\sim^{iid} \mathcal{N}(0, \sigma^2) & \nonumber
+& u_{ij} = \mu + Condition_i + \epsilon_{ij}, \text{where } \sum\limits_{i=1}^{I} Condition_i = 0, \epsilon_{ij} \mathop\sim^{iid} \mathcal{N}(0, \sigma^2)
 \label{eq:limma_model}
@@
 The $MSstats$ framework has a number advantages over the methods above. 
-First, unlike \todo{t-test (or ANOVA?) and $Limma$}, the family of linear mixed effects models is applicable to many complex experimental designs, including designs with multiple sources of variation, and unbalanced designs.
+First, unlike Anova and $Limma$, the family of linear mixed effects models is applicable to many complex experimental designs, including designs with multiple sources of variation, and unbalanced designs.
 Second, the approach is applicable to various data acquisition types, including label-free DDA and DIA, and experiments with TMT labeling, and is compatible with various data processing tools such as Skyline, Spectronaut, MaxQuant, Progenesis, Proteome Discoverer, and OpenMS.
@@
 The evaluations were based on the $MSstatsPTM$, and were performed without imputing missing values or using Empirical Bayes moderation.
-$MSstatsPTM$ was compared with \todo{$t$-test?} and Limma.
+$MSstatsPTM$ was compared with Anova and Limma.
 We did not compare to IsobarPTM because it is only applicable to TMT experiments, meaning it could not be applied to our ground truth data.
@@
 After the adjustment, the FDR of all methods improved.
-In a clean simulation, recall and accuracy of $MSstatsPTM$ and $Limma$ performed similarly.
+In a clean simulation, recall and accuracy of $MSstatsPTM$ and Limma performed similarly.
 As expected when not using Empiricial Bayes moderation in MSstatsPTM, Limma performed slightly better in simulations with fewer biological replicates.
-In contrast, the performance of the two-sample $t$-test method lagged behind.
-This was due to the fact that the $t$-test only uses data within the groups of interest while ignoring the remainder of the data.
-In contrast, $MSstatsPTM$ and Limma leverage information across comparisons, resulting in improved power in simulations with more comparisons.
-Additionally, since this simulation is a group comparison design that doesn't require many variance components, it is favorable to Limma and t-test
+In contrast, the performance of Anova lagged behind, especially in cases with higher simulated variance.
+This was due to Anova's more simplistic variance estimation.
+In contrast, $MSstatsPTM$ and Limma implement methods to provide better estimates of the variance, resulting in improved power.
+Finally, since this simulation is a group comparison design that doesn't require many variance components, it is favorable to Limma and Anova.
+As discussed previously, these methods are not directly applicable to experiments with additional variance components.
 
@@
 
-In simulations with missing values and few features, $MSstatsPTM$ outperformed $t$-test and $Limma$.
+In simulations with missing values and few features, $MSstatsPTM$ outperformed Anova and Limma.
 Changes in unmodified protein level still needed to be accounted for to control the FDR.
 Once controlling for changes in the unmodified protein, the proposed method outperformed the other methods, as seen in~\figref{sim2_acc}.
-The proposed method calibrated model accuracy and recall rate well, even when the number of replicates were low.
-The proposed approach can handle low feature counts due to using Tukey's Median Polish for summarization.
+The proposed method calibrated model accuracy  well, even when the number of replicates were low.
+$MSstatsPTM$ can handle low feature counts better than the existing methods due to using Tukey's Median Polish for summarization.
 The existing methods summation approach requires more observations to correctly summarize the intensities in the run, whereas TMP is robust to few measurements.
@@
 Additionally, when comparing the fold change estimation across all modified peptides, the proposed method showed a tighter distribution of estimated fold changes around the true fold change.
-Specifically, the inner quartile range (IQR) of the estimated fold change for the proposed method was on average 21.8\% smaller than $Limma$ IQR and 10.4\% smaller than $t$-test's IQR.
+Specifically, the inner quartile range (IQR) of the estimated fold change for the proposed method was on average 32.5\% smaller than Limma and Anova's IQR.
 This means that while the mean of the estimated fold changes was generally correct for all methods, the proposed approach correctly estimated the fold change more often across all PTMs.
@@
 In this experiment all models incorrectly estimated the fold change of the modified spike-in peptides before adjusting for changes in unmodified protein abundance.
-After adjustment, the spike-in peptides' fold change was generally in line with expectation in all methods, however the distribution of estimated fold changes was visibly wider (\figref{spikein_boxplot}).
+After adjustment, the spike-in peptides' fold change was generally in line with expectation in all methods, however the existing method's distribution of estimated fold changes was visibly wider (\figref{spikein_boxplot}).
 Of the three approaches, the proposed method showed the tightest distribution around the true log fold change.
-Comparing the IQR of the spike-in peptide's log fold change, the proposed method's IQR was 36.78\% smaller than $Limma$ and 32.98\% smaller than $t$-test's.
-While Limma and $t$-test generally estimated the correct fold change, the proposed method's estimation was more consistent across all modifications.
+Comparing the IQR of the spike-in peptide's log fold change, the proposed method's IQR was 32.86\% smaller than Limma and Anova's IQR.
+While Limma and Anova overall estimated the correct fold change, they were inconsistent in their estimation, resulting in a higher error rate.
+The proposed method's estimation was more consistent, with a smaller error.
+As discussed in the previous section, the summarization methods of the proposed methods excel when the number of observations are low.
+This was the case here, with each PTM only averaging $1.37$ features.
+The existing approaches used a non-robust summarization method, that was very sensitive to few observations, causing their fold change estimation to be more inconsistent, with a larger error compared to the proposed method.
 
@@
 Specifically, for the proposed method, the number of false positives went from 20.88\% to 1.84\% after adjustment was applied.
-While the proposed method and $Limma$ both correctly estimated the fold change of the spike-in peptides, using $Limma$ resulted in many large adjusted p-values.
-Using $Limma$ would have resulted in missing the majority of differential PTMs.
+While the proposed method and Limma both correctly estimated the fold change of the spike-in peptides, using Limma resulted in many large adjusted p-values.
+This was due to Limma estimating a 35.7\% larger variance compared to the proposed method when estimating the spike-in peptides.
 This experiment introduced more variance components, which is challenging for the existing methods and therefore the difference in the results is more pronounced than in the simulations.
@@
 When the test was applied to Dataset 4, only one PTM became non-differentially abundant due to an increase in variance.
-When applied to Dataset 5, 548 PTMs became non-differentially abundant due to an increase in variance (only 3.4\% of all PTMs that became became non-differentially abundant).
+When applied to Dataset 5, 548 PTMs became non-differentially abundant due to an increase in variance (only 3.4\% of all PTMs that became non-differentially abundant).
 Thus we can conclude that the drop off in differentially abundant PTMs was due to changes in global protein abundance, not simply increased model variance.
@@
 There are many non-differentially abundant PTMs, and having to comb through each one individually to determine false negatives is generally an unfeasible task.
-The proposed approach removes the confounding on an experimental level, revealing interesting PTMs that are entirely masked due to changes in the unmodified protein.
+The proposed approach removes the confounding on an experimental level, revealing modifications of interest that are entirely masked due to changes in the unmodified protein.
 
@@
 Here only 25 PTMs became insignificant due to increased variance.
-When there was no global profiling run the methods were less effective at removing confounding with the unmodified protein.
-This was entirely due to not being able to perform adjustment without a corresponding global protein.
+Without a global profiling run, the method was less effective at removing confounding with the unmodified protein.
+This was due to not being able to perform adjustment without a corresponding global protein.
 In this case adjustment caused us to toss out a large number of the PTMs in the original model.
@@
 The framework is general and is applicable to a variety of experimental designs.
-It outperforms the ad-hoc methods underlying $t$-test and Limma, and yields accurate results in the broad type of experimental circumstances, including the presence of missing values, changes in protein abundance, few representative peptides, and different acquisition methods.
+It outperforms the ad-hoc methods underlying Anova and Limma, and yields accurate results in the broad type of experimental circumstances, including the presence of missing values, changes in protein abundance, few representative peptides, and different acquisition methods.
 The framework allows us to plan for subsequent experiments, and choose the appropriate number of replicates in consideration of adjustment with respect to protein abundance.
@@
 b) The advantage of using the proposed approach was apparent when including limited observations and missing values.
-Looking at accuracy, the proposed method outperformed Limma and $t$-test in nearly every model.
+Looking at accuracy, the proposed method outperformed Anova and Limma in nearly every model.
 }
@@
 a) Before adjustment the fold change of the spike-in peptides' were systematically different from the expected fold change in all models.
-After adjustment, this systemic difference was removed, however the inner quartile range of the Limma and $t$-test models was wider than the proposed method.
+After adjustment, this systemic difference was removed, however the inner quartile range of the Limma and Anova models was wider than the proposed method.
 b) Before adjustment the spike-in peptides (colored red) did not follow the expected log fold change; after adjustment, the spike-in peptides were more in line with expectation.

### 2022-07-15 supplementary/ptm_sm.tex
replaced ttest with anova

@@
 
-The results are summarized from Figure \ref{fig:sim1_fpr} to Figure \ref{fig:sim1_recall}.
-These results include the false discovery rate with and without adjusting for changes in unmodified protein abundance, false positive rate, and overall accuracy.
-Additionally, the results are compared to simulation 2 described in the next section.
+The results are summarized in Figure \ref{fig:sim1_recall}.
 
@@
   \centering
-	\includegraphics[width=.8\textwidth]{sim_new/sim1_FDR_all_models}\\
-	\caption{All the considered methods in simulation 1 correctly calibrated FDR  when adjusting for changes in protein abundance.
-In comparison, the methods without accounting for the protein-level changes resulted in off-target, high false discovery rates.
-The performance of the models without adjustment was much lower than those with adjustment, thus only models with adjustment are compared going forward.}
- \end{subfigure}
- \begin{subfigure}{\textwidth}
-  \centering
-	\includegraphics[width=.8\textwidth]{sim_new/sim1_FDR}
-	\caption{The considered methods with protein adjustment were compared in detail. All three methods with adjustment generally performed similarly in terms of FDR.}
- \end{subfigure}
- \caption{FDR of Simulation 1.}
-\label{fig:sim1_fpr}
-\end{figure}
-
-\begin{figure}[h!]
-\centering
- \begin{subfigure}{\textwidth}
-  \centering
 	\includegraphics[width=.8\textwidth]{sim_new/sim1_Recall}\\
@@
 \centering
-\includegraphics[width=.85\textwidth]{sim_new/spike_in_limma_volcano}
+\includegraphics[width=1\textwidth]{sim_new/spike_in_limma_volcano}
 \caption{When modeling the experiment with the Limma method, the spike-in peptides again follow the expected log fold change better after adjusting for changes in protein level.
@@
 \centering
-\includegraphics[width=.85\textwidth]{sim_new/spike_in_ttest_volcano}
+\includegraphics[width=1\textwidth]{sim_new/spike_in_anova_volcano}
 \caption{Using the two sample $t$-test, none of the comparisons either before or after adjustment show any significant peptides.

### 2022-07-15 code/MSstatsPTM_SpikeIn_Analysis.R

@@ -927,8 +927,8 @@ fit_anova = function(data, contrast){
   ptms = unique(data$PTM)
   test_results = data.table()
   unadj_test_results = data.table()
-  for (i in seq_along(ptms)){
-
+  for (i in seq_along(ptms)){#
+    print(ptms[[i]])
     ptm_data_adj = data %>% filter(PTM == ptms[[i]] & is.finite(Adj_Abundance))
     ptm_data_unadj = data %>% filter(PTM == ptms[[i]] & is.finite(Abundance.x))
 
@@ -937,13 +937,13 @@ fit_anova = function(data, contrast){
       missing = setdiff(c("mix1", "mix2", "mix3", "mix4"),
                         unique(ptm_data_adj$Condition))
       keep = c(TRUE,TRUE,TRUE,TRUE,TRUE,TRUE)
-      for (i in seq_along(missing)){
-        check = !str_detect(rownames(contrast), missing[[i]])
+      for (k in seq_along(missing)){
+        check = !str_detect(rownames(contrast), missing[[k]])
         keep = ifelse(!check, check, keep)
         temp_contrast = contrast[keep,, drop = FALSE]
       }
-      print(temp_contrast)
-      print(model_conditions)
+      # print(temp_contrast)
+      # print(model_conditions)
 
       temp_contrast = temp_contrast[, model_conditions]
 

### 2022-07-15 code/Simulate_data_old.R

@@ -1,0 +1,462 @@
+
+## Load packages
+library(MSstatsPTM)
+library(data.table)
+library(tidyverse)
+library(limma)
+
+## Source simulation code
+source("PTMsimulateExperiment.R")
+
+## Define Modeling Functions ---------------------------------------------------
+## Format simulation data into MSstatsPTM input
+convert_to_msstats_format <- function(df){
+
+  ## modified data, edit column names and add missing
+  df$PTM$protein <- paste(df$PTM$protein, df$PTM$site, sep = "_")
+  df$PTM$PrecursorCharge <- NA
+  df$PTM$FragmentIon <- NA
+  df$PTM$ProductCharge <- NA
+  df$PTM$IsotopeLabelType <- "L"
+  df$PTM$BioReplicate <- paste0(df$PTM$run, "rep")
+  df$PTM$log2inty <- 2**df$PTM$log2inty
+  df$PTM$PeptideSequence <- paste(df$PTM$protein, df$PTM$feature, sep = "_")
+
+  setnames(df$PTM, c("protein", "group", "run", "log2inty"),
+           c("ProteinName", "Condition", "Run", "Intensity"))
+
+  ## unmodified data, edit column names and add missing
+  df$PROTEIN$PrecursorCharge <- NA
+  df$PROTEIN$FragmentIon <- NA
+  df$PROTEIN$ProductCharge <- NA
+  df$PROTEIN$IsotopeLabelType <- "L"
+  df$PROTEIN$log2inty <- 2**df$PROTEIN$log2inty
+  df$PROTEIN$BioReplicate <- paste0(df$PROTEIN$run, "rep")
+  df$PROTEIN$PeptideSequence <- paste(df$PROTEIN$protein, df$PROTEIN$feature,
+                                      sep = "_")
+  setnames(df$PROTEIN, c("protein", "group", "run", "log2inty"),
+           c("ProteinName", "Condition", "Run", "Intensity"))
+
+  return(df)
+}
+
+## Run ttest model using run level summarized data
+run_ttest <- function(summarized_data){
+
+  ptms <- unique(summarized_data$PTM)
+
+  ttest_temp <- data.table()
+  ttest_adj_temp <- data.table()
+  ## Run ttest for each ptm
+  for (p in seq_along(ptms)){
+    temp_joined <- summarized_data %>% filter(PTM == ptms[[p]])
+    groups <- (temp_joined %>% distinct(Condition))[[1]]
+    t <- 2
+
+    ## Loop over groups
[…]

### 2022-07-15 code/Simulated_Analysis.R

@@ -36,49 +36,72 @@ convert_to_msstats_format <- function(df){
   return(df)
 }
 
-## Run ttest using run level summared data
-run_ttest <- function(summarized_data){
+## Run anova using run level summared data
+fit_anova = function(data, contrast){
 
-  ptms <- unique(summarized_data$PTM)
+  ptms = unique(data$PTM)
+  test_results = data.table()
+  unadj_test_results = data.table()
+  for (i in seq_along(ptms)){#
+    # print(ptms[[i]])
+    ptm_data_adj = data %>% filter(PTM == ptms[[i]] & is.finite(Adj_Abundance))
+    ptm_data_unadj = data %>% filter(PTM == ptms[[i]] & is.finite(Abundance.x))
 
-  ttest_temp <- data.table()
-  ttest_adj_temp <- data.table()
-  ## Run ttest for each ptm
-  for (p in seq_along(ptms)){
-    temp_joined <- summarized_data %>% filter(PTM == ptms[[p]])
-    groups <- (temp_joined %>% distinct(Condition))[[1]]
-    t <- 2
-
-    ## Loop over groups
-    for (g in 1:(length(groups)-1)){
-      for (g2 in 2:length(groups)){
-        tryCatch({
-        ttest_ptm <- t.test((temp_joined %>% filter(Condition == groups[g]) %>% select(Abundance.x))[[1]],
-                                      (temp_joined %>% filter(Condition == groups[g2]) %>% select(Abundance.x))[[1]])
-        ttest_temp <- rbindlist(list(ttest_temp, data.table(ptm = ptms[[p]],
-                                                            label = paste(groups[g], "vs", groups[g2], sep = " "),
-                                                            pval = ttest_ptm$p.value,
-                                                            tstat = ttest_ptm$statistic[[1]],
-                                                            SE = ttest_ptm$stderr,
-                                                            df = ttest_ptm$parameter[[1]],
-                                                            estimate = ttest_ptm$estimate[[2]] - ttest_ptm$estimate[[1]])))},
-                 error=function(e){cat("ERROR :", as.character(p), "\n")})
-        tryCatch({ttest_adj_ptm <- t.test((temp_joined %>% filter(Condition == groups[g]) %>% select(Adj_Abundance))[[1]],
-                                          (temp_joined %>% filter(Condition == groups[g2]) %>% select(Adj_Abundance))[[1]])
[…]

### 2022-07-15 main/MSstatsPTM_Main.tex
minor change to df calc

@@
 The model allows us to estimate $\hat{\Delta}$ and its standard error.
-Based on the model, the approach tests the null hypothesis in \eqref{conv_null_hyp}, by comparing the test statistic against the Student distribution with $df=IJ-I$ degrees of freedom in balanced designs.
+Based on the model, the approach tests the null hypothesis in \eqref{conv_null_hyp}, by comparing the test statistic against the Student distribution with $df=IJK-I$ \todo{should we replace IJK with N (number obs)?} degrees of freedom in balanced designs.
 Unfortunately, this approach is fundamentally flawed as it does not account for the confounding between changes in the PTM abundance and the overall changes in the abundance of the unmodified protein.

### 2022-07-15 supplementary/ptm_sm.tex
supp text changes

@@
 
-The proposed statistical approach was evaluated and compared to the $t$-test and Limma methods using computer simulations.
+The proposed statistical approach was evaluated and compared to the Anova and Limma methods using computer simulations.
 Specifically, their statistical properties with respect to protein-level adjustment and real world experimental conditions were evaluated.
@@
 Approaches without considering protein-level adjustment lose track of an important aspect in interpreting observed changes in PTM abundance, which may result in misleading conclusions.
-To highlight the necessity of the adjustment, we compared six different approaches as follows: a) proposed approach, b) proposed approach without adjusting for unmodified peptides c) $t$-test (with adjustment), d) $t$-test (no adjustment), e) Limma (with adjustment), and f) Limma (no adjustment).
+To highlight the necessity of the adjustment, we compared six different approaches as follows: a) proposed approach, b) proposed approach without adjusting for unmodified peptides c) Anova (with adjustment), d) Anova (no adjustment), e) Limma (with adjustment), and f) Limma (no adjustment).
 
 In experiments of complex designs, multiple inter-related conditions are often compared together.
-The proposed approach and Limma leverage measurements in all conditions for the inference of the underlying abundance, whereas $t$-test uses measurements from the two conditions being compared.
-To highlight this distinction, multiple conditions of data were generated.
-While multiple conditions were generated, all comparisons were still made with the same fold change between conditions ($0.75$).
+All comparisons were made with the same fold change between conditions ($1$).
 This ensured that all peptides were differentially abundant by the same amount and the only difference in calculations was using more conditions in the modeling.
@@
 \item Standard deviations of log-intensities for modified and unmodified peptides: $0.2$, $0.3$
-\item Difference in PTM abundance between conditions: $0$, $0.75$, $1.5$, $2.25$
-\item Difference in protein abundance between conditions: $0$, $0.75$, $1.5$, $2.25$
+\item Difference in PTM abundance between conditions: $0$, $1.$, $2.$, $3.$
+\item Difference in protein abundance between conditions: $0$, $1.$, $2.$, $3.$
 \item Number of replicates: $2$, $3$, $5$, $10$
@@
 At higher replicates the performance of the proposed methods and Limma were comparable.
-$t$-test clearly performed worse across all methods.}
+Anova clearly performed worse across all methods.}
  \end{subfigure}
@@
 Even at lower replicates the proposed method still outperformed Limma.
-Again the lowest performing method was $t$-test.}
+Again the lowest performing method was Anova.}
 \label{fig:sim2_recall}
@@
 	\includegraphics[width=.8\textwidth]{sim_new/sim1_FC_boxplot}
-	\caption{In simulation 1 all considered methods correctly estimated the fold change between conditions, with a median fold change estimation of .75.
+	\caption{In simulation 1 all considered methods correctly estimated the fold change between conditions, with a median fold change estimation of 1.
 The distributions around the median were also consistent across all methods.}
@@
 	\includegraphics[width=.8\textwidth]{sim_new/sim3_FC_boxplot}
-	\caption{In simulation 2 all methods correctly estimated the fold change with a median log change of .75.
+	\caption{In simulation 2 all methods correctly estimated the fold change with a median log change of 1.
 The proposed method in this simulation had a visibly tighter distribution around the median.
-Both Limma and $t$-test showed a wider range around the fold change.
-The inner quartile range of the Proposed method was on average 10.4\% smaller than t-test and 21.8\% smaller than Limma.}
+Both Limma and Anova showed a wider range around the fold change.}
  \end{subfigure}
@@
 
-Again we consider three different methods and assess their performance: the method proposed in this paper, Limma, and two sample $t$-test.
+Again we consider three different methods and assess their performance: the method proposed in this paper, Limma, and Anova.
 All methods are analyzed after adjusting for changes in overall protein level.
@@
 \includegraphics[width=1\textwidth]{sim_new/spike_in_anova_volcano}
-\caption{Using the two sample $t$-test, none of the comparisons either before or after adjustment show any significant peptides.
-With that being said, the fold change of the spike-in peptides was much closer to expectation after adjusting for global protein abundance.
-\label{fig:spike_volcano_ttest}}
+\caption{Using Anova the fold change of the spike-in peptides was much closer to expectation after adjusting for global protein abundance.
+The log FC estimation was the same as Limma, however the pvalues were different.
+We actually see a few more True Positives using Anova compared to Limma.\label{fig:spike_volcano_ttest}}
 \end{figure}
@@
 \caption{Comparing the summary statistics between methods, the proposed method performed the strongest.
-In terms of accuracy and specificity the three methods were close, with Limma and $t$-test showing slightly higher values.
+In terms of accuracy and specificity the three methods were close, with Limma and Anova showing slightly higher values.
 Accuracy and specificity were dominated by the large number of true negatives (background peptides) compared to the true positives (spike-in peptides).

### 2022-07-15 code/Shigella_Analysis.R
fixed plot titles

@@ -12,13 +12,13 @@ load(file = "shigella_global_TMTFormat.rda")#shigella_global
 load(file = "shigella_pY_TMTFormat.rda")#shigella_peptide_pY
 
 ## Setup for summarization
-shigella_global <- shigella_global %>% group_by(ProteinName, PeptideSequence, 
+shigella_global <- shigella_global %>% group_by(ProteinName, PeptideSequence,
   Charge, PSM, Mixture, TechRepMixture, Run, Channel, Condition, BioReplicate) %>%
   summarize(Intensity = max(Intensity))
 
-pY_input <- list(PTM = data.frame(shigella_peptide_pY), 
+pY_input <- list(PTM = data.frame(shigella_peptide_pY),
                  PROTEIN = data.frame(shigella_global))
-pST_input <- list(PTM = data.frame(shigella_peptide_pST), 
+pST_input <- list(PTM = data.frame(shigella_peptide_pST),
                   PROTEIN = data.frame(shigella_global))
 
 ## Data summarization ----------------------------------------------------------
@@ -42,7 +42,7 @@ global_feature_data <- pST_summarization$PROTEIN$FeatureLevelData
 ## Create pYST dataset
 ## Create pY/pST Dataset
 ## check the overlapped sites between pST and pY
-shared_sites <- intersect(unique(py_summarized_data$Protein), 
+shared_sites <- intersect(unique(py_summarized_data$Protein),
                           unique(pst_summarized_data$Protein))
 shared_ST <- shared_sites[grepl("_S", shared_sites)|grepl("_T", shared_sites)]
 shared_Y <- shared_sites[grepl("_Y", shared_sites)]
@@ -170,9 +170,9 @@ test[test$FeatureType == 'Model'][['FeatureType']] <- "Protein Summarized"
 test[test$FeatureType == 'Peptide'][['FeatureType']] <- "Protein Feature"
 test$Mixture = paste0("Mixture ", test$Mixture)
 p2 <- test %>% ggplot() +
-  geom_line(aes(x = BioReplicate, y = log2Intensity, group = PSM, 
+  geom_line(aes(x = BioReplicate, y = log2Intensity, group = PSM,
                 color = FeatureType), size = 2) +
-  geom_point(aes(x = BioReplicate, y = log2Intensity, group = PSM, 
[…]

### 2022-07-15 main/MSstatsPTM_Main.tex
fixed plot titles
minor formula typos

--- 2022-07-15 main/MSstatsPTM_Main.tex
@@
 The model allows us to estimate $\hat{\Delta}$ and its standard error.
-Based on the model, the approach tests the null hypothesis in \eqref{conv_null_hyp}, by comparing the test statistic against the Student distribution with $df=IJ-I$ degrees of freedom in balanced designs.
+Based on the model, the approach tests the null hypothesis in \eqref{conv_null_hyp}, by comparing the test statistic against the Student distribution with $df=I(JK-1)$ degrees of freedom in balanced designs.
 Unfortunately, this approach is fundamentally flawed as it does not account for the confounding between changes in the PTM abundance and the overall changes in the abundance of the unmodified protein.
@@
 The model allows us to estimate $\hat{\Delta}_{adj}$ and its standard error.
-Based on this model, we can test the more relevant null hypothesis in~\eqref{null_hyp}, by comparing the test statistic against the Student distribution with $df=IJ-I$ degrees of freedom in balanced designs.
+Based on this model, we can test the more relevant null hypothesis in~\eqref{null_hyp}, by comparing the test statistic against the Student distribution with $df=I(J-1)$ degrees of freedom in balanced designs.
 
@@
 \begin{eqnarray}
-& \widehat{SE}(\hat{\Delta})_{adj} = \sqrt{\widehat{SE}(\hat{\Delta}_{PTM})^2 + \widehat{SE}(\hat{\Delta}_{protein})^2} = &\\ 
+& \widehat{SE}(\hat{\Delta}_{adj}) = \sqrt{\widehat{SE}(\hat{\Delta}_{PTM})^2 + \widehat{SE}(\hat{\Delta}_{protein})^2} = &\\ 
 & \sqrt{\frac{1}{J}\hat{\sigma}^2_{PTM} + \frac{1}{J}\hat{\sigma}^2_{protein}} = \sqrt{\frac{2}{J}(\hat{\sigma}^2_{PTM} + \hat{\sigma}^2_{protein})} = \sqrt{\hat{\sigma}^2_{PTM} + \hat{\sigma}^2_{protein}} & \nonumber
@@
 \centering
-\begin{subfigure}[c]{0.5\linewidth}
+\begin{subfigure}[c]{0.55\linewidth}
 \includegraphics[width=1\textwidth]{images/ipah_venn_diagramm.png}
@@
 \end{subfigure}
-\begin{subfigure}[c]{0.5\linewidth}
+\begin{subfigure}[c]{0.55\linewidth}
 \includegraphics[width=1\textwidth]{images/shig_venn_diagramm.png}
@@
 \caption{
-\todo{Update the figure titles}
-%
 a) Dataset 4: Human - Ubiquitination - 1mix-TMT.
@@
  \centering
-	\includegraphics[height=.525\textwidth]{images/usp30_venn_diagramm}
+	\includegraphics[height=.55\textwidth]{images/usp30_venn_diagramm}
 	\caption{}
@@
  \centering
-	\includegraphics[height=.525\textwidth]{images/usp30_venn_diagramm_matching_only}
+	\includegraphics[height=.55\textwidth]{images/usp30_venn_diagramm_matching_only}
 	\caption{}

### 2022-07-19 main/MSstatsPTM_Main.tex
Cosmetics

@@
 For example, in the simple case of \figref{data-structure} with $J=2$ conditions, where $\hat{\sigma}^2_{PTM}$ and $\hat{\sigma}^2_{protein}$ are the estimates of the error variance for the PTM and protein model described in \eqref{msstats_model}, the standard error would be calculated as
-
-\begin{eqnarray}
-& \widehat{SE}(\hat{\Delta}_{adj}) = \sqrt{\widehat{SE}(\hat{\Delta}_{PTM})^2 + \widehat{SE}(\hat{\Delta}_{protein})^2} = &\\ 
-& \sqrt{\frac{1}{J}\hat{\sigma}^2_{PTM} + \frac{1}{J}\hat{\sigma}^2_{protein}} = \sqrt{\frac{2}{J}(\hat{\sigma}^2_{PTM} + \hat{\sigma}^2_{protein})} = \sqrt{\hat{\sigma}^2_{PTM} + \hat{\sigma}^2_{protein}} & \nonumber
+\begin{eqnarray}
+ \widehat{SE}(\hat{\Delta}_{adj}) &=& \sqrt{\widehat{SE}(\hat{\Delta}_{PTM})^2 + \widehat{SE}(\hat{\Delta}_{protein})^2} \\ 
+&=& \sqrt{\frac{1}{J}\hat{\sigma}^2_{PTM} + \frac{1}{J}\hat{\sigma}^2_{protein}} = \sqrt{\frac{2}{J}(\hat{\sigma}^2_{PTM} + \hat{\sigma}^2_{protein})} = \sqrt{\hat{\sigma}^2_{PTM} + \hat{\sigma}^2_{protein}} \nonumber
 \label{eq:se_calc}

### 2022-07-21 main/MSstatsPTM_Main.tex
Existing methods
Have not yet finished

@@
 Consider a label-free LC-MS/MS experiment in the special case of a balanced design with $I$ conditions and $J$ biological replicates per condition.
-For simplicity, we assume that the experiment has no technical replicates, such that each biological replicate corresponds to an LC-MS/MS run.
-\figref{data-structure} schematically illustrates this data structure for one PTM site, in a special case of $I=2$ conditions and $J=2$ biological replicates per condition.
+For simplicity, we assume that the experiment has no technical replicates, such that each biological replicate is represented by a single LC-MS/MS run.
+\figref{data-structure} schematically illustrates this data structure for one PTM site, in a special case of a label-free experiment with $I=2$ conditions and $J=2$ biological replicates per condition.
 For one protein, the PTM site is represented by $K$ spectral features (i.e., peptide ions, distinguished by their cleavage residues and charge states). 
@@
 
-The population parameter of interest is the difference between the abundances of a PTM site in Condition $i$ and Condition $i^{\prime}$, denoted by $\mu_i^{\ast}$ and $\mu_{i^{\prime}}^{\ast}$ respectively.
+The population quantity of interest is the difference between the log$_2$ abundances of a PTM site in Condition $i$ and Condition $i^{\prime}$, denoted by $\mu_i^{\ast}$ and $\mu_{i^{\prime}}^{\ast}$ respectively.
 We are interested in testing the null hypothesis
@@
 \end{eqnarray}
-Unfortunately, the population parameter is inherently confounded with the overall changes in protein abundance. To account for this, it is advantageous to consider a different null hypothesis:
+Unfortunately, this population quantity is inherently confounded with the overall changes in protein abundance. To account for this, it is advantageous to consider a different null hypothesis:
 \begin{eqnarray}
@@
 \end{eqnarray}
-where $\mu_i$ and $\mu_{i^{\prime}}$ reflect the overall protein abundances in Condition $i$ and Condition $i^{\prime}$.
+where $\mu_i$ and $\mu_{i^{\prime}}$ reflect the overall log$_2$ protein abundances in Condition $i$ and Condition $i^{\prime}$.
 These quantities are estimated using protein features without with and without the modification site.
@@
 
-\noindent {\bf Anova based on modified log$_2$-intensities}
+\noindent {\bf ANOVA on modified log$_2$-intensities}
 
@@
 
-The basic Anova~\cite{kutner_etal_04a} is based on the model
-\begin{eqnarray}
-y^*_{ijk} = \hat{\mu}^{*}_{i} + \epsilon^{\ast}_{ij}, \ \epsilon^{\ast}_{ij} \mathop\sim^{iid} \mathcal{N}(0, \sigma^{*2}),\ i=1,\ldots,I,\ j=1,\ldots,J,\ k=1,\ldots,K
+The basic Analysis of Variance (ANOVA) model~\cite{kutner_etal_04a}  is
+\begin{eqnarray}
+y^*_{ijk} = \mu^{*}_{i} + \epsilon^{\ast}_{ij}, \ \epsilon^{\ast}_{ij} \mathop\sim^{iid} \mathcal{N}(0, \sigma^{*2}),\ i=1,\ldots,I,\ j=1,\ldots,J,\ k=1,\ldots,K
 \label{eq:ttest_model}
@@
 Based on the model, the approach tests the null hypothesis in \eqref{conv_null_hyp}, by comparing the test statistic against the Student distribution with $df=I(JK-1)$ degrees of freedom in balanced designs.
+\todo{I do not think this is right.
+I believe there should be a summarization first, such that the data are $\hat{y}_{ij}$, the model becomes $\hat{y}^*_{ij} = \mu^{*}_{i} + \epsilon^{\ast}_{ij}, \ \epsilon^{\ast}_{ij} \mathop\sim^{iid} \mathcal{N}(0, \sigma^{*2})$ and the df are $df=I(J-1)$.
+Could you introduce the summarization first?}
 Unfortunately, this approach is fundamentally flawed as it does not account for the confounding between changes in the PTM abundance and the overall changes in the abundance of the unmodified protein.
@@
 \medskip \noindent 
-The basic Anova can be extended to account for the confounding of changes in PTM abundance and overall changes in protein abundance~\cite{Schwammle2015,THOMAS2020,Mertins:2013}.
-Typically this is done by first summarizing the feature intensities in each run, separately for modified and unmodified features, e.g.
+The basic ANOVA can be extended to account for the confounding of changes in PTM abundance and overall changes in protein abundance~\cite{Schwammle2015,THOMAS2020,Mertins:2013}.
+Typically this is done by first summarizing the feature intensities in each run, \todo{This is a good place to remind of the summary notation $\hat{y}_{ij}$ that drops subscript $k$} separately for modified and unmodified features, e.g.
 with the sum on the original scale, dividing the summary of the PTM abundance by the summary of the protein abundance and then taking a log.
@@
 Technological artifacts such as missing values further undermine the calculation of $u_{ij}$ in~\eqref{adj_estimation}.
-Finally, there is no self contained, straightforward implementation of the method, such as in the form of a coding package, requiring a manual application.
+Finally, there is no self contained, straightforward implementation of the method, such as in the form of a coding package, and therefore the approach requires a manual implementation.
 
@@
 
-\medskip \noindent The two versions of Anova above can be expanded with Empirical Bayes moderation in Limma~\cite{ Ritchie_15a, Schwammle2015, Smyth:2004, Smyth:2005, Zhu,Chappell:2021}.
-In particular, the ratio-based version which models the ratios $\hat{u}_{ij}$ in~\eqref{adj_estimation} can be used to answer the null hypothesis in \eqref{null_hyp}.
+\medskip \noindent The two versions of ANOVA above can be expanded with Empirical Bayes moderation in Limma~\cite{ Ritchie_15a, Schwammle2015, Smyth:2004, Smyth:2005, Zhu,Chappell:2021}.
+In particular, the ratio-based version which models the ratios \todo{do you need hat on top of u?} $\hat{u}_{ij}$ in~\eqref{adj_estimation} can be used to answer the null hypothesis in \eqref{null_hyp}.
 Limma uses the transformed inputs to fit a linear model with a fixed condition effect for each PTM.
@@
 \includegraphics[scale=.45]{images/fig3.png}
-\caption{Schematic representation of one PTM site, in a special case of $I=2$ conditions and $J=2$ biological replicates per condition.
-We are interested in estimating the difference between the population-level PTM abundance between Condition 1 and Condition 2 (i.e, $\mu^*_1-\mu^*_2$), relative to the population-level difference of the overall protein abundance (i.e., $\mu_1-\mu_2$).
+\caption{Schematic representation of one PTM site, in a special case of a label-free experiment with $I=2$ conditions and $J=2$ biological replicates per condition.
+After a log$_2$ transform, we are interested in estimating the difference between the population-level PTM abundance between Condition 1 and Condition 2 (i.e, $\mu^*_1-\mu^*_2$), relative to the population-level difference of the overall protein abundance (i.e., $\mu_1-\mu_2$).
 These quantities are characterized by the observed spectral Features (boxes), i.e.

### 2022-07-21 code/Spike-in_Vizualization.Rmd
minor changes

--- 2022-07-15 code/Spike-in_Analysis.Rmd
@@
 ## Spike-in Dataset Method Comparison
-
-This markdown file examines the custom made spike-in analysis targeting PTMs.
-This allows us to know which PTMs are actually differential and which are not.
-The models used in this analysis are: MSstatsPTM, limma, and two condition t-test.
-For MSstatsPTM the full workflow is used for run-level summarization and modeling.
-Limma and t-test use feature averaging for run summarization and then their coresponding models are fit.
-The model results are shown first and then a comparison is made.
 
@@
 
-First a comparison between the adjusted and unadjusted models are shown.
-The spike-in peptides are colored red and their expected log Fold Change is indicated in the subplot title.
-The background peptides are colored grey and should all be insignificant.
+```{r pressure, echo=FALSE, fig.width = 10, fig.height = 7}
+load(file = "MSstatsPTM_Summarized.rda")
 
-```{r pressure, echo=FALSE, fig.width = 10, fig.height = 7}
-load(file = "D:\\Northeastern\\Research\\MSstats\\PTM_Refractor\\Label_Free_PTM_Data\\benchmark-spike\\MSstatsPTM_Summarized.rda")
-
-load(file = "../data/Spike_in_MSstatsPTM_Model.rda")#model_df
+load(file = "MSstatsPTM_Model.rda")#model_df
 
@@
 # Volcano plot
-png("../supplementary/sim_new/spike_in_msstatsptm_volcano.png", width = 1000, height = 500)
 plot_df %>% arrange(labeling) %>% 
@@
   labs(title = "Proposed Method VolcanoPlot")
-dev.off()
 ```
-
-We can clearly see the effect of the adjustment in the volcano plots.
-Before adjustment, the red spike-in peptides did not follow the expected fold change.
-Additionally there are a large amount of false positives, shown as grey dots above the significance threshold.
-After applying protein level adjustment the red spike-in peptides are much more in line with expectation and there are many less false positives.
 
@@
 
-In the summary statistics we can clearly see the advantage of the protein adjustment.
-Overall accuracy increases to near 100% (although we need to keep in mind it is dominated by the large number of true negatives).
-The False Positive Rate decreases substantially.
-Precision increases but is still quiet low.
-This is mainly due to the very large number of true negatives.
-While we were able to correctly categorize a large portion of the true negatives, there are still a large number that were missed and dominate this statistic.
-Recall only saw a slight increase in the adjusted model.
-This is because, while the fold change does now follow what we expected, a large number of the spike-in peptides still had an adjusted p value over .05.
-This is most likely due to increases in the variance when combined with the protein model.
-Finally specificity also saw a large increase when adding in the protein adjustment, indicating we are correctly identifying most of the background peptides as negatives.
-
 #### Log Fold Change - Spike-in Peptides
-
-Lastly, the plot below looks at the distribution spike-in peptides log fold changes. The expected log Fold Change is indicated by the red `X` marks.
 
@@
 ```
-
-We can see in the plot above that the fold change of the spike-in peptides falls in line with expectation much better after adjustment.
-The expected change falls very close to the median value, and is at least within the 1st and 3rd quartile boxes in every comparison.
-In the unadjusted plot, the expected falls outside the quartile boxes in the majority of comparisons.
 
@@
 
-First a comparison between the adjusted and unadjusted models are shown.
-Again the spike-in peptides are colored red and their expected log Fold Change is indicated in the subplot title.
-The background peptides are colored grey and should all be insignificant.
-
 ```{r, echo=FALSE, fig.width = 10, fig.height = 7}
-load(file = "../data/Spike_in_Limma_Model.rda")#limma_model
+load(file = "Limma_Model.rda")#limma_model
 
@@
 # Volcano plot
-png("../supplementary/sim_new/spike_in_limma_volcano.png", width = 1000, height = 500)
 plot_df %>% arrange(labeling) %>% 
@@
   labs(title = "Limma VolcanoPlot")
-dev.off()
 ```
-
-The limma model shows an improvement in terms of the false positives, but loses insight into the majority of spike-in peptides.
-After adjustment, the spike-in peptides are much more inline with the expected log2FC, however they are mostly below the significant pvalue threshold.
-This is most likely due to an increase in standard error or lower degrees of freedom compared to MSstatsPTM.
 
@@
 
-The summary statistics repeat what we saw in the volcano plots.
-The accuracy and specificity are both much higher after adjustment due to the decrease in false positives.
-Of course this is also seen in the lower FPR.
-Recall after adjustment is much lower.
-We are correctly labeling very few of the true positives after adjustment.
-This is in contrast to MSstatsPTM which actually saw an increase in recall after adjustment.
-
 #### Log Fold Change - Spike-in Peptides
-
-Again the last thing we will look at is the distribution spike-in peptides log fold changes. The expected log Fold Change is indicated by the red `X` marks.
 
@@
 
-Once again we see a strong improvement in the Log2FC of spike-in peptides after adjustment.
-In contrast to MSstatsPTM the distribution after adjustment is much wider, with a median close to the true FC.
-This larger variance around the median indicates that while we are generally seeing the true FC, the results are not as strong as MSstatsPTM.
-
 ### Anova
@@
 ```{r, echo=FALSE, fig.width = 10, fig.height = 7}
-load(file = "../data/anova_results.rda")#anova_results
+load(file = "Anova_Model.rda")#anova_results
 
@@
 # Volcano plot
-png("../supplementary/sim_new/spike_in_anova_volcano.png", width = 1000, height = 500)
 anova_results %>% arrange(labeling) %>% 
@@
   labs(title = "Anova VolcanoPlot")
-dev.off()
 ```
 
-#### Summary Statistics
-
-The summary statistics for this model are skipped because the adjusted pvalue shows no positive results.
-
 #### Log Fold Change - Spike-in Peptides
-
-Lastly, we will again look at the distribution of the spike-in peptides log fold changes. The expected log Fold Change is indicated by the red `X` marks.
 
@@
 
-This plots is very similar to the limma plot.
-The spike-in peptides follow the expected FC much better after adjustment.
-However the effect is much wider around the true FC than MSstatsPTM, meaning the effect is less accurate.
-
 ### Summary
 
-From the analysis above it is clear MSstatsPTM is the most robust modeling method of the three tested.
-In the computer simulations we saw MSstatsPTM and Limma having similar results, however in the real world experiment Limma performanced significantly worse.
-This could be in part due to the robust TMP summary methods used by MSstatsPTM compared the log sum of abundance used for Limma.
-An extension to this analysis could look into the performance of Limma using the same summarization methods as used in MSstatsPTM.
-In either case it is clear that the full workflow of MSstatsPTM performs the best of the observed models.
-
-In the following plot we compare the summary statsitics of the adjusted models. Note T-test is shown but all values are zero.
 ```{r, echo = FALSE, fig.width = 8}
@@
 
-png("../supplementary/sim_new/spike_in_statistic_comparison.png", width = 750, height = 500)
 summary %>% mutate(Model = factor(Model, levels=c("Proposed", "Limma", "Anova"))) %>% 
@@
        legend.position = "right")
-dev.off()
 ```
-
-This plot confirms that MSstatsPTM and limma are comparable in terms of FDR, however MSstatsPTM performs much better when identifying the true positives, spike-in, peptides.
-
-In the following plot we compare the log2FC of the spike-in peptides across models.
 
@@
   labs(title = "Fold Change of Spike-in Peptides", x = "Comparison with Expected FC")
-dev.off()
-
-spike
-setDT(spike)[,list(IQR=IQR(log2FC, na.rm = TRUE)),by=list(Model, Label, protadj)] %>% 
-  filter(protadj == 'protein adjustment') %>% 
-  group_by(Model) %>% summarize(mean(IQR))
-
-spike_se = setDT(spike)[,list(SE=mean(SE, na.rm = TRUE)),by=list(Model, Label, protadj)] %>% 
-  filter(protadj == 'protein adjustment') %>% 
-  group_by(Model) %>% summarize(mean(SE))
-
-all_se = setDT(spike)[,list(SE=mean(SE, na.rm = TRUE)),by=list(Model, Label, protadj)] %>% 
-  filter(protadj == 'protein adjustment') %>% 
-  group_by(Model) %>% summarize(mean(SE))
-
-(0.5984246 - 0.9309134) / 0.9309134
 
 ```
-
-In this plot we can see that the unadjusted models are very similar and do not follow the expected FC.
-In comparison the adjusted models all generally follow the expectation, however MSstatsPTM has a tighter distribution around the expected FC.
-In contrast the other two models have a wider range around the true FC.

### 2022-07-21 code/IpaH_analysis.R

@@ -78,7 +78,7 @@ gsmd_62_sum[gsmd_62_sum$FeatureType == 'Peptide'][['FeatureType']] <- "PTM Featu
 gsmd_sum[gsmd_sum$FeatureType == 'Model'][['FeatureType']] <- "Protein Summarized"
 gsmd_sum[gsmd_sum$FeatureType == 'Peptide'][['FeatureType']] <- "Protein Feature"
 
-p1 <- gsmd_62_sum %>% ggplot() + geom_line(aes(x = BioReplicate, y = Abundance, 
+p1 <- gsmd_62_sum %>% ggplot() + geom_line(aes(x = BioReplicate, y = Abundance,
                                                color = FeatureType,
                                                group = PSM, size = FeatureType)) +
   geom_point(aes(x = BioReplicate, y = Abundance, color = FeatureType), size = 5) +
@@ -86,7 +86,7 @@ p1 <- gsmd_62_sum %>% ggplot() + geom_line(aes(x = BioReplicate, y = Abundance,
              aes(xintercept=as.numeric(x)), linetype = "dashed") +
   scale_colour_manual(values = c("#C3C3C3", "#D55E00")) +
   scale_size_manual(values = c(1.5,1)) +
-  labs(title = "GSDMD with ubiquitination at site K62", x = "BioReplicate", 
+  labs(title = "GSDMD with ubiquitination at site K62", x = "BioReplicate",
        y = "Abundance") +
   theme_bw() +
   theme(axis.text.x = element_text(angle = 60, hjust=1, size = 16),
@@ -107,7 +107,7 @@ p1 <- gsmd_62_sum %>% ggplot() + geom_line(aes(x = BioReplicate, y = Abundance,
   annotate("text", x = 8.5, y = 22.75, label = "Dox4hr", size = 8) +
   annotate("text", x = 10.5, y = 22., label = "Dox6hr", size = 8)
 
-p2 <- gsmd_sum %>% ggplot() + geom_line(aes(x = BioReplicate, y = Abundance, 
+p2 <- gsmd_sum %>% ggplot() + geom_line(aes(x = BioReplicate, y = Abundance,
                                             color = FeatureType,
                                          group = PSM, size = FeatureType)) +
   geom_point(aes(x = BioReplicate, y = Abundance, color = FeatureType), size = 5) +
[…]

### 2022-07-21 code/MSstatsPTM_SpikeIn_Analysis.R

@@ -467,11 +467,6 @@ aqua_summarized <- aqua_summarized %>%
 
 ## Summarize and model ---------------------------------------------------------
 ## Combine into one list
-aqua_df <- list(
-  PTM = aqua_summarized,
-  PROTEIN = NULL
-)
-
 df_site <- df_site %>% select(-feature, -Fraction)
 
 input_df <- list(
@@ -485,6 +480,7 @@ summary_df <- dataSummarizationPTM(input_df,
                                    normalization.PTM = FALSE,
                                    MBimpute = FALSE, MBimpute.PTM = FALSE)
 
+## Add Aqua peptides into summarized data
 aqua_summarized$Protein  = aqua_summarized$ProteinName
 aqua_summarized$originalRUN = aqua_summarized$Run
 aqua_summarized$GROUP = aqua_summarized$Condition
@@ -508,11 +504,6 @@ summary_df$PTM$ProteinLevelData = rbindlist(
 save(summary_df,
      file = "D:\\Northeastern\\Research\\MSstats\\PTM_Refractor\\Label_Free_PTM_Data\\benchmark-spike\\MSstatsPTM_Summarized.rda")
 
-# summary_df <- temp_summary_df
-# summary_df$PTM$ProteinLevelData <- summary_df$PTM$ProteinLevelData %>%
-#   mutate(LogIntensities = ifelse(str_detect(originalRUN, "mix3|mix4"),
-#                                  LogIntensities -1, LogIntensities))
-
 # Contrast
 comparison = matrix(c(-1,1,0,0,
                        -1,0,1,0,
@@ -529,31 +520,7 @@ model_df <- groupComparisonPTM(summary_df, data.type = "LabelFree",
                                contrast.matrix = comparison,
                                use_log_file=FALSE)
 
-save(model_df,
-     file = "D:\\Northeastern\\Research\\MSstats\\MSstatsPTM-manuscript\\data\\Spike_in_MSstatsPTM_Model.rda")
-
-# model_df$PTM.Model <- model_df$PTM.Model %>% mutate(
-#   Label = ifelse(Label == "mix2 vs mix1", "mix1-mix2",
-#                ifelse(Label == "mix3 vs mix1", "mix1-mix3",
-#                       ifelse(Label == "mix4 vs mix1", "mix1-mix4",
-#                              ifelse(Label == "mix2 vs mix3", "mix2-mix3",
-#                                     ifelse(Label == "mix2 vs mix4", "mix2-mix4",
[…]

### 2022-07-21 main/MSstatsPTM_Main.tex
More on existing methods

@@
 I believe there should be a summarization first, such that the data are $\hat{y}_{ij}$, the model becomes $\hat{y}^*_{ij} = \mu^{*}_{i} + \epsilon^{\ast}_{ij}, \ \epsilon^{\ast}_{ij} \mathop\sim^{iid} \mathcal{N}(0, \sigma^{*2})$ and the df are $df=I(J-1)$.
-Could you introduce the summarization first?}
+The index k should drop.
+Could you introduce the summarization first?
+What do you think re keeping or removing the hat on $\hat{y}_{ij}$?}
 Unfortunately, this approach is fundamentally flawed as it does not account for the confounding between changes in the PTM abundance and the overall changes in the abundance of the unmodified protein.
@@
 
-\noindent {\bf Anova based on ratios of modified and unmodified log$_2$-intensities}
+\noindent {\bf ANOVA based on ratios of modified and unmodified log$_2$-intensities}
 
@@
 The basic ANOVA can be extended to account for the confounding of changes in PTM abundance and overall changes in protein abundance~\cite{Schwammle2015,THOMAS2020,Mertins:2013}.
-Typically this is done by first summarizing the feature intensities in each run, \todo{This is a good place to remind of the summary notation $\hat{y}_{ij}$ that drops subscript $k$} separately for modified and unmodified features, e.g.
+Typically this is done by first summarizing the feature intensities in each run, separately for modified and unmodified features, e.g.
 with the sum on the original scale, dividing the summary of the PTM abundance by the summary of the protein abundance and then taking a log.
@@
 \end{eqnarray}
-The approach then models these values as
+The approach then models these values as \todo{if we keep a hat on $\hat{y}_{ij}$, we should probably keep a hat on $\hat{u}_{ij}$}
 \begin{eqnarray}
@@
 Although effective, the approach is somewhat simplistic.
-It is not applicable to experimental designs with more complex sources of biological and technological variation, such as experiments with repeated measurements, experiments with multiple batches or experiments with multi-run multiplexing.
+It is not applicable to experimental designs with more complex sources of biological and technological variation, such as experiments with repeated measurements, experiments with multiple batches or experiments with TMT labeling.
 Since \eqref{adj_estimation} performs the adjustment on the replicate level, the experiment must contain a matching number of replicates in both the modified and unmodified runs.
@@
 \medskip \noindent The two versions of ANOVA above can be expanded with Empirical Bayes moderation in Limma~\cite{ Ritchie_15a, Schwammle2015, Smyth:2004, Smyth:2005, Zhu,Chappell:2021}.
-In particular, the ratio-based version which models the ratios \todo{do you need hat on top of u?} $\hat{u}_{ij}$ in~\eqref{adj_estimation} can be used to answer the null hypothesis in \eqref{null_hyp}.
-Limma uses the transformed inputs to fit a linear model with a fixed condition effect for each PTM.
-
-\begin{eqnarray}
-& u_{ij} = \mu + Condition_i + \epsilon_{ij}, \text{where } \sum\limits_{i=1}^{I} Condition_i = 0, \epsilon_{ij} \mathop\sim^{iid} \mathcal{N}(0, \sigma^2)
-\label{eq:limma_model}
-\end{eqnarray}
-
-The method than uses Empirical Bayes moderation to combine the variance estimations across all PTMs modeled.
-This increases the degrees of freedom with which the individual PTM variances are estimated~\cite{Ritchie_15a}.
-The goal of Limma is to increase the sensitivity of detection of differential abundance when the experiment has few biological replicates.
+A typical application of Limma takes as input $u_{ij}$, and for each PTM fits the linear model in~\eqref{ttest_model}.
+The method differs from the ratio-based ANOVA in specifying additional prior distributions for the model parameters.
+The prior distributions are estimated from the data across all the proteins and all the PTM as described in \todo{XXX}. 
+With this approach, testing the null hypothesis in~\eqref{null_hyp} amounts to combining the PTM-specific and the consensus variance. \todo{I am wondering if an equation would help}
+In  experiments with few biological replicates, the resulting standard errors are often smaller, and the degrees of freedom are often larger than in absence of the moderation~\cite{Ritchie_15a}. 
+Thus the approach tends to increase the sensitivity of detection of differential abundance.
 
 Since Limma only improves upon the methods above in estimation of variation, it has the same limitations in terms of the experimental designs.
-The method is only directly applicable to experiments with at most two variance components, meaning that in more complex experiments it cannot account for all sources of variation.
-Additionally, it still requires the same experimental design for the modified and unmodified features so that the ratio calculation can be performed.
+The method is only directly applicable to experiments with at most two variance components, and cannot account for all the sources of variation in more complex experiments.
+Additionally, it requires the same experimental design for the modified and unmodified features so that the ratio calculation can be performed.
 Finally, there is no self contained implementation of the methods to PTMs, requiring manual transformation and application by the user.
@@
 \end{eqnarray}
+\todo{Is it true that we have a separate variance parameter for each run $ \sigma^2_j$? I don't think so. I think it should be a separate sigma, but same for all the runs}
 Parameters of the model are estimated using restricted maximum likelihood (REML)~\cite{Kenward}.
@@
 \end{eqnarray}
-
+\todo{There is no $\mu$ in Eq 6.
+Could you explain how $\mu$ relates to $Condition$?
+Alternatively, we can re-write the ANOVA models so that we use zero sum parametrization rather than the cell means parametriation}
 Similarly to $Limma$, $MSstatsTMT$ includes an optional Empirical Bayes moderation of the standard error~\cite{Huang:2020}, which increases the sensitivity of the model when the number of biological replicates in each condition is small.

### 2022-07-23 main/MSstatsPTM_Main.tex
Fine-tune existing methods

@@
 For simplicity, we assume that the experiment has no technical replicates, such that each biological replicate is represented by a single LC-MS/MS run.
-\figref{data-structure} schematically illustrates this data structure for one PTM site, in a special case of a label-free experiment with $I=2$ conditions and $J=2$ biological replicates per condition.
+\figref{data-structure} schematically illustrates this data structure for one protein and one PTM site, $I=2$ and $J=2$.
 For one protein, the PTM site is represented by $K$ spectral features (i.e., peptide ions, distinguished by their cleavage residues and charge states). 
-The number of modified and unmodified features typically varies across proteins. Some log-intensities may be outliers, and some spectral features can be missing. 
+The number of modified and unmodified features typically varies across proteins. Some log-intensities can be outliers, and some spectral features can be missing. 
 The log$_2$-intensity of Feature $k$, in Replicate $j$ of Condition $i$ is denoted by $y_{ijk}^{\ast}$.
 Conversely, the unmodified protein is represented by $L$ spectral features, and the log-intensity of Feature $l$ from the unmodified protein in the same run is denoted by $y_{ijl}$.
-The features can be quantified as part of a same mass spectrometry run, or in separate enrichment and global proteome profiling runs.
+The features can be quantified as part of a same mass spectrometry run, or in a separate enrichment and global proteome profiling run.
 
@@
 where $\mu_i$ and $\mu_{i^{\prime}}$ reflect the overall log$_2$ protein abundances in Condition $i$ and Condition $i^{\prime}$.
-These quantities are estimated using protein features without with and without the modification site.
+These quantities are estimated using protein features with and without the modification site.
 
@@
 
+\todo{I do not think this is right.
+I believe there should be a summarization first, such that the data are $\hat{y}_{ij}$, the model becomes $\hat{y}^*_{ij} = \mu^{*}_{i} + \epsilon^{\ast}_{ij}, \ \epsilon^{\ast}_{ij} \mathop\sim^{iid} \mathcal{N}(0, \sigma^{*2})$ and the df are $df=I(J-1)$.
+The index k should drop.
+Could you start by introducing the summarization?
+What do you think re keeping or removing the hat on $\hat{y}_{ij}$?}
 The basic Analysis of Variance (ANOVA) model~\cite{kutner_etal_04a}  is
@@
 \end{eqnarray}
-The model allows us to estimate $\hat{\Delta}$ and its standard error.
-Based on the model, the approach tests the null hypothesis in \eqref{conv_null_hyp}, by comparing the test statistic against the Student distribution with $df=I(JK-1)$ degrees of freedom in balanced designs.
-\todo{I do not think this is right.
-I believe there should be a summarization first, such that the data are $\hat{y}_{ij}$, the model becomes $\hat{y}^*_{ij} = \mu^{*}_{i} + \epsilon^{\ast}_{ij}, \ \epsilon^{\ast}_{ij} \mathop\sim^{iid} \mathcal{N}(0, \sigma^{*2})$ and the df are $df=I(J-1)$.
-The index k should drop.
-Could you introduce the summarization first?
-What do you think re keeping or removing the hat on $\hat{y}_{ij}$?}
+The model allows us to estimate $\hat{\Delta}$ and its standard error, and test the null hypothesis in \eqref{conv_null_hyp} by comparing the model-based test statistic against the Student distribution with $df=I(JK-1)$ \todo{$df=I(J-1)$} degrees of freedom in balanced designs.
 Unfortunately, this approach is fundamentally flawed as it does not account for the confounding between changes in the PTM abundance and the overall changes in the abundance of the unmodified protein.
@@
 Typically this is done by first summarizing the feature intensities in each run, separately for modified and unmodified features, e.g.
-with the sum on the original scale, dividing the summary of the PTM abundance by the summary of the protein abundance and then taking a log.
-Specifically, for condition $i$ and biological replicate $j$, the approach calculates
+with the sum on the original scale, and considering sample-wise ratios of the summaries.
+On the log$_2$ scale, for condition $i$ and biological replicate $j$ the approach is equivalent to calculating
 \begin{eqnarray}
-u_{ij} = \log \left( \sum_{k=1}^{K} 2^{y_{ijk}^{\ast}} \right) - \log \left( \sum_{l=1}^{L} 2^{y_{ijl}} \right)
+u_{ij} = \log_2 \frac{ \sum_{k=1}^{K} 2^{y_{ijk}^{\ast}} }{ \sum_{l=1}^{L} 2^{y_{ijl}} } = \log_2 \left( \sum_{k=1}^{K} 2^{y_{ijk}^{\ast}} \right) - \log_2 \left( \sum_{l=1}^{L} 2^{y_{ijl}} \right)
 \label{eq:adj_estimation}
 \end{eqnarray}
-The approach then models these values as \todo{if we keep a hat on $\hat{y}_{ij}$, we should probably keep a hat on $\hat{u}_{ij}$}
+The approach then models these values with the basic ANOVA, which corresponds to \todo{if we keep a hat on $\hat{y}_{ij}$, we should probably keep a hat on $\hat{u}_{ij}$}
 \begin{eqnarray}
@@
 
-\medskip \noindent The two versions of ANOVA above can be expanded with Empirical Bayes moderation in Limma~\cite{ Ritchie_15a, Schwammle2015, Smyth:2004, Smyth:2005, Zhu,Chappell:2021}.
+\medskip \noindent The estimation of nuisance variation of the ratio-based ANOVA model above is often further expanded with Empirical Bayes moderation implemented in Limma~\cite{ Ritchie_15a, Schwammle2015, Smyth:2004, Smyth:2005, Zhu,Chappell:2021}.
 A typical application of Limma takes as input $u_{ij}$, and for each PTM fits the linear model in~\eqref{ttest_model}.
-The method differs from the ratio-based ANOVA in specifying additional prior distributions for the model parameters.
+The model differs from~\eqref{ttest_model} in that it specifies additional prior distributions for the model parameters.
 The prior distributions are estimated from the data across all the proteins and all the PTM as described in \todo{XXX}. 
-With this approach, testing the null hypothesis in~\eqref{null_hyp} amounts to combining the PTM-specific and the consensus variance. \todo{I am wondering if an equation would help}
-In  experiments with few biological replicates, the resulting standard errors are often smaller, and the degrees of freedom are often larger than in absence of the moderation~\cite{Ritchie_15a}. 
-Thus the approach tends to increase the sensitivity of detection of differential abundance.
-
-Since Limma only improves upon the methods above in estimation of variation, it has the same limitations in terms of the experimental designs.
-The method is only directly applicable to experiments with at most two variance components, and cannot account for all the sources of variation in more complex experiments.
-Additionally, it requires the same experimental design for the modified and unmodified features so that the ratio calculation can be performed.
-Finally, there is no self contained implementation of the methods to PTMs, requiring manual transformation and application by the user.
-
-\medskip \noindent {\bf Isobar-PTM} \medskip \noindent Isobar-PTM was also proposed for experiments with LC-MS/MS quantitative strategies that employ isobaric labels such as TMT or isobaric tag for relative and absolute quantification (iTRAQ)\cite{Breitwieser:2013}.
+With this approach, testing the null hypothesis in~\eqref{null_hyp} amounts to combining the PTM- and protein-specific estimates of variation with a consensus estimate obtained from all the PTM and all the proteins.
+\todo{It's a bit vague.
+I am wondering if an equation would help}
+As the result, in experiments with few biological replicates the standard errors are often smaller, and the degrees of freedom are often larger than without moderation~\cite{Ritchie_15a}. 
+Thus the approach tends to increase the sensitivity of detecting differential abundance.
+
+Since Limma only improves upon the estimation of variation, it has the same limitations in terms of the experimental designs.
+Furthermore, the method is only directly applicable to experiments with at most two variance components, and cannot account for all the sources of variation in more complex experiments.
+There is no self contained implementation of the methods to PTMs, requiring manual transformation and application by the user.
+
+\medskip \noindent {\bf Isobar-PTM}
+
+\medskip \noindent Isobar-PTM was also proposed for experiments with LC-MS/MS quantitative strategies that employ isobaric labels such as TMT, or isobaric tag for relative and absolute quantification (iTRAQ)\cite{Breitwieser:2013}.
 Isobar-PTM expresses MS measurements with a linear model and performs adjustment with respect to protein abundance using the difference between log-ratio of modified peptides in two channels and log-ratio of protein level.

### 2022-07-23 main/MSstatsPTM_Main.tex
More existing methods

@@
 For example, to model the unmodified protein in \figref{data-structure} $MSstats$ fits the following model
+\todo{I rewrote the model so that it uses $\mu_i$ directly rather than conditions}
 \begin{eqnarray}
-& y_{ij} = \mu + Condition_i + Run_{j} + \epsilon_{ij}, & \\ 
-& \text{where } \sum\limits_{i=1}^{I} Condition_i = 0,\ Run \mathop\sim^{iid} \mathcal{N}(0, \sigma^2_j), \epsilon_{ij} \mathop\sim^{iid} \mathcal{N}(0, \sigma^2) & \nonumber
+& y_{ij} = \mu_i + Run_{j} + \epsilon_{ij}, \  \text{where } Run \stackrel{iid}{\sim} \mathcal{N}(0, \sigma^2_j), \epsilon_{ij} \stackrel{iid}{\sim} \mathcal{N}(0, \sigma^2) & 
 \label{eq:msstats_model}
 \end{eqnarray}
-\todo{Is it true that we have a separate variance parameter for each run $ \sigma^2_j$? I don't think so. I think it should be a separate sigma, but same for all the runs}
+\todo{Is it true that we have a separate variance parameter for each run $ \sigma^2_j$? I don't think so. I think it should be a separate sigma (or maybe $\tau^2$), but same for all the runs}
 Parameters of the model are estimated using restricted maximum likelihood (REML)~\cite{Kenward}.
 In turn, the parameters allow us to estimate the relevant quantity $\hat{\Delta}_{Protein}$ and its standard error. Where $\hat{\Delta}_{Protein}$ is estimated as
-
 \begin{eqnarray}
@@
 \end{eqnarray}
-\todo{There is no $\mu$ in Eq 6.
-Could you explain how $\mu$ relates to $Condition$?
-Alternatively, we can re-write the ANOVA models so that we use zero sum parametrization rather than the cell means parametriation}
 Similarly to $Limma$, $MSstatsTMT$ includes an optional Empirical Bayes moderation of the standard error~\cite{Huang:2020}, which increases the sensitivity of the model when the number of biological replicates in each condition is small.

### 2022-07-23 main/MSstatsPTM_Main.tex
MSstats

@@
 Isobar-PTM expresses MS measurements with a linear model and performs adjustment with respect to protein abundance using the difference between log-ratio of modified peptides in two channels and log-ratio of protein level.
-Unfortunately, this statistical modeling framework is not applicable to either label-free workflows or experiments with complex designs.
+\todo{This seems somewhat similar to the ANOVA on ratios.
+Could you contrast with the ANOVA?
+Or just provide an equation in the same notation?} Unfortunately, this statistical modeling framework is not applicable to either label-free workflows or experiments with complex designs.
 
@@
 
-$MSstats$~\cite{Choi:2014} and $MSstatsTMT$~\cite{Huang:2020} are a series of R/Bioconductor packages designed for protein significance analysis and statistical relative quantification of proteins and peptides in global, targeted and data-independent proteomics.
-Models in these packages takes as input a list of log$_2$-intensities $y_{ijk}$.
-For each protein, the feature log$_2$-intensities are first summarized into a single value per protein per run using Tukey's median polish~\cite{Tukey:1977}.
-These summarized values are then used as input to fit a flexible family of linear mixed-effects models \cite{McLean:1991, Faraway:2006, Bolker2009}.
-The specific model instances depend on the experimental design and data acquisition.
-For example, to model the unmodified protein in \figref{data-structure} $MSstats$ fits the following model
+$MSstats$~\cite{Choi:2014} and $MSstatsTMT$~\cite{Huang:2020} are a family of R/Bioconductor packages for statistical relative quantification of proteins and peptides in global, targeted and data-independent proteomics.
+The packages take as input log$_2$-intensities $y_{ijk}$.
+For each protein, the log$_2$-intensities are first summarized into a single value per protein per run using Tukey's median polish~\cite{Tukey:1977}.
+The summaries are then used as input to fit a flexible family of linear mixed-effects models \cite{McLean:1991, Faraway:2006, Bolker2009}.
+The specific model fit for the individual protein depends on the design of the experiment, labeling type and data acquisition type, as well as on the pattern of missing values in that particular protein.
+For example, the unmodified protein features in \figref{data-structure} are modeled as
 \todo{I rewrote the model so that it uses $\mu_i$ directly rather than conditions}
+\todo{Should we use a hat on top of y, to indicate that we model the summaries?}
 \begin{eqnarray}
@@
 \end{eqnarray}
-\todo{Is it true that we have a separate variance parameter for each run $ \sigma^2_j$? I don't think so. I think it should be a separate sigma (or maybe $\tau^2$), but same for all the runs}
+\todo{Is it true that we add run in label-free experiments?
+I thought it was only with TMT.
+But maybe I forgot.
+Could you check?
+If we don't include run, then give a model for label-free, and then contrast with TMT}
+\todo{Is it true that we have a separate variance parameter for each run $ \sigma^2_j$? I don't think so. I think it should be a separate sigma (maybe denote it $\tau^2$), but same for all the runs}
 Parameters of the model are estimated using restricted maximum likelihood (REML)~\cite{Kenward}.
-In turn, the parameters allow us to estimate the relevant quantity $\hat{\Delta}_{Protein}$ and its standard error. Where $\hat{\Delta}_{Protein}$ is estimated as
-\begin{eqnarray}
-\hat{\Delta}_{Protein} = \hat{\mu}_{i} - \hat{\mu}_{i^{\prime}}
-\label{eq:protein_delta}
-\end{eqnarray}
-Similarly to $Limma$, $MSstatsTMT$ includes an optional Empirical Bayes moderation of the standard error~\cite{Huang:2020}, which increases the sensitivity of the model when the number of biological replicates in each condition is small.
-Based on the model, the approach tests the null hypothesis in \eqref{conv_null_hyp}, by comparing the test statistic against the Student distribution.
+In turn, the parameters allow us to estimate the pairwise comparison $\hat{\Delta}_{Protein} = \hat{\mu}_{i} - \hat{\mu}_{i^{\prime}}$ and its standard error.
+Similarly to $Limma$, $MSstatsTMT$ includes an optional Empirical Bayes moderation of the standard error~\cite{Huang:2020}, which increases the sensitivity of detecting differential abundance when the number of biological replicates in each condition is small.
+
+In addition to detecting differential abundance at the protein level, $MSstats$ and $MSstatsTMT$ can be also used at the feature or at the modification cite levels.
+For example, summarizing the features per PTM cite instead of per protein, the approach allows us to test the null hypothesis in \eqref{conv_null_hyp}.
 
 The $MSstats$ framework has a number advantages over the methods above. 
-First, unlike Anova and $Limma$, the family of linear mixed effects models is applicable to many complex experimental designs, including designs with multiple sources of variation, and unbalanced designs.
+First, unlike ANOVA and $Limma$, $MSstats$ and $MSstatsTMT$ are applicable to many complex experimental designs, including designs with multiple sources of variation, and unbalanced designs. 
 Second, the approach is applicable to various data acquisition types, including label-free DDA and DIA, and experiments with TMT labeling, and is compatible with various data processing tools such as Skyline, Spectronaut, MaxQuant, Progenesis, Proteome Discoverer, and OpenMS.
 Third, the run-level summarization has been shown to be more robust to outliers and missing values as compared to feature averages~\cite{Choi:2014}. 
-Finally, the custom $MSstats$ and $MSstatsTMT$ implementation accounts for potential data artifacts, and is numerically scalable and stable, and is available through both command line and a dedicated graphical user interface.
-
-Unfortunately, MSstats estimates the difference in overall protein abundance and tests the null hypothesis in~\eqref{conv_null_hyp}.
+Finally, the custom $MSstats$ and $MSstatsTMT$ implementation accounts for potential data artifacts, is numerically scalable and stable, and is available through both command line and a dedicated graphical user interface.
+
+Unfortunately, the $MSstats$ framework focuses on overall protein abundance, and as the result tests the null hypothesis in~\eqref{conv_null_hyp}.
 It does not account for the confounding between the changes in PTM abundance and the overall changes in protein abundance.
-In other words, it cannot directly test the null hypothesis in~\eqref{null_hyp}.
+This manuscript extends the methodology in $MSstats$ and $MSstatsTMT$, to enable testing of the relevant the null hypothesis in~\eqref{null_hyp}.
 

### 2022-07-23 main/MSstatsPTM_Main.tex
Proposed approach

@@
 It does not account for the confounding between the changes in PTM abundance and the overall changes in protein abundance.
-This manuscript extends the methodology in $MSstats$ and $MSstatsTMT$, to enable testing of the relevant the null hypothesis in~\eqref{null_hyp}.
+This manuscript proposes a simple extension to the methodology in $MSstats$ and $MSstatsTMT$, to enable testing of the relevant the null hypothesis in~\eqref{null_hyp}.
 
@@
 
-\medskip \noindent {\bf Detection of changes in PTMs in balanced design with one source of variation}
+\medskip \noindent {\bf Detecting changes in PTMs, adjusted for global changes in protein abundance}
 
 \medskip \noindent
-We propose a statistical approach for detecting changes in PTMs.
 The method takes as input the modified spectral features $y_{ijk}^{\ast}$, and the corresponding unmodified features $y_{ijk}$.
@@
 We leverage the existing methods in $MSstats$ and $MSstatsTMT$ and extend them to remove confounding between changes in the PTM and the overall changes in protein abundance.
-Each feature type is summarized separately using $MSstats$ or $MSstatsTMT$, resulting in run level values for the PTM, $\hat{y}_{ij}^{\ast}$, and unmodified protein.
-The estimated summaries of the modified features are then used as the input to the model, such as in \eqref{msstats_model}, in $MSstats$ or $MSstatsTMT$.
-Similarly, the estimated summaries of the unmodified features are used as the input to the corresponding model in $MSstats$ or $MSstatsTMT$.
-
-For the modified features, the model-based estimates in $MSstats$ and $MSstatsTMT$ include $\hat{\Delta}_{PTM}=\hat{\mu}^{\ast}_{i}-\hat{\mu}_{i^{\prime}}^{\ast}$, and its standard error $\widehat{SE}(\hat{\Delta}_{PTM})$.
-Similarly, for the unmodified features, the model-based estimates in $MSstats$ and $MSstatsTMT$ incluse $\hat{\Delta}_{protein}=\hat{\mu}_{i}-\hat{\mu}_{i^{\prime}}$, and its standard error $\widehat{SE}(\hat{\Delta}_{protein})$.
-Therefore, the adjusted difference $\hat{\Delta}_{adj}$ in~\eqref{null_hyp} can be estimated as
+Each feature type is first analyzed separately using $MSstats$ or $MSstatsTMT$.
+In particular, the modified features are summarized into run-level summaries $\hat{y}_{ij}^{\ast}$ \todo{check the use of hat}.
+The estimated summaries of the modified features are then used as the input to the appropriate model in $MSstats$ or $MSstatsTMT$,  such as in \eqref{msstats_model}.
+The resulting model-based estimates include $\hat{\Delta}_{PTM}=\hat{\mu}^{\ast}_{i}-\hat{\mu}_{i^{\prime}}^{\ast}$, and its standard error $\widehat{SE}(\hat{\Delta}_{PTM})$.
+Similarly, the unmodified features of each protein are summarized for each run into $\hat{y}_{ij}$, the summaries are used as input to a separate analysis by $MSstats$ or $MSstatsTMT$, producing $\hat{\Delta}_{protein}=\hat{\mu}_{i}-\hat{\mu}_{i^{\prime}}$ and $\widehat{SE}(\hat{\Delta}_{protein})$.
+
+From the summaries above, the proposed approach estimates the adjusted difference $\hat{\Delta}_{adj}$ in~\eqref{null_hyp}
 \begin{eqnarray}
@@
 \end{eqnarray}
-Assuming that the unexplained by the model sources of variation in the modified features are independent from the corresponding sources of variation in the modified features, the standard error $\widehat{SE}(\hat{\Delta}_{adj})$ is obtained by combining the standard errors from the unmodified and modified model fits.
+Assuming that the sources of variation in the modified features unexplained by the model are independent from the corresponding sources of variation in the modified features, the standard error $\widehat{SE}(\hat{\Delta}_{adj})$ is obtained by combining the standard errors from the unmodified and modified model fits
 \begin{eqnarray}
@@
 \end{eqnarray}
-
 For example, in the simple case of \figref{data-structure} with $J=2$ conditions, where $\hat{\sigma}^2_{PTM}$ and $\hat{\sigma}^2_{protein}$ are the estimates of the error variance for the PTM and protein model described in \eqref{msstats_model}, the standard error would be calculated as
 \begin{eqnarray}
- \widehat{SE}(\hat{\Delta}_{adj}) &=& \sqrt{\widehat{SE}(\hat{\Delta}_{PTM})^2 + \widehat{SE}(\hat{\Delta}_{protein})^2} \\ 
-&=& \sqrt{\frac{1}{J}\hat{\sigma}^2_{PTM} + \frac{1}{J}\hat{\sigma}^2_{protein}} = \sqrt{\frac{2}{J}(\hat{\sigma}^2_{PTM} + \hat{\sigma}^2_{protein})} = \sqrt{\hat{\sigma}^2_{PTM} + \hat{\sigma}^2_{protein}} \nonumber
+ \widehat{SE}(\hat{\Delta}_{adj}) &=& \sqrt{\widehat{SE}(\hat{\Delta}_{PTM})^2 + \widehat{SE}(\hat{\Delta}_{protein})^2} = \sqrt{\frac{1}{J}\hat{\sigma}^2_{PTM} + \frac{1}{J}\hat{\sigma}^2_{protein}}
 \label{eq:se_calc}
 \end{eqnarray}
-
 The estimated standard error is larger than the standard errors associated with each distinct feature type, reflecting the increased uncertainty inherent to combining the uncertainty in two estimators.
@@
 \begin{eqnarray}
-\left( \widehat{SE}(\hat{\Delta}_{PTM})^{2} + \widehat{SE}(\hat{\Delta}_{protein})^{2} \right)^2 \bigg/
+df \left(  \widehat{SE}(\hat{\Delta}_{adj}) \right)=  \left( \widehat{SE}(\hat{\Delta}_{PTM})^{2} + \widehat{SE}(\hat{\Delta}_{protein})^{2} \right)^2 \bigg/
 \left( \frac{\widehat{SE}(\hat{\Delta}_{PTM})^{4}}{\mathrm{df}(\widehat{SE}(\hat{\Delta}_{PTM}))} + \frac{\widehat{SE}(\hat{\Delta}_{protein})^{4}}{ \mathrm{df}(\widehat{SE}(\hat{\Delta}_{protein}))} \right)
@@
 
-By leveraging the implementations in $MSstats$ and $MSstatsTMT$, the proposed approach is very versatile.
+\todo{This paragraph relates more to the implementation.
+Could you integrate it with the implementation text?} By leveraging the implementations in $MSstats$ and $MSstatsTMT$, the proposed approach is very versatile.
 It is applicable to a wide variety of experimental designs, including group comparison, paired designs, time course designs and unbalanced designs.

### 2022-07-26 main/MSstatsPTM_Main.tex
Small

@@
 
-We leverage the existing methods in $MSstats$ and $MSstatsTMT$ and extend them to remove confounding between changes in the PTM and the overall changes in protein abundance.
+We leverage the existing methods in $MSstats$ and $MSstatsTMT$, and extend them to remove confounding between changes in the PTM and the overall changes in protein abundance.
 Each feature type is first analyzed separately using $MSstats$ or $MSstatsTMT$.
@@
 The resulting model-based estimates include $\hat{\Delta}_{PTM}=\hat{\mu}^{\ast}_{i}-\hat{\mu}_{i^{\prime}}^{\ast}$, and its standard error $\widehat{SE}(\hat{\Delta}_{PTM})$.
-Similarly, the unmodified features of each protein are summarized for each run into $\hat{y}_{ij}$, the summaries are used as input to a separate analysis by $MSstats$ or $MSstatsTMT$, producing $\hat{\Delta}_{protein}=\hat{\mu}_{i}-\hat{\mu}_{i^{\prime}}$ and $\widehat{SE}(\hat{\Delta}_{protein})$.
+Similarly, the unmodified features of each protein are summarized for each run into $\hat{y}_{ij}$, and the summaries are used as input to a separate analysis by $MSstats$ or $MSstatsTMT$ producing $\hat{\Delta}_{protein}=\hat{\mu}_{i}-\hat{\mu}_{i^{\prime}}$ and $\widehat{SE}(\hat{\Delta}_{protein})$.
 
@@
 \end{eqnarray}
-Assuming that the sources of variation in the modified features unexplained by the model are independent from the corresponding sources of variation in the modified features, the standard error $\widehat{SE}(\hat{\Delta}_{adj})$ is obtained by combining the standard errors from the unmodified and modified model fits
+Assuming that the sources of variation in the modified features that are unexplained by the model are independent from the corresponding sources of variation in the modified features, the standard error $\widehat{SE}(\hat{\Delta}_{adj})$ is obtained by combining the standard errors from the unmodified and modified model fits
 \begin{eqnarray}
@@
 \end{eqnarray}
-For example, in the simple case of \figref{data-structure} with $J=2$ conditions, where $\hat{\sigma}^2_{PTM}$ and $\hat{\sigma}^2_{protein}$ are the estimates of the error variance for the PTM and protein model described in \eqref{msstats_model}, the standard error would be calculated as
+For example, in the simple case of \figref{data-structure} with $J=2$ conditions, where $\hat{\sigma}^2_{PTM}$ and $\hat{\sigma}^2_{protein}$ are respectively  the estimates of the error variance for the PTM and protein model described in \eqref{msstats_model}, the standard error is calculated as
 \begin{eqnarray}
@@
 \end{eqnarray}
-The estimated standard error is larger than the standard errors associated with each distinct feature type, reflecting the increased uncertainty inherent to combining the uncertainty in two estimators.
-
+The estimated standard error is larger than the standard errors associated with each individual feature type, reflecting the combined uncertainty in the two estimates. 
 Finally, the degrees of freedom associated with~\eqref{prop_se} are obtained via the Satterthwaite approximation~\cite{kutner_etal_04a, satterthwaite:1946}
@@
 \todo{This paragraph relates more to the implementation.
-Could you integrate it with the implementation text?} By leveraging the implementations in $MSstats$ and $MSstatsTMT$, the proposed approach is very versatile.
+Could you integrate it with the implementation section?} By leveraging the implementations in $MSstats$ and $MSstatsTMT$, the proposed approach is very versatile.
 It is applicable to a wide variety of experimental designs, including group comparison, paired designs, time course designs and unbalanced designs.
@@
 
-\subsubsection*{Design of balanced PTM experiments with one source of variation}
+\subsubsection*{Sample size calculation for future PTM experiments}
 \label{sec:design}
 
-The proposed statistical framework allows for design of PTM experiments in terms of sample size calculation and power analysis. 
-%Sample size calculation based on linear models has been described for general applications~\cite{kutner_etal_04a} and specifically for protein significance analysis~\cite{oberg_vitek_09a}. 
-Sample size calculation takes as input a) $q$, the desired false discovery rate, b) $\beta$, the average Type II error rate, c) $\Delta_{adj}$, the minimal log-fold change in adjusted PTM abundance that we would like to detect, d) $m_0 / (m_0 + m_1)$, the fraction of truly differentially modified PTM sites in the comparison, and e) $\sigma_{PTM}^{2}$ and $\sigma_{protein}^{2}$, the anticipated variances associated to modified and unmodified peptide features, respectively.
-The variances can be derived based on the dataset being analyzed, assuming similar quantitative properties and variations.
-With these values and a user-specified number of conditions, the corresponding number of technical replicates per condition can then be derived, as described in~\cite{kutner_etal_04a}.
-Given the above quantities, the minimal number of replicates $J$ is determined by the variance of the estimated log-fold change $\mathrm{SE}^{2}(\hat{\Delta}_{adj})$ as
-
+The proposed statistical framework enables sample size calculation for future PTM experiments. 
+The procedure has been described in general in~\cite{kutner_etal_04a}, and specifically for protein significance analysis in~\cite{oberg_vitek_09a}. 
+It requires us to specify the desired levels of the following quantities: a) $q$, the False Discovery Rate of detecting differential abundance, b) $\beta$, the average Type II error rate, c) $\Delta_{adj}$, the minimal log$_2$-fold change in adjusted PTM abundance  of interest, d) $m_0 / (m_0 + m_1)$, the fraction of truly differentially modified PTM sites in the comparison, and e) $\sigma_{PTM}^{2}$ and $\sigma_{protein}^{2}$, the anticipated variances associated with the summaries of the modified and unmodified features, respectively.
+Typically, the variances are estimated from an existing experiment, conducted with the same biological material and measurement workflow.
+Given the above quantities, and assuming a balanced design, the minimal number of replicates $J$ across $I$ conditions is chosen to bound the variance of the estimated log$_2$-fold change $\mathrm{SE}^{2}(\hat{\Delta}_{adj})$: \todo{There should be no hats in equations 11-13.
+The calculations assume that the variances are given externally, and are known.
+The input quantities above have no hats}
 \begin{equation}
@@
 \end{equation}
-
-and $t_{1-\beta, df}$ and $t_{1-\alpha /2, df}$ are the $100(1-\beta)^{\text{th}}$ and the $100(1-\alpha /2)^{\text{th}}$ percentiles of the $t$ distribution, with $df = I(J-1)$ degrees of freedom in balanced designs.
-
-Solving for $J$, the number of biological replicates per group is calculated as
-
+and $t_{1-\beta, df}$ and $t_{1-\alpha /2, df}$ are the $100(1-\beta)^{\text{th}}$ and the $100(1-\alpha /2)^{\text{th}}$ percentiles of the Student distribution with $df = I(J-1)$ degrees of freedom.
+Solving for $J$, the number of biological replicates per condition is
 \begin{equation}
@@
 \end{equation}
-
-Details on how the proposed method can be used to run sample size calculations and power analysis on experiments with complex design can be seen in {\bf Supplementary Sec.
+\todo{Comment on the fact that since the numerator has two sources of variation, reflecting a larger uncertainty in the adjusted calculation, the adjustment typically requires a larger sample size to gain the same sensitivity as the unadjusted estimation.} Details of applications of this approach to experiments with complex design are in {\bf Supplementary Sec.
 \ref{supp-sec:complex_design}}.
-More details on sample size calculation can be found in~\cite{oberg_vitek_09a}.
 
@@
 $MSstatsPTM$ takes as input lists of identified and quantified spectral features, produced by spectral processing tools such as MaxQuant, Progenesis, or Spectronaut.
-We require the processing tool to identify the amino acid in the protein sequence where the modification occurred (i.e.
+
+$MSstatsPTM$ requires the processing tool to identify the amino acid in the protein sequence where the modification occurred (i.e.
 the modification site).
@@
 For example, a modification on a serine, which was the 70th amino acid in the sequence would be marked ``S70''.
+
 Occasionally tools will only include the peptide sequence with the modified amino acid highlighted.

### 2022-07-26 main/MSstatsPTM_Main.tex
Implementation

@@
 \end{equation}
-\todo{Comment on the fact that since the numerator has two sources of variation, reflecting a larger uncertainty in the adjusted calculation, the adjustment typically requires a larger sample size to gain the same sensitivity as the unadjusted estimation.} Details of applications of this approach to experiments with complex design are in {\bf Supplementary Sec.
+Since the numerator has two sources of variation, reflecting a larger uncertainty in the adjusted calculation, the adjustment typically requires a larger sample size to gain the same sensitivity as the unadjusted estimation.
+Details of applications of this approach to experiments with complex design are in {\bf Supplementary Sec.
 \ref{supp-sec:complex_design}}.
@@
 \subsection*{Implementation of MSstatsPTM}
+
+\todo{I suggest that you follow the same order as Figure 3, one paragraph per step.
+Describe each step, and refer to both the figure, the quantity in the equations and the MSstats functions.
+Specifically: design; data acquisition; data processing and requirements; converters; steps of statistical analysis.
+I would state upfront that all the extols before last are performed separately for each data type, and not repeat it in later paragraphs}
 
@@
 This will generally include the amino acid abbreviation, plus it's number in the protein sequence.
-For example, a modification on a serine, which was the 70th amino acid in the sequence would be marked ``S70''.
-
-Occasionally tools will only include the peptide sequence with the modified amino acid highlighted.
-If there is no way to determine where in the protein sequence the modification occurred, $MSstatsPTM$ includes functionality to do this manually, given the modified peptide sequence and a FASTA file with the entire protein sequence.
-The package includes converters for seamless integration with multiple such tools.
-The converters can optionally keep peptides with both single and multiple modification sites, or filter out only those with multiple sites.
-If multiple sites are kept, these are grouped separately from the singly modified peptide so as not to confound the effect.
-Additionally, peptides with oxidation can also be removed if the effect is believed to confound the modifications of interest.
+For example, a modification on a 70th amino acid in the sequence serine should be marked as ``S70''.
+
+Occasionally the outputs of data processing tools only include the peptide sequence with the modified amino acid highlighted, without indicating where in the protein sequence the modification occurred.
+For these cases $MSstatsPTM$ includes functionality for identifying this location \todo{what do you mean by 'manually'?
+The users have to manually input it?} manually, given the modified peptide sequence and a FASTA file with the entire protein sequence.
+The package includes converters for seamless integration \todo{integration of what?} with multiple such tools.
+\todo{Maybe put all the discussion of converters in one place - integrate with the first paragraph?
+} The converters can optionally keep peptides with both single and multiple modification sites, or filter out only those with multiple sites.
+If multiple sites are kept, these are grouped separately from the singly modified peptide so as not to confound the effect \todo{confound how?
+The effect of what?
+We use confounding to indicate the changes in protein abundance.
+This looks like a different confounding.
+Could you clarify?}.
+Additionally, peptides with oxidation can also be removed if the effect is believed to confound the modifications of interest \todo{confound how?
+The effect of what?}.
 For experimental designs that include separate runs enriched in modified peptides and global profiling runs, conversion is performed separately for the runs enriched in modified peptides and for the global profiling runs.

### 2022-07-26 main/MSstatsPTM_Main.tex
Evaluation

@@
 
-\todo{I suggest that you follow the same order as Figure 3, one paragraph per step.
+\todo{I suggest that you follow the same order as Figure 3 (my suggested version), one paragraph per step.
 Describe each step, and refer to both the figure, the quantity in the equations and the MSstats functions.
@@
 
-The evaluations were based on the $MSstatsPTM$, and were performed without imputing missing values or using Empirical Bayes moderation.
-$MSstatsPTM$ was compared with Anova and Limma.
-We did not compare to IsobarPTM because it is only applicable to TMT experiments, meaning it could not be applied to our ground truth data.
-All the evaluations were done at the adjusted p-value cutoff of $\alpha=.05$.
+We compared the performance of  $MSstatsPTM$ to that of ANOVA and of Limma.
+Since IsobarPTM is only applicable to experiments with TMT labeling, it could not be applied to the datasets with known ground truth in this manuscript, and was excluded from the comparisons.
+$MSstatsPTM$ was used without imputing missing values, and without Empirical Bayes moderation.
+All the evaluations were done at the FDR-adjusted p-value cutoff of \todo{typically adjusted p-values are denoted $q$, and $\alpha$ is left for the unadjusted p-values.
+E.f., Eq.
+12 uses $q$.
+Could you check the entire manuscript for consistency} $\alpha=.05$.
 More details are in {\bf Supplementary Sec.

### 2022-07-26 main/MSstatsPTM_Main.tex
Small suggestions for results

@@
 Frequently data from these experiments are analyzed using statistical methods that were not originally designed for this task.
-Researchers use methods such as $t$-test\cite{Kalpic:2011}, Analysis of Variance (Anova)\cite{girden:1992}, or Limma\cite{Ritchie_15a}, by taking as input the intensity ratios of modified and unmodified peptide features, and comparing the mean abundance of different PTM sites.
+Researchers use methods such as $t$-test\cite{Kalpic:2011}, Analysis of Variance (ANOVA)\cite{girden:1992}, or Limma\cite{Ritchie_15a}, by taking as input the intensity ratios of modified and unmodified peptide features, and comparing the mean abundance of different PTM sites.
 Such approaches do not fully account for all the sources of uncertainty.
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsubsection*{Evaluation criteria}
+
 We evaluated $MSstatsPTM$ on simulated and spike-in datasets with known ground truth in terms of true positives ($TP$), false positives ($FP$), true negatives ($TN$), and false negatives ($FN$) differentially abundant PTMs.
@@
 All the evaluations were done at the FDR-adjusted p-value cutoff of \todo{typically adjusted p-values are denoted $q$, and $\alpha$ is left for the unadjusted p-values.
-E.f., Eq.
+E.g., Eq.
 12 uses $q$.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection*{In clean simulations, MSstatsPTM corrected for high FDR of differentially abundant PTM}
+\subsubsection*{In clean simulations, MSstatsPTM corrected the excess FDR in unadjusted differentially abundant PTM}
 
@@
 In a clean simulation, recall and accuracy of $MSstatsPTM$ and Limma performed similarly.
-As expected when not using Empiricial Bayes moderation in MSstatsPTM, Limma performed slightly better in simulations with fewer biological replicates.
-In contrast, the performance of Anova lagged behind, especially in cases with higher simulated variance.
-This was due to Anova's more simplistic variance estimation.
+As expected when not using Empiricial Bayes moderation in MSstatsPTM, Limma performed slightly better in simulations with fewer biological replicates \todo{Hmm...
+figure 4a doesn't show the sample size.
+Did you mean Fig 4b?
+Then it should be in the next paragraph.
+If you refer to supplementary figures, say so first}.
+In contrast, the performance of ANOVA lagged behind, especially in cases with higher simulated variance.
+This was due to ANOVA's more simplistic variance estimation.
 In contrast, $MSstatsPTM$ and Limma implement methods to provide better estimates of the variance, resulting in improved power.
-Finally, since this simulation is a group comparison design that doesn't require many variance components, it is favorable to Limma and Anova.
+Finally, since this simulation is a group comparison design that doesn't require many variance components, it is favorable to Limma and ANOVA.
 As discussed previously, these methods are not directly applicable to experiments with additional variance components.
@@
 
-In simulations with missing values and few features, $MSstatsPTM$ outperformed Anova and Limma.
+In simulations with missing values and few features, $MSstatsPTM$ outperformed ANOVA and Limma.
 Changes in unmodified protein level still needed to be accounted for to control the FDR.
@@
 Additionally, when comparing the fold change estimation across all modified peptides, the proposed method showed a tighter distribution of estimated fold changes around the true fold change.
-Specifically, the inner quartile range (IQR) of the estimated fold change for the proposed method was on average 32.5\% smaller than Limma and Anova's IQR.
+Specifically, the inner quartile range (IQR) of the estimated fold change for the proposed method was on average 32.5\% smaller than Limma and ANOVA's IQR.
 This means that while the mean of the estimated fold changes was generally correct for all methods, the proposed approach correctly estimated the fold change more often across all PTMs.
@@
 Of the three approaches, the proposed method showed the tightest distribution around the true log fold change.
-Comparing the IQR of the spike-in peptide's log fold change, the proposed method's IQR was 32.86\% smaller than Limma and Anova's IQR.
-While Limma and Anova overall estimated the correct fold change, they were inconsistent in their estimation, resulting in a higher error rate.
+Comparing the IQR of the spike-in peptide's log fold change, the proposed method's IQR was 32.86\% smaller than Limma and ANOVA's IQR.
+While Limma and ANOVA overall estimated the correct fold change, they were inconsistent in their estimation, resulting in a higher error rate.
 The proposed method's estimation was more consistent, with a smaller error.
@@
 The framework is general and is applicable to a variety of experimental designs.
-It outperforms the ad-hoc methods underlying Anova and Limma, and yields accurate results in the broad type of experimental circumstances, including the presence of missing values, changes in protein abundance, few representative peptides, and different acquisition methods.
+It outperforms the ad-hoc methods underlying ANOVA and Limma, and yields accurate results in the broad type of experimental circumstances, including the presence of missing values, changes in protein abundance, few representative peptides, and different acquisition methods.
 The framework allows us to plan for subsequent experiments, and choose the appropriate number of replicates in consideration of adjustment with respect to protein abundance.
@@
 \end{subfigure}
-\caption{Dataset 1 \& 2: Computer simulation.
+\caption{\todo{In the figure titles and in the caption, explain which simulation is 'clean' and which is 'noisy'.
+Use the same names for the datasets as in the Datasets section throughout.
+ANOVA is an abbreviation and needs all capital letters - could you check throughout in the figures} Dataset 1 \& 2: Computer simulation.
 a) All the considered methods in the first computer simulation correctly calibrated FDR when adjusting for changes in protein abundance.
@@
 b) The advantage of using the proposed approach was apparent when including limited observations and missing values.
-Looking at accuracy, the proposed method outperformed Anova and Limma in nearly every model.
+Looking at accuracy, the proposed method outperformed ANOVA and Limma in nearly every model.
 }
@@
 a) Before adjustment the fold change of the spike-in peptides' were systematically different from the expected fold change in all models.
-After adjustment, this systemic difference was removed, however the inner quartile range of the Limma and Anova models was wider than the proposed method.
+After adjustment, this systemic difference was removed, however the inner quartile range of the Limma and ANOVA models was wider than the proposed method.
 b) Before adjustment the spike-in peptides (colored red) did not follow the expected log fold change; after adjustment, the spike-in peptides were more in line with expectation.

### 2022-07-26 main/MSstatsPTM_Main.tex
evaluation changes

@@
 Typically, the variances are estimated from an existing experiment, conducted with the same biological material and measurement workflow.
-Given the above quantities, and assuming a balanced design, the minimal number of replicates $J$ across $I$ conditions is chosen to bound the variance of the estimated log$_2$-fold change $\mathrm{SE}^{2}(\hat{\Delta}_{adj})$: \todo{There should be no hats in equations 11-13.
-The calculations assume that the variances are given externally, and are known.
-The input quantities above have no hats}
+Given the above quantities, and assuming a balanced design, the minimal number of replicates $J$ across $I$ conditions is chosen to bound the variance of the estimated log$_2$-fold change $\mathrm{SE}^{2}(\Delta_{adj})$:
+
 \begin{equation}
-\widehat{SE}(\hat{\Delta}_{adj})^2 = \left[ \frac{2}{J} \left( \hat{\sigma}_{PTM}^{2} + \hat{\sigma}_{protein}^{2} \right) \right]
+SE(\Delta_{adj})^2 = \left[ \frac{2}{J} \left( \sigma_{PTM}^{2} + \sigma_{protein}^{2} \right) \right]
 \leq \left( \frac{\Delta_{adj}}{t_{1-\beta, df} + t_{1-\alpha /2, df}} \right)^{2}
@@
 \begin{equation}
-J \geq \frac{(2\hat{\sigma}_{PTM}^{2} + 2\hat{\sigma}_{protein}^{2})(t_{1-\beta, df} + t_{1-\alpha /2, df})^2}{\Delta_{adj}^2}
+J \geq \frac{(2\sigma_{PTM}^{2} + 2\sigma_{protein}^{2})(t_{1-\beta, df} + t_{1-\alpha /2, df})^2}{\Delta_{adj}^2}
 \end{equation}
@@
 $MSstatsPTM$ was used without imputing missing values, and without Empirical Bayes moderation.
-All the evaluations were done at the FDR-adjusted p-value cutoff of \todo{typically adjusted p-values are denoted $q$, and $\alpha$ is left for the unadjusted p-values.
-E.g., Eq.
-12 uses $q$.
-Could you check the entire manuscript for consistency} $\alpha=.05$.
+All the evaluations were done at the FDR-adjusted p-value cutoff of $q=.05$.
 More details are in {\bf Supplementary Sec.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection*{In clean simulations, MSstatsPTM corrected the excess FDR in unadjusted differentially abundant PTM}
+\subsubsection*{MSstatsPTM corrected the excess FDR in unadjusted differentially abundant PTMs in clean simulations}
 
 \figref{sim1_fdr} illustrated the high FDR observed on the simulated datasets when not accounting for the overall changes in protein abundance.
-After the adjustment, the FDR of all methods improved.
-In a clean simulation, recall and accuracy of $MSstatsPTM$ and Limma performed similarly.
-As expected when not using Empiricial Bayes moderation in MSstatsPTM, Limma performed slightly better in simulations with fewer biological replicates \todo{Hmm...
-figure 4a doesn't show the sample size.
-Did you mean Fig 4b?
-Then it should be in the next paragraph.
-If you refer to supplementary figures, say so first}.
-In contrast, the performance of ANOVA lagged behind, especially in cases with higher simulated variance.
-This was due to ANOVA's more simplistic variance estimation.
-In contrast, $MSstatsPTM$ and Limma implement methods to provide better estimates of the variance, resulting in improved power.
-Finally, since this simulation is a group comparison design that doesn't require many variance components, it is favorable to Limma and ANOVA.
+After the adjustment, the FDR of all methods improved, and were in line with the excepted FDR of $.05$.
+Adjusting for changes in the unmodified protein removed differentially abundant PTMs whose abundance change was caused by the protein.
+%Finally, since this simulation is a group comparison design that doesn't require many variance components, it is favorable to Limma and ANOVA.
 As discussed previously, these methods are not directly applicable to experiments with additional variance components.
@@
 Specifically, the inner quartile range (IQR) of the estimated fold change for the proposed method was on average 32.5\% smaller than Limma and ANOVA's IQR.
-This means that while the mean of the estimated fold changes was generally correct for all methods, the proposed approach correctly estimated the fold change more often across all PTMs.
+While the mean of the estimated fold changes was generally correct for all methods, the proposed approach correctly estimated the fold change more often across all PTMs.
 This fold change comparison can be seen in {\bf Supplementary Figure~\ref{supp-fig:fc_boxplot}}.
@@
 This experiment introduced more variance components, which is challenging for the existing methods and therefore the difference in the results is more pronounced than in the simulations.
+Volcano plots for all methods and comparisons can be seen in {\bf Supplementary Section~\ref{supp-sec:benchmark}}.
 
@@
 The results of these experiments are summarized in~\figref{venn_diagrams} and~\figref{profile_plots}.
-\figref{data4_venn_diagram} and~\figref{data5_venn_diagram} show the number of differentially abundant modified peptides before and after adjustment.
+\figref{venn_diagrams} shows the number of differentially abundant modified peptides before and after adjustment for Datasets 4 \& 5.
 Adjusting for changes in the unmodified protein caused more PTMs to change from differentially abundant to non-differentially abundant than vice versa.
@@
 Originally the modification at $S178$ of protein $TTP$ was shown to be differentially abundant between KO\_Uninfect and KO\_Early, with a log fold change of $2.9$.
-However the unmodified protein was shown to contribute $2.014$ of this change, meaning that the modification only accounted for $.886$ after adjustment.
+However the unmodified protein was shown to contribute $2.014$ of this change, meaning that the modification only accounted for $.886$ after adjustment (seen in {\bf Supplementary Figure~\ref{supp-fig:No_Diff_Shigella_PTM}}).
 This caused the modification to go from differentially abundant to non-differentially abundant.

### 2022-07-26 supplementary/ptm_sm.tex
evaluation changes

@@
 
-\begin{figure}[h!]
-\centering
- \begin{subfigure}{\textwidth}
-  \centering
-	\includegraphics[width=.8\textwidth]{sim_new/sim1_Recall}\\
-	\caption{All methods with adjustment were compared by comparing recall in simulation 1.
+%\begin{figure}[h!]
+%\centering
+% \begin{subfigure}{\textwidth}
+%  \centering
+%	\includegraphics[width=.8\textwidth]{sim_new/sim1_Recall}\\
+%	\caption{All methods with adjustment were compared by comparing recall in simulation 1.
 Limma performed the strongest here when the number of replicates were low.
@@
 Anova clearly performed worse across all methods.}
- \end{subfigure}
- \begin{subfigure}{\textwidth}
-  \centering
-	\includegraphics[width=.8\textwidth]{sim_new/sim1_accuracy}
-	\caption{The overall accuracy plot mimicked the observations in the recall plot. Limma performed strongest when replicates were low, but was comparable to the proposed method with more replicates.}
- \end{subfigure}
-\caption{Recall and Accuracy results of Simulation 1.}
-\label{fig:sim1_recall}
-\end{figure}
-
-\begin{figure}[h!]
-\centering
-	\includegraphics[width=.8\textwidth]{sim_new/sim3_Recall}
-	\caption{}
-\caption{Recall results of Simulation 2.
+% \end{subfigure}
+% \begin{subfigure}{\textwidth}
+%  \centering
+%	\includegraphics[width=.8\textwidth]{sim_new/sim1_accuracy}
+%	\caption{The overall accuracy plot mimicked the observations in the recall plot. Limma performed strongest when replicates were low, but was comparable to the proposed method with more replicates.}
+% \end{subfigure}
+%\caption{Recall and Accuracy results of Simulation 1.}
+%\label{fig:sim1_recall}
+%\end{figure}
+
+%\begin{figure}[h!]
+%\centering
+%	\includegraphics[width=.8\textwidth]{sim_new/sim3_Recall}
+%	\caption{}
+%\caption{Recall results of Simulation 2.
 The advantage of using the proposed approach were apparent when looking at simulation 2, which included limited observations and the presence of missing values.
@@
 Again the lowest performing method was Anova.}
-\label{fig:sim2_recall}
-\end{figure}
+%\label{fig:sim2_recall}
+%\end{figure}
 
@@
  \centering
- \begin{subfigure}{\textwidth}
- \centering
-	\includegraphics[width=.9\textwidth]{sim_new/IpaH_prof_plot}
-	\caption{Here the global profiling of protein $GSDMD\_HUMAN|P57764$ with the ubiquitination of the protein at site $K62$ was compared.
-The individual PSM features areshown in grey, while the feature summarization is shown in red.
-The summary of the modification and global protein showed that the conditions followed different trends.
-Specifically, there appeared to be no change in abundance between Dox1hr and Dox4hr in the modified plot, however there was a large negative change in the unmodified plot.
-This indicates the modification was confounded with changes in the unmodified protein.}
- \end{subfigure}
- \begin{subfigure}{\textwidth}
- \centering
 	\includegraphics[width=.9\textwidth]{sim_new/IpaH_volcano_plot}
@@
 In this case the proposed method allowed us to identify a differential modified peptide that could have otherwise been missed.}
- \end{subfigure}
-\caption{Summary plots for modification of protein GSDMD at site K62.}
 \label{fig:ipah_figures}
@@
 \centering
- \begin{subfigure}{\textwidth}
- \centering
-	\includegraphics[width=.95\textwidth]{sim_new/No_Difference_Shigella_Profile_Plot}
-	\caption{The global profiling of protein $TTP\_MOUSE|P22893$ with the modification of the protein at site $S178$ was compared.
-The individual PSM features are shown in grey, while the feature summarization is shown in red.
-The summary of the modification and global protein showed that the conditions followed the same trend.
-Specifically, there was a positive adjustment in abundance when comparing WT\_Uninfect to WT\_Late in both the modification and global profiling run.
-This indicated the movement is driven by changes in global protein.}
- \end{subfigure}
- \begin{subfigure}{\textwidth}
- \centering
 	\includegraphics[width=.9\textwidth]{sim_new/No_Difference_Shigella_Volcano}
@@
 After adjustment the fold change was much smaller and the adjusted pvalue was insignificant.}
- \end{subfigure}
-\caption{Summary plots for modification of protein TTP at site S178.}
 \label{fig:No_Diff_Shigella_PTM}

### 2022-07-27 main/MSstatsPTM_Main.tex
updated implementation figure
Small

@@
 Specifically: design; data acquisition; data processing and requirements; converters; steps of statistical analysis.
-I would state upfront that all the extols before last are performed separately for each data type, and not repeat it in later paragraphs}
+I would state upfront that all the steps before last are performed separately for each data type, and not repeat it in later paragraphs}
 
@@
 \centering
-\includegraphics[scale=.6]{images/MSstatsPTM_design.png}
-\caption{The $MSstatsPTM$ workflow.
-The names of the R functions used for each step are highlighted in red.
+\includegraphics[scale=.5]{images/MSstatsPTM_design.png}
+\caption{\todo{Looks great.
+Maybe to the left add vertical text 'Upstream experiments and computation' and  'MSstatsPTM' next to the corresponding boxes.
+No need for a capital L in lysate.
+'Group comparison among PTM' and 'Group comparison among unmodified proteins'.
+I would add () next to the function names, to clarify that these are functions} The $MSstatsPTM$ workflow.
+The names of the R functions used for each step are highlighted in purple and the output notations are highlighted in red.
 The package is applicable to label-free data acquisitions such as DDA, DIA, SRM and label-based data acquisitions such as TMT.
-The package's workflow starts after modified and unmodified peptide quantification.
 It takes as input lists of identified and quantified spectral features for the PTM and unmodified protein, produced by spectral processing tools such as MaxQuant, Progenesis, or Spectronaut.
-Separate parallel analysis for the PTM and unmodified protein is performed, including feature summarization and modeling.
-After a model is fit to both the PTM and unmodified protein, the PTM model is adjusted for changes in unmodified protein abundance.
-Plots can optionally be created at each step to summarize the results.}
+Conversion, summarization and statistical modeling are performed separately for the PTM and for the unmodified proteins.
+At the end, the model-based summaries are combined to adjust the changes in the PTM abundance for changes in unmodified protein abundance.
+Data visualizations can optionally be created at each step.}
 \label{fig:msstatsptm_design}
@@
 \centering
-\begin{subfigure}[c]{0.825\linewidth}
+\begin{subfigure}[c]{0.7\linewidth}
 \includegraphics[width=1\textwidth]{images/sim1_FDR_all_models.png}

### 2022-07-28 code/Spike-in_Vizualization.Rmd
addressed all comments in main

@@
 ```{r pressure, echo=FALSE, fig.width = 10, fig.height = 7}
-load(file = "MSstatsPTM_Summarized.rda")
+load(file = "../data/MSstatsPTM_Summarized.rda")
 
-load(file = "MSstatsPTM_Model.rda")#model_df
+load(file = "../data/Spike_in_MSstatsPTM_Model.rda")#model_df
 
@@
 plot_df %>% arrange(labeling) %>% 
-  filter(abs(log2FC) < Inf) %>% # & Label %in% c("mix1-mix3", "mix1-mix4", "mix2-mix3", "mix2-mix4")
+  filter(abs(log2FC) < Inf) %>% 
   ggplot(aes(log2FC, -log10(adj.pvalue))) + 
@@
   scale_alpha_manual(values = c(0.2, 1)) + 
-  #scale_size_manual(values = c(2, 1)) + 
   geom_hline(yintercept = 0) + 
@@
 ```{r, echo=FALSE, fig.width = 10, fig.height = 7}
-load(file = "Limma_Model.rda")#limma_model
+load(file = "../data/Spike_in_Limma_Model.rda")#limma_model
 
@@
 ```{r, echo=FALSE, fig.width = 10, fig.height = 7}
-load(file = "Anova_Model.rda")#anova_results
+load(file = "../data/anova_results.rda")#anova_results
 
@@
 # Volcano plot
+png("../supplementary/sim_new/spike_in_anova_volcano.png", width = 1000, height = 500)
 anova_results %>% arrange(labeling) %>% 
@@
   guides(colour = guide_legend(override.aes = list(size=5))) + 
-  labs(title = "Anova VolcanoPlot")
+  labs(title = "ANOVA VolcanoPlot")
+dev.off()
+
 ```
@@
 anova_summary$Adjustment <- "protein adjustment"
-anova_summary$Model <- "Anova"
+anova_summary$Model <- "ANOVA"
 
@@
 unadj_spike_ttest$Adjustment <- "no adjustment"
-unadj_spike_ttest$Model <- "Anova"
+unadj_spike_ttest$Model <- "ANOVA"
 
@@
 adj_spike_ttest$Adjustment <- "protein adjustment"
-adj_spike_ttest$Model <- "Anova"
+adj_spike_ttest$Model <- "ANOVA"
 
@@
 
+png("../main/images/spike_in_fc.png", width = 750, height = 500)
 spike %>% mutate(Comparison = str_split(contrast_truth, " ", simplify = TRUE)[,1],
-      Model = factor(Model, levels=c("Proposed", "Limma", "Anova"))) %>%
+      Model = factor(Model, levels=c("Proposed", "Limma", "ANOVA"))) %>%
   filter(!Label %in% c("mix1-mix2", "mix3-mix4")) %>% 
@@
   labs(title = "Fold Change of Spike-in Peptides", x = "Comparison with Expected FC")
+dev.off()
 

### 2022-07-28 main/MSstatsPTM_Main.tex
addressed all comments in main

@@
 For example, meaningful patterns of changes in post-translational modifications (PTMs) abundance can serve as biomarkers of a disease~\cite{Petushkova_2017}.
-Alternatively, distinguishing the quantitative changes in a PTM from the overall changes of the protein abundance helps gain insight into biological and physiological processes operating on a very short timescale~\cite{Chandramouli:2009}\cite{Kim2016}.
+Alternatively, distinguishing the quantitative changes in a PTM from the overall changes of the protein abundance helps gain insight into biological and physiological processes operating on a very short timescale~\cite{Wu:2011, Chandramouli:2009, Kim2016}.
 This helps to distinguish between relative site occupancy changes at steady-state protein levels, typical for short time-scale signaling events, and observed relative changes of PTMs as a result of underlying gene expression or protein abundance levels.
@@
 Second, contrary to the often large number of identified peptides that can be used to quantify protein abundance, there are relatively few representative peptides that span a modification site, and there may be multiple modified sites on a single peptide~\cite{Mann}.
-Third, unless early signaling events are interrogated, the interpretation of the relative changes in modification occupancy are inherently confounded with changes in the overall protein abundance, complicating the interpretation of the results \cite{Olsen:2013}\cite{Wu:2011}.
+Third, unless early signaling events are interrogated, the interpretation of the relative changes in modification occupancy are inherently confounded with changes in the overall protein abundance, complicating the interpretation of the results \cite{Wu:2011, Olsen:2013}.
 Finally, technological aspects of bottom-up MS experiments, such as presence of labeling by tandem mass tag (TMT), introduce additional sources of uncertainty and variation.
@@
 
-\todo{I do not think this is right.
-I believe there should be a summarization first, such that the data are $\hat{y}_{ij}$, the model becomes $\hat{y}^*_{ij} = \mu^{*}_{i} + \epsilon^{\ast}_{ij}, \ \epsilon^{\ast}_{ij} \mathop\sim^{iid} \mathcal{N}(0, \sigma^{*2})$ and the df are $df=I(J-1)$.
-The index k should drop.
-Could you start by introducing the summarization?
-What do you think re keeping or removing the hat on $\hat{y}_{ij}$?}
 The basic Analysis of Variance (ANOVA) model~\cite{kutner_etal_04a}  is
 \begin{eqnarray}
-y^*_{ijk} = \mu^{*}_{i} + \epsilon^{\ast}_{ij}, \ \epsilon^{\ast}_{ij} \mathop\sim^{iid} \mathcal{N}(0, \sigma^{*2}),\ i=1,\ldots,I,\ j=1,\ldots,J,\ k=1,\ldots,K
+\hat{u}^*_{ij} = \mu^{*}_{i} + \epsilon^{\ast}_{ij}, \ \epsilon^{\ast}_{ij} \mathop\sim^{iid} \mathcal{N}(0, \sigma^{*2}),\ i=1,\ldots,I,\ j=1,\ldots,J
 \label{eq:ttest_model}
 \end{eqnarray}
-The model allows us to estimate $\hat{\Delta}$ and its standard error, and test the null hypothesis in \eqref{conv_null_hyp} by comparing the model-based test statistic against the Student distribution with $df=I(JK-1)$ \todo{$df=I(J-1)$} degrees of freedom in balanced designs.
+
+Where $\hat{u}^*_{ij}$ is the run-level summarization of modified features. For example, on the log$_2$ scale for condition $i$ and biological replicate $j$, $\hat{u}^*_{ij}$ is
+
+\begin{eqnarray}
+\hat{u}^*_{ij} =  \log_2 \sum_{k=1}^{K} 2^{y_{ijk}^{\ast}}
+\label{eq:anova_summ}
+\end{eqnarray}
+
+The model allows us to estimate $\hat{\Delta}$ and its standard error, and test the null hypothesis in \eqref{conv_null_hyp} by comparing the model-based test statistic against the Student distribution with $df=I(J-1)$ degrees of freedom in balanced designs.
 Unfortunately, this approach is fundamentally flawed as it does not account for the confounding between changes in the PTM abundance and the overall changes in the abundance of the unmodified protein.
 
-\medskip \noindent
+%\medskip \noindent
 
@@
 \begin{eqnarray}
-u_{ij} = \log_2 \frac{ \sum_{k=1}^{K} 2^{y_{ijk}^{\ast}} }{ \sum_{l=1}^{L} 2^{y_{ijl}} } = \log_2 \left( \sum_{k=1}^{K} 2^{y_{ijk}^{\ast}} \right) - \log_2 \left( \sum_{l=1}^{L} 2^{y_{ijl}} \right)
+\hat{u}_{ij} = \log_2 \frac{ \sum_{k=1}^{K} 2^{y_{ijk}^{\ast}} }{ \sum_{l=1}^{L} 2^{y_{ijl}} } = \log_2 \left( \sum_{k=1}^{K} 2^{y_{ijk}^{\ast}} \right) - \log_2 \left( \sum_{l=1}^{L} 2^{y_{ijl}} \right)
 \label{eq:adj_estimation}
 \end{eqnarray}
-The approach then models these values with the basic ANOVA, which corresponds to \todo{if we keep a hat on $\hat{y}_{ij}$, we should probably keep a hat on $\hat{u}_{ij}$}
-\begin{eqnarray}
-u_{ij} = (\mu^{\ast}_{i}-\mu_i) + \epsilon^{\prime}_{ij}, \text{where } \epsilon^{\prime}_{ij} \mathop\sim^{iid} \mathcal{N}(0, \sigma^{\prime 2}),\ i=1,\ldots,I,\ j=1,\ldots,J
+The approach then models these values with the basic ANOVA, which corresponds to 
+\begin{eqnarray}
+\hat{u}_{ij} = (\mu^{\ast}_{i}-\mu_i) + \epsilon^{\prime}_{ij}, \text{where } \epsilon^{\prime}_{ij} \mathop\sim^{iid} \mathcal{N}(0, \sigma^{\prime 2}),\ i=1,\ldots,I,\ j=1,\ldots,J
 \label{eq:ttest_model}
@@
 The model differs from~\eqref{ttest_model} in that it specifies additional prior distributions for the model parameters.
-The prior distributions are estimated from the data across all the proteins and all the PTM as described in \todo{XXX}. 
+The prior distributions are estimated from the data across all the proteins and all the PTM as described in \cite{Smyth:2004}. 
 With this approach, testing the null hypothesis in~\eqref{null_hyp} amounts to combining the PTM- and protein-specific estimates of variation with a consensus estimate obtained from all the PTM and all the proteins.
-\todo{It's a bit vague.
-I am wondering if an equation would help}
 As the result, in experiments with few biological replicates the standard errors are often smaller, and the degrees of freedom are often larger than without moderation~\cite{Ritchie_15a}. 
@@
 Isobar-PTM expresses MS measurements with a linear model and performs adjustment with respect to protein abundance using the difference between log-ratio of modified peptides in two channels and log-ratio of protein level.
-\todo{This seems somewhat similar to the ANOVA on ratios.
-Could you contrast with the ANOVA?
-Or just provide an equation in the same notation?} Unfortunately, this statistical modeling framework is not applicable to either label-free workflows or experiments with complex designs.
+Unfortunately, this statistical modeling framework is not applicable to either label-free workflows or experiments with complex designs.
 
@@
 For example, the unmodified protein features in \figref{data-structure} are modeled as
-\todo{I rewrote the model so that it uses $\mu_i$ directly rather than conditions}
-\todo{Should we use a hat on top of y, to indicate that we model the summaries?}
-\begin{eqnarray}
-& y_{ij} = \mu_i + Run_{j} + \epsilon_{ij}, \  \text{where } Run \stackrel{iid}{\sim} \mathcal{N}(0, \sigma^2_j), \epsilon_{ij} \stackrel{iid}{\sim} \mathcal{N}(0, \sigma^2) & 
+
+\begin{eqnarray}
+\hat{y}_{ij} = \mu_i + \epsilon_{ij}, \text{where } \epsilon_{ij} \stackrel{iid}{\sim} \mathcal{N}(0, \sigma^2)
 \label{eq:msstats_model}
 \end{eqnarray}
-\todo{Is it true that we add run in label-free experiments?
-I thought it was only with TMT.
-But maybe I forgot.
-Could you check?
-If we don't include run, then give a model for label-free, and then contrast with TMT}
-\todo{Is it true that we have a separate variance parameter for each run $ \sigma^2_j$? I don't think so. I think it should be a separate sigma (maybe denote it $\tau^2$), but same for all the runs}
+
+In the simplified case of \figref{data-structure} this amounts to the same model as in \eqref{ttest_model}.
+However, the model can be extended to more complex cases with additional sources of variation, which simpler models, such as ANOVA, cannot account for.
+For example, given a group comparison TMT experiment, with the addition of mixture as a source of variation, the model would extend to
+
+\begin{eqnarray}
+\hat{y}_{imj} = \mu_i + Mixture_m + \epsilon_{imj}, \text{where } Mixture_m \stackrel{iid}{\sim} \mathcal{N}(0, \sigma_M^2), \epsilon_{imj} \stackrel{iid}{\sim} \mathcal{N}(0, \sigma^2)
+\label{eq:msstatstmt_model}
+\end{eqnarray}
+
+Additional models for different experimental designs, including time series, can be seen in {\bf Supplementary Figure~\ref{supp-fig:statistical_inference}}.
+
 Parameters of the model are estimated using restricted maximum likelihood (REML)~\cite{Kenward}.
@@
 Each feature type is first analyzed separately using $MSstats$ or $MSstatsTMT$.
-In particular, the modified features are summarized into run-level summaries $\hat{y}_{ij}^{\ast}$ \todo{check the use of hat}.
+In particular, the modified features are summarized into run-level summaries $\hat{y}_{ij}^{\ast}$.
 The estimated summaries of the modified features are then used as the input to the appropriate model in $MSstats$ or $MSstatsTMT$,  such as in \eqref{msstats_model}.
@@
 The p-values of the comparison are adjusted for multiple testing using the approach by Benjamin and Hochberg~\cite{Benjamini:1995}.
-
-\todo{This paragraph relates more to the implementation.
-Could you integrate it with the implementation section?} By leveraging the implementations in $MSstats$ and $MSstatsTMT$, the proposed approach is very versatile.
-It is applicable to a wide variety of experimental designs, including group comparison, paired designs, time course designs and unbalanced designs.
-It is applicable to label-free data acquisition strategies such as DDA and DIA and experiments with TMT labeling.
-The proposed approach is also applicable to experiments where the experimental designs for PTM profiling and global proteome profiling vary in properties such as number of biological replicates, data acquisition strategies and runs.
 
@@
 
-\todo{I suggest that you follow the same order as Figure 3 (my suggested version), one paragraph per step.
-Describe each step, and refer to both the figure, the quantity in the equations and the MSstats functions.
-Specifically: design; data acquisition; data processing and requirements; converters; steps of statistical analysis.
-I would state upfront that all the steps before last are performed separately for each data type, and not repeat it in later paragraphs}
-
 The proposed methods are implemented in the open source R package $MSstatsPTM$, overviewed in~\figref{msstatsptm_design}.
-$MSstatsPTM$ is applicable to most experimental designs, label-free data acquisitions such as DDA, DIA, SRM and label-based data acquisitions such as TMT.
-$MSstatsPTM$ takes as input lists of identified and quantified spectral features, produced by spectral processing tools such as MaxQuant, Progenesis, or Spectronaut.
-
-$MSstatsPTM$ requires the processing tool to identify the amino acid in the protein sequence where the modification occurred (i.e.
+By leveraging the implementations in $MSstats$ and $MSstatsTMT$, the proposed approach is very versatile.
+It is applicable to a wide variety of experimental designs, including group comparison, paired designs, time course designs and unbalanced designs.
+It is applicable to label-free data acquisitions such as DDA, DIA, SRM and label-based data acquisitions such as TMT.
+It can model experiments where the experimental designs for PTM profiling and global proteome profiling vary in properties such as number of biological replicates, data acquisition strategies and runs.
+
+$MSstatsPTM$ takes as input lists of identified and quantified spectral features, produced by spectral processing tools, such as MaxQuant, Progenesis, or Spectronaut (step 3 of~\figref{msstatsptm_design}).
+Conversion is performed separately for the runs enriched in modified peptides and for the global profiling runs.
+We require the processing tool to identify the amino acid in the protein sequence where the modification occurred (i.e.
 the modification site).
 This will generally include the amino acid abbreviation, plus it's number in the protein sequence.
-For example, a modification on a 70th amino acid in the sequence serine should be marked as ``S70''.
-
+For example, a modification on a 70th amino acid in the sequence, serine should be marked as ``S70''.
 Occasionally the outputs of data processing tools only include the peptide sequence with the modified amino acid highlighted, without indicating where in the protein sequence the modification occurred.
-For these cases $MSstatsPTM$ includes functionality for identifying this location \todo{what do you mean by 'manually'?
-The users have to manually input it?} manually, given the modified peptide sequence and a FASTA file with the entire protein sequence.
-The package includes converters for seamless integration \todo{integration of what?} with multiple such tools.
-\todo{Maybe put all the discussion of converters in one place - integrate with the first paragraph?
-} The converters can optionally keep peptides with both single and multiple modification sites, or filter out only those with multiple sites.
-If multiple sites are kept, these are grouped separately from the singly modified peptide so as not to confound the effect \todo{confound how?
-The effect of what?
-We use confounding to indicate the changes in protein abundance.
-This looks like a different confounding.
-Could you clarify?}.
-Additionally, peptides with oxidation can also be removed if the effect is believed to confound the modifications of interest \todo{confound how?
-The effect of what?}.
-For experimental designs that include separate runs enriched in modified peptides and global profiling runs, conversion is performed separately for the runs enriched in modified peptides and for the global profiling runs.
-
-The next step is PTM/protein summarization.
+For these cases $MSstatsPTM$ includes functionality for identifying the location, given the modified peptide sequence and a FASTA file with the entire protein sequence.
+The converters output the modified spectral features $y_{ijk}^{\ast}$ and the corresponding unmodified features $y_{ijk}$ in the format required for summarization.
+
+The next step is PTM/protein summarization using the $dataSummarizationPTM()$ function (step 4 of~\figref{msstatsptm_design}).
 Summarization is performed separately for the PTM and unmodified protein.
@@
 These issues can violate the assumptions underlying the imputation, and lead to numerically unstable results.
-Next, separate statistical models are fit to both the summaries of the modified features and the to summaries of the unmodified features.
+The outputs of this step are the run-level summaries for the modified, $\hat{y}_{ij}^{\ast}$, and unmodified, $\hat{y}_{ij}$, features.
+
+Finally, separate statistical models are fit to both feature summaries using the $groupComparisonPTM()$ function (step 5 of~\figref{msstatsptm_design}).
 The models are automatically selected to reflect the experimental design and the data acquisition.
@@
 due to missing data, a simplified model is fit.
-
+The output of the models are the estimates $\hat{\Delta}_{PTM}$ and $\hat{\Delta}_{protein}$, and their standard errors $\widehat{SE}(\hat{\Delta}_{PTM})$ and $\widehat{SE}(\hat{\Delta}_{protein})$.
 After modeling, the modified model is adjusted for changes in unmodified protein abundance, using the methods described above.
@@
 In this case the implementation reverts to testing the null hypothesis in~\eqref{conv_null_hyp} using the statistical methods seen in $MSstats$, applied separately to each modified peptide.
-In addition to the statistical functionalities, the implementation includes visualizations for quality control and assessment of the quality of model fit.
+The final output is the estimate $\hat{\Delta}_{adj}$ and it's standard error $\widehat{SE}(\hat{\Delta}_{adj})$.
+
+In addition to the above functionalities, the implementation includes visualizations for quality control, $dataProcessPlotsPTM()$, and assessment of the quality of model fit, $groupComparisonPlotsPTM()$.
 
@@
 \includegraphics[scale=.5]{images/MSstatsPTM_design.png}
-\caption{\todo{Looks great.
-Maybe to the left add vertical text 'Upstream experiments and computation' and  'MSstatsPTM' next to the corresponding boxes.
-No need for a capital L in lysate.
-'Group comparison among PTM' and 'Group comparison among unmodified proteins'.
-I would add () next to the function names, to clarify that these are functions} The $MSstatsPTM$ workflow.
+\caption{The $MSstatsPTM$ workflow.
 The names of the R functions used for each step are highlighted in purple and the output notations are highlighted in red.
+The workflow begins with the acquisition of the enriched and global proteome lysates.
 The package is applicable to label-free data acquisitions such as DDA, DIA, SRM and label-based data acquisitions such as TMT.
@@
 At the end, the model-based summaries are combined to adjust the changes in the PTM abundance for changes in unmodified protein abundance.
-Data visualizations can optionally be created at each step.}
+Data visualizations can optionally be created after steps 4 and 5 of the workflow.}
 \label{fig:msstatsptm_design}
@@
 \end{subfigure}
-\caption{\todo{In the figure titles and in the caption, explain which simulation is 'clean' and which is 'noisy'.
-Use the same names for the datasets as in the Datasets section throughout.
-ANOVA is an abbreviation and needs all capital letters - could you check throughout in the figures} Dataset 1 \& 2: Computer simulation.
-a) All the considered methods in the first computer simulation correctly calibrated FDR when adjusting for changes in protein abundance.
+\caption{Dataset 1 \& 2: Computer simulation.
+a) In a clean simulation, all the considered methods in the first computer simulation correctly calibrated FDR when adjusting for changes in protein abundance.
 In comparison, the methods without accounting for the protein-level changes resulted in off-target, high false positive rates.
-b) The advantage of using the proposed approach was apparent when including limited observations and missing values.
+b) In a noisy simulation, including limited feature observations and missing values, the advantage of the proposed approach was apparent.
 Looking at accuracy, the proposed method outperformed ANOVA and Limma in nearly every model.

### 2022-07-28 supplementary/ptm_sm.tex
addressed all comments in main

@@
 
-\begin{figure}[h!]
-\centering
-\includegraphics[width=.85\textwidth]{sim_new/spike_in_statistic_comparison}
-\caption{Comparing the summary statistics between methods, the proposed method performed the strongest.
+%\begin{figure}[h!]
+%\centering
+%\includegraphics[width=.85\textwidth]{sim_new/spike_in_statistic_comparison}
+%\caption{Comparing the summary statistics between methods, the proposed method performed the strongest.
 In terms of accuracy and specificity the three methods were close, with Limma and Anova showing slightly higher values.
@@
 \label{fig:spike_stat}}
-\end{figure}
+%\end{figure}
 

### 2022-07-28 main/MSstatsPTM_Main.tex
moved summation anova section

@@
 
-The basic Analysis of Variance (ANOVA) model~\cite{kutner_etal_04a}  is
+First run-level summarization of modified features is performed.
+This is generally done by taking the log sum of the original modified features.
+For example, on the log$_2$ scale for condition $i$ and biological replicate $j$, the summarization would be
+
+\begin{eqnarray}
+\hat{y}^*_{ij} =  \log_2 \sum_{k=1}^{K} 2^{y_{ijk}^{\ast}}
+\label{eq:anova_summ}
+\end{eqnarray}
+
+Then the basic Analysis of Variance (ANOVA) model~\cite{kutner_etal_04a}  is
 \begin{eqnarray}
@@
 \label{eq:ttest_model}
-\end{eqnarray}
-
-Where $\hat{u}^*_{ij}$ is the run-level summarization of modified features. For example, on the log$_2$ scale for condition $i$ and biological replicate $j$, $\hat{u}^*_{ij}$ is
-
-\begin{eqnarray}
-\hat{u}^*_{ij} =  \log_2 \sum_{k=1}^{K} 2^{y_{ijk}^{\ast}}
-\label{eq:anova_summ}
 \end{eqnarray}
@@
 \begin{eqnarray}
-\hat{u}_{ij} = \log_2 \frac{ \sum_{k=1}^{K} 2^{y_{ijk}^{\ast}} }{ \sum_{l=1}^{L} 2^{y_{ijl}} } = \log_2 \left( \sum_{k=1}^{K} 2^{y_{ijk}^{\ast}} \right) - \log_2 \left( \sum_{l=1}^{L} 2^{y_{ijl}} \right)
+\hat{y}_{ij} = \log_2 \frac{ \sum_{k=1}^{K} 2^{y_{ijk}^{\ast}} }{ \sum_{l=1}^{L} 2^{y_{ijl}} } = \log_2 \left( \sum_{k=1}^{K} 2^{y_{ijk}^{\ast}} \right) - \log_2 \left( \sum_{l=1}^{L} 2^{y_{ijl}} \right)
 \label{eq:adj_estimation}
@@
 \begin{eqnarray}
-\hat{u}_{ij} = (\mu^{\ast}_{i}-\mu_i) + \epsilon^{\prime}_{ij}, \text{where } \epsilon^{\prime}_{ij} \mathop\sim^{iid} \mathcal{N}(0, \sigma^{\prime 2}),\ i=1,\ldots,I,\ j=1,\ldots,J
+\hat{y}_{ij} = (\mu^{\ast}_{i}-\mu_i) + \epsilon^{\prime}_{ij}, \text{where } \epsilon^{\prime}_{ij} \mathop\sim^{iid} \mathcal{N}(0, \sigma^{\prime 2}),\ i=1,\ldots,I,\ j=1,\ldots,J
 \label{eq:ttest_model}

### 2022-07-28 main/MSstatsPTM_Main.tex
minor notation fix
update supp formulas, removed supp text

@@
 The second simulation introduced real-world characteristics, such as limited modified features and missing values.
-Details of computer simulations are available in Supplementary Section \ref{supp-sec:sim} and on GitHub (\url{https://github.com/devonjkohler/MSstatsPTM_simulations}).
+Details of computer simulations are available in {\bf Supplementary Sec.
+\ref{supp-sec:sim}} and on GitHub (\url{https://github.com/devonjkohler/MSstatsPTM_simulations}).
 
@@
 \begin{eqnarray}
-\hat{u}^*_{ij} = \mu^{*}_{i} + \epsilon^{\ast}_{ij}, \ \epsilon^{\ast}_{ij} \mathop\sim^{iid} \mathcal{N}(0, \sigma^{*2}),\ i=1,\ldots,I,\ j=1,\ldots,J
+\hat{y}^*_{ij} = \mu^{*}_{i} + \epsilon^{\ast}_{ij}, \ \epsilon^{\ast}_{ij} \mathop\sim^{iid} \mathcal{N}(0, \sigma^{*2}),\ i=1,\ldots,I,\ j=1,\ldots,J
 \label{eq:ttest_model}
@@
 \ref{supp-sec:complex_design}}.
+A power analysis of different sample sizes can be seen in {\bf Supplementary Sec.
+\ref{supp-sec:sample_size_analysis}}.
 

### 2022-07-28 supplementary/ptm_sm.tex
update supp formulas, removed supp text

@@
 
-The proposed statistical approach was evaluated and compared to the Anova and Limma methods using computer simulations.
-Specifically, their statistical properties with respect to protein-level adjustment and real world experimental conditions were evaluated.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection{Computer simulation}
-\label{sec:comp_sim}
-Differential levels of modified peptides may be due to differential modifications and/or changes in protein abundance.
-Approaches without considering protein-level adjustment lose track of an important aspect in interpreting observed changes in PTM abundance, which may result in misleading conclusions.
-To highlight the necessity of the adjustment, we compared six different approaches as follows: a) proposed approach, b) proposed approach without adjusting for unmodified peptides c) Anova (with adjustment), d) Anova (no adjustment), e) Limma (with adjustment), and f) Limma (no adjustment).
-
-In experiments of complex designs, multiple inter-related conditions are often compared together.
-All comparisons were made with the same fold change between conditions ($1$).
-This ensured that all peptides were differentially abundant by the same amount and the only difference in calculations was using more conditions in the modeling.
+We compared six different approaches as follows: a) proposed approach, b) proposed approach without adjusting for unmodified peptides c) ANOVA (with adjustment), d) ANOVA (no adjustment), e) Limma (with adjustment), and f) Limma (no adjustment).
 
@@
 \label{sec:dataset1}
-In the first simulation an experiment with many features per PTM and unmodified protein was created.
-Additionally this simulation contained no missing data.
-These attributes are not representative of a real experiment, but provide a baseline for model performance.
+
+In the first simulation an experiment with many features per PTM and unmodified protein was created. Additionally this simulation contained no missing data.
 
@@
 
-The results are summarized in Figure \ref{fig:sim1_recall}.
-
 \subsubsection{Dataset 2 : Computer simulation 2 - Label-free missing values and low features}
@@
 In the second simulation we introduced limited feature observations per PTM as well as masking a portion of the observation to simulate missing values.
-This is more in line with what we would expect in a biological experiment and provides a more realistic expectation of model performance.
-The number of PTM features and missing data percentage where determined by looking at the features and missing data in the biological experiments in this paper.
 
@@
 \end{itemize}
-
-The results are summarized from Figure \ref{fig:sim2_recall} to Figure \ref{fig:fc_boxplot}.
 
@@
 The proposed method in this simulation had a visibly tighter distribution around the median.
-Both Limma and Anova showed a wider range around the fold change.}
+Both Limma and ANOVA showed a wider range around the fold change.}
  \end{subfigure}
@@
 
-Again we consider three different methods and assess their performance: the method proposed in this paper, Limma, and Anova.
-All methods are analyzed after adjusting for changes in overall protein level.
-The results are summarized from Figure \ref{fig:spike_volcano_msstats} to Figure \ref{fig:spike_stat}, including volcano plots, model summary statistics, and fold change analysis.
-
 \begin{figure}[h!]
@@
 \includegraphics[width=1\textwidth]{sim_new/spike_in_anova_volcano}
-\caption{Using Anova the fold change of the spike-in peptides was much closer to expectation after adjusting for global protein abundance.
+\caption{Using ANOVA the fold change of the spike-in peptides was much closer to expectation after adjusting for global protein abundance.
 The log FC estimation was the same as Limma, however the pvalues were different.
@@
 
-$$Y_{mcb} = \mu + Condition_c + Subject_{mcb} + \epsilon_{mcb}$$
+$$Y_{mij} = \mu_i + \epsilon_{mij}$$
 
-$$\sum_{c=1}^C{Condition_c} = 0 ,\: Subject_{mcb} ~ \sim N(0, \sigma^2_S) ,\: \epsilon_{mcb} ~ \sim N(0, \sigma^2)$$
-
-The results of the proposed method to this experiment can be seen in Figure \ref{fig:ipah_figures}.
+$$\sum_{i=1}^I{\mu_i} = 0 ,\: \epsilon_{mij} ~ \sim N(0, \sigma^2)$$
 
@@
 
-In this study, the correlation between the gene Atg16L1 and killing of Shigella flexneri (S.flexneri) was assessed~\cite{Maculins}.
-Multiplex proteomics was used to quantify the abundance of total protein, phosphorylation, and ubiquitination in wild type (WT) and ATG16L1-deficient (cKO) samples, uninfected and uninfected with S.flexneri.
-The abundance of total protein and post-translation modifications were quantified at three time points, uninfected, early infection (45-60 minutes), and late infection (3-3.5 hours).
-Quantifying the total protein along with the post-translational modifications allowed us to adjust for changes in total protein and see the true impact of the site specific modifications.
-Two mixtures using 11-plex were ran over the six conditions.
-The six conditions were split between 11 channels leading to an unbalanced experimental design.
-Each mixture contained two replicates per early and late WT and KO conditions.
-Mixture one contained one replicate of uninfected WT and two replicates of uninfected KO.
-Mixture two contained one replicate of uninfected KO and two uninfected WT.
 The experimental design can be seen in Table~\ref{table:shigella_design}.
@@
 
-A model was fit for the total protein, phosphorylation, and ubiquitination separately, as described previously for TMT experiments. The model formula can be seen below.
+A model was fit for the total protein, and phosphorylation. The model formula can be seen below.
 
-$$Y_{mcb} = \mu + Mixture_m + Condition_c + Subject_{mcb} + \epsilon_{mcb}$$
+$$Y_{mij} = \mu_i + Mixture_m + \epsilon_{mij}$$
 
-$$Mixture_m ~ \sim N(0, \sigma^2_M) ,\: \sum_{c=1}^C{Condition_c} = 0 ,\: Subject_{mcb} ~ \sim N(0, \sigma^2_S) ,\: \epsilon_{mcb} ~ \sim N(0, \sigma^2)$$
-
-The results of the proposed method to this experiment can be seen in Figure \ref{fig:No_Diff_Shigella_PTM} and Figure \ref{fig:Diff_Shigella_PTM}.
+$$Mixture_m ~ \sim N(0, \sigma^2_M) ,\: \sum_{i=1}^I{\mu_i} = 0 ,\: \epsilon_{mij} ~ \sim N(0, \sigma^2)$$
 
@@
 
-This experiment looked into the relationship between USP30 and protein kinase PINK1, and their association with Parkinson's Disease.
-Ubiquitination site profiling was performed and the modified site abundance was analyzed.
-Four conditions were tested with two biological replicates per condition.
-The conditions were as follows: CCCP, USP30 overexpression (USP30\_OE), Combo, and Control.
-Label-free mass spectrometry quantification was used to quantify the abundance of modified peptides.
-A corresponding mixed effects model was fit per modification and global protein as described previously in this supplementary.
-The experiment was modeled as a group comparison.
-
-In contrast to other experiments analyzed in this paper, there was no unmodified global protein profiling run performed in this experiment.
-Once identification and quantification of the Ubiquitinated profiling was performed, peptides which were unmodified were extracted and used in place of a global profiling run.
-This resulted in a significant lack of overlap between modified and unmodified peptides and low feature counts for the unmodified protein model.
-
 An example profile plot for this experiment can be seen in  Figure \ref{fig:USP30_profile_plot}.
@@
 \end{figure}
+
 \clearpage
@@
 \section{Sample size calculation and power analysis}
+\label{sec:sample_size_analysis}
 

### 2022-07-28 supplementary/ptm_sm.tex
added ref to supp

@@
 
-We compared six different approaches as follows: a) proposed approach, b) proposed approach without adjusting for unmodified peptides c) ANOVA (with adjustment), d) ANOVA (no adjustment), e) Limma (with adjustment), and f) Limma (no adjustment).
+We compared six different approaches as follows: a) proposed approach, b) proposed approach without adjusting for unmodified peptides c) ANOVA~\cite{kutner_etal_04a} (with adjustment), d) ANOVA (no adjustment), e) Limma~\cite{Smyth:2004} (with adjustment), and f) Limma (no adjustment).
 
@@
 \clearpage
-\subsubsection{Dataset 5 : Mouse - Phosphorylation - 2mix-TMT}
+\subsubsection{Dataset 5 : Mouse - Phosphorylation - 2mix-TMT~\cite{Maculins}}
 \label{sec:shigella}
@@
 
-\subsubsection{Dataset 6 : Human - Ubiquitination - Label-free no global profiling run}
+\subsubsection{Dataset 6 : Human - Ubiquitination - Label-free no global profiling run~\cite{Cunningham2015}}
 \label{sec:usp30}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\clearpage
+\clearpage
 
-%\printbibliography
+\printbibliography
 

### 2022-07-29 main/MSstatsPTM_Main.tex
format typo
minor typo in implementation
Methods revisited

@@
 
-\noindent {\bf ANOVA on modified log$_2$-intensities}
+\noindent {\bf ANOVA on summarized modified log$_2$-intensities}
 
@@
 
-First run-level summarization of modified features is performed.
-This is generally done by taking the log sum of the original modified features.
-For example, on the log$_2$ scale for condition $i$ and biological replicate $j$, the summarization would be
-
-\begin{eqnarray}
-\hat{y}^*_{ij} =  \log_2 \sum_{k=1}^{K} 2^{y_{ijk}^{\ast}}
+Analysis of Variance (ANOVA)~\cite{kutner_etal_04a} is the simplest statistical model for summarized modified features in each biological replicate.
+The summarization often consists of averaging (or taking the median or other robust summary) of the log$_2$ intensities of the modified features in each replicate, e.g.
+$\hat{y}^*_{ij} = \sum_{k=1}^K{y_{ijk}^{\ast}}/K$.
+Alternatively, summarization sums the intensities of the modified features on the original scale, and then takes the log
+\begin{eqnarray}
+\hat{y}^*_{ij} =  \log_2 \left(\sum_{k=1}^{K} 2^{y_{ijk}^{\ast}}\right)
 \label{eq:anova_summ}
 \end{eqnarray}
-
-Then the basic Analysis of Variance (ANOVA) model~\cite{kutner_etal_04a}  is
+The basic ANOVA model  is then
 \begin{eqnarray}
@@
 \end{eqnarray}
-
-The model allows us to estimate $\hat{\Delta}$ and its standard error, and test the null hypothesis in \eqref{conv_null_hyp} by comparing the model-based test statistic against the Student distribution with $df=I(J-1)$ degrees of freedom in balanced designs.
+The model allows us to estimate $\hat{\Delta}$ and its standard error.
+The estimates are used to test the null hypothesis in \eqref{conv_null_hyp} by comparing the model-based test statistic against the Student distribution with $df=I(J-1)$ degrees of freedom in balanced designs.
 Unfortunately, this approach is fundamentally flawed as it does not account for the confounding between changes in the PTM abundance and the overall changes in the abundance of the unmodified protein.
 
-%\medskip \noindent
+\medskip \noindent
 
@@
 The basic ANOVA can be extended to account for the confounding of changes in PTM abundance and overall changes in protein abundance~\cite{Schwammle2015,THOMAS2020,Mertins:2013}.
-Typically this is done by first summarizing the feature intensities in each run, separately for modified and unmodified features, e.g.
-with the sum on the original scale, and considering sample-wise ratios of the summaries.
-On the log$_2$ scale, for condition $i$ and biological replicate $j$ the approach is equivalent to calculating
-\begin{eqnarray}
-\hat{y}_{ij} = \log_2 \frac{ \sum_{k=1}^{K} 2^{y_{ijk}^{\ast}} }{ \sum_{l=1}^{L} 2^{y_{ijl}} } = \log_2 \left( \sum_{k=1}^{K} 2^{y_{ijk}^{\ast}} \right) - \log_2 \left( \sum_{l=1}^{L} 2^{y_{ijl}} \right)
+Typically this is done by first calculating sums of the intensities of the modified and unmodified features on the original scale, and then considering replicate-wise ratios of the sums and taking the log
+\begin{eqnarray}
+u_{ij} = \log_2 \frac{ \sum_{k=1}^{K} 2^{y_{ijk}^{\ast}} }{ \sum_{l=1}^{L} 2^{y_{ijl}} } = \log_2 \left( \sum_{k=1}^{K} 2^{y_{ijk}^{\ast}} \right) - \log_2 \left( \sum_{l=1}^{L} 2^{y_{ijl}} \right)
 \label{eq:adj_estimation}
@@
 \begin{eqnarray}
-\hat{y}_{ij} = (\mu^{\ast}_{i}-\mu_i) + \epsilon^{\prime}_{ij}, \text{where } \epsilon^{\prime}_{ij} \mathop\sim^{iid} \mathcal{N}(0, \sigma^{\prime 2}),\ i=1,\ldots,I,\ j=1,\ldots,J
+u_{ij} = (\mu^{\ast}_{i}-\mu_i) + \epsilon^{\prime}_{ij}, \text{where } \epsilon^{\prime}_{ij} \mathop\sim^{iid} \mathcal{N}(0, \sigma^{\prime 2}),\ i=1,\ldots,I,\ j=1,\ldots,J
 \label{eq:ttest_model}
@@
 The prior distributions are estimated from the data across all the proteins and all the PTM as described in \cite{Smyth:2004}. 
-With this approach, testing the null hypothesis in~\eqref{null_hyp} amounts to combining the PTM- and protein-specific estimates of variation with a consensus estimate obtained from all the PTM and all the proteins.
+With this approach, testing the null hypothesis in~\eqref{null_hyp} is enhanced by combining the PTM- and protein-specific estimates of variation with a consensus estimate obtained from all the PTM and all the proteins.
 As the result, in experiments with few biological replicates the standard errors are often smaller, and the degrees of freedom are often larger than without moderation~\cite{Ritchie_15a}. 
@@
 
-Since Limma only improves upon the estimation of variation, it has the same limitations in terms of the experimental designs.
-Furthermore, the method is only directly applicable to experiments with at most two variance components, and cannot account for all the sources of variation in more complex experiments.
+Since Limma only improves upon the estimation of variation, its limitations are similar to those of ANOVA.
+The method is only directly applicable to experiments with at most two variance components, and cannot account for all the sources of variation in experiments with either isobaric labeling or complex designs.
 There is no self contained implementation of the methods to PTMs, requiring manual transformation and application by the user.
@@
 
-\subsection*{Statistical modeling and parameter estimation in MSstats}
+\subsection*{Relative protein quantification in MSstats}
 
@@
 The summaries are then used as input to fit a flexible family of linear mixed-effects models \cite{McLean:1991, Faraway:2006, Bolker2009}.
-The specific model fit for the individual protein depends on the design of the experiment, labeling type and data acquisition type, as well as on the pattern of missing values in that particular protein.
-For example, the unmodified protein features in \figref{data-structure} are modeled as
-
+The models are fit separately for each protein.
+The specific model depends on the design of the experiment, labeling type and data acquisition type as summarized in {\bf Supplementary Figure~\ref{supp-fig:statistical_inference}}.
+For example, the unmodified protein features in the simple design in \figref{data-structure} are modeled with one-way ANOVA
 \begin{eqnarray}
@@
 \end{eqnarray}
-
-In the simplified case of \figref{data-structure} this amounts to the same model as in \eqref{ttest_model}.
-However, the model can be extended to more complex cases with additional sources of variation, which simpler models, such as ANOVA, cannot account for.
-For example, given a group comparison TMT experiment, with the addition of mixture as a source of variation, the model would extend to
-
+In contrast, a group comparison experiment with multiple TMT mixtures is modeled as
 \begin{eqnarray}
@@
 \end{eqnarray}
-
-Additional models for different experimental designs, including time series, can be seen in {\bf Supplementary Figure~\ref{supp-fig:statistical_inference}}.
+Moreover, the model fit for a particular protein depends on the pattern of missing values in that protein.
+If some of the terms of the model reflecting the experimental design are not estimable, a simpler model is fit for that protein instead.
 
 Parameters of the model are estimated using restricted maximum likelihood (REML)~\cite{Kenward}.
-In turn, the parameters allow us to estimate the pairwise comparison $\hat{\Delta}_{Protein} = \hat{\mu}_{i} - \hat{\mu}_{i^{\prime}}$ and its standard error.
-Similarly to $Limma$, $MSstatsTMT$ includes an optional Empirical Bayes moderation of the standard error~\cite{Huang:2020}, which increases the sensitivity of detecting differential abundance when the number of biological replicates in each condition is small.
-
-In addition to detecting differential abundance at the protein level, $MSstats$ and $MSstatsTMT$ can be also used at the feature or at the modification cite levels.
+The parameters allow us to estimate the pairwise comparison $\hat{\Delta}_{Protein} = \hat{\mu}_{i} - \hat{\mu}_{i^{\prime}}$ and its standard error.
+Similarly to $Limma$, $MSstatsTMT$ includes an optional Empirical Bayes moderation of the standard error~\cite{Huang:2020},  increasing the sensitivity of detecting differential abundance when the number of biological replicates in each condition is small.
+
+$MSstats$ and $MSstatsTMT$ can  also be used at the feature or at the modification cite level, as opposed to protein level.
 For example, summarizing the features per PTM cite instead of per protein, the approach allows us to test the null hypothesis in \eqref{conv_null_hyp}.
@@
 The $MSstats$ framework has a number advantages over the methods above. 
-First, unlike ANOVA and $Limma$, $MSstats$ and $MSstatsTMT$ are applicable to many complex experimental designs, including designs with multiple sources of variation, and unbalanced designs. 
-Second, the approach is applicable to various data acquisition types, including label-free DDA and DIA, and experiments with TMT labeling, and is compatible with various data processing tools such as Skyline, Spectronaut, MaxQuant, Progenesis, Proteome Discoverer, and OpenMS.
-Third, the run-level summarization has been shown to be more robust to outliers and missing values as compared to feature averages~\cite{Choi:2014}. 
+First, unlike ANOVA and $Limma$, $MSstats$ and $MSstatsTMT$ are applicable to arbitrary complex experimental designs, including designs with multiple sources of variation, and unbalanced designs. 
+Second, the approach is applicable to various data acquisition types, including label-free DDA and DIA, and experiments with TMT labeling. 
+Third, the $MSstats$ packages are compatible with various data processing tools such as Skyline, Spectronaut, MaxQuant, Progenesis, Proteome Discoverer, and OpenMS. 
 Finally, the custom $MSstats$ and $MSstatsTMT$ implementation accounts for potential data artifacts, is numerically scalable and stable, and is available through both command line and a dedicated graphical user interface.
@@
 It does not account for the confounding between the changes in PTM abundance and the overall changes in protein abundance.
-This manuscript proposes a simple extension to the methodology in $MSstats$ and $MSstatsTMT$, to enable testing of the relevant the null hypothesis in~\eqref{null_hyp}.
+This manuscript proposes a simple extension to the methodology in $MSstats$ and $MSstatsTMT$, to test the null hypothesis in~\eqref{null_hyp}.
 
@@
 After modeling, the modified model is adjusted for changes in unmodified protein abundance, using the methods described above.
-Modification sites for which lacking the corresponding global profiling information cannot be adjusted for changes in protein abundance.
+Modification sites which lack corresponding global profiling information cannot be adjusted for changes in protein abundance.
 In this case the implementation reverts to testing the null hypothesis in~\eqref{conv_null_hyp} using the statistical methods seen in $MSstats$, applied separately to each modified peptide.

### 2022-07-29 main/MSstatsPTM_Main.tex
Proposed methods

@@
 \medskip \noindent
-The method takes as input the modified spectral features $y_{ijk}^{\ast}$, and the corresponding unmodified features $y_{ijk}$.
+The overall statistical analysis workflow and its implementation are summarized in~\figref{msstatsptm_design}.
+$MSstatsPTM$ takes as input the modified spectral features $y_{ijk}^{\ast}$, and the corresponding unmodified features $y_{ijk}$.
 Ideally, the modified features are acquired separately, after an enrichment to maximize the information content in the resulting dataset, and the unmodified features are acquired separately as part of a global proteome profiling.
@@
 
-We leverage the existing methods in $MSstats$ and $MSstatsTMT$, and extend them to remove confounding between changes in the PTM and the overall changes in protein abundance.
 Each feature type is first analyzed separately using $MSstats$ or $MSstatsTMT$.
 In particular, the modified features are summarized into run-level summaries $\hat{y}_{ij}^{\ast}$.
-The estimated summaries of the modified features are then used as the input to the appropriate model in $MSstats$ or $MSstatsTMT$,  such as in \eqref{msstats_model}.
+The estimated summaries of the modified features are then used as the input to the appropriate model in $MSstats$ or $MSstatsTMT$,  such as in~\eqref{msstats_model} or~\eqref{msstatstmt_model} \todo{Could we show in Fig 3 where $MSstats$ or $MSstatsTMT$ come into place?}.
 The resulting model-based estimates include $\hat{\Delta}_{PTM}=\hat{\mu}^{\ast}_{i}-\hat{\mu}_{i^{\prime}}^{\ast}$, and its standard error $\widehat{SE}(\hat{\Delta}_{PTM})$.
@@
 The p-values of the comparison are adjusted for multiple testing using the approach by Benjamin and Hochberg~\cite{Benjamini:1995}.
+\todo{Add 'df' to Step 5 of Fig 3?}
 
@@
 The proposed statistical framework enables sample size calculation for future PTM experiments. 
-The procedure has been described in general in~\cite{kutner_etal_04a}, and specifically for protein significance analysis in~\cite{oberg_vitek_09a}. 
+The procedure has been described in general in~\cite{kutner_etal_04a}, and for protein significance analysis specifically in~\cite{oberg_vitek_09a}. 
 It requires us to specify the desired levels of the following quantities: a) $q$, the False Discovery Rate of detecting differential abundance, b) $\beta$, the average Type II error rate, c) $\Delta_{adj}$, the minimal log$_2$-fold change in adjusted PTM abundance  of interest, d) $m_0 / (m_0 + m_1)$, the fraction of truly differentially modified PTM sites in the comparison, and e) $\sigma_{PTM}^{2}$ and $\sigma_{protein}^{2}$, the anticipated variances associated with the summaries of the modified and unmodified features, respectively.
@@
 SE(\Delta_{adj})^2 = \left[ \frac{2}{J} \left( \sigma_{PTM}^{2} + \sigma_{protein}^{2} \right) \right]
-\leq \left( \frac{\Delta_{adj}}{t_{1-\beta, df} + t_{1-\alpha /2, df}} \right)^{2}
+\leq \left( \frac{\Delta_{adj}}{z_{1-\beta} + z_{1-\alpha /2}} \right)^{2}
 \end{equation}
@@
 \end{equation}
-and $t_{1-\beta, df}$ and $t_{1-\alpha /2, df}$ are the $100(1-\beta)^{\text{th}}$ and the $100(1-\alpha /2)^{\text{th}}$ percentiles of the Student distribution with $df = I(J-1)$ degrees of freedom.
+and $z_{1-\beta}$ and $z_{1-\alpha /2}$ are the $100(1-\beta)^{\text{th}}$ and the $100(1-\alpha /2)^{\text{th}}$ percentiles of the Standard Normal distribution.
 Solving for $J$, the number of biological replicates per condition is
 \begin{equation}
-J \geq \frac{(2\sigma_{PTM}^{2} + 2\sigma_{protein}^{2})(t_{1-\beta, df} + t_{1-\alpha /2, df})^2}{\Delta_{adj}^2}
+J \geq \frac{(2\sigma_{PTM}^{2} + 2\sigma_{protein}^{2})(z_{1-\beta} + z_{1-\alpha /2})^2}{\Delta_{adj}^2}
 \end{equation}
-Since the numerator has two sources of variation, reflecting a larger uncertainty in the adjusted calculation, the adjustment typically requires a larger sample size to gain the same sensitivity as the unadjusted estimation.
+The numerator has two sources of variation, reflecting a larger uncertainty in the adjusted calculation.
+Therefore, the adjustment typically requires a larger sample size to gain the same sensitivity as the unadjusted estimation.
 Details of applications of this approach to experiments with complex design are in {\bf Supplementary Sec.
 \ref{supp-sec:complex_design}}.
-A power analysis of different sample sizes can be seen in {\bf Supplementary Sec.
+Examples of power analyses are in {\bf Supplementary Sec.
 \ref{supp-sec:sample_size_analysis}}.
@@
 
-We proposed a general statistical modeling framework and implementation for PTM characterization.
+\todo{When you have a chance, could you do minor clean ups of the references?
+Remove the doi; check that the same journal has the same abbreviation in all the references; keep the reference format consistent (e.g., remove capitalization in article titles; italic for book/volume titles etc)} We proposed a general statistical modeling framework and implementation for PTM characterization.
 The framework is designed for bottom-up MS workflows, which are characterized with variations from multiple confounded sources, frequent missing data, and associated uncertainty in the conclusions.
@@
 \includegraphics[scale=.5]{images/MSstatsPTM_design.png}
-\caption{The $MSstatsPTM$ workflow.
+\caption{\todo{Where does the sample size calculation fit into that?} \todo{Could you rotate the 'Upstream...' Text and 'MSstatsPTM text and place it vertically to the left of the corresponding steps?
+It will be a bit more clear} \todo{It is not very clear which steps leverage MSstats/MSstatsTMT.
+Should we say in relevant places that these are MSstats implementations?} The $MSstatsPTM$ workflow.
 The names of the R functions used for each step are highlighted in purple and the output notations are highlighted in red.

### 2022-07-29 main/MSstatsPTM_Main.tex
Begin evaluations

@@
 
-The proposed methods are implemented in the open source R package $MSstatsPTM$, overviewed in~\figref{msstatsptm_design}.
+The implementation of the open source R package $MSstatsPTM$ is also overviewed in~\figref{msstatsptm_design}.
 By leveraging the implementations in $MSstats$ and $MSstatsTMT$, the proposed approach is very versatile.
@@
 
-$MSstatsPTM$ takes as input lists of identified and quantified spectral features, produced by spectral processing tools, such as MaxQuant, Progenesis, or Spectronaut (step 3 of~\figref{msstatsptm_design}).
-Conversion is performed separately for the runs enriched in modified peptides and for the global profiling runs.
-We require the processing tool to identify the amino acid in the protein sequence where the modification occurred (i.e.
-the modification site).
+$MSstatsPTM$ takes as input lists of identified and quantified spectral features, produced by spectral processing tools such as MaxQuant, Progenesis, or Spectronaut (Step 3 of~\figref{msstatsptm_design}).
+Conversion is performed separately for the runs enriched in modified peptides, and separately for the global profiling runs.
+We require the processing tools to identify the modification site (i.e., the amino acid in the protein sequence where the modification occurred).
 This will generally include the amino acid abbreviation, plus it's number in the protein sequence.
 For example, a modification on a 70th amino acid in the sequence, serine should be marked as ``S70''.
-Occasionally the outputs of data processing tools only include the peptide sequence with the modified amino acid highlighted, without indicating where in the protein sequence the modification occurred.
+Occasionally the outputs of data processing tools only include the peptide sequence with the modified amino acid highlighted, without indicating the location in the protein sequence.
 For these cases $MSstatsPTM$ includes functionality for identifying the location, given the modified peptide sequence and a FASTA file with the entire protein sequence.
@@
 
-The next step is PTM/protein summarization using the $dataSummarizationPTM()$ function (step 4 of~\figref{msstatsptm_design}).
-Summarization is performed separately for the PTM and unmodified protein.
+The next step is PTM/protein summarization using the $dataSummarizationPTM()$ function (Step 4 of~\figref{msstatsptm_design}).
+Summarization is performed separately for the PTM and the unmodified protein.
 When summarizing the PTM, modified peptide features that span the same modification site are summarized together.
@@
 
-Finally, separate statistical models are fit to both feature summaries using the $groupComparisonPTM()$ function (step 5 of~\figref{msstatsptm_design}).
+Finally, separate statistical models are fit to both feature summaries using the $groupComparisonPTM()$ function (Step 5 of~\figref{msstatsptm_design}).
 The models are automatically selected to reflect the experimental design and the data acquisition.
@@
 The output of the models are the estimates $\hat{\Delta}_{PTM}$ and $\hat{\Delta}_{protein}$, and their standard errors $\widehat{SE}(\hat{\Delta}_{PTM})$ and $\widehat{SE}(\hat{\Delta}_{protein})$.
+
 After modeling, the modified model is adjusted for changes in unmodified protein abundance, using the methods described above.
@@
 In this case the implementation reverts to testing the null hypothesis in~\eqref{conv_null_hyp} using the statistical methods seen in $MSstats$, applied separately to each modified peptide.
-The final output is the estimate $\hat{\Delta}_{adj}$ and it's standard error $\widehat{SE}(\hat{\Delta}_{adj})$.
+\todo{This is indicated in the `issues' column in the output?} The final output is the estimate $\hat{\Delta}_{adj}$ and it's standard error $\widehat{SE}(\hat{\Delta}_{adj})$.
 
@@
 
-We compared the performance of  $MSstatsPTM$ to that of ANOVA and of Limma.
+We compared the performance of  $MSstatsPTM$ to that of \todo{ratio-based or both?} ANOVA and of Limma.
 Since IsobarPTM is only applicable to experiments with TMT labeling, it could not be applied to the datasets with known ground truth in this manuscript, and was excluded from the comparisons.
-$MSstatsPTM$ was used without imputing missing values, and without Empirical Bayes moderation.
+\todo{ANOVA and Limma took as input protein summaries XXX}  $MSstatsPTM$ was used without imputing missing values, and without Empirical Bayes moderation.
 All the evaluations were done at the FDR-adjusted p-value cutoff of $q=.05$.
@@
 \end{subfigure}
-\caption{Dataset 1 \& 2: Computer simulation.
-a) In a clean simulation, all the considered methods in the first computer simulation correctly calibrated FDR when adjusting for changes in protein abundance.
-In comparison, the methods without accounting for the protein-level changes resulted in off-target, high false positive rates.
+\caption{\todo{Could you clarify what 'with' and 'without' adjustment means for each model?
+E.g., what is 'proposed with no adjustment'?
+We did not propose anything without adjustment - did you mean base MSstats?
+Also, maybe clarify what you mean by limma with and without adjustment, since we technically didn't introduce limma without the adjustment.
+Maybe explain the inputs to each version of the model?} Computer simulations.
+a) Dataset 1, clean simulation, analyzed to control FDR at at most 5\%.
+The methods not accounting for the confounding between changes in PTM and overall changes in protein abundance produced exceedingly high numbers of false positives.
+In contrast, the methods accounting for the confounding correctly calibrated FDR.
 b) In a noisy simulation, including limited feature observations and missing values, the advantage of the proposed approach was apparent.

### 2022-07-29 main/MSstatsPTM_Main.tex
Small evaluations

@@
 Since IsobarPTM is only applicable to experiments with TMT labeling, it could not be applied to the datasets with known ground truth in this manuscript, and was excluded from the comparisons.
-\todo{ANOVA and Limma took as input protein summaries XXX}  $MSstatsPTM$ was used without imputing missing values, and without Empirical Bayes moderation.
+\todo{ANOVA and Limma took as input protein summaries XXX - define the variants of the same models (e.g., used in Fig 4)}  $MSstatsPTM$ was used without imputing missing values, and without Empirical Bayes moderation.
 All the evaluations were done at the FDR-adjusted p-value cutoff of $q=.05$.

### 2022-07-30 main/MSstatsPTM_Main.tex
More evaluation

--- 2022-07-29 main/MSstatsPTM_Main.tex
@@
 
-\subsection*{Dataset 1: Computer simulation 1 - Label-free}
+\subsection*{Dataset 1: Computer simulation 1 - Label-free clean}
 \label{sec:comp_sim_procedure1}
@@
 
-\subsection*{Dataset 2: Computer simulation 2 - Label-free missing values and low features}
+\subsection*{Dataset 2: Computer simulation 2 - Label-free with few low feature counts and missing values}
 \label{sec:comp_sim_procedure2}
@@
 u_{ij} = (\mu^{\ast}_{i}-\mu_i) + \epsilon^{\prime}_{ij}, \text{where } \epsilon^{\prime}_{ij} \mathop\sim^{iid} \mathcal{N}(0, \sigma^{\prime 2}),\ i=1,\ldots,I,\ j=1,\ldots,J
-\label{eq:ttest_model}
+\label{eq:ttest_ratio_model}
 \end{eqnarray}
@@
 
-\medskip \noindent The estimation of nuisance variation of the ratio-based ANOVA model above is often further expanded with Empirical Bayes moderation implemented in Limma~\cite{ Ritchie_15a, Schwammle2015, Smyth:2004, Smyth:2005, Zhu,Chappell:2021}.
-A typical application of Limma takes as input $u_{ij}$, and for each PTM fits the linear model in~\eqref{ttest_model}.
-The model differs from~\eqref{ttest_model} in that it specifies additional prior distributions for the model parameters.
-The prior distributions are estimated from the data across all the proteins and all the PTM as described in \cite{Smyth:2004}. 
-With this approach, testing the null hypothesis in~\eqref{null_hyp} is enhanced by combining the PTM- and protein-specific estimates of variation with a consensus estimate obtained from all the PTM and all the proteins.
+\medskip \noindent The estimation of nuisance variation of the two ANOVA models above is often further expanded with Empirical Bayes moderation implemented in Limma~\cite{ Ritchie_15a, Schwammle2015, Smyth:2004, Smyth:2005, Zhu,Chappell:2021}.
+A typical application of Limma on summarized modified log$_2$-intensities takes as input $\hat{y}_{ij}$, and for each PTM fits the linear model in~\eqref{ttest_model}. 
+A typical application of ratio-based Limma takes as input $u_{ij}$, and for each PTM fits the linear model in~\eqref{ttest_ratio_model}. 
+The Limma versions of the models differ in that they specifies additional prior distributions for the model parameters. 
+The priors are estimated from the same data by combining the information across all the proteins and all the PTM as described in \cite{Smyth:2004}. 
+With this approach, testing the null hypothesis is enhanced by combining the PTM- and protein-specific estimates of variation with a consensus estimate obtained from all the PTM and all the proteins.
 As the result, in experiments with few biological replicates the standard errors are often smaller, and the degrees of freedom are often larger than without moderation~\cite{Ritchie_15a}. 
@@
 Since Limma only improves upon the estimation of variation, its limitations are similar to those of ANOVA.
-The method is only directly applicable to experiments with at most two variance components, and cannot account for all the sources of variation in experiments with either isobaric labeling or complex designs.
+In particular, the method is only directly applicable to experiments with at most two variance components, and cannot account for all the sources of variation in experiments with either isobaric labeling or complex designs.
 There is no self contained implementation of the methods to PTMs, requiring manual transformation and application by the user.
@@
 The true negatives were defined as PTMs which, after accounting for the changes in the overall protein abundance, were not differentially abundant.
-Additional summaries were the false discovery rate $FDR=FP/(TP+FP)$, $Recall=TP/(TP + FN)$, and $Accuracy=(TP + TN) / (TP + TN + FP + FN)$.
+Additional summaries were the false discovery rate $FDR=FP/(TP+FP)$ \todo{I am not sure we should call this FDR.
+By definition FDR is a population-level quantity (it is an expected value of the proportion of false positives.
+Here it is an observed value).
+Maybe replace with 1-PPV?}, $Recall=TP/(TP + FN)$, and $Accuracy=(TP + TN) / (TP + TN + FP + FN)$.
 For biological experiments with unknown ground truth, we compared the differentially abundant PTMs with and without adjusting for changes in unmodified protein abundance.
 
-We compared the performance of  $MSstatsPTM$ to that of \todo{ratio-based or both?} ANOVA and of Limma.
+We compared the performance of  $MSstatsPTM$ to that of \todo{Need more details on the 'adjusted' vs 'unadjusted' versions ANOVA and Limma.
+Give each version a name that you'll later use in the figures, and describe what type of summaries each version takes.} ANOVA and of Limma.
 Since IsobarPTM is only applicable to experiments with TMT labeling, it could not be applied to the datasets with known ground truth in this manuscript, and was excluded from the comparisons.
-\todo{ANOVA and Limma took as input protein summaries XXX}  $MSstatsPTM$ was used without imputing missing values, and without Empirical Bayes moderation.
+$MSstatsPTM$ \todo{Also describe the 'adjusted' version (MSstatsPTM) and the 'unadjusted' version (I guess it is MSstats/MSstatsTMT)} was used without imputing missing values, and without Empirical Bayes moderation.
 All the evaluations were done at the FDR-adjusted p-value cutoff of $q=.05$.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection*{MSstatsPTM corrected the excess FDR in unadjusted differentially abundant PTMs in clean simulations}
-
-\figref{sim1_fdr} illustrated the high FDR observed on the simulated datasets when not accounting for the overall changes in protein abundance.
-After the adjustment, the FDR of all methods improved, and were in line with the excepted FDR of $.05$.
-Adjusting for changes in the unmodified protein removed differentially abundant PTMs whose abundance change was caused by the protein.
-%Finally, since this simulation is a group comparison design that doesn't require many variance components, it is favorable to Limma and ANOVA.
-As discussed previously, these methods are not directly applicable to experiments with additional variance components.
+%\subsubsection*{MSstatsPTM corrected the excess FDR in unadjusted differentially abundant PTMs in clean simulations}
+\subsubsection*{Protein-level adjustment was required to control FDR in differentially abundant PTM}
+
+\figref{sim1_fdr}(a) summarizes the FDR \todo{I am not sure we should call this FDR.
+By definition FDR is a population-level quantity (it is an expected value of the proportion of false positives.
+Here it is an observed value).
+Maybe replace with 1-PPV?} reported on Computer Simulation 1 dataset by the unadjusted and adjusted versions of methods \todo{XXX}.
+All the analyses were performed to control the FDR at at most 5\%.
+The  simulation mimics a ``clean" label-free experiment, not compromised by issues such as deviations from model assumptions, missing values and outliers.
+Yet, even in these favorable situations, all the versions of the models that did not adjust for confounding from changes in overall protein abundance produced an excessive number of false positives.
+The versions of the models that accounted for this confounding produced much better calibrated error rates.
 
@@
 Also, maybe clarify what you mean by limma with and without adjustment, since we technically didn't introduce limma without the adjustment.
-Maybe explain the inputs to each version of the model?} Computer simulations.
+Maybe explain the inputs to each version of the model?} \todo{In (a), could you add a horizontal line at 5\%?} Computer simulations.
 a) Dataset 1, clean simulation, analyzed to control FDR at at most 5\%.
 The methods not accounting for the confounding between changes in PTM and overall changes in protein abundance produced exceedingly high numbers of false positives.
-In contrast, the methods accounting for the confounding correctly calibrated FDR.
+In contrast, the methods accounting for the confounding correctly calibrated the proportion of false positive differentially abundant PTM.
 b) In a noisy simulation, including limited feature observations and missing values, the advantage of the proposed approach was apparent.

### 2022-07-30 main/MSstatsPTM_Main.tex
More evaluation

@@
 The  simulation mimics a ``clean" label-free experiment, not compromised by issues such as deviations from model assumptions, missing values and outliers.
-Yet, even in these favorable situations, all the versions of the models that did not adjust for confounding from changes in overall protein abundance produced an excessive number of false positives.
-The versions of the models that accounted for this confounding produced much better calibrated error rates.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection*{In noisy simulations, MSstatsPTM better calibrated the true fold change over the existing methods}
+Yet, even under these favorable circumstances, the models that did not adjust for confounding from changes in overall protein abundance produced an excessive number of false positives.
+The versions of the models that accounted for the confounding produced error rates that were much better calibrated at the desired level.
+Since the simulation mimicked the assumptions of \todo{model XXX}, the error rates of this model were calibrated best.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsubsection*{In noisy simulations, MSstatsPTM improved the estimation of fold change over the existing methods \todo{Hmm... Fig 4b is not about the fold change... Rephrase?}}
+
+\todo{I suggest this order for this section.
+(Fig 4b has multiple panels, but the current narrative doesn't make it quite clear what is the difference between the panels) Fig 4b summarizes the results on Sim 2 dataset.
+The dataset is more realistic, and includes XXX.
+For low noise, methods compared in terms of sample size and number of conditions XXX.
+As the noise increased, XXX.
+The difference is due primarily to the robust nature of summarization in MSstats.
+Supplementary Fig XXX details the estimation of fold change}
 
@@
 After adjustment, this systemic difference was removed, however the inner quartile range of the Limma and ANOVA models was wider than the proposed method.
-b) Before adjustment the spike-in peptides (colored red) did not follow the expected log fold change; after adjustment, the spike-in peptides were more in line with expectation.
+b) \todo{The title of the figure reads as if it is accuracy-adjusted.
+Maybe rephrase?} Before adjustment the spike-in peptides (colored red) did not follow the expected log fold change; after adjustment, the spike-in peptides were more in line with expectation.
 Using Limma, the spike-in peptides followed the expected log fold change after adjustment, however the majority of spike-in peptides did not have a significant adjusted p-value.}

### 2022-07-30 main/MSstatsPTM_Main.tex
Small: discussion

@@
 \todo{When you have a chance, could you do minor clean ups of the references?
-Remove the doi; check that the same journal has the same abbreviation in all the references; keep the reference format consistent (e.g., remove capitalization in article titles; italic for book/volume titles etc)} We proposed a general statistical modeling framework and implementation for PTM characterization.
+Remove the doi; check that the same journal has the same abbreviation in all the references; keep the reference format consistent (e.g., remove capitalization in article titles; italic for book/volume titles etc)}
+
+\todo{Somewhere in the discussion it would be good to comment that the proposed approach relies on the commutative property of the null hypothesis in Eq 2.
+While the ratio-based approaches look at the differences within one sample first, we look at the differences between conditions first.
+This provides us with a greater flexibility as far as modeling, accounting for outliers and missing values, and handling complex designs and robust implementations.}
+
+We proposed a general statistical modeling framework and implementation for PTM characterization.
 The framework is designed for bottom-up MS workflows, which are characterized with variations from multiple confounded sources, frequent missing data, and associated uncertainty in the conclusions.

### 2022-07-30 main/MSstatsPTM_Main.tex
Small

@@
 
-Additionally, when comparing the fold change estimation across all modified peptides, the proposed method showed a tighter distribution of estimated fold changes around the true fold change.
-Specifically, the inner quartile range (IQR) of the estimated fold change for the proposed method was on average 32.5\% smaller than Limma and ANOVA's IQR.
+{\bf Supplementary Figure~\ref{supp-fig:fc_boxplot}} further compares the fold change estimation across all modified peptides.
+$MSstatsPTM$ showed a tighter distribution of estimated fold changes around the true fold change.
+Specifically, the inter-quartile range (IQR) of the estimated fold change for the proposed method was on average 32.5\% smaller than Limma and ANOVA's IQR.
 While the mean of the estimated fold changes was generally correct for all methods, the proposed approach correctly estimated the fold change more often across all PTMs.
-This fold change comparison can be seen in {\bf Supplementary Figure~\ref{supp-fig:fc_boxplot}}.
 

### 2022-08-02 main/MSstatsPTM_Main.tex
Edits on zoom

@@
 The p-values of the comparison are adjusted for multiple testing using the approach by Benjamin and Hochberg~\cite{Benjamini:1995}.
-\todo{Add 'df' to Step 5 of Fig 3?}
 
@@
 In this case the implementation reverts to testing the null hypothesis in~\eqref{conv_null_hyp} using the statistical methods seen in $MSstats$, applied separately to each modified peptide.
-\todo{This is indicated in the `issues' column in the output?} The final output is the estimate $\hat{\Delta}_{adj}$ and it's standard error $\widehat{SE}(\hat{\Delta}_{adj})$.
+The final output is the estimate $\hat{\Delta}_{adj}$ and it's standard error $\widehat{SE}(\hat{\Delta}_{adj})$.
 

### 2022-08-02 main/MSstatsPTM_Main.tex
eval comments

@@
 
+We compared the performance of $MSstatsPTM$ to that of$Limma$ and $ANOVA$.
+Since IsobarPTM is only applicable to experiments with TMT labeling, it could not be applied to the datasets with known ground truth in this manuscript, and was excluded from the comparisons.
+All methods were compared both before and after adjusting for changes in the unmodified protein.
+$MSstatsPTM$ without adjustment is the same as running $MSstats$ on the peptide level.
+$ANOVA$ without adjustment is shown in \eqref{ttest_model}, and will be referred to as basic $ANOVA$.
+Finally, $Limma$ without adjustment uses the same model formula in \eqref{ttest_model}, while including a moderated variance estimation, and is referred to as basic $Limma$.
+$MSstatsPTM$ was used without imputing missing values, and without Empirical Bayes moderation.
+All the evaluations were done at the FDR-adjusted p-value cutoff of $q=.05$.
+More details are in {\bf Supplementary Sec.
+\ref{supp-sec:experiments}}.
+
 We evaluated $MSstatsPTM$ on simulated and spike-in datasets with known ground truth in terms of true positives ($TP$), false positives ($FP$), true negatives ($TN$), and false negatives ($FN$) differentially abundant PTMs.
@@
 
-We compared the performance of  $MSstatsPTM$ to that of \todo{Need more details on the 'adjusted' vs 'unadjusted' versions ANOVA and Limma.
-Give each version a name that you'll later use in the figures, and describe what type of summaries each version takes.} ANOVA and of Limma.
-Since IsobarPTM is only applicable to experiments with TMT labeling, it could not be applied to the datasets with known ground truth in this manuscript, and was excluded from the comparisons.
-$MSstatsPTM$ \todo{Also describe the 'adjusted' version (MSstatsPTM) and the 'unadjusted' version (I guess it is MSstats/MSstatsTMT)} was used without imputing missing values, and without Empirical Bayes moderation.
-All the evaluations were done at the FDR-adjusted p-value cutoff of $q=.05$.
-More details are in {\bf Supplementary Sec.
-\ref{supp-sec:experiments}}.
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 Here it is an observed value).
-Maybe replace with 1-PPV?} reported on Computer Simulation 1 dataset by the unadjusted and adjusted versions of methods \todo{XXX}.
+Maybe replace with 1-PPV?} reported on Computer Simulation 1 dataset by $MSstatsPTM$, $ANOVA$, $Limma$ and the basic $MSstats$, $ANOVA$ and $Limma$ methods.
 All the analyses were performed to control the FDR at at most 5\%.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection*{In noisy simulations, MSstatsPTM improved the estimation of fold change over the existing methods \todo{Hmm... Fig 4b is not about the fold change... Rephrase?}}
-
-\todo{I suggest this order for this section.
-(Fig 4b has multiple panels, but the current narrative doesn't make it quite clear what is the difference between the panels) Fig 4b summarizes the results on Sim 2 dataset.
-The dataset is more realistic, and includes XXX.
-For low noise, methods compared in terms of sample size and number of conditions XXX.
-As the noise increased, XXX.
-The difference is due primarily to the robust nature of summarization in MSstats.
-Supplementary Fig XXX details the estimation of fold change}
-
-In simulations with missing values and few features, $MSstatsPTM$ outperformed ANOVA and Limma.
-Changes in unmodified protein level still needed to be accounted for to control the FDR.
-Once controlling for changes in the unmodified protein, the proposed method outperformed the other methods, as seen in~\figref{sim2_acc}.
-The proposed method calibrated model accuracy  well, even when the number of replicates were low.
-$MSstatsPTM$ can handle low feature counts better than the existing methods due to using Tukey's Median Polish for summarization.
+\subsubsection*{In noisy simulations, MSstatsPTM more accurately detected differentially modified modifications over the existing methods}
+
+Figure \figref{sim2_acc} summarizes the results of the second computer simulation.
+The simulation is more realistic, and includes low modified feature counts and missing values.
+Differing numbers of conditions and replicates were compared in the presence of low and high noise.
+For low noise, $MSstatsPTM$ outperformed the existing methods across all conditions and number of replicates, with near 100\% accuracy when the replicates were high.
+As the noise increased the accuracy of all the methods decreased, however $MStatsPTM$ still outperformed the existing methods.
+The difference is primarily due to the robust nature of summarization in $MSstatsPTM$ through Tukey's Median Polish.
 The existing methods summation approach requires more observations to correctly summarize the intensities in the run, whereas TMP is robust to few measurements.
@@
 In this experiment all models incorrectly estimated the fold change of the modified spike-in peptides before adjusting for changes in unmodified protein abundance.
-After adjustment, the spike-in peptides' fold change was generally in line with expectation in all methods, however the existing method's distribution of estimated fold changes was visibly wider (\figref{spikein_boxplot}).
-Of the three approaches, the proposed method showed the tightest distribution around the true log fold change.
+After adjustment, the spike-in peptides' fold change was generally in line with expectation in all methods, however $MSstatsPTM$'s distribution of estimated fold changes was visibly tighter than the existing methods (\figref{spikein_boxplot}).
 Comparing the IQR of the spike-in peptide's log fold change, the proposed method's IQR was 32.86\% smaller than Limma and ANOVA's IQR.
-While Limma and ANOVA overall estimated the correct fold change, they were inconsistent in their estimation, resulting in a higher error rate.
+While Limma and ANOVA generally estimated the correct fold change, they were inconsistent in their estimation, resulting in a higher error rate.
 The proposed method's estimation was more consistent, with a smaller error.
@@
 In contrast, the methods accounting for the confounding correctly calibrated the proportion of false positive differentially abundant PTM.
-b) In a noisy simulation, including limited feature observations and missing values, the advantage of the proposed approach was apparent.
+b)  \todo{The title of the figure reads as if it is accuracy-adjusted.
+Maybe rephrase?} In a noisy simulation, including limited feature observations and missing values, the advantage of the proposed approach was apparent.
 Looking at accuracy, the proposed method outperformed ANOVA and Limma in nearly every model.
@@
 After adjustment, this systemic difference was removed, however the inner quartile range of the Limma and ANOVA models was wider than the proposed method.
-b) \todo{The title of the figure reads as if it is accuracy-adjusted.
-Maybe rephrase?} Before adjustment the spike-in peptides (colored red) did not follow the expected log fold change; after adjustment, the spike-in peptides were more in line with expectation.
+b) Before adjustment the spike-in peptides (colored red) did not follow the expected log fold change; after adjustment, the spike-in peptides were more in line with expectation.
 Using Limma, the spike-in peptides followed the expected log fold change after adjustment, however the majority of spike-in peptides did not have a significant adjusted p-value.}

### 2022-08-09 code/Spike-in_Vizualization.Rmd
implementation and eval updates

@@
 
+model_df$ADJUSTED.Model$protadj = "MSstatsPTM"
+model_df$PTM.Model$protadj = "MSstats"
+
 plot_df <- rbind(model_df$ADJUSTED.Model %>% select(-GlobalProtein), 
               model_df$PTM.Model %>% select(-c(MissingPercentage,ImputationPercentage, issue)))
-plot_df$protadj <- factor(plot_df$protadj, levels = c('no adjustment','protein adjustment'))
+plot_df$protadj <- factor(plot_df$protadj, levels = c('MSstats','MSstatsPTM'))
 
@@
 # Volcano plot
+png("../supplementary/sim_new/spike_in_msstatsptm_volcano.png", width = 1000, height = 500)
 plot_df %>% arrange(labeling) %>% 
@@
   guides(colour = guide_legend(override.aes = list(size=5))) +
-  labs(title = "Proposed Method VolcanoPlot")
+  labs(title = "MSstatsPTM VolcanoPlot")
+dev.off()
 ```
@@
 unadj_spike <- model_df$PTM.Model %>% filter(labeling == "heavy")
-unadj_spike$Adjustment <- "no adjustment"
+unadj_spike$Adjustment <- "MSstats"
 adj_spike <- model_df$ADJUSTED.Model %>% filter(labeling == "heavy")
-adj_spike$Adjustment <- "protein adjustment"
+adj_spike$Adjustment <- "MSstatsPTM"
 spike <- rbindlist(list(unadj_spike, adj_spike), fill = TRUE)
@@
 
+limma_model$ADJUSTED.Model$protadj = "Limma"
+limma_model$PTM.Model$protadj = "basic Limma"
+
 plot_df <- rbind(limma_model$ADJUSTED.Model, limma_model$PTM.Model, fill = TRUE)
-plot_df$protadj <- factor(plot_df$protadj, levels = c('no adjustment','protein adjustment'))
+plot_df$protadj <- factor(plot_df$protadj, levels = c('basic Limma','Limma'))
 
@@
 # Volcano plot
+png("../supplementary/sim_new/spike_in_limma_volcano.png", width = 1000, height = 500)
 plot_df %>% arrange(labeling) %>% 
@@
   labs(title = "Limma VolcanoPlot")
+dev.off()
 ```
@@
 
-proposed_plot_df$Model <- "Proposed Method"
+proposed_plot_df$Model <- "MSstatsPTM"
+proposed_plot_df[proposed_plot_df$protadj == "MSstats"]$protadj = "no adjustment"
+proposed_plot_df[proposed_plot_df$protadj == "MSstatsPTM"]$protadj = "protein adjustment"
 limma_plot_df$Model <- "Limma"
@@
 limma_plot_df$log2FC <- limma_plot_df$Log2FC
+limma_plot_df[limma_plot_df$protadj == "basic Limma"]$protadj = "no adjustment"
+limma_plot_df[limma_plot_df$protadj == "Limma"]$protadj = "protein adjustment"
 
@@
                                             "protadj", "Model")])
-combo_plot_df$Model <- factor(combo_plot_df$Model, levels=c('Proposed Method','Limma'))
+combo_plot_df$Model <- factor(combo_plot_df$Model, levels=c('MSstatsPTM','Limma'))
 
+png("../main/images/spike_in_volcano.png", width = 750, height = 500)
 combo_plot_df %>% arrange(Model, labeling) %>% 
@@
   labs(title = "SpikeIn VolcanoPlot", x= "Log2FC")
+dev.off()
 
@@
 
+anova_results[anova_results$protadj == "no adjustment"]$protadj = "basic ANOVA"
+anova_results[anova_results$protadj == "protein adjustment"]$protadj = "ANOVA"
+
 ## Update some naming conventions
@@
 png("../supplementary/sim_new/spike_in_anova_volcano.png", width = 1000, height = 500)
-anova_results %>% arrange(labeling) %>% 
+anova_results %>% arrange(protadj, labeling) %>% 
   filter(is.finite(log2FC)) %>%
@@
 spike <- as.data.table(spike)
-spike[spike$Model == "MSstatsPTM", Model := "Proposed"]
+# spike[spike$Model == "MSstatsPTM", Model := "Proposed"]
 #plot_df[plot_df$Label == 'mix1-mix3'][['contrast_truth']]
@@
 spike %>% mutate(Comparison = str_split(contrast_truth, " ", simplify = TRUE)[,1],
-      Model = factor(Model, levels=c("Proposed", "Limma", "ANOVA"))) %>%
+      Model = factor(Model, levels=c("MSstatsPTM", "Limma", "ANOVA"))) %>%
   filter(!Label %in% c("mix1-mix2", "mix3-mix4")) %>% 

### 2022-08-09 main/MSstatsPTM_Main.tex
typo
implementation and eval updates

@@
 Frequently data from these experiments are analyzed using statistical methods that were not originally designed for this task.
-Researchers use methods such as $t$-test\cite{Kalpic:2011}, Analysis of Variance (ANOVA)\cite{girden:1992}, or Limma\cite{Ritchie_15a}, by taking as input the intensity ratios of modified and unmodified peptide features, and comparing the mean abundance of different PTM sites.
+Researchers use methods such as $t$-test\cite{Kalpic:2011}, Analysis of Variance ($ANOVA$)\cite{girden:1992}, or $Limma$\cite{Ritchie_15a}, by taking as input the intensity ratios of modified and unmodified peptide features, and comparing the mean abundance of different PTM sites.
 Such approaches do not fully account for all the sources of uncertainty.
@@
 
-Analysis of Variance (ANOVA)~\cite{kutner_etal_04a} is the simplest statistical model for summarized modified features in each biological replicate.
+Analysis of Variance ($ANOVA$)~\cite{kutner_etal_04a} is the simplest statistical model for summarized modified features in each biological replicate.
 The summarization often consists of averaging (or taking the median or other robust summary) of the log$_2$ intensities of the modified features in each replicate, e.g.
@@
 \end{eqnarray}
-The basic ANOVA model  is then
+The basic $ANOVA$ model  is then
 \begin{eqnarray}
@@
 \medskip \noindent 
-The basic ANOVA can be extended to account for the confounding of changes in PTM abundance and overall changes in protein abundance~\cite{Schwammle2015,THOMAS2020,Mertins:2013}.
+The basic $ANOVA$ can be extended to account for the confounding of changes in PTM abundance and overall changes in protein abundance~\cite{Schwammle2015,THOMAS2020,Mertins:2013}.
 Typically this is done by first calculating sums of the intensities of the modified and unmodified features on the original scale, and then considering replicate-wise ratios of the sums and taking the log
@@
 \end{eqnarray}
-The approach then models these values with the basic ANOVA, which corresponds to 
+The approach then models these values with the basic $ANOVA$, which corresponds to 
 \begin{eqnarray}
@@
 
-\medskip \noindent The estimation of nuisance variation of the two ANOVA models above is often further expanded with Empirical Bayes moderation implemented in Limma~\cite{ Ritchie_15a, Schwammle2015, Smyth:2004, Smyth:2005, Zhu,Chappell:2021}.
-A typical application of Limma on summarized modified log$_2$-intensities takes as input $\hat{y}_{ij}$, and for each PTM fits the linear model in~\eqref{ttest_model}. 
-A typical application of ratio-based Limma takes as input $u_{ij}$, and for each PTM fits the linear model in~\eqref{ttest_ratio_model}. 
-The Limma versions of the models differ in that they specifies additional prior distributions for the model parameters. 
+\medskip \noindent The estimation of nuisance variation of the two $ANOVA$ models above is often further expanded with Empirical Bayes moderation implemented in $Limma$~\cite{ Ritchie_15a, Schwammle2015, Smyth:2004, Smyth:2005, Zhu,Chappell:2021}.
+A typical application of $Limma$ on summarized modified log$_2$-intensities takes as input $\hat{y}_{ij}$, and for each PTM fits the linear model in~\eqref{ttest_model}. 
+A typical application of ratio-based $Limma$ takes as input $u_{ij}$, and for each PTM fits the linear model in~\eqref{ttest_ratio_model}. 
+The $Limma$ versions of the models differ in that they specifies additional prior distributions for the model parameters. 
 The priors are estimated from the same data by combining the information across all the proteins and all the PTM as described in \cite{Smyth:2004}. 
@@
 
-Since Limma only improves upon the estimation of variation, its limitations are similar to those of ANOVA.
+Since $Limma$ only improves upon the estimation of variation, its limitations are similar to those of $ANOVA$.
 In particular, the method is only directly applicable to experiments with at most two variance components, and cannot account for all the sources of variation in experiments with either isobaric labeling or complex designs.
@@
 The specific model depends on the design of the experiment, labeling type and data acquisition type as summarized in {\bf Supplementary Figure~\ref{supp-fig:statistical_inference}}.
-For example, the unmodified protein features in the simple design in \figref{data-structure} are modeled with one-way ANOVA
+For example, the unmodified protein features in the simple design in \figref{data-structure} are modeled with one-way $ANOVA$
 \begin{eqnarray}
@@
 The $MSstats$ framework has a number advantages over the methods above. 
-First, unlike ANOVA and $Limma$, $MSstats$ and $MSstatsTMT$ are applicable to arbitrary complex experimental designs, including designs with multiple sources of variation, and unbalanced designs. 
+First, unlike $ANOVA$ and $Limma$, $MSstats$ and $MSstatsTMT$ are applicable to arbitrary complex experimental designs, including designs with multiple sources of variation, and unbalanced designs. 
 Second, the approach is applicable to various data acquisition types, including label-free DDA and DIA, and experiments with TMT labeling. 
@@
 In particular, the modified features are summarized into run-level summaries $\hat{y}_{ij}^{\ast}$.
-The estimated summaries of the modified features are then used as the input to the appropriate model in $MSstats$ or $MSstatsTMT$,  such as in~\eqref{msstats_model} or~\eqref{msstatstmt_model} \todo{Could we show in Fig 3 where $MSstats$ or $MSstatsTMT$ come into place?}.
+The estimated summaries of the modified features are then used as the input to the appropriate model in $MSstats$ or $MSstatsTMT$,  such as in~\eqref{msstats_model} or~\eqref{msstatstmt_model}.
 The resulting model-based estimates include $\hat{\Delta}_{PTM}=\hat{\mu}^{\ast}_{i}-\hat{\mu}_{i^{\prime}}^{\ast}$, and its standard error $\widehat{SE}(\hat{\Delta}_{PTM})$.
@@
 
-Finally, separate statistical models are fit to both feature summaries using the $groupComparisonPTM()$ function (Step 5 of~\figref{msstatsptm_design}).
-The models are automatically selected to reflect the experimental design and the data acquisition.
+Separate statistical models are fit to both feature summaries using the $groupComparisonPTM()$ function (Step 5 of~\figref{msstatsptm_design}).
+$MSstats$ or $MSstatsTMT$ models are leveraged to automatically reflect the experimental design and the data acquisition.
 If the base model is not applicable for a particular PTM or protein, e.g.
@@
 
+Finally, the statistical models can be used to calculate the required sample size for future PTM experiments using the $designSampleSizePTM$ function (Step 6 of~\figref{msstatsptm_design}).
+
 In addition to the above functionalities, the implementation includes visualizations for quality control, $dataProcessPlotsPTM()$, and assessment of the quality of model fit, $groupComparisonPlotsPTM()$.
@@
 
-We compared the performance of $MSstatsPTM$ to that of$Limma$ and $ANOVA$.
-Since IsobarPTM is only applicable to experiments with TMT labeling, it could not be applied to the datasets with known ground truth in this manuscript, and was excluded from the comparisons.
+We compared the performance of $MSstatsPTM$ to that of $Limma$ and $ANOVA$.
+Since Isobar-PTM is only applicable to experiments with TMT labeling, it could not be applied to the datasets with known ground truth in this manuscript, and was excluded from the comparisons.
 All methods were compared both before and after adjusting for changes in the unmodified protein.
 $MSstatsPTM$ without adjustment is the same as running $MSstats$ on the peptide level.
-$ANOVA$ without adjustment is shown in \eqref{ttest_model}, and will be referred to as basic $ANOVA$.
-Finally, $Limma$ without adjustment uses the same model formula in \eqref{ttest_model}, while including a moderated variance estimation, and is referred to as basic $Limma$.
+$ANOVA$ without adjustment is shown in \eqref{ttest_model}, and will be referred to as $basic$ $ANOVA$.
+Finally, $Limma$ without adjustment uses the same model formula in \eqref{ttest_model}, while including a moderated variance estimation, and is referred to as $basic$ $Limma$.
 $MSstatsPTM$ was used without imputing missing values, and without Empirical Bayes moderation.
@@
 The true negatives were defined as PTMs which, after accounting for the changes in the overall protein abundance, were not differentially abundant.
-Additional summaries were the false discovery rate $FDR=FP/(TP+FP)$ \todo{I am not sure we should call this FDR.
-By definition FDR is a population-level quantity (it is an expected value of the proportion of false positives.
-Here it is an observed value).
-Maybe replace with 1-PPV?}, $Recall=TP/(TP + FN)$, and $Accuracy=(TP + TN) / (TP + TN + FP + FN)$.
+Additional summaries were performed including accuracy, recall, and positive predictive value (PPV)/empirical False Discovery Rate (eFDR) as described in \eqref{summary_stats}.
+
+\begin{equation}\label{eq:summary_stats}
+\begin{alignedat}{2}
+&\text{Accuracy} = \frac{TP + TN}{TP+TN+FP+FN}, \text{ Recall} = \frac{TP}{TP+FN} & \\
+&1 - \text{PPV} = \frac{FP}{TP+FP} = \text{empirical False Discovery Rate (eFDR)}
+\end{alignedat}
+\end{equation}
+
 For biological experiments with unknown ground truth, we compared the differentially abundant PTMs with and without adjusting for changes in unmodified protein abundance.
@@
 %\subsubsection*{MSstatsPTM corrected the excess FDR in unadjusted differentially abundant PTMs in clean simulations}
-\subsubsection*{Protein-level adjustment was required to control FDR in differentially abundant PTM}
-
-\figref{sim1_fdr}(a) summarizes the FDR \todo{I am not sure we should call this FDR.
-By definition FDR is a population-level quantity (it is an expected value of the proportion of false positives.
-Here it is an observed value).
-Maybe replace with 1-PPV?} reported on Computer Simulation 1 dataset by $MSstatsPTM$, $ANOVA$, $Limma$ and the basic $MSstats$, $ANOVA$ and $Limma$ methods.
-All the analyses were performed to control the FDR at at most 5\%.
+\subsubsection*{Protein-level adjustment was required to control eFDR in differentially abundant PTM}
+
+\figref{sim1_fdr}(a) summarizes the eFDR reported on Computer Simulation 1 dataset by $MSstatsPTM$, $ANOVA$, $Limma$ and the basic $MSstats$, $ANOVA$ and $Limma$ methods.
+All the analyses were performed to control the eFDR at at most 5\%.
 The  simulation mimics a ``clean" label-free experiment, not compromised by issues such as deviations from model assumptions, missing values and outliers.
@@
 $MSstatsPTM$ showed a tighter distribution of estimated fold changes around the true fold change.
-Specifically, the inter-quartile range (IQR) of the estimated fold change for the proposed method was on average 32.5\% smaller than Limma and ANOVA's IQR.
+Specifically, the inter-quartile range (IQR) of the estimated fold change for the proposed method was on average 32.5\% smaller than $Limma$ and $ANOVA$'s IQR.
 While the mean of the estimated fold changes was generally correct for all methods, the proposed approach correctly estimated the fold change more often across all PTMs.
@@
 After adjustment, the spike-in peptides' fold change was generally in line with expectation in all methods, however $MSstatsPTM$'s distribution of estimated fold changes was visibly tighter than the existing methods (\figref{spikein_boxplot}).
-Comparing the IQR of the spike-in peptide's log fold change, the proposed method's IQR was 32.86\% smaller than Limma and ANOVA's IQR.
-While Limma and ANOVA generally estimated the correct fold change, they were inconsistent in their estimation, resulting in a higher error rate.
+Comparing the IQR of the spike-in peptide's log fold change, the proposed method's IQR was 32.86\% smaller than $Limma$ and $ANOVA$'s IQR.
+While $Limma$ and $ANOVA$ generally estimated the correct fold change, they were inconsistent in their estimation, resulting in a higher error rate.
 The proposed method's estimation was more consistent, with a smaller error.
@@
 Specifically, for the proposed method, the number of false positives went from 20.88\% to 1.84\% after adjustment was applied.
-While the proposed method and Limma both correctly estimated the fold change of the spike-in peptides, using Limma resulted in many large adjusted p-values.
-This was due to Limma estimating a 35.7\% larger variance compared to the proposed method when estimating the spike-in peptides.
+While the proposed method and $Limma$ both correctly estimated the fold change of the spike-in peptides, using $Limma$ resulted in many large adjusted p-values.
+This was due to $Limma$ estimating a 35.7\% larger variance compared to the proposed method when estimating the spike-in peptides.
 This experiment introduced more variance components, which is challenging for the existing methods and therefore the difference in the results is more pronounced than in the simulations.
@@
 The framework is general and is applicable to a variety of experimental designs.
-It outperforms the ad-hoc methods underlying ANOVA and Limma, and yields accurate results in the broad type of experimental circumstances, including the presence of missing values, changes in protein abundance, few representative peptides, and different acquisition methods.
+It outperforms the ad-hoc methods underlying $ANOVA$ and $Limma$, and yields accurate results in the broad type of experimental circumstances, including the presence of missing values, changes in protein abundance, few representative peptides, and different acquisition methods.
 The framework allows us to plan for subsequent experiments, and choose the appropriate number of replicates in consideration of adjustment with respect to protein abundance.
@@
 \centering
-\includegraphics[scale=.5]{images/MSstatsPTM_design.png}
-\caption{\todo{Where does the sample size calculation fit into that?} \todo{Could you rotate the 'Upstream...' Text and 'MSstatsPTM text and place it vertically to the left of the corresponding steps?
-It will be a bit more clear} \todo{It is not very clear which steps leverage MSstats/MSstatsTMT.
-Should we say in relevant places that these are MSstats implementations?} The $MSstatsPTM$ workflow.
+\includegraphics[scale=.55]{images/MSstatsPTM_design.png}
+\caption{The $MSstatsPTM$ workflow.
 The names of the R functions used for each step are highlighted in purple and the output notations are highlighted in red.
@@
 Conversion, summarization and statistical modeling are performed separately for the PTM and for the unmodified proteins.
-At the end, the model-based summaries are combined to adjust the changes in the PTM abundance for changes in unmodified protein abundance.
-Data visualizations can optionally be created after steps 4 and 5 of the workflow.}
+Steps 4 and 5 leverage the summarization and modeling functions from $MSstats$ and $MSstatsTMT$.
+Model-based summaries are combined to adjust the changes in the PTM abundance for changes in unmodified protein abundance.
+Finally, sample size calculation for future experiments can be performed using the modeling output.}
 \label{fig:msstatsptm_design}
@@
 \end{subfigure}
-\caption{\todo{Could you clarify what 'with' and 'without' adjustment means for each model?
-E.g., what is 'proposed with no adjustment'?
-We did not propose anything without adjustment - did you mean base MSstats?
-Also, maybe clarify what you mean by limma with and without adjustment, since we technically didn't introduce limma without the adjustment.
-Maybe explain the inputs to each version of the model?} \todo{In (a), could you add a horizontal line at 5\%?} Computer simulations.
-a) Dataset 1, clean simulation, analyzed to control FDR at at most 5\%.
+\caption{Computer simulations.
+a) Dataset 1, clean simulation, analyzed to control eFDR at at most 5\%.
 The methods not accounting for the confounding between changes in PTM and overall changes in protein abundance produced exceedingly high numbers of false positives.
 In contrast, the methods accounting for the confounding correctly calibrated the proportion of false positive differentially abundant PTM.
-b)  \todo{The title of the figure reads as if it is accuracy-adjusted.
-Maybe rephrase?} In a noisy simulation, including limited feature observations and missing values, the advantage of the proposed approach was apparent.
-Looking at accuracy, the proposed method outperformed ANOVA and Limma in nearly every model.
+b) In a noisy simulation, including limited feature observations and missing values, the advantage of the proposed approach was apparent.
+Looking at accuracy, the proposed method outperformed $ANOVA$ and $Limma$ in nearly every model.
 }
@@
 a) Before adjustment the fold change of the spike-in peptides' were systematically different from the expected fold change in all models.
-After adjustment, this systemic difference was removed, however the inner quartile range of the Limma and ANOVA models was wider than the proposed method.
+After adjustment, this systemic difference was removed, however the inner quartile range of the $Limma$ and $ANOVA$ models was wider than the proposed method.
 b) Before adjustment the spike-in peptides (colored red) did not follow the expected log fold change; after adjustment, the spike-in peptides were more in line with expectation.
-Using Limma, the spike-in peptides followed the expected log fold change after adjustment, however the majority of spike-in peptides did not have a significant adjusted p-value.}
+Using $Limma$, the spike-in peptides followed the expected log fold change after adjustment, however the majority of spike-in peptides did not have a significant adjusted p-value.}
 \label{fig:spikein_volcano}

### 2022-08-10 main/MSstatsPTM_Main.tex
Methods

@@
 \end{eqnarray}
-The model allows us to estimate $\hat{\Delta}$ and its standard error.
+The model allows us to estimate $\hat{\Delta}$ \todo{should be $\hat{\Delta}_{PTM}?$} and its standard error.
 The estimates are used to test the null hypothesis in \eqref{conv_null_hyp} by comparing the model-based test statistic against the Student distribution with $df=I(J-1)$ degrees of freedom in balanced designs.
@@
 \medskip \noindent The estimation of nuisance variation of the two $ANOVA$ models above is often further expanded with Empirical Bayes moderation implemented in $Limma$~\cite{ Ritchie_15a, Schwammle2015, Smyth:2004, Smyth:2005, Zhu,Chappell:2021}.
-A typical application of $Limma$ on summarized modified log$_2$-intensities takes as input $\hat{y}_{ij}$, and for each PTM fits the linear model in~\eqref{ttest_model}. 
+A typical application of $Limma$ on summarized modified log$_2$-intensities takes as input $\hat{y}_{ij}$, \todo{should it be $\hat{y}^*_{ij}$?
+And maybe clarify that it is obtained as in Eq 3 and 4} and for each PTM fits the linear model in~\eqref{ttest_model}.
 A typical application of ratio-based $Limma$ takes as input $u_{ij}$, and for each PTM fits the linear model in~\eqref{ttest_ratio_model}. 
-The $Limma$ versions of the models differ in that they specifies additional prior distributions for the model parameters. 
+The $Limma$ versions of the models differ \todo{from the models XXX above} in that they specifies additional prior distributions for the model parameters. 
 The priors are estimated from the same data by combining the information across all the proteins and all the PTM as described in \cite{Smyth:2004}. 
-With this approach, testing the null hypothesis is enhanced by combining the PTM- and protein-specific estimates of variation with a consensus estimate obtained from all the PTM and all the proteins.
+Testing the null hypothesis is enhanced by combining the PTM- and protein-specific estimates of variation with a consensus estimate obtained from the estimated priors.
 As the result, in experiments with few biological replicates the standard errors are often smaller, and the degrees of freedom are often larger than without moderation~\cite{Ritchie_15a}. 
@@
 The packages take as input log$_2$-intensities $y_{ijk}$.
-For each protein, the log$_2$-intensities are first summarized into a single value per protein per run using Tukey's median polish~\cite{Tukey:1977}.
+For each protein, the log$_2$-intensities are first summarized into a single value per protein per run $\hat{y}_{ij}$ using Tukey's median polish~\cite{Tukey:1977}.
 The summaries are then used as input to fit a flexible family of linear mixed-effects models \cite{McLean:1991, Faraway:2006, Bolker2009}.
@@
 The overall statistical analysis workflow and its implementation are summarized in~\figref{msstatsptm_design}.
-$MSstatsPTM$ takes as input the modified spectral features $y_{ijk}^{\ast}$, and the corresponding unmodified features $y_{ijk}$.
-Ideally, the modified features are acquired separately, after an enrichment to maximize the information content in the resulting dataset, and the unmodified features are acquired separately as part of a global proteome profiling.
+$MSstatsPTM$ takes as input the modified spectral features $y_{ijk}^{\ast}$, and the corresponding unmodified features $y_{ijk}$ (\todo{Steps 1-3 in the Figure}).
+Ideally, the modified features are acquired separately after an enrichment, to maximize the information content in the resulting dataset, and the unmodified features are acquired separately as part of a global proteome profiling.
 However the method can also take as input a combination of modified and unmodified features acquired within a same run.
 
-Each feature type is first analyzed separately using $MSstats$ or $MSstatsTMT$.
-In particular, the modified features are summarized into run-level summaries $\hat{y}_{ij}^{\ast}$.
-The estimated summaries of the modified features are then used as the input to the appropriate model in $MSstats$ or $MSstatsTMT$,  such as in~\eqref{msstats_model} or~\eqref{msstatstmt_model}.
+Each feature type is first analyzed separately, with $MSstatsPTM$ functions calling the relevant functionalities in $MSstats$ (for label-free experiments) or $MSstatsTMT$ (for experiments with TMT labels).
+In particular, the modified features are summarized into run-level summaries $\hat{y}_{ij}^{\ast}$ (\todo{Step 4 in the Figure}).
+
+\todo{In Step 5}, the estimated summaries of the modified features are used as the input to models  such as in~\eqref{msstats_model} or~\eqref{msstatstmt_model}.
 The resulting model-based estimates include $\hat{\Delta}_{PTM}=\hat{\mu}^{\ast}_{i}-\hat{\mu}_{i^{\prime}}^{\ast}$, and its standard error $\widehat{SE}(\hat{\Delta}_{PTM})$.
 Similarly, the unmodified features of each protein are summarized for each run into $\hat{y}_{ij}$, and the summaries are used as input to a separate analysis by $MSstats$ or $MSstatsTMT$ producing $\hat{\Delta}_{protein}=\hat{\mu}_{i}-\hat{\mu}_{i^{\prime}}$ and $\widehat{SE}(\hat{\Delta}_{protein})$.
-
-From the summaries above, the proposed approach estimates the adjusted difference $\hat{\Delta}_{adj}$ in~\eqref{null_hyp}
+From these summaries, the proposed approach estimates the adjusted difference $\hat{\Delta}_{adj}$ in~\eqref{null_hyp}
 \begin{eqnarray}
@@
 \caption{The $MSstatsPTM$ workflow.
-The names of the R functions used for each step are highlighted in purple and the output notations are highlighted in red.
+\todo{in 6, maybe call it $J^{\prime}$?}The names of the $MSstatsPTM$ R functions used for each step are highlighted in purple \todo{distinguish bf and plain?} and the output notations are highlighted in red.
 The workflow begins with the acquisition of the enriched and global proteome lysates.

### 2022-08-10 main/MSstatsPTM_Main.tex
Small

@@
 
-Each feature type is first analyzed separately, with $MSstatsPTM$ functions calling the relevant functionalities in $MSstats$ (for label-free experiments) or $MSstatsTMT$ (for experiments with TMT labels).
+Each feature type is first analyzed separately, with $MSstatsPTM$ methods calling the relevant functionalities in $MSstats$ (for label-free experiments) or $MSstatsTMT$ (for experiments with TMT labels).
 In particular, the modified features are summarized into run-level summaries $\hat{y}_{ij}^{\ast}$ (\todo{Step 4 in the Figure}).
@@
 
-The proposed statistical framework enables sample size calculation for future PTM experiments. 
+The statistical framework in $MSstatsPTM$ enables sample size calculation for future experiments studying changes in PTM (\todo{Step 6 of Fig 3}). 
 The procedure has been described in general in~\cite{kutner_etal_04a}, and for protein significance analysis specifically in~\cite{oberg_vitek_09a}. 
@@
 The implementation of the open source R package $MSstatsPTM$ is also overviewed in~\figref{msstatsptm_design}.
-By leveraging the implementations in $MSstats$ and $MSstatsTMT$, the proposed approach is very versatile.
+By leveraging the implementations in $MSstats$ and $MSstatsTMT$, the proposed approach is versatile.
 It is applicable to a wide variety of experimental designs, including group comparison, paired designs, time course designs and unbalanced designs.
@@
 For these cases $MSstatsPTM$ includes functionality for identifying the location, given the modified peptide sequence and a FASTA file with the entire protein sequence.
-The converters output the modified spectral features $y_{ijk}^{\ast}$ and the corresponding unmodified features $y_{ijk}$ in the format required for summarization.
+The converters output the modified spectral features $y_{ijk}^{\ast}$, and the corresponding unmodified features $y_{ijk}$ in the format required for summarization.
 
@@
 These issues can violate the assumptions underlying the imputation, and lead to numerically unstable results.
-The outputs of this step are the run-level summaries for the modified, $\hat{y}_{ij}^{\ast}$, and unmodified, $\hat{y}_{ij}$, features.
+The outputs of this step are the run-level summaries for the modified $\hat{y}_{ij}^{\ast}$, and unmodified $\hat{y}_{ij}$, features.
 
@@
 We compared the performance of $MSstatsPTM$ to that of $Limma$ and $ANOVA$.
-Since Isobar-PTM is only applicable to experiments with TMT labeling, it could not be applied to the datasets with known ground truth in this manuscript, and was excluded from the comparisons.
+Since $Isobar$-$PTM$ is only applicable to experiments with TMT labeling, it could not be applied to the datasets with known ground truth in this manuscript, and was excluded from the comparisons.
 All methods were compared both before and after adjusting for changes in the unmodified protein.

### 2022-08-11 main/MSstatsPTM_Main.tex
Small
updated discussion and refs

@@
 \end{eqnarray}
-The model allows us to estimate $\hat{\Delta}$ \todo{should be $\hat{\Delta}_{PTM}?$} and its standard error.
+The model allows us to estimate $\hat{\Delta}_{PTM}$ and its standard error.
 The estimates are used to test the null hypothesis in \eqref{conv_null_hyp} by comparing the model-based test statistic against the Student distribution with $df=I(J-1)$ degrees of freedom in balanced designs.
@@
 \medskip \noindent The estimation of nuisance variation of the two $ANOVA$ models above is often further expanded with Empirical Bayes moderation implemented in $Limma$~\cite{ Ritchie_15a, Schwammle2015, Smyth:2004, Smyth:2005, Zhu,Chappell:2021}.
-A typical application of $Limma$ on summarized modified log$_2$-intensities takes as input $\hat{y}_{ij}$, \todo{should it be $\hat{y}^*_{ij}$?
-And maybe clarify that it is obtained as in Eq 3 and 4} and for each PTM fits the linear model in~\eqref{ttest_model}.
-A typical application of ratio-based $Limma$ takes as input $u_{ij}$, and for each PTM fits the linear model in~\eqref{ttest_ratio_model}. 
-The $Limma$ versions of the models differ \todo{from the models XXX above} in that they specifies additional prior distributions for the model parameters. 
+A typical application of $Limma$ on summarized modified log$_2$-intensities takes as input $\hat{y}^*_{ij}$, obtained as in \eqref{anova_summ}, and for each PTM fits the linear model in~\eqref{ttest_model}.
+A typical application of ratio-based $Limma$ takes as input $u_{ij}$, obtained as in \eqref{adj_estimation}, and for each PTM fits the linear model in~\eqref{ttest_ratio_model}. 
+The $Limma$ versions of the models differ from the models in~\eqref{ttest_model} and~\eqref{ttest_ratio_model} in that they specify additional prior distributions for the model parameters. 
 The priors are estimated from the same data by combining the information across all the proteins and all the PTM as described in \cite{Smyth:2004}. 
@@
 The overall statistical analysis workflow and its implementation are summarized in~\figref{msstatsptm_design}.
-$MSstatsPTM$ takes as input the modified spectral features $y_{ijk}^{\ast}$, and the corresponding unmodified features $y_{ijk}$ (\todo{Steps 1-3 in the Figure}).
+$MSstatsPTM$ takes as input the modified spectral features $y_{ijk}^{\ast}$, and the corresponding unmodified features $y_{ijk}$.
 Ideally, the modified features are acquired separately after an enrichment, to maximize the information content in the resulting dataset, and the unmodified features are acquired separately as part of a global proteome profiling.
@@
 Each feature type is first analyzed separately, with $MSstatsPTM$ methods calling the relevant functionalities in $MSstats$ (for label-free experiments) or $MSstatsTMT$ (for experiments with TMT labels).
-In particular, the modified features are summarized into run-level summaries $\hat{y}_{ij}^{\ast}$ (\todo{Step 4 in the Figure}).
-
-\todo{In Step 5}, the estimated summaries of the modified features are used as the input to models  such as in~\eqref{msstats_model} or~\eqref{msstatstmt_model}.
+In particular, the modified features are summarized into run-level summaries $\hat{y}_{ij}^{\ast}$.
+
+The estimated summaries of the modified features are used as the input to models  such as in~\eqref{msstats_model} or~\eqref{msstatstmt_model}.
 The resulting model-based estimates include $\hat{\Delta}_{PTM}=\hat{\mu}^{\ast}_{i}-\hat{\mu}_{i^{\prime}}^{\ast}$, and its standard error $\widehat{SE}(\hat{\Delta}_{PTM})$.
@@
 
-The statistical framework in $MSstatsPTM$ enables sample size calculation for future experiments studying changes in PTM (\todo{Step 6 of Fig 3}). 
+The statistical framework in $MSstatsPTM$ enables sample size calculation for future experiments studying changes in PTM. 
 The procedure has been described in general in~\cite{kutner_etal_04a}, and for protein significance analysis specifically in~\cite{oberg_vitek_09a}. 
@@
 
-The implementation of the open source R package $MSstatsPTM$ is also overviewed in~\figref{msstatsptm_design}.
+The implementation of the open source R package $MSstatsPTM$ is overviewed in~\figref{msstatsptm_design}.
 By leveraging the implementations in $MSstats$ and $MSstatsTMT$, the proposed approach is versatile.
@@
 All methods were compared both before and after adjusting for changes in the unmodified protein.
-$MSstatsPTM$ without adjustment is the same as running $MSstats$ on the peptide level.
-$ANOVA$ without adjustment is shown in \eqref{ttest_model}, and will be referred to as $basic$ $ANOVA$.
+$MSstatsPTM$ was compared to to base $MSstats$, modeled as seen in \eqref{msstats_model} on the peptide level.
+$ANOVA$ without adjustment was modeled as shown in \eqref{ttest_model}, and will be referred to as $basic$ $ANOVA$.
+$ANOVA$ with adjustment is referred to as $ANOVA$ and was modeled as in \eqref{ttest_ratio_model}.
 Finally, $Limma$ without adjustment uses the same model formula in \eqref{ttest_model}, while including a moderated variance estimation, and is referred to as $basic$ $Limma$.
+$Limma$ with adjustment is referred to as $Limma$ and was modeled as in \eqref{ttest_ratio_model} including moderated variance estimation.
 $MSstatsPTM$ was used without imputing missing values, and without Empirical Bayes moderation.
@@
 
-\todo{When you have a chance, could you do minor clean ups of the references?
-Remove the doi; check that the same journal has the same abbreviation in all the references; keep the reference format consistent (e.g., remove capitalization in article titles; italic for book/volume titles etc)}
-
-\todo{Somewhere in the discussion it would be good to comment that the proposed approach relies on the commutative property of the null hypothesis in Eq 2.
-While the ratio-based approaches look at the differences within one sample first, we look at the differences between conditions first.
-This provides us with a greater flexibility as far as modeling, accounting for outliers and missing values, and handling complex designs and robust implementations.}
-
 We proposed a general statistical modeling framework and implementation for PTM characterization.
@@
 The framework is general and is applicable to a variety of experimental designs.
-It outperforms the ad-hoc methods underlying $ANOVA$ and $Limma$, and yields accurate results in the broad type of experimental circumstances, including the presence of missing values, changes in protein abundance, few representative peptides, and different acquisition methods.
+While the ratio-based approaches of $ANOVA$ and $Limma$ look at the differences within samples first, the proposed approach looks at the differences between conditions first.
+This provides us with greater flexibility in terms of modeling, accounting for outliers and missing values, and handling complex designs and robust implementations.
+Additionally, the proposed approach outperforms $ANOVA$ and $Limma$, and yields accurate results in the broad type of experimental circumstances.
 The framework allows us to plan for subsequent experiments, and choose the appropriate number of replicates in consideration of adjustment with respect to protein abundance.
@@
 \caption{Computer simulations.
-a) Dataset 1, clean simulation, analyzed to control eFDR at at most 5\%.
+a) Dataset 1, clean simulation, analyzed to control eFDR at at most 5\% (horizontal line).
 The methods not accounting for the confounding between changes in PTM and overall changes in protein abundance produced exceedingly high numbers of false positives.

### 2022-08-11 main/MSstatsPTM_Main.tex
minor change to implementation fig

@@
 \caption{The $MSstatsPTM$ workflow.
-\todo{in 6, maybe call it $J^{\prime}$?}The names of the $MSstatsPTM$ R functions used for each step are highlighted in purple \todo{distinguish bf and plain?} and the output notations are highlighted in red.
+The names of the $MSstatsPTM$ R functions used for each step are highlighted in purple and the output notations are highlighted in red.
 The workflow begins with the acquisition of the enriched and global proteome lysates.

### 2022-08-11 supplementary/ptm_sm.tex
minor change to implementation fig

@@
 The estimation of standard error and degrees of freedom for each model is shown.
-\label{fig:statistical_inference}}
+Only one side of the PTM and unmodified protein is shown, variance must be multiplied by two to combine.}
+\label{fig:statistical_inference}
 \end{figure}
@@
 Here we analyzed the sample size needed to achieve a desired statistical power.
-The proposed approach corrected for confounding between the modified peptide and unmodified protein at the cost of increased variation.
-This can be seen in the calculation for variance.
-Increased variation required a larger number of replicates to reach the same power.
-Thus the statistical power was dependent on the variance from both the modified peptide and unmodifed protein.
-
 We compared the statistical power in experiments with differing numbers of replicates, variance, and fold change for both the modified and unmodified runs.
@@
 When the PTM and protein were the same we chose a variance of $.15$, whereas when the PTM was higher than the protein we chose a PTM variance of $.2$ and a protein variance of $.1$.
-
 The results of the power and sample size analysis can be seen in Figure \ref{fig:power_sd_combo}.
-When the variance and replicates were equal, higher replicates predictably lead to higher power.
-In cases where the replicates were unbalanced, but the variance was still the same, it did not matter if there were more replicates in the modified or unmodified runs.
-In comparison, with differing variance and equal replicates, higher replicates still lead to higher power.
-When the replicates were unbalanced and the variance was higher in the PTM, there was more power when more replicates were allocated to the PTM run than the unmodified protein run.
-The results lead to the takeaway that in cases where the number of replicates have to be limited, it is better to allocate more to the PTM run.
 

### 2022-08-12 main/MSstatsPTM_Main.tex
Results half-way through

@@
 
-We compared the performance of $MSstatsPTM$ to that of $Limma$ and $ANOVA$.
-Since $Isobar$-$PTM$ is only applicable to experiments with TMT labeling, it could not be applied to the datasets with known ground truth in this manuscript, and was excluded from the comparisons.
-All methods were compared both before and after adjusting for changes in the unmodified protein.
-$MSstatsPTM$ was compared to to base $MSstats$, modeled as seen in \eqref{msstats_model} on the peptide level.
-$ANOVA$ without adjustment was modeled as shown in \eqref{ttest_model}, and will be referred to as $basic$ $ANOVA$.
-$ANOVA$ with adjustment is referred to as $ANOVA$ and was modeled as in \eqref{ttest_ratio_model}.
-Finally, $Limma$ without adjustment uses the same model formula in \eqref{ttest_model}, while including a moderated variance estimation, and is referred to as $basic$ $Limma$.
-$Limma$ with adjustment is referred to as $Limma$ and was modeled as in \eqref{ttest_ratio_model} including moderated variance estimation.
-$MSstatsPTM$ was used without imputing missing values, and without Empirical Bayes moderation.
-All the evaluations were done at the FDR-adjusted p-value cutoff of $q=.05$.
-More details are in {\bf Supplementary Sec.
-\ref{supp-sec:experiments}}.
+We compared the performance of $MSstatsPTM$ to that of $Limma$ and $ANOVA$, both before and after adjusting for changes in the unmodified protein.
+Since $Isobar$-$PTM$ is only applicable to experiments with TMT labeling, it could not be applied to the datasets with known ground truth in this manuscript, and was therefore excluded from the comparisons.
+
+\todo{I am wondering if 'basic' is the right vocabulary here.
+Would you like to switch to 'unadjusted ANOVA', 'adjusted ANOVA', 'unadjusted Limma' and 'adjusted Limma', everywhere in the text and in the figures?
+}
+
+$MSstatsPTM$ before adjusting for changes in the unmodified protein corresponds to base $MSstats$ or $MSstatsTMT$, modeled on the peptide level \todo{as described in Suppl Table XXX, as appropriate for the experimental design}.
+$MSstatsPTM$ with the adjustment, \todo{described in Sec XXX}, was used without imputing missing values, and without Empirical Bayes moderation.
+
+$Basic$ $ANOVA$, i.e. $ANOVA$ before adjusting for changes in the unmodified protein, was as in \eqref{ttest_model}. $ANOVA$ with the adjustment was modeled as in \eqref{ttest_ratio_model}. 
+Finally, $basic$ $Limma$ (i.e.
+without $Limma$ adjustment) used the same model formula in \eqref{ttest_model}, while including a moderated variance estimation.
+$Limma$ with adjustment was modeled as in \eqref{ttest_ratio_model} including moderated variance estimation.
+All the evaluations were done at the FDR-adjusted p-value cutoff of $q=.05$. More details are in {\bf Supplementary Sec. \ref{supp-sec:experiments}}.
 
@@
 \end{equation}
-
 For biological experiments with unknown ground truth, we compared the differentially abundant PTMs with and without adjusting for changes in unmodified protein abundance.
@@
 
-\figref{sim1_fdr}(a) summarizes the eFDR reported on Computer Simulation 1 dataset by $MSstatsPTM$, $ANOVA$, $Limma$ and the basic $MSstats$, $ANOVA$ and $Limma$ methods.
+\figref{sim1_fdr} summarizes the eFDR reported on Computer Simulation 1 dataset by $MSstatsPTM$, $ANOVA$, $Limma$ and the basic $MSstats$, $ANOVA$ and $Limma$ methods.
+The  simulation mimicked a ``clean" label-free group comparison experiment, not compromised by issues such as deviations from model assumptions, missing values and outliers.
 All the analyses were performed to control the eFDR at at most 5\%.
-The  simulation mimics a ``clean" label-free experiment, not compromised by issues such as deviations from model assumptions, missing values and outliers.
 Yet, even under these favorable circumstances, the models that did not adjust for confounding from changes in overall protein abundance produced an excessive number of false positives.
 The versions of the models that accounted for the confounding produced error rates that were much better calibrated at the desired level.
-Since the simulation mimicked the assumptions of \todo{model XXX}, the error rates of this model were calibrated best.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection*{In noisy simulations, MSstatsPTM more accurately detected differentially modified modifications over the existing methods}
-
-Figure \figref{sim2_acc} summarizes the results of the second computer simulation.
-The simulation is more realistic, and includes low modified feature counts and missing values.
-Differing numbers of conditions and replicates were compared in the presence of low and high noise.
+\todo{Since the simulation mimicked the assumptions of model XXX?, the error rates of this model were calibrated best.}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsubsection*{In noisy simulations, MSstatsPTM more accurately detected differentially abundant PTM}
+
+\figref{sim2_acc} summarizes the overall accuracy of methods adjusting for changes in the unmodified protein for Computer Simulation 2.
+The simulation mimicked a more realistic label-free group comparison experiment, including low counts of modified features and missing values.
+We evaluated the impact of low versus high noise, of the number of biological replicates, and of the number of conditions.
 For low noise, $MSstatsPTM$ outperformed the existing methods across all conditions and number of replicates, with near 100\% accuracy when the replicates were high.
@@
 The difference is primarily due to the robust nature of summarization in $MSstatsPTM$ through Tukey's Median Polish.
-The existing methods summation approach requires more observations to correctly summarize the intensities in the run, whereas TMP is robust to few measurements.
 
@@
 $MSstatsPTM$ showed a tighter distribution of estimated fold changes around the true fold change.
-Specifically, the inter-quartile range (IQR) of the estimated fold change for the proposed method was on average 32.5\% smaller than $Limma$ and $ANOVA$'s IQR.
-While the mean of the estimated fold changes was generally correct for all methods, the proposed approach correctly estimated the fold change more often across all PTMs.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection*{In the label-free benchmark experiment, MSstatsPTM had a higher sensitivity than the existing methods}
-
-In this experiment all models incorrectly estimated the fold change of the modified spike-in peptides before adjusting for changes in unmodified protein abundance.
-After adjustment, the spike-in peptides' fold change was generally in line with expectation in all methods, however $MSstatsPTM$'s distribution of estimated fold changes was visibly tighter than the existing methods (\figref{spikein_boxplot}).
-Comparing the IQR of the spike-in peptide's log fold change, the proposed method's IQR was 32.86\% smaller than $Limma$ and $ANOVA$'s IQR.
-While $Limma$ and $ANOVA$ generally estimated the correct fold change, they were inconsistent in their estimation, resulting in a higher error rate.
-The proposed method's estimation was more consistent, with a smaller error.
-As discussed in the previous section, the summarization methods of the proposed methods excel when the number of observations are low.
-This was the case here, with each PTM only averaging $1.37$ features.
-The existing approaches used a non-robust summarization method, that was very sensitive to few observations, causing their fold change estimation to be more inconsistent, with a larger error compared to the proposed method.
-
-In~\figref{spikein_prop_volcano} we can clearly see the fold change of the red labeled spike-in peptides was only correctly estimated when accounting for changes in the unmodified protein abundance.
+Specifically, the inter-quartile range (IQR) of the estimated fold change for $MSstatsPTM$ was on average 32.5\% smaller than $Limma$ and $ANOVA$'s IQR.
+While the mean of the estimated fold changes was generally correct for all the methods, the proposed approach correctly estimated the fold change more often across all PTMs.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsubsection*{In the label-free benchmark experiment, MSstatsPTM had a higher sensitivity}
+
+\figref{spikein_boxplot} summarizes the evaluation on the label-free spike-in group comparison experiment, for all the methods, with and without adjusting for changes in the unmodified protein. 
+Without adjusting for changes in unmodified protein abundance, all the approaches incorrectly estimated the log$_2$-fold change of the modified spike-in peptides.
+After adjustment, the estimation was generally in line with the ground truth for all methods, however $MSstatsPTM$'s distribution of estimated fold changes was tighter. 
+\todo{On average over what?} the inter-quartile range of the estimates by $MSstatsPTM$ was 32.86\% smaller than that by $Limma$ or $ANOVA$. 
+As seen in the previous section, the robust summarization by $MSstatsPTM$ was particularly useful when the number of \todo{features? Biological replicates?} was low. 
+This was also the case in this experiment, with an average of $1.37$ features per PTM.
+
+\figref{spikein_prop_volcano} details the detection of differentially abundant PTM for $MSstatsPTM$ and $Limma$, with and without the adjustment for changes in the unmodified protein.
+As above, the log$_2$-fold changes of the spike-in peptides were only correctly estimated when accounting for changes in the unmodified protein abundance. 
 Additionally, the background peptides, serving as the null model, show many false positives before adjustment. 
-After adjustment the number of false positives substantially decreased.
-Specifically, for the proposed method, the number of false positives went from 20.88\% to 1.84\% after adjustment was applied.
-While the proposed method and $Limma$ both correctly estimated the fold change of the spike-in peptides, using $Limma$ resulted in many large adjusted p-values.
-This was due to $Limma$ estimating a 35.7\% larger variance compared to the proposed method when estimating the spike-in peptides.
-This experiment introduced more variance components, which is challenging for the existing methods and therefore the difference in the results is more pronounced than in the simulations.
+After adjustment the extent of false positives substantially decreased. 
+Specifically, for $MSstatsPTM$ the number of false positives went from 20.88\% to 1.84\% after the adjustment. \todo{the numbers for Limma?} 
+While the proposed method and $Limma$ both correctly estimated the fold change of the spike-in peptides, using $Limma$ resulted in many large adjusted p-values, and lower sensitivity. 
+This experiment introduced more variance components \todo{not sure what you mean. The design is group comparison, and the model has only one variance component. Rephrase?}. 
+On average \todo{over all the proteins in this experiment?}, the variance components estimates by $Limma$ were 35.7\% larger than for $MSstatsPTM$.
 Volcano plots for all methods and comparisons can be seen in {\bf Supplementary Section~\ref{supp-sec:benchmark}}.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection*{In two group comparison TMT experiments, MSstatsPTM corrected for confounding with the unmodified protein}
-
-The results of these experiments are summarized in~\figref{venn_diagrams} and~\figref{profile_plots}.
+\subsubsection*{In two biological experiments with TMT labeling, MSstatsPTM corrected for confounding with the unmodified protein}
+
+The two results of these experiments are summarized in~\figref{venn_diagrams} and~\figref{profile_plots}.
 \figref{venn_diagrams} shows the number of differentially abundant modified peptides before and after adjustment for Datasets 4 \& 5.
@@
 \caption{The $MSstatsPTM$ workflow.
-The names of the $MSstatsPTM$ R functions used for each step are highlighted in purple and the output notations are highlighted in red.
+\todo{in 6, maybe call it $J^{\prime}$ - in the figure and in the text?}The names of the $MSstatsPTM$ R functions used for each step are highlighted in purple \todo{distinguish bf and plain?} and the output notations are highlighted in red.
 The workflow begins with the acquisition of the enriched and global proteome lysates.
@@
 In contrast, the methods accounting for the confounding correctly calibrated the proportion of false positive differentially abundant PTM.
-b) In a noisy simulation, including limited feature observations and missing values, the advantage of the proposed approach was apparent.
-Looking at accuracy, the proposed method outperformed $ANOVA$ and $Limma$ in nearly every model.
+b) In a noisy simulation, including limited feature observations and missing values, $MSstatsPTM$ had higher accuracy than $ANOVA$ and $Limma$ given the same noise, number of biological replicates, and number of conditions.
 }
@@
 \caption{Dataset 3: Spike-in benchmark - Ubiquitination - Label-free.
-a) Before adjustment the fold change of the spike-in peptides' were systematically different from the expected fold change in all models.
+\todo{explain the meaning of vertical solid vs dashed lines; and of the horizontal line} a) Before adjustment the fold change of the spike-in peptides' were systematically different from the expected fold change in all models.
 After adjustment, this systemic difference was removed, however the inner quartile range of the $Limma$ and $ANOVA$ models was wider than the proposed method.

### 2022-08-12 supplementary/ptm_sm.tex
Results half-way through

--- 2022-07-28 supplementary/ptm_sm.tex
@@
 
-\title{MSstatsPTM statistical relative quantification of post-translational modifications in global proteomics experiments
-
+\title{MSstatsPTM statistical relative quantification of post-translational modifications in global proteomics experiments\\
 \textbf{Supplementary Information}
@@
 \caption{Different models that are fit depending on the experimental design.
-Models are fit for Label-free and TMT acquisition methods, as well as group comparison and time course experimental designs.
+Models are fit for label-free and TMT acquisition methods, as well as group comparison and time course experimental designs.
 The estimation of standard error and degrees of freedom for each model is shown.
@@
 
-We compared six different approaches as follows: a) proposed approach, b) proposed approach without adjusting for unmodified peptides c) ANOVA~\cite{kutner_etal_04a} (with adjustment), d) ANOVA (no adjustment), e) Limma~\cite{Smyth:2004} (with adjustment), and f) Limma (no adjustment).
+\todo{Could you check that we have the same method names and fonts for the models here as in the main text?} We compared six different approaches as follows: a) proposed approach, b) proposed approach without adjusting for unmodified peptides c) ANOVA~\cite{kutner_etal_04a} (with adjustment), d) ANOVA (no adjustment), e) Limma~\cite{Smyth:2004} (with adjustment), and f) Limma (no adjustment).
 
@@
 	\includegraphics[width=.8\textwidth]{sim_new/sim1_FC_boxplot}
-	\caption{In simulation 1 all considered methods correctly estimated the fold change between conditions, with a median fold change estimation of 1.
+	\caption{\todo{not sure about the a and b labeling style.
+Could yo label the subfigures as (a) and (b), and combine the captions into a single caption, as in the main text?} In simulation 1 all considered methods correctly estimated the fold change between conditions, with a median fold change estimation of 1.
 The distributions around the median were also consistent across all methods.}

### 2022-08-12 main/MSstatsPTM_Main.tex

@@
 \caption{The $MSstatsPTM$ workflow.
-\todo{in 6, maybe call it $J^{\prime}$ - in the figure and in the text?}The names of the $MSstatsPTM$ R functions used for each step are highlighted in purple \todo{distinguish bf and plain?} and the output notations are highlighted in red.
+\todo{in 6, maybe call it $J^{\prime}$ - in the figure and in the text?} The names of the $MSstatsPTM$ R functions used for each step are highlighted in purple and the output notations are highlighted in red.
 The workflow begins with the acquisition of the enriched and global proteome lysates.

### 2022-08-12 supplementary/ptm_sm.tex

@@
 The estimation of standard error and degrees of freedom for each model is shown.
+Only one side of the PTM and unmodified protein is shown, variance must be multiplied by two to combine.
 \label{fig:statistical_inference}}

### 2022-08-12 main/MSstatsPTM_Main.tex
More conflict resolve ;-)

@@
 \caption{The $MSstatsPTM$ workflow.
-\todo{in 6, maybe call it $J^{\prime}$ - in the figure and in the text?} The names of the $MSstatsPTM$ R functions used for each step are highlighted in purple and the output notations are highlighted in red.
+The names of the $MSstatsPTM$ R functions used for each step are highlighted in purple and the output notations are highlighted in red.
 The workflow begins with the acquisition of the enriched and global proteome lysates.

### 2022-08-12 main/MSstatsPTM_Main.tex
Small

@@
 In contrast, the methods accounting for the confounding correctly calibrated the proportion of false positive differentially abundant PTM.
-b) In a noisy simulation, including limited feature observations and missing values, $MSstatsPTM$ had higher accuracy than $ANOVA$ and $Limma$ given the same noise, number of biological replicates, and number of conditions.
+b) In a noisy simulation, which included limited feature observations and missing values, $MSstatsPTM$ had higher accuracy than $ANOVA$ and $Limma$ given the same noise, number of biological replicates, and number of conditions.
 }

### 2022-08-12 main/MSstatsPTM_Main.tex
Small

@@
 \caption{Dataset 3: Spike-in benchmark - Ubiquitination - Label-free.
-\todo{explain the meaning of vertical solid vs dashed lines; and of the horizontal line} a) Before adjustment the fold change of the spike-in peptides' were systematically different from the expected fold change in all models.
+a) Before adjustment the fold change of the spike-in peptides' were systematically different from the expected fold change in all models.
 After adjustment, this systemic difference was removed, however the inner quartile range of the $Limma$ and $ANOVA$ models was wider than the proposed method.
-b) Before adjustment the spike-in peptides (colored red) did not follow the expected log fold change; after adjustment, the spike-in peptides were more in line with expectation.
+b) \todo{explain the meaning of vertical solid vs dashed lines; and of the horizontal line}  Before adjustment the spike-in peptides (colored red) did not follow the expected log fold change; after adjustment, the spike-in peptides were more in line with expectation.
 Using $Limma$, the spike-in peptides followed the expected log fold change after adjustment, however the majority of spike-in peptides did not have a significant adjusted p-value.}

### 2022-08-15 main/MSstatsPTM_Main.tex
Results

@@
 The difference is primarily due to the robust nature of summarization in $MSstatsPTM$ through Tukey's Median Polish.
+\todo{In Background we also said that a disadvantage of the ratios is in losing data when a ratio could not be constructed.
+Could you comment on how much less data was used by ratio-based approaches, as compared to our approach?}
 
@@
 
-\figref{spikein_prop_volcano} details the detection of differentially abundant PTM for $MSstatsPTM$ and $Limma$, with and without the adjustment for changes in the unmodified protein.
+\figref{spikein_prop_volcano} details the detection of differentially abundant PTM for $MSstatsPTM$ and $Limma$, with and without the adjustment for changes in the unmodified protein, \todo{for which comparisons?}.
 As above, the log$_2$-fold changes of the spike-in peptides were only correctly estimated when accounting for changes in the unmodified protein abundance. 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection*{In two biological experiments with TMT labeling, MSstatsPTM corrected for confounding with the unmodified protein}
-
-The two results of these experiments are summarized in~\figref{venn_diagrams} and~\figref{profile_plots}.
+\subsubsection*{In two biological experiments with TMT labeling, MSstatsPTM corrected for confounding between changes in the PTM and changes in the unmodified protein}
+
+\todo{Could you separate the stories?
+One paragraph for one dataset; another paragraph (or two?) for the other dataset.}The two results of these experiments are summarized in~\figref{venn_diagrams} and~\figref{profile_plots}.
 \figref{venn_diagrams} shows the number of differentially abundant modified peptides before and after adjustment for Datasets 4 \& 5.
 Adjusting for changes in the unmodified protein caused more PTMs to change from differentially abundant to non-differentially abundant than vice versa.
-A question that must be addressed is if the decrease in differentially abundant peptides is due to the increased variance that comes from adjustment.
+A question that must be addressed is if the decrease in differentially abundant peptides is due to the increased variance \todo{variance of what?
+estimation of the variance parameter?
+Clarify} that comes from adjustment.
 This was tested by looking for modified peptides whose adjusted log fold change was within 10\% of the unadjusted log fold change but became statistically insignificant after adjustment.
@@
 \caption{Dataset 3: Spike-in benchmark - Ubiquitination - Label-free.
-a) Before adjustment the fold change of the spike-in peptides' were systematically different from the expected fold change in all models.
+\todo{Describe the details of what we see.
+The high-level conclusions are best moved to the main text} a) \todo{First explain what we see in the figure, before interpreting the conclusions} Before adjustment the fold change of the spike-in peptides' were systematically different from the expected fold change in all models.
 After adjustment, this systemic difference was removed, however the inner quartile range of the $Limma$ and $ANOVA$ models was wider than the proposed method.
-b) \todo{explain the meaning of vertical solid vs dashed lines; and of the horizontal line}  Before adjustment the spike-in peptides (colored red) did not follow the expected log fold change; after adjustment, the spike-in peptides were more in line with expectation.
-Using $Limma$, the spike-in peptides followed the expected log fold change after adjustment, however the majority of spike-in peptides did not have a significant adjusted p-value.}
+b) \todo{explain the meaning of vertical solid vs dashed lines; and of the horizontal line; explain that only a subset of comparisons and methods}  Before adjustment the spike-in peptides (colored red) did not follow the expected log fold change; after adjustment, the spike-in peptides were more in line with expectation.
+Using $Limma$, the spike-in peptides followed the expected log fold change after adjustment, however the adjusted p-value for the majority of spike-in peptides were below the cutoff.}
 \label{fig:spikein_volcano}
@@
 \caption{
+The overlap of differential modified peptides for the PTM model with and without global protein level adjustment across all pairwise comparisons.
 a) Dataset 4: Human - Ubiquitination - 1mix-TMT.
-The overlap of differential modified peptides for the PTM model with and without global protein level adjustment across all comparisons.
-3,236 modified peptides became insignificant, 1,352 became significant, while 4,282 were significant in both models.
-b) Dataset 5: Mouse - Phosphorylation - 2mix-TMT.
-The overlap of differentially modified peptides between the PTM model with and without global protein level adjustment across all comparisons.
-19,286 peptides became insignificant, 4,947 became significant, and 41,552 were significant in both models.
+T b) Dataset 5: Mouse - Phosphorylation - 2mix-TMT.
 }
@@
 	 \end{subfigure}
-\caption{a) Dataset 5: Mouse - Phosphorylation - 2mix-TMT.
+\caption{ \todo{Describe the details of what we see.
+The high-level conclusions are best moved to the main text} a) Dataset 5: Mouse - Phosphorylation - 2mix-TMT.
 Comparing the global profiling of protein $TTP$ with the modification at site $S178$.

### 2022-08-15 supplementary/ptm_sm.tex
Results

@@
 	\caption{\todo{not sure about the a and b labeling style.
-Could yo label the subfigures as (a) and (b), and combine the captions into a single caption, as in the main text?} In simulation 1 all considered methods correctly estimated the fold change between conditions, with a median fold change estimation of 1.
+Could yo label the subfigures as (a) and (b), and combine the captions into a single caption, as in the main text?
+Use the same code names for the datasets as in the main text} In simulation 1 all considered methods correctly estimated the fold change between conditions, with a median fold change estimation of 1.
 The distributions around the median were also consistent across all methods.}

### 2022-08-16 main/MSstatsPTM_Main.tex
evaluation comments

@@
 
-\todo{I am wondering if 'basic' is the right vocabulary here.
-Would you like to switch to 'unadjusted ANOVA', 'adjusted ANOVA', 'unadjusted Limma' and 'adjusted Limma', everywhere in the text and in the figures?
-}
-
-$MSstatsPTM$ before adjusting for changes in the unmodified protein corresponds to base $MSstats$ or $MSstatsTMT$, modeled on the peptide level \todo{as described in Suppl Table XXX, as appropriate for the experimental design}.
-$MSstatsPTM$ with the adjustment, \todo{described in Sec XXX}, was used without imputing missing values, and without Empirical Bayes moderation.
-
-$Basic$ $ANOVA$, i.e. $ANOVA$ before adjusting for changes in the unmodified protein, was as in \eqref{ttest_model}. $ANOVA$ with the adjustment was modeled as in \eqref{ttest_ratio_model}. 
-Finally, $basic$ $Limma$ (i.e.
-without $Limma$ adjustment) used the same model formula in \eqref{ttest_model}, while including a moderated variance estimation.
-$Limma$ with adjustment was modeled as in \eqref{ttest_ratio_model} including moderated variance estimation.
+$MSstatsPTM$ before adjusting for changes in the unmodified protein corresponds to base $MSstats$ or $MSstatsTMT$, modeled on the peptide level as described in {\bf Supplementary Table \ref{supp-fig:statistical_inference}}, as appropriate for the experimental design.
+$MSstatsPTM$ with the adjustment, described in \secref{adjustment_summary}, was used without imputing missing values, and without Empirical Bayes moderation.
+
+$Unadjusted$ $ANOVA$, i.e.
+$ANOVA$ before adjusting for changes in the unmodified protein, was as in \eqref{ttest_model}.
+$Adjusted$ $ANOVA$, i.e.
+$ANOVA$ with the adjustment was modeled as in \eqref{ttest_ratio_model}.
+Finally, $unadjusted$ $Limma$ used the same model formula in \eqref{ttest_model}, while including a moderated variance estimation.
+$Adjusted$ $Limma$ was modeled as in \eqref{ttest_ratio_model} including moderated variance estimation.
 All the evaluations were done at the FDR-adjusted p-value cutoff of $q=.05$. More details are in {\bf Supplementary Sec. \ref{supp-sec:experiments}}.
@@
 
-\figref{sim1_fdr} summarizes the eFDR reported on Computer Simulation 1 dataset by $MSstatsPTM$, $ANOVA$, $Limma$ and the basic $MSstats$, $ANOVA$ and $Limma$ methods.
+\figref{sim1_fdr} summarizes the eFDR reported on Computer Simulation 1 dataset by $MSstatsPTM$, $adjusted$ $ANOVA$, $adjusted$ $Limma$ and base $MSstats$, $unadjusted$ $ANOVA$ and $unadjusted$ $Limma$ methods.
 The simulation mimicked a ``clean" label-free group comparison experiment, not compromised by issues such as deviations from model assumptions, missing values and outliers.
@@
 The versions of the models that accounted for the confounding produced error rates that were much better calibrated at the desired level.
-\todo{Since the simulation mimicked the assumptions of model XXX?, the error rates of this model were calibrated best.}
+
+%Since the simulation mimicked the assumptions of model XXX?, the error rates of this model were calibrated best.
 
@@
 The difference is primarily due to the robust nature of summarization in $MSstatsPTM$ through Tukey's Median Polish.
-\todo{In Background we also said that a disadvantage of the ratios is in losing data when a ratio could not be constructed.
-Could you comment on how much less data was used by ratio-based approaches, as compared to our approach?}
+The ratio-based summarization approach used by $Limma$ and $ANOVA$ requires measurements for both the PTM and unmodified protein.
+$MSstatsPTM$ in contrast can still leverage the information in the PTM or unmodified protein if one is missing.
+On average over all simulations, 3.94\% of the total unmodified protein data used by $MSstatsPTM$ was thrown out by $Limma$ and $ANOVA$ due to missing PTM data.
+No PTM data was thrown out due to missing protein data.
 
@@
 After adjustment, the estimation was generally in line with the ground truth for all methods, however $MSstatsPTM$'s distribution of estimated fold changes was tighter. 
-\todo{On average over what?} the inter-quartile range of the estimates by $MSstatsPTM$ was 32.86\% smaller than that by $Limma$ or $ANOVA$. 
-As seen in the previous section, the robust summarization by $MSstatsPTM$ was particularly useful when the number of \todo{features? Biological replicates?} was low. 
+On average over all comparisons the inter-quartile range of the estimates by $MSstatsPTM$ was 32.86\% smaller than that by $Limma$ and $ANOVA$. 
+As seen in the previous section, the robust summarization by $MSstatsPTM$ was particularly useful when the number of features was low. 
 This was also the case in this experiment, with an average of $1.37$ features per PTM.
-
-\figref{spikein_prop_volcano} details the detection of differentially abundant PTM for $MSstatsPTM$ and $Limma$, with and without the adjustment for changes in the unmodified protein, \todo{for which comparisons?}.
+Additionally, the ratio based summarization used by $Limma$ and $ANOVA$ discarded more data than in the simulations.
+The ratio based summarization lost 9.63\% of the unmodified protein and 1.94\% of the PTM data used by $MSstatsPTM$.
+This additional information lead $MSstatsPTM$ to a more accurate fold change estimation and better calibrated variance.
+
+\figref{spikein_prop_volcano} details the detection of differentially abundant PTM for $MSstatsPTM$ and $Limma$, with and without the adjustment for changes in the unmodified protein, for the $mix3$-$mix2$ and $mix4$-$mix2$ comparisons.
 As above, the log$_2$-fold changes of the spike-in peptides were only correctly estimated when accounting for changes in the unmodified protein abundance. 
@@
 After adjustment the extent of false positives substantially decreased. 
-Specifically, for $MSstatsPTM$ the number of false positives went from 20.88\% to 1.84\% after the adjustment. \todo{the numbers for Limma?} 
+Specifically, for $MSstatsPTM$ the number of false positives went from 20.88\% to 1.84\% after the adjustment, and for $Limma$ went from 26.04\% to 1.18\%.
 While the proposed method and $Limma$ both correctly estimated the fold change of the spike-in peptides, using $Limma$ resulted in many large adjusted p-values, and lower sensitivity. 
-This experiment introduced more variance components \todo{not sure what you mean. The design is group comparison, and the model has only one variance component. Rephrase?}. 
-On average \todo{over all the proteins in this experiment?}, the variance components estimates by $Limma$ were 35.7\% larger than for $MSstatsPTM$.
+This was mainly due to $Limma$ estimating a higher variance, even when including variance moderation.
+On average, over all the PTMs in this experiment, the variance components estimated by $Limma$ were 35.7\% larger than for $MSstatsPTM$.
 Volcano plots for all methods and comparisons can be seen in {\bf Supplementary Section~\ref{supp-sec:benchmark}}.
@@
 
-\todo{Could you separate the stories?
-One paragraph for one dataset; another paragraph (or two?) for the other dataset.}The two results of these experiments are summarized in~\figref{venn_diagrams} and~\figref{profile_plots}.
-\figref{venn_diagrams} shows the number of differentially abundant modified peptides before and after adjustment for Datasets 4 \& 5.
+The results of Dataset 4: Human - Ubiquitination - 1mix-TMT are summarized in~\figref{data4_venn_diagram} and~\figref{data4_profile_plot}.
+\figref{data4_venn_diagram} shows the number of differentially abundant modified peptides before and after adjustment.
 Adjusting for changes in the unmodified protein caused more PTMs to change from differentially abundant to non-differentially abundant than vice versa.
-A question that must be addressed is if the decrease in differentially abundant peptides is due to the increased variance \todo{variance of what?
-estimation of the variance parameter?
-Clarify} that comes from adjustment.
+A question that must be addressed is if the decrease in differentially abundant peptides is due to the increased standard error that comes from adjustment, as in \eqref{prop_se}.
 This was tested by looking for modified peptides whose adjusted log fold change was within 10\% of the unadjusted log fold change but became statistically insignificant after adjustment.
-In other words, the fold change was the same between models but variance increased.
-When the test was applied to Dataset 4, only one PTM became non-differentially abundant due to an increase in variance.
-When applied to Dataset 5, 548 PTMs became non-differentially abundant due to an increase in variance (only 3.4\% of all PTMs that became non-differentially abundant).
+In other words, the fold change was the same between models but the standard error increased.
+When the test was applied to Dataset 4, only one PTM became non-differentially abundant due to an increase in standard error.
 Thus we can conclude that the drop off in differentially abundant PTMs was due to changes in global protein abundance, not simply increased model variance.
 
-\figref{venn_diagrams} demonstrates that the unmodified protein was more likely to cause the observed changes in PTM abundance, rather than mask changes in the PTM.
-The later case can occur when the unmodified protein exhibits a change in abundance, while the modification does not.
-Additionally, this case may come about due to the high variability of PTMs compared to the global protein.
-An example of this can be seen in \figref{data5_profile_plot}.
-Originally the modification at $S178$ of protein $TTP$ was shown to be differentially abundant between KO\_Uninfect and KO\_Early, with a log fold change of $2.9$.
-However the unmodified protein was shown to contribute $2.014$ of this change, meaning that the modification only accounted for $.886$ after adjustment (seen in {\bf Supplementary Figure~\ref{supp-fig:No_Diff_Shigella_PTM}}).
-This caused the modification to go from differentially abundant to non-differentially abundant.
-These cases occur frequently and can be thought of as false positives.
-The proposed approach corrects for this confounding, allowing us to see the true fold change of PTMs, while excluding those that do not exhibit differential changes.
-
+\figref{data4_venn_diagram} demonstrates that the unmodified protein was more likely to cause the observed changes in PTM abundance, rather than mask changes in the PTM.
 Conversely, cases where the unmodified protein masks changes in the PTM are more infrequent.
@@
 
+The results of Dataset 5: Mouse - Phosphorylation - 2mix-TMT are summarized in~\figref{data5_venn_diagram} and~\figref{data5_profile_plot}.
+\figref{data5_venn_diagram} showed that adjusting for changes in the unmodified protein caused more PTMs to change from differentially abundant to non-differentially abundant.
+Again we tested if this change was due to fold change adjustment or an increased standard error.
+When applied to Dataset 5, 548 PTMs became non-differentially abundant due to an increase in standard error (only 3.4\% of all PTMs that became non-differentially abundant).
+
+\figref{data5_profile_plot} shows a case where the unmodified protein caused the observed changes in PTM abundance.
+This occurs when the protein exhibits a change in abundance while the modification does not.
+Originally the modification at $S178$ of protein $TTP$ was shown to be differentially abundant between KO\_Uninfect and KO\_Early, with a log fold change of $2.9$.
+However the unmodified protein was shown to contribute $2.014$ of this change, meaning that the modification only accounted for $.886$ after adjustment (seen in {\bf Supplementary Figure~\ref{supp-fig:No_Diff_Shigella_PTM}}).
+This caused the modification to go from differentially abundant to non-differentially abundant.
+These cases occur frequently and can be thought of as false positives.
+The proposed approach corrects for this confounding, allowing us to see the true fold change of PTMs, while excluding those that do not exhibit differential changes.
+
+%Adjusting for changes in the unmodified protein caused more PTMs to change from differentially abundant to non-differentially abundant than vice versa.
+A question that must be addressed is if the decrease in differentially abundant peptides is due to the increased variance \todo{variance of what?
+estimation of the variance parameter?
+Clarify} that comes from adjustment.
+This was tested by looking for modified peptides whose adjusted log fold change was within 10\% of the unadjusted log fold change but became statistically insignificant after adjustment.
+In other words, the fold change was the same between models but variance increased.
+When the test was applied to Dataset 4, only one PTM became non-differentially abundant due to an increase in variance.
+When applied to Dataset 5, 548 PTMs became non-differentially abundant due to an increase in variance (only 3.4\% of all PTMs that became non-differentially abundant).
+Thus we can conclude that the drop off in differentially abundant PTMs was due to changes in global protein abundance, not simply increased model variance.
+
+%\figref{venn_diagrams} demonstrates that the unmodified protein was more likely to cause the observed changes in PTM abundance, rather than mask changes in the PTM.
+The later case can occur when the unmodified protein exhibits a change in abundance, while the modification does not.
+Additionally, this case may come about due to the high variability of PTMs compared to the global protein.
+An example of this can be seen in \figref{data5_profile_plot}.
+Originally the modification at $S178$ of protein $TTP$ was shown to be differentially abundant between KO\_Uninfect and KO\_Early, with a log fold change of $2.9$.
+However the unmodified protein was shown to contribute $2.014$ of this change, meaning that the modification only accounted for $.886$ after adjustment (seen in {\bf Supplementary Figure~\ref{supp-fig:No_Diff_Shigella_PTM}}).
+This caused the modification to go from differentially abundant to non-differentially abundant.
+These cases occur frequently and can be thought of as false positives.
+The proposed approach corrects for this confounding, allowing us to see the true fold change of PTMs, while excluding those that do not exhibit differential changes.
+
 %\subsubsection*{In the two-mi[xture TMT experiment, MSstatsPTM removed confounding \todo{OV: between what and what? Is it similar to the previous section?}}
@@
 When looking only at PTMs with a corresponding unmodified protein, the number of differentially modified PTMs exhibited a small drop, but was more consistent.
-As above, we tested if this drop in differentially abundant PTMs was due to an increase in variance.
-Here only 25 PTMs became insignificant due to increased variance.
+As above, we tested if this drop in differentially abundant PTMs was due to an increase in standard error.
+Here only 25 PTMs became insignificant due to increased standard error.
 Without a global profiling run, the method was less effective at removing confounding with the unmodified protein.
@@
 \caption{Dataset 3: Spike-in benchmark - Ubiquitination - Label-free.
-\todo{Describe the details of what we see.
-The high-level conclusions are best moved to the main text} a) \todo{First explain what we see in the figure, before interpreting the conclusions} Before adjustment the fold change of the spike-in peptides' were systematically different from the expected fold change in all models.
-After adjustment, this systemic difference was removed, however the inner quartile range of the $Limma$ and $ANOVA$ models was wider than the proposed method.
-b) \todo{explain the meaning of vertical solid vs dashed lines; and of the horizontal line; explain that only a subset of comparisons and methods}  Before adjustment the spike-in peptides (colored red) did not follow the expected log fold change; after adjustment, the spike-in peptides were more in line with expectation.
-Using $Limma$, the spike-in peptides followed the expected log fold change after adjustment, however the adjusted p-value for the majority of spike-in peptides were below the cutoff.}
+a) The distribution of spike-in peptides log$_2$ fold change estimated by $MSstatsPTM$, $Limma$, and $ANOVA$ with and without adjustment.
+The expected log$_2$ fold change is highlighted by a red `X'.
+Protein adjustment removed systematic differences from the expected log$_2$ fold change in all models.
+b) The statistical results of $MSstatsPTM$ and $Limma$ modeling the $mix3$-$mix2$ and $mix4$-$mix2$ comparisons before and after adjustment.
+The spike-in peptides are colored red, while the background peptides are highlighted grey.
+The solid horizontal line shows the adjusted pvalue cutoff of .05.
+The solid vertical line shows log$_2$ fold change of 0.
+The dashed vertical line shows the expected log$_2$ fold change of the spike-in peptides.
+Before adjustment the spike-in peptides (colored red) did not follow the expected log fold change; after adjustment, the spike-in peptides were more in line with expectation.}
 \label{fig:spikein_volcano}
@@
  \centering
-\includegraphics[width=1.0\textwidth]{images/No_Difference_Shigella_Profile_Plot}
+\includegraphics[width=1.0\textwidth]{images/IpaH_prof_plot.png}
 \caption{}
@@
  \centering
-	\includegraphics[width=1.0\textwidth]{images/IpaH_prof_plot.png}
+	\includegraphics[width=1.0\textwidth]{images/No_Difference_Shigella_Profile_Plot}
 	\caption{}
@@
 	 \end{subfigure}
-\caption{ \todo{Describe the details of what we see.
-The high-level conclusions are best moved to the main text} a) Dataset 5: Mouse - Phosphorylation - 2mix-TMT.
-Comparing the global profiling of protein $TTP$ with the modification at site $S178$.
-The modification and unmodified protein abundance followed the same trend across conditions.
-Specifically, there was a positive adjustment in abundance when comparing WT\_Uninfect to WT\_Late in both the modification and global profiling run.
-b) Dataset 4: Human - Ubiquitination - 1mix-TMT.
-Comparing the global profiling of protein $GSDMD$ with the ubiquitination of the protein at site $K62$.
-Here the modification and global protein follow different trends.
-There appeared to be no change in abundance between Dox1hr and Dox4hr in the modified plot, however there was a large negative change when looking at the unmodified plot.
-This indicated the modification was confounded with changes in the unmodified protein.}
+\caption{a) Dataset 4: Human - Ubiquitination - 1mix-TMT.
+Comparing the $GSDMD$ protein and its ubiquitination at site $K62$ across all conditions.
+The individual PTM and protein features are shown in grey, while the summarization is highlighted in red.
+b) Dataset 5: Mouse - Phosphorylation - 2mix-TMT.
+Comparing the profiling of protein $TTP$ with the phosphorylation at site $S178$.
+The individual PTM and protein features are shown in grey, while the summarization is highlighted in red.
+The plots are separated into TMT Mixtures 1 and 2.}
 \label{fig:profile_plots}
@@
  \caption{Dataset 6: Human - Ubiquitination - Label-free no global profiling run.
-a) The overlap of differential modified peptides for the PTM model with and without global protein level adjustment across all comparisons.
-More PTMs became insignificant than became significant after adjustment.
-In total, 2709 modified peptides became insignificant, while only 547 became significant.
-b) Here we made the same comparison but only looked at modified peptides where adjustment could be performed, ie they had a matching unmodified protein.
-In this case there were significantly less peptides that became insignificant after adjustment.
-726 modified peptides became insignificant, 547 became significant, and 1,078 were significant in both models.}
+a) The overlap of differential modified peptides for the PTM model with and without global protein level adjustment across all comparisons, including all measured PTMs.
+b) The overlap of differential modified peptides with and without global protein level adjustment, including only PTMs with a corresponding unmodified protein.}
 \label{fig:data6_plots}

### 2022-08-16 supplementary/ptm_sm.tex
evaluation comments

--- 2022-08-12 supplementary/ptm_sm.tex
@@
 
-\todo{Could you check that we have the same method names and fonts for the models here as in the main text?} We compared six different approaches as follows: a) proposed approach, b) proposed approach without adjusting for unmodified peptides c) ANOVA~\cite{kutner_etal_04a} (with adjustment), d) ANOVA (no adjustment), e) Limma~\cite{Smyth:2004} (with adjustment), and f) Limma (no adjustment).
+We compared six different approaches as follows: a) $MSstatsPTM$, b) base $MSstats$ c) $adjusted$ $ANOVA$~\cite{kutner_etal_04a}, d) $Unadjusted$ $ANOVA$, e) $adjusted$ $Limma$~\cite{Smyth:2004}, and f) $unadjusted$ $Limma$.
 
@@
 
-\begin{figure}[h!]
+\begin{figure}[ht]
 \centering
@@
 	\includegraphics[width=.8\textwidth]{sim_new/sim1_FC_boxplot}
-	\caption{\todo{not sure about the a and b labeling style.
-Could yo label the subfigures as (a) and (b), and combine the captions into a single caption, as in the main text?} In simulation 1 all considered methods correctly estimated the fold change between conditions, with a median fold change estimation of 1.
-The distributions around the median were also consistent across all methods.}
+	\caption{}
+	\label{fig:fc_boxplota}
  \end{subfigure}
@@
 	\includegraphics[width=.8\textwidth]{sim_new/sim3_FC_boxplot}
-	\caption{In simulation 2 all methods correctly estimated the fold change with a median log change of 1.
-The proposed method in this simulation had a visibly tighter distribution around the median.
-Both Limma and ANOVA showed a wider range around the fold change.}
+	\caption{}
+	\label{fig:fc_boxplotb}
  \end{subfigure}
-\caption{Fold change distribution comparison between Simulations 1 and 2.}
+\caption{Fold change distribution comparison between Simulations 1 and 2.
+a) In simulation 1 all considered methods correctly estimated the fold change between conditions, with a median fold change estimation of 1.
+The distributions around the median were also consistent across all methods.
+b) In simulation 2 all methods correctly estimated the fold change with a median log change of 1.
+$MSstatsPTM$ in this simulation had a visibly tighter distribution around the median.
+Both $Limma$ and $ANOVA$ showed a wider range around the fold change.}
 \label{fig:fc_boxplot}
@@
 \includegraphics[width=1\textwidth]{sim_new/spike_in_msstatsptm_volcano}
-\caption{The modeling results of the proposed method both before and after adjustment.
+\caption{The modeling results of the $MSstatsPTM$ both before and after adjustment.
 The spike-in peptides are colored red and the background peptides are colored grey.
@@
 \includegraphics[width=1\textwidth]{sim_new/spike_in_limma_volcano}
-\caption{When modeling the experiment with the Limma method, the spike-in peptides again follow the expected log fold change better after adjusting for changes in protein level.
+\caption{When modeling the experiment with the $Limma$ method, the spike-in peptides again follow the expected log fold change better after adjusting for changes in protein level.
 However, while the fold change was more accurate, the majority of spike-in peptides did not have a significant adjusted pvalue.
@@
 \includegraphics[width=1\textwidth]{sim_new/spike_in_anova_volcano}
-\caption{Using ANOVA the fold change of the spike-in peptides was much closer to expectation after adjusting for global protein abundance.
-The log FC estimation was the same as Limma, however the pvalues were different.
-We actually see a few more True Positives using Anova compared to Limma.\label{fig:spike_volcano_ttest}}
+\caption{Using $ANOVA$ the fold change of the spike-in peptides was much closer to expectation after adjusting for global protein abundance.
+The log FC estimation was the same as $Limma$, however the pvalues were different.
+We actually see a few more True Positives using $ANOVA$ compared to $Limma$.\label{fig:spike_volcano_ttest}}
 \end{figure}
@@
 	\includegraphics[width=.9\textwidth]{sim_new/Difference_Shigella_Profile_Plot}
-	\caption{The global profiling of protein $KI67\_MOUSE|E9PVX6$ with the modification at site $T215$ was compared.
-In this case the modification and global protein appeared to show small or no difference between conditions, however after adjusting for change in global protein abundance, the modification was statistically significant.
-Additionally, this profile plot showed the large difference in available features between modified peptides and global proteins.}
+	\caption{}
  \end{subfigure}
@@
 	\includegraphics[width=.9\textwidth]{sim_new/Difference_Shigella_Volcano}
-	\caption{The volcano plot of the WT\_Uninfect and WT\_Early comparison showed the specifics of the adjustment.
+	\caption{}
+ \end{subfigure}
+ \caption{Summary plots for modification of protein KI67 at site T215.
+a) The global profiling of protein $KI67\_MOUSE|E9PVX6$ with the modification at site $T215$ was compared.
+In this case the modification and global protein appeared to show small or no difference between conditions, however after adjusting for change in global protein abundance, the modification was statistically significant.
+Additionally, this profile plot showed the large difference in available features between modified peptides and global proteins.
+b) The volcano plot of the WT\_Uninfect and WT\_Early comparison showed the specifics of the adjustment.
 The modification looked to be flat, with a log fold change of $.185$, while the global profiling showed a small negative fold change of $-.619$.
 While both exhibit small changes, when combined we saw a log fold change of $.804$ and adjusted p-value of $.0156$.}
- \end{subfigure}
- \caption{Summary plots for modification of protein KI67 at site T215.}
 \label{fig:Diff_Shigella_PTM}
@@
 In datasets 4 and 5 the variance of the PTM was higher than the global protein.
-In dataset 6 the variance of the PTM and Protein were generally the same.
-We mimicked these scenarios and analyzed the power of experiments when the PTM variance was higher than the protein and when they were equal.
-When the PTM and protein were the same we chose a variance of $.15$, whereas when the PTM was higher than the protein we chose a PTM variance of $.2$ and a protein variance of $.1$.
+In dataset 6 the variance of the PTM and global protein were generally the same.
+We mimicked these scenarios and analyzed the power of experiments when the PTM variance was higher than the global protein and when they were equal.
+When the PTM and global protein were the same we chose a variance of $.15$, whereas when the PTM was higher than the global protein we chose a PTM variance of $.2$ and protein variance of $.1$.
 The results of the power and sample size analysis can be seen in Figure \ref{fig:power_sd_combo}.
@@
 	\includegraphics[width=.75\textwidth]{sim_new/same_var_power}
-	\caption{The power of an experiment targeting PTMs with the same variance, .15, for the modified and unmodified peptides.
-Predictably when the replicates were high for both modified and unmodified peptides the power was much higher.
-Conversely at low replicates for each the power was much lower.
-With equal variance, it did not matter if the PTM replicates or protein replicates were higher.}
+	\caption{}
  \end{subfigure}\vspace{-1mm}
@@
 	\includegraphics[width=.75\textwidth]{sim_new/high_ptm_var_power}
-	\caption{In this chart the variance for the PTM was higher than the unmodified protein.
+	\caption{}
+ \end{subfigure}\vspace{-1mm}
+ \caption{Power analysis of experiments with differing variances.
+a) The power of an experiment targeting PTMs with the same variance, .15, for the modified and unmodified peptides.
+Predictably when the replicates were high for both modified and unmodified peptides the power was much higher.
+Conversely at low replicates for each the power was much lower.
+With equal variance, it did not matter if the PTM replicates or protein replicates were higher.
+b) In this chart the variance for the PTM was higher than the unmodified protein.
 The PTM variance was .2, while the unmodified protein variance was .1.
@@
 When the replicates were not equal, more replicates allocated to the PTM runs lead to higher power.}
- \end{subfigure}\vspace{-1mm}
- \caption{Power analysis of experiments with differing variances.}
 \label{fig:power_sd_combo}

### 2022-08-16 code/MSstatsPTM_SpikeIn_Analysis.R

@@ -521,6 +521,7 @@ model_df <- groupComparisonPTM(summary_df, data.type = "LabelFree",
                                use_log_file=FALSE)
 
 save(model_df, file = "D:\\Northeastern\\Research\\MSstats\\MSstatsPTM-manuscript\\data\\Spike_in_MSstatsPTM_Model.rda")
+load(file = "D:\\Northeastern\\Research\\MSstats\\MSstatsPTM-manuscript\\data\\Spike_in_MSstatsPTM_Model.rda")
 
 
 ## Model Plots -----------------------------------------------------------------
@@ -665,16 +666,30 @@ for (r in seq_along(runs)){
   summarized_proteins <- rbindlist(list(summarized_proteins, sum_runs_prot))
 }
 
-# summarized_ptm <- summarized_ptm %>%
-#   mutate(Abundance = ifelse(str_detect(Run, "mix3|mix4"),
-#                             Abundance - 1, Abundance))
-
 summarized_ptm$PTM <- summarized_ptm$ProteinName
 summarized_ptm$ProteinName <- sapply(summarized_ptm$PTM,
                                      function(x) {str_split(x, "_",2)[[1]][1]})
 joined <- merge(summarized_ptm, summarized_proteins,
                 by = c("ProteinName", "Run", "Condition"), all.x = TRUE)
 joined$Adj_Abundance <- joined$Abundance.x - joined$Abundance.y
+
+adjustable = model_df$ADJUSTED.Model %>% distinct(Protein, Label) %>% separate(Label, c("Condition1", "Condition2"), sep = " vs ")
+adjustable1 = adjustable %>% select(Protein, Condition1)
+adjustable1$Condition = adjustable1$Condition1
+adjustable2 = adjustable %>% select(Protein, Condition2)
+adjustable2$Condition = adjustable2$Condition2
+
+adjustable = rbind(adjustable1, adjustable2, fill = TRUE) %>% select(Protein, Condition)
+
+## PTM missing rows. non-unique
+joined %>% filter(is.infinite(Abundance.x) & is.finite(Abundance.y)
+                  & PTM %in% adjustable$Protein & Condition %in% adjustable$Condition) %>% nrow() / joined %>% nrow()
+
+## PTM missing rows
+joined %>% filter(is.infinite(Abundance.y) & is.finite(Abundance.x)
[…]

### 2022-08-19 main/MSstatsPTM_Main.tex
First final

@@
 Frequently data from these experiments are analyzed using statistical methods that were not originally designed for this task.
-Researchers use methods such as $t$-test\cite{Kalpic:2011}, Analysis of Variance ($ANOVA$)\cite{girden:1992}, or $Limma$\cite{Ritchie_15a}, by taking as input the intensity ratios of modified and unmodified peptide features, and comparing the mean abundance of different PTM sites.
+Researchers use methods such as $t$-test\cite{Kalpic:2011}, Analysis of Variance ($ANOVA$)\cite{girden:1992}, or $Limma$\cite{Ritchie_15a}, by taking as input the intensity ratios of modified and unmodified peptide summaries, and comparing the mean abundance of different PTM sites.
 Such approaches do not fully account for all the sources of uncertainty.
@@
 The statistical methodology separately summarizes the abundances of peptides with and without the modification sites, and fits separate linear mixed effects models that reflect the biological and technological aspects of the experimental design.
-Next, model-based inferences regarding the PTM and the unmodified protein-level abundances are combined to account for the confounding between these two sources.
+Next, model-based inferences regarding the PTM and the unmodified protein-level summaries are combined to account for the confounding between these two sources.
 
@@
 The datasets illustrate a diverse set of organisms, modification types, acquisition methods and experimental designs, showing the applicability of the framework to a variety of situations.
-By appropriately leveraging the information from the unmodified proteins, the proposed approach improved the accuracy of the estimates of PTM fold changes, and produced a better calibrated false positive rate of detecting differentially abundant PTMs as compared to existing methods.
+By appropriately leveraging the information from the unmodified portion of the protein sequence, the proposed approach improved the accuracy of the estimates of PTM fold changes, and produced a better calibrated false positive rate of detecting differentially abundant PTMs as compared to existing methods.
 In particular, accounting for the confounding from unmodified protein abundance allowed us to characterize the true effect of the modification, avoiding the need for more manual and time intensive follow-up investigation.
@@
 The PTMs of 500 proteins had a differential fold change between conditions, while the other 500 proteins were generated with no changes in abundance between conditions.
-Furthermore, the fold changes of half of the 500 differential PTMs were fully masked by changes in the unmodified protein.
-Finally, the fold change of half the 500 non-differential PTMs was entirely due to changes in the unmodified protein.
+Furthermore, the fold changes of half of the 500 differential PTMs were fully masked by changes in the unmodified portion of the protein.
+Finally, the fold change of half the 500 non-differential PTMs was entirely due to changes in the unmodified portion of the protein.
 All the differential PTMs were generated with an expected log base 2 fold change of 0.75 between conditions.
@@
 The feature counts and the proportion of missing values were as observed on average over all the the experimental datasets in this manuscript.
-Specifically, PTMs were simulated with 2 modified peptide features, and unmodified proteins were simulated with 10 features.
+Specifically, PTMs were simulated with 2 modified peptide features, and unmodified portions of the protein were simulated with 10 features.
 Additionally, 20\% of observations for both modified and unmodified peptides were missing completely at random.
@@
 
-\medskip \noindent {\bf Data acquisition:} This experiment did not include a separate global profiling run to measure unmodified peptides.
+\medskip \noindent {\bf Data acquisition:} This experiment did not include a separate global profiling run to quantify unmodified peptides.
 In addition to low feature counts for unmodified peptides,  this lead to substantially fewer matches between modified and unmodified peptides.
@@
 The log$_2$-intensity of Feature $k$, in Replicate $j$ of Condition $i$ is denoted by $y_{ijk}^{\ast}$.
-Conversely, the unmodified protein is represented by $L$ spectral features, and the log-intensity of Feature $l$ from the unmodified protein in the same run is denoted by $y_{ijl}$.
+Conversely, the unmodified portion of the protein is represented by $L$ spectral features, and the log-intensity of Feature $l$ from the unmodified portion of the protein in the same run is denoted by $y_{ijl}$.
 The features can be quantified as part of a same mass spectrometry run, or in a separate enrichment and global proteome profiling run.
@@
 The estimates are used to test the null hypothesis in \eqref{conv_null_hyp} by comparing the model-based test statistic against the Student distribution with $df=I(J-1)$ degrees of freedom in balanced designs.
-Unfortunately, this approach is fundamentally flawed as it does not account for the confounding between changes in the PTM abundance and the overall changes in the abundance of the unmodified protein.
+Unfortunately, this approach is fundamentally flawed as it does not account for the confounding between changes in the PTM abundance and the overall changes in the abundance of the unmodified portion of the protein.
 
@@
 \end{eqnarray}
-Assuming that the sources of variation in the modified features that are unexplained by the model are independent from the corresponding sources of variation in the modified features, the standard error $\widehat{SE}(\hat{\Delta}_{adj})$ is obtained by combining the standard errors from the unmodified and modified model fits
+Assuming that the sources of variation in the summaries of modified features that are unexplained by the model are independent from the sources of variation in the summaries of modified features, the standard error $\widehat{SE}(\hat{\Delta}_{adj})$ is obtained by combining the standard errors from the two model fits
 \begin{eqnarray}
@@
 The procedure has been described in general in~\cite{kutner_etal_04a}, and for protein significance analysis specifically in~\cite{oberg_vitek_09a}. 
-It requires us to specify the desired levels of the following quantities: a) $q$, the False Discovery Rate of detecting differential abundance, b) $\beta$, the average Type II error rate, c) $\Delta_{adj}$, the minimal log$_2$-fold change in adjusted PTM abundance  of interest, d) $m_0 / (m_0 + m_1)$, the fraction of truly differentially modified PTM sites in the comparison, and e) $\sigma_{PTM}^{2}$ and $\sigma_{protein}^{2}$, the anticipated variances associated with the summaries of the modified and unmodified features, respectively.
-Typically, the variances are estimated from an existing experiment, conducted with the same biological material and measurement workflow.
-Given the above quantities, and assuming a balanced design, the minimal number of replicates $J$ across $I$ conditions is chosen to bound the variance of the estimated log$_2$-fold change $\mathrm{SE}^{2}(\Delta_{adj})$:
+It requires us to specify the desired levels of the following quantities: a) $q$, the False Discovery Rate of detecting differential abundance, b) $\beta$, the average Type II error rate, c) $\Delta_{adj}$, the minimal log$_2$-fold change in adjusted PTM abundance  of interest, d) $m_0 / (m_0 + m_1)$, the fraction of truly differentially modified PTM sites in the comparison, and e) $\sigma_{PTM}^{2}$ and $\sigma_{protein}^{2}$, the anticipated variances associated with the pairwise comparisons of conditions for the modified and unmodified protein summaries, respectively.
+Typically, these variances are estimated from an existing experiment, conducted with the same biological material and measurement workflow.
+\todo{The estimation of the variance components for pairwise comparisons of conditions depends on the experimental design, and is detailed in Supply Figure S1}
+
+Given the above quantities, and assuming a balanced design that allocates the same number of replicates $J^{\prime}$ to both modified and unmodified profiling runs, the minimal number of replicates $J^{\prime}$ for each of $I$ conditions is chosen to bound the variance of the estimated log$_2$-fold change $\mathrm{SE}^{2}(\Delta_{adj})$:
 
 \begin{equation}
-SE(\Delta_{adj})^2 = \left[ \frac{2}{J} \left( \sigma_{PTM}^{2} + \sigma_{protein}^{2} \right) \right]
+SE(\Delta_{adj})^2 = \left[ \frac{2}{J^{\prime}} \left( \sigma_{PTM}^{2} + \sigma_{protein}^{2} \right) \right]
 \leq \left( \frac{\Delta_{adj}}{z_{1-\beta} + z_{1-\alpha /2}} \right)^{2}
@@
 and $z_{1-\beta}$ and $z_{1-\alpha /2}$ are the $100(1-\beta)^{\text{th}}$ and the $100(1-\alpha /2)^{\text{th}}$ percentiles of the Standard Normal distribution.
-Solving for $J$, the number of biological replicates per condition is
+Solving for $J^{\prime}$, the number of biological replicates per condition is
 \begin{equation}
-J \geq \frac{(2\sigma_{PTM}^{2} + 2\sigma_{protein}^{2})(z_{1-\beta} + z_{1-\alpha /2})^2}{\Delta_{adj}^2}
+J^{\prime} \geq \frac{(2\sigma_{PTM}^{2} + 2\sigma_{protein}^{2})(z_{1-\beta} + z_{1-\alpha /2})^2}{\Delta_{adj}^2}
 \end{equation}
@@
 Therefore, the adjustment typically requires a larger sample size to gain the same sensitivity as the unadjusted estimation.
-Details of applications of this approach to experiments with complex design are in {\bf Supplementary Sec.
-\ref{supp-sec:complex_design}}.
-Examples of power analyses are in {\bf Supplementary Sec.
-\ref{supp-sec:sample_size_analysis}}.
+
+\todo{Here we calculate sample size, but in the results we calculate power.
+Could you briefly comment on how we can solve for the power given the sample size (say that this is using the same equation, but solving for $q$ instead.) In fact we can fix all value but one and solve for the remainder}
+
+\todo{I rearranged the supplementary sections. Could you check throughout the manuscript that all the supplementary sections are appropriately referenced?}
 
@@
 The next step is PTM/protein summarization using the $dataSummarizationPTM()$ function (Step 4 of~\figref{msstatsptm_design}).
-Summarization is performed separately for the PTM and the unmodified protein.
+Summarization is performed separately for the PTM features, and for the features representing the unmodified portion of the protein.
 When summarizing the PTM, modified peptide features that span the same modification site are summarized together.
@@
 
-After modeling, the modified model is adjusted for changes in unmodified protein abundance, using the methods described above.
+After modeling, the model for the modified peptides is adjusted for changes in the abunance of the unmodified portion of the protein, using the methods described above.
 Modification sites which lack corresponding global profiling information cannot be adjusted for changes in protein abundance.
@@
 
-We compared the performance of $MSstatsPTM$ to that of $Limma$ and $ANOVA$, both before and after adjusting for changes in the unmodified protein.
-Since $Isobar$-$PTM$ is only applicable to experiments with TMT labeling, it could not be applied to the datasets with known ground truth in this manuscript, and was therefore excluded from the comparisons.
-
-$MSstatsPTM$ before adjusting for changes in the unmodified protein corresponds to base $MSstats$ or $MSstatsTMT$, modeled on the peptide level as described in {\bf Supplementary Table \ref{supp-fig:statistical_inference}}, as appropriate for the experimental design.
-$MSstatsPTM$ with the adjustment, described in \secref{adjustment_summary}, was used without imputing missing values, and without Empirical Bayes moderation.
+We compared the performance of $MSstatsPTM$ to that of $Limma$ and $ANOVA$, both before and after adjusting for changes in the unmodified portion of the protein.
+Since $Isobar$-$PTM$ is only applicable to experiments with TMT labeling it could not be applied to the datasets with known ground truth in this manuscript, and was therefore excluded from the comparisons.
+
+$MSstatsPTM$ before adjusting for changes in the unmodified portion of the protein corresponds to base $MSstats$ or $MSstatsTMT$, modeled on the peptide level as described in {\bf Supplementary Table \ref{supp-fig:statistical_inference}}, as appropriate for the experimental design.
+$MSstatsPTM$ with the adjustment described in \secref{adjustment_summary} was used without imputing missing values, and without Empirical Bayes moderation.
 
 $Unadjusted$ $ANOVA$, i.e.
-$ANOVA$ before adjusting for changes in the unmodified protein, was as in \eqref{ttest_model}.
+$ANOVA$ before adjusting for changes in the unmodified portion of the protein, was as in \eqref{ttest_model}.
 $Adjusted$ $ANOVA$, i.e.
@@
 We evaluated $MSstatsPTM$ on simulated and spike-in datasets with known ground truth in terms of true positives ($TP$), false positives ($FP$), true negatives ($TN$), and false negatives ($FN$) differentially abundant PTMs.
-The true positives were defined as PTMs with changes distinct from the overall changes in unmodified protein abundance.
+The true positives were defined as PTMs with changes distinct from the overall changes in abundance of the unmodified portion of the protein.
 The true negatives were defined as PTMs which, after accounting for the changes in the overall protein abundance, were not differentially abundant.
@@
 
-\figref{sim1_fdr} summarizes the eFDR reported on Computer Simulation 1 dataset by $MSstatsPTM$, $adjusted$ $ANOVA$, $adjusted$ $Limma$ and base $MSstats$, $unadjusted$ $ANOVA$ and $unadjusted$ $Limma$ methods.
+\figref{computer_sim}(a) summarizes the eFDR reported on Computer Simulation 1 dataset by $MSstatsPTM$, $adjusted$ $ANOVA$, $adjusted$ $Limma$ and base $MSstats$, $unadjusted$ $ANOVA$ and $unadjusted$ $Limma$ methods.
 The simulation mimicked a ``clean" label-free group comparison experiment, not compromised by issues such as deviations from model assumptions, missing values and outliers.
@@
 
-\figref{sim2_acc} summarizes the overall accuracy of methods adjusting for changes in the unmodified protein for Computer Simulation 2.
+\figref{computer_sim}(b) summarizes the overall accuracy of methods adjusting for changes in abundance of the unmodified portion of the proteins in Computer Simulation 2.
 The simulation mimicked a more realistic label-free group comparison experiment, including low counts of modified features and missing values.
@@
 As the noise increased the accuracy of all the methods decreased, however $MStatsPTM$ still outperformed the existing methods.
-The difference is primarily due to the robust nature of summarization in $MSstatsPTM$ through Tukey's Median Polish.
-The ratio-based summarization approach used by $Limma$ and $ANOVA$ requires measurements for both the PTM and unmodified protein.
-$MSstatsPTM$ in contrast can still leverage the information in the PTM or unmodified protein if one is missing.
-On average over all simulations, 3.94\% of the total unmodified protein data used by $MSstatsPTM$ was thrown out by $Limma$ and $ANOVA$ due to missing PTM data.
-No PTM data was thrown out due to missing protein data.
+The difference was primarily due to two reasons.
+First, the ratio-based summarization approach used by $Limma$ and $ANOVA$ requires measurements for both the PTM and unmodified protein.
+In contrast, $MSstatsPTM$ can leverage the information in the PTM or unmodified portion of the protein if one of the two is missing.
+On average over all simulations, 3.94\% of the total unmodified protein data \todo{what do you mean by 'data'?
+Runs?
+Combination of features and runs?} used by $MSstatsPTM$ was discarded by $Limma$ and $ANOVA$ due to missing PTM data.
+No PTM data was discarded due to missing protein data.
+Second, the robust Turkey Median Polish summarization in $MSstatsPTM$ was more resistant to outliers.
 
@@
 
-\figref{spikein_boxplot} summarizes the evaluation on the label-free spike-in group comparison experiment, for all the methods, with and without adjusting for changes in the unmodified protein. 
+\figref{spikein}(a) summarizes the evaluation on the label-free spike-in group comparison experiment, for all the methods, with and without adjusting for changes in abundance of the unmodified portion of the protein.
 Without adjusting for changes in unmodified protein abundance, all the approaches incorrectly estimated the log$_2$-fold change of the modified spike-in peptides.
@@
 Additionally, the ratio based summarization used by $Limma$ and $ANOVA$ discarded more data than in the simulations.
-The ratio based summarization lost 9.63\% of the unmodified protein and 1.94\% of the PTM data used by $MSstatsPTM$.
+The ratio based summarization lost 9.63\% of the unmodified protein and 1.94\% of the PTM data \todo{clarify what you mean by 'data'} used by $MSstatsPTM$.
 This additional information lead $MSstatsPTM$ to a more accurate fold change estimation and better calibrated variance.
 
-\figref{spikein_prop_volcano} details the detection of differentially abundant PTM for $MSstatsPTM$ and $Limma$, with and without the adjustment for changes in the unmodified protein, for the $mix3$-$mix2$ and $mix4$-$mix2$ comparisons.
+\figref{spikein}(b) details the detection of differentially abundant PTM for $MSstatsPTM$ and $Limma$, with and without the adjustment for changes in abundance in the unmodified portion of the protein, for the $mix3$-$mix2$ and $mix4$-$mix2$ comparisons.
 As above, the log$_2$-fold changes of the spike-in peptides were only correctly estimated when accounting for changes in the unmodified protein abundance. 
@@
 
-The results of Dataset 4: Human - Ubiquitination - 1mix-TMT are summarized in~\figref{data4_venn_diagram} and~\figref{data4_profile_plot}.
-\figref{data4_venn_diagram} shows the number of differentially abundant modified peptides before and after adjustment.
-Adjusting for changes in the unmodified protein caused more PTMs to change from differentially abundant to non-differentially abundant than vice versa.
-A question that must be addressed is if the decrease in differentially abundant peptides is due to the increased standard error that comes from adjustment, as in \eqref{prop_se}.
-This was tested by looking for modified peptides whose adjusted log fold change was within 10\% of the unadjusted log fold change but became statistically insignificant after adjustment.
-In other words, the fold change was the same between models but the standard error increased.
-When the test was applied to Dataset 4, only one PTM became non-differentially abundant due to an increase in standard error.
-Thus we can conclude that the drop off in differentially abundant PTMs was due to changes in global protein abundance, not simply increased model variance.
-
-\figref{data4_venn_diagram} demonstrates that the unmodified protein was more likely to cause the observed changes in PTM abundance, rather than mask changes in the PTM.
-Conversely, cases where the unmodified protein masks changes in the PTM are more infrequent.
-In this case, the modification and unmodified protein exhibit contradictory changes in abundance, moving in opposite directions.
-One such case can be seen in~\figref{data4_profile_plot}.
-Luchetti et al.~\cite{LUCHETTI2021} showed that $GSDMD$ was actively degraded when IpaH7.8 expression was induced by dox treatment.
-Our reanalysis confirmed this, with the $GSDMD$ protein being down-regulated when Dox treatments reached the 4 and 6 hour marks.
+\todo{I rearranged the figures and the captions for the rest of the section.
+Could you please check that.I did not introduce any mistakes?} \todo{Please check for appropriate use (or lack thereof) of articles 'a' and 'the' throughout the rest of the section and discussion.
+This remains a mistery to me ;-) }
+
+\figref{Dataset4}(a) summarizes the results of Dataset 4: Human - Ubiquitination - 1mix-TMT in terms of number of differentially abundant PTMs before and after adjustment.
+Adjusting for changes in abundance in the unmodified portion of the protein caused fewer PTMs to be detected as differentially abundant.
+A question is whether this was due to adjustment in the log$_2$-fold estimation, or to increase in standard errors during the adjustment (\eqref{prop_se}).
+To check that, we considered modified peptides for which the adjusted log$_2$-fold change was within 10\% of the unadjusted log$_2$-fold fold change, but which lost statistical significance after the adjustment.
+In the case of Dataset 4, only one PTM became non-differentially abundant due to an increase in standard error.
+In other words, the decrease in differentially abundant PTMs was primarily due to removing the confounding with global protein abundance, and not to larger variance estimates.
+
+Cases where true changes in PTM  were masked by changes in abundance in the unmodified portion of the protein were less frequent.
+One such case is in~\figref{Dataset4}(b).
+Luchetti et al.~\cite{LUCHETTI2021} showed that $GSDMD$ was actively degraded when IpaH7.8 expression was induced by Dox treatment.
+Our reanalysis confirmed that the $GSDMD$ protein was down-regulated when Dox treatments reached the 4 and 6 hour marks.
 Conversely, ubiquitination of $GSDMD$ at site $K62$ up-regulated abundance between the same conditions.
 This up-regulation was originally confounded by the down-regulation of unmodified $GSDMD$, and made the modification appear to have little change between no Dox and Dox 4 and 6 hour conditions.
-The proposed approach accounted for this confounding and the modification was shown to be differentially abundant, with an absolute fold change of $1.415$ between the Dox 1 hour and Dox 4 hour conditions.
-This was an interesting result, as the modification contradicts what previous research showed for the unmodified protein.
-This effect can be thought of as a false negative and would have been challenging to observe without the proposed approach.
-There are many non-differentially abundant PTMs, and having to comb through each one individually to determine false negatives is generally an unfeasible task.
-The proposed approach removes the confounding on an experimental level, revealing modifications of interest that are entirely masked due to changes in the unmodified protein.
-
-The results of Dataset 5: Mouse - Phosphorylation - 2mix-TMT are summarized in~\figref{data5_venn_diagram} and~\figref{data5_profile_plot}.
-\figref{data5_venn_diagram} showed that adjusting for changes in the unmodified protein caused more PTMs to change from differentially abundant to non-differentially abundant.
-Again we tested if this change was due to fold change adjustment or an increased standard error.
-When applied to Dataset 5, 548 PTMs became non-differentially abundant due to an increase in standard error (only 3.4\% of all PTMs that became non-differentially abundant).
-
-\figref{data5_profile_plot} shows a case where the unmodified protein caused the observed changes in PTM abundance.
-This occurs when the protein exhibits a change in abundance while the modification does not.
-Originally the modification at $S178$ of protein $TTP$ was shown to be differentially abundant between KO\_Uninfect and KO\_Early, with a log fold change of $2.9$.
-However the unmodified protein was shown to contribute $2.014$ of this change, meaning that the modification only accounted for $.886$ after adjustment (seen in {\bf Supplementary Figure~\ref{supp-fig:No_Diff_Shigella_PTM}}).
-This caused the modification to go from differentially abundant to non-differentially abundant.
-These cases occur frequently and can be thought of as false positives.
-The proposed approach corrects for this confounding, allowing us to see the true fold change of PTMs, while excluding those that do not exhibit differential changes.
-
-%Adjusting for changes in the unmodified protein caused more PTMs to change from differentially abundant to non-differentially abundant than vice versa.
-A question that must be addressed is if the decrease in differentially abundant peptides is due to the increased variance \todo{variance of what?
-estimation of the variance parameter?
-Clarify} that comes from adjustment.
-This was tested by looking for modified peptides whose adjusted log fold change was within 10\% of the unadjusted log fold change but became statistically insignificant after adjustment.
-In other words, the fold change was the same between models but variance increased.
-When the test was applied to Dataset 4, only one PTM became non-differentially abundant due to an increase in variance.
-When applied to Dataset 5, 548 PTMs became non-differentially abundant due to an increase in variance (only 3.4\% of all PTMs that became non-differentially abundant).
-Thus we can conclude that the drop off in differentially abundant PTMs was due to changes in global protein abundance, not simply increased model variance.
-
-%\figref{venn_diagrams} demonstrates that the unmodified protein was more likely to cause the observed changes in PTM abundance, rather than mask changes in the PTM.
-The later case can occur when the unmodified protein exhibits a change in abundance, while the modification does not.
-Additionally, this case may come about due to the high variability of PTMs compared to the global protein.
-An example of this can be seen in \figref{data5_profile_plot}.
-Originally the modification at $S178$ of protein $TTP$ was shown to be differentially abundant between KO\_Uninfect and KO\_Early, with a log fold change of $2.9$.
-However the unmodified protein was shown to contribute $2.014$ of this change, meaning that the modification only accounted for $.886$ after adjustment (seen in {\bf Supplementary Figure~\ref{supp-fig:No_Diff_Shigella_PTM}}).
-This caused the modification to go from differentially abundant to non-differentially abundant.
-These cases occur frequently and can be thought of as false positives.
-The proposed approach corrects for this confounding, allowing us to see the true fold change of PTMs, while excluding those that do not exhibit differential changes.
-
-%\subsubsection*{In the two-mi[xture TMT experiment, MSstatsPTM removed confounding \todo{OV: between what and what? Is it similar to the previous section?}}
-
-%\todo{This section just describes the figure.
-Could you emphasize why we see these results?
-What are the reasons for differences in performance?} The results of this experiment are summarized in Figure \ref{fig:data5_plots}.
-Figure \ref{fig:data5_venn_diagram} shows the number of significant modified peptides before and after adjustment.
-Again more PTMs became insignificant after adjustment than became significant.
-Again we tested if the decrease in significant peptides was due to removing confounding with changes in the umodified protein or increased variance after adjustment was applied.
-This was tested by looking for modified peptides whose adjusted log fold change was within 10\% of the unadjusted log fold change but became insignificant after adjustment.
-When this test was applied, 548 peptides became insignificant due to an increase in variance.
-This is only 3.4\% of the total peptides that became insignificant after adjustment.
-Thus we can conclude that the drop off in significant peptides was mainly due to changes in global protein abundance.
-
-%In Figure \ref{fig:data5_profile_plot} the profile plot of protein $TTP$ modified at site $S178$ showed the power of the proposed method.
-Without adjustment, there was a large positive log fold change of $2.9$ between the WT\_Late and WT\_Uninfected conditions.
-However, the global profiling run showed a similar log fold change of $2.014$ between the same conditions.
-This indicated that the abundance change in the modified peptide is nearly entirely due to changes in the global protein.
-When adjusting for the global protein the modified peptide's adjusted p-value became insignificant, going from $.0009$ to $.248$.
-Correcting for changes in global protein abundance allowed us to see the true impact of the modification at $S178$ which would have otherwise been challenging to perceive.
+The proposed approach accounted for this confounding and the modification was detected as differentially abundant, with an absolute fold change of $1.415$ between the Dox 1 hour and Dox 4 hour conditions.
+The effect \todo{on what?} would have been challenging to observe without the proposed approach.
+Indeed, the modification contradicts the previous research focusing on global profiling of that protein \todo{not sure I understand the previous sentence.
+Also, is there a reference?}.
+
+\figref{Dataset5}(a) illustrates a similar result for Dataset 5: Mouse - Phosphorylation - 2mix-TMT.
+Adjusting for changes in abundance in the unmodified portion of the protein caused fewer PTMs to be detected as differentially abundant.
+As above, we checked whether this change was due to fold change adjustment or an increased standard error.
+In the case of Dataset 5, 548 PTMs were not detected as differential abundant due to an increased standard error.
+This corresponded to 3.4\% of all the PTMs that became non-differentially abundant.
+
+\figref{Dataset5}(b) shows an opposite case where the protein exhibits a change in abundance while the modification does not.
+Originally \todo{What do you mean by 'originally'?
+In a prior publication?
+Could you cite a reference?} the modification at $S178$ of protein $TTP$ was shown to be differentially abundant between KO\_Uninfect and KO\_Early, with a log fold change of $2.9$.
+However the unmodified protein was shown to contribute $2.014$ of this change, i.e.
+the modification only accounted for $.886$ \todo{on which scale?
+Maybe report as a \%?} after adjustment ({\bf Supplementary Figure~\ref{supp-fig:No_Diff_Shigella_PTM}}).
+This caused the modification to lose differential abundance.
+
+\todo{Should we compare the results to limma and anova.
+Is there something to comment on?
+Maybe in supplementary?
+In particular, how many runs were lost when calculation the ratios?
+We don't have the ground truth, but maybe to say that qualitatively (in terms of relative number of diff.
+abundance) the results look similar?}
 
@@
 
-As discussed in \secref{exp_proc_dataset6}, there was no unmodified global profiling run performed in this experiment.
-Once identification and quantification of the Ubiquitination profiling was performed, peptides which were unmodified were extracted and used in place of a global profiling run.
-This resulted in a lack of overlap between PTMs and unmodified proteins.
-Any PTM without a corresponding unmodified protein could not be adjusted.
-Of the 10,799 ubiquitin sites identified, only 4,526 had a corresponding unmodified protein and could be adjusted.
-Additionally, the lack of a separate global profiling run resulted in low feature counts for the unmodified protein model compared to other experiments, seen in Table \ref{tab:dataDescription}.
-
-The number of differentially abundant PTMs before and after adjustment are shown in~\figref{data6_plots}.
-After adjusting for changes in the unmodified protein, there were fewer differentially modified peptides.
-However, this was mainly due to not having a global profiling run.
-When looking only at PTMs with a corresponding unmodified protein, the number of differentially modified PTMs exhibited a small drop, but was more consistent.
+Unlike the other datasets in this manuscript, Dataset 6: Human - Ubiquitination - Label-free had no unmodified global profiling run.
+Therefore, after peptide identification and quantification, data from unmodified peptides were used separately in place of a global profiling run.
+This resulted in a sparse coverage of the modified features by the unmodified protein counterparts.
+Of the 10,799 identified ubiquitination sites, only 4,526 had features from unmodified portion of the same protein.
+A PTM without features from the unmodified portion of the protein could not be adjusted for the confounding.
+Additionally, the lack of a separate global profiling run resulted in low feature counts, and noisier measurements of the unmodified peptides as compared to the other experiments (Table \ref{tab:dataDescription}).
+
+\figref{Dataset6} shows the number of differentially abundant PTMs before and after adjustment.
+After adjusting for changes in abundance in the unmodified portion of the protein, the number of differentially abundant PTMs decreased.
+However this was mainly due to the lack of a global profiling run.
+The decrease in differentially modified PTMs was smaller among PTMs with available unmodified protein counterparts.
 As above, we tested if this drop in differentially abundant PTMs was due to an increase in standard error.
-Here only 25 PTMs became insignificant due to increased standard error.
-Without a global profiling run, the method was less effective at removing confounding with the unmodified protein.
-This was due to not being able to perform adjustment without a corresponding global protein.
-In this case adjustment caused us to toss out a large number of the PTMs in the original model.
-In cases without a global profiling run, the proposed method can still be used, but it should be used in conjunction with the unadjusted model so that information is not lost.
-
-%In Figure \ref{fig:usp30_profile} the profile plot of peptide $P52209\_K059$ is shown.
-The abundance between conditions appears flat in the modified peptide profile plot, however this is contrast with a large negative change in the corresponding global protein.
-Specifically looking at the comparison between the CCCP and Control conditions, the unadjusted model showed a small log fold change of $-.255$ and insignificant adjusted pvalue of $.104$.
-However when adjusting for changes in the undmodified protein, the log fold change increased to $-2.73$ and adjusted pvalue became significant at $1.067e-04$.
-Using the proposed method we get increased information that would otherwise have been missed.
+Here only 25 PTMs lost differential abundance due to increased standard error.
+
+\todo{I see that Supplementary Fig 8 and 9 are not referenced in the main text.
+Could you check that every supplementary fig or section is mentioned somewhere in the main text?
+Otherwise it is unclear what the supplementary figure is doing there.
+(I rearranged the order/number of Supplies sections, could you check that they are cited correctly?)}
+
+\subsubsection*{Noisy PTM measurements benefited from additional biological replicates}
+
+Sample size calculations and statistical power analyses help evaluate the benefit of additional biological replicates, in particular when measurements are noisy.
+Since the proposed approach does not have the restriction of the equal number of modified and unmodified global profiling runs,  it is also interesting to evaluate the interplay between the number of replicates of each type in presence of differing amounts of noise.
+In Datasets 4 and 5 the variance of the PTMs was higher than that of the unmodified protein summaries, \todo{with median values of PTM variance of $.2$, and protein variance of $.1$}.
+In Dataset 6 the variance of the PTM and the unmodified protein summaries were comparable, \todo{with the median value of around $.15$}.
+\todo{The power analysis took as input these median variance values.}
+
+\figref{power_sd_combo} shows that regardless of the relative amount of variation, larger adjusted \todo{ log$_2$-fold changes or FC?} and larger number of replicates enabled larger statistical power.
+When the variance of PTMs was equal to the variance of global protein profiles, it did not matter whether we increased the number of PTM runs or the number of unmodified protein runs.
+However, when the variance of the PTM summaries was higher than that of the unmodified protein summaries, allocating more biological replicates to the PTM profiles lead to a more efficient increase of statistical power.
 
@@
 
-We proposed a general statistical modeling framework and implementation for PTM characterization.
-The framework is designed for bottom-up MS workflows, which are characterized with variations from multiple confounded sources, frequent missing data, and associated uncertainty in the conclusions.
-The framework is general and is applicable to a variety of experimental designs.
-While the ratio-based approaches of $ANOVA$ and $Limma$ look at the differences within samples first, the proposed approach looks at the differences between conditions first.
-This provides us with greater flexibility in terms of modeling, accounting for outliers and missing values, and handling complex designs and robust implementations.
-Additionally, the proposed approach outperforms $ANOVA$ and $Limma$, and yields accurate results in the broad type of experimental circumstances.
-The framework allows us to plan for subsequent experiments, and choose the appropriate number of replicates in consideration of adjustment with respect to protein abundance.
-The implementation allows for straightforward application of the methods discussed and allows for reproducible experimental analysis.
-
-Our results show that the proposed approach for modeling and summarization leads to more sensitive PTM significance analysis and more accurate and precise quantification.
-The gain is due to a more efficient use of the data, and to a more accurate understanding of the systematic and random variations.
-The proposed framework can be extended beyond the experimental designs with variation from multiple sources discussed above.
+We proposed a statistical modeling framework for detecting differentially abundant PTM, and its implementation in $MSstatsPTM$.
+The proposed approach removes the confounding of changes in PTMs with changes in the unmodified protein summaries, and can reveal modifications of interest that are otherwise entirely masked by changes in abundance in the unmodified portion of the  protein.
+This is valuable, because many PTMs are often found non-differentially abundant prior to the adjustment, and validating them manually to establish false negatives is generally unfeasible.
+
+% stats
+Our results show that $MSstatsPTM$ is more accurate and more sensitive than the existing approaches.
+The gain is due to a more efficient use of the data, and to a more accurate representation of the systematic and random variations.
+While the ratio-based approaches of $ANOVA$ and $Limma$ first consider differences within samples, $MSstatsPTM$ first considers the differences between conditions.
+This enables a greater flexibility in terms of modeling complex designs, accounting for outliers and missing values, and planning subsequent experiments.
+
+% compatibility with designs
+The implementation of $MSstatsPTM$ is a straightforward extension of $MSstats$, and is therefore applicable to the full breadth of experiment types supported by $MSstats$.
 Although demonstrated here on DDA, is also applicable to DIA, SRM and PRM acquisitions. 
-Additionally, the approach can handle experiments with modified peptides processed using label-free methods and unmodified peptides processed using TMT labeling, or vice versa.
-In this case summarization and modeling is still done separately for both the modified and unmodified data, and then combined after modeling.
-
-A potential limitation of the proposed framework is the assumption that all the peptides are correctly mapped to the underlying proteins and PTM sites, and the features are informative of the abundances of underlying protein and PTM.
-Also, characterizing PTMs with current data-dependent acquisition workflows is prone to being under sampled, leading to a sparse dataset with a large number of missing values for the analysis.
-Statistical methods accounting for effects due to experimental units and missing values introduced in this manuscript help interpret the data in a more objective manner.
-The latest development of targeted acquisition and data-independent acquisition methods are expected to further alleviate these issues.
-
-Additionally, abundance levels of PTM sites can be confounded with each other if there are multiple modification sites per peptide, or confounded with changes in the unmodified peptide (as opposed to the unmodified protein).
-In the current implementation the effect of a specific modification in a peptide with multiple modifications cannot be quantified.
-One potential solution to this is to measure the abundance of peptides with one modification and use this to adjust the peptide with multiple sites to remove the confounding.
-However, this method would likely run into challenges due to sparsity of features for modified peptides with both a single and multiple modification sites.
-A more complex approach to addressing this problem is likely necessary.
-
-Overall, the proposed approach balances accuracy and practicality, and enables the analysis of complex experiments in high throughput.
+\todo{Additionally, the approach can handle experiments that use different strategies for modified versus unmodified peptides, e.f.
+using label-free methods for unmodified peptides and TMT labeling for modified peptides, or vice versa.}
+
+% limitation
+The proposed approach assumes that all the peptides are correctly mapped to the underlying proteins and PTM sites, and that the features are informative of the protein and PTM abundances.
+However, multiple modification sites per peptide can confound the abundance of each PTM site.
+Changes in the unmodified peptide (as opposed to the unmodified protein) can also confound changes in PTM abundance.
+One potential solution is to quantify the abundance of peptides with one modification and use this to adjust the peptide with multiple sites to remove the confounding.
+However, this method would likely be challenged by scarsity of modified peptide features containing both a single and multiple modification sites.
+As the result, peptides with multiple modifications are currently beyond the $MSstatsPTM$ scope.
+
+Overall, $MSstatsPTM$ balances accuracy and practicality, and enables the analysis of complex experiments in high throughput.
 Future work is to carry out the inference and testing for not only the relative change of PTM abundance, but also the fraction of the protein that is modified at the particular site (site occupancy, or stoichiometry), and attempt to remove the confounding of individual PTMs in peptides with multiple modifications.
@@
 
-%\begin{table}[]
-%\begin{tabular}{|c|c|c|}
-%\hline
-%Comparison & Unadj. true log$_{2}$FC & Adj. true log$_{2}$FC \\ %\hline \hline
-%mix2 vs mix1  & -1 & -1 \\ \hline
-%mix3 vs mix1  & 0 & 1 \\ \hline
-%mix4 vs mix1  & -1 & 0 \\ \hline
-%mix3 vs mix2  & 1 & 2 \\ \hline
-%mix4 vs mix2  & 0 & 1 \\ \hline
-%mix4 vs mix3  & -1 & -1 \\ \hline
-%\end{tabular}
-%\end{table}
-
 \begin{figure}[ht]
@@
 The package is applicable to label-free data acquisitions such as DDA, DIA, SRM and label-based data acquisitions such as TMT.
-It takes as input lists of identified and quantified spectral features for the PTM and unmodified protein, produced by spectral processing tools such as MaxQuant, Progenesis, or Spectronaut.
-Conversion, summarization and statistical modeling are performed separately for the PTM and for the unmodified proteins.
+It takes as input lists of identified and quantified spectral features for the PTM and for the unmodified portion of the protein, produced by spectral processing tools such as MaxQuant, Progenesis, or Spectronaut.
+Conversion, summarization and statistical modeling are performed separately for the PTM and for the unmodified portions of the proteins.
 Steps 4 and 5 leverage the summarization and modeling functions from $MSstats$ and $MSstatsTMT$.
-Model-based summaries are combined to adjust the changes in the PTM abundance for changes in unmodified protein abundance.
+Model-based summaries are combined to adjust the changes in the PTM abundance for changes in abundance of the unmodified portion of the protein.
 Finally, sample size calculation for future experiments can be performed using the modeling output.}
@@
 \centering
-\begin{subfigure}[c]{0.7\linewidth}
-\includegraphics[width=1\textwidth]{images/sim1_FDR_all_models.png}
-\caption{}
-\label{fig:sim1_fdr}
-\end{subfigure}
-\begin{subfigure}[c]{0.825\linewidth}
-\includegraphics[width=1\textwidth]{images/sim3_Accuracy.png}
-\caption{}
-\label{fig:sim2_acc}
-\end{subfigure}
+\begin{tabular}{c}
+\includegraphics[width=0.8\textwidth]{images/sim1_FDR_all_models.png}\\
+(a)\\
+\includegraphics[width=0.8\textwidth]{images/sim3_Accuracy.png}\\
+(b)
+\end{tabular}
 \caption{Computer simulations.
@@
 In contrast, the methods accounting for the confounding correctly calibrated the proportion of false positive differentially abundant PTM.
-b) In a noisy simulation, which included limited feature observations and missing values, $MSstatsPTM$ had higher accuracy than $ANOVA$ and $Limma$ given the same noise, number of biological replicates, and number of conditions.
+b) Dataset 2, a noisy simulation, which included limited feature observations and missing values.
+$MSstatsPTM$ had higher accuracy than $ANOVA$ and $Limma$ given the same noise, number of biological replicates, and number of conditions.
 }
@@
 \centering
-\begin{subfigure}[c]{0.825\linewidth}
-\includegraphics[width=1\textwidth]{images/spike_in_fc.png}
-\caption{}
-\label{fig:spikein_boxplot}
-\end{subfigure}
-\begin{subfigure}[c]{0.825\linewidth}
-\includegraphics[width=1\textwidth]{images/spike_in_volcano.png}
-\caption{}
-\label{fig:spikein_prop_volcano}
-\end{subfigure}
+\begin{tabular}{c}
+\includegraphics[width=0.8\textwidth]{images/spike_in_fc.png}\\
+(a)\\
+\includegraphics[width=0.8\textwidth]{images/spike_in_volcano.png}\\
+(b)
+\end{tabular}
 \caption{Dataset 3: Spike-in benchmark - Ubiquitination - Label-free.
@@
 b) The statistical results of $MSstatsPTM$ and $Limma$ modeling the $mix3$-$mix2$ and $mix4$-$mix2$ comparisons before and after adjustment.
-The spike-in peptides are colored red, while the background peptides are highlighted grey.
 The solid horizontal line shows the adjusted pvalue cutoff of .05.
@@
 The dashed vertical line shows the expected log$_2$ fold change of the spike-in peptides.
-Before adjustment the spike-in peptides (colored red) did not follow the expected log fold change; after adjustment, the spike-in peptides were more in line with expectation.}
-\label{fig:spikein_volcano}
+Before adjustment the spike-in peptides did not follow the expected log fold change, but were more in line with expectation after adjustment.}
+\label{fig:spikein}
 \end{figure}
@@
 \centering
-\begin{subfigure}[c]{0.55\linewidth}
-\includegraphics[width=1\textwidth]{images/ipah_venn_diagramm.png}
-\caption{}
-\label{fig:data4_venn_diagram}
-\end{subfigure}
-\begin{subfigure}[c]{0.55\linewidth}
-\includegraphics[width=1\textwidth]{images/shig_venn_diagramm.png}
-\caption{}
-\label{fig:data5_venn_diagram}
-\end{subfigure}
+\begin{tabular}{ccc}
+\includegraphics[width=0.32\textwidth]{images/ipah_venn_diagramm.png}
+& $~~~$ &
+\includegraphics[width=0.62\textwidth]{images/IpaH_prof_plot.png}\\
+(a) && (b)
+\end{tabular}
 \caption{
-The overlap of differential modified peptides for the PTM model with and without global protein level adjustment across all pairwise comparisons.
-a) Dataset 4: Human - Ubiquitination - 1mix-TMT.
-T b) Dataset 5: Mouse - Phosphorylation - 2mix-TMT.
-}
-\label{fig:venn_diagrams}
+Dataset 4: Human - Ubiquitination - 1mix-TMT, analysis with $MSstatsPTM$.
+(a) The overlap of differential modified peptides for the PTM model with and without global protein level adjustment across all pairwise comparisons.
+(b)   \todo{The caption of the sub-figure was missing (maybe I deleted it accidentally?
+Please add back)}}
+\label{fig:Dataset4}
+\end{figure}
+
+\begin{figure}[ht]
+\centering
+\begin{tabular}{ccc}
+\includegraphics[width=0.32\textwidth]{images/shig_venn_diagramm.png}
+& $~~~$ &
+\includegraphics[width=0.62\textwidth]{images/No_Difference_Shigella_Profile_Plot}\\
+(a) && (b)
+\end{tabular}
+\caption{
+Dataset 5: Mouse - Phosphorylation - 2mix-TMT, analysis with $MSstatsPTM$.
+(a) The overlap of differential modified peptides for the PTM model with and without global protein level adjustment across all pairwise comparisons.
+(b)  Comparing the profiling of protein $TTP$ with the phosphorylation at site $S178$.
+The individual PTM and protein features are shown in grey, while the summarization is highlighted in red.
+The plots are separated according to TMT Mixtures 1 and 2.}
+\label{fig:Dataset5}
 \end{figure}
@@
 \centering
-\begin{subfigure}{\textwidth}
- \centering
-\includegraphics[width=1.0\textwidth]{images/IpaH_prof_plot.png}
-\caption{}
-\label{fig:data4_profile_plot}
- \end{subfigure}
- \begin{subfigure}{\textwidth}
- \centering
-	\includegraphics[width=1.0\textwidth]{images/No_Difference_Shigella_Profile_Plot}
-	\caption{}
-	\label{fig:data5_profile_plot}
-	 \end{subfigure}
-\caption{a) Dataset 4: Human - Ubiquitination - 1mix-TMT.
-Comparing the $GSDMD$ protein and its ubiquitination at site $K62$ across all conditions.
-The individual PTM and protein features are shown in grey, while the summarization is highlighted in red.
-b) Dataset 5: Mouse - Phosphorylation - 2mix-TMT.
-Comparing the profiling of protein $TTP$ with the phosphorylation at site $S178$.
-The individual PTM and protein features are shown in grey, while the summarization is highlighted in red.
-The plots are separated into TMT Mixtures 1 and 2.}
-\label{fig:profile_plots}
+\begin{tabular}{ccc}
+\includegraphics[width=0.32\textwidth]{images/usp30_venn_diagramm}
+& $~~~$ &
+\includegraphics[width=0.325\textwidth]{images/usp30_venn_diagramm_matching_only}\\
+(a) && (b)
+\end{tabular}
+ \caption{Dataset 6: Human - Ubiquitination - Label-free no global profiling run, analysis with $MSstatsPTM$.
+a) The overlap of differential modified peptides for the PTM model with and without global protein level adjustment across all comparisons, including all measured PTMs.
+b) The overlap of differential modified peptides with and without global protein level adjustment, including only PTMs for which features from the unmodified portion of the same protein were also available.}
+\label{fig:Dataset6}
 \end{figure}
@@
 \centering
- \begin{subfigure}{\textwidth}
- \centering
-	\includegraphics[height=.55\textwidth]{images/usp30_venn_diagramm}
-	\caption{}
-	\label{fig:data6_vd1}
- \end{subfigure}
- \begin{subfigure}{\textwidth}
- \centering
-	\includegraphics[height=.55\textwidth]{images/usp30_venn_diagramm_matching_only}
-	\caption{}
-	\label{fig:data6_vd2}
- \end{subfigure}
- \caption{Dataset 6: Human - Ubiquitination - Label-free no global profiling run.
-a) The overlap of differential modified peptides for the PTM model with and without global protein level adjustment across all comparisons, including all measured PTMs.
-b) The overlap of differential modified peptides with and without global protein level adjustment, including only PTMs with a corresponding unmodified protein.}
-\label{fig:data6_plots}
+ \begin{tabular}{cc}
+	\includegraphics[width=.45\textwidth]{images/same_var_power}
+&
+	\includegraphics[width=.45\textwidth]{images/high_ptm_var_power}
+\\
+(a) & (b)
+\end{tabular}
+ \caption{\todo{Could you check that I copied the correct figures?} Power analysis with with $MSstatsPTM$, \todo{based on variance components from what datasets?}.
+Y-axis: statistical power, i.e.
+the probability of rejecting the null hypotheses of no protein-adjusted change in a PTM between conditions, when in fact the change exists.
+X-axis: \todo{clarify whether the x axis are adjusted FC.} \todo{Check whether the x axis is logFC or the FC on the original scale - looks like the original scale to me} a) The power of detecting differential abundant PTMs, when  the modified features and the unmodified protein summaries have the same variance .15.
+b) The power of detecting differential abundant PTMs, when the modified features had a larger variance (.2) as compared to the unmodified protein summaries (.1).}
+\label{fig:power_sd_combo}
 \end{figure}
 
-%\begin{figure}[h!]
-%\centering
-% \begin{subfigure}{\textwidth}
-% \centering
-%	\includegraphics[width=.75\textwidth]{images/same_var_power}
-%	\caption{}
-% \end{subfigure}\vspace{5mm}
-% \begin{subfigure}{\textwidth}
-% \centering
-%	\includegraphics[width=.75\textwidth]{images/high_ptm_var_power}
-%	\caption{}
-% \end{subfigure}
-% \caption{Power analysis of experiments with differing variances.
-a) The power of an experiment targeting PTMs with the same variance, .15, for the modified and unmodified peptides.
-Predictably when the replicates are high for both modified and unmodified peptides the power is much higher.
-Conversely at low replicates for each the power is much lower.
-The interesting part of this chart is when the replicates are different between runs.
-With equal variance, it does not matter if the PTM replicates or protein replicates are higher.
-b) In this chart the variance for the PTM is higher than the unmodified protein.
-The PTM variance is .2, while the unmodified protein variance is .1.
-With equal replicates the results are the same as above, more replicates equals more power.
-When the replicates are not the same we can clearly see that having more replicates for the PTM leads to higher power.}
-%\label{fig:power_sd_combo}
-%\end{figure}
-
 \end{document}

### 2022-08-19 supplementary/ptm_sm.tex
First final

@@
 \clearpage
-%\section{Overview of proposed approach and notation}
-%\label{sec:intro}
-
-%This study addresses three major goals in the characterization of post-translational modifications (PTMs): a) relative PTM quantification, b) PTM significance analysis, i.e., to detect PTM sites that are differentially modified across experimental conditions, and c) statistical design and analysis of PTM experiments.
-
-%Unlike protein-level analysis, in which multiple quantified peptides are often observed, phospho-peptides are more frequently detected and quantified only once.
-There is a strong correlation between signal strength and reproducibility.
-When selecting sites for further study, investigators should pay close attention to the number of PSMs and the signal strength for peptides harboring each site.
-%Simply calculating averages or medians of all peptides containing a given site might not reval the full complexity of cellular phosphorylation patterns.
-Singly and doubly phosphorylated forms of a peptide might be present at different levels.
-One must also not forget that changes in total protein level are not refelcted in the phosphopeptide ratios.
-Whenever possible, separate protein-level measurements made from unmodified peptides should be performed and used to normalize phosphopeptide ratios.
-
-%\subsection{Data structure of PTM quantification experiments.}
-%A set of fully-cleaved and/or partially-cleaved peptides containing a same PTM (e.g., ubiquitination/phosphorylation) at one site are considered together.
-There are $I$ conditions and $J$ mass spectrometry runs (technical replicates) per condition in the experiment.
-The PTM site is represented by $K$ spectral features (peptide ions, distinguished by their cleavage residues and charge states).
-The log-intensity (base 2) of Feature $k$, in Run $j$ of Condition $i$ is denoted by $y_{ijk}^{\ast}$.
-To account for the underlying protein abundance, features corresponding to the unmodified peptides from the same protein are considered together, except those unmodified peptides containing a modified site to avoid the confounding effect due to the PTM.
-The log-intensity of Feature $l$ from the unmodified peptides in the same run is denoted by $y_{ijl}$.
-Figure \ref{fig:dtable} shows an example data representation of modified peptide ions at one site and unmodified peptide ions of the same protein.
-Unmodified peptides from the same protein provide additional evidence on the underlying protein abundance, which needs to be integrated for PTM characterization.
-To address the goals of PTM characterization, statistical analysis needs to summarize values in this table using appropriate statistical models, translate the goal into a model-based quantity of interest, and draw inference (i.e., characterize the uncertainty) about the quantity.
-
-%\begin{figure}[h!]
-%\centering
-%\begin{footnotesize}
-%\[
-%\begin{array}{l l c c c c c c c c c} 
-%\toprule
-% &  & \multicolumn{4}{c}{\text{Condition }1} & \dots & \multicolumn{4}{c}{\text{Condition }I} \\ 
-%\cmidrule{3-6} 
-%\cmidrule{8-11} 
-% &  &  \text{Run }1 & \text{Run }2 & \dots & \text{Run }J & \dots & \text{Run }1 & \text{Run }2 & \dots & \text{Run }J \\
-%\midrule
-%\multirow{4}{*}{Modified} & \text{Feature }1 & y_{111}^{\ast} & y_{121}^{\ast} & \dots & y_{1J1}^{\ast} & \dots & y_{I11}^{\ast} & y_{I21}^{\ast} & \dots & y_{IJ1}^{\ast} \\
-% & \text{Feature }2 & y_{112}^{\ast} & - & \dots & y_{1J2}^{\ast} & \dots & y_{I12}^{\ast} & y_{I22}^{\ast} & \dots & y_{IJ2}^{\ast}  \\
-% & \dots & \dots & \dots & \dots & \dots & \dots & \dots & \dots & \dots & \dots \\
-% & \text{Feature }K & y_{11K}^{\ast} & y_{12K}^{\ast} & \dots & y_{1JK}^{\ast} & \dots & y_{I1K}^{\ast} & y_{I2K}^{\ast} & \dots & y_{IJK}^{\ast} \\
-%\midrule
-%\multirow{4}{*}{Unmodified} & \text{Feature }1 & y_{111} & y_{121} & \dots & y_{1J1} & \dots & y_{I11} & y_{I21} & \dots & y_{IJ1} \\
-% & \text{Feature }2 & y_{112} & y_{122} & \dots & y_{1J2} & \dots & - & - & \dots & -  \\
-%% & \text{Feature }2 & y_{112} & y_{122} & \dots & y_{1J2} & \dots & y_{I12} & y_{I22} & \dots & y_{IJ2}  \\
-% & \dots & \dots & \dots & \dots & \dots & \dots & \dots & \dots & \dots & \dots \\
-% & \text{Feature }L & y_{11L} & y_{12L} & \dots & y_{1JL} & \dots & y_{I1L} & y_{I2L} & \dots & y_{IJL} \\
-%\bottomrule
-%\end{array}
-%\]
-%\[
-%\begin{array}{l l c c c c c} 
-%\toprule
-% &  & \multicolumn{2}{c}{\text{Condition }1} &  & \multicolumn{2}{c}{\text{Condition }2} \\ 
-%\cmidrule{3-4} 
-%\cmidrule{6-7} 
-% &  &  \text{Run }1 & \text{Run }2 &  & \text{Run }1 & \text{Run }2 \\
-%\midrule
-%\multirow{4}{*}{\begin{turn}{90} {\text{Modified}} \end{turn}} & \text{Feature }1 & y_{111} & y_{121} &  & y_{211} & y_{221} \\
-%%\multirow{4}{*}{Modified} & \text{Feature }1 & y_{111} & y_{121} &  & y_{211} & y_{221} \\
-% & \text{Feature }2 & y_{112} & - &  & y_{212} & y_{222} \\
-% & \dots & \dots & \dots &  & \dots & \dots \\
-% & \text{Feature }5 & y_{115} & y_{125} &  & y_{215} & y_{225} \\
-%\midrule
-%\multirow{7}{*}{\begin{turn}{90} {\text{Unmodified}} \end{turn}} & \text{Feature }1 & y_{111}^{\ast} & y_{121}^{\ast} &  & y_{211}^{\ast} & y_{221}^{\ast} \\
-% & \text{Feature }2 & y_{112}^{\ast} & y_{122}^{\ast} &  & - & y_{222}^{\ast} \\
-% & \dots & \dots & \dots &  & \dots & \dots \\
-% & \text{Feature }7 & y_{117}^{\ast} & y_{127}^{\ast} &  & y_{217}^{\ast} & y_{227}^{\ast} \\
-% & \text{Feature }8 & y_{118}^{\ast} & y_{128}^{\ast} &  & y_{218}^{\ast} & y_{228}^{\ast} \\
-% & \dots & \dots & \dots &  & \dots & \dots \\
-% & \text{Feature }11 & y_{1,1,11}^{\ast} & y_{1,2,11}^{\ast} &  & y_{2,1,11}^{\ast} & y_{2,2,11}^{\ast} \\
-%\bottomrule
-%\end{array}
-%\]
-%\end{footnotesize}
-%\caption{Representation of the data of modified peptides at one site and unmodified peptides of the same protein, with $I$ conditions and $J$ replicate runs.
-Abundances of the PTM and protein are quantified by multiple spectral features (peptide ions, $K$ for modified peptides and $L$ for unmodified peptides).
-Some spectral features can be missing (shown as $-$), either randomly in individual runs or completely in certain conditions.
-In real practice, the number of runs can vary across conditions.
-\label{fig:dtable}}
-%\end{figure}
-%
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\section{Existing methods} \label{sec:ttest}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\subsection{$t$-test for data in batches}
-%
-%The test is not directly applicable to a problem with batches of data.
-Several \textit{ad-hoc} approaches may be used.
-Three commonly used approaches are 1) $t$-test (no batch): ignoring batch effects when applying $t$-test, 2) $t$-test (most significant batch): applying $t$-test in each batch and drawing conclusions based on the most significant batch, and 3) $t$-test (all batch): applying $t$-test in each batch and drawing conclusions based on all batches.
-Although simple, these \textit{ad-hoc} methods lack statistical justification.
-We reconstruct their meanings below and characterize their statistical properties under various forms of batch effects in \secref{sec:sim}.
-The abundance in each run is taken as input each of the three methods and is often estimated by sum of peak intensities, e.g., the abundance estimate for Run $j$ of Condition $i$ in Batch $b$ is given by
-%\[
-%y_{b, ij+} = \sum_{k=1}^{K_b} y_{b, ijk}
-%\]
-%
-%\paragraph{Two-sample $t$-test (no batch).}
-%In this approach, the modification abundance is estimated as the average over runs in all batches. For example, the estimate for Group $i$ is given by
-%\[
-%\hat{\mu}_{i}^{\ast} = \frac{1}{BJ} \sum_{b=1}^{B} \sum_{j=1}^{J} y_{b, ij+}.
-%\]
-%This is equivalent to fitting a linear model with only an intercept, which does not account for different characteristics across batches. 
-%
-%
-%\paragraph{Two-sample $t$-test (most significant batch).} 
-%The modification abundance is estimated in all batches, i.e., 
-%\[
-%\hat{\mu}_{b, i}^{\ast} = \frac{1}{J} \sum_{j=1}^{J} y_{b, ij+},
-%\]
-%where $\mathrm{SE} (\hat{\mu}_{b, i}^{\ast})$ is the standard error associated with the estimate. 
-%The test statistic in Batch $b$ is 
-%\[
-%t_{b} = \frac{\hat{\mu}_{b, i}^{\ast}}{\mathrm{SE} (\hat{\mu}_{b, i}^{\ast})}
-%\]
-%To summarize the evidence across batches, one strategy is to use the most significant batch. With balanced design, the batch with the largest test statistic across all batch is selected, i.e., 
-%\[
-%b_{\text{max}} = \{ b \mid t_b = \max_{b'=1, \ldots, B} | t_{b'} | \}.
-%\] 
-%The change in modification abundance is estimated as
-%\[
-%\hat{\mu}_{i}^{\ast} - \hat{\mu}_{i'}^{\ast} = \hat{\mu}_{b_{\text{max}}, i}^{\ast} - \hat{\mu}_{b_{\text{max}}, i'}^{\ast}
-%\]
-%and the test statistic for the $t$-test is given by
-%\[
-%\frac{\hat{\mu}_{i}^{\ast} - \hat{\mu}_{i'}^{\ast}}{\mathrm{SE}\left( \hat{\mu}_{i}^{\ast} - \hat{\mu}_{i'}^{\ast} \right)} 
-%= \frac{ \hat{\mu}_{b_{\text{max}}, i}^{\ast} - \hat{\mu}_{b_{\text{max}}, i'}^{\ast} }{\left[ \mathrm{SE}(\hat{\mu}_{b_{\text{max}}, i}^{\ast})^2 + \mathrm{SE}(\hat{\mu}_{b_{\text{max}}, i'}^{\ast})^2 \right]^{1/2}},
-%\]
-%%where $\hat{\sigma}_{\pi, ijk}^{2}$ and $\hat{\sigma}_{\pi, ijk'}^{2}$ are estimated as sample variances. 
-%
-%
-%\paragraph{Two-sample $t$-test (all batch).}
-%A change in modification abundance is considered as statistically significant only if such change is significant in all batches.
-This is equivalent to take the smallest test statistic across batches for the $t$-test:
-%\[
-%\frac{\hat{\mu}_{i}^{\ast} - \hat{\mu}_{i'}^{\ast}}{\mathrm{SE}\left( \hat{\mu}_{i}^{\ast} - \hat{\mu}_{i'}^{\ast} \right)} 
-%= \frac{ \hat{\mu}_{b_{\text{min}}, i}^{\ast} - \hat{\mu}_{b_{\text{min}}, i'}^{\ast} }{\left[ \mathrm{SE}(\hat{\mu}_{b_{\text{min}}, i}^{\ast})^2 + \mathrm{SE}(\hat{\mu}_{b_{\text{min}}, i'}^{\ast})^2 \right]^{1/2}},
-%\]
-%where 
-%\[
-%b_{\text{min}} = \{ b \mid t_b = \min_{b'=1, \ldots, B} | t_{b'} | \}.
-%\]
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Proposed approach}
+\section{Details of the proposed approach}
 \label{sec:prop}
-
-%To characterize the observed feature intensities, different levels of variations are expressed using linear mixed models in consideration of the following factors in a simple experimental design: modification, condition, run, and feature.
-As different degrees of variability are present in the feature intensities of modified and unmodified peptides, they are expressed by separate models.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\subsection{Statistical modeling and inference}
-%\label{sec:model}
-
-%The observed log-intensity of a modified peptide feature is denoted by $y_{ijk}$ and represented as
-%\[
-%y_{ijk} = \psi + C_{i} + R_{j(i)} + F_{k} + (R \times F)_{ijk},
-%\]
-%where the effects of condition and feature are modeled as fixed effects: 
-%\[
-%\sum_{i=1}^{I} C_{i} = 0, \quad \sum_{k=1}^{K} F_{k} = 0,
-%\]
-%and the effects of run and its interaction with feature are considered as random effects arising from normal distribution with mean $0$: 
-%\[
-%R_{j(i)} = \gamma_{j(i)} \stackrel{\text{iid}}{\sim} \mathcal{N}(0, \sigma_{\gamma}^{2}), \qquad
-%(R \times F)_{ijk} = \epsilon_{ijk} \stackrel{\text{iid}}{\sim} \mathcal{N}(0, \sigma_{\epsilon}^{2}).
-%\]
-%Similarly, the observed log-intensity of an unmodified peptide feature is denoted by $y_{ijl}^{\ast}$ and represented as 
-%\[
-%y_{ijl}^{\ast} = \psi^{\ast} + C_{i}^{\ast} + R_{j(i)}^{\ast} + F_{l}^{\ast} + (R \times F)_{ijl}^{\ast}, 
-%\]
-%where the effects of condition and feature are modeled as fixed effects: 
-%\[
-%\sum_{i=1}^{I} C_{i}^{\ast} = 0, \quad \sum_{l=1}^{L} F_{l}^{\ast} = 0,
-%\]
-%and 
-%\[
-%R_{j(i)}^{\ast} = \gamma_{j(i)}^{\ast} \stackrel{\text{iid}}{\sim} \mathcal{N}(0, \sigma_{\gamma^{\ast}}^{2}), \qquad
-%(R \times F)_{ijl}^{\ast} = \epsilon_{ijl}^{\ast} \stackrel{\text{iid}}{\sim} \mathcal{N}(0, \sigma_{\epsilon^{\ast}}^{2}).
-%\]
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\subsection{Run-level summarization of feature intensities}
-%\label{sec:sum}
-
-%Run-level summarization of feature intensities for each PTM site is carried out as in the sub-plot model of MSstats~\cite{choi_etal_14a} or MSstatsTMT~\cite{Huang:2020} depending on acquisition strategy, which involves a) imputation of censored missing values, and b) summarization of feature intensities using Tukey's median polish.
-\cite{tukey_77a} The run-level summary for the PTM in Run $j$ of Condition $i$ is denoted by $\hat{y}_{ij}^{\ast}$.
-From here forward the input to the method assumes that all features are summarized.
-
-%\subsection{Balanced design with one source of variation}
-%\label{sec:model}
-%The methods when applied to a balanced design with multiple conditions and one technical replicate are shown in this section.
-In practice the experimental design can be unbalanced and variation can come from multiple sources.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\subsubsection{Model-based inference of the underlying abundance}
-%\label{sec:infer}
-
-%The PTM abundance in each run is represented as 
-%\[
-%\hat{y}_{ij}^{\ast} = \psi^{\ast} + C_{i}^{\ast} + R_{j(i)}^{\ast},
-%\]
-%where $\sum_{i=1}^{I} C_{i}^{\ast} = 0$, $R_{j(i)}^{\ast} = \gamma_{j(i)}^{\ast} \stackrel{\text{iid}}{\sim} \mathcal{N}(0, \sigma_{\gamma^{\ast}}^{2})$. 
-%Similarly, the protein abundance in each run is expressed as
-%\[
-%\hat{y}_{ij} = \psi + C_{i} + R_{j(i)},
-%\]
-%where $\sum_{i=1}^{I} C_{i} = 0$, $R_{j(i)} = \gamma_{j(i)} \stackrel{\text{iid}}{\sim} \mathcal{N}(0, \sigma_{\gamma}^{2})$.
-The expected values of log-abundances of the PTM and protein in Condition $i$ are denoted by $\mu_{i}^{\ast}$ and $\mu_{i}$, respectively, and the values are estimated as:
-%\begin{align*}
-%\hat{\mu}_{i}^{\ast} &= \hat{\psi}^{\ast} + \hat{C}_{i}^{\ast} = \frac{1}{J} \hat{y}_{i+}^{\ast} \\
-%\hat{\mu}_{i} &= \hat{\psi} + \hat{C}_{i} = \frac{1}{J} \hat{y}_{i+},
-%\end{align*}
-%where the standard errors of the estimates are $\mathrm{SE^{\ast}}(\hat{\mu}_{i}) = (\hat{\sigma}_{\gamma^{\ast}}^{2} / J)^{1/2}$ and $\mathrm{SE}(\hat{\mu}_{i}) = (\hat{\sigma}_{\gamma^{\ast}}^{2} / J)^{1/2}$.
-%\[
-%\left( \frac{\hat{\sigma}_{\gamma}^{2}}{J} \right)^{1/2}, \quad \left( \frac{\hat{\sigma}_{\gamma^{\ast}}^{2}}{J} \right) ^ {1/2}.
-%\]
-%\[
-%\left( \frac{1}{J} \hat{\sigma}_{\gamma}^{2} \right)^{1/2}, \quad \left( \frac{\hat{\sigma}_{\gamma^{\ast}}^{2}}{J} \right) ^ {1/2}.
-%\]
-%Based on the estimates $\hat{\mu}_{i}^{\ast}$ and $\hat{\mu}_{i}$, the adjusted log-abundance of the PTM is given by $(\hat{\mu}_{i}^{\ast} - \hat{\mu}_{i})$ and the standard error of the estimate is
-%\[
-%\left[ \frac{1}{J} \left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right) \right]^{1/2}.
-%\]
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\subsubsection{PTM significance analysis}
-%\label{sec:test}
-
-%With protein-level adjustment, the model-based testing is based on the hypothesis that there is no difference in adjusted PTM abundance between Conditions $i$ and $i'$
-%Adjusted by protein abundance, the hypothesis states that there is no difference in log-abundance of the modification between groups $i$ and $i'$
-%\begin{align*}
-%H_{0}: \Delta = \left( \mu_{i}^{\ast} - \mu_{i'}^{\ast} \right) - \left( \mu_{i} - \mu_{i'} \right) &= 0 \\
-%H_{a}: \Delta = \left( \mu_{i}^{\ast} - \mu_{i'}^{\ast} \right) - \left( \mu_{i} - \mu_{i'} \right) &\neq 0
-%\end{align*}
-%In a balanced design, the log-fold change in the adjusted PTM abundance, $\Delta$, is estimated by 
-%\[
-%\hat{\Delta} = \left[ \frac{1}{J} \left( \hat{y}_{i+}^{\ast} - \hat{y}_{i'+}^{\ast} \right) \right] - \left[ \frac{1}{J} \left( \hat{y}_{i+} - \hat{y}_{i'+} \right) \right],
-%\]
-%and the standard error of the estimate $\mathrm{SE}(\hat{\Delta})$ is 
-%\[
-%SE(\hat{\Delta}) = \left[ \frac{2}{J} \left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right) \right]^{1/2}.
-%\]
-%\[
-%\left[ \left( \frac{2}{J} \right) \cdot \left( \hat{\sigma}_{\gamma}^{2} + \hat{\sigma}_{\gamma^{\ast}}^{2} \right) \right]^{1/2}.
-%\]
-%The test statistic $\hat{\Delta} / \mathrm{SE}(\hat{\Delta})$ is compared against the $t$ distribution, with degrees of freedom approximated by
-%\[
-%\left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right)^2 \bigg/
-%\left( \frac{\hat{\sigma}_{\gamma^{\ast}}^{4}}{\mathrm{df}(\gamma^{\ast})} + \frac{\hat{\sigma}_{\gamma}^{4}}{ \mathrm{df}(\gamma)} \right).
-%\]
-%A distinctive property of the proposed model-based testing to the two-sample $t$-test is that even only the PTM abundances in Conditions $i$ and $i'$ are compared, measurements from all conditions are used for the modeling and inference.
-
-%The formulations for $\hat{\Delta}$, $SE(\hat{\Delta})$, and degrees of freedom shown above are for a simple balanced experiment.
-They can also be extended to more realistic experiments with complex designs.
-This is discussed further in Section \ref{sec:complex_methods}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\subsubsection{Design of PTM experiments}
-%\label{sec:design}
-
-%The proposed statistical framework allows for design of PTM experiments in terms of sample size calculation and power analysis. 
-%Sample size calculation based on linear models has been described for general applications~\cite{kutner_etal_04a} and specifically for protein significance analysis~\cite{oberg_vitek_09a}. 
-%Sample size calculation takes as input a) $q$, the desired false discovery rate, b) $\beta$, the average Type II error rate, c) $\Delta$, the minimal log-fold change in adjusted PTM abundance that we would like to detect, d) $m_0 / (m_0 + m_1)$, the fraction of truly differentially modified PTM sites in the comparison, and e) $\sigma_{\gamma^{\ast}}^{2}$ and $\sigma_{\gamma}^{2}$, the anticipated variances associated to modified and unmodified peptide features, respectively.
-The variances can be derived based on the dataset being analyzed, assuming similar quantitative properties and variations.
-With these values and a user-specified number of conditions, the corresponding number of technical replicates per condition can then be derived, as described in~\cite{kutner_etal_04a}.
-Given the above quantities, the minimal number of replicates $J$ is determined by the variance of the estimated log-fold change $\mathrm{SE}^{2}(\hat{\Delta})$ as
-%\[
-%\mathrm{SE}^{2}(\hat{\Delta}) = \left[ \frac{2}{J} \left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right) \right]
-%\leq \left( \frac{\Delta}{t_{1-\beta, df} + t_{1-\alpha /2, df}} \right)^{2},
-%\]
-%where 
-%\[
-%\alpha = (1 - \beta) \cdot \frac{q}{1 + (1-q) \cdot m_0 / m_1},
-%\]
-%and $t_{1-\beta, df}$ and $t_{1-\alpha /2, df}$ are the $100(1-\beta)^{\text{th}}$ and the $100(1-\alpha /2)^{\text{th}}$ percentiles of the $t$ distribution, with $df = I(J-1)$ degrees of freedom in balanced designs.
-More details can be found in~\cite{oberg_vitek_09a}.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\subsection{Extension to complex designs}
-%\label{sec:complex}
-%
-%The proposed statistical framework can be extended to analyze data from experiments of complex designs, such as factorial design and time series.
-We discuss below a specific design commonly applied in PTM experiments, in which data are acquired in multiple batches.
-%
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\subsubsection{Batch effects}
-%\label{sec:batch}
-%
-%As in \secref{sec:test}, hypothesis testing on the adjusted log-abundances of the PTM in Conditions $i$ and $i'$ is performed to detect differentially modified PTM sites.
-The hypothesis is that there is no difference in adjusted log-abundance of the PTM between Conditions $i$ and $i'$
-%\begin{align*}
-%H_{0}: \Delta = (\mu_{i} - \mu_{i'}) - (\mu_{i}^{\ast} - \mu_{i'}^{\ast}) &= 0 \\
-%H_{a}: \Delta = (\mu_{i} - \mu_{i'}) - (\mu_{i}^{\ast} - \mu_{i'}^{\ast}) &\neq 0
-%\end{align*}
-%For batch-wise data, we consider two ways to estimate the difference in adjusted PTM abundance, $\Delta = (\mu_{i} - \mu_{i'}) - (\mu_{i}^{\ast} - \mu_{i'}^{\ast})$ based on different assumptions about the properties of batch effects, namely per-batch model (proposed approach) and all-batch model.
-For the following discussion, we denote the log-intensity of a modified peptide feature in Run $j$ of Condition $i$ and Batch $b$ by $y_{b, ijk}$, where $b = 1, \ldots, B$.
-Similarly, $y_{b, ijl}^{\ast}$ is denoted for the unmodified peptide feature in the same run.
-%
-%\paragraph{Per-batch model (proposed approach).} 
-%The model assumes different levels of variability are present in different batches and the differences between conditions vary across batches (i.e., there is an interaction effect between condition and batch).
-Difference between conditions is estimated in each batch, and the overall log-fold change in adjusted PTM abundance is estimated as the average over batches
-%\[
-%\hat{\Delta} = \frac{1}{B} \sum_{b=1}^{B} \left[ \frac{1}{J} \left( \hat{y}_{b,i+} - \hat{y}_{b,i'+} \right) \right]
-%- \frac{1}{B} \sum_{b=1}^{B} \left[ \frac{1}{J} \left( \hat{y}_{b,i+}^{\ast} - \hat{y}_{b,i'+}^{\ast} \right) \right].
-%\]
-%The standard error associated to the estimate is 
-%\[
-%\left[ \left(\frac{1}{B}\right)^2 \cdot \left( \frac{2}{J} \right) \cdot \sum_{b=1}^{B} \left( \hat{\sigma}_{\gamma_b}^{2} + \hat{\sigma}_{\gamma_b^{\ast}}^{2} \right) \right]^{1/2}.
-%\]
-%The test statistic $\hat{\Delta} / \mathrm{SE}(\hat{\Delta})$ is compared against the $t$ distribution, where the degrees of freedom are approximated as
-%\[
-%\left[ \sum_{b=1}^{B} \left( \hat{\sigma}_{\gamma_{b}}^{2} + \hat{\sigma}_{\gamma_{b}^{\ast}}^{2} \right) \right]^2 \bigg/
-%\sum_{b=1}^{B} \left[ \frac{\hat{\sigma}_{\gamma_{b}}^{4}}{  \mathrm{df}(\gamma_b)} + \frac{\hat{\sigma}_{\gamma_{b}^{\ast}}^{4}}{\mathrm{df}(\gamma_b^{\ast})} \right].
-%\]
-%
-%\paragraph{All-batch model.}
-%In contrast to the per-batch model, the all-batch model assumes identical variance and difference between conditions in all batches.
-The log-fold change in adjusted PTM abundance is estimated as the average over runs and batches
-%\[
-%\hat{\Delta} = \frac{1}{BJ} \left( \hat{y}_{+,i+} - \hat{y}_{+,i'+} \right) - \frac{1}{BJ} \left( \hat{y}_{+,i+}^{\ast} - \hat{y}_{+,i'+}^{\ast} \right),
-%\]
-%and the standard error of the estimate is 
-%\[
-%\left[ \frac{2}{BJ} \left( \hat{\sigma}_{\gamma}^{2} + \hat{\sigma}_{\gamma^{\ast}}^{2} \right) \right] ^{1/2}.
-%\]
-%The test statistic $\hat{\Delta} / \mathrm{SE}(\hat{\Delta})$ is compared against the $t$ distribution, with degrees of freedom approximated by
-%\[
-%\left( \hat{\sigma}_{\gamma}^{2} + \hat{\sigma}_{\gamma^{\ast}}^{2} \right)^2 \bigg/
-%\left( \frac{\hat{\sigma}_{\gamma}^{4}}{  \mathrm{df}(\gamma)} + \frac{\hat{\sigma}_{\gamma^{\ast}}^{4}}{\mathrm{df}(\gamma^{\ast})} \right).
-%\]
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\subsection{Extension to complex designs}
-%\label{sec:complex_methods}
-
-%The methods described previously in Section \ref{sec:model} can be extrapolated to experiments with complex designs targeting PTMs.
-This includes experiments with additional sources of variation and unbalanced designs.
-
-%\subsubsection{Model-based inference and significance analysis}
-
-%Model inference is done in the same way to how MSstats~\cite{choi_etal_14a} and MSstatsTMT~\cite{Huang:2020} target the global protein.
-The null hypothesis is still that there is no difference in mean PTM abundance between Conditions $i$ and $i'$ after adjusting for changes in mean unmodified peptide abundance between Conditions $i$ and $i'$.
-%\begin{align*}
-%H_{0}: \Delta = (\mu_{i}^{\ast} - \mu_{i'}^{\ast}) - (\mu_{i} - \mu_{i'}) &= 0 \\
-%H_{a}: \Delta = (\mu_{i}^{\ast} - \mu_{i'}^{\ast}) - (\mu_{i} - \mu_{i'}) &\neq 0
-%\end{align*}
-
-%In cases where the design in unbalanced, a linear mixed effects model is fit, which takes into account all potential sources of variation, and restricted maximum likelihood is used to estimate the parameters of each model.
-Once the parameters are estimated, we can combine the models using modified versions of the formulas in Section \ref{sec:test}.
-
-%The log-fold change in the adjusted PTM abundance, $\Delta$, is estimated by 
-%\[
-%\hat{\Delta} = (\hat{\mu}_{RML_i}^{\ast} - \hat{\mu}_{RML_{i'}}^{\ast}) - (\hat{\mu}_{RML_i} - \hat{\mu}_{RML_{i'}})
-%\]
-%and the standard error of the estimate $\mathrm{SE}(\hat{\Delta})$ is 
-%\[
-%SE(\hat{\Delta}) = \left[ \left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right) \right]^{1/2}.
-%\]
-%\[
-%\left[ \left( \frac{2}{J} \right) \cdot \left( \hat{\sigma}_{\gamma}^{2} + \hat{\sigma}_{\gamma^{\ast}}^{2} \right) \right]^{1/2}.
-%\]
-%The test statistic $\hat{\Delta} / \mathrm{SE}(\hat{\Delta})$ is compared against the $t$ distribution, with degrees of freedom approximated by
-%\[
-%\left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right)^2 \bigg/
-%\left( \frac{\hat{\sigma}_{\gamma^{\ast}}^{4}}{\mathrm{df}(\gamma^{\ast})} + \frac{\hat{\sigma}_{\gamma}^{4}}{ \mathrm{df}(\gamma)} \right).
-%\]
-%
-
-%An example of this is in TMT experiments where additional variation comes from the mixtures used.
-Going forward we will denote the log-intensity of a modified peptide for Technical Replicate j of Mixture m for Condition i by $y_{ijm}$, where $m = 1,....,M$.
-This is done in the same way for the unmodified peptide $y^*_{ijm}$.
-
-%The model assumes the variability of the mixture is a random effect and is not convoluted with the other variables in the model.
-The model also assumes that the difference between conditions is the same between mixtures.
-For a given modified peptide and corresponding unmodified protein a mixed effects model is fit as follows.
-%\begin{align*}
-%y^*_{mtcb} = \mu^* + M^*_m + T(M)^*_{tm} + C^*_c + S^*_{mcb} + \epsilon^*_{mtcb}\\
-%y_{mtcb} = \mu + M_m + T(M)_{tm} + C_c + S_{mcb} + \epsilon_{mtcb}
-%\end{align*}
-
-%Where $M_m \sim N(0, \sigma^2_m)$, $T(M)_{tm} \sim N(0, \sigma^2_{tm})$, $S\sim N(0, \sigma^2_{s})$ are random effects, and $\sum_{c=1}^{I} C_{c}^{\ast} = 0$ is a fixed effect.
-Finally the error is expressed as $\epsilon_{mtcb} \sim N(0, \sigma^2)$.
-
-\subsection{Design of PTM experiments - Extension to complex designs}
-\label{sec:complex_design}
-
-%The proposed approach applied to experiments with complex designs still allows for sample size calculation and power analysis.
-To make the calculation the standard deviation and degrees of freedom are needed.
-The estimation of these quantities is dependent on the experimental design.
-In Figure \ref{fig:statistical_inference} different experimental designs that the proposed approach can be applied to, along with their estimation of standard error and degrees of freedom is shown.
 
@@
 \makebox[\textwidth][c]{\includegraphics[width=1.2\textwidth]{sim_new/Statistical_inference}}%
-\caption{Different models that are fit depending on the experimental design.
-Models are fit for label-free and TMT acquisition methods, as well as group comparison and time course experimental designs.
-The estimation of standard error and degrees of freedom for each model is shown.
-Only one side of the PTM and unmodified protein is shown, variance must be multiplied by two to combine.
+\caption{Different models that are fit depending on the experimental design (group comparison and time course) and quantification workflow (label-free versus TMT).
+The table shows the true values of the standard errors, along with their estimates and the associated degrees of freedom.
+The same formulas holds when comparing changes in PTM, or hcnges in the unmodified portion of the protein.
+When combining the two comparisons for an adjustment, the variance must be multiplied by 2.
 \label{fig:statistical_inference}}
@@
 
-%The standard error and degrees of freedom are estimated separately for both the PTM and unmodified protein models.
-Once this is done we can use these these values to calculate the sample size and perform power analysis.
-Given $q$, the desired false discovery rate, $\beta$, the average Type II error rate, $m_0 / (m_0 + m_1)$, the fraction of truly differentially modified PTM sites in the comparison.
-The minimal number of replicates is determined by the variance of the estimated log-fold change $\mathrm{SE}^{2}(\hat{\Delta})$ as
-%\[
-%\mathrm{SE}^{2}(\hat{\Delta}) = \left[ \left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right) \right]
-%\leq \left( \frac{\Delta}{t_{1-\beta, df} + t_{1-\alpha /2, df}} \right)^{2},
-%\]
-%Where 
-%\[
-%\alpha = (1 - \beta) \cdot \frac{q}{1 + (1-q) \cdot m_0 / m_1},
-%\]
-%and $t_{1-\beta, df}$ and $t_{1-\alpha /2, df}$ are the $100(1-\beta)^{\text{th}}$ and the $100(1-\alpha /2)^{\text{th}}$ percentiles of the $t$ distribution.
-The degrees of freedom are estimated for the appropriate model using restricted maximum likelihood.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\clearpage
+\section{Details on the generation of simulated datasets}
+\label{sec:experiments}
 
-\clearpage
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Experimental datasets}
-\label{sec:experiments}
-\subsection{Experiments with known ground truth}
-\label{sec:sim}
-
-We compared six different approaches as follows: a) $MSstatsPTM$, b) base $MSstats$ c) $adjusted$ $ANOVA$~\cite{kutner_etal_04a}, d) $Unadjusted$ $ANOVA$, e) $adjusted$ $Limma$~\cite{Smyth:2004}, and f) $unadjusted$ $Limma$.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection{Dataset 1 : Computer simulation 1 - Label-free}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%e%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Dataset 1 : Computer simulation 1 - Label-free}
 \label{sec:dataset1}
@@
 
-\subsubsection{Dataset 2 : Computer simulation 2 - Label-free missing values and low features}
+\subsection{Dataset 2 : Computer simulation 2 - Label-free missing values and low features}
 \label{sec:dataset2}
@@
 
-%\begin{figure}[h!]
-%\centering
-% \begin{subfigure}{\textwidth}
-%  \centering
-%	\includegraphics[width=.8\textwidth]{sim_new/sim1_Recall}\\
-%	\caption{All methods with adjustment were compared by comparing recall in simulation 1.
-Limma performed the strongest here when the number of replicates were low.
-At higher replicates the performance of the proposed methods and Limma were comparable.
-Anova clearly performed worse across all methods.}
-% \end{subfigure}
-% \begin{subfigure}{\textwidth}
-%  \centering
-%	\includegraphics[width=.8\textwidth]{sim_new/sim1_accuracy}
-%	\caption{The overall accuracy plot mimicked the observations in the recall plot. Limma performed strongest when replicates were low, but was comparable to the proposed method with more replicates.}
-% \end{subfigure}
-%\caption{Recall and Accuracy results of Simulation 1.}
-%\label{fig:sim1_recall}
-%\end{figure}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\clearpage
+\section{Additional evaluation results}
+\label{sec:experiments}
 
-%\begin{figure}[h!]
-%\centering
-%	\includegraphics[width=.8\textwidth]{sim_new/sim3_Recall}
-%	\caption{}
-%\caption{Recall results of Simulation 2.
-The advantage of using the proposed approach were apparent when looking at simulation 2, which included limited observations and the presence of missing values.
-In the case of recall the proposed method performed stronger than Limma and t-test in nearly every model.
-Even at lower replicates the proposed method still outperformed Limma.
-Again the lowest performing method was Anova.}
-%\label{fig:sim2_recall}
-%\end{figure}
+\subsection{Simulated Datasets 1 and 2}
 
@@
 \centering
- \begin{subfigure}{\textwidth}
- \centering
-	\includegraphics[width=.8\textwidth]{sim_new/sim1_FC_boxplot}
-	\caption{}
-	\label{fig:fc_boxplota}
- \end{subfigure}
- \begin{subfigure}{\textwidth}
-  \centering
-	\includegraphics[width=.8\textwidth]{sim_new/sim3_FC_boxplot}
-	\caption{}
-	\label{fig:fc_boxplotb}
- \end{subfigure}
-\caption{Fold change distribution comparison between Simulations 1 and 2.
-a) In simulation 1 all considered methods correctly estimated the fold change between conditions, with a median fold change estimation of 1.
+ \begin{tabular}{cc}
+	\includegraphics[width=.445\textwidth]{sim_new/sim1_FC_boxplot}
+	&
+	\includegraphics[width=.56\textwidth]{sim_new/sim3_FC_boxplot}\\
+	(a)&(b)
+\end{tabular}
+\caption{\todo{In (b) could you indicate in panel labels that we are still talking about conditions and SD?}Simulated Datasets 1 and 2 : fold change distributions.
+(a) In Simulated Dataset 1 all considered methods correctly estimated the fold change between conditions, with a median fold change estimation of 1.
 The distributions around the median were also consistent across all methods.
-b) In simulation 2 all methods correctly estimated the fold change with a median log change of 1.
-$MSstatsPTM$ in this simulation had a visibly tighter distribution around the median.
-Both $Limma$ and $ANOVA$ showed a wider range around the fold change.}
+(b) In Simulated Dataset 2 all methods correctly estimated the fold change with a median log change of 1.
+$MSstatsPTM$ in this simulation had a  tighter distribution around the median.
+Both $Limma$ and $ANOVA$ showed a wider range around the fold change.
+\todo{In is not clear from the text whether Limma and ANOVA were used with or without adjustments for unmodified proteins.
+Could you clarify?}}
 \label{fig:fc_boxplot}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \clearpage
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection{Dataset 3 : SpikeIn benchmark - Ubiquitination - Label-free}
+\subsection{Dataset 3 : SpikeIn benchmark - Ubiquitination - Label-free}
 \label{sec:benchmark}
-
-%A custom designed experiment with labeling was used to assess the performance of the proposed method in a real experimental setting.
-Heavy-labeled KGG modified peptides were used as spike-in peptides.
-The spike-in peptides were mixed with human lysate to create four mixture conditions.
-Two sets of data were acquired for each mixture: KGG enriched + LC-MS, and LC-MS only.
-The KGG enriched dataset included the spike-in peptides, as well as modified and unmodified human lysate.
-The LC-MS dataset included only unmodified peptides.
-The spike-in peptides have a known fold change between conditions and are expected to be differential in most cases.
-Additionally, background E coli lysate was used to normalize total protein levels prior to the enrichment and global protein profiling.
-The background lysate was treated as controls and are not expected to be differential in any comparison.
-
-%\begin{figure}[h!]
-%\centering
-%\includegraphics[width=.8\textwidth]{sim_new/SpikeIn_Overview}
-%\caption{An overview of the spike-in benchmark experimental design.
-The corresponding levels of spike-in peptide, human lysate, and background lysate for each mixture are shown.
-The spike-in peptide levels are adjusted for the human lysate levels.
-For example, the true spike-in abundance of mixture 1 is entirely accounted for by human lysate, whereas mix 2 has double the amount human lysate compared to spike-in abundance.
-When comparing mix 1 vs mix 2 we would expect to see a log fold change of -1.
-In this way we know the expected log fold change between mixture comparisons.
-\label{fig:spikein_overview}}
-%\end{figure}
 
@@
 \includegraphics[width=1\textwidth]{sim_new/spike_in_msstatsptm_volcano}
-\caption{The modeling results of the $MSstatsPTM$ both before and after adjustment.
+\caption{\todo{Clarify that MSstats is the basic version before the adjustment, and MSstatsPTM after the adjustment} \todo{Start the caption with the name of the dataset} The modeling results of the $MSstatsPTM$ both before and after adjustment.
 The spike-in peptides are colored red and the background peptides are colored grey.
-All grey peptides are expected to be insignificant.
-Using the proposed method to model the benchmark experiment, the spike-in peptides (colored red) did not follow the expected log fold change before adjustment.
+All grey peptides are expected to not be detected as differentially abundant.
+The spike-in peptides (colored red) did not follow the expected log fold change before adjustment.
 After adjusting for changes in overall protein abundance the spike-in peptides were more in line with expectation.
@@
 \includegraphics[width=1\textwidth]{sim_new/spike_in_limma_volcano}
-\caption{When modeling the experiment with the $Limma$ method, the spike-in peptides again follow the expected log fold change better after adjusting for changes in protein level.
-However, while the fold change was more accurate, the majority of spike-in peptides did not have a significant adjusted pvalue.
-In this case, the known differential peptides were missed by the model.
-In terms of false positives, the results were very similar to the proposed method, with many false positives before adjustment and fewer after.
+\caption{\todo{Start the caption with the name of the dataset} When modeling the experiment with the $Limma$ method, the spike-in peptides again follow the expected log fold change better after adjusting for changes in protein level.
+However, while the fold change was more accurate, the majority of spike-in peptides were not detected as differentially abundant.
+There were more false positive differentially abundant PTM before than after adjustment.
 \label{fig:spike_volcano_limma}}
@@
 \includegraphics[width=1\textwidth]{sim_new/spike_in_anova_volcano}
-\caption{Using $ANOVA$ the fold change of the spike-in peptides was much closer to expectation after adjusting for global protein abundance.
-The log FC estimation was the same as $Limma$, however the pvalues were different.
-We actually see a few more True Positives using $ANOVA$ compared to $Limma$.\label{fig:spike_volcano_ttest}}
+\caption{\todo{Start the caption with the name of the dataset} Using $ANOVA$ the fold change of the spike-in peptides was much closer to expectation after adjusting for global protein abundance.
+The log FC estimation was the same as $Limma$, however the p-values were different.
+In this particular case we detect a few more true positives using $ANOVA$ compared to $Limma$.\label{fig:spike_volcano_ttest}}
 \end{figure}
 
-%\begin{figure}[h!]
-%\centering
-%\includegraphics[width=.85\textwidth]{sim_new/spike_in_statistic_comparison}
-%\caption{Comparing the summary statistics between methods, the proposed method performed the strongest.
-In terms of accuracy and specificity the three methods were close, with Limma and Anova showing slightly higher values.
-Accuracy and specificity were dominated by the large number of true negatives (background peptides) compared to the true positives (spike-in peptides).
-In terms of recall, the proposed approach out performed the other two methods, showing that it correctly labeled the most spike-in peptides.
-\label{fig:spike_stat}}
-%\end{figure}
-
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \clearpage
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Biological investigations}
-\label{sec:bio_investigations}
-
-\subsubsection{Dataset 4 : Human - Ubiquitination - 1mix-TMT~\cite{LUCHETTI2021}}
+\subsection{Dataset 4 : Human - Ubiquitination - 1mix-TMT}
 \label{sec:ipah}
 
-The experimental design can be seen in Table~\ref{table:ipah_design}.
+The experiment had a simple group comparison design, shown in Table~\ref{table:ipah_design}.
 
@@
 \hline
-
 \end{tabular}
@@
 \end{table}
-
-A model was fit for the total protein and ubiquitination separately. The model formula can be seen below.
-
-$$Y_{mij} = \mu_i + \epsilon_{mij}$$
-
-$$\sum_{i=1}^I{\mu_i} = 0 ,\: \epsilon_{mij} ~ \sim N(0, \sigma^2)$$
-
+The following model was fit separately for ubiquitinated peptides and for unmodified protein 
+\begin{eqnarray*}
+& Y_{mij} = \mu_i + \epsilon_{mij},\ \sum_{i=1}^I{\mu_i} = 0 ,\ \epsilon_{mij} ~ \sim N(0, \sigma^2) &
+\end{eqnarray*}
 \begin{figure}[h!]
  \centering
-	\includegraphics[width=.9\textwidth]{sim_new/IpaH_volcano_plot}
-	\caption{Volcano plots of Dox4hr vs Dox1hr both before and after protein adjustment.
+	\includegraphics[width=.8\textwidth]{sim_new/IpaH_volcano_plot}
+	\caption{\todo{Could you separate the panels, and add labels (a) and (b) in the subfigures.
+(Could you use tabular as I did everywhere?) The title of the second panel is cut off.
+It is not clear which method is used - is it MSstats and MSstatsPTM?} \todo{Start the caption with the name of the dataset} Volcano plots of Dox4hr vs Dox1hr both before and after protein adjustment, \todo{with $MSstatsPTM$}.
 The $GSDMD\_HUMAN|P57764\_K62$ modification is highlighted.
-Before adjustment the modification had a small fold change and insignificant adjusted pvalue.
-After adjustment the fold change was much larger and the adjusted pvalue was significant.
-In this case the proposed method allowed us to identify a differential modified peptide that could have otherwise been missed.}
+(a) Before adjustment the modification had a small fold change and was not detected as differentially abundant.
+(b) After adjustment the fold change was much larger, and the modification was detected as differentially abundant.
+In this case \todo{$MSstatsPTM$?} allowed us to identify a differential modified peptide that could have otherwise been missed.}
 \label{fig:ipah_figures}
@@
 \clearpage
-\subsubsection{Dataset 5 : Mouse - Phosphorylation - 2mix-TMT~\cite{Maculins}}
+\subsection{Dataset 5 : Mouse - Phosphorylation - 2mix-TMT}
 \label{sec:shigella}
 
-The experimental design can be seen in Table~\ref{table:shigella_design}.
+The experiment had a group comparison design, and the data were acquired in two mixtures, as shown in Table~\ref{table:shigella_design}.
 
@@
 \end{table}
-
-A model was fit for the total protein, and phosphorylation. The model formula can be seen below.
-
-$$Y_{mij} = \mu_i + Mixture_m + \epsilon_{mij}$$
-
-$$Mixture_m ~ \sim N(0, \sigma^2_M) ,\: \sum_{i=1}^I{\mu_i} = 0 ,\: \epsilon_{mij} ~ \sim N(0, \sigma^2)$$
+The following model was fit separately for phosphorylated peptides and for unmodified protein 
+\begin{eqnarray*}
+& Y_{mij} = \mu_i + Mixture_m + \epsilon_{mij},\ Mixture_m ~ \sim N(0, \sigma^2_M) ,\  \sum_{i=1}^I{\mu_i} = 0 ,\ \epsilon_{mij} ~ \sim N(0, \sigma^2)&
+\end{eqnarray*}
 
@@
 \centering
-	\includegraphics[width=.9\textwidth]{sim_new/No_Difference_Shigella_Volcano}
-	\caption{Volcano plots of WT\_Late vs WT\_Uninfect both before and after protein adjustment.
+	\includegraphics[width=.8\textwidth]{sim_new/No_Difference_Shigella_Volcano}
+	\caption{\todo{Start the caption with the name of the dataset} \todo{Separate the labels and add labels (a) and (b) for the subpanels.
+(Could you use tabular as I did everywhere?)}Volcano plots of WT\_Late vs WT\_Uninfect both before and after protein adjustment \todo{with $MSstatsPTM$}.
 The $TTP\_MOUSE|P22893\_S178$ modification is highlighted.
-Before adjustment the modification had a large fold change and significant adjusted pvalue.
-After adjustment the fold change was much smaller and the adjusted pvalue was insignificant.}
+(a) Before adjustment the modification had a large fold change and a small p-value.
+(b) After adjustment the fold change was much smaller and the modification was not detected as differentially abundant.}
 \label{fig:No_Diff_Shigella_PTM}
@@
 \centering
- \begin{subfigure}{\textwidth}
- \centering
-	\includegraphics[width=.9\textwidth]{sim_new/Difference_Shigella_Profile_Plot}
-	\caption{}
- \end{subfigure}
- \begin{subfigure}{\textwidth}
- \centering
-	\includegraphics[width=.9\textwidth]{sim_new/Difference_Shigella_Volcano}
-	\caption{}
- \end{subfigure}
- \caption{Summary plots for modification of protein KI67 at site T215.
-a) The global profiling of protein $KI67\_MOUSE|E9PVX6$ with the modification at site $T215$ was compared.
-In this case the modification and global protein appeared to show small or no difference between conditions, however after adjusting for change in global protein abundance, the modification was statistically significant.
+ \begin{tabular}{c}
+	\includegraphics[width=.9\textwidth]{sim_new/Difference_Shigella_Profile_Plot}\\
+	(a)\\
+	\includegraphics[width=.8\textwidth]{sim_new/Difference_Shigella_Volcano}\\
+	(b)
+ \end{tabular}
+ \caption{\todo{Start the caption with the name of the dataset} Protein $KI67\_MOUSE|E9PVX6$ with the modification at site $T215$.
+(a) The modification and global protein appeared to show small or no difference between conditions when considered separately.
+However after adjusting for change in global protein abundance, the modification was statistically significant.
 Additionally, this profile plot showed the large difference in available features between modified peptides and global proteins.
-b) The volcano plot of the WT\_Uninfect and WT\_Early comparison showed the specifics of the adjustment.
-The modification looked to be flat, with a log fold change of $.185$, while the global profiling showed a small negative fold change of $-.619$.
-While both exhibit small changes, when combined we saw a log fold change of $.804$ and adjusted p-value of $.0156$.}
+(b) The volcano plot of the WT\_Uninfect and WT\_Early comparison showed the specifics of the adjustment \todo{with $MSstatsPTM$}.
+The profile of the modified peptides appeared to be flat, with a log fold change of $.185$, while the global profiling showed a small negative fold change of $-.619$.
+Since both exhibit small changes in opposite directions, their combination produced a log$_2$-fold change of $.804$ and adjusted p-value of $.0156$.}
 \label{fig:Diff_Shigella_PTM}
@@
 
-\subsubsection{Dataset 6 : Human - Ubiquitination - Label-free no global profiling run~\cite{Cunningham2015}}
+\subsection{Dataset 6 : Human - Ubiquitination - Label-free no global profiling run}
 \label{sec:usp30}
@@
 \includegraphics[width=\textwidth]{sim_new/USP30_profile_plot}
-\caption{The global profiling of protein $P52209$ with the modification of the protein at site $K059$ was compared.
-The modification appeared generally unchanged between all conditions, whereas the global profiling run showed the CCCP and Combo conditions had a higher relative abundance compared to the Control and USP30\_OE.
-This indicated that the modification actually had an effect when comparing CCCP and Combo to Control and USP30\_OE.
+\caption{\todo{Start the caption with the name of the dataset} Protein $P52209$ with the modification of the protein at site $K059$.
+The modification appeared generally unchanged between all conditions, whereas the global profiling run showed that the CCCP and Combo conditions had a higher relative abundance compared to the Control and USP30\_OE.
+This indicated that the modification had in fact an effect when comparing CCCP and Combo to Control and USP30\_OE.
 However it was not entirely clear as one unmodified peptide feature appeared to be changed, while the other did not.
-This uncertainty was another result of not running a separate global profiling run.
-With a global profiling run, many unmodified features are generally quantified, removing the uncertainty that comes with low feature counts.}
+This uncertainty was another result of not acquiring data from a separate global profiling run.
+A separate global profiling run would have likely resulted in more and better quality unmodified features for that protein.
+}
 \label{fig:USP30_profile_plot}
 \end{figure}
-
-\clearpage
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Sample size calculation and power analysis}
-\label{sec:sample_size_analysis}
-
-\subsubsection*{Noisy PTM measurements benefited from additional biological replicates}
-
-Here we analyzed the sample size needed to achieve a desired statistical power.
-We compared the statistical power in experiments with differing numbers of replicates, variance, and fold change for both the modified and unmodified runs.
-In terms of the number of replicates, we tested scenarios with equal replicates in both the modified and unmodified runs, as well as scenarios where the replicates differed between runs.
-We used the biological experiments to determine what variance values to test.
-In datasets 4 and 5 the variance of the PTM was higher than the global protein.
-In dataset 6 the variance of the PTM and global protein were generally the same.
-We mimicked these scenarios and analyzed the power of experiments when the PTM variance was higher than the global protein and when they were equal.
-When the PTM and global protein were the same we chose a variance of $.15$, whereas when the PTM was higher than the global protein we chose a PTM variance of $.2$ and protein variance of $.1$.
-The results of the power and sample size analysis can be seen in Figure \ref{fig:power_sd_combo}.
-
-\begin{figure}[h!]
-\centering
- \begin{subfigure}{\textwidth}
- \centering
-	\includegraphics[width=.75\textwidth]{sim_new/same_var_power}
-	\caption{}
- \end{subfigure}\vspace{-1mm}
- \begin{subfigure}{\textwidth}
- \centering
-	\includegraphics[width=.75\textwidth]{sim_new/high_ptm_var_power}
-	\caption{}
- \end{subfigure}\vspace{-1mm}
- \caption{Power analysis of experiments with differing variances.
-a) The power of an experiment targeting PTMs with the same variance, .15, for the modified and unmodified peptides.
-Predictably when the replicates were high for both modified and unmodified peptides the power was much higher.
-Conversely at low replicates for each the power was much lower.
-With equal variance, it did not matter if the PTM replicates or protein replicates were higher.
-b) In this chart the variance for the PTM was higher than the unmodified protein.
-The PTM variance was .2, while the unmodified protein variance was .1.
-With equal replicates the results were the same as above.
-When the replicates were not equal, more replicates allocated to the PTM runs lead to higher power.}
-\label{fig:power_sd_combo}
-\end{figure}
-
-%The proposed approach allows us to calculate the sample size needed to achieve a desired statistical power as described in Section \ref{sec:design}.
-The proposed method adjusts for the underlying protein abundance in the PTM significance analysis, which corrects the confounding factor with a cost of increased variation.
-When the variation is increased, the experiment requires a larger number of replicates to reach the same power.
-Thus the statistical power is dependent on the variance from both the modified peptide and unmodifed protein as well as the number of replicates in both runs.
-
-%We compared the statistical power in experiments with differing numbers of replicates, variance, and fold change for both the modified and unmodified runs.
-For variance we used the variance from the biological experiments to determine what values to set.
-In dataset 4 and 5 the variance of the PTM was generally higher than the global protein.
-In contrast, in dataset 6 the variance of the PTM and Protein was generally the same.
-Because of this we will analyze the power of our experiments when the PTM variance is higher than the protein, and also when they are the same.
-For the variance numbers, in the case when the PTM and protein are the same we chose a variance of .15.
-When the PTM is higher than the protein we chose a PTM variance of .2 and a protein variance of .1.
-These numbers were derived from the biological experiments in this paper.
-
-%The results of the power and sample size analysis can be seen in \sfigref{fig:power_sd_combo}.
-
-%\begin{figure}[h!]
-%\centering
-% \begin{subfigure}{\textwidth}
-% \centering
-%	\includegraphics[width=.725\textwidth]{sim_new/same_var_power}
-%	\caption{Here we can see the power of an experiment targeting PTMs with the same variance, .15, for the modified and unmodified peptides.
-Predictably when the replicates are high for both modified and unmodified peptides the power is much higher.
-Conversely at low replicates for each the power is much lower.
-The interesting part of this chart is when the replicates are different between runs.
-With equal variance, it does not matter if the PTM replicates or protein replicates are higher.
-Both cases result in the same power with equal variance.}
-% \end{subfigure}\vspace{-5mm}
-% \begin{subfigure}{\textwidth}
-% \centering
-%	\includegraphics[width=.725\textwidth]{sim_new/high_ptm_var_power}
-%	\caption{In this chart the variance for the PTM is higher than the unmodified protein.
-The PTM variance is .2, while the unmodified protein variance is .1.
-With equal replicates the results are the same as above, more replicates equals more power.
-When the replicates are not the same we can clearly see that having more replicates for the PTM leads to more power.
-It is clearly more important to have high replicates for the PTM run than the unmodified protein, when the PTM variance is higher.}
-% \end{subfigure}
-% \caption{Power analysis of experiments with differing variances.}
-%\label{fig:power_sd_combo}
-%\end{figure}
-
-%\begin{figure}[h!]
-%\centering
-%\includegraphics[width=\textwidth]{sim/size_prot}
-%\caption{Protein-level adjustment relies on the inference of protein abundance, %which introduces additional uncertainty in the estimate of PTM difference.
-%Therefore, the required sample size to detect a systematic change is higher than as %expected for standard differential analysis without adjustment.
-The discrepancy can %be profound if the uncertainty associated with the protein abundance estimate is %greater than that of PTM abundance estimate.
-Sample size calculation without %accounting for the uncertainty would lead to under-powered studies.
-%\label{fig:size_prot}}
-%\end{figure}
-
-%\begin{figure}[h!]
-%\centering
-%\includegraphics[width=\textwidth]{sim/size_grp}
-%\includegraphics[width=.725\textwidth]{sim/size_grp}
-%\caption{In complex designs, simultaneously analyzing all the conditions% effectively increases the degrees of freedom and requires fewer replicates. %\label{fig:size_grp}}
-%\end{figure}
-
-%\begin{figure}[h!]
-%\centering
-%\includegraphics[width=\textwidth]{sim/pwr_grp}
-%\includegraphics[width=.725\textwidth]{sim/pwr_grp}
-%\caption{Statistical power can be improved by increasing the sample size and %analyzing multiple conditions together. \label{fig:pwr_grp}}
-%\end{figure}
 

### 2022-08-26 main/MSstatsPTM_Main.tex
minor changes for final submission

@@
 
-Liquid chromatography coupled with bottom up mass spectrometry (LC-MS/MS)-based proteomics is increasingly used to detect changes in post-translational modifications (PTMs) in samples conditions.
+Liquid chromatography coupled with bottom up mass spectrometry (LC-MS/MS)-based proteomics is increasingly used to detect changes in post-translational modifications (PTMs) in samples from different conditions.
 Analysis of data from such experiments faces numerous statistical challenges.
@@
 The statistical methodology separately summarizes the abundances of peptides with and without the modification sites, and fits separate linear mixed effects models that reflect the biological and technological aspects of the experimental design.
+Unmodified peptides may or may not span a modifiable site.
 Next, model-based inferences regarding the PTM and the unmodified protein-level summaries are combined to account for the confounding between these two sources.
@@
 Details of computer simulations are available in {\bf Supplementary Sec.
-\ref{supp-sec:sim}} and on GitHub (\url{https://github.com/devonjkohler/MSstatsPTM_simulations}).
+\ref{supp-sec:dataset1} \& \ref{supp-sec:dataset2}}, and on GitHub (\url{https://github.com/devonjkohler/MSstatsPTM_simulations}).
 
@@
 \medskip \noindent {\bf Evaluation:} We evaluated the ability of the statistical methods to correctly detect differentially abundant PTMs.
-We gauged the methods ability to avoid false positives (i.e.
+We gauged the ability of the methods to avoid false positives (i.e.
 specificity), accurately estimate the fold change between conditions, and analyzed the sensitivity of detecting differentially abundant PTMs.
@@
 Four mixtures (i.e., conditions) were created with varying amounts of human lysate, background {\it E.
-Coli} lysate, and human spike-in ub-peptide mixture.
+Coli} lysates, and human spike-in ub-peptide mixture.
 Unmodified peptides from human lysate were viewed as the global proteome.
@@
 Coli} lysate peptides were not expected to change in abundance in comparison, after accounting for adjustment, and were viewed as additional negative controls.
-We evaluated the statistical methods ability to avoid false positives, as well as their sensitivity in detecting the differentially abundant spike-in peptides and accurately estimate their expected fold change.
+We evaluated the ability of the statistical methods to avoid false positives, as well as their sensitivity in detecting the differentially abundant spike-in peptides and accurately estimate their expected fold change.
 
@@
 \medskip \noindent {\bf Evaluation:} We evaluated the ability of the statistical methods to detect changes in the abundance of modified peptides both before and after adjusting for changes in global protein abundance.
-The six condition were labeled Dox1hr, Dox2hr, Dox4hr, Dox6hr, NoDox0hr, and NoDox6hr.
+The six conditions were labeled Dox1hr, Dox2hr, Dox4hr, Dox6hr, NoDox0hr, and NoDox6hr.
 All conditions were compared with each other, resulting in 15 pairwise comparisons.
@@
 For one protein, the PTM site is represented by $K$ spectral features (i.e., peptide ions, distinguished by their cleavage residues and charge states). 
-The number of modified and unmodified features typically varies across proteins. Some log-intensities can be outliers, and some spectral features can be missing. 
+The number of modified and unmodified features typically varies across proteins. Some log$_2$-intensities can be outliers, and some spectral features can be missing. 
 The log$_2$-intensity of Feature $k$, in Replicate $j$ of Condition $i$ is denoted by $y_{ijk}^{\ast}$.
-Conversely, the unmodified portion of the protein is represented by $L$ spectral features, and the log-intensity of Feature $l$ from the unmodified portion of the protein in the same run is denoted by $y_{ijl}$.
+Conversely, the unmodified portion of the protein is represented by $L$ spectral features, and the log$_2$-intensity of Feature $l$ from the unmodified portion of the protein in the same run is denoted by $y_{ijl}$.
 The features can be quantified as part of a same mass spectrometry run, or in a separate enrichment and global proteome profiling run.
@@
 $\hat{y}^*_{ij} = \sum_{k=1}^K{y_{ijk}^{\ast}}/K$.
-Alternatively, summarization sums the intensities of the modified features on the original scale, and then takes the log
+Alternatively, summarization sums the intensities of the modified features on the original scale, and then takes the log$_2$
 \begin{eqnarray}
@@
 The basic $ANOVA$ can be extended to account for the confounding of changes in PTM abundance and overall changes in protein abundance~\cite{Schwammle2015,THOMAS2020,Mertins:2013}.
-Typically this is done by first calculating sums of the intensities of the modified and unmodified features on the original scale, and then considering replicate-wise ratios of the sums and taking the log
+Typically this is done by first calculating sums of the intensities of the modified and unmodified features on the original scale, and then considering replicate-wise ratios of the sums and taking the log$_2$
 \begin{eqnarray}
@@
 
-$MSstats$ and $MSstatsTMT$ can  also be used at the feature or at the modification cite level, as opposed to protein level.
-For example, summarizing the features per PTM cite instead of per protein, the approach allows us to test the null hypothesis in \eqref{conv_null_hyp}.
-
-The $MSstats$ framework has a number advantages over the methods above. 
+$MSstats$ and $MSstatsTMT$ can  also be used at the feature or at the modification site level, as opposed to protein level.
+For example, summarizing the features per PTM site instead of per protein, the approach allows us to test the null hypothesis in \eqref{conv_null_hyp}.
+
+The $MSstats$ framework has a number of advantages over the methods above. 
 First, unlike $ANOVA$ and $Limma$, $MSstats$ and $MSstatsTMT$ are applicable to arbitrary complex experimental designs, including designs with multiple sources of variation, and unbalanced designs. 
@@
 \end{eqnarray}
-For example, in the simple case of \figref{data-structure} with $J=2$ conditions, where $\hat{\sigma}^2_{PTM}$ and $\hat{\sigma}^2_{protein}$ are respectively  the estimates of the error variance for the PTM and protein model described in \eqref{msstats_model}, the standard error is calculated as
+For example, in the simple case of \figref{data-structure} with $J=2$ replicates, where $\hat{\sigma}^2_{PTM}$ and $\hat{\sigma}^2_{protein}$ are respectively  the estimates of the error variance for the PTM and protein model described in \eqref{msstats_model}, the standard error is calculated as
 \begin{eqnarray}
@@
 Typically, these variances are estimated from an existing experiment, conducted with the same biological material and measurement workflow.
-\todo{The estimation of the variance components for pairwise comparisons of conditions depends on the experimental design, and is detailed in Supply Figure S1}
 
@@
 \leq \left( \frac{\Delta_{adj}}{z_{1-\beta} + z_{1-\alpha /2}} \right)^{2}
+\label{eq:sample_size_calc}
 \end{equation}
@@
 J^{\prime} \geq \frac{(2\sigma_{PTM}^{2} + 2\sigma_{protein}^{2})(z_{1-\beta} + z_{1-\alpha /2})^2}{\Delta_{adj}^2}
+\label{eq:sample_size_calc_solved}
 \end{equation}
@@
 
-\todo{Here we calculate sample size, but in the results we calculate power.
-Could you briefly comment on how we can solve for the power given the sample size (say that this is using the same equation, but solving for $q$ instead.) In fact we can fix all value but one and solve for the remainder}
-
-\todo{I rearranged the supplementary sections. Could you check throughout the manuscript that all the supplementary sections are appropriately referenced?}
+In \eqref{sample_size_calc_solved} we calculated the required sample size, however in some cases the statistical power of an experiment may be of interest.
+In this case, \eqref{sample_size_calc} is still used, however the number of replicates, $J^{\prime}$, is fixed and False Discovery Rate, $q$, is solved for.
 
@@
 We require the processing tools to identify the modification site (i.e., the amino acid in the protein sequence where the modification occurred).
-This will generally include the amino acid abbreviation, plus it's number in the protein sequence.
+This will generally include the amino acid abbreviation, plus its number in the protein sequence.
 For example, a modification on a 70th amino acid in the sequence, serine should be marked as ``S70''.
@@
 In contrast, $MSstatsPTM$ can leverage the information in the PTM or unmodified portion of the protein if one of the two is missing.
-On average over all simulations, 3.94\% of the total unmodified protein data \todo{what do you mean by 'data'?
-Runs?
-Combination of features and runs?} used by $MSstatsPTM$ was discarded by $Limma$ and $ANOVA$ due to missing PTM data.
-No PTM data was discarded due to missing protein data.
+On average over all simulations, 3.94\% of the total unmodified protein run summarizations used by $MSstatsPTM$ were discarded by $Limma$ and $ANOVA$ due to missing PTM data.
+No PTM run summarizations were discarded due to missing protein data.
 Second, the robust Turkey Median Polish summarization in $MSstatsPTM$ was more resistant to outliers.
@@
 Additionally, the ratio based summarization used by $Limma$ and $ANOVA$ discarded more data than in the simulations.
-The ratio based summarization lost 9.63\% of the unmodified protein and 1.94\% of the PTM data \todo{clarify what you mean by 'data'} used by $MSstatsPTM$.
+The ratio based summarization lost 9.63\% of the unmodified protein and 1.94\% of the PTM run summarizations used by $MSstatsPTM$.
 This additional information lead $MSstatsPTM$ to a more accurate fold change estimation and better calibrated variance.
@@
 
-\todo{I rearranged the figures and the captions for the rest of the section.
-Could you please check that.I did not introduce any mistakes?} \todo{Please check for appropriate use (or lack thereof) of articles 'a' and 'the' throughout the rest of the section and discussion.
-This remains a mistery to me ;-) }
-
 \figref{Dataset4}(a) summarizes the results of Dataset 4: Human - Ubiquitination - 1mix-TMT in terms of number of differentially abundant PTMs before and after adjustment.
 Adjusting for changes in abundance in the unmodified portion of the protein caused fewer PTMs to be detected as differentially abundant.
-A question is whether this was due to adjustment in the log$_2$-fold estimation, or to increase in standard errors during the adjustment (\eqref{prop_se}).
+A question is whether this was due to adjustment in the log$_2$-fold estimation, or to an increase in standard errors during the adjustment (\eqref{prop_se}).
 To check that, we considered modified peptides for which the adjusted log$_2$-fold change was within 10\% of the unadjusted log$_2$-fold fold change, but which lost statistical significance after the adjustment.
@@
 
-Cases where true changes in PTM  were masked by changes in abundance in the unmodified portion of the protein were less frequent.
+Cases where true changes in PTMs were masked by changes in abundance in the unmodified portion of the protein were less frequent.
 One such case is in~\figref{Dataset4}(b).
@@
 This up-regulation was originally confounded by the down-regulation of unmodified $GSDMD$, and made the modification appear to have little change between no Dox and Dox 4 and 6 hour conditions.
-The proposed approach accounted for this confounding and the modification was detected as differentially abundant, with an absolute fold change of $1.415$ between the Dox 1 hour and Dox 4 hour conditions.
-The effect \todo{on what?} would have been challenging to observe without the proposed approach.
-Indeed, the modification contradicts the previous research focusing on global profiling of that protein \todo{not sure I understand the previous sentence.
-Also, is there a reference?}.
+The proposed approach accounted for this confounding and the modification was detected as differentially abundant, with a log$_2$ fold change of $2.79$ between the Dox 1 hour and Dox 4 hour conditions ({\bf Supplementary Figure~\ref{supp-fig:ipah_figures}}).
+The change in PTM abundance would have been challenging to observe without the proposed approach.
+Indeed, the modification contradicts the previous research focusing on global profiling of the $GSDMD$ protein~\cite{LUCHETTI2021}.
 
@@
 \figref{Dataset5}(b) shows an opposite case where the protein exhibits a change in abundance while the modification does not.
-Originally \todo{What do you mean by 'originally'?
-In a prior publication?
-Could you cite a reference?} the modification at $S178$ of protein $TTP$ was shown to be differentially abundant between KO\_Uninfect and KO\_Early, with a log fold change of $2.9$.
-However the unmodified protein was shown to contribute $2.014$ of this change, i.e.
-the modification only accounted for $.886$ \todo{on which scale?
-Maybe report as a \%?} after adjustment ({\bf Supplementary Figure~\ref{supp-fig:No_Diff_Shigella_PTM}}).
+Before adjusting for changes in the unmodified protein, the modification at $S178$ of protein $TTP$ was shown to be differentially abundant between WT\_Uninfect and WT\_Late, with a log$_2$ fold change of $2.9$.
+However the unmodified protein was shown to contribute 69.5\% of this change, while the modification only accounted for 30.5\% after adjustment ({\bf Supplementary Figure~\ref{supp-fig:No_Diff_Shigella_PTM}}).
 This caused the modification to lose differential abundance.
 
-\todo{Should we compare the results to limma and anova.
+%\todo{Should we compare the results to limma and anova.
 Is there something to comment on?
@@
 A PTM without features from the unmodified portion of the protein could not be adjusted for the confounding.
-Additionally, the lack of a separate global profiling run resulted in low feature counts, and noisier measurements of the unmodified peptides as compared to the other experiments (Table \ref{tab:dataDescription}).
+Additionally, the lack of a separate global profiling run resulted in low feature counts and noisier measurements of the unmodified peptides as compared to the other experiments (Table \ref{tab:dataDescription}).
 
@@
 
-\todo{I see that Supplementary Fig 8 and 9 are not referenced in the main text.
-Could you check that every supplementary fig or section is mentioned somewhere in the main text?
-Otherwise it is unclear what the supplementary figure is doing there.
-(I rearranged the order/number of Supplies sections, could you check that they are cited correctly?)}
-
 \subsubsection*{Noisy PTM measurements benefited from additional biological replicates}
@@
 Since the proposed approach does not have the restriction of the equal number of modified and unmodified global profiling runs,  it is also interesting to evaluate the interplay between the number of replicates of each type in presence of differing amounts of noise.
-In Datasets 4 and 5 the variance of the PTMs was higher than that of the unmodified protein summaries, \todo{with median values of PTM variance of $.2$, and protein variance of $.1$}.
-In Dataset 6 the variance of the PTM and the unmodified protein summaries were comparable, \todo{with the median value of around $.15$}.
-\todo{The power analysis took as input these median variance values.}
-
-\figref{power_sd_combo} shows that regardless of the relative amount of variation, larger adjusted \todo{ log$_2$-fold changes or FC?} and larger number of replicates enabled larger statistical power.
+In Datasets 4 and 5 the variance of the PTMs was higher than that of the unmodified protein summaries, with median values of PTM variance of $.45$, and protein variance of $.3$.
+In Dataset 6 the variance of the PTM and the unmodified protein summaries were comparable, with a median variance of $.45$.
+The power analysis took as input these median variance values.
+
+\figref{power_sd_combo} shows that regardless of the relative amount of variation, larger adjusted log$_2$-fold changes and larger number of replicates enabled larger statistical power.
 When the variance of PTMs was equal to the variance of global protein profiles, it did not matter whether we increased the number of PTM runs or the number of unmodified protein runs.
@@
 The implementation of $MSstatsPTM$ is a straightforward extension of $MSstats$, and is therefore applicable to the full breadth of experiment types supported by $MSstats$.
-Although demonstrated here on DDA, is also applicable to DIA, SRM and PRM acquisitions. 
-\todo{Additionally, the approach can handle experiments that use different strategies for modified versus unmodified peptides, e.f.
-using label-free methods for unmodified peptides and TMT labeling for modified peptides, or vice versa.}
+Although demonstrated here on DDA, it is also applicable to DIA, SRM and PRM acquisitions. 
+Additionally, the approach can handle experiments that use different strategies for modified versus unmodified peptides, e.g.
+using label-free methods for unmodified peptides and TMT labeling for modified peptides, or vice versa.
 
@@
 Future work is to carry out the inference and testing for not only the relative change of PTM abundance, but also the fraction of the protein that is modified at the particular site (site occupancy, or stoichiometry), and attempt to remove the confounding of individual PTMs in peptides with multiple modifications.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Acknowledgments}
+
+\section{Abbreviations}
+
+MS: mass spectrometry
+
+\noindent PTM: Post-translational modifications
+
+\noindent TMT: Tandem mass tag
+
+\noindent FDR: false discovery rate
+
+\noindent PPV: positive predictive value
+
+\noindent ANOVA: Analysis of bariance
+
+\noindent DDA: Data-dependent acquisitions
+
+\noindent DIA: Data-independent acquisition
+
+\noindent SRM: Selected reaction monitoring
+
+\noindent PRM: Parallel reaction monitoring
+
+\noindent KGG: Ub-remnant diglycyl-lysine
+
+\noindent GSDMD: Gasdermin D
 
@@
 \centering
-\includegraphics[scale=.5]{images/benchmark_fig.png}
+\includegraphics[scale=.5]{images/Figure1.png}
 \caption{{\bf Dataset 3: Spike-in benchmark - Ubiquitination - Label-free}. 
@@
 \centering
-\includegraphics[scale=.45]{images/fig3.png}
+\includegraphics[scale=.45]{images/Figure2.png}
 \caption{Schematic representation of one PTM site, in a special case of a label-free experiment with $I=2$ conditions and $J=2$ biological replicates per condition.
@@
 The peptides can be fully cleaved (solid lines), or partially cleaved (dashed lines).
+Unmodified features, colored blue, in the enriched runs are removed.
 The log$_2$-intensities of the modified peptides in Condition $i$, Run $j$, and Feature $k$ are denoted by $y_{ijk}^{\ast}$.
@@
 \centering
-\includegraphics[scale=.55]{images/MSstatsPTM_design.png}
+\includegraphics[scale=.55]{images/Figure3.png}
 \caption{The $MSstatsPTM$ workflow.
@@
 \begin{tabular}{c}
-\includegraphics[width=0.8\textwidth]{images/sim1_FDR_all_models.png}\\
+\includegraphics[width=0.8\textwidth]{images/Figure4a.png}\\
 (a)\\
-\includegraphics[width=0.8\textwidth]{images/sim3_Accuracy.png}\\
+\includegraphics[width=0.8\textwidth]{images/Figure4b.png}\\
 (b)
@@
 \begin{tabular}{c}
-\includegraphics[width=0.8\textwidth]{images/spike_in_fc.png}\\
+\includegraphics[width=0.8\textwidth]{images/Figure5a.png}\\
 (a)\\
-\includegraphics[width=0.8\textwidth]{images/spike_in_volcano.png}\\
+\includegraphics[width=0.8\textwidth]{images/Figure5b.png}\\
 (b)
@@
 The dashed vertical line shows the expected log$_2$ fold change of the spike-in peptides.
-Before adjustment the spike-in peptides did not follow the expected log fold change, but were more in line with expectation after adjustment.}
+Before adjustment the spike-in peptides did not follow the expected log$_2$ fold change, but were more in line with expectation after adjustment.}
 \label{fig:spikein}
@@
 \begin{tabular}{ccc}
-\includegraphics[width=0.32\textwidth]{images/ipah_venn_diagramm.png}
+\includegraphics[width=0.32\textwidth]{images/Figure6a.png}
 & $~~~$ &
-\includegraphics[width=0.62\textwidth]{images/IpaH_prof_plot.png}\\
+\includegraphics[width=0.62\textwidth]{images/Figure6b.png}\\
 (a) && (b)
@@
 (a) The overlap of differential modified peptides for the PTM model with and without global protein level adjustment across all pairwise comparisons.
-(b)   \todo{The caption of the sub-figure was missing (maybe I deleted it accidentally?
-Please add back)}}
+(b)   Comparing the profiling of protein $GSDMD$ with the ubiquitination at site $K62$.
+The individual PTM and protein features are shown in grey, while the summarization is highlighted in red.}
 \label{fig:Dataset4}
@@
 \begin{tabular}{ccc}
-\includegraphics[width=0.32\textwidth]{images/shig_venn_diagramm.png}
+\includegraphics[width=0.32\textwidth]{images/Figure7a.png}
 & $~~~$ &
-\includegraphics[width=0.62\textwidth]{images/No_Difference_Shigella_Profile_Plot}\\
+\includegraphics[width=0.62\textwidth]{images/Figure7b.png}\\
 (a) && (b)
@@
 \begin{tabular}{ccc}
-\includegraphics[width=0.32\textwidth]{images/usp30_venn_diagramm}
+\includegraphics[width=0.32\textwidth]{images/Figure8a.png}
 & $~~~$ &
-\includegraphics[width=0.325\textwidth]{images/usp30_venn_diagramm_matching_only}\\
+\includegraphics[width=0.325\textwidth]{images/Figure8b.png}\\
 (a) && (b)
@@
  \begin{tabular}{cc}
-	\includegraphics[width=.45\textwidth]{images/same_var_power}
+	\includegraphics[width=.5\textwidth]{images/Figure9a.png}
 &
-	\includegraphics[width=.45\textwidth]{images/high_ptm_var_power}
+	\includegraphics[width=.5\textwidth]{images/Figure9b.png}
 \\
@@
 \end{tabular}
- \caption{\todo{Could you check that I copied the correct figures?} Power analysis with with $MSstatsPTM$, \todo{based on variance components from what datasets?}.
+\caption{Power analysis with with $MSstatsPTM$, based on variance components from Datasets 4, 5, \& 6.
 Y-axis: statistical power, i.e.
 the probability of rejecting the null hypotheses of no protein-adjusted change in a PTM between conditions, when in fact the change exists.
-X-axis: \todo{clarify whether the x axis are adjusted FC.} \todo{Check whether the x axis is logFC or the FC on the original scale - looks like the original scale to me} a) The power of detecting differential abundant PTMs, when  the modified features and the unmodified protein summaries have the same variance .15.
-b) The power of detecting differential abundant PTMs, when the modified features had a larger variance (.2) as compared to the unmodified protein summaries (.1).}
+X-axis: log$_2$ fold change, after for adjusting for changes in the unmodified protein.
+a) The power of detecting differential abundant PTMs, when  the modified features and the unmodified protein summaries have the same variance .45.
+b) The power of detecting differential abundant PTMs, when the modified features had a larger variance (.45) as compared to the unmodified protein summaries (.3).}
 \label{fig:power_sd_combo}

### 2022-08-26 supplementary/ptm_sm.tex
minor changes for final submission

@@
  \begin{tabular}{cc}
-	\includegraphics[width=.445\textwidth]{sim_new/sim1_FC_boxplot}
+	\includegraphics[width=.445\textwidth]{sim_new/SuppFigure2a.png}
 	&
-	\includegraphics[width=.56\textwidth]{sim_new/sim3_FC_boxplot}\\
+	\includegraphics[width=.56\textwidth]{sim_new/SuppFigure2b.png}\\
 	(a)&(b)
 \end{tabular}
-\caption{\todo{In (b) could you indicate in panel labels that we are still talking about conditions and SD?}Simulated Datasets 1 and 2 : fold change distributions.
+\caption{Simulated Datasets 1 and 2 : fold change distributions.
+All models are shown after adjusting for changes in unmodified protein abundance.
 (a) In Simulated Dataset 1 all considered methods correctly estimated the fold change between conditions, with a median fold change estimation of 1.
@@
 $MSstatsPTM$ in this simulation had a  tighter distribution around the median.
-Both $Limma$ and $ANOVA$ showed a wider range around the fold change.
-\todo{In is not clear from the text whether Limma and ANOVA were used with or without adjustments for unmodified proteins.
-Could you clarify?}}
+Both $Limma$ and $ANOVA$ showed a wider range around the fold change.}
+
 \label{fig:fc_boxplot}
@@
 \centering
-\includegraphics[width=1\textwidth]{sim_new/spike_in_msstatsptm_volcano}
-\caption{\todo{Clarify that MSstats is the basic version before the adjustment, and MSstatsPTM after the adjustment} \todo{Start the caption with the name of the dataset} The modeling results of the $MSstatsPTM$ both before and after adjustment.
+\includegraphics[width=1\textwidth]{sim_new/SuppFigure3.png}
+\caption{Dataset 3 : SpikeIn benchmark - Ubiquitination - Label-free.
+The modeling results of the $MSstatsPTM$ both before and after adjustment.
+The unadjusted model is the basic version of $MSstats$ before adjustment.
 The spike-in peptides are colored red and the background peptides are colored grey.
@@
 \centering
-\includegraphics[width=1\textwidth]{sim_new/spike_in_limma_volcano}
-\caption{\todo{Start the caption with the name of the dataset} When modeling the experiment with the $Limma$ method, the spike-in peptides again follow the expected log fold change better after adjusting for changes in protein level.
+\includegraphics[width=1\textwidth]{sim_new/SuppFigure4.png}
+\caption{Dataset 3 : SpikeIn benchmark - Ubiquitination - Label-free.
+When modeling the experiment with the $Limma$ method, the spike-in peptides again follow the expected log fold change better after adjusting for changes in protein level.
 However, while the fold change was more accurate, the majority of spike-in peptides were not detected as differentially abundant.
@@
 \centering
-\includegraphics[width=1\textwidth]{sim_new/spike_in_anova_volcano}
-\caption{\todo{Start the caption with the name of the dataset} Using $ANOVA$ the fold change of the spike-in peptides was much closer to expectation after adjusting for global protein abundance.
+\includegraphics[width=1\textwidth]{sim_new/SuppFigure5.png}
+\caption{Dataset 3 : SpikeIn benchmark - Ubiquitination - Label-free.
+Using $ANOVA$ the fold change of the spike-in peptides was much closer to expectation after adjusting for global protein abundance.
 The log FC estimation was the same as $Limma$, however the p-values were different.
@@
 \end{eqnarray*}
-\begin{figure}[h!]
+
+\begin{figure}[ht]
 \centering
-	\includegraphics[width=.8\textwidth]{sim_new/IpaH_volcano_plot}
-	\caption{\todo{Could you separate the panels, and add labels (a) and (b) in the subfigures.
-(Could you use tabular as I did everywhere?) The title of the second panel is cut off.
-It is not clear which method is used - is it MSstats and MSstatsPTM?} \todo{Start the caption with the name of the dataset} Volcano plots of Dox4hr vs Dox1hr both before and after protein adjustment, \todo{with $MSstatsPTM$}.
+ \begin{tabular}{cc}
+	\includegraphics[width=.475\textwidth]{sim_new/SuppFigure6a.png}
+	&
+	\includegraphics[width=.475\textwidth]{sim_new/SuppFigure6b.png}\\
+	(a)&(b)
+\end{tabular}
+\caption{Dataset 4 : Human - Ubiquitination - 1mix-TMT.
+Volcano plots of Dox4hr vs Dox1hr both before and after protein adjustment with $MSstatsPTM$.
 The $GSDMD\_HUMAN|P57764\_K62$ modification is highlighted.
@@
 (b) After adjustment the fold change was much larger, and the modification was detected as differentially abundant.
-In this case \todo{$MSstatsPTM$?} allowed us to identify a differential modified peptide that could have otherwise been missed.}
+In this case $MSstatsPTM$ allowed us to identify a differential modified peptide that could have otherwise been missed.}
+
 \label{fig:ipah_figures}
@@
 
-\begin{figure}[h!]
+\begin{figure}[ht]
 \centering
-	\includegraphics[width=.8\textwidth]{sim_new/No_Difference_Shigella_Volcano}
-	\caption{\todo{Start the caption with the name of the dataset} \todo{Separate the labels and add labels (a) and (b) for the subpanels.
-(Could you use tabular as I did everywhere?)}Volcano plots of WT\_Late vs WT\_Uninfect both before and after protein adjustment \todo{with $MSstatsPTM$}.
+ \begin{tabular}{cc}
+	\includegraphics[width=.475\textwidth]{sim_new/SuppFigure7a.png}
+	&
+	\includegraphics[width=.475\textwidth]{sim_new/SuppFigure7b.png}\\
+	(a)&(b)
+\end{tabular}
+\caption{Dataset 5 : Mouse - Phosphorylation - 2mix-TMT.
+Volcano plots of WT\_Late vs WT\_Uninfect both before and after protein adjustment with $MSstatsPTM$.
 The $TTP\_MOUSE|P22893\_S178$ modification is highlighted.
@@
 (b) After adjustment the fold change was much smaller and the modification was not detected as differentially abundant.}
+
 \label{fig:No_Diff_Shigella_PTM}
@@
 
-\begin{figure}[h!]
-\centering
- \begin{tabular}{c}
-	\includegraphics[width=.9\textwidth]{sim_new/Difference_Shigella_Profile_Plot}\\
-	(a)\\
-	\includegraphics[width=.8\textwidth]{sim_new/Difference_Shigella_Volcano}\\
-	(b)
- \end{tabular}
- \caption{\todo{Start the caption with the name of the dataset} Protein $KI67\_MOUSE|E9PVX6$ with the modification at site $T215$.
+%\begin{figure}[h!]
+%\centering
+% \begin{tabular}{c}
+%	\includegraphics[width=.9\textwidth]{sim_new/Difference_Shigella_Profile_Plot}\\
+%	(a)\\
+%	\includegraphics[width=.8\textwidth]{sim_new/%Difference_Shigella_Volcano}\\%z
+%	(b)
+% \end{tabular}
+% \caption{\todo{Start the caption with the name of the dataset} Protein $KI67\_MOUSE|E9PVX6$ with the modification at site $T215$.
 (a) The modification and global protein appeared to show small or no difference between conditions when considered separately.
@@
 Since both exhibit small changes in opposite directions, their combination produced a log$_2$-fold change of $.804$ and adjusted p-value of $.0156$.}
-\label{fig:Diff_Shigella_PTM}
-\end{figure}
+%\label{fig:Diff_Shigella_PTM}
+%\end{figure}
 
-\clearpage
+%\clearpage
 
-\subsection{Dataset 6 : Human - Ubiquitination - Label-free no global profiling run}
-\label{sec:usp30}
+%\subsection{Dataset 6 : Human - Ubiquitination - Label-free no global profiling run}
+%\label{sec:usp30}
 
-An example profile plot for this experiment can be seen in  Figure \ref{fig:USP30_profile_plot}.
+%An example profile plot for this experiment can be seen in  Figure \ref{fig:USP30_profile_plot}.
 
-\begin{figure}[h!]
-\centering
-\includegraphics[width=\textwidth]{sim_new/USP30_profile_plot}
-\caption{\todo{Start the caption with the name of the dataset} Protein $P52209$ with the modification of the protein at site $K059$.
+%\begin{figure}[h!]
+%\centering
+%\includegraphics[width=\textwidth]{sim_new/USP30_profile_plot}
+%\caption{\todo{Start the caption with the name of the dataset} Protein $P52209$ with the modification of the protein at site $K059$.
 The modification appeared generally unchanged between all conditions, whereas the global profiling run showed that the CCCP and Combo conditions had a higher relative abundance compared to the Control and USP30\_OE.
@@
 }
-\label{fig:USP30_profile_plot}
-\end{figure}
+%\label{fig:USP30_profile_plot}
+%\end{figure}
 

### 2022-08-26 code/IpaH_analysis.R

@@ -13,6 +13,53 @@ protein_input <- read.csv("../data/Ipah_Global_Input.txt", sep = "\t")
 protein_input$FeatureType <- "Peptide"
 protein_input$Abundance <- log2(protein_input$Intensity)
 
+ptm_df <- ptm_input
+protein_df <- protein_input
+
+summarized_ptm <- data.table()
+summarized_proteins <- data.table()
+runs <- unique(ptm_df$Run)
+for (r in seq_along(runs)){
+  sum_runs <- ptm_df %>% filter(Run == runs[[r]]) %>%
+    group_by(ProteinName, Condition, Run) %>%
+    summarize(Abundance = log2(sum(Intensity, na.rm = TRUE)))
+  summarized_ptm <- rbindlist(list(summarized_ptm, sum_runs))
+  
+  sum_runs_prot <- protein_df %>% filter(Run == runs[[r]]) %>%
+    group_by(ProteinName, Condition, Run) %>%
+    summarize(Abundance = log2(sum(Intensity, na.rm = TRUE)))
+  summarized_proteins <- rbindlist(list(summarized_proteins, sum_runs_prot))
+}
+
+summarized_ptm$PTM <- summarized_ptm$ProteinName
+summarized_ptm$ProteinName <- sapply(summarized_ptm$PTM,
+                                     function(x) {paste(str_split(x, "_", 3)[[1]][1], 
+                                                        str_split(x, "_", 3)[[1]][2], sep="_")})
+joined <- merge(summarized_ptm, summarized_proteins,
+                by = c("ProteinName", "Run", "Condition"), all = TRUE)
+
+adjustable = adj_ptm_model %>% 
+  distinct(Protein, Label) %>% 
+  separate(Label, c("Condition1", "Condition2"), sep = "_v_")
+adjustable1 = adjustable %>% select(Protein, Condition1)
+adjustable1$Condition = adjustable1$Condition1
+adjustable2 = adjustable %>% select(Protein, Condition2)
+adjustable2$Condition = adjustable2$Condition2
+
+adjustable = rbindlist(list(adjustable1, adjustable2), fill = TRUE) %>% select(Protein, Condition)
+
+
+joined %>% filter(!is.finite(Abundance.x) & is.finite(Abundance.y)) %>% 
+  nrow() / joined %>% nrow()
+
[…]

### 2022-08-26 code/USP30_MSstatsPTM_Analysis.R

@@ -254,5 +254,3 @@ combo %>% filter(log2FC.x < log2FC.y*1.09 & log2FC.x > log2FC.y*.91 &
                    adj.pvalue.x >= .05 & adj.pvalue.y < .05)
 
 
-
-

### 2022-08-26 code/sample_size_analysis.R

@@ -13,9 +13,9 @@ sigma_model <- function(ptm_var, ptm_n, prot_var, prot_n){
 }
 
 ## Define adjustable metrics
-low <- .1
-mid <- .15
-high <- .2
+low <- .3
+mid <- .45
+high <- .45
 n_sample_list <- c(2,5)
 
 ## Hard metrics
@@ -59,8 +59,8 @@ high_var_df <- rbindlist(bind_list)
 cbPalette <- c("#E69F00", "#56B4E9", "#009E73", "#999999", 
                "#0072B2", "#D55E00", "#CC79A7", "#999999")
 
-p1 <- same_var_df %>%
-  ggplot() + geom_line(aes(x = Log2FC, y = Power, color = `PTM-Protein Replicates`), 
+p1 <- same_var_df %>% mutate(log2FC = log2(Log2FC)) %>% 
+  ggplot() + geom_line(aes(x = log2FC, y = Power, color = `PTM-Protein Replicates`), 
                        position = "jitter", size = 1.2) +
   scale_colour_manual(values=cbPalette) +
   labs(title = "Power analysis with same variance") +
@@ -71,11 +71,12 @@ p1 <- same_var_df %>%
         axis.title.y = element_text(size = 22),
         axis.title.x = element_text(size = 22),
         title = element_text(size = 22),
-        strip.text = element_text(size = 16))
+        strip.text = element_text(size = 16)) + 
+  ylim(0, 1)
 p1
 
-p2 <- high_ptm_var_df %>%
-  ggplot() + geom_line(aes(x = Log2FC, y = Power, color = `PTM-Protein Replicates`), size = 1.2) +
+p2 <- high_ptm_var_df %>% mutate(log2FC = log2(Log2FC)) %>% 
+  ggplot() + geom_line(aes(x = log2FC, y = Power, color = `PTM-Protein Replicates`), size = 1.2) +
   scale_colour_manual(values=cbPalette) +
   labs(title = "Power analysis with high PTM variance") +
   theme_bw() +
@@ -85,7 +86,8 @@ p2 <- high_ptm_var_df %>%
         axis.title.y = element_text(size = 22),
         axis.title.x = element_text(size = 22),
         title = element_text(size = 22),
-        strip.text = element_text(size = 16))
+        strip.text = element_text(size = 16)) + 
+  ylim(0, 1)
 p2
 
 png("../supplementary/sim_new/same_var_power.png", width = 750, height = 500)
